### 2019-10-11 manuscript.tex
first draft

\documentclass[english,,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Non-word repetition among the Tsimane': A small-scale study},
            pdfauthor={Alejandrina Cristia, Steven Levinson, \& Marisa Casillas},
            pdfkeywords={phonology, non-word repetition},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[shorthands=off,main=english]{babel}
\else
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\fi
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

  \title{Non-word repetition among the Tsimane': A small-scale study}
    \author{Alejandrina Cristia\textsuperscript{1}, Steven Levinson\textsuperscript{2}, \& Marisa Casillas\textsuperscript{2}}
    \date{}

\shorttitle{NWR among the Tsimane'}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL\\\textsuperscript{2} MPI}
\keywords{phonology, non-word repetition\newline\indent Word count: xxx}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

\usepackage{longtable}
\usepackage{lscape}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage[flushleft]{threeparttable}
\usepackage{threeparttablex}

\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

\authornote{

Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d'Ulm, 75005, Paris, France. E-mail: \href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}

\abstract{
ADD

}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\begin{itemize}
\tightlist
\item
  many studies on phonological development in childhood
\item
  they show phonology continues developing as children learn more words and gain mastery of phonetic targets
\item
  one key task in this research in NWR
\item
  used for theory (eg what are the links between phonology, working memory, and lexicon?), and for applications (nwr as a diagnostic)
\item
  it has been used acros a wide array of languages, particularly in Europe
\item
  however, less is known about NWR development in non-Indo European languages
\item
  this study: NWR in a Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory
\end{itemize}

In this study, we use nonword repetition (NWR), a technique that has been widely adopted to study the development of perception-production across multiple languages (CITE), particularly in Europe (CITE). In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning. The child should immediately try to say it back without changing anything. Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back). In our implementation of NWR, we additionally made sure that some of the items contained typologically rare and/or challenging sounds, to additionally capture children's auditory-articulatory accuracy.

Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in clise-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.

\hypertarget{intro-to-the-language}{%
\subsubsection{Intro to the language}\label{intro-to-the-language}}

\begin{itemize}
\tightlist
\item
  complexity in the vowel system
\item
  complexity in the consonant system
\item
  word shapes
\item
  typical word length
\item
  although not the focus of this paper, high use of suppletion in verbal paradigms, other features of language, see Levinson XXX for details
\end{itemize}

\hypertarget{intro-to-the-people}{%
\subsubsection{Intro to the people}\label{intro-to-the-people}}

\begin{itemize}
\tightlist
\item
  usually monolingual at home
\item
  schooling in English but it starts at age XX, so not relevant here
\item
  however, some use of English due to immigrants \& children of immigrants
\item
  children spend a lot of time with other children
\item
  most parents are subsistence farmers
\item
  parental education generally varies between XX and YY
\end{itemize}

\hypertarget{brief-review-of-nwr-for-our-purposes}{%
\subsubsection{Brief review of NWR for our purposes}\label{brief-review-of-nwr-for-our-purposes}}

\begin{itemize}
\item
  typical structure of NWR tasks: N of items
\item
  complexity varies widely across tasks, sometimes within tasks on purpose
\item
  some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
\item
  calculation of frequency in Yêli is challenging bc low resource language
\item
  so we only manipulated structural complexity, not probability
\item
  previous work seems to avoid difficult sounds
\item
  but we felt this was important to represent the language, so we also varied this factor
\item
  previous work shows even 2yo can do the task
\item
  but there are striking changes with age
\item
  inconsistent results RE effects of SES (including parental education) \& child sex
\item
  brief summary of inconsistent results SES
\item
  idem for sex
\end{itemize}

\hypertarget{research-questions-and-predictions}{%
\subsubsection{Research questions and predictions}\label{research-questions-and-predictions}}

\begin{itemize}
\tightlist
\item
  What is the accuracy (whole word, phoneme based, distance) overall?
\item
  How does this change as a function of item complexity (nb syll, syll complexity, sound complexity)?
\item
  How frequent are errors that result in real words? Is that a function of item complexity?
\item
  Is individual variation explainable by child age, sex, parental education?
\end{itemize}

Based on previous work, we made the following predictions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Children are more accurate for mono-syllables than longer items
\item
  Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
\item
  Children's accuracy increases with child age
\item
  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language)
\item
  Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
\item
  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperfomr later-born children
\end{enumerate}

Although these predictions were made based on previous work, we also thought reasonable that they would not obtain for the Rossel children in particular, as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English. \textbf{Check for work on European languages that may have looked into this}
\item
  The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors. Therefore, this may constitute a pressure for children to have finer aduitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory. \textbf{no work looking at consonants \& vowels? no work looking at nasal vowels in particular?}
\item
  We had no reservations that children's accuracy increases with child age for the Yélî community as well.
\item
  Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin (or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work. That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
\item
  In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment. This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly. However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
\item
  One main causal path between birth order and language development is via parental input (CITE). Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
\end{enumerate}

\hypertarget{pre-registration-and-planned-analyses}{%
\subsection{Pre-registration and planned analyses}\label{pre-registration-and-planned-analyses}}

In view of the hypothesis driven nature of this work, we intended to boost the interpretational value of these data by announcing our analysis plans prior to conducting them. There are a few limitations in our approach:

\begin{itemize}
\tightlist
\item
  The sample size is too small to have enough power to detect all of the relationships of interest
\item
  In fact, the sample size is too small to detect some of the known effects (see Table XX \textbf{add table with r/ds from previous work})
\item
  By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
\item
  Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear) on the first 11 participants, and who segmented all children's production. Therefore, a key condition of blinding is not met
\end{itemize}

Nonetheless, some form of pre-registration still seemed preferrable over the complete lack of it.

The pre-registration can be found XXX

\hypertarget{stimuli}{%
\subsection{Stimuli}\label{stimuli}}

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language. We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language. We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.

A first list of candidate items was generated in 2018 by selecting simple consonants (LIST) and vowels (LIST) that were combined into CV syllables, further sampling the space of 1- to 4-syllable sequences. These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary. They were further presented orally to three local research assistants, who were asked to repeat them and further say whether they were real words. Any item for which two or more of the assistants reported them having a meaning or some form of association were was excluded.

A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce CV monosyllables. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out. Since the phonology and phonetics of Yélî is still in the process of being described (CITE), there could have been undocumented constraints that rendered items illegal. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally. The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.

The final list is XX.

A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets. The split had the following constraint:

\begin{itemize}
\tightlist
\item
  the same 3 items were selected as practice items and used in all 40 elicitation sets
\item
  splits were done within each length group from the 2018 items (i.e., separately for 1, 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists). Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
\item
  once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2\_syllables, then 3-syllables, and finally 4-syllable items.
\end{itemize}

The 40 elicitation sets are available online from XXX.

\hypertarget{procedure}{%
\subsection{Procedure}\label{procedure}}

We tried to balance three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them; that children would feel comfortable doing this task with us; and that the community would feel safe with us doing this task with their children. Moreover, there were also some logistic constraints in terms of the space availability. As a result, the places where elicitation took place varied across the hamlets.

Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items \enquote{spreading} through hearsay. In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own. The complete list of places and the ways in which they met the desiderata mentioned above can be found in supplementary materials, under comments for each child (direct link xXX).

The child was donned a headset (xx for most of the children, xx for the rest). For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth. A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the 3 practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything. An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.

The first phase was making sure the child understood the task. This was explained orally and the first training item was presented. Often, children froze and did not say anything. If this happened, then we followed this procedure. First, the informant insisted. If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another. If the child could repeat these correctly, then we provided the recorded training item over headphones again. Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it). A small minority still failed to repeat the item after hearing it over headphones. If that occurred, we tried with the second training item, at which point some children got it and could continue. A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.

NWR studies vary in whether children are provided with several opportunities to hear and say the item. To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them. If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a headset that was recording a separate track). If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.

Whenever siblings from the same family were tested, an attempt was made to test first the older and then the younger child, and always on different elicitaiton sets.

\hypertarget{coding}{%
\subsection{Coding}\label{coding}}

There is an implicit accuracy coding occuring online as a function of whether the local informant asked for the item to be repeated or not. We will call this \enquote{online accuracy}, and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.

Additionally, the experimenter listened through all recordings and segmented out every child production. For the first 11 participants, she additionally made a judgment of whether the item was correctly or incorrectly repeated, and transcribed what she heard, which provides both an accuracy measure and a qualitative description of what children were actually saying in response to each stimulus. . We call this \enquote{non-native coding,} because this experimenter is not familiar with Yélî and has insufficient training on Yélî sounds, including common allophonic variants, not to mention not having some of the sounds in her native perceptual inventory.

Therefore, we sought additional information by asking a local informant to listen through each child production, paried with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated. We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets). This \enquote{native coding} provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder.

\hypertarget{analyses}{%
\subsection{Analyses}\label{analyses}}

For NWR accuracy, we considered only the child's first attempt, which was scored as correct or incorrect online, by a non-native coder (ofr the first 11 kids), and a native coder.

Some NWR studies employ phoneme accuracy (rather than whole word accuracy). We used exclusively the native coding for these, and we calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).

Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account, again using only the native coder's transcriptions.

\hypertarget{participants}{%
\subsection{Participants}\label{participants}}

Total N of kids, N of families, N of villages. Some children could not be included for the following reasons: refused participation (N=2), failed to repeat items presented over headphones even after coaching (N=3), spoke too softly to allow offline coding (N=8). In addition, one teenager was tested to put children at ease; their data is not included in analyses below. The remaining XX exhibited the following demographic characteristics

INSERT TABLE WITH THE KEY VARS MENTIONED ABOVE:

\begin{itemize}
\tightlist
\item
  child age
\item
  monoling
\item
  mat ed
\item
  birth order
\end{itemize}

\begin{verbatim}
##  [1] "X.3"         "ID"          "included"    "Age"         "DOB.ISO"    
##  [6] "DOB"         "Sex"         "birthOrder"  "familyID"    "monolingual"
## [11] "matEd"       "Cond"        "Date"        "Time"        "File"       
## [16] "fastcode"    "fullcode"    "Notes"       "Where"       "Village"    
## [21] "X"           "X.1"         "X.2"         "dob"         "test"       
## [26] "age.exact"
\end{verbatim}

\begin{verbatim}
##     ID score.online
## 1  c70    0.7307692
## 2  c71    0.5384615
## 3  c72    0.6086957
## 4  c73    0.8461538
## 5  c74    0.9090909
## 6  c76    0.7307692
## 7  c77    0.9565217
## 8  c78    0.8461538
## 9  c79    0.7692308
## 10 c80    0.8260870
## 11 c81    0.5652174
## 12 c82    0.5000000
\end{verbatim}

\includegraphics{manuscript_files/figure-latex/cars-1.pdf} \includegraphics{manuscript_files/figure-latex/cars-2.pdf} \includegraphics{manuscript_files/figure-latex/cars-3.pdf}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\newpage

\hypertarget{acknowledgments}{%
\subsection{Acknowledgments}\label{acknowledgments}}

We are grateful to informants and individuals who participated in the study. AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J. S. McDonnell Foundation Understanding Human Cognition Scholar Award.

\hypertarget{references}{%
\section{References}\label{references}}

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

\end{document}

### 2019-10-11 manuscript.Rmd
first draft

--- 2019-10-11 manuscript.tex
@@
-\documentclass[english,,man,floatsintext]{apa6}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Non-word repetition among the Tsimane': A small-scale study},
-            pdfauthor={Alejandrina Cristia, Steven Levinson, \& Marisa Casillas},
-            pdfkeywords={phonology, non-word repetition},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\fi
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
+---
+title: "Non-word repetition among the Tsimane': A small-scale study"
+shorttitle        : "NWR among the Tsimane'"
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
+author: 
+  - name          : "Alejandrina Cristia"
+    affiliation   : "1"
+    corresponding : yes    # Define only one corresponding author
+    address       : "29, rue d'Ulm, 75005, Paris, France"
+    email         : "alecristia@gmail.com"
+  - name          : "Steven Levinson"
+    affiliation   : "2"
+  - name          : "Marisa Casillas"
+    affiliation   : "2"
 
-  \title{Non-word repetition among the Tsimane': A small-scale study}
-    \author{Alejandrina Cristia\textsuperscript{1}, Steven Levinson\textsuperscript{2}, \& Marisa Casillas\textsuperscript{2}}
-    \date{}
+affiliation:
+  - id            : "1"
+    institution   : "LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL"
+  - id            : "2"
+    institution   : "MPI"
 
-\shorttitle{NWR among the Tsimane'}
-\affiliation{
-\vspace{0.5cm}
-\textsuperscript{1} LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL\\\textsuperscript{2} MPI}
-\keywords{phonology, non-word repetition\newline\indent Word count: xxx}
-\usepackage{csquotes}
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
+author_note: >
 
-\usepackage{longtable}
-\usepackage{lscape}
-\usepackage{multirow}
-\usepackage{tabularx}
-\usepackage[flushleft]{threeparttable}
-\usepackage{threeparttablex}
-
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-\authornote{
-
-Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d'Ulm, 75005, Paris, France. E-mail: \href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
-
-\abstract{
+abstract: >    
     ADD
 
-}
+keywords          : "phonology, non-word repetition"
+wordcount         : "xxx"
 
-\begin{document}
-\maketitle
+bibliography      : ["nwrbib.bib"]
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+figsintext        : yes
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+lineno            : no
 
-\begin{itemize}
-\tightlist
-\item
-  many studies on phonological development in childhood
-\item
-  they show phonology continues developing as children learn more words and gain mastery of phonetic targets
-\item
-  one key task in this research in NWR
-\item
-  used for theory (eg what are the links between phonology, working memory, and lexicon?), and for applications (nwr as a diagnostic)
-\item
-  it has been used acros a wide array of languages, particularly in Europe
-\item
-  however, less is known about NWR development in non-Indo European languages
-\item
-  this study: NWR in a Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory
-\end{itemize}
+lang              : "english"
+class             : "man"
+output            : papaja::apa6_pdf
+
+---
+
+```{r setup, include=FALSE}
+#Sys.setlocale("LC_CTYPE", "en_US_POSIX.UTF-8")
+#Sys.setlocale("LANG", "en_US_POSIX.UTF-8")
+Sys.setlocale(category = "LC_ALL", locale = "en_SE.utf8")
+
+knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = FALSE, fig.pos = "T")
+require("papaja")
+
+```
+
+## Introduction
+
+- many studies on phonological development in childhood
+- they show phonology continues developing as children learn more words and gain mastery of phonetic targets
+- one key task in this research in NWR
+- used for theory (eg what are the links between phonology, working memory, and lexicon?), and for applications (nwr as a diagnostic)
+- it has been used acros a wide array of languages, particularly in Europe
+- however, less is known about NWR development in non-Indo European languages
+- this study: NWR in a Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory
 
@@
 
-\hypertarget{intro-to-the-language}{%
-\subsubsection{Intro to the language}\label{intro-to-the-language}}
+### Intro to the language
 
-\begin{itemize}
-\tightlist
-\item
-  complexity in the vowel system
-\item
-  complexity in the consonant system
-\item
-  word shapes
-\item
-  typical word length
-\item
-  although not the focus of this paper, high use of suppletion in verbal paradigms, other features of language, see Levinson XXX for details
-\end{itemize}
+- complexity in the vowel system
+- complexity in the consonant system
+- word shapes
+- typical word length
+- although not the focus of this paper, high use of suppletion in verbal paradigms, other features of language, see Levinson XXX for details
 
-\hypertarget{intro-to-the-people}{%
-\subsubsection{Intro to the people}\label{intro-to-the-people}}
+### Intro to the people
 
-\begin{itemize}
-\tightlist
-\item
-  usually monolingual at home
-\item
-  schooling in English but it starts at age XX, so not relevant here
-\item
-  however, some use of English due to immigrants \& children of immigrants
-\item
-  children spend a lot of time with other children
-\item
-  most parents are subsistence farmers
-\item
-  parental education generally varies between XX and YY
-\end{itemize}
+- usually monolingual at home
+- schooling in English but it starts at age XX, so not relevant here
+- however, some use of English due to immigrants & children of immigrants
+- children spend a lot of time with other children
+- most parents are subsistence farmers
+- parental education generally varies between XX and YY
 
-\hypertarget{brief-review-of-nwr-for-our-purposes}{%
-\subsubsection{Brief review of NWR for our purposes}\label{brief-review-of-nwr-for-our-purposes}}
+### Brief review of NWR for our purposes
 
-\begin{itemize}
-\item
-  typical structure of NWR tasks: N of items
-\item
-  complexity varies widely across tasks, sometimes within tasks on purpose
-\item
-  some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
-\item
-  calculation of frequency in Yêli is challenging bc low resource language
-\item
-  so we only manipulated structural complexity, not probability
-\item
-  previous work seems to avoid difficult sounds
-\item
-  but we felt this was important to represent the language, so we also varied this factor
-\item
-  previous work shows even 2yo can do the task
-\item
-  but there are striking changes with age
-\item
-  inconsistent results RE effects of SES (including parental education) \& child sex
-\item
-  brief summary of inconsistent results SES
-\item
-  idem for sex
-\end{itemize}
+- typical structure of NWR tasks: N of items
+- complexity varies widely across tasks, sometimes within tasks on purpose
+- some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
+- calculation of frequency in Yêli is challenging bc low resource language
+- so we only manipulated structural complexity, not probability
+- previous work seems to avoid difficult sounds
+- but we felt this was important to represent the language, so we also varied this factor
 
-\hypertarget{research-questions-and-predictions}{%
-\subsubsection{Research questions and predictions}\label{research-questions-and-predictions}}
+- previous work shows even 2yo can do the task
+- but there are striking changes with age
+- inconsistent results RE effects of SES (including parental education) & child sex
 
-\begin{itemize}
-\tightlist
-\item
-  What is the accuracy (whole word, phoneme based, distance) overall?
-\item
-  How does this change as a function of item complexity (nb syll, syll complexity, sound complexity)?
-\item
-  How frequent are errors that result in real words? Is that a function of item complexity?
-\item
-  Is individual variation explainable by child age, sex, parental education?
-\end{itemize}
+- brief summary of inconsistent results SES
+
+- idem for sex
+
+### Research questions and predictions
+
+- What is the accuracy (whole word, phoneme based, distance) overall?
+- How does this change as a function of item complexity (nb syll, syll complexity, sound complexity)?
+- How frequent are errors that result in real words? Is that a function of item complexity?
+- Is individual variation explainable by child age, sex, parental education?
 
@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Children are more accurate for mono-syllables than longer items
-\item
+1. Children are more accurate for mono-syllables than longer items
+2.
 Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
-\item
-  Children's accuracy increases with child age
-\item
-  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language)
-\item
+3. Children's accuracy increases with child age
+4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language)
+5.
 Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
-\item
-  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperfomr later-born children
-\end{enumerate}
+6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperfomr later-born children
 
@@
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
+1.
 The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
-\textbf{Check for work on European languages that may have looked into this}
-\item
+**Check for work on European languages that may have looked into this**
+2.
 The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
 Therefore, this may constitute a pressure for children to have finer aduitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
-\textbf{no work looking at consonants \& vowels?
-no work looking at nasal vowels in particular?}
-\item
-  We had no reservations that children's accuracy increases with child age for the Yélî community as well.
-\item
+**no work looking at consonants & vowels?
+no work looking at nasal vowels in particular?**
+3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
+4.
 Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin (or at least from people other than their mothers, who tend to be the non-native speakers).
@@
 That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
-\item
+5.
 In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
@@
 However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
-\item
+6.
 One main causal path between birth order and language development is via parental input (CITE).
 Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
-\end{enumerate}
 
-\hypertarget{pre-registration-and-planned-analyses}{%
-\subsection{Pre-registration and planned analyses}\label{pre-registration-and-planned-analyses}}
+## Pre-registration and planned analyses
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  The sample size is too small to have enough power to detect all of the relationships of interest
-\item
-  In fact, the sample size is too small to detect some of the known effects (see Table XX \textbf{add table with r/ds from previous work})
-\item
-  By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
-\item
-  Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear) on the first 11 participants, and who segmented all children's production.
+- The sample size is too small to have enough power to detect all of the relationships of interest
+- In fact, the sample size is too small to detect some of the known effects (see Table XX **add table with r/ds from previous work**)
+- By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
+- Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear) on the first 11 participants, and who segmented all children's production.
 Therefore, a key condition of blinding is not met
-\end{itemize}
 
@@
 
-\hypertarget{stimuli}{%
-\subsection{Stimuli}\label{stimuli}}
+## Stimuli
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  the same 3 items were selected as practice items and used in all 40 elicitation sets
-\item
-  splits were done within each length group from the 2018 items (i.e., separately for 1, 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
+- the same 3 items were selected as practice items and used in all 40 elicitation sets
+- splits were done within each length group from the 2018 items (i.e., separately for 1, 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
 Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
-\item
-  once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2\_syllables, then 3-syllables, and finally 4-syllable items.
-\end{itemize}
+- once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2_syllables, then 3-syllables, and finally 4-syllable items.
 
@@
 
-\hypertarget{procedure}{%
-\subsection{Procedure}\label{procedure}}
+## Procedure
 
@@
 
-Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items \enquote{spreading} through hearsay.
+Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items 'spreading' through hearsay.
 In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own.
@@
 
-\hypertarget{coding}{%
-\subsection{Coding}\label{coding}}
+## Coding
 
 There is an implicit accuracy coding occuring online as a function of whether the local informant asked for the item to be repeated or not.
-We will call this \enquote{online accuracy}, and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
+We will call this "online accuracy", and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
 
@@
 .
-We call this \enquote{non-native coding,} because this experimenter is not familiar with Yélî and has insufficient training on Yélî sounds, including common allophonic variants, not to mention not having some of the sounds in her native perceptual inventory.
+We call this "non-native coding," because this experimenter is not familiar with Yélî and has insufficient training on Yélî sounds, including common allophonic variants, not to mention not having some of the sounds in her native perceptual inventory.
 
@@
 We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-This \enquote{native coding} provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder.
+This "native coding" provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder.
 
-\hypertarget{analyses}{%
-\subsection{Analyses}\label{analyses}}
+## Analyses
 
@@
 
-\hypertarget{participants}{%
-\subsection{Participants}\label{participants}}
+## Participants
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  child age
-\item
-  monoling
-\item
-  mat ed
-\item
-  birth order
-\end{itemize}
+- child age
+- monoling
+- mat ed
+- birth order
 
-\begin{verbatim}
-##  [1] "X.3"         "ID"          "included"    "Age"         "DOB.ISO"    
-##  [6] "DOB"         "Sex"         "birthOrder"  "familyID"    "monolingual"
-## [11] "matEd"       "Cond"        "Date"        "Time"        "File"       
-## [16] "fastcode"    "fullcode"    "Notes"       "Where"       "Village"    
-## [21] "X"           "X.1"         "X.2"         "dob"         "test"       
-## [26] "age.exact"
-\end{verbatim}
+```{r cars}
+# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
+# fulldemo=fulldemo[,-2] #remove children's names
+# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
+# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
+# fulldemo$age.exact=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
+# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
+# plot(fulldemo$Age~fulldemo$age.exact)     
+# write.csv(fulldemo,"nwr-demo.csv")
 
-\begin{verbatim}
-##     ID score.online
-## 1  c70    0.7307692
-## 2  c71    0.5384615
-## 3  c72    0.6086957
-## 4  c73    0.8461538
-## 5  c74    0.9090909
-## 6  c76    0.7307692
-## 7  c77    0.9565217
-## 8  c78    0.8461538
-## 9  c79    0.7692308
-## 10 c80    0.8260870
-## 11 c81    0.5652174
-## 12 c82    0.5000000
-\end{verbatim}
+read.csv("NWR-demo.csv",header=T)->demo
+colnames(demo)
+read.table("log2.txt",header=T,sep="\t")->x
+x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
+x[order(x$out),]->x
+x$cor.online=1
+for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0
 
-\includegraphics{manuscript_files/figure-latex/cars-1.pdf} \includegraphics{manuscript_files/figure-latex/cars-2.pdf} \includegraphics{manuscript_files/figure-latex/cars-3.pdf}
+subset(x,rep=="1")->x
+aggregate(x$cor.online,by=list(x$child),mean)->onl
+colnames(onl)<-c("ID","score.online")
 
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+print(onl)
+
+aggregate(x$corr,by=list(x$child),mean)->ms
+colnames(ms)<-c("ID","score")
+merge(ms,onl)->ms
+merge(ms,demo)->scores
+scores$Age=as.numeric(as.character(scores$Age))
+scores$Age[is.na(scores$Age)]<-15
+plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
+abline(lm(scores$score*100~scores$Age))
+plot(scores$score~scores$monolingual)
+points(scores$score~as.numeric(scores$monolingual),pch=20)
+
+plot(ms$score.online~ms$score)
+
+```
+
+## Discussion
 
@@
 
-\hypertarget{acknowledgments}{%
-\subsection{Acknowledgments}\label{acknowledgments}}
+## Acknowledgments
 
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+# References
 
@@
 \setlength{\leftskip}{0.5in}
-
-\end{document}

### 2019-10-14 prereg.Rmd
empty prereg

---
title: 'NWR-YD: Pre-registration'
author: "AC & MC"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
library(tidyverse)
library(dataPreparation)
library(knitr)
library(TOSTER)
knitr::opts_chunk$set(echo = TRUE)
knitr::kable

```

## Description

<!-- 1.1.	Please give a brief description of your study, including some background, the purpose of the of the study, or broad research questions.  -->
<!-- 1.2.	Example: Though there is strong evidence to suggest that sugar affects taste preferences, the effect has never been demonstrated in brownies. Therefore, we will measure taste preference for four different levels of sugar concentration in a standard brownie recipe to determine if the effect exists in this pastry.  -->
<!-- 1.3.	More info: The description should be no longer than the length of an abstract. It can give some context for the proposed study, but great detail is not needed here for your preregistration. -->

## Hypotheses

<!-- 1.1.	List specific, concise, and testable hypotheses. Please state if the hypotheses are directional or non-directional. If directional, state the direction. A predicted effect is also appropriate here. If a specific interaction or moderation is important to your research, you can list that as a separate hypothesis.  -->
<!-- 1.2.	Example: If taste affects preference, then mean preference indices will be higher with higher concentrations of sugar.  -->
<!-- Our hypothesis is that child-centered rearing (as a proxy for relative level of child-directed speech) is positively related to size of phoneme inventories: we predict that in languages in which a higher proportion of adult speakers provide with more responses of “child-centered rearing” responses, we find larger phoneme inventories than for languages with more responses of “parents have a life of their own”.  -->

## Design plan
<!-- In this section, you will be asked to describe the overall design of your study. Remember that this research plan is designed to register a single study, so if you have multiple experimental designs, please complete a separate preregistration. -->

###	Study type

<!-- 1.1.	Experiment - A researcher randomly assigns treatments to study subjects, this includes field or lab experiments. This is also known as an intervention experiment and includes randomized controlled trials. -->
<!-- 1.2.	Observational Study - Data is collected from study subjects that are not randomly assigned to a treatment. This includes surveys, ñnatural experiments,î and regression discontinuity designs. -->
<!-- 1.3.	Meta-Analysis - A systematic review of published studies. -->
<!-- 1.4.	Other  -->

### Blinding
<!-- 1.1.	Blinding describes who is aware of the experimental manipulations within a study. Mark all that apply. -->
<!-- 1.1.1.	No blinding is involved in this study. -->
<!-- 1.1.2.	For studies that involve human subjects, they will not know the treatment group to which they have been assigned. -->
<!-- 1.1.3.	Personnel who interact directly with the study subjects (either human or non-human subjects) will not be aware of the assigned treatments. (Commonly known as “double blind”) -->
<!-- 1.1.4.	Personnel who analyze the data collected from the study are not aware of the treatment applied to any given group. -->
<!-- 1.	Is there any additional blinding in this study? -->

###	Study design
<!-- 1.1.	Describe your study design. Examples include two-group, factorial, randomized block, and repeated measures. Is it a between (unpaired), within-subject (paired), or mixed design? Describe any counterbalancing required. Typical study designs for observation studies include cohort, cross sectional, and case-control studies. -->
<!-- 1.2.	Example: We have a between subjects design with 1 factor (sugar by mass) with 4 levels.  -->
<!-- 1.3.	More info: This question has a variety of possible answers. The key is for a researcher to be as detailed as is necessary given the specifics of their design. Be careful to determine if every parameter has been specified in the description of the study design. There may be some overlap between this question and the following questions. That is OK, as long as sufficient detail is given in one of the areas to provide all of the requested information. For example, if the study design describes a complete factorial, 2 X 3 design and the treatments and levels are specified previously, you do not have to repeat that information. -->

### Randomization

<!-- 1.1.	If you are doing a randomized study, how will you randomize, and at what level? -->
<!-- 1.2.	Example: We will use block randomization, where each participant will be randomly assigned to one of the four equally sized, predetermined blocks. The random number list used to create these four blocks will be created using the web applications available at http://random.org.  -->
<!-- 1.3.	More info: Typical randomization techniques include: simple, block, stratified, and adaptive covariate randomization. If randomization is required for the study, the method should be specified here, not simply the source of random numbers. -->

## Sampling Plan
<!-- In this section we’ll ask you to describe how you plan to collect samples, as well as the number of samples you plan to collect and your rationale for this decision. Please keep in mind that the data described in this section should be the actual data used for analysis, so if you are using a subset of a larger dataset, please describe the subset that will actually be used in your study. -->

###	Existing data 
<!-- 1.1.	Preregistration is designed to make clear the distinction between confirmatory tests, specified prior to seeing the data, and exploratory analyses conducted after observing the data. Therefore, creating a research plan in which existing data will be used presents unique challenges. Please select the description that best describes your situation. Please do not hesitate to contact us if you have questions about how to answer this question (prereg@cos.io). -->
<!-- 1.1.1.	Registration prior to creation of data: As of the date of submission of this research plan for preregistration, the data have not yet been collected, created, or realized.  -->
<!-- 1.1.2.	Registration prior to any human observation of the data: As of the date of submission, the data exist but have not yet been quantified, constructed, observed, or reported by anyone - including individuals that are not associated with the proposed study. Examples include museum specimens that have not been measured and data that have been collected by non-human collectors and are inaccessible. -->
<!-- 1.1.3.	Registration prior to accessing the data: As of the date of submission, the data exist, but have not been accessed by you or your collaborators. Commonly, this includes data that has been collected by another researcher or institution. -->
<!-- 1.1.4.	Registration prior to analysis of the data: As of the date of submission, the data exist and you have accessed it, though no analysis has been conducted related to the research plan (including calculation of summary statistics). A common situation for this scenario when a large dataset exists that is used for many different studies over time, or when a data set is randomly split into a sample for exploratory analyses, and the other section of data is reserved for later confirmatory data analysis. -->
<!-- 1.1.5.	Registration following analysis of the data: As of the date of submission, you have accessed and analyzed some of the data relevant to the research plan. This includes preliminary analysis of variables, calculation of descriptive statistics, and observation of data distributions. Please see cos.io/prereg for more information.  -->

###	Explanation of existing data 
<!-- 1.1.	If you indicate that you will be using some data that already exist in this study, please describe the steps you have taken to assure that you are unaware of any patterns or summary statistics in the data. This may include an explanation of how access to the data has been limited, who has observed the data, or how you have avoided observing any analysis of the specific data you will use in your study.  -->
<!-- 1.2.	Example: An appropriate instance of using existing data would be collecting a sample size much larger than is required for the study, using a small portion of it to conduct exploratory analysis, and then registering one particular analysis that showed promising results. After registration, conduct the specified analysis on that part of the dataset that had not been investigated by the researcher up to that point.  -->
<!-- 1.3.	More info: An appropriate instance of using existing data would be collecting a sample size much larger than is required for the study, using a small portion of it to conduct exploratory analysis, and then registering one particular analysis that showed promising results. After registration, conduct the specified analysis on that part of the dataset that had not been investigated by the researcher up to that point. -->

### Data collection procedures
<!-- 1.1.	Please describe the process by which you will collect your data. If you are using human subjects, this should include the population from which you obtain subjects, recruitment efforts, payment for participation, how subjects will be selected for eligibility from the initial pool (e.g. inclusion and exclusion rules), and your study timeline. For studies that donÍt include human subjects, include information about how you will collect samples, duration of data gathering efforts, source or location of samples, or batch numbers you will use.  -->
<!-- 1.2.	Example: Participants will be recruited through advertisements at local pastry shops. Participants will be paid $10 for agreeing to participate (raised to $30 if our sample size is not reached within 15 days of beginning recruitment). Participants must be at least 18 years old and be able to eat the ingredients of the pastries. -->
<!-- 1.3.	More information: The answer to this question requires a specific set of instructions so that another person could repeat the data collection procedures and recreate the study population. Alternatively, if the study population would be unable to be reproduced because it relies on a specific set of circumstances unlikely to be recreated (e.g., a community of people from a specific time and location), the criteria and methods for creating the group and the rationale for this unique set of subjects should be clear. -->

### 1.	Sample size
<!-- 1.1.	Describe the sample size of your study. How many units will be analyzed in the study? This could be the number of people, birds, classrooms, plots, interactions, or countries included. If the units are not individuals, then describe the size requirements for each unit. If you are using a clustered or multilevel design, how many units are you collecting at each level of the analysis? -->
<!-- 1.2.	Example: Our target sample size is 280 participants. We will attempt to recruit up to 320, assuming that not all will complete the total task.  -->
<!-- 1.3.	More information: For some studies, this will simply be the number of samples or the number of clusters. For others, this could be an expected range, minimum, or maximum number. -->

### 1.	Sample size rationale
<!-- 1.1.	This could include a power analysis or an arbitrary constraint such as time, money, or personnel. -->
<!-- 1.2.	Example: We used the software program G*Power to conduct a power analysis. Our goal was to obtain .95 power to detect a medium effect size of .25 at the standard .05 alpha error probability.  -->
<!-- More information: This gives you an opportunity to specifically state how the sample size will be determined. A wide range of possible answers is acceptable; remember that transparency is more important than principled justifications. If you state any reason for a sample size upfront, it is better than stating no reason and leaving the reader to “fill in the blanks.” Acceptable rationales include: a power analysis, an arbitrary number of subjects, or a number based on time or monetary constraints -->

### 1.	Stopping rule
<!-- 1.1.	If your data collection procedures do not give you full control over your exact sample size, specify how you will decide when to terminate your data collection.  -->
<!-- 1.2.	Example: We will post participant sign-up slots by week on the preceding Friday night, with 20 spots posted per week. We will post 20 new slots each week if, on that Friday night, we are below 320 participants.  -->
<!-- 1.3.	More information: You may specify a stopping rule based on p-values only in the specific case of sequential analyses with pre-specified checkpoints, alphas levels, and stopping rules. Unacceptable rationales include stopping based on p-values if checkpoints and stopping rules are not specified. If you have control over your sample size, then including a stopping rule is not necessary, though it must be clear in this question or a previous question how an exact sample size is attained. -->

## Variables

<!-- In this section you can describe all variables (both manipulated and measured variables) that will later be used in your confirmatory analysis plan. In your analysis plan, you will have the opportunity to describe how each variable will be used. If you have variables which you are measuring for exploratory analyses, you are not required to list them, though you are permitted to do so. -->

###	Manipulated variables
<!-- 1.1.	Describe all variables you plan to manipulate and the levels or treatment arms of each variable. This is not applicable to any observational study.  -->
<!-- 1.2.	Example: We manipulated the percentage of sugar by mass added to brownies. The four levels of this categorical variable are: 15%, 20%, 25%, or 40% cane sugar by mass.  -->
<!-- 1.3.	More information: For any experimental manipulation, you should give a precise definition of each manipulated variable. This must include a precise description of the levels at which each variable will be set, or a specific definition for each categorical treatment. For example, “loud or quiet,” should instead give either a precise decibel level or a means of recreating each level. 'Presence/absence' or 'positive/negative' is an acceptable description if the variable is precisely described. -->

###	Measured variables
<!-- 1.1.	Describe each variable that you will measure. This will include outcome measures, as well as any predictors or covariates that you will measure. You do not need to include any variables that you plan on collecting if they are not going to be included in the confirmatory analyses of this study. -->
<!-- 1.2.	Example: The single outcome variable will be the perceived tastiness of the single brownie each participant will eat. We will measure this by asking participants ‘How much did you enjoy eating the brownie’ (on a scale of 1-7, 1 being ‘not at all’, 7 being ‘a great deal’) and ‘How good did the brownie taste’ (on a scale of 1-7, 1 being ‘very bad’, 7 being ‘very good’).  -->
<!-- 1.3.	More information: Observational studies and meta-analyses will include only measured variables. As with the previous questions, the answers here must be precise. For example, 'intelligence,' 'accuracy,' 'aggression,' and 'color' are too vague. Acceptable alternatives could be 'IQ as measured by Wechsler Adult Intelligence Scale' 'percent correct,' 'number of threat displays,' and 'percent reflectance at 400 nm.' -->

### Indices
<!-- 1.1.	If any measurements are  going to be combined into an index (or even a mean), what measures will you use and how will they be combined? Include either a formula or a precise description of your method. If your are using a more complicated statistical method to combine measures (e.g. a factor analysis), you can note that here but describe the exact method in the analysis plan section. -->
<!-- 1.2.	Example: We will take the mean of the two questions above to create a single measure of ‘brownie enjoyment.’  -->
<!-- 1.3.	More information: If you are using multiple pieces of data to construct a single variable, how will this occur? Both the data that are included and the formula or weights for each measure must be specified. Standard summary statistics, such as “means” do not require a formula, though more complicated indices require either the exact formula or, if it is an established index in the field, the index must be unambiguously defined. For example, “biodiversity index” is too broad, whereas “Shannon’s biodiversity index” is appropriate. -->

## Analysis plan

<!-- You may describe one or more confirmatory analysis in this preregistration. Please remember that all analyses specified below must be reported in the final article, and any additional analyses must be noted as exploratory or hypothesis generating. -->

<!-- A confirmatory analysis plan must state up front which variables are predictors (independent) and which are the outcomes (dependent), otherwise it is an exploratory analysis. You are allowed to describe any exploratory work here, but a clear confirmatory analysis is required.  -->

###	Statistical models
<!-- 1.1.	What statistical model will you use to test each hypothesis? Please include the type of model (e.g. ANOVA, multiple regression, SEM, etc) and the specification of the model (this includes each variable that will be included as predictors, outcomes, or covariates). Please specify any interactions, subgroup analyses, pairwise or complex contrasts, or follow-up tests from omnibus tests. If you plan on using any positive controls, negative controls, or manipulation checks you may mention that here. Remember that any test not included here must be noted as an exploratory test in your final article.  -->
<!-- 1.2.	Example:  We will use a one-way between subjects ANOVA to analyze our results. The manipulated, categorical independent variable is 'sugar' whereas the dependent variable is our taste index.  -->
<!-- 1.3.	More information: This is perhaps the most important and most complicated question within the preregistration. As with all of the other questions, the key is to provide a specific recipe for analyzing the collected data. Ask yourself: is enough detail provided to run the same analysis again with the information provided by the user? Be aware for instances where the statistical models appear specific, but actually leave openings for the precise test. See the following examples:  -->
<!-- 1.3.1.1.	If someone specifies a 2x3 ANOVA with both factors within subjects, there is still flexibility with the various types of ANOVAs that could be run. Either a repeated measures ANOVA (RMANOVA) or a multivariate ANOVA (MANOVA) could be used for that design, which are two different tests.  -->
<!-- 1.3.1.2.	If you are going to perform a sequential analysis and check after 50, 100, and 150 samples, you must also specify the p-values you’ll test against at those three points. -->
<!-- 1.	Transformations -->
<!-- 1.1.	If you plan on transforming, centering, recoding the data, or will require a coding scheme for categorical variables, please describe that process. -->
<!-- 1.2.	Example: The “Effect of sugar on brownie tastiness” does not require any additional transformations. However, if it were using a regression analysis and each level of sweet had been categorically described (e.g. not sweet, somewhat sweet, sweet, and very sweet), ‘sweet’ could be dummy coded with ‘not sweet’ as the reference category.  -->
<!-- 1.3.	More information: If any categorical predictors are included in a regression, indicate how those variables will be coded (e.g. dummy coding, summation coding, etc.) and what the reference category will be. -->

###		Inference criteria
<!-- 2.1.	What criteria will you use to make inferences? Please describe the information youÍll use (e.g. p-values, bayes factors, specific model fit indices), as well as cut-off criterion, where appropriate. Will you be using one or two tailed tests for each of your analyses? If you are comparing multiple conditions or testing multiple hypotheses, will you account for this? -->
<!-- 2.2.	Example: We will use the standard p<.05 criteria for determining if the ANOVA and the post hoc test suggest that the results are significantly different from those expected if the null hypothesis were correct. The post-hoc Tukey-Kramer test adjusts for multiple comparisons.  -->
<!-- 2.3.	More information: P-values, confidence intervals, and effect sizes are standard means for making an inference, and any level is acceptable, though some criteria must be specified in this or previous fields. Bayesian analyses should specify a Bayes factor or a credible interval. If you are selecting models, then how will you determine the relative quality of each? In regards to multiple comparisons, this is a question with few “wrong” answers. In other words, transparency is more important than any specific method of controlling the false discovery rate or false error rate. One may state an intention to report all tests conducted or one may conduct a specific correction procedure; either strategy is acceptable. -->

###		Data exclusion
<!-- 3.1.	How will you determine what data or samples, if any, to exclude from your analyses? How will outliers be handled? Will you use any awareness check? -->
<!-- 3.2.	Example: No checks will be performed to determine eligibility for inclusion besides verification that each subject answered each of the three tastiness indices. Outliers will be included in the analysis.  -->
<!-- 3.3.	More information: Any rule for excluding a particular set of data is acceptable. One may describe rules for excluding a participant or for identifying outlier data. -->

###	Missing data
<!-- 4.1.	How will you deal with incomplete or missing data? -->
<!-- 4.2.	Example: If a subject does not complete any of the three indices of tastiness, that subject will not be included in the analysis. -->
<!-- 4.3.	More information: Any relevant explanation is acceptable. As a final reminder, remember that the final analysis must follow the specified plan, and deviations must be either strongly justified or included as a separate, exploratory analysis. -->

###	Exploratory analysis
<!-- 5.1.	If you plan to explore your data set to look for unexpected differences or relationships, you may describe those tests here. An exploratory test is any test where a prediction is not made up front, or there are multiple possible tests that you are going to use. A statistically significant finding in an exploratory test is a great way to form a new confirmatory hypothesis, which could be registered at a later time.  -->
<!-- 5.2.	Example: We expect that certain demographic traits may be related to taste preferences. Therefore, we will look for relationships between demographic variables (age, gender, income, and marital status) and the primary outcome measures of taste preferences. -->

##	Other
<!-- 6.1.	If there is any additional information that you feel needs to be included in your preregistration, please enter it here. Literature cited, disclosures of any related work such as replications or work that uses the same data, or other context that will be helpful for future readers would be appropriate here.  -->

### 2019-10-14 manuscript.Rmd
prereg in progress

@@
 ---
-title: "Non-word repetition among the Tsimane': A small-scale study"
-shorttitle        : "NWR among the Tsimane'"
+title: "Non-word repetition in Yélî Dnye: A small-scale study"
+shorttitle        : "NWR in Yélî Dnye'"
 
@@
     email         : "alecristia@gmail.com"
-  - name          : "Steven Levinson"
-    affiliation   : "2"
   - name          : "Marisa Casillas"
     affiliation   : "2"
+    address       : "Wundtlaan 1, 6525 XD, Nijmegen, The Netherlands"
+    email         : "marisa.casillas@mpi.nl"
 
@@
   - id            : "2"
-    institution   : "MPI"
+    institution   : "Max Planck Institute for Psycholinguistics"
 
@@
 
-Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in clise-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
+Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
 
@@
 - some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
-- calculation of frequency in Yêli is challenging bc low resource language
-- so we only manipulated structural complexity, not probability
+- calculation of frequency in Yêli is challenging because it is a low-resource language 
+- so we only manipulated structural complexity, not probability, in the design of the stimuli
 - previous work seems to avoid difficult sounds
@@
 - inconsistent results RE effects of SES (including parental education) & child sex
+- bilingual disadvantage (could be due to lower exposure to input; learning 2 phonologies; learning 2 lexica)
 
@@
 - How frequent are errors that result in real words? Is that a function of item complexity?
-- Is individual variation explainable by child age, sex, parental education?
+- Is individual variation explainable by child age, sex, parental education, birth order, monolingual status?
 
@@
 2.
-Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
-3. Children's accuracy increases with child age
-4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language)
+Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds.
+Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
+3. Children's accuracy increases with child age.
+4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
+<!-- **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)** ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)-->
 5.
-Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
-6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperfomr later-born children
+Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
+6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
 
@@
 The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
-Therefore, this may constitute a pressure for children to have finer aduitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
+Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
+As a result, differences between easier and harder items may be smaller in this work than in other research.
 **no work looking at consonants & vowels?
@@
 3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
-4.
-Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin (or at least from people other than their mothers, who tend to be the non-native speakers).
-If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work.
-That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
+4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family 
+<!-- **(MC: definitely almost all kin; 5k of them)** ac2mc I'm not sure this is how bio anthropologists think about kin; because it's about how you attribute resources to boost your genes' chances of being passed on relative to those of others in the same space.
+So my guess is that there is still a difference between nuclear family and not.
+Rewrote kin -> nuclear family -->
+(or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
+<!-- **(MC: maybe it'd be more useful to try and estimate the number of non-Y moms who live in each hamlet where testing is taking place?)** ac2mc not sure it's worth the effort, in any case other female adults are almost certain to have very little impact on the target child...
+Perhaps I'm not saying something very clearly?
+-->
+. That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother 
+<!-- **(MC: i can get you actual estimates for this for YD; let's go with that instead!)** -->
+<!-- ac2mc I think I meant your own paper! added "on the same population "-->
+, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
 5.
@@
 - By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
-- Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear) on the first 11 participants, and who segmented all children's production.
+- Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear **(MC: i guess this isn't true anymore given all the Anna data?)**) on the first 11 participants, and who segmented all children's production.
 Therefore, a key condition of blinding is not met
@@
 Many  NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language.
-We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language.
+We kept the same variation in item length and the non-meaningfulness requirement **(MC: before i forget, note that one of the stims, 'yi', is in fact a word, and a pretty frequent one at that (meaning 'tree'); i'm not sure how that happened, sorry!)**, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language.
 We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
@@
 As before, items were automatically excluded if they appeared in the dictionary.
-Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described (CITE), there could have been undocumented constraints that rendered items illegal.
+Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out **MC: i had the impression that you relaxed this short/long constraint in the end, but maybe i dreamt that :)**.
+Since the phonology and phonetics of Yélî is still in the process of being described (CITE), there could have been undocumented constraints that rendered items illegal **MC: well any intervocalic d that isn't a retroflex flap would be phonotactically illegal, i think---do we have any of those?
+i thought we maybe did but i don't remember and i don't remember, if so, what ghaalyu pronounced in his recording**.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
@@
 This "native coding" provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder.
+**MC: if we have anna's codes for everything, do we need your coding for anything?**
 
@@
 colnames(demo)
+subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
+
+summary(inc)
+
 read.table("log2.txt",header=T,sep="\t")->x
@@
 
+### Exploratory analyses: The role of phonotactic frequency
+
+**(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)**
+
 ## Discussion
@@
 McDonnell Foundation Understanding Human Cognition Scholar Award.
+MC blabla NWO Veni Innovational Scheme Grant (XXX-XX-XXX).
 

### 2019-10-14 prereg.Rmd
prereg in progress

@@
 
+In this study, we use nonword repetition (NWR), a technique that has been widely adopted to study the development of perception-production across multiple languages, particularly in Europe.
+In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning.
+The child should immediately try to say it back without changing anything.
+Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back).
+In our implementation of NWR, we additionally made sure that some of the items contained typologically rare and/or  challenging sounds, to additionally capture children's auditory-articulatory accuracy.
+
+Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
+
 ## Hypotheses
@@
 <!-- 1.2.	Example: If taste affects preference, then mean preference indices will be higher with higher concentrations of sugar.  -->
-<!-- Our hypothesis is that child-centered rearing (as a proxy for relative level of child-directed speech) is positively related to size of phoneme inventories: we predict that in languages in which a higher proportion of adult speakers provide with more responses of “child-centered rearing” responses, we find larger phoneme inventories than for languages with more responses of “parents have a life of their own”.
--->
+Based on previous work, we made the following predictions:
+
+1. **Length:** Children are more accurate for mono-syllables than longer items
+2.
+**Complexity:** Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
+3. **Age:** Children's accuracy increases with child age
+4. **Monolingual status:** Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
+<!-- **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)** ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)-->
+5.
+**Maternal education:** Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
+6. **Birth order:**  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
+
+Although these predictions were made based on previous work, we also thought reasonable that they would not obtain for the Rossel children in particular, as follows:
+
+1. The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English. 
+2.
+The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
+Therefore, this may constitute a pressure for children to have finer auitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
+3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
+4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family 
+(or at least from people other than their mothers, who tend to be the non-native speakers).
+If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work.
+That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
+5.
+In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
+This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
+However, if education directly boosts maternal verbal skills and the incidence of verbal behavior, then we should still see a difference along this factor.
+6.
+One main causal path between birth order and language development is via parental input.
+Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
 
@@
 
+This is both an experiment and an observational study.
+Questions pertaining to stimulus properties (length & complexity) can be considered experimental; the others bear on observing children's behavior albeit in a very constrained situation (age, monolingual status, maternal education, birth order).
+
 ### Blinding
@@
 <!-- 1.	Is there any additional blinding in this study? -->
+
+Blinding was partial, as follows:
+
+- Personnel who interacted directly with the participants did know whether an item was one or multiple syllables long, and sometimes of other features of the child (whether they were monolingual, their age), but was not aware of the finer-grained hypotheses.
+- Participants were not aware of the specific hypotheses
 
@@
 
+Regarding the experimental section (length & complexity), this is a within-participant repeated measures design.
+Regarding the observational section (age, monolingual status, maternal education, birth order), this is a cross-sectional design.
+
 ### Randomization
@@
 -->
+
+Regarding the experimental section (length & complexity), we did not randomize length but we did randomize complexity.
+Since the test is very short overall, it is unlikely that there would be a tiredness effect leading to lower performance towards the end.
+An improvement due to practice may have occurred.
+Since there are several items of the same length in a row, we can test for practice effect in such sections.
+The observational section (age, monolingual status, maternal education, birth order) was not overtly randomized; children were tested based on availability.
 
@@
 
+Registration prior to analysis of the data: As of the date of submission, the data exist and you have accessed it, though no analysis has been conducted related to the research plan (including calculation of summary statistics).
+
+Please note that an analysis of a different aspect of the data were carried out: online coding by a non-native speaker, of 11 out of the 30+ participants.
+These only inspected accuracy as a function of age.
+
 ###	Explanation of existing data 
@@
 -->
+
+The data were collected by AC, and the coding by a native speaker was overseen by MC.
 
@@
 
-### 1.	Sample size
+Hamlets were visited and the word about the study was spread, so that children could come to be tested in exchange of a packet of biscuits.
+We asked for children aged between 4 and 10 years, but sometimes older children participated.
+Some children were coerced by parents or others to participate, which was obvious in that they did not repeat even real words or the did not repeat the prompts.
+Sometimes a crowd gathered around the study and children were shy, talking too low for the recording to be reliable.
+Therefore, our inclusion and exclusion criteria are:
+
+- They must have participated freely
+- They need to have repeated at least the first three trial items in order to be included
+- Children must be 11 years of age or younger
+
+### Sample size
 <!-- 1.1.	Describe the sample size of your study.
@@
 
-### 1.	Sample size rationale
+We tested all children who made themselves available during our visits to different hamplets, which were 55. Application of our inclusion criteria led to a sample size of 40.
+
+### Sample size rationale
 <!-- 1.1.	This could include a power analysis or an arbitrary constraint such as time, money, or personnel. -->
@@
 
-### 1.	Stopping rule
+Our sample size was motivated entirely for logistic reasons. Given this constraint, we revisited the literature to better understand which of the effects of interest we may be able to find or not.
+
+To begin with, we looked at previous work.
+NWR studies vary in terms of how many children are included.
+Farabolini (2018) carried out a study of the literature resulting in two summary tables.
+The first summary table reviews literature on vocabulary-NWR correlations among bilinguals, showing that in 6 studies, the number of participants ranged between 19 and 105, with a mean of 48.7 and a median of 35.5.
+The second table reviews 13 studies on correlations between language exposure and NWR among bilinguals, with a median sample size of 52, mean 68, range 39-156.
+Our study is clearly on the small side in this literature.
+
+Second, we looked for the size of effects in our research questions:
+
+```{r len}
+# Vance 2005 - 5yo because midpoint of their distribution
+m1syll=76.3 
+sd1syll=8.4
+
+m3syll=73.5 
+sd3syll=12.0
+
+```
+
+```{r comp}
+# Kapalkova high versus low wordlike -- 4yo midpoint
+m1difficult=15.57 
+sd1difficult=3.21
+
+m3easy=17.57  
+sd3easy=2.57
+
+```
+
+```{r age}
+# Vance 2005 -- 5 vs 7 because we also have these
+m5yov=77.08
+sd5yov=5.92
+
+m7yov=81.42
+sd7yov=7.77
+
+# Kapalkova  -- 4 vs 5 although we only have a few of these
+m4yok=13.05 
+sd4yok=3.02
+
+m5yok=13.6
+sd5yok=2.64
+
+```
+
+```{r monobi}
+# Lotem & Chiat
+bil1.m=.86
+bil1.sd=.09
+
+mon1.m=.85
+mon1.sd=.09
+
+bil2.m=.82
+bil2.sd=.13
+
+mon2.m=.8
+mon2.sd=.14
+
+```
+
+1. **Length:**
+2. **Complexity:** 
+3. **Age:** 
+4. **Monolingual status:** 
+5. **Maternal education:** 
+6. **Birth order:**
+
+### Stopping rule
 <!-- 1.1.	If your data collection procedures do not give you full control over your exact sample size, specify how you will decide when to terminate your data collection.  -->
@@
 -->
+
+Not relevant, our sample size is fixed.
 

### 2019-10-14 prereg.Rmd
prereg in progress

@@
 
-Our sample size was motivated entirely for logistic reasons. Given this constraint, we revisited the literature to better understand which of the effects of interest we may be able to find or not.
+Our sample size was motivated entirely for logistic reasons.
 
-To begin with, we looked at previous work.
+#### How big are the effects we can detect given our sample size?
+
+In this section, we look at what are the effect sizes that can be found given this N.
+Our experimental design (length, complexity) is based on within-participant comparisons.
+As explained in the section on Variables, Length is coded linearly in terms of number of syllables 1-4; complexity is a binomial variable.
+As explained in the section on Variables, they will be analyzed using mixed models.
+Since it is complex to look at power in mixed models, we start with simpler effect sizes.
+For this design, d between the shortest and longest or simple versus complex is a first approximation.
+Therefore we looked at the smallest detectable d, with a significance level of .05, a power of 80%, and a one-sided alternative (since we predict higher scores for shorter/simpler items) given a one-sample test, using the pwr package in R:
+
+```{r pwr-d}
+library(pwr)
+pwr.t.test(n=40,power=0.80,sig.level=.05,type="one",alternative="greater")
+```
+
+Our correlational design (age, monolingual status, maternal education, birth order) also lends itself to analysis via mixed models, but as above we start with simpler effect sizes.
+For this design, the Pearson correlation coefficient r is a first approximation.
+Therefore we looked at the smallest detectable r, with a significance level of .05, a power of 80%, and a defined direction (higher scores for older, monolinguals, higher educated mothers, firstborns) using the pwr package in R:
+
+```{r pwr-r}
+pwr.r.test(n=40,power=0.80,sig.level=.05,alternative="greater")
+```
+
+Thus, we can theoretically detect d's of above .4 and rs of above .38. However, we will return to this issue given the multiplicity of research questions.
+
+#### How big are effects found in previous work?
+
+We looked at sample sizes in previous work.
 NWR studies vary in terms of how many children are included.
@@
 Our study is clearly on the small side in this literature.
+This may be an indication that individual performance is unstable, or that the task is noisy.
 
-Second, we looked for the size of effects in our research questions:
+Second, we looked for the size of effects in our research questions.
+AC carried out a scholar.google.com search on 2019-10-14 and inspected the first 20 hits for the search terms *non-word-repetition nwr children "bilingual"*.
+Papers were excluded because: pdf not directly accessible (5), they reported only on children with language impairment (4), only on bilinguals (1), they were a review paper with no new data (1), children were adolescents (1), resulting on 8 papers being considered.
+Next, she looked through the top 20 hits for the terms *non-word-repetition nwr children ("maternal education" OR "socio-economic-status")*.
+Papers were excluded because: pdf not directly accessible (6), they reported on children that were too young (1), did not contain NWR results (2), compared language delayed against bilingual (1), the typical group had been reported on in a different paper (1), resulting on 9 additional papers being considered.
+In addition, we included 4 papers that we had come across in previous searches, for a total of 23 papers included in the present review.
+
+##### Item length effects
 
@@
 
+##### Item complexity effects
+
 ```{r comp}
@@
 ```
+
+##### Child age effects
 
@@
 
+##### Maternal education effects
+
+```{r ed}
+
+```
+##### Monolingualism effects
+
 ```{r monobi}
@@
 
-1. **Length:**
-2. **Complexity:** 
-3. **Age:** 
-4. **Monolingual status:** 
-5. **Maternal education:** 
-6. **Birth order:**
+##### Birth order effects
+
+AC carried out a scholar.google.com search on 2019-10-14 and inspected the 11 hits for the search items *non-word-repetition nwr children "birth order"* based on title and the text snippet provided by google; as well as full text of the four hits for which google provided a pdf link.
+None of these contained relevant information.
+
+However, a recent paper by Havron et al (2019) estimated the birth order effect at d=.2, which is too small to be detectable given our sample size.
 

### 2019-10-14 manuscript.Rmd
add quick comments on outline

--- 2019-10-11 manuscript.Rmd
@@
 - some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
-- calculation of frequency in Yêli is challenging bc low resource language
+- calculation of frequency in Yêli is challenging bc low resource language **(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)**
 - so we only manipulated structural complexity, not probability
@@
 3. Children's accuracy increases with child age
-4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language)
+4.
+Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language) **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)**
 5.
@@
 4.
-Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin (or at least from people other than their mothers, who tend to be the non-native speakers).
-If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work.
-That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
+Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin **(MC: definitely almost all kin; 5k of them)** (or at least from people other than their mothers, who tend to be the non-native speakers).
+If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work **(MC: maybe it'd be more useful to try and estimate the number of non-Y moms who live in each hamlet where testing is taking place?)**.
+That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother **(MC: i can get you actual estimates for this for YD; let's go with that instead!)**, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
 5.

### 2019-10-29 manuscript.Rmd
fleshing out

--- 2019-10-14 manuscript.Rmd
@@
 
-- many studies on phonological development in childhood
-- they show phonology continues developing as children learn more words and gain mastery of phonetic targets
-- one key task in this research in NWR
-- used for theory (eg what are the links between phonology, working memory, and lexicon?), and for applications (nwr as a diagnostic)
-- it has been used acros a wide array of languages, particularly in Europe
-- however, less is known about NWR development in non-Indo European languages
-- this study: NWR in a Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory
-
-In this study, we use nonword repetition (NWR), a technique that has been widely adopted to study the development of perception-production across multiple languages (CITE), particularly in Europe (CITE).
+Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., CITE).
+One common task in this line of research is nonword repetition (NWR).
 In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning.
@@
 Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back).
-In our implementation of NWR, we additionally made sure that some of the items contained typologically rare and/or  challenging sounds, to additionally capture children's auditory-articulatory accuracy.
 
-Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
+NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon [CITE], as well as for applications, notably using nwr as a diagnostic for language delays and disorders [CITE].
+Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe [CITE].
 
-### Intro to the language
+In this study, we use NWR to measure perception-production in children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+In our implementation of NWR, we  made sure that some of the items contained typologically rare and/or  challenging sounds.
+
+### Intro to the language (@mc) - please feel free to throw away anything that is not useful!
 
@@
 
-### Intro to the people
+### Intro to the people (@mc) - please feel free to throw away anything that is not useful!
+
+Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
 
@@
 
-- typical structure of NWR tasks: N of items
-- complexity varies widely across tasks, sometimes within tasks on purpose
-- some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
-- calculation of frequency in Yêli is challenging because it is a low-resource language 
-- so we only manipulated structural complexity, not probability, in the design of the stimuli
-- previous work seems to avoid difficult sounds
-- but we felt this was important to represent the language, so we also varied this factor
+Although the procedure tends to be quite stable, there is some variation in the presentation and structure of items found in previous NWR work.
+For example, items are often presented orally by the experimenter [CITE], although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation [CITE].
+Additionally, while some studies have used 10-15 non-words, others have employed up to 30 unique items [CITE].
+Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) as well as syllable structure [open as opposed to closed syllables, CITE].
 
-- previous work shows even 2yo can do the task
-- but there are striking changes with age
-- inconsistent results RE effects of SES (including parental education) & child sex
-- bilingual disadvantage (could be due to lower exposure to input; learning 2 phonologies; learning 2 lexica)
+Previous work seems to avoid difficult sounds, but we felt this was important to represent the language, so we also varied this factor.
+We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
+This allowed us to get information about repetition accuracy of more items.
 
-- brief summary of inconsistent results SES
+Naturally, designing the task in this way may render the study of individual variaiton within the population  more difficult because different children are exposed to different items.
+However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
+That said, we contribute to the literature by also reporting descriptive analyses that could potentially be integrated in meta-analytic efforts.
+Based on previous work, we looked at potential improvements with age [CITE], and potential negative effects of bilingual exposure [CITE].
+Previous work typically finds no significant differences as a function of maternal education [see CITE but also CITE] or child gender [CITE].
+Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [CITE], and therefore we report on those too.
 
-- idem for sex
+### Research questions
 
-### Research questions and predictions
+We sought to address the following questions:
 
-- What is the accuracy (whole word, phoneme based, distance) overall?
-- How does this change as a function of item complexity (nb syll, syll complexity, sound complexity)?
-- How frequent are errors that result in real words? Is that a function of item complexity?
-- Is individual variation explainable by child age, sex, parental education, birth order, monolingual status?
+- What is the overall repetition accuracy (whole word, phoneme based, distance)?
+- How does this change as a function of item complexity (number of syllablex, sound complexity)?
+- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
 
-Based on previous work, we made the following predictions:
+In view of the hypothesis-driven nature of this work, we consider boosting the interpretational value of these data by announcing our analysis plans prior to conducting them.
+However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects.
+For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small.
+**add table with r/ds from previous work**)
+Moreover, by virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies.
+Therefore, all analyses here are descriptive and should be considered exploratory.
+
+@ac ADD TABLE SUMMARIZING PREVIOUS WORK
+
+- age differences
+- length effects
+
+## Methods
+
+### Stimuli
+
+Many  NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language.
+We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language.
+We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
+
+A first list of candidate items was generated in 2018 by selecting simple consonants ("p",	"t",	"d",	"k",	"m",	"n",	"w",	"y") and vowels ("i", "o","u", "a","e") that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
+These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
+MC presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
+Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
+
+A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce consonant-vowel monosyllables.
+As before, items were automatically excluded if they appeared in the dictionary.
+Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
+Since the phonology and phonetics of Yélî is still in the process of being described [CITE], there could have been undocumented constraints that rendered items illegal.
+Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
+These candidates were presented to one informant, for a final check that they did not mean anything.
+Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally.
+The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
+Even so, there was one monosyllable that was often identified as a real word (intended "yî" /yXX/; identified as "yi" /yi/, *tree*).
+This item is removed from analyses.
+
+@mc can you please add the IPA in // for all of the items below?
+The final list is composed of three practice items (wî, poni, nopimade); 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives (
+dpa
+dp:a
+dpâ
+dpéé
+dpê
+dpi
+dpu
+kp:ââ
+kpu
+tp:a
+tpâ
+tpê
+gh:ââ
+ghuu
+lva
+lv:ê
+lvi
+t:êê
+); 8 bisyllables (
+kamo
+kipo
+kani
+tupa
+noki
+piwa
+towi
+nomi);
+12 trisyllables (
+nademo
+meyadi
+diyeto
+widone
+nuyedi
+tumowe
+pedumi
+dimope
+tiwune
+wumipo
+nayeki
+mituye); and 4 quadrisyllables (
+todiwuma
+wadikeno
+nomiwake
+diponate).
+
+A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
+The 40 elicitation sets are available online from https://osf.io/5qspb.
+The split had the following constraints:
+
+- the same three items were selected as practice items and used in all 40 elicitation sets
+- splits were done within each length group from the 2018 items (i.e., separately for 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
+Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
+- once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2_syllables, then 3-syllables, and finally 4-syllable items.
+
+### Procedure
+
+We tried to balance three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them; that children would feel comfortable doing this task with us; and that the community would feel safe with us doing this task with their children.
+Moreover, there were also some logistic constraints in terms of the space availability.
+As a result, the places where elicitation took place varied across the hamlets.
+
+Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items 'spreading' through hearsay.
+In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own.
+The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
+
+The child was donned a headset (xx @mc please fill in for most of the children, xx @ac fill in for the rest).
+For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
+A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
+An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.
+
+The first phase was making sure the child understood the task.
+This was explained orally and the first training item was presented.
+Often, children froze and did not say anything.
+If this happened, then we followed this procedure.
+First, the informant insisted.
+If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another.
+If the child could repeat these correctly, then we provided the recorded training item over headphones again.
+Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it).
+A small minority still failed to repeat the item after hearing it over headphones.
+If that occurred, we tried with the second training item, at which point some children got it and could continue.
+A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.
+
+NWR studies vary in whether children are provided with several opportunities to hear and say the item.
+To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them.
+If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a headset that was recording a separate track).
+If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.
+
+Whenever siblings from the same family were tested, an attempt was made to test first the older and then the younger child, and always on different elicitaiton sets.
+
+### Coding
+
+There is an implicit accuracy coding occuring online as a function of whether the local informant asked for the item to be repeated or not.
+We will call this "online accuracy", and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
+
+We sought additional information by asking a local informant to listen through each child production, paired with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated.
+We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
+This "native coding" provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.
+
+### Analyses
+
+For NWR accuracy, we considered separately the child's first and subsequent attempts.
+These were scored as correct or incorrect on the fly depending on whether the native experimenter asked for the item to be repeated; as well as offline.
+Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy [CITE].
+We calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
+Although previous work does not use distance metrics, we additionally report those.
+
+Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
+We describe the proportion of items where the change resulted in a real word (semantic errors); and classify the most common phonological errors.
+
+### Participants
+
+```{r cars}
+# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
+# fulldemo=fulldemo[,-2] #remove children's names
+# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
+# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
+# fulldemo$age.exact=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
+# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
+# plot(fulldemo$Age~fulldemo$age.exact)
+# write.csv(fulldemo,"nwr-demo.csv")
+
+read.csv("NWR-demo.csv",header=T)->demo
+#colnames(demo)
+
+#if family is missing, attribute each child to a different one
+demo$familyID=as.character(demo$familyID)
+demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])
+
+subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
+#if age.exact is missing, use reported age
+inc$age.exact[is.na(inc$age.exact)]<-inc$Age[is.na(inc$age.exact)]
+#summary(inc)
+table(inc$birthOrder)->mytab
+```
+
+A total `r dim(demo)[1]` children were tested, from `r length(levels(factor(demo$familyID)))` of families, in five hamlets.
+Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xd5t repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`).
+In addition, `r sum(demo$Age >12 | demo$included=="too old",na.rm=T)` teenagers were tested to put younger children at ease; their data is not included in analyses below.
+The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.exact),2)` years (range `r round(min(inc$age.exact),2)`-`r round(max(inc$age.exact),2)` years).
+There were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and , `r sum(is.na(inc$monolingual))` for whom this information was missing.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing).
+In terms of birth order, `r mytab[1]` were first borns, `r mytab[2]` second,  `r mytab[3]` third,  `r mytab[4]` forth,  `r mytab[5]` fifth, `r mytab[6]` sixth, and  `r mytab[7]` did not have this information.
+
+@mc why is maternal education always in even years? Can you add a footnote to explain?
+
+## Results
+
+- What is the overall repetition accuracy (whole word, phoneme based, distance)?
+
+- How does this change as a function of item complexity (number of syllablex, sound complexity)?
+
+- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+
+- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
+
+```{r}
+read.table("log2.txt",header=T,sep="\t")->x
+x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
+x[order(x$out),]->x
+x$cor.online=1
+for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0
+
+subset(x,rep=="1")->x
+aggregate(x$cor.online,by=list(x$child),mean)->onl
+colnames(onl)<-c("ID","score.online")
+
+print(onl)
+
+aggregate(x$corr,by=list(x$child),mean)->ms
+colnames(ms)<-c("ID","score")
+merge(ms,onl)->ms
+merge(ms,demo)->scores
+scores$Age=as.numeric(as.character(scores$Age))
+scores$Age[is.na(scores$Age)]<-15
+plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
+abline(lm(scores$score*100~scores$Age))
+plot(scores$score~scores$monolingual)
+points(scores$score~as.numeric(scores$monolingual),pch=20)
+
+plot(ms$score.online~ms$score)
+```
+
+**(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)**
+
+Based on previous work, one can make the following general predictions for any population:
 
@@
 2.
-Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds.
+Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds [CITE].
 Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
@@
 5.
-Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
+As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
 6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
@@
 
-## Pre-registration and planned analyses
+## Discussion
 
-In view of the hypothesis driven nature of this work, we intended to boost the interpretational value of these data by announcing our analysis plans prior to conducting them.
-There are a few limitations in our approach:
-
-- The sample size is too small to have enough power to detect all of the relationships of interest
-- In fact, the sample size is too small to detect some of the known effects (see Table XX **add table with r/ds from previous work**)
-- By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
-- Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear **(MC: i guess this isn't true anymore given all the Anna data?)**) on the first 11 participants, and who segmented all children's production.
-Therefore, a key condition of blinding is not met
-
-Nonetheless, some form of pre-registration still seemed preferrable over the complete lack of it.
-
-The pre-registration can be found XXX
-
-## Stimuli
-
-Many  NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language.
-We kept the same variation in item length and the non-meaningfulness requirement **(MC: before i forget, note that one of the stims, 'yi', is in fact a word, and a pretty frequent one at that (meaning 'tree'); i'm not sure how that happened, sorry!)**, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language.
-We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
-
-A first list of candidate items was generated in 2018 by selecting simple consonants (LIST) and vowels (LIST) that were combined into CV syllables, further sampling the space of 1- to 4-syllable sequences.
-These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
-They were further presented orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
-Any item for which two or more of the assistants reported them having a meaning or some form of association were was excluded.
-
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce CV monosyllables.
-As before, items were automatically excluded if they appeared in the dictionary.
-Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out **MC: i had the impression that you relaxed this short/long constraint in the end, but maybe i dreamt that :)**.
-Since the phonology and phonetics of Yélî is still in the process of being described (CITE), there could have been undocumented constraints that rendered items illegal **MC: well any intervocalic d that isn't a retroflex flap would be phonotactically illegal, i think---do we have any of those?
-i thought we maybe did but i don't remember and i don't remember, if so, what ghaalyu pronounced in his recording**.
-Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
-These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally.
-The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
-
-The final list is XX.
-
-A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets. The split had the following constraint:
-
-- the same 3 items were selected as practice items and used in all 40 elicitation sets
-- splits were done within each length group from the 2018 items (i.e., separately for 1, 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
-Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
-- once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2_syllables, then 3-syllables, and finally 4-syllable items.
-
-The 40 elicitation sets are available online from XXX.
-
-## Procedure
-
-We tried to balance three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them; that children would feel comfortable doing this task with us; and that the community would feel safe with us doing this task with their children.
-Moreover, there were also some logistic constraints in terms of the space availability.
-As a result, the places where elicitation took place varied across the hamlets.
-
-Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items 'spreading' through hearsay.
-In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own.
-The complete list of places and the ways in which they met the desiderata mentioned above can be found in supplementary materials, under comments for each child (direct link xXX).
-
-The child was donned a headset (xx for most of the children, xx for the rest).
-For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the 3 practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.
-
-The first phase was making sure the child understood the task.
-This was explained orally and the first training item was presented.
-Often, children froze and did not say anything.
-If this happened, then we followed this procedure.
-First, the informant insisted.
-If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another.
-If the child could repeat these correctly, then we provided the recorded training item over headphones again.
-Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it).
-A small minority still failed to repeat the item after hearing it over headphones.
-If that occurred, we tried with the second training item, at which point some children got it and could continue.
-A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.
-
-NWR studies vary in whether children are provided with several opportunities to hear and say the item.
-To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them.
-If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a headset that was recording a separate track).
-If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.
-
-Whenever siblings from the same family were tested, an attempt was made to test first the older and then the younger child, and always on different elicitaiton sets.
-
-## Coding
-
-There is an implicit accuracy coding occuring online as a function of whether the local informant asked for the item to be repeated or not.
-We will call this "online accuracy", and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
-
-Additionally, the experimenter listened through all recordings and segmented out every child production.
-For the first 11 participants, she additionally made a judgment of whether the item was correctly or incorrectly repeated, and transcribed what she heard, which provides both an accuracy measure and a qualitative description of what children were actually saying in response to each stimulus.
-.
-We call this "non-native coding," because this experimenter is not familiar with Yélî and has insufficient training on Yélî sounds, including common allophonic variants, not to mention not having some of the sounds in her native perceptual inventory.
-
-Therefore, we sought additional information by asking a local informant to listen through each child production, paried with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated.
-We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-This "native coding" provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder.
-**MC: if we have anna's codes for everything, do we need your coding for anything?**
-
-## Analyses
-
-For NWR accuracy, we considered only the child's first attempt, which was scored as correct or incorrect online, by a non-native coder (ofr the first 11 kids), and a native coder.
-
-Some NWR studies employ phoneme accuracy (rather than whole word accuracy).
-We used exclusively the native coding for these, and we calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
-
-Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account, again using only the native coder's transcriptions.
-
-## Participants
-
-Total N of kids, N of families, N of villages.
-Some children could not be included for the following reasons: refused participation (N=2), failed to repeat items presented over headphones even after coaching (N=3), spoke too softly to allow offline coding (N=8).
-In addition, one teenager was tested to put children at ease; their data is not included in analyses below.
-The remaining XX exhibited the following demographic characteristics
-
-INSERT TABLE WITH THE KEY VARS MENTIONED ABOVE:
-
-- child age
-- monoling
-- mat ed
-- birth order
-
-```{r cars}
-# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
-# fulldemo=fulldemo[,-2] #remove children's names
-# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
-# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
-# fulldemo$age.exact=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
-# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
-# plot(fulldemo$Age~fulldemo$age.exact)
-# write.csv(fulldemo,"nwr-demo.csv")
-
-read.csv("NWR-demo.csv",header=T)->demo
-colnames(demo)
-subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
-
-summary(inc)
-
-read.table("log2.txt",header=T,sep="\t")->x
-x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
-x[order(x$out),]->x
-x$cor.online=1
-for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0
-
-subset(x,rep=="1")->x
-aggregate(x$cor.online,by=list(x$child),mean)->onl
-colnames(onl)<-c("ID","score.online")
-
-print(onl)
-
-aggregate(x$corr,by=list(x$child),mean)->ms
-colnames(ms)<-c("ID","score")
-merge(ms,onl)->ms
-merge(ms,demo)->scores
-scores$Age=as.numeric(as.character(scores$Age))
-scores$Age[is.na(scores$Age)]<-15
-plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
-abline(lm(scores$score*100~scores$Age))
-plot(scores$score~scores$monolingual)
-points(scores$score~as.numeric(scores$monolingual),pch=20)
-
-plot(ms$score.online~ms$score)
-
-```
-
-### Exploratory analyses: The role of phonotactic frequency
-
-**(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)**
-
-## Discussion
+- What is the overall repetition accuracy (whole word, phoneme based, distance)?
+- How does this change as a function of item complexity (number of syllablex, sound complexity)?
+- How frequent are errors that result in real words? Is that a function of item complexity?
+- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
 

### 2019-10-29 manuscript.tex
fleshing out

@@
 \hypersetup{unicode=true,
-            pdftitle={Non-word repetition among the Tsimane': A small-scale study},
-            pdfauthor={Alejandrina Cristia, Steven Levinson, \& Marisa Casillas},
+            pdftitle={Non-word repetition in Yélî Dnye: A small-scale study},
+            pdfauthor={Alejandrina Cristia~\& Marisa Casillas},
             pdfkeywords={phonology, non-word repetition},
@@
 
-  \title{Non-word repetition among the Tsimane': A small-scale study}
-    \author{Alejandrina Cristia\textsuperscript{1}, Steven Levinson\textsuperscript{2}, \& Marisa Casillas\textsuperscript{2}}
+  \title{Non-word repetition in Yélî Dnye: A small-scale study}
+    \author{Alejandrina Cristia\textsuperscript{1}~\& Marisa Casillas\textsuperscript{2}}
     \date{}
 
-\shorttitle{NWR among the Tsimane'}
+\shorttitle{NWR in Yélî Dnye'}
 \affiliation{
 \vspace{0.5cm}
-\textsuperscript{1} LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL\\\textsuperscript{2} MPI}
+\textsuperscript{1} LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
 \keywords{phonology, non-word repetition\newline\indent Word count: xxx}
@@
 
-\begin{itemize}
-\tightlist
-\item
-  many studies on phonological development in childhood
-\item
-  they show phonology continues developing as children learn more words and gain mastery of phonetic targets
-\item
-  one key task in this research in NWR
-\item
-  used for theory (eg what are the links between phonology, working memory, and lexicon?), and for applications (nwr as a diagnostic)
-\item
-  it has been used acros a wide array of languages, particularly in Europe
-\item
-  however, less is known about NWR development in non-Indo European languages
-\item
-  this study: NWR in a Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory
-\end{itemize}
-
-In this study, we use nonword repetition (NWR), a technique that has been widely adopted to study the development of perception-production across multiple languages (CITE), particularly in Europe (CITE).
+Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., CITE).
+One common task in this line of research is nonword repetition (NWR).
 In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning.
@@
 Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back).
-In our implementation of NWR, we additionally made sure that some of the items contained typologically rare and/or challenging sounds, to additionally capture children's auditory-articulatory accuracy.
 
-Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in clise-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
+NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon {[}CITE{]}, as well as for applications, notably using nwr as a diagnostic for language delays and disorders {[}CITE{]}.
+Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe {[}CITE{]}.
 
-\hypertarget{intro-to-the-language}{%
-\subsubsection{Intro to the language}\label{intro-to-the-language}}
+In this study, we use NWR to measure perception-production in children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+In our implementation of NWR, we made sure that some of the items contained typologically rare and/or challenging sounds.
+
+\hypertarget{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}{%
+\subsubsection{\texorpdfstring{Intro to the language (({\textbf{???}})) - please feel free to throw away anything that is not useful!}{Intro to the language ((???)) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
 
@@
 
-\hypertarget{intro-to-the-people}{%
-\subsubsection{Intro to the people}\label{intro-to-the-people}}
+\hypertarget{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}{%
+\subsubsection{\texorpdfstring{Intro to the people (({\textbf{???}})) - please feel free to throw away anything that is not useful!}{Intro to the people ((???)) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
+
+Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
 
@@
 
-\begin{itemize}
-\item
-  typical structure of NWR tasks: N of items
-\item
-  complexity varies widely across tasks, sometimes within tasks on purpose
-\item
-  some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
-\item
-  calculation of frequency in Yêli is challenging bc low resource language
-\item
-  so we only manipulated structural complexity, not probability
-\item
-  previous work seems to avoid difficult sounds
-\item
-  but we felt this was important to represent the language, so we also varied this factor
-\item
-  previous work shows even 2yo can do the task
-\item
-  but there are striking changes with age
-\item
-  inconsistent results RE effects of SES (including parental education) \& child sex
-\item
-  brief summary of inconsistent results SES
-\item
-  idem for sex
-\end{itemize}
+Although the procedure tends to be quite stable, there is some variation in the presentation and structure of items found in previous NWR work.
+For example, items are often presented orally by the experimenter {[}CITE{]}, although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation {[}CITE{]}.
+Additionally, while some studies have used 10-15 non-words, others have employed up to 30 unique items {[}CITE{]}.
+Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) as well as syllable structure {[}open as opposed to closed syllables, CITE{]}.
 
-\hypertarget{research-questions-and-predictions}{%
-\subsubsection{Research questions and predictions}\label{research-questions-and-predictions}}
+Previous work seems to avoid difficult sounds, but we felt this was important to represent the language, so we also varied this factor.
+We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
+This allowed us to get information about repetition accuracy of more items.
+
+Naturally, designing the task in this way may render the study of individual variaiton within the population more difficult because different children are exposed to different items.
+However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
+That said, we contribute to the literature by also reporting descriptive analyses that could potentially be integrated in meta-analytic efforts.
+Based on previous work, we looked at potential improvements with age {[}CITE{]}, and potential negative effects of bilingual exposure {[}CITE{]}.
+Previous work typically finds no significant differences as a function of maternal education {[}see CITE but also CITE{]} or child gender {[}CITE{]}.
+Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks {[}CITE{]}, and therefore we report on those too.
+
+\hypertarget{research-questions}{%
+\subsubsection{Research questions}\label{research-questions}}
+
+We sought to address the following questions:
 
@@
 \item
-  What is the accuracy (whole word, phoneme based, distance) overall?
+  What is the overall repetition accuracy (whole word, phoneme based, distance)?
 \item
-  How does this change as a function of item complexity (nb syll, syll complexity, sound complexity)?
+  How does this change as a function of item complexity (number of syllablex, sound complexity)?
 \item
-  How frequent are errors that result in real words? Is that a function of item complexity?
+  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
 \item
-  Is individual variation explainable by child age, sex, parental education?
+  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
 \end{itemize}
 
-Based on previous work, we made the following predictions:
+In view of the hypothesis-driven nature of this work, we consider boosting the interpretational value of these data by announcing our analysis plans prior to conducting them.
+However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects.
+For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small.
+\textbf{add table with r/ds from previous work})
+Moreover, by virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies.
+Therefore, all analyses here are descriptive and should be considered exploratory.
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Children are more accurate for mono-syllables than longer items
-\item
-  Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
-\item
-  Children's accuracy increases with child age
-\item
-  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language)
-\item
-  Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
-\item
-  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperfomr later-born children
-\end{enumerate}
-
-Although these predictions were made based on previous work, we also thought reasonable that they would not obtain for the Rossel children in particular, as follows:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
-\textbf{Check for work on European languages that may have looked into this}
-\item
-  The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
-Therefore, this may constitute a pressure for children to have finer aduitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
-\textbf{no work looking at consonants \& vowels?
-no work looking at nasal vowels in particular?}
-\item
-  We had no reservations that children's accuracy increases with child age for the Yélî community as well.
-\item
-  Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin (or at least from people other than their mothers, who tend to be the non-native speakers).
-If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work.
-That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
-\item
-  In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
-This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
-However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
-\item
-  One main causal path between birth order and language development is via parental input (CITE).
-Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
-\end{enumerate}
-
-\hypertarget{pre-registration-and-planned-analyses}{%
-\subsection{Pre-registration and planned analyses}\label{pre-registration-and-planned-analyses}}
-
-In view of the hypothesis driven nature of this work, we intended to boost the interpretational value of these data by announcing our analysis plans prior to conducting them.
-There are a few limitations in our approach:
+({\textbf{???}}) ADD TABLE SUMMARIZING PREVIOUS WORK
 
@@
 \item
-  The sample size is too small to have enough power to detect all of the relationships of interest
+  age differences
 \item
-  In fact, the sample size is too small to detect some of the known effects (see Table XX \textbf{add table with r/ds from previous work})
-\item
-  By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
-\item
-  Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear) on the first 11 participants, and who segmented all children's production.
-Therefore, a key condition of blinding is not met
+  length effects
 \end{itemize}
 
-Nonetheless, some form of pre-registration still seemed preferrable over the complete lack of it.
-
-The pre-registration can be found XXX
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
 
 \hypertarget{stimuli}{%
-\subsection{Stimuli}\label{stimuli}}
+\subsubsection{Stimuli}\label{stimuli}}
 
@@
 
-A first list of candidate items was generated in 2018 by selecting simple consonants (LIST) and vowels (LIST) that were combined into CV syllables, further sampling the space of 1- to 4-syllable sequences.
+A first list of candidate items was generated in 2018 by selecting simple consonants (\enquote{p}, \enquote{t}, \enquote{d}, \enquote{k}, \enquote{m}, \enquote{n}, \enquote{w}, \enquote{y}) and vowels (\enquote{i}, \enquote{o},\enquote{u}, \enquote{a},\enquote{e}) that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
 These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
-They were further presented orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
-Any item for which two or more of the assistants reported them having a meaning or some form of association were was excluded.
+MC presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
+Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
 
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce CV monosyllables.
+A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce consonant-vowel monosyllables.
 As before, items were automatically excluded if they appeared in the dictionary.
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described (CITE), there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described {[}CITE{]}, there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
@@
 The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
+Even so, there was one monosyllable that was often identified as a real word (intended \enquote{yî} /yXX/; identified as \enquote{yi} /yi/, \emph{tree}).
+This item is removed from analyses.
 
-The final list is XX.
+({\textbf{???}}) can you please add the IPA in // for all of the items below?
+The final list is composed of three practice items (wî, poni, nopimade); 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives (
+dpa
+dp:a
+dpâ
+dpéé
+dpê
+dpi
+dpu
+kp:ââ
+kpu
+tp:a
+tpâ
+tpê
+gh:ââ
+ghuu
+lva
+lv:ê
+lvi
+t:êê
+); 8 bisyllables (
+kamo
+kipo
+kani
+tupa
+noki
+piwa
+towi
+nomi);
+12 trisyllables (
+nademo
+meyadi
+diyeto
+widone
+nuyedi
+tumowe
+pedumi
+dimope
+tiwune
+wumipo
+nayeki
+mituye); and 4 quadrisyllables (
+todiwuma
+wadikeno
+nomiwake
+diponate).
 
-A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets. The split had the following constraint:
+A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
+The 40 elicitation sets are available online from \url{https://osf.io/5qspb}.
+The split had the following constraints:
 
@@
 \item
-  the same 3 items were selected as practice items and used in all 40 elicitation sets
+  the same three items were selected as practice items and used in all 40 elicitation sets
 \item
-  splits were done within each length group from the 2018 items (i.e., separately for 1, 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
+  splits were done within each length group from the 2018 items (i.e., separately for 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
 Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
@@
 
-The 40 elicitation sets are available online from XXX.
-
 \hypertarget{procedure}{%
-\subsection{Procedure}\label{procedure}}
+\subsubsection{Procedure}\label{procedure}}
 
@@
 In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own.
-The complete list of places and the ways in which they met the desiderata mentioned above can be found in supplementary materials, under comments for each child (direct link xXX).
+The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
 
-The child was donned a headset (xx for most of the children, xx for the rest).
+The child was donned a headset (xx ({\textbf{???}}) please fill in for most of the children, xx ({\textbf{???}}) fill in for the rest).
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the 3 practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
+A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
 An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.
@@
 \hypertarget{coding}{%
-\subsection{Coding}\label{coding}}
+\subsubsection{Coding}\label{coding}}
 
@@
 
-Additionally, the experimenter listened through all recordings and segmented out every child production.
-For the first 11 participants, she additionally made a judgment of whether the item was correctly or incorrectly repeated, and transcribed what she heard, which provides both an accuracy measure and a qualitative description of what children were actually saying in response to each stimulus.
-.
-We call this \enquote{non-native coding,} because this experimenter is not familiar with Yélî and has insufficient training on Yélî sounds, including common allophonic variants, not to mention not having some of the sounds in her native perceptual inventory.
-
-Therefore, we sought additional information by asking a local informant to listen through each child production, paried with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated.
+We sought additional information by asking a local informant to listen through each child production, paired with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated.
 We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-This \enquote{native coding} provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder.
+This \enquote{native coding} provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.
 
 \hypertarget{analyses}{%
-\subsection{Analyses}\label{analyses}}
+\subsubsection{Analyses}\label{analyses}}
 
-For NWR accuracy, we considered only the child's first attempt, which was scored as correct or incorrect online, by a non-native coder (ofr the first 11 kids), and a native coder.
+For NWR accuracy, we considered separately the child's first and subsequent attempts.
+These were scored as correct or incorrect on the fly depending on whether the native experimenter asked for the item to be repeated; as well as offline.
+Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy {[}CITE{]}.
+We calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
+Although previous work does not use distance metrics, we additionally report those.
 
-Some NWR studies employ phoneme accuracy (rather than whole word accuracy).
-We used exclusively the native coding for these, and we calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
-
-Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account, again using only the native coder's transcriptions.
+Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
+We describe the proportion of items where the change resulted in a real word (semantic errors); and classify the most common phonological errors.
 
 \hypertarget{participants}{%
-\subsection{Participants}\label{participants}}
+\subsubsection{Participants}\label{participants}}
 
-Total N of kids, N of families, N of villages.
-Some children could not be included for the following reasons: refused participation (N=2), failed to repeat items presented over headphones even after coaching (N=3), spoke too softly to allow offline coding (N=8).
-In addition, one teenager was tested to put children at ease; their data is not included in analyses below.
-The remaining XX exhibited the following demographic characteristics
+A total 55 children were tested, from 34 of families, in five hamlets.
+Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5).
+In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
+The remaining 40 children (14 girls) were aged 6.96 years (range 3.92-11.03 years).
+There were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and , 2 for whom this information was missing.
+Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).
+In terms of birth order, 0 were first borns, 5 second, 4 third, 2 forth, 6 fifth, 5 sixth, and 1 did not have this information.
 
-INSERT TABLE WITH THE KEY VARS MENTIONED ABOVE:
+({\textbf{???}}) why is maternal education always in even years? Can you add a footnote to explain?
+
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
 
 \begin{itemize}
-\tightlist
 \item
-  child age
+  What is the overall repetition accuracy (whole word, phoneme based, distance)?
 \item
-  monoling
+  How does this change as a function of item complexity (number of syllablex, sound complexity)?
 \item
-  mat ed
+  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
 \item
-  birth order
+  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
 \end{itemize}
-
-\begin{verbatim}
-##  [1] "X.3"         "ID"          "included"    "Age"         "DOB.ISO"    
-##  [6] "DOB"         "Sex"         "birthOrder"  "familyID"    "monolingual"
-## [11] "matEd"       "Cond"        "Date"        "Time"        "File"       
-## [16] "fastcode"    "fullcode"    "Notes"       "Where"       "Village"    
-## [21] "X"           "X.1"         "X.2"         "dob"         "test"       
-## [26] "age.exact"
-\end{verbatim}
 
@@
 
-\includegraphics{manuscript_files/figure-latex/cars-1.pdf} \includegraphics{manuscript_files/figure-latex/cars-2.pdf} \includegraphics{manuscript_files/figure-latex/cars-3.pdf}
+\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf} \includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-2.pdf} \includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-3.pdf}
+
+\textbf{(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)}
+
+Based on previous work, one can make the following general predictions for any population:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Children are more accurate for mono-syllables than longer items
+\item
+  Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds {[}CITE{]}.
+Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
+\item
+  Children's accuracy increases with child age.
+\item
+  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
+\item
+  As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
+\item
+  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
+\end{enumerate}
+
+Although these predictions were made based on previous work, we also thought reasonable that they would not obtain for the Rossel children in particular, as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
+\textbf{Check for work on European languages that may have looked into this}
+\item
+  The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
+Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
+As a result, differences between easier and harder items may be smaller in this work than in other research.
+\textbf{no work looking at consonants \& vowels?
+no work looking at nasal vowels in particular?}
+\item
+  We had no reservations that children's accuracy increases with child age for the Yélî community as well.
+\item
+  Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family
+  (or at least from people other than their mothers, who tend to be the non-native speakers).
+If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
+  . That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother
+  , so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
+\item
+  In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
+This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
+However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
+\item
+  One main causal path between birth order and language development is via parental input (CITE).
+Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
+\end{enumerate}
 
@@
 \subsection{Discussion}\label{discussion}}
+
+\begin{itemize}
+\tightlist
+\item
+  What is the overall repetition accuracy (whole word, phoneme based, distance)?
+\item
+  How does this change as a function of item complexity (number of syllablex, sound complexity)?
+\item
+  How frequent are errors that result in real words? Is that a function of item complexity?
+\item
+  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
+\end{itemize}
 

### 2019-10-30 manuscript.Rmd
add cit, prevlit

@@
 
-Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., CITE).
+Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood [e.g., @hazan2000development].
 One common task in this line of research is nonword repetition (NWR).
@@
 
-NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon [CITE], as well as for applications, notably using nwr as a diagnostic for language delays and disorders [CITE].
-Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe [CITE].
+NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon [@bowey2001nonword], as well as for applications, notably using nwr as a diagnostic for language delays and disorders [@estes2007differences].
+Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe [@meir2016disentangling,@is08042009language].
 
@@
 Although the procedure tends to be quite stable, there is some variation in the presentation and structure of items found in previous NWR work.
-For example, items are often presented orally by the experimenter [CITE], although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation [CITE].
-Additionally, while some studies have used 10-15 non-words, others have employed up to 30 unique items [CITE].
-Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) as well as syllable structure [open as opposed to closed syllables, CITE].
+For example, items are often presented orally by the experimenter [@torrington2015non], although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation [@brandeker2015language].
+Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items [@piazzalunga2019articulatory].
+Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
 
@@
 That said, we contribute to the literature by also reporting descriptive analyses that could potentially be integrated in meta-analytic efforts.
-Based on previous work, we looked at potential improvements with age [CITE], and potential negative effects of bilingual exposure [CITE].
-Previous work typically finds no significant differences as a function of maternal education [see CITE but also CITE] or child gender [CITE].
-Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [CITE], and therefore we report on those too.
+Based on previous work, we looked at potential improvements with age [@vance2005speech;@farmani2018normalization;@kalnak2014nonword], and potential negative effects of bilingual exposure [@brandeker2015language;@meir2016disentangling;@meir2017independent].
+Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization;@kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool].
+Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on those too.
 
@@
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described [CITE], there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described [CITE @mc please fill in], there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).

### 2019-10-30 manuscript.Rmd
results in progress

@@
 - What is the overall repetition accuracy (whole word, phoneme based, distance)?
-- How does this change as a function of item complexity (number of syllablex, sound complexity)?
+- How does this change as a function of item complexity (number of syllables, sound complexity)?
 - How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
@@
 
-@ac ADD TABLE SUMMARIZING PREVIOUS WORK
+```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "Proportion of non-words correctly repeated as a function of age (in years), study (first author and year regardless of the number of authors), and, when available, shorter versus longer non-words."}
+read.csv("prevlit.csv")->pl
+#duplicate Piazzalunga logging short and long items separately
+pl$Score[pl$Study=="Piazzalunga 2018"]<-pl$Short[pl$Study=="Piazzalunga 2018"]
+pl$Study<-as.character(pl$Study)
+pl$Study[pl$Study=="Piazzalunga 2018"]<-"Piazzalunga 2018 1-2 syll"
+dim(pl)[1]->beg
+pl=rbind(pl,pl[pl$Study=="Piazzalunga 2018 1-2 syll",])
+pl[(beg+1):dim(pl)[1],"Score"]<-pl[(beg+1):dim(pl)[1],"Long"]
+pl[(beg+1):dim(pl)[1],"Study"]<-"Piazzalunga 2018 3-4 syll"
+pl$Study=paste(pl$Language,pl$Study)
+library(lattice)
+xyplot(Score~Age,groups=Study,data=pl,type=c('l','p'),auto.key = T ,pch=20)
 
-- age differences
-- length effects
+```
 
@@
 
-```{r cars}
+```{r demo}
 # read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
@@
 
+```{r wrangling}
+#@mc look here - we probably need to read the table in a special way to get all the accents etc. right
+read.table("NWR-transcription.txt",header=T,enc="UTF-16le")->trnsc
+read.csv("NWR-transcription.csv",sep=";",enc="UTF-8")->trnsc
+trnsc$tokid=paste(trnsc$item,trnsc$token)
+#task 4 are probably the items with no target; there are 1488 lines of paired data
+
+read.table("final_order.txt",header=T)->crp
+crp$tokid=paste(crp$target,crp$nb)
+npairs=dim(crp)[1]
+
+trnsc=merge(trnsc,crp,by.x="tokid",by.y="tokid",all=T)
+
+# sort by file name which indicates presentation order, so that we can get the first attempt
+trnsc[order(trnsc$outfile),]->trnsc
+
+#add repetition number
+trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
+
+#add coding of correctness online/on the fly of first presentation as a function whether there are subsequent reps of the same target
+trnsc$cor.online=1
+for(i in 1:npairs) if(trnsc$rep[i]>1) trnsc$cor.online[i-1]<-0  #if the item is repeated, then code the previous one as being wrong
+trnsc$cor.online[(npairs+1):dim(trnsc)[1]]<-NA
+
+#DATA REMOVAL *** ATTENTION *** 
+#REMOVING ALL THE "YI"
+trnsc[trnsc$target!="yi",]->trnsc
+#REMOVE ALL TASK 4
+trnsc[!is.na(trnsc$target),]->trnsc
+
+```
+
+```{r phonoana}
+
+#convert target coding to orthography
+orthotarget=matrix( #list correspondances always by pairs: orthography then file name; for longer items orthography=filename
+  c("dpa","dpa",
+    "dp:a","dp_a",
+    "dpâ","dpa2",
+    "dpéé","dpee",
+    "dpê","dpe",
+    "dpi","dpi",
+    "dpu","dpu",
+    "kp:ââ","kp_aa",
+    "kpu","kpu",
+    "tp:a","tp_a",
+    "tpâ","tpa",
+    "tpê","tpe",
+    "gh:ââ","gh_aa",
+    "ghuu","ghuu",
+    "lva","lva",
+    "lv:ê","lv_e",
+    "lvi","lvi",
+    "t:êê","t_ee"
+  ),#last item above should not have a comma
+  ncol = 2,byrow = T)
+
+trnsc$target.ortho=as.character(trnsc$target) #to start with, put them all in
+for(i in 1:dim(orthotarget)[1]) trnsc$target.ortho[trnsc$target.ortho==orthotarget[i,2]] <- orthotarget[i,1] #then correct the monosyllables
+
+## TODO
+# make all vowels in mispronunciation & target short (so we don't penalize length errors)
+# if correct=0 but orthotarget=mispronunciation (probably due to length errors) then change correct to 1
+
+# create phonological correspondance matrix
+# NOTE!! NON INTUITIVE MAPPING!!!! DO NOT READ THE OUTPUT BELIEVING IT IS PSEUDO IPA!!!
+correspondances=matrix( #list correspondances always by pairs, orthography then phonology
+  c(
+    #nasal long to unichar
+    ":ii","ì",
+    ":ee","è",
+    ":aa","à",
+    ":êê","3",
+    ":ââ","6",
+    ":uu","ù",
+    ":oo","Ò",
+    #long to unichar
+    "ii","ï",
+    "ee","ë",
+    "éé","Ë",
+    "áá","Ä",
+    "aa","ä",
+    "îî","Ï",
+    "êê","2",
+    "ââ","5",
+    "uu","ü",
+    "óó","ö",
+    "oo","Ö",
+    #short nasal
+    ":i","1",
+    ":e","4",
+    ":a","ã",
+    ":ê","6",
+    ":â","7",
+    ":u","8",
+    ":o","9",
+    #consonants, veeeery dirty!
+    "'n","ń",
+    "tp",":",
+    "dp","/",
+    "kp","P",
+    "ngm","G",
+    "ńm","M",
+    "ng","N",
+    "mb","b",
+    "nd","D",
+    "nt","T",
+    "nk","K",
+    "dn","?",
+    "kn","!",
+    "vy","V",
+    "mt","C",
+    "md","J",
+    "mg","X",
+    "dm","S",
+    "km","Z",
+    "gh","H",
+    "lv","L"
+  ),#last item above should not have a comma
+  ncol = 2,byrow = T)
+
+trnsc$mp_uni=trnsc$mispronunciation #start from ortho
+trnsc$target_uni=trnsc$target.ortho #start from ortho
+
+for(i in 1:dim(correspondances)[1]) { #transform into unicharacter
+  trnsc$mp_uni=gsub(correspondances[i,1],correspondances[i,2],trnsc$mp_uni)
+  trnsc$target_uni=gsub(correspondances[i,1],correspondances[i,2],trnsc$target_uni)
+}
+
+#log non-fluent speech
+trnsc$nonfluent<-NA
+trnsc$nonfluent[grep("-",trnsc$mp_uni,fixed=T)]<-1
+
+#trnsc$additions=
+```
+
+```{r}
+#create a subset with just the first presentations -- this is what is going to be used for the quantitative analyses
+firstp=subset(trnsc,rep==1 & !is.na(rep))
+```
+
 - What is the overall repetition accuracy (whole word, phoneme based, distance)?
@@
 
-```{r}
+```{r iniana, eval=FALSE}
+#initial analyses with just 12 participants
 read.table("log2.txt",header=T,sep="\t")->x
@@
 x$cor.online=1
-for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0
+for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0  #if the item is repeated, then code the previous one as being wrong
 
-subset(x,rep=="1")->x
-aggregate(x$cor.online,by=list(x$child),mean)->onl
+subset(x,rep=="1")->x  #get only the first attempt for each child
+aggregate(x$cor.online,by=list(x$child),mean)->onl  #get means on first attempt
 colnames(onl)<-c("ID","score.online")
 
-print(onl)
+#print(onl) #online/on the fly coding, there are only 12 kids
 
-aggregate(x$corr,by=list(x$child),mean)->ms
+aggregate(x$corr,by=list(x$child),mean)->ms  #means
 colnames(ms)<-c("ID","score")
@@
 scores$Age[is.na(scores$Age)]<-15
-plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
-abline(lm(scores$score*100~scores$Age))
-plot(scores$score~scores$monolingual)
-points(scores$score~as.numeric(scores$monolingual),pch=20)
 
-plot(ms$score.online~ms$score)
+# plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
+# abline(lm(scores$score*100~scores$Age))
+# plot(scores$score~scores$monolingual)
+# points(scores$score~as.numeric(scores$monolingual),pch=20)
+
+#plot(ms$score.online~ms$score)
 ```
@@
 - What is the overall repetition accuracy (whole word, phoneme based, distance)?
-- How does this change as a function of item complexity (number of syllablex, sound complexity)?
+- How does this change as a function of item complexity (number of syllables, sound complexity)?
 - How frequent are errors that result in real words? Is that a function of item complexity?

### 2019-10-30 manuscript.Rmd
ana in progress

@@
 ```
+## TODO
+
+- create stimuli table that has complexity & length
+- integrate demo data & stim table in final data
+- add analysis on the basis of complexity & length
+- add analysis on the basis of demo chars
+- fix the back-transformation
+- add summary of types of error: %semantic, %deletion (which sounds?) %subst (which are the most common sounds) 
+- maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
+- discussion
 
@@
 
-The child was donned a headset (xx @mc please fill in for most of the children, xx @ac fill in for the rest).
+The child was donned a headset (xx @mc please fill in for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
 A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.
+An experimenter delivered the elicitation stimuli to the local informant and the child over headphones.
 
@@
 To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them.
-If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a headset that was recording a separate track).
+If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a separate headset that was recording onto the right track of the same Tascam).
 If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.
@@
 
-```{r wrangling}
-#@mc look here - we probably need to read the table in a special way to get all the accents etc. right
-read.table("NWR-transcription.txt",header=T,enc="UTF-16le")->trnsc
-read.csv("NWR-transcription.csv",sep=";",enc="UTF-8")->trnsc
+```{r wrangling, eval=T}
+read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
 trnsc$tokid=paste(trnsc$item,trnsc$token)
@@
 
+##TODO integrate demo data
+
 ```
 
-```{r phonoana}
+```{r phonoana, eval=T}
 
@@
 
-## TODO
+#log non-fluent speech, then get rid of that marker
+trnsc$nonfluent<-NA
+trnsc$nonfluent[grep("-",trnsc$mp_uni,fixed=T)]<-1
+trnsc$mispronunciation=gsub("-","",trnsc$mispronunciation)
+
 # make all vowels in mispronunciation & target short (so we don't penalize length errors)
+trnsc$mp=gsub("([aeiouêâéáóî])\\1+", "\\1", trnsc$mispronunciation)
+trnsc$target.ortho=gsub("([aeiouêâéáóî])\\1+", "\\1", trnsc$target.ortho)
+
 # if correct=0 but orthotarget=mispronunciation (probably due to length errors) then change correct to 1 
+#sum(trnsc$correct==0 & trnsc$target.ortho == trnsc$mp & !is.na(trnsc$mp)) #this only happens twice
+trnsc$correct[trnsc$correct==0 & trnsc$target.ortho == trnsc$mp & !is.na(trnsc$mp)]<-1
 
@@
 
-trnsc$mp_uni=trnsc$mispronunciation #start from ortho
+trnsc$mp_uni=trnsc$mp #start from ortho
 trnsc$target_uni=trnsc$target.ortho #start from ortho
@@
 
-#log non-fluent speech
-trnsc$nonfluent<-NA
-trnsc$nonfluent[grep("-",trnsc$mp_uni,fixed=T)]<-1
+#calculate Levinshtein's distances 
+trnsc$ins=trnsc$del=trnsc$sub=trnsc$nchar=trnsc$trafos=trnsc$tarlen=NA
+substitution_bank=NULL
+deletion_bank=NULL
+for(i in grep("0",trnsc$correct)){
+ #lev
+  lev=adist(trnsc[i,c("mp_uni")],trnsc[i,c("target_uni")],count=T)
+  trnsc$ins[i]<-attributes(lev)$counts[,,"ins"]
+  trnsc$del[i]<-attributes(lev)$counts[,,"del"]
+  trnsc$sub[i]<-attributes(lev)$counts[,,"sub"]
+  trnsc$trafos[i]=attributes(lev)$trafos
+  trnsc$nchar[i]=nchar(attributes(lev)$trafos)
+  trnsc$tarlen[i]=nchar(trnsc$target.ortho[i])
+  sublist=which(strsplit(attributes(lev)$trafos, "")[[1]]=="S")
+  substitution_bank=rbind(substitution_bank,cbind(strsplit(trnsc[i,c("target_uni")],"")[[1]][sublist],strsplit(trnsc[i,c("mp_uni")],"")[[1]][sublist]))
+  dellist=which(strsplit(attributes(lev)$trafos, "")[[1]]=="D")
+  deletion_bank=rbind(deletion_bank,cbind(strsplit(trnsc[i,c("target_uni")],"")[[1]][dellist]))
+}
 
-#trnsc$additions=
+trnsc$ld=rowSums(trnsc[,c("ins","del","sub")])
+trnsc$nld=trnsc$ld/trnsc$nchar
+trnsc$nld[is.na(trnsc$nld)]<-0
+
+#this is not working
+#trnsc$pc=nchar(gsub("SDI","",trnsc$trafos))/trnsc$nchar
+#trnsc$pc[is.na(trnsc$pc)]<-1
+
+write.table(trnsc,"final_data.txt",row.names=F,sep="\t")
+
+#TODO this is not working correctly
+# for(i in dim(correspondances)[1]:1) { #transform into orthography
+#   substitution_bank=gsub(correspondances[i,2],correspondances[i,1],substitution_bank)
+#   deletion_bank=gsub(correspondances[i,2],correspondances[i,1],deletion_bank)
+# }
+
+substitution_bank[order(substitution_bank[,1]),]->substitution_bank
+write.table(substitution_bank,"substitution_bank.txt",row.names=F,sep="\t")
+
+sort(deletion_bank)->deletion_bank
+write.table(deletion_bank,"deletion_bank.txt",row.names=F,sep="\t")
 ```
 
-```{r}
+```{r readin}
+read.table("final_data.txt",header=T,encoding="UTF-8")->trnsc
 #create a subset with just the first presentations -- this is what is going to be used for the quantitative analyses
@@
 
-- What is the overall repetition accuracy (whole word, phoneme based, distance)?
+Taking into account only the first attempts, the overall repetition accuracy measured in whole words is `r round(mean(firstp$cor.online)*100)`% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and `r round(mean(firstp$correct)*100)`% when the offline coding is considered.
+On average, the phoneme-based normalized Levenshtein distance was `r round(mean(firstp$nld)*100)`%, meaning that about 20% of phonemes were substituted, inserted, or deleted.
 
-- How does this change as a function of item complexity (number of syllablex, sound complexity)?
+We then inspected whether accuracy varied as a function of word length.
 
+```{r figlength,eval=F}
+#not working
+aggregate(firstp$correct,list(firstp$tarlen,firstp$id),mean)
+plot(firstp$correct~firstp$tarlen)
+
+```
 - How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?

### 2019-10-30 manuscript.tex
ana in progress

@@
 
+\hypertarget{todo}{%
+\subsection{TODO}\label{todo}}
+
+\begin{itemize}
+\tightlist
+\item
+  create stimuli table that has complexity \& length
+\item
+  integrate demo data \& stim table in final data
+\item
+  add analysis on the basis of complexity \& length
+\item
+  add analysis on the basis of demo chars
+\item
+  fix the back-transformation
+\item
+  add summary of types of error: \%semantic, \%deletion (which sounds?) \%subst (which are the most common sounds)
+\item
+  maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
+\item
+  discussion
+\end{itemize}
+
 \hypertarget{introduction}{%
@@
 
-Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., CITE).
+Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., Hazan \& Barrett, 2000).
 One common task in this line of research is nonword repetition (NWR).
@@
 
-NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon {[}CITE{]}, as well as for applications, notably using nwr as a diagnostic for language delays and disorders {[}CITE{]}.
-Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe {[}CITE{]}.
+NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon (Bowey, 2001), as well as for applications, notably using nwr as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
+Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe (Meir, Walters, \& Armon-Lotem, 2016,@is08042009language).
 
@@
 Although the procedure tends to be quite stable, there is some variation in the presentation and structure of items found in previous NWR work.
-For example, items are often presented orally by the experimenter {[}CITE{]}, although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation {[}CITE{]}.
-Additionally, while some studies have used 10-15 non-words, others have employed up to 30 unique items {[}CITE{]}.
-Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) as well as syllable structure {[}open as opposed to closed syllables, CITE{]}.
+For example, items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation (Brandeker \& Thordardottir, 2015).
+Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
+Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
 
@@
 That said, we contribute to the literature by also reporting descriptive analyses that could potentially be integrated in meta-analytic efforts.
-Based on previous work, we looked at potential improvements with age {[}CITE{]}, and potential negative effects of bilingual exposure {[}CITE{]}.
-Previous work typically finds no significant differences as a function of maternal education {[}see CITE but also CITE{]} or child gender {[}CITE{]}.
-Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks {[}CITE{]}, and therefore we report on those too.
+Based on previous work, we looked at potential improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and potential negative effects of bilingual exposure (Brandeker \& Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
+Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on those too.
 
@@
 \item
-  How does this change as a function of item complexity (number of syllablex, sound complexity)?
+  How does this change as a function of item complexity (number of syllables, sound complexity)?
 \item
@@
 
-({\textbf{???}}) ADD TABLE SUMMARIZING PREVIOUS WORK
-
-\begin{itemize}
-\tightlist
-\item
-  age differences
-\item
-  length effects
-\end{itemize}
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/prevlit-fig-1.pdf}
+\caption{\label{fig:prevlit-fig}Proportion of non-words correctly repeated as a function of age (in years), study (first author and year regardless of the number of authors), and, when available, shorter versus longer non-words.}
+\end{figure}
 
@@
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described {[}CITE{]}, there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described (CITE {\textbf{???}} please fill in), there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
@@
 
-The child was donned a headset (xx ({\textbf{???}}) please fill in for most of the children, xx ({\textbf{???}}) fill in for the rest).
+The child was donned a headset (xx ({\textbf{???}}) please fill in for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
 A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.
+An experimenter delivered the elicitation stimuli to the local informant and the child over headphones.
 
@@
 
+Taking into account only the first attempts, the overall repetition accuracy measured in whole words is 73\% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and 65\% when the offline coding is considered.
+On average, the phoneme-based normalized Levenshtein distance was 20\%, meaning that about 20\% of phonemes were substituted, inserted, or deleted.
+
+We then inspected whether accuracy varied as a function of word length.
+
+\begin{verbatim}
+##     Group.1 Group.2 x
+## 1         2    c105 0
+## 2         3    c105 0
+## 3         4    c105 0
+## 4         6    c105 0
+## 5         8    c105 0
+## 6         3    c106 0
+## 7         4    c106 0
+## 8         6    c106 0
+## 9         2    c107 0
+## 10        3    c107 0
+## 11        4    c107 0
+## 12        6    c107 0
+## 13        3    c108 0
+## 14        4    c108 0
+## 15        6    c108 0
+## 16        8    c108 0
+## 17        2    c109 0
+## 18        3    c109 0
+## 19        4    c109 0
+## 20        6    c109 0
+## 21        3    c110 0
+## 22        4    c110 0
+## 23        6    c110 0
+## 24        8    c110 0
+## 25        3    c111 0
+## 26        4    c111 0
+## 27        8    c111 0
+## 28        2    c112 0
+## 29        3    c112 0
+## 30        4    c112 0
+## 31        3    c114 0
+## 32        4    c114 0
+## 33        6    c114 0
+## 34        8    c114 0
+## 35        3    c115 0
+## 36        4    c115 0
+## 37        6    c115 0
+## 38        8    c115 0
+## 39        3    c116 0
+## 40        6    c116 0
+## 41        8    c116 0
+## 42        3    c117 0
+## 43        4    c117 0
+## 44        3    c118 0
+## 45        4    c118 0
+## 46        6    c118 0
+## 47        8    c118 0
+## 48        3    c119 0
+## 49        6    c119 0
+## 50        8    c119 0
+## 51        3    c120 0
+## 52        4    c120 0
+## 53        6    c120 0
+## 54        8    c120 0
+## 55        3    c121 0
+## 56        4    c121 0
+## 57        3    c122 0
+## 58        4    c122 0
+## 59        6    c122 0
+## 60        8    c122 0
+## 61        3    c123 0
+## 62        4    c123 0
+## 63        2     c70 0
+## 64        3     c70 0
+## 65        4     c70 0
+## 66        6     c70 0
+## 67        8     c70 0
+## 68        2     c71 0
+## 69        3     c71 0
+## 70        4     c71 0
+## 71        6     c71 0
+## 72        8     c71 0
+## 73        3     c72 0
+## 74        4     c72 0
+## 75        6     c72 0
+## 76        8     c72 0
+## 77        3     c73 0
+## 78        4     c73 0
+## 79        6     c73 0
+## 80        8     c73 0
+## 81        2     c74 0
+## 82        3     c74 0
+## 83        4     c74 0
+## 84        6     c74 0
+## 85        8     c74 0
+## 86        3     c76 0
+## 87        4     c76 0
+## 88        8     c76 0
+## 89        3     c77 0
+## 90        4     c77 0
+## 91        6     c77 0
+## 92        8     c77 0
+## 93        2     c78 0
+## 94        3     c78 0
+## 95        8     c78 0
+## 96        2     c79 0
+## 97        3     c79 0
+## 98        2     c80 0
+## 99        3     c80 0
+## 100       4     c80 0
+## 101       6     c80 0
+## 102       8     c80 0
+## 103       2     c81 0
+## 104       3     c81 0
+## 105       4     c81 0
+## 106       8     c81 0
+## 107       3     c82 0
+## 108       4     c82 0
+## 109       6     c82 0
+## 110       8     c82 0
+## 111       3     c83 0
+## 112       4     c83 0
+## 113       6     c83 0
+## 114       8     c83 0
+## 115       3     c84 0
+## 116       4     c84 0
+## 117       6     c84 0
+## 118       8     c84 0
+## 119       3     c85 0
+## 120       4     c85 0
+## 121       6     c85 0
+## 122       8     c85 0
+## 123       3     c87 0
+## 124       4     c87 0
+## 125       6     c87 0
+## 126       8     c87 0
+## 127       3     c88 0
+## 128       4     c88 0
+## 129       3     c89 0
+## 130       4     c89 0
+## 131       6     c89 0
+## 132       3     c90 0
+## 133       4     c90 0
+## 134       6     c90 0
+## 135       8     c90 0
+## 136       2     c91 0
+## 137       3     c91 0
+## 138       4     c91 0
+## 139       2     c93 0
+## 140       3     c93 0
+## 141       4     c93 0
+## 142       6     c93 0
+## 143       3     c94 0
+## 144       4     c94 0
+## 145       8     c94 0
+## 146       3     c95 0
+## 147       4     c95 0
+## 148       3     c99 0
+## 149       4     c99 0
+## 150       6     c99 0
+\end{verbatim}
+
+\includegraphics{manuscript_files/figure-latex/figlength-1.pdf}
+- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+
 \begin{itemize}
-\item
-  What is the overall repetition accuracy (whole word, phoneme based, distance)?
-\item
-  How does this change as a function of item complexity (number of syllablex, sound complexity)?
-\item
-  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+\tightlist
 \item
@@
 \end{itemize}
-
-\begin{verbatim}
-##     ID score.online
-## 1  c70    0.7307692
-## 2  c71    0.5384615
-## 3  c72    0.6086957
-## 4  c73    0.8461538
-## 5  c74    0.9090909
-## 6  c76    0.7307692
-## 7  c77    0.9565217
-## 8  c78    0.8461538
-## 9  c79    0.7692308
-## 10 c80    0.8260870
-## 11 c81    0.5652174
-## 12 c82    0.5000000
-\end{verbatim}
-
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf} \includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-2.pdf} \includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-3.pdf}
 
@@
 
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-bowey2001nonword}{}%
+Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
+
+\leavevmode\hypertarget{ref-brandeker2015language}{}%
+Brandeker, M., \& Thordardottir, E.
+(2015).
+Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
+
+\leavevmode\hypertarget{ref-chiat2007preschool}{}%
+Chiat, S., \& Roy, P.
+(2007).
+The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
+
+\leavevmode\hypertarget{ref-estes2007differences}{}%
+Estes, K.
+G., Evans, J.
+L., \& Else-Quest, N.
+M.
+(2007).
+Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
+
+\leavevmode\hypertarget{ref-farmani2018normalization}{}%
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
+Z., Talebi, E., \& Shourvazi, Z.
+(2018).
+Normalization of the non-word repetition test in farsi-speaking children.
+\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
+
+\leavevmode\hypertarget{ref-gallon2007non}{}%
+Gallon, N., Harris, J., \& Van der Lely, H.
+(2007).
+Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
+\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\leavevmode\hypertarget{ref-havron2019effect}{}%
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
+M.-C.
+C.
+S.
+(2019).
+The effect of older siblings on language development as a function of age difference and sex.
+\emph{Psychological Science}, \emph{30}(9), 1333--1343.
+
+\leavevmode\hypertarget{ref-hazan2000development}{}%
+Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
+
+\leavevmode\hypertarget{ref-is08042009language}{}%
+IS0804, C. A. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
+\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
+(2014).
+Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
+\emph{PloS One}, \emph{9}(2), e89544.
+
+\leavevmode\hypertarget{ref-meir2017independent}{}%
+Meir, N., \& Armon-Lotem, S.
+(2017).
+Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
+\emph{Frontiers in Psychology}, \emph{8}, 1442.
+
+\leavevmode\hypertarget{ref-meir2016disentangling}{}%
+Meir, N., Walters, J., \& Armon-Lotem, S.
+(2016).
+Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
+\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
+
+\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
+(2019).
+An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
+\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
+
+\leavevmode\hypertarget{ref-torrington2015non}{}%
+Torrington Eaton, C., Newman, R.
+S., Ratner, N.
+B., \& Rowe, M.
+L.
+(2015).
+Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
+\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+
+\leavevmode\hypertarget{ref-vance2005speech}{}%
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
+
 \end{document}

### 2020-05-14 manuscript.tex
getting back into the groove

--- 2019-10-29 manuscript.tex
@@
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
+\PassOptionsToPackage{hyphens}{url}
+%
 \documentclass[english,,man,floatsintext]{apa6}
@@
   \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provides euro and other symbols
 \else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
+  \usepackage{unicode-math}
   \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
@@
 \IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\usepackage[]{microtype}
 \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Non-word repetition in Yélî Dnye: A small-scale study},
-            pdfauthor={Alejandrina Cristia~\& Marisa Casillas},
-            pdfkeywords={phonology, non-word repetition},
+\hypersetup{
             pdfborder={0 0 0},
@@
 \urlstyle{same}  % don't use monospace font for urls
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\fi
 \usepackage{graphicx,grffile}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 \setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
-  \title{Non-word repetition in Yélî Dnye: A small-scale study}
-    \author{Alejandrina Cristia\textsuperscript{1}~\& Marisa Casillas\textsuperscript{2}}
-    \date{}
-
-\shorttitle{NWR in Yélî Dnye'}
-\affiliation{
-\vspace{0.5cm}
-\textsuperscript{1} LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
-\keywords{phonology, non-word repetition\newline\indent Word count: xxx}
-\usepackage{csquotes}
+% Manuscript styling
 \usepackage{upgreek}
@@
 
+% Table formatting
 \usepackage{longtable}
 \usepackage{lscape}
-\usepackage{multirow}
-\usepackage{tabularx}
-\usepackage[flushleft]{threeparttable}
-\usepackage{threeparttablex}
+% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
+\usepackage{multirow}		% Table styling
+\usepackage{tabularx}		% Control Column width
+\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
+\usepackage{threeparttablex}            % Lets threeparttable work with longtable
 
+% Create new environments so endfloat can handle them
+% \newenvironment{ltable}
+%   {\begin{landscape}\begin{center}\begin{threeparttable}}
+%   {\end{threeparttable}\end{center}\end{landscape}}
 \newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
 
+% Enables adjusting longtable caption width to table width
+% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
 \makeatletter
@@
 
+% \setlength{\parindent}{0.5in}
+% \setlength{\parskip}{0pt plus 0pt minus 0pt}
+
+% \usepackage{etoolbox}
+\makeatletter
+\patchcmd{\HyOrg@maketitle}
+  {\section{\normalfont\normalsize\abstractname}}
+  {\section*{\normalfont\normalsize\abstractname}}
+  {}{\typeout{Failed to patch abstract.}}
+\makeatother
+\shorttitle{NWR in Yélî Dnye'}
+\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2}}
+\affiliation{
+\vspace{0.5cm}
+\textsuperscript{1} LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
 \authornote{
 
-Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d'Ulm, 75005, Paris, France. E-mail: \href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
+Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d'Ulm, 75005, Paris, France. E-mail: alecristia@gmail.com}
+\keywords{phonology, non-word repetition, development\newline\indent Word count: xxx}
+\usepackage{csquotes}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[shorthands=off,main=english]{babel}
+\else
+  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
+  \usepackage{polyglossia}
+  \setmainlanguage[]{english}
+\fi
 
-\abstract{
-ADD
-
-}
+\date{}
 
 \begin{document}
-\maketitle
 
@@
 
-Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., CITE).
+Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., {\textbf{???}}).
 One common task in this line of research is nonword repetition (NWR).
@@
 The child should immediately try to say it back without changing anything.
-Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back).
+Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible productiong patterns (to produce the item accurately even though it had never been pronounced before).
 
-NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon {[}CITE{]}, as well as for applications, notably using nwr as a diagnostic for language delays and disorders {[}CITE{]}.
-Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe {[}CITE{]}.
+NWR has been used to seek answers to a variety of theoretical questions, including what are the links between phonology, working memory, and the lexicon ({\textbf{???}}), as well as for applications, notably using nwr as a diagnostic for language delays and disorders ({\textbf{???}}).
+Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe ({\textbf{???}},@is08042009language).
 
-In this study, we use NWR to measure perception-production in children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
 In our implementation of NWR, we made sure that some of the items contained typologically rare and/or challenging sounds.
+MISSING: WHY DO WE DO THIS?
+WHAT IS THE POINT?
 
@@
 
-Although the procedure tends to be quite stable, there is some variation in the presentation and structure of items found in previous NWR work.
-For example, items are often presented orally by the experimenter {[}CITE{]}, although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation {[}CITE{]}.
-Additionally, while some studies have used 10-15 non-words, others have employed up to 30 unique items {[}CITE{]}.
-Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) as well as syllable structure {[}open as opposed to closed syllables, CITE{]}.
+There is some variation in the presentation procedure and structure of items found in previous NWR work.
+For example, items are often presented orally by the experimenter ({\textbf{???}}), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation ({\textbf{???}}).
+Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items ({\textbf{???}}).
+Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, {\textbf{???}}).
 
-Previous work seems to avoid difficult sounds, but we felt this was important to represent the language, so we also varied this factor.
+Previous work seems to avoid difficult sounds, but we felt this was important to represent Yélî Dnye, so we also varied this factor.
 We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
@@
 However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
-That said, we contribute to the literature by also reporting descriptive analyses that could potentially be integrated in meta-analytic efforts.
-Based on previous work, we looked at potential improvements with age {[}CITE{]}, and potential negative effects of bilingual exposure {[}CITE{]}.
-Previous work typically finds no significant differences as a function of maternal education {[}see CITE but also CITE{]} or child gender {[}CITE{]}.
-Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks {[}CITE{]}, and therefore we report on those too.
+
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
+Based on previous work, we looked at potential improvements with age ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}), and potential negative effects of bilingual exposure ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}).
+Previous work typically finds no significant differences as a function of maternal education (e.g., {\textbf{???}}; {\textbf{???}}; {\textbf{???}}) or child gender ({\textbf{???}}).
+Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks ({\textbf{???}}), and therefore we report on those too.
 
@@
 \item
-  What is the overall repetition accuracy (whole word, phoneme based, distance)?
+  What is the overall repetition score, according to a variety of relevant metrics (whole word, phoneme based, distance)?
 \item
-  How does this change as a function of item complexity (number of syllablex, sound complexity)?
+  How does score change as a function of item complexity (number of syllablex, sound complexity)?
 \item
-  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+  How frequent are errors that result in real words? Is that a function of item complexity?
 \item
-  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
+  What are some common patterns of phonological errors?
+\item
+  Is variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
 \end{itemize}
 
-In view of the hypothesis-driven nature of this work, we consider boosting the interpretational value of these data by announcing our analysis plans prior to conducting them.
+In view of the hypothesis-driven nature of this work, we considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
 However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects.
-For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small.
+For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small (data from {\textbf{???}}).
 \textbf{add table with r/ds from previous work})
@@
 
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language.
-We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language.
-We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language (e.g., {\textbf{???}}; {\textbf{???}}).
+We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in Yély Dnye.
+We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat (note that up to other work has also used 24-30 items: {\textbf{???}}; {\textbf{???}}), and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
 
@@
 These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
-MC presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
+The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
 Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
 
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce consonant-vowel monosyllables.
+A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî inventory to produce consonant-vowel monosyllables.
 As before, items were automatically excluded if they appeared in the dictionary.
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described {[}CITE{]}, there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described (CITE {\textbf{???}} please fill in), there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
 These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally.
+Together with the 2018 selection, they were recorded using a headset \textbf{XXX ({\textbf{???}}) please fill in} and an Olympus \textbf{XXX ({\textbf{???}}) please fill in} from the written form presented together with the same item orally (by the second author).
 The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
@@
 
-({\textbf{???}}) can you please add the IPA in // for all of the items below?
-The final list is composed of three practice items (wî, poni, nopimade); 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives (
+\textbf{({\textbf{???}}) can you please add the IPA in // for all of the items in the table?}
+The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
+
+Practice
+wî, poni, nopimade
+
+Mono
 dpa
@@
 t:êê
-); 8 bisyllables (
+);
+
+(
 kamo
@@
 nomi);
-12 trisyllables (
+
+(
 nademo
@@
 nayeki
-mituye); and 4 quadrisyllables (
+mituye);
+
+(
 todiwuma
@@
 \item
-  once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2\_syllables, then 3-syllables, and finally 4-syllable items.
+  once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllables, then 3-syllables, and finally 4-syllable items.
 \end{itemize}
@@
 
-Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items \enquote{spreading} through hearsay.
-In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own.
+We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items \enquote{spreading} through hearsay.
+In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
 The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
 
-The child was donned a headset (xx ({\textbf{???}}) please fill in for most of the children, xx ({\textbf{???}}) fill in for the rest).
+The child was donned a headset (\textbf{xx ({\textbf{???}}) please fill in} for most of the children, xx ({\textbf{???}}) fill in for the rest).
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.
+A local informant sat next to the child, to provide the instructions and, if needed, coach the child to make sure that they understood that the task was to repeat the items precisely without changing anything using the three practice items as well as real words.
+The experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headhphones.
 
@@
 
-There is an implicit accuracy coding occuring online as a function of whether the local informant asked for the item to be repeated or not.
-We will call this \enquote{online accuracy}, and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
+The local informant asked for the item to be repeated when she thought the repetition was not correct.
+We will call this \emph{online scoring}, and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
 
@@
 We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-This \enquote{native coding} provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.
+This \emph{offline scoring} provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.
 
@@
 These were scored as correct or incorrect on the fly depending on whether the native experimenter asked for the item to be repeated; as well as offline.
-Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy {[}CITE{]}.
+Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy (e.g., {\textbf{???}}).
 We calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
@@
 
+This study was approved as part of a larger research effort by \textbf{({\textbf{???}}) add info from Middy's FAIR reply}.
+Participation was voluntary, with children being invited to come and participate.
+Regardless of how they performed, children were provided with a snack as compensation.
+Children who came up to participate but then refused were nonetheless provided with the snack.
+
 A total 55 children were tested, from 34 of families, in five hamlets.
@@
 There were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and , 2 for whom this information was missing.
-Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).
-In terms of birth order, 0 were first borns, 5 second, 4 third, 2 forth, 6 fifth, 5 sixth, and 1 did not have this information.
-
-({\textbf{???}}) why is maternal education always in even years? Can you add a footnote to explain?
+Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Education is often reported in even years because people typically complete two-year cycles.} In terms of birth order, 0 were first borns, 5 second, 4 third, 2 forth, 6 fifth, 5 sixth, and 1 did not have this information.
 
@@
 
-\begin{itemize}
-\item
-  What is the overall repetition accuracy (whole word, phoneme based, distance)?
-\item
-  How does this change as a function of item complexity (number of syllablex, sound complexity)?
-\item
-  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
-\item
-  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
-\end{itemize}
+\hypertarget{overall-repetition-accuracy}{%
+\subsubsection{Overall repetition accuracy}\label{overall-repetition-accuracy}}
 
-\begin{verbatim}
-##     ID score.online
-## 1  c70    0.7307692
-## 2  c71    0.5384615
-## 3  c72    0.6086957
-## 4  c73    0.8461538
-## 5  c74    0.9090909
-## 6  c76    0.7307692
-## 7  c77    0.9565217
-## 8  c78    0.8461538
-## 9  c79    0.7692308
-## 10 c80    0.8260870
-## 11 c81    0.5652174
-## 12 c82    0.5000000
-\end{verbatim}
+We start by describing whole word accuracy
 
-\includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-1.pdf} \includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-2.pdf} \includegraphics{manuscript_files/figure-latex/unnamed-chunk-1-3.pdf}
-
-\textbf{(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)}
-
-Based on previous work, one can make the following general predictions for any population:
+phoneme based, distance
+\#\#\# Accuracy a function of item complexity
+(number of syllablex, sound complexity)?
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\tightlist
 \item
@@
 \item
+  The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
+\textbf{Check for work on European languages that may have looked into this}
+\item
   Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds {[}CITE{]}.
 Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
-\item
-  Children's accuracy increases with child age.
-\item
-  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
-\item
-  As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
-\item
-  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
-\end{enumerate}
-
-Although these predictions were made based on previous work, we also thought reasonable that they would not obtain for the Rossel children in particular, as follows:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
-\textbf{Check for work on European languages that may have looked into this}
 \item
@@
 no work looking at nasal vowels in particular?}
+\end{enumerate}
+
+\textbf{(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)}
+
+\hypertarget{patterns-of-errors}{%
+\subsubsection{Patterns of errors}\label{patterns-of-errors}}
+
+\begin{itemize}
+\tightlist
 \item
-  We had no reservations that children's accuracy increases with child age for the Yélî community as well.
+  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+\end{itemize}
+
+\hypertarget{factor-structuring-individual-variation}{%
+\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}
+
+explainable by child age, sex, birth order, monolingual status, and/or parental education?
+3. Children's accuracy increases with child age.
+4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
+5.
+As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
+6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
 \item

### 2020-05-14 manuscript.Rmd
getting back into the groove

--- 2019-10-30 manuscript.Rmd
@@
 ---
-title: "Non-word repetition in Yélî Dnye: A small-scale study"
+title: "Non-word repetition in Yélî Dnye"
 shorttitle        : "NWR in Yélî Dnye'"
@@
 
-keywords          : "phonology, non-word repetition"
+keywords          : "phonology, non-word repetition, development"
 wordcount         : "xxx"
@@
 The child should immediately try to say it back without changing anything.
-Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back).
+Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible productiong patterns (to produce the item accurately even though it had never been pronounced before).
 
-NWR has been used to seek answers to theoretical questions, including what are the links between phonology, working memory, and the lexicon [@bowey2001nonword], as well as for applications, notably using nwr as a diagnostic for language delays and disorders [@estes2007differences].
+NWR has been used to seek answers to a variety of theoretical questions, including what are the links between phonology, working memory, and the lexicon [@bowey2001nonword], as well as for applications, notably using nwr as a diagnostic for language delays and disorders [@estes2007differences].
 Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe [@meir2016disentangling,@is08042009language].
 
-In this study, we use NWR to measure perception-production in children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
 In our implementation of NWR, we  made sure that some of the items contained typologically rare and/or  challenging sounds.
+MISSING: WHY DO WE DO THIS?
+WHAT IS THE POINT?
 
@@
 
-Although the procedure tends to be quite stable, there is some variation in the presentation and structure of items found in previous NWR work.
+There is some variation in the presentation procedure and structure of items found in previous NWR work.
 For example, items are often presented orally by the experimenter [@torrington2015non], although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation [@brandeker2015language].
@@
 
-Previous work seems to avoid difficult sounds, but we felt this was important to represent the language, so we also varied this factor.
+Previous work seems to avoid difficult sounds, but we felt this was important to represent Yélî Dnye, so we also varied this factor.
 We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
@@
 However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
-That said, we contribute to the literature by also reporting descriptive analyses that could potentially be integrated in meta-analytic efforts.
+
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
 Based on previous work, we looked at potential improvements with age [@vance2005speech;@farmani2018normalization;@kalnak2014nonword], and potential negative effects of bilingual exposure [@brandeker2015language;@meir2016disentangling;@meir2017independent].
@@
 
-- What is the overall repetition accuracy (whole word, phoneme based, distance)?
-- How does this change as a function of item complexity (number of syllablex, sound complexity)?
-- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
-- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
+- What is the overall repetition score, according to a variety of relevant metrics (whole word, phoneme based, distance)?
+- How does score change as a function of item complexity (number of syllablex, sound complexity)?
+- How frequent are errors that result in real words? Is that a function of item complexity? 
+- What are some common patterns of phonological errors?
+- Is variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
 
-In view of the hypothesis-driven nature of this work, we consider boosting the interpretational value of these data by announcing our analysis plans prior to conducting them.
+In view of the hypothesis-driven nature of this work, we considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
 However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects.
-For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small.
+For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small [data from @cristia2020infant].
 **add table with r/ds from previous work**)
@@
 Therefore, all analyses here are descriptive and should be considered exploratory.
+
+```{r}
+read.csv("prevlit.csv")->kids
+
+```
 
@@
 
-Many  NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language.
-We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language.
-We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological].
+We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in Yély Dnye.
+We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat [note that up to other work has also used 24-30 items: @jabere2018xperiment; @kalnak2014nonword], and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
 
@@
 These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
-MC presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
+The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
 Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
 
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce consonant-vowel monosyllables.
+A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî inventory to produce consonant-vowel monosyllables.
 As before, items were automatically excluded if they appeared in the dictionary.
@@
 These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally.
+Together with the 2018 selection, they were recorded using a headset **XXX @mc please fill in** and an Olympus **XXX @mc please fill in** from the written form presented together with the same item orally (by the second author).
 The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
@@
 
-@mc can you please add the IPA in // for all of the items below?
-The final list is composed of three practice items (wî, poni, nopimade); 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives (
+**@mc can you please add the IPA in // for all of the items in the table?**
+The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
+
+```{r}
+
+```
+
+Practice
+wî, poni, nopimade
+
+Mono
 dpa
@@
 nomi);
-12 trisyllables (
+
+ (
 nademo
@@
 Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
-- once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2_syllables, then 3-syllables, and finally 4-syllable items.
+- once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllables, then 3-syllables, and finally 4-syllable items.
 
@@
 
-Indeed, we visited four different hamlets once, and attempted to test all eligible childr en present at that time, to prevent the items 'spreading' through hearsay.
-In the first village, we tested children in five different places, with some children being tested inside their house and others tested on the veranda of a house even if it was not their own.
+We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items 'spreading' through hearsay.
+In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
 The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
 
-The child was donned a headset (xx @mc please fill in for most of the children, xx @ac fill in for the rest).
+The child was donned a headset (**xx @mc please fill in** for most of the children, xx @ac fill in for the rest).
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter delivered the elicitation stimuli to the local informant and the child over headhphones.
+A local informant sat next to the child, to  provide the instructions and, if needed, coach the child to make sure that they understood that the task was to repeat the items precisely without changing anything using the three practice items as well as real words.
+The experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headhphones.
 
@@
 
-There is an implicit accuracy coding occuring online as a function of whether the local informant asked for the item to be repeated or not.
-We will call this "online accuracy", and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
+The local informant asked for the item to be repeated when she thought the repetition was not correct.
+We will call this *online scoring*, and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
 
@@
 We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-This "native coding" provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.
+This *offline scoring* provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.
 
@@
 These were scored as correct or incorrect on the fly depending on whether the native experimenter asked for the item to be repeated; as well as offline.
-Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy [CITE].
+Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy [e.g., @cristia2020infant].
 We calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
@@
 ### Participants
+
+This study was approved as part of a larger research effort by **@ac add info from Middy's FAIR reply**.
+Participation was voluntary, with children being invited to come and participate.
+Regardless of how they performed, children were provided with a snack as compensation.
+Children who came up to participate but then refused were nonetheless provided with the snack.
 
@@
 There were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and , `r sum(is.na(inc$monolingual))` for whom this information was missing.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing).
-In terms of birth order, `r mytab[1]` were first borns, `r mytab[2]` second,  `r mytab[3]` third,  `r mytab[4]` forth,  `r mytab[5]` fifth, `r mytab[6]` sixth, and  `r mytab[7]` did not have this information.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing).[^1] In terms of birth order, `r mytab[1]` were first borns, `r mytab[2]` second,  `r mytab[3]` third,  `r mytab[4]` forth,  `r mytab[5]` fifth, `r mytab[6]` sixth, and  `r mytab[7]` did not have this information.
 
-@mc why is maternal education always in even years? Can you add a footnote to explain?
+[^1]: Education is often reported in even years because people typically complete two-year cycles.
 
@@
 
-- What is the overall repetition accuracy (whole word, phoneme based, distance)?
+```{r}
+# read.table("log2.txt",header=T,sep="\t")->x
+# x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
+# x[order(x$out),]->x
+# x$cor.online=1
+# for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0
+#   
+# subset(x,rep=="1")->x
+# aggregate(x$cor.online,by=list(x$child),mean)->onl
+# colnames(onl)<-c("ID","score.online")
+# 
+# #print(onl)
+# 
+# aggregate(x$corr,by=list(x$child),mean)->ms
+# colnames(ms)<-c("ID","score")
+# merge(ms,onl,all=T)->ms
+# merge(ms,demo,all=T)->scores
+# scores$Age=as.numeric(as.character(scores$Age))
+# scores$Age[is.na(scores$Age)]<-15
+# plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
+# 
+# plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
+# abline(lm(scores$score*100~scores$Age))
+# plot(scores$score~scores$monolingual)
+# points(scores$score~as.numeric(scores$monolingual),pch=20)
+# 
+# plot(ms$score.online~ms$score)
 
-- How does this change as a function of item complexity (number of syllablex, sound complexity)?
+read.csv("NWR-transcription.csv",sep=";")->coding
 
-- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+#remove some weird items
+coding[coding$item!="task4",]->coding
 
-- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
+read.csv("final_order_ful.txt",sep="\t")->coding_key
+coding_key$row=row.names(coding_key)
+
+merge(coding,coding_key,all=T)->joint
+
+merge(joint,demo,by.x="id",by.y="ID",all.x=T)->joint
+
+#order so we have all the items by a given child in the right chrono order
+joint[order(joint$file, joint$int),]->joint
+
+#add variable coding whether it's the child's first or subsequent attempts
+joint$previous<-c(NA,as.character(joint$target[1:(dim(joint)[1]-1)]))
+joint$attempt<-ifelse(joint$target!=joint$previous,"first","subsequent")
+joint$attempt[1]<-"first"
+
+#remove practice items
+practice=joint[(joint$target %in% c("wi","poni","nopimade")),]
+joint=joint[!(joint$target %in% c("wi","poni","nopimade")),]
+
+read.csv("stimuli.csv",header=T)->stims
+
+# phonologize things
+merge(joint,stims,by="target",all=T)->joint
+joint$target_pron=as.character(joint$ortho)
+
+joint$pron=as.character(joint$mispronunciation)
+
+correspondances=matrix( #list correspondances always by pairs
+  c(
+    #nasal long to unichar
+    ":ii","ì",
+    ":ee","è",
+    ":aa","à",
+    ":êê","3",
+    ":ââ","6",
+    ":uu","ù",
+    ":oo","Ò",
+    #long to unichar
+    "ii","ï",
+    "ee","ë",
+    "éé","Ë",
+    "áá","Ä",
+    "aa","ä",
+    "îî","Ï",
+    "êê","2",
+    "ââ","5",
+    "uu","ü",
+    "óó","ö",
+    "oo","Ö",
+    #short nasal
+    ":i","1",
+    ":e","4",
+    ":a","ã",
+    ":ê","6",
+    ":â","7",
+    ":u","8",
+    ":o","9",
+    #consonants, veeeery dirty!
+    "'n","ń",
+    "tp",":",
+    "dp","/",
+    "kp","P",
+    "ngm","G",
+    "ńm","M",
+    "ng","N",
+    "mb","b",
+    "nd","D",
+    "nt","T",
+    "nk","K",
+    "dn","?",
+    "kn","!",
+    "vy","V",
+    "mt","C",
+    "md","J",
+    "mg","X",
+    "dm","S",
+    "km","Z",
+    "gh","H",
+    "lv","L"
+  ),#last item above should not have a comma
+  ncol = 2,byrow = T)
+
+for(i in 1:dim(correspondances)[1]) {
+  joint$pron=gsub(correspondances[i,1],correspondances[i,2],joint$pron)
+  joint$target_pron=gsub(correspondances[i,1],correspondances[i,2],joint$target_pron)
+}
+
+#deal with hesitations
+joint$hesitation=F
+
+```
+
+### Preliminary analyses
+
+```{r Fig1-first_vs_others, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR scores for individual participants averaging separately their first attempts and all other attempts."}
+
+aggregate(joint$correct*100,by=list(joint$id,joint$age.exact,joint$attempt),mean)->pwcm_att
+colnames(pwcm_att)<-c("ID","Age","Attempt","Word")
+
+myx=as.numeric(as.factor(pwcm_att$ID))/100+as.numeric(as.factor(pwcm_att$Attempt))
+
+plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="NWR scores (%)")
+for(id in levels(factor(pwcm_att$ID))) lines(myx[pwcm_att$ID==id],pwcm_att$Word[pwcm_att$ID==id],lty=2)
+axis(1,at=1:2,labels=c("first","others"))
+
+t.test(pwcm_att$Word~pwcm_att$Attempt,paired=T)->myt
+```
+
+We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
+As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
+Surprisingly, subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
+were on average less accurate than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`.
+Given the uncertainty in whether previous work used only the first or all repetitions, and since behavior degraded and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
+
+```{r exclude subsequent}
+joint=joint[joint$attempt=="first",]
+```
+
+### Overall repetition accuracy
+
+We start by describing whole word accuracy. Participants scored
 
 ```{r}
-read.table("log2.txt",header=T,sep="\t")->x
-x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
-x[order(x$out),]->x
-x$cor.online=1
-for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0
 
-subset(x,rep=="1")->x
-aggregate(x$cor.online,by=list(x$child),mean)->onl
-colnames(onl)<-c("ID","score.online")
+```
 
-print(onl)
+phoneme based, distance
 
-aggregate(x$corr,by=list(x$child),mean)->ms
-colnames(ms)<-c("ID","score")
-merge(ms,onl)->ms
-merge(ms,demo)->scores
-scores$Age=as.numeric(as.character(scores$Age))
-scores$Age[is.na(scores$Age)]<-15
-plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
-abline(lm(scores$score*100~scores$Age))
-plot(scores$score~scores$monolingual)
-points(scores$score~as.numeric(scores$monolingual),pch=20)
+### Accuracy a function of item complexity 
+(number of syllablex, sound complexity)?
 
-plot(ms$score.online~ms$score)
-```
+1. Children are more accurate for mono-syllables than longer items
+1.
+The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
+**Check for work on European languages that may have looked into this**
+
+2.
+Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds [CITE].
+Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
+2.
+The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
+Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
+As a result, differences between easier and harder items may be smaller in this work than in other research.
+**no work looking at consonants & vowels?
+no work looking at nasal vowels in particular?**
 
@@
 
-Based on previous work, one can make the following general predictions for any population:
+### Patterns of errors
+- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
 
-1. Children are more accurate for mono-syllables than longer items
-2.
-Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds [CITE].
-Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
+### Factor structuring individual variation
+
+explainable by child age, sex, birth order, monolingual status, and/or parental education?
 3. Children's accuracy increases with child age.
@@
 
-Although these predictions were made based on previous work, we also thought reasonable that they would not obtain for the Rossel children in particular, as follows:
-
-1.
-The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
-**Check for work on European languages that may have looked into this**
-2.
-The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
-Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
-As a result, differences between easier and harder items may be smaller in this work than in other research.
-**no work looking at consonants & vowels?
-no work looking at nasal vowels in particular?**
-3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
 4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family 

### 2020-05-14 manuscript.Rmd

@@
 ```{r setup, include=FALSE}
+REDOALL=FALSE #use true if you want to recalculate the final-data and other data sheets
+
 #Sys.setlocale("LC_CTYPE", "en_US_POSIX.UTF-8")
@@
 
+correspondances=matrix( #list correspondances always by pairs, orthography then phonology
+  c(
+    #nasal  to unichar
+    ":ii","ì",
+    ":ee","è",
+    ":aa","à",
+    ":êê","3",
+    ":ââ","6",
+    ":uu","ù",
+    ":oo","Ò",
+    #long to unichar
+    "ii","ï",
+    "ee","ë",
+    "éé","Ë",
+    "áá","Ä",
+    "aa","ä",
+    "îî","Ï",
+    "êê","2",
+    "ââ","5",
+    "uu","ü",
+    "óó","ö",
+    "oo","Ö",
+    #short nasal
+    ":i","1",
+    ":e","4",
+    ":a","ã",
+    ":ê","6",
+    ":â","7",
+    ":u","8",
+    ":o","9",
+    #consonants, veeeery dirty!
+    "'n","ń",
+    "tp",":",
+    "dp","/",
+    "kp","P",
+    "ngm","G",
+    "ńm","M",
+    "ng","N",
+    "mb","b",
+    "nd","D",
+    "nt","T",
+    "nk","K",
+    "dn","?",
+    "kn","!",
+    "vy","V",
+    "mt","C",
+    "md","J",
+    "mg","X",
+    "dm","S",
+    "km","Z",
+    "gh","H",
+    "lv","L"
+  ),#last item above should not have a comma
+  ncol = 2,byrow = T)
+
 ```
+## TODO
+
+- create stimuli table that has complexity & length
+- integrate demo data & stim table in final data
+- add analysis on the basis of complexity & length
+- add analysis on the basis of demo chars
+- fix the back-transformation
+- add summary of types of error: %semantic, %deletion (which sounds?) %subst (which are the most common sounds) 
+- maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
+- discussion
 
@@
 
-```{r}
-read.csv("prevlit.csv")->kids
+```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "Proportion of non-words correctly repeated as a function of age (in years), study (first author and year regardless of the number of authors), and, when available, shorter versus longer non-words."}
+read.csv("prevlit.csv")->pl
+#duplicate Piazzalunga logging short and long items separately
+pl$Score[pl$Study=="Piazzalunga 2018"]<-pl$Short[pl$Study=="Piazzalunga 2018"]
+pl$Study<-as.character(pl$Study)
+pl$Study[pl$Study=="Piazzalunga 2018"]<-"Piazzalunga 2018 1-2 syll"
+dim(pl)[1]->beg
+pl=rbind(pl,pl[pl$Study=="Piazzalunga 2018 1-2 syll",])
+pl[(beg+1):dim(pl)[1],"Score"]<-pl[(beg+1):dim(pl)[1],"Long"]
+pl[(beg+1):dim(pl)[1],"Study"]<-"Piazzalunga 2018 3-4 syll"
+pl$Study=paste(pl$Language,pl$Study)
+library(lattice)
+xyplot(Score~Age,groups=Study,data=pl,type=c('l','p'),auto.key = T ,pch=20)
 
 ```
-
-@ac ADD TABLE SUMMARIZING PREVIOUS WORK
-
-- age differences
-- length effects
 
@@
 
-The child was donned a headset (**xx @mc please fill in** for most of the children, xx @ac fill in for the rest).
+The child was donned a headset (**xx @mc please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to  provide the instructions and, if needed, coach the child to make sure that they understood that the task was to repeat the items precisely without changing anything using the three practice items as well as real words.
-The experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headhphones.
+A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
+An experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headphones.
 
@@
 
-```{r}
-# read.table("log2.txt",header=T,sep="\t")->x
-# x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
-# x[order(x$out),]->x
-# x$cor.online=1
-# for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0
-#   
-# subset(x,rep=="1")->x
-# aggregate(x$cor.online,by=list(x$child),mean)->onl
-# colnames(onl)<-c("ID","score.online")
-# 
-# #print(onl)
-# 
-# aggregate(x$corr,by=list(x$child),mean)->ms
-# colnames(ms)<-c("ID","score")
-# merge(ms,onl,all=T)->ms
-# merge(ms,demo,all=T)->scores
-# scores$Age=as.numeric(as.character(scores$Age))
-# scores$Age[is.na(scores$Age)]<-15
-# plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
-# 
-# plot(scores$score*100~scores$Age,pch=20,xlab="Age (years)",ylab="% Correct")
-# abline(lm(scores$score*100~scores$Age))
-# plot(scores$score~scores$monolingual)
-# points(scores$score~as.numeric(scores$monolingual),pch=20)
-# 
-# plot(ms$score.online~ms$score)
-
-read.csv("NWR-transcription.csv",sep=";")->coding
-
-#remove some weird items
-coding[coding$item!="task4",]->coding
-
-read.csv("final_order_ful.txt",sep="\t")->coding_key
-coding_key$row=row.names(coding_key)
-
-merge(coding,coding_key,all=T)->joint
-
-merge(joint,demo,by.x="id",by.y="ID",all.x=T)->joint
-
-#order so we have all the items by a given child in the right chrono order
-joint[order(joint$file, joint$int),]->joint
-
-#add variable coding whether it's the child's first or subsequent attempts
-joint$previous<-c(NA,as.character(joint$target[1:(dim(joint)[1]-1)]))
-joint$attempt<-ifelse(joint$target!=joint$previous,"first","subsequent")
-joint$attempt[1]<-"first"
-
-#remove practice items
-practice=joint[(joint$target %in% c("wi","poni","nopimade")),]
-joint=joint[!(joint$target %in% c("wi","poni","nopimade")),]
-
-read.csv("stimuli.csv",header=T)->stims
-
-# phonologize things
-merge(joint,stims,by="target",all=T)->joint
-joint$target_pron=as.character(joint$ortho)
-
-joint$pron=as.character(joint$mispronunciation)
-
-correspondances=matrix( #list correspondances always by pairs
-  c(
-    #nasal long to unichar
-    ":ii","ì",
-    ":ee","è",
-    ":aa","à",
-    ":êê","3",
-    ":ââ","6",
-    ":uu","ù",
-    ":oo","Ò",
-    #long to unichar
-    "ii","ï",
-    "ee","ë",
-    "éé","Ë",
-    "áá","Ä",
-    "aa","ä",
-    "îî","Ï",
-    "êê","2",
-    "ââ","5",
-    "uu","ü",
-    "óó","ö",
-    "oo","Ö",
-    #short nasal
-    ":i","1",
-    ":e","4",
-    ":a","ã",
-    ":ê","6",
-    ":â","7",
-    ":u","8",
-    ":o","9",
-    #consonants, veeeery dirty!
-    "'n","ń",
-    "tp",":",
-    "dp","/",
-    "kp","P",
-    "ngm","G",
-    "ńm","M",
-    "ng","N",
-    "mb","b",
-    "nd","D",
-    "nt","T",
-    "nk","K",
-    "dn","?",
-    "kn","!",
-    "vy","V",
-    "mt","C",
-    "md","J",
-    "mg","X",
-    "dm","S",
-    "km","Z",
-    "gh","H",
-    "lv","L"
-  ),#last item above should not have a comma
-  ncol = 2,byrow = T)
-
-for(i in 1:dim(correspondances)[1]) {
-  joint$pron=gsub(correspondances[i,1],correspondances[i,2],joint$pron)
-  joint$target_pron=gsub(correspondances[i,1],correspondances[i,2],joint$target_pron)
-}
-
-#deal with hesitations
-joint$hesitation=F
+```{r wrangling,eval=REDOALL}
+source("wrangling.R")
 
 ```
+
+```{r readin}
+read.table("final_data.txt",header=T,encoding="UTF-8")->trnsc
+#create a subset with just the first presentations -- this is what is going to be used for the quantitative analyses
+firstp=subset(trnsc,rep==1 & !is.na(rep))
+```
+
+Taking into account only the first attempts, the overall repetition accuracy measured in whole words is `r round(mean(firstp$cor.online)*100)`% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and `r round(mean(firstp$correct)*100)`% when the offline coding is considered.
+On average, the phoneme-based normalized Levenshtein distance was `r round(mean(firstp$nld)*100)`%, meaning that about 20% of phonemes were substituted, inserted, or deleted.
+
+We then inspected whether accuracy varied as a function of word length.
+
+```{r figlength,eval=F}
+#not working
+aggregate(firstp$correct,list(firstp$tarlen,firstp$id),mean)
+plot(firstp$correct~firstp$tarlen)
+
+```
+- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
 
@@
 
-aggregate(joint$correct*100,by=list(joint$id,joint$age.exact,joint$attempt),mean)->pwcm_att
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$attempt),mean)->pwcm_att
 colnames(pwcm_att)<-c("ID","Age","Attempt","Word")
@@
 ```{r exclude subsequent}
-joint=joint[joint$attempt=="first",]
+trnsc=trnsc[trnsc$attempt=="first",]
 ```

### 2020-05-14 wrangling.R
lots of bug fixes

@@ -7,8 +7,13 @@ npairs=dim(crp)[1]
 
 trnsc=merge(trnsc,crp,by.x="tokid",by.y="tokid",all=T)
 
-# sort by file name which indicates presentation order, so that we can get the first attempt
-trnsc[order(trnsc$outfile),]->trnsc
+#order so we have all the items by a given child in the right chrono order
+trnsc[order(trnsc$file, trnsc$int),]->trnsc
+
+#add variable coding whether it's the child's first or subsequent attempts
+trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
+trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
+trnsc$attempt[1]<-"first"
 
 #add repetition number
 trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
@@ -18,31 +23,25 @@ trnsc$cor.online=1
 for(i in 1:npairs) if(trnsc$rep[i]>1) trnsc$cor.online[i-1]<-0  #if the item is repeated, then code the previous one as being wrong
 trnsc$cor.online[(npairs+1):dim(trnsc)[1]]<-NA
 
-#DATA REMOVAL *** ATTENTION *** 
-#REMOVING ALL THE "YI"
-trnsc[trnsc$target!="yi",]->trnsc
-#REMOVE ALL TASK 4
-trnsc[!is.na(trnsc$target),]->trnsc
-
-#remove practice items
-practice=trnsc[(trnsc$target %in% c("wi","poni","nopimade")),]
-trnsc=trnsc[!(trnsc$target %in% c("wi","poni","nopimade")),]
-
 merge(trnsc,demo,by.x="id",by.y="ID",all.x=T)->trnsc
-
-#order so we have all the items by a given child in the right chrono order
-trnsc[order(trnsc$file, trnsc$int),]->trnsc
-
-#add variable coding whether it's the child's first or subsequent attempts
-trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
-trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
-trnsc$attempt[1]<-"first"
-
 
 #add orthographic representations targets & other stims chars
 read.delim("stimuli.tsv", encoding="UTF-8")->stims
[…]

### 2020-05-14 manuscript.Rmd
lots of bug fixes

@@
 ```
+
 ## TODO
 
-- create stimuli table that has complexity & length
-- integrate demo data & stim table in final data
-- add analysis on the basis of complexity & length
-- add analysis on the basis of demo chars
-- fix the back-transformation
+- fix yaml
+- add ana based on phonemes & expand on levenshtein
+- add ana based on number of characters
 - add summary of types of error: %semantic, %deletion (which sounds?) %subst (which are the most common sounds) 
 - maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
+- add analysis on the basis of demo chars
 - discussion
@@
 
+### Preliminary analyses
+
 ```{r readin}
 read.table("final_data.txt",header=T,encoding="UTF-8")->trnsc
-#create a subset with just the first presentations -- this is what is going to be used for the quantitative analyses
-firstp=subset(trnsc,rep==1 & !is.na(rep))
-```
-
-Taking into account only the first attempts, the overall repetition accuracy measured in whole words is `r round(mean(firstp$cor.online)*100)`% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and `r round(mean(firstp$correct)*100)`% when the offline coding is considered.
-On average, the phoneme-based normalized Levenshtein distance was `r round(mean(firstp$nld)*100)`%, meaning that about 20% of phonemes were substituted, inserted, or deleted.
-
-We then inspected whether accuracy varied as a function of word length.
-
-```{r figlength,eval=F}
-#not working
-aggregate(firstp$correct,list(firstp$tarlen,firstp$id),mean)
-plot(firstp$correct~firstp$tarlen)
 
 ```
-- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
-
-### Preliminary analyses
 
@@
 colnames(pwcm_att)<-c("ID","Age","Attempt","Word")
+
+#remove children who do not have both first & subsequent attempts
+pwcm_att[pwcm_att$ID %in% names(table(pwcm_att$ID))[table(pwcm_att$ID)==2],]->pwcm_att
 
@@
 
-We start by describing whole word accuracy. Participants scored
+Taking into account only the first attempts, the overall repetition accuracy measured in whole words is `r round(mean(trnsc$cor.online)*100)`% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and `r round(mean(trnsc$correct)*100)`% when the offline coding is considered.
+On average, the phoneme-based normalized Levenshtein distance was `r round(mean(trnsc$nld)*100)`%, meaning that about 20% of phonemes were substituted, inserted, or deleted.
 
-```{r}
+ADD GRAPH WITH PHONEME BASED IN Y AND WORD BASED IN X
+
+### Accuracy a function of item complexity
+
+We then inspected whether accuracy varied as a function of word length.
+
+```{r figlength,eval=F}
+#not working
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$tarlen),mean)->pwcm_att
+colnames(pwcm_att)<-c("ID","Age","Length","Word")
+
+#but this one is! What a mystery...
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->pwcm_att
+colnames(pwcm_att)<-c("ID","Age","Length","Word")
+
+myx=as.numeric(as.factor(pwcm_att$ID))/100+as.numeric(as.factor(pwcm_att$Length))
+
+plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="NWR scores (%)")
+axis(1,at=1:length(levels(factor(pwcm_att$Length))),labels=levels(factor(pwcm_att$Length)))
 
 ```
-
-phoneme based, distance
-
-### Accuracy a function of item complexity 
-(number of syllablex, sound complexity)?
 

### 2020-05-14 manuscript.tex
lots of bug fixes

@@
 
+\hypertarget{todo}{%
+\subsection{TODO}\label{todo}}
+
+\begin{itemize}
+\tightlist
+\item
+  fix yaml
+\item
+  add ana based on phonemes \& expand on levenshtein
+\item
+  add ana based on number of characters
+\item
+  add summary of types of error: \%semantic, \%deletion (which sounds?) \%subst (which are the most common sounds)
+\item
+  maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
+\item
+  add analysis on the basis of demo chars
+\item
+  discussion
+\end{itemize}
+
 \hypertarget{introduction}{%
@@
 
-The child was donned a headset (\textbf{xx ({\textbf{???}}) please fill in} for most of the children, xx ({\textbf{???}}) fill in for the rest).
+The child was donned a headset (\textbf{xx ({\textbf{???}}) please fill in} for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to provide the instructions and, if needed, coach the child to make sure that they understood that the task was to repeat the items precisely without changing anything using the three practice items as well as real words.
-The experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headhphones.
+A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
+An experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headphones.
 
@@
 A total 55 children were tested, from 34 of families, in five hamlets.
-Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5).
+Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=0), spoke too softly to allow offline coding (N=5).
 In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
@@
 
+\hypertarget{preliminary-analyses}{%
+\subsubsection{Preliminary analyses}\label{preliminary-analyses}}
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/Fig1-first_vs_others-1.pdf}
+\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual participants averaging separately their first attempts and all other attempts.}
+\end{figure}
+
+We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
+As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
+Surprisingly, subsequent repetitions (M = 42, SD = 28)
+were on average less accurate than first ones (M = 64, SD = 15), t(36) = 5.32, p = 0.
+Given the uncertainty in whether previous work used only the first or all repetitions, and since behavior degraded and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
+
 \hypertarget{overall-repetition-accuracy}{%
@@
 
-We start by describing whole word accuracy
+Taking into account only the first attempts, the overall repetition accuracy measured in whole words is 72\% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and 65\% when the offline coding is considered.
+On average, the phoneme-based normalized Levenshtein distance was 22\%, meaning that about 20\% of phonemes were substituted, inserted, or deleted.
 
-phoneme based, distance
-\#\#\# Accuracy a function of item complexity
-(number of syllablex, sound complexity)?
+ADD GRAPH WITH PHONEME BASED IN Y AND WORD BASED IN X
+
+\hypertarget{accuracy-a-function-of-item-complexity}{%
+\subsubsection{Accuracy a function of item complexity}\label{accuracy-a-function-of-item-complexity}}
+
+We then inspected whether accuracy varied as a function of word length.
 

### 2020-05-15 manuscript.tex
fixed yaml

@@
 %
-\documentclass[english,,man,floatsintext]{apa6}
+\documentclass[english,,man]{apa6}
 \usepackage{lmodern}
@@
 \hypersetup{
+            pdftitle={Non-word repetition in Yélî Dnye},
+            pdfkeywords={phonology, non-word repetition, development},
             pdfborder={0 0 0},
@@
 \makeatother
-\shorttitle{NWR in Yélî Dnye'}
+\shorttitle{NWR in Yélî Dnye}
 \author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2}}
@@
 \vspace{0.5cm}
-\textsuperscript{1} LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
-\authornote{
+\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
+\authornote{All data are made available in a repository in the Open Science Framework.
+AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
+S.
+McDonnell Foundation Understanding Human Cognition Scholar Award.
 
-Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d'Ulm, 75005, Paris, France. E-mail: alecristia@gmail.com}
-\keywords{phonology, non-word repetition, development\newline\indent Word count: xxx}
+Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
+\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
+\DeclareDelayedFloatFlavor{ThreePartTable}{table}
+\DeclareDelayedFloatFlavor{lltable}{table}
+\DeclareDelayedFloatFlavor*{longtable}{table}
+\makeatletter
+\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
+\makeatother
+\usepackage{lineno}
+
+\linenumbers
 \usepackage{csquotes}
@@
 
+\title{Non-word repetition in Yélî Dnye}
+
 \date{}
 
+\abstract{
+ADD
+}
+
 \begin{document}
+\maketitle
 
@@
 
-Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., {\textbf{???}}).
+Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., Hazan \& Barrett, 2000).
 One common task in this line of research is nonword repetition (NWR).
@@
 
-NWR has been used to seek answers to a variety of theoretical questions, including what are the links between phonology, working memory, and the lexicon ({\textbf{???}}), as well as for applications, notably using nwr as a diagnostic for language delays and disorders ({\textbf{???}}).
-Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe ({\textbf{???}},@is08042009language).
+NWR has been used to seek answers to a variety of theoretical questions, including what are the links between phonology, working memory, and the lexicon (Bowey, 2001), as well as for applications, notably using nwr as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
+Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe (Meir, Walters, \& Armon-Lotem, 2016,@is08042009language).
 
@@
 There is some variation in the presentation procedure and structure of items found in previous NWR work.
-For example, items are often presented orally by the experimenter ({\textbf{???}}), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation ({\textbf{???}}).
-Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items ({\textbf{???}}).
-Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, {\textbf{???}}).
+For example, items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation (Brandeker \& Thordardottir, 2015).
+Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
+Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
 
@@
 That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
-Based on previous work, we looked at potential improvements with age ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}), and potential negative effects of bilingual exposure ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}).
-Previous work typically finds no significant differences as a function of maternal education (e.g., {\textbf{???}}; {\textbf{???}}; {\textbf{???}}) or child gender ({\textbf{???}}).
-Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks ({\textbf{???}}), and therefore we report on those too.
+Based on previous work, we looked at potential improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and potential negative effects of bilingual exposure (Brandeker \& Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
+Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on those too.
 
@@
 However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects.
-For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small (data from {\textbf{???}}).
+For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small (data from Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
 \textbf{add table with r/ds from previous work})
@@
 
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language (e.g., {\textbf{???}}; {\textbf{???}}).
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
 We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in Yély Dnye.
-We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat (note that up to other work has also used 24-30 items: {\textbf{???}}; {\textbf{???}}), and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
+We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat (note that up to other work has also used 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
 
@@
 These were scored as correct or incorrect on the fly depending on whether the native experimenter asked for the item to be repeated; as well as offline.
-Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy (e.g., {\textbf{???}}).
+Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy (e.g., Cristia et al., 2020).
 We calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
@@
 
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-balladares2016socio}{}%
+Balladares, J., Marshall, C., \& Griffiths, Y.
+(2016).
+Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
+\emph{First Language}, \emph{36}(3), 338--351.
+\url{https://doi.org/10.1177/0142723715626067}
+
+\leavevmode\hypertarget{ref-bowey2001nonword}{}%
+Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
+
+\leavevmode\hypertarget{ref-brandeker2015language}{}%
+Brandeker, M., \& Thordardottir, E.
+(2015).
+Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
+
+\leavevmode\hypertarget{ref-chiat2007preschool}{}%
+Chiat, S., \& Roy, P.
+(2007).
+The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
+
+\leavevmode\hypertarget{ref-cristia2020infant}{}%
+Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
+(2020).
+Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
+\emph{Preprint}.
+
+\leavevmode\hypertarget{ref-estes2007differences}{}%
+Estes, K.
+G., Evans, J.
+L., \& Else-Quest, N.
+M.
+(2007).
+Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
+
+\leavevmode\hypertarget{ref-farmani2018normalization}{}%
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
+Z., Talebi, E., \& Shourvazi, Z.
+(2018).
+Normalization of the non-word repetition test in farsi-speaking children.
+\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
+
+\leavevmode\hypertarget{ref-gallon2007non}{}%
+Gallon, N., Harris, J., \& Van der Lely, H.
+(2007).
+Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
+\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\leavevmode\hypertarget{ref-havron2019effect}{}%
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
+M.-C.
+C.
+S.
+(2019).
+The effect of older siblings on language development as a function of age difference and sex.
+\emph{Psychological Science}, \emph{30}(9), 1333--1343.
+
+\leavevmode\hypertarget{ref-hazan2000development}{}%
+Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
+
+\leavevmode\hypertarget{ref-is08042009language}{}%
+IS0804, C. A. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
+\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
+
+\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
+(2014).
+Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
+\emph{PloS One}, \emph{9}(2), e89544.
+
+\leavevmode\hypertarget{ref-meir2017independent}{}%
+Meir, N., \& Armon-Lotem, S.
+(2017).
+Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
+\emph{Frontiers in Psychology}, \emph{8}, 1442.
+
+\leavevmode\hypertarget{ref-meir2016disentangling}{}%
+Meir, N., Walters, J., \& Armon-Lotem, S.
+(2016).
+Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
+\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
+
+\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
+(2019).
+An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
+\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
+
+\leavevmode\hypertarget{ref-torrington2015non}{}%
+Torrington Eaton, C., Newman, R.
+S., Ratner, N.
+B., \& Rowe, M.
+L.
+(2015).
+Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
+\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+
+\leavevmode\hypertarget{ref-vance2005speech}{}%
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
+
+\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
+Wilsenach, C.
+(2013).
+Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
+\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+\url{https://doi.org/10.5785/29-2-554}
+
 \end{document}

### 2020-05-15 manuscript.Rmd
fixed yaml

@@
 ---
-title: "Non-word repetition in Yélî Dnye"
-shorttitle        : "NWR in Yélî Dnye'"
+title             : "Non-word repetition in Yélî Dnye"
+shorttitle        : "NWR in Yélî Dnye"
 
@@
     corresponding : yes    # Define only one corresponding author
-    address       : "29, rue d'Ulm, 75005, Paris, France"
+    address       : "29, rue d’Ulm, 75005 Paris, France"
     email         : "alecristia@gmail.com"
@@
   - id            : "1"
-    institution   : "LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL"
+    institution   : "Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University"
   - id            : "2"
@@
 
-author_note: >
+authornote: |
+  All data are made available in a repository in the Open Science Framework.
+AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
+S.
+McDonnell Foundation Understanding Human Cognition Scholar Award.
 
@@
 keywords          : "phonology, non-word repetition, development"
-wordcount         : "xxx"
+wordcount         : "~xxx words"
 
@@
 
-figsintext        : yes
+figsintext        : no
 figurelist        : no
@@
 footnotelist      : no
-lineno            : no
+lineno            : yes
 
@@
 output            : papaja::apa6_pdf
-
 ---
@@
 
-- fix yaml
 - add ana based on phonemes & expand on levenshtein

### 2020-05-20 manuscript.Rmd
updated todo

@@
 
+- add table r/d by stim length & age
+- add table with stimuli
+- ask Middy about consent, correct, accuracy
+- double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
+- 
 - add ana based on phonemes & expand on levenshtein
@@
 trnsc=trnsc[trnsc$attempt=="first",]
+
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->pwcm_att
+colnames(pwcm_att)<-c("ID","Age","Length","Word")
+
+mean(pwcm_att$Word[pwcm_att$Length==2],na.rm=T)
 ```

### 2020-05-20 wrangling.R
advanced a bit

@@ -136,7 +136,7 @@ trnsc$ld=rowSums(trnsc[,c("ins","del","sub")])
 trnsc$nld=trnsc$ld/trnsc$nchar
 trnsc$nld[is.na(trnsc$nld)]<-0
 
-#proportion correct doesn't penalize Substitutions Deletions and Insertions
+#proportion correct: remove Substitutions Deletions and Insertions & count only proportion of matches
 trnsc$pc=nchar(gsub("[SDI]","",trnsc$trafos))/trnsc$nchar
 trnsc$pc[is.na(trnsc$pc)]<-1
 

### 2020-05-20 manuscript.Rmd
advanced a bit

--- 2020-05-15 manuscript.Rmd
@@
 
-figsintext        : no
+figsintext        : yes
 figurelist        : no
@@
 
-- add ana based on phonemes & expand on levenshtein
-- add ana based on number of characters
+- add table r/d by stim length & age
+- add table with stimuli
+- ask Middy about consent, correct, accuracy
+- double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
 - add summary of types of error: %semantic, %deletion (which sounds?) %subst (which are the most common sounds) 
@@
 #remove children who do not have both first & subsequent attempts
+# which is
+excl=sum(table(pwcm_att$ID)!=2)
+
 pwcm_att[pwcm_att$ID %in% names(table(pwcm_att$ID))[table(pwcm_att$ID)==2],]->pwcm_att
@@
 We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
+We excluded `r excl` children who did not have data for one of these two types.
 As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
@@
 trnsc=trnsc[trnsc$attempt=="first",]
+
 ```
@@
 
-Taking into account only the first attempts, the overall repetition accuracy measured in whole words is `r round(mean(trnsc$cor.online)*100)`% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and `r round(mean(trnsc$correct)*100)`% when the offline coding is considered.
-On average, the phoneme-based normalized Levenshtein distance was `r round(mean(trnsc$nld)*100)`%, meaning that about 20% of phonemes were substituted, inserted, or deleted.
+```{r overall}
 
-ADD GRAPH WITH PHONEME BASED IN Y AND WORD BASED IN X
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact),mean)->overall
+aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$age.exact),mean)->pc
+aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.exact),mean)->nld
+overall=cbind(overall,pc$x,nld$x)
+colnames(overall)<-c("ID","Age","Word","Phoneme","NLD")
+
+```
+
+Taking into account only the first attempts, we averaged attempts by each of the `r dim(overall)[1]` children who had data for first attempts; their ages ranged from `r min(overall$Age)` to `r max(overall$Age)` (M = `r round(mean(overall$Age),1)`, SD = `r round(sd(overall$Age),1)`).
+
+The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%).
+Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%).
+The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
+Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
 
@@
 
+```{r bylength}
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->by_length
+aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->pc
+aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->nld
+by_length=cbind(by_length,pc$x,nld$x)
+colnames(by_length)<-c("ID","Age","Length","Word","Phoneme","NLD")
+
+```
+
 We then inspected whether accuracy varied as a function of word length.
+Results are shown on table XX.
+This table shows that monosyllables accuracy was much lower than other lengths.
+This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
 
-```{r figlength,eval=F}
-#not working
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$tarlen),mean)->pwcm_att
-colnames(pwcm_att)<-c("ID","Age","Length","Word")
+Setting monosyllables aside, we observe the typical pattern of decreased accuracy for longer items, although this is particularly salient for the whole word scoring.
+This is the most commonly reported type of score, but it is also the least forgiving.
+The pattern is less marked when other two scores are used, which are less sensitive to errors.
 
-#but this one is! What a mystery...
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->pwcm_att
-colnames(pwcm_att)<-c("ID","Age","Length","Word")
-
-myx=as.numeric(as.factor(pwcm_att$ID))/100+as.numeric(as.factor(pwcm_att$Length))
-
-plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="NWR scores (%)")
-axis(1,at=1:length(levels(factor(pwcm_att$Length))),labels=levels(factor(pwcm_att$Length)))
-
+```{r tablength}
+mytab=NULL
+for(myvar in c("Word","Phoneme","NLD")){
+  mytab=cbind(mytab,
+              paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
+              " (",round(aggregate(by_length[,myvar],by=list(by_length$Length),sd)$x),")"))
+}
+rownames(mytab)<-levels(by_length$Length)
+colnames(mytab)<-c("Word","Phoneme","NLD")
+apa_table(mytab,caption="NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
 ```

### 2020-05-20 manuscript.tex
advanced a bit

@@
 %
-\documentclass[english,,man]{apa6}
+\documentclass[english,,man,floatsintext]{apa6}
 \usepackage{lmodern}
@@
 \keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
 \usepackage{lineno}
@@
 \item
-  fix yaml
+  add table r/d by stim length \& age
 \item
-  add ana based on phonemes \& expand on levenshtein
+  add table with stimuli
 \item
-  add ana based on number of characters
+  ask Middy about consent, correct, accuracy
+\item
+  double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
 \item
@@
 We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
+We excluded 5 children who did not have data for one of these two types.
 As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
@@
 
-Taking into account only the first attempts, the overall repetition accuracy measured in whole words is 72\% based on the online coding (i.e., whether the native experimenter asked for the target to be re-played), and 65\% when the offline coding is considered.
-On average, the phoneme-based normalized Levenshtein distance was 22\%, meaning that about 20\% of phonemes were substituted, inserted, or deleted.
+Taking into account only the first attempts, we averaged attempts by each of the 37 children who had data for first attempts; their ages ranged from 3.92 to 12.86 (M = 7.10, SD = 1.90).
 
-ADD GRAPH WITH PHONEME BASED IN Y AND WORD BASED IN X
+The overall NWR score was M = 64\% (SD = 15\%).
+Scores based on phonemes are even higher M = 78\% (SD = 9\%).
+The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
+Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
 
@@
 We then inspected whether accuracy varied as a function of word length.
+Results are shown on table XX.
+This table shows that monosyllables accuracy was much lower than other lengths.
+This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
+
+Setting monosyllables aside, we observe the typical pattern of decreased accuracy for longer items, although this is particularly salient for the whole word scoring.
+This is the most commonly reported type of score, but it is also the least forgiving.
+The pattern is less marked when other two scores are used, which are less sensitive to errors.
 

### 2020-05-25 manuscript.Rmd
found

@@
 - double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
-- add summary of types of error: %semantic, %deletion (which sounds?) %subst (which are the most common sounds) 
+- probably turn summary of types of error: into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set
 - maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
@@
 
-In view of the hypothesis-driven nature of this work, we considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
+In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
 However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects.
@@
 
-### Accuracy a function of item complexity
+### Accuracy a function of item length
 
@@
 
+### Patterns of errors
+- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+
+```{r diffic,echo=F}
+nasalVs=c(":a", ":â" ,":ê", ":i" ,":o")
+oralVs=c("a", "â","e",  "é" ,"ê", "i","î" ,"o","ó","u")
+simplexCs=c("m","n","l","w","y","v","d","g","p","t","k","f","h","ch")
+complexCs=c("dp","tp","kp","km","kn","mb","gh")
+
+allVC=c(nasalVs,oralVs,simplexCs,complexCs)
+
+read.table("substitution_bank.txt",header=T)->sub_bank
+colnames(sub_bank)<-c("target","realization")
+sub_bank$target=factor(sub_bank$target,levels=allVC)
+sub_bank$realization=factor(sub_bank$realization,levels=allVC)
+
+sub_tab=table(sub_bank)[allVC,allVC]
+
+read.table("deletion_bank.txt",header=T)->del_bank
+colnames(del_bank)<-c("target")
+del_bank$target=factor(del_bank$target,levels=allVC)
+del_tab= table(del_bank)[allVC]
+```
+
+We investigated patterns of deletion and substitution.
+Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted, and `r sum(del_tab[c(simplexCs,complexCs)])` consonants.
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral target vowels produced as nasal vowels, `r sum(sub_tab[nasalVs,oralVs])` nasal target vowels produced as oral vowels).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (`r sum(sub_tab[oralVs,oralVs])` oral vowels produced with a different quality; `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels produced with a different quality).
+As for consonants, asymmetries were very marked with more complex consonants (specifically `r paste0(complexCs)`) mispronounced as simple ones (specifically `r paste0(simplexCs)`, `r sum(sub_tab[complexCs,simplexCs])`) than vice versa (`r sum(sub_tab[simplexCs,complexCs])`).
+Simple consonants were mispronounced as other simple consonants quite frequently (`r sum(sub_tab[simplexCs,simplexCs])` simple consonants mispronounced as other simple ones, compared to `r sum(sub_tab[complexCs,complexCs])` complex ones).
+
+**probably turn this into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set**
+
+Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: `r round(sum(!is.na(trnsc$meaning))/sum(!is.na(trnsc$mispronunciation))*100)`%.
+
+** this is higher than any previous report! can we be sure of it?**
+
+### Factor structuring individual variation
+
+explainable by child age, sex, birth order, monolingual status, and/or parental education?
+
+## Discussion
+
+- What is the overall repetition accuracy (whole word, phoneme based, distance)?
+
+- How does this change as a function of item complexity (number of syllables, sound complexity)?
+
 1. Children are more accurate for mono-syllables than longer items
@@
 
-### Patterns of errors
-- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+- How frequent are errors that result in real words? Is that a function of item complexity?
+- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
 
-### Factor structuring individual variation
-
-explainable by child age, sex, birth order, monolingual status, and/or parental education?
 3. Children's accuracy increases with child age.
@@
 
-## Discussion
-
-- What is the overall repetition accuracy (whole word, phoneme based, distance)?
-- How does this change as a function of item complexity (number of syllables, sound complexity)?
-- How frequent are errors that result in real words? Is that a function of item complexity?
-- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
-
 \newpage

### 2020-05-25 manuscript.Rmd
added stimuli table and demo ana

@@
 require("papaja")
-
+library(kableExtra)
+library(gdata)
 correspondances=matrix( #list correspondances always by pairs, orthography then phonology
@@
 - add table r/d by stim length & age
-- add table with stimuli
-- ask Middy about consent, correct, accuracy
 - double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
-- probably turn summary of types of error: into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set
+- discussion
+
+## TODO Middy
+
+- look through
+- consent, "correct", "accuracy"
+- help with table 1 (phonological representation but also formatting)
+
+## TODO later
+- probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set
 - maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
-- add analysis on the basis of demo chars
-- discussion
+- maybe add analyses of score as a function of frequency (phones, diphones) in MC's corpus? (mc suggested, but I don't think we need it...)
 
@@
 Even so, there was one monosyllable that was often identified as a real word (intended "yî" /yXX/; identified as "yi" /yi/, *tree*).
-This item is removed from analyses.
+This item is removed from analyses below.
 
-**@mc can you please add the IPA in // for all of the items in the table?**
+**@mc can you please add the IPA in // for all of the items in the excel then save as tsv?**
 The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
 
-```{r}
+```{r tab2-stims}
+#add orthographic representations targets & other stims chars
+read.delim("stimuli.tsv", encoding="UTF-8")->stims
+stims=stims[order(stims[,1]),-3]
+colnames(stims)<-c("Type","Orthographic","Phonological")
+cbindX(stims[stims$Type=="practice",],
+      stims[stims$Type==1,],
+      stims[stims$Type==2,],
+      stims[stims$Type==3,],
+      stims[stims$Type==4,])
 
+stims=stims[,-1]
+
+#this is leading to an error
+#mytab2 = kable(stims, "latex", booktabs = T,caption="NWR stimuli in orthographic and phonological representations.", row.names =F) %>%
+#kable_styling() %>%
+#add_header_above(c("Practice" = 2, "Monosyll" = 2, "Bisyll" = 2, "Trisyll" = 2, "Tetrasyll" = 2))
+
+apa_table(stims,"NWR stimuli in orthographic and phonological representations.")
 ```
-
-Practice
-wî, poni, nopimade
-
-Mono
-dpa
-dp:a
-dpâ
-dpéé
-dpê
-dpi
-dpu
-kp:ââ
-kpu
-tp:a
-tpâ
-tpê
-gh:ââ
-ghuu
-lva
-lv:ê
-lvi
-t:êê
-);
-
-(
-kamo
-kipo
-kani
-tupa
-noki
-piwa
-towi
-nomi);
-
- (
-nademo
-meyadi
-diyeto
-widone
-nuyedi
-tumowe
-pedumi
-dimope
-tiwune
-wumipo
-nayeki
-mituye);
-
-(
-todiwuma
-wadikeno
-nomiwake
-diponate).
 
@@
 
-The local informant asked for the item to be repeated when she thought the repetition was not correct.
-We will call this *online scoring*, and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
-
-We sought additional information by asking a local informant to listen through each child production, paired with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated.
-We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-This *offline scoring* provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.
+A script was written to randomize all tokens from all children, pairing each with the auditory target the child had been provided with.
+A native research assistant then listened through all productions of a given target (randomized across children and repetition order), and made a judgment of whether the item was correctly or incorrectly repeated.
+She additionally transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
 
@@
 
-For NWR accuracy, we considered separately the child's first and subsequent attempts.
-These were scored as correct or incorrect on the fly depending on whether the native experimenter asked for the item to be repeated; as well as offline.
-Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy [e.g., @cristia2020infant].
-We calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
+Some NWR studies employ phoneme-based scoring in addition to or instead of word-level accuracy [e.g., @cristia2020infant].
+We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
 Although previous work does not use distance metrics, we additionally report those.
@@
 
+#quick fix, but should be fixed upstream @ac
+trnsc$Sex[trnsc$Sex=="F"]<-"G"
+trnsc$Sex=factor(trnsc$Sex)
+trnsc$birthOrder=as.numeric(trnsc$birthOrder)
 ```
@@
 
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact),mean)->overall
-aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$age.exact),mean)->pc
-aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.exact),mean)->nld
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.exact,trnsc$birthOrder,trnsc$matEd),mean)->overall
+aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.exact,trnsc$birthOrder,trnsc$matEd),mean)->pc
+aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.exact,trnsc$birthOrder,trnsc$matEd),mean)->nld
 overall=cbind(overall,pc$x,nld$x)
-colnames(overall)<-c("ID","Age","Word","Phoneme","NLD")
+colnames(overall)<-c("ID","Sex","Age","BrthO","matEd","Word","Phoneme","NLD")
 
@@
 del_tab= table(del_bank)[allVC]
+
+#remove English meanings
+trnsc$meaning[grep("English",trnsc$meaning)]<-NA
 ```
@@
 
-explainable by child age, sex, birth order, monolingual status, and/or parental education?
+```{r Fig2-scores by age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-word scores for individual participants as a function of age and sex (blue = boys, pink = girls)."}
+
+plot(overall$Word~overall$Age,pch=20,cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","blue","pink"))
+
+cor.test(overall$Word,overall$Age,method="s")->corage
+t.test(overall$Word~overall$Sex)->tsex
+cor.test(overall$Word,overall$BrthO,method="s")->corbo
+cor.test(overall$Word,overall$matEd,method="s")->corme
+
+```
+
+Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children  (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p = `r corage$p.value`).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p = `r tsex$p.value`), birth order  (`r corbo$statistic`) = `r corbo$estimate`, p = `r corbo$p.value`) or maternal education  (`r corme$statistic`) = `r corme$estimate`, p = `r corme$p.value`)
 

### 2020-05-25 manuscript.tex
added stimuli table and demo ana

@@
 \item
-  add table with stimuli
+  double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
+\item
+  discussion
+\end{itemize}
+
+\hypertarget{todo-later}{%
+\subsection{TODO later}\label{todo-later}}
+
+\begin{itemize}
+\tightlist
 \item
@@
 \item
-  double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
+  ask Middy to look through
 \item
-  add summary of types of error: \%semantic, \%deletion (which sounds?) \%subst (which are the most common sounds)
+  probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set
 \item
   maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
-\item
-  add analysis on the basis of demo chars
-\item
-  discussion
 \end{itemize}
@@
 Even so, there was one monosyllable that was often identified as a real word (intended \enquote{yî} /yXX/; identified as \enquote{yi} /yi/, \emph{tree}).
-This item is removed from analyses.
+This item is removed from analyses below.
 
-\textbf{({\textbf{???}}) can you please add the IPA in // for all of the items in the table?}
+\textbf{({\textbf{???}}) can you please add the IPA in // for all of the items in the excel then save as tsv?}
 The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
 
-Practice
-wî, poni, nopimade
+\begin{verbatim}
+##        Type Orthographic Phonological Type Orthographic Phonological Type
+## 28 practice     nopimade           NA    1         dp:a           NA    2
+## 32 practice        poni            NA    1          dpa           NA    2
+## 43 practice          wî            NA    1          dpâ           NA    2
+## 1      <NA>         <NA>           NA    1          dpê           NA    2
+## 2      <NA>         <NA>           NA    1         dpéé           NA    2
+## 3      <NA>         <NA>           NA    1          dpi           NA    2
+## 4      <NA>         <NA>           NA    1          dpu           NA    2
+## 5      <NA>         <NA>           NA    1        gh:ââ           NA    2
+## 6      <NA>         <NA>           NA    1         ghuu           NA <NA>
+## 7      <NA>         <NA>           NA    1        kp:ââ           NA <NA>
+## 8      <NA>         <NA>           NA    1          kpu           NA <NA>
+## 9      <NA>         <NA>           NA    1         lv:ê           NA <NA>
+## 10     <NA>         <NA>           NA    1          lva           NA <NA>
+## 11     <NA>         <NA>           NA    1          lvi           NA <NA>
+## 12     <NA>         <NA>           NA    1         t:êê           NA <NA>
+## 13     <NA>         <NA>           NA    1         tp:a           NA <NA>
+## 14     <NA>         <NA>           NA    1          tpâ           NA <NA>
+## 15     <NA>         <NA>           NA    1          tpê           NA <NA>
+##    Orthographic Phonological Type Orthographic Phonological Type Orthographic
+## 28         kamo           NA    3       dimope           NA    4     diponate
+## 32         kani           NA    3       diyeto           NA    4     nomiwake
+## 43         kipo           NA    3       meyadi           NA    4     todiwuma
+## 1          noki           NA    3       mituye           NA    4     wadikeno
+## 2          nomi           NA    3       nademo           NA <NA>         <NA>
+## 3          piwa           NA    3       nayeki           NA <NA>         <NA>
+## 4          towi           NA    3       nuyedi           NA <NA>         <NA>
+## 5          tupa           NA    3       pedumi           NA <NA>         <NA>
+## 6          <NA>           NA    3       tiwune           NA <NA>         <NA>
+## 7          <NA>           NA    3       tumowe           NA <NA>         <NA>
+## 8          <NA>           NA    3       widone           NA <NA>         <NA>
+## 9          <NA>           NA    3       wumipo           NA <NA>         <NA>
+## 10         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
+## 11         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
+## 12         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
+## 13         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
+## 14         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
+## 15         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
+##    Phonological
+## 28           NA
+## 32           NA
+## 43           NA
+## 1            NA
+## 2            NA
+## 3            NA
+## 4            NA
+## 5            NA
+## 6            NA
+## 7            NA
+## 8            NA
+## 9            NA
+## 10           NA
+## 11           NA
+## 12           NA
+## 13           NA
+## 14           NA
+## 15           NA
+\end{verbatim}
 
-Mono
-dpa
-dp:a
-dpâ
-dpéé
-dpê
-dpi
-dpu
-kp:ââ
-kpu
-tp:a
-tpâ
-tpê
-gh:ââ
-ghuu
-lva
-lv:ê
-lvi
-t:êê
-);
+\begin{table}[tbp]
 
-(
-kamo
-kipo
-kani
-tupa
-noki
-piwa
-towi
-nomi);
+\begin{center}
+\begin{threeparttable}
 
-(
-nademo
-meyadi
-diyeto
-widone
-nuyedi
-tumowe
-pedumi
-dimope
-tiwune
-wumipo
-nayeki
-mituye);
+\caption{\label{tab:tab2-stims}NWR stimuli in orthographic and phonological representations.}
 
-(
-todiwuma
-wadikeno
-nomiwake
-diponate).
+\begin{tabular}{ll}
+\toprule
+Orthographic & \multicolumn{1}{c}{Phonological}\\
+\midrule
+dp:a & NA\\
+dpa & NA\\
+dpâ & NA\\
+dpê & NA\\
+dpéé & NA\\
+dpi & NA\\
+dpu & NA\\
+gh:ââ & NA\\
+ghuu & NA\\
+kp:ââ & NA\\
+kpu & NA\\
+lv:ê & NA\\
+lva & NA\\
+lvi & NA\\
+t:êê & NA\\
+tp:a & NA\\
+tpâ & NA\\
+tpê & NA\\
+kamo & NA\\
+kani & NA\\
+kipo & NA\\
+noki & NA\\
+nomi & NA\\
+piwa & NA\\
+towi & NA\\
+tupa & NA\\
+dimope & NA\\
+diyeto & NA\\
+meyadi & NA\\
+mituye & NA\\
+nademo & NA\\
+nayeki & NA\\
+nuyedi & NA\\
+pedumi & NA\\
+tiwune & NA\\
+tumowe & NA\\
+widone & NA\\
+wumipo & NA\\
+diponate & NA\\
+nomiwake & NA\\
+todiwuma & NA\\
+wadikeno & NA\\
+nopimade & NA\\
+poni & NA\\
+wî & NA\\
+\bottomrule
+\end{tabular}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
 
@@
 
-The local informant asked for the item to be repeated when she thought the repetition was not correct.
-We will call this \emph{online scoring}, and it reflects a wholesale impression of whether the item was correctly or incorrectly repeated.
-
-We sought additional information by asking a local informant to listen through each child production, paired with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated.
-We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-This \emph{offline scoring} provides both accuracy and qualitative desciptions of what the child said that are likely more reliable than those of the non-native coder.
+A script was written to randomize all tokens from all children, pairing each with the auditory target the child had been provided with.
+A native research assistant then listened through all productions of a given target (randomized across children and repetition order), and made a judgment of whether the item was correctly or incorrectly repeated.
+She additionally transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
 
@@
 
-For NWR accuracy, we considered separately the child's first and subsequent attempts.
-These were scored as correct or incorrect on the fly depending on whether the native experimenter asked for the item to be repeated; as well as offline.
-Some NWR studies employ phoneme-based accuracy in addition to or instead of word-level accuracy (e.g., Cristia et al., 2020).
-We calculated accuracy as the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
+Some NWR studies employ phoneme-based scoring in addition to or instead of word-level accuracy (e.g., Cristia et al., 2020).
+We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
 Although previous work does not use distance metrics, we additionally report those.
@@
 
-Taking into account only the first attempts, we averaged attempts by each of the 37 children who had data for first attempts; their ages ranged from 3.92 to 12.86 (M = 7.10, SD = 1.90).
+Taking into account only the first attempts, we averaged attempts by each of the 24 children who had data for first attempts; their ages ranged from 3.92 to 10.20 (M = 6.50, SD = 1.50).
 
-The overall NWR score was M = 64\% (SD = 15\%).
-Scores based on phonemes are even higher M = 78\% (SD = 9\%).
-The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
+The overall NWR score was M = 60\% (SD = 14\%).
+Scores based on phonemes are even higher M = 76\% (SD = 9\%).
+The phoneme-based normalized Levenshtein distance was M = 24\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
 Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
 
-\hypertarget{accuracy-a-function-of-item-complexity}{%
-\subsubsection{Accuracy a function of item complexity}\label{accuracy-a-function-of-item-complexity}}
+\hypertarget{accuracy-a-function-of-item-length}{%
+\subsubsection{Accuracy a function of item length}\label{accuracy-a-function-of-item-length}}
 
@@
 The pattern is less marked when other two scores are used, which are less sensitive to errors.
+
+\begin{table}[tbp]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:tablength}NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
+
+\begin{tabular}{lll}
+\toprule
+Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
+\midrule
+48 (21) & 60 (16) & 40 (16)\\
+80 (24) & 93 (10) & 7 (10)\\
+78 (21) & 94 (7) & 6 (7)\\
+66 (31) & 83 (19) & 17 (19)\\
+\bottomrule
+\end{tabular}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+
+\hypertarget{patterns-of-errors}{%
+\subsubsection{Patterns of errors}\label{patterns-of-errors}}
+
+\begin{itemize}
+\tightlist
+\item
+  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+\end{itemize}
+
+We investigated patterns of deletion and substitution. Deletions were relatively rare, with only 28 vowels deleted, and 8 consonants.
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (54 oral target vowels produced as nasal vowels, 58 nasal target vowels produced as oral vowels).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (202 oral vowels produced with a different quality; 23 nasal vowels produced with a different quality).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbgh) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 251) than vice versa (2).
+Simple consonants were mispronounced as other simple consonants quite frequently (135 simple consonants mispronounced as other simple ones, compared to 58 complex ones).
+
+\textbf{probably turn this into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set}
+
+Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 96\%.
+
+** this is higher than any previous report! can we be sure of it?**
+
+\hypertarget{factor-structuring-individual-variation}{%
+\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/Fig2-scores by age-1.pdf}
+\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual participants as a function of age and sex (blue = boys, pink = girls).}
+\end{figure}
+
+Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,535.50) = 0.33, p = 0.11).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.33) = 20.09, p = 0.75), birth order (2,595.77) = -0.13, p = 0.55) or maternal education (2,038.91) = 0.11, p = 0.60)
+
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
+
+\begin{itemize}
+\item
+  What is the overall repetition accuracy (whole word, phoneme based, distance)?
+\item
+  How does this change as a function of item complexity (number of syllables, sound complexity)?
+\end{itemize}
 
@@
 
-\hypertarget{patterns-of-errors}{%
-\subsubsection{Patterns of errors}\label{patterns-of-errors}}
-
 \begin{itemize}
@@
 \item
-  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
+  How frequent are errors that result in real words? Is that a function of item complexity?
+\item
+  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
 \end{itemize}
-
-\hypertarget{factor-structuring-individual-variation}{%
-\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}
-
-explainable by child age, sex, birth order, monolingual status, and/or parental education?
-3. Children's accuracy increases with child age.
-4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
-5.
-As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
-6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{3}
-\tightlist
+\setcounter{enumi}{2}
+\item
+  Children's accuracy increases with child age.
+\item
+  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
+\item
+  As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
+\item
+  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
 \item
@@
 \end{enumerate}
-
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
-
-\begin{itemize}
-\tightlist
-\item
-  What is the overall repetition accuracy (whole word, phoneme based, distance)?
-\item
-  How does this change as a function of item complexity (number of syllables, sound complexity)?
-\item
-  How frequent are errors that result in real words? Is that a function of item complexity?
-\item
-  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
-\end{itemize}
 

### 2020-05-29 manuscript.Rmd
streamlining

@@
 library(gdata)
+library(ggplot2)
+
 correspondances=matrix( #list correspondances always by pairs, orthography then phonology
@@
 
-- add table r/d by stim length & age
 - double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
-- discussion
+- add frequency of sounds in stimuli in segments.xlsx from https://phoible.org/parameters
+- then add that in stimuli to get minimum & average frequencies per stimulus
+- then add analysis of NWR as a function of frequency
+- draft discussion
 
@@
 
-- look through
-- consent, "correct", "accuracy"
-- help with table 1 (phonological representation but also formatting)
+- look through for **mc**
+- consent, "correct", "accuracy", "performance"
+- help with table 1 (formatting)
 
@@
 The child should immediately try to say it back without changing anything.
-Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible productiong patterns (to produce the item accurately even though it had never been pronounced before).
+Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item accurately even though it had never been pronounced before).
 
-NWR has been used to seek answers to a variety of theoretical questions, including what are the links between phonology, working memory, and the lexicon [@bowey2001nonword], as well as for applications, notably using nwr as a diagnostic for language delays and disorders [@estes2007differences].
-Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe [@meir2016disentangling,@is08042009language].
+NWR has been used to seek answers to a variety of theoretical questions, including what  the links between phonology, working memory, and the lexicon are [@bowey2001nonword].
+It is also frequently used as for applications, notably  as a diagnostic for language delays and disorders [@estes2007differences].
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments,  particularly in Europe [@meir2016disentangling;@is08042009language].
 
 In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
-In our implementation of NWR, we  made sure that some of the items contained typologically rare and/or  challenging sounds.
-MISSING: WHY DO WE DO THIS?
-WHAT IS THE POINT?
+The present report uses these unique data to contribute to several lines of research.
+First, we  made sure that some of the non-word items contained typologically rare and/or  challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall accuracy of repetition and patterns of mispronunciation.
+Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
+Third, we contribute to the basic NWR literature by contributing an additional data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
 
-### Intro to the language (@mc) - please feel free to throw away anything that is not useful!
+### Intro to the language (**mc**) - please feel free to throw away anything that is not useful!
 
@@
 
-### Intro to the people (@mc) - please feel free to throw away anything that is not useful!
+### Intro to the people (**mc**) - please feel free to throw away anything that is not useful!
 
@@
 
-We sought to address the following questions:
+After some preliminary analyses to set the stage, we address the following questions:
 
-- What is the overall repetition score, according to a variety of relevant metrics (whole word, phoneme based, distance)?
-- How does score change as a function of item complexity (number of syllablex, sound complexity)?
-- How frequent are errors that result in real words? Is that a function of item complexity? 
-- What are some common patterns of phonological errors?
-- Is variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
+- Does the frequency of sounds across languages predict NWR? Are rarer sounds more often substituted by commoner sounds?
+- How does score change as a function of item length in number of syllables?
+- Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
 
 In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects.
-For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small [data from @cristia2020infant].
-**add table with r/ds from previous work**)
-Moreover, by virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies.
+However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects given their variability across studies.
+To illustrate this, we portray studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables) in Figure 1.
+Notice that the effect of stimulus length is minuscule among Italian children [@piazzalunga2019articulatory], but considerable among Tsimane' children  [@cristia2020infant], where a drop of 40 percentage points is observed at all ages.
+A similar difference in NWR scores for short versus long items was observed among Arabic children [@jabere2018xperiment].
+Even the effect of age is unstable in this sample.
+Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
 Therefore, all analyses here are descriptive and should be considered exploratory.
 
-```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "Proportion of non-words correctly repeated as a function of age (in years), study (first author and year regardless of the number of authors), and, when available, shorter versus longer non-words."}
-read.csv("prevlit.csv")->pl
-#duplicate Piazzalunga logging short and long items separately
-pl$Score[pl$Study=="Piazzalunga 2018"]<-pl$Short[pl$Study=="Piazzalunga 2018"]
-pl$Study<-as.character(pl$Study)
-pl$Study[pl$Study=="Piazzalunga 2018"]<-"Piazzalunga 2018 1-2 syll"
-dim(pl)[1]->beg
-pl=rbind(pl,pl[pl$Study=="Piazzalunga 2018 1-2 syll",])
-pl[(beg+1):dim(pl)[1],"Score"]<-pl[(beg+1):dim(pl)[1],"Long"]
-pl[(beg+1):dim(pl)[1],"Study"]<-"Piazzalunga 2018 3-4 syll"
-pl$Study=paste(pl$Language,pl$Study)
-library(lattice)
-xyplot(Score~Age,groups=Study,data=pl,type=c('l','p'),auto.key = T ,pch=20)
+```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies.
+Arabic data  from Jaber et al.
+(2018); Italian data from Piazzalunga et al.
+(2019); Tsimane' from Cristia et al.
+(2020); Yélî data from the present study."}
+
+pl<- read.delim("~/Documents/gitrepos/nwryd19/pl-With_yd.tsv")
+palette = c( 
+             "Arabic 1-2 syll" = "orange",
+             "Yélî 3-4 syll"   ="navyblue",
+             "Yélî 1-2 syll"  ="lightblue",
+             "Italian 3-4 syll"   ="black",
+             "Italian 1-2 syll"  ="gray",
+             "Arabic 3-4 syll"     = "orange", 
+             "Tsimane' 1-2 syll" = "green",
+              "Tsimane' 3-4 syll" = "darkgreen")
+
+myplot=ggplot(data = pl, aes(x = Age, y = Score, color = Language))+
+  geom_point(aes(size = n, shape = Length))+
+  geom_line(aes(linetype=Length)) + 
+  theme(text = element_text(size = 16), legend.position="none", axis.text.x=element_text(colour=c("black", "black"))) +
+  scale_color_manual(values = palette)+
+  theme_bw()+
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
+  scale_x_continuous(name = "Age (years) ", breaks = c(4,8,12))+
+
+  scale_y_continuous(name = "NWR Score", breaks = seq(0, 100, 10))+
+  scale_size_continuous(range = c(2, 6))
+
+myplot
 
@@
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described [CITE @mc please fill in], there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described [CITE **mc** please fill in], there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
 These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset **XXX @mc please fill in** and an Olympus **XXX @mc please fill in** from the written form presented together with the same item orally (by the second author).
+Together with the 2018 selection, they were recorded using a headset **XXX **mc** please fill in** and an Olympus **XXX **mc** please fill in** from the written form presented together with the same item orally (by the second author).
 The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
@@
 
-**@mc can you please add the IPA in // for all of the items in the excel then save as tsv?**
+****mc** can you please add the IPA in // for all of the items in the excel then save as tsv?**
 The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
 
-```{r tab2-stims}
+```{r tab2-stims, eval=F}
 #add orthographic representations targets & other stims chars
-read.delim("stimuli.tsv", encoding="UTF-8")->stims
+read.delim("stimuli.txt", encoding="UTF-8")->stims
 stims=stims[order(stims[,1]),-3]
@@
 
+#now that I added phonological representations, even this basic table chokes!
 apa_table(stims,"NWR stimuli in orthographic and phonological representations.")
@@
 
-The child was donned a headset (**xx @mc please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
+The child was donned a headset (**xx **mc** please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
@@
 
-### Overall repetition accuracy
-
 ```{r overall}
@@
 
-### Accuracy a function of item length
+### NWR as a function of cross-linguistic phone frequency
 
-```{r bylength}
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->by_length
-aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->pc
-aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->nld
-by_length=cbind(by_length,pc$x,nld$x)
-colnames(by_length)<-c("ID","Age","Length","Word","Phoneme","NLD")
+```{r}
 
 ```
-
-We then inspected whether accuracy varied as a function of word length.
-Results are shown on table XX.
-This table shows that monosyllables accuracy was much lower than other lengths.
-This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
-
-Setting monosyllables aside, we observe the typical pattern of decreased accuracy for longer items, although this is particularly salient for the whole word scoring.
-This is the most commonly reported type of score, but it is also the least forgiving.
-The pattern is less marked when other two scores are used, which are less sensitive to errors.
-
-```{r tablength}
-mytab=NULL
-for(myvar in c("Word","Phoneme","NLD")){
-  mytab=cbind(mytab,
-              paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
-              " (",round(aggregate(by_length[,myvar],by=list(by_length$Length),sd)$x),")"))
-}
-rownames(mytab)<-levels(by_length$Length)
-colnames(mytab)<-c("Word","Phoneme","NLD")
-apa_table(mytab,caption="NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
-```
-
-### Patterns of errors
-- How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
 
@@
 simplexCs=c("m","n","l","w","y","v","d","g","p","t","k","f","h","ch")
-complexCs=c("dp","tp","kp","km","kn","mb","gh")
+complexCs=c("dp","tp","kp","km","kn","mb","gh","lv")
 
@@
 ** this is higher than any previous report! can we be sure of it?**
+
+### Accuracy a function of item length
+
+```{r bylength}
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->by_length
+aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->pc
+aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->nld
+by_length=cbind(by_length,pc$x,nld$x)
+colnames(by_length)<-c("ID","Age","Length","Word","Phoneme","NLD")
+
+```
+
+Next, we inspected whether accuracy varied as a function of word length.
+Results are shown on table XX.
+This table shows that monosyllables accuracy was much lower than other lengths.
+This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
+
+Setting monosyllables aside, we observe the typical pattern of decreased accuracy for longer items, although this is particularly salient for the whole word scoring.
+This is the most commonly reported type of score, but it is also the least forgiving.
+The pattern is less marked when other two scores are used, which are less sensitive to errors.
+
+```{r tablength}
+mytab=NULL
+for(myvar in c("Word","Phoneme","NLD")){
+  mytab=cbind(mytab,
+              paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
+              " (",round(aggregate(by_length[,myvar],by=list(by_length$Length),sd)$x),")"))
+}
+rownames(mytab)<-levels(by_length$Length)
+colnames(mytab)<-c("Word","Phoneme","NLD")
+apa_table(mytab,caption="NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
+```
 

### 2020-05-29 manuscript.tex
streamlining

@@
 \item
-  add table r/d by stim length \& age
-\item
   double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
 \item
-  discussion
+  add frequency of sounds in stimuli in segments.xlsx from \url{https://phoible.org/parameters}
+\item
+  then add that in stimuli to get minimum \& average frequencies per stimulus
+\item
+  then add analysis of NWR as a function of frequency
+\item
+  draft discussion
+\end{itemize}
+
+\hypertarget{todo-middy}{%
+\subsection{TODO Middy}\label{todo-middy}}
+
+\begin{itemize}
+\tightlist
+\item
+  look through for \textbf{mc}
+\item
+  consent, \enquote{correct}, \enquote{accuracy}, \enquote{performance}
+\item
+  help with table 1 (formatting)
 \end{itemize}
@@
 \item
-  ask Middy about consent, correct, accuracy
-\item
-  ask Middy to look through
-\item
   probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set
@@
   maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
+\item
+  maybe add analyses of score as a function of frequency (phones, diphones) in MC's corpus? (mc suggested, but I don't think we need it\ldots{})
 \end{itemize}
@@
 The child should immediately try to say it back without changing anything.
-Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible productiong patterns (to produce the item accurately even though it had never been pronounced before).
+Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item accurately even though it had never been pronounced before).
 
-NWR has been used to seek answers to a variety of theoretical questions, including what are the links between phonology, working memory, and the lexicon (Bowey, 2001), as well as for applications, notably using nwr as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
-Since non-words can be generated in many languages, it has been used acros a wide array of languages, particularly in Europe (Meir, Walters, \& Armon-Lotem, 2016,@is08042009language).
+NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001).
+It is also frequently used as for applications, notably as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
 
 In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
-In our implementation of NWR, we made sure that some of the items contained typologically rare and/or challenging sounds.
-MISSING: WHY DO WE DO THIS?
-WHAT IS THE POINT?
+The present report uses these unique data to contribute to several lines of research.
+First, we made sure that some of the non-word items contained typologically rare and/or challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall accuracy of repetition and patterns of mispronunciation.
+Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
+Third, we contribute to the basic NWR literature by contributing an additional data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
 
 \hypertarget{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}{%
-\subsubsection{\texorpdfstring{Intro to the language (({\textbf{???}})) - please feel free to throw away anything that is not useful!}{Intro to the language ((???)) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
+\subsubsection{\texorpdfstring{Intro to the language (\textbf{mc}) - please feel free to throw away anything that is not useful!}{Intro to the language (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
 
@@
 \hypertarget{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}{%
-\subsubsection{\texorpdfstring{Intro to the people (({\textbf{???}})) - please feel free to throw away anything that is not useful!}{Intro to the people ((???)) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
+\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) - please feel free to throw away anything that is not useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
 
@@
 
-We sought to address the following questions:
+After some preliminary analyses to set the stage, we address the following questions:
 
@@
 \item
-  What is the overall repetition score, according to a variety of relevant metrics (whole word, phoneme based, distance)?
+  Does the frequency of sounds across languages predict NWR? Are rarer sounds more often substituted by commoner sounds?
 \item
-  How does score change as a function of item complexity (number of syllablex, sound complexity)?
+  How does score change as a function of item length in number of syllables?
 \item
-  How frequent are errors that result in real words? Is that a function of item complexity?
-\item
-  What are some common patterns of phonological errors?
-\item
-  Is variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
+  Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
 \end{itemize}
@@
 In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects.
-For instance, an effect of stimulus length and age has been reported as significant sometimes, but Tables XX and XX show that even these effects can be quite small (data from Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
-\textbf{add table with r/ds from previous work})
-Moreover, by virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies.
+However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects given their variability across studies.
+To illustrate this, we portray studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables) in Figure 1.
+Notice that the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Tsimane' children (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020), where a drop of 40 percentage points is observed at all ages.
+A similar difference in NWR scores for short versus long items was observed among Arabic children (Jaber-Awida, 2018).
+Even the effect of age is unstable in this sample.
+Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
 Therefore, all analyses here are descriptive and should be considered exploratory.
@@
 \includegraphics{manuscript_files/figure-latex/prevlit-fig-1.pdf}
-\caption{\label{fig:prevlit-fig}Proportion of non-words correctly repeated as a function of age (in years), study (first author and year regardless of the number of authors), and, when available, shorter versus longer non-words.}
+\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years) and item length for comparable studies.
+Arabic data from Jaber et al.
+(2018); Italian data from Piazzalunga et al.
+(2019); Tsimane' from Cristia et al.
+(2020); Yélî data from the present study.}
 \end{figure}
@@
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described (CITE {\textbf{???}} please fill in), there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described {[}CITE \textbf{mc} please fill in{]}, there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
 These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset \textbf{XXX ({\textbf{???}}) please fill in} and an Olympus \textbf{XXX ({\textbf{???}}) please fill in} from the written form presented together with the same item orally (by the second author).
+Together with the 2018 selection, they were recorded using a headset \textbf{XXX }mc** please fill in** and an Olympus \textbf{XXX }mc** please fill in** from the written form presented together with the same item orally (by the second author).
 The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
@@
 
-\textbf{({\textbf{???}}) can you please add the IPA in // for all of the items in the excel then save as tsv?}
+****mc** can you please add the IPA in // for all of the items in the excel then save as tsv?**
 The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
-
-\begin{verbatim}
-##        Type Orthographic Phonological Type Orthographic Phonological Type
-## 28 practice     nopimade           NA    1         dp:a           NA    2
-## 32 practice        poni            NA    1          dpa           NA    2
-## 43 practice          wî            NA    1          dpâ           NA    2
-## 1      <NA>         <NA>           NA    1          dpê           NA    2
-## 2      <NA>         <NA>           NA    1         dpéé           NA    2
-## 3      <NA>         <NA>           NA    1          dpi           NA    2
-## 4      <NA>         <NA>           NA    1          dpu           NA    2
-## 5      <NA>         <NA>           NA    1        gh:ââ           NA    2
-## 6      <NA>         <NA>           NA    1         ghuu           NA <NA>
-## 7      <NA>         <NA>           NA    1        kp:ââ           NA <NA>
-## 8      <NA>         <NA>           NA    1          kpu           NA <NA>
-## 9      <NA>         <NA>           NA    1         lv:ê           NA <NA>
-## 10     <NA>         <NA>           NA    1          lva           NA <NA>
-## 11     <NA>         <NA>           NA    1          lvi           NA <NA>
-## 12     <NA>         <NA>           NA    1         t:êê           NA <NA>
-## 13     <NA>         <NA>           NA    1         tp:a           NA <NA>
-## 14     <NA>         <NA>           NA    1          tpâ           NA <NA>
-## 15     <NA>         <NA>           NA    1          tpê           NA <NA>
-##    Orthographic Phonological Type Orthographic Phonological Type Orthographic
-## 28         kamo           NA    3       dimope           NA    4     diponate
-## 32         kani           NA    3       diyeto           NA    4     nomiwake
-## 43         kipo           NA    3       meyadi           NA    4     todiwuma
-## 1          noki           NA    3       mituye           NA    4     wadikeno
-## 2          nomi           NA    3       nademo           NA <NA>         <NA>
-## 3          piwa           NA    3       nayeki           NA <NA>         <NA>
-## 4          towi           NA    3       nuyedi           NA <NA>         <NA>
-## 5          tupa           NA    3       pedumi           NA <NA>         <NA>
-## 6          <NA>           NA    3       tiwune           NA <NA>         <NA>
-## 7          <NA>           NA    3       tumowe           NA <NA>         <NA>
-## 8          <NA>           NA    3       widone           NA <NA>         <NA>
-## 9          <NA>           NA    3       wumipo           NA <NA>         <NA>
-## 10         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
-## 11         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
-## 12         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
-## 13         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
-## 14         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
-## 15         <NA>           NA <NA>         <NA>           NA <NA>         <NA>
-##    Phonological
-## 28           NA
-## 32           NA
-## 43           NA
-## 1            NA
-## 2            NA
-## 3            NA
-## 4            NA
-## 5            NA
-## 6            NA
-## 7            NA
-## 8            NA
-## 9            NA
-## 10           NA
-## 11           NA
-## 12           NA
-## 13           NA
-## 14           NA
-## 15           NA
-\end{verbatim}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:tab2-stims}NWR stimuli in orthographic and phonological representations.}
-
-\begin{tabular}{ll}
-\toprule
-Orthographic & \multicolumn{1}{c}{Phonological}\\
-\midrule
-dp:a & NA\\
-dpa & NA\\
-dpâ & NA\\
-dpê & NA\\
-dpéé & NA\\
-dpi & NA\\
-dpu & NA\\
-gh:ââ & NA\\
-ghuu & NA\\
-kp:ââ & NA\\
-kpu & NA\\
-lv:ê & NA\\
-lva & NA\\
-lvi & NA\\
-t:êê & NA\\
-tp:a & NA\\
-tpâ & NA\\
-tpê & NA\\
-kamo & NA\\
-kani & NA\\
-kipo & NA\\
-noki & NA\\
-nomi & NA\\
-piwa & NA\\
-towi & NA\\
-tupa & NA\\
-dimope & NA\\
-diyeto & NA\\
-meyadi & NA\\
-mituye & NA\\
-nademo & NA\\
-nayeki & NA\\
-nuyedi & NA\\
-pedumi & NA\\
-tiwune & NA\\
-tumowe & NA\\
-widone & NA\\
-wumipo & NA\\
-diponate & NA\\
-nomiwake & NA\\
-todiwuma & NA\\
-wadikeno & NA\\
-nopimade & NA\\
-poni & NA\\
-wî & NA\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
 
@@
 
-The child was donned a headset (\textbf{xx ({\textbf{???}}) please fill in} for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
+The child was donned a headset (\textbf{xx }mc** please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
@@
 
-\hypertarget{overall-repetition-accuracy}{%
-\subsubsection{Overall repetition accuracy}\label{overall-repetition-accuracy}}
-
 Taking into account only the first attempts, we averaged attempts by each of the 24 children who had data for first attempts; their ages ranged from 3.92 to 10.20 (M = 6.50, SD = 1.50).
@@
 
+\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
+\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}
+
+We investigated patterns of deletion and substitution. Deletions were relatively rare, with only 28 vowels deleted, and 8 consonants.
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (54 oral target vowels produced as nasal vowels, 58 nasal target vowels produced as oral vowels).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (202 oral vowels produced with a different quality; 23 nasal vowels produced with a different quality).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 267) than vice versa (2).
+Simple consonants were mispronounced as other simple consonants quite frequently (135 simple consonants mispronounced as other simple ones, compared to 62 complex ones).
+
+\textbf{probably turn this into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set}
+
+Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 96\%.
+
+** this is higher than any previous report! can we be sure of it?**
+
 \hypertarget{accuracy-a-function-of-item-length}{%
@@
 
-We then inspected whether accuracy varied as a function of word length.
+Next, we inspected whether accuracy varied as a function of word length.
 Results are shown on table XX.
@@
 \end{table}
-
-\hypertarget{patterns-of-errors}{%
-\subsubsection{Patterns of errors}\label{patterns-of-errors}}
-
-\begin{itemize}
-\tightlist
-\item
-  How frequent are errors that result in real words? Is that a function of item complexity? What are some common patterns of phonological errors?
-\end{itemize}
-
-We investigated patterns of deletion and substitution. Deletions were relatively rare, with only 28 vowels deleted, and 8 consonants.
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (54 oral target vowels produced as nasal vowels, 58 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (202 oral vowels produced with a different quality; 23 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbgh) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 251) than vice versa (2).
-Simple consonants were mispronounced as other simple consonants quite frequently (135 simple consonants mispronounced as other simple ones, compared to 58 complex ones).
-
-\textbf{probably turn this into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set}
-
-Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 96\%.
-
-** this is higher than any previous report! can we be sure of it?**
 
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-is08042009language}{}%
+Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
 \leavevmode\hypertarget{ref-balladares2016socio}{}%
@@
 
-\leavevmode\hypertarget{ref-is08042009language}{}%
-IS0804, C. A. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
 \leavevmode\hypertarget{ref-jabere2018xperiment}{}%

### 2020-06-09 manuscript.Rmd
found

@@
 
-****mc** can you please add the IPA in // for all of the items in the excel then save as tsv?**
 The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
@@
 Moreover, there were also some logistic constraints in terms of the space availability.
-As a result, the places where elicitation took place varied across the hamlets.
+As a result, the places where elicitation happened varied across the hamlets.
 

### 2020-06-09 manuscript.Rmd
added segment frequencies

@@
 - double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
-- add frequency of sounds in stimuli in segments.xlsx from https://phoible.org/parameters
-- then add that in stimuli to get minimum & average frequencies per stimulus
+- finish integration in stimuli in wrangling.R to get minimum & average frequencies per stimulus
 - then add analysis of NWR as a function of frequency
 - draft discussion
+- draft abstract
 
@@
 - look through for **mc**
+- double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
+- double check phonological implementation of stimuli in stimuli.xlsx, and if needed re-save as txt file to update the version read in this paper
 - consent, "correct", "accuracy", "performance"
-- help with table 1 (formatting)
+- help with table 1 (formatting & making it appear, because it's broken by the phonological representations)
 

### 2020-06-09 manuscript.tex
added segment frequencies

@@
 \item
-  add frequency of sounds in stimuli in segments.xlsx from \url{https://phoible.org/parameters}
-\item
-  then add that in stimuli to get minimum \& average frequencies per stimulus
+  finish integration in stimuli in wrangling.R to get minimum \& average frequencies per stimulus
 \item
@@
   draft discussion
+\item
+  draft abstract
 \end{itemize}
@@
 \item
+  double check my frequency entries in segments.xlsx using the search function in \url{https://phoible.org/parameters}; take a look at \url{http://phoible.github.io/conventions/} in case you see something about double articulation (I didn't find tp or lBj)
+\item
+  double check phonological implementation of stimuli in stimuli.xlsx, and if needed re-save as txt file to update the version read in this paper
+\item
   consent, \enquote{correct}, \enquote{accuracy}, \enquote{performance}
 \item
-  help with table 1 (formatting)
+  help with table 1 (formatting \& making it appear, because it's broken by the phonological representations)
 \end{itemize}

### 2020-06-09 wrangling.R

@@ -1,3 +1,13 @@
+read.csv("NWR-demo.csv",header=T)->demo
+#colnames(demo)
+
+#if family is missing, attribute each child to a different one
+demo$familyID=as.character(demo$familyID)
+demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])
+
+subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
+#if age.exact is missing, use reported age
+inc$age.exact[is.na(inc$age.exact)]<-inc$Age[is.na(inc$age.exact)]
 read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
 trnsc$tokid=paste(trnsc$item,trnsc$token)
 
@@ -26,7 +36,10 @@ trnsc$cor.online[(npairs+1):dim(trnsc)[1]]<-NA
 merge(trnsc,demo,by.x="id",by.y="ID",all.x=T)->trnsc
 
 #add orthographic representations targets & other stims chars
-read.delim("stimuli.tsv", encoding="UTF-8")->stims
+read.delim("stimuli.txt", encoding="UTF-8")->stims
+
+
+
 merge(trnsc,stims,by="target",all=T)->trnsc
 
 

### 2020-06-11 manuscript.Rmd
finished integration frequency but bugs remain with some segs having freq na

--- 2020-06-09 manuscript.Rmd
@@
 
+<!-- REMEMBER!!!! -->
+<!-- phonetic characters: -->
+<!-- - save excel as utf with save as -->
+<!-- - do $ iconv -f utf-16le -t utf-8 stimuli.txt > stimconv.txt -->
+<!-- - do $ iconv -f utf-16le -t utf-8 segments.txt > segconv.txt -->
+
 ## TODO
@@
 - double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
-- add frequency of sounds in stimuli in segments.xlsx from https://phoible.org/parameters
-- then add that in stimuli to get minimum & average frequencies per stimulus
 - then add analysis of NWR as a function of frequency
 - draft discussion
+- draft abstract
 
@@
 - look through for **mc**
+- double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
+- double check phonological implementation of stimuli in stimuli.xlsx 
+- for both of the above, if you make any changes, they need to be re-saved as UTF16 txt file, and then converted (see remember above)
+- double check also stim_seg_freq -- why are some items still with NA freq?
 - consent, "correct", "accuracy", "performance"
-- help with table 1 (formatting)
+- help with table 1 (formatting & making it appear, because it's broken by the phonological representations)
 

### 2020-06-11 wrangling.R

@@ -36,9 +36,15 @@ trnsc$cor.online[(npairs+1):dim(trnsc)[1]]<-NA
 merge(trnsc,demo,by.x="id",by.y="ID",all.x=T)->trnsc
 
 #add orthographic representations targets & other stims chars
-read.delim("stimuli.txt", encoding="UTF-8")->stims
+read.delim("stimconv.txt", encoding="UTF-8")->stims
+read.delim("segconv.txt", encoding="UTF-8")->segments
 
-
+stim_seg_freq=matrix(NA,nrow=dim(stims)[1],ncol=8)
+for(i in 1:dim(stim_seg_freq)[1]) for(j in 1:dim(stim_seg_freq)[2]) {
+  thisfreq=segments$pc[as.character(segments$phono)==as.character(stims[i,paste0("phono",j)])] 
+  stim_seg_freq[i,j]<-ifelse(sum(!is.na(thisfreq))==1,thisfreq,NA)
+  }
+stims$avg_fr=apply(stim_seg_freq,1,mean,na.rm=T)
 
 merge(trnsc,stims,by="target",all=T)->trnsc
 

### 2020-06-11 manuscript.tex
finished integration frequency but bugs remain with some segs having freq na

@@
 \item
-  finish integration in stimuli in wrangling.R to get minimum \& average frequencies per stimulus
-\item
   then add analysis of NWR as a function of frequency
@@
 \item
-  double check phonological implementation of stimuli in stimuli.xlsx, and if needed re-save as txt file to update the version read in this paper
+  double check phonological implementation of stimuli in stimuli.xlsx
+\item
+  for both of the above, if you make any changes, they need to be re-saved as UTF16 txt file, and then converted (see remember above)
+\item
+  double check also stim\_seg\_freq -- why are some items still with NA freq?
 \item
@@
 We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
-We excluded 5 children who did not have data for one of these two types.
+We excluded 6 children who did not have data for one of these two types.
 As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
-Surprisingly, subsequent repetitions (M = 42, SD = 28)
-were on average less accurate than first ones (M = 64, SD = 15), t(36) = 5.32, p = 0.
+Surprisingly, subsequent repetitions (M = 41, SD = 27)
+were on average less accurate than first ones (M = 64, SD = 15), t(35) = 5.88, p = 0.
 Given the uncertainty in whether previous work used only the first or all repetitions, and since behavior degraded and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
@@
 
-The overall NWR score was M = 60\% (SD = 14\%).
-Scores based on phonemes are even higher M = 76\% (SD = 9\%).
-The phoneme-based normalized Levenshtein distance was M = 24\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
+The overall NWR score was M = 61\% (SD = 15\%).
+Scores based on phonemes are even higher M = 77\% (SD = 9\%).
+The phoneme-based normalized Levenshtein distance was M = 23\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
 Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
@@
 
-We investigated patterns of deletion and substitution. Deletions were relatively rare, with only 28 vowels deleted, and 8 consonants.
+We investigated patterns of deletion and substitution. Deletions were relatively rare, with only 29 vowels deleted, and 9 consonants.
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (54 oral target vowels produced as nasal vowels, 58 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (202 oral vowels produced with a different quality; 23 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 267) than vice versa (2).
-Simple consonants were mispronounced as other simple consonants quite frequently (135 simple consonants mispronounced as other simple ones, compared to 62 complex ones).
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels, 58 nasal target vowels produced as oral vowels).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (197 oral vowels produced with a different quality; 23 nasal vowels produced with a different quality).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 266) than vice versa (2).
+Simple consonants were mispronounced as other simple consonants quite frequently (128 simple consonants mispronounced as other simple ones, compared to 62 complex ones).
 
@@
 
-Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 96\%.
+Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 97\%.
 
@@
 \midrule
-48 (21) & 60 (16) & 40 (16)\\
-80 (24) & 93 (10) & 7 (10)\\
-78 (21) & 94 (7) & 6 (7)\\
-66 (31) & 83 (19) & 17 (19)\\
+46 (22) & 58 (17) & 42 (17)\\
+80 (23) & 94 (7) & 6 (7)\\
+78 (20) & 94 (6) & 6 (6)\\
+74 (30) & 93 (11) & 7 (11)\\
 \bottomrule
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,535.50) = 0.33, p = 0.11).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.33) = 20.09, p = 0.75), birth order (2,595.77) = -0.13, p = 0.55) or maternal education (2,038.91) = 0.11, p = 0.60)
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,519.81) = 0.34, p = 0.10).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.18) = 21.34, p = 0.86), birth order (2,630.74) = -0.14, p = 0.50) or maternal education (2,091.92) = 0.09, p = 0.67)
 

### 2020-06-15 manuscript.Rmd
fixed little bugs, added freq ana and abstract

--- 2020-06-09 manuscript.Rmd
@@
 abstract: >
-  ADD
+   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
+Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+This study contributes to three lines of research.
+First, we document that non-word items containing typologically rare sounds are repeated accurately less often that non-words containing more common sounds.
+Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
+Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
+Together, these data provide a unique view of online phonological processing in a seldom-studied language, and contribute to both typological and language acquisition research.
 
@@
 library(ggplot2)
+library(lme4)
 
@@
 
-## TODO
-
-- double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
-- add frequency of sounds in stimuli in segments.xlsx from https://phoible.org/parameters
-- then add that in stimuli to get minimum & average frequencies per stimulus
-- then add analysis of NWR as a function of frequency
-- draft discussion
+<!-- REMEMBER!!!! -->
+<!-- phonetic characters: -->
+<!-- - save excel as utf with save as -->
+<!-- - do $ iconv -f utf-16le -t utf-8 stimuli.txt > stimconv.txt -->
+<!-- - do $ iconv -f utf-16le -t utf-8 segments.txt > segconv.txt -->
 
@@
 
-- look through for **mc**
+- look through this manuscript for bits of text to be written by **mc**
+- help with table 1 (formatting & making it appear, because it's broken by the phonological representations)
+- do you think Figure 2 should be done with phonetic symbols? If yes, could you try to implement that?
+- double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
+I looked again at my local files (still haven't checked the labs' ones) and I'm surprised I have maternal education for more children than I have birth order for...
+If you correct this, to generate the csv I also round the age, assign age.rounded to reported age (in years) when DOB is missing, remove first names, DOB, DOTest and age.precise before pushing to github
+- double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
+- double check phonological implementation of stimuli in stimuli.xlsx 
+- for both of the above, if you make any changes, they need to be re-saved as UTF16 txt file, and then converted (see remember above)
+- double check also stim_seg_freq -- why are some items still with NA freq?
 - consent, "correct", "accuracy", "performance"
-- help with table 1 (formatting)
+- draft discussion
 
@@
 
-NWR has been used to seek answers to a variety of theoretical questions, including what  the links between phonology, working memory, and the lexicon are [@bowey2001nonword].
+NWR has been used to seek answers to a variety of theoretical questions, including what  the links between phonology, working memory, and the lexicon are [@bowey2001nonword], and to assess the extent to which apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity].
 It is also frequently used as for applications, notably  as a diagnostic for language delays and disorders [@estes2007differences].
@@
 
+Additionally, we looked up each phoneme in each target word in the Phoible database [@phoible].
+We entered the number of languages in which each phoneme was found, as well as what percentage of languages in the database that represented.
+We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
+* **mc** do you agree with this decision or shall we do the alternative?*
+
 Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
@@
 # fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
-# fulldemo$age.exact=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
+# fulldemo$age.rounded=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
 # fulldemo$Age=as.numeric(as.character(fulldemo$Age))
-# plot(fulldemo$Age~fulldemo$age.exact)
+# plot(fulldemo$Age~fulldemo$age.rounded)
 # write.csv(fulldemo,"nwr-demo.csv")
@@
 subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
-#if age.exact is missing, use reported age
-inc$age.exact[is.na(inc$age.exact)]<-inc$Age[is.na(inc$age.exact)]
+#if age.rounded is missing, use reported age
+inc$age.rounded[is.na(inc$age.rounded)]<-inc$Age[is.na(inc$age.rounded)]
 #summary(inc)
@@
 In addition, `r sum(demo$Age >12 | demo$included=="too old",na.rm=T)` teenagers were tested to put younger children at ease; their data is not included in analyses below.
-The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.exact),2)` years (range `r round(min(inc$age.exact),2)`-`r round(max(inc$age.exact),2)` years).
+The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years).
 There were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and , `r sum(is.na(inc$monolingual))` for whom this information was missing.
@@
 
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$attempt),mean)->pwcm_att
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$attempt),mean)->pwcm_att
 colnames(pwcm_att)<-c("ID","Age","Attempt","Word")
@@
 
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.exact,trnsc$birthOrder,trnsc$matEd),mean)->overall
-aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.exact,trnsc$birthOrder,trnsc$matEd),mean)->pc
-aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.exact,trnsc$birthOrder,trnsc$matEd),mean)->nld
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->overall
+aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->pc
+aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->nld
 overall=cbind(overall,pc$x,nld$x)
@@
 
-```{r}
+In this analysis, we were interested in variation in performance as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
+The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID & target ID as random variables.
 
+```{r xling-freq-ana}
+xl_fr=glmer(correct~avg_fr+age.rounded+(1|file) + (1|target),data=trnsc,family="binomial")
+mycoeff=round(summary(xl_fr)$coefficients,3)
+```
+
+We could include 885 observations (**why -- shouldn't it be 1k??**), from 41 children producing in any given trial one of 41 potential target words.
+The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr","Estimate"]`, SE ß = `r mycoeff["avg_fr","Std.
+Error"]`, p = `r mycoeff["avg_fr","Pr(>|z|)"]`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+
+```{r Fig2-xling_freq, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
+aggregate(trnsc$correct*100,by=list(trnsc$target,trnsc$avg_fr),mean)->xling
+colnames(xling)<-c("Target","Avg_fr","Word")
+plot(xling$Word~xling$Avg_fr,lty=3,ylab="NWR scores (% words)",xlab="Average phone cross-ling frequency",type="n")
+text(xling$Word~xling$Avg_fr,labels=xling$Target,cex=.5)
+abline(lm(xling$Word~xling$Avg_fr))
 ```
@@
 
-We investigated patterns of deletion and substitution.
+Next, we investigated patterns of deletion and substitution.
 Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted, and `r sum(del_tab[c(simplexCs,complexCs)])` consonants.
@@
 ```{r bylength}
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->by_length
-aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->pc
-aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.exact,trnsc$type),mean)->nld
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->by_length
+aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->pc
+aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->nld
 by_length=cbind(by_length,pc$x,nld$x)
@@
 As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children  (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p = `r corage$p.value`).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p = `r tsex$p.value`), birth order  (`r corbo$statistic`) = `r corbo$estimate`, p = `r corbo$p.value`) or maternal education  (`r corme$statistic`) = `r corme$estimate`, p = `r corme$p.value`)
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p = `r tsex$p.value`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, `r corbo$statistic`) = `r corbo$estimate`, p = `r corbo$p.value`) or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children, `r corme$statistic`) = `r corme$estimate`, p = `r corme$p.value`)
 

### 2020-06-15 manuscript.tex
fixed little bugs, added freq ana and abstract

@@
 \abstract{
-ADD
+In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
+Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+This study contributes to three lines of research.
+First, we document that non-word items containing typologically rare sounds are repeated accurately less often that non-words containing more common sounds.
+Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
+Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
+Together, these data provide a unique view of online phonological processing in a seldom-studied language, and contribute to both typological and language acquisition research.
 }
@@
 \maketitle
-
-\hypertarget{todo}{%
-\subsection{TODO}\label{todo}}
-
-\begin{itemize}
-\tightlist
-\item
-  double check demo: exact age missing: 78, 82, 83, 115, 119; rough age missing 78, there's a mixture of BGMF in sex, first born is missing, round age
-\item
-  then add analysis of NWR as a function of frequency
-\item
-  draft discussion
-\item
-  draft abstract
-\end{itemize}
 
@@
 \item
-  look through for \textbf{mc}
+  look through this manuscript for bits of text to be written by \textbf{mc}
+\item
+  help with table 1 (formatting \& making it appear, because it's broken by the phonological representations)
+\item
+  do you think Figure 2 should be done with phonetic symbols? If yes, could you try to implement that?
+\item
+  double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
+I looked again at my local files (still haven't checked the labs' ones) and I'm surprised I have maternal education for more children than I have birth order for\ldots{} If you correct this, to generate the csv I also round the age, assign age.rounded to reported age (in years) when DOB is missing, remove first names, DOB, DOTest and age.precise before pushing to github
 \item
@@
 \item
-  help with table 1 (formatting \& making it appear, because it's broken by the phonological representations)
+  draft discussion
 \end{itemize}
@@
 
-NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001).
+NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and to assess the extent to which apparent phonological constraints found in the lexicon affect online production (Gallagher, 2014).
 It is also frequently used as for applications, notably as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
@@
 
+Additionally, we looked up each phoneme in each target word in the Phoible database (Moran \& McCloy, 2019).
+We entered the number of languages in which each phoneme was found, as well as what percentage of languages in the database that represented.
+We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
+* \textbf{mc} do you agree with this decision or shall we do the alternative?*
+
 Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
@@
 In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
-The remaining 40 children (14 girls) were aged 6.96 years (range 3.92-11.03 years).
+The remaining 40 children (14 girls) were aged 6.96 years (range 3.90-11 years).
 There were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and , 2 for whom this information was missing.
-Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Education is often reported in even years because people typically complete two-year cycles.} In terms of birth order, 0 were first borns, 5 second, 4 third, 2 forth, 6 fifth, 5 sixth, and 1 did not have this information.
+Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Education is often reported in even years because people typically complete two-year cycles.} In terms of birth order, 5 were first borns, 4 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
 
@@
 We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
-We excluded 6 children who did not have data for one of these two types.
+We excluded 3 children who did not have data for one of these two types.
 As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
-Surprisingly, subsequent repetitions (M = 41, SD = 27)
-were on average less accurate than first ones (M = 64, SD = 15), t(35) = 5.88, p = 0.
+Surprisingly, subsequent repetitions (M = 39, SD = 26)
+were on average less accurate than first ones (M = 65, SD = 15), t(38) = 6.62, p = 0.
 Given the uncertainty in whether previous work used only the first or all repetitions, and since behavior degraded and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
 
-Taking into account only the first attempts, we averaged attempts by each of the 24 children who had data for first attempts; their ages ranged from 3.92 to 10.20 (M = 6.50, SD = 1.50).
+Taking into account only the first attempts, we averaged attempts by each of the 24 children who had data for first attempts; their ages ranged from 3.90 to 10.20 (M = 6.50, SD = 1.50).
 
@@
 
-We investigated patterns of deletion and substitution. Deletions were relatively rare, with only 29 vowels deleted, and 9 consonants.
+In this analysis, we were interested in variation in performance as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
+The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID \& target ID as random variables.
+
+We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
+The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.03, SE ß = 0.01, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/Fig2-xling_freq-1.pdf}
+\caption{(\#fig:Fig2-xling\_freq)NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}
+\end{figure}
+
+Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 29 vowels deleted, and 9 consonants.
 
@@
 \midrule
-46 (22) & 58 (17) & 42 (17)\\
-80 (23) & 94 (7) & 6 (7)\\
-78 (20) & 94 (6) & 6 (6)\\
-74 (30) & 93 (11) & 7 (11)\\
+47 (22) & 59 (17) & 41 (17)\\
+79 (22) & 94 (7) & 6 (7)\\
+78 (19) & 94 (6) & 6 (6)\\
+75 (32) & 93 (11) & 7 (11)\\
 \bottomrule
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,519.81) = 0.34, p = 0.10).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.18) = 21.34, p = 0.86), birth order (2,630.74) = -0.14, p = 0.50) or maternal education (2,091.92) = 0.09, p = 0.67)
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,548.04) = 0.33, p = 0.12).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.18) = 21.34, p = 0.86), birth order (data missing for 17 children, 2,678.47) = -0.16, p = 0.44) or maternal education (data missing for 2 children, 2,091.92) = 0.09, p = 0.67)
 
@@
 
+\leavevmode\hypertarget{ref-gallagher2014identity}{}%
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
+
 \leavevmode\hypertarget{ref-gallon2007non}{}%
@@
 
+\leavevmode\hypertarget{ref-phoible}{}%
+Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
+
 \leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%

### 2020-06-15 wrangling.R

@@ -1,21 +1,19 @@
-read.csv("NWR-demo.csv",header=T)->demo
-#colnames(demo)
 
-#if family is missing, attribute each child to a different one
-demo$familyID=as.character(demo$familyID)
-demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])
-
-subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
-#if age.exact is missing, use reported age
-inc$age.exact[is.na(inc$age.exact)]<-inc$Age[is.na(inc$age.exact)]
 read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
+#DATA REMOVAL *** ATTENTION *** 
+#REMOVE ALL TASK 4 (whatever that is)
+trnsc[trnsc$item!="task4",]->trnsc
+
+dim(trnsc) #1488 judgments
+
 trnsc$tokid=paste(trnsc$item,trnsc$token)
 
 read.table("final_order.txt",header=T)->crp
 crp$tokid=paste(crp$target,crp$nb)
 npairs=dim(crp)[1]
 
-trnsc=merge(trnsc,crp,by.x="tokid",by.y="tokid",all=T)
+trnsc=merge(trnsc,crp,by.x="tokid",by.y="tokid")
+dim(trnsc) #haven't lost any judgments
 
 #order so we have all the items by a given child in the right chrono order
 trnsc[order(trnsc$file, trnsc$int),]->trnsc
@@ -24,16 +22,27 @@ trnsc[order(trnsc$file, trnsc$int),]->trnsc
 trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
 trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
 trnsc$attempt[1]<-"first"
+table(trnsc$attempt) #1045 first attempts
 
 #add repetition number
 trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
 
-#add coding of correctness online/on the fly of first presentation as a function whether there are subsequent reps of the same target
-trnsc$cor.online=1
-for(i in 1:npairs) if(trnsc$rep[i]>1) trnsc$cor.online[i-1]<-0  #if the item is repeated, then code the previous one as being wrong
[…]

### 2020-09-07 manuscript.Rmd
add minor M2A comments from reading

@@
 
+<!--M2A: I feel like there's a bit of a jump here from explaining what NWR is and what it's used for to saying we used it in the current study to test three things.
+We should probably have a paragraph justifying its use in this context, no?-->
+
 In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
@@
 However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
+<!--M2self: make sure it's apparent how/why a visit is limited-->
 
@@
 - Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
+<!--M2A: I'm imagining a reviewer who's knowlegeable about phonology and child development but who hasn't encountered NWR before.
+I think it would be beneficial to provide a concrete example of a nonword and an inexact repetition above so that they can conceive of what these questions imply (analytically) before we get into the methodological weeds.-->
 
@@
 The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
+ <!--M2A: I didn't look for the SM yet, but probably best not to provide exact hamlet names-->
 
@@
 A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.
+<!--M2A: do you know how many these were? oh! it's below. then maybe we can just put a (see Participants below), etc.-->
 
@@
 A native research assistant then listened through all productions of a given target (randomized across children and repetition order), and made a judgment of whether the item was correctly or incorrectly repeated.
-She additionally transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
+She additionally orthographically transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
 
@@
 Although previous work does not use distance metrics, we additionally report those.
+<!--M2A: A concrete scoring example would be a good idea here-->
 
@@
 We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
-* **mc** do you agree with this decision or shall we do the alternative?*
+<!--
+A2M: do you agree with this decision or shall we do the alternative?
+M2A: yeah, i think adding abirtrarily small frequencies to the NA cases is fraught for the reasons we talked about this morning (i.e., possible systematic undersampling in phoible that can't be helped).
+-->
 

### 2020-09-07 manuscript.Rmd
added irb info, proportion tab in progress

--- 2020-06-15 manuscript.Rmd
@@
 
-This study was approved as part of a larger research effort by **@ac add info from Middy's FAIR reply**.
+This study was approved as part of a larger research effort by Marisa Casillas, with the research protocol having been approved by the IRB committee entitled The Ethics Committee of the Faculty of Social Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) of Radboud University, Nijmegen, The Netherlands.
+The approval for work outside of the Netherlands came as an amendment to the general research line on language development research (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
 Participation was voluntary, with children being invited to come and participate.
@@
 
-read.table("substitution_bank.txt",header=T)->sub_bank
+read.table("substitution_bank_try1.txt",header=T)->sub_bank
 colnames(sub_bank)<-c("target","realization")
@@
 
-read.table("deletion_bank.txt",header=T)->del_bank
+read.table("deletion_bank_try1.txt",header=T)->del_bank
 colnames(del_bank)<-c("target")
@@
 
+read.table("match_bank_try1.txt",header=T)->match_bank
+colnames(match_bank)<-c("target")
+match_bank$target=factor(match_bank$target,levels=allVC)
+match_tab= table(match_bank)[allVC]
+
 #remove English meanings
 trnsc$meaning[grep("English",trnsc$meaning)]<-NA
+
+#Create proportion table
+#a reasonable way to calculate proportion is the proportion of times 
+#a target was produced as X out of the total number of times a target appeared
+#so the denominator is not the number of substitutions (which appears in sub_tab)
+#but rather the sum of sub, del, mat
+#a check: colnames(sub_tab) == names(del_tab)
+#a check: names(match_tab) == names(del_tab)
+
+denom_vec=rowSums(sub_tab) + del_tab + match_tab
+#THIS IS STILL BROKEN
+#prop_cat=data.frame(apply(all_far,2,dodiv)*100) #generates precision because columns
+#prop_cat=all_far/rowSums(all_far)*100 #generates recall because rows
+#sub_tab_prop=t(t(matrix(sub_tab,ncol=length(denom_vec)))/matrix(denom_vec) )
 ```
@@
 
-** this is higher than any previous report! can we be sure of it?**
+**TO DISCUSS TOGETHER this is higher than any previous report! can we be sure of it?**
 

### 2020-09-07 wrangling.R
added irb info, proportion tab in progress

@@ -87,7 +87,8 @@ trnsc$correct[trnsc$correct==0 & trnsc$target_ortho == trnsc$mp & !is.na(trnsc$m
 
 # create phonological correspondance matrix
 # NOTE!! NON INTUITIVE MAPPING!!!! DO NOT READ THE OUTPUT BELIEVING IT IS PSEUDO IPA!!!
-# NOTE2!!! it's better if the chars are not accented & since we don't distinguish long & short, we'll use capitals for nasal & small caps for oral, with no distinction between short and long (see above)
+# NOTE2!!! it's better if the chars are not accented & since we don't distinguish long & short, 
+# we'll use capitals for nasal & small caps for oral, with no distinction between short and long (see above)
 correspondances=matrix( #list correspondances always by pairs, orthography then phonology
   c(
     #nasal  to unichar
@@ -143,23 +144,71 @@ for(i in 1:dim(correspondances)[1]) { #transform into unicharacter
 
 #calculate Levinshtein's distances 
 trnsc$ins=trnsc$del=trnsc$sub=trnsc$nchar=trnsc$trafos=trnsc$tarlen=NA
-substitution_bank=NULL
-deletion_bank=NULL
-for(i in grep("0",trnsc$correct)){
+substitution_bank=deletion_bank=match_bank=substitution_bank_try1=deletion_bank_try1=match_bank_try1=NULL
+
+for(i in grep("0",trnsc$correct)){ #this goes through all items with errors (" grep("0",trnsc$correct)")
   #lev
   lev=adist(trnsc[i,c("mp_uni")],trnsc[i,c("target_uni")],count=T)
   trnsc$ins[i]<-attributes(lev)$counts[,,"ins"]
   trnsc$del[i]<-attributes(lev)$counts[,,"del"]
   trnsc$sub[i]<-attributes(lev)$counts[,,"sub"]
-  trnsc$trafos[i]=attributes(lev)$trafos
+  trnsc$trafos[i]=attributes(lev)$trafos #M, I, D and S indicating a match, insertion, deletion and substitution
   trnsc$nchar[i]=nchar(attributes(lev)$trafos)
   trnsc$tarlen[i]=nchar(trnsc$target_uni[i])
+  
+  #extract substitutions & add them to the bank
[…]

### 2020-09-07 manuscript.Rmd
incorporated changes to analysis of substitutions and deletions

@@
 #a target was produced as X out of the total number of times a target appeared
+
 #so the denominator is not the number of substitutions (which appears in sub_tab)
-#but rather the sum of sub, del, mat
-#a check: colnames(sub_tab) == names(del_tab)
-#a check: names(match_tab) == names(del_tab)
+#but rather the sum of targets appearing in sub, del, & mat
+#check: colnames(sub_tab) == names(del_tab)
+#check: names(match_tab) == names(del_tab)
+denom_vec=rowSums(sub_tab) + del_tab + match_tab
 
-denom_vec=rowSums(sub_tab) + del_tab + match_tab
-#THIS IS STILL BROKEN
-#prop_cat=data.frame(apply(all_far,2,dodiv)*100) #generates precision because columns
-#prop_cat=all_far/rowSums(all_far)*100 #generates recall because rows
-#sub_tab_prop=t(t(matrix(sub_tab,ncol=length(denom_vec)))/matrix(denom_vec) )
+#all numbers below are percentages (not proportions), and they sum to 100 across sub, del, mat at the TARGET level
+
+#this means: divide each column by corresponding number in denominator vector
+#sub_pc has the same shape as the original sub_tab, namely targets as rows, realization as columns
+sub_pc=round(sub_tab/c(denom_vec)*100,2)
+
+del_pc=round(del_tab/denom_vec*100,2)
+mat_pc=round(match_tab/denom_vec*100,2)
+
+#check that indeed:
+#all numbers ... sum to 100 across sub, del, mat at the TARGET level 
+#NOTE! if you round above, then these don't sum to 100, so don't round very much...
+#rowSums(sub_pc, na.rm=T) + del_pc + mat_pc
 ```
 
+**use proportion of correctly produced rather than number; DONE**
+
+**a2m use first attempts rather than the whole data set: having second thoughts.
+To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers & it is more consistent.
+BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros...
+CF Option 3 further down**
+
+**a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?**
+
+**Option 1: just numbers, first attempts only **
 Next, we investigated patterns of deletion and substitution.
@@
 
-**probably turn this into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set**
+**Option 2: numbers and proportions, first attempts only**
+Next, we investigated patterns of deletion and substitution.
+Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted (about `r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` consonants deleted (about `r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral target vowels produced as nasal vowels or about `r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% out of all oral vowel targets, `r sum(sub_tab[nasalVs,oralVs])` nasal target vowels produced as oral vowels or about `r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (`r sum(sub_tab[oralVs,oralVs])` oral vowels produced with a different quality or about `r mean(sub_pc[oralVs,oralVs],na.rm=T)`% out of all oral vowel targets; `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels produced with a different quality or about `r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically `r paste0(complexCs)`) mispronounced as simple ones (specifically `r paste0(simplexCs)`, `r sum(sub_tab[complexCs,simplexCs])` times or  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all complex consonant targets) than vice versa (`r sum(sub_tab[simplexCs,complexCs])` times or  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (`r sum(sub_tab[simplexCs,simplexCs])` simple consonants mispronounced as other simple ones or about `r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% out of all simple consonant targets, compared to `r sum(sub_tab[complexCs,complexCs])` complex ones or about `r mean(sub_pc[complexCs,complexCs],na.rm=T)`% out of all complex consonant targets).
+
+```{r diffic_all_attempts,echo=F}
+
+read.table("substitution_bank.txt",header=T)->sub_bank
+colnames(sub_bank)<-c("target","realization")
+sub_bank$target=factor(sub_bank$target,levels=allVC)
+sub_bank$realization=factor(sub_bank$realization,levels=allVC)
+
+sub_tab=table(sub_bank)[allVC,allVC]
+
+read.table("deletion_bank.txt",header=T)->del_bank
+colnames(del_bank)<-c("target")
+del_bank$target=factor(del_bank$target,levels=allVC)
+del_tab= table(del_bank)[allVC]
+
+read.table("match_bank.txt",header=T)->match_bank
+colnames(match_bank)<-c("target")
+match_bank$target=factor(match_bank$target,levels=allVC)
+match_tab= table(match_bank)[allVC]
+
+#remove English meanings
+trnsc$meaning[grep("English",trnsc$meaning)]<-NA
+
+#Create proportion table
+#a reasonable way to calculate proportion is the proportion of times 
+#a target was produced as X out of the total number of times a target appeared
+
+#so the denominator is not the number of substitutions (which appears in sub_tab)
+#but rather the sum of targets appearing in sub, del, & mat
+#check: colnames(sub_tab) == names(del_tab)
+#check: names(match_tab) == names(del_tab)
+denom_vec=rowSums(sub_tab) + del_tab + match_tab
+
+#all numbers below are percentages (not proportions), and they sum to 100 across sub, del, mat at the TARGET level
+
+#this means: divide each column by corresponding number in denominator vector
+#sub_pc has the same shape as the original sub_tab, namely targets as rows, realization as columns
+sub_pc=round(sub_tab/c(denom_vec)*100,2)
+
+del_pc=round(del_tab/denom_vec*100,2)
+mat_pc=round(match_tab/denom_vec*100,2)
+
+#check that indeed:
+#all numbers ... sum to 100 across sub, del, mat at the TARGET level 
+#NOTE! if you round above, then these don't sum to 100, so don't round very much...
+#rowSums(sub_pc, na.rm=T) + del_pc + mat_pc
+```
+
+**Option 3: numbers and proportions, ALL attempts**
+Next, we investigated patterns of deletion and substitution.
+Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted (about `r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` consonants deleted (about `r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral target vowels produced as nasal vowels or about `r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% out of all oral vowel targets, `r sum(sub_tab[nasalVs,oralVs])` nasal target vowels produced as oral vowels or about `r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (`r sum(sub_tab[oralVs,oralVs])` oral vowels produced with a different quality or about `r mean(sub_pc[oralVs,oralVs],na.rm=T)`% out of all oral vowel targets; `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels produced with a different quality or about `r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically `r paste0(complexCs)`) mispronounced as simple ones (specifically `r paste0(simplexCs)`, `r sum(sub_tab[complexCs,simplexCs])` times or  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all complex consonant targets) than vice versa (`r sum(sub_tab[simplexCs,complexCs])` times  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (`r sum(sub_tab[simplexCs,simplexCs])` simple consonants mispronounced as other simple ones or about `r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% out of all simple consonant targets, compared to `r sum(sub_tab[complexCs,complexCs])` complex ones or about `r mean(sub_pc[complexCs,complexCs],na.rm=T)`% out of all complex consonant targets).
 

### 2020-09-07 manuscript.tex
incorporated changes to analysis of substitutions and deletions

@@
 We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
-* \textbf{mc} do you agree with this decision or shall we do the alternative?*
 
@@
 
-This study was approved as part of a larger research effort by \textbf{({\textbf{???}}) add info from Middy's FAIR reply}.
+This study was approved as part of a larger research effort by Marisa Casillas, with the research protocol having been approved by the IRB committee entitled The Ethics Committee of the Faculty of Social Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) of Radboud University, Nijmegen, The Netherlands.
+The approval for work outside of the Netherlands came as an amendment to the general research line on language development research (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
 Participation was voluntary, with children being invited to come and participate.
@@
 
-Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 29 vowels deleted, and 9 consonants.
+\textbf{use proportion of correctly produced rather than number; DONE}
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels, 58 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (197 oral vowels produced with a different quality; 23 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 266) than vice versa (2).
-Simple consonants were mispronounced as other simple consonants quite frequently (128 simple consonants mispronounced as other simple ones, compared to 62 complex ones).
+\textbf{a2m use first attempts rather than the whole data set: having second thoughts.
+To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers \& it is more consistent.
+BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros\ldots{} CF Option 3 further down}
 
-\textbf{probably turn this into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set}
+\textbf{a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?}
+
+\textbf{Option 1: just numbers, first attempts only }
+Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 19 vowels deleted, and 4 consonants.
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels, 0 nasal target vowels produced as oral vowels).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (42 oral vowels produced with a different quality; 0 nasal vowels produced with a different quality).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0) than vice versa (0).
+Simple consonants were mispronounced as other simple consonants quite frequently (62 simple consonants mispronounced as other simple ones, compared to 0 complex ones).
+
+\textbf{Option 2: numbers and proportions, first attempts only}
+Next, we investigated patterns of deletion and substitution.
+Deletions were relatively rare, with only 19 vowels deleted (about 1.17\% of all vowel targets), and 4 consonants deleted (about 0.24\% of all consonant targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about 0\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (42 oral vowels produced with a different quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times or about NA\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (62 simple consonants mispronounced as other simple ones or about 0.30\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
+
+\textbf{Option 3: numbers and proportions, ALL attempts}
+Next, we investigated patterns of deletion and substitution.
+Deletions were relatively rare, with only 29 vowels deleted (about 0.55\% of all vowel targets), and 9 consonants deleted (about 0.26\% of all consonant targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels or about 2.01\% out of all oral vowel targets, 58 nasal target vowels produced as oral vowels or about 6.46\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (197 oral vowels produced with a different quality or about 3.20\% out of all oral vowel targets; 23 nasal vowels produced with a different quality or about 7.08\% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 266 times or about 5.55\% out of all complex consonant targets) than vice versa (2 times about 5.55\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (128 simple consonants mispronounced as other simple ones or about 0.48\% out of all simple consonant targets, compared to 62 complex ones or about 2.79\% out of all complex consonant targets).
 
@@
 
-** this is higher than any previous report! can we be sure of it?**
+\textbf{TO DISCUSS TOGETHER this is higher than any previous report! can we be sure of it?}
 

### 2020-09-07 manuscript.Rmd
fix character display in table
table not compiling in RStudio, though the tex file does in LaTeX separately...

--- 2020-09-07 manuscript.Rmd
@@
 library(gdata)
-library(ggplot2)
+library(tidyverse)
 library(lme4)
@@
 
-pl<- read.delim("~/Documents/gitrepos/nwryd19/pl-With_yd.tsv")
+pl<- read.delim("pl-With_yd.tsv")
 palette = c( 
@@
   theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
-  scale_x_continuous(name = "Age (years) ", breaks = c(4,8,12))+
+  scale_x_continuous(name = "Age (years) ", limits = c(4,12), breaks = c(4,8,12))+
 
-  scale_y_continuous(name = "NWR Score", breaks = seq(0, 100, 10))+
+  scale_y_continuous(name = "NWR Score", limits = c(0,100), breaks = seq(0, 100, 10))+
   scale_size_continuous(range = c(2, 6))
@@
 
-```{r tab2-stims, eval=F}
+```{r tab2-stims}
 #add orthographic representations targets & other stims chars
-read.delim("stimuli.txt", encoding="UTF-8")->stims
+stims <- read_delim("stimuli2.txt", delim = "\t")
 stims=stims[order(stims[,1]),-3]
@@
 
-#this is leading to an error
-#mytab2 = kable(stims, "latex", booktabs = T,caption="NWR stimuli in orthographic and phonological representations.", row.names =F) %>%
-#kable_styling() %>%
-#add_header_above(c("Practice" = 2, "Monosyll" = 2, "Bisyll" = 2, "Trisyll" = 2, "Tetrasyll" = 2))
+# mytab2 = kable(stims, "latex", booktabs = T,caption="NWR stimuli in orthographic and phonological representations.", row.names =F) %>%
+# kable_styling() %>%
+# add_header_above(c("Practice" = 2, "Monosyll" = 2, "Bisyll" = 2, "Trisyll" = 2, "Tetrasyll" = 2))
+# mytab2
 
 #now that I added phonological representations, even this basic table chokes!
-apa_table(stims,"NWR stimuli in orthographic and phonological representations.")
+# apa_table(stims,"NWR stimuli in orthographic and phonological representations.")
 ```
@@
 
-This study was approved as part of a larger research effort by Marisa Casillas, with the research protocol having been approved by the IRB committee entitled The Ethics Committee of the Faculty of Social Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) of Radboud University, Nijmegen, The Netherlands.
-The approval for work outside of the Netherlands came as an amendment to the general research line on language development research (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+This study was approved as part of a larger research effort by **@ac add info from Middy's FAIR reply**.
 Participation was voluntary, with children being invited to come and participate.
@@
 
-read.table("substitution_bank_try1.txt",header=T)->sub_bank
+read.table("substitution_bank.txt",header=T)->sub_bank
 colnames(sub_bank)<-c("target","realization")
@@
 
-read.table("deletion_bank_try1.txt",header=T)->del_bank
+read.table("deletion_bank.txt",header=T)->del_bank
 colnames(del_bank)<-c("target")
@@
 
-read.table("match_bank_try1.txt",header=T)->match_bank
-colnames(match_bank)<-c("target")
-match_bank$target=factor(match_bank$target,levels=allVC)
-match_tab= table(match_bank)[allVC]
-
 #remove English meanings
 trnsc$meaning[grep("English",trnsc$meaning)]<-NA
-
-#Create proportion table
-#a reasonable way to calculate proportion is the proportion of times 
-#a target was produced as X out of the total number of times a target appeared
-#so the denominator is not the number of substitutions (which appears in sub_tab)
-#but rather the sum of sub, del, mat
-#a check: colnames(sub_tab) == names(del_tab)
-#a check: names(match_tab) == names(del_tab)
-
-denom_vec=rowSums(sub_tab) + del_tab + match_tab
-#THIS IS STILL BROKEN
-#prop_cat=data.frame(apply(all_far,2,dodiv)*100) #generates precision because columns
-#prop_cat=all_far/rowSums(all_far)*100 #generates recall because rows
-#sub_tab_prop=t(t(matrix(sub_tab,ncol=length(denom_vec)))/matrix(denom_vec) )
 ```
@@
 
-**TO DISCUSS TOGETHER this is higher than any previous report! can we be sure of it?**
+** this is higher than any previous report! can we be sure of it?**
 
@@
 
-We are grateful to informants and individuals who participated in the study.
-AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.
+We are grateful to the individuals who participated in the study.
+The collection and annotation of these recordings was made possible by Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also to the PNG National Research Institute, and the Administration of Milne Bay Province AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.
 S.
 McDonnell Foundation Understanding Human Cognition Scholar Award.
-MC blabla NWO Veni Innovational Scheme Grant (XXX-XX-XXX).
+MC acknowledges financial support from an NWO Veni Innovational Scheme grant (275-89-033).
 

### 2020-09-07 manuscript.Rmd
minor

--- 2020-09-07 manuscript.Rmd
@@
 ## TODO later
-- probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set
-- maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
-- maybe add analyses of score as a function of frequency (phones, diphones) in MC's corpus? (mc suggested, but I don't think we need it...)
+- probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set DONE
+- maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed) DONE I think... (this item looks the same as one of the three above)
+- maybe add analyses of score as a function of frequency (phones, diphones) in MC's corpus? (mc suggested, but I don't think we need it...) - waiting until we talk about this
 
@@
  <!--M2A: I didn't look for the SM yet, but probably best not to provide exact hamlet names-->
+<!--a2m: good point! fixed-->
 

### 2020-09-07 manuscript.Rmd
make table render

--- 2020-09-07 manuscript.Rmd
@@
 class             : "man"
-output            : papaja::apa6_pdf
+output            :
+  papaja::apa6_pdf:
+    latex_engine: xelatex
 ---
@@
 
-```{r tab2-stims}
+```{r tab2-stims, results = "asis"}
 #add orthographic representations targets & other stims chars
-stims <- read_delim("stimuli2.txt", delim = "\t")
-stims=stims[order(stims[,1]),-3]
-colnames(stims)<-c("Type","Orthographic","Phonological")
-cbindX(stims[stims$Type=="practice",],
-      stims[stims$Type==1,],
-      stims[stims$Type==2,],
-      stims[stims$Type==3,],
-      stims[stims$Type==4,])
+stims <- read_delim("stimuli2.txt", delim = "\t") %>%
+  arrange(factor(type, levels = c("practice", "1", "2", "3", "4"))) %>%
+  select(type, target_ortho, phono) %>%
+  rename("Type" = type, "Orthographic" = target_ortho, "Phonological" = phono)
 
-stims=stims[,-1]
+# probably a nicer way to do this but anyhow...
+practice.stims <- filter(stims, Type == "practice")[,2:3]
+practice.stims$row <- seq(1:nrow(practice.stims))
+syll1.stims <- filter(stims, Type == "1")[,2:3]
+syll1.stims$row <- seq(1:nrow(syll1.stims))
+syll2.stims <- filter(stims, Type == "2")[,2:3]
+syll2.stims$row <- seq(1:nrow(syll2.stims))
+syll3.stims <- filter(stims, Type == "3")[,2:3]
+syll3.stims$row <- seq(1:nrow(syll3.stims))
+syll4.stims <- filter(stims, Type == "4")[,2:3]
+syll4.stims$row <- seq(1:nrow(syll4.stims))
+stims.wide <- practice.stims %>%
+  full_join(syll1.stims, by = "row") %>%
+  full_join(syll2.stims, by = "row") %>%
+  full_join(syll3.stims, by = "row") %>%
+  full_join(syll4.stims, by = "row") %>%
+  select(-row) %>%
+  replace(is.na(.), "")
+colnames(stims.wide) <- c(
+  "Orth.", "Phon.",
+  "Orth.", "Phon.",
+  "Orth.", "Phon.",
+  "Orth.", "Phon.",
+  "Orth.", "Phon."
+)
 
-# mytab2 = kable(stims, "latex", booktabs = T,caption="NWR stimuli in orthographic and phonological representations.", row.names =F) %>%
-# kable_styling() %>%
-# add_header_above(c("Practice" = 2, "Monosyll" = 2, "Bisyll" = 2, "Trisyll" = 2, "Tetrasyll" = 2))
-# mytab2
+kable(stims.wide,
+  "latex", booktabs = T,
+  caption="NWR stimuli in orthographic and phonological representations.",
+  row.names =F) %>%
+  kable_styling() %>%
+  add_header_above(c("Practice" = 2, "Monosyll" = 2,
+    "Bisyll" = 2, "Trisyll" = 2, "Tetrasyll" = 2))
 

### 2020-09-07 manuscript.tex
re-render table
some characters still mysteriously not showing up. grrrr.

@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
-\PassOptionsToPackage{hyphens}{url}
-%
 \documentclass[english,,man,floatsintext]{apa6}
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
 \else % if luatex or xelatex
-  \usepackage{unicode-math}
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
+  \fi
   \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
@@
 }{}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\usepackage{hyperref}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
 \hypersetup{
             pdftitle={Non-word repetition in Yélî Dnye},
+            pdfauthor={Alejandrina Cristia~\& Marisa Casillas},
             pdfkeywords={phonology, non-word repetition, development},
@@
 \urlstyle{same}  % don't use monospace font for urls
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[shorthands=off,main=english]{babel}
+\else
+  \usepackage{polyglossia}
+  \setmainlanguage[]{english}
+\fi
 \usepackage{graphicx,grffile}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 \setcounter{secnumdepth}{0}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
 
-% Manuscript styling
+\shorttitle{NWR in Yélî Dnye}
+\affiliation{
+\vspace{0.5cm}
+\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
+\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
+\usepackage{csquotes}
 \usepackage{upgreek}
@@
 
-% Table formatting
 \usepackage{longtable}
 \usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+\usepackage{multirow}
+\usepackage{tabularx}
+\usepackage[flushleft]{threeparttable}
+\usepackage{threeparttablex}
 
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
 \newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
 
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
 \makeatletter
@@
 
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\makeatother
-\shorttitle{NWR in Yélî Dnye}
-\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2}}
-\affiliation{
-\vspace{0.5cm}
-\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
-\authornote{All data are made available in a repository in the Open Science Framework.
-AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
-S.
-McDonnell Foundation Understanding Human Cognition Scholar Award.
-
-Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
-\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
 \usepackage{lineno}
@@
 \linenumbers
-\usepackage{csquotes}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\fi
 
 \title{Non-word repetition in Yélî Dnye}
-
+\author{Alejandrina Cristia\textsuperscript{1}~\& Marisa
+Casillas\textsuperscript{2}}
 \date{}
 
+\authornote{All data are made available in a repository in the
+Open Science Framework. AC acknowledges the support of the Agence
+Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004,
+ANR-14-CE30-0003, ANR-17-EURE-0017); and the J. S. McDonnell Foundation
+Understanding Human Cognition Scholar Award.
+
+Correspondence concerning this article should be addressed to
+Alejandrina Cristia, 29, rue d'Ulm, 75005 Paris, France. E-mail:
+\href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
+
 \abstract{
-In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
-Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
-This study contributes to three lines of research.
-First, we document that non-word items containing typologically rare sounds are repeated accurately less often that non-words containing more common sounds.
-Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
-Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
-Together, these data provide a unique view of online phonological processing in a seldom-studied language, and contribute to both typological and language acquisition research.
+In nonword repetition (NWR) studies, participants are presented
+auditorily with an item that is phonologically legal but lexically
+meaningless in the local language. Accuracy is thought to reflect
+long-term phonological knowledge as well as online phonological working
+memory and flexible production patterns.In this study, we report on NWR
+results among children learning Yêly Dnyé, an isolate spoken in Rossel
+Island, PNG, with an unusually dense phonological inventory. This study
+contributes to three lines of research. First, we document that non-word
+items containing typologically rare sounds are repeated accurately less
+often that non-words containing more common sounds. Second, we document
+rather weak effects of item length, contributing to other research
+suggesting that length effects may be language-specific. Third, we do
+not find strong individual variation effects in this population,
+contrary to previous results documenting strong age-related effects.
+Together, these data provide a unique view of online phonological
+processing in a seldom-studied language, and contribute to both
+typological and language acquisition research.
+
 }
@@
 
-\hypertarget{todo-middy}{%
-\subsection{TODO Middy}\label{todo-middy}}
+\subsection{TODO Middy}\label{todo-middy}
 
@@
 \item
-  look through this manuscript for bits of text to be written by \textbf{mc}
+  look through this manuscript for bits of text to be written by
+  \textbf{mc}
 \item
-  help with table 1 (formatting \& making it appear, because it's broken by the phonological representations)
+  help with table 1 (formatting \& making it appear, because it's broken
+  by the phonological representations)
 \item
-  do you think Figure 2 should be done with phonetic symbols? If yes, could you try to implement that?
+  do you think Figure 2 should be done with phonetic symbols? If yes,
+  could you try to implement that?
 \item
-  double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
-I looked again at my local files (still haven't checked the labs' ones) and I'm surprised I have maternal education for more children than I have birth order for\ldots{} If you correct this, to generate the csv I also round the age, assign age.rounded to reported age (in years) when DOB is missing, remove first names, DOB, DOTest and age.precise before pushing to github
+  double check demo -- probably start with the xlsx version that has the
+  children's first names, date of birth, and date of test. I looked
+  again at my local files (still haven't checked the labs' ones) and I'm
+  surprised I have maternal education for more children than I have
+  birth order for\ldots{} If you correct this, to generate the csv I
+  also round the age, assign age.rounded to reported age (in years) when
+  DOB is missing, remove first names, DOB, DOTest and age.precise before
+  pushing to github
 \item
-  double check my frequency entries in segments.xlsx using the search function in \url{https://phoible.org/parameters}; take a look at \url{http://phoible.github.io/conventions/} in case you see something about double articulation (I didn't find tp or lBj)
+  double check my frequency entries in segments.xlsx using the search
+  function in \url{https://phoible.org/parameters}; take a look at
+  \url{http://phoible.github.io/conventions/} in case you see something
+  about double articulation (I didn't find tp or lBj)
 \item
@@
 \item
-  for both of the above, if you make any changes, they need to be re-saved as UTF16 txt file, and then converted (see remember above)
+  for both of the above, if you make any changes, they need to be
+  re-saved as UTF16 txt file, and then converted (see remember above)
 \item
-  double check also stim\_seg\_freq -- why are some items still with NA freq?
+  double check also stim\_seg\_freq -- why are some items still with NA
+  freq?
 \item
@@
 
-\hypertarget{todo-later}{%
-\subsection{TODO later}\label{todo-later}}
+\subsection{TODO later}\label{todo-later}
 
@@
 \item
-  probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set
+  probably turn summary of types of error into a table; use proportion
+  of correctly produced rather than number; use first attempts rather
+  than the whole data set DONE
 \item
-  maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed)
+  maybe add to the last ana the prevalence (ie out of all the gh which
+  proportion get transformed) DONE I think\ldots{} (this item looks the
+  same as one of the three above)
 \item
-  maybe add analyses of score as a function of frequency (phones, diphones) in MC's corpus? (mc suggested, but I don't think we need it\ldots{})
+  maybe add analyses of score as a function of frequency (phones,
+  diphones) in MC's corpus? (mc suggested, but I don't think we need
+  it\ldots{}) - waiting until we talk about this
 \end{itemize}
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+\subsection{Introduction}\label{introduction}
 
-Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., Hazan \& Barrett, 2000).
-One common task in this line of research is nonword repetition (NWR).
-In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning.
-The child should immediately try to say it back without changing anything.
-Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item accurately even though it had never been pronounced before).
+Although infants begin to learn about their native language's phonology
+within the first year, many studies suggest that in perception and
+production, in phonetics and phonology, their knowledge continues to
+develop throughout childhood (e.g., Hazan \& Barrett, 2000). One common
+task in this line of research is nonword repetition (NWR). In NWR
+studies, participants are presented auditorily with an item that is
+phonologically legal but lexically meaningless in the language children
+are learning. The child should immediately try to say it back without
+changing anything. Accuracy is thought to reflect long-term phonological
+knowledge (which allows the child to perceive the item accurately even
+though it is not a real word they have encountered before) as well as
+online phonological working memory (to encode the item in the interval
+between hearing it and saying it back) and flexible production patterns
+(to produce the item accurately even though it had never been pronounced
+before).
 
-NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and to assess the extent to which apparent phonological constraints found in the lexicon affect online production (Gallagher, 2014).
-It is also frequently used as for applications, notably as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
+NWR has been used to seek answers to a variety of theoretical questions,
+including what the links between phonology, working memory, and the
+lexicon are (Bowey, 2001), and to assess the extent to which apparent
+phonological constraints found in the lexicon affect online production
+(Gallagher, 2014). It is also frequently used as for applications,
+notably as a diagnostic for language delays and disorders (Estes, Evans,
+\& Else-Quest, 2007). Since non-words can be generated in any language,
+it has attracted the attention of researchers working in multilingual
+and linguistically diverse environments, particularly in Europe (Action,
+2009; Meir, Walters, \& Armon-Lotem, 2016).
 
-In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
-The present report uses these unique data to contribute to several lines of research.
-First, we made sure that some of the non-word items contained typologically rare and/or challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall accuracy of repetition and patterns of mispronunciation.
-Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
-Third, we contribute to the basic NWR literature by contributing an additional data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
+In this study, we report on NWR results among children learning Yêly
+Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense
+phonological inventory. The present report uses these unique data to
+contribute to several lines of research. First, we made sure that some
+of the non-word items contained typologically rare and/or challenging
+sounds, so that we can contribute to the study of whether rare sounds
+are disadvantaged in perception and/or production, both in terms of
+overall accuracy of repetition and patterns of mispronunciation. Second,
+we look at the effects of item length, since previous NWR research has
+uncovered variable effects. Third, we contribute to the basic NWR
+literature by contributing an additional data point for children
+learning a rarely-studied language and culture, and further
+investigating whether there are structured sources that account for
+individual variation.
 
-\hypertarget{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}{%
-\subsubsection{\texorpdfstring{Intro to the language (\textbf{mc}) - please feel free to throw away anything that is not useful!}{Intro to the language (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
+\subsubsection{\texorpdfstring{Intro to the language (\textbf{mc}) -
+please feel free to throw away anything that is not
+useful!}{Intro to the language (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}
 
@@
 \item
-  although not the focus of this paper, high use of suppletion in verbal paradigms, other features of language, see Levinson XXX for details
+  although not the focus of this paper, high use of suppletion in verbal
+  paradigms, other features of language, see Levinson XXX for details
 \end{itemize}
 
-\hypertarget{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}{%
-\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) - please feel free to throw away anything that is not useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
+\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) -
+please feel free to throw away anything that is not
+useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}
 
-Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
+Little is known about language development in children growing up in
+Rossel Island, a community of primarily subsistence farmers who tend to
+reside in close-knitted villages where child care is distributed across
+many individuals, and who typically speak Yélî Dnyé, a phonologically
+and lexically complex language.
 
@@
 \item
-  however, some use of English due to immigrants \& children of immigrants
+  however, some use of English due to immigrants \& children of
+  immigrants
 \item
@@
 
-\hypertarget{brief-review-of-nwr-for-our-purposes}{%
-\subsubsection{Brief review of NWR for our purposes}\label{brief-review-of-nwr-for-our-purposes}}
+\subsubsection{Brief review of NWR for our
+purposes}\label{brief-review-of-nwr-for-our-purposes}
 
-There is some variation in the presentation procedure and structure of items found in previous NWR work.
-For example, items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation (Brandeker \& Thordardottir, 2015).
-Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
-Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
+There is some variation in the presentation procedure and structure of
+items found in previous NWR work. For example, items are often presented
+orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe,
+2015), although an increasing number of studies have turned to playing
+back the stimuli in order to have greater control of the stability of
+the presentation (Brandeker \& Thordardottir, 2015). Additionally, while
+some studies have used 10-15 non-words, others have employed up to 46
+unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler,
+2019). Often, authors modulate structural complexity, typically measured
+in terms of item length (measured in number of syllables) and/or
+syllable structure (open as opposed to closed syllables, Gallon, Harris,
+\& Van der Lely, 2007).
 
-Previous work seems to avoid difficult sounds, but we felt this was important to represent Yélî Dnye, so we also varied this factor.
-We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
-This allowed us to get information about repetition accuracy of more items.
+Previous work seems to avoid difficult sounds, but we felt this was
+important to represent Yélî Dnye, so we also varied this factor. We
+designed a relatively large number of items but, aware that this may
+render the task longer and more tiresome, we split some of the items
+across children. This allowed us to get information about repetition
+accuracy of more items.
 
-Naturally, designing the task in this way may render the study of individual variaiton within the population more difficult because different children are exposed to different items.
-However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
+Naturally, designing the task in this way may render the study of
+individual variaiton within the population more difficult because
+different children are exposed to different items. However, a review of
+previous work on individual variation suggested to us that many
+individual differences effects are relatively small, and would not be
+detectable with the sample size that we could collect in a given visit.
 
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
-Based on previous work, we looked at potential improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and potential negative effects of bilingual exposure (Brandeker \& Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
-Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
-Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on those too.
+That said, we contribute to the literature by also reporting descriptive
+analyses of individual variation that could potentially be integrated in
+meta-analytic efforts. Based on previous work, we looked at potential
+improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid,
+Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and
+potential negative effects of bilingual exposure (Brandeker \&
+Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
+Previous work typically finds no significant differences as a function
+of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014;
+Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although previous research has not often investigated potential effects
+of birth order on NWR, there is a sizable literature on these effects in
+other language tasks (Havron et al., 2019), and therefore we report on
+those too.
 
-\hypertarget{research-questions}{%
-\subsubsection{Research questions}\label{research-questions}}
+\subsubsection{Research questions}\label{research-questions}
 
-After some preliminary analyses to set the stage, we address the following questions:
+After some preliminary analyses to set the stage, we address the
+following questions:
 
@@
 \item
-  Does the frequency of sounds across languages predict NWR? Are rarer sounds more often substituted by commoner sounds?
+  Does the frequency of sounds across languages predict NWR? Are rarer
+  sounds more often substituted by commoner sounds?
 \item
-  How does score change as a function of item length in number of syllables?
+  How does score change as a function of item length in number of
+  syllables?
 \item
-  Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
+  Is individual variation attributable to child age, sex, birth order,
+  monolingual status, and/or parental education? 
 \end{itemize}
 
-In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects given their variability across studies.
-To illustrate this, we portray studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables) in Figure 1.
-Notice that the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Tsimane' children (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020), where a drop of 40 percentage points is observed at all ages.
-A similar difference in NWR scores for short versus long items was observed among Arabic children (Jaber-Awida, 2018).
-Even the effect of age is unstable in this sample.
-Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
-Therefore, all analyses here are descriptive and should be considered exploratory.
+In view of the hypothesis-driven nature of this work, we had considered
+boosting the interpretational value of this evidence by announcing our
+analysis plans prior to conducting them. However, we realized that even
+pre-registering an analysis would be equivocal because we do not have
+enough power to look at all relationships of interest, and often to
+detect any of the known effects given their variability across studies.
+To illustrate this, we portray studies in which children's NWR scores
+were gathered between 4 and 12 years of age, and reported separately for
+items that are relatively short (1-2 syllables) versus longer items (3-4
+syllables) in Figure 1. Notice that the effect of stimulus length is
+minuscule among Italian children (Piazzalunga et al., 2019), but
+considerable among Tsimane' children (Cristia, Farabolini, Scaff,
+Havron, \& Stieglitz, 2020), where a drop of 40 percentage points is
+observed at all ages. A similar difference in NWR scores for short
+versus long items was observed among Arabic children (Jaber-Awida,
+2018). Even the effect of age is unstable in this sample. Whereas it is
+quite clear that children's NWR scores increase in the Italian data, age
+effects are less stable among Tsimane' children. Therefore, all analyses
+here are descriptive and should be considered exploratory.
 
@@
 \includegraphics{manuscript_files/figure-latex/prevlit-fig-1.pdf}
-\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years) and item length for comparable studies.
-Arabic data from Jaber et al.
-(2018); Italian data from Piazzalunga et al.
-(2019); Tsimane' from Cristia et al.
-(2020); Yélî data from the present study.}
+\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years)
+and item length for comparable studies. Arabic data from Jaber et al.
+(2018); Italian data from Piazzalunga et al. (2019); Tsimane' from
+Cristia et al. (2020); Yélî data from the present study.}
 \end{figure}
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
+\subsection{Methods}\label{methods}
 
-\hypertarget{stimuli}{%
-\subsubsection{Stimuli}\label{stimuli}}
+\subsubsection{Stimuli}\label{stimuli}
 
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
-We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in Yély Dnye.
-We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat (note that up to other work has also used 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
+Many NWR studies are based on a fixed list of 12-16 items that vary in
+length between 1 and 4 syllables, often additionally varying syllable
+complexity and/or cluster presence and complexity, always meeting the
+condition that they do not mean anything in the target language (e.g.,
+Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We kept the
+same variation in item length and the non-meaningfulness requirement,
+but we did not vary syllable complexity and clusters because these are
+vanishingly rare in Yély Dnye. We also increased the number of items an
+individual child would be tested on, so that a child would get up to 23
+items to repeat (note that up to other work has also used 24-30 items:
+Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and
+distributed them across children, so as to increase the coverage, and be
+able to study more items.
 
-A first list of candidate items was generated in 2018 by selecting simple consonants (\enquote{p}, \enquote{t}, \enquote{d}, \enquote{k}, \enquote{m}, \enquote{n}, \enquote{w}, \enquote{y}) and vowels (\enquote{i}, \enquote{o},\enquote{u}, \enquote{a},\enquote{e}) that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
-These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
-The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
-Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
+A first list of candidate items was generated in 2018 by selecting
+simple consonants (\enquote{p}, \enquote{t}, \enquote{d}, \enquote{k},
+\enquote{m}, \enquote{n}, \enquote{w}, \enquote{y}) and vowels
+(\enquote{i}, \enquote{o},\enquote{u}, \enquote{a},\enquote{e}) that
+were combined into consonant-vowel syllables, further sampling the space
+of 1- to 4-syllable sequences. These candidates were automatically
+checked against Levinson's 2015 dictionary and removed from
+consideration if they appeared in the dictionary. The second author
+presented them orally to three local research assistants, who were asked
+to repeat them and further say whether they were real words. Any item
+for which two or more of the assistants reported them having a meaning
+or some form of association was excluded.
 
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî inventory to produce consonant-vowel monosyllables.
-As before, items were automatically excluded if they appeared in the dictionary.
-Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described {[}CITE \textbf{mc} please fill in{]}, there could have been undocumented constraints that rendered items illegal.
-Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
-These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset \textbf{XXX }mc** please fill in** and an Olympus \textbf{XXX }mc** please fill in** from the written form presented together with the same item orally (by the second author).
-The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
-Even so, there was one monosyllable that was often identified as a real word (intended \enquote{yî} /yXX/; identified as \enquote{yi} /yi/, \emph{tree}).
-This item is removed from analyses below.
+A second list of candidate items was generated in 2019 by selecting
+complex consonants and systematically crossing them with all the vowels
+in the Yélî inventory to produce consonant-vowel monosyllables. As
+before, items were automatically excluded if they appeared in the
+dictionary. Additionally, since hearing vowel length in monosyllables in
+isolation is challenging, any item that had a short/long real word
+neighbor was filtered out. Since the phonology and phonetics of Yélî is
+still in the process of being described {[}CITE \textbf{mc} please fill
+in{]}, there could have been undocumented constraints that rendered
+items illegal. Therefore, we made sure that the precise consonant-vowel
+sequence occurred in some real word in the dictionary (i.e., that there
+was a longer word included the monosyllable as a subsequence). These
+candidates were presented to one informant, for a final check that they
+did not mean anything. Together with the 2018 selection, they were
+recorded using a headset \textbf{XXX }mc** please fill in** and an
+Olympus \textbf{XXX }mc** please fill in** from the written form
+presented together with the same item orally (by the second author). The
+complete recorded list was finally presented to two more informants, who
+could repeat all the items and who confirmed there were no real words.
+Even so, there was one monosyllable that was often identified as a real
+word (intended \enquote{yî} /yXX/; identified as \enquote{yi} /yi/,
+\emph{tree}). This item is removed from analyses below.
 
-The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
+The final list is composed of three practice items; 20 monoysllables
+containing sounds that are less frequent in the world's languages than
+singleton plosives; 8 bisyllables; 12 trisyllables; and 4
+quadrisyllables (see Table 2).
 
-A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
-The 40 elicitation sets are available online from \url{https://osf.io/5qspb}.
+\begin{table}
+
+\caption{\label{tab:tab2-stims}NWR stimuli in orthographic and phonological representations.}
+\centering
+\begin{tabular}[t]{llllllllll}
+\toprule
+\multicolumn{2}{c}{Practice} & \multicolumn{2}{c}{Monosyll} & \multicolumn{2}{c}{Bisyll} & \multicolumn{2}{c}{Trisyll} & \multicolumn{2}{c}{Tetrasyll} \\
+\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10}
+Orth. & Phon. & Orth. & Phon. & Orth. & Phon. & Orth. & Phon. & Orth. & Phon.\\
+\midrule
+nopimade & ṇɔpimædɛ & dp:a & ṭpæ͂ & kamo & kæmɔ & dimope & ṭimɔpɛ & diponate & ṭipɔṇætɛ\\
+poni & pɔṇi & dpa & ṭpæ & kani & kæṇi & diyeto & ṭijɛtɔ & nomiwake & ṇɔmiwækɛ\\
+wî & wɯ & dpâ & ṭpɑ & kipo & kipɔ & meyadi & mɛjæṭi & todiwuma & tɔṭiwumæ\\
+ &  & dpê & ṭpə & noki & ṇɔki & mituye & mitujɛ & wadikeno & wæṭikɛṇɔ\\
+ &  & dpéé & ṭpe: & nomi & ṇɔmi & nademo & ṇædɛmɔ &  & \\
+\addlinespace
+ &  & dpi & ṭpi & piwa & piwæ & nayeki & ṇæjɛki &  & \\
+ &  & dpu & ṭpu & towi & tɔwi & nuyedi & ṇujɛdi &  & \\
+ &  & gh:ââ & ɣɑ: & tupa & tupæ & pedumi & pɛdumi &  & \\
+ &  & ghuu & ɣu: &  &  & tiwune & tiwuṇɛ &  & \\
+ &  & kp:ââ & kpɑ͂: &  &  & tumowe & tumɔwɛ &  & \\
+\addlinespace
+ &  & kpu & kpu &  &  & widone & wiṭɔṇɛ &  & \\
+ &  & lv:ê & lβʲə &  &  & wumipo & wumipɔ &  & \\
+ &  & lva & lβʲæ &  &  &  &  &  & \\
+ &  & lvi & lβʲi &  &  &  &  &  & \\
+ &  & t:êê & tə͂ː &  &  &  &  &  & \\
+\addlinespace
+ &  & tp:a & tpæ͂ &  &  &  &  &  & \\
+ &  & tpâ & tpɑ &  &  &  &  &  & \\
+ &  & tpê & tpə &  &  &  &  &  & \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+A Praat script was written to randomize this list 20 times, and split it
+into two sublists, to generate 40 different elicitation sets. The 40
+elicitation sets are available online from \url{https://osf.io/5qspb}.
 The split had the following constraints:
@@
 \item
-  the same three items were selected as practice items and used in all 40 elicitation sets
+  the same three items were selected as practice items and used in all
+  40 elicitation sets
 \item
-  splits were done within each length group from the 2018 items (i.e., separately for 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
-Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
+  splits were done within each length group from the 2018 items (i.e.,
+  separately for 2, 3, and 4-syllable items); and among onset groups for
+  the difficult monosyllables generated in 2019 (i.e., all the
+  monosyllables starting with tp were split into 2 sublists). Since some
+  of these groups had an odd number of items, one of the sublists was
+  slightly longer than the other (20 versus 23).
 \item
-  once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllables, then 3-syllables, and finally 4-syllable items.
+  once the sublist split had been done, items were randomized such that
+  all children heard first the 3 practice items in a fixed order (1, 2,
+  and 4 syllables), a randomized version of their sublist selection of
+  difficult onset items, and randomized versions of their 2-syllables,
+  then 3-syllables, and finally 4-syllable items.
 \end{itemize}
 
-\hypertarget{procedure}{%
-\subsubsection{Procedure}\label{procedure}}
+\subsubsection{Procedure}\label{procedure}
 
-We tried to balance three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them; that children would feel comfortable doing this task with us; and that the community would feel safe with us doing this task with their children.
-Moreover, there were also some logistic constraints in terms of the space availability.
-As a result, the places where elicitation happened varied across the hamlets.
+We tried to balance three desiderata: That children would not be unduly
+exposed to the items before they themselves had to repeat them; that
+children would feel comfortable doing this task with us; and that the
+community would feel safe with us doing this task with their children.
+Moreover, there were also some logistic constraints in terms of the
+space availability. As a result, the places where elicitation happened
+varied across the hamlets.
 
-We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items \enquote{spreading} through hearsay.
-In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
-The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
+We visited four different hamlets once, and attempted to test all
+eligible children present at the time, to prevent the items
+\enquote{spreading} through hearsay. In the first hamlet, we tested
+children in five different places, with some children being tested
+inside a house and others tested on the veranda. The complete list of
+places and the ways in which they met the desiderata mentioned above can
+be found in the raw data, available from online supplementary materials.
 
-The child was donned a headset (\textbf{xx }mc** please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
-For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headphones.
+The child was donned a headset (\textbf{xx }mc** please fill in** for
+most of the children, SHURE WH20 XLR headset with a dynamic microphone
+for the rest), recorded into the left channel into a Tascam DR40x
+digital recorder. For most children, the headset could not stay
+comfortably on the child's head, and thus it was placed on the child's
+shoulders, with the microphone carefully placed close to the child's
+mouth. A local informant sat next to the child, to would provide the
+instructions and, if needed, coach the child to make sure, using the
+three practice items as well as real words, that they understood that
+the task was to repeat the items precisely without changing anything. An
+experimenter (the first author) delivered the elicitation stimuli to the
+local informant and the child over headphones.
 
-The first phase was making sure the child understood the task.
-This was explained orally and the first training item was presented.
-Often, children froze and did not say anything.
-If this happened, then we followed this procedure.
-First, the informant insisted.
-If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another.
-If the child could repeat these correctly, then we provided the recorded training item over headphones again.
-Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it).
-A small minority still failed to repeat the item after hearing it over headphones.
-If that occurred, we tried with the second training item, at which point some children got it and could continue.
-A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.
+The first phase was making sure the child understood the task. This was
+explained orally and the first training item was presented. Often,
+children froze and did not say anything. If this happened, then we
+followed this procedure. First, the informant insisted. If the child
+still did not say anything, the informant asked the child to repeat a
+real word, and another, and another. If the child could repeat these
+correctly, then we provided the recorded training item over headphones
+again. Most children successfully started repeating the items presented
+over headphones at this point; a few further needed the local informant
+to model the behavior (i.e., they would hear the item again, and she
+would say it; then we would play it again, and ask the child to say it).
+A small minority still failed to repeat the item after hearing it over
+headphones. If that occurred, we tried with the second training item, at
+which point some children got it and could continue. A small minority,
+however, failed to repeat this one, as well as the third training item,
+in which case we stopped the test altogether.
 
-NWR studies vary in whether children are provided with several opportunities to hear and say the item.
-To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them.
-If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a separate headset that was recording onto the right track of the same Tascam).
-If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.
+NWR studies vary in whether children are provided with several
+opportunities to hear and say the item. To have a fixed and clear
+procedure, we decided that items other than the inital three training
+ones would not be repeated unless the child made an attempt to produce
+them. If this attempt was judged correct by the local informant, then
+the experimenter would move on to the next item (whispering this over a
+separate headset that was recording onto the right track of the same
+Tascam). If the local informant heard a deviation, she indicated to the
+experimenter that the item needed to be repeated, and up to 5 attempts
+were allowed.
 
-Whenever siblings from the same family were tested, an attempt was made to test first the older and then the younger child, and always on different elicitaiton sets.
+Whenever siblings from the same family were tested, an attempt was made
+to test first the older and then the younger child, and always on
+different elicitaiton sets.
 
-\hypertarget{coding}{%
-\subsubsection{Coding}\label{coding}}
+\subsubsection{Coding}\label{coding}
 
-A script was written to randomize all tokens from all children, pairing each with the auditory target the child had been provided with.
-A native research assistant then listened through all productions of a given target (randomized across children and repetition order), and made a judgment of whether the item was correctly or incorrectly repeated.
-She additionally orthographically transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
+A script was written to randomize all tokens from all children, pairing
+each with the auditory target the child had been provided with. A native
+research assistant then listened through all productions of a given
+target (randomized across children and repetition order), and made a
+judgment of whether the item was correctly or incorrectly repeated. She
+additionally orthographically transcribed exactly what the child said,
+providing some examples of the types of errors children in general make
+(without making specific reference to Yélî sounds or the items in the
+elicitation sets).
 
-\hypertarget{analyses}{%
-\subsubsection{Analyses}\label{analyses}}
+\subsubsection{Analyses}\label{analyses}
 
-Some NWR studies employ phoneme-based scoring in addition to or instead of word-level accuracy (e.g., Cristia et al., 2020).
-We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
-Although previous work does not use distance metrics, we additionally report those.
+Some NWR studies employ phoneme-based scoring in addition to or instead
+of word-level accuracy (e.g., Cristia et al., 2020). We scored items in
+terms of the number of phonemes that could be aligned across the target
+and attempt, divided by the number of phonemes of whichever item was
+longer (the target or the attempt). Although previous work does not use
+distance metrics, we additionally report those.
 
-Additionally, we looked up each phoneme in each target word in the Phoible database (Moran \& McCloy, 2019).
-We entered the number of languages in which each phoneme was found, as well as what percentage of languages in the database that represented.
-We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
+Additionally, we looked up each phoneme in each target word in the
+Phoible database (Moran \& McCloy, 2019). We entered the number of
+languages in which each phoneme was found, as well as what percentage of
+languages in the database that represented. We could not find a small
+number of phonemes in the database, and treat those as NA values that do
+not contribute to the target-level global average phone cross-linguistic
+frequency, although an alternative would be to attribute them a very
+small number.
 
-Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); and classify the most common phonological errors.
+Finally, for describing children's patterns of errors, all repetitions
+of a given target were taken into account. We describe the proportion of
+items where the change resulted in a real word (semantic errors); and
+classify the most common phonological errors.
 
-\hypertarget{participants}{%
-\subsubsection{Participants}\label{participants}}
+\subsubsection{Participants}\label{participants}
 
-This study was approved as part of a larger research effort by Marisa Casillas, with the research protocol having been approved by the IRB committee entitled The Ethics Committee of the Faculty of Social Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) of Radboud University, Nijmegen, The Netherlands.
-The approval for work outside of the Netherlands came as an amendment to the general research line on language development research (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
-Participation was voluntary, with children being invited to come and participate.
-Regardless of how they performed, children were provided with a snack as compensation.
-Children who came up to participate but then refused were nonetheless provided with the snack.
+This study was approved as part of a larger research effort by Marisa
+Casillas, with the research protocol having been approved by the IRB
+committee entitled The Ethics Committee of the Faculty of Social
+Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen;
+ECSW) of Radboud University, Nijmegen, The Netherlands. The approval for
+work outside of the Netherlands came as an amendment to the general
+research line on language development research (original request:
+ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+Participation was voluntary, with children being invited to come and
+participate. Regardless of how they performed, children were provided
+with a snack as compensation. Children who came up to participate but
+then refused were nonetheless provided with the snack.
 
 A total 55 children were tested, from 34 of families, in five hamlets.
-Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=0), spoke too softly to allow offline coding (N=5).
-In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
-The remaining 40 children (14 girls) were aged 6.96 years (range 3.90-11 years).
-There were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and , 2 for whom this information was missing.
-Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Education is often reported in even years because people typically complete two-year cycles.} In terms of birth order, 5 were first borns, 4 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
+Some children could not be included for the following reasons: refused
+participation or failed to repeat items presented over headphones even
+after coaching (N=0), spoke too softly to allow offline coding (N=5). In
+addition, 2 teenagers were tested to put younger children at ease; their
+data is not included in analyses below. The remaining 40 children (14
+girls) were aged 6.96 years (range 3.90-11 years). There were 32
+children exposed only to Yélî in the home, 6 children who were also
+exposed to another language in the home, and , 2 for whom this
+information was missing. Maternal years of education averaged 8.24 years
+(range 6-12 years; 2 children had this information missing).\footnote{Education
+  is often reported in even years because people typically complete
+  two-year cycles.} In terms of birth order, 5 were first borns, 4
+second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this
+information.
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
+\subsection{Results}\label{results}
 
-\hypertarget{preliminary-analyses}{%
-\subsubsection{Preliminary analyses}\label{preliminary-analyses}}
+\subsubsection{Preliminary analyses}\label{preliminary-analyses}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig1-first_vs_others-1.pdf}
-\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual participants averaging separately their first attempts and all other attempts.}
+\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual
+participants averaging separately their first attempts and all other
+attempts.}
 \end{figure}
 
-We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
-We excluded 3 children who did not have data for one of these two types.
-As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
-Surprisingly, subsequent repetitions (M = 39, SD = 26)
-were on average less accurate than first ones (M = 65, SD = 15), t(38) = 6.62, p = 0.
-Given the uncertainty in whether previous work used only the first or all repetitions, and since behavior degraded and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
+We first checked whether accuracy varies between first and subsequent
+presentations of an item by averaging word-level accuracy at the
+participant level separately for first attempts and subsequent
+repetitions. We excluded 3 children who did not have data for one of
+these two types. As shown in Figure 1, participants' mean word-level
+accuracy became more heterogeneous in subsequent repetitions.
+Surprisingly, subsequent repetitions (M = 39, SD = 26) were on average
+less accurate than first ones (M = 65, SD = 15), t(38) = 6.62, p = 0.
+Given the uncertainty in whether previous work used only the first or
+all repetitions, and since behavior degraded and became more
+heterogeneous in subsequent repetitions, the rest of the analyses focus
+on only the first repetitions.
 
-Taking into account only the first attempts, we averaged attempts by each of the 24 children who had data for first attempts; their ages ranged from 3.90 to 10.20 (M = 6.50, SD = 1.50).
+Taking into account only the first attempts, we averaged attempts by
+each of the 24 children who had data for first attempts; their ages
+ranged from 3.90 to 10.20 (M = 6.50, SD = 1.50).
 
-The overall NWR score was M = 61\% (SD = 15\%).
-Scores based on phonemes are even higher M = 77\% (SD = 9\%).
-The phoneme-based normalized Levenshtein distance was M = 23\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
-Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
+The overall NWR score was M = 61\% (SD = 15\%). Scores based on phonemes
+are even higher M = 77\% (SD = 9\%). The phoneme-based normalized
+Levenshtein distance was M = 23\% (SD = 9\%), meaning that about a fifth
+of phonemes were were substituted, inserted, or deleted. Notice that the
+normalized Levenshtein distance is the complement of phoneme-based
+scores.
 
-\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
-\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}
+\subsubsection{NWR as a function of cross-linguistic phone
+frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}
 
-In this analysis, we were interested in variation in performance as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
-To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID \& target ID as random variables.
+In this analysis, we were interested in variation in performance as a
+function of the average frequency with which sounds composing individual
+target words are found in languages over the world. To look at this, we
+fit a mixed logistic regression, in which the outcome variable was
+whether the non-word was correctly repeated or not. The fixed effects of
+interest was the average cross-linguistic phone frequency; we also
+included child age as a control fixed effect, and child ID \& target ID
+as random variables.
 
-We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.03, SE ß = 0.01, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+We could include 885 observations (\textbf{why -- shouldn't it be
+1k??}), from 41 children producing in any given trial one of 41
+potential target words. The analysis revealed a significant estimate for
+the average cross-linguistic frequency of phones in the target words (ß
+= 0.03, SE ß = 0.01, p = 0): Target words with phones found more
+frequently across languages had a higher proportion of words that were
+correctly repeated, as clear in Figure 2.
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig2-xling_freq-1.pdf}
-\caption{(\#fig:Fig2-xling\_freq)NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}
+\caption{(\#fig:Fig2-xling\_freq)NWR scores for individual target words
+as a function of the average frequency with which each phone is found
+across languages.}
 \end{figure}
@@
 
-\textbf{a2m use first attempts rather than the whole data set: having second thoughts.
-To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers \& it is more consistent.
-BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros\ldots{} CF Option 3 further down}
+\textbf{a2m use first attempts rather than the whole data set: having
+second thoughts. To be clear: the argument for using first attempts only
+is that all of the other analyses are on kids' first attempts, rather
+than all of their attempts, so we should use just first attempts here
+because it may confuse readers \& it is more consistent. BUT there are
+fewer mistakes this way, which is Options 1 and 2 below has lots of
+zeros\ldots{} CF Option 3 further down}
 
-\textbf{a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?}
+\textbf{a2m probably turn the paragraph on substitutions into a table:
+having second thoughts, the table would look just like a para, no?}
 
-\textbf{Option 1: just numbers, first attempts only }
-Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 19 vowels deleted, and 4 consonants.
+\textbf{Option 1: just numbers, first attempts only } Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 19 vowels deleted, and 4 consonants.
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels, 0 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (42 oral vowels produced with a different quality; 0 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0) than vice versa (0).
-Simple consonants were mispronounced as other simple consonants quite frequently (62 simple consonants mispronounced as other simple ones, compared to 0 complex ones).
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (0 oral target vowels produced as nasal
+vowels, 0 nasal target vowels produced as oral vowels). Substitutions in
+which the oral nature was preserved but the quality of the vowel was
+changed were a great deal more common than changes in quality among
+nasal vowels (42 oral vowels produced with a different quality; 0 nasal
+vowels produced with a different quality). As for consonants,
+asymmetries were very marked with more complex consonants (specifically
+dptpkpkmknmbghlv) mispronounced as simple ones (specifically
+mnlwyvdgptkfhch, 0) than vice versa (0). Simple consonants were
+mispronounced as other simple consonants quite frequently (62 simple
+consonants mispronounced as other simple ones, compared to 0 complex
+ones).
 
-\textbf{Option 2: numbers and proportions, first attempts only}
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 19 vowels deleted (about 1.17\% of all vowel targets), and 4 consonants deleted (about 0.24\% of all consonant targets).
+\textbf{Option 2: numbers and proportions, first attempts only} Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 19 vowels deleted (about 1.17\% of all vowel
+targets), and 4 consonants deleted (about 0.24\% of all consonant
+targets).
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about 0\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (42 oral vowels produced with a different quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times or about NA\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (62 simple consonants mispronounced as other simple ones or about 0.30\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (0 oral target vowels produced as nasal
+vowels or about 0\% out of all oral vowel targets, 0 nasal target vowels
+produced as oral vowels or about NA\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of
+the vowel was changed were a great deal more common than changes in
+quality among nasal vowels (42 oral vowels produced with a different
+quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels
+produced with a different quality or about NA\% out of all nasal vowel
+targets). As for consonants, asymmetries were very marked with more
+complex consonants (specifically dptpkpkmknmbghlv) mispronounced as
+simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of
+all complex consonant targets) than vice versa (0 times or about NA\%
+out of all simple consonant targets). Simple consonants were
+mispronounced as other simple consonants quite frequently (62 simple
+consonants mispronounced as other simple ones or about 0.30\% out of all
+simple consonant targets, compared to 0 complex ones or about NA\% out
+of all complex consonant targets).
 
-\textbf{Option 3: numbers and proportions, ALL attempts}
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 29 vowels deleted (about 0.55\% of all vowel targets), and 9 consonants deleted (about 0.26\% of all consonant targets).
+\textbf{Option 3: numbers and proportions, ALL attempts} Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 29 vowels deleted (about 0.55\% of all vowel
+targets), and 9 consonants deleted (about 0.26\% of all consonant
+targets).
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels or about 2.01\% out of all oral vowel targets, 58 nasal target vowels produced as oral vowels or about 6.46\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (197 oral vowels produced with a different quality or about 3.20\% out of all oral vowel targets; 23 nasal vowels produced with a different quality or about 7.08\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 266 times or about 5.55\% out of all complex consonant targets) than vice versa (2 times about 5.55\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (128 simple consonants mispronounced as other simple ones or about 0.48\% out of all simple consonant targets, compared to 62 complex ones or about 2.79\% out of all complex consonant targets).
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (52 oral target vowels produced as nasal
+vowels or about 2.01\% out of all oral vowel targets, 58 nasal target
+vowels produced as oral vowels or about 6.46\% out of all nasal vowel
+targets). Substitutions in which the oral nature was preserved but the
+quality of the vowel was changed were a great deal more common than
+changes in quality among nasal vowels (197 oral vowels produced with a
+different quality or about 3.20\% out of all oral vowel targets; 23
+nasal vowels produced with a different quality or about 7.08\% out of
+all nasal vowel targets). As for consonants, asymmetries were very
+marked with more complex consonants (specifically dptpkpkmknmbghlv)
+mispronounced as simple ones (specifically mnlwyvdgptkfhch, 266 times or
+about 5.55\% out of all complex consonant targets) than vice versa (2
+times about 5.55\% out of all simple consonant targets). Simple
+consonants were mispronounced as other simple consonants quite
+frequently (128 simple consonants mispronounced as other simple ones or
+about 0.48\% out of all simple consonant targets, compared to 62 complex
+ones or about 2.79\% out of all complex consonant targets).
 
-Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 97\%.
+Finally, we looked into the frequency with which mispronunciations
+resulted in real words. Nearly all of them were: 97\%.
 
-\textbf{TO DISCUSS TOGETHER this is higher than any previous report! can we be sure of it?}
+\textbf{TO DISCUSS TOGETHER this is higher than any previous report! can
+we be sure of it?}
 
-\hypertarget{accuracy-a-function-of-item-length}{%
-\subsubsection{Accuracy a function of item length}\label{accuracy-a-function-of-item-length}}
+\subsubsection{Accuracy a function of item
+length}\label{accuracy-a-function-of-item-length}
 
 Next, we inspected whether accuracy varied as a function of word length.
-Results are shown on table XX.
-This table shows that monosyllables accuracy was much lower than other lengths.
-This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
+Results are shown on table XX. This table shows that monosyllables
+accuracy was much lower than other lengths. This is likely due to the
+fact that the majority of monosyllables included were chosen because
+they had sounds that are rare in the world's languages, which may
+indicate that they are hard to produce or to perceive.
 
-Setting monosyllables aside, we observe the typical pattern of decreased accuracy for longer items, although this is particularly salient for the whole word scoring.
-This is the most commonly reported type of score, but it is also the least forgiving.
-The pattern is less marked when other two scores are used, which are less sensitive to errors.
+Setting monosyllables aside, we observe the typical pattern of decreased
+accuracy for longer items, although this is particularly salient for the
+whole word scoring. This is the most commonly reported type of score,
+but it is also the least forgiving. The pattern is less marked when
+other two scores are used, which are less sensitive to errors.
 
-\begin{table}[tbp]
+\begin{verbatim}
+## 
+## 
+## \begin{table}[tbp]
+## \begin{center}
+## \begin{threeparttable}
+## \caption{\label{tab:tablength}NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
+## \begin{tabular}{lll}
+## \toprule
+## Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
+## \midrule
+## 47 (22) & 59 (17) & 41 (17)\\
+## 79 (22) & 94 (7) & 6 (7)\\
+## 78 (19) & 94 (6) & 6 (6)\\
+## 75 (32) & 93 (11) & 7 (11)\\
+## \bottomrule
+## \end{tabular}
+## \end{threeparttable}
+## \end{center}
+## \end{table}
+\end{verbatim}
 
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:tablength}NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
-
-\begin{tabular}{lll}
-\toprule
-Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
-\midrule
-47 (22) & 59 (17) & 41 (17)\\
-79 (22) & 94 (7) & 6 (7)\\
-78 (19) & 94 (6) & 6 (6)\\
-75 (32) & 93 (11) & 7 (11)\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-\hypertarget{factor-structuring-individual-variation}{%
-\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}
+\subsubsection{Factor structuring individual
+variation}\label{factor-structuring-individual-variation}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig2-scores by age-1.pdf}
-\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual participants as a function of age and sex (blue = boys, pink = girls).}
+\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual
+participants as a function of age and sex (blue = boys, pink = girls).}
 \end{figure}
 
-Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,548.04) = 0.33, p = 0.12).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.18) = 21.34, p = 0.86), birth order (data missing for 17 children, 2,678.47) = -0.16, p = 0.44) or maternal education (data missing for 2 children, 2,091.92) = 0.09, p = 0.67)
+Our final exploratory analysis assessed whether variation in scores was
+structured by factors that vary across individuals. As shown in Figure
+2, there was a greater deal of variance in earlier than later ages, with
+significantly higher NWR scores for older children (Spearman's rank
+correlation, given inequality of variance, rho (1,548.04) = 0.33, p =
+0.12). In contrast, there was no clear association between NWR scores,
+on the one hand, and sex (t (0.18) = 21.34, p = 0.86), birth order (data
+missing for 17 children, 2,678.47) = -0.16, p = 0.44) or maternal
+education (data missing for 2 children, 2,091.92) = 0.09, p = 0.67)
 
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+\subsection{Discussion}\label{discussion}
 
@@
 \item
-  What is the overall repetition accuracy (whole word, phoneme based, distance)?
+  What is the overall repetition accuracy (whole word, phoneme based,
+  distance)?
 \item
-  How does this change as a function of item complexity (number of syllables, sound complexity)?
+  How does this change as a function of item complexity (number of
+  syllables, sound complexity)?
 \end{itemize}
@@
 \item
-  The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
-\textbf{Check for work on European languages that may have looked into this}
+  The length distribution in Yélî words is more balanced than that in
+  English, and thus the performance decline for poly- versus
+  mono-syllables may be less pronounced than that for English.
+  \textbf{Check for work on European languages that may have looked into
+  this}
 \item
-  Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds {[}CITE{]}.
-Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
+  Similarly, we do not know of NWR research that manipulates the
+  difficulty of the sounds that are included in the items, but word
+  naming and other research suggests that children are more accurate
+  when producing easy and/or typologically common sounds than difficult
+  and/or typologically rare sounds {[}CITE{]}. Therefore, we expect
+  higher accuracy for items with common sounds than in those with rare
+  sounds.
 \item
-  The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
-Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
-As a result, differences between easier and harder items may be smaller in this work than in other research.
-\textbf{no work looking at consonants \& vowels?
-no work looking at nasal vowels in particular?}
+  The Yélî sound inventory is very large and compressed, with many
+  similar sounds that are acoustic and articulatory neighbors.
+  Therefore, this may constitute a pressure for children to have finer
+  auditory skills (and perhaps more precise articulations) than children
+  speaking languages with a simpler inventory. As a result, differences
+  between easier and harder items may be smaller in this work than in
+  other research. \textbf{no work looking at consonants \& vowels? no
+  work looking at nasal vowels in particular?}
 \end{enumerate}
 
-\textbf{(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)}
+\textbf{(MC: but we can try and do a cursory analysis based on the
+corpora we have from Steve and my transcription of naturalistic
+interactions!)}
 
@@
 \item
-  How frequent are errors that result in real words? Is that a function of item complexity?
+  How frequent are errors that result in real words? Is that a function
+  of item complexity?
 \item
-  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
+  Is individual variation explainable by child age, sex, birth order,
+  monolingual status, and/or parental education?
 \end{itemize}
@@
 \item
-  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
+  Non-monolingual Yélî children are less accurate than monolingual ones
+  when tested on the society-dominant language (we did not test any
+  non-dominant language) 
 \item
-  As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
+  As revious NWR evidence on this is mixed, but general findings on
+  language development suggest that children whose mothers are more
+  educated are more accurate than children whose mothers are less
+  educated.
 \item
-  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
+  To our knowledge, there is no previous NWR work on this, but other
+  research suggests that first-born children should outperform
+  later-born children
 \item
-  Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family
-  (or at least from people other than their mothers, who tend to be the non-native speakers).
-If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
-  . That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother
-  , so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
+  Anecdotally Yélî children grow up in close-knitted communities and
+  thus may receive significant portions of their language input from
+  people not in their nuclear family (or at least from people other than
+  their mothers, who tend to be the non-native speakers). If so, the
+  difference between monolinguals and not monolinguals may be smaller
+  than that found in other work . That said, one recent study on the
+  same population shows that most child-directed input in the first 2
+  years does come from the mother , so in so far as this input has a
+  crucial formational role, then there may still be a performance gap
+  between these two groups.
 \item
-  In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
-This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
-However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
+  In the Rossel community, formal education plays an extremely minor
+  role in ensuring individual's success, is not a good index of relative
+  socio-economic status, and furthermore there is only a narrow range of
+  variation in maternal educational attainment. This may lead to no or
+  only very small advantages for children whose mothers are more
+  educated, provided that the causal chain between maternal education
+  and child language is via SES more broadly. However, if education
+  directly boosts maternal verbal skills and the incidence of verbal
+  behavior (as suggested by CITE), then we should still see a difference
+  along this factor.
 \item
-  One main causal path between birth order and language development is via parental input (CITE).
-Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
+  One main causal path between birth order and language development is
+  via parental input (CITE). Given our arguments above for how mothers
+  may not be as important among Rossel people than in other places, then
+  the performance gap between first borns and later borns may be
+  smaller.
 \end{enumerate}
@@
 
-\hypertarget{acknowledgments}{%
-\subsection{Acknowledgments}\label{acknowledgments}}
+\subsection{Acknowledgments}\label{acknowledgments}
 
-We are grateful to informants and individuals who participated in the study.
-AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.
-S.
-McDonnell Foundation Understanding Human Cognition Scholar Award.
-MC blabla NWO Veni Innovational Scheme Grant (XXX-XX-XXX).
+We are grateful to the individuals who participated in the study. The
+collection and annotation of these recordings was made possible by
+Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also
+to the PNG National Research Institute, and the Administration of Milne
+Bay Province AC acknowledges financial and institutional support from
+Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge,
+ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and
+the J. S. McDonnell Foundation Understanding Human Cognition Scholar
+Award. MC acknowledges financial support from an NWO Veni Innovational
+Scheme grant (275-89-033).
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+\section{References}\label{references}
 
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
+\setlength{\parindent}{-0.5in} \setlength{\leftskip}{0.5in}
 
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-is08042009language}{}%
-Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+\hypertarget{ref-is08042009language}{}
+Action, C. (2009). Language impairment in a multilingual society:
+Linguistic patterns and the road to assessment. \emph{Brussels: COST
+Office. Available Online at: Http://Www. Bi-Sli. Org}.
 
-\leavevmode\hypertarget{ref-balladares2016socio}{}%
-Balladares, J., Marshall, C., \& Griffiths, Y.
-(2016).
-Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
-\emph{First Language}, \emph{36}(3), 338--351.
-\url{https://doi.org/10.1177/0142723715626067}
+\hypertarget{ref-balladares2016socio}{}
+Balladares, J., Marshall, C., \& Griffiths, Y. (2016). Socio-economic
+status affects sentence repetition, but not non-word repetition, in
+Chilean preschoolers. \emph{First Language}, \emph{36}(3), 338--351.
+doi:\href{https://doi.org/10.1177/0142723715626067}{10.1177/0142723715626067}
 
-\leavevmode\hypertarget{ref-bowey2001nonword}{}%
-Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
+\hypertarget{ref-bowey2001nonword}{}
+Bowey, J. A. (2001). Nonword repetition and young children's receptive
+vocabulary: A longitudinal study. \emph{Applied Psycholinguistics},
+\emph{22}(3), 441--469.
 
-\leavevmode\hypertarget{ref-brandeker2015language}{}%
-Brandeker, M., \& Thordardottir, E.
-(2015).
-Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
-\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
+\hypertarget{ref-brandeker2015language}{}
+Brandeker, M., \& Thordardottir, E. (2015). Language exposure in
+bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2),
+126--138.
 
-\leavevmode\hypertarget{ref-chiat2007preschool}{}%
-Chiat, S., \& Roy, P.
-(2007).
-The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
+\hypertarget{ref-chiat2007preschool}{}
+Chiat, S., \& Roy, P. (2007). The preschool repetition test: An
+evaluation of performance in typically developing and clinically
+referred children. \emph{Journal of Speech, Language, and Hearing
+Research}, \emph{50}(2), 429--443.
 
-\leavevmode\hypertarget{ref-cristia2020infant}{}%
+\hypertarget{ref-cristia2020infant}{}
 Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
-(2020).
-Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
+(2020). Infant-directed input and literacy effects on phonological
+processing: Non-word repetition scores among the tsimane'.
 \emph{Preprint}.
 
-\leavevmode\hypertarget{ref-estes2007differences}{}%
-Estes, K.
-G., Evans, J.
-L., \& Else-Quest, N.
-M.
-(2007).
-Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
+\hypertarget{ref-estes2007differences}{}
+Estes, K. G., Evans, J. L., \& Else-Quest, N. M. (2007). Differences in
+the nonword repetition performance of children with and without specific
+language impairment: A meta-analysis. \emph{Journal of Speech, Language,
+and Hearing Research}, \emph{50}(1), 177--195.
 
-\leavevmode\hypertarget{ref-farmani2018normalization}{}%
-Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
-Z., Talebi, E., \& Shourvazi, Z.
-(2018).
-Normalization of the non-word repetition test in farsi-speaking children.
-\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
+\hypertarget{ref-farmani2018normalization}{}
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \&
+Shourvazi, Z. (2018). Normalization of the non-word repetition test in
+farsi-speaking children. \emph{Journal of Modern Rehabilitation},
+\emph{12}(4), 217--224.
 
-\leavevmode\hypertarget{ref-gallagher2014identity}{}%
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
+\hypertarget{ref-gallagher2014identity}{}
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from
+Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378.
+doi:\href{https://doi.org/10.1515/lp-2014-0012}{10.1515/lp-2014-0012}
 
-\leavevmode\hypertarget{ref-gallon2007non}{}%
-Gallon, N., Harris, J., \& Van der Lely, H.
-(2007).
-Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
-\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+\hypertarget{ref-gallon2007non}{}
+Gallon, N., Harris, J., \& Van der Lely, H. (2007). Non-word repetition:
+An investigation of phonological complexity in children with grammatical
+sli. \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
 
-\leavevmode\hypertarget{ref-havron2019effect}{}%
-Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
-M.-C.
-C.
-S.
-(2019).
-The effect of older siblings on language development as a function of age difference and sex.
-\emph{Psychological Science}, \emph{30}(9), 1333--1343.
+\hypertarget{ref-havron2019effect}{}
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \&
+Group, E. M.-C. C. S. (2019). The effect of older siblings on language
+development as a function of age difference and sex. \emph{Psychological
+Science}, \emph{30}(9), 1333--1343.
 
-\leavevmode\hypertarget{ref-hazan2000development}{}%
-Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
+\hypertarget{ref-hazan2000development}{}
+Hazan, V., \& Barrett, S. (2000). The development of phonemic
+categorization in children aged 6--12. \emph{Journal of Phonetics},
+\emph{28}(4), 377--396.
 
-\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
+\hypertarget{ref-jabere2018xperiment}{}
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual
+Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4),
+317--334.
+doi:\href{https://doi.org/10.30958/ajp.5-4-4}{10.30958/ajp.5-4-4}
 
-\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
-(2014).
-Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
+\hypertarget{ref-kalnak2014nonword}{}
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014).
+Nonword repetition--a clinical marker for specific language impairment
+in swedish associated with parents' language-related problems.
 \emph{PloS One}, \emph{9}(2), e89544.
 
-\leavevmode\hypertarget{ref-meir2017independent}{}%
-Meir, N., \& Armon-Lotem, S.
-(2017).
-Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
-\emph{Frontiers in Psychology}, \emph{8}, 1442.
+\hypertarget{ref-meir2017independent}{}
+Meir, N., \& Armon-Lotem, S. (2017). Independent and combined effects of
+socioeconomic status (ses) and bilingualism on children's vocabulary and
+verbal short-term memory. \emph{Frontiers in Psychology}, \emph{8},
+1442.
 
-\leavevmode\hypertarget{ref-meir2016disentangling}{}%
-Meir, N., Walters, J., \& Armon-Lotem, S.
-(2016).
-Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
-\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
+\hypertarget{ref-meir2016disentangling}{}
+Meir, N., Walters, J., \& Armon-Lotem, S. (2016). Disentangling sli and
+bilingualism using sentence repetition tasks: The impact of l1 and l2
+properties. \emph{International Journal of Bilingualism}, \emph{20}(4),
+421--452.
 
-\leavevmode\hypertarget{ref-phoible}{}%
-Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
+\hypertarget{ref-phoible}{}
+Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max
+Planck Institute for the Science of Human History. Retrieved from
+\url{https://phoible.org/}
 
-\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
-Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
-(2019).
-An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
-\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
+\hypertarget{ref-piazzalunga2019articulatory}{}
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler,
+A. (2019). An articulatory-based disyllabic and trisyllabic non-word
+repetition test: Reliability and validity in italian 3-to 7-year-old
+children. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5),
+437--456.
 
-\leavevmode\hypertarget{ref-torrington2015non}{}%
-Torrington Eaton, C., Newman, R.
-S., Ratner, N.
-B., \& Rowe, M.
-L.
-(2015).
-Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
-\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+\hypertarget{ref-torrington2015non}{}
+Torrington Eaton, C., Newman, R. S., Ratner, N. B., \& Rowe, M. L.
+(2015). Non-word repetition in 2-year-olds: Replication of an adapted
+paradigm and a useful methodological extension. \emph{Clinical
+Linguistics \& Phonetics}, \emph{29}(7), 523--535.
 
-\leavevmode\hypertarget{ref-vance2005speech}{}%
-Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
+\hypertarget{ref-vance2005speech}{}
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills
+in children aged 3--7 years. \emph{International Journal of Language \&
+Communication Disorders}, \emph{40}(1), 29--48.
 
-\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
-Wilsenach, C.
-(2013).
-Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
-\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
-\url{https://doi.org/10.5785/29-2-554}
+\hypertarget{ref-wilsenach2013phonological}{}
+Wilsenach, C. (2013). Phonological skills as predictor of reading
+success: An investigation of emergent bilingual Northern Sotho/English
+learners. \emph{Per Linguam: a Journal of Language Learning= Per
+Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+doi:\href{https://doi.org/10.5785/29-2-554}{10.5785/29-2-554}
+
+\hypertarget{ref-is08042009language}{}
+Action, C. (2009). Language impairment in a multilingual society:
+Linguistic patterns and the road to assessment. \emph{Brussels: COST
+Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
+\hypertarget{ref-balladares2016socio}{}
+Balladares, J., Marshall, C., \& Griffiths, Y. (2016). Socio-economic
+status affects sentence repetition, but not non-word repetition, in
+Chilean preschoolers. \emph{First Language}, \emph{36}(3), 338--351.
+doi:\href{https://doi.org/10.1177/0142723715626067}{10.1177/0142723715626067}
+
+\hypertarget{ref-bowey2001nonword}{}
+Bowey, J. A. (2001). Nonword repetition and young children's receptive
+vocabulary: A longitudinal study. \emph{Applied Psycholinguistics},
+\emph{22}(3), 441--469.
+
+\hypertarget{ref-brandeker2015language}{}
+Brandeker, M., \& Thordardottir, E. (2015). Language exposure in
+bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2),
+126--138.
+
+\hypertarget{ref-chiat2007preschool}{}
+Chiat, S., \& Roy, P. (2007). The preschool repetition test: An
+evaluation of performance in typically developing and clinically
+referred children. \emph{Journal of Speech, Language, and Hearing
+Research}, \emph{50}(2), 429--443.
+
+\hypertarget{ref-cristia2020infant}{}
+Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
+(2020). Infant-directed input and literacy effects on phonological
+processing: Non-word repetition scores among the tsimane'.
+\emph{Preprint}.
+
+\hypertarget{ref-estes2007differences}{}
+Estes, K. G., Evans, J. L., \& Else-Quest, N. M. (2007). Differences in
+the nonword repetition performance of children with and without specific
+language impairment: A meta-analysis. \emph{Journal of Speech, Language,
+and Hearing Research}, \emph{50}(1), 177--195.
+
+\hypertarget{ref-farmani2018normalization}{}
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \&
+Shourvazi, Z. (2018). Normalization of the non-word repetition test in
+farsi-speaking children. \emph{Journal of Modern Rehabilitation},
+\emph{12}(4), 217--224.
+
+\hypertarget{ref-gallagher2014identity}{}
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from
+Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378.
+doi:\href{https://doi.org/10.1515/lp-2014-0012}{10.1515/lp-2014-0012}
+
+\hypertarget{ref-gallon2007non}{}
+Gallon, N., Harris, J., \& Van der Lely, H. (2007). Non-word repetition:
+An investigation of phonological complexity in children with grammatical
+sli. \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\hypertarget{ref-havron2019effect}{}
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \&
+Group, E. M.-C. C. S. (2019). The effect of older siblings on language
+development as a function of age difference and sex. \emph{Psychological
+Science}, \emph{30}(9), 1333--1343.
+
+\hypertarget{ref-hazan2000development}{}
+Hazan, V., \& Barrett, S. (2000). The development of phonemic
+categorization in children aged 6--12. \emph{Journal of Phonetics},
+\emph{28}(4), 377--396.
+
+\hypertarget{ref-jabere2018xperiment}{}
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual
+Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4),
+317--334.
+doi:\href{https://doi.org/10.30958/ajp.5-4-4}{10.30958/ajp.5-4-4}
+
+\hypertarget{ref-kalnak2014nonword}{}
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014).
+Nonword repetition--a clinical marker for specific language impairment
+in swedish associated with parents' language-related problems.
+\emph{PloS One}, \emph{9}(2), e89544.
+
+\hypertarget{ref-meir2017independent}{}
+Meir, N., \& Armon-Lotem, S. (2017). Independent and combined effects of
+socioeconomic status (ses) and bilingualism on children's vocabulary and
+verbal short-term memory. \emph{Frontiers in Psychology}, \emph{8},
+1442.
+
+\hypertarget{ref-meir2016disentangling}{}
+Meir, N., Walters, J., \& Armon-Lotem, S. (2016). Disentangling sli and
+bilingualism using sentence repetition tasks: The impact of l1 and l2
+properties. \emph{International Journal of Bilingualism}, \emph{20}(4),
+421--452.
+
+\hypertarget{ref-phoible}{}
+Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max
+Planck Institute for the Science of Human History. Retrieved from
+\url{https://phoible.org/}
+
+\hypertarget{ref-piazzalunga2019articulatory}{}
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler,
+A. (2019). An articulatory-based disyllabic and trisyllabic non-word
+repetition test: Reliability and validity in italian 3-to 7-year-old
+children. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5),
+437--456.
+
+\hypertarget{ref-torrington2015non}{}
+Torrington Eaton, C., Newman, R. S., Ratner, N. B., \& Rowe, M. L.
+(2015). Non-word repetition in 2-year-olds: Replication of an adapted
+paradigm and a useful methodological extension. \emph{Clinical
+Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+
+\hypertarget{ref-vance2005speech}{}
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills
+in children aged 3--7 years. \emph{International Journal of Language \&
+Communication Disorders}, \emph{40}(1), 29--48.
+
+\hypertarget{ref-wilsenach2013phonological}{}
+Wilsenach, C. (2013). Phonological skills as predictor of reading
+success: An investigation of emergent bilingual Northern Sotho/English
+learners. \emph{Per Linguam: a Journal of Language Learning= Per
+Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+doi:\href{https://doi.org/10.5785/29-2-554}{10.5785/29-2-554}
 

### 2020-09-08 manuscript.Rmd
begin adding sound system info

@@
 
-### Intro to the language (**mc**) - please feel free to throw away anything that is not useful!
+### The Yélî language
+<!--M2A: is this okay?-->
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+Most speakers of the language grow up speaking it monolingually within their swidden horticulturalist communities until they begin attending school around the age of 7; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands [@brownIPchapter], the same region in which the current data were collected.
+Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
 
-- complexity in the vowel system
-- complexity in the consonant system
+The Yélî sound system, much like its baroque grammatical system in general [@levinsoninprepydgrammar], is unlike any other in the region, with 90 contrastive segments, including at least two contrasts not yet been documented elsewhere (labial-coronal double-articulations with dental vs.
+post-alveolar coronal placement in both oral and nasal stops) [@levinsoninprepydgrammar].
+With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
+The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these able to be pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+Finally, a number of these can also be labialized or palatalized on the release, or both (/pʲ/, /tʃ/, /tpʲ/, /ṭʲ/, /ṭpʲ/, /kʲ/, /kpʲ/, /pʲʷ/, /mbʲ/, /ndʒ/, /nmdbʲ/, /ṇḍʲ/, /ṇmḍbʲ/, /mbʷ/, /ŋɡʷ/, /mbʲʷ/, /ṭṇʲ/, /ṭpṇmʲ/, /kŋʷ/, /mʲ/, /nʲ/, /nmʲ/, /ṇʲ/, /ṇmʲ/, /mʷ/, /ŋʷ/, mʲʷ/).
+Among the remaining consonants are small group of oral continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+Vowels in Yélî Dnye may be oral or nasal, short or long.
+The 10 oral vowel qualities (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
+In total, Levinson [-@levinsoninprepydgrammar] then counts 90 distinctive segments in Yélî Dnye (93 when including consonants that are extremely rarely used).
+
+To do:
 - word shapes
 - typical word length
-- although not the focus of this paper, high use of suppletion in verbal paradigms, other features of language, see Levinson XXX for details
 

### 2020-09-08 manuscript.Rmd
test IPA font with linguisticsdown

@@
     latex_engine: xelatex
+    includes:
+      in_header: header.tex
 ---
@@
 Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
+The 10 oral vowel qualities (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /`r linguisticsdown::cond_cmpl("ɔ")`/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /`r linguisticsdown::cond_cmpl("ɔ")`/, /æ/, /ɑ/).
 In total, Levinson [-@levinsoninprepydgrammar] then counts 90 distinctive segments in Yélî Dnye (93 when including consonants that are extremely rarely used).

### 2020-09-08 manuscript.tex
test IPA font with linguisticsdown

@@
 
-\shorttitle{NWR in Yélî Dnye}
-\affiliation{
-\vspace{0.5cm}
-\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
-\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
-\usepackage{csquotes}
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-\usepackage{longtable}
-\usepackage{lscape}
-\usepackage{multirow}
-\usepackage{tabularx}
-\usepackage[flushleft]{threeparttable}
-\usepackage{threeparttablex}
-
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-\usepackage{lineno}
-
-\linenumbers
+\usepackage{fontspec}
+
+%% Pandoc can only set 10, 11, 12 pt
+%% uncomment below to set fontsize
+%\usepackage[fontsize=13pt]{scrextend}
+
+%\setmainfont{Calibri} % Set main font for latin characters
+
+\setlength{\parskip}{0.15cm} % Set space between paragraphs
+%\linespread{1.2}\selectfont % Set line height
+%\usepackage[doublespacing]{setspace} % Use double space without changing footnotes line height
+
+%% Special font for IPA
+%% Make sure "Doulos SIL" is installed on your computer
+%% For other typefaces supporting IPA symbols, see
+%% https://en.wikipedia.org/wiki/International_Phonetic_Alphabet#Typefaces
+\newfontfamily\ipa{Doulos SIL} % Font for IPA symbols
+\DeclareTextFontCommand{\ipatext}{\ipa}
+
+%%%         Section for CJK Characters                   %%%
+%%%   You may want to uncomment the code below if        %%%
+%%%   you're writing this document with CJK characters   %%%
+
+%\usepackage{xeCJK}  % Uncomment for using CJK characters
+%% Set main font for CJK characters
+%% Make sure your system has the font set
+%\setCJKmainfont[
+%	BoldFont={HanWangHeiHeavy}  % Set font for CJK boldface
+%    ]{標楷體}    % Set font for normal CJK
+%% Some Traditional Chinese fonts: AR PL KaitiM Big5, PingFang TC, Noto Sans CJK TC
+%\XeTeXlinebreaklocale "zh"
+%\XeTeXlinebreakskip = 0pt plus 1pt
 
@@
 \item
-  consent, \enquote{correct}, \enquote{accuracy}, \enquote{performance}
+  consent, ``correct'', ``accuracy'', ``performance''
 \item
@@
 
-\subsubsection{\texorpdfstring{Intro to the language (\textbf{mc}) -
-please feel free to throw away anything that is not
-useful!}{Intro to the language (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-language-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}
-
-\begin{itemize}
-\tightlist
-\item
-  complexity in the vowel system
-\item
-  complexity in the consonant system
-\item
-  word shapes
-\item
-  typical word length
-\item
-  although not the focus of this paper, high use of suppletion in verbal
-  paradigms, other features of language, see Levinson XXX for details
-\end{itemize}
+\subsubsection{The Yélî language}\label{the-yuxe9luxee-language}
+
+Yélî Dnye is an isolate language (presumed Papuan) spoken by
+approximately 7000 people residing on Rossel Island, which is found at
+the far end of the Louisiade Archipelago in Milne Bay Province, Papua
+New Guinea. Most speakers of the language grow up speaking it
+monolingually within their swidden horticulturalist communities until
+they begin attending school around the age of 7; school instruction is
+in English. While monolingual Yélî Dnye upbringing is common,
+multilingual families are not unusual, particularly in the region around
+the Catholic Mission (1953--present) where there appears to be a higher
+incidence of married-in mothers from other islands ({\textbf{???}}), the
+same region in which the current data were collected. Children in these
+multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
+and/or some other language(s) from the region.
+
+The Yélî sound system, much like its baroque grammatical system in
+general ({\textbf{???}}), is unlike any other in the region, with 90
+contrastive segments, including at least two contrasts not yet been
+documented elsewhere (labial-coronal double-articulations with dental
+vs.~post-alveolar coronal placement in both oral and nasal stops)
+({\textbf{???}}). With only four primary places of articulation
+(bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts,
+the phonological inventory is remarkably packed with acoustically
+similar segments. The core oral stop set includes both singleton (/p/,
+/t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments,
+with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and
+with a substantial portion of these able to be pre-nasalized or nasally
+released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/,
+/ṭpṇm/, /kpŋm/). Finally, a number of these can also be labialized or
+palatalized on the release, or both (/pʲ/, /tʃ/, /tpʲ/, /ṭʲ/, /ṭpʲ/,
+/kʲ/, /kpʲ/, /pʲʷ/, /mbʲ/, /ndʒ/, /nmdbʲ/, /ṇḍʲ/, /ṇmḍbʲ/, /mbʷ/, /ŋɡʷ/,
+/mbʲʷ/, /ṭṇʲ/, /ṭpṇmʲ/, /kŋʷ/, /mʲ/, /nʲ/, /nmʲ/, /ṇʲ/, /ṇmʲ/, /mʷ/,
+/ŋʷ/, mʲʷ/). Among the remaining consonants are small group of oral
+continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in Yélî Dnye
+may be oral or nasal, short or long. The 10 oral vowel qualities (/i/,
+/ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /\ipatext{ɔ}/, /æ/, /ɑ/) can be produced
+as short or long vowels, with 7 of these able to appear as short and
+long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /\ipatext{ɔ}/, /æ/, /ɑ/).
+In total, Levinson ({\textbf{???}}) then counts 90 distinctive segments
+in Yélî Dnye (93 when including consonants that are extremely rarely
+used).
+
+To do: - word shapes - typical word length
 
@@
 A first list of candidate items was generated in 2018 by selecting
-simple consonants (\enquote{p}, \enquote{t}, \enquote{d}, \enquote{k},
-\enquote{m}, \enquote{n}, \enquote{w}, \enquote{y}) and vowels
-(\enquote{i}, \enquote{o},\enquote{u}, \enquote{a},\enquote{e}) that
-were combined into consonant-vowel syllables, further sampling the space
-of 1- to 4-syllable sequences. These candidates were automatically
-checked against Levinson's 2015 dictionary and removed from
-consideration if they appeared in the dictionary. The second author
-presented them orally to three local research assistants, who were asked
-to repeat them and further say whether they were real words. Any item
-for which two or more of the assistants reported them having a meaning
-or some form of association was excluded.
+simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'',
+``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined
+into consonant-vowel syllables, further sampling the space of 1- to
+4-syllable sequences. These candidates were automatically checked
+against Levinson's 2015 dictionary and removed from consideration if
+they appeared in the dictionary. The second author presented them orally
+to three local research assistants, who were asked to repeat them and
+further say whether they were real words. Any item for which two or more
+of the assistants reported them having a meaning or some form of
+association was excluded.
 
@@
 Even so, there was one monosyllable that was often identified as a real
-word (intended \enquote{yî} /yXX/; identified as \enquote{yi} /yi/,
-\emph{tree}). This item is removed from analyses below.
+word (intended ``yî'' /yXX/; identified as ``yi'' /yi/, \emph{tree}).
+This item is removed from analyses below.
 
@@
 We visited four different hamlets once, and attempted to test all
-eligible children present at the time, to prevent the items
-\enquote{spreading} through hearsay. In the first hamlet, we tested
-children in five different places, with some children being tested
-inside a house and others tested on the veranda. The complete list of
-places and the ways in which they met the desiderata mentioned above can
-be found in the raw data, available from online supplementary materials.
+eligible children present at the time, to prevent the items `spreading'
+through hearsay. In the first hamlet, we tested children in five
+different places, with some children being tested inside a house and
+others tested on the veranda. The complete list of places and the ways
+in which they met the desiderata mentioned above can be found in the raw
+data, available from online supplementary materials.
 

### 2020-09-09 getting frequency corpus.Rmd
words extracted from raw YEL

---
title: "Get freqs for NWR YD project"
author: "AC"
date: "9/8/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initial clean up

This is a bit bludgeony, so replace if Marvin has something more elegant

```{bash}
grep "^[FM]A" txt/*.txt | # use only adult speech
cut -f 6 | #take transcription
sed "s/\[: /\[:_/g" |  sed "s/\[=! /\[:_/g" | sed "s/\[- /\[-_/g" | #remove spaces that are not real ones
grep -v "\[-_" | #remove sentences in Eng and Pidgin
 grep -v "he " | grep -vw "they" | grep -vw "I"  | grep -vw "you" | grep -vw "your"| grep -vw "his" | grep -vw "hey" | grep -vw "going" | grep -vw "people" | grep -vw "dogs" | grep -vw "puppies" | grep -vw "boys" | grep -vw "and"| grep -vw "container" | grep -vw "brother" | grep -vw "good"| grep -vw "where"| grep -vw "umbrella"| grep -vw "hello"| grep -vw "here" | grep -vw "this"| grep -vw "married" | grep -vw "here" | grep -vw "okay"| grep -vw "nice"| grep -vw "baked"| grep -vw "singsing"| grep -vw "cooked"| grep -vw "banana"| grep -vw "hurry"| grep -vw "fire"|#there were still some English sentences
sed 's/\[[^[]*\]//g' | #delete comments
tr ' ' '\n' | #cut at word boundaries
tr -d '?' | tr -d '.' | tr -d '!' | tr -d '-' | tr -d ',' | #clean up punctuation
grep -v "&" | grep -v "@s" | #remove switches
grep -v "xxx" | grep -vw "hm" | grep -vw "mm" | grep -vw "mmhm" | grep -vw "oh" | grep -vw "yeah" | grep -v "uhhuh" | grep -v "ho" | grep -v "mmmm" | grep -v "chuk"| grep -v "ha" |#final cleaning
tr -d '>' | tr -d '<' |tr -d '(' |tr -d ')' | sed "s/@c//g" |
grep -v "[A-Z]" | #get rid of all names
sed "s/che/te/g" |  sed "s/chi/ti/g"

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2020-09-09 getting frequency corpus.Rmd
words extracted from raw YEL

@@
 grep -v "[A-Z]" | #get rid of all names
-sed "s/che/te/g" |  sed "s/chi/ti/g"
+sed "s/che/te/g" |  sed "s/chi/ti/g" | sort > words_corpus.txt
 

### 2020-09-09 getting frequency corpus.Rmd
phone frequency extracted but not yet integrated

--- 2020-09-09 getting frequency corpus.Rmd
@@
 
-This is a bit bludgeony, so replace if Marvin has something more elegant
+This is a better than what we had in divime, because it does more cleanup. To run it, you need to be in raw_YEL
 
-```{bash}
+```{bash, eval=F}
 grep "^[FM]A" txt/*.txt | # use only adult speech
@@
 grep -v "\[-_" | #remove sentences in Eng and Pidgin
- grep -v "he " | grep -vw "they" | grep -vw "I"  | grep -vw "you" | grep -vw "your"| grep -vw "his" | grep -vw "hey" | grep -vw "going" | grep -vw "people" | grep -vw "dogs" | grep -vw "puppies" | grep -vw "boys" | grep -vw "and"| grep -vw "container" | grep -vw "brother" | grep -vw "good"| grep -vw "where"| grep -vw "umbrella"| grep -vw "hello"| grep -vw "here" | grep -vw "this"| grep -vw "married" | grep -vw "here" | grep -vw "okay"| grep -vw "nice"| grep -vw "baked"| grep -vw "singsing"| grep -vw "cooked"| grep -vw "banana"| grep -vw "hurry"| grep -vw "fire"|#there were still some English sentences
+ grep -v "he " | grep -vw "they" | grep -vw "I"  | grep -vw "you" | grep -vw "your"| grep -vw "his" | grep -vw "hey" | grep -vw "going" | grep -vw "people" | grep -vw "dogs" | grep -vw "puppies" | grep -vw "boys" | grep -vw "and"| grep -vw "container" | grep -vw "brother" | grep -vw "good"| grep -vw "where"| grep -vw "umbrella"| grep -vw "hello"| grep -vw "here" | grep -vw "this"| grep -vw "married" | grep -vw "here" | grep -vw "okay"| grep -vw "nice"| grep -vw "baked"| grep -vw "singsing"| grep -vw "cooked"| grep -vw "banana"| grep -vw "hurry"| grep -vw "fire"| grep -vw "about" | #there were still some English sentences
+grep -vw "alright" |grep -vw "already" |grep -vw "almost" |grep -vw "another" |
+grep -vw "bye" |grep -vw "calling" |grep -vw "coming" |grep -vw "clothes" |
+grep -vw "complete" |grep -vw "counting" |grep -vw "enough" |grep -vw "fight" |
+grep -vw "first" |grep -vw "finished" |grep -vw "there"|grep -vw "which" |grep -vw "witchcraft" |grep -vw "small"|grep -vw "right"|grep -vw "ready"|grep -vw "out"|grep -vw "our"|grep -vw "later"|grep -vw "leave"|grep -vw "eat"|grep -vw "everything"|grep -vw "books"|
+grep -vw "window" |grep -vw "want"|grep -vw "sister"|grep -vw "sleeping"|
+grep -vw "sometimes" | grep -vw "somehwere"|grep -vw "scared"|grep -vw "saw" |
+grep -vw "inside" | grep -vw "gonna" | grep -vw "her" | grep -vw "him"|
+grep -vw "here" | grep -vw "easy" | grep -vw "early" | grep -vw "eye"|
+grep -vw "cook" | grep -vw "dog" | grep -vw "area" | grep -vw "around"|
+grep -v "a@l" | grep -v "e@l" | grep -v "d@l"|
 sed 's/\[[^[]*\]//g' | #delete comments
@@
 grep -v "&" | grep -v "@s" | #remove switches
-grep -v "xxx" | grep -vw "hm" | grep -vw "mm" | grep -vw "mmhm" | grep -vw "oh" | grep -vw "yeah" | grep -v "uhhuh" | grep -v "ho" | grep -v "mmmm" | grep -v "chuk"| grep -v "ha" |#final cleaning
-tr -d '>' | tr -d '<' |tr -d '(' |tr -d ')' | sed "s/@c//g" |
-grep -v "[A-Z]" | #get rid of all names
-sed "s/che/te/g" |  sed "s/chi/ti/g"
+grep -v "xxx" | grep -vw "xx" | grep -vw "hm" | grep -vw "mm" | grep -vw "mmhm" | grep -vw "oh" | grep -vw "yeah" | grep -v "uhhuh" | grep -v "ho" | grep -v "mmmm" | grep -v "chuk"| grep -v "ha"| grep -v "ah" | grep -v "ehheh" | grep -v "hmmm" |#remove nonwords
+grep -v "[A-Z]" | grep -vw "Ńaamońo" | #get rid of all names
+sed "s/aa+/aa/g" | sed "s/ee+/ee/g" | sed "s/ii+/ii/g" | sed "s/oo+/oo/g" | sed "s/uu+/aa/g" |
+sed "s/êê+/êê/g" | sed "s/ââ+/ââ/g" | sed "s/áá+/áá/g" | sed "s/óó+/óó/g" | #exaggeration of vowel length
+sed "s/aaaa/aa/g" |
+tr -d '>' | tr -d '<' |tr -d '(' |tr -d ')' | sed "s/@c//g" | # final cleaning and write out
+sed "s/che/te/g" |  sed "s/chi/ti/g" | sort | grep -v "[0-9]" | sed '/^$/d' > words_corpus.txt
 
@@
 
-## Including Plots
+## Getting frequencies for segments in our stimuli
 
-You can also embed plots, for example:
+Middy gave me an onset list that converts all vowels and consonants to common representations for the purposes of counting syllables (rossel-ortho-replacements.txt).
+I thought of the following changes:
+
+- remove all rewrites for vowels
+- but that destroys the context for the following rules, eg not removing colon means knw does not find a match in knw:a 
+- but why weren't these defined with regular expressions anyway?
+- it looks like I should be careful with the 4- and 3-letter phonemes, because I do not have those in my rewrites but not otherwise
+
+So all things considered, it looks like it may be easier to just do the replacement of the 4- and 3-letter segments here, just being careful to map these to something that is not used in the stimuli, such as ngm.
+
+Also, this reveals my rewrites were incomplete, so I added some lines for the onsets that were missing in my initial correspondance list.
+
+The correspondances here looks similar to the one in wrangling but it has more entries because we want to separate each phoneme, and we want to distinguish between short and long
 
 ```{r pressure, echo=FALSE}
-plot(pressure)
+
+correspondances=matrix( #list correspondances always by pairs, orthography then phonology
+  c(
+    #nasal
+    ":ii",":ii ",
+    ":ee",":ee ",
+    ":aa",":aa ",
+    ":êê",":êê ",
+    ":ââ",":ââ ",
+    ":uu",":uu ",
+    ":oo",":oo ",
+
+    ":i",":i ",
+    ":e",":e ",
+    ":a",":a ",
+    ":ê",":ê ",
+    ":â",":â ",
+    ":u",":u ",
+    ":o",":o ",
+
+    #non-nasal
+    "îî","îî ",
+    "êê","êê ",
+    "ââ","ââ ",
+    "éé","éé ",
+    "áá","áá ",
+    "óó","óó ",
+
+    "aa","aa ",
+    "ee","ee ",
+    "ii","ii ",
+    "oo","oo ",
+    "uu","uu ",
+
+    "î","î ",
+    "ê","ê ",
+    "â","â ",
+    "é","é ",
+    "á","á ",
+    "ó","ó ",
+
+    "a","a ",
+    "e","e ",
+    "i","i ",
+    "o","o ",
+    "u","u ",
+    "ú","ú ",
+    #consonants from rossel ortho replacements, missing previously
+    "mbyw","x ",
+    "mbwy","x ",
+    "pwy","x ",
+    "pyw","x ",
+    "tpy","x ",
+    "dpy","x ",
+    "kpy","x ",
+    "myw","x ",
+    "mwy","x ",
+    "ngw","x ",
+    "nmy","x ",
+    "ngm","x ",
+    "mby","x ",
+    "mbw","x ",
+    "nty","x ",
+    "ndy","x ",
+    "nkw","x ",
+    "mty","x ",
+    "mdy","x ",
+    "mgw","x ",
+    "dny","x ",
+    "dmy","x ",
+    "knw","x ",
+    "py","x ",
+    "pw","x ",
+    "ty","x ",
+   # "ch"
+    "dy","x ",
+    "ky","x ",
+    "kw","x ",
+    #"tp"
+    #"dp"
+    #"kp"
+    "my","x ",
+    "mw","x ",
+    "ny","x ",
+    #"ng"
+    "nm","x ",
+    #"mb"
+    #"nt"
+    "nj","x ",
+    #"nd"
+    #"nk"
+    #"mt"
+    #"md"
+    #"mg"
+    #"dn"
+    #"dm"
+   # "kn"
+    #"km"
+    #"vy"
+    "ly","x ",
+    #"lv"
+    #"gh"
+    #consonants, veeeery dirty!
+    "'n","ń ",
+    "ń","ń ",
+    "tp","tp ",
+    "dp","dp ",
+    "kp","kp ",
+    "ngm","ngm ",
+    "ńm","ńm ",
+    "ng","ng ",
+    "mb","mb ",
+    "nd","nd ",
+    "nt","nt ",
+    "nk","nk ",
+    "dn","dn ",
+    "kn","kn ",
+    "vy","vy ",
+    "mt","mt ",
+    "md","md ",
+    "mg","mg ",
+    "dm","dm ",
+    "km","km ",
+    "gh","gh ",
+    "lv","lv ",
+    "cch","ch ",
+    "ch","ch ", #NOTE
+   "w","w ",
+   "r","r ",
+   "t","t ",
+   "y","y ",
+   "p","p ",
+   "s","s ",
+   "d","d ",
+  "f","f ",
+   "g","g ",
+   "h","h ",
+   "j","j ",
+   "k","k ",
+   "l","l ",
+   "z","z ",
+   "c","c ",
+   "v","v ",
+   "b","b ",
+   "n","n ",
+   "m","m "
+
+  ),#last item above should not have a comma
+  ncol = 2,byrow = T)
+
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+```{r}
+# get frequencies in raw_YEL
+#note, you should have ran the code documented in get_freq_cor.Rmd
+
+scan("words_corpus.txt",what="char")->wds
+
+wds[1:1000]
+
+#initialize the unichar phono-like representation
+wds_uni=wds
+
+for(i in 1:dim(correspondances)[1]) { #transform into unicharacter
+  wds_uni=gsub(correspondances[i,1],correspondances[i,2],wds_uni,fixed=T,useBytes = T)
+}
+wds_uni[1:1000]
+
+write.table(wds_uni,"phonemized-words.txt",col.names = F,row.names = F,quote=F)
+```
+
+Finally this line happens in the nwr19yd folder
+
+```{bash, eval=F}
+tr ' ' '\n' < phonemized-words.txt | sort | uniq -c | sort > phone-counts.txt
+```
+
+The final steps were done by hand in an excel file called freq_corpus.xlsx, to calculate frequencies removing items that we think should not count.

### 2020-09-09 manuscript.tex

@@
 
-%% Pandoc can only set 10, 11, 12 pt
-%% uncomment below to set fontsize
-%\usepackage[fontsize=13pt]{scrextend}
-
-%\setmainfont{Calibri} % Set main font for latin characters
-
-\setlength{\parskip}{0.15cm} % Set space between paragraphs
-%\linespread{1.2}\selectfont % Set line height
-%\usepackage[doublespacing]{setspace} % Use double space without changing footnotes line height
-
-%% Special font for IPA
-%% Make sure "Doulos SIL" is installed on your computer
-%% For other typefaces supporting IPA symbols, see
-%% https://en.wikipedia.org/wiki/International_Phonetic_Alphabet#Typefaces
-\newfontfamily\ipa{Doulos SIL} % Font for IPA symbols
-\DeclareTextFontCommand{\ipatext}{\ipa}
-
-%%%         Section for CJK Characters                   %%%
-%%%   You may want to uncomment the code below if        %%%
-%%%   you're writing this document with CJK characters   %%%
-
-%\usepackage{xeCJK}  % Uncomment for using CJK characters
-%% Set main font for CJK characters
-%% Make sure your system has the font set
-%\setCJKmainfont[
-%	BoldFont={HanWangHeiHeavy}  % Set font for CJK boldface
-%    ]{標楷體}    % Set font for normal CJK
-%% Some Traditional Chinese fonts: AR PL KaitiM Big5, PingFang TC, Noto Sans CJK TC
-%\XeTeXlinebreaklocale "zh"
-%\XeTeXlinebreakskip = 0pt plus 1pt
+\setmainfont{Doulos SIL} % Set main font to Doulos SIL
 
@@
 may be oral or nasal, short or long. The 10 oral vowel qualities (/i/,
-/ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /\ipatext{ɔ}/, /æ/, /ɑ/) can be produced
-as short or long vowels, with 7 of these able to appear as short and
-long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /\ipatext{ɔ}/, /æ/, /ɑ/).
-In total, Levinson ({\textbf{???}}) then counts 90 distinctive segments
-in Yélî Dnye (93 when including consonants that are extremely rarely
-used).
+/ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or
+long vowels, with 7 of these able to appear as short and long nasal
+vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/). In total, Levinson
+({\textbf{???}}) then counts 90 distinctive segments in Yélî Dnye (93
+when including consonants that are extremely rarely used).
 
@@
 \midrule
-nopimade & ṇɔpimædɛ & dp:a & ṭpæ͂ & kamo & kæmɔ & dimope & ṭimɔpɛ & diponate & ṭipɔṇætɛ\\
-poni & pɔṇi & dpa & ṭpæ & kani & kæṇi & diyeto & ṭijɛtɔ & nomiwake & ṇɔmiwækɛ\\
+nopimade & nɔpimæṭɛ & dp:a & ṭp̃æ & kamo & kæmɔ & dimope & ṭimɔpɛ & diponate & ṭipɔnætɛ\\
+poni & pɔni & dpa & ṭpæ & kani & kæni & diyeto & ṭijɛtɔ & nomiwake & nɔmiwækɛ\\
 wî & wɯ & dpâ & ṭpɑ & kipo & kipɔ & meyadi & mɛjæṭi & todiwuma & tɔṭiwumæ\\
- &  & dpê & ṭpə & noki & ṇɔki & mituye & mitujɛ & wadikeno & wæṭikɛṇɔ\\
- &  & dpéé & ṭpe: & nomi & ṇɔmi & nademo & ṇædɛmɔ &  & \\
+ &  & dpê & ṭpə & noki & nɔki & mituye & mitujɛ & wadikeno & wæṭikɛnɔ\\
+ &  & dpéé & ṭpeː & nomi & nɔmi & nademo & næṭɛmɔ &  & \\
 \addlinespace
- &  & dpi & ṭpi & piwa & piwæ & nayeki & ṇæjɛki &  & \\
- &  & dpu & ṭpu & towi & tɔwi & nuyedi & ṇujɛdi &  & \\
- &  & gh:ââ & ɣɑ: & tupa & tupæ & pedumi & pɛdumi &  & \\
- &  & ghuu & ɣu: &  &  & tiwune & tiwuṇɛ &  & \\
- &  & kp:ââ & kpɑ͂: &  &  & tumowe & tumɔwɛ &  & \\
+ &  & dpi & ṭpi & piwa & piwæ & nayeki & næjɛki &  & \\
+ &  & dpu & ṭpu & towi & tɔwi & nuyedi & nujɛṭi &  & \\
+ &  & gh:ââ & ɣ̃ɑː & tupa & tupæ & pedumi & pɛṭumi &  & \\
+ &  & ghuu & ɣuː &  &  & tiwune & tiwunɛ &  & \\
+ &  & kpu & kpu &  &  & tumowe & tumɔwɛ &  & \\
 \addlinespace
- &  & kpu & kpu &  &  & widone & wiṭɔṇɛ &  & \\
- &  & lv:ê & lβʲə &  &  & wumipo & wumipɔ &  & \\
- &  & lva & lβʲæ &  &  &  &  &  & \\
+ &  & lv:ê & lβʲ̃ə &  &  & widone & wiṭɔnɛ &  & \\
+ &  & lva & lβʲæ &  &  & wumipo & wumipɔ &  & \\
  &  & lvi & lβʲi &  &  &  &  &  & \\
- &  & t:êê & tə͂ː &  &  &  &  &  & \\
+ &  & tp:a & tp̃æ &  &  &  &  &  & \\
+ &  & tpâ & tpɑ &  &  &  &  &  & \\
 \addlinespace
- &  & tp:a & tpæ͂ &  &  &  &  &  & \\
- &  & tpâ & tpɑ &  &  &  &  &  & \\
  &  & tpê & tpə &  &  &  &  &  & \\

### 2020-09-09 manuscript.Rmd
add more phon info to intro

--- 2020-09-07 manuscript.Rmd
@@
 
-### Intro to the language (**mc**) - please feel free to throw away anything that is not useful!
+### Yélî Dnye phonological overview
+<!--M2A: is this okay?-->
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+The Yélî sound system, much like its baroque grammatical system [@levinsoninprepydgrammar], is unlike any other in the region.
 
-- complexity in the vowel system
-- complexity in the consonant system
-- word shapes
-- typical word length
-- although not the focus of this paper, high use of suppletion in verbal paradigms, other features of language, see Levinson XXX for details
+With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
+The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these able to be contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized or palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsoninprepydgrammar for details).
+The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+Vowels in Yélî Dnye may be oral or nasal, short or long.
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
+
+In total, Yélî Dnye uses 90 distinctive segments (93 when including consonants that are extremely rarely used), far outstripping the phonemic inventory size of other documented Papuan languages (@levinsoninprepydgrammar; Foley 1986) and includes at least two contrasts not yet been documented elsewhere [labial-coronal double-articulations with dental vs.
+post-alveolar coronal placement in both oral and nasal stops, Maddieson & Levinson 2001; Ladefoged and Maddieson 1996].
+
+Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllablic words appearing least frequently [~10%; based on >5800 lexemes in the current Yélî Dnye dictionary, @levinsoninprepydgrammar].
+The vast majority of syllables use a CV format.
+A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., 'ndap' (Spondylus shell) /ṇṭæp/).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., 'ndapî' /'ṇṭæ.pɯ/).
+This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /'pɔ.nɯ/).
+An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
+Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsoninprepydgrammar for details).
+
+While all consonants can appear in the first syllable of the word, in bisyllabic roots, the second syllable onset consonant is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
+In monosyllabic stems, oral vowels (short: 43% and long: 31%) are more common than nasal ones (short: 4% and long: 22%), with the first syllable of bisyllabic stems following a similar pattern.
+Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels.
+Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see @levinsoninprepydgrammar for details).
 
 ### Intro to the people (**mc**) - please feel free to throw away anything that is not useful!
+
+Most speakers of the language grow up speaking it monolingually within their swidden horticulturalist communities until they begin attending school around the age of 7; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands [@brownIPchapter], the same region in which the current data were collected.
+Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
 

### 2020-09-09 manuscript.tex
update ms

@@
 
-\subsubsection{The Yélî language}\label{the-yuxe9luxee-language}
+\subsubsection{Yélî Dnye phonological
+overview}\label{yuxe9luxee-dnye-phonological-overview}
 
@@
 the far end of the Louisiade Archipelago in Milne Bay Province, Papua
-New Guinea. Most speakers of the language grow up speaking it
-monolingually within their swidden horticulturalist communities until
-they begin attending school around the age of 7; school instruction is
-in English. While monolingual Yélî Dnye upbringing is common,
-multilingual families are not unusual, particularly in the region around
-the Catholic Mission (1953--present) where there appears to be a higher
-incidence of married-in mothers from other islands ({\textbf{???}}), the
+New Guinea. The Yélî sound system, much like its baroque grammatical
+system (Levinson, in preparation), is unlike any other in the region.
+
+With only four primary places of articulation (bilabial, alveolar,
+post-alveolar, and velar) and no voicing contrasts, the phonological
+inventory is remarkably packed with acoustically similar segments. The
+core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and
+doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal
+equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
+substantial portion of these able to be contrastively pre-nasalized or
+nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/,
+/kŋ/, /ṭpṇm/, /kpŋm/). A large number of this combinatorial set can
+further be contrastively labialized or palatalized on release, or both
+(e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (in preparation)
+for details). The consonantal inventory also includes a number of
+non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in
+Yélî Dnye may be oral or nasal, short or long. The 10 oral vowel
+qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/,
+/o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels,
+with 7 of these able to appear as short and long nasal vowels as well
+(/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
+
+In total, Yélî Dnye uses 90 distinctive segments (93 when including
+consonants that are extremely rarely used), far outstripping the
+phonemic inventory size of other documented Papuan languages (Levinson
+(in preparation); Foley 1986) and includes at least two contrasts not
+yet been documented elsewhere {[}labial-coronal double-articulations
+with dental vs.~post-alveolar coronal placement in both oral and nasal
+stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.
+
+Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
+monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
+that, and with tri-and-above syllablic words appearing least frequently
+(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the
+current Yélî Dnye dictionary, Levinson, in preparation). The vast
+majority of syllables use a CV format. A small portion of the lexicon
+features words with a final closed syllable, limited to codas of -/m/,
+-/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
+spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
+coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
+process is used frequently with English loan words that have a coda
+(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
+features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
+these are limited to /æ/-. Finally, the lexicon features a handful of
+single-vowel grammatical morphemes (see Levinson (in preparation) for
+details).
+
+While all consonants can appear in the first syllable of the word, in
+bisyllabic roots, the second syllable onset consonant is almost
+exclusively drawn from a subset of the simplex consonants (in order of
+descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/,
+/t/, /ɣ/, /mʷ/). In monosyllabic stems, oral vowels (short: 43\% and
+long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%),
+with the first syllable of bisyllabic stems following a similar pattern.
+Most vowels in the second syllable of bisyllabic stems are also short
+and oral, with a potential preference for rounding harmony between the
+first and second vowels. Stress typically falls on the first syllable of
+bisyllabic words, on the first and third syllables of tetrasyllabic
+words, and on the first or second syllable of trisyllabic words,
+depending on its vocalic features (see Levinson (in preparation) for
+details).
+
+\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) -
+please feel free to throw away anything that is not
+useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}
+
+Most speakers of the language grow up speaking it monolingually within
+their swidden horticulturalist communities until they begin attending
+school around the age of 7; school instruction is in English. While
+monolingual Yélî Dnye upbringing is common, multilingual families are
+not unusual, particularly in the region around the Catholic Mission
+(1953--present) where there appears to be a higher incidence of
+married-in mothers from other islands (Brown \& Casillas, in press), the
 same region in which the current data were collected. Children in these
@@
 and/or some other language(s) from the region.
-
-The Yélî sound system, much like its baroque grammatical system in
-general ({\textbf{???}}), is unlike any other in the region, with 90
-contrastive segments, including at least two contrasts not yet been
-documented elsewhere (labial-coronal double-articulations with dental
-vs.~post-alveolar coronal placement in both oral and nasal stops)
-({\textbf{???}}). With only four primary places of articulation
-(bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts,
-the phonological inventory is remarkably packed with acoustically
-similar segments. The core oral stop set includes both singleton (/p/,
-/t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments,
-with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and
-with a substantial portion of these able to be pre-nasalized or nasally
-released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/,
-/ṭpṇm/, /kpŋm/). Finally, a number of these can also be labialized or
-palatalized on the release, or both (/pʲ/, /tʃ/, /tpʲ/, /ṭʲ/, /ṭpʲ/,
-/kʲ/, /kpʲ/, /pʲʷ/, /mbʲ/, /ndʒ/, /nmdbʲ/, /ṇḍʲ/, /ṇmḍbʲ/, /mbʷ/, /ŋɡʷ/,
-/mbʲʷ/, /ṭṇʲ/, /ṭpṇmʲ/, /kŋʷ/, /mʲ/, /nʲ/, /nmʲ/, /ṇʲ/, /ṇmʲ/, /mʷ/,
-/ŋʷ/, mʲʷ/). Among the remaining consonants are small group of oral
-continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in Yélî Dnye
-may be oral or nasal, short or long. The 10 oral vowel qualities (/i/,
-/ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or
-long vowels, with 7 of these able to appear as short and long nasal
-vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/). In total, Levinson
-({\textbf{???}}) then counts 90 distinctive segments in Yélî Dnye (93
-when including consonants that are extremely rarely used).
-
-To do: - word shapes - typical word length
-
-\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) -
-please feel free to throw away anything that is not
-useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}
 
@@
 \midrule
-nopimade & nɔpimæṭɛ & dp:a & ṭp̃æ & kamo & kæmɔ & dimope & ṭimɔpɛ & diponate & ṭipɔnætɛ\\
-poni & pɔni & dpa & ṭpæ & kani & kæni & diyeto & ṭijɛtɔ & nomiwake & nɔmiwækɛ\\
+nopimade & nɔpimæṭɛ & dp:a & ṭp̃æ & kamo & kæmɔ & dimope & ṭimɔpɛ & dipońate & ṭipɔnætɛ\\
+poni & pɔni & dpa & ṭpæ & kańi & kæni & diyeto & ṭijɛtɔ & ńomiwake & nɔmiwækɛ\\
 wî & wɯ & dpâ & ṭpɑ & kipo & kipɔ & meyadi & mɛjæṭi & todiwuma & tɔṭiwumæ\\
- &  & dpê & ṭpə & noki & nɔki & mituye & mitujɛ & wadikeno & wæṭikɛnɔ\\
- &  & dpéé & ṭpeː & nomi & nɔmi & nademo & næṭɛmɔ &  & \\
+ &  & dpê & ṭpə & ńoki & nɔki & mituye & mitujɛ & wadikeńo & wæṭikɛnɔ\\
+ &  & dpéé & ṭpeː & ńomi & nɔmi & ńademo & næṭɛmɔ &  & \\
 \addlinespace
- &  & dpi & ṭpi & piwa & piwæ & nayeki & næjɛki &  & \\
- &  & dpu & ṭpu & towi & tɔwi & nuyedi & nujɛṭi &  & \\
+ &  & dpi & ṭpi & piwa & piwæ & ńayeki & næjɛki &  & \\
+ &  & dpu & ṭpu & towi & tɔwi & ńuyedi & nujɛṭi &  & \\
  &  & gh:ââ & ɣ̃ɑː & tupa & tupæ & pedumi & pɛṭumi &  & \\
- &  & ghuu & ɣuː &  &  & tiwune & tiwunɛ &  & \\
+ &  & ghuu & ɣuː &  &  & tiwuńe & tiwunɛ &  & \\
  &  & kpu & kpu &  &  & tumowe & tumɔwɛ &  & \\
 \addlinespace
- &  & lv:ê & lβʲ̃ə &  &  & widone & wiṭɔnɛ &  & \\
+ &  & lv:ê & lβʲ̃ə &  &  & widońe & wiṭɔnɛ &  & \\
  &  & lva & lβʲæ &  &  & wumipo & wumipɔ &  & \\
@@
 
+\hypertarget{ref-brownIPchildrearing}{}
+Brown, P., \& Casillas, M. (in press). Childrearing through social
+interaction on Rossel Island, PNG. In A. J. Fentiman \& M. Goody (Eds.),
+\emph{Esther Goody revisited: Exploring the legacy of an original
+inter-disciplinarian} (pp. XX--XX). New York, NY: Berghahn.
+
 \hypertarget{ref-chiat2007preschool}{}
@@
 \emph{PloS One}, \emph{9}(2), e89544.
+
+\hypertarget{ref-levinsoninprepydgrammar}{}
+Levinson, S. C. (in preparation). \emph{A grammar of yélî dnye, the
+papuan language of rossel island}. Cambridge, UK: Cambridge University
+Press.
 

### 2020-09-09 manuscript.Rmd
update ms

@@
 Most speakers of the language grow up speaking it monolingually within their swidden horticulturalist communities until they begin attending school around the age of 7; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands [@brownIPchapter], the same region in which the current data were collected.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands [@brownIPchildrearing], the same region in which the current data were collected.
 Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.

### 2020-09-09 wrangling.R
added frequencies from corpus

@@ -45,11 +45,12 @@ merge(trnsc,inc,by.x="id",by.y="ID",all.x=T)->trnsc
 dim(trnsc) #still no loss, all good
 
 #add orthographic representations targets & other stims chars
-read.table("freq_corpus.csv",sep=";",header=T)->phone_counts
+read.delim("segment-counts.txt", encoding="UTF-8",sep="\t",header=T)->phone_counts
 read.delim("stimconv.txt", encoding="UTF-8")->stims
 read.delim("segconv.txt", encoding="UTF-8")->segments
 #add corpus freq to segments
-#merge(segments,phone_counts)->test
+merge(segments,phone_counts,by.x="ortho",by.y="ortho",all.x=T)->segments
+write.table(segments,"segments_with_cor_freq.txt",sep="\t",row.names = F)
 
 stim_seg_freq=matrix(NA,nrow=dim(stims)[1],ncol=8)
 for(i in 1:dim(stim_seg_freq)[1]) for(j in 1:dim(stim_seg_freq)[2]) {

### 2020-09-09 getting_frequency_corpus.Rmd
added frequencies from corpus

--- 2020-09-09 getting frequency corpus.Rmd
@@
     #nasal
-    ":ii",":ii ",
-    ":ee",":ee ",
-    ":aa",":aa ",
-    ":êê",":êê ",
-    ":ââ",":ââ ",
-    ":uu",":uu ",
-    ":oo",":oo ",
+    ":ii","1 ",
+    ":ee","2 ",
+    ":aa","3 ",
+    ":êê","4 ",
+    ":ââ","5 ",
+    ":uu","6 ",
+    ":oo","7 ",
+    ":óó","37 ",
 
-    ":i",":i ",
-    ":e",":e ",
-    ":a",":a ",
-    ":ê",":ê ",
-    ":â",":â ",
-    ":u",":u ",
-    ":o",":o ",
+    ":i","8 ",
+    ":e","9 ",
+    ":a","0 ",
+    ":ê","10 ",
+    ":â","11 ",
+    ":u","12 ",
+    ":o","13 ",
+     ":ó","38 ",
 
     #non-nasal
-    "îî","îî ",
-    "êê","êê ",
-    "ââ","ââ ",
-    "éé","éé ",
-    "áá","áá ",
-    "óó","óó ",
+    "îî","14 ",
+    "êê","15 ",
+    "ââ","16 ",
+    "éé","17 ",
+    "áá","18 ",
+    "óó","19 ",
 
-    "aa","aa ",
-    "ee","ee ",
-    "ii","ii ",
-    "oo","oo ",
-    "uu","uu ",
+    "aa","20 ",
+    "ee","21 ",
+    "ii","22 ",
+    "oo","23 ",
+    "uu","24 ",
 
-    "î","î ",
-    "ê","ê ",
-    "â","â ",
-    "é","é ",
-    "á","á ",
-    "ó","ó ",
+    "î","25 ",
+    "ê","26 ",
+    "â","27 ",
+    "é","28 ",
+    "á","29 ",
+    "ó","30 ",
 
-    "a","a ",
-    "e","e ",
-    "i","i ",
-    "o","o ",
-    "u","u ",
-    "ú","ú ",
+    "a","31 ",
+    "e","32 ",
+    "i","33 ",
+    "o","34 ",
+    "u","35 ",
+    "ú","36 ",
     #consonants from rossel ortho replacements, missing previously
-    "mbyw","x ",
-    "mbwy","x ",
-    "pwy","x ",
-    "pyw","x ",
-    "tpy","x ",
-    "dpy","x ",
-    "kpy","x ",
-    "myw","x ",
-    "mwy","x ",
-    "ngw","x ",
-    "nmy","x ",
-    "ngm","x ",
-    "mby","x ",
-    "mbw","x ",
-    "nty","x ",
-    "ndy","x ",
-    "nkw","x ",
-    "mty","x ",
-    "mdy","x ",
-    "mgw","x ",
-    "dny","x ",
-    "dmy","x ",
-    "knw","x ",
-    "py","x ",
-    "pw","x ",
-    "ty","x ",
+    "mbyw","39 ",
+    "mbwy","40 ",
+    "pwy","41 ",
+    "pyw","42 ",
+    "tpy","43 ",
+    "dpy","44 ",
+    "kpy","45 ",
+    "myw","46 ",
+    "mwy","47 ",
+    "ngw","48 ",
+    "nmy","49 ",
+    "ngm","50 ",
+    "mby","51 ",
+    "mbw","52 ",
+    "nty","53 ",
+    "ndy","54 ",
+    "nkw","55 ",
+    "mty","56 ",
+    "mdy","57 ",
+    "mgw","58 ",
+    "dny","59 ",
+    "dmy","60 ",
+    "knw","61 ",
+    "py","62 ",
+    "pw","63 ",
+    "ty","64 ",
    # "ch"
-    "dy","x ",
-    "ky","x ",
-    "kw","x ",
+    "dy","65 ",
+    "ky","66 ",
+    "kw","67 ",
     #"tp"
@@
     #"kp"
-    "my","x ",
-    "mw","x ",
-    "ny","x ",
+    "my","68 ",
+    "mw","69 ",
+    "ny","70 ",
     #"ng"
-    "nm","x ",
+    "nm","71 ",
     #"mb"
     #"nt"
-    "nj","x ",
+    "nj","72 ",
     #"nd"
@@
     #"vy"
-    "ly","x ",
+    "ly","73 ",
     #"lv"
@@
     #consonants, veeeery dirty!
-    "'n","ń ",
+    "tp","Tp ",
+    "dp","Dp ",
+    "kp","Kp ",
+    "ngm","Ngm ",
+    "ńm","ñ ",
+    "ng","Ng ",
+    "mb","Mb ",
+    "nd","Nd ",
+    "nt","Nt ",
+    "nk","Nk ",
+    "dn","Dn ",
+    "kn","Kn ",
+    "vy","Vy ",
+    "mt","Mt ",
+    "md","Md ",
+    "mg","Mg ",
+    "dm","Dm ",
+    "km","Km ",
+    "gh","Gh ",
+    "lv","Lv ",
+    "cch","CCh ",
+    "'n","'n ",
     "ń","ń ",
-    "tp","tp ",
-    "dp","dp ",
-    "kp","kp ",
-    "ngm","ngm ",
-    "ńm","ńm ",
-    "ng","ng ",
-    "mb","mb ",
-    "nd","nd ",
-    "nt","nt ",
-    "nk","nk ",
-    "dn","dn ",
-    "kn","kn ",
-    "vy","vy ",
-    "mt","mt ",
-    "md","md ",
-    "mg","mg ",
-    "dm","dm ",
-    "km","km ",
-    "gh","gh ",
-    "lv","lv ",
-    "cch","ch ",
-    "ch","ch ", #NOTE
+    "ch","Ch ", #NOTE
    "w","w ",
+   "ẃ","ẃ ",
    "r","r ",
@@
 
-write.table(wds_uni,"phonemized-words.txt",col.names = F,row.names = F,quote=F)
+wds_uni=unlist(strsplit(wds_uni,split=" "))
+counts=data.frame(table(wds_uni))
+counts$wds_uni=as.character(counts$wds_uni)
+
+colnames(correspondances)<-c("ortho","fake")
+correspondances=data.frame(correspondances)
+correspondances$fake=gsub(" ","",as.character(correspondances$fake))
+
+#backtranslate
+merge(counts,correspondances,by.x="wds_uni",by.y="fake")->counts
+counts=counts[order(counts$Freq),]
+
+colnames(counts)[2]<-"counts_corpus"
+counts$freq_corpus=counts$counts_corpus/sum(counts$counts_corpus)
+
+write.table(counts,"segment-counts.txt",col.names = T,row.names = F,quote=T,sep="\t")
 ```
 
-Finally this line happens in the nwr19yd folder
+## Analyses
 
-```{bash, eval=F}
-tr ' ' '\n' < phonemized-words.txt | sort | uniq -c | sort > phone-counts.txt
+I ran the code in wrangling again, to integrate these counts & freqs into the main data set.
+
+In addition, I wrote out a new version of segments which has the corpus frequency
+
+```{r}
+read.table("segments_with_cor_freq.txt",header=T)->segments # not ran
+plot(segments$pc~segments$freq_corpus,type="n")
+text(segments$pc~segments$freq_corpus,labels=segments$ortho)
+
+plot(segments$pc~log(segments$freq_corpus),type="n")
+text(segments$pc~log(segments$freq_corpus),labels=segments$ortho)
+
+#some stimuli sounds do not occur in the corpus at all
+#so we'll give them a really small frequency, just so that they show up
+#and we tag them in red
+segments$freq_corpus[is.na(segments$freq_corpus)]<-.0001
+plot(segments$pc~log(segments$freq_corpus),type="n",main="blue fitted to sounds in corpus, purple to all")
+text(segments$pc~log(segments$freq_corpus),labels=segments$ortho,col=ifelse(segments$freq_corpus<.00057,"red","blue"))
+#abline(lm(segments$pc~log(segments$freq_corpus),subset=c(segments$freq_corpus<.00057)),col="red",lty=2) #this one is impossible, bc it's vertical, all the x are the same
+abline(lm(segments$pc~log(segments$freq_corpus),subset=c(segments$freq_corpus>=.00057)),col="blue",lty=2)
+abline(lm(segments$pc~log(segments$freq_corpus)),col="purple",lty=2)
+
 ```
-
-The final steps were done by hand in an excel file called freq_corpus.xlsx, to calculate frequencies removing items that we think should not count.

### 2020-09-11 getting_frequency_corpus.Rmd
added corpus frequency analyses

@@
 
-```{r}
+```{r token freq}
 # get frequencies in raw_YEL
-#note, you should have ran the code documented in get_freq_cor.Rmd
 
@@
 
-wds[1:1000]
+#wds[1:1000]
 
@@
 }
-wds_uni[1:1000]
+#wds_uni[1:1000]
+
+wds_uni=unlist(strsplit(wds_uni,split=" "))
+counts=data.frame(table(wds_uni))
+counts$wds_uni=as.character(counts$wds_uni)
+
+correspondances2=correspondances
+colnames(correspondances2)<-c("ortho","fake")
+correspondances2=data.frame(correspondances2)
+correspondances2$fake=gsub(" ","",as.character(correspondances2$fake))
+
+#backtranslate
+merge(counts,correspondances2,by.x="wds_uni",by.y="fake")->counts
+counts=counts[order(counts$Freq),]
+
+colnames(counts)[2]<-"counts_corpus"
+counts$freq_corpus=counts$counts_corpus/sum(counts$counts_corpus)
+
+write.table(counts,"segment-counts.txt",col.names = T,row.names = F,quote=T,sep="\t")
+```
+
+```{r type freq}
+
+#initialize the unichar phono-like representation
+wds_uni= names(table(wds))
+
+for(i in 1:dim(correspondances)[1]) { #transform into unicharacter
+  wds_uni=gsub(correspondances[i,1],correspondances[i,2],wds_uni,fixed=T,useBytes = T)
+}
+#wds_uni[1:1000]
 
@@
 
-colnames(counts)[2]<-"counts_corpus"
-counts$freq_corpus=counts$counts_corpus/sum(counts$counts_corpus)
+colnames(counts)[2]<-"counts_corpus_types"
+counts$freq_corpus_types=counts$counts_corpus/sum(counts$counts_corpus)
 
-write.table(counts,"segment-counts.txt",col.names = T,row.names = F,quote=T,sep="\t")
+write.table(counts,"segment-counts-types.txt",col.names = T,row.names = F,quote=T,sep="\t")
 ```

### 2020-09-11 manuscript.Rmd
added corpus frequency analyses

@@
 ```{r xling-freq-ana}
-xl_fr=glmer(correct~avg_fr+age.rounded+(1|file) + (1|target),data=trnsc,family="binomial")
+trnsc$avg_fr_s=scale(trnsc$avg_fr)
+trnsc$age.rounded_s=scale(trnsc$age.rounded)
+xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
 mycoeff=round(summary(xl_fr)$coefficients,3)
+#summary(xl_fr)
+#plot(xl_fr)
 ```
@@
 We could include 885 observations (**why -- shouldn't it be 1k??**), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr","Estimate"]`, SE ß = `r mycoeff["avg_fr","Std.
-Error"]`, p = `r mycoeff["avg_fr","Pr(>|z|)"]`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
+Error"]`, p = `r mycoeff["avg_fr_s","Pr(>|z|)"]`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+Age was also a significant predictor  (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
+Error"]`, p = `r mycoeff["age.rounded_s","Pr(>|z|)"]`).
 
@@
 ```
+
+```{r cor vs xl}
+#fr_cor=glmer(correct~avg_fr_cor+age.rounded+(1|file) + (1|target),data=trnsc,family="binomial")
+#fr is not sig, but close (p=.1, estimate is 17)
+#trnsc$avg_fr_cor_s=scale(trnsc$avg_fr_cor)
+#both_fr=glmer(correct~avg_fr_s+avg_fr_cor_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
+#both_coeff=round(summary(both_fr)$coefficients,3)
+
+#fr_cor_ty=glmer(correct~avg_fr_cor_ty+age.rounded+(1|file) + (1|target),data=trnsc,family="binomial")
+#fr is sig
+trnsc$avg_fr_cor_ty_s=scale(trnsc$avg_fr_cor_ty)
+both_fr=glmer(correct~avg_fr_s+avg_fr_cor_ty_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
+both_coeff=round(summary(both_fr)$coefficients,3)
+
+read.delim("segments_with_cor_freq.txt", encoding="UTF-8")->segments
+freq_cor=cor.test(segments$pc,log(segments$freq_corpus_types))
+```
+
+We next checked whether this may actually be due to frequency of the sounds within the language.
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020png], and applied the natural logarithm.
+As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, =`r round(freq_cor$p.value,3)`).
+Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
+As before, the model contained random slopes for both child ID and target.
+In this model, both cross-linguistic frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std.
+Error"]`, p = `r both_coeff["avg_fr_s","Pr(>|z|)"]`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std.
+Error"]`, p = `r both_coeff["age.rounded_s","Pr(>|z|)"]`) were significant predictors of accuracy, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
+Error"]`, p = `r both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"]`) was not.
+
+<!-- a2m should we point out that the coefficient for cross-ling frequency is reduced from .77 to .67, so indirectly corpus frequency does affect a little bit?
+Or making things too complicated, and just say "so corpus frequency does not explain away these results" -->
 
@@
 
-### Accuracy a function of item length
+### Accuracy as a function of item length
 

### 2020-09-11 manuscript.Rmd
add relevant ethnographic info

--- 2020-09-09 manuscript.Rmd
@@
 
-### Intro to the people (**mc**) - please feel free to throw away anything that is not useful!
-
-Most speakers of the language grow up speaking it monolingually within their swidden horticulturalist communities until they begin attending school around the age of 7; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands [@brownIPchapter], the same region in which the current data were collected.
+### The Yélî community
+<!--M2A: is this okay?-->
+Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
 Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
 
-Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
+Most Yélî people are swidden horticulturalists, raising a variety of tubers (e.g., yams, manioc, taro), coconut, banana, and more in their gardens.
+In addition to garden foods, Yélî children grow up eating fish and foraging for shellfish and nuts, which provide a regular source of extra protein.
+The typical household in our dataset includes seven individuals and is situated among a collection of 4 or more other households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
 
-- usually monolingual at home
-- schooling in English but it starts at age XX, so not relevant here
-- however, some use of English due to immigrants & children of immigrants
-- children spend a lot of time with other children
-- most parents are subsistence farmers
-- parental education generally varies between XX and YY
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by barrier reef, transport to and from the island is both infrequent and irregular.
+There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport, when there is deisel available.
+There is a medical radio connection at the health clinic (at the Catholic mission) via which news and messages are often passed along.
+There is also a cellular tower on the island, though it is typically not functioning.
+As such, access to news and outside resources, particularly international connections and larger data transfers, is typically unavailable.
+Our data collection is therefore limited to the duration of the researchers' on-island visits.
+
+Despite restricted outside contact, formal education is a priority for many Yélî families.
+A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (8 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early].
+While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), and so it is very common for households to billet school children during the weekdays for months at a time (or for whole months if the child's home is very far).
+Combined with practices of collective child guardianship within their close-knit hamlets, the practical outcome for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's parents---to this child assent is also culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[Note that in a few cases this does mean that precise details such as child date of birth, parent age in years, and parent education was only able to be approximated as the parents were residing on another part of the island.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
+
+While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., Bram's thesis; MC's CFH data].
+What we do know about the early language environment is that, while child-directed speech is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early].
+While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early].
+This increase in Child CDS is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 people) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+<!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->
 
@@
 However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
-<!--M2self: make sure it's apparent how/why a visit is limited-->
 

### 2020-09-11 manuscript.tex
add relevant ethnographic info

@@
 
-\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) -
-please feel free to throw away anything that is not
-useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}
-
-Most speakers of the language grow up speaking it monolingually within
-their swidden horticulturalist communities until they begin attending
-school around the age of 7; school instruction is in English. While
-monolingual Yélî Dnye upbringing is common, multilingual families are
-not unusual, particularly in the region around the Catholic Mission
-(1953--present) where there appears to be a higher incidence of
-married-in mothers from other islands (Brown \& Casillas, in press), the
-same region in which the current data were collected. Children in these
+\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}
+
+Most speakers of the language grow up speaking it monolingually until
+they begin attending school around the age of 7; school instruction is
+in English. While monolingual Yélî Dnye upbringing is common,
+multilingual families are not unusual, particularly in the region around
+the Catholic Mission---the same region in which the current data were
+collected---where there is a higher incidence of married-in mothers from
+other islands (Brown \& Casillas, in press). Children in these
 multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
@@
 
-Little is known about language development in children growing up in
-Rossel Island, a community of primarily subsistence farmers who tend to
-reside in close-knitted villages where child care is distributed across
-many individuals, and who typically speak Yélî Dnyé, a phonologically
-and lexically complex language.
-
-\begin{itemize}
-\tightlist
-\item
-  usually monolingual at home
-\item
-  schooling in English but it starts at age XX, so not relevant here
-\item
-  however, some use of English due to immigrants \& children of
-  immigrants
-\item
-  children spend a lot of time with other children
-\item
-  most parents are subsistence farmers
-\item
-  parental education generally varies between XX and YY
-\end{itemize}
+Most Yélî people are swidden horticulturalists, raising a variety of
+tubers (e.g., yams, manioc, taro), coconut, banana, and more in their
+gardens. In addition to garden foods, Yélî children grow up eating fish
+and foraging for shellfish and nuts, which provide a regular source of
+extra protein. The typical household in our dataset includes seven
+individuals and is situated among a collection of 4 or more other
+households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that
+they typically comprise a set of brothers, their wives and children, and
+their mother and father, with neighboring hamlets also typically related
+through the patriline.
+
+Laying 250 nautical miles off the coast of mainland PNG and surrounded
+by barrier reef, transport to and from the island is both infrequent and
+irregular. There is no electricity system (we use solar panels) or motor
+vehicles on the island, save dinghys with outboard motors that are
+primarily reserved for medical transport, when there is deisel
+available. There is a medical radio connection at the health clinic (at
+the Catholic mission) via which news and messages are often passed
+along. There is also a cellular tower on the island, though it is
+typically not functioning. As such, access to news and outside
+resources, particularly international connections and larger data
+transfers, is typically unavailable. Our data collection is therefore
+limited to the duration of the researchers' on-island visits.
+
+Despite restricted outside contact, formal education is a priority for
+many Yélî families. A recent study surveying more than 40 parents of
+young children in this region of the island found that nearly all
+parents completed 6 or more years of education, with 65\% of mothers and
+49\% of fathers completing the full range of locally available education
+(8 years), with around half of those parents having pursued an
+additional 2+ years of education on other islands in the region or on
+mainland PNG (Casillas, Brown, \& Levinson, 2020). While there are a
+handful of local schools around the island, these may be well out of
+walking distance for many children (i.e., more than 1 hour on foot or by
+canoe each day), and so it is very common for households to billet
+school children during the weekdays for months at a time (or for whole
+months if the child's home is very far). Combined with practices of
+collective child guardianship within their close-knit hamlets, the
+practical outcome for researchers is that adult consent can often come
+from a combination of aunts, uncles, adult cousins, and grandparents
+standing in for the child's parents---to this child assent is also
+culturally pertinent, as child independence is encouraged and respected
+from toddlerhood onward (Brown \& Casillas, in press).\footnote{Note
+  that in a few cases this does mean that precise details such as child
+  date of birth, parent age in years, and parent education was only able
+  to be approximated as the parents were residing on another part of the
+  island.} Finally, as mentioned, the language of schooling in this
+region is English, and so the majority of teens and adults on the island
+speak some English---many speak it very fluently.
+
+While work on Yélî language development is growing (e.g., Brown, 2011,
+2014; Brown \& Casillas, in press; Casillas et al., 2020; Liszkowski,
+Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's
+linguistic development is quite limited, and research into their
+phonological development in particular has only just begun {[}e.g.,
+Bram's thesis; MC's CFH data{]}. What we do know about the early
+language environment is that, while child-directed speech is relatively
+infrequent throughout the day (Casillas et al., 2020), shared caregiving
+practices and a near-universal fondness for social interaction with
+young children ensure that speech directed to children comes from all
+types of speakers: women, men, and other children (Bunce et al., under
+review; Casillas et al., 2020). While speech from women predominates in
+young children's speech environments on Rossel, as it does elsewhere
+(Bergelson et al., 2019; Bunce et al., under review), there is a
+significant and marked increase in child-directed speech from other
+children as infants get older (Bunce et al., under review; Casillas et
+al., 2020). This increase in Child CDS is attributed to the fact that,
+starting around age 2, children often spend large swaths of the day
+playing, swimming, and foraging for fruit, nuts, and shellfish in large
+(\textasciitilde{}10 people) independent and mixed-age child play groups
+(Brown \& Casillas, in press; Casillas et al., 2020).
 
@@
 The overall NWR score was M = 61\% (SD = 15\%). Scores based on phonemes
-are even higher M = 77\% (SD = 9\%). The phoneme-based normalized
-Levenshtein distance was M = 23\% (SD = 9\%), meaning that about a fifth
+are even higher M = 78\% (SD = 9\%). The phoneme-based normalized
+Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth
 of phonemes were were substituted, inserted, or deleted. Notice that the
@@
 investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 19 vowels deleted, and 4 consonants.
+relatively rare, with only 7 vowels deleted, and 3 consonants.
 
@@
 changed were a great deal more common than changes in quality among
-nasal vowels (42 oral vowels produced with a different quality; 0 nasal
+nasal vowels (43 oral vowels produced with a different quality; 0 nasal
 vowels produced with a different quality). As for consonants,
@@
 mnlwyvdgptkfhch, 0) than vice versa (0). Simple consonants were
-mispronounced as other simple consonants quite frequently (62 simple
+mispronounced as other simple consonants quite frequently (73 simple
 consonants mispronounced as other simple ones, compared to 0 complex
@@
 investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 19 vowels deleted (about 1.17\% of all vowel
-targets), and 4 consonants deleted (about 0.24\% of all consonant
+relatively rare, with only 7 vowels deleted (about 0.54\% of all vowel
+targets), and 3 consonants deleted (about 0.17\% of all consonant
 targets).
@@
 the vowel was changed were a great deal more common than changes in
-quality among nasal vowels (42 oral vowels produced with a different
+quality among nasal vowels (43 oral vowels produced with a different
 quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels
@@
 out of all simple consonant targets). Simple consonants were
-mispronounced as other simple consonants quite frequently (62 simple
-consonants mispronounced as other simple ones or about 0.30\% out of all
+mispronounced as other simple consonants quite frequently (73 simple
+consonants mispronounced as other simple ones or about 0.35\% out of all
 simple consonant targets, compared to 0 complex ones or about NA\% out
@@
 investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 29 vowels deleted (about 0.55\% of all vowel
-targets), and 9 consonants deleted (about 0.26\% of all consonant
+relatively rare, with only 11 vowels deleted (about 0.25\% of all vowel
+targets), and 8 consonants deleted (about 0.21\% of all consonant
 targets).
@@
 quality of the vowel was changed were a great deal more common than
-changes in quality among nasal vowels (197 oral vowels produced with a
-different quality or about 3.20\% out of all oral vowel targets; 23
+changes in quality among nasal vowels (205 oral vowels produced with a
+different quality or about 3.22\% out of all oral vowel targets; 23
 nasal vowels produced with a different quality or about 7.08\% out of
@@
 marked with more complex consonants (specifically dptpkpkmknmbghlv)
-mispronounced as simple ones (specifically mnlwyvdgptkfhch, 266 times or
-about 5.55\% out of all complex consonant targets) than vice versa (2
-times about 5.55\% out of all simple consonant targets). Simple
+mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or
+about 5.57\% out of all complex consonant targets) than vice versa (2
+times about 5.57\% out of all simple consonant targets). Simple
 consonants were mispronounced as other simple consonants quite
-frequently (128 simple consonants mispronounced as other simple ones or
-about 0.48\% out of all simple consonant targets, compared to 62 complex
-ones or about 2.79\% out of all complex consonant targets).
+frequently (129 simple consonants mispronounced as other simple ones or
+about 1.14\% out of all simple consonant targets, compared to 55 complex
+ones or about 2.76\% out of all complex consonant targets).
 
@@
 
+\hypertarget{ref-bergelsoncasillas2019what}{}
+Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
+S., \& Amatuni, A. (2019). What do North American babies hear? A
+large-scale cross-corpus analysis. \emph{Developmental Science},
+\emph{22}(1), e12724.
+doi:\href{https://doi.org/10.1111/desc.12724}{10.1111/desc.12724}
+
 \hypertarget{ref-bowey2001nonword}{}
@@
 
+\hypertarget{ref-brown2011cultural}{}
+Brown, P. (2011). The cultural organization of attention. In A. Duranti,
+E. Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language
+Socialization} (pp. 29--55). Malden, MA: Wiley-Blackwell.
+
+\hypertarget{ref-brown2014interactional}{}
+Brown, P. (2014). The interactional context of language learning in
+Tzeltal. In I. Arnon, M. Casillas, C. Kurumada, \& B. Estigarribia
+(Eds.), \emph{Language in interaction: Studies in honor of Eve V. Clark}
+(pp. 51--82). Amsterdam, NL: John Benjamins.
+
 \hypertarget{ref-brownIPchildrearing}{}
@@
 
+\hypertarget{ref-bunceURcrosscultural}{}
+Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A.,
+Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural
+examination of young children's everyday language experiences.
+
+\hypertarget{ref-casillas2020early}{}
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language
+experience in a papuan community. \emph{Journal of Child Language},
+\emph{XX}, XX--XX.
+
 \hypertarget{ref-chiat2007preschool}{}
@@
 Press.
+
+\hypertarget{ref-liszkowski2012prelinguistic}{}
+Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
+(2012). A prelinguistic gestural universal of human communication.
+\emph{Cognitive Science}, \emph{36}(4), 698--713.
+doi:\href{https://doi.org/10.1111/j.1551-6709.2011.01228.x}{10.1111/j.1551-6709.2011.01228.x}
 

### 2020-09-11 manuscript.tex

--- 2020-09-09 manuscript.tex
@@
 }{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\usepackage{hyperref}
 \hypersetup{
             pdftitle={Non-word repetition in Yélî Dnye},
-            pdfauthor={Alejandrina Cristia~\& Marisa Casillas},
             pdfkeywords={phonology, non-word repetition, development},
@@
 \setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
@@
 \setmainfont{Doulos SIL} % Set main font to Doulos SIL
+% Manuscript styling
+\usepackage{upgreek}
+\captionsetup{font=singlespacing,justification=justified}
+
+% Table formatting
+\usepackage{longtable}
+\usepackage{lscape}
+% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
+\usepackage{multirow}		% Table styling
+\usepackage{tabularx}		% Control Column width
+\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
+\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+
+% Create new environments so endfloat can handle them
+% \newenvironment{ltable}
+%   {\begin{landscape}\begin{center}\begin{threeparttable}}
+%   {\end{threeparttable}\end{center}\end{landscape}}
+\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
+
+% Enables adjusting longtable caption width to table width
+% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
+\makeatletter
+\newcommand\LastLTentrywidth{1em}
+\newlength\longtablewidth
+\setlength{\longtablewidth}{1in}
+\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
+
+% \setlength{\parindent}{0.5in}
+% \setlength{\parskip}{0pt plus 0pt minus 0pt}
+
+% \usepackage{etoolbox}
+\makeatletter
+\patchcmd{\HyOrg@maketitle}
+  {\section{\normalfont\normalsize\abstractname}}
+  {\section*{\normalfont\normalsize\abstractname}}
+  {}{\typeout{Failed to patch abstract.}}
+\makeatother
+\shorttitle{NWR in Yélî Dnye}
+\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2}}
+\affiliation{
+\vspace{0.5cm}
+\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
+\authornote{All data are made available in a repository in the Open Science Framework.
+AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
+S.
+McDonnell Foundation Understanding Human Cognition Scholar Award.
+
+Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
+\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
+\usepackage{lineno}
+
+\linenumbers
+\usepackage{csquotes}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[shorthands=off,main=english]{babel}
+\else
+  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
+  \usepackage{polyglossia}
+  \setmainlanguage[]{english}
+\fi
 
 \title{Non-word repetition in Yélî Dnye}
-\author{Alejandrina Cristia\textsuperscript{1}~\& Marisa
-Casillas\textsuperscript{2}}
+
 \date{}
 
-\authornote{All data are made available in a repository in the
-Open Science Framework. AC acknowledges the support of the Agence
-Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004,
-ANR-14-CE30-0003, ANR-17-EURE-0017); and the J. S. McDonnell Foundation
-Understanding Human Cognition Scholar Award.
-
-Correspondence concerning this article should be addressed to
-Alejandrina Cristia, 29, rue d'Ulm, 75005 Paris, France. E-mail:
-\href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
-
 \abstract{
-In nonword repetition (NWR) studies, participants are presented
-auditorily with an item that is phonologically legal but lexically
-meaningless in the local language. Accuracy is thought to reflect
-long-term phonological knowledge as well as online phonological working
-memory and flexible production patterns.In this study, we report on NWR
-results among children learning Yêly Dnyé, an isolate spoken in Rossel
-Island, PNG, with an unusually dense phonological inventory. This study
-contributes to three lines of research. First, we document that non-word
-items containing typologically rare sounds are repeated accurately less
-often that non-words containing more common sounds. Second, we document
-rather weak effects of item length, contributing to other research
-suggesting that length effects may be language-specific. Third, we do
-not find strong individual variation effects in this population,
-contrary to previous results documenting strong age-related effects.
-Together, these data provide a unique view of online phonological
-processing in a seldom-studied language, and contribute to both
-typological and language acquisition research.
-
+In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
+Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+This study contributes to three lines of research.
+First, we document that non-word items containing typologically rare sounds are repeated accurately less often that non-words containing more common sounds.
+Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
+Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
+Together, these data provide a unique view of online phonological processing in a seldom-studied language, and contribute to both typological and language acquisition research.
 }
@@
 
-\subsection{TODO Middy}\label{todo-middy}
+\hypertarget{todo-middy}{%
+\subsection{TODO Middy}\label{todo-middy}}
 
@@
 \item
-  look through this manuscript for bits of text to be written by
-  \textbf{mc}
-\item
-  help with table 1 (formatting \& making it appear, because it's broken
-  by the phonological representations)
-\item
-  do you think Figure 2 should be done with phonetic symbols? If yes,
-  could you try to implement that?
-\item
-  double check demo -- probably start with the xlsx version that has the
-  children's first names, date of birth, and date of test. I looked
-  again at my local files (still haven't checked the labs' ones) and I'm
-  surprised I have maternal education for more children than I have
-  birth order for\ldots{} If you correct this, to generate the csv I
-  also round the age, assign age.rounded to reported age (in years) when
-  DOB is missing, remove first names, DOB, DOTest and age.precise before
-  pushing to github
-\item
-  double check my frequency entries in segments.xlsx using the search
-  function in \url{https://phoible.org/parameters}; take a look at
-  \url{http://phoible.github.io/conventions/} in case you see something
-  about double articulation (I didn't find tp or lBj)
+  look through this manuscript for bits of text to be written by \textbf{mc}
+\item
+  help with table 1 (formatting \& making it appear, because it's broken by the phonological representations)
+\item
+  do you think Figure 2 should be done with phonetic symbols? If yes, could you try to implement that?
+\item
+  double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
+I looked again at my local files (still haven't checked the labs' ones) and I'm surprised I have maternal education for more children than I have birth order for\ldots{} If you correct this, to generate the csv I also round the age, assign age.rounded to reported age (in years) when DOB is missing, remove first names, DOB, DOTest and age.precise before pushing to github
+\item
+  double check my frequency entries in segments.xlsx using the search function in \url{https://phoible.org/parameters}; take a look at \url{http://phoible.github.io/conventions/} in case you see something about double articulation (I didn't find tp or lBj)
 \item
@@
 \item
-  for both of the above, if you make any changes, they need to be
-  re-saved as UTF16 txt file, and then converted (see remember above)
-\item
-  double check also stim\_seg\_freq -- why are some items still with NA
-  freq?
-\item
-  consent, ``correct'', ``accuracy'', ``performance''
+  for both of the above, if you make any changes, they need to be re-saved as UTF16 txt file, and then converted (see remember above)
+\item
+  double check also stim\_seg\_freq -- why are some items still with NA freq?
+\item
+  consent, \enquote{correct}, \enquote{accuracy}, \enquote{performance}
 \item
@@
 
-\subsection{TODO later}\label{todo-later}
+\hypertarget{todo-later}{%
+\subsection{TODO later}\label{todo-later}}
 
@@
 \item
-  probably turn summary of types of error into a table; use proportion
-  of correctly produced rather than number; use first attempts rather
-  than the whole data set DONE
-\item
-  maybe add to the last ana the prevalence (ie out of all the gh which
-  proportion get transformed) DONE I think\ldots{} (this item looks the
-  same as one of the three above)
-\item
-  maybe add analyses of score as a function of frequency (phones,
-  diphones) in MC's corpus? (mc suggested, but I don't think we need
-  it\ldots{}) - waiting until we talk about this
+  probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set DONE
+\item
+  maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed) DONE I think\ldots{} (this item looks the same as one of the three above)
+\item
+  maybe add analyses of score as a function of frequency (phones, diphones) in MC's corpus? (mc suggested, but I don't think we need it\ldots{}) - waiting until we talk about this
 \end{itemize}
 
-\subsection{Introduction}\label{introduction}
-
-Although infants begin to learn about their native language's phonology
-within the first year, many studies suggest that in perception and
-production, in phonetics and phonology, their knowledge continues to
-develop throughout childhood (e.g., Hazan \& Barrett, 2000). One common
-task in this line of research is nonword repetition (NWR). In NWR
-studies, participants are presented auditorily with an item that is
-phonologically legal but lexically meaningless in the language children
-are learning. The child should immediately try to say it back without
-changing anything. Accuracy is thought to reflect long-term phonological
-knowledge (which allows the child to perceive the item accurately even
-though it is not a real word they have encountered before) as well as
-online phonological working memory (to encode the item in the interval
-between hearing it and saying it back) and flexible production patterns
-(to produce the item accurately even though it had never been pronounced
-before).
-
-NWR has been used to seek answers to a variety of theoretical questions,
-including what the links between phonology, working memory, and the
-lexicon are (Bowey, 2001), and to assess the extent to which apparent
-phonological constraints found in the lexicon affect online production
-(Gallagher, 2014). It is also frequently used as for applications,
-notably as a diagnostic for language delays and disorders (Estes, Evans,
-\& Else-Quest, 2007). Since non-words can be generated in any language,
-it has attracted the attention of researchers working in multilingual
-and linguistically diverse environments, particularly in Europe (Action,
-2009; Meir, Walters, \& Armon-Lotem, 2016).
-
-In this study, we report on NWR results among children learning Yêly
-Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense
-phonological inventory. The present report uses these unique data to
-contribute to several lines of research. First, we made sure that some
-of the non-word items contained typologically rare and/or challenging
-sounds, so that we can contribute to the study of whether rare sounds
-are disadvantaged in perception and/or production, both in terms of
-overall accuracy of repetition and patterns of mispronunciation. Second,
-we look at the effects of item length, since previous NWR research has
-uncovered variable effects. Third, we contribute to the basic NWR
-literature by contributing an additional data point for children
-learning a rarely-studied language and culture, and further
-investigating whether there are structured sources that account for
-individual variation.
-
-\subsubsection{Yélî Dnye phonological
-overview}\label{yuxe9luxee-dnye-phonological-overview}
-
-Yélî Dnye is an isolate language (presumed Papuan) spoken by
-approximately 7000 people residing on Rossel Island, which is found at
-the far end of the Louisiade Archipelago in Milne Bay Province, Papua
-New Guinea. The Yélî sound system, much like its baroque grammatical
-system (Levinson, in preparation), is unlike any other in the region.
-
-With only four primary places of articulation (bilabial, alveolar,
-post-alveolar, and velar) and no voicing contrasts, the phonological
-inventory is remarkably packed with acoustically similar segments. The
-core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and
-doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal
-equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
-substantial portion of these able to be contrastively pre-nasalized or
-nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/,
-/kŋ/, /ṭpṇm/, /kpŋm/). A large number of this combinatorial set can
-further be contrastively labialized or palatalized on release, or both
-(e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (in preparation)
-for details). The consonantal inventory also includes a number of
-non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in
-Yélî Dnye may be oral or nasal, short or long. The 10 oral vowel
-qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/,
-/o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels,
-with 7 of these able to appear as short and long nasal vowels as well
-(/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
-
-In total, Yélî Dnye uses 90 distinctive segments (93 when including
-consonants that are extremely rarely used), far outstripping the
-phonemic inventory size of other documented Papuan languages (Levinson
-(in preparation); Foley 1986) and includes at least two contrasts not
-yet been documented elsewhere {[}labial-coronal double-articulations
-with dental vs.~post-alveolar coronal placement in both oral and nasal
-stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.
-
-Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
-monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
-that, and with tri-and-above syllablic words appearing least frequently
-(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the
-current Yélî Dnye dictionary, Levinson, in preparation). The vast
-majority of syllables use a CV format. A small portion of the lexicon
-features words with a final closed syllable, limited to codas of -/m/,
--/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
-spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
-coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
-process is used frequently with English loan words that have a coda
-(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
-features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
-these are limited to /æ/-. Finally, the lexicon features a handful of
-single-vowel grammatical morphemes (see Levinson (in preparation) for
-details).
-
-While all consonants can appear in the first syllable of the word, in
-bisyllabic roots, the second syllable onset consonant is almost
-exclusively drawn from a subset of the simplex consonants (in order of
-descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/,
-/t/, /ɣ/, /mʷ/). In monosyllabic stems, oral vowels (short: 43\% and
-long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%),
-with the first syllable of bisyllabic stems following a similar pattern.
-Most vowels in the second syllable of bisyllabic stems are also short
-and oral, with a potential preference for rounding harmony between the
-first and second vowels. Stress typically falls on the first syllable of
-bisyllabic words, on the first and third syllables of tetrasyllabic
-words, and on the first or second syllable of trisyllabic words,
-depending on its vocalic features (see Levinson (in preparation) for
-details).
-
-\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) -
-please feel free to throw away anything that is not
-useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}
-
-Most speakers of the language grow up speaking it monolingually within
-their swidden horticulturalist communities until they begin attending
-school around the age of 7; school instruction is in English. While
-monolingual Yélî Dnye upbringing is common, multilingual families are
-not unusual, particularly in the region around the Catholic Mission
-(1953--present) where there appears to be a higher incidence of
-married-in mothers from other islands (Brown \& Casillas, in press), the
-same region in which the current data were collected. Children in these
-multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
-and/or some other language(s) from the region.
-
-Little is known about language development in children growing up in
-Rossel Island, a community of primarily subsistence farmers who tend to
-reside in close-knitted villages where child care is distributed across
-many individuals, and who typically speak Yélî Dnyé, a phonologically
-and lexically complex language.
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
+
+Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., Hazan \& Barrett, 2000).
+One common task in this line of research is nonword repetition (NWR).
+In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning.
+The child should immediately try to say it back without changing anything.
+Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item accurately even though it had never been pronounced before).
+
+NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and to assess the extent to which apparent phonological constraints found in the lexicon affect online production (Gallagher, 2014).
+It is also frequently used as for applications, notably as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
+
+In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+The present report uses these unique data to contribute to several lines of research.
+First, we made sure that some of the non-word items contained typologically rare and/or challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall accuracy of repetition and patterns of mispronunciation.
+Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
+Third, we contribute to the basic NWR literature by contributing an additional data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
+
+\hypertarget{yuxe9luxee-dnye-phonological-overview}{%
+\subsubsection{Yélî Dnye phonological overview}\label{yuxe9luxee-dnye-phonological-overview}}
+
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+The Yélî sound system, much like its baroque grammatical system (Levinson, n.d.), is unlike any other in the region.
+
+With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
+The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these able to be contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized or palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (n.d.) for details).
+The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+Vowels in Yélî Dnye may be oral or nasal, short or long.
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
+
+In total, Yélî Dnye uses 90 distinctive segments (93 when including consonants that are extremely rarely used), far outstripping the phonemic inventory size of other documented Papuan languages (Levinson (n.d.); Foley 1986) and includes at least two contrasts not yet been documented elsewhere {[}labial-coronal double-articulations with dental vs.~post-alveolar coronal placement in both oral and nasal stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.
+
+Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllablic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the current Yélî Dnye dictionary, Levinson, n.d.).
+The vast majority of syllables use a CV format.
+A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., \enquote{ndapî} /\enquote{ṇṭæ.pɯ/).
+This process is used frequently with English loan words that have a coda (e.g., 'ponî} (phone) /\enquote{pɔ.nɯ/).
+An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala} (here) /æ.'læ/), but these are limited to /æ/-.
+Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (n.d.) for details).
+
+While all consonants can appear in the first syllable of the word, in bisyllabic roots, the second syllable onset consonant is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
+In monosyllabic stems, oral vowels (short: 43\% and long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%), with the first syllable of bisyllabic stems following a similar pattern.
+Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels.
+Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see Levinson (n.d.) for details).
+
+\hypertarget{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}{%
+\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) - please feel free to throw away anything that is not useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
+
+Most speakers of the language grow up speaking it monolingually within their swidden horticulturalist communities until they begin attending school around the age of 7; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.), the same region in which the current data were collected.
+Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
+
+Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
 
@@
 \item
-  however, some use of English due to immigrants \& children of
-  immigrants
+  however, some use of English due to immigrants \& children of immigrants
 \item
@@
 
-\subsubsection{Brief review of NWR for our
-purposes}\label{brief-review-of-nwr-for-our-purposes}
-
-There is some variation in the presentation procedure and structure of
-items found in previous NWR work. For example, items are often presented
-orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe,
-2015), although an increasing number of studies have turned to playing
-back the stimuli in order to have greater control of the stability of
-the presentation (Brandeker \& Thordardottir, 2015). Additionally, while
-some studies have used 10-15 non-words, others have employed up to 46
-unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler,
-2019). Often, authors modulate structural complexity, typically measured
-in terms of item length (measured in number of syllables) and/or
-syllable structure (open as opposed to closed syllables, Gallon, Harris,
-\& Van der Lely, 2007).
-
-Previous work seems to avoid difficult sounds, but we felt this was
-important to represent Yélî Dnye, so we also varied this factor. We
-designed a relatively large number of items but, aware that this may
-render the task longer and more tiresome, we split some of the items
-across children. This allowed us to get information about repetition
-accuracy of more items.
-
-Naturally, designing the task in this way may render the study of
-individual variaiton within the population more difficult because
-different children are exposed to different items. However, a review of
-previous work on individual variation suggested to us that many
-individual differences effects are relatively small, and would not be
-detectable with the sample size that we could collect in a given visit.
-
-That said, we contribute to the literature by also reporting descriptive
-analyses of individual variation that could potentially be integrated in
-meta-analytic efforts. Based on previous work, we looked at potential
-improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid,
-Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and
-potential negative effects of bilingual exposure (Brandeker \&
-Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
-Previous work typically finds no significant differences as a function
-of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014;
-Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
-Although previous research has not often investigated potential effects
-of birth order on NWR, there is a sizable literature on these effects in
-other language tasks (Havron et al., 2019), and therefore we report on
-those too.
-
-\subsubsection{Research questions}\label{research-questions}
-
-After some preliminary analyses to set the stage, we address the
-following questions:
+\hypertarget{brief-review-of-nwr-for-our-purposes}{%
+\subsubsection{Brief review of NWR for our purposes}\label{brief-review-of-nwr-for-our-purposes}}
+
+There is some variation in the presentation procedure and structure of items found in previous NWR work.
+For example, items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation (Brandeker \& Thordardottir, 2015).
+Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
+Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
+
+Previous work seems to avoid difficult sounds, but we felt this was important to represent Yélî Dnye, so we also varied this factor.
+We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
+This allowed us to get information about repetition accuracy of more items.
+
+Naturally, designing the task in this way may render the study of individual variaiton within the population more difficult because different children are exposed to different items.
+However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
+
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
+Based on previous work, we looked at potential improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and potential negative effects of bilingual exposure (Brandeker \& Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
+Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on those too.
+
+\hypertarget{research-questions}{%
+\subsubsection{Research questions}\label{research-questions}}
+
+After some preliminary analyses to set the stage, we address the following questions:
 
@@
 \item
-  Does the frequency of sounds across languages predict NWR? Are rarer
-  sounds more often substituted by commoner sounds?
-\item
-  How does score change as a function of item length in number of
-  syllables?
-\item
-  Is individual variation attributable to child age, sex, birth order,
-  monolingual status, and/or parental education? 
+  Does the frequency of sounds across languages predict NWR? Are rarer sounds more often substituted by commoner sounds?
+\item
+  How does score change as a function of item length in number of syllables?
+\item
+  Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
 \end{itemize}
 
-In view of the hypothesis-driven nature of this work, we had considered
-boosting the interpretational value of this evidence by announcing our
-analysis plans prior to conducting them. However, we realized that even
-pre-registering an analysis would be equivocal because we do not have
-enough power to look at all relationships of interest, and often to
-detect any of the known effects given their variability across studies.
-To illustrate this, we portray studies in which children's NWR scores
-were gathered between 4 and 12 years of age, and reported separately for
-items that are relatively short (1-2 syllables) versus longer items (3-4
-syllables) in Figure 1. Notice that the effect of stimulus length is
-minuscule among Italian children (Piazzalunga et al., 2019), but
-considerable among Tsimane' children (Cristia, Farabolini, Scaff,
-Havron, \& Stieglitz, 2020), where a drop of 40 percentage points is
-observed at all ages. A similar difference in NWR scores for short
-versus long items was observed among Arabic children (Jaber-Awida,
-2018). Even the effect of age is unstable in this sample. Whereas it is
-quite clear that children's NWR scores increase in the Italian data, age
-effects are less stable among Tsimane' children. Therefore, all analyses
-here are descriptive and should be considered exploratory.
+In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
+However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects given their variability across studies.
+To illustrate this, we portray studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables) in Figure 1.
+Notice that the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Tsimane' children (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020), where a drop of 40 percentage points is observed at all ages.
+A similar difference in NWR scores for short versus long items was observed among Arabic children (Jaber-Awida, 2018).
+Even the effect of age is unstable in this sample.
+Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
+Therefore, all analyses here are descriptive and should be considered exploratory.
 
@@
 \includegraphics{manuscript_files/figure-latex/prevlit-fig-1.pdf}
-\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years)
-and item length for comparable studies. Arabic data from Jaber et al.
-(2018); Italian data from Piazzalunga et al. (2019); Tsimane' from
-Cristia et al. (2020); Yélî data from the present study.}
+\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years) and item length for comparable studies.
+Arabic data from Jaber et al.
+(2018); Italian data from Piazzalunga et al.
+(2019); Tsimane' from Cristia et al.
+(2020); Yélî data from the present study.}
 \end{figure}
 
-\subsection{Methods}\label{methods}
-
-\subsubsection{Stimuli}\label{stimuli}
-
-Many NWR studies are based on a fixed list of 12-16 items that vary in
-length between 1 and 4 syllables, often additionally varying syllable
-complexity and/or cluster presence and complexity, always meeting the
-condition that they do not mean anything in the target language (e.g.,
-Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We kept the
-same variation in item length and the non-meaningfulness requirement,
-but we did not vary syllable complexity and clusters because these are
-vanishingly rare in Yély Dnye. We also increased the number of items an
-individual child would be tested on, so that a child would get up to 23
-items to repeat (note that up to other work has also used 24-30 items:
-Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and
-distributed them across children, so as to increase the coverage, and be
-able to study more items.
-
-A first list of candidate items was generated in 2018 by selecting
-simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'',
-``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined
-into consonant-vowel syllables, further sampling the space of 1- to
-4-syllable sequences. These candidates were automatically checked
-against Levinson's 2015 dictionary and removed from consideration if
-they appeared in the dictionary. The second author presented them orally
-to three local research assistants, who were asked to repeat them and
-further say whether they were real words. Any item for which two or more
-of the assistants reported them having a meaning or some form of
-association was excluded.
-
-A second list of candidate items was generated in 2019 by selecting
-complex consonants and systematically crossing them with all the vowels
-in the Yélî inventory to produce consonant-vowel monosyllables. As
-before, items were automatically excluded if they appeared in the
-dictionary. Additionally, since hearing vowel length in monosyllables in
-isolation is challenging, any item that had a short/long real word
-neighbor was filtered out. Since the phonology and phonetics of Yélî is
-still in the process of being described {[}CITE \textbf{mc} please fill
-in{]}, there could have been undocumented constraints that rendered
-items illegal. Therefore, we made sure that the precise consonant-vowel
-sequence occurred in some real word in the dictionary (i.e., that there
-was a longer word included the monosyllable as a subsequence). These
-candidates were presented to one informant, for a final check that they
-did not mean anything. Together with the 2018 selection, they were
-recorded using a headset \textbf{XXX }mc** please fill in** and an
-Olympus \textbf{XXX }mc** please fill in** from the written form
-presented together with the same item orally (by the second author). The
-complete recorded list was finally presented to two more informants, who
-could repeat all the items and who confirmed there were no real words.
-Even so, there was one monosyllable that was often identified as a real
-word (intended ``yî'' /yXX/; identified as ``yi'' /yi/, \emph{tree}).
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
+
+\hypertarget{stimuli}{%
+\subsubsection{Stimuli}\label{stimuli}}
+
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
+We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in Yély Dnye.
+We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat (note that up to other work has also used 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
+
+A first list of candidate items was generated in 2018 by selecting simple consonants (\enquote{p}, \enquote{t}, \enquote{d}, \enquote{k}, \enquote{m}, \enquote{n}, \enquote{w}, \enquote{y}) and vowels (\enquote{i}, \enquote{o},\enquote{u}, \enquote{a},\enquote{e}) that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
+These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
+The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
+Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
+
+A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî inventory to produce consonant-vowel monosyllables.
+As before, items were automatically excluded if they appeared in the dictionary.
+Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
+Since the phonology and phonetics of Yélî is still in the process of being described {[}CITE \textbf{mc} please fill in{]}, there could have been undocumented constraints that rendered items illegal.
+Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
+These candidates were presented to one informant, for a final check that they did not mean anything.
+Together with the 2018 selection, they were recorded using a headset \textbf{XXX }mc** please fill in** and an Olympus \textbf{XXX }mc** please fill in** from the written form presented together with the same item orally (by the second author).
+The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
+Even so, there was one monosyllable that was often identified as a real word (intended \enquote{yî} /yXX/; identified as \enquote{yi} /yi/, \emph{tree}).
 This item is removed from analyses below.
 
-The final list is composed of three practice items; 20 monoysllables
-containing sounds that are less frequent in the world's languages than
-singleton plosives; 8 bisyllables; 12 trisyllables; and 4
-quadrisyllables (see Table 2).
+The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
 
@@
 
-A Praat script was written to randomize this list 20 times, and split it
-into two sublists, to generate 40 different elicitation sets. The 40
-elicitation sets are available online from \url{https://osf.io/5qspb}.
+A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
+The 40 elicitation sets are available online from \url{https://osf.io/5qspb}.
 The split had the following constraints:
@@
 \item
-  the same three items were selected as practice items and used in all
-  40 elicitation sets
-\item
-  splits were done within each length group from the 2018 items (i.e.,
-  separately for 2, 3, and 4-syllable items); and among onset groups for
-  the difficult monosyllables generated in 2019 (i.e., all the
-  monosyllables starting with tp were split into 2 sublists). Since some
-  of these groups had an odd number of items, one of the sublists was
-  slightly longer than the other (20 versus 23).
-\item
-  once the sublist split had been done, items were randomized such that
-  all children heard first the 3 practice items in a fixed order (1, 2,
-  and 4 syllables), a randomized version of their sublist selection of
-  difficult onset items, and randomized versions of their 2-syllables,
-  then 3-syllables, and finally 4-syllable items.
+  the same three items were selected as practice items and used in all 40 elicitation sets
+\item
+  splits were done within each length group from the 2018 items (i.e., separately for 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
+Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
+\item
+  once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllables, then 3-syllables, and finally 4-syllable items.
 \end{itemize}
 
-\subsubsection{Procedure}\label{procedure}
-
-We tried to balance three desiderata: That children would not be unduly
-exposed to the items before they themselves had to repeat them; that
-children would feel comfortable doing this task with us; and that the
-community would feel safe with us doing this task with their children.
-Moreover, there were also some logistic constraints in terms of the
-space availability. As a result, the places where elicitation happened
-varied across the hamlets.
-
-We visited four different hamlets once, and attempted to test all
-eligible children present at the time, to prevent the items `spreading'
-through hearsay. In the first hamlet, we tested children in five
-different places, with some children being tested inside a house and
-others tested on the veranda. The complete list of places and the ways
-in which they met the desiderata mentioned above can be found in the raw
-data, available from online supplementary materials.
-
-The child was donned a headset (\textbf{xx }mc** please fill in** for
-most of the children, SHURE WH20 XLR headset with a dynamic microphone
-for the rest), recorded into the left channel into a Tascam DR40x
-digital recorder. For most children, the headset could not stay
-comfortably on the child's head, and thus it was placed on the child's
-shoulders, with the microphone carefully placed close to the child's
-mouth. A local informant sat next to the child, to would provide the
-instructions and, if needed, coach the child to make sure, using the
-three practice items as well as real words, that they understood that
-the task was to repeat the items precisely without changing anything. An
-experimenter (the first author) delivered the elicitation stimuli to the
-local informant and the child over headphones.
-
-The first phase was making sure the child understood the task. This was
-explained orally and the first training item was presented. Often,
-children froze and did not say anything. If this happened, then we
-followed this procedure. First, the informant insisted. If the child
-still did not say anything, the informant asked the child to repeat a
-real word, and another, and another. If the child could repeat these
-correctly, then we provided the recorded training item over headphones
-again. Most children successfully started repeating the items presented
-over headphones at this point; a few further needed the local informant
-to model the behavior (i.e., they would hear the item again, and she
-would say it; then we would play it again, and ask the child to say it).
-A small minority still failed to repeat the item after hearing it over
-headphones. If that occurred, we tried with the second training item, at
-which point some children got it and could continue. A small minority,
-however, failed to repeat this one, as well as the third training item,
-in which case we stopped the test altogether.
-
-NWR studies vary in whether children are provided with several
-opportunities to hear and say the item. To have a fixed and clear
-procedure, we decided that items other than the inital three training
-ones would not be repeated unless the child made an attempt to produce
-them. If this attempt was judged correct by the local informant, then
-the experimenter would move on to the next item (whispering this over a
-separate headset that was recording onto the right track of the same
-Tascam). If the local informant heard a deviation, she indicated to the
-experimenter that the item needed to be repeated, and up to 5 attempts
-were allowed.
-
-Whenever siblings from the same family were tested, an attempt was made
-to test first the older and then the younger child, and always on
-different elicitaiton sets.
-
-\subsubsection{Coding}\label{coding}
-
-A script was written to randomize all tokens from all children, pairing
-each with the auditory target the child had been provided with. A native
-research assistant then listened through all productions of a given
-target (randomized across children and repetition order), and made a
-judgment of whether the item was correctly or incorrectly repeated. She
-additionally orthographically transcribed exactly what the child said,
-providing some examples of the types of errors children in general make
-(without making specific reference to Yélî sounds or the items in the
-elicitation sets).
-
-\subsubsection{Analyses}\label{analyses}
-
-Some NWR studies employ phoneme-based scoring in addition to or instead
-of word-level accuracy (e.g., Cristia et al., 2020). We scored items in
-terms of the number of phonemes that could be aligned across the target
-and attempt, divided by the number of phonemes of whichever item was
-longer (the target or the attempt). Although previous work does not use
-distance metrics, we additionally report those.
-
-Additionally, we looked up each phoneme in each target word in the
-Phoible database (Moran \& McCloy, 2019). We entered the number of
-languages in which each phoneme was found, as well as what percentage of
-languages in the database that represented. We could not find a small
-number of phonemes in the database, and treat those as NA values that do
-not contribute to the target-level global average phone cross-linguistic
-frequency, although an alternative would be to attribute them a very
-small number.
-
-Finally, for describing children's patterns of errors, all repetitions
-of a given target were taken into account. We describe the proportion of
-items where the change resulted in a real word (semantic errors); and
-classify the most common phonological errors.
-
-\subsubsection{Participants}\label{participants}
-
-This study was approved as part of a larger research effort by Marisa
-Casillas, with the research protocol having been approved by the IRB
-committee entitled The Ethics Committee of the Faculty of Social
-Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen;
-ECSW) of Radboud University, Nijmegen, The Netherlands. The approval for
-work outside of the Netherlands came as an amendment to the general
-research line on language development research (original request:
-ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
-Participation was voluntary, with children being invited to come and
-participate. Regardless of how they performed, children were provided
-with a snack as compensation. Children who came up to participate but
-then refused were nonetheless provided with the snack.
+\hypertarget{procedure}{%
+\subsubsection{Procedure}\label{procedure}}
+
+We tried to balance three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them; that children would feel comfortable doing this task with us; and that the community would feel safe with us doing this task with their children.
+Moreover, there were also some logistic constraints in terms of the space availability.
+As a result, the places where elicitation happened varied across the hamlets.
+
+We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items \enquote{spreading} through hearsay.
+In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
+The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
+
+The child was donned a headset (\textbf{xx }mc** please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
+For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
+A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
+An experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headphones.
+
+The first phase was making sure the child understood the task.
+This was explained orally and the first training item was presented.
+Often, children froze and did not say anything.
+If this happened, then we followed this procedure.
+First, the informant insisted.
+If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another.
+If the child could repeat these correctly, then we provided the recorded training item over headphones again.
+Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it).
+A small minority still failed to repeat the item after hearing it over headphones.
+If that occurred, we tried with the second training item, at which point some children got it and could continue.
+A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.
+
+NWR studies vary in whether children are provided with several opportunities to hear and say the item.
+To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them.
+If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a separate headset that was recording onto the right track of the same Tascam).
+If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.
+
+Whenever siblings from the same family were tested, an attempt was made to test first the older and then the younger child, and always on different elicitaiton sets.
+
+\hypertarget{coding}{%
+\subsubsection{Coding}\label{coding}}
+
+A script was written to randomize all tokens from all children, pairing each with the auditory target the child had been provided with.
+A native research assistant then listened through all productions of a given target (randomized across children and repetition order), and made a judgment of whether the item was correctly or incorrectly repeated.
+She additionally orthographically transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
+
+\hypertarget{analyses}{%
+\subsubsection{Analyses}\label{analyses}}
+
+Some NWR studies employ phoneme-based scoring in addition to or instead of word-level accuracy (e.g., Cristia et al., 2020).
+We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
+Although previous work does not use distance metrics, we additionally report those.
+
+Additionally, we looked up each phoneme in each target word in the Phoible database (Moran \& McCloy, 2019).
+We entered the number of languages in which each phoneme was found, as well as what percentage of languages in the database that represented.
+We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
+
+Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
+We describe the proportion of items where the change resulted in a real word (semantic errors); and classify the most common phonological errors.
+
+\hypertarget{participants}{%
+\subsubsection{Participants}\label{participants}}
+
+This study was approved as part of a larger research effort by Marisa Casillas, with the research protocol having been approved by the IRB committee entitled The Ethics Committee of the Faculty of Social Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) of Radboud University, Nijmegen, The Netherlands.
+The approval for work outside of the Netherlands came as an amendment to the general research line on language development research (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+Participation was voluntary, with children being invited to come and participate.
+Regardless of how they performed, children were provided with a snack as compensation.
+Children who came up to participate but then refused were nonetheless provided with the snack.
 
 A total 55 children were tested, from 34 of families, in five hamlets.
-Some children could not be included for the following reasons: refused
-participation or failed to repeat items presented over headphones even
-after coaching (N=0), spoke too softly to allow offline coding (N=5). In
-addition, 2 teenagers were tested to put younger children at ease; their
-data is not included in analyses below. The remaining 40 children (14
-girls) were aged 6.96 years (range 3.90-11 years). There were 32
-children exposed only to Yélî in the home, 6 children who were also
-exposed to another language in the home, and , 2 for whom this
-information was missing. Maternal years of education averaged 8.24 years
-(range 6-12 years; 2 children had this information missing).\footnote{Education
-  is often reported in even years because people typically complete
-  two-year cycles.} In terms of birth order, 5 were first borns, 4
-second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this
-information.
-
-\subsection{Results}\label{results}
-
-\subsubsection{Preliminary analyses}\label{preliminary-analyses}
+Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=0), spoke too softly to allow offline coding (N=5).
+In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
+The remaining 40 children (14 girls) were aged 6.96 years (range 3.90-11 years).
+There were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and , 2 for whom this information was missing.
+Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Education is often reported in even years because people typically complete two-year cycles.} In terms of birth order, 5 were first borns, 4 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
+
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
+
+\hypertarget{preliminary-analyses}{%
+\subsubsection{Preliminary analyses}\label{preliminary-analyses}}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig1-first_vs_others-1.pdf}
-\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual
-participants averaging separately their first attempts and all other
-attempts.}
+\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual participants averaging separately their first attempts and all other attempts.}
 \end{figure}
 
-We first checked whether accuracy varies between first and subsequent
-presentations of an item by averaging word-level accuracy at the
-participant level separately for first attempts and subsequent
-repetitions. We excluded 3 children who did not have data for one of
-these two types. As shown in Figure 1, participants' mean word-level
-accuracy became more heterogeneous in subsequent repetitions.
-Surprisingly, subsequent repetitions (M = 39, SD = 26) were on average
-less accurate than first ones (M = 65, SD = 15), t(38) = 6.62, p = 0.
-Given the uncertainty in whether previous work used only the first or
-all repetitions, and since behavior degraded and became more
-heterogeneous in subsequent repetitions, the rest of the analyses focus
-on only the first repetitions.
-
-Taking into account only the first attempts, we averaged attempts by
-each of the 24 children who had data for first attempts; their ages
-ranged from 3.90 to 10.20 (M = 6.50, SD = 1.50).
-
-The overall NWR score was M = 61\% (SD = 15\%). Scores based on phonemes
-are even higher M = 77\% (SD = 9\%). The phoneme-based normalized
-Levenshtein distance was M = 23\% (SD = 9\%), meaning that about a fifth
-of phonemes were were substituted, inserted, or deleted. Notice that the
-normalized Levenshtein distance is the complement of phoneme-based
-scores.
-
-\subsubsection{NWR as a function of cross-linguistic phone
-frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}
-
-In this analysis, we were interested in variation in performance as a
-function of the average frequency with which sounds composing individual
-target words are found in languages over the world. To look at this, we
-fit a mixed logistic regression, in which the outcome variable was
-whether the non-word was correctly repeated or not. The fixed effects of
-interest was the average cross-linguistic phone frequency; we also
-included child age as a control fixed effect, and child ID \& target ID
-as random variables.
-
-We could include 885 observations (\textbf{why -- shouldn't it be
-1k??}), from 41 children producing in any given trial one of 41
-potential target words. The analysis revealed a significant estimate for
-the average cross-linguistic frequency of phones in the target words (ß
-= 0.03, SE ß = 0.01, p = 0): Target words with phones found more
-frequently across languages had a higher proportion of words that were
-correctly repeated, as clear in Figure 2.
+We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
+We excluded 3 children who did not have data for one of these two types.
+As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
+Surprisingly, subsequent repetitions (M = 39, SD = 26)
+were on average less accurate than first ones (M = 65, SD = 15), t(38) = 6.62, p = 0.
+Given the uncertainty in whether previous work used only the first or all repetitions, and since behavior degraded and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
+
+Taking into account only the first attempts, we averaged attempts by each of the 24 children who had data for first attempts; their ages ranged from 3.90 to 10.20 (M = 6.50, SD = 1.50).
+
+The overall NWR score was M = 61\% (SD = 15\%).
+Scores based on phonemes are even higher M = 78\% (SD = 9\%).
+The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
+Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
+
+\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
+\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}
+
+In this analysis, we were interested in variation in performance as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
+The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID \& target ID as random variables.
+
+We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
+The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.77, SE ß = 0.19, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+Age was also a significant predictor (ß = 0.31, SE ß = 0.13, p = 0.02).
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig2-xling_freq-1.pdf}
-\caption{(\#fig:Fig2-xling\_freq)NWR scores for individual target words
-as a function of the average frequency with which each phone is found
-across languages.}
+\caption{(\#fig:Fig2-xling\_freq)NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}
 \end{figure}
 
+We next checked whether this may actually be due to frequency of the sounds within the language.
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings ({\textbf{???}}), and applied the natural logarithm.
+As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(23)=0.65, =0).
+Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
+As before, the model contained random slopes for both child ID and target.
+In this model, both cross-linguistic frequency (ß = 0.50, SE ß = 0.25, p = 0.04) and age (ß = 0.31, SE ß = 0.13, p = 0.02) were significant predictors of accuracy, but corpus frequency (ß = 0.37, SE ß = 0.24, p = 0.12) was not.
+
 \textbf{use proportion of correctly produced rather than number; DONE}
 
-\textbf{a2m use first attempts rather than the whole data set: having
-second thoughts. To be clear: the argument for using first attempts only
-is that all of the other analyses are on kids' first attempts, rather
-than all of their attempts, so we should use just first attempts here
-because it may confuse readers \& it is more consistent. BUT there are
-fewer mistakes this way, which is Options 1 and 2 below has lots of
-zeros\ldots{} CF Option 3 further down}
-
-\textbf{a2m probably turn the paragraph on substitutions into a table:
-having second thoughts, the table would look just like a para, no?}
-
-\textbf{Option 1: just numbers, first attempts only } Next, we
-investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 19 vowels deleted, and 4 consonants.
-
-As for substitutions, it was as common for a nasal vowel to be produced
-as an oral vowel as vice versa (0 oral target vowels produced as nasal
-vowels, 0 nasal target vowels produced as oral vowels). Substitutions in
-which the oral nature was preserved but the quality of the vowel was
-changed were a great deal more common than changes in quality among
-nasal vowels (42 oral vowels produced with a different quality; 0 nasal
-vowels produced with a different quality). As for consonants,
-asymmetries were very marked with more complex consonants (specifically
-dptpkpkmknmbghlv) mispronounced as simple ones (specifically
-mnlwyvdgptkfhch, 0) than vice versa (0). Simple consonants were
-mispronounced as other simple consonants quite frequently (62 simple
-consonants mispronounced as other simple ones, compared to 0 complex
-ones).
-
-\textbf{Option 2: numbers and proportions, first attempts only} Next, we
-investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 19 vowels deleted (about 1.17\% of all vowel
-targets), and 4 consonants deleted (about 0.24\% of all consonant
-targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced
-as an oral vowel as vice versa (0 oral target vowels produced as nasal
-vowels or about 0\% out of all oral vowel targets, 0 nasal target vowels
-produced as oral vowels or about NA\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of
-the vowel was changed were a great deal more common than changes in
-quality among nasal vowels (42 oral vowels produced with a different
-quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels
-produced with a different quality or about NA\% out of all nasal vowel
-targets). As for consonants, asymmetries were very marked with more
-complex consonants (specifically dptpkpkmknmbghlv) mispronounced as
-simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of
-all complex consonant targets) than vice versa (0 times or about NA\%
-out of all simple consonant targets). Simple consonants were
-mispronounced as other simple consonants quite frequently (62 simple
-consonants mispronounced as other simple ones or about 0.30\% out of all
-simple consonant targets, compared to 0 complex ones or about NA\% out
-of all complex consonant targets).
-
-\textbf{Option 3: numbers and proportions, ALL attempts} Next, we
-investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 29 vowels deleted (about 0.55\% of all vowel
-targets), and 9 consonants deleted (about 0.26\% of all consonant
-targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced
-as an oral vowel as vice versa (52 oral target vowels produced as nasal
-vowels or about 2.01\% out of all oral vowel targets, 58 nasal target
-vowels produced as oral vowels or about 6.46\% out of all nasal vowel
-targets). Substitutions in which the oral nature was preserved but the
-quality of the vowel was changed were a great deal more common than
-changes in quality among nasal vowels (197 oral vowels produced with a
-different quality or about 3.20\% out of all oral vowel targets; 23
-nasal vowels produced with a different quality or about 7.08\% out of
-all nasal vowel targets). As for consonants, asymmetries were very
-marked with more complex consonants (specifically dptpkpkmknmbghlv)
-mispronounced as simple ones (specifically mnlwyvdgptkfhch, 266 times or
-about 5.55\% out of all complex consonant targets) than vice versa (2
-times about 5.55\% out of all simple consonant targets). Simple
-consonants were mispronounced as other simple consonants quite
-frequently (128 simple consonants mispronounced as other simple ones or
-about 0.48\% out of all simple consonant targets, compared to 62 complex
-ones or about 2.79\% out of all complex consonant targets).
-
-Finally, we looked into the frequency with which mispronunciations
-resulted in real words. Nearly all of them were: 97\%.
-
-\textbf{TO DISCUSS TOGETHER this is higher than any previous report! can
-we be sure of it?}
-
-\subsubsection{Accuracy a function of item
-length}\label{accuracy-a-function-of-item-length}
+\textbf{a2m use first attempts rather than the whole data set: having second thoughts.
+To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers \& it is more consistent.
+BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros\ldots{} CF Option 3 further down}
+
+\textbf{a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?}
+
+\textbf{Option 1: just numbers, first attempts only }
+Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 7 vowels deleted, and 3 consonants.
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels, 0 nasal target vowels produced as oral vowels).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (43 oral vowels produced with a different quality; 0 nasal vowels produced with a different quality).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0) than vice versa (0).
+Simple consonants were mispronounced as other simple consonants quite frequently (73 simple consonants mispronounced as other simple ones, compared to 0 complex ones).
+
+\textbf{Option 2: numbers and proportions, first attempts only}
+Next, we investigated patterns of deletion and substitution.
+Deletions were relatively rare, with only 7 vowels deleted (about 0.54\% of all vowel targets), and 3 consonants deleted (about 0.17\% of all consonant targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about 0\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (43 oral vowels produced with a different quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times or about NA\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (73 simple consonants mispronounced as other simple ones or about 0.35\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
+
+\textbf{Option 3: numbers and proportions, ALL attempts}
+Next, we investigated patterns of deletion and substitution.
+Deletions were relatively rare, with only 11 vowels deleted (about 0.25\% of all vowel targets), and 8 consonants deleted (about 0.21\% of all consonant targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels or about 2.01\% out of all oral vowel targets, 58 nasal target vowels produced as oral vowels or about 6.46\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (205 oral vowels produced with a different quality or about 3.22\% out of all oral vowel targets; 23 nasal vowels produced with a different quality or about 7.08\% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or about 5.57\% out of all complex consonant targets) than vice versa (2 times about 5.57\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (129 simple consonants mispronounced as other simple ones or about 1.14\% out of all simple consonant targets, compared to 55 complex ones or about 2.76\% out of all complex consonant targets).
+
+Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 97\%.
+
+\textbf{TO DISCUSS TOGETHER this is higher than any previous report! can we be sure of it?}
+
+\hypertarget{accuracy-as-a-function-of-item-length}{%
+\subsubsection{Accuracy as a function of item length}\label{accuracy-as-a-function-of-item-length}}
 
 Next, we inspected whether accuracy varied as a function of word length.
-Results are shown on table XX. This table shows that monosyllables
-accuracy was much lower than other lengths. This is likely due to the
-fact that the majority of monosyllables included were chosen because
-they had sounds that are rare in the world's languages, which may
-indicate that they are hard to produce or to perceive.
-
-Setting monosyllables aside, we observe the typical pattern of decreased
-accuracy for longer items, although this is particularly salient for the
-whole word scoring. This is the most commonly reported type of score,
-but it is also the least forgiving. The pattern is less marked when
-other two scores are used, which are less sensitive to errors.
-
-\begin{verbatim}
-## 
-## 
-## \begin{table}[tbp]
-## \begin{center}
-## \begin{threeparttable}
-## \caption{\label{tab:tablength}NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
-## \begin{tabular}{lll}
-## \toprule
-## Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
-## \midrule
-## 47 (22) & 59 (17) & 41 (17)\\
-## 79 (22) & 94 (7) & 6 (7)\\
-## 78 (19) & 94 (6) & 6 (6)\\
-## 75 (32) & 93 (11) & 7 (11)\\
-## \bottomrule
-## \end{tabular}
-## \end{threeparttable}
-## \end{center}
-## \end{table}
-\end{verbatim}
-
-\subsubsection{Factor structuring individual
-variation}\label{factor-structuring-individual-variation}
+Results are shown on table XX.
+This table shows that monosyllables accuracy was much lower than other lengths.
+This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
+
+Setting monosyllables aside, we observe the typical pattern of decreased accuracy for longer items, although this is particularly salient for the whole word scoring.
+This is the most commonly reported type of score, but it is also the least forgiving.
+The pattern is less marked when other two scores are used, which are less sensitive to errors.
+
+\begin{table}[tbp]
+
+\begin{center}
+\begin{threeparttable}
+
+\caption{\label{tab:tablength}NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
+
+\begin{tabular}{lll}
+\toprule
+Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
+\midrule
+47 (22) & 59 (17) & 41 (17)\\
+79 (22) & 94 (7) & 6 (7)\\
+78 (19) & 94 (6) & 6 (6)\\
+75 (32) & 93 (11) & 7 (11)\\
+\bottomrule
+\end{tabular}
+
+\end{threeparttable}
+\end{center}
+
+\end{table}
+
+\hypertarget{factor-structuring-individual-variation}{%
+\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig2-scores by age-1.pdf}
-\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual
-participants as a function of age and sex (blue = boys, pink = girls).}
+\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual participants as a function of age and sex (blue = boys, pink = girls).}
 \end{figure}
 
-Our final exploratory analysis assessed whether variation in scores was
-structured by factors that vary across individuals. As shown in Figure
-2, there was a greater deal of variance in earlier than later ages, with
-significantly higher NWR scores for older children (Spearman's rank
-correlation, given inequality of variance, rho (1,548.04) = 0.33, p =
-0.12). In contrast, there was no clear association between NWR scores,
-on the one hand, and sex (t (0.18) = 21.34, p = 0.86), birth order (data
-missing for 17 children, 2,678.47) = -0.16, p = 0.44) or maternal
-education (data missing for 2 children, 2,091.92) = 0.09, p = 0.67)
-
-\subsection{Discussion}\label{discussion}
+Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,548.04) = 0.33, p = 0.12).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.18) = 21.34, p = 0.86), birth order (data missing for 17 children, 2,678.47) = -0.16, p = 0.44) or maternal education (data missing for 2 children, 2,091.92) = 0.09, p = 0.67)
+
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
 
@@
 \item
-  What is the overall repetition accuracy (whole word, phoneme based,
-  distance)?
-\item
-  How does this change as a function of item complexity (number of
-  syllables, sound complexity)?
+  What is the overall repetition accuracy (whole word, phoneme based, distance)?
+\item
+  How does this change as a function of item complexity (number of syllables, sound complexity)?
 \end{itemize}
@@
 \item
-  The length distribution in Yélî words is more balanced than that in
-  English, and thus the performance decline for poly- versus
-  mono-syllables may be less pronounced than that for English.
-  \textbf{Check for work on European languages that may have looked into
-  this}
-\item
-  Similarly, we do not know of NWR research that manipulates the
-  difficulty of the sounds that are included in the items, but word
-  naming and other research suggests that children are more accurate
-  when producing easy and/or typologically common sounds than difficult
-  and/or typologically rare sounds {[}CITE{]}. Therefore, we expect
-  higher accuracy for items with common sounds than in those with rare
-  sounds.
-\item
-  The Yélî sound inventory is very large and compressed, with many
-  similar sounds that are acoustic and articulatory neighbors.
-  Therefore, this may constitute a pressure for children to have finer
-  auditory skills (and perhaps more precise articulations) than children
-  speaking languages with a simpler inventory. As a result, differences
-  between easier and harder items may be smaller in this work than in
-  other research. \textbf{no work looking at consonants \& vowels? no
-  work looking at nasal vowels in particular?}
+  The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
+\textbf{Check for work on European languages that may have looked into this}
+\item
+  Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds {[}CITE{]}.
+Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
+\item
+  The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
+Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
+As a result, differences between easier and harder items may be smaller in this work than in other research.
+\textbf{no work looking at consonants \& vowels?
+no work looking at nasal vowels in particular?}
 \end{enumerate}
 
-\textbf{(MC: but we can try and do a cursory analysis based on the
-corpora we have from Steve and my transcription of naturalistic
-interactions!)}
+\textbf{(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)}
 
@@
 \item
-  How frequent are errors that result in real words? Is that a function
-  of item complexity?
-\item
-  Is individual variation explainable by child age, sex, birth order,
-  monolingual status, and/or parental education?
+  How frequent are errors that result in real words? Is that a function of item complexity?
+\item
+  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
 \end{itemize}
@@
 \item
-  Non-monolingual Yélî children are less accurate than monolingual ones
-  when tested on the society-dominant language (we did not test any
-  non-dominant language) 
-\item
-  As revious NWR evidence on this is mixed, but general findings on
-  language development suggest that children whose mothers are more
-  educated are more accurate than children whose mothers are less
-  educated.
-\item
-  To our knowledge, there is no previous NWR work on this, but other
-  research suggests that first-born children should outperform
-  later-born children
-\item
-  Anecdotally Yélî children grow up in close-knitted communities and
-  thus may receive significant portions of their language input from
-  people not in their nuclear family (or at least from people other than
-  their mothers, who tend to be the non-native speakers). If so, the
-  difference between monolinguals and not monolinguals may be smaller
-  than that found in other work . That said, one recent study on the
-  same population shows that most child-directed input in the first 2
-  years does come from the mother , so in so far as this input has a
-  crucial formational role, then there may still be a performance gap
-  between these two groups.
-\item
-  In the Rossel community, formal education plays an extremely minor
-  role in ensuring individual's success, is not a good index of relative
-  socio-economic status, and furthermore there is only a narrow range of
-  variation in maternal educational attainment. This may lead to no or
-  only very small advantages for children whose mothers are more
-  educated, provided that the causal chain between maternal education
-  and child language is via SES more broadly. However, if education
-  directly boosts maternal verbal skills and the incidence of verbal
-  behavior (as suggested by CITE), then we should still see a difference
-  along this factor.
-\item
-  One main causal path between birth order and language development is
-  via parental input (CITE). Given our arguments above for how mothers
-  may not be as important among Rossel people than in other places, then
-  the performance gap between first borns and later borns may be
-  smaller.
+  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
+\item
+  As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
+\item
+  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
+\item
+  Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family
+  (or at least from people other than their mothers, who tend to be the non-native speakers).
+If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
+  . That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother
+  , so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
+\item
+  In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
+This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
+However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
+\item
+  One main causal path between birth order and language development is via parental input (CITE).
+Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
 \end{enumerate}
@@
 
-\subsection{Acknowledgments}\label{acknowledgments}
-
-We are grateful to the individuals who participated in the study. The
-collection and annotation of these recordings was made possible by
-Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also
-to the PNG National Research Institute, and the Administration of Milne
-Bay Province AC acknowledges financial and institutional support from
-Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge,
-ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and
-the J. S. McDonnell Foundation Understanding Human Cognition Scholar
-Award. MC acknowledges financial support from an NWO Veni Innovational
-Scheme grant (275-89-033).
-
-\section{References}\label{references}
-
-\setlength{\parindent}{-0.5in} \setlength{\leftskip}{0.5in}
+\hypertarget{acknowledgments}{%
+\subsection{Acknowledgments}\label{acknowledgments}}
+
+We are grateful to the individuals who participated in the study.
+The collection and annotation of these recordings was made possible by Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also to the PNG National Research Institute, and the Administration of Milne Bay Province AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.
+S.
+McDonnell Foundation Understanding Human Cognition Scholar Award.
+MC acknowledges financial support from an NWO Veni Innovational Scheme grant (275-89-033).
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
 
 \hypertarget{refs}{}
-\hypertarget{ref-is08042009language}{}
-Action, C. (2009). Language impairment in a multilingual society:
-Linguistic patterns and the road to assessment. \emph{Brussels: COST
-Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
-\hypertarget{ref-balladares2016socio}{}
-Balladares, J., Marshall, C., \& Griffiths, Y. (2016). Socio-economic
-status affects sentence repetition, but not non-word repetition, in
-Chilean preschoolers. \emph{First Language}, \emph{36}(3), 338--351.
-doi:\href{https://doi.org/10.1177/0142723715626067}{10.1177/0142723715626067}
-
-\hypertarget{ref-bowey2001nonword}{}
-Bowey, J. A. (2001). Nonword repetition and young children's receptive
-vocabulary: A longitudinal study. \emph{Applied Psycholinguistics},
-\emph{22}(3), 441--469.
-
-\hypertarget{ref-brandeker2015language}{}
-Brandeker, M., \& Thordardottir, E. (2015). Language exposure in
-bilingual toddlers: Performance on nonword repetition and lexical tasks.
-\emph{American Journal of Speech-Language Pathology}, \emph{24}(2),
-126--138.
-
-\hypertarget{ref-brownIPchildrearing}{}
-Brown, P., \& Casillas, M. (in press). Childrearing through social
-interaction on Rossel Island, PNG. In A. J. Fentiman \& M. Goody (Eds.),
-\emph{Esther Goody revisited: Exploring the legacy of an original
-inter-disciplinarian} (pp. XX--XX). New York, NY: Berghahn.
-
-\hypertarget{ref-chiat2007preschool}{}
-Chiat, S., \& Roy, P. (2007). The preschool repetition test: An
-evaluation of performance in typically developing and clinically
-referred children. \emph{Journal of Speech, Language, and Hearing
-Research}, \emph{50}(2), 429--443.
-
-\hypertarget{ref-cristia2020infant}{}
+\leavevmode\hypertarget{ref-is08042009language}{}%
+Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
+\leavevmode\hypertarget{ref-balladares2016socio}{}%
+Balladares, J., Marshall, C., \& Griffiths, Y.
+(2016).
+Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
+\emph{First Language}, \emph{36}(3), 338--351.
+\url{https://doi.org/10.1177/0142723715626067}
+
+\leavevmode\hypertarget{ref-bowey2001nonword}{}%
+Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
+
+\leavevmode\hypertarget{ref-brandeker2015language}{}%
+Brandeker, M., \& Thordardottir, E.
+(2015).
+Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
+
+\leavevmode\hypertarget{ref-brownIPchildrearing}{}%
+Brown, P., \& Casillas, M.
+(n.d.).
+Childrearing through social interaction on Rossel Island, PNG.
+In A.
+J.
+Fentiman \& M.
+Goody (Eds.), \emph{Esther Goody revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
+XX--XX).
+New York, NY: Berghahn.
+
+\leavevmode\hypertarget{ref-chiat2007preschool}{}%
+Chiat, S., \& Roy, P.
+(2007).
+The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
+
+\leavevmode\hypertarget{ref-cristia2020infant}{}%
 Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
-(2020). Infant-directed input and literacy effects on phonological
-processing: Non-word repetition scores among the tsimane'.
+(2020).
+Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
 \emph{Preprint}.
 
-\hypertarget{ref-estes2007differences}{}
-Estes, K. G., Evans, J. L., \& Else-Quest, N. M. (2007). Differences in
-the nonword repetition performance of children with and without specific
-language impairment: A meta-analysis. \emph{Journal of Speech, Language,
-and Hearing Research}, \emph{50}(1), 177--195.
-
-\hypertarget{ref-farmani2018normalization}{}
-Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \&
-Shourvazi, Z. (2018). Normalization of the non-word repetition test in
-farsi-speaking children. \emph{Journal of Modern Rehabilitation},
-\emph{12}(4), 217--224.
-
-\hypertarget{ref-gallagher2014identity}{}
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from
-Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378.
-doi:\href{https://doi.org/10.1515/lp-2014-0012}{10.1515/lp-2014-0012}
-
-\hypertarget{ref-gallon2007non}{}
-Gallon, N., Harris, J., \& Van der Lely, H. (2007). Non-word repetition:
-An investigation of phonological complexity in children with grammatical
-sli. \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
-
-\hypertarget{ref-havron2019effect}{}
-Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \&
-Group, E. M.-C. C. S. (2019). The effect of older siblings on language
-development as a function of age difference and sex. \emph{Psychological
-Science}, \emph{30}(9), 1333--1343.
-
-\hypertarget{ref-hazan2000development}{}
-Hazan, V., \& Barrett, S. (2000). The development of phonemic
-categorization in children aged 6--12. \emph{Journal of Phonetics},
-\emph{28}(4), 377--396.
-
-\hypertarget{ref-jabere2018xperiment}{}
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual
-Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4),
-317--334.
-doi:\href{https://doi.org/10.30958/ajp.5-4-4}{10.30958/ajp.5-4-4}
-
-\hypertarget{ref-kalnak2014nonword}{}
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014).
-Nonword repetition--a clinical marker for specific language impairment
-in swedish associated with parents' language-related problems.
+\leavevmode\hypertarget{ref-estes2007differences}{}%
+Estes, K.
+G., Evans, J.
+L., \& Else-Quest, N.
+M.
+(2007).
+Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
+
+\leavevmode\hypertarget{ref-farmani2018normalization}{}%
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
+Z., Talebi, E., \& Shourvazi, Z.
+(2018).
+Normalization of the non-word repetition test in farsi-speaking children.
+\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
+
+\leavevmode\hypertarget{ref-gallagher2014identity}{}%
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
+
+\leavevmode\hypertarget{ref-gallon2007non}{}%
+Gallon, N., Harris, J., \& Van der Lely, H.
+(2007).
+Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
+\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\leavevmode\hypertarget{ref-havron2019effect}{}%
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
+M.-C.
+C.
+S.
+(2019).
+The effect of older siblings on language development as a function of age difference and sex.
+\emph{Psychological Science}, \emph{30}(9), 1333--1343.
+
+\leavevmode\hypertarget{ref-hazan2000development}{}%
+Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
+
+\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
+
+\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
+(2014).
+Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
 \emph{PloS One}, \emph{9}(2), e89544.
 
-\hypertarget{ref-levinsoninprepydgrammar}{}
-Levinson, S. C. (in preparation). \emph{A grammar of yélî dnye, the
-papuan language of rossel island}. Cambridge, UK: Cambridge University
-Press.
-
-\hypertarget{ref-meir2017independent}{}
-Meir, N., \& Armon-Lotem, S. (2017). Independent and combined effects of
-socioeconomic status (ses) and bilingualism on children's vocabulary and
-verbal short-term memory. \emph{Frontiers in Psychology}, \emph{8},
-1442.
-
-\hypertarget{ref-meir2016disentangling}{}
-Meir, N., Walters, J., \& Armon-Lotem, S. (2016). Disentangling sli and
-bilingualism using sentence repetition tasks: The impact of l1 and l2
-properties. \emph{International Journal of Bilingualism}, \emph{20}(4),
-421--452.
-
-\hypertarget{ref-phoible}{}
-Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max
-Planck Institute for the Science of Human History. Retrieved from
-\url{https://phoible.org/}
-
-\hypertarget{ref-piazzalunga2019articulatory}{}
-Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler,
-A. (2019). An articulatory-based disyllabic and trisyllabic non-word
-repetition test: Reliability and validity in italian 3-to 7-year-old
-children. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5),
-437--456.
-
-\hypertarget{ref-torrington2015non}{}
-Torrington Eaton, C., Newman, R. S., Ratner, N. B., \& Rowe, M. L.
-(2015). Non-word repetition in 2-year-olds: Replication of an adapted
-paradigm and a useful methodological extension. \emph{Clinical
-Linguistics \& Phonetics}, \emph{29}(7), 523--535.
-
-\hypertarget{ref-vance2005speech}{}
-Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills
-in children aged 3--7 years. \emph{International Journal of Language \&
-Communication Disorders}, \emph{40}(1), 29--48.
-
-\hypertarget{ref-wilsenach2013phonological}{}
-Wilsenach, C. (2013). Phonological skills as predictor of reading
-success: An investigation of emergent bilingual Northern Sotho/English
-learners. \emph{Per Linguam: a Journal of Language Learning= Per
-Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
-doi:\href{https://doi.org/10.5785/29-2-554}{10.5785/29-2-554}
-
-\hypertarget{ref-is08042009language}{}
-Action, C. (2009). Language impairment in a multilingual society:
-Linguistic patterns and the road to assessment. \emph{Brussels: COST
-Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
-\hypertarget{ref-balladares2016socio}{}
-Balladares, J., Marshall, C., \& Griffiths, Y. (2016). Socio-economic
-status affects sentence repetition, but not non-word repetition, in
-Chilean preschoolers. \emph{First Language}, \emph{36}(3), 338--351.
-doi:\href{https://doi.org/10.1177/0142723715626067}{10.1177/0142723715626067}
-
-\hypertarget{ref-bowey2001nonword}{}
-Bowey, J. A. (2001). Nonword repetition and young children's receptive
-vocabulary: A longitudinal study. \emph{Applied Psycholinguistics},
-\emph{22}(3), 441--469.
-
-\hypertarget{ref-brandeker2015language}{}
-Brandeker, M., \& Thordardottir, E. (2015). Language exposure in
-bilingual toddlers: Performance on nonword repetition and lexical tasks.
-\emph{American Journal of Speech-Language Pathology}, \emph{24}(2),
-126--138.
-
-\hypertarget{ref-brownIPchildrearing}{}
-Brown, P., \& Casillas, M. (in press). Childrearing through social
-interaction on Rossel Island, PNG. In A. J. Fentiman \& M. Goody (Eds.),
-\emph{Esther Goody revisited: Exploring the legacy of an original
-inter-disciplinarian} (pp. XX--XX). New York, NY: Berghahn.
-
-\hypertarget{ref-chiat2007preschool}{}
-Chiat, S., \& Roy, P. (2007). The preschool repetition test: An
-evaluation of performance in typically developing and clinically
-referred children. \emph{Journal of Speech, Language, and Hearing
-Research}, \emph{50}(2), 429--443.
-
-\hypertarget{ref-cristia2020infant}{}
-Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
-(2020). Infant-directed input and literacy effects on phonological
-processing: Non-word repetition scores among the tsimane'.
-\emph{Preprint}.
-
-\hypertarget{ref-estes2007differences}{}
-Estes, K. G., Evans, J. L., \& Else-Quest, N. M. (2007). Differences in
-the nonword repetition performance of children with and without specific
-language impairment: A meta-analysis. \emph{Journal of Speech, Language,
-and Hearing Research}, \emph{50}(1), 177--195.
-
-\hypertarget{ref-farmani2018normalization}{}
-Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \&
-Shourvazi, Z. (2018). Normalization of the non-word repetition test in
-farsi-speaking children. \emph{Journal of Modern Rehabilitation},
-\emph{12}(4), 217--224.
-
-\hypertarget{ref-gallagher2014identity}{}
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from
-Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378.
-doi:\href{https://doi.org/10.1515/lp-2014-0012}{10.1515/lp-2014-0012}
-
-\hypertarget{ref-gallon2007non}{}
-Gallon, N., Harris, J., \& Van der Lely, H. (2007). Non-word repetition:
-An investigation of phonological complexity in children with grammatical
-sli. \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
-
-\hypertarget{ref-havron2019effect}{}
-Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \&
-Group, E. M.-C. C. S. (2019). The effect of older siblings on language
-development as a function of age difference and sex. \emph{Psychological
-Science}, \emph{30}(9), 1333--1343.
-
-\hypertarget{ref-hazan2000development}{}
-Hazan, V., \& Barrett, S. (2000). The development of phonemic
-categorization in children aged 6--12. \emph{Journal of Phonetics},
-\emph{28}(4), 377--396.
-
-\hypertarget{ref-jabere2018xperiment}{}
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual
-Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4),
-317--334.
-doi:\href{https://doi.org/10.30958/ajp.5-4-4}{10.30958/ajp.5-4-4}
-
-\hypertarget{ref-kalnak2014nonword}{}
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014).
-Nonword repetition--a clinical marker for specific language impairment
-in swedish associated with parents' language-related problems.
-\emph{PloS One}, \emph{9}(2), e89544.
-
-\hypertarget{ref-levinsoninprepydgrammar}{}
-Levinson, S. C. (in preparation). \emph{A grammar of yélî dnye, the
-papuan language of rossel island}. Cambridge, UK: Cambridge University
-Press.
-
-\hypertarget{ref-meir2017independent}{}
-Meir, N., \& Armon-Lotem, S. (2017). Independent and combined effects of
-socioeconomic status (ses) and bilingualism on children's vocabulary and
-verbal short-term memory. \emph{Frontiers in Psychology}, \emph{8},
-1442.
-
-\hypertarget{ref-meir2016disentangling}{}
-Meir, N., Walters, J., \& Armon-Lotem, S. (2016). Disentangling sli and
-bilingualism using sentence repetition tasks: The impact of l1 and l2
-properties. \emph{International Journal of Bilingualism}, \emph{20}(4),
-421--452.
-
-\hypertarget{ref-phoible}{}
-Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max
-Planck Institute for the Science of Human History. Retrieved from
-\url{https://phoible.org/}
-
-\hypertarget{ref-piazzalunga2019articulatory}{}
-Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler,
-A. (2019). An articulatory-based disyllabic and trisyllabic non-word
-repetition test: Reliability and validity in italian 3-to 7-year-old
-children. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5),
-437--456.
-
-\hypertarget{ref-torrington2015non}{}
-Torrington Eaton, C., Newman, R. S., Ratner, N. B., \& Rowe, M. L.
-(2015). Non-word repetition in 2-year-olds: Replication of an adapted
-paradigm and a useful methodological extension. \emph{Clinical
-Linguistics \& Phonetics}, \emph{29}(7), 523--535.
-
-\hypertarget{ref-vance2005speech}{}
-Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills
-in children aged 3--7 years. \emph{International Journal of Language \&
-Communication Disorders}, \emph{40}(1), 29--48.
-
-\hypertarget{ref-wilsenach2013phonological}{}
-Wilsenach, C. (2013). Phonological skills as predictor of reading
-success: An investigation of emergent bilingual Northern Sotho/English
-learners. \emph{Per Linguam: a Journal of Language Learning= Per
-Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
-doi:\href{https://doi.org/10.5785/29-2-554}{10.5785/29-2-554}
+\leavevmode\hypertarget{ref-levinsoninprepydgrammar}{}%
+Levinson, S. C. (n.d.). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Cambridge, UK: Cambridge University Press.
+
+\leavevmode\hypertarget{ref-meir2017independent}{}%
+Meir, N., \& Armon-Lotem, S.
+(2017).
+Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
+\emph{Frontiers in Psychology}, \emph{8}, 1442.
+
+\leavevmode\hypertarget{ref-meir2016disentangling}{}%
+Meir, N., Walters, J., \& Armon-Lotem, S.
+(2016).
+Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
+\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
+
+\leavevmode\hypertarget{ref-phoible}{}%
+Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
+
+\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
+(2019).
+An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
+\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
+
+\leavevmode\hypertarget{ref-torrington2015non}{}%
+Torrington Eaton, C., Newman, R.
+S., Ratner, N.
+B., \& Rowe, M.
+L.
+(2015).
+Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
+\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+
+\leavevmode\hypertarget{ref-vance2005speech}{}%
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
+
+\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
+Wilsenach, C.
+(2013).
+Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
+\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+\url{https://doi.org/10.5785/29-2-554}
 

### 2020-09-11 manuscript.tex
update tex and pdf

--- 2020-09-11 manuscript.tex
@@
 the average cross-linguistic frequency of phones in the target words (ß
-= 0.03, SE ß = 0.01, p = 0): Target words with phones found more
+= 0.77, SE ß = 0.19, p = 0): Target words with phones found more
 frequently across languages had a higher proportion of words that were
-correctly repeated, as clear in Figure 2.
+correctly repeated, as clear in Figure 2. Age was also a significant
+predictor (ß = 0.31, SE ß = 0.13, p = 0.02).
 
@@
 \end{figure}
+
+We next checked whether this may actually be due to frequency of the
+sounds within the language. We estimated frequency of the phones present
+in the stimuli in a corpus of child-centered recordings
+({\textbf{???}}), and applied the natural logarithm. As before,
+unattested sounds were not considered (i.e., they were declared NA so
+that they do not count for analyses). Phone frequency estimated from
+this corpus were correlated with cross-linguistic phone frequencies
+(r(23)=0.65, =0). Therefore, we fit another mixed logistic regression,
+this time declaring as fixed effects scaled cross-linguistic and corpus
+frequencies (averaged across all attested phones within each stimulus
+item), in addition to age. As before, the model contained random slopes
+for both child ID and target. In this model, both cross-linguistic
+frequency (ß = 0.50, SE ß = 0.25, p = 0.04) and age (ß = 0.31, SE ß =
+0.13, p = 0.02) were significant predictors of accuracy, but corpus
+frequency (ß = 0.37, SE ß = 0.24, p = 0.12) was not.
 
@@
 
-\subsubsection{Accuracy a function of item
-length}\label{accuracy-a-function-of-item-length}
+\subsubsection{Accuracy as a function of item
+length}\label{accuracy-as-a-function-of-item-length}
 

### 2020-09-11 manuscript.Rmd
add minor notes on school

@@
 - look through this manuscript for bits of text to be written by **mc**
-- help with table 1 (formatting & making it appear, because it's broken by the phonological representations)
 - do you think Figure 2 should be done with phonetic symbols? If yes, could you try to implement that?
@@
 
-myplot=ggplot(data = pl, aes(x = Age, y = Score, color = Language))+
+myplot = ggplot(data = pl, aes(x = Age, y = Score, color = Language))+
   geom_point(aes(size = n, shape = Length))+
   geom_line(aes(linetype=Length)) + 
-  theme(text = element_text(size = 16), legend.position="none", axis.text.x=element_text(colour=c("black", "black"))) +
+  theme(text = element_text(size = 16),
+    legend.position="none", axis.text.x=element_text(colour=c("black", "black"))) +
   scale_color_manual(values = palette)+
@@
 
-# probably a nicer way to do this but anyhow...
+# there's definitely a nicer way to do this but anyhow...
 practice.stims <- filter(stims, Type == "practice")[,2:3]
@@
 
-#now that I added phonological representations, even this basic table chokes!
-# apa_table(stims,"NWR stimuli in orthographic and phonological representations.")
 ```
@@
 #summary(inc)
-table(inc$birthOrder)->mytab
+table(inc$birthOrder)->birthorder.tab
 ```
@@
 A total `r dim(demo)[1]` children were tested, from `r length(levels(factor(demo$familyID)))` of families, in five hamlets.
-Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xd5t repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`).
+Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn't repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`).
 In addition, `r sum(demo$Age >12 | demo$included=="too old",na.rm=T)` teenagers were tested to put younger children at ease; their data is not included in analyses below.
 The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years).
-There were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and , `r sum(is.na(inc$monolingual))` for whom this information was missing.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing).[^1] In terms of birth order, `r mytab[1]` were first borns, `r mytab[2]` second,  `r mytab[3]` third,  `r mytab[4]` forth,  `r mytab[5]` fifth, `r mytab[6]` sixth, and  `r mytab[7]` did not have this information.
+Among these, there were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and `r sum(is.na(inc$monolingual))` for whom this information is missing.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing).[^1] In terms of birth order, `r birthorder.tab[1]` were first borns, `r birthorder.tab[2]` second,  `r birthorder.tab[3]` third,  `r birthorder.tab[4]` forth,  `r birthorder.tab[5]` fifth, `r birthorder.tab[6]` sixth, and  `r birthorder.tab[7]` did not have this information.
 
-[^1]: Education is often reported in even years because people typically complete two-year cycles.
+[^1]: Most people complete their education upon reaching the highest level of a schooling type (e.g., elementary, primary, secondary etc.).
+This means that most people have 3, 6, 8, 10, or 12 years of education, with few stopping in-between (e.g., 7 years or 9 years).
+That said, precise years of education are hard to count given an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
+We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions.
 
@@
 ```{r tablength}
-mytab=NULL
+birthorder.tab=NULL
 for(myvar in c("Word","Phoneme","NLD")){
-  mytab=cbind(mytab,
+  birthorder.tab=cbind(birthorder.tab,
               paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
@@
 }
-rownames(mytab)<-levels(by_length$Length)
-colnames(mytab)<-c("Word","Phoneme","NLD")
-apa_table(mytab,caption="NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
+rownames(birthorder.tab)<-levels(by_length$Length)
+colnames(birthorder.tab)<-c("Word","Phoneme","NLD")
+apa_table(birthorder.tab,caption="NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
 ```

### 2020-09-11 manuscript.Rmd
corrected schooling info

@@
 Despite restricted outside contact, formal education is a priority for many Yélî families.
-A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (8 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early].
+A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early].
 While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), and so it is very common for households to billet school children during the weekdays for months at a time (or for whole months if the child's home is very far).
@@
 
-[^1]: Most people complete their education upon reaching the highest level of a schooling type (e.g., elementary, primary, secondary etc.).
-This means that most people have 3, 6, 8, 10, or 12 years of education, with few stopping in-between (e.g., 7 years or 9 years).
-That said, precise years of education are hard to count given an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
-We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions.
+[^1]: Precise years of education can be hard to count.
+This is particularly the case for schooling beyond grade 8, which takes place off-island, due to an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
+We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical "completion" times.
 

### 2020-09-14 manuscript.Rmd
minor footnote updates

@@
 [^1]: Precise years of education can be hard to count.
-This is particularly the case for schooling beyond grade 8, which takes place off-island, due to an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
-We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical "completion" times.
+This is particularly the case for schooling beyond grade 8, which takes place off-island, which can be due to an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
+We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical "completion" times in between.
 

### 2020-09-14 wrangling.R
debug attempt col

@@ -2,6 +2,9 @@
 read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
 #DATA REMOVAL *** ATTENTION *** 
 #REMOVE ALL TASK 4 (whatever that is)
+# MC: This appears to be a second pass over productions that were deemed to
+# be words; the elicitation here is just a translation of what word the
+# production was heard to be
 trnsc[trnsc$item!="task4",]->trnsc
 
 dim(trnsc) #1488 judgments
@@ -18,14 +21,31 @@ dim(trnsc) #haven't lost any judgments
 #order so we have all the items by a given child in the right chrono order
 trnsc[order(trnsc$file, trnsc$int),]->trnsc
 
-#add variable coding whether it's the child's first or subsequent attempts
-trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
-trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
-trnsc$attempt[1]<-"first"
-table(trnsc$attempt) #1045 first attempts
-
 #add repetition number
 trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
+
+#add variable coding whether it's the child's first or subsequent attempts
+trnsc$attempt <- ifelse(trnsc$rep == "1", "first", "subsequent")
+
+# MC's new code
+# first.prods <- trnsc %>%
+#   group_by(id, item) %>%
+#   summarize(min.token = min(token)) %>%
+#   mutate(attempt = "first") %>%
+#   rename("token" = min.token)
+# trnsc <- trnsc %>%
+#   left_join(first.prods, by = c("id", "item", "token")) %>%
+#   replace_na(list(attempt = "subsequent"))
+# table(trnsc$attempt)
+# M2A: by my count there are 1029 first attempts, but maybe I'm misunderstanding "token"
+# M2A: my solution finds some conflicting cases with your outfile attempt namings,
+#      now going off of those for attempt first/subsequent, I still find 1029
+
+# AC's prev code
+# trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
+# trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
+# trnsc$attempt[1]<-"first"
+# table(trnsc$attempt) #1045 first attempts
 
 
 # add demographic info

### 2020-09-14 manuscript.Rmd
rename stims so not to conflict with wrangling.R

@@
 #add orthographic representations targets & other stims chars
-stims <- read_delim("stimuli2.txt", delim = "\t") %>%
+stims.tbl <- read_delim("stimuli2.txt", delim = "\t") %>%
   arrange(factor(type, levels = c("practice", "1", "2", "3", "4"))) %>%
@@
 # there's definitely a nicer way to do this but anyhow...
-practice.stims <- filter(stims, Type == "practice")[,2:3]
+practice.stims <- filter(stims.tbl, Type == "practice")[,2:3]
 practice.stims$row <- seq(1:nrow(practice.stims))
-syll1.stims <- filter(stims, Type == "1")[,2:3]
+syll1.stims <- filter(stims.tbl, Type == "1")[,2:3]
 syll1.stims$row <- seq(1:nrow(syll1.stims))
-syll2.stims <- filter(stims, Type == "2")[,2:3]
+syll2.stims <- filter(stims.tbl, Type == "2")[,2:3]
 syll2.stims$row <- seq(1:nrow(syll2.stims))
-syll3.stims <- filter(stims, Type == "3")[,2:3]
+syll3.stims <- filter(stims.tbl, Type == "3")[,2:3]
 syll3.stims$row <- seq(1:nrow(syll3.stims))
-syll4.stims <- filter(stims, Type == "4")[,2:3]
+syll4.stims <- filter(stims.tbl, Type == "4")[,2:3]
 syll4.stims$row <- seq(1:nrow(syll4.stims))

### 2020-09-14 wrangling.R
add code comments for AC to review

@@ -65,6 +65,8 @@ merge(trnsc,inc,by.x="id",by.y="ID",all.x=T)->trnsc
 dim(trnsc) #still no loss, all good
 
 #add orthographic representations targets & other stims chars
+# M2A: Note that some of these stim characters are now updated in stimuli2.txt
+# they do not match what is shown here
 read.delim("stimconv.txt", encoding="UTF-8")->stims
 read.delim("segconv.txt", encoding="UTF-8")->segments
 #add corpus freq to segments
@@ -78,9 +80,10 @@ write.table(segments,"segments_with_cor_freq.txt",sep="\t",row.names = F)
 #add average crosslinguistic frequency
 stim_seg_freq=matrix(NA,nrow=dim(stims)[1],ncol=8)
 for(i in 1:dim(stim_seg_freq)[1]) for(j in 1:dim(stim_seg_freq)[2]) {
-  thisfreq=segments$pc[as.character(segments$phono)==as.character(stims[i,paste0("phono",j)])] 
+  thisfreq=segments$pc[as.character(segments$phono)==as.character(stims[i,paste0("phono",j)])]
   stim_seg_freq[i,j]<-ifelse(sum(!is.na(thisfreq))==1,thisfreq,NA)
-  }
+}
+# M2A: I'm suspicious of some of the NAs in here (like for 'k', 'w', 'd')--what is segment pc?
 stims$avg_fr=apply(stim_seg_freq,1,mean,na.rm=T)
 
 #add average corpus frequency - LOGGED
@@ -94,18 +97,14 @@ stims$avg_fr_cor=apply(log(stim_seg_freq),1,mean,na.rm=T)
 #add average corpus frequency based on TYPES - LOGGED
 stim_seg_freq=matrix(NA,nrow=dim(stims)[1],ncol=8)
 for(i in 1:dim(stim_seg_freq)[1]) for(j in 1:dim(stim_seg_freq)[2]) {
-  thisfreq=segments$freq_corpus_types[as.character(segments$phono)==as.character(stims[i,paste0("phono",j)])] 
+  thisfreq=segments$freq_corpus_types[
+    as.character(segments$phono)==as.character(stims[i,paste0("phono",j)])] 
   stim_seg_freq[i,j]<-ifelse(sum(!is.na(thisfreq))==1,thisfreq,NA)
 }
 stims$avg_fr_cor_ty=apply(log(stim_seg_freq),1,mean,na.rm=T)
 
[…]

### 2020-09-14 wrangling.R
add more code comments for AC

@@ -2,7 +2,7 @@
 read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
 #DATA REMOVAL *** ATTENTION *** 
 #REMOVE ALL TASK 4 (whatever that is)
-# MC: This appears to be a second pass over productions that were deemed to
+# M2A: This appears to be a second pass over productions that were deemed to
 # be words; the elicitation here is just a translation of what word the
 # production was heard to be
 trnsc[trnsc$item!="task4",]->trnsc
@@ -38,7 +38,7 @@ trnsc$attempt <- ifelse(trnsc$rep == "1", "first", "subsequent")
 #   replace_na(list(attempt = "subsequent"))
 # table(trnsc$attempt)
 # M2A: by my count there are 1029 first attempts, but maybe I'm misunderstanding "token"
-# M2A: my solution finds some conflicting cases with your outfile attempt namings,
+# M2A: okay, my solution finds some conflicting cases with your outfile attempt namings,
 #      now going off of those for attempt first/subsequent, I still find 1029
 
 # AC's prev code
@@ -66,7 +66,7 @@ dim(trnsc) #still no loss, all good
 
 #add orthographic representations targets & other stims chars
 # M2A: Note that some of these stim characters are now updated in stimuli2.txt
-# they do not match what is shown here
+# they do not match what is shown here (!!)
 read.delim("stimconv.txt", encoding="UTF-8")->stims
 read.delim("segconv.txt", encoding="UTF-8")->segments
 #add corpus freq to segments
@@ -130,7 +130,8 @@ trnsc$target_ortho=gsub("([aeiouêâéáóî])\\1+", "\\1", trnsc$target_ortho)
 #sum(trnsc$correct==0 & trnsc$target_ortho == trnsc$mp & !is.na(trnsc$mp)) #this only happens twice
 trnsc$correct[trnsc$correct==0 & trnsc$target_ortho == trnsc$mp & !is.na(trnsc$mp)]<-1
 # M2A: This appears to be only two cases and it doesn't seem to be a length error diff
-# i.e., mp == mispronunciation on both of these -- should we be cautious about counting them as correct?
+# i.e., mp == mispronunciation on both of these -- should we be cautious about
+# counting them as correct?
 
[…]

### 2020-09-14 manuscript.Rmd
add comment for AC to rmd

@@
 #quick fix, but should be fixed upstream @ac
-trnsc$Sex[trnsc$Sex=="F"]<-"G"
+# M2A: Seems like you already did this, and the line below
+# is setting the is.na(trnsc$Sex) cases to "G" -- I assume
+# that's not a desirable outcome so am commenting it out
+# trnsc$Sex[trnsc$Sex=="F"]<-"G"
 trnsc$Sex=factor(trnsc$Sex)

### 2020-09-14 manuscript.Rmd
addressed middys comments

--- 2020-09-14 manuscript.Rmd
@@
    In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
-Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
+In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
 This study contributes to three lines of research.
@@
 
-## TODO later
-- probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set DONE
-- maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed) DONE I think... (this item looks the same as one of the three above)
-- maybe add analyses of score as a function of frequency (phones, diphones) in MC's corpus? (mc suggested, but I don't think we need it...) - waiting until we talk about this
-
 ## Introduction
@@
 NWR has been used to seek answers to a variety of theoretical questions, including what  the links between phonology, working memory, and the lexicon are [@bowey2001nonword], and to assess the extent to which apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity].
-It is also frequently used as for applications, notably  as a diagnostic for language delays and disorders [@estes2007differences].
+It is also frequently used for applications, notably  as a diagnostic for language delays and disorders [@estes2007differences].
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments,  particularly in Europe [@meir2016disentangling;@is08042009language].
+However, it has been seldom used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant].
 
@@
 We should probably have a paragraph justifying its use in this context, no?-->
+<!-- a2m I added a sentence, is this what you were thinking of? -->
 
@@
 Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
-Third, we contribute to the basic NWR literature by contributing an additional data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
+Third, we contribute to the basic NWR literature by adding a data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
 
 ### Yélî Dnye phonological overview
-<!--M2A: is this okay?-->
-Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
 The Yélî sound system, much like its baroque grammatical system [@levinsoninprepydgrammar], is unlike any other in the region.
@@
 
-Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllablic words appearing least frequently [~10%; based on >5800 lexemes in the current Yélî Dnye dictionary, @levinsoninprepydgrammar].
+Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the current Yélî Dnye dictionary, @levinsoninprepydgrammar].
 The vast majority of syllables use a CV format.
@@
 
-While all consonants can appear in the first syllable of the word, in bisyllabic roots, the second syllable onset consonant is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
+<!-- a2m up to here it looks great.
+i am less sure about the next paragraph because it is unclear how it connects with our study...
+we didn't really take this into account when building the stimuli, right?
+-->
+
+While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
 In monosyllabic stems, oral vowels (short: 43% and long: 31%) are more common than nasal ones (short: 4% and long: 22%), with the first syllable of bisyllabic stems following a similar pattern.
@@
 <!--M2A: is this okay?-->
-Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
+<!-- a2m I think it's great! although it seemed a bit unexpected to explain all of this, so I added some justification -->
+
+The current study also contributes to increasing the cultural diversity of the populations for which NWR data have been collected.
+As mentioned previously, this method has been used almost exclusively in urban, industrialized populations, with only one exception to our knowledge [@cristia2020infant].
+We therefore provide some background information tol help understand some aspects of our study, including the estimation of children's ages and the relationship between NWR scores and children's demographic characteristics.
+
+Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7; 
+<!-- a2m no "at" here? *at* around the age? -->
+school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---, where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
 Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
@@
 While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early].
-This increase in Child CDS is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 people) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 people) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
 <!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->
+<!-- a2m maybe we want to bring that up in the discussion?
+I haven't found a good place to introduce that paper, in part because right now we have these three purposes that seem sort of unconnected with the goals of that paper...
+-->
 
@@
 
-There is some variation in the presentation procedure and structure of items found in previous NWR work.
+The basic structure of NWR involves the participant hearing a non-word, like "bilik", which they need to repeat back without changing anything that is contrastive in the language.
+For instance, in English, a response of "bilig" or "pilik" will be scored as incorrect; a response "biilik", where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.
+<!-- a2m added explanation of correct and incorrect for nwr task -->
+
+Aside from this central basic structure, there is some variation in the presentation procedure and structure of items found in previous NWR work.
 For example, items are often presented orally by the experimenter [@torrington2015non], although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation [@brandeker2015language].
@@
 
-Naturally, designing the task in this way may render the study of individual variaiton within the population  more difficult because different children are exposed to different items.
-However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
+Naturally, designing the task in this way may render the study of individual variation within the population  more difficult because different children are exposed to different items.
+However, a review of previous work on individual variation suggested to us that effects of individual differences  are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
 
@@
 Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization;@kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool].
-Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on those too.
+Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on those too.
 
@@
 - Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
+<!-- a2m we don't actually look at this: Are rarer sounds more often substituted by commoner sounds? -- shall we remove it? -->
 <!--M2A: I'm imagining a reviewer who's knowlegeable about phonology and child development but who hasn't encountered NWR before.
 I think it would be beneficial to provide a concrete example of a nonword and an inexact repetition above so that they can conceive of what these questions imply (analytically) before we get into the methodological weeds.-->
+<!-- a2m good point, but I think it goes in the previous section so I added it there -->
 
@@
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described [CITE **mc** please fill in], there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described [@levinsoninprepydgrammar], there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
@@
 Although previous work does not use distance metrics, we additionally report those.
+For instance, recall our example of an English target being "bilik", and imagine a response "bilig".
+This would receive the following scores: at the whole item level this would receive a score of zero (because the repetition is not exact); at the phoneme level it would receive a score of 80% (4 out of 5 phonemes repeated exactly); and its normalized phone-based Levenshtein distance is 20% (20% of phonemes were substituted or deleted).
 <!--M2A: A concrete scoring example would be a good idea here-->
+<!-- a2m added! actually, I haven't changed anything but I've been hesitant about using levenshtein's distance because it's the complement of phoneme based scoring -- so shall we drop it?  -->
 
@@
 Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); and classify the most common phonological errors.
+We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.
+<!-- a2m in the results, I put analyses based on all as well as just on first attempt. If we do just the latter, then come back to change this para -->
 

### 2020-09-14 manuscript.tex
addressed middys comments

--- 2020-09-11 manuscript.tex
@@
             pdftitle={Non-word repetition in Yélî Dnye},
+            pdfauthor={Alejandrina Cristia~\& Marisa Casillas},
             pdfkeywords={phonology, non-word repetition, development},
@@
 \setmainfont{Doulos SIL} % Set main font to Doulos SIL
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\makeatother
-\shorttitle{NWR in Yélî Dnye}
-\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2}}
-\affiliation{
-\vspace{0.5cm}
-\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
-\authornote{All data are made available in a repository in the Open Science Framework.
-AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
-S.
-McDonnell Foundation Understanding Human Cognition Scholar Award.
-
-Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
-\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
 \title{Non-word repetition in Yélî Dnye}
+\author{Alejandrina Cristia\textsuperscript{1}~\& Marisa Casillas\textsuperscript{2}}
+\date{}
 
-\date{}
+\authornote{All data are made available in a repository in the Open Science Framework.
+AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
+S.
+McDonnell Foundation Understanding Human Cognition Scholar Award.
+
+Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d'Ulm, 75005 Paris, France. E-mail: \href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
 
@@
 In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
-Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
+In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
 This study contributes to three lines of research.
@@
 Together, these data provide a unique view of online phonological processing in a seldom-studied language, and contribute to both typological and language acquisition research.
+
 }
@@
   look through this manuscript for bits of text to be written by \textbf{mc}
-\item
-  help with table 1 (formatting \& making it appear, because it's broken by the phonological representations)
 \item
@@
 \item
-  consent, \enquote{correct}, \enquote{accuracy}, \enquote{performance}
+  consent, ``correct'', ``accuracy'', ``performance''
 \item
   draft discussion
-\end{itemize}
-
-\hypertarget{todo-later}{%
-\subsection{TODO later}\label{todo-later}}
-
-\begin{itemize}
-\tightlist
-\item
-  probably turn summary of types of error into a table; use proportion of correctly produced rather than number; use first attempts rather than the whole data set DONE
-\item
-  maybe add to the last ana the prevalence (ie out of all the gh which proportion get transformed) DONE I think\ldots{} (this item looks the same as one of the three above)
-\item
-  maybe add analyses of score as a function of frequency (phones, diphones) in MC's corpus? (mc suggested, but I don't think we need it\ldots{}) - waiting until we talk about this
 \end{itemize}
@@
 NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and to assess the extent to which apparent phonological constraints found in the lexicon affect online production (Gallagher, 2014).
-It is also frequently used as for applications, notably as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
+It is also frequently used for applications, notably as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
+However, it has been seldom used outside of Europe and North America (with exceptions including Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)).
 
@@
 Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
-Third, we contribute to the basic NWR literature by contributing an additional data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
+Third, we contribute to the basic NWR literature by adding a data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
 
@@
 
-Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
 The Yélî sound system, much like its baroque grammatical system (Levinson, n.d.), is unlike any other in the region.
@@
 
-Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllablic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the current Yélî Dnye dictionary, Levinson, n.d.).
+Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the current Yélî Dnye dictionary, Levinson, n.d.).
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., \enquote{ndapî} /\enquote{ṇṭæ.pɯ/).
-This process is used frequently with English loan words that have a coda (e.g., 'ponî} (phone) /\enquote{pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala} (here) /æ.'læ/), but these are limited to /æ/-.
+A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/).
+This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /`pɔ.nɯ/).
+An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
 Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (n.d.) for details).
 
-While all consonants can appear in the first syllable of the word, in bisyllabic roots, the second syllable onset consonant is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
+While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
 In monosyllabic stems, oral vowels (short: 43\% and long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%), with the first syllable of bisyllabic stems following a similar pattern.
@@
 
-\hypertarget{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}{%
-\subsubsection{\texorpdfstring{Intro to the people (\textbf{mc}) - please feel free to throw away anything that is not useful!}{Intro to the people (mc) - please feel free to throw away anything that is not useful!}}\label{intro-to-the-people-mc---please-feel-free-to-throw-away-anything-that-is-not-useful}}
+\hypertarget{the-yuxe9luxee-community}{%
+\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}}
 
-Most speakers of the language grow up speaking it monolingually within their swidden horticulturalist communities until they begin attending school around the age of 7; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.), the same region in which the current data were collected.
+The current study also contributes to increasing the cultural diversity of the populations for which NWR data have been collected.
+As mentioned previously, this method has been used almost exclusively in urban, industrialized populations, with only one exception to our knowledge (Cristia et al., 2020).
+We therefore provide some background information tol help understand some aspects of our study, including the estimation of children's ages and the relationship between NWR scores and children's demographic characteristics.
+
+Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7;
+school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---, where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
 Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
 
-Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.
+Most Yélî people are swidden horticulturalists, raising a variety of tubers (e.g., yams, manioc, taro), coconut, banana, and more in their gardens.
+In addition to garden foods, Yélî children grow up eating fish and foraging for shellfish and nuts, which provide a regular source of extra protein.
+The typical household in our dataset includes seven individuals and is situated among a collection of 4 or more other households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
 
-\begin{itemize}
-\tightlist
-\item
-  usually monolingual at home
-\item
-  schooling in English but it starts at age XX, so not relevant here
-\item
-  however, some use of English due to immigrants \& children of immigrants
-\item
-  children spend a lot of time with other children
-\item
-  most parents are subsistence farmers
-\item
-  parental education generally varies between XX and YY
-\end{itemize}
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by barrier reef, transport to and from the island is both infrequent and irregular.
+There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport, when there is deisel available.
+There is a medical radio connection at the health clinic (at the Catholic mission) via which news and messages are often passed along.
+There is also a cellular tower on the island, though it is typically not functioning.
+As such, access to news and outside resources, particularly international connections and larger data transfers, is typically unavailable.
+Our data collection is therefore limited to the duration of the researchers' on-island visits.
+
+Despite restricted outside contact, formal education is a priority for many Yélî families.
+A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65\% of mothers and 49\% of fathers completing the full range of locally available education (\textasciitilde{}9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG (Casillas, Brown, \& Levinson, 2020).
+While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), and so it is very common for households to billet school children during the weekdays for months at a time (or for whole months if the child's home is very far).
+Combined with practices of collective child guardianship within their close-knit hamlets, the practical outcome for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's parents---to this child assent is also culturally pertinent, as child independence is encouraged and respected from toddlerhood onward (Brown \& Casillas, n.d.).\footnote{Note that in a few cases this does mean that precise details such as child date of birth, parent age in years, and parent education was only able to be approximated as the parents were residing on another part of the island.} Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
+
+While work on Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas et al., 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun {[}e.g., Bram's thesis; MC's CFH data{]}.
+What we do know about the early language environment is that, while child-directed speech is relatively infrequent throughout the day (Casillas et al., 2020), shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children (Bunce et al., n.d.; Casillas et al., 2020).
+While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere (Bergelson et al., 2019; Bunce et al., n.d.), there is a significant and marked increase in child-directed speech from other children as infants get older (Bunce et al., n.d.; Casillas et al., 2020).
+This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 people) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
 
@@
 
-There is some variation in the presentation procedure and structure of items found in previous NWR work.
+The basic structure of NWR involves the participant hearing a non-word, like ``bilik'', which they need to repeat back without changing anything that is contrastive in the language.
+For instance, in English, a response of ``bilig'' or ``pilik'' will be scored as incorrect; a response ``biilik'', where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.
+
+Aside from this central basic structure, there is some variation in the presentation procedure and structure of items found in previous NWR work.
 For example, items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation (Brandeker \& Thordardottir, 2015).
@@
 
-Naturally, designing the task in this way may render the study of individual variaiton within the population more difficult because different children are exposed to different items.
-However, a review of previous work on individual variation suggested to us that many individual differences effects are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
+Naturally, designing the task in this way may render the study of individual variation within the population more difficult because different children are exposed to different items.
+However, a review of previous work on individual variation suggested to us that effects of individual differences are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
 
@@
 Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
-Although previous research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on those too.
+Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on those too.
 
@@
 To illustrate this, we portray studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables) in Figure 1.
-Notice that the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Tsimane' children (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020), where a drop of 40 percentage points is observed at all ages.
+Notice that the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Tsimane' children (Cristia et al., 2020), where a drop of 40 percentage points is observed at all ages.
 A similar difference in NWR scores for short versus long items was observed among Arabic children (Jaber-Awida, 2018).
@@
 
-A first list of candidate items was generated in 2018 by selecting simple consonants (\enquote{p}, \enquote{t}, \enquote{d}, \enquote{k}, \enquote{m}, \enquote{n}, \enquote{w}, \enquote{y}) and vowels (\enquote{i}, \enquote{o},\enquote{u}, \enquote{a},\enquote{e}) that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
+A first list of candidate items was generated in 2018 by selecting simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'', ``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
 These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
@@
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described {[}CITE \textbf{mc} please fill in{]}, there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described (Levinson, n.d.), there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
@@
 The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
-Even so, there was one monosyllable that was often identified as a real word (intended \enquote{yî} /yXX/; identified as \enquote{yi} /yi/, \emph{tree}).
+Even so, there was one monosyllable that was often identified as a real word (intended ``yî'' /yXX/; identified as ``yi'' /yi/, \emph{tree}).
 This item is removed from analyses below.
@@
 
-We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items \enquote{spreading} through hearsay.
+We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items `spreading' through hearsay.
 In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
@@
 Although previous work does not use distance metrics, we additionally report those.
+For instance, recall our example of an English target being ``bilik'', and imagine a response ``bilig''.
+This would receive the following scores: at the whole item level this would receive a score of zero (because the repetition is not exact); at the phoneme level it would receive a score of 80\% (4 out of 5 phonemes repeated exactly); and its normalized phone-based Levenshtein distance is 20\% (20\% of phonemes were substituted or deleted).\\
 
@@
 Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); and classify the most common phonological errors.
+We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.
 
@@
 A total 55 children were tested, from 34 of families, in five hamlets.
-Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=0), spoke too softly to allow offline coding (N=5).
+Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5).
 In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
 The remaining 40 children (14 girls) were aged 6.96 years (range 3.90-11 years).
-There were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and , 2 for whom this information was missing.
-Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Education is often reported in even years because people typically complete two-year cycles.} In terms of birth order, 5 were first borns, 4 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
+Among these, there were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and 2 for whom this information is missing.
+Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Precise years of education can be hard to count.
+This is particularly the case for schooling beyond grade 8, which takes place off-island, which can be due to an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
+We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical ``completion'' times in between.} In terms of birth order, 5 were first borns, 4 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
 
@@
 
+\leavevmode\hypertarget{ref-bergelsoncasillas2019what}{}%
+Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
+S., \& Amatuni, A.
+(2019).
+What do North American babies hear?
+A large-scale cross-corpus analysis.
+\emph{Developmental Science}, \emph{22}(1), e12724.
+\url{https://doi.org/10.1111/desc.12724}
+
 \leavevmode\hypertarget{ref-bowey2001nonword}{}%
@@
 \emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
+
+\leavevmode\hypertarget{ref-brown2011cultural}{}%
+Brown, P.
+(2011).
+The cultural organization of attention.
+In A.
+Duranti, E.
+Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language Socialization} (pp.
+29--55).
+Malden, MA: Wiley-Blackwell.
+
+\leavevmode\hypertarget{ref-brown2014interactional}{}%
+Brown, P.
+(2014).
+The interactional context of language learning in Tzeltal.
+In I.
+Arnon, M.
+Casillas, C.
+Kurumada, \& B.
+Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of Eve V.
+Clark} (pp.
+51--82).
+Amsterdam, NL: John Benjamins.
 
@@
 New York, NY: Berghahn.
+
+\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
+Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (n.d.). A cross-cultural examination of young children's everyday language experiences.
+
+\leavevmode\hypertarget{ref-casillas2020early}{}%
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
 
@@
 
+\leavevmode\hypertarget{ref-liszkowski2012prelinguistic}{}%
+Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
+(2012).
+A prelinguistic gestural universal of human communication.
+\emph{Cognitive Science}, \emph{36}(4), 698--713.
+\url{https://doi.org/10.1111/j.1551-6709.2011.01228.x}
+
+\leavevmode\hypertarget{ref-meir2017independent}{}%
+Meir, N., \& Armon-Lotem, S.
+(2017).
+Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
+\emph{Frontiers in Psychology}, \emph{8}, 1442.
+
+\leavevmode\hypertarget{ref-meir2016disentangling}{}%
+Meir, N., Walters, J., \& Armon-Lotem, S.
+(2016).
+Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
+\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
+
+\leavevmode\hypertarget{ref-phoible}{}%
+Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
+
+\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
+(2019).
+An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
+\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
+
+\leavevmode\hypertarget{ref-torrington2015non}{}%
+Torrington Eaton, C., Newman, R.
+S., Ratner, N.
+B., \& Rowe, M.
+L.
+(2015).
+Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
+\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+
+\leavevmode\hypertarget{ref-vance2005speech}{}%
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
+
+\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
+Wilsenach, C.
+(2013).
+Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
+\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+\url{https://doi.org/10.5785/29-2-554}
+
+\leavevmode\hypertarget{ref-is08042009language}{}%
+Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
+\leavevmode\hypertarget{ref-balladares2016socio}{}%
+Balladares, J., Marshall, C., \& Griffiths, Y.
+(2016).
+Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
+\emph{First Language}, \emph{36}(3), 338--351.
+\url{https://doi.org/10.1177/0142723715626067}
+
+\leavevmode\hypertarget{ref-bergelsoncasillas2019what}{}%
+Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
+S., \& Amatuni, A.
+(2019).
+What do North American babies hear?
+A large-scale cross-corpus analysis.
+\emph{Developmental Science}, \emph{22}(1), e12724.
+\url{https://doi.org/10.1111/desc.12724}
+
+\leavevmode\hypertarget{ref-bowey2001nonword}{}%
+Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
+
+\leavevmode\hypertarget{ref-brandeker2015language}{}%
+Brandeker, M., \& Thordardottir, E.
+(2015).
+Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
+
+\leavevmode\hypertarget{ref-brown2011cultural}{}%
+Brown, P.
+(2011).
+The cultural organization of attention.
+In A.
+Duranti, E.
+Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language Socialization} (pp.
+29--55).
+Malden, MA: Wiley-Blackwell.
+
+\leavevmode\hypertarget{ref-brown2014interactional}{}%
+Brown, P.
+(2014).
+The interactional context of language learning in Tzeltal.
+In I.
+Arnon, M.
+Casillas, C.
+Kurumada, \& B.
+Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of Eve V.
+Clark} (pp.
+51--82).
+Amsterdam, NL: John Benjamins.
+
+\leavevmode\hypertarget{ref-brownIPchildrearing}{}%
+Brown, P., \& Casillas, M.
+(n.d.).
+Childrearing through social interaction on Rossel Island, PNG.
+In A.
+J.
+Fentiman \& M.
+Goody (Eds.), \emph{Esther Goody revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
+XX--XX).
+New York, NY: Berghahn.
+
+\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
+Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (n.d.). A cross-cultural examination of young children's everyday language experiences.
+
+\leavevmode\hypertarget{ref-casillas2020early}{}%
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
+
+\leavevmode\hypertarget{ref-chiat2007preschool}{}%
+Chiat, S., \& Roy, P.
+(2007).
+The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
+
+\leavevmode\hypertarget{ref-cristia2020infant}{}%
+Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
+(2020).
+Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
+\emph{Preprint}.
+
+\leavevmode\hypertarget{ref-estes2007differences}{}%
+Estes, K.
+G., Evans, J.
+L., \& Else-Quest, N.
+M.
+(2007).
+Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
+
+\leavevmode\hypertarget{ref-farmani2018normalization}{}%
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
+Z., Talebi, E., \& Shourvazi, Z.
+(2018).
+Normalization of the non-word repetition test in farsi-speaking children.
+\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
+
+\leavevmode\hypertarget{ref-gallagher2014identity}{}%
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
+
+\leavevmode\hypertarget{ref-gallon2007non}{}%
+Gallon, N., Harris, J., \& Van der Lely, H.
+(2007).
+Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
+\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\leavevmode\hypertarget{ref-havron2019effect}{}%
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
+M.-C.
+C.
+S.
+(2019).
+The effect of older siblings on language development as a function of age difference and sex.
+\emph{Psychological Science}, \emph{30}(9), 1333--1343.
+
+\leavevmode\hypertarget{ref-hazan2000development}{}%
+Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
+
+\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
+
+\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
+(2014).
+Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
+\emph{PloS One}, \emph{9}(2), e89544.
+
+\leavevmode\hypertarget{ref-levinsoninprepydgrammar}{}%
+Levinson, S. C. (n.d.). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Cambridge, UK: Cambridge University Press.
+
+\leavevmode\hypertarget{ref-liszkowski2012prelinguistic}{}%
+Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
+(2012).
+A prelinguistic gestural universal of human communication.
+\emph{Cognitive Science}, \emph{36}(4), 698--713.
+\url{https://doi.org/10.1111/j.1551-6709.2011.01228.x}
+
 \leavevmode\hypertarget{ref-meir2017independent}{}%

### 2020-09-14 manuscript.tex
addressed middys comments code review

@@
 In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
-Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
+NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
 In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
 This study contributes to three lines of research.
-First, we document that non-word items containing typologically rare sounds are repeated accurately less often that non-words containing more common sounds.
+First, we document that non-word items containing typologically rare sounds are repeated without changes less often that non-words containing more common sounds.
 Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
@@
 \item
-  consent, ``correct'', ``accuracy'', ``performance''
+  draft discussion
+\end{itemize}
+
+\hypertarget{todo-alex}{%
+\subsection{TODO Alex}\label{todo-alex}}
+
+\begin{itemize}
+\tightlist
 \item
-  draft discussion
+  add effect sizes to all analyses
+\item
+  reply to questions
 \end{itemize}
@@
 The child should immediately try to say it back without changing anything.
-Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item accurately even though it had never been pronounced before).
+NWR scores are thought to reflect long-term phonological knowledge (which allows the child to perceive the item precisely even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely even though it had never been pronounced before).
 
@@
 The present report uses these unique data to contribute to several lines of research.
-First, we made sure that some of the non-word items contained typologically rare and/or challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall accuracy of repetition and patterns of mispronunciation.
+First, we made sure that some of the non-word items contained typologically rare and/or challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
 Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
@@
 We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
-This allowed us to get information about repetition accuracy of more items.
+This allowed us to get NWR scores for more items.
 
@@
  &  & ghuu & ɣuː &  &  & tiwuńe & tiwunɛ &  & \\
- &  & kpu & kpu &  &  & tumowe & tumɔwɛ &  & \\
+ &  & kp:ââ & kp̃ɑː &  &  & tumowe & tumɔwɛ &  & \\
 \addlinespace
- &  & lv:ê & lβʲ̃ə &  &  & widońe & wiṭɔnɛ &  & \\
- &  & lva & lβʲæ &  &  & wumipo & wumipɔ &  & \\
+ &  & kpu & kpu &  &  & widońe & wiṭɔnɛ &  & \\
+ &  & lv:ê & lβʲ̃ə &  &  & wumipo & wumipɔ &  & \\
+ &  & lva & lβʲæ &  &  &  &  &  & \\
  &  & lvi & lβʲi &  &  &  &  &  & \\
  &  & tp:a & tp̃æ &  &  &  &  &  & \\
+\addlinespace
  &  & tpâ & tpɑ &  &  &  &  &  & \\
-\addlinespace
  &  & tpê & tpə &  &  &  &  &  & \\
@@
 
-Some NWR studies employ phoneme-based scoring in addition to or instead of word-level accuracy (e.g., Cristia et al., 2020).
+Some NWR studies employ phoneme-based scoring in addition to or instead of word-level scoring (e.g., Cristia et al., 2020).
 We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
@@
 Participation was voluntary, with children being invited to come and participate.
-Regardless of how they performed, children were provided with a snack as compensation.
+Regardless of whether they did any of the task, children were provided with a snack as compensation.
 Children who came up to participate but then refused were nonetheless provided with the snack.
@@
 
-We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
+We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
 We excluded 3 children who did not have data for one of these two types.
-As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
-Surprisingly, subsequent repetitions (M = 39, SD = 26)
-were on average less accurate than first ones (M = 65, SD = 15), t(38) = 6.62, p = 0.
-Given the uncertainty in whether previous work used only the first or all repetitions, and since behavior degraded and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
+As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
+Surprisingly, NWR scores for subsequent repetitions (M = 40, SD = 26)
+were on average lower than first ones (M = 64, SD = 15), t(38) = 6.30, p = 0.
+Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
 
@@
 The overall NWR score was M = 61\% (SD = 15\%).
-Scores based on phonemes are even higher M = 78\% (SD = 9\%).
-The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
+Scores based on phonemes are even higher M = 100\% (SD = 0\%).
+The phoneme-based normalized Levenshtein distance was M = 0\% (SD = 0\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
 Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
@@
 
-In this analysis, we were interested in variation in performance as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+In this analysis, we were interested in variation in NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
 To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
@@
 We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.77, SE ß = 0.19, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Age was also a significant predictor (ß = 0.31, SE ß = 0.13, p = 0.02).
+The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.79, SE ß = 0.19, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+Age was also a significant predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
 
@@
 As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic frequency (ß = 0.50, SE ß = 0.25, p = 0.04) and age (ß = 0.31, SE ß = 0.13, p = 0.02) were significant predictors of accuracy, but corpus frequency (ß = 0.37, SE ß = 0.24, p = 0.12) was not.
+In this model, both cross-linguistic frequency (ß = 0.51, SE ß = 0.25, p = 0.04) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0.39, SE ß = 0.24, p = 0.11) was not.
 
@@
 \textbf{Option 1: just numbers, first attempts only }
-Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 7 vowels deleted, and 3 consonants.
+Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 0 vowels deleted, and 0 consonants.
 
 As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels, 0 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (43 oral vowels produced with a different quality; 0 nasal vowels produced with a different quality).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality; 0 nasal vowels produced with a different quality).
 As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0) than vice versa (0).
-Simple consonants were mispronounced as other simple consonants quite frequently (73 simple consonants mispronounced as other simple ones, compared to 0 complex ones).
+Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones, compared to 0 complex ones).
 
@@
 Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 7 vowels deleted (about 0.54\% of all vowel targets), and 3 consonants deleted (about 0.17\% of all consonant targets).
+Deletions were relatively rare, with only 0 vowels deleted (about NA\% of all vowel targets), and 0 consonants deleted (about NA\% of all consonant targets).
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about 0\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (43 oral vowels produced with a different quality or about 0.29\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about NA\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality or about NA\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
 As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times or about NA\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (73 simple consonants mispronounced as other simple ones or about 0.35\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones or about NA\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
 
@@
 Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 11 vowels deleted (about 0.25\% of all vowel targets), and 8 consonants deleted (about 0.21\% of all consonant targets).
+Deletions were relatively rare, with only 0 vowels deleted (about NA\% of all vowel targets), and 0 consonants deleted (about NA\% of all consonant targets).
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels or about 2.01\% out of all oral vowel targets, 58 nasal target vowels produced as oral vowels or about 6.46\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (205 oral vowels produced with a different quality or about 3.22\% out of all oral vowel targets; 23 nasal vowels produced with a different quality or about 7.08\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or about 5.57\% out of all complex consonant targets) than vice versa (2 times about 5.57\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (129 simple consonants mispronounced as other simple ones or about 1.14\% out of all simple consonant targets, compared to 55 complex ones or about 2.76\% out of all complex consonant targets).
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about NA\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality or about NA\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times about NA\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones or about NA\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
 
@@
 
-\hypertarget{accuracy-as-a-function-of-item-length}{%
-\subsubsection{Accuracy as a function of item length}\label{accuracy-as-a-function-of-item-length}}
+\hypertarget{nwr-scores-as-a-function-of-item-length}{%
+\subsubsection{NWR scores as a function of item length}\label{nwr-scores-as-a-function-of-item-length}}
 
-Next, we inspected whether accuracy varied as a function of word length.
+Next, we inspected whether NWR scores varied as a function of word length.
 Results are shown on table XX.
-This table shows that monosyllables accuracy was much lower than other lengths.
+This table shows that scores for monosyllables was much lower than for other lengths.
 This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
 
-Setting monosyllables aside, we observe the typical pattern of decreased accuracy for longer items, although this is particularly salient for the whole word scoring.
+Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole word scoring.
 This is the most commonly reported type of score, but it is also the least forgiving.
@@
 
-\caption{\label{tab:tablength}NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
+\caption{\label{tab:tablength}NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
 
@@
 \midrule
-47 (22) & 59 (17) & 41 (17)\\
-79 (22) & 94 (7) & 6 (7)\\
-78 (19) & 94 (6) & 6 (6)\\
-75 (32) & 93 (11) & 7 (11)\\
+47 (22) & 100 (0) & 0 (0)\\
+79 (22) & 100 (0) & 0 (0)\\
+78 (19) & 100 (0) & 0 (0)\\
+75 (32) & 100 (0) & 0 (0)\\
 \bottomrule
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,548.04) = 0.33, p = 0.12).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.18) = 21.34, p = 0.86), birth order (data missing for 17 children, 2,678.47) = -0.16, p = 0.44) or maternal education (data missing for 2 children, 2,091.92) = 0.09, p = 0.67)
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,485.88) = 0.35, p = 0.09).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.08) = 21.39, p = 0.94), birth order (data missing for 17 children, 2,730.78) = -0.19, p = 0.38) or maternal education (data missing for 2 children, 2,147.84) = 0.07, p = 0.76)
 
@@
 \item
-  What is the overall repetition accuracy (whole word, phoneme based, distance)?
+  What is the overall repetition scores (whole word, phoneme based, distance)?
 \item
@@
 \item
-  Children are more accurate for mono-syllables than longer items
+  Prediction based on previous work: Children have higher scores for shorter items
 \item
-  The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
-\textbf{Check for work on European languages that may have looked into this}
+  Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.
 \item
-  Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds {[}CITE{]}.
-Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
+  it turns out we were right! the scores for 2-4 syll words decline only slightly with length
 \item
-  The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
+  Prediction based on previous work: Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children achieve higher scores when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds {[}CITE{]}.
+Therefore, we expect higher scores for items with common sounds than in those with rare sounds.
+\item
+  Prediction for Yélî made before seeing the data: The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
 Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
 As a result, differences between easier and harder items may be smaller in this work than in other research.
-\textbf{no work looking at consonants \& vowels?
-no work looking at nasal vowels in particular?}
+\item
+  it turns out we do see
 \end{enumerate}
@@
 \item
-  Children's accuracy increases with child age.
+  Children's scores increases with child age.
 \item
-  Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
+  General prediction: Non-monolingual Yélî children have lower scores than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
 \item
-  As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
-\item
-  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
-\item
-  Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family
+  Local prediction: Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family
   (or at least from people other than their mothers, who tend to be the non-native speakers).
@@
   . That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother
-  , so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
+  , so in so far as this input has a crucial formational role, then there may still be a score difference between these two groups.
 \item
-  In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
+  We don't see a difference
+\item
+  General prediction: previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated score higher in standardized language tests (eg ppvt) than children whose mothers are less educated.
+\item
+  Prediction made before seeing data: In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
 This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
@@
 \item
-  One main causal path between birth order and language development is via parental input (CITE).
-Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
+  We did not find this
+\item
+  General prediction: To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children score higher in standardized language tests later-born children.
+\item
+  Prediction before seeing data: One main causal path between birth order and language development is via parental input (CITE).
+Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the difference in scores between first borns and later borns may be small
+\item
+  we did not find a sig effect, but this is a small effect (eg d=.2 in Havron et al)
 \end{enumerate}

### 2020-09-14 manuscript.Rmd
addressed middys comments code review

@@
    In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
-Accuracy is thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
+NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
 In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
 This study contributes to three lines of research.
-First, we document that non-word items containing typologically rare sounds are repeated accurately less often that non-words containing more common sounds.
+First, we document that non-word items containing typologically rare sounds are repeated without changes less often that non-words containing more common sounds.
 Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
@@
 - double check also stim_seg_freq -- why are some items still with NA freq?
-- consent, "correct", "accuracy", "performance"
 - draft discussion
+
+## TODO Alex
+
+- add effect sizes to all analyses
+- reply to questions
 
@@
 The present report uses these unique data to contribute to several lines of research.
-First, we  made sure that some of the non-word items contained typologically rare and/or  challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall accuracy of repetition and patterns of mispronunciation.
+First, we  made sure that some of the non-word items contained typologically rare and/or  challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
 Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
@@
 
-Some NWR studies employ phoneme-based scoring in addition to or instead of word-level accuracy [e.g., @cristia2020infant].
+Some NWR studies employ phoneme-based scoring in addition to or instead of word-level scoring [e.g., @cristia2020infant].
 We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
@@
 # that's not a desirable outcome so am commenting it out
-# trnsc$Sex[trnsc$Sex=="F"]<-"G"
+# a2m fixed, but I think it is better to fix all of these error upstream, in the demo.xlsx file
+trnsc$Sex[trnsc$Sex=="F" & !is.na(trnsc$Sex)]<-"G"
 trnsc$Sex=factor(trnsc$Sex)
@@
 
-We first checked whether accuracy varies between first and subsequent presentations of an item by averaging word-level accuracy at the participant level separately for first attempts and subsequent repetitions.
+We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
 We excluded `r excl` children who did not have data for one of these two types.
-As shown in Figure 1, participants' mean word-level accuracy became more heterogeneous in subsequent repetitions.
-Surprisingly, subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
-were on average less accurate than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`.
-Given the uncertainty in whether previous work used only the first or all repetitions, and since behavior degraded and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
+As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
+Surprisingly, NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
+were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`.
+Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
 
@@
 aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->overall
-aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->pc
+aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->phon_score
 aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->nld
-overall=cbind(overall,pc$x,nld$x)
+overall=cbind(overall,phon_score$x,nld$x)
 colnames(overall)<-c("ID","Sex","Age","BrthO","matEd","Word","Phoneme","NLD")
@@
 Error"]`, p = `r both_coeff["avg_fr_s","Pr(>|z|)"]`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std.
-Error"]`, p = `r both_coeff["age.rounded_s","Pr(>|z|)"]`) were significant predictors of accuracy, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
+Error"]`, p = `r both_coeff["age.rounded_s","Pr(>|z|)"]`) were significant predictors of word-level NWR scores, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
 Error"]`, p = `r both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"]`) was not.
@@
 
-### Accuracy as a function of item length
+### NWR scores as a function of item length
 
@@
 aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->by_length
-aggregate(trnsc$pc*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->pc
+aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->phon_score
 aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->nld
-by_length=cbind(by_length,pc$x,nld$x)
+by_length=cbind(by_length,phon_score$x,nld$x)
 colnames(by_length)<-c("ID","Age","Length","Word","Phoneme","NLD")
@@
 
-Next, we inspected whether accuracy varied as a function of word length.
+Next, we inspected whether NWR scores varied as a function of word length.
 Results are shown on table XX.
-This table shows that monosyllables accuracy was much lower than other lengths.
+This table shows that scores for monosyllables was much lower than for other lengths.
 This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
 
-Setting monosyllables aside, we observe the typical pattern of decreased accuracy for longer items, although this is particularly salient for the whole word scoring.
+Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole word scoring.
 This is the most commonly reported type of score, but it is also the least forgiving.
@@
 colnames(birthorder.tab)<-c("Word","Phoneme","NLD")
-apa_table(birthorder.tab,caption="NWR scores measured in whole word accuracy, phoneme-based accuracy, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
+apa_table(birthorder.tab,caption="NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
 ```
@@
 
-- What is the overall repetition accuracy (whole word, phoneme based, distance)?
+- What is the overall repetition scores (whole word, phoneme based, distance)?
 
@@
 
-1. Children are more accurate for mono-syllables than longer items
+1. Prediction based on previous work: Children have higher scores for shorter items
 1.
-The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English.
-**Check for work on European languages that may have looked into this**
+Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.
+1. it turns out we were right! the scores for 2-4 syll words decline only slightly with length
 
 2.
-Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds [CITE].
-Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
+Prediction based on previous work: Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children achieve higher scores when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds [CITE].
+Therefore, we expect higher scores for items with common sounds than in those with rare sounds.
 2.
-The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
+Prediction for Yélî made before seeing the data: The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
 Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
 As a result, differences between easier and harder items may be smaller in this work than in other research.
-**no work looking at consonants & vowels?
-no work looking at nasal vowels in particular?**
+2. it turns out we do see
 
@@
 
-3. Children's accuracy increases with child age.
-4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
+3. Children's scores increases with child age.
+
+4. General prediction: Non-monolingual Yélî children have lower scores than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
 <!-- **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)** ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)-->
-5.
-As revious NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
-6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children
-
-4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family 
+4. Local prediction: Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family 
 <!-- **(MC: definitely almost all kin; 5k of them)** ac2mc I'm not sure this is how bio anthropologists think about kin; because it's about how you attribute resources to boost your genes' chances of being passed on relative to those of others in the same space.
@@
 <!-- ac2mc I think I meant your own paper! added "on the same population "-->
-, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
+, so in so far as this input has a crucial formational role, then there may still be a score difference between these two groups.
+4. We don't see a difference
+
 5.
-In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
+General prediction: previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated score higher in standardized language tests (eg ppvt) than children whose mothers are less educated.
+5.
+Prediction made before seeing data: In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
 This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
 However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
+5. We did not find this
+
+6. General prediction: To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children score higher in standardized language tests later-born children. 
 6.
-One main causal path between birth order and language development is via parental input (CITE).
-Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
+Prediction before seeing data: One main causal path between birth order and language development is via parental input (CITE).
+Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the difference in scores between first borns and later borns may be small
+6. we did not find a sig effect, but this is a small effect (eg d=.2 in Havron et al)
 

### 2020-09-14 wrangling.R

@@ -28,6 +28,10 @@ trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
 trnsc$attempt <- ifelse(trnsc$rep == "1", "first", "subsequent")
 
 # MC's new code
+# A2m please look at lines 21-22, those are the lines that determine which items are first versus subsequent. 
+# int is the interval in praat, so it is a certain measure of time; I don't remember
+# what token is, but there is a file called notes_for_supmat.txt which contains the explanation
+# for all columns, which may say for this one and others
 # first.prods <- trnsc %>%
 #   group_by(id, item) %>%
 #   summarize(min.token = min(token)) %>%
@@ -67,6 +71,16 @@ dim(trnsc) #still no loss, all good
 #add orthographic representations targets & other stims chars
 # M2A: Note that some of these stim characters are now updated in stimuli2.txt
 # they do not match what is shown here (!!)
+# a2m: I'm confused, what do you mean? 
+# my pipeline was: stimuli.txt I couldn't read here, so I put these instructions in the rmd:
+# <!-- REMEMBER!!!! -->
+#   <!-- phonetic characters: -->
+#   <!-- - save excel as utf with save as -->
+#   <!-- - do $ iconv -f utf-16le -t utf-8 stimuli.txt > stimconv.txt -->
+#   <!-- - do $ iconv -f utf-16le -t utf-8 segments.txt > segconv.txt -->
+#   
+# also, notice that stims and segments are merged using ortho, so as long as ortho is matched
+# across the two, the merge should carry over whatever changes you make
 read.delim("stimconv.txt", encoding="UTF-8")->stims
 read.delim("segconv.txt", encoding="UTF-8")->segments
 #add corpus freq to segments
@@ -84,6 +98,10 @@ for(i in 1:dim(stim_seg_freq)[1]) for(j in 1:dim(stim_seg_freq)[2]) {
   stim_seg_freq[i,j]<-ifelse(sum(!is.na(thisfreq))==1,thisfreq,NA)
 }
 # M2A: I'm suspicious of some of the NAs in here (like for 'k', 'w', 'd')--what is segment pc?
+# a2m please note I didn't look up all sounds, only those used in our stimuli
+# w is in our stimuli for sure, so that may not explain away all the NAs
+# pc is percentage of languages in which the sound appears in phoible
[…]

### 2020-09-16 wrangling.R
respond to AC's comments

@@ -24,32 +24,32 @@ trnsc[order(trnsc$file, trnsc$int),]->trnsc
 #add repetition number
 trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
 
-#add variable coding whether it's the child's first or subsequent attempts
-trnsc$attempt <- ifelse(trnsc$rep == "1", "first", "subsequent")
-
-# MC's new code
-# A2m please look at lines 21-22, those are the lines that determine which items are first versus subsequent. 
-# int is the interval in praat, so it is a certain measure of time; I don't remember
-# what token is, but there is a file called notes_for_supmat.txt which contains the explanation
-# for all columns, which may say for this one and others
-# first.prods <- trnsc %>%
-#   group_by(id, item) %>%
-#   summarize(min.token = min(token)) %>%
-#   mutate(attempt = "first") %>%
-#   rename("token" = min.token)
-# trnsc <- trnsc %>%
-#   left_join(first.prods, by = c("id", "item", "token")) %>%
-#   replace_na(list(attempt = "subsequent"))
-# table(trnsc$attempt)
-# M2A: by my count there are 1029 first attempts, but maybe I'm misunderstanding "token"
-# M2A: okay, my solution finds some conflicting cases with your outfile attempt namings,
-#      now going off of those for attempt first/subsequent, I still find 1029
-
 # AC's prev code
 # trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
 # trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
 # trnsc$attempt[1]<-"first"
 # table(trnsc$attempt) #1045 first attempts
+
+# MC's new code
+# M2A: by my count there are 1029 first attempts, but maybe I'm misunderstanding "token"
+# M2A: okay, my solution finds some conflicting cases with your outfile attempt namings,
+#      now going off of those for attempt first/subsequent, I still find 1029
+# A2m please look at lines 21-22, those are the lines that determine which items are first versus subsequent. 
+# int is the interval in praat, so it is a certain measure of time; I don't remember
+# what token is, but there is a file called notes_for_supmat.txt which contains the explanation
+# for all columns, which may say for this one and others
[…]

### 2020-09-16 manuscript.tex
add comments for AC + fix fig IPA, table on length, and one ref

@@
 }{}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\usepackage{hyperref}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
 \hypersetup{
@@
 \setmainfont{Doulos SIL} % Set main font to Doulos SIL
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\fi
 
 \title{Non-word repetition in Yélî Dnye}
-\author{Alejandrina Cristia\textsuperscript{1}~\& Marisa Casillas\textsuperscript{2}}
+\author{Alejandrina Cristia\textsuperscript{1}~\& Marisa
+Casillas\textsuperscript{2}}
 \date{}
 
-\authornote{All data are made available in a repository in the Open Science Framework.
-AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
-S.
-McDonnell Foundation Understanding Human Cognition Scholar Award.
-
-Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d'Ulm, 75005 Paris, France. E-mail: \href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
+\authornote{All data are made available in a repository in the
+Open Science Framework. AC acknowledges the support of the Agence
+Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004,
+ANR-14-CE30-0003, ANR-17-EURE-0017); and the J. S. McDonnell Foundation
+Understanding Human Cognition Scholar Award.
+
+Correspondence concerning this article should be addressed to
+Alejandrina Cristia, 29, rue d'Ulm, 75005 Paris, France. E-mail:
+\href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
 
 \abstract{
-In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
-NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
-In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
-This study contributes to three lines of research.
-First, we document that non-word items containing typologically rare sounds are repeated without changes less often that non-words containing more common sounds.
-Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
-Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
-Together, these data provide a unique view of online phonological processing in a seldom-studied language, and contribute to both typological and language acquisition research.
+In nonword repetition (NWR) studies, participants are presented
+auditorily with an item that is phonologically legal but lexically
+meaningless in the local language. NWR scores are thought to reflect
+long-term phonological knowledge as well as online phonological working
+memory and flexible production patterns. In this study, we report on NWR
+results among children learning Yêly Dnyé, an isolate spoken in Rossel
+Island, PNG, with an unusually dense phonological inventory. This study
+contributes to three lines of research. First, we document that non-word
+items containing typologically rare sounds are repeated without changes
+less often that non-words containing more common sounds. Second, we
+document rather weak effects of item length, contributing to other
+research suggesting that length effects may be language-specific. Third,
+we do not find strong individual variation effects in this population,
+contrary to previous results documenting strong age-related effects.
+Together, these data provide a unique view of online phonological
+processing in a seldom-studied language, and contribute to both
+typological and language acquisition research.
 
@@
 \item
-  look through this manuscript for bits of text to be written by \textbf{mc}
-\item
-  do you think Figure 2 should be done with phonetic symbols? If yes, could you try to implement that?
-\item
-  double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
-I looked again at my local files (still haven't checked the labs' ones) and I'm surprised I have maternal education for more children than I have birth order for\ldots{} If you correct this, to generate the csv I also round the age, assign age.rounded to reported age (in years) when DOB is missing, remove first names, DOB, DOTest and age.precise before pushing to github
-\item
-  double check my frequency entries in segments.xlsx using the search function in \url{https://phoible.org/parameters}; take a look at \url{http://phoible.github.io/conventions/} in case you see something about double articulation (I didn't find tp or lBj)
-\item
-  double check phonological implementation of stimuli in stimuli.xlsx
-\item
-  for both of the above, if you make any changes, they need to be re-saved as UTF16 txt file, and then converted (see remember above)
-\item
-  double check also stim\_seg\_freq -- why are some items still with NA freq?
+  look through this manuscript for bits of text to be written by
+  \textbf{mc}
+\item
+  double check demo -- probably start with the xlsx version that has the
+  children's first names, date of birth, and date of test. I looked
+  again at my local files (still haven't checked the labs' ones) and I'm
+  surprised I have maternal education for more children than I have
+  birth order for\ldots{} If you correct this, to generate the csv I
+  also round the age, assign age.rounded to reported age (in years) when
+  DOB is missing, remove first names, DOB, DOTest and age.precise before
+  pushing to github
+\item
+  double check my frequency entries in segments.xlsx using the search
+  function in \url{https://phoible.org/parameters}; take a look at
+  \url{http://phoible.github.io/conventions/} in case you see something
+  about double articulation (I didn't find tp or lBj)
+\item
+  double check also stim\_seg\_freq -- why are some items still with NA
+  freq?
 \item
@@
 
-\hypertarget{todo-alex}{%
-\subsection{TODO Alex}\label{todo-alex}}
+\subsection{TODO Alex}\label{todo-alex}
 
@@
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
-
-Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., Hazan \& Barrett, 2000).
-One common task in this line of research is nonword repetition (NWR).
-In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning.
-The child should immediately try to say it back without changing anything.
-NWR scores are thought to reflect long-term phonological knowledge (which allows the child to perceive the item precisely even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely even though it had never been pronounced before).
-
-NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and to assess the extent to which apparent phonological constraints found in the lexicon affect online production (Gallagher, 2014).
-It is also frequently used for applications, notably as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-However, it has been seldom used outside of Europe and North America (with exceptions including Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)).
-
-In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
-The present report uses these unique data to contribute to several lines of research.
-First, we made sure that some of the non-word items contained typologically rare and/or challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
-Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
-Third, we contribute to the basic NWR literature by adding a data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
-
-\hypertarget{yuxe9luxee-dnye-phonological-overview}{%
-\subsubsection{Yélî Dnye phonological overview}\label{yuxe9luxee-dnye-phonological-overview}}
-
-Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
-The Yélî sound system, much like its baroque grammatical system (Levinson, n.d.), is unlike any other in the region.
-
-With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these able to be contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized or palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (n.d.) for details).
-The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
-Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
-
-In total, Yélî Dnye uses 90 distinctive segments (93 when including consonants that are extremely rarely used), far outstripping the phonemic inventory size of other documented Papuan languages (Levinson (n.d.); Foley 1986) and includes at least two contrasts not yet been documented elsewhere {[}labial-coronal double-articulations with dental vs.~post-alveolar coronal placement in both oral and nasal stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.
-
-Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the current Yélî Dnye dictionary, Levinson, n.d.).
-The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/).
-This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /`pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
-Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (n.d.) for details).
-
-While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
-In monosyllabic stems, oral vowels (short: 43\% and long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%), with the first syllable of bisyllabic stems following a similar pattern.
-Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels.
-Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see Levinson (n.d.) for details).
-
-\hypertarget{the-yuxe9luxee-community}{%
-\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}}
-
-The current study also contributes to increasing the cultural diversity of the populations for which NWR data have been collected.
-As mentioned previously, this method has been used almost exclusively in urban, industrialized populations, with only one exception to our knowledge (Cristia et al., 2020).
-We therefore provide some background information tol help understand some aspects of our study, including the estimation of children's ages and the relationship between NWR scores and children's demographic characteristics.
-
-Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7;
-school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---, where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
-Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
-
-Most Yélî people are swidden horticulturalists, raising a variety of tubers (e.g., yams, manioc, taro), coconut, banana, and more in their gardens.
-In addition to garden foods, Yélî children grow up eating fish and foraging for shellfish and nuts, which provide a regular source of extra protein.
-The typical household in our dataset includes seven individuals and is situated among a collection of 4 or more other households, with structures often arranged around an open grassy area.
-These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
-
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by barrier reef, transport to and from the island is both infrequent and irregular.
-There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport, when there is deisel available.
-There is a medical radio connection at the health clinic (at the Catholic mission) via which news and messages are often passed along.
-There is also a cellular tower on the island, though it is typically not functioning.
-As such, access to news and outside resources, particularly international connections and larger data transfers, is typically unavailable.
-Our data collection is therefore limited to the duration of the researchers' on-island visits.
-
-Despite restricted outside contact, formal education is a priority for many Yélî families.
-A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65\% of mothers and 49\% of fathers completing the full range of locally available education (\textasciitilde{}9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG (Casillas, Brown, \& Levinson, 2020).
-While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), and so it is very common for households to billet school children during the weekdays for months at a time (or for whole months if the child's home is very far).
-Combined with practices of collective child guardianship within their close-knit hamlets, the practical outcome for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's parents---to this child assent is also culturally pertinent, as child independence is encouraged and respected from toddlerhood onward (Brown \& Casillas, n.d.).\footnote{Note that in a few cases this does mean that precise details such as child date of birth, parent age in years, and parent education was only able to be approximated as the parents were residing on another part of the island.} Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
-
-While work on Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas et al., 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun {[}e.g., Bram's thesis; MC's CFH data{]}.
-What we do know about the early language environment is that, while child-directed speech is relatively infrequent throughout the day (Casillas et al., 2020), shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children (Bunce et al., n.d.; Casillas et al., 2020).
-While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere (Bergelson et al., 2019; Bunce et al., n.d.), there is a significant and marked increase in child-directed speech from other children as infants get older (Bunce et al., n.d.; Casillas et al., 2020).
-This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 people) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
-
-\hypertarget{brief-review-of-nwr-for-our-purposes}{%
-\subsubsection{Brief review of NWR for our purposes}\label{brief-review-of-nwr-for-our-purposes}}
-
-The basic structure of NWR involves the participant hearing a non-word, like ``bilik'', which they need to repeat back without changing anything that is contrastive in the language.
-For instance, in English, a response of ``bilig'' or ``pilik'' will be scored as incorrect; a response ``biilik'', where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.
-
-Aside from this central basic structure, there is some variation in the presentation procedure and structure of items found in previous NWR work.
-For example, items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation (Brandeker \& Thordardottir, 2015).
-Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
-Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
-
-Previous work seems to avoid difficult sounds, but we felt this was important to represent Yélî Dnye, so we also varied this factor.
-We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
-This allowed us to get NWR scores for more items.
-
-Naturally, designing the task in this way may render the study of individual variation within the population more difficult because different children are exposed to different items.
-However, a review of previous work on individual variation suggested to us that effects of individual differences are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
-
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
-Based on previous work, we looked at potential improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and potential negative effects of bilingual exposure (Brandeker \& Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
-Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
-Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on those too.
-
-\hypertarget{research-questions}{%
-\subsubsection{Research questions}\label{research-questions}}
-
-After some preliminary analyses to set the stage, we address the following questions:
+\subsection{Introduction}\label{introduction}
+
+Although infants begin to learn about their native language's phonology
+within the first year, many studies suggest that in perception and
+production, in phonetics and phonology, their knowledge continues to
+develop throughout childhood (e.g., Hazan \& Barrett, 2000). One common
+task in this line of research is nonword repetition (NWR). In NWR
+studies, participants are presented auditorily with an item that is
+phonologically legal but lexically meaningless in the language children
+are learning. The child should immediately try to say it back without
+changing anything. NWR scores are thought to reflect long-term
+phonological knowledge (which allows the child to perceive the item
+precisely even though it is not a real word they have encountered
+before) as well as online phonological working memory (to encode the
+item in the interval between hearing it and saying it back) and flexible
+production patterns (to produce the item precisely even though it had
+never been pronounced before).
+
+NWR has been used to seek answers to a variety of theoretical questions,
+including what the links between phonology, working memory, and the
+lexicon are (Bowey, 2001), and to assess the extent to which apparent
+phonological constraints found in the lexicon affect online production
+(Gallagher, 2014). It is also frequently used for applications, notably
+as a diagnostic for language delays and disorders (Estes, Evans, \&
+Else-Quest, 2007). Since non-words can be generated in any language, it
+has attracted the attention of researchers working in multilingual and
+linguistically diverse environments, particularly in Europe (Action,
+2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has been seldom
+used outside of Europe and North America (with exceptions including
+Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz
+(2020)).
+
+In this study, we report on NWR results among children learning Yêly
+Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense
+phonological inventory. The present report uses these unique data to
+contribute to several lines of research. First, we made sure that some
+of the non-word items contained typologically rare and/or challenging
+sounds, so that we can contribute to the study of whether rare sounds
+are disadvantaged in perception and/or production, both in terms of
+overall repetition scores and patterns of mispronunciation. Second, we
+look at the effects of item length, since previous NWR research has
+uncovered variable effects. Third, we contribute to the basic NWR
+literature by adding a data point for children learning a rarely-studied
+language and culture, and further investigating whether there are
+structured sources that account for individual variation.
+
+\subsubsection{Yélî Dnye phonological
+overview}\label{yuxe9luxee-dnye-phonological-overview}
+
+Yélî Dnye is an isolate language (presumed Papuan) spoken by
+approximately 7,000 people residing on Rossel Island, which is found at
+the far end of the Louisiade Archipelago in Milne Bay Province, Papua
+New Guinea. The Yélî sound system, much like its baroque grammatical
+system (Levinson, in preparation), is unlike any other in the region.
+
+With only four primary places of articulation (bilabial, alveolar,
+post-alveolar, and velar) and no voicing contrasts, the phonological
+inventory is remarkably packed with acoustically similar segments. The
+core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and
+doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal
+equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
+substantial portion of these able to be contrastively pre-nasalized or
+nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/,
+/kŋ/, /ṭpṇm/, /kpŋm/). A large number of this combinatorial set can
+further be contrastively labialized or palatalized on release, or both
+(e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (in preparation)
+for details). The consonantal inventory also includes a number of
+non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in
+Yélî Dnye may be oral or nasal, short or long. The 10 oral vowel
+qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/,
+/o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels,
+with 7 of these able to appear as short and long nasal vowels as well
+(/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
+
+In total, Yélî Dnye uses 90 distinctive segments (93 when including
+consonants that are extremely rarely used), far outstripping the
+phonemic inventory size of other documented Papuan languages (Levinson
+(in preparation); Foley 1986) and includes at least two contrasts not
+yet been documented elsewhere {[}labial-coronal double-articulations
+with dental vs.~post-alveolar coronal placement in both oral and nasal
+stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.
+
+Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
+monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
+that, and with tri-and-above syllabic words appearing least frequently
+(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the
+current Yélî Dnye dictionary, Levinson, in preparation). The vast
+majority of syllables use a CV format. A small portion of the lexicon
+features words with a final closed syllable, limited to codas of -/m/,
+-/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
+spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
+coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
+process is used frequently with English loan words that have a coda
+(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
+features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
+these are limited to /æ/-. Finally, the lexicon features a handful of
+single-vowel grammatical morphemes (see Levinson (in preparation) for
+details).
+
+While all consonants can appear in the first syllable of the word, the
+second syllable onset consonant in bisyllabic roots is almost
+exclusively drawn from a subset of the simplex consonants (in order of
+descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/,
+/t/, /ɣ/, /mʷ/). In monosyllabic stems, oral vowels (short: 43\% and
+long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%),
+with the first syllable of bisyllabic stems following a similar pattern.
+Most vowels in the second syllable of bisyllabic stems are also short
+and oral, with a potential preference for rounding harmony between the
+first and second vowels. Stress typically falls on the first syllable of
+bisyllabic words, on the first and third syllables of tetrasyllabic
+words, and on the first or second syllable of trisyllabic words,
+depending on its vocalic features (see Levinson (in preparation) for
+details).
+
+\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}
+
+The current study also contributes to increasing the cultural diversity
+of the populations for which NWR data have been collected. As mentioned
+previously, this method has been used almost exclusively in urban,
+industrialized populations, with only one exception to our knowledge
+(Cristia et al., 2020). We therefore provide some background information
+tol help understand some aspects of our study, including the estimation
+of children's ages and the relationship between NWR scores and
+children's demographic characteristics.
+
+Most speakers of the language grow up speaking it monolingually until
+they begin attending school around the age of 7; school instruction is
+in English. While monolingual Yélî Dnye upbringing is common,
+multilingual families are not unusual, particularly in the region around
+the Catholic Mission---the same region in which the current data were
+collected---, where there is a higher incidence of married-in mothers
+from other islands (Brown \& Casillas, in press). Children in these
+multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
+and/or some other language(s) from the region.
+
+Most Yélî people are swidden horticulturalists, raising a variety of
+tubers (e.g., yams, manioc, taro), coconut, banana, and more in their
+gardens. In addition to garden foods, Yélî children grow up eating fish
+and foraging for shellfish and nuts, which provide a regular source of
+extra protein. The typical household in our dataset includes seven
+individuals and is situated among a collection of 4 or more other
+households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that
+they typically comprise a set of brothers, their wives and children, and
+their mother and father, with neighboring hamlets also typically related
+through the patriline.
+
+Laying 250 nautical miles off the coast of mainland PNG and surrounded
+by barrier reef, transport to and from the island is both infrequent and
+irregular. There is no electricity system (we use solar panels) or motor
+vehicles on the island, save dinghys with outboard motors that are
+primarily reserved for medical transport, when there is deisel
+available. There is a medical radio connection at the health clinic (at
+the Catholic mission) via which news and messages are often passed
+along. There is also a cellular tower on the island, though it is
+typically not functioning. As such, access to news and outside
+resources, particularly international connections and larger data
+transfers, is typically unavailable. Our data collection is therefore
+limited to the duration of the researchers' on-island visits.
+
+Despite restricted outside contact, formal education is a priority for
+many Yélî families. A recent study surveying more than 40 parents of
+young children in this region of the island found that nearly all
+parents completed 6 or more years of education, with 65\% of mothers and
+49\% of fathers completing the full range of locally available education
+(\textasciitilde{}9 years), with around half of those parents having
+pursued an additional 2+ years of education on other islands in the
+region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While
+there are a handful of local schools around the island, these may be
+well out of walking distance for many children (i.e., more than 1 hour
+on foot or by canoe each day), and so it is very common for households
+to billet school children during the weekdays for months at a time (or
+for whole months if the child's home is very far). Combined with
+practices of collective child guardianship within their close-knit
+hamlets, the practical outcome for researchers is that adult consent can
+often come from a combination of aunts, uncles, adult cousins, and
+grandparents standing in for the child's parents---to this child assent
+is also culturally pertinent, as child independence is encouraged and
+respected from toddlerhood onward (Brown \& Casillas, in
+press).\footnote{Note that in a few cases this does mean that precise
+  details such as child date of birth, parent age in years, and parent
+  education was only able to be approximated as the parents were
+  residing on another part of the island.} Finally, as mentioned, the
+language of schooling in this region is English, and so the majority of
+teens and adults on the island speak some English---many speak it very
+fluently.
+
+While work on Yélî language development is growing (e.g., Brown, 2011,
+2014; Brown \& Casillas, in press; Casillas et al., 2020; Liszkowski,
+Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's
+linguistic development is quite limited, and research into their
+phonological development in particular has only just begun {[}e.g.,
+Bram's thesis; MC's CFH data{]}. What we do know about the early
+language environment is that, while child-directed speech is relatively
+infrequent throughout the day (Casillas et al., 2020), shared caregiving
+practices and a near-universal fondness for social interaction with
+young children ensure that speech directed to children comes from all
+types of speakers: women, men, and other children (Bunce et al., under
+review; Casillas et al., 2020). While speech from women predominates in
+young children's speech environments on Rossel, as it does elsewhere
+(Bergelson et al., 2019; Bunce et al., under review), there is a
+significant and marked increase in child-directed speech from other
+children as infants get older (Bunce et al., under review; Casillas et
+al., 2020). This increase in child-directed speech from other children
+is attributed to the fact that, starting around age 2, children often
+spend large swaths of the day playing, swimming, and foraging for fruit,
+nuts, and shellfish in large (\textasciitilde{}10 people) independent
+and mixed-age child play groups (Brown \& Casillas, in press; Casillas
+et al., 2020).
+
+\subsubsection{Brief review of NWR for our
+purposes}\label{brief-review-of-nwr-for-our-purposes}
+
+The basic structure of NWR involves the participant hearing a non-word,
+like ``bilik'', which they need to repeat back without changing anything
+that is contrastive in the language. For instance, in English, a
+response of ``bilig'' or ``pilik'' will be scored as incorrect; a
+response ``biilik'', where the vowel is lengthened without change of
+quality would be scored as correct, because English does not have a pure
+length contrast.
+
+Aside from this central basic structure, there is some variation in the
+presentation procedure and structure of items found in previous NWR
+work. For example, items are often presented orally by the experimenter
+(Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an
+increasing number of studies have turned to playing back the stimuli in
+order to have greater control of the stability of the presentation
+(Brandeker \& Thordardottir, 2015). Additionally, while some studies
+have used 10-15 non-words, others have employed up to 46 unique items
+(Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019). Often,
+authors modulate structural complexity, typically measured in terms of
+item length (measured in number of syllables) and/or syllable structure
+(open as opposed to closed syllables, Gallon, Harris, \& Van der Lely,
+2007).
+
+Previous work seems to avoid difficult sounds, but we felt this was
+important to represent Yélî Dnye, so we also varied this factor. We
+designed a relatively large number of items but, aware that this may
+render the task longer and more tiresome, we split some of the items
+across children. This allowed us to get NWR scores for more items.
+
+Naturally, designing the task in this way may render the study of
+individual variation within the population more difficult because
+different children are exposed to different items. However, a review of
+previous work on individual variation suggested to us that effects of
+individual differences are relatively small, and would not be detectable
+with the sample size that we could collect in a given visit.
+
+That said, we contribute to the literature by also reporting descriptive
+analyses of individual variation that could potentially be integrated in
+meta-analytic efforts. Based on previous work, we looked at potential
+improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid,
+Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and
+potential negative effects of bilingual exposure (Brandeker \&
+Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
+Previous work typically finds no significant differences as a function
+of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014;
+Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although past research has not often investigated potential effects of
+birth order on NWR, there is a sizable literature on these effects in
+other language tasks (Havron et al., 2019), and therefore we report on
+those too.
+
+\subsubsection{Research questions}\label{research-questions}
+
+After some preliminary analyses to set the stage, we address the
+following questions:
 
@@
 \item
-  Does the frequency of sounds across languages predict NWR? Are rarer sounds more often substituted by commoner sounds?
-\item
-  How does score change as a function of item length in number of syllables?
-\item
-  Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
+  Does the frequency of sounds across languages predict NWR? Are rarer
+  sounds more often substituted by commoner sounds?
+\item
+  How does score change as a function of item length in number of
+  syllables?
+\item
+  Is individual variation attributable to child age, sex, birth order,
+  monolingual status, and/or parental education? 
 \end{itemize}
 
-In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects given their variability across studies.
-To illustrate this, we portray studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables) in Figure 1.
-Notice that the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Tsimane' children (Cristia et al., 2020), where a drop of 40 percentage points is observed at all ages.
-A similar difference in NWR scores for short versus long items was observed among Arabic children (Jaber-Awida, 2018).
-Even the effect of age is unstable in this sample.
-Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
-Therefore, all analyses here are descriptive and should be considered exploratory.
+In view of the hypothesis-driven nature of this work, we had considered
+boosting the interpretational value of this evidence by announcing our
+analysis plans prior to conducting them. However, we realized that even
+pre-registering an analysis would be equivocal because we do not have
+enough power to look at all relationships of interest, and often to
+detect any of the known effects given their variability across studies.
+To illustrate this, we portray studies in which children's NWR scores
+were gathered between 4 and 12 years of age, and reported separately for
+items that are relatively short (1-2 syllables) versus longer items (3-4
+syllables) in Figure 1. Notice that the effect of stimulus length is
+minuscule among Italian children (Piazzalunga et al., 2019), but
+considerable among Tsimane' children (Cristia et al., 2020), where a
+drop of 40 percentage points is observed at all ages. A similar
+difference in NWR scores for short versus long items was observed among
+Arabic children (Jaber-Awida, 2018). Even the effect of age is unstable
+in this sample. Whereas it is quite clear that children's NWR scores
+increase in the Italian data, age effects are less stable among Tsimane'
+children. Therefore, all analyses here are descriptive and should be
+considered exploratory.
 
@@
 \includegraphics{manuscript_files/figure-latex/prevlit-fig-1.pdf}
-\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years) and item length for comparable studies.
-Arabic data from Jaber et al.
-(2018); Italian data from Piazzalunga et al.
-(2019); Tsimane' from Cristia et al.
-(2020); Yélî data from the present study.}
+\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years)
+and item length for comparable studies. Arabic data from Jaber et al.
+(2018); Italian data from Piazzalunga et al. (2019); Tsimane' from
+Cristia et al. (2020); Yélî data from the present study.}
 \end{figure}
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
-
-\hypertarget{stimuli}{%
-\subsubsection{Stimuli}\label{stimuli}}
-
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
-We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in Yély Dnye.
-We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat (note that up to other work has also used 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
-
-A first list of candidate items was generated in 2018 by selecting simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'', ``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
-These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
-The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
-Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
-
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî inventory to produce consonant-vowel monosyllables.
-As before, items were automatically excluded if they appeared in the dictionary.
-Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described (Levinson, n.d.), there could have been undocumented constraints that rendered items illegal.
-Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
-These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset \textbf{XXX }mc** please fill in** and an Olympus \textbf{XXX }mc** please fill in** from the written form presented together with the same item orally (by the second author).
-The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
-Even so, there was one monosyllable that was often identified as a real word (intended ``yî'' /yXX/; identified as ``yi'' /yi/, \emph{tree}).
+\subsection{Methods}\label{methods}
+
+\subsubsection{Stimuli}\label{stimuli}
+
+Many NWR studies are based on a fixed list of 12-16 items that vary in
+length between 1 and 4 syllables, often additionally varying syllable
+complexity and/or cluster presence and complexity, always meeting the
+condition that they do not mean anything in the target language (e.g.,
+Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We kept the
+same variation in item length and the non-meaningfulness requirement,
+but we did not vary syllable complexity and clusters because these are
+vanishingly rare in Yély Dnye. We also increased the number of items an
+individual child would be tested on, so that a child would get up to 23
+items to repeat (note that up to other work has also used 24-30 items:
+Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and
+distributed them across children, so as to increase the coverage, and be
+able to study more items.
+
+A first list of candidate items was generated in 2018 by selecting
+simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'',
+``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined
+into consonant-vowel syllables, further sampling the space of 1- to
+4-syllable sequences. These candidates were automatically checked
+against Levinson's 2015 dictionary and removed from consideration if
+they appeared in the dictionary. The second author presented them orally
+to three local research assistants, who were asked to repeat them and
+further say whether they were real words. Any item for which two or more
+of the assistants reported them having a meaning or some form of
+association was excluded.
+
+A second list of candidate items was generated in 2019 by selecting
+complex consonants and systematically crossing them with all the vowels
+in the Yélî inventory to produce consonant-vowel monosyllables. As
+before, items were automatically excluded if they appeared in the
+dictionary. Additionally, since hearing vowel length in monosyllables in
+isolation is challenging, any item that had a short/long real word
+neighbor was filtered out. Since the phonology and phonetics of Yélî is
+still in the process of being described (Levinson, in preparation),
+there could have been undocumented constraints that rendered items
+illegal. Therefore, we made sure that the precise consonant-vowel
+sequence occurred in some real word in the dictionary (i.e., that there
+was a longer word included the monosyllable as a subsequence). These
+candidates were presented to one informant, for a final check that they
+did not mean anything. Together with the 2018 selection, they were
+recorded using a headset \textbf{XXX }mc** please fill in** and an
+Olympus \textbf{XXX }mc** please fill in** from the written form
+presented together with the same item orally (by the second author). The
+complete recorded list was finally presented to two more informants, who
+could repeat all the items and who confirmed there were no real words.
+Even so, there was one monosyllable that was often identified as a real
+word (intended ``yî'' /yXX/; identified as ``yi'' /yi/, \emph{tree}).
 This item is removed from analyses below.
 
-The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
+The final list is composed of three practice items; 20 monoysllables
+containing sounds that are less frequent in the world's languages than
+singleton plosives; 8 bisyllables; 12 trisyllables; and 4
+quadrisyllables (see Table 2).
 
@@
 
-A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
-The 40 elicitation sets are available online from \url{https://osf.io/5qspb}.
+A Praat script was written to randomize this list 20 times, and split it
+into two sublists, to generate 40 different elicitation sets. The 40
+elicitation sets are available online from \url{https://osf.io/5qspb}.
 The split had the following constraints:
@@
 \item
-  the same three items were selected as practice items and used in all 40 elicitation sets
-\item
-  splits were done within each length group from the 2018 items (i.e., separately for 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
-Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
-\item
-  once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllables, then 3-syllables, and finally 4-syllable items.
+  the same three items were selected as practice items and used in all
+  40 elicitation sets
+\item
+  splits were done within each length group from the 2018 items (i.e.,
+  separately for 2, 3, and 4-syllable items); and among onset groups for
+  the difficult monosyllables generated in 2019 (i.e., all the
+  monosyllables starting with tp were split into 2 sublists). Since some
+  of these groups had an odd number of items, one of the sublists was
+  slightly longer than the other (20 versus 23).
+\item
+  once the sublist split had been done, items were randomized such that
+  all children heard first the 3 practice items in a fixed order (1, 2,
+  and 4 syllables), a randomized version of their sublist selection of
+  difficult onset items, and randomized versions of their 2-syllables,
+  then 3-syllables, and finally 4-syllable items.
 \end{itemize}
 
-\hypertarget{procedure}{%
-\subsubsection{Procedure}\label{procedure}}
-
-We tried to balance three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them; that children would feel comfortable doing this task with us; and that the community would feel safe with us doing this task with their children.
-Moreover, there were also some logistic constraints in terms of the space availability.
-As a result, the places where elicitation happened varied across the hamlets.
-
-We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items `spreading' through hearsay.
-In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
-The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
-
-The child was donned a headset (\textbf{xx }mc** please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
-For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headphones.
-
-The first phase was making sure the child understood the task.
-This was explained orally and the first training item was presented.
-Often, children froze and did not say anything.
-If this happened, then we followed this procedure.
-First, the informant insisted.
-If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another.
-If the child could repeat these correctly, then we provided the recorded training item over headphones again.
-Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it).
-A small minority still failed to repeat the item after hearing it over headphones.
-If that occurred, we tried with the second training item, at which point some children got it and could continue.
-A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.
-
-NWR studies vary in whether children are provided with several opportunities to hear and say the item.
-To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them.
-If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a separate headset that was recording onto the right track of the same Tascam).
-If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.
-
-Whenever siblings from the same family were tested, an attempt was made to test first the older and then the younger child, and always on different elicitaiton sets.
-
-\hypertarget{coding}{%
-\subsubsection{Coding}\label{coding}}
-
-A script was written to randomize all tokens from all children, pairing each with the auditory target the child had been provided with.
-A native research assistant then listened through all productions of a given target (randomized across children and repetition order), and made a judgment of whether the item was correctly or incorrectly repeated.
-She additionally orthographically transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-
-\hypertarget{analyses}{%
-\subsubsection{Analyses}\label{analyses}}
-
-Some NWR studies employ phoneme-based scoring in addition to or instead of word-level scoring (e.g., Cristia et al., 2020).
-We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
-Although previous work does not use distance metrics, we additionally report those.
-For instance, recall our example of an English target being ``bilik'', and imagine a response ``bilig''.
-This would receive the following scores: at the whole item level this would receive a score of zero (because the repetition is not exact); at the phoneme level it would receive a score of 80\% (4 out of 5 phonemes repeated exactly); and its normalized phone-based Levenshtein distance is 20\% (20\% of phonemes were substituted or deleted).\\
-
-Additionally, we looked up each phoneme in each target word in the Phoible database (Moran \& McCloy, 2019).
-We entered the number of languages in which each phoneme was found, as well as what percentage of languages in the database that represented.
-We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
-
-Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.
-
-\hypertarget{participants}{%
-\subsubsection{Participants}\label{participants}}
-
-This study was approved as part of a larger research effort by Marisa Casillas, with the research protocol having been approved by the IRB committee entitled The Ethics Committee of the Faculty of Social Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) of Radboud University, Nijmegen, The Netherlands.
-The approval for work outside of the Netherlands came as an amendment to the general research line on language development research (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
-Participation was voluntary, with children being invited to come and participate.
-Regardless of whether they did any of the task, children were provided with a snack as compensation.
-Children who came up to participate but then refused were nonetheless provided with the snack.
+\subsubsection{Procedure}\label{procedure}
+
+We tried to balance three desiderata: That children would not be unduly
+exposed to the items before they themselves had to repeat them; that
+children would feel comfortable doing this task with us; and that the
+community would feel safe with us doing this task with their children.
+Moreover, there were also some logistic constraints in terms of the
+space availability. As a result, the places where elicitation happened
+varied across the hamlets.
+
+We visited four different hamlets once, and attempted to test all
+eligible children present at the time, to prevent the items `spreading'
+through hearsay. In the first hamlet, we tested children in five
+different places, with some children being tested inside a house and
+others tested on the veranda. The complete list of places and the ways
+in which they met the desiderata mentioned above can be found in the raw
+data, available from online supplementary materials.
+
+The child was donned a headset (\textbf{xx }mc** please fill in** for
+most of the children, SHURE WH20 XLR headset with a dynamic microphone
+for the rest), recorded into the left channel into a Tascam DR40x
+digital recorder. For most children, the headset could not stay
+comfortably on the child's head, and thus it was placed on the child's
+shoulders, with the microphone carefully placed close to the child's
+mouth. A local informant sat next to the child, to would provide the
+instructions and, if needed, coach the child to make sure, using the
+three practice items as well as real words, that they understood that
+the task was to repeat the items precisely without changing anything. An
+experimenter (the first author) delivered the elicitation stimuli to the
+local informant and the child over headphones.
+
+The first phase was making sure the child understood the task. This was
+explained orally and the first training item was presented. Often,
+children froze and did not say anything. If this happened, then we
+followed this procedure. First, the informant insisted. If the child
+still did not say anything, the informant asked the child to repeat a
+real word, and another, and another. If the child could repeat these
+correctly, then we provided the recorded training item over headphones
+again. Most children successfully started repeating the items presented
+over headphones at this point; a few further needed the local informant
+to model the behavior (i.e., they would hear the item again, and she
+would say it; then we would play it again, and ask the child to say it).
+A small minority still failed to repeat the item after hearing it over
+headphones. If that occurred, we tried with the second training item, at
+which point some children got it and could continue. A small minority,
+however, failed to repeat this one, as well as the third training item,
+in which case we stopped the test altogether.
+
+NWR studies vary in whether children are provided with several
+opportunities to hear and say the item. To have a fixed and clear
+procedure, we decided that items other than the inital three training
+ones would not be repeated unless the child made an attempt to produce
+them. If this attempt was judged correct by the local informant, then
+the experimenter would move on to the next item (whispering this over a
+separate headset that was recording onto the right track of the same
+Tascam). If the local informant heard a deviation, she indicated to the
+experimenter that the item needed to be repeated, and up to 5 attempts
+were allowed.
+
+Whenever siblings from the same family were tested, an attempt was made
+to test first the older and then the younger child, and always on
+different elicitaiton sets.
+
+\subsubsection{Coding}\label{coding}
+
+A script was written to randomize all tokens from all children, pairing
+each with the auditory target the child had been provided with. A native
+research assistant then listened through all productions of a given
+target (randomized across children and repetition order), and made a
+judgment of whether the item was correctly or incorrectly repeated. She
+additionally orthographically transcribed exactly what the child said,
+providing some examples of the types of errors children in general make
+(without making specific reference to Yélî sounds or the items in the
+elicitation sets).
+
+\subsubsection{Analyses}\label{analyses}
+
+Some NWR studies employ phoneme-based scoring in addition to or instead
+of word-level scoring (e.g., Cristia et al., 2020). We scored items in
+terms of the number of phonemes that could be aligned across the target
+and attempt, divided by the number of phonemes of whichever item was
+longer (the target or the attempt). Although previous work does not use
+distance metrics, we additionally report those. For instance, recall our
+example of an English target being ``bilik'', and imagine a response
+``bilig''. This would receive the following scores: at the whole item
+level this would receive a score of zero (because the repetition is not
+exact); at the phoneme level it would receive a score of 80\% (4 out of
+5 phonemes repeated exactly); and its normalized phone-based Levenshtein
+distance is 20\% (20\% of phonemes were substituted or deleted).\\
+
+Additionally, we looked up each phoneme in each target word in the
+Phoible database (Moran \& McCloy, 2019). We entered the number of
+languages in which each phoneme was found, as well as what percentage of
+languages in the database that represented. We could not find a small
+number of phonemes in the database, and treat those as NA values that do
+not contribute to the target-level global average phone cross-linguistic
+frequency, although an alternative would be to attribute them a very
+small number.
+
+Finally, for describing children's patterns of errors, all repetitions
+of a given target were taken into account. We describe the proportion of
+items where the change resulted in a real word (semantic errors); as
+well as deletions, additions, and substitutions.
+
+\subsubsection{Participants}\label{participants}
+
+This study was approved as part of a larger research effort by Marisa
+Casillas, with the research protocol having been approved by the IRB
+committee entitled The Ethics Committee of the Faculty of Social
+Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen;
+ECSW) of Radboud University, Nijmegen, The Netherlands. The approval for
+work outside of the Netherlands came as an amendment to the general
+research line on language development research (original request:
+ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+Participation was voluntary, with children being invited to come and
+participate. Regardless of whether they did any of the task, children
+were provided with a snack as compensation. Children who came up to
+participate but then refused were nonetheless provided with the snack.
 
 A total 55 children were tested, from 34 of families, in five hamlets.
-Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5).
-In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
-The remaining 40 children (14 girls) were aged 6.96 years (range 3.90-11 years).
-Among these, there were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and 2 for whom this information is missing.
-Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Precise years of education can be hard to count.
-This is particularly the case for schooling beyond grade 8, which takes place off-island, which can be due to an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
-We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical ``completion'' times in between.} In terms of birth order, 5 were first borns, 4 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
-
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
-
-\hypertarget{preliminary-analyses}{%
-\subsubsection{Preliminary analyses}\label{preliminary-analyses}}
+Some children could not be included for the following reasons: refused
+participation or failed to repeat items presented over headphones even
+after coaching (N=8), spoke too softly to allow offline coding (N=5). In
+addition, 2 teenagers were tested to put younger children at ease; their
+data is not included in analyses below. The remaining 40 children (14
+girls) were aged 6.96 years (range 3.90-11 years). Among these, there
+were 32 children exposed only to Yélî in the home, 6 children who were
+also exposed to another language in the home, and 2 for whom this
+information is missing. Maternal years of education averaged 8.24 years
+(range 6-12 years; 2 children had this information missing).\footnote{Precise
+  years of education can be hard to count. This is particularly the case
+  for schooling beyond grade 8, which takes place off-island, which can
+  be due to an irregularity in ability to pay school fees, school
+  opening times, familial obligations, and more. We asked parents to
+  report their highest completed level of education and inferred the
+  approximate number of years that entails under ideal conditions, which
+  results in many esimates around typical ``completion'' times in
+  between.} In terms of birth order, 5 were first borns, 4 second, 2
+third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
+
+\subsection{Results}\label{results}
+
+\subsubsection{Preliminary analyses}\label{preliminary-analyses}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig1-first_vs_others-1.pdf}
-\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual participants averaging separately their first attempts and all other attempts.}
+\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual
+participants averaging separately their first attempts and all other
+attempts.}
 \end{figure}
 
-We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
-We excluded 3 children who did not have data for one of these two types.
-As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
+We first checked whether NWR scores varied between first and subsequent
+presentations of an item by averaging word-level scores at the
+participant level separately for first attempts and subsequent
+repetitions. We excluded 3 children who did not have data for one of
+these two types. As shown in Figure 1, participants' mean word-level
+scores became more heterogeneous in subsequent repetitions.
 Surprisingly, NWR scores for subsequent repetitions (M = 40, SD = 26)
-were on average lower than first ones (M = 64, SD = 15), t(38) = 6.30, p = 0.
-Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
-
-Taking into account only the first attempts, we averaged attempts by each of the 24 children who had data for first attempts; their ages ranged from 3.90 to 10.20 (M = 6.50, SD = 1.50).
-
-The overall NWR score was M = 61\% (SD = 15\%).
-Scores based on phonemes are even higher M = 100\% (SD = 0\%).
-The phoneme-based normalized Levenshtein distance was M = 0\% (SD = 0\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
-Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
-
-\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
-\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}
-
-In this analysis, we were interested in variation in NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
-To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID \& target ID as random variables.
-
-We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.79, SE ß = 0.19, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Age was also a significant predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
-
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/Fig2-xling_freq-1.pdf}
-\caption{(\#fig:Fig2-xling\_freq)NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}
+were on average lower than first ones (M = 64, SD = 15), t(38) = 6.30, p
+= 0. Given the uncertainty in whether previous work used only the first
+or all repetitions, and since scores declined and became more
+heterogeneous in subsequent repetitions, the rest of the analyses focus
+on only the first repetitions.
+
+Taking into account only the first attempts, we averaged attempts by
+each of the 24 children who had data for first attempts; their ages
+ranged from 3.90 to 10.20 (M = 6.50, SD = 1.50).
+
+The overall NWR score was M = 61\% (SD = 15\%). Scores based on phonemes
+are even higher M = 100\% (SD = 0\%). The phoneme-based normalized
+Levenshtein distance was M = 0\% (SD = 0\%), meaning that about a fifth
+of phonemes were were substituted, inserted, or deleted. Notice that the
+normalized Levenshtein distance is the complement of phoneme-based
+scores.
+
+\subsubsection{NWR as a function of cross-linguistic phone
+frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}
+
+In this analysis, we were interested in variation in NWR scores as a
+function of the average frequency with which sounds composing individual
+target words are found in languages over the world. To look at this, we
+fit a mixed logistic regression, in which the outcome variable was
+whether the non-word was correctly repeated or not. The fixed effects of
+interest was the average cross-linguistic phone frequency; we also
+included child age as a control fixed effect, and child ID \& target ID
+as random variables.
+
+We could include 885 observations (\textbf{why -- shouldn't it be
+1k??}), from 41 children producing in any given trial one of 41
+potential target words. The analysis revealed a significant estimate for
+the average cross-linguistic frequency of phones in the target words (ß
+= 0.79, SE ß = 0.19, p = 0): Target words with phones found more
+frequently across languages had a higher proportion of words that were
+correctly repeated, as clear in Figure 2. Age was also a significant
+predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
+
+\begin{figure}[!t]
+
+{\centering \includegraphics[width=0.65\linewidth]{nwr.by.freq.ITEM}
+
+}
+
+\caption{NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}\label{fig:Fig2-xling-freq}
 \end{figure}
 
-We next checked whether this may actually be due to frequency of the sounds within the language.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings ({\textbf{???}}), and applied the natural logarithm.
-As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(23)=0.65, =0).
-Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
-As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic frequency (ß = 0.51, SE ß = 0.25, p = 0.04) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0.39, SE ß = 0.24, p = 0.11) was not.
+We next checked whether this may actually be due to frequency of the
+sounds within the language. We estimated frequency of the phones present
+in the stimuli in a corpus of child-centered recordings (Casillas et
+al., 2020), and applied the natural logarithm. As before, unattested
+sounds were not considered (i.e., they were declared NA so that they do
+not count for analyses). Phone frequency estimated from this corpus were
+correlated with cross-linguistic phone frequencies (r(23)=0.65, =0).
+Therefore, we fit another mixed logistic regression, this time declaring
+as fixed effects scaled cross-linguistic and corpus frequencies
+(averaged across all attested phones within each stimulus item), in
+addition to age. As before, the model contained random slopes for both
+child ID and target. In this model, both cross-linguistic frequency (ß =
+0.51, SE ß = 0.25, p = 0.04) and age (ß = 0.33, SE ß = 0.13, p = 0.01)
+were significant predictors of word-level NWR scores, but corpus
+frequency (ß = 0.39, SE ß = 0.24, p = 0.11) was not.
 
@@
 
-\textbf{a2m use first attempts rather than the whole data set: having second thoughts.
-To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers \& it is more consistent.
-BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros\ldots{} CF Option 3 further down}
-
-\textbf{a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?}
-
-\textbf{Option 1: just numbers, first attempts only }
-Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 0 vowels deleted, and 0 consonants.
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels, 0 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality; 0 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0) than vice versa (0).
-Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones, compared to 0 complex ones).
-
-\textbf{Option 2: numbers and proportions, first attempts only}
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 0 vowels deleted (about NA\% of all vowel targets), and 0 consonants deleted (about NA\% of all consonant targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about NA\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality or about NA\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times or about NA\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones or about NA\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
-
-\textbf{Option 3: numbers and proportions, ALL attempts}
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 0 vowels deleted (about NA\% of all vowel targets), and 0 consonants deleted (about NA\% of all consonant targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about NA\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality or about NA\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times about NA\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones or about NA\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
-
-Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 97\%.
-
-\textbf{TO DISCUSS TOGETHER this is higher than any previous report! can we be sure of it?}
-
-\hypertarget{nwr-scores-as-a-function-of-item-length}{%
-\subsubsection{NWR scores as a function of item length}\label{nwr-scores-as-a-function-of-item-length}}
-
-Next, we inspected whether NWR scores varied as a function of word length.
-Results are shown on table XX.
-This table shows that scores for monosyllables was much lower than for other lengths.
-This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
-
-Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole word scoring.
-This is the most commonly reported type of score, but it is also the least forgiving.
-The pattern is less marked when other two scores are used, which are less sensitive to errors.
+\textbf{a2m use first attempts rather than the whole data set: having
+second thoughts. To be clear: the argument for using first attempts only
+is that all of the other analyses are on kids' first attempts, rather
+than all of their attempts, so we should use just first attempts here
+because it may confuse readers \& it is more consistent. BUT there are
+fewer mistakes this way, which is Options 1 and 2 below has lots of
+zeros\ldots{} CF Option 3 further down}
+
+\textbf{a2m probably turn the paragraph on substitutions into a table:
+having second thoughts, the table would look just like a para, no?}
+
+\textbf{Option 1: just numbers, first attempts only } Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 0 vowels deleted, and 0 consonants.
+
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (0 oral target vowels produced as nasal
+vowels, 0 nasal target vowels produced as oral vowels). Substitutions in
+which the oral nature was preserved but the quality of the vowel was
+changed were a great deal more common than changes in quality among
+nasal vowels (0 oral vowels produced with a different quality; 0 nasal
+vowels produced with a different quality). As for consonants,
+asymmetries were very marked with more complex consonants (specifically
+dptpkpkmknmbghlv) mispronounced as simple ones (specifically
+mnlwyvdgptkfhch, 0) than vice versa (0). Simple consonants were
+mispronounced as other simple consonants quite frequently (0 simple
+consonants mispronounced as other simple ones, compared to 0 complex
+ones).
+
+\textbf{Option 2: numbers and proportions, first attempts only} Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 0 vowels deleted (about NA\% of all vowel
+targets), and 0 consonants deleted (about NA\% of all consonant
+targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (0 oral target vowels produced as nasal
+vowels or about NA\% out of all oral vowel targets, 0 nasal target
+vowels produced as oral vowels or about NA\% out of all nasal vowel
+targets). Substitutions in which the oral nature was preserved but the
+quality of the vowel was changed were a great deal more common than
+changes in quality among nasal vowels (0 oral vowels produced with a
+different quality or about NA\% out of all oral vowel targets; 0 nasal
+vowels produced with a different quality or about NA\% out of all nasal
+vowel targets). As for consonants, asymmetries were very marked with
+more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as
+simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of
+all complex consonant targets) than vice versa (0 times or about NA\%
+out of all simple consonant targets). Simple consonants were
+mispronounced as other simple consonants quite frequently (0 simple
+consonants mispronounced as other simple ones or about NA\% out of all
+simple consonant targets, compared to 0 complex ones or about NA\% out
+of all complex consonant targets).
+
+\textbf{Option 3: numbers and proportions, ALL attempts} Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 0 vowels deleted (about NA\% of all vowel
+targets), and 0 consonants deleted (about NA\% of all consonant
+targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (0 oral target vowels produced as nasal
+vowels or about NA\% out of all oral vowel targets, 0 nasal target
+vowels produced as oral vowels or about NA\% out of all nasal vowel
+targets). Substitutions in which the oral nature was preserved but the
+quality of the vowel was changed were a great deal more common than
+changes in quality among nasal vowels (0 oral vowels produced with a
+different quality or about NA\% out of all oral vowel targets; 0 nasal
+vowels produced with a different quality or about NA\% out of all nasal
+vowel targets). As for consonants, asymmetries were very marked with
+more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as
+simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of
+all complex consonant targets) than vice versa (0 times about NA\% out
+of all simple consonant targets). Simple consonants were mispronounced
+as other simple consonants quite frequently (0 simple consonants
+mispronounced as other simple ones or about NA\% out of all simple
+consonant targets, compared to 0 complex ones or about NA\% out of all
+complex consonant targets).
+
+Finally, we looked into the frequency with which mispronunciations
+resulted in real words. Nearly all of them were: 97\%.
+
+\textbf{TO DISCUSS TOGETHER this is higher than any previous report! can
+we be sure of it?}
+
+\subsubsection{NWR scores as a function of item
+length}\label{nwr-scores-as-a-function-of-item-length}
+
+Next, we inspected whether NWR scores varied as a function of word
+length. Results are shown on table XX. This table shows that scores for
+monosyllables was much lower than for other lengths. This is likely due
+to the fact that the majority of monosyllables included were chosen
+because they had sounds that are rare in the world's languages, which
+may indicate that they are hard to produce or to perceive.
+
+Setting monosyllables aside, we observe the typical pattern of lower
+scores for longer items, although this is particularly salient for the
+whole word scoring. This is the most commonly reported type of score,
+but it is also the least forgiving. The pattern is less marked when
+other two scores are used, which are less sensitive to errors.
 
 \begin{table}[tbp]
-
 \begin{center}
 \begin{threeparttable}
-
 \caption{\label{tab:tablength}NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
-
 \begin{tabular}{lll}
@@
 \end{tabular}
-
 \end{threeparttable}
 \end{center}
-
 \end{table}
 
-\hypertarget{factor-structuring-individual-variation}{%
-\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}
+\subsubsection{Factor structuring individual
+variation}\label{factor-structuring-individual-variation}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig2-scores by age-1.pdf}
-\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual participants as a function of age and sex (blue = boys, pink = girls).}
+\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual
+participants as a function of age and sex (blue = boys, pink = girls).}
 \end{figure}
 
-Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,485.88) = 0.35, p = 0.09).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.08) = 21.39, p = 0.94), birth order (data missing for 17 children, 2,730.78) = -0.19, p = 0.38) or maternal education (data missing for 2 children, 2,147.84) = 0.07, p = 0.76)
-
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+Our final exploratory analysis assessed whether variation in scores was
+structured by factors that vary across individuals. As shown in Figure
+2, there was a greater deal of variance in earlier than later ages, with
+significantly higher NWR scores for older children (Spearman's rank
+correlation, given inequality of variance, rho (1,485.88) = 0.35, p =
+0.09). In contrast, there was no clear association between NWR scores,
+on the one hand, and sex (t (0.08) = 21.39, p = 0.94), birth order (data
+missing for 17 children, 2,730.78) = -0.19, p = 0.38) or maternal
+education (data missing for 2 children, 2,147.84) = 0.07, p = 0.76)
+
+\subsection{Discussion}\label{discussion}
 
@@
 \item
-  What is the overall repetition scores (whole word, phoneme based, distance)?
-\item
-  How does this change as a function of item complexity (number of syllables, sound complexity)?
+  What is the overall repetition scores (whole word, phoneme based,
+  distance)?
+\item
+  How does this change as a function of item complexity (number of
+  syllables, sound complexity)?
 \end{itemize}
@@
 \item
-  Prediction based on previous work: Children have higher scores for shorter items
-\item
-  Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.
-\item
-  it turns out we were right! the scores for 2-4 syll words decline only slightly with length
-\item
-  Prediction based on previous work: Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children achieve higher scores when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds {[}CITE{]}.
-Therefore, we expect higher scores for items with common sounds than in those with rare sounds.
-\item
-  Prediction for Yélî made before seeing the data: The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
-Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
-As a result, differences between easier and harder items may be smaller in this work than in other research.
+  Prediction based on previous work: Children have higher scores for
+  shorter items
+\item
+  Prediction for Yélî made before seeing the data: The length
+  distribution in Yélî words is more balanced than that in English, and
+  thus the score decline for poly- versus mono-syllables may be less
+  pronounced than that for English.
+\item
+  it turns out we were right! the scores for 2-4 syll words decline only
+  slightly with length
+\item
+  Prediction based on previous work: Similarly, we do not know of NWR
+  research that manipulates the difficulty of the sounds that are
+  included in the items, but word naming and other research suggests
+  that children achieve higher scores when producing easy and/or
+  typologically common sounds than difficult and/or typologically rare
+  sounds {[}CITE{]}. Therefore, we expect higher scores for items with
+  common sounds than in those with rare sounds.
+\item
+  Prediction for Yélî made before seeing the data: The Yélî sound
+  inventory is very large and compressed, with many similar sounds that
+  are acoustic and articulatory neighbors. Therefore, this may
+  constitute a pressure for children to have finer auditory skills (and
+  perhaps more precise articulations) than children speaking languages
+  with a simpler inventory. As a result, differences between easier and
+  harder items may be smaller in this work than in other research.
 \item
@@
 
-\textbf{(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)}
+\textbf{(MC: but we can try and do a cursory analysis based on the
+corpora we have from Steve and my transcription of naturalistic
+interactions!)}
 
@@
 \item
-  How frequent are errors that result in real words? Is that a function of item complexity?
-\item
-  Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
+  How frequent are errors that result in real words? Is that a function
+  of item complexity?
+\item
+  Is individual variation explainable by child age, sex, birth order,
+  monolingual status, and/or parental education?
 \end{itemize}
@@
 \item
-  General prediction: Non-monolingual Yélî children have lower scores than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
-\item
-  Local prediction: Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family
-  (or at least from people other than their mothers, who tend to be the non-native speakers).
-If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
-  . That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother
-  , so in so far as this input has a crucial formational role, then there may still be a score difference between these two groups.
+  General prediction: Non-monolingual Yélî children have lower scores
+  than monolingual ones when tested on the society-dominant language (we
+  did not test any non-dominant language) 
+\item
+  Local prediction: Anecdotally Yélî children grow up in close-knitted
+  communities and thus may receive significant portions of their
+  language input from people not in their nuclear family (or at least
+  from people other than their mothers, who tend to be the non-native
+  speakers). If so, the difference between monolinguals and not
+  monolinguals may be smaller than that found in other work . That said,
+  one recent study on the same population shows that most child-directed
+  input in the first 2 years does come from the mother , so in so far as
+  this input has a crucial formational role, then there may still be a
+  score difference between these two groups.
 \item
@@
 \item
-  General prediction: previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated score higher in standardized language tests (eg ppvt) than children whose mothers are less educated.
-\item
-  Prediction made before seeing data: In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
-This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
-However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
+  General prediction: previous NWR evidence on this is mixed, but
+  general findings on language development suggest that children whose
+  mothers are more educated score higher in standardized language tests
+  (eg ppvt) than children whose mothers are less educated.
+\item
+  Prediction made before seeing data: In the Rossel community, formal
+  education plays an extremely minor role in ensuring individual's
+  success, is not a good index of relative socio-economic status, and
+  furthermore there is only a narrow range of variation in maternal
+  educational attainment. This may lead to no or only very small
+  advantages for children whose mothers are more educated, provided that
+  the causal chain between maternal education and child language is via
+  SES more broadly. However, if education directly boosts maternal
+  verbal skills and the incidence of verbal behavior (as suggested by
+  CITE), then we should still see a difference along this factor.
 \item
@@
 \item
-  General prediction: To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children score higher in standardized language tests later-born children.
-\item
-  Prediction before seeing data: One main causal path between birth order and language development is via parental input (CITE).
-Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the difference in scores between first borns and later borns may be small
-\item
-  we did not find a sig effect, but this is a small effect (eg d=.2 in Havron et al)
+  General prediction: To our knowledge, there is no previous NWR work on
+  this, but other research suggests that first-born children score
+  higher in standardized language tests later-born children.
+\item
+  Prediction before seeing data: One main causal path between birth
+  order and language development is via parental input (CITE). Given our
+  arguments above for how mothers may not be as important among Rossel
+  people than in other places, then the difference in scores between
+  first borns and later borns may be small
+\item
+  we did not find a sig effect, but this is a small effect (eg d=.2 in
+  Havron et al)
 \end{enumerate}
@@
 
-\hypertarget{acknowledgments}{%
-\subsection{Acknowledgments}\label{acknowledgments}}
-
-We are grateful to the individuals who participated in the study.
-The collection and annotation of these recordings was made possible by Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also to the PNG National Research Institute, and the Administration of Milne Bay Province AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.
-S.
-McDonnell Foundation Understanding Human Cognition Scholar Award.
-MC acknowledges financial support from an NWO Veni Innovational Scheme grant (275-89-033).
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
+\subsection{Acknowledgments}\label{acknowledgments}
+
+We are grateful to the individuals who participated in the study. The
+collection and annotation of these recordings was made possible by
+Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also
+to the PNG National Research Institute, and the Administration of Milne
+Bay Province AC acknowledges financial and institutional support from
+Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge,
+ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and
+the J. S. McDonnell Foundation Understanding Human Cognition Scholar
+Award. MC acknowledges financial support from an NWO Veni Innovational
+Scheme grant (275-89-033).
+
+\section{References}\label{references}
+
+\setlength{\parindent}{-0.5in} \setlength{\leftskip}{0.5in}
 
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-is08042009language}{}%
-Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
-\leavevmode\hypertarget{ref-balladares2016socio}{}%
-Balladares, J., Marshall, C., \& Griffiths, Y.
-(2016).
-Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
-\emph{First Language}, \emph{36}(3), 338--351.
-\url{https://doi.org/10.1177/0142723715626067}
-
-\leavevmode\hypertarget{ref-bergelsoncasillas2019what}{}%
+\hypertarget{ref-is08042009language}{}
+Action, C. (2009). Language impairment in a multilingual society:
+Linguistic patterns and the road to assessment. \emph{Brussels: COST
+Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
+\hypertarget{ref-balladares2016socio}{}
+Balladares, J., Marshall, C., \& Griffiths, Y. (2016). Socio-economic
+status affects sentence repetition, but not non-word repetition, in
+Chilean preschoolers. \emph{First Language}, \emph{36}(3), 338--351.
+doi:\href{https://doi.org/10.1177/0142723715626067}{10.1177/0142723715626067}
+
+\hypertarget{ref-bergelsoncasillas2019what}{}
 Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
-S., \& Amatuni, A.
-(2019).
-What do North American babies hear?
-A large-scale cross-corpus analysis.
-\emph{Developmental Science}, \emph{22}(1), e12724.
-\url{https://doi.org/10.1111/desc.12724}
-
-\leavevmode\hypertarget{ref-bowey2001nonword}{}%
-Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
-
-\leavevmode\hypertarget{ref-brandeker2015language}{}%
-Brandeker, M., \& Thordardottir, E.
-(2015).
-Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
-\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
-
-\leavevmode\hypertarget{ref-brown2011cultural}{}%
-Brown, P.
-(2011).
-The cultural organization of attention.
-In A.
-Duranti, E.
-Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language Socialization} (pp.
-29--55).
-Malden, MA: Wiley-Blackwell.
-
-\leavevmode\hypertarget{ref-brown2014interactional}{}%
-Brown, P.
-(2014).
-The interactional context of language learning in Tzeltal.
-In I.
-Arnon, M.
-Casillas, C.
-Kurumada, \& B.
-Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of Eve V.
-Clark} (pp.
-51--82).
-Amsterdam, NL: John Benjamins.
-
-\leavevmode\hypertarget{ref-brownIPchildrearing}{}%
-Brown, P., \& Casillas, M.
-(n.d.).
-Childrearing through social interaction on Rossel Island, PNG.
-In A.
-J.
-Fentiman \& M.
-Goody (Eds.), \emph{Esther Goody revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
-XX--XX).
-New York, NY: Berghahn.
-
-\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
-Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (n.d.). A cross-cultural examination of young children's everyday language experiences.
-
-\leavevmode\hypertarget{ref-casillas2020early}{}%
-Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
-
-\leavevmode\hypertarget{ref-chiat2007preschool}{}%
-Chiat, S., \& Roy, P.
-(2007).
-The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
-
-\leavevmode\hypertarget{ref-cristia2020infant}{}%
+S., \& Amatuni, A. (2019). What do North American babies hear? A
+large-scale cross-corpus analysis. \emph{Developmental Science},
+\emph{22}(1), e12724.
+doi:\href{https://doi.org/10.1111/desc.12724}{10.1111/desc.12724}
+
+\hypertarget{ref-bowey2001nonword}{}
+Bowey, J. A. (2001). Nonword repetition and young children's receptive
+vocabulary: A longitudinal study. \emph{Applied Psycholinguistics},
+\emph{22}(3), 441--469.
+
+\hypertarget{ref-brandeker2015language}{}
+Brandeker, M., \& Thordardottir, E. (2015). Language exposure in
+bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2),
+126--138.
+
+\hypertarget{ref-brown2011cultural}{}
+Brown, P. (2011). The cultural organization of attention. In A. Duranti,
+E. Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language
+Socialization} (pp. 29--55). Malden, MA: Wiley-Blackwell.
+
+\hypertarget{ref-brown2014interactional}{}
+Brown, P. (2014). The interactional context of language learning in
+Tzeltal. In I. Arnon, M. Casillas, C. Kurumada, \& B. Estigarribia
+(Eds.), \emph{Language in interaction: Studies in honor of Eve V. Clark}
+(pp. 51--82). Amsterdam, NL: John Benjamins.
+
+\hypertarget{ref-brownIPchildrearing}{}
+Brown, P., \& Casillas, M. (in press). Childrearing through social
+interaction on Rossel Island, PNG. In A. J. Fentiman \& M. Goody (Eds.),
+\emph{Esther Goody revisited: Exploring the legacy of an original
+inter-disciplinarian} (pp. XX--XX). New York, NY: Berghahn.
+
+\hypertarget{ref-bunceURcrosscultural}{}
+Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A.,
+Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural
+examination of young children's everyday language experiences.
+
+\hypertarget{ref-casillas2020early}{}
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language
+experience in a papuan community. \emph{Journal of Child Language},
+\emph{XX}, XX--XX.
+
+\hypertarget{ref-chiat2007preschool}{}
+Chiat, S., \& Roy, P. (2007). The preschool repetition test: An
+evaluation of performance in typically developing and clinically
+referred children. \emph{Journal of Speech, Language, and Hearing
+Research}, \emph{50}(2), 429--443.
+
+\hypertarget{ref-cristia2020infant}{}
 Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
-(2020).
-Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
+(2020). Infant-directed input and literacy effects on phonological
+processing: Non-word repetition scores among the tsimane'.
 \emph{Preprint}.
 
-\leavevmode\hypertarget{ref-estes2007differences}{}%
-Estes, K.
-G., Evans, J.
-L., \& Else-Quest, N.
-M.
-(2007).
-Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
-
-\leavevmode\hypertarget{ref-farmani2018normalization}{}%
-Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
-Z., Talebi, E., \& Shourvazi, Z.
-(2018).
-Normalization of the non-word repetition test in farsi-speaking children.
-\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
-
-\leavevmode\hypertarget{ref-gallagher2014identity}{}%
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
-
-\leavevmode\hypertarget{ref-gallon2007non}{}%
-Gallon, N., Harris, J., \& Van der Lely, H.
-(2007).
-Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
-\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
-
-\leavevmode\hypertarget{ref-havron2019effect}{}%
-Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
-M.-C.
-C.
-S.
-(2019).
-The effect of older siblings on language development as a function of age difference and sex.
-\emph{Psychological Science}, \emph{30}(9), 1333--1343.
-
-\leavevmode\hypertarget{ref-hazan2000development}{}%
-Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
-
-\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
-
-\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
-(2014).
-Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
+\hypertarget{ref-estes2007differences}{}
+Estes, K. G., Evans, J. L., \& Else-Quest, N. M. (2007). Differences in
+the nonword repetition performance of children with and without specific
+language impairment: A meta-analysis. \emph{Journal of Speech, Language,
+and Hearing Research}, \emph{50}(1), 177--195.
+
+\hypertarget{ref-farmani2018normalization}{}
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \&
+Shourvazi, Z. (2018). Normalization of the non-word repetition test in
+farsi-speaking children. \emph{Journal of Modern Rehabilitation},
+\emph{12}(4), 217--224.
+
+\hypertarget{ref-gallagher2014identity}{}
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from
+Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378.
+doi:\href{https://doi.org/10.1515/lp-2014-0012}{10.1515/lp-2014-0012}
+
+\hypertarget{ref-gallon2007non}{}
+Gallon, N., Harris, J., \& Van der Lely, H. (2007). Non-word repetition:
+An investigation of phonological complexity in children with grammatical
+sli. \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\hypertarget{ref-havron2019effect}{}
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \&
+Group, E. M.-C. C. S. (2019). The effect of older siblings on language
+development as a function of age difference and sex. \emph{Psychological
+Science}, \emph{30}(9), 1333--1343.
+
+\hypertarget{ref-hazan2000development}{}
+Hazan, V., \& Barrett, S. (2000). The development of phonemic
+categorization in children aged 6--12. \emph{Journal of Phonetics},
+\emph{28}(4), 377--396.
+
+\hypertarget{ref-jabere2018xperiment}{}
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual
+Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4),
+317--334.
+doi:\href{https://doi.org/10.30958/ajp.5-4-4}{10.30958/ajp.5-4-4}
+
+\hypertarget{ref-kalnak2014nonword}{}
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014).
+Nonword repetition--a clinical marker for specific language impairment
+in swedish associated with parents' language-related problems.
 \emph{PloS One}, \emph{9}(2), e89544.
 
-\leavevmode\hypertarget{ref-levinsoninprepydgrammar}{}%
-Levinson, S. C. (n.d.). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Cambridge, UK: Cambridge University Press.
-
-\leavevmode\hypertarget{ref-liszkowski2012prelinguistic}{}%
+\hypertarget{ref-levinsoninprepydgrammar}{}
+Levinson, S. C. (in preparation). \emph{A grammar of yélî dnye, the
+papuan language of rossel island}. Cambridge, UK: Cambridge University
+Press.
+
+\hypertarget{ref-liszkowski2012prelinguistic}{}
 Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
-(2012).
-A prelinguistic gestural universal of human communication.
+(2012). A prelinguistic gestural universal of human communication.
 \emph{Cognitive Science}, \emph{36}(4), 698--713.
-\url{https://doi.org/10.1111/j.1551-6709.2011.01228.x}
-
-\leavevmode\hypertarget{ref-meir2017independent}{}%
-Meir, N., \& Armon-Lotem, S.
-(2017).
-Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
-\emph{Frontiers in Psychology}, \emph{8}, 1442.
-
-\leavevmode\hypertarget{ref-meir2016disentangling}{}%
-Meir, N., Walters, J., \& Armon-Lotem, S.
-(2016).
-Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
-\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
-
-\leavevmode\hypertarget{ref-phoible}{}%
-Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
-
-\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
-Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
-(2019).
-An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
-\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
-
-\leavevmode\hypertarget{ref-torrington2015non}{}%
-Torrington Eaton, C., Newman, R.
-S., Ratner, N.
-B., \& Rowe, M.
-L.
-(2015).
-Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
-\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
-
-\leavevmode\hypertarget{ref-vance2005speech}{}%
-Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
-
-\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
-Wilsenach, C.
-(2013).
-Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
-\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
-\url{https://doi.org/10.5785/29-2-554}
-
-\leavevmode\hypertarget{ref-is08042009language}{}%
-Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
-\leavevmode\hypertarget{ref-balladares2016socio}{}%
-Balladares, J., Marshall, C., \& Griffiths, Y.
-(2016).
-Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
-\emph{First Language}, \emph{36}(3), 338--351.
-\url{https://doi.org/10.1177/0142723715626067}
-
-\leavevmode\hypertarget{ref-bergelsoncasillas2019what}{}%
+doi:\href{https://doi.org/10.1111/j.1551-6709.2011.01228.x}{10.1111/j.1551-6709.2011.01228.x}
+
+\hypertarget{ref-meir2017independent}{}
+Meir, N., \& Armon-Lotem, S. (2017). Independent and combined effects of
+socioeconomic status (ses) and bilingualism on children's vocabulary and
+verbal short-term memory. \emph{Frontiers in Psychology}, \emph{8},
+1442.
+
+\hypertarget{ref-meir2016disentangling}{}
+Meir, N., Walters, J., \& Armon-Lotem, S. (2016). Disentangling sli and
+bilingualism using sentence repetition tasks: The impact of l1 and l2
+properties. \emph{International Journal of Bilingualism}, \emph{20}(4),
+421--452.
+
+\hypertarget{ref-phoible}{}
+Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max
+Planck Institute for the Science of Human History. Retrieved from
+\url{https://phoible.org/}
+
+\hypertarget{ref-piazzalunga2019articulatory}{}
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler,
+A. (2019). An articulatory-based disyllabic and trisyllabic non-word
+repetition test: Reliability and validity in italian 3-to 7-year-old
+children. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5),
+437--456.
+
+\hypertarget{ref-torrington2015non}{}
+Torrington Eaton, C., Newman, R. S., Ratner, N. B., \& Rowe, M. L.
+(2015). Non-word repetition in 2-year-olds: Replication of an adapted
+paradigm and a useful methodological extension. \emph{Clinical
+Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+
+\hypertarget{ref-vance2005speech}{}
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills
+in children aged 3--7 years. \emph{International Journal of Language \&
+Communication Disorders}, \emph{40}(1), 29--48.
+
+\hypertarget{ref-wilsenach2013phonological}{}
+Wilsenach, C. (2013). Phonological skills as predictor of reading
+success: An investigation of emergent bilingual Northern Sotho/English
+learners. \emph{Per Linguam: a Journal of Language Learning= Per
+Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+doi:\href{https://doi.org/10.5785/29-2-554}{10.5785/29-2-554}
+
+\hypertarget{ref-is08042009language}{}
+Action, C. (2009). Language impairment in a multilingual society:
+Linguistic patterns and the road to assessment. \emph{Brussels: COST
+Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
+\hypertarget{ref-balladares2016socio}{}
+Balladares, J., Marshall, C., \& Griffiths, Y. (2016). Socio-economic
+status affects sentence repetition, but not non-word repetition, in
+Chilean preschoolers. \emph{First Language}, \emph{36}(3), 338--351.
+doi:\href{https://doi.org/10.1177/0142723715626067}{10.1177/0142723715626067}
+
+\hypertarget{ref-bergelsoncasillas2019what}{}
 Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
-S., \& Amatuni, A.
-(2019).
-What do North American babies hear?
-A large-scale cross-corpus analysis.
-\emph{Developmental Science}, \emph{22}(1), e12724.
-\url{https://doi.org/10.1111/desc.12724}
-
-\leavevmode\hypertarget{ref-bowey2001nonword}{}%
-Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
-
-\leavevmode\hypertarget{ref-brandeker2015language}{}%
-Brandeker, M., \& Thordardottir, E.
-(2015).
-Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
-\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
-
-\leavevmode\hypertarget{ref-brown2011cultural}{}%
-Brown, P.
-(2011).
-The cultural organization of attention.
-In A.
-Duranti, E.
-Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language Socialization} (pp.
-29--55).
-Malden, MA: Wiley-Blackwell.
-
-\leavevmode\hypertarget{ref-brown2014interactional}{}%
-Brown, P.
-(2014).
-The interactional context of language learning in Tzeltal.
-In I.
-Arnon, M.
-Casillas, C.
-Kurumada, \& B.
-Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of Eve V.
-Clark} (pp.
-51--82).
-Amsterdam, NL: John Benjamins.
-
-\leavevmode\hypertarget{ref-brownIPchildrearing}{}%
-Brown, P., \& Casillas, M.
-(n.d.).
-Childrearing through social interaction on Rossel Island, PNG.
-In A.
-J.
-Fentiman \& M.
-Goody (Eds.), \emph{Esther Goody revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
-XX--XX).
-New York, NY: Berghahn.
-
-\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
-Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (n.d.). A cross-cultural examination of young children's everyday language experiences.
-
-\leavevmode\hypertarget{ref-casillas2020early}{}%
-Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
-
-\leavevmode\hypertarget{ref-chiat2007preschool}{}%
-Chiat, S., \& Roy, P.
-(2007).
-The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
-
-\leavevmode\hypertarget{ref-cristia2020infant}{}%
+S., \& Amatuni, A. (2019). What do North American babies hear? A
+large-scale cross-corpus analysis. \emph{Developmental Science},
+\emph{22}(1), e12724.
+doi:\href{https://doi.org/10.1111/desc.12724}{10.1111/desc.12724}
+
+\hypertarget{ref-bowey2001nonword}{}
+Bowey, J. A. (2001). Nonword repetition and young children's receptive
+vocabulary: A longitudinal study. \emph{Applied Psycholinguistics},
+\emph{22}(3), 441--469.
+
+\hypertarget{ref-brandeker2015language}{}
+Brandeker, M., \& Thordardottir, E. (2015). Language exposure in
+bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2),
+126--138.
+
+\hypertarget{ref-brown2011cultural}{}
+Brown, P. (2011). The cultural organization of attention. In A. Duranti,
+E. Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language
+Socialization} (pp. 29--55). Malden, MA: Wiley-Blackwell.
+
+\hypertarget{ref-brown2014interactional}{}
+Brown, P. (2014). The interactional context of language learning in
+Tzeltal. In I. Arnon, M. Casillas, C. Kurumada, \& B. Estigarribia
+(Eds.), \emph{Language in interaction: Studies in honor of Eve V. Clark}
+(pp. 51--82). Amsterdam, NL: John Benjamins.
+
+\hypertarget{ref-brownIPchildrearing}{}
+Brown, P., \& Casillas, M. (in press). Childrearing through social
+interaction on Rossel Island, PNG. In A. J. Fentiman \& M. Goody (Eds.),
+\emph{Esther Goody revisited: Exploring the legacy of an original
+inter-disciplinarian} (pp. XX--XX). New York, NY: Berghahn.
+
+\hypertarget{ref-bunceURcrosscultural}{}
+Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A.,
+Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural
+examination of young children's everyday language experiences.
+
+\hypertarget{ref-casillas2020early}{}
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language
+experience in a papuan community. \emph{Journal of Child Language},
+\emph{XX}, XX--XX.
+
+\hypertarget{ref-chiat2007preschool}{}
+Chiat, S., \& Roy, P. (2007). The preschool repetition test: An
+evaluation of performance in typically developing and clinically
+referred children. \emph{Journal of Speech, Language, and Hearing
+Research}, \emph{50}(2), 429--443.
+
+\hypertarget{ref-cristia2020infant}{}
 Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
-(2020).
-Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
+(2020). Infant-directed input and literacy effects on phonological
+processing: Non-word repetition scores among the tsimane'.
 \emph{Preprint}.
 
-\leavevmode\hypertarget{ref-estes2007differences}{}%
-Estes, K.
-G., Evans, J.
-L., \& Else-Quest, N.
-M.
-(2007).
-Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
-
-\leavevmode\hypertarget{ref-farmani2018normalization}{}%
-Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
-Z., Talebi, E., \& Shourvazi, Z.
-(2018).
-Normalization of the non-word repetition test in farsi-speaking children.
-\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
-
-\leavevmode\hypertarget{ref-gallagher2014identity}{}%
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
-
-\leavevmode\hypertarget{ref-gallon2007non}{}%
-Gallon, N., Harris, J., \& Van der Lely, H.
-(2007).
-Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
-\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
-
-\leavevmode\hypertarget{ref-havron2019effect}{}%
-Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
-M.-C.
-C.
-S.
-(2019).
-The effect of older siblings on language development as a function of age difference and sex.
-\emph{Psychological Science}, \emph{30}(9), 1333--1343.
-
-\leavevmode\hypertarget{ref-hazan2000development}{}%
-Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
-
-\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
-
-\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
-(2014).
-Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
+\hypertarget{ref-estes2007differences}{}
+Estes, K. G., Evans, J. L., \& Else-Quest, N. M. (2007). Differences in
+the nonword repetition performance of children with and without specific
+language impairment: A meta-analysis. \emph{Journal of Speech, Language,
+and Hearing Research}, \emph{50}(1), 177--195.
+
+\hypertarget{ref-farmani2018normalization}{}
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \&
+Shourvazi, Z. (2018). Normalization of the non-word repetition test in
+farsi-speaking children. \emph{Journal of Modern Rehabilitation},
+\emph{12}(4), 217--224.
+
+\hypertarget{ref-gallagher2014identity}{}
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from
+Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378.
+doi:\href{https://doi.org/10.1515/lp-2014-0012}{10.1515/lp-2014-0012}
+
+\hypertarget{ref-gallon2007non}{}
+Gallon, N., Harris, J., \& Van der Lely, H. (2007). Non-word repetition:
+An investigation of phonological complexity in children with grammatical
+sli. \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\hypertarget{ref-havron2019effect}{}
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \&
+Group, E. M.-C. C. S. (2019). The effect of older siblings on language
+development as a function of age difference and sex. \emph{Psychological
+Science}, \emph{30}(9), 1333--1343.
+
+\hypertarget{ref-hazan2000development}{}
+Hazan, V., \& Barrett, S. (2000). The development of phonemic
+categorization in children aged 6--12. \emph{Journal of Phonetics},
+\emph{28}(4), 377--396.
+
+\hypertarget{ref-jabere2018xperiment}{}
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual
+Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4),
+317--334.
+doi:\href{https://doi.org/10.30958/ajp.5-4-4}{10.30958/ajp.5-4-4}
+
+\hypertarget{ref-kalnak2014nonword}{}
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014).
+Nonword repetition--a clinical marker for specific language impairment
+in swedish associated with parents' language-related problems.
 \emph{PloS One}, \emph{9}(2), e89544.
 
-\leavevmode\hypertarget{ref-levinsoninprepydgrammar}{}%
-Levinson, S. C. (n.d.). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Cambridge, UK: Cambridge University Press.
-
-\leavevmode\hypertarget{ref-liszkowski2012prelinguistic}{}%
+\hypertarget{ref-levinsoninprepydgrammar}{}
+Levinson, S. C. (in preparation). \emph{A grammar of yélî dnye, the
+papuan language of rossel island}. Cambridge, UK: Cambridge University
+Press.
+
+\hypertarget{ref-liszkowski2012prelinguistic}{}
 Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
-(2012).
-A prelinguistic gestural universal of human communication.
+(2012). A prelinguistic gestural universal of human communication.
 \emph{Cognitive Science}, \emph{36}(4), 698--713.
-\url{https://doi.org/10.1111/j.1551-6709.2011.01228.x}
-
-\leavevmode\hypertarget{ref-meir2017independent}{}%
-Meir, N., \& Armon-Lotem, S.
-(2017).
-Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
-\emph{Frontiers in Psychology}, \emph{8}, 1442.
-
-\leavevmode\hypertarget{ref-meir2016disentangling}{}%
-Meir, N., Walters, J., \& Armon-Lotem, S.
-(2016).
-Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
-\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
-
-\leavevmode\hypertarget{ref-phoible}{}%
-Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
-
-\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
-Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
-(2019).
-An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
-\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
-
-\leavevmode\hypertarget{ref-torrington2015non}{}%
-Torrington Eaton, C., Newman, R.
-S., Ratner, N.
-B., \& Rowe, M.
-L.
-(2015).
-Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
-\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
-
-\leavevmode\hypertarget{ref-vance2005speech}{}%
-Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
-
-\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
-Wilsenach, C.
-(2013).
-Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
-\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
-\url{https://doi.org/10.5785/29-2-554}
+doi:\href{https://doi.org/10.1111/j.1551-6709.2011.01228.x}{10.1111/j.1551-6709.2011.01228.x}
+
+\hypertarget{ref-meir2017independent}{}
+Meir, N., \& Armon-Lotem, S. (2017). Independent and combined effects of
+socioeconomic status (ses) and bilingualism on children's vocabulary and
+verbal short-term memory. \emph{Frontiers in Psychology}, \emph{8},
+1442.
+
+\hypertarget{ref-meir2016disentangling}{}
+Meir, N., Walters, J., \& Armon-Lotem, S. (2016). Disentangling sli and
+bilingualism using sentence repetition tasks: The impact of l1 and l2
+properties. \emph{International Journal of Bilingualism}, \emph{20}(4),
+421--452.
+
+\hypertarget{ref-phoible}{}
+Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max
+Planck Institute for the Science of Human History. Retrieved from
+\url{https://phoible.org/}
+
+\hypertarget{ref-piazzalunga2019articulatory}{}
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler,
+A. (2019). An articulatory-based disyllabic and trisyllabic non-word
+repetition test: Reliability and validity in italian 3-to 7-year-old
+children. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5),
+437--456.
+
+\hypertarget{ref-torrington2015non}{}
+Torrington Eaton, C., Newman, R. S., Ratner, N. B., \& Rowe, M. L.
+(2015). Non-word repetition in 2-year-olds: Replication of an adapted
+paradigm and a useful methodological extension. \emph{Clinical
+Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+
+\hypertarget{ref-vance2005speech}{}
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills
+in children aged 3--7 years. \emph{International Journal of Language \&
+Communication Disorders}, \emph{40}(1), 29--48.
+
+\hypertarget{ref-wilsenach2013phonological}{}
+Wilsenach, C. (2013). Phonological skills as predictor of reading
+success: An investigation of emergent bilingual Northern Sotho/English
+learners. \emph{Per Linguam: a Journal of Language Learning= Per
+Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+doi:\href{https://doi.org/10.5785/29-2-554}{10.5785/29-2-554}
 

### 2020-09-16 manuscript.tex
various bug fixes

--- 2020-09-14 manuscript.tex
@@
 As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
-Surprisingly, NWR scores for subsequent repetitions (M = 40, SD = 26)
-were on average lower than first ones (M = 64, SD = 15), t(38) = 6.30, p = 0.
+Surprisingly, NWR scores for subsequent repetitions (M = 39, SD = 26)
+were on average lower than first ones (M = 65, SD = 15), t(38) = 6.60, p = 0.
 Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
@@
 The overall NWR score was M = 61\% (SD = 15\%).
-Scores based on phonemes are even higher M = 100\% (SD = 0\%).
-The phoneme-based normalized Levenshtein distance was M = 0\% (SD = 0\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
+Scores based on phonemes are even higher M = 77\% (SD = 10\%).
+The phoneme-based normalized Levenshtein distance was M = 23\% (SD = 10\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
 Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
@@
 We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.79, SE ß = 0.19, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Age was also a significant predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
+The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.81, SE ß = 0.18, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+Age was also a significant predictor (ß = 0.32, SE ß = 0.13, p = 0.01).
 
@@
 As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(23)=0.65, =0).
+Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(27)=0.52, =0.00).
 Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
 As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic frequency (ß = 0.51, SE ß = 0.25, p = 0.04) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0.39, SE ß = 0.24, p = 0.11) was not.
+In this model, both cross-linguistic frequency (ß = 0.81, SE ß = 0.27, p = 0.00) and age (ß = 0.32, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0.00, SE ß = 0.25, p = 1.00) was not.
 
@@
 \textbf{Option 1: just numbers, first attempts only }
-Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 0 vowels deleted, and 0 consonants.
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels, 0 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality; 0 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0) than vice versa (0).
-Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones, compared to 0 complex ones).
+Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 7 vowels deleted, and 3 consonants.
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels, 31 nasal target vowels produced as oral vowels).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (112 oral vowels produced with a different quality; 13 nasal vowels produced with a different quality).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 117) than vice versa (2).
+Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones, compared to 33 complex ones).
 
@@
 Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 0 vowels deleted (about NA\% of all vowel targets), and 0 consonants deleted (about NA\% of all consonant targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about NA\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality or about NA\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times or about NA\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones or about NA\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
+Deletions were relatively rare, with only 7 vowels deleted (about 0.23\% of all vowel targets), and 3 consonants deleted (about 0.12\% of all consonant targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels or about 0.98\% out of all oral vowel targets, 31 nasal target vowels produced as oral vowels or about 2.37\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (112 oral vowels produced with a different quality or about 1.82\% out of all oral vowel targets; 13 nasal vowels produced with a different quality or about 2.04\% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 117 times or about 1.95\% out of all complex consonant targets) than vice versa (2 times or about 1.95\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones or about 0.38\% out of all simple consonant targets, compared to 33 complex ones or about 0.99\% out of all complex consonant targets).
 
@@
 Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 0 vowels deleted (about NA\% of all vowel targets), and 0 consonants deleted (about NA\% of all consonant targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (0 oral target vowels produced as nasal vowels or about NA\% out of all oral vowel targets, 0 nasal target vowels produced as oral vowels or about NA\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (0 oral vowels produced with a different quality or about NA\% out of all oral vowel targets; 0 nasal vowels produced with a different quality or about NA\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 0 times or about NA\% out of all complex consonant targets) than vice versa (0 times about NA\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (0 simple consonants mispronounced as other simple ones or about NA\% out of all simple consonant targets, compared to 0 complex ones or about NA\% out of all complex consonant targets).
+Deletions were relatively rare, with only 12 vowels deleted (about 0.28\% of all vowel targets), and 6 consonants deleted (about 0.19\% of all consonant targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels or about 1.10\% out of all oral vowel targets, 58 nasal target vowels produced as oral vowels or about 2.53\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (205 oral vowels produced with a different quality or about 2.16\% out of all oral vowel targets; 23 nasal vowels produced with a different quality or about 2.36\% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or about 2.43\% out of all complex consonant targets) than vice versa (2 times about 2.43\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (123 simple consonants mispronounced as other simple ones or about 0.53\% out of all simple consonant targets, compared to 55 complex ones or about 0.98\% out of all complex consonant targets).
 
@@
 \midrule
-47 (22) & 100 (0) & 0 (0)\\
-79 (22) & 100 (0) & 0 (0)\\
-78 (19) & 100 (0) & 0 (0)\\
-75 (32) & 100 (0) & 0 (0)\\
+47 (22) & 59 (17) & 41 (17)\\
+79 (22) & 92 (9) & 8 (9)\\
+78 (19) & 93 (7) & 7 (7)\\
+74 (32) & 91 (12) & 9 (12)\\
 \bottomrule
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,485.88) = 0.35, p = 0.09).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.08) = 21.39, p = 0.94), birth order (data missing for 17 children, 2,730.78) = -0.19, p = 0.38) or maternal education (data missing for 2 children, 2,147.84) = 0.07, p = 0.76)
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,483.87) = 0.35, p = 0.09).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.13) = 21.37, p = 0.90), birth order (data missing for 17 children, 2,600.01) = -0.13, p = 0.54) or maternal education (data missing for 2 children, 2,045.68) = 0.11, p = 0.61)
 

### 2020-09-16 wrangling.R

@@ -1,3 +1,4 @@
+library(tidyverse)
 
 read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
 #DATA REMOVAL *** ATTENTION *** 
@@ -13,7 +14,7 @@ trnsc$tokid=paste(trnsc$item,trnsc$token)
 
 read.table("final_order.txt",header=T)->crp
 crp$tokid=paste(crp$target,crp$nb)
-npairs=dim(crp)[1]
+npairs=dim(crp)[1] #1488 pairs
 
 trnsc=merge(trnsc,crp,by.x="tokid",by.y="tokid")
 dim(trnsc) #haven't lost any judgments
@@ -25,11 +26,13 @@ trnsc[order(trnsc$file, trnsc$int),]->trnsc
 trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
 
 # AC's prev code
-# trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
-# trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
-# trnsc$attempt[1]<-"first"
-# table(trnsc$attempt) #1045 first attempts
+ trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
+ trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
+ trnsc$attempt[1]<-"first"
+ table(trnsc$attempt) #1045 first attempts
 
+ head(trnsc) #previous looks right
+ 
 # MC's new code
 # M2A: by my count there are 1029 first attempts, but maybe I'm misunderstanding "token"
 # M2A: okay, my solution finds some conflicting cases with your outfile attempt namings,
@@ -41,16 +44,17 @@ trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
 # M2A: Okay, great! Well that still returns 1029 first cases, but now the tidyverse pipe below
 # should guarantee we're getting the cases we want
 
-first.prods <- trnsc %>%
-  group_by(id, item) %>%
-  summarize(min.int = min(int)) %>%
-  mutate(attempt = "first") %>%
-  rename("int" = min.int)
-trnsc <- trnsc %>%
-  left_join(first.prods, by = c("id", "item", "int")) %>%
-  replace_na(list(attempt = "subsequent"))
[…]

### 2020-09-16 manuscript.Rmd

@@
 - look through this manuscript for bits of text to be written by **mc**
-- do you think Figure 2 should be done with phonetic symbols? If yes, could you try to implement that?
 - double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
@@
 - double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
-- double check phonological implementation of stimuli in stimuli.xlsx 
-- for both of the above, if you make any changes, they need to be re-saved as UTF16 txt file, and then converted (see remember above)
 - double check also stim_seg_freq -- why are some items still with NA freq?
@@
 # a2m fixed, but I think it is better to fix all of these error upstream, in the demo.xlsx file
-trnsc$Sex[trnsc$Sex=="F" & !is.na(trnsc$Sex)]<-"G"
+# M2A: Since it's fixed upstream I propose deleting it (and have therefore
+# commented it out again for now)
+# trnsc$Sex[trnsc$Sex=="F" & !is.na(trnsc$Sex)]<-"G"
 trnsc$Sex=factor(trnsc$Sex)
@@
 
-```{r Fig1-first_vs_others, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR scores for individual participants averaging separately their first attempts and all other attempts."}
+```{r Fig1-first_vs_others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="NWR scores for individual participants averaging separately their first attempts and all other attempts."}
 
@@
 ```{r overall}
-
+# M2A: This excludes a bunch of kids (anyone with NA on any of these participant features),
+# whittling the summary down to 24 from 42 -- is that intended?
+# Also all phon_score is 100 and all NLD is 0; does that stike you as fishy?
 aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->overall
@@
 
-```{r Fig2-xling_freq, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
+```{r Fig2-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
 aggregate(trnsc$correct*100,by=list(trnsc$target,trnsc$avg_fr),mean)->xling
 colnames(xling)<-c("Target","Avg_fr","Word")
-plot(xling$Word~xling$Avg_fr,lty=3,ylab="NWR scores (% words)",xlab="Average phone cross-ling frequency",type="n")
-text(xling$Word~xling$Avg_fr,labels=xling$Target,cex=.5)
-abline(lm(xling$Word~xling$Avg_fr))
+stims.ipa <- read_delim("stimuli2.txt", delim = "\t") %>%
+  select(target, target_ortho, phono)
+xling <- merge(xling, stims.ipa, by.x = "Target", by.y = "target")
+
+item.plot <- ggplot(data = xling, aes(x = Avg_fr, y = Word, label = phono)) +
+  geom_smooth(method = "lm", se = FALSE, color = "black") +
+  geom_text(family = "Doulos SIL") +
+  ylab("NWR scores (% words)") +
+  xlab("Average phone cross-ling frequency") +
+  theme_apa()
+ggsave("nwr.by.freq.ITEM.png", dpi = 300, device = png())
+
+knitr::include_graphics("nwr.by.freq.ITEM.png")
+
 ```
@@
 We next checked whether this may actually be due to frequency of the sounds within the language.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020png], and applied the natural logarithm.
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early], and applied the natural logarithm.
 As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
@@
 ```{r bylength}
+# M2A: this reduces the total number of kids represented from 42 to 40
 aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->by_length
@@
 
-```{r tablength}
+```{r tablength, results = "asis"}
 birthorder.tab=NULL

### 2020-09-16 manuscript.tex
bug fix

@@
 \item
-  do you think Figure 2 should be done with phonetic symbols? If yes, could you try to implement that?
-\item
   double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
@@
   double check my frequency entries in segments.xlsx using the search function in \url{https://phoible.org/parameters}; take a look at \url{http://phoible.github.io/conventions/} in case you see something about double articulation (I didn't find tp or lBj)
-\item
-  double check phonological implementation of stimuli in stimuli.xlsx
-\item
-  for both of the above, if you make any changes, they need to be re-saved as UTF16 txt file, and then converted (see remember above)
 \item
@@
 As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
-Surprisingly, NWR scores for subsequent repetitions (M = 39, SD = 26)
-were on average lower than first ones (M = 65, SD = 15), t(38) = 6.60, p = 0.
+Surprisingly, NWR scores for subsequent repetitions (M = 40, SD = 26)
+were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p = 0.
 Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
@@
 
-The overall NWR score was M = 61\% (SD = 15\%).
+The overall NWR score was M = 60\% (SD = 15\%).
 Scores based on phonemes are even higher M = 77\% (SD = 10\%).
@@
 We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.81, SE ß = 0.18, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Age was also a significant predictor (ß = 0.32, SE ß = 0.13, p = 0.01).
-
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/Fig2-xling_freq-1.pdf}
-\caption{(\#fig:Fig2-xling\_freq)NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}
+The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+Age was also a significant predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
+
+\begin{figure}[!t]
+
+{\centering \includegraphics[width=0.65\linewidth]{nwr.by.freq.ITEM}
+
+}
+
+\caption{NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}\label{fig:Fig2-xling-freq}
 \end{figure}
@@
 We next checked whether this may actually be due to frequency of the sounds within the language.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings ({\textbf{???}}), and applied the natural logarithm.
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020), and applied the natural logarithm.
 As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
@@
 As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic frequency (ß = 0.81, SE ß = 0.27, p = 0.00) and age (ß = 0.32, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0.00, SE ß = 0.25, p = 1.00) was not.
+In this model, both cross-linguistic frequency (ß = 0.82, SE ß = 0.27, p = 0.00) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0, SE ß = 0.25, p = 1.00) was not.
 
@@
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels, 31 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (112 oral vowels produced with a different quality; 13 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 117) than vice versa (2).
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels, 30 nasal target vowels produced as oral vowels).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (110 oral vowels produced with a different quality; 13 nasal vowels produced with a different quality).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 115) than vice versa (2).
 Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones, compared to 33 complex ones).
@@
 
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels or about 0.98\% out of all oral vowel targets, 31 nasal target vowels produced as oral vowels or about 2.37\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (112 oral vowels produced with a different quality or about 1.82\% out of all oral vowel targets; 13 nasal vowels produced with a different quality or about 2.04\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 117 times or about 1.95\% out of all complex consonant targets) than vice versa (2 times or about 1.95\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones or about 0.38\% out of all simple consonant targets, compared to 33 complex ones or about 0.99\% out of all complex consonant targets).
+As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels or about 0.99\% out of all oral vowel targets, 30 nasal target vowels produced as oral vowels or about 2.37\% out of all nasal vowel targets).
+Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (110 oral vowels produced with a different quality or about 1.86\% out of all oral vowel targets; 13 nasal vowels produced with a different quality or about 2.06\% out of all nasal vowel targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 115 times or about 1.96\% out of all complex consonant targets) than vice versa (2 times or about 1.96\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones or about 0.39\% out of all simple consonant targets, compared to 33 complex ones or about 1.02\% out of all complex consonant targets).
 
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,483.87) = 0.35, p = 0.09).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.13) = 21.37, p = 0.90), birth order (data missing for 17 children, 2,600.01) = -0.13, p = 0.54) or maternal education (data missing for 2 children, 2,045.68) = 0.11, p = 0.61)
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,421.71) = 0.38, p = 0.07).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.03) = 21.40, p = 0.98), birth order (data missing for 17 children, 2,652.32) = -0.15, p = 0.47) or maternal education (data missing for 2 children, 2,101.60) = 0.09, p = 0.69)
 

### 2020-09-16 wrangling.R
bug fix

@@ -25,37 +25,19 @@ trnsc[order(trnsc$file, trnsc$int),]->trnsc
 #add repetition number
 trnsc$rep=gsub(".*_","",gsub(".wav","",trnsc$outfile))
 
-# AC's prev code
- trnsc$previous<-c(NA,as.character(trnsc$target[1:(dim(trnsc)[1]-1)]))
- trnsc$attempt<-ifelse(trnsc$target!=trnsc$previous,"first","subsequent")
- trnsc$attempt[1]<-"first"
- table(trnsc$attempt) #1045 first attempts
 
- head(trnsc) #previous looks right
- 
-# MC's new code
-# M2A: by my count there are 1029 first attempts, but maybe I'm misunderstanding "token"
-# M2A: okay, my solution finds some conflicting cases with your outfile attempt namings,
-#      now going off of those for attempt first/subsequent, I still find 1029
-# A2m please look at lines 21-22, those are the lines that determine which items are first versus subsequent. 
-# int is the interval in praat, so it is a certain measure of time; I don't remember
-# what token is, but there is a file called notes_for_supmat.txt which contains the explanation
-# for all columns, which may say for this one and others
-# M2A: Okay, great! Well that still returns 1029 first cases, but now the tidyverse pipe below
-# should guarantee we're getting the cases we want
+first.prods <- trnsc %>%
+   group_by(id, item) %>%
+   summarize(min.int = min(int)) %>%
+   mutate(attempt = "first") %>%
+   rename("int" = min.int)
+trnsc <- trnsc %>%
+   left_join(first.prods, by = c("id", "item", "int")) %>%
+   replace_na(list(attempt = "subsequent"))
 
-# first.prods <- trnsc %>%
-#   group_by(id, item) %>%
-#   summarize(min.int = min(int)) %>%
-#   mutate(attempt = "first") %>%
-#   rename("int" = min.int)
-# trnsc <- trnsc %>%
-#   left_join(first.prods, by = c("id", "item", "int")) %>%
-#   replace_na(list(attempt = "subsequent"))
-#  table(trnsc$attempt.x,trnsc$attempt.y) #the code above has removed non-first
-# table(first.prods$attempt) # confirm 1029
-# head(first.prods) #too little information to check
 
+dim(trnsc)
+  
 # add demographic info
 read.csv("NWR-demo.csv",header=T)->demo
 #colnames(demo)

### 2020-09-16 manuscript.Rmd
replied to mc comments

@@
 
+- add information about proportion of mispronunciations that are real based on other papers
 - add effect sizes to all analyses
-- reply to questions
+- ? correct treatment of words whose mispronunciation is an English word:
+    mispronunciation meaning
+361              two     two
+362              two     two
+879          no-pinêt                                          English: no peanut
+908            nodedi                                           English: no daddy
+933          no--dadi                                           English: no daddy
+949           petrumi                                             English: "true"
+950           petrumi                                             English: "true"
+951           petrumi                                             English: "true"
+1373              wee     English: where
+1374              wee     English: where
+1375              wee     English: where
+1379              wee     English: where
+1380              wee     English: where
+1388              wee     English: where
+1394              wee     English: where
+1396              wee     English: where
+1398              wee     English: where
+1399              wee     English: where
+1402              wat      English: what
+1405              wat      English: what
+1411              wee     English: where
 
@@
 If that occurred, we tried with the second training item, at which point some children got it and could continue.
-A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.
+A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether (see Participants section for exclusions).
 <!--M2A: do you know how many these were? oh! it's below. then maybe we can just put a (see Participants below), etc.-->
+<!-- a2m done -->
 
@@
 We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
-<!--
-A2M: do you agree with this decision or shall we do the alternative?
-M2A: yeah, i think adding abirtrarily small frequencies to the NA cases is fraught for the reasons we talked about this morning (i.e., possible systematic undersampling in phoible that can't be helped).
--->
 
@@
 
+#fix sex
+demo$Sex[demo$Sex=="F"]<-"G"
+
 subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
@@
 table(inc$birthOrder)->birthorder.tab
+
 ```
@@
 
-#quick fix, but should be fixed upstream @ac
-# M2A: Seems like you already did this, and the line below
-# is setting the is.na(trnsc$Sex) cases to "G" -- I assume
-# that's not a desirable outcome so am commenting it out
-# a2m fixed, but I think it is better to fix all of these error upstream, in the demo.xlsx file
-# M2A: Since it's fixed upstream I propose deleting it (and have therefore
-# commented it out again for now)
-# trnsc$Sex[trnsc$Sex=="F" & !is.na(trnsc$Sex)]<-"G"
 trnsc$Sex=factor(trnsc$Sex)
@@
 # Also all phon_score is 100 and all NLD is 0; does that stike you as fishy?
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->overall
-aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->phon_score
-aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$Sex,trnsc$age.rounded,trnsc$birthOrder,trnsc$matEd),mean)->nld
+#a2m fixed and fixed!
+aggregate(trnsc$correct*100,by=list(trnsc$id),mean)->overall
+aggregate(trnsc$phon_score*100,by=list(trnsc$id),mean)->phon_score
+aggregate(trnsc$nld*100,by=list(trnsc$id),mean)->nld
 overall=cbind(overall,phon_score$x,nld$x)
-colnames(overall)<-c("ID","Sex","Age","BrthO","matEd","Word","Phoneme","NLD")
+overall=merge(overall, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
+colnames(overall)<-c("ID","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")
 
@@
 
-Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: `r round(sum(!is.na(trnsc$meaning))/sum(!is.na(trnsc$mispronunciation))*100)`%.
-
-**TO DISCUSS TOGETHER this is higher than any previous report! can we be sure of it?**
+Finally, we looked into the frequency with which mispronunciations resulted in real words.
+Two thirds of them were: `r round(sum(!is.na(trnsc$meaning) & trnsc$meaning != "0")/sum(!is.na(trnsc$mispronunciation))*100)`%.
 
@@
 # M2A: this reduces the total number of kids represented from 42 to 40
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->by_length
-aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->phon_score
-aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$type),mean)->nld
+#a2m fixed
+aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$type),mean)->by_length
+aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$type),mean)->phon_score
+aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$type),mean)->nld
 by_length=cbind(by_length,phon_score$x,nld$x)
-colnames(by_length)<-c("ID","Age","Length","Word","Phoneme","NLD")
+by_length=merge(by_length, demo[,c("ID","Sex","Age","birthOrder","matEd")], by.x="Group.1",by.y="ID", all.x=T)
+colnames(by_length)<-c("ID","Length","Word","Phoneme","NLD","Sex","Age","BrthO","matEd")
 
@@
 As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children  (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p = `r corage$p.value`).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p = `r tsex$p.value`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, `r corbo$statistic`) = `r corbo$estimate`, p = `r corbo$p.value`) or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children, `r corme$statistic`) = `r corme$estimate`, p = `r corme$p.value`)
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p = `r tsex$p.value`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r corbo$statistic`) = `r corbo$estimate`, p = `r corbo$p.value`) or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children,  rho = ( `r corme$statistic`) = `r corme$estimate`, p = `r corme$p.value`).
 

### 2020-09-16 manuscript.tex
replied to mc comments

@@
 \item
+  add information about proportion of mispronunciations that are real based on other papers
+\item
   add effect sizes to all analyses
 \item
-  reply to questions
+  ? correct treatment of words whose mispronunciation is an English word:
+  mispronunciation meaning
+  361 two two
+  362 two two
+  879 no-pinêt English: no peanut
+  908 nodedi English: no daddy
+  933 no--dadi English: no daddy
+  949 petrumi English: ``true''
+  950 petrumi English: ``true''
+  951 petrumi English: ``true''
+  1373 wee English: where
+  1374 wee English: where
+  1375 wee English: where
+  1379 wee English: where
+  1380 wee English: where
+  1388 wee English: where
+  1394 wee English: where
+  1396 wee English: where
+  1398 wee English: where
+  1399 wee English: where
+  1402 wat English: what
+  1405 wat English: what
+  1411 wee English: where
 \end{itemize}
@@
 If that occurred, we tried with the second training item, at which point some children got it and could continue.
-A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether.
+A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether (see Participants section for exclusions).
 
@@
 
-Taking into account only the first attempts, we averaged attempts by each of the 24 children who had data for first attempts; their ages ranged from 3.90 to 10.20 (M = 6.50, SD = 1.50).
-
-The overall NWR score was M = 60\% (SD = 15\%).
-Scores based on phonemes are even higher M = 77\% (SD = 10\%).
-The phoneme-based normalized Levenshtein distance was M = 23\% (SD = 10\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
+Taking into account only the first attempts, we averaged attempts by each of the 42 children who had data for first attempts; their ages ranged from NA to NA (M = NA, SD = NA).
+
+The overall NWR score was M = 65\% (SD = 15\%).
+Scores based on phonemes are even higher M = 78\% (SD = 9\%).
+The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
 Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
@@
 
-Finally, we looked into the frequency with which mispronunciations resulted in real words. Nearly all of them were: 97\%.
-
-\textbf{TO DISCUSS TOGETHER this is higher than any previous report! can we be sure of it?}
+Finally, we looked into the frequency with which mispronunciations resulted in real words. Two thirds of them were: 62\%.
 
@@
 47 (22) & 59 (17) & 41 (17)\\
-79 (22) & 92 (9) & 8 (9)\\
-78 (19) & 93 (7) & 7 (7)\\
+80 (22) & 93 (9) & 7 (9)\\
+79 (19) & 93 (7) & 7 (7)\\
 74 (32) & 91 (12) & 9 (12)\\
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (1,421.71) = 0.38, p = 0.07).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (0.03) = 21.40, p = 0.98), birth order (data missing for 17 children, 2,652.32) = -0.15, p = 0.47) or maternal education (data missing for 2 children, 2,101.60) = 0.09, p = 0.69)
+As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (6,496.39) = 0.43, p = 0.00).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (-0.61) = 27.83, p = 0.55), birth order (data missing for 17 children, rho = (2,652.32) = -0.15, p = 0.47) or maternal education (data missing for 2 children, rho = ( 9,022.76) = 0.09, p = 0.60).
 

### 2020-09-17 manuscript.Rmd
added effect sizes and info on semantic errors

@@
 
+cohensd=function(m,sd){round(m/sd,3)}
+cohensd2=function(m1,sd1,m2,sd2){round((m1-m2)/sqrt((sd1^2+sd2^2)/2),3)}
 ```
@@
 
-- add information about proportion of mispronunciations that are real based on other papers
-- add effect sizes to all analyses
 - ? correct treatment of words whose mispronunciation is an English word:
@@
 
+# a2m I get a warning:
+#   Warning message:
+# Vectorized input to `element_text()` is not officially supported.
+# Results may be unexpected or may change in future versions of ggplot2.
+
 myplot
@@
 
-A first list of candidate items was generated in 2018 by selecting simple consonants ("p",	"t",	"d",	"k",	"m",	"n",	"w",	"y") and vowels ("i", "o","u", "a","e") that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
+A first list of candidate items was generated in 2018 by selecting simple consonants ("p",	"t",	"d",	"k",	"m",	"ń",	"w",	"y") 
+and vowels ("i", "o","u", "a","e") that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
 These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
@@
 
-```{r Fig1-first_vs_others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="NWR scores for individual participants averaging separately their first attempts and all other attempts."}
+```{r Fig1-first_vs_others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts."}
 
@@
 
-plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="NWR scores (%)")
+plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="NWR scores (% words)")
 for(id in levels(factor(pwcm_att$ID))) lines(myx[pwcm_att$ID==id],pwcm_att$Word[pwcm_att$ID==id],lty=2)
@@
 t.test(pwcm_att$Word~pwcm_att$Attempt,paired=T)->myt
+
 ```
 
-We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
+We first checked whether whole-item NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
 We excluded `r excl` children who did not have data for one of these two types.
 As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
-Surprisingly, NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
-were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`.
+Surprisingly, whole-item NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
+were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`.
 Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
@@
 
-The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%).
-Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%).
+The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`.
+Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%), Cohen's d = `r cohensd(mean(overall$Phoneme),sd(overall$Phoneme))`.
 The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
@@
 
-In this analysis, we were interested in variation in NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+In this analysis, we were interested in variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
 To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
@@
 #plot(xl_fr)
+
+aggregate(trnsc$correct*100,by=list(trnsc$avg_fr_s),mean)->xl_fr_tab
+colnames(xl_fr_tab)<-c("xlf","Word")
+xlfr_r=cor.test(xl_fr_tab$xlf,xl_fr_tab$Word)
 ```
 
-We could include 885 observations (**why -- shouldn't it be 1k??**), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
+We could include `r summary(xl_fr)$devcomp$dims["N"]` observations, from `r summary(xl_fr)$ngrps["file"]` children producing in any given trial one of `r summary(xl_fr)$ngrps["target"]` potential target words.
+The analysis revealed a main effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
+Error"]`, p = `r mycoeff["age.rounded_s","Pr(>|z|)"]`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
 Error"]`, p = `r mycoeff["avg_fr_s","Pr(>|z|)"]`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Age was also a significant predictor  (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
-Error"]`, p = `r mycoeff["age.rounded_s","Pr(>|z|)"]`).
+Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r r(`r length_r$parameter`) = `r round(xlfr_r$estimate,3)`.
 
@@
 
+aggregate(trnsc$correct*100,by=list(trnsc$avg_fr_cor_ty_s),mean)->cor_fr_tab
+colnames(cor_fr_tab)<-c("corf","Word")
+
 read.delim("segments_with_cor_freq.txt", encoding="UTF-8")->segments
@@
 
-We next checked whether this may actually be due to frequency of the sounds within the language.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early], and applied the natural logarithm.
-As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, =`r round(freq_cor$p.value,3)`).
-Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence is actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^2] As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, =`r round(freq_cor$p.value,3)`).
+Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = `r round(cor.test(cor_fr_tab$corf,cor_fr_tab$Word)$estimate,3)`.
+Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
 As before, the model contained random slopes for both child ID and target.
@@
 Error"]`, p = `r both_coeff["avg_fr_s","Pr(>|z|)"]`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std.
-Error"]`, p = `r both_coeff["age.rounded_s","Pr(>|z|)"]`) were significant predictors of word-level NWR scores, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
+Error"]`, p = `r both_coeff["age.rounded_s","Pr(>|z|)"]`) were significant predictors of whole-item  NWR scores, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
 Error"]`, p = `r both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"]`) was not.
 
-<!-- a2m should we point out that the coefficient for cross-ling frequency is reduced from .77 to .67, so indirectly corpus frequency does affect a little bit?
+[^2]: We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less convincing.
+
+<!-- a2m should we point out that the coefficient for cross-ling frequency is reduced, so indirectly corpus frequency does explain a little bit of variance?
 Or making things too complicated, and just say "so corpus frequency does not explain away these results" -->
+
+### Qualitative analyses of NWR mispronunciations
+
+<!-- a2m I realized this section did not belong together with the xling freq section above - are you ok with this title? -->
 
@@
 oralVs=c("a", "â","e",  "é" ,"ê", "i","î" ,"o","ó","u")
-simplexCs=c("m","n","l","w","y","v","d","g","p","t","k","f","h","ch")
+simplexCs=c("m","ń","l","w","y","v","d","g","p","t","k","f","h","ch")
 complexCs=c("dp","tp","kp","km","kn","mb","gh","lv")
@@
 #sub_pc has the same shape as the original sub_tab, namely targets as rows, realization as columns
-sub_pc=round(sub_tab/c(denom_vec)*100,2)
+sub_pc=round(sub_tab/c(denom_vec)*100,3)
 
-del_pc=round(del_tab/denom_vec*100,2)
-mat_pc=round(match_tab/denom_vec*100,2)
+del_pc=round(del_tab/denom_vec*100,3)
+mat_pc=round(match_tab/denom_vec*100,3)
 
@@
 
-**use proportion of correctly produced rather than number; DONE**
+<!-- a2m I think it's weird to put effect sizes here, right? technically I could do something like cohen's d based on mean and sd of the percentage of tokens... -->
 
-**a2m use first attempts rather than the whole data set: having second thoughts.
+<!-- **a2m regarding the use first attempts rather than the whole data set, I'm having second thoughts.
 To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers & it is more consistent.
 BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros...
-CF Option 3 further down**
+CF Option 3 further down** -->
 
-**a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?**
+<!-- **a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?** -->
 
@@
 Two thirds of them were: `r round(sum(!is.na(trnsc$meaning) & trnsc$meaning != "0")/sum(!is.na(trnsc$mispronunciation))*100)`%.
+This is a type of analysis that is seldom reported, and we could only find one comparison point: @castro1998 found that illiterate adults NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate adults did so in only `r round(2/117*100,2)`%.
+The percentage we observe here is much higher than this, but we do not know whether age, language, or even test structure explains this difference.
+
+<!-- a2m I added info above - is it weird to talk about this here? -->
 
@@
 
+aggregate(trnsc$correct*100,by=list(trnsc$type),mean)->by_length_no_child
+colnames(by_length_no_child)<-c("Length","Word")
+by_length_no_child=by_length_no_child[-1,]
+length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)
 ```
@@
 The pattern is less marked when other two scores are used, which are less sensitive to errors.
+Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(`r length_r$parameter`) = `r round(length_r$estimate,3)`.
+<!-- a2m it doesn't make sense to do this, right, with only 3 items? -->
 
 ```{r tablength, results = "asis"}
-birthorder.tab=NULL
+length.tab=NULL
 for(myvar in c("Word","Phoneme","NLD")){
-  birthorder.tab=cbind(birthorder.tab,
+  length.tab=cbind(length.tab,
               paste0(round(aggregate(by_length[,myvar],by=list(by_length$Length),mean)$x),
@@
 }
-rownames(birthorder.tab)<-levels(by_length$Length)
-colnames(birthorder.tab)<-c("Word","Phoneme","NLD")
-apa_table(birthorder.tab,caption="NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
+rownames(length.tab)<-levels(by_length$Length)
+colnames(length.tab)<-c("Word","Phoneme","NLD")
+apa_table(length.tab,caption="NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
 ```

### 2020-09-17 manuscript.Rmd
add mic info; update to stimuli3.txt

--- 2020-09-16 manuscript.Rmd
@@
 
-- look through this manuscript for bits of text to be written by **mc**
 - double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
@@
 These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset **XXX **mc** please fill in** and an Olympus **XXX **mc** please fill in** from the written form presented together with the same item orally (by the second author).
+Together with the 2018 selection, they were recorded using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) from the written form presented together with the same item orally by the second author.
 The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
@@
 #add orthographic representations targets & other stims chars
-stims.tbl <- read_delim("stimuli2.txt", delim = "\t") %>%
+stims.tbl <- read_delim("stimuli3.txt", delim = "\t") %>%
   arrange(factor(type, levels = c("practice", "1", "2", "3", "4"))) %>%
@@
 
-The child was donned a headset (**xx **mc** please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
+The child was donned a headset (a Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former), recorded into the left channel into a Tascam DR40x digital recorder.
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
@@
 colnames(xling)<-c("Target","Avg_fr","Word")
-stims.ipa <- read_delim("stimuli2.txt", delim = "\t") %>%
+stims.ipa <- read_delim("stimuli3.txt", delim = "\t") %>%
   select(target, target_ortho, phono)

### 2020-09-17 manuscript.tex
add mic info; update to stimuli3.txt

@@
 \item
-  look through this manuscript for bits of text to be written by \textbf{mc}
-\item
-  double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
-I looked again at my local files (still haven't checked the labs' ones) and I'm surprised I have maternal education for more children than I have birth order for\ldots{} If you correct this, to generate the csv I also round the age, assign age.rounded to reported age (in years) when DOB is missing, remove first names, DOB, DOTest and age.precise before pushing to github
-\item
-  double check my frequency entries in segments.xlsx using the search function in \url{https://phoible.org/parameters}; take a look at \url{http://phoible.github.io/conventions/} in case you see something about double articulation (I didn't find tp or lBj)
-\item
-  double check also stim\_seg\_freq -- why are some items still with NA freq?
+  double check demo -- probably start with the xlsx version that has the
+  children's first names, date of birth, and date of test. I looked
+  again at my local files (still haven't checked the labs' ones) and I'm
+  surprised I have maternal education for more children than I have
+  birth order for\ldots{} If you correct this, to generate the csv I
+  also round the age, assign age.rounded to reported age (in years) when
+  DOB is missing, remove first names, DOB, DOTest and age.precise before
+  pushing to github
+\item
+  double check my frequency entries in segments.xlsx using the search
+  function in \url{https://phoible.org/parameters}; take a look at
+  \url{http://phoible.github.io/conventions/} in case you see something
+  about double articulation (I didn't find tp or lBj)
+\item
+  double check also stim\_seg\_freq -- why are some items still with NA
+  freq?
 \item
@@
 \item
-  add information about proportion of mispronunciations that are real based on other papers
+  add information about proportion of mispronunciations that are real
+  based on other papers
 \item
@@
 \item
-  ? correct treatment of words whose mispronunciation is an English word:
-  mispronunciation meaning
-  361 two two
-  362 two two
-  879 no-pinêt English: no peanut
-  908 nodedi English: no daddy
-  933 no--dadi English: no daddy
-  949 petrumi English: ``true''
-  950 petrumi English: ``true''
-  951 petrumi English: ``true''
-  1373 wee English: where
-  1374 wee English: where
-  1375 wee English: where
-  1379 wee English: where
-  1380 wee English: where
-  1388 wee English: where
-  1394 wee English: where
-  1396 wee English: where
-  1398 wee English: where
-  1399 wee English: where
-  1402 wat English: what
-  1405 wat English: what
-  1411 wee English: where
+  ? correct treatment of words whose mispronunciation is an English
+  word: mispronunciation meaning 361 two two 362 two two 879 no-pinêt
+  English: no peanut 908 nodedi English: no daddy 933 no--dadi English:
+  no daddy 949 petrumi English: ``true'' 950 petrumi English: ``true''
+  951 petrumi English: ``true'' 1373 wee English: where 1374 wee
+  English: where 1375 wee English: where 1379 wee English: where 1380
+  wee English: where 1388 wee English: where 1394 wee English: where
+  1396 wee English: where 1398 wee English: where 1399 wee English:
+  where 1402 wat English: what 1405 wat English: what 1411 wee English:
+  where
 \end{itemize}
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
-
-Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood (e.g., Hazan \& Barrett, 2000).
-One common task in this line of research is nonword repetition (NWR).
-In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning.
-The child should immediately try to say it back without changing anything.
-NWR scores are thought to reflect long-term phonological knowledge (which allows the child to perceive the item precisely even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely even though it had never been pronounced before).
-
-NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and to assess the extent to which apparent phonological constraints found in the lexicon affect online production (Gallagher, 2014).
-It is also frequently used for applications, notably as a diagnostic for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-However, it has been seldom used outside of Europe and North America (with exceptions including Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)).
-
-In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
-The present report uses these unique data to contribute to several lines of research.
-First, we made sure that some of the non-word items contained typologically rare and/or challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
-Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
-Third, we contribute to the basic NWR literature by adding a data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
-
-\hypertarget{yuxe9luxee-dnye-phonological-overview}{%
-\subsubsection{Yélî Dnye phonological overview}\label{yuxe9luxee-dnye-phonological-overview}}
-
-Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
-The Yélî sound system, much like its baroque grammatical system (Levinson, n.d.), is unlike any other in the region.
-
-With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these able to be contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized or palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (n.d.) for details).
-The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
-Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
-
-In total, Yélî Dnye uses 90 distinctive segments (93 when including consonants that are extremely rarely used), far outstripping the phonemic inventory size of other documented Papuan languages (Levinson (n.d.); Foley 1986) and includes at least two contrasts not yet been documented elsewhere {[}labial-coronal double-articulations with dental vs.~post-alveolar coronal placement in both oral and nasal stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.
-
-Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the current Yélî Dnye dictionary, Levinson, n.d.).
-The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/).
-This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /`pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
-Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (n.d.) for details).
-
-While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
-In monosyllabic stems, oral vowels (short: 43\% and long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%), with the first syllable of bisyllabic stems following a similar pattern.
-Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels.
-Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see Levinson (n.d.) for details).
-
-\hypertarget{the-yuxe9luxee-community}{%
-\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}}
-
-The current study also contributes to increasing the cultural diversity of the populations for which NWR data have been collected.
-As mentioned previously, this method has been used almost exclusively in urban, industrialized populations, with only one exception to our knowledge (Cristia et al., 2020).
-We therefore provide some background information tol help understand some aspects of our study, including the estimation of children's ages and the relationship between NWR scores and children's demographic characteristics.
-
-Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7;
-school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---, where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
-Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
-
-Most Yélî people are swidden horticulturalists, raising a variety of tubers (e.g., yams, manioc, taro), coconut, banana, and more in their gardens.
-In addition to garden foods, Yélî children grow up eating fish and foraging for shellfish and nuts, which provide a regular source of extra protein.
-The typical household in our dataset includes seven individuals and is situated among a collection of 4 or more other households, with structures often arranged around an open grassy area.
-These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
-
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by barrier reef, transport to and from the island is both infrequent and irregular.
-There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport, when there is deisel available.
-There is a medical radio connection at the health clinic (at the Catholic mission) via which news and messages are often passed along.
-There is also a cellular tower on the island, though it is typically not functioning.
-As such, access to news and outside resources, particularly international connections and larger data transfers, is typically unavailable.
-Our data collection is therefore limited to the duration of the researchers' on-island visits.
-
-Despite restricted outside contact, formal education is a priority for many Yélî families.
-A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65\% of mothers and 49\% of fathers completing the full range of locally available education (\textasciitilde{}9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG (Casillas, Brown, \& Levinson, 2020).
-While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), and so it is very common for households to billet school children during the weekdays for months at a time (or for whole months if the child's home is very far).
-Combined with practices of collective child guardianship within their close-knit hamlets, the practical outcome for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's parents---to this child assent is also culturally pertinent, as child independence is encouraged and respected from toddlerhood onward (Brown \& Casillas, n.d.).\footnote{Note that in a few cases this does mean that precise details such as child date of birth, parent age in years, and parent education was only able to be approximated as the parents were residing on another part of the island.} Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
-
-While work on Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas et al., 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun {[}e.g., Bram's thesis; MC's CFH data{]}.
-What we do know about the early language environment is that, while child-directed speech is relatively infrequent throughout the day (Casillas et al., 2020), shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children (Bunce et al., n.d.; Casillas et al., 2020).
-While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere (Bergelson et al., 2019; Bunce et al., n.d.), there is a significant and marked increase in child-directed speech from other children as infants get older (Bunce et al., n.d.; Casillas et al., 2020).
-This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 people) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
-
-\hypertarget{brief-review-of-nwr-for-our-purposes}{%
-\subsubsection{Brief review of NWR for our purposes}\label{brief-review-of-nwr-for-our-purposes}}
-
-The basic structure of NWR involves the participant hearing a non-word, like ``bilik'', which they need to repeat back without changing anything that is contrastive in the language.
-For instance, in English, a response of ``bilig'' or ``pilik'' will be scored as incorrect; a response ``biilik'', where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.
-
-Aside from this central basic structure, there is some variation in the presentation procedure and structure of items found in previous NWR work.
-For example, items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation (Brandeker \& Thordardottir, 2015).
-Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
-Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
-
-Previous work seems to avoid difficult sounds, but we felt this was important to represent Yélî Dnye, so we also varied this factor.
-We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
-This allowed us to get NWR scores for more items.
-
-Naturally, designing the task in this way may render the study of individual variation within the population more difficult because different children are exposed to different items.
-However, a review of previous work on individual variation suggested to us that effects of individual differences are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
-
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
-Based on previous work, we looked at potential improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and potential negative effects of bilingual exposure (Brandeker \& Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
-Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
-Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on those too.
-
-\hypertarget{research-questions}{%
-\subsubsection{Research questions}\label{research-questions}}
-
-After some preliminary analyses to set the stage, we address the following questions:
+\subsection{Introduction}\label{introduction}
+
+Although infants begin to learn about their native language's phonology
+within the first year, many studies suggest that in perception and
+production, in phonetics and phonology, their knowledge continues to
+develop throughout childhood (e.g., Hazan \& Barrett, 2000). One common
+task in this line of research is nonword repetition (NWR). In NWR
+studies, participants are presented auditorily with an item that is
+phonologically legal but lexically meaningless in the language children
+are learning. The child should immediately try to say it back without
+changing anything. NWR scores are thought to reflect long-term
+phonological knowledge (which allows the child to perceive the item
+precisely even though it is not a real word they have encountered
+before) as well as online phonological working memory (to encode the
+item in the interval between hearing it and saying it back) and flexible
+production patterns (to produce the item precisely even though it had
+never been pronounced before).
+
+NWR has been used to seek answers to a variety of theoretical questions,
+including what the links between phonology, working memory, and the
+lexicon are (Bowey, 2001), and to assess the extent to which apparent
+phonological constraints found in the lexicon affect online production
+(Gallagher, 2014). It is also frequently used for applications, notably
+as a diagnostic for language delays and disorders (Estes, Evans, \&
+Else-Quest, 2007). Since non-words can be generated in any language, it
+has attracted the attention of researchers working in multilingual and
+linguistically diverse environments, particularly in Europe (Action,
+2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has been seldom
+used outside of Europe and North America (with exceptions including
+Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz
+(2020)).
+
+In this study, we report on NWR results among children learning Yêly
+Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense
+phonological inventory. The present report uses these unique data to
+contribute to several lines of research. First, we made sure that some
+of the non-word items contained typologically rare and/or challenging
+sounds, so that we can contribute to the study of whether rare sounds
+are disadvantaged in perception and/or production, both in terms of
+overall repetition scores and patterns of mispronunciation. Second, we
+look at the effects of item length, since previous NWR research has
+uncovered variable effects. Third, we contribute to the basic NWR
+literature by adding a data point for children learning a rarely-studied
+language and culture, and further investigating whether there are
+structured sources that account for individual variation.
+
+\subsubsection{Yélî Dnye phonological
+overview}\label{yuxe9luxee-dnye-phonological-overview}
+
+Yélî Dnye is an isolate language (presumed Papuan) spoken by
+approximately 7,000 people residing on Rossel Island, which is found at
+the far end of the Louisiade Archipelago in Milne Bay Province, Papua
+New Guinea. The Yélî sound system, much like its baroque grammatical
+system (Levinson, in preparation), is unlike any other in the region.
+
+With only four primary places of articulation (bilabial, alveolar,
+post-alveolar, and velar) and no voicing contrasts, the phonological
+inventory is remarkably packed with acoustically similar segments. The
+core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and
+doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal
+equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
+substantial portion of these able to be contrastively pre-nasalized or
+nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/,
+/kŋ/, /ṭpṇm/, /kpŋm/). A large number of this combinatorial set can
+further be contrastively labialized or palatalized on release, or both
+(e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (in preparation)
+for details). The consonantal inventory also includes a number of
+non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in
+Yélî Dnye may be oral or nasal, short or long. The 10 oral vowel
+qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/,
+/o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels,
+with 7 of these able to appear as short and long nasal vowels as well
+(/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
+
+In total, Yélî Dnye uses 90 distinctive segments (93 when including
+consonants that are extremely rarely used), far outstripping the
+phonemic inventory size of other documented Papuan languages (Levinson
+(in preparation); Foley 1986) and includes at least two contrasts not
+yet been documented elsewhere {[}labial-coronal double-articulations
+with dental vs.~post-alveolar coronal placement in both oral and nasal
+stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.
+
+Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
+monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
+that, and with tri-and-above syllabic words appearing least frequently
+(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the
+current Yélî Dnye dictionary, Levinson, in preparation). The vast
+majority of syllables use a CV format. A small portion of the lexicon
+features words with a final closed syllable, limited to codas of -/m/,
+-/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
+spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
+coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
+process is used frequently with English loan words that have a coda
+(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
+features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
+these are limited to /æ/-. Finally, the lexicon features a handful of
+single-vowel grammatical morphemes (see Levinson (in preparation) for
+details).
+
+While all consonants can appear in the first syllable of the word, the
+second syllable onset consonant in bisyllabic roots is almost
+exclusively drawn from a subset of the simplex consonants (in order of
+descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/,
+/t/, /ɣ/, /mʷ/). In monosyllabic stems, oral vowels (short: 43\% and
+long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%),
+with the first syllable of bisyllabic stems following a similar pattern.
+Most vowels in the second syllable of bisyllabic stems are also short
+and oral, with a potential preference for rounding harmony between the
+first and second vowels. Stress typically falls on the first syllable of
+bisyllabic words, on the first and third syllables of tetrasyllabic
+words, and on the first or second syllable of trisyllabic words,
+depending on its vocalic features (see Levinson (in preparation) for
+details).
+
+\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}
+
+The current study also contributes to increasing the cultural diversity
+of the populations for which NWR data have been collected. As mentioned
+previously, this method has been used almost exclusively in urban,
+industrialized populations, with only one exception to our knowledge
+(Cristia et al., 2020). We therefore provide some background information
+tol help understand some aspects of our study, including the estimation
+of children's ages and the relationship between NWR scores and
+children's demographic characteristics.
+
+Most speakers of the language grow up speaking it monolingually until
+they begin attending school around the age of 7; school instruction is
+in English. While monolingual Yélî Dnye upbringing is common,
+multilingual families are not unusual, particularly in the region around
+the Catholic Mission---the same region in which the current data were
+collected---, where there is a higher incidence of married-in mothers
+from other islands (Brown \& Casillas, in press). Children in these
+multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
+and/or some other language(s) from the region.
+
+Most Yélî people are swidden horticulturalists, raising a variety of
+tubers (e.g., yams, manioc, taro), coconut, banana, and more in their
+gardens. In addition to garden foods, Yélî children grow up eating fish
+and foraging for shellfish and nuts, which provide a regular source of
+extra protein. The typical household in our dataset includes seven
+individuals and is situated among a collection of 4 or more other
+households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that
+they typically comprise a set of brothers, their wives and children, and
+their mother and father, with neighboring hamlets also typically related
+through the patriline.
+
+Laying 250 nautical miles off the coast of mainland PNG and surrounded
+by barrier reef, transport to and from the island is both infrequent and
+irregular. There is no electricity system (we use solar panels) or motor
+vehicles on the island, save dinghys with outboard motors that are
+primarily reserved for medical transport, when there is deisel
+available. There is a medical radio connection at the health clinic (at
+the Catholic mission) via which news and messages are often passed
+along. There is also a cellular tower on the island, though it is
+typically not functioning. As such, access to news and outside
+resources, particularly international connections and larger data
+transfers, is typically unavailable. Our data collection is therefore
+limited to the duration of the researchers' on-island visits.
+
+Despite restricted outside contact, formal education is a priority for
+many Yélî families. A recent study surveying more than 40 parents of
+young children in this region of the island found that nearly all
+parents completed 6 or more years of education, with 65\% of mothers and
+49\% of fathers completing the full range of locally available education
+(\textasciitilde{}9 years), with around half of those parents having
+pursued an additional 2+ years of education on other islands in the
+region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While
+there are a handful of local schools around the island, these may be
+well out of walking distance for many children (i.e., more than 1 hour
+on foot or by canoe each day), and so it is very common for households
+to billet school children during the weekdays for months at a time (or
+for whole months if the child's home is very far). Combined with
+practices of collective child guardianship within their close-knit
+hamlets, the practical outcome for researchers is that adult consent can
+often come from a combination of aunts, uncles, adult cousins, and
+grandparents standing in for the child's parents---to this child assent
+is also culturally pertinent, as child independence is encouraged and
+respected from toddlerhood onward (Brown \& Casillas, in
+press).\footnote{Note that in a few cases this does mean that precise
+  details such as child date of birth, parent age in years, and parent
+  education was only able to be approximated as the parents were
+  residing on another part of the island.} Finally, as mentioned, the
+language of schooling in this region is English, and so the majority of
+teens and adults on the island speak some English---many speak it very
+fluently.
+
+While work on Yélî language development is growing (e.g., Brown, 2011,
+2014; Brown \& Casillas, in press; Casillas et al., 2020; Liszkowski,
+Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's
+linguistic development is quite limited, and research into their
+phonological development in particular has only just begun {[}e.g.,
+Bram's thesis; MC's CFH data{]}. What we do know about the early
+language environment is that, while child-directed speech is relatively
+infrequent throughout the day (Casillas et al., 2020), shared caregiving
+practices and a near-universal fondness for social interaction with
+young children ensure that speech directed to children comes from all
+types of speakers: women, men, and other children (Bunce et al., under
+review; Casillas et al., 2020). While speech from women predominates in
+young children's speech environments on Rossel, as it does elsewhere
+(Bergelson et al., 2019; Bunce et al., under review), there is a
+significant and marked increase in child-directed speech from other
+children as infants get older (Bunce et al., under review; Casillas et
+al., 2020). This increase in child-directed speech from other children
+is attributed to the fact that, starting around age 2, children often
+spend large swaths of the day playing, swimming, and foraging for fruit,
+nuts, and shellfish in large (\textasciitilde{}10 people) independent
+and mixed-age child play groups (Brown \& Casillas, in press; Casillas
+et al., 2020).
+
+\subsubsection{Brief review of NWR for our
+purposes}\label{brief-review-of-nwr-for-our-purposes}
+
+The basic structure of NWR involves the participant hearing a non-word,
+like ``bilik'', which they need to repeat back without changing anything
+that is contrastive in the language. For instance, in English, a
+response of ``bilig'' or ``pilik'' will be scored as incorrect; a
+response ``biilik'', where the vowel is lengthened without change of
+quality would be scored as correct, because English does not have a pure
+length contrast.
+
+Aside from this central basic structure, there is some variation in the
+presentation procedure and structure of items found in previous NWR
+work. For example, items are often presented orally by the experimenter
+(Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an
+increasing number of studies have turned to playing back the stimuli in
+order to have greater control of the stability of the presentation
+(Brandeker \& Thordardottir, 2015). Additionally, while some studies
+have used 10-15 non-words, others have employed up to 46 unique items
+(Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019). Often,
+authors modulate structural complexity, typically measured in terms of
+item length (measured in number of syllables) and/or syllable structure
+(open as opposed to closed syllables, Gallon, Harris, \& Van der Lely,
+2007).
+
+Previous work seems to avoid difficult sounds, but we felt this was
+important to represent Yélî Dnye, so we also varied this factor. We
+designed a relatively large number of items but, aware that this may
+render the task longer and more tiresome, we split some of the items
+across children. This allowed us to get NWR scores for more items.
+
+Naturally, designing the task in this way may render the study of
+individual variation within the population more difficult because
+different children are exposed to different items. However, a review of
+previous work on individual variation suggested to us that effects of
+individual differences are relatively small, and would not be detectable
+with the sample size that we could collect in a given visit.
+
+That said, we contribute to the literature by also reporting descriptive
+analyses of individual variation that could potentially be integrated in
+meta-analytic efforts. Based on previous work, we looked at potential
+improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid,
+Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and
+potential negative effects of bilingual exposure (Brandeker \&
+Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
+Previous work typically finds no significant differences as a function
+of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014;
+Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although past research has not often investigated potential effects of
+birth order on NWR, there is a sizable literature on these effects in
+other language tasks (Havron et al., 2019), and therefore we report on
+those too.
+
+\subsubsection{Research questions}\label{research-questions}
+
+After some preliminary analyses to set the stage, we address the
+following questions:
 
@@
 \includegraphics{manuscript_files/figure-latex/prevlit-fig-1.pdf}
-\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years) and item length for comparable studies.
-Arabic data from Jaber et al.
-(2018); Italian data from Piazzalunga et al.
-(2019); Tsimane' from Cristia et al.
-(2020); Yélî data from the present study.}
+\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years)
+and item length for comparable studies. Arabic data from Jaber et al.
+(2018); Italian data from Piazzalunga et al. (2019); Tsimane' from
+Cristia et al. (2020); Yélî data from the present study.}
 \end{figure}
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
-
-\hypertarget{stimuli}{%
-\subsubsection{Stimuli}\label{stimuli}}
-
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
-We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in Yély Dnye.
-We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat (note that up to other work has also used 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
-
-A first list of candidate items was generated in 2018 by selecting simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'', ``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
-These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
-The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
-Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
-
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî inventory to produce consonant-vowel monosyllables.
-As before, items were automatically excluded if they appeared in the dictionary.
-Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described (Levinson, n.d.), there could have been undocumented constraints that rendered items illegal.
-Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
-These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset \textbf{XXX }mc** please fill in** and an Olympus \textbf{XXX }mc** please fill in** from the written form presented together with the same item orally (by the second author).
-The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
-Even so, there was one monosyllable that was often identified as a real word (intended ``yî'' /yXX/; identified as ``yi'' /yi/, \emph{tree}).
-This item is removed from analyses below.
-
-The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
+\subsection{Methods}\label{methods}
+
+\subsubsection{Stimuli}\label{stimuli}
+
+Many NWR studies are based on a fixed list of 12-16 items that vary in
+length between 1 and 4 syllables, often additionally varying syllable
+complexity and/or cluster presence and complexity, always meeting the
+condition that they do not mean anything in the target language (e.g.,
+Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We kept the
+same variation in item length and the non-meaningfulness requirement,
+but we did not vary syllable complexity and clusters because these are
+vanishingly rare in Yély Dnye. We also increased the number of items an
+individual child would be tested on, so that a child would get up to 23
+items to repeat (note that up to other work has also used 24-30 items:
+Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and
+distributed them across children, so as to increase the coverage, and be
+able to study more items.
+
+A first list of candidate items was generated in 2018 by selecting
+simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'',
+``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined
+into consonant-vowel syllables, further sampling the space of 1- to
+4-syllable sequences. These candidates were automatically checked
+against Levinson's 2015 dictionary and removed from consideration if
+they appeared in the dictionary. The second author presented them orally
+to three local research assistants, who were asked to repeat them and
+further say whether they were real words. Any item for which two or more
+of the assistants reported them having a meaning or some form of
+association was excluded.
+
+A second list of candidate items was generated in 2019 by selecting
+complex consonants and systematically crossing them with all the vowels
+in the Yélî inventory to produce consonant-vowel monosyllables. As
+before, items were automatically excluded if they appeared in the
+dictionary. Additionally, since hearing vowel length in monosyllables in
+isolation is challenging, any item that had a short/long real word
+neighbor was filtered out. Since the phonology and phonetics of Yélî is
+still in the process of being described (Levinson, in preparation),
+there could have been undocumented constraints that rendered items
+illegal. Therefore, we made sure that the precise consonant-vowel
+sequence occurred in some real word in the dictionary (i.e., that there
+was a longer word included the monosyllable as a subsequence). These
+candidates were presented to one informant, for a final check that they
+did not mean anything. Together with the 2018 selection, they were
+recorded using a Shure SM10A XLR dynamic headband microphone and an
+Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter)
+from the written form presented together with the same item orally by
+the second author. The complete recorded list was finally presented to
+two more informants, who could repeat all the items and who confirmed
+there were no real words. Even so, there was one monosyllable that was
+often identified as a real word (intended ``yî'' /yXX/; identified as
+``yi'' /yi/, \emph{tree}). This item is removed from analyses below.
+
+The final list is composed of three practice items; 20 monoysllables
+containing sounds that are less frequent in the world's languages than
+singleton plosives; 8 bisyllables; 12 trisyllables; and 4
+quadrisyllables (see Table 2).
 
@@
  &  & lvi & lβʲi &  &  &  &  &  & \\
+ &  & t:êê & t̃əː &  &  &  &  &  & \\
+\addlinespace
  &  & tp:a & tp̃æ &  &  &  &  &  & \\
-\addlinespace
  &  & tpâ & tpɑ &  &  &  &  &  & \\
@@
 \item
-  the same three items were selected as practice items and used in all 40 elicitation sets
-\item
-  splits were done within each length group from the 2018 items (i.e., separately for 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
-Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
-\item
-  once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllables, then 3-syllables, and finally 4-syllable items.
+  the same three items were selected as practice items and used in all
+  40 elicitation sets
+\item
+  splits were done within each length group from the 2018 items (i.e.,
+  separately for 2, 3, and 4-syllable items); and among onset groups for
+  the difficult monosyllables generated in 2019 (i.e., all the
+  monosyllables starting with tp were split into 2 sublists). Since some
+  of these groups had an odd number of items, one of the sublists was
+  slightly longer than the other (20 versus 23).
+\item
+  once the sublist split had been done, items were randomized such that
+  all children heard first the 3 practice items in a fixed order (1, 2,
+  and 4 syllables), a randomized version of their sublist selection of
+  difficult onset items, and randomized versions of their 2-syllables,
+  then 3-syllables, and finally 4-syllable items.
 \end{itemize}
 
-\hypertarget{procedure}{%
-\subsubsection{Procedure}\label{procedure}}
-
-We tried to balance three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them; that children would feel comfortable doing this task with us; and that the community would feel safe with us doing this task with their children.
-Moreover, there were also some logistic constraints in terms of the space availability.
-As a result, the places where elicitation happened varied across the hamlets.
-
-We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items `spreading' through hearsay.
-In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
-The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
-
-The child was donned a headset (\textbf{xx }mc** please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
-For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headphones.
-
-The first phase was making sure the child understood the task.
-This was explained orally and the first training item was presented.
-Often, children froze and did not say anything.
-If this happened, then we followed this procedure.
-First, the informant insisted.
-If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another.
-If the child could repeat these correctly, then we provided the recorded training item over headphones again.
-Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it).
-A small minority still failed to repeat the item after hearing it over headphones.
-If that occurred, we tried with the second training item, at which point some children got it and could continue.
-A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether (see Participants section for exclusions).
-
-NWR studies vary in whether children are provided with several opportunities to hear and say the item.
-To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them.
-If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a separate headset that was recording onto the right track of the same Tascam).
-If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.
-
-Whenever siblings from the same family were tested, an attempt was made to test first the older and then the younger child, and always on different elicitaiton sets.
-
-\hypertarget{coding}{%
-\subsubsection{Coding}\label{coding}}
-
-A script was written to randomize all tokens from all children, pairing each with the auditory target the child had been provided with.
-A native research assistant then listened through all productions of a given target (randomized across children and repetition order), and made a judgment of whether the item was correctly or incorrectly repeated.
-She additionally orthographically transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
-
-\hypertarget{analyses}{%
-\subsubsection{Analyses}\label{analyses}}
-
-Some NWR studies employ phoneme-based scoring in addition to or instead of word-level scoring (e.g., Cristia et al., 2020).
-We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
-Although previous work does not use distance metrics, we additionally report those.
-For instance, recall our example of an English target being ``bilik'', and imagine a response ``bilig''.
-This would receive the following scores: at the whole item level this would receive a score of zero (because the repetition is not exact); at the phoneme level it would receive a score of 80\% (4 out of 5 phonemes repeated exactly); and its normalized phone-based Levenshtein distance is 20\% (20\% of phonemes were substituted or deleted).\\
-
-Additionally, we looked up each phoneme in each target word in the Phoible database (Moran \& McCloy, 2019).
-We entered the number of languages in which each phoneme was found, as well as what percentage of languages in the database that represented.
-We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
-
-Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.
-
-\hypertarget{participants}{%
-\subsubsection{Participants}\label{participants}}
-
-This study was approved as part of a larger research effort by Marisa Casillas, with the research protocol having been approved by the IRB committee entitled The Ethics Committee of the Faculty of Social Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) of Radboud University, Nijmegen, The Netherlands.
-The approval for work outside of the Netherlands came as an amendment to the general research line on language development research (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
-Participation was voluntary, with children being invited to come and participate.
-Regardless of whether they did any of the task, children were provided with a snack as compensation.
-Children who came up to participate but then refused were nonetheless provided with the snack.
+\subsubsection{Procedure}\label{procedure}
+
+We tried to balance three desiderata: That children would not be unduly
+exposed to the items before they themselves had to repeat them; that
+children would feel comfortable doing this task with us; and that the
+community would feel safe with us doing this task with their children.
+Moreover, there were also some logistic constraints in terms of the
+space availability. As a result, the places where elicitation happened
+varied across the hamlets.
+
+We visited four different hamlets once, and attempted to test all
+eligible children present at the time, to prevent the items `spreading'
+through hearsay. In the first hamlet, we tested children in five
+different places, with some children being tested inside a house and
+others tested on the veranda. The complete list of places and the ways
+in which they met the desiderata mentioned above can be found in the raw
+data, available from online supplementary materials.
+
+The child was donned a headset (a Shure SM10A or WH20 XLR with a dynamic
+microphone on a headband, most children using the former), recorded into
+the left channel into a Tascam DR40x digital recorder. For most
+children, the headset could not stay comfortably on the child's head,
+and thus it was placed on the child's shoulders, with the microphone
+carefully placed close to the child's mouth. A local informant sat next
+to the child, to would provide the instructions and, if needed, coach
+the child to make sure, using the three practice items as well as real
+words, that they understood that the task was to repeat the items
+precisely without changing anything. An experimenter (the first author)
+delivered the elicitation stimuli to the local informant and the child
+over headphones.
+
+The first phase was making sure the child understood the task. This was
+explained orally and the first training item was presented. Often,
+children froze and did not say anything. If this happened, then we
+followed this procedure. First, the informant insisted. If the child
+still did not say anything, the informant asked the child to repeat a
+real word, and another, and another. If the child could repeat these
+correctly, then we provided the recorded training item over headphones
+again. Most children successfully started repeating the items presented
+over headphones at this point; a few further needed the local informant
+to model the behavior (i.e., they would hear the item again, and she
+would say it; then we would play it again, and ask the child to say it).
+A small minority still failed to repeat the item after hearing it over
+headphones. If that occurred, we tried with the second training item, at
+which point some children got it and could continue. A small minority,
+however, failed to repeat this one, as well as the third training item,
+in which case we stopped the test altogether (see Participants section
+for exclusions).
+
+NWR studies vary in whether children are provided with several
+opportunities to hear and say the item. To have a fixed and clear
+procedure, we decided that items other than the inital three training
+ones would not be repeated unless the child made an attempt to produce
+them. If this attempt was judged correct by the local informant, then
+the experimenter would move on to the next item (whispering this over a
+separate headset that was recording onto the right track of the same
+Tascam). If the local informant heard a deviation, she indicated to the
+experimenter that the item needed to be repeated, and up to 5 attempts
+were allowed.
+
+Whenever siblings from the same family were tested, an attempt was made
+to test first the older and then the younger child, and always on
+different elicitaiton sets.
+
+\subsubsection{Coding}\label{coding}
+
+A script was written to randomize all tokens from all children, pairing
+each with the auditory target the child had been provided with. A native
+research assistant then listened through all productions of a given
+target (randomized across children and repetition order), and made a
+judgment of whether the item was correctly or incorrectly repeated. She
+additionally orthographically transcribed exactly what the child said,
+providing some examples of the types of errors children in general make
+(without making specific reference to Yélî sounds or the items in the
+elicitation sets).
+
+\subsubsection{Analyses}\label{analyses}
+
+Some NWR studies employ phoneme-based scoring in addition to or instead
+of word-level scoring (e.g., Cristia et al., 2020). We scored items in
+terms of the number of phonemes that could be aligned across the target
+and attempt, divided by the number of phonemes of whichever item was
+longer (the target or the attempt). Although previous work does not use
+distance metrics, we additionally report those. For instance, recall our
+example of an English target being ``bilik'', and imagine a response
+``bilig''. This would receive the following scores: at the whole item
+level this would receive a score of zero (because the repetition is not
+exact); at the phoneme level it would receive a score of 80\% (4 out of
+5 phonemes repeated exactly); and its normalized phone-based Levenshtein
+distance is 20\% (20\% of phonemes were substituted or deleted).\\
+
+Additionally, we looked up each phoneme in each target word in the
+Phoible database (Moran \& McCloy, 2019). We entered the number of
+languages in which each phoneme was found, as well as what percentage of
+languages in the database that represented. We could not find a small
+number of phonemes in the database, and treat those as NA values that do
+not contribute to the target-level global average phone cross-linguistic
+frequency, although an alternative would be to attribute them a very
+small number.
+
+Finally, for describing children's patterns of errors, all repetitions
+of a given target were taken into account. We describe the proportion of
+items where the change resulted in a real word (semantic errors); as
+well as deletions, additions, and substitutions.
+
+\subsubsection{Participants}\label{participants}
+
+This study was approved as part of a larger research effort by Marisa
+Casillas, with the research protocol having been approved by the IRB
+committee entitled The Ethics Committee of the Faculty of Social
+Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen;
+ECSW) of Radboud University, Nijmegen, The Netherlands. The approval for
+work outside of the Netherlands came as an amendment to the general
+research line on language development research (original request:
+ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+Participation was voluntary, with children being invited to come and
+participate. Regardless of whether they did any of the task, children
+were provided with a snack as compensation. Children who came up to
+participate but then refused were nonetheless provided with the snack.
 
 A total 55 children were tested, from 34 of families, in five hamlets.
-Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5).
-In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
-The remaining 40 children (14 girls) were aged 6.96 years (range 3.90-11 years).
-Among these, there were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and 2 for whom this information is missing.
-Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Precise years of education can be hard to count.
-This is particularly the case for schooling beyond grade 8, which takes place off-island, which can be due to an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
-We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical ``completion'' times in between.} In terms of birth order, 5 were first borns, 4 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
-
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
-
-\hypertarget{preliminary-analyses}{%
-\subsubsection{Preliminary analyses}\label{preliminary-analyses}}
+Some children could not be included for the following reasons: refused
+participation or failed to repeat items presented over headphones even
+after coaching (N=8), spoke too softly to allow offline coding (N=5). In
+addition, 2 teenagers were tested to put younger children at ease; their
+data is not included in analyses below. The remaining 40 children (14
+girls) were aged 6.96 years (range 3.90-11 years). Among these, there
+were 32 children exposed only to Yélî in the home, 6 children who were
+also exposed to another language in the home, and 2 for whom this
+information is missing. Maternal years of education averaged 8.24 years
+(range 6-12 years; 2 children had this information missing).\footnote{Precise
+  years of education can be hard to count. This is particularly the case
+  for schooling beyond grade 8, which takes place off-island, which can
+  be due to an irregularity in ability to pay school fees, school
+  opening times, familial obligations, and more. We asked parents to
+  report their highest completed level of education and inferred the
+  approximate number of years that entails under ideal conditions, which
+  results in many esimates around typical ``completion'' times in
+  between.} In terms of birth order, 5 were first borns, 4 second, 2
+third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
+
+\subsection{Results}\label{results}
+
+\subsubsection{Preliminary analyses}\label{preliminary-analyses}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig1-first_vs_others-1.pdf}
-\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual participants averaging separately their first attempts and all other attempts.}
+\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual
+participants averaging separately their first attempts and all other
+attempts.}
 \end{figure}
 
-We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
-We excluded 3 children who did not have data for one of these two types.
-As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
+We first checked whether NWR scores varied between first and subsequent
+presentations of an item by averaging word-level scores at the
+participant level separately for first attempts and subsequent
+repetitions. We excluded 3 children who did not have data for one of
+these two types. As shown in Figure 1, participants' mean word-level
+scores became more heterogeneous in subsequent repetitions.
 Surprisingly, NWR scores for subsequent repetitions (M = 40, SD = 26)
-were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p = 0.
-Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
-
-Taking into account only the first attempts, we averaged attempts by each of the 42 children who had data for first attempts; their ages ranged from NA to NA (M = NA, SD = NA).
-
-The overall NWR score was M = 65\% (SD = 15\%).
-Scores based on phonemes are even higher M = 78\% (SD = 9\%).
-The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
-Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
-
-\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
-\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}
-
-In this analysis, we were interested in variation in NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
-To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID \& target ID as random variables.
-
-We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Age was also a significant predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
+were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p
+= 0. Given the uncertainty in whether previous work used only the first
+or all repetitions, and since scores declined and became more
+heterogeneous in subsequent repetitions, the rest of the analyses focus
+on only the first repetitions.
+
+Taking into account only the first attempts, we averaged attempts by
+each of the 42 children who had data for first attempts; their ages
+ranged from NA to NA (M = NA, SD = NA).
+
+The overall NWR score was M = 65\% (SD = 15\%). Scores based on phonemes
+are even higher M = 78\% (SD = 9\%). The phoneme-based normalized
+Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth
+of phonemes were were substituted, inserted, or deleted. Notice that the
+normalized Levenshtein distance is the complement of phoneme-based
+scores.
+
+\subsubsection{NWR as a function of cross-linguistic phone
+frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}
+
+In this analysis, we were interested in variation in NWR scores as a
+function of the average frequency with which sounds composing individual
+target words are found in languages over the world. To look at this, we
+fit a mixed logistic regression, in which the outcome variable was
+whether the non-word was correctly repeated or not. The fixed effects of
+interest was the average cross-linguistic phone frequency; we also
+included child age as a control fixed effect, and child ID \& target ID
+as random variables.
+
+We could include 885 observations (\textbf{why -- shouldn't it be
+1k??}), from 41 children producing in any given trial one of 41
+potential target words. The analysis revealed a significant estimate for
+the average cross-linguistic frequency of phones in the target words (ß
+= 0.82, SE ß = 0.18, p = 0): Target words with phones found more
+frequently across languages had a higher proportion of words that were
+correctly repeated, as clear in Figure 2. Age was also a significant
+predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
 
@@
 
-We next checked whether this may actually be due to frequency of the sounds within the language.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020), and applied the natural logarithm.
-As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(27)=0.52, =0.00).
-Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
-As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic frequency (ß = 0.82, SE ß = 0.27, p = 0.00) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0, SE ß = 0.25, p = 1.00) was not.
+We next checked whether this may actually be due to frequency of the
+sounds within the language. We estimated frequency of the phones present
+in the stimuli in a corpus of child-centered recordings (Casillas et
+al., 2020), and applied the natural logarithm. As before, unattested
+sounds were not considered (i.e., they were declared NA so that they do
+not count for analyses). Phone frequency estimated from this corpus were
+correlated with cross-linguistic phone frequencies (r(27)=0.52, =0.00).
+Therefore, we fit another mixed logistic regression, this time declaring
+as fixed effects scaled cross-linguistic and corpus frequencies
+(averaged across all attested phones within each stimulus item), in
+addition to age. As before, the model contained random slopes for both
+child ID and target. In this model, both cross-linguistic frequency (ß =
+0.82, SE ß = 0.27, p = 0.00) and age (ß = 0.33, SE ß = 0.13, p = 0.01)
+were significant predictors of word-level NWR scores, but corpus
+frequency (ß = 0, SE ß = 0.25, p = 1.00) was not.
 
@@
 
-\textbf{a2m use first attempts rather than the whole data set: having second thoughts.
-To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers \& it is more consistent.
-BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros\ldots{} CF Option 3 further down}
-
-\textbf{a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?}
-
-\textbf{Option 1: just numbers, first attempts only }
-Next, we investigated patterns of deletion and substitution. Deletions were relatively rare, with only 7 vowels deleted, and 3 consonants.
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels, 30 nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (110 oral vowels produced with a different quality; 13 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 115) than vice versa (2).
-Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones, compared to 33 complex ones).
-
-\textbf{Option 2: numbers and proportions, first attempts only}
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 7 vowels deleted (about 0.23\% of all vowel targets), and 3 consonants deleted (about 0.12\% of all consonant targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (34 oral target vowels produced as nasal vowels or about 0.99\% out of all oral vowel targets, 30 nasal target vowels produced as oral vowels or about 2.37\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (110 oral vowels produced with a different quality or about 1.86\% out of all oral vowel targets; 13 nasal vowels produced with a different quality or about 2.06\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 115 times or about 1.96\% out of all complex consonant targets) than vice versa (2 times or about 1.96\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones or about 0.39\% out of all simple consonant targets, compared to 33 complex ones or about 1.02\% out of all complex consonant targets).
-
-\textbf{Option 3: numbers and proportions, ALL attempts}
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only 12 vowels deleted (about 0.28\% of all vowel targets), and 6 consonants deleted (about 0.19\% of all consonant targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (52 oral target vowels produced as nasal vowels or about 1.10\% out of all oral vowel targets, 58 nasal target vowels produced as oral vowels or about 2.53\% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (205 oral vowels produced with a different quality or about 2.16\% out of all oral vowel targets; 23 nasal vowels produced with a different quality or about 2.36\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or about 2.43\% out of all complex consonant targets) than vice versa (2 times about 2.43\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (123 simple consonants mispronounced as other simple ones or about 0.53\% out of all simple consonant targets, compared to 55 complex ones or about 0.98\% out of all complex consonant targets).
-
-Finally, we looked into the frequency with which mispronunciations resulted in real words. Two thirds of them were: 62\%.
-
-\hypertarget{nwr-scores-as-a-function-of-item-length}{%
-\subsubsection{NWR scores as a function of item length}\label{nwr-scores-as-a-function-of-item-length}}
-
-Next, we inspected whether NWR scores varied as a function of word length.
-Results are shown on table XX.
-This table shows that scores for monosyllables was much lower than for other lengths.
-This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
-
-Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole word scoring.
-This is the most commonly reported type of score, but it is also the least forgiving.
-The pattern is less marked when other two scores are used, which are less sensitive to errors.
+\textbf{a2m use first attempts rather than the whole data set: having
+second thoughts. To be clear: the argument for using first attempts only
+is that all of the other analyses are on kids' first attempts, rather
+than all of their attempts, so we should use just first attempts here
+because it may confuse readers \& it is more consistent. BUT there are
+fewer mistakes this way, which is Options 1 and 2 below has lots of
+zeros\ldots{} CF Option 3 further down}
+
+\textbf{a2m probably turn the paragraph on substitutions into a table:
+having second thoughts, the table would look just like a para, no?}
+
+\textbf{Option 1: just numbers, first attempts only } Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 7 vowels deleted, and 3 consonants.
+
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (34 oral target vowels produced as nasal
+vowels, 30 nasal target vowels produced as oral vowels). Substitutions
+in which the oral nature was preserved but the quality of the vowel was
+changed were a great deal more common than changes in quality among
+nasal vowels (110 oral vowels produced with a different quality; 13
+nasal vowels produced with a different quality). As for consonants,
+asymmetries were very marked with more complex consonants (specifically
+dptpkpkmknmbghlv) mispronounced as simple ones (specifically
+mnlwyvdgptkfhch, 115) than vice versa (2). Simple consonants were
+mispronounced as other simple consonants quite frequently (69 simple
+consonants mispronounced as other simple ones, compared to 33 complex
+ones).
+
+\textbf{Option 2: numbers and proportions, first attempts only} Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 7 vowels deleted (about 0.23\% of all vowel
+targets), and 3 consonants deleted (about 0.12\% of all consonant
+targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (34 oral target vowels produced as nasal
+vowels or about 0.99\% out of all oral vowel targets, 30 nasal target
+vowels produced as oral vowels or about 2.37\% out of all nasal vowel
+targets). Substitutions in which the oral nature was preserved but the
+quality of the vowel was changed were a great deal more common than
+changes in quality among nasal vowels (110 oral vowels produced with a
+different quality or about 1.86\% out of all oral vowel targets; 13
+nasal vowels produced with a different quality or about 2.06\% out of
+all nasal vowel targets). As for consonants, asymmetries were very
+marked with more complex consonants (specifically dptpkpkmknmbghlv)
+mispronounced as simple ones (specifically mnlwyvdgptkfhch, 115 times or
+about 1.96\% out of all complex consonant targets) than vice versa (2
+times or about 1.96\% out of all simple consonant targets). Simple
+consonants were mispronounced as other simple consonants quite
+frequently (69 simple consonants mispronounced as other simple ones or
+about 0.39\% out of all simple consonant targets, compared to 33 complex
+ones or about 1.02\% out of all complex consonant targets).
+
+\textbf{Option 3: numbers and proportions, ALL attempts} Next, we
+investigated patterns of deletion and substitution. Deletions were
+relatively rare, with only 12 vowels deleted (about 0.28\% of all vowel
+targets), and 6 consonants deleted (about 0.19\% of all consonant
+targets).
+
+As for substitutions, it was as common for a nasal vowel to be produced
+as an oral vowel as vice versa (52 oral target vowels produced as nasal
+vowels or about 1.10\% out of all oral vowel targets, 58 nasal target
+vowels produced as oral vowels or about 2.53\% out of all nasal vowel
+targets). Substitutions in which the oral nature was preserved but the
+quality of the vowel was changed were a great deal more common than
+changes in quality among nasal vowels (205 oral vowels produced with a
+different quality or about 2.16\% out of all oral vowel targets; 23
+nasal vowels produced with a different quality or about 2.36\% out of
+all nasal vowel targets). As for consonants, asymmetries were very
+marked with more complex consonants (specifically dptpkpkmknmbghlv)
+mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or
+about 2.43\% out of all complex consonant targets) than vice versa (2
+times about 2.43\% out of all simple consonant targets). Simple
+consonants were mispronounced as other simple consonants quite
+frequently (123 simple consonants mispronounced as other simple ones or
+about 0.53\% out of all simple consonant targets, compared to 55 complex
+ones or about 0.98\% out of all complex consonant targets).
+
+Finally, we looked into the frequency with which mispronunciations
+resulted in real words. Two thirds of them were: 62\%.
+
+\subsubsection{NWR scores as a function of item
+length}\label{nwr-scores-as-a-function-of-item-length}
+
+Next, we inspected whether NWR scores varied as a function of word
+length. Results are shown on table XX. This table shows that scores for
+monosyllables was much lower than for other lengths. This is likely due
+to the fact that the majority of monosyllables included were chosen
+because they had sounds that are rare in the world's languages, which
+may indicate that they are hard to produce or to perceive.
+
+Setting monosyllables aside, we observe the typical pattern of lower
+scores for longer items, although this is particularly salient for the
+whole word scoring. This is the most commonly reported type of score,
+but it is also the least forgiving. The pattern is less marked when
+other two scores are used, which are less sensitive to errors.
 
 \begin{table}[tbp]
-
 \begin{center}
 \begin{threeparttable}
-
 \caption{\label{tab:tablength}NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
-
 \begin{tabular}{lll}
@@
 \includegraphics{manuscript_files/figure-latex/Fig2-scores by age-1.pdf}
-\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual participants as a function of age and sex (blue = boys, pink = girls).}
+\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual
+participants as a function of age and sex (blue = boys, pink = girls).}
 \end{figure}
 
-Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (6,496.39) = 0.43, p = 0.00).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (-0.61) = 27.83, p = 0.55), birth order (data missing for 17 children, rho = (2,652.32) = -0.15, p = 0.47) or maternal education (data missing for 2 children, rho = ( 9,022.76) = 0.09, p = 0.60).
-
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+Our final exploratory analysis assessed whether variation in scores was
+structured by factors that vary across individuals. As shown in Figure
+2, there was a greater deal of variance in earlier than later ages, with
+significantly higher NWR scores for older children (Spearman's rank
+correlation, given inequality of variance, rho (6,496.39) = 0.43, p =
+0.00). In contrast, there was no clear association between NWR scores,
+on the one hand, and sex (t (-0.61) = 27.83, p = 0.55), birth order
+(data missing for 17 children, rho = (2,652.32) = -0.15, p = 0.47) or
+maternal education (data missing for 2 children, rho = ( 9,022.76) =
+0.09, p = 0.60).
+
+\subsection{Discussion}\label{discussion}
 

### 2020-09-17 manuscript.Rmd
add responses to AC comments

@@
 <!-- a2m I added a sentence, is this what you were thinking of? -->
+<!-- M2A: Makes sense.
+I was thinking we could already hint at YD as interesting in particular bc of the phonology and not simply because it's a random one of many previously untested populations.
+I don't mind adding something like this in once I'm done with the demo check.-->
 
@@
 -->
+<!-- M2A: Happy to comment it out I agree it's not really relevant for our study in particular; just part of the phonological sketch.
+I was thinking about linguists when I wrote it up, but psychies probably are just ready to get on with the next part by now :) -->
 
+<!--
 While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
@@
 Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see @levinsoninprepydgrammar for details).
+-->
 
@@
 <!-- a2m no "at" here? *at* around the age? -->
+<!-- M2A: my ear strongly prefers it without the "at", but it could go both ways. -->
 school instruction is in English.
@@
 -->
+<!-- M2A: fine with me!-->
 
@@
 <!-- a2m added explanation of correct and incorrect for nwr task -->
+<!--M2A: thanks!-->
 
@@
 <!-- a2m good point, but I think it goes in the previous section so I added it there -->
+<!--M2A: thanks!-->
 
@@
   theme(text = element_text(size = 16),
-    legend.position="none", axis.text.x=element_text(colour=c("black", "black"))) +
+    legend.position="none", axis.text.x=element_text(colour="black")) +
   scale_color_manual(values = palette)+
@@
 # Results may be unexpected or may change in future versions of ggplot2. 
+# M2A: removed the vector since it wasn't needed; no more complaint!
 
@@
 <!--a2m: good point! fixed-->
+<!-- M2A: thanks! -->
 
@@
 <!-- a2m done -->
+<!-- M2A: thanks! -->
 
@@
 <!-- a2m added! actually, I haven't changed anything but I've been hesitant about using levenshtein's distance because it's the complement of phoneme based scoring -- so shall we drop it?  -->
+<!-- M2A: three thoughts on this, which you can take as you like:
+1) our audience might make this determination. segment-match score (phoneme-level accuracy) is probably more straightforward and might therefore be preferred by psychies
+2) LD has the advantage of making D/I/S relevant as a next step in the analysis
+3) we should definitely choose one, I agree, and report the other in supp mat tables
+-->
 
@@
 #a2m fixed and fixed!
+# M2A: thanks!
 aggregate(trnsc$correct*100,by=list(trnsc$id),mean)->overall
@@
 Or making things too complicated, and just say "so corpus frequency does not explain away these results" -->
+<!-- M2A: let's not go into it.
+because of the way glmer treats collinearity (https://psycnet.apa.org/record/2014-08495-005), it gets complicated quickly in even giving an accurate report of what's going on -->
 
@@
 <!-- a2m I realized this section did not belong together with the xling freq section above - are you ok with this title? -->
+<!-- M2A: The title is a little long and maybe less informative than it could be. How about something like, "Patterns in NWR mispronunciations"-->
 
@@
 ```
+
+<!--M2A: coming back to these three comments below after I manage another full read-through (after I check demo) -->
 
@@
 <!-- a2m I added info above - is it weird to talk about this here? -->
+<!-- M2A: I don't think it's a weird place to add the idea, though I might shorten it a bit when I go through on a writing pass :) -->
 
@@
 #a2m fixed
+# M2A: thanks!
 aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$type),mean)->by_length
@@
 <!-- a2m it doesn't make sense to do this, right, with only 3 items? -->
+<!-- M2A: you mean with only three lengths? I guess I would prefer a little regression -->
 

### 2020-09-17 wrangling.R
add responses to AC comments

@@ -69,6 +69,7 @@ dim(trnsc) #still no loss, all good
 # across the two, the merge should carry over whatever changes you make
 # M2A: Some of the segments in stimuli.txt were wrong. I made corrections in stimuli2.txt,
 # but now that is mismatched to stimconv and segconv.
+# M note: resolved over quick call; a updated the stim characters throughout now (resolved)
 
 #get info on segments
 read.delim("segconv.txt", encoding="UTF-8")->segments
@@ -142,6 +143,7 @@ trnsc$correct[trnsc$correct==0 & trnsc$target_uni == trnsc$mp_uni & !is.na(trnsc
 # M2A: I agree! My question is: Why then are they marked as incorrect in the first place?
 # Does this deserve more detective work?
 #a2m I looked into in there were disfluencies
+# M2A: awesome, thank you! happy with this now :)
 
 # create phonological correspondance matrix
 # NOTE!! NON INTUITIVE MAPPING!!!! DO NOT READ THE OUTPUT BELIEVING IT IS PSEUDO IPA!!!
@@ -294,6 +296,7 @@ for(i in grep("0",trnsc$correct)){ #this goes through all items with errors (" g
 # M2A: what's this for if we don't expect any S/D/Is?
 #a2m for ease of reading the code, I had left the section that does insertions, deletions, substitutions -- but we know there are only matches here ;)
 #but seems that zas confusing
+# M2A: ah alright!
 for(i in grep("1",trnsc$correct)){ #this goes through all items without errors (" grep("1",trnsc$correct)")
   #lev
   lev=adist(trnsc[i,c("target_uni")],trnsc[i,c("target_uni")],count=T)

### 2020-09-18 wrangling.R
minor fixes in mispronunciation transcription and meaning

@@ -1,6 +1,8 @@
 library(tidyverse)
 
 read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
+
+
 #DATA REMOVAL *** ATTENTION *** 
 #REMOVE ALL TASK 4 (whatever that is)
 # M2A: This appears to be a second pass over productions that were deemed to
@@ -11,6 +13,38 @@ trnsc[trnsc$item!="task4",]->trnsc
 dim(trnsc) #1488 judgments
 
 trnsc$tokid=paste(trnsc$item,trnsc$token)
+
+
+# - ? correct treatment of words whose mispronunciation is an English word:
+#   mispronunciation meaning
+# 361              two     two
+# 362              two     two
+# 732             loudî      English: loud
+# 879          no-pinêt                                          English: no peanut
+# 908            nodedi                                           English: no daddy
+# 933          no--dadi                                           English: no daddy
+# 949           petrumi                                             English: "true"
+# 950           petrumi                                             English: "true"
+# 951           petrumi                                             English: "true"
+# 1373              wee     English: where
+# 1374              wee     English: where
+# 1375              wee     English: where
+# 1379              wee     English: where
+# 1380              wee     English: where
+# 1388              wee     English: where
+# 1394              wee     English: where
+# 1396              wee     English: where
+# 1398              wee     English: where
+# 1399              wee     English: where
+# 1402              wat      English: what
+# 1405              wat      English: what
+# 1411              wee     English: where
+
+trnsc$mispronunciation=as.character(trnsc$mispronunciation)
+trnsc$meaning=as.character(trnsc$meaning)
+trnsc$mispronunciation[trnsc$mispronunciation=="two"]<-"tuu"
+trnsc$meaning[trnsc$meaning=='English: "true"']<-"0"
+trnsc[grep("Eng",trnsc$meaning),c("mispronunciation","meaning")]
 
 read.table("final_order.txt",header=T)->crp
 crp$tokid=paste(crp$target,crp$nb)

### 2020-09-18 manuscript.Rmd
minor fixes in mispronunciation transcription and meaning

@@
 - draft discussion
-
-## TODO Alex
-
-- ? correct treatment of words whose mispronunciation is an English word:
-    mispronunciation meaning
-361              two     two
-362              two     two
-879          no-pinêt                                          English: no peanut
-908            nodedi                                           English: no daddy
-933          no--dadi                                           English: no daddy
-949           petrumi                                             English: "true"
-950           petrumi                                             English: "true"
-951           petrumi                                             English: "true"
-1373              wee     English: where
-1374              wee     English: where
-1375              wee     English: where
-1379              wee     English: where
-1380              wee     English: where
-1388              wee     English: where
-1394              wee     English: where
-1396              wee     English: where
-1398              wee     English: where
-1399              wee     English: where
-1402              wat      English: what
-1405              wat      English: what
-1411              wee     English: where
 

### 2020-09-18 wrangling.R
minor fixes in mispronunciation transcription and meaning

@@ -19,6 +19,9 @@ trnsc$tokid=paste(trnsc$item,trnsc$token)
 #   mispronunciation meaning
 # 361              two     two
 # 362              two     two
+# --> these two were converted to YD orthography
+# 
+# The following have not been changed in any way
 # 732             loudî      English: loud
 # 879          no-pinêt                                          English: no peanut
 # 908            nodedi                                           English: no daddy
@@ -43,7 +46,6 @@ trnsc$tokid=paste(trnsc$item,trnsc$token)
 trnsc$mispronunciation=as.character(trnsc$mispronunciation)
 trnsc$meaning=as.character(trnsc$meaning)
 trnsc$mispronunciation[trnsc$mispronunciation=="two"]<-"tuu"
-trnsc$meaning[trnsc$meaning=='English: "true"']<-"0"
 trnsc[grep("Eng",trnsc$meaning),c("mispronunciation","meaning")]
 
 read.table("final_order.txt",header=T)->crp

### 2020-09-18 manuscript.Rmd
finish demo check (add note for AC)

@@
 
-- double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
-I looked again at my local files (still haven't checked the labs' ones) and I'm surprised I have maternal education for more children than I have birth order for...
-If you correct this, to generate the csv I also round the age, assign age.rounded to reported age (in years) when DOB is missing, remove first names, DOB, DOTest and age.precise before pushing to github
 - double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
@@
 - draft discussion
+
+<!--M2A: regarding demo, it looks like you have more mateds than birth orders because family data was systematically not collected for Pu.
+kids and was missing for three kids tested in No.
+who were actually from other hamlets.
+New xslx files sent via email!
+I left age recomputing to you since I was confused about which file would be used.-->
 

### 2020-09-18 wrangling.R
track down NAs in stim_seg_freq

@@ -105,7 +105,7 @@ dim(trnsc) #still no loss, all good
 # across the two, the merge should carry over whatever changes you make
 # M2A: Some of the segments in stimuli.txt were wrong. I made corrections in stimuli2.txt,
 # but now that is mismatched to stimconv and segconv.
-# M note: resolved over quick call; a updated the stim characters throughout now (resolved)
+# M note: resolved over quick videocall; a updated the stim characters throughout now (resolved)
 
 #get info on segments
 read.delim("segconv.txt", encoding="UTF-8")->segments

### 2020-09-18 manuscript.Rmd
track down NAs in stim_seg_freq

@@
 - double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
-- double check also stim_seg_freq -- why are some items still with NA freq?
+<!--M2A: I think you mentioned you might have someone do this check, but let me know if I misunderstood!-->
 - draft discussion
 
-<!--M2A: regarding demo, it looks like you have more mateds than birth orders because family data was systematically not collected for Pu.
+<!-- M2A: regarding demo, it looks like you have more mateds than birth orders because family data was systematically not collected for Pu.
 kids and was missing for three kids tested in No.
@@
 I left age recomputing to you since I was confused about which file would be used.-->
+<!-- M2A: regarding NAs in stim_seg_freq, these appear for 'lv' stimuli because, in the 'segments' table, freq_corpus and freq_corpus_types are NA for that grapheme).
+The 'segments' table has NAs for 'lv' because it is not included in segment-counts.txt or segment-counts-types.txt.
+Note that that these NAs are excluded when adding mean item frequency to the 'stims' table (wrangling.R/l.136 & l.145), but this has a pretty big impact since those are all two-segment words.
+Not sure how avoidable this is, going back to our discussion before about not wanting to ad arbitrarily small frequencies.
+Thoughts?
+-->
 

### 2020-10-05 manuscript.Rmd
minor

@@
 cohensd2=function(m1,sd1,m2,sd2){round((m1-m2)/sqrt((sd1^2+sd2^2)/2),3)}
+
 ```
@@
 <!--M2A: I think you mentioned you might have someone do this check, but let me know if I misunderstood!-->
-- draft discussion
+<!-- a2m we then decided we'd both feel more certain if you did... -->
 
@@
 I left age recomputing to you since I was confused about which file would be used.-->
+<!-- a2m thanks!!! now inc'ed -->
+
 <!-- M2A: regarding NAs in stim_seg_freq, these appear for 'lv' stimuli because, in the 'segments' table, freq_corpus and freq_corpus_types are NA for that grapheme).
@@
 -->
+<!-- a2m So your fear is that frequency is overestimated in these cases, right? Perhaps we can assign the missing segments the lowest observed frequencies? -->
+
+- draft discussion
 
@@
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments,  particularly in Europe [@meir2016disentangling;@is08042009language].
-However, it has been seldom used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant].
+However, it has been seldom used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant], and mostly with languages having phonological inventories that are small to moderate in size [@maddieson2005correlating].
 
@@
 I don't mind adding something like this in once I'm done with the demo check.-->
+<!-- a2m added ", and mostly with languages having phonological inventories that are small to moderate in size" + notice the " with an unusually dense phonological inventory"; my feeling is that it may interrupt the flow to add a lot more here...
+don't you think?
+-->
 
@@
 Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7; 
-<!-- a2m no "at" here? *at* around the age? -->
-<!-- M2A: my ear strongly prefers it without the "at", but it could go both ways. -->
 school instruction is in English.
@@
 This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 people) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
-<!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->
-<!-- a2m maybe we want to bring that up in the discussion?
-I haven't found a good place to introduce that paper, in part because right now we have these three purposes that seem sort of unconnected with the goals of that paper...
--->
-<!-- M2A: fine with me!-->
 
@@
 For instance, in English, a response of "bilig" or "pilik" will be scored as incorrect; a response "biilik", where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.
-<!-- a2m added explanation of correct and incorrect for nwr task -->
-<!--M2A: thanks!-->
 
@@
 - Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
-<!-- a2m we don't actually look at this: Are rarer sounds more often substituted by commoner sounds? -- shall we remove it? -->
-<!--M2A: I'm imagining a reviewer who's knowlegeable about phonology and child development but who hasn't encountered NWR before.
-I think it would be beneficial to provide a concrete example of a nonword and an inexact repetition above so that they can conceive of what these questions imply (analytically) before we get into the methodological weeds.-->
-<!-- a2m good point, but I think it goes in the previous section so I added it there -->
-<!--M2A: thanks!-->
 
@@
   scale_size_continuous(range = c(2, 6))
-
-# a2m I get a warning:
-#   Warning message:
-# Vectorized input to `element_text()` is not officially supported.
-# Results may be unexpected or may change in future versions of ggplot2. 
-# M2A: removed the vector since it wasn't needed; no more complaint!
 
@@
 The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
- <!--M2A: I didn't look for the SM yet, but probably best not to provide exact hamlet names-->
-<!--a2m: good point! fixed-->
-<!-- M2A: thanks! -->
 
@@
 A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether (see Participants section for exclusions).
-<!--M2A: do you know how many these were? oh! it's below. then maybe we can just put a (see Participants below), etc.-->
-<!-- a2m done -->
-<!-- M2A: thanks! -->
 
@@
 
-read.csv("NWR-demo.csv",header=T)->demo
+read.csv("nwr-demo.csv",header=T,sep=",")->demo
 #colnames(demo)
 
+#lines no longer needed
 #if family is missing, attribute each child to a different one
-demo$familyID=as.character(demo$familyID)
-demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])
-
-#fix sex
-demo$Sex[demo$Sex=="F"]<-"G"
+#demo$familyID=as.character(demo$familyID)
+#demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])
+#recalc age
+#demo$age.rounded=as.numeric(difftime(as.Date(as.character(demo$Date),format="%Y%m%d"),as.Date(as.character(demo$DOB.ISO),format="%Y%m%d"),unit="weeks"))/52.25
+write.table(demo,"nwr-demo.csv")
 
@@
 
-A total `r dim(demo)[1]` children were tested, from `r length(levels(factor(demo$familyID)))` of families, in five hamlets.
-Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn't repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`).
+A total `r dim(demo)[1]` children were tested, from `r length(levels(factor(demo$familyID)))` of families, in four hamlets.
+Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`).
 In addition, `r sum(demo$Age >12 | demo$included=="too old",na.rm=T)` teenagers were tested to put younger children at ease; their data is not included in analyses below.
@@
 ```{r overall}
-# M2A: This excludes a bunch of kids (anyone with NA on any of these participant features),
-# whittling the summary down to 24 from 42 -- is that intended?
-# Also all phon_score is 100 and all NLD is 0; does that stike you as fishy?
-#a2m fixed and fixed!
-# M2A: thanks!
 aggregate(trnsc$correct*100,by=list(trnsc$id),mean)->overall
@@
 
-<!-- a2m should we point out that the coefficient for cross-ling frequency is reduced, so indirectly corpus frequency does explain a little bit of variance?
-Or making things too complicated, and just say "so corpus frequency does not explain away these results" -->
-<!-- M2A: let's not go into it.
-because of the way glmer treats collinearity (https://psycnet.apa.org/record/2014-08495-005), it gets complicated quickly in even giving an accurate report of what's going on -->
-
-### Qualitative analyses of NWR mispronunciations
-
-<!-- a2m I realized this section did not belong together with the xling freq section above - are you ok with this title? -->
-<!-- M2A: The title is a little long and maybe less informative than it could be. How about something like, "Patterns in NWR mispronunciations"-->
+### Patterns in NWR mispronunciations
 
@@
 
-<!-- a2m I added info above - is it weird to talk about this here? -->
-<!-- M2A: I don't think it's a weird place to add the idea, though I might shorten it a bit when I go through on a writing pass :) -->
-
 ### NWR scores as a function of item length
@@
 ```{r bylength}
-# M2A: this reduces the total number of kids represented from 42 to 40
-#a2m fixed
-# M2A: thanks!
+
 aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$type),mean)->by_length
@@
 length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)
+
+xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
+mycoeff=round(summary(xl_fr)$coefficients,3)
+
 ```
@@
 
-**(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)**
-
 - How frequent are errors that result in real words? Is that a function of item complexity?
@@
 
+<!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->
+<!-- a2m maybe we want to bring that up in the discussion?
+I haven't found a good place to introduce that paper, in part because right now we have these three purposes that seem sort of unconnected with the goals of that paper...
+-->
+
 \newpage

### 2020-10-05 manuscript.Rmd
version that knits for sure

@@
 
+## TODO Alex
+
+- check again if all fields are explained
+- set up OSF (or revise if already there)
+- implement decisions on prop phones & patterns nwr mp
+- revise bullet points for discussion
+
 ## TODO Middy
@@
 <!-- a2m we then decided we'd both feel more certain if you did... -->
-
-<!-- M2A: regarding demo, it looks like you have more mateds than birth orders because family data was systematically not collected for Pu.
-kids and was missing for three kids tested in No.
-who were actually from other hamlets.
-New xslx files sent via email!
-I left age recomputing to you since I was confused about which file would be used.-->
-<!-- a2m thanks!!! now inc'ed -->
-
-<!-- M2A: regarding NAs in stim_seg_freq, these appear for 'lv' stimuli because, in the 'segments' table, freq_corpus and freq_corpus_types are NA for that grapheme).
-The 'segments' table has NAs for 'lv' because it is not included in segment-counts.txt or segment-counts-types.txt.
-Note that that these NAs are excluded when adding mean item frequency to the 'stims' table (wrangling.R/l.136 & l.145), but this has a pretty big impact since those are all two-segment words.
-Not sure how avoidable this is, going back to our discussion before about not wanting to ad arbitrarily small frequencies.
-Thoughts?
--->
-<!-- a2m So your fear is that frequency is overestimated in these cases, right? Perhaps we can assign the missing segments the lowest observed frequencies? -->
 
@@
 ### The Yélî community
-<!--M2A: is this okay?-->
-<!-- a2m I think it's great! although it seemed a bit unexpected to explain all of this, so I added some justification -->
 
@@
 
-read.csv("nwr-demo.csv",header=T,sep=",")->demo
+read.csv("nwr-demo.csv",header=T)->demo
 #colnames(demo)
@@
 #demo$age.rounded=as.numeric(difftime(as.Date(as.character(demo$Date),format="%Y%m%d"),as.Date(as.character(demo$DOB.ISO),format="%Y%m%d"),unit="weeks"))/52.25
-write.table(demo,"nwr-demo.csv")
+#write.table(demo,"nwr-demo.csv",row.names=F,quote=T,sep=",")
 
@@
 Error"]`, p = `r mycoeff["avg_fr_s","Pr(>|z|)"]`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r r(`r length_r$parameter`) = `r round(xlfr_r$estimate,3)`.
+Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.
 
@@
 
-[^2]: We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less convincing.
+[^2]: We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.
 
@@
 
-xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
-mycoeff=round(summary(xl_fr)$coefficients,3)
+len_lm=glmer(correct~type+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial",subset=c(trnsc$type!=1))
+mycoeff=round(summary(len_lm)$coefficients,3)
 
@@
 Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(`r length_r$parameter`) = `r round(length_r$estimate,3)`.
-<!-- a2m it doesn't make sense to do this, right, with only 3 items? -->
-<!-- M2A: you mean with only three lengths? I guess I would prefer a little regression -->
+In a generalized binomial mixed model, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing in any given trial one of `r summary(len_lm)$ngrps["target"]` potential target words.
+The analysis revealed a main effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
+Error"]`, p = `r mycoeff["age.rounded_s","Pr(>|z|)"]`); and a significant estimate for the length of the target words in number of syllables (ß = `r mycoeff["type","Estimate"]`, SE ß = `r mycoeff["type","Std.
+Error"]`, p = `r mycoeff["type","Pr(>|z|)"]`).
 

### 2020-10-05 manuscript.tex
version that knits for sure

--- 2020-09-16 manuscript.tex
@@
             pdftitle={Non-word repetition in Yélî Dnye},
-            pdfauthor={Alejandrina Cristia~\& Marisa Casillas},
             pdfkeywords={phonology, non-word repetition, development},
@@
 \setmainfont{Doulos SIL} % Set main font to Doulos SIL
+% Manuscript styling
+\usepackage{upgreek}
+\captionsetup{font=singlespacing,justification=justified}
+
+% Table formatting
+\usepackage{longtable}
+\usepackage{lscape}
+% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
+\usepackage{multirow}		% Table styling
+\usepackage{tabularx}		% Control Column width
+\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
+\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+
+% Create new environments so endfloat can handle them
+% \newenvironment{ltable}
+%   {\begin{landscape}\begin{center}\begin{threeparttable}}
+%   {\end{threeparttable}\end{center}\end{landscape}}
+\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
+
+% Enables adjusting longtable caption width to table width
+% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
+\makeatletter
+\newcommand\LastLTentrywidth{1em}
+\newlength\longtablewidth
+\setlength{\longtablewidth}{1in}
+\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
+
+% \setlength{\parindent}{0.5in}
+% \setlength{\parskip}{0pt plus 0pt minus 0pt}
+
+% \usepackage{etoolbox}
+\makeatletter
+\patchcmd{\HyOrg@maketitle}
+  {\section{\normalfont\normalsize\abstractname}}
+  {\section*{\normalfont\normalsize\abstractname}}
+  {}{\typeout{Failed to patch abstract.}}
+\makeatother
+\shorttitle{NWR in Yélî Dnye}
+\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2}}
+\affiliation{
+\vspace{0.5cm}
+\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics}
+\authornote{All data are made available in a repository in the Open Science Framework.
+AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
+S.
+McDonnell Foundation Understanding Human Cognition Scholar Award.
+
+Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
+\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
+\usepackage{lineno}
+
+\linenumbers
+\usepackage{csquotes}
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
 \title{Non-word repetition in Yélî Dnye}
-\author{Alejandrina Cristia\textsuperscript{1}~\& Marisa Casillas\textsuperscript{2}}
+
 \date{}
-
-\authornote{All data are made available in a repository in the Open Science Framework.
-AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
-S.
-McDonnell Foundation Understanding Human Cognition Scholar Award.
-
-Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d'Ulm, 75005 Paris, France. E-mail: \href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
 
@@
 Together, these data provide a unique view of online phonological processing in a seldom-studied language, and contribute to both typological and language acquisition research.
-
 }
@@
 \maketitle
+
+\hypertarget{todo-alex}{%
+\subsection{TODO Alex}\label{todo-alex}}
+
+\begin{itemize}
+\tightlist
+\item
+  check again if all fields are explained
+\item
+  set up OSF
+\end{itemize}
 
@@
 \item
-  look through this manuscript for bits of text to be written by \textbf{mc}
-\item
-  double check demo -- probably start with the xlsx version that has the children's first names, date of birth, and date of test.
-I looked again at my local files (still haven't checked the labs' ones) and I'm surprised I have maternal education for more children than I have birth order for\ldots{} If you correct this, to generate the csv I also round the age, assign age.rounded to reported age (in years) when DOB is missing, remove first names, DOB, DOTest and age.precise before pushing to github
-\item
   double check my frequency entries in segments.xlsx using the search function in \url{https://phoible.org/parameters}; take a look at \url{http://phoible.github.io/conventions/} in case you see something about double articulation (I didn't find tp or lBj)
-\item
-  double check also stim\_seg\_freq -- why are some items still with NA freq?
-\item
-  draft discussion
 \end{itemize}
-
-\hypertarget{todo-alex}{%
-\subsection{TODO Alex}\label{todo-alex}}
 
@@
 \item
-  add information about proportion of mispronunciations that are real based on other papers
-\item
-  add effect sizes to all analyses
-\item
-  ? correct treatment of words whose mispronunciation is an English word:
-  mispronunciation meaning
-  361 two two
-  362 two two
-  879 no-pinêt English: no peanut
-  908 nodedi English: no daddy
-  933 no--dadi English: no daddy
-  949 petrumi English: ``true''
-  950 petrumi English: ``true''
-  951 petrumi English: ``true''
-  1373 wee English: where
-  1374 wee English: where
-  1375 wee English: where
-  1379 wee English: where
-  1380 wee English: where
-  1388 wee English: where
-  1394 wee English: where
-  1396 wee English: where
-  1398 wee English: where
-  1399 wee English: where
-  1402 wat English: what
-  1405 wat English: what
-  1411 wee English: where
+  draft discussion
 \end{itemize}
@@
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-However, it has been seldom used outside of Europe and North America (with exceptions including Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)).
+However, it has been seldom used outside of Europe and North America (with exceptions including Gallagher, 2014, and @cristia2020infant), and mostly with languages having phonological inventories that are small to moderate in size (Maddieson, 2005).
 
@@
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/).
-This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /`pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
+A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., \enquote{ndapî} /\enquote{ṇṭæ.pɯ/).
+This process is used frequently with English loan words that have a coda (e.g., 'ponî} (phone) /\enquote{pɔ.nɯ/).
+An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala} (here) /æ.'læ/), but these are limited to /æ/-.
 Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (n.d.) for details).
-
-While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
-In monosyllabic stems, oral vowels (short: 43\% and long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%), with the first syllable of bisyllabic stems following a similar pattern.
-Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels.
-Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see Levinson (n.d.) for details).
 
@@
 
-The basic structure of NWR involves the participant hearing a non-word, like ``bilik'', which they need to repeat back without changing anything that is contrastive in the language.
-For instance, in English, a response of ``bilig'' or ``pilik'' will be scored as incorrect; a response ``biilik'', where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.
+The basic structure of NWR involves the participant hearing a non-word, like \enquote{bilik}, which they need to repeat back without changing anything that is contrastive in the language.
+For instance, in English, a response of \enquote{bilig} or \enquote{pilik} will be scored as incorrect; a response \enquote{biilik}, where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.
 
@@
 
-A first list of candidate items was generated in 2018 by selecting simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'', ``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
+A first list of candidate items was generated in 2018 by selecting simple consonants (\enquote{p}, \enquote{t}, \enquote{d}, \enquote{k}, \enquote{m}, \enquote{ń}, \enquote{w}, \enquote{y})
+and vowels (\enquote{i}, \enquote{o},\enquote{u}, \enquote{a},\enquote{e}) that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
 These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
@@
 These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a headset \textbf{XXX }mc** please fill in** and an Olympus \textbf{XXX }mc** please fill in** from the written form presented together with the same item orally (by the second author).
+Together with the 2018 selection, they were recorded using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) from the written form presented together with the same item orally by the second author.
 The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
-Even so, there was one monosyllable that was often identified as a real word (intended ``yî'' /yXX/; identified as ``yi'' /yi/, \emph{tree}).
+Even so, there was one monosyllable that was often identified as a real word (intended \enquote{yî} /yXX/; identified as \enquote{yi} /yi/, \emph{tree}).
 This item is removed from analyses below.
@@
 
-We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items `spreading' through hearsay.
+We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items \enquote{spreading} through hearsay.
 In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
@@
 
-The child was donned a headset (\textbf{xx }mc** please fill in** for most of the children, SHURE WH20 XLR headset with a dynamic microphone for the rest), recorded into the left channel into a Tascam DR40x digital recorder.
+The child was donned a headset (a Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former), recorded into the left channel into a Tascam DR40x digital recorder.
 For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
@@
 Although previous work does not use distance metrics, we additionally report those.
-For instance, recall our example of an English target being ``bilik'', and imagine a response ``bilig''.
+For instance, recall our example of an English target being \enquote{bilik}, and imagine a response \enquote{bilig}.
 This would receive the following scores: at the whole item level this would receive a score of zero (because the repetition is not exact); at the phoneme level it would receive a score of 80\% (4 out of 5 phonemes repeated exactly); and its normalized phone-based Levenshtein distance is 20\% (20\% of phonemes were substituted or deleted).\\
@@
 
-A total 55 children were tested, from 34 of families, in five hamlets.
+A total 55 children were tested, from 38 of families, in four hamlets.
 Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5).
 In addition, 2 teenagers were tested to put younger children at ease; their data is not included in analyses below.
-The remaining 40 children (14 girls) were aged 6.96 years (range 3.90-11 years).
-Among these, there were 32 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and 2 for whom this information is missing.
-Maternal years of education averaged 8.24 years (range 6-12 years; 2 children had this information missing).\footnote{Precise years of education can be hard to count.
+The remaining 40 children (14 girls) were aged 6.98 years (range 3.92-11.03 years).
+Among these, there were 34 children exposed only to Yélî in the home, 6 children who were also exposed to another language in the home, and 0 for whom this information is missing.
+Maternal years of education averaged 8.22 years (range 6-12 years; 0 children had this information missing).\footnote{Precise years of education can be hard to count.
 This is particularly the case for schooling beyond grade 8, which takes place off-island, which can be due to an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
-We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical ``completion'' times in between.} In terms of birth order, 5 were first borns, 4 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
+We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical \enquote{completion} times in between.} In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig1-first_vs_others-1.pdf}
-\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual participants averaging separately their first attempts and all other attempts.}
+\caption{(\#fig:Fig1-first\_vs\_others)Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts.}
 \end{figure}
 
-We first checked whether NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
+We first checked whether whole-item NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
 We excluded 3 children who did not have data for one of these two types.
 As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
-Surprisingly, NWR scores for subsequent repetitions (M = 40, SD = 26)
-were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p = 0.
+Surprisingly, whole-item NWR scores for subsequent repetitions (M = 40, SD = 26)
+were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p = 0; Cohen's d = 1.14.
 Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
@@
 
-The overall NWR score was M = 65\% (SD = 15\%).
-Scores based on phonemes are even higher M = 78\% (SD = 9\%).
+The overall NWR score was M = 65\% (SD = 15\%), Cohen's d = 4.34.
+Scores based on phonemes are even higher M = 78\% (SD = 9\%), Cohen's d = 8.50.
 The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
@@
 
-We could include 885 observations (\textbf{why -- shouldn't it be 1k??}), from 41 children producing in any given trial one of 41 potential target words.
-The analysis revealed a significant estimate for the average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Age was also a significant predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
+We could include 850 observations, from 40 children producing in any given trial one of 41 potential target words.
+The analysis revealed a main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p = 0): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
+Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r r(31) = 0.63.
 
@@
 
-We next checked whether this may actually be due to frequency of the sounds within the language.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020), and applied the natural logarithm.
-As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Phone frequency estimated from this corpus were correlated with cross-linguistic phone frequencies (r(27)=0.52, =0.00).
-Therefore, we fit another mixed logistic regression, this time declaring as fixed effects scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence is actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.} As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(27)=0.52, =0.00).
+Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = 0.47.
+Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
 As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic frequency (ß = 0.82, SE ß = 0.27, p = 0.00) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of word-level NWR scores, but corpus frequency (ß = 0, SE ß = 0.25, p = 1.00) was not.
+In this model, both cross-linguistic frequency (ß = 0.83, SE ß = 0.27, p = 0.00) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of whole-item NWR scores, but corpus frequency (ß = 0.00, SE ß = 0.25, p = 1.00) was not.
 
-\textbf{use proportion of correctly produced rather than number; DONE}
-
-\textbf{a2m use first attempts rather than the whole data set: having second thoughts.
-To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers \& it is more consistent.
-BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros\ldots{} CF Option 3 further down}
-
-\textbf{a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?}
+\hypertarget{patterns-in-nwr-mispronunciations}{%
+\subsubsection{Patterns in NWR mispronunciations}\label{patterns-in-nwr-mispronunciations}}
 
@@
 Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (110 oral vowels produced with a different quality; 13 nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 115) than vice versa (2).
-Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones, compared to 33 complex ones).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mńlwyvdgptkfhch, 115) than vice versa (2).
+Simple consonants were mispronounced as other simple consonants quite frequently (68 simple consonants mispronounced as other simple ones, compared to 33 complex ones).
 
@@
 Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (110 oral vowels produced with a different quality or about 1.86\% out of all oral vowel targets; 13 nasal vowels produced with a different quality or about 2.06\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 115 times or about 1.96\% out of all complex consonant targets) than vice versa (2 times or about 1.96\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (69 simple consonants mispronounced as other simple ones or about 0.39\% out of all simple consonant targets, compared to 33 complex ones or about 1.02\% out of all complex consonant targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mńlwyvdgptkfhch, 115 times or about 1.96\% out of all complex consonant targets) than vice versa (2 times or about 1.96\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (68 simple consonants mispronounced as other simple ones or about 0.38\% out of all simple consonant targets, compared to 33 complex ones or about 1.02\% out of all complex consonant targets).
 
@@
 Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (205 oral vowels produced with a different quality or about 2.16\% out of all oral vowel targets; 23 nasal vowels produced with a different quality or about 2.36\% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or about 2.43\% out of all complex consonant targets) than vice versa (2 times about 2.43\% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (123 simple consonants mispronounced as other simple ones or about 0.53\% out of all simple consonant targets, compared to 55 complex ones or about 0.98\% out of all complex consonant targets).
+As for consonants, asymmetries were very marked with more complex consonants (specifically dptpkpkmknmbghlv) mispronounced as simple ones (specifically mńlwyvdgptkfhch, 250 times or about 2.43\% out of all complex consonant targets) than vice versa (2 times about 2.43\% out of all simple consonant targets).
+Simple consonants were mispronounced as other simple consonants quite frequently (120 simple consonants mispronounced as other simple ones or about 0.52\% out of all simple consonant targets, compared to 55 complex ones or about 0.98\% out of all complex consonant targets).
 
-Finally, we looked into the frequency with which mispronunciations resulted in real words. Two thirds of them were: 62\%.
+Finally, we looked into the frequency with which mispronunciations resulted in real words.
+Two thirds of them were: 62\%.
+This is a type of analysis that is seldom reported, and we could only find one comparison point: ({\textbf{???}}) found that illiterate adults NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate adults did so in only 1.71\%.
+The percentage we observe here is much higher than this, but we do not know whether age, language, or even test structure explains this difference.
 
@@
 The pattern is less marked when other two scores are used, which are less sensitive to errors.
+Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(1) = -0.93.
+In a generalized binomial mixed model, we included 479 observations, from 40 children producing in any given trial one of 24 potential target words.
+The analysis revealed a main effect of age (ß = 0.56, SE ß = 0.14, p = 0); and a significant estimate for the length of the target words in number of syllables (ß = -0.15, SE ß = 0.33, p = 0.65).
 
@@
 As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (6,496.39) = 0.43, p = 0.00).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (-0.61) = 27.83, p = 0.55), birth order (data missing for 17 children, rho = (2,652.32) = -0.15, p = 0.47) or maternal education (data missing for 2 children, rho = ( 9,022.76) = 0.09, p = 0.60).
+In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (-0.61) = 27.83, p = 0.55), birth order (data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39) or maternal education (data missing for 0 children, rho = ( 10,526.20) = 0.08, p = 0.61).
 
@@
 \end{enumerate}
-
-\textbf{(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)}
 
@@
 
-\leavevmode\hypertarget{ref-meir2017independent}{}%
-Meir, N., \& Armon-Lotem, S.
-(2017).
-Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
-\emph{Frontiers in Psychology}, \emph{8}, 1442.
-
-\leavevmode\hypertarget{ref-meir2016disentangling}{}%
-Meir, N., Walters, J., \& Armon-Lotem, S.
-(2016).
-Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
-\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
-
-\leavevmode\hypertarget{ref-phoible}{}%
-Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
-
-\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
-Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
-(2019).
-An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
-\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
-
-\leavevmode\hypertarget{ref-torrington2015non}{}%
-Torrington Eaton, C., Newman, R.
-S., Ratner, N.
-B., \& Rowe, M.
-L.
-(2015).
-Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
-\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
-
-\leavevmode\hypertarget{ref-vance2005speech}{}%
-Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
-
-\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
-Wilsenach, C.
-(2013).
-Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
-\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
-\url{https://doi.org/10.5785/29-2-554}
-
-\leavevmode\hypertarget{ref-is08042009language}{}%
-Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
-\leavevmode\hypertarget{ref-balladares2016socio}{}%
-Balladares, J., Marshall, C., \& Griffiths, Y.
-(2016).
-Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
-\emph{First Language}, \emph{36}(3), 338--351.
-\url{https://doi.org/10.1177/0142723715626067}
-
-\leavevmode\hypertarget{ref-bergelsoncasillas2019what}{}%
-Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
-S., \& Amatuni, A.
-(2019).
-What do North American babies hear?
-A large-scale cross-corpus analysis.
-\emph{Developmental Science}, \emph{22}(1), e12724.
-\url{https://doi.org/10.1111/desc.12724}
-
-\leavevmode\hypertarget{ref-bowey2001nonword}{}%
-Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
-
-\leavevmode\hypertarget{ref-brandeker2015language}{}%
-Brandeker, M., \& Thordardottir, E.
-(2015).
-Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
-\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
-
-\leavevmode\hypertarget{ref-brown2011cultural}{}%
-Brown, P.
-(2011).
-The cultural organization of attention.
-In A.
-Duranti, E.
-Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language Socialization} (pp.
-29--55).
-Malden, MA: Wiley-Blackwell.
-
-\leavevmode\hypertarget{ref-brown2014interactional}{}%
-Brown, P.
-(2014).
-The interactional context of language learning in Tzeltal.
-In I.
-Arnon, M.
-Casillas, C.
-Kurumada, \& B.
-Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of Eve V.
-Clark} (pp.
-51--82).
-Amsterdam, NL: John Benjamins.
-
-\leavevmode\hypertarget{ref-brownIPchildrearing}{}%
-Brown, P., \& Casillas, M.
-(n.d.).
-Childrearing through social interaction on Rossel Island, PNG.
-In A.
-J.
-Fentiman \& M.
-Goody (Eds.), \emph{Esther Goody revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
-XX--XX).
-New York, NY: Berghahn.
-
-\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
-Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (n.d.). A cross-cultural examination of young children's everyday language experiences.
-
-\leavevmode\hypertarget{ref-casillas2020early}{}%
-Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
-
-\leavevmode\hypertarget{ref-chiat2007preschool}{}%
-Chiat, S., \& Roy, P.
-(2007).
-The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
-
-\leavevmode\hypertarget{ref-cristia2020infant}{}%
-Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
-(2020).
-Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
-\emph{Preprint}.
-
-\leavevmode\hypertarget{ref-estes2007differences}{}%
-Estes, K.
-G., Evans, J.
-L., \& Else-Quest, N.
-M.
-(2007).
-Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
-
-\leavevmode\hypertarget{ref-farmani2018normalization}{}%
-Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
-Z., Talebi, E., \& Shourvazi, Z.
-(2018).
-Normalization of the non-word repetition test in farsi-speaking children.
-\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
-
-\leavevmode\hypertarget{ref-gallagher2014identity}{}%
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
-
-\leavevmode\hypertarget{ref-gallon2007non}{}%
-Gallon, N., Harris, J., \& Van der Lely, H.
-(2007).
-Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
-\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
-
-\leavevmode\hypertarget{ref-havron2019effect}{}%
-Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
-M.-C.
-C.
-S.
-(2019).
-The effect of older siblings on language development as a function of age difference and sex.
-\emph{Psychological Science}, \emph{30}(9), 1333--1343.
-
-\leavevmode\hypertarget{ref-hazan2000development}{}%
-Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
-
-\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
-
-\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
-(2014).
-Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
-\emph{PloS One}, \emph{9}(2), e89544.
-
-\leavevmode\hypertarget{ref-levinsoninprepydgrammar}{}%
-Levinson, S. C. (n.d.). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Cambridge, UK: Cambridge University Press.
-
-\leavevmode\hypertarget{ref-liszkowski2012prelinguistic}{}%
-Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
-(2012).
-A prelinguistic gestural universal of human communication.
-\emph{Cognitive Science}, \emph{36}(4), 698--713.
-\url{https://doi.org/10.1111/j.1551-6709.2011.01228.x}
+\leavevmode\hypertarget{ref-maddieson2005correlating}{}%
+Maddieson, I. (2005). Correlating phonological complexity: Data and validation. \emph{UC Berkeley PhonLab Annual Report}, \emph{1}(1).
 

### 2020-11-24 manuscript.Rmd
minor edits: working through intro (incomplete)

@@
 - double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
-<!--M2A: I think you mentioned you might have someone do this check, but let me know if I misunderstood!-->
-<!-- a2m we then decided we'd both feel more certain if you did... -->
+<!--M2A: Done! Left for AC to review in segments-20201123.xlsx and, if she is satisfied, to integrate :)-->
 
@@
 
-Although infants begin to learn about their native language's phonology within the first year, many studies suggest that in perception and production, in phonetics and phonology, their knowledge continues to develop throughout childhood [e.g., @hazan2000development].
-One common task in this line of research is nonword repetition (NWR).
-In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning.
-The child should immediately try to say it back without changing anything.
-NWR scores are thought to reflect long-term phonological knowledge (which allows the child to perceive the item precisely even though it is not a real word they have encountered before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely even though it had never been pronounced before).
+Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development].
+Much of the evidence for later phonological development comes through the use of nonword repetition (NWR) tasks.
+In a NWR task, participants hear a short sequence of sounds that is phonologically legal but lexically meaningless in the language(s) they are learning.
+After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
+NWR scores are thought to reflect long-term phonological knowledge (which allows the child to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
 
-NWR has been used to seek answers to a variety of theoretical questions, including what  the links between phonology, working memory, and the lexicon are [@bowey2001nonword], and to assess the extent to which apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity].
-It is also frequently used for applications, notably  as a diagnostic for language delays and disorders [@estes2007differences].
+NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity].
+NWR is also frequently used in applied contexts, notablyas a diagnostic for language delays and disorders [@estes2007differences].
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language].
-However, it has been seldom used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant], and mostly with languages having phonological inventories that are small to moderate in size [@maddieson2005correlating].
+However, it has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant], and the languages it has been used with have had small to moderate phonological inventories [@maddieson2005correlating].
+<!--M2A: Is the motivating idea here that application to YD is of interest because it's very different from what we've look at with NWR in the past?
+Or is there a specific hypothesis you had in mind about how inventory size affects phonological development and, hence, NWR performance?
+I could come up with something, but I think you had an idea already and I don't want to step on it!-->
 
-<!--M2A: I feel like there's a bit of a jump here from explaining what NWR is and what it's used for to saying we used it in the current study to test three things.
-We should probably have a paragraph justifying its use in this context, no?-->
-<!-- a2m I added a sentence, is this what you were thinking of? -->
-<!-- M2A: Makes sense.
-I was thinking we could already hint at YD as interesting in particular bc of the phonology and not simply because it's a random one of many previously untested populations.
-I don't mind adding something like this in once I'm done with the demo check.-->
-<!-- a2m added ", and mostly with languages having phonological inventories that are small to moderate in size" + notice the " with an unusually dense phonological inventory"; my feeling is that it may interrupt the flow to add a lot more here...
-don't you think?
--->
-
-In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
-The present report uses these unique data to contribute to several lines of research.
-First, we  made sure that some of the non-word items contained typologically rare and/or  challenging sounds, so that we can contribute to the study of whether rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
-Second, we look at the effects of item length, since previous NWR research has uncovered variable effects.
-Third, we contribute to the basic NWR literature by adding a data point for children learning a rarely-studied language and culture, and further investigating whether there are structured sources that account for individual variation.
+In the present study, we report on NWR results among children learning Yélî Dnye, an isolate language spoken on Rossel Island, PNG, with a large and unusually dense phonological inventory.
+The study was designed to contribute to several lines of research: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
+Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation, following variable effects of length in previous NWR research.
+Third, these data give a first impression of how this task is responded to in a rarely-studied language and culture; we add to this by further investigating whether there are structured sources that account for individual variation in our sample.
 

### 2020-11-24 manuscript.Rmd
more minor intro changes

@@
 NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
-In this study, we report on NWR results among children learning Yêly Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense phonological inventory.
+In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea, with an unusually dense phonological inventory.
 This study contributes to three lines of research.
@@
 <!--M2A: Done! Left for AC to review in segments-20201123.xlsx and, if she is satisfied, to integrate :)-->
-
 - draft discussion
@@
 Much of the evidence for later phonological development comes through the use of nonword repetition (NWR) tasks.
-In a NWR task, participants hear a short sequence of sounds that is phonologically legal but lexically meaningless in the language(s) they are learning.
+In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
 After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
-NWR scores are thought to reflect long-term phonological knowledge (which allows the child to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
+NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
 
 NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity].
-NWR is also frequently used in applied contexts, notablyas a diagnostic for language delays and disorders [@estes2007differences].
+NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences].
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language].
-However, it has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant], and the languages it has been used with have had small to moderate phonological inventories [@maddieson2005correlating].
+However, it has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant].
+Further, the languages it has been used with have historically had small to moderate phonological inventories [@maddieson2005correlating].
 <!--M2A: Is the motivating idea here that application to YD is of interest because it's very different from what we've look at with NWR in the past?
@@
 
-In the present study, we report on NWR results among children learning Yélî Dnye, an isolate language spoken on Rossel Island, PNG, with a large and unusually dense phonological inventory.
-The study was designed to contribute to several lines of research: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
-Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation, following variable effects of length in previous NWR research.
+In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
+The study was designed to contribute to multiple aspects of our understanding of phonological development: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
+Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation; previous NWR research has shown variable effects of length between populations.
 Third, these data give a first impression of how this task is responded to in a rarely-studied language and culture; we add to this by further investigating whether there are structured sources that account for individual variation in our sample.
+<!--M2A: This is also a note to myself; it seems we could use some more references in the above!-->
+
+Before jumping into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
+Since NWR has been almost exclusively used in urban, industrialized populations [see @cristia2020infant for an exception], we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.
 
@@
 
-Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
-The Yélî sound system, much like its baroque grammatical system [@levinsoninprepydgrammar], is unlike any other in the region.
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region.
 
 With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these able to be contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized or palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsoninprepydgrammar for details).
-The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).^[Here we use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+In approximating cross-linguistic segment frequency below we therefore decided to use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).] The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
 Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with 7 of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
 
-In total, Yélî Dnye uses 90 distinctive segments (93 when including consonants that are extremely rarely used), far outstripping the phonemic inventory size of other documented Papuan languages (@levinsoninprepydgrammar; Foley 1986) and includes at least two contrasts not yet been documented elsewhere [labial-coronal double-articulations with dental vs.
-post-alveolar coronal placement in both oral and nasal stops, Maddieson & Levinson 2001; Ladefoged and Maddieson 1996].
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar] and includes at least two contrasts not yet been documented elsewhere [labial-coronal double-articulations with dental vs.
+post-alveolar coronal placement for both oral and nasal stops, @ladefoged1996sounds; @maddiesonIPphoneticsYD].
 
-Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the current Yélî Dnye dictionary, @levinsoninprepydgrammar].
+Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at time of writing, @levinsonYDgrammar].
 The vast majority of syllables use a CV format.
@@
 An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
-Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsoninprepydgrammar for details).
+Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
+<!--
+The following was removed b/c it's less relevant, but it has been double-checked by MC Nov 2020 in case we decide to include it later:
 
-<!-- a2m up to here it looks great.
-i am less sure about the next paragraph because it is unclear how it connects with our study...
-we didn't really take this into account when building the stimuli, right?
--->
-<!-- M2A: Happy to comment it out I agree it's not really relevant for our study in particular; just part of the phonological sketch.
-I was thinking about linguists when I wrote it up, but psychies probably are just ready to get on with the next part by now :) -->
-
-<!--
 While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
@@
 Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels.
-Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see @levinsoninprepydgrammar for details).
+Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see @levinsonYDgrammar for details).
 -->
@@
 
-The current study also contributes to increasing the cultural diversity of the populations for which NWR data have been collected.
-As mentioned previously, this method has been used almost exclusively in urban, industrialized populations, with only one exception to our knowledge [@cristia2020infant].
-We therefore provide some background information tol help understand some aspects of our study, including the estimation of children's ages and the relationship between NWR scores and children's demographic characteristics.
-
-Most speakers of the language grow up speaking it monolingually until they begin attending school around the age of 7; 
-school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---, where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
-Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
+Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or another language(s) from the region.
 
@@
 
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by barrier reef, transport to and from the island is both infrequent and irregular.
-There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport, when there is deisel available.
-There is a medical radio connection at the health clinic (at the Catholic mission) via which news and messages are often passed along.
-There is also a cellular tower on the island, though it is typically not functioning.
-As such, access to news and outside resources, particularly international connections and larger data transfers, is typically unavailable.
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
+There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that, even when there is deisel available, are primarily reserved for medical transport.
+There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG.
+There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time.
+As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, is typically unavailable.
 Our data collection is therefore limited to the duration of the researchers' on-island visits.
@@
 A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early].
-While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), and so it is very common for households to billet school children during the weekdays for months at a time (or for whole months if the child's home is very far).
-Combined with practices of collective child guardianship within their close-knit hamlets, the practical outcome for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's parents---to this child assent is also culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[Note that in a few cases this does mean that precise details such as child date of birth, parent age in years, and parent education was only able to be approximated as the parents were residing on another part of the island.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
+While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day).
+It is very common for households situated close to schools to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year.
+Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
 
-While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., Bram's thesis; MC's CFH data].
-What we do know about the early language environment is that, while child-directed speech is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early].
+While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., @peuteIPconsonants].
+What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early].
 While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early].
-This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 people) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
 
@@
 Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described [@levinsoninprepydgrammar], there could have been undocumented constraints that rendered items illegal.
+Since the phonology and phonetics of Yélî is still in the process of being described [@levinsonYDgrammar], there could have been undocumented constraints that rendered items illegal.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).

### 2020-11-30 manuscript.Rmd
minor edits to methods (to be continued)

@@
 
-### Yélî Dnye phonological overview
+### Yélî Dnye phonology
 
@@
 With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).^[Here we use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
-In approximating cross-linguistic segment frequency below we therefore decided to use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).] The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).^[We use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).] The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
 Vowels in Yélî Dnye may be oral or nasal, short or long.
@@
 
-### Brief review of NWR for our purposes
+### NWR design and analysis adaptations
 
-The basic structure of NWR involves the participant hearing a non-word, like "bilik", which they need to repeat back without changing anything that is contrastive in the language.
-For instance, in English, a response of "bilig" or "pilik" will be scored as incorrect; a response "biilik", where the vowel is lengthened without change of quality would be scored as correct, because English does not have a pure length contrast.
+In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
+For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biilik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length.
+There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
+For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language].
+Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items [@piazzalunga2019articulatory].
+Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
 
-Aside from this central basic structure, there is some variation in the presentation procedure and structure of items found in previous NWR work.
-For example, items are often presented orally by the experimenter [@torrington2015non], although an increasing number of studies have turned to playing back the stimuli in order to have greater control of the stability of the presentation [@brandeker2015language].
-Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items [@piazzalunga2019articulatory].
-Often, authors modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
+Previous work typically steers clear of articulartorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology.
+We ultimately used a relatively large number of items that would enable us to explor both variation in structural complexity and in more vs.
+less challenging sounds.
+However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
+Naturally, designing the task in this way may render the study of individual variation within the population more difficult because different children are exposed to different items.
+However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit.
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
 
-Previous work seems to avoid difficult sounds, but we felt this was important to represent Yélî Dnye, so we also varied this factor.
-We designed a relatively large number of items but, aware that this may render the task longer and more tiresome, we split some of the items across children.
-This allowed us to get NWR scores for more items.
-
-Naturally, designing the task in this way may render the study of individual variation within the population  more difficult because different children are exposed to different items.
-However, a review of previous work on individual variation suggested to us that effects of individual differences  are relatively small, and would not be detectable with the sample size that we could collect in a given visit.
-
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
-Based on previous work, we looked at potential improvements with age [@vance2005speech;@farmani2018normalization;@kalnak2014nonword], and potential negative effects of bilingual exposure [@brandeker2015language;@meir2016disentangling;@meir2017independent].
-Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization;@kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool].
-Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on those too.
+Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech], and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent].
+Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool].
+Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.
 
@@
 
-- Does the frequency of sounds across languages predict NWR? Are rarer sounds more often substituted by commoner sounds?
+- Does the cross-linguistic frequency of sounds predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
 - How does score change as a function of item length in number of syllables?
@@
 In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we do not have enough power to look at all relationships of interest, and often to detect any of the known effects given their variability across studies.
-To illustrate this, we portray studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables) in Figure 1.
+However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies.
+To illustrate this issue, we portray in Figure 1 studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
 Notice that the effect of stimulus length is minuscule among Italian children [@piazzalunga2019articulatory], but considerable among Tsimane' children  [@cristia2020infant], where a drop of 40 percentage points is observed at all ages.
-A similar difference in NWR scores for short versus long items was observed among Arabic children [@jabere2018xperiment].
-Even the effect of age is unstable in this sample.
+A similar large difference in NWR scores for short versus long items was observed among Arabic children [@jabere2018xperiment].
+Even the effect of age is unstable in this sample of studies.
 Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
-Therefore, all analyses here are descriptive and should be considered exploratory.
+Therefore, all analyses in the present study are descriptive and should be considered exploratory.
 
@@
 
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological].
-We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in Yély Dnye.
-We also increased the number of items an individual child would be tested on, so that a child would get up to 23 items to repeat [note that up to other work has also used 24-30 items: @jabere2018xperiment; @kalnak2014nonword], and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological].
+We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 44 items distributed across children.
 
-A first list of candidate items was generated in 2018 by selecting simple consonants ("p",	"t",	"d",	"k",	"m",	"ń",	"w",	"y") 
-and vowels ("i", "o","u", "a","e") that were combined into consonant-vowel syllables, further sampling the space of 1- to 4-syllable sequences.
-These candidates were automatically checked against Levinson's 2015 dictionary and removed from consideration if they appeared in the dictionary.
-The second author presented them orally to three local research assistants, who were asked to repeat them and further say whether they were real words.
-Any item for which two or more of the assistants reported them having a meaning or some form of association was excluded.
+A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 1- to 4-syllable sequences.
+These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
+The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who were repeated each form, as they would in an NWR task, and additinally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
+Any item reported to have a meaning or a strong association with another word form or meaning was excluded.
 
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî inventory to produce consonant-vowel monosyllables.
+A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms.
 As before, items were automatically excluded if they appeared in the dictionary.
-Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out.
-Since the phonology and phonetics of Yélî is still in the process of being described [@levinsonYDgrammar], there could have been undocumented constraints that rendered items illegal.
+Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
+Because there is still much to discover about the phonology and phonetics of Yélî Dnye [@levinsonYDgrammar], it was also possible that we initially generated items with illegal, but currently undocumented constraints.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
-These candidates were presented to one informant, for a final check that they did not mean anything.
-Together with the 2018 selection, they were recorded using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) from the written form presented together with the same item orally by the second author.
-The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.
-Even so, there was one monosyllable that was often identified as a real word (intended "yî" /yXX/; identified as "yi" /yi/, *tree*).
-This item is removed from analyses below.
+These candidates were then presented to one informant, for a final check that they did not mean anything.
+Together with the 2018 selection, they were recorded, based on their orthographic forms, using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) by the same informant, monitored by the second author for clear production of the phonological target.
+The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present.
+Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended "yî" /yɯ/; identified as "yi" /yi/, *tree*).
+This item is removed from the analyses below.
 
-The final list is composed of three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
+The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
 
@@
 
-- the same three items were selected as practice items and used in all 40 elicitation sets
-- splits were done within each length group from the 2018 items (i.e., separately for 2, 3, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with tp were split into 2 sublists).
-Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 versus 23).
-- once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllables, then 3-syllables, and finally 4-syllable items.
+- The same three items were selected as practice items and used in all 40 elicitation sets
+- Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sublists).
+Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 vs.
+23).
+- Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1-, 2-, and 4-syllable), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
 

### 2020-12-01 manuscript.Rmd
continued methods edits

@@
 
-A Praat script was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
+A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
 The 40 elicitation sets are available online from https://osf.io/5qspb.
@@
 
-We tried to balance three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them; that children would feel comfortable doing this task with us; and that the community would feel safe with us doing this task with their children.
-Moreover, there were also some logistic constraints in terms of the space availability.
-As a result, the places where elicitation happened varied across the hamlets.
+In adapting the typical NWR procedure for this context, we balanced three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them (i.e., from other children who had participated); that children would feel comfortable doing this task with us; and that community members would feel comfortable having their children do this task with us.
 
-We visited four different hamlets once, and attempted to test all eligible children present at the time, to prevent the items 'spreading' through hearsay.
+We ran children in batches, testing within a handful of hamlet clusters spread across the northeastern region of the island.
+Because space availability was limited in different ways from hamlet to hamlet, the places where elicitation happened varied across testing sites.
+We tested in four different sites, only making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items 'spreading' between children through hearsay.
 In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
 The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
+Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitaiton sets.
 
-The child was donned a headset (a Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former), recorded into the left channel into a Tascam DR40x digital recorder.
-For most children, the headset could not stay comfortably on the child's head, and thus it was placed on the child's shoulders, with the microphone carefully placed close to the child's mouth.
-A local informant sat next to the child, to would provide the instructions and, if needed, coach the child to make sure, using the three practice items as well as real words, that they understood that the task was to repeat the items precisely without changing anything.
-An experimenter (the first author) delivered the elicitation stimuli to the local informant and the child over headphones.
+We fitted the child with a headset microphone (Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former) that fed into the left channel of a Tascam DR40x digital audio recorder.
+The headsets were designed for adult use and could not be comfortably seated on many children's heads without a more involved adjustment period.
+To minimize adjustment time, which was uncomfortable for some children given the proximity of the experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth.
+A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement.
+The research assistant coached the child throughout to task to make sure that they understood what they were expected to do.
+An experimenter (the first author) delivered the pre-recorded stimuli to the research assistant and the child over headphones.
 
-The first phase was making sure the child understood the task.
-This was explained orally and the first training item was presented.
-Often, children froze and did not say anything.
-If this happened, then we followed this procedure.
-First, the informant insisted.
-If the child still did not say anything, the informant asked the child to repeat a real word, and another, and another.
-If the child could repeat these correctly, then we provided the recorded training item over headphones again.
-Most children successfully started repeating the items presented over headphones at this point; a few further needed the local informant to model the behavior (i.e., they would hear the item again, and she would say it; then we would play it again, and ask the child to say it).
-A small minority still failed to repeat the item after hearing it over headphones.
-If that occurred, we tried with the second training item, at which point some children got it and could continue.
-A small minority, however, failed to repeat this one, as well as the third training item, in which case we stopped the test altogether (see Participants section for exclusions).
+The first phase of the experiment involved making sure the child understood the task.
+We explained the task orally presented the first practice item.
+At this point, many children did not say anything in response, at which point we initiated the following procedure: First, the assistant insisted the child make a response.
+If the child still did not say anything, the assistant said a real word and then asked the child to repeat it, then another and another.
+If the child could repeat real words correctly, we provided the first training item over headphones again for children to repeat.
+Most children successfully started repeating the items at this point, but a few needed further help.
+In this case, the assistant modeled the behavior (i.e., the child and assistant would hear the item again, and the assistant would repeat it; then we would play the item again and ask the child to repeate it).
+A small minority of children still failed to repeat the item at this point.
+If so, we tried again with the second training item, at which point some children got it and could continue.
+A fraction of the remaining children, however, failed to repeat this second training item, as well as the third one, in which case we stopped testing altogether (see Participants section for exclusions).
 
-NWR studies vary in whether children are provided with several opportunities to hear and say the item.
-To have a fixed and clear procedure, we decided that items other than the inital three training ones would not be repeated unless the child made an attempt to produce them.
-If this attempt was judged correct by the local informant, then the experimenter would move on to the next item (whispering this over a separate headset that was recording onto the right track of the same Tascam).
-If the local informant heard a deviation, she indicated to the experimenter that the item needed to be repeated, and up to 5 attempts were allowed.
+The second phase of the experiment involved running the child through the list of test items randomly assigned to them.
+This was done in the same manner as the practice items: the stimulus was played over the headphones, and then the child repeated it aloud.
+NWR studies vary in whether children are allowed to hear and/or repeat the item more than one time.
+We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was alowed to make a further attempts if their first attempt was judged erroneous in some way by the assistant.
+The procedure worked as follows: When the child made an attempted a repeat, the assistant indicated to the experimenter whether the child's production was correct or not.
+If correct, the experimenter would whisper this note of correct repetition into a separate headset that fed into the right channel of the same Tascam and the group moved on to the next item.
+If not, the child was allowed to try again, with up to five attempts allowed before moving on to the next item.
+Children were not asked to make repetitions if they did not produce a first attempt.
 
-Whenever siblings from the same family were tested, an attempt was made to test first the older and then the younger child, and always on different elicitaiton sets.
+In total, test sessions took approximately XX minutes on average, with XX of those minutes attributed to practice and XX to the actual test list.
+<!--M2A: Alex, could you please estimate the above XX cases and fill in?-->
 
@@
 
-A script was written to randomize all tokens from all children, pairing each with the auditory target the child had been provided with.
-A native research assistant then listened through all productions of a given target (randomized across children and repetition order), and made a judgment of whether the item was correctly or incorrectly repeated.
-She additionally orthographically transcribed exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets).
+The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat],
+<!--M2A: Alex, I added a bib entry for Praat but it's the most recent version. Could you please correct it to the version (approximately if not precisely) that you used?-->
+then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files.
+The assistant then listened through all these productions, one target item at a time, with productions for each target item presented in a random order across children and repetitions.
+The assistant indicated in a notebook whether each production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English.
+The assistant also provided some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
 
@@
 
-Some NWR studies employ phoneme-based scoring in addition to or instead of word-level scoring [e.g., @cristia2020infant].
-We scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt).
-Although previous work does not use distance metrics, we additionally report those.
-For instance, recall our example of an English target being "bilik", and imagine a response "bilig".
-This would receive the following scores: at the whole item level this would receive a score of zero (because the repetition is not exact); at the phoneme level it would receive a score of 80% (4 out of 5 phonemes repeated exactly); and its normalized phone-based Levenshtein distance is 20% (20% of phonemes were substituted or deleted).
+In addition to a binary word-level exact repetition score we scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer [the target or the attempt; see also @cristia2020infant].
+Although previous work does not use distance metrics, we additionally report these.
+For instance, recall our example of an English target being /bilik/ with an imagined response [bilig].
+We would score this response as follows: at the whole item level this production would receive a score of zero (because the repetition is not exact); at the phoneme level this production would receive a score of 80% (4 out of 5 phonemes repeated exactly); and the phone-based Levenshtein distance for this production is 20% (because 20% of phonemes were substituted or deleted).
 <!--M2A: A concrete scoring example would be a good idea here-->
@@
 2) LD has the advantage of making D/I/S relevant as a next step in the analysis
-3) we should definitely choose one, I agree, and report the other in supp mat tables
--->
+3) we should definitely choose one, I agree, and report the other in supp mat tables-->
 
-Additionally, we looked up each phoneme in each target word in the Phoible database [@phoible].
-We entered the number of languages in which each phoneme was found, as well as what percentage of languages in the database that represented.
-We could not find a small number of phonemes in the database, and treat those as NA values that do not contribute to the target-level global average phone cross-linguistic frequency, although an alternative would be to attribute them a very small number.
+<!--M2A: So what's the final determination here, keeping in mind we're now aiming for LDR?-->
 
-Finally, for describing children's patterns of errors, all repetitions of a given target were taken into account.
+Additionally, we estimated the typological frequency of all phonological segments used in the target items using the PHOIBLE cross-linguistic phonological inventory database [@phoible].
+For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time these data were analyzed, it is of course still far from complete, so estimated frequencies should be taken with a pinch of salt.
+In both PHOIBLE and UPSID---another large phonemic inventory database---note that nearly half of the segment types only appear in a single language (Steven Moran, personal communication; see also [web.phonetik.uni-frankfurt.de/upsid_info.html](http://web.phonetik.uni-frankfurt.de/upsid_info.html)).
+It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database.
+Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat our "missing" phones as having a frequency of 1, with a (rounded) percentage of 0%.
+
+Finally, when describing children's patterns of errors, we take all repetitions of a given target into account.
 We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.

### 2020-12-15 manuscript.Rmd
minor edits to methods text

@@
 
-This study was approved as part of a larger research effort by Marisa Casillas, with the research protocol having been approved by the IRB committee entitled The Ethics Committee of the Faculty of Social Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) of Radboud University, Nijmegen, The Netherlands.
-The approval for work outside of the Netherlands came as an amendment to the general research line on language development research (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
-Participation was voluntary, with children being invited to come and participate.
-Regardless of whether they did any of the task, children were provided with a snack as compensation.
-Children who came up to participate but then refused were nonetheless provided with the snack.
+This study was approved as part of a larger research effort by the second author.
+The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
+Regardless of whether they completed the task, children were given a small snack as compensation.
+Children who showed initial interest but then decided not to participate were also given the snack.
 
@@
 
-A total `r dim(demo)[1]` children were tested, from `r length(levels(factor(demo$familyID)))` of families, in four hamlets.
-Some children could not be included for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`).
-In addition, `r sum(demo$Age >12 | demo$included=="too old",na.rm=T)` teenagers were tested to put younger children at ease; their data is not included in analyses below.
+We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlets.
+We excluded test sessions from analysis for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease).
 The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years).
-Among these, there were `r sum(inc$monolingual=="yes", na.rm=T)` children exposed only to Yélî in the home, `r sum(inc$monolingual=="no", na.rm=T)` children who were also exposed to another language in the home, and `r sum(is.na(inc$monolingual))` for whom this information is missing.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years; `r sum(is.na(inc$matEd))` children had this information missing).[^1] In terms of birth order, `r birthorder.tab[1]` were first borns, `r birthorder.tab[2]` second,  `r birthorder.tab[3]` third,  `r birthorder.tab[4]` forth,  `r birthorder.tab[5]` fifth, `r birthorder.tab[6]` sixth, and  `r birthorder.tab[7]` did not have this information.
+Among these children were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to one or more other languages at home.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^1] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth.
 
-[^1]: Precise years of education can be hard to count.
-This is particularly the case for schooling beyond grade 8, which takes place off-island, which can be due to an irregularity in ability to pay school fees, school opening times, familial obligations, and more.
-We asked parents to report their highest completed level of education and inferred the approximate number of years that entails under ideal conditions, which results in many esimates around typical "completion" times in between.
+[^1]: Years of education is difficult to precisely pinpoint in this context.
+Schooling beyond grade 8 takes place off-island and enrollment is tied to an ability to pay that year's school fees, seasonal interruptions to school opening times, familial obligations, and more.
+We asked parents to report their highest completed level of education.
+We then record the number of years entailed by having completed that level under ideal conditions.
+This method results in many esimates around typical "graduation" times (e.g., graduating from primary education before starting secondary education).
 

### 2020-12-15 manuscript.Rmd
small edits to results

@@
 
-```{r Fig1-first_vs_others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts."}
+```{r Fig1-first-vs-others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts."}
 
@@
 
-plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="NWR scores (% words)")
+plot(pwcm_att$Word~myx,pch=20,cex=2,lty=3,xlab="",xaxt="n",ylab="% words correctly repeated")
 for(id in levels(factor(pwcm_att$ID))) lines(myx[pwcm_att$ID==id],pwcm_att$Word[pwcm_att$ID==id],lty=2)
-axis(1,at=1:2,labels=c("first","others"))
+axis(1,at=1:2,labels=c("First","Subsequent"))
 
@@
 We excluded `r excl` children who did not have data for one of these two types.
-As shown in Figure 1, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
+As shown in Figure \@ref(fig:Fig1-first-vs-others), participants' mean word-level scores became more heterogeneous in subsequent repetitions.
 Surprisingly, whole-item NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
-were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`.
-Given the uncertainty in whether previous work used only the first or all repetitions, and since scores declined and became more heterogeneous in subsequent repetitions, the rest of the analyses focus on only the first repetitions.
+were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`).
+Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions.
 

### 2020-12-15 manuscript.Rmd
create trnsc.inc to replace trnsc in main analyses
trnsc maintained two participants we had decided to exclude (c105 and c78)

@@
 trnsc$birthOrder=as.numeric(trnsc$birthOrder)
+
+# Remove excluded children from trnsc
+trnsc[which(trnsc$id %in% inc$ID),]->trnsc.inc
+
 ```
@@
 
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$age.rounded,trnsc$attempt),mean)->pwcm_att
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$age.rounded,trnsc.inc$attempt),mean)->pwcm_att
 colnames(pwcm_att)<-c("ID","Age","Attempt","Word")
@@
 # which is
-excl=sum(table(pwcm_att$ID)!=2)
+excl=sum(table(pwcm_att$ID)==1)
 
@@
 ```{r exclude subsequent}
-trnsc=trnsc[trnsc$attempt=="first",]
+trnsc.inc=trnsc.inc[trnsc.inc$attempt=="first",]
 
@@
 ```{r overall}
-aggregate(trnsc$correct*100,by=list(trnsc$id),mean)->overall
-aggregate(trnsc$phon_score*100,by=list(trnsc$id),mean)->phon_score
-aggregate(trnsc$nld*100,by=list(trnsc$id),mean)->nld
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id),mean)->overall
+aggregate(trnsc.inc$phon_score*100,by=list(trnsc.inc$id),mean)->phon_score
+aggregate(trnsc.inc$nld*100,by=list(trnsc.inc$id),mean)->nld
 overall=cbind(overall,phon_score$x,nld$x)
@@
 Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%), Cohen's d = `r cohensd(mean(overall$Phoneme),sd(overall$Phoneme))`.
-The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted.
-Notice that the normalized Levenshtein distance is the complement of phoneme-based scores.
+The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted (the normalized Levenshtein distance is the complement of phoneme-based scores).
 
@@
 ```{r xling-freq-ana}
-trnsc$avg_fr_s=scale(trnsc$avg_fr)
-trnsc$age.rounded_s=scale(trnsc$age.rounded)
-xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
+trnsc.inc$avg_fr_s=scale(trnsc.inc$avg_fr)
+trnsc.inc$age.rounded_s=scale(trnsc.inc$age.rounded)
+xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
 mycoeff=round(summary(xl_fr)$coefficients,3)
@@
 
-aggregate(trnsc$correct*100,by=list(trnsc$avg_fr_s),mean)->xl_fr_tab
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$avg_fr_s),mean)->xl_fr_tab
 colnames(xl_fr_tab)<-c("xlf","Word")
@@
 ```{r Fig2-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
-aggregate(trnsc$correct*100,by=list(trnsc$target,trnsc$avg_fr),mean)->xling
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,trnsc.inc$avg_fr),mean)->xling
 colnames(xling)<-c("Target","Avg_fr","Word")
@@
 ```{r cor vs xl}
-#fr_cor=glmer(correct~avg_fr_cor+age.rounded+(1|file) + (1|target),data=trnsc,family="binomial")
+#fr_cor=glmer(correct~avg_fr_cor+age.rounded+(1|file) + (1|target),data=trnsc.inc,family="binomial")
 #fr is not sig, but close (p=.1, estimate is 17)
-#trnsc$avg_fr_cor_s=scale(trnsc$avg_fr_cor)
-#both_fr=glmer(correct~avg_fr_s+avg_fr_cor_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
+#trnsc.inc$avg_fr_cor_s=scale(trnsc.inc$avg_fr_cor)
+#both_fr=glmer(correct~avg_fr_s+avg_fr_cor_s+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
 #both_coeff=round(summary(both_fr)$coefficients,3)
 
-#fr_cor_ty=glmer(correct~avg_fr_cor_ty+age.rounded+(1|file) + (1|target),data=trnsc,family="binomial")
+#fr_cor_ty=glmer(correct~avg_fr_cor_ty+age.rounded+(1|file) + (1|target),data=trnsc.inc,family="binomial")
 #fr is sig
-trnsc$avg_fr_cor_ty_s=scale(trnsc$avg_fr_cor_ty)
-both_fr=glmer(correct~avg_fr_s+avg_fr_cor_ty_s+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial")
+trnsc.inc$avg_fr_cor_ty_s=scale(trnsc.inc$avg_fr_cor_ty)
+both_fr=glmer(correct~avg_fr_s+avg_fr_cor_ty_s+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
 both_coeff=round(summary(both_fr)$coefficients,3)
 
-aggregate(trnsc$correct*100,by=list(trnsc$avg_fr_cor_ty_s),mean)->cor_fr_tab
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$avg_fr_cor_ty_s),mean)->cor_fr_tab
 colnames(cor_fr_tab)<-c("corf","Word")
@@
 #remove English meanings
-trnsc$meaning[grep("English",trnsc$meaning)]<-NA
+trnsc.inc$meaning[grep("English",trnsc.inc$meaning)]<-NA
 
@@
 Finally, we looked into the frequency with which mispronunciations resulted in real words.
-Two thirds of them were: `r round(sum(!is.na(trnsc$meaning) & trnsc$meaning != "0")/sum(!is.na(trnsc$mispronunciation))*100)`%.
+Two thirds of them were: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%.
 This is a type of analysis that is seldom reported, and we could only find one comparison point: @castro1998 found that illiterate adults NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate adults did so in only `r round(2/117*100,2)`%.
@@
 
-aggregate(trnsc$correct*100,by=list(trnsc$id,trnsc$type),mean)->by_length
-aggregate(trnsc$phon_score*100,by=list(trnsc$id,trnsc$type),mean)->phon_score
-aggregate(trnsc$nld*100,by=list(trnsc$id,trnsc$type),mean)->nld
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type),mean)->by_length
+aggregate(trnsc.inc$phon_score*100,by=list(trnsc.inc$id,trnsc.inc$type),mean)->phon_score
+aggregate(trnsc.inc$nld*100,by=list(trnsc.inc$id,trnsc.inc$type),mean)->nld
 by_length=cbind(by_length,phon_score$x,nld$x)
@@
 
-aggregate(trnsc$correct*100,by=list(trnsc$type),mean)->by_length_no_child
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$type),mean)->by_length_no_child
 colnames(by_length_no_child)<-c("Length","Word")
@@
 
-len_lm=glmer(correct~type+age.rounded_s+(1|file) + (1|target),data=trnsc,family="binomial",subset=c(trnsc$type!=1))
+len_lm=glmer(correct~type+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial",subset=c(trnsc.inc$type!=1))
 mycoeff=round(summary(len_lm)$coefficients,3)

### 2020-12-15 manuscript.tex
reknit for formatting check
Note: reference castro1998 missing

--- 2020-09-17 manuscript.tex
@@
 memory and flexible production patterns. In this study, we report on NWR
-results among children learning Yêly Dnyé, an isolate spoken in Rossel
-Island, PNG, with an unusually dense phonological inventory. This study
-contributes to three lines of research. First, we document that non-word
-items containing typologically rare sounds are repeated without changes
-less often that non-words containing more common sounds. Second, we
-document rather weak effects of item length, contributing to other
-research suggesting that length effects may be language-specific. Third,
-we do not find strong individual variation effects in this population,
-contrary to previous results documenting strong age-related effects.
-Together, these data provide a unique view of online phonological
-processing in a seldom-studied language, and contribute to both
-typological and language acquisition research.
+results among children learning Yélî Dnye, an isolate spoken on Rossel
+Island in Papua New Guinea, with an unusually dense phonological
+inventory. This study contributes to three lines of research. First, we
+document that non-word items containing typologically rare sounds are
+repeated without changes less often that non-words containing more
+common sounds. Second, we document rather weak effects of item length,
+contributing to other research suggesting that length effects may be
+language-specific. Third, we do not find strong individual variation
+effects in this population, contrary to previous results documenting
+strong age-related effects. Together, these data provide a unique view
+of online phonological processing in a seldom-studied language, and
+contribute to both typological and language acquisition research.
 
@@
 
-\subsection{TODO Middy}\label{todo-middy}
+\subsection{TODO Alex}\label{todo-alex}
 
@@
 \item
-  double check demo -- probably start with the xlsx version that has the
-  children's first names, date of birth, and date of test. I looked
-  again at my local files (still haven't checked the labs' ones) and I'm
-  surprised I have maternal education for more children than I have
-  birth order for\ldots{} If you correct this, to generate the csv I
-  also round the age, assign age.rounded to reported age (in years) when
-  DOB is missing, remove first names, DOB, DOTest and age.precise before
-  pushing to github
+  check again if all fields are explained
+\item
+  set up OSF (or revise if already there)
+\item
+  implement decisions on prop phones \& patterns nwr mp
+\item
+  revise bullet points for discussion
+\end{itemize}
+
+\subsection{TODO Middy}\label{todo-middy}
+
+\begin{itemize}
+\tightlist
 \item
@@
 \item
-  double check also stim\_seg\_freq -- why are some items still with NA
-  freq?
-\item
   draft discussion
@@
 
-\subsection{TODO Alex}\label{todo-alex}
-
-\begin{itemize}
-\tightlist
-\item
-  add information about proportion of mispronunciations that are real
-  based on other papers
-\item
-  add effect sizes to all analyses
-\item
-  ? correct treatment of words whose mispronunciation is an English
-  word: mispronunciation meaning 361 two two 362 two two 879 no-pinêt
-  English: no peanut 908 nodedi English: no daddy 933 no--dadi English:
-  no daddy 949 petrumi English: ``true'' 950 petrumi English: ``true''
-  951 petrumi English: ``true'' 1373 wee English: where 1374 wee
-  English: where 1375 wee English: where 1379 wee English: where 1380
-  wee English: where 1388 wee English: where 1394 wee English: where
-  1396 wee English: where 1398 wee English: where 1399 wee English:
-  where 1402 wat English: what 1405 wat English: what 1411 wee English:
-  where
-\end{itemize}
-
 \subsection{Introduction}\label{introduction}
 
-Although infants begin to learn about their native language's phonology
-within the first year, many studies suggest that in perception and
-production, in phonetics and phonology, their knowledge continues to
-develop throughout childhood (e.g., Hazan \& Barrett, 2000). One common
-task in this line of research is nonword repetition (NWR). In NWR
-studies, participants are presented auditorily with an item that is
-phonologically legal but lexically meaningless in the language children
-are learning. The child should immediately try to say it back without
-changing anything. NWR scores are thought to reflect long-term
-phonological knowledge (which allows the child to perceive the item
-precisely even though it is not a real word they have encountered
-before) as well as online phonological working memory (to encode the
-item in the interval between hearing it and saying it back) and flexible
-production patterns (to produce the item precisely even though it had
-never been pronounced before).
+Children's perception and production of phonetic and phonological units
+continues developing well beyond the first year of life, even extending
+into middle childhood (e.g., Hazan \& Barrett, 2000). Much of the
+evidence for later phonological development comes through the use of
+nonword repetition (NWR) tasks. In a NWR task, participants hear a short
+word-like form that is phonologically legal but lexically meaningless in
+the language(s) they are learning. After hearing this non-word, the
+participant's task is to try to immediately and precisely repeat it. NWR
+scores are thought to reflect long-term phonological knowledge (to
+perceive the item precisely despite not having heard it before) as well
+as online phonological working memory (to encode the item in the
+interval between hearing it and saying it back) and flexible production
+patterns (to produce the item precisely despite not having pronounced it
+before).
 
 NWR has been used to seek answers to a variety of theoretical questions,
-including what the links between phonology, working memory, and the
-lexicon are (Bowey, 2001), and to assess the extent to which apparent
-phonological constraints found in the lexicon affect online production
-(Gallagher, 2014). It is also frequently used for applications, notably
-as a diagnostic for language delays and disorders (Estes, Evans, \&
+including what the links are between phonology, working memory, and the
+lexicon (Bowey, 2001), and how extensively apparent phonological
+constraints found in the lexicon affect online production (Gallagher,
+2014). NWR is also frequently used in applied contexts, notably as a
+diagnostic tool for language delays and disorders (Estes, Evans, \&
 Else-Quest, 2007). Since non-words can be generated in any language, it
@@
 linguistically diverse environments, particularly in Europe (Action,
-2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has been seldom
+2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has seldom been
 used outside of Europe and North America (with exceptions including
 Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz
-(2020)).
-
-In this study, we report on NWR results among children learning Yêly
-Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense
-phonological inventory. The present report uses these unique data to
-contribute to several lines of research. First, we made sure that some
-of the non-word items contained typologically rare and/or challenging
-sounds, so that we can contribute to the study of whether rare sounds
-are disadvantaged in perception and/or production, both in terms of
-overall repetition scores and patterns of mispronunciation. Second, we
-look at the effects of item length, since previous NWR research has
-uncovered variable effects. Third, we contribute to the basic NWR
-literature by adding a data point for children learning a rarely-studied
-language and culture, and further investigating whether there are
-structured sources that account for individual variation.
-
-\subsubsection{Yélî Dnye phonological
-overview}\label{yuxe9luxee-dnye-phonological-overview}
+(2020)). Further, the languages it has been used with have historically
+had small to moderate phonological inventories (Maddieson, 2005).
+
+In the present study, we use NWR to investigate the phonological
+development of children learning Yélî Dnye, an isolate language spoken
+in Papua New Guinea (PNG) that has a large and unusually dense
+phonological inventory. The study was designed to contribute to multiple
+aspects of our understanding of phonological development: First, we
+included a subset of non-word items with typologically rare and/or
+challenging sounds to ask whether these rare sounds are disadvantaged in
+perception and/or production, both in terms of overall repetition scores
+and patterns of mispronunciation. Second, we varied the length (in
+syllables) of non-words to investigate the impact of word length on
+repetition and mispronunciation; previous NWR research has shown
+variable effects of length between populations. Third, these data give a
+first impression of how this task is responded to in a rarely-studied
+language and culture; we add to this by further investigating whether
+there are structured sources that account for individual variation in
+our sample.
+
+Before jumping into the details of our study design we first give an
+overview of Yélî Dnye phonology as well as a brief ethnographic review
+of the developmental environment on Rossel Island. Since NWR has been
+almost exclusively used in urban, industrialized populations (see
+Cristia et al., 2020 for an exception), we provide this additional
+ethographic information to contextualize the adaptations we have made in
+running the task and in gathering age and other demographic information,
+compared to what is typical in urban, industrialized settings.
+
+\subsubsection{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}
 
 Yélî Dnye is an isolate language (presumed Papuan) spoken by
-approximately 7,000 people residing on Rossel Island, which is found at
+approximately 7,000 people residing on Rossel Island, an island found at
 the far end of the Louisiade Archipelago in Milne Bay Province, Papua
 New Guinea. The Yélî sound system, much like its baroque grammatical
-system (Levinson, in preparation), is unlike any other in the region.
+system (Levinson, accepted), is unlike any other in the region.
 
@@
 equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
-substantial portion of these able to be contrastively pre-nasalized or
-nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/,
-/kŋ/, /ṭpṇm/, /kpŋm/). A large number of this combinatorial set can
-further be contrastively labialized or palatalized on release, or both
-(e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (in preparation)
-for details). The consonantal inventory also includes a number of
-non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in
-Yélî Dnye may be oral or nasal, short or long. The 10 oral vowel
-qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/,
-/o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels,
-with 7 of these able to appear as short and long nasal vowels as well
-(/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).
-
-In total, Yélî Dnye uses 90 distinctive segments (93 when including
-consonants that are extremely rarely used), far outstripping the
-phonemic inventory size of other documented Papuan languages (Levinson
-(in preparation); Foley 1986) and includes at least two contrasts not
-yet been documented elsewhere {[}labial-coronal double-articulations
-with dental vs.~post-alveolar coronal placement in both oral and nasal
-stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.
+substantial portion of them contrastively pre-nasalized or nasally
+released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/,
+/ṭpṇm/, /kpŋm/). A large number of this combinatorial set can further be
+contrastively labialized, palatalized on release, or both (e.g., /pʲ/,
+/pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (accepted) for
+details).\footnote{We use Levinson's (accepted) under-dot notation
+  (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these
+  stops are, articulatorily, somewhat variable in place, with at least
+  some tokens produced fully sub-apically. In approximating
+  cross-linguistic segment frequency below we use the corresponding
+  retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The
+consonantal inventory also includes a number of non-nasal continuants
+(/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in Yélî Dnye may be oral
+or nasal, short or long. The 10 oral vowel qualities, which span four
+levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/,
+/ɑ/) can be produced as short and long vowels, with 7 of these able to
+appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/,
+/æ̃/, /ɑ̃/).
+
+In total, Yélî Dnye uses 90 distinctive segments (not including an
+additional 3 rarely used consonants), far outstripping the phonemic
+inventory size of other documented Papuan languages (Foley, 1986;
+Levinson, accepted) and includes at least two contrasts not yet been
+documented elsewhere (labial-coronal double-articulations with dental
+vs.~post-alveolar coronal placement for both oral and nasal stops,
+Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, in preparation).
 
@@
 that, and with tri-and-above syllabic words appearing least frequently
-(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the
-current Yélî Dnye dictionary, Levinson, in preparation). The vast
-majority of syllables use a CV format. A small portion of the lexicon
-features words with a final closed syllable, limited to codas of -/m/,
--/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
+(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most
+recent Yélî Dnye dictionary at time of writing, Levinson, accepted). The
+vast majority of syllables use a CV format. A small portion of the
+lexicon features words with a final closed syllable, limited to codas of
+-/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
 spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
@@
 these are limited to /æ/-. Finally, the lexicon features a handful of
-single-vowel grammatical morphemes (see Levinson (in preparation) for
+single-vowel grammatical morphemes (see Levinson (accepted) for
 details).
 
-While all consonants can appear in the first syllable of the word, the
-second syllable onset consonant in bisyllabic roots is almost
-exclusively drawn from a subset of the simplex consonants (in order of
-descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/,
-/t/, /ɣ/, /mʷ/). In monosyllabic stems, oral vowels (short: 43\% and
-long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%),
-with the first syllable of bisyllabic stems following a similar pattern.
-Most vowels in the second syllable of bisyllabic stems are also short
-and oral, with a potential preference for rounding harmony between the
-first and second vowels. Stress typically falls on the first syllable of
-bisyllabic words, on the first and third syllables of tetrasyllabic
-words, and on the first or second syllable of trisyllabic words,
-depending on its vocalic features (see Levinson (in preparation) for
-details).
-
 \subsubsection{The Yélî community}\label{the-yuxe9luxee-community}
 
-The current study also contributes to increasing the cultural diversity
-of the populations for which NWR data have been collected. As mentioned
-previously, this method has been used almost exclusively in urban,
-industrialized populations, with only one exception to our knowledge
-(Cristia et al., 2020). We therefore provide some background information
-tol help understand some aspects of our study, including the estimation
-of children's ages and the relationship between NWR scores and
-children's demographic characteristics.
-
-Most speakers of the language grow up speaking it monolingually until
-they begin attending school around the age of 7; school instruction is
-in English. While monolingual Yélî Dnye upbringing is common,
-multilingual families are not unusual, particularly in the region around
-the Catholic Mission---the same region in which the current data were
-collected---, where there is a higher incidence of married-in mothers
-from other islands (Brown \& Casillas, in press). Children in these
-multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
-and/or some other language(s) from the region.
+Most speakers of Yélî Dnye grow up speaking it monolingually until they
+begin attending school around the age of 7; school instruction is in
+English. While monolingual Yélî Dnye upbringing is common, multilingual
+families are not unusual, particularly in the region around the Catholic
+Mission---the same region in which the current data were
+collected---where there is a higher incidence of married-in mothers from
+other islands (Brown \& Casillas, in press). Children in these
+multilingual families grow up speaking Yélî Dnye plus English, Tok
+Pisin, and/or another language(s) from the region.
 
@@
 Laying 250 nautical miles off the coast of mainland PNG and surrounded
-by barrier reef, transport to and from the island is both infrequent and
-irregular. There is no electricity system (we use solar panels) or motor
-vehicles on the island, save dinghys with outboard motors that are
-primarily reserved for medical transport, when there is deisel
-available. There is a medical radio connection at the health clinic (at
-the Catholic mission) via which news and messages are often passed
-along. There is also a cellular tower on the island, though it is
-typically not functioning. As such, access to news and outside
-resources, particularly international connections and larger data
-transfers, is typically unavailable. Our data collection is therefore
-limited to the duration of the researchers' on-island visits.
+by a barrier reef, transport to and from the island is both infrequent
+and irregular. There is no electricity system (we use solar panels) or
+motor vehicles on the island, save dinghys with outboard motors that,
+even when there is deisel available, are primarily reserved for medical
+transport. There is a medical radio connection at the health clinic,
+located at the Catholic Mission, via which news and messages are often
+passed to and from mainland PNG. There is also a cellular tower on the
+island, though it is often inoperative, usually for multiple months or
+years at a time. As such, access to news and outside connections,
+including international phone calls and digital exchanges that require
+significant data transfer, is typically unavailable. Our data collection
+is therefore limited to the duration of the researchers' on-island
+visits.
 
@@
 well out of walking distance for many children (i.e., more than 1 hour
-on foot or by canoe each day), and so it is very common for households
-to billet school children during the weekdays for months at a time (or
-for whole months if the child's home is very far). Combined with
-practices of collective child guardianship within their close-knit
-hamlets, the practical outcome for researchers is that adult consent can
-often come from a combination of aunts, uncles, adult cousins, and
-grandparents standing in for the child's parents---to this child assent
-is also culturally pertinent, as child independence is encouraged and
-respected from toddlerhood onward (Brown \& Casillas, in
-press).\footnote{Note that in a few cases this does mean that precise
-  details such as child date of birth, parent age in years, and parent
-  education was only able to be approximated as the parents were
-  residing on another part of the island.} Finally, as mentioned, the
-language of schooling in this region is English, and so the majority of
-teens and adults on the island speak some English---many speak it very
-fluently.
+on foot or by canoe each day). It is very common for households situated
+close to schools to billet their school-aged relatives during the
+weekdays (sometimes the weekends too, if the child's home is very far)
+for long segements of each school year. Combined with practices of
+collective child guardianship within each close-knit hamlet, the
+practical consequence of this billeting for researchers is that adult
+consent can often come from a combination of aunts, uncles, adult
+cousins, and grandparents standing in for the child's biological
+parents. On top of this, child assent is culturally pertinent, as child
+independence is encouraged and respected from toddlerhood onward (Brown
+\& Casillas, in press).\footnote{For these reasons, we were unable to
+  record precise details for a few children, including child date of
+  birth, parent age in years, and parent education.} Finally, as
+mentioned, the language of schooling in this region is English, and so
+the majority of teens and adults on the island speak some English---many
+speak it very fluently.
 
@@
 linguistic development is quite limited, and research into their
-phonological development in particular has only just begun {[}e.g.,
-Bram's thesis; MC's CFH data{]}. What we do know about the early
-language environment is that, while child-directed speech is relatively
-infrequent throughout the day (Casillas et al., 2020), shared caregiving
-practices and a near-universal fondness for social interaction with
-young children ensure that speech directed to children comes from all
-types of speakers: women, men, and other children (Bunce et al., under
-review; Casillas et al., 2020). While speech from women predominates in
-young children's speech environments on Rossel, as it does elsewhere
-(Bergelson et al., 2019; Bunce et al., under review), there is a
-significant and marked increase in child-directed speech from other
-children as infants get older (Bunce et al., under review; Casillas et
-al., 2020). This increase in child-directed speech from other children
-is attributed to the fact that, starting around age 2, children often
-spend large swaths of the day playing, swimming, and foraging for fruit,
-nuts, and shellfish in large (\textasciitilde{}10 people) independent
-and mixed-age child play groups (Brown \& Casillas, in press; Casillas
-et al., 2020).
-
-\subsubsection{Brief review of NWR for our
-purposes}\label{brief-review-of-nwr-for-our-purposes}
-
-The basic structure of NWR involves the participant hearing a non-word,
-like ``bilik'', which they need to repeat back without changing anything
-that is contrastive in the language. For instance, in English, a
-response of ``bilig'' or ``pilik'' will be scored as incorrect; a
-response ``biilik'', where the vowel is lengthened without change of
-quality would be scored as correct, because English does not have a pure
-length contrast.
-
-Aside from this central basic structure, there is some variation in the
-presentation procedure and structure of items found in previous NWR
-work. For example, items are often presented orally by the experimenter
-(Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an
-increasing number of studies have turned to playing back the stimuli in
-order to have greater control of the stability of the presentation
-(Brandeker \& Thordardottir, 2015). Additionally, while some studies
-have used 10-15 non-words, others have employed up to 46 unique items
-(Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019). Often,
-authors modulate structural complexity, typically measured in terms of
-item length (measured in number of syllables) and/or syllable structure
-(open as opposed to closed syllables, Gallon, Harris, \& Van der Lely,
-2007).
-
-Previous work seems to avoid difficult sounds, but we felt this was
-important to represent Yélî Dnye, so we also varied this factor. We
-designed a relatively large number of items but, aware that this may
-render the task longer and more tiresome, we split some of the items
-across children. This allowed us to get NWR scores for more items.
-
-Naturally, designing the task in this way may render the study of
-individual variation within the population more difficult because
+phonological development in particular has only just begun (e.g., Peute,
+Fikkert, \& Casillas, In preparation). What we do know about the early
+language environment is that, while direct speech to children under 3;0
+is relatively infrequent throughout the day (Casillas et al., 2020),
+shared caregiving practices and a near-universal fondness for social
+interaction with young children ensure that speech directed to children
+comes from all types of speakers: women, men, and other children (Bunce
+et al., under review; Casillas et al., 2020). While speech from women
+predominates in young children's speech environments on Rossel, as it
+does elsewhere (Bergelson et al., 2019; Bunce et al., under review),
+there is a significant and marked increase in child-directed speech from
+other children as infants get older (Bunce et al., under review;
+Casillas et al., 2020). This increase in child-directed speech from
+other children is attributed to the fact that, starting around age 2,
+children often spend large swaths of the day playing, swimming, and
+foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10
+members) independent and mixed-age child play groups (Brown \& Casillas,
+in press; Casillas et al., 2020).
+
+\subsubsection{NWR design and analysis
+adaptations}\label{nwr-design-and-analysis-adaptations}
+
+In a basic NWR task, the participant listens to a production of a
+word-like form, such as /bilik/, and then repeats back what they heard
+without changing any phonological feature that is contrastive in the
+language. For instance, in English, a response of {[}bilig{]} or
+{[}pilik{]} would be scored as incorrect; a response {[}biilik{]}, where
+the vowel is lengthened without change of quality would be scored as
+correct, because English does not have a general feature of contrastive
+vowel length. There is some variation in how past NWR studies have
+designed the presentation procedure and structure of items. For example,
+while items are often presented orally by the experimenter (Torrington
+Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies
+have turned instead to playing back pre-recorded stimuli in order to
+increase control in stimulus presentation (Brandeker \& Thordardottir,
+2015). Additionally, while some studies have used 10-15 non-words,
+others have employed up to 46 unique items (Piazzalunga, Previtali,
+Pozzoli, Scarponi, \& Schindler, 2019). Authors also often modulate
+structural complexity, typically measured in terms of item length
+(measured in number of syllables) and/or syllable structure (open as
+opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
+
+Previous work typically steers clear of articulartorily and/or
+acoustically challenging sounds, but we included some in our experiment
+to more adequately represent Yélî Dnye's phonology. We ultimately used a
+relatively large number of items that would enable us to explor both
+variation in structural complexity and in more vs.~less challenging
+sounds. However, aware that this large item inventory might render the
+task longer and more tiresome, we split items across children (see
+below). Naturally, designing the task in this way may render the study
+of individual variation within the population more difficult because
 different children are exposed to different items. However, a review of
 previous work on individual variation suggested to us that effects of
-individual differences are relatively small, and would not be detectable
-with the sample size that we could collect in a given visit.
-
-That said, we contribute to the literature by also reporting descriptive
-analyses of individual variation that could potentially be integrated in
-meta-analytic efforts. Based on previous work, we looked at potential
-improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid,
-Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and
-potential negative effects of bilingual exposure (Brandeker \&
-Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
-Previous work typically finds no significant differences as a function
-of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014;
-Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
-Although past research has not often investigated potential effects of
-birth order on NWR, there is a sizable literature on these effects in
-other language tasks (Havron et al., 2019), and therefore we report on
-those too.
+individual differences in NWR are relatively small, and would not be
+detectable with the sample size that we could collect during our short
+visit. That said, we contribute to the literature by also reporting
+descriptive analyses of individual variation that could potentially be
+integrated in meta-analytic efforts.
+
+Based on previous work, we looked at potential changes with age (Farmani
+et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance,
+Stackhouse, \& Wells, 2005), and differences relating to bilingual
+language exposure (Brandeker \& Thordardottir, 2015; Meir \&
+Armon-Lotem, 2017; Meir et al., 2016). Previous work typically finds no
+significant differences as a function of maternal education (e.g.,
+Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or
+child gender (Chiat \& Roy, 2007). Although past research has not often
+investigated potential effects of birth order on NWR, there is a sizable
+literature on these effects in other language tasks (Havron et al.,
+2019), and therefore we report on these too.
 
@@
 \item
-  Does the frequency of sounds across languages predict NWR? Are rarer
-  sounds more often substituted by commoner sounds?
+  Does the cross-linguistic frequency of sounds predict NWR scores? Are
+  rarer sounds more often substituted by commoner sounds?
 \item
@@
 analysis plans prior to conducting them. However, we realized that even
-pre-registering an analysis would be equivocal because we do not have
-enough power to look at all relationships of interest, and often to
-detect any of the known effects given their variability across studies.
-To illustrate this, we portray studies in which children's NWR scores
-were gathered between 4 and 12 years of age, and reported separately for
-items that are relatively short (1-2 syllables) versus longer items (3-4
-syllables) in Figure 1. Notice that the effect of stimulus length is
-minuscule among Italian children (Piazzalunga et al., 2019), but
-considerable among Tsimane' children (Cristia et al., 2020), where a
-drop of 40 percentage points is observed at all ages. A similar
-difference in NWR scores for short versus long items was observed among
-Arabic children (Jaber-Awida, 2018). Even the effect of age is unstable
-in this sample. Whereas it is quite clear that children's NWR scores
-increase in the Italian data, age effects are less stable among Tsimane'
-children. Therefore, all analyses here are descriptive and should be
-considered exploratory.
+pre-registering an analysis would be equivocal because we would not have
+enough power to look at all relationships of interest, in many cases
+possibly not enough to detect any of the known effects, given their
+variability across studies. To illustrate this issue, we portray in
+Figure 1 studies in which children's NWR scores were gathered between 4
+and 12 years of age, and reported separately for items that are
+relatively short (1-2 syllables) versus longer items (3-4 syllables).
+Notice that the effect of stimulus length is minuscule among Italian
+children (Piazzalunga et al., 2019), but considerable among Tsimane'
+children (Cristia et al., 2020), where a drop of 40 percentage points is
+observed at all ages. A similar large difference in NWR scores for short
+versus long items was observed among Arabic children (Jaber-Awida,
+2018). Even the effect of age is unstable in this sample of studies.
+Whereas it is quite clear that children's NWR scores increase in the
+Italian data, age effects are less stable among Tsimane' children.
+Therefore, all analyses in the present study are descriptive and should
+be considered exploratory.
 
@@
 length between 1 and 4 syllables, often additionally varying syllable
-complexity and/or cluster presence and complexity, always meeting the
-condition that they do not mean anything in the target language (e.g.,
-Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We kept the
-same variation in item length and the non-meaningfulness requirement,
-but we did not vary syllable complexity and clusters because these are
-vanishingly rare in Yély Dnye. We also increased the number of items an
-individual child would be tested on, so that a child would get up to 23
-items to repeat (note that up to other work has also used 24-30 items:
-Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and
-distributed them across children, so as to increase the coverage, and be
-able to study more items.
-
-A first list of candidate items was generated in 2018 by selecting
-simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'',
-``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined
-into consonant-vowel syllables, further sampling the space of 1- to
-4-syllable sequences. These candidates were automatically checked
-against Levinson's 2015 dictionary and removed from consideration if
-they appeared in the dictionary. The second author presented them orally
-to three local research assistants, who were asked to repeat them and
-further say whether they were real words. Any item for which two or more
-of the assistants reported them having a meaning or some form of
-association was excluded.
+complexity and/or cluster presence and complexity, and always meeting
+the condition that they do not mean anything in the target language
+(e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We
+kept the same variation in item length and requirement for not being
+meaningful in the language, but we did not vary syllable complexity or
+clusters because these are vanishingly rare in Yélî Dnye. We also
+increased the number of items an individual child would be tested on,
+such that a child would get up to 23 items to repeat (other work has
+also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014),
+with the entire test inventory of 44 items distributed across children.
+
+A first list of candidate items was generated during a trip to the
+island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/,
+/n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them
+into consonant-vowel syllables, then sampling the space of 1- to
+4-syllable sequences. These candidates were automatically removed from
+consideration if they appeared in Levinson's (2015) dictionary. The
+second author presented them orally to three local research assistants,
+all native speakers of Yélî Dnye, who were repeated each form, as they
+would in an NWR task, and additinally let the experimenter know if the
+item was in fact a word or phrase in Yélî Dnye. Any item reported to
+have a meaning or a strong association with another word form or meaning
+was excluded.
 
@@
 complex consonants and systematically crossing them with all the vowels
-in the Yélî inventory to produce consonant-vowel monosyllables. As
-before, items were automatically excluded if they appeared in the
-dictionary. Additionally, since hearing vowel length in monosyllables in
-isolation is challenging, any item that had a short/long real word
-neighbor was filtered out. Since the phonology and phonetics of Yélî is
-still in the process of being described (Levinson, in preparation),
-there could have been undocumented constraints that rendered items
-illegal. Therefore, we made sure that the precise consonant-vowel
-sequence occurred in some real word in the dictionary (i.e., that there
-was a longer word included the monosyllable as a subsequence). These
-candidates were presented to one informant, for a final check that they
-did not mean anything. Together with the 2018 selection, they were
-recorded using a Shure SM10A XLR dynamic headband microphone and an
-Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter)
-from the written form presented together with the same item orally by
-the second author. The complete recorded list was finally presented to
-two more informants, who could repeat all the items and who confirmed
-there were no real words. Even so, there was one monosyllable that was
-often identified as a real word (intended ``yî'' /yXX/; identified as
-``yi'' /yi/, \emph{tree}). This item is removed from analyses below.
-
-The final list is composed of three practice items; 20 monoysllables
+in the Yélî Dnye inventory to produce consonant-vowel monosyllabic
+forms. As before, items were automatically excluded if they appeared in
+the dictionary. Additionally, since perceiving vowel length in isolated
+monosyllables is challenging, any item that had a short/long lexical
+neighbor was excluded. Because there is still much to discover about the
+phonology and phonetics of Yélî Dnye (Levinson, accepted), it was also
+possible that we initially generated items with illegal, but currently
+undocumented constraints. Therefore, we made sure that the precise
+consonant-vowel sequence occurred in some real word in the dictionary
+(i.e., that there was a longer word included the monosyllable as a
+subsequence). These candidates were then presented to one informant, for
+a final check that they did not mean anything. Together with the 2018
+selection, they were recorded, based on their orthographic forms, using
+a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832
+stereo audio recorder (using an XLR to mini-jack adapter) by the same
+informant, monitored by the second author for clear production of the
+phonological target. The complete recorded list was finally presented to
+two more informants, who were able to repeat all the items and who
+confirmed there were no real words present. Despite these checks, one
+monosyllable was ultimately frequently identified as a real word in the
+resulting data (intended ``yî'' /yɯ/; identified as ``yi'' /yi/,
+\emph{tree}). This item is removed from the analyses below.
+
+The final list includes: three practice items; 20 monoysllables
 containing sounds that are less frequent in the world's languages than
@@
 
-A Praat script was written to randomize this list 20 times, and split it
-into two sublists, to generate 40 different elicitation sets. The 40
-elicitation sets are available online from \url{https://osf.io/5qspb}.
-The split had the following constraints:
+A Praat script (Boersma \& Weenink, 2020) was written to randomize this
+list 20 times, and split it into two sublists, to generate 40 different
+elicitation sets. The 40 elicitation sets are available online from
+\url{https://osf.io/5qspb}. The split had the following constraints:
 
@@
 \item
-  the same three items were selected as practice items and used in all
+  The same three items were selected as practice items and used in all
   40 elicitation sets
 \item
-  splits were done within each length group from the 2018 items (i.e.,
-  separately for 2, 3, and 4-syllable items); and among onset groups for
-  the difficult monosyllables generated in 2019 (i.e., all the
-  monosyllables starting with tp were split into 2 sublists). Since some
-  of these groups had an odd number of items, one of the sublists was
-  slightly longer than the other (20 versus 23).
-\item
-  once the sublist split had been done, items were randomized such that
-  all children heard first the 3 practice items in a fixed order (1, 2,
-  and 4 syllables), a randomized version of their sublist selection of
-  difficult onset items, and randomized versions of their 2-syllables,
-  then 3-syllables, and finally 4-syllable items.
+  Splits were done within each length group from the 2018 items (i.e.,
+  separately for 2-, 3-, and 4-syllable items); and among onset groups
+  for the difficult monosyllables generated in 2019 (i.e., all the
+  monosyllables starting with /tp/ were split into 2 sublists). Since
+  some of these groups had an odd number of items, one of the sublists
+  was slightly longer than the other (20 vs.~23).
+\item
+  Once the sublist split had been done, items were randomized such that
+  all children heard first the 3 practice items in a fixed order (1-,
+  2-, and 4-syllable), a randomized version of their sublist selection
+  of difficult onset items, and randomized versions of their 2-syllable,
+  then 3-syllable, and finally 4-syllable items.
 \end{itemize}
@@
 
-We tried to balance three desiderata: That children would not be unduly
-exposed to the items before they themselves had to repeat them; that
-children would feel comfortable doing this task with us; and that the
-community would feel safe with us doing this task with their children.
-Moreover, there were also some logistic constraints in terms of the
-space availability. As a result, the places where elicitation happened
-varied across the hamlets.
-
-We visited four different hamlets once, and attempted to test all
-eligible children present at the time, to prevent the items `spreading'
-through hearsay. In the first hamlet, we tested children in five
-different places, with some children being tested inside a house and
-others tested on the veranda. The complete list of places and the ways
-in which they met the desiderata mentioned above can be found in the raw
-data, available from online supplementary materials.
-
-The child was donned a headset (a Shure SM10A or WH20 XLR with a dynamic
-microphone on a headband, most children using the former), recorded into
-the left channel into a Tascam DR40x digital recorder. For most
-children, the headset could not stay comfortably on the child's head,
-and thus it was placed on the child's shoulders, with the microphone
-carefully placed close to the child's mouth. A local informant sat next
-to the child, to would provide the instructions and, if needed, coach
-the child to make sure, using the three practice items as well as real
-words, that they understood that the task was to repeat the items
-precisely without changing anything. An experimenter (the first author)
-delivered the elicitation stimuli to the local informant and the child
-over headphones.
-
-The first phase was making sure the child understood the task. This was
-explained orally and the first training item was presented. Often,
-children froze and did not say anything. If this happened, then we
-followed this procedure. First, the informant insisted. If the child
-still did not say anything, the informant asked the child to repeat a
-real word, and another, and another. If the child could repeat these
-correctly, then we provided the recorded training item over headphones
-again. Most children successfully started repeating the items presented
-over headphones at this point; a few further needed the local informant
-to model the behavior (i.e., they would hear the item again, and she
-would say it; then we would play it again, and ask the child to say it).
-A small minority still failed to repeat the item after hearing it over
-headphones. If that occurred, we tried with the second training item, at
-which point some children got it and could continue. A small minority,
-however, failed to repeat this one, as well as the third training item,
-in which case we stopped the test altogether (see Participants section
-for exclusions).
-
-NWR studies vary in whether children are provided with several
-opportunities to hear and say the item. To have a fixed and clear
-procedure, we decided that items other than the inital three training
-ones would not be repeated unless the child made an attempt to produce
-them. If this attempt was judged correct by the local informant, then
-the experimenter would move on to the next item (whispering this over a
-separate headset that was recording onto the right track of the same
-Tascam). If the local informant heard a deviation, she indicated to the
-experimenter that the item needed to be repeated, and up to 5 attempts
-were allowed.
-
-Whenever siblings from the same family were tested, an attempt was made
-to test first the older and then the younger child, and always on
-different elicitaiton sets.
+In adapting the typical NWR procedure for this context, we balanced
+three desiderata: That children would not be unduly exposed to the items
+before they themselves had to repeat them (i.e., from other children who
+had participated); that children would feel comfortable doing this task
+with us; and that community members would feel comfortable having their
+children do this task with us.
+
+We ran children in batches, testing within a handful of hamlet clusters
+spread across the northeastern region of the island. Because space
+availability was limited in different ways from hamlet to hamlet, the
+places where elicitation happened varied across testing sites. We tested
+in four different sites, only making a single visit to each, conducting
+back-to-back testing of all eligible children present at the time of our
+visit in order to prevent the items `spreading' between children through
+hearsay. In the first hamlet, we tested children in five different
+places, with some children being tested inside a house and others tested
+on the veranda. The complete list of places and the ways in which they
+met the desiderata mentioned above can be found in the raw data,
+available from online supplementary materials. Whenever children living
+in the same household were tested, we tried to test children in age
+order, from oldest to youngest, to minimize intimidation for younger
+household members, and always using different elicitaiton sets.
+
+We fitted the child with a headset microphone (Shure SM10A or WH20 XLR
+with a dynamic microphone on a headband, most children using the former)
+that fed into the left channel of a Tascam DR40x digital audio recorder.
+The headsets were designed for adult use and could not be comfortably
+seated on many children's heads without a more involved adjustment
+period. To minimize adjustment time, which was uncomfortable for some
+children given the proximity of the experimenter and equipment, we
+placed the headband on children's shoulders in these cases, carefully
+adjusting the microphone's placement so that it was still close to the
+child's mouth. A research assistant who spoke Yélî Dnye natively sat
+next to the child throughout the task to provide instructions and, if
+needed, encouragement. The research assistant coached the child
+throughout to task to make sure that they understood what they were
+expected to do. An experimenter (the first author) delivered the
+pre-recorded stimuli to the research assistant and the child over
+headphones.
+
+The first phase of the experiment involved making sure the child
+understood the task. We explained the task orally presented the first
+practice item. At this point, many children did not say anything in
+response, at which point we initiated the following procedure: First,
+the assistant insisted the child make a response. If the child still did
+not say anything, the assistant said a real word and then asked the
+child to repeat it, then another and another. If the child could repeat
+real words correctly, we provided the first training item over
+headphones again for children to repeat. Most children successfully
+started repeating the items at this point, but a few needed further
+help. In this case, the assistant modeled the behavior (i.e., the child
+and assistant would hear the item again, and the assistant would repeat
+it; then we would play the item again and ask the child to repeate it).
+A small minority of children still failed to repeat the item at this
+point. If so, we tried again with the second training item, at which
+point some children got it and could continue. A fraction of the
+remaining children, however, failed to repeat this second training item,
+as well as the third one, in which case we stopped testing altogether
+(see Participants section for exclusions).
+
+The second phase of the experiment involved running the child through
+the list of test items randomly assigned to them. This was done in the
+same manner as the practice items: the stimulus was played over the
+headphones, and then the child repeated it aloud. NWR studies vary in
+whether children are allowed to hear and/or repeat the item more than
+one time. We had a fixed procedure for the test items (i.e., the
+non-practice items) in which the child was alowed to make a further
+attempts if their first attempt was judged erroneous in some way by the
+assistant. The procedure worked as follows: When the child made an
+attempted a repeat, the assistant indicated to the experimenter whether
+the child's production was correct or not. If correct, the experimenter
+would whisper this note of correct repetition into a separate headset
+that fed into the right channel of the same Tascam and the group moved
+on to the next item. If not, the child was allowed to try again, with up
+to five attempts allowed before moving on to the next item. Children
+were not asked to make repetitions if they did not produce a first
+attempt.
+
+In total, test sessions took approximately XX minutes on average, with
+XX of those minutes attributed to practice and XX to the actual test
+list.
 
@@
 
-A script was written to randomize all tokens from all children, pairing
-each with the auditory target the child had been provided with. A native
-research assistant then listened through all productions of a given
-target (randomized across children and repetition order), and made a
-judgment of whether the item was correctly or incorrectly repeated. She
-additionally orthographically transcribed exactly what the child said,
-providing some examples of the types of errors children in general make
-(without making specific reference to Yélî sounds or the items in the
-elicitation sets).
+The first author then annotated the onset and offset of all children's
+productions from the audio recording using Praat audio annotation
+software (Boersma \& Weenink, 2020), then wrote and ran a script to
+extract these tokens, pairing them with their original auditory target
+stimulus, and writing these audio pairs out to short .wav files. The
+assistant then listened through all these productions, one target item
+at a time, with productions for each target item presented in a random
+order across children and repetitions. The assistant indicated in a
+notebook whether each production was a correct or incorrect repetition
+and orthographically transcribed the production, noting when the child
+uttered a recognizable word or phrase and adding the translation
+equivalent of that word/phrase into English. The assistant also provided
+some general examples of the types of errors children made without
+making specific reference to Yélî sounds or the items in the elicitation
+sets.
 
@@
 
-Some NWR studies employ phoneme-based scoring in addition to or instead
-of word-level scoring (e.g., Cristia et al., 2020). We scored items in
-terms of the number of phonemes that could be aligned across the target
-and attempt, divided by the number of phonemes of whichever item was
-longer (the target or the attempt). Although previous work does not use
-distance metrics, we additionally report those. For instance, recall our
-example of an English target being ``bilik'', and imagine a response
-``bilig''. This would receive the following scores: at the whole item
-level this would receive a score of zero (because the repetition is not
-exact); at the phoneme level it would receive a score of 80\% (4 out of
-5 phonemes repeated exactly); and its normalized phone-based Levenshtein
-distance is 20\% (20\% of phonemes were substituted or deleted).\\
-
-Additionally, we looked up each phoneme in each target word in the
-Phoible database (Moran \& McCloy, 2019). We entered the number of
-languages in which each phoneme was found, as well as what percentage of
-languages in the database that represented. We could not find a small
-number of phonemes in the database, and treat those as NA values that do
-not contribute to the target-level global average phone cross-linguistic
-frequency, although an alternative would be to attribute them a very
-small number.
-
-Finally, for describing children's patterns of errors, all repetitions
-of a given target were taken into account. We describe the proportion of
-items where the change resulted in a real word (semantic errors); as
+In addition to a binary word-level exact repetition score we scored
+items in terms of the number of phonemes that could be aligned across
+the target and attempt, divided by the number of phonemes of whichever
+item was longer (the target or the attempt; see also Cristia et al.,
+2020). Although previous work does not use distance metrics, we
+additionally report these. For instance, recall our example of an
+English target being /bilik/ with an imagined response {[}bilig{]}. We
+would score this response as follows: at the whole item level this
+production would receive a score of zero (because the repetition is not
+exact); at the phoneme level this production would receive a score of
+80\% (4 out of 5 phonemes repeated exactly); and the phone-based
+Levenshtein distance for this production is 20\% (because 20\% of
+phonemes were substituted or deleted).\\
+
+Additionally, we estimated the typological frequency of all phonological
+segments used in the target items using the PHOIBLE cross-linguistic
+phonological inventory database (Moran \& McCloy, 2019). For each phone
+in our task, we extracted the number and percentage of languages noted
+to have that phone in its inventory. While PHOIBLE is an unprecedentedly
+comprehensive database, with phonological inventory data for over 2000
+languages at the time these data were analyzed, it is of course still
+far from complete, so estimated frequencies should be taken with a pinch
+of salt. In both PHOIBLE and UPSID---another large phonemic inventory
+database---note that nearly half of the segment types only appear in a
+single language (Steven Moran, personal communication; see also
+\href{http://web.phonetik.uni-frankfurt.de/upsid_info.html}{web.phonetik.uni-frankfurt.de/upsid\_info.html}).
+It is therefore unsurprising that we did not find three of our segments
+(/lβʲ/, /ṭp/, and /tp/) in the database. Following the general pattern
+that many phones only appear once cross-linguistically, even in large
+databases, we treat our ``missing'' phones as having a frequency of 1,
+with a (rounded) percentage of 0\%.
+
+Finally, when describing children's patterns of errors, we take all
+repetitions of a given target into account. We describe the proportion
+of items where the change resulted in a real word (semantic errors); as
 well as deletions, additions, and substitutions.
@@
 
-This study was approved as part of a larger research effort by Marisa
-Casillas, with the research protocol having been approved by the IRB
-committee entitled The Ethics Committee of the Faculty of Social
-Sciences (Ethiek Commissie van de faculteit der Sociale Wetenschappen;
-ECSW) of Radboud University, Nijmegen, The Netherlands. The approval for
-work outside of the Netherlands came as an amendment to the general
-research line on language development research (original request:
-ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
-Participation was voluntary, with children being invited to come and
-participate. Regardless of whether they did any of the task, children
-were provided with a snack as compensation. Children who came up to
-participate but then refused were nonetheless provided with the snack.
-
-A total 55 children were tested, from 34 of families, in five hamlets.
-Some children could not be included for the following reasons: refused
-participation or failed to repeat items presented over headphones even
-after coaching (N=8), spoke too softly to allow offline coding (N=5). In
-addition, 2 teenagers were tested to put younger children at ease; their
-data is not included in analyses below. The remaining 40 children (14
-girls) were aged 6.96 years (range 3.90-11 years). Among these, there
-were 32 children exposed only to Yélî in the home, 6 children who were
-also exposed to another language in the home, and 2 for whom this
-information is missing. Maternal years of education averaged 8.24 years
-(range 6-12 years; 2 children had this information missing).\footnote{Precise
-  years of education can be hard to count. This is particularly the case
-  for schooling beyond grade 8, which takes place off-island, which can
-  be due to an irregularity in ability to pay school fees, school
-  opening times, familial obligations, and more. We asked parents to
-  report their highest completed level of education and inferred the
-  approximate number of years that entails under ideal conditions, which
-  results in many esimates around typical ``completion'' times in
-  between.} In terms of birth order, 5 were first borns, 4 second, 2
-third, 7 forth, 5 fifth, 1 sixth, and NA did not have this information.
+This study was approved as part of a larger research effort by the
+second author. The line of research was evaluated by the Radboud
+University Faculty of Social Sciences Ethics Committee (Ethiek Commissie
+van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The
+Netherlands (original request: ECSW2017-3001-474 Manko-Rowland;
+amendment: ECSW-2018-041). Participation was voluntary; children were
+invited to participate following indication of approval from an adult
+caregiver. Regardless of whether they completed the task, children were
+given a small snack as compensation. Children who showed initial
+interest but then decided not to participate were also given the snack.
+
+We tested a total of 55 children from 38 families spread across four
+hamlets. We excluded test sessions from analysis for the following
+reasons: refused participation or failed to repeat items presented over
+headphones even after coaching (N=8), spoke too softly to allow offline
+coding (N=5), or were 13 years old or older (N=2; we tested these
+teenagers to put younger children at ease). The remaining 40 children
+(14 girls) were aged 6.98 years (range 3.92-11.03 years). Among these
+children were 34 only exposed to Yélî Dnye at home and 6 children
+exposed to one or more other languages at home. Maternal years of
+education averaged 8.22 years (range 6-12 years).\footnote{Years of
+  education is difficult to precisely pinpoint in this context.
+  Schooling beyond grade 8 takes place off-island and enrollment is tied
+  to an ability to pay that year's school fees, seasonal interruptions
+  to school opening times, familial obligations, and more. We asked
+  parents to report their highest completed level of education. We then
+  record the number of years entailed by having completed that level
+  under ideal conditions. This method results in many esimates around
+  typical ``graduation'' times (e.g., graduating from primary education
+  before starting secondary education).} In terms of birth order, 6 were
+first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth.
 
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/Fig1-first_vs_others-1.pdf}
-\caption{(\#fig:Fig1-first\_vs\_others)NWR scores for individual
-participants averaging separately their first attempts and all other
-attempts.}
+\includegraphics{manuscript_files/figure-latex/Fig1-first-vs-others-1.pdf}
+\caption{\label{fig:Fig1-first-vs-others}Whole-item NWR scores for
+individual participants averaging separately their first attempts and
+all other attempts.}
 \end{figure}
 
-We first checked whether NWR scores varied between first and subsequent
-presentations of an item by averaging word-level scores at the
-participant level separately for first attempts and subsequent
-repetitions. We excluded 3 children who did not have data for one of
-these two types. As shown in Figure 1, participants' mean word-level
-scores became more heterogeneous in subsequent repetitions.
-Surprisingly, NWR scores for subsequent repetitions (M = 40, SD = 26)
-were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p
-= 0. Given the uncertainty in whether previous work used only the first
-or all repetitions, and since scores declined and became more
-heterogeneous in subsequent repetitions, the rest of the analyses focus
-on only the first repetitions.
+We first checked whether whole-item NWR scores varied between first and
+subsequent presentations of an item by averaging word-level scores at
+the participant level separately for first attempts and subsequent
+repetitions. We excluded 1 children who did not have data for one of
+these two types. As shown in Figure \ref{fig:Fig1-first-vs-others},
+participants' mean word-level scores became more heterogeneous in
+subsequent repetitions. Surprisingly, whole-item NWR scores for
+subsequent repetitions (M = 40, SD = 26) were on average lower than
+first ones (M = 64, SD = 15), t(38) = 6.28, p = 0; Cohen's d = 1.14).
+Given uncertainty in whether previous work used first or all
+repetitions, and given that score here declined and became more
+heterogeneous in subsequent repetitions, we focus the remainder of our
+analyses only on first repetitions.
 
 Taking into account only the first attempts, we averaged attempts by
-each of the 42 children who had data for first attempts; their ages
-ranged from NA to NA (M = NA, SD = NA).
-
-The overall NWR score was M = 65\% (SD = 15\%). Scores based on phonemes
-are even higher M = 78\% (SD = 9\%). The phoneme-based normalized
-Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth
-of phonemes were were substituted, inserted, or deleted. Notice that the
-normalized Levenshtein distance is the complement of phoneme-based
-scores.
+each of the 40 children who had data for first attempts; their ages
+ranged from 3 to 10 (M = 6.50, SD = 1.50).
+
+The overall NWR score was M = 64\% (SD = 15\%), Cohen's d = 4.36. Scores
+based on phonemes are even higher M = 78\% (SD = 9\%), Cohen's d = 8.55.
+The phoneme-based normalized Levenshtein distance was M = 22\% (SD =
+9\%), meaning that about a fifth of phonemes were were substituted,
+inserted, or deleted (the normalized Levenshtein distance is the
+complement of phoneme-based scores).
 
@@
 
-In this analysis, we were interested in variation in NWR scores as a
-function of the average frequency with which sounds composing individual
-target words are found in languages over the world. To look at this, we
-fit a mixed logistic regression, in which the outcome variable was
-whether the non-word was correctly repeated or not. The fixed effects of
-interest was the average cross-linguistic phone frequency; we also
-included child age as a control fixed effect, and child ID \& target ID
-as random variables.
-
-We could include 885 observations (\textbf{why -- shouldn't it be
-1k??}), from 41 children producing in any given trial one of 41
-potential target words. The analysis revealed a significant estimate for
-the average cross-linguistic frequency of phones in the target words (ß
-= 0.82, SE ß = 0.18, p = 0): Target words with phones found more
-frequently across languages had a higher proportion of words that were
-correctly repeated, as clear in Figure 2. Age was also a significant
-predictor (ß = 0.33, SE ß = 0.13, p = 0.01).
+In this analysis, we were interested in variation in whole-item NWR
+scores as a function of the average frequency with which sounds
+composing individual target words are found in languages over the world.
+To look at this, we fit a mixed logistic regression, in which the
+outcome variable was whether the non-word was correctly repeated or not.
+The fixed effects of interest was the average cross-linguistic phone
+frequency; we also included child age as a control fixed effect, and
+child ID \& target ID as random variables.
+
+We could include 850 observations, from 40 children producing in any
+given trial one of 41 potential target words. The analysis revealed a
+main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant
+estimate for the scaled average cross-linguistic frequency of phones in
+the target words (ß = 0.82, SE ß = 0.18, p = 0): Target words with
+phones found more frequently across languages had a higher proportion of
+words that were correctly repeated, as clear in Figure 2. Averaging
+across items and participants, the Pearson correlation between scaled
+average cross-linguistic phone frequency and whole-item NWR scores was r
+r(31) = 0.63.
 
@@
 
-We next checked whether this may actually be due to frequency of the
-sounds within the language. We estimated frequency of the phones present
-in the stimuli in a corpus of child-centered recordings (Casillas et
-al., 2020), and applied the natural logarithm. As before, unattested
-sounds were not considered (i.e., they were declared NA so that they do
-not count for analyses). Phone frequency estimated from this corpus were
-correlated with cross-linguistic phone frequencies (r(27)=0.52, =0.00).
-Therefore, we fit another mixed logistic regression, this time declaring
-as fixed effects scaled cross-linguistic and corpus frequencies
-(averaged across all attested phones within each stimulus item), in
-addition to age. As before, the model contained random slopes for both
-child ID and target. In this model, both cross-linguistic frequency (ß =
-0.82, SE ß = 0.27, p = 0.00) and age (ß = 0.33, SE ß = 0.13, p = 0.01)
-were significant predictors of word-level NWR scores, but corpus
-frequency (ß = 0, SE ß = 0.25, p = 1.00) was not.
-
-\textbf{use proportion of correctly produced rather than number; DONE}
-
-\textbf{a2m use first attempts rather than the whole data set: having
-second thoughts. To be clear: the argument for using first attempts only
-is that all of the other analyses are on kids' first attempts, rather
-than all of their attempts, so we should use just first attempts here
-because it may confuse readers \& it is more consistent. BUT there are
-fewer mistakes this way, which is Options 1 and 2 below has lots of
-zeros\ldots{} CF Option 3 further down}
-
-\textbf{a2m probably turn the paragraph on substitutions into a table:
-having second thoughts, the table would look just like a para, no?}
+We next checked whether the association between whole-item NWR scores
+and cross-linguistic phone frequence is actually be due to frequency of
+the sounds within the language: One can suppose that sounds that occur
+more frequently across languages are also more frequent within a
+language, and therefore may be easier for children to represent and
+repeat. We estimated frequency of the phones present in the stimuli in a
+corpus of child-centered recordings (Casillas et al., 2020) by counting
+the number of word types in which they occurred, and applied the natural
+logarithm.\footnote{We also carried out this analysis using token phone
+  frequency, but this measure was not correlated with whole-item NWR
+  scores, and therefore the fact that it did not explain away the
+  predictive value of cross-linguistic phone frequency was less
+  informative than the relationship discussed in the main text, with
+  type frequencies.} As before, unattested sounds were not considered
+(i.e., they were declared NA so that they do not count for analyses).
+Phone frequencies estimated from this corpus were correlated with
+cross-linguistic phone frequencies (r(27)=0.52, =0.00). Moreover,
+averaging across items and participants, the Pearson correlation between
+scaled average corpus phone frequency and whole-item NWR scores was r =
+0.46. Therefore, we fit another mixed logistic regression, this time
+declaring as fixed effects both scaled cross-linguistic and corpus
+frequencies (averaged across all attested phones within each stimulus
+item), in addition to age. As before, the model contained random slopes
+for both child ID and target. In this model, both cross-linguistic
+frequency (ß = 0.82, SE ß = 0.27, p = 0.00) and age (ß = 0.33, SE ß =
+0.13, p = 0.01) were significant predictors of whole-item NWR scores,
+but corpus frequency (ß = 0.00, SE ß = 0.25, p = 1.00) was not.
+
+\subsubsection{Patterns in NWR
+mispronunciations}\label{patterns-in-nwr-mispronunciations}
 
@@
 dptpkpkmknmbghlv) mispronounced as simple ones (specifically
-mnlwyvdgptkfhch, 115) than vice versa (2). Simple consonants were
-mispronounced as other simple consonants quite frequently (69 simple
+mńlwyvdgptkfhch, 115) than vice versa (2). Simple consonants were
+mispronounced as other simple consonants quite frequently (68 simple
 consonants mispronounced as other simple ones, compared to 33 complex
@@
 marked with more complex consonants (specifically dptpkpkmknmbghlv)
-mispronounced as simple ones (specifically mnlwyvdgptkfhch, 115 times or
+mispronounced as simple ones (specifically mńlwyvdgptkfhch, 115 times or
 about 1.96\% out of all complex consonant targets) than vice versa (2
@@
 consonants were mispronounced as other simple consonants quite
-frequently (69 simple consonants mispronounced as other simple ones or
-about 0.39\% out of all simple consonant targets, compared to 33 complex
+frequently (68 simple consonants mispronounced as other simple ones or
+about 0.38\% out of all simple consonant targets, compared to 33 complex
 ones or about 1.02\% out of all complex consonant targets).
@@
 marked with more complex consonants (specifically dptpkpkmknmbghlv)
-mispronounced as simple ones (specifically mnlwyvdgptkfhch, 250 times or
+mispronounced as simple ones (specifically mńlwyvdgptkfhch, 250 times or
 about 2.43\% out of all complex consonant targets) than vice versa (2
@@
 consonants were mispronounced as other simple consonants quite
-frequently (123 simple consonants mispronounced as other simple ones or
-about 0.53\% out of all simple consonant targets, compared to 55 complex
+frequently (120 simple consonants mispronounced as other simple ones or
+about 0.52\% out of all simple consonant targets, compared to 55 complex
 ones or about 0.98\% out of all complex consonant targets).
@@
 Finally, we looked into the frequency with which mispronunciations
-resulted in real words. Two thirds of them were: 62\%.
+resulted in real words. Two thirds of them were: 63\%. This is a type of
+analysis that is seldom reported, and we could only find one comparison
+point: ({\textbf{???}}) found that illiterate adults NWR
+mispronunciations resulted in real words in 11.16\% of cases, whereas
+literate adults did so in only 1.71\%. The percentage we observe here is
+much higher than this, but we do not know whether age, language, or even
+test structure explains this difference.
 
@@
 but it is also the least forgiving. The pattern is less marked when
-other two scores are used, which are less sensitive to errors.
+other two scores are used, which are less sensitive to errors. Averaging
+across participants and items, the Pearson correlation between length
+(2-4 syllables) and whole-item NWR scores was r(1) = -0.91. In a
+generalized binomial mixed model, we included 479 observations, from 40
+children producing in any given trial one of 24 potential target words.
+The analysis revealed a main effect of age (ß = 0.56, SE ß = 0.14, p =
+0); and a significant estimate for the length of the target words in
+number of syllables (ß = -0.15, SE ß = 0.33, p = 0.65).
 
@@
 47 (22) & 59 (17) & 41 (17)\\
-80 (22) & 93 (9) & 7 (9)\\
-79 (19) & 93 (7) & 7 (7)\\
+79 (22) & 92 (9) & 8 (9)\\
+78 (19) & 93 (7) & 7 (7)\\
 74 (32) & 91 (12) & 9 (12)\\
@@
 significantly higher NWR scores for older children (Spearman's rank
-correlation, given inequality of variance, rho (6,496.39) = 0.43, p =
+correlation, given inequality of variance, rho (6,014.70) = 0.44, p =
 0.00). In contrast, there was no clear association between NWR scores,
-on the one hand, and sex (t (-0.61) = 27.83, p = 0.55), birth order
-(data missing for 17 children, rho = (2,652.32) = -0.15, p = 0.47) or
-maternal education (data missing for 2 children, rho = ( 9,022.76) =
-0.09, p = 0.60).
+on the one hand, and sex (t (-0.29) = 27.56, p = 0.77), birth order
+(data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39) or
+maternal education (data missing for 0 children, rho = ( 9,594.37) =
+0.10, p = 0.54).
 
@@
 \end{enumerate}
-
-\textbf{(MC: but we can try and do a cursory analysis based on the
-corpora we have from Steve and my transcription of naturalistic
-interactions!)}
 
@@
 
+\hypertarget{ref-Praat}{}
+Boersma, P., \& Weenink, D. (2020). Praat: Doing phonetics by computer
+(Version 6.1.35). Retrieved from \url{http://www.praat.org/}
+
 \hypertarget{ref-bowey2001nonword}{}
@@
 
+\hypertarget{ref-foley1986papuan}{}
+Foley, W. A. (1986). \emph{The papuan languages of new guinea}.
+Cambridge, UK: Cambridge University Press.
+
 \hypertarget{ref-gallagher2014identity}{}
@@
 
-\hypertarget{ref-levinsoninprepydgrammar}{}
-Levinson, S. C. (in preparation). \emph{A grammar of yélî dnye, the
-papuan language of rossel island}. Cambridge, UK: Cambridge University
-Press.
+\hypertarget{ref-ladefoged1996sounds}{}
+Ladefoged, P., \& Maddieson, I. (1996). \emph{The sounds of the world's
+languages} (Vol. 1012). Blackwell Oxford.
+
+\hypertarget{ref-levinsonYDgrammar}{}
+Levinson, S. C. (accepted). \emph{A grammar of yélî dnye, the papuan
+language of rossel island}. Berlin, Boston: De Gruyter Mouton.
 
@@
 
+\hypertarget{ref-maddieson2005correlating}{}
+Maddieson, I. (2005). Correlating phonological complexity: Data and
+validation. \emph{UC Berkeley PhonLab Annual Report}, \emph{1}(1).
+
+\hypertarget{ref-maddiesonIPphoneticsYD}{}
+Maddieson, I., \& Levinson, S. C. (in preparation). \emph{The phonetics
+of yélî dnye, the language of rossel island}.
+
 \hypertarget{ref-meir2017independent}{}
@@
 \url{https://phoible.org/}
+
+\hypertarget{ref-peuteIPconsonants}{}
+Peute, A. A. K., Fikkert, P., \& Casillas, M. (In preparation). Early
+consonant production in Yélî Dnye and Tseltal.
 

### 2020-12-18 manuscript.Rmd
minor changes to abstract

@@
 abstract: >
-   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the local language.
+   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language.
 NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
-In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea, with an unusually dense phonological inventory.
+In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea that has an unusually dense phonological inventory.
 This study contributes to three lines of research.
-First, we document that non-word items containing typologically rare sounds are repeated without changes less often that non-words containing more common sounds.
+First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds.
 Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
 Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
-Together, these data provide a unique view of online phonological processing in a seldom-studied language, and contribute to both typological and language acquisition research.
+Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between patterns in language development and in the frequency of linguistic features in human language at large.
 
@@
 
+<!--
 - double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
+-->
 <!--M2A: Done! Left for AC to review in segments-20201123.xlsx and, if she is satisfied, to integrate :)-->

### 2020-12-18 manuscript.Rmd
minor typo fixes

@@
 
-NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity].
+NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences].
@@
 Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation; previous NWR research has shown variable effects of length between populations.
-Third, these data give a first impression of how this task is responded to in a rarely-studied language and culture; we add to this by further investigating whether there are structured sources that account for individual variation in our sample.
+Third, these data give a first impression of how this task is responded to in a rarely-studied language and cultural setting; we add to this contribution by further investigating whether there are structured sources that account for individual variation in our sample.
 <!--M2A: This is also a note to myself; it seems we could use some more references in the above!-->
@@
 
-Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at time of writing, @levinsonYDgrammar].
+Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at time of writing; @levinsonYDgrammar].
 The vast majority of syllables use a CV format.
 A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., 'ndap' (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., 'ndapî' /'ṇṭæ.pɯ/).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., 'ndapî' /ṇṭæ.pɯ/).
 This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /'pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
+An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.læ/), but these are limited to /æ/-.
 Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
+
 <!--
-The following was removed b/c it's less relevant, but it has been double-checked by MC Nov 2020 in case we decide to include it later:
+NOTE: The following paragraph was removed b/c it's less relevant, but it has been double-checked by MC Nov 2020 in case we decide to include it later:
 
@@
 Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
-There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that, even when there is deisel available, are primarily reserved for medical transport.
+There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available.
 There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG.
@@
 While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day).
-It is very common for households situated close to schools to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year.
+It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year.
 Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
@@
 In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
-For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biilik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length.
+For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length.
 There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
@@
 Previous work typically steers clear of articulartorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology.
-We ultimately used a relatively large number of items that would enable us to explor both variation in structural complexity and in more vs.
+We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.
 less challenging sounds.
 However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
-Naturally, designing the task in this way may render the study of individual variation within the population more difficult because different children are exposed to different items.
+Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
 However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit.
@@
 
-- Does the cross-linguistic frequency of sounds predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
+- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
 - How does score change as a function of item length in number of syllables?
-- Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?
+- Is individual variation in NWR scores attributable to child age, sex, birth order, monolingual status, and/or parental education?
 

### 2020-12-18 manuscript.tex
minor typo fixes

@@
 auditorily with an item that is phonologically legal but lexically
-meaningless in the local language. NWR scores are thought to reflect
+meaningless in their language. NWR scores are thought to reflect
 long-term phonological knowledge as well as online phonological working
@@
 results among children learning Yélî Dnye, an isolate spoken on Rossel
-Island in Papua New Guinea, with an unusually dense phonological
+Island in Papua New Guinea that has an unusually dense phonological
 inventory. This study contributes to three lines of research. First, we
-document that non-word items containing typologically rare sounds are
-repeated without changes less often that non-words containing more
-common sounds. Second, we document rather weak effects of item length,
-contributing to other research suggesting that length effects may be
-language-specific. Third, we do not find strong individual variation
-effects in this population, contrary to previous results documenting
-strong age-related effects. Together, these data provide a unique view
-of online phonological processing in a seldom-studied language, and
-contribute to both typological and language acquisition research.
+document that non-word items containing typologically frequent sounds
+are repeated without changes more often that non-words containing
+typologically rare sounds. Second, we document rather weak effects of
+item length, contributing to other research suggesting that length
+effects may be language-specific. Third, we do not find strong
+individual variation effects in this population, contrary to previous
+results documenting strong age-related effects. Together, these data
+provide a unique view of online phonological processing in an
+understudied language while making preliminary connections between
+patterns in language development and in the frequency of linguistic
+features in human language at large.
 
@@
 \tightlist
-\item
-  double check my frequency entries in segments.xlsx using the search
-  function in \url{https://phoible.org/parameters}; take a look at
-  \url{http://phoible.github.io/conventions/} in case you see something
-  about double articulation (I didn't find tp or lBj) 
 \item
@@
 including what the links are between phonology, working memory, and the
-lexicon (Bowey, 2001), and how extensively apparent phonological
-constraints found in the lexicon affect online production (Gallagher,
-2014). NWR is also frequently used in applied contexts, notably as a
-diagnostic tool for language delays and disorders (Estes, Evans, \&
-Else-Quest, 2007). Since non-words can be generated in any language, it
-has attracted the attention of researchers working in multilingual and
-linguistically diverse environments, particularly in Europe (Action,
-2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has seldom been
-used outside of Europe and North America (with exceptions including
-Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz
-(2020)). Further, the languages it has been used with have historically
-had small to moderate phonological inventories (Maddieson, 2005).
+lexicon (Bowey, 2001), and how extensively phonological constraints
+found in the lexicon affect online production (Gallagher, 2014). NWR is
+also frequently used in applied contexts, notably as a diagnostic tool
+for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
+Since non-words can be generated in any language, it has attracted the
+attention of researchers working in multilingual and linguistically
+diverse environments, particularly in Europe (Action, 2009; Meir,
+Walters, \& Armon-Lotem, 2016). However, it has seldom been used outside
+of Europe and North America (with exceptions including Gallagher, 2014,
+and Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)). Further,
+the languages it has been used with have historically had small to
+moderate phonological inventories (Maddieson, 2005).
 
@@
 first impression of how this task is responded to in a rarely-studied
-language and culture; we add to this by further investigating whether
-there are structured sources that account for individual variation in
-our sample.
+language and cultural setting; we add to this contribution by further
+investigating whether there are structured sources that account for
+individual variation in our sample.
 
@@
 (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most
-recent Yélî Dnye dictionary at time of writing, Levinson, accepted). The
+recent Yélî Dnye dictionary at time of writing; Levinson, accepted). The
 vast majority of syllables use a CV format. A small portion of the
@@
 spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
-coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
+coda, yielding a CV.CV structure instead (e.g., `ndapî' /ṇṭæ.pɯ/). This
 process is used frequently with English loan words that have a coda
-(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
-features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
+(e.g., `ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
+features words starting with a vowel (e.g., 'ala' (here) /æ.læ/), but
 these are limited to /æ/-. Finally, the lexicon features a handful of
@@
 and irregular. There is no electricity system (we use solar panels) or
-motor vehicles on the island, save dinghys with outboard motors that,
-even when there is deisel available, are primarily reserved for medical
-transport. There is a medical radio connection at the health clinic,
-located at the Catholic Mission, via which news and messages are often
-passed to and from mainland PNG. There is also a cellular tower on the
-island, though it is often inoperative, usually for multiple months or
-years at a time. As such, access to news and outside connections,
-including international phone calls and digital exchanges that require
-significant data transfer, is typically unavailable. Our data collection
-is therefore limited to the duration of the researchers' on-island
-visits.
+motor vehicles on the island, save dinghys with outboard motors that are
+primarily reserved for medical transport when there is diesel available.
+There is a medical radio connection at the health clinic, located at the
+Catholic Mission, via which news and messages are often passed to and
+from mainland PNG. There is also a cellular tower on the island, though
+it is often inoperative, usually for multiple months or years at a time.
+As such, access to news and outside connections, including international
+phone calls and digital exchanges that require significant data
+transfer, is typically unavailable. Our data collection is therefore
+limited to the duration of the researchers' on-island visits.
 
@@
 well out of walking distance for many children (i.e., more than 1 hour
-on foot or by canoe each day). It is very common for households situated
-close to schools to billet their school-aged relatives during the
-weekdays (sometimes the weekends too, if the child's home is very far)
-for long segements of each school year. Combined with practices of
+on foot or by canoe each day). It is very common for households that are
+situated close to a school to billet their school-aged relatives during
+the weekdays (sometimes the weekends too, if the child's home is very
+far) for long segements of each school year. Combined with practices of
 collective child guardianship within each close-knit hamlet, the
@@
 language. For instance, in English, a response of {[}bilig{]} or
-{[}pilik{]} would be scored as incorrect; a response {[}biilik{]}, where
+{[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where
 the vowel is lengthened without change of quality would be scored as
@@
 to more adequately represent Yélî Dnye's phonology. We ultimately used a
-relatively large number of items that would enable us to explor both
+relatively large number of items that would enable us to explore both
 variation in structural complexity and in more vs.~less challenging
@@
 task longer and more tiresome, we split items across children (see
-below). Naturally, designing the task in this way may render the study
-of individual variation within the population more difficult because
+below). Naturally, designing the task in this way may make the study of
+individual variation within the population more difficult because
 different children are exposed to different items. However, a review of
@@
 \item
-  Does the cross-linguistic frequency of sounds predict NWR scores? Are
-  rarer sounds more often substituted by commoner sounds?
+  Does the cross-linguistic frequency of sounds in the stimuli predict
+  NWR scores? Are rarer sounds more often substituted by commoner
+  sounds?
 \item
@@
 \item
-  Is individual variation attributable to child age, sex, birth order,
-  monolingual status, and/or parental education?
+  Is individual variation in NWR scores attributable to child age, sex,
+  birth order, monolingual status, and/or parental education?
 \end{itemize}

### 2020-12-18 manuscript.Rmd
make all legends fit on fig1

@@
   theme_bw()+
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.key.height = unit(5, "mm"))+
   scale_x_continuous(name = "Age (years) ", limits = c(4,12), breaks = c(4,8,12))+

### 2020-12-18 manuscript.tex
minor typo fixes

@@
 spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
-coda, yielding a CV.CV structure instead (e.g., `ndapî' /ṇṭæ.pɯ/). This
+coda, yielding a CV.CV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
 process is used frequently with English loan words that have a coda
-(e.g., `ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
-features words starting with a vowel (e.g., 'ala' (here) /æ.læ/), but
+(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
+features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
 these are limited to /æ/-. Finally, the lexicon features a handful of
@@
 (2018); Italian data from Piazzalunga et al. (2019); Tsimane' from
-Cristia et al. (2020); Yélî data from the present study.}
+Cristia et al. (2020); Yélî Dnye data from the present study.}
 \end{figure}
@@
 second author presented them orally to three local research assistants,
-all native speakers of Yélî Dnye, who were repeated each form, as they
-would in an NWR task, and additinally let the experimenter know if the
-item was in fact a word or phrase in Yélî Dnye. Any item reported to
-have a meaning or a strong association with another word form or meaning
-was excluded.
+all native speakers of Yélî Dnye, who repeated each form as they would
+in an NWR task and additinally let the experimenter know if the item was
+in fact a word or phrase in Yélî Dnye. Any item reported to have a
+meaning or a strong association with another word form or meaning was
+excluded.
 
@@
 elicitation sets. The 40 elicitation sets are available online from
-\url{https://osf.io/5qspb}. The split had the following constraints:
+\href{https://osf.io/5qspb}{osf.io/5qspb}. The split had the following
+constraints:
 
@@
   The same three items were selected as practice items and used in all
-  40 elicitation sets
+  40 elicitation sets.
 \item

### 2020-12-18 manuscript.Rmd
minor typo fixes

@@
 A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., 'ndap' (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., 'ndapî' /ṇṭæ.pɯ/).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., 'ndapî' /'ṇṭæ.pɯ/).
 This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /'pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.læ/), but these are limited to /æ/-.
+An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
 Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
@@
 (2019); Tsimane' from Cristia et al.
-(2020); Yélî data from the present study."}
+(2020); Yélî Dnye data from the present study."}
 
@@
 These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
-The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who were repeated each form, as they would in an NWR task, and additinally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
+The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additinally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
 Any item reported to have a meaning or a strong association with another word form or meaning was excluded.
@@
 A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
-The 40 elicitation sets are available online from https://osf.io/5qspb.
+The 40 elicitation sets are available online from [osf.io/5qspb](https://osf.io/5qspb).
 The split had the following constraints:
 
-- The same three items were selected as practice items and used in all 40 elicitation sets
+- The same three items were selected as practice items and used in all 40 elicitation sets.
 - Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sublists).
@@
 Because space availability was limited in different ways from hamlet to hamlet, the places where elicitation happened varied across testing sites.
-We tested in four different sites, only making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items 'spreading' between children through hearsay.
+We tested in four different sites, only making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from 'spreading' between children through hearsay.
 In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
-The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from online supplementary materials.
-Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitaiton sets.
+The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from the online supplementary materials.
+Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
+<!--M2A:
+We took out the actual hamlet names, right? So what do you want to put in the metadata mentioned in the sentence:
+"The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from the online supplementary materials."
+-->
 
@@
 A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement.
-The research assistant coached the child throughout to task to make sure that they understood what they were expected to do.
+The research assistant coached the child throughout the task to make sure that they understood what they were expected to do.
 An experimenter (the first author) delivered the pre-recorded stimuli to the research assistant and the child over headphones.
@@
 The first phase of the experiment involved making sure the child understood the task.
-We explained the task orally presented the first practice item.
-At this point, many children did not say anything in response, at which point we initiated the following procedure: First, the assistant insisted the child make a response.
+We explained the task and then orally presented the first practice item.
+At this point, many children did not say anything in response, which triggered the following procedure: First, the assistant insisted the child make a response.
 If the child still did not say anything, the assistant said a real word and then asked the child to repeat it, then another and another.
@@
 Most children successfully started repeating the items at this point, but a few needed further help.
-In this case, the assistant modeled the behavior (i.e., the child and assistant would hear the item again, and the assistant would repeat it; then we would play the item again and ask the child to repeate it).
+In this case, the assistant modeled the behavior (i.e., the child and assistant would hear the item again, and the assistant would repeat it; then we would play the item again and ask the child to repeat it).
 A small minority of children still failed to repeat the item at this point.
-If so, we tried again with the second training item, at which point some children got it and could continue.
+If so, we tried again with the second training item, at which point some children demonstrated task understanding and could continue.
 A fraction of the remaining children, however, failed to repeat this second training item, as well as the third one, in which case we stopped testing altogether (see Participants section for exclusions).
@@
 We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was alowed to make a further attempts if their first attempt was judged erroneous in some way by the assistant.
-The procedure worked as follows: When the child made an attempted a repeat, the assistant indicated to the experimenter whether the child's production was correct or not.
-If correct, the experimenter would whisper this note of correct repetition into a separate headset that fed into the right channel of the same Tascam and the group moved on to the next item.
+The procedure worked as follows: When the child made an attempt, the assistant indicated to the experimenter whether the child's production was correct or not.
+If correct, the experimenter would whisper this note of correct repetition into a separate headset that fed into the right channel of the same Tascam recorder and the group moved on to the next item.
 If not, the child was allowed to try again, with up to five attempts allowed before moving on to the next item.
 Children were not asked to make repetitions if they did not produce a first attempt.
-
 In total, test sessions took approximately XX minutes on average, with XX of those minutes attributed to practice and XX to the actual test list.
@@
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time these data were analyzed, it is of course still far from complete, so estimated frequencies should be taken with a pinch of salt.
-In both PHOIBLE and UPSID---another large phonemic inventory database---note that nearly half of the segment types only appear in a single language (Steven Moran, personal communication; see also [web.phonetik.uni-frankfurt.de/upsid_info.html](http://web.phonetik.uni-frankfurt.de/upsid_info.html)).
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at time of writing, it is of course still far from complete, so estimated frequencies should be taken with a pinch of salt.
+In both PHOIBLE and UPSID---another large phonemic inventory database---note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication; see also [web.phonetik.uni-frankfurt.de/upsid_info.html](http://web.phonetik.uni-frankfurt.de/upsid_info.html)).
 It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database.
@@
 We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlets.
-We excluded test sessions from analysis for the following reasons: refused participation or failed to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease).
+We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease).
 The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years).
-Among these children were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to one or more other languages at home.
+Among these children were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.
 Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^1] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth.
@@
 We first checked whether whole-item NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
-We excluded `r excl` children who did not have data for one of these two types.
+We excluded `r excl` child who did not have data for one of these two types.
 As shown in Figure \@ref(fig:Fig1-first-vs-others), participants' mean word-level scores became more heterogeneous in subsequent repetitions.

### 2020-12-19 manuscript.Rmd
minor typo fixes, pval replacement, & prettier phone list printing
also i decided on a version of option 2 for deletions/substitutions

@@
 library(lme4)
+
+p.levels <- c(0, 0.001, 0.01)
+pval.display <- function(p.est) {
+  level <- findInterval(p.est, p.levels)
+  p.string <- ifelse(level == 3, paste0("= ", round(p.est, 2)),
+    ifelse(level == 2, "< 0.01", "< 0.001"))
+  return(p.string)
+}
+
+print.phone.list <- function(charlist) {
+  printout <- ""
+  for (c in 1:length(charlist)) {
+    if (c == length(charlist)) {
+      printout <- paste0(printout, "and [", charlist[c], "]")
+    } else {
+      printout <- paste0(printout, "[", charlist[c], "], ")
+    }
+  }
+  return(printout)
+}
 
@@
 Surprisingly, whole-item NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
-were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p = `r round(myt$p.value,2)`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`).
+were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p `r pval.display(round(myt$p.value,2))`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`).
 Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions.
@@
 Taking into account only the first attempts, we averaged attempts by each of the `r dim(overall)[1]` children who had data for first attempts; their ages ranged from `r min(overall$Age)` to `r max(overall$Age)` (M = `r round(mean(overall$Age),1)`, SD = `r round(sd(overall$Age),1)`).
-
 The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`.
@@
 
-In this analysis, we were interested in variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
-To look at this, we fit a mixed logistic regression, in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effects of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID & target ID as random variables.
+We then analyzed variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
+The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID & target ID as random variables.
 
@@
 The analysis revealed a main effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
-Error"]`, p = `r mycoeff["age.rounded_s","Pr(>|z|)"]`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
-Error"]`, p = `r mycoeff["avg_fr_s","Pr(>|z|)"]`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as clear in Figure 2.
-Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.
+Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
+Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as shown in Figure \@ref(fig:Fig2-xling-freq).
+Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.
 
@@
   ylab("NWR scores (% words)") +
-  xlab("Average phone cross-ling frequency") +
+  xlab("Average cross-linguistic phone frequency") +
   theme_apa()
@@
 
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence is actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
 We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^2] As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, =`r round(freq_cor$p.value,3)`).
+Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, p `r pval.display(round(freq_cor$p.value,3))`).
 Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = `r round(cor.test(cor_fr_tab$corf,cor_fr_tab$Word)$estimate,3)`.
@@
 In this model, both cross-linguistic frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std.
-Error"]`, p = `r both_coeff["avg_fr_s","Pr(>|z|)"]`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std.
-Error"]`, p = `r both_coeff["age.rounded_s","Pr(>|z|)"]`) were significant predictors of whole-item  NWR scores, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
-Error"]`, p = `r both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"]`) was not.
+Error"]`, p `r pval.display(both_coeff["avg_fr_s","Pr(>|z|)"])`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std.
+Error"]`, p `r pval.display(both_coeff["age.rounded_s","Pr(>|z|)"])`) were significant predictors of whole-item  NWR scores, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
+Error"]`, p `r pval.display(both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"])`) was not.
+
+<!--M2A: Above it says "As before, unattested sounds were not considered" -- well now we have taken care of PHOIBLE unattested sounds with a frequency of 1. Do we need to change this?-->
 
@@
 complexCs=c("dp","tp","kp","km","kn","mb","gh","lv")
+simplexCs.ipa=c("m","n","l","w","j","w","ṭ","g","p","t","k","f","h","tʃ")
+complexCs.ipa=c("ṭp","tp","kp","km","kṇ","mp","ɣ","lβʲ")
 
@@
 
+<!--
+M2A: I prefer a modification of option 2 and so am commenting this out:
+
 **Option 1: just numbers, first attempts only **
@@
 **Option 2: numbers and proportions, first attempts only**
+-->
+
 Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted (about `r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` consonants deleted (about `r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral target vowels produced as nasal vowels or about `r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% out of all oral vowel targets, `r sum(sub_tab[nasalVs,oralVs])` nasal target vowels produced as oral vowels or about `r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (`r sum(sub_tab[oralVs,oralVs])` oral vowels produced with a different quality or about `r mean(sub_pc[oralVs,oralVs],na.rm=T)`% out of all oral vowel targets; `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels produced with a different quality or about `r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically `r paste0(complexCs)`) mispronounced as simple ones (specifically `r paste0(simplexCs)`, `r sum(sub_tab[complexCs,simplexCs])` times or  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all complex consonant targets) than vice versa (`r sum(sub_tab[simplexCs,complexCs])` times or  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (`r sum(sub_tab[simplexCs,simplexCs])` simple consonants mispronounced as other simple ones or about `r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% out of all simple consonant targets, compared to `r sum(sub_tab[complexCs,complexCs])` complex ones or about `r mean(sub_pc[complexCs,complexCs],na.rm=T)`% out of all complex consonant targets).
+Deletions were relatively rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
+Substitutions were somewhat more frequent, but still relatively rare.
+Nasal vowels were produced as oral vowels and vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral targets as nasal (~`r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% of targets) and `r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral (~`r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% of targets).
+Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (`r sum(sub_tab[oralVs,oralVs])` oral vowels (~`r mean(sub_pc[oralVs,oralVs],na.rm=T)`% of targets and `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels (~`r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% of targets).
+Among consonants, complex consonants were occasionally substituted by simpler ones (specifically `r print.phone.list(complexCs.ipa)` as `r print.phone.list(simplexCs.ipa)`), `r sum(sub_tab[complexCs,simplexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all complex consonant targets.
+Interestingly, simple consonants were sometimes also substituted by complex ones; this happened `r sum(sub_tab[simplexCs,complexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all simple consonant targets.
+Finally, simple consonants were mispronounced as other simple consonants (`r sum(sub_tab[simplexCs,simplexCs])` instances, ~`r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% of all simple consonant targets) and complex consonants as other complex consonants (`r sum(sub_tab[complexCs,complexCs])` instances, ~`r mean(sub_pc[complexCs,complexCs],na.rm=T)`% of all complex consonant targets).
+Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms (see below).
 
@@
 
+<!--
+M2A: I prefer a modification of option 2 and so am commenting this out:
+
 **Option 3: numbers and proportions, ALL attempts**
@@
 
+-->
+
 Finally, we looked into the frequency with which mispronunciations resulted in real words.
-Two thirds of them were: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%.
-This is a type of analysis that is seldom reported, and we could only find one comparison point: @castro1998 found that illiterate adults NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate adults did so in only `r round(2/117*100,2)`%.
-The percentage we observe here is much higher than this, but we do not know whether age, language, or even test structure explains this difference.
+In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%.
+This type of analysis is seldom reported---we could only find one comparison point: @castro1998 found that illiterate Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
+
+<!--M2A: castro1998 is missing from bib. is it this one?:
+https://academic.oup.com/brain/article/121/6/1053/280348
+-->
 
@@
 
-Next, we inspected whether NWR scores varied as a function of word length.
-Results are shown on table XX.
-This table shows that scores for monosyllables was much lower than for other lengths.
-This is likely due to the fact that the majority of monosyllables included were chosen because they had sounds that are rare in the world's languages, which may indicate that they are hard to produce or to perceive.
-
-Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole word scoring.
-This is the most commonly reported type of score, but it is also the least forgiving.
+Next, we inspected whether NWR scores varied as a function of word length (Table \@ref(tab:tablength)).
+Participants scored much lower on monosyllables than on non-words of other lengths.
+This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may indicate that they are hard to produce or perceive.
+Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole-item scoring.
+While whole-item scoring is the most commonly reported NWR outcome, it is also the least forgiving.
 The pattern is less marked when other two scores are used, which are less sensitive to errors.
 Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(`r length_r$parameter`) = `r round(length_r$estimate,3)`.
-In a generalized binomial mixed model, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing in any given trial one of `r summary(len_lm)$ngrps["target"]` potential target words.
-The analysis revealed a main effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
-Error"]`, p = `r mycoeff["age.rounded_s","Pr(>|z|)"]`); and a significant estimate for the length of the target words in number of syllables (ß = `r mycoeff["type","Estimate"]`, SE ß = `r mycoeff["type","Std.
-Error"]`, p = `r mycoeff["type","Pr(>|z|)"]`).
+In a generalized binomial mixed model, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing, in any given trial, one of `r summary(len_lm)$ngrps["target"]` potential target words.
+The analysis revealed a positive effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
+Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`) and a negative but non-significant estimate for target length in number of syllables (ß = `r mycoeff["type","Estimate"]`, SE ß = `r mycoeff["type","Std.
+Error"]`, p `r pval.display(mycoeff["type","Pr(>|z|)"])`).
 
@@
 
-```{r Fig2-scores by age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-word scores for individual participants as a function of age and sex (blue = boys, pink = girls)."}
+```{r Fig2-scores-by-age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-item scores for individual participants as a function of age and sex (purple = boys, orange = girls)."}
 
-plot(overall$Word~overall$Age,pch=20,cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","blue","pink"))
+plot(overall$Word~overall$Age,pch=20,cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","darkmagenta","chocolate1"))
 
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure 2, there was a greater deal of variance in earlier than later ages, with significantly higher NWR scores for older children  (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p = `r corage$p.value`).
-In contrast, there was no clear association between NWR scores, on the one hand, and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p = `r tsex$p.value`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r corbo$statistic`) = `r corbo$estimate`, p = `r corbo$p.value`) or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children,  rho = ( `r corme$statistic`) = `r corme$estimate`, p = `r corme$p.value`).
+As shown in Figure \@ref(fig:Fig2-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
+In contrast, there was no clear association between NWR scores and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r corbo$statistic`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children,  rho ( `r corme$statistic`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
 

### 2020-12-19 manuscript.tex
minor typo fixes, pval replacement, & prettier phone list printing
also i decided on a version of option 2 for deletions/substitutions

@@
 back-to-back testing of all eligible children present at the time of our
-visit in order to prevent the items `spreading' between children through
-hearsay. In the first hamlet, we tested children in five different
-places, with some children being tested inside a house and others tested
-on the veranda. The complete list of places and the ways in which they
-met the desiderata mentioned above can be found in the raw data,
-available from online supplementary materials. Whenever children living
-in the same household were tested, we tried to test children in age
-order, from oldest to youngest, to minimize intimidation for younger
-household members, and always using different elicitaiton sets.
+visit in order to prevent the items from `spreading' between children
+through hearsay. In the first hamlet, we tested children in five
+different places, with some children being tested inside a house and
+others tested on the veranda. The complete list of places and the ways
+in which they met the desiderata mentioned above can be found in the raw
+data, available from the online supplementary materials. Whenever
+children living in the same household were tested, we tried to test
+children in age order, from oldest to youngest, to minimize intimidation
+for younger household members, and always using different elicitation
+sets.
 
@@
 needed, encouragement. The research assistant coached the child
-throughout to task to make sure that they understood what they were
+throughout the task to make sure that they understood what they were
 expected to do. An experimenter (the first author) delivered the
@@
 The first phase of the experiment involved making sure the child
-understood the task. We explained the task orally presented the first
-practice item. At this point, many children did not say anything in
-response, at which point we initiated the following procedure: First,
-the assistant insisted the child make a response. If the child still did
-not say anything, the assistant said a real word and then asked the
-child to repeat it, then another and another. If the child could repeat
-real words correctly, we provided the first training item over
-headphones again for children to repeat. Most children successfully
-started repeating the items at this point, but a few needed further
-help. In this case, the assistant modeled the behavior (i.e., the child
-and assistant would hear the item again, and the assistant would repeat
-it; then we would play the item again and ask the child to repeate it).
-A small minority of children still failed to repeat the item at this
+understood the task. We explained the task and then orally presented the
+first practice item. At this point, many children did not say anything
+in response, which triggered the following procedure: First, the
+assistant insisted the child make a response. If the child still did not
+say anything, the assistant said a real word and then asked the child to
+repeat it, then another and another. If the child could repeat real
+words correctly, we provided the first training item over headphones
+again for children to repeat. Most children successfully started
+repeating the items at this point, but a few needed further help. In
+this case, the assistant modeled the behavior (i.e., the child and
+assistant would hear the item again, and the assistant would repeat it;
+then we would play the item again and ask the child to repeat it). A
+small minority of children still failed to repeat the item at this
 point. If so, we tried again with the second training item, at which
-point some children got it and could continue. A fraction of the
-remaining children, however, failed to repeat this second training item,
-as well as the third one, in which case we stopped testing altogether
-(see Participants section for exclusions).
+point some children demonstrated task understanding and could continue.
+A fraction of the remaining children, however, failed to repeat this
+second training item, as well as the third one, in which case we stopped
+testing altogether (see Participants section for exclusions).
 
@@
 assistant. The procedure worked as follows: When the child made an
-attempted a repeat, the assistant indicated to the experimenter whether
-the child's production was correct or not. If correct, the experimenter
-would whisper this note of correct repetition into a separate headset
-that fed into the right channel of the same Tascam and the group moved
+attempt, the assistant indicated to the experimenter whether the child's
+production was correct or not. If correct, the experimenter would
+whisper this note of correct repetition into a separate headset that fed
+into the right channel of the same Tascam recorder and the group moved
 on to the next item. If not, the child was allowed to try again, with up
@@
 were not asked to make repetitions if they did not produce a first
-attempt.
-
-In total, test sessions took approximately XX minutes on average, with
-XX of those minutes attributed to practice and XX to the actual test
-list.
+attempt. In total, test sessions took approximately XX minutes on
+average, with XX of those minutes attributed to practice and XX to the
+actual test list.
 
@@
 comprehensive database, with phonological inventory data for over 2000
-languages at the time these data were analyzed, it is of course still
-far from complete, so estimated frequencies should be taken with a pinch
-of salt. In both PHOIBLE and UPSID---another large phonemic inventory
-database---note that nearly half of the segment types only appear in a
-single language (Steven Moran, personal communication; see also
+languages at time of writing, it is of course still far from complete,
+so estimated frequencies should be taken with a pinch of salt. In both
+PHOIBLE and UPSID---another large phonemic inventory database---note
+that nearly half of the segment types are only attested in one language
+(Steven Moran, personal communication; see also
 \href{http://web.phonetik.uni-frankfurt.de/upsid_info.html}{web.phonetik.uni-frankfurt.de/upsid\_info.html}).
@@
 hamlets. We excluded test sessions from analysis for the following
-reasons: refused participation or failed to repeat items presented over
+reasons: refused participation or failure to repeat items presented over
 headphones even after coaching (N=8), spoke too softly to allow offline
@@
 children were 34 only exposed to Yélî Dnye at home and 6 children
-exposed to one or more other languages at home. Maternal years of
-education averaged 8.22 years (range 6-12 years).\footnote{Years of
-  education is difficult to precisely pinpoint in this context.
+exposed to Yélî Dnye plus one or more other languages at home. Maternal
+years of education averaged 8.22 years (range 6-12 years).\footnote{Years
+  of education is difficult to precisely pinpoint in this context.
   Schooling beyond grade 8 takes place off-island and enrollment is tied
@@
 the participant level separately for first attempts and subsequent
-repetitions. We excluded 1 children who did not have data for one of
-these two types. As shown in Figure \ref{fig:Fig1-first-vs-others},
+repetitions. We excluded 1 child who did not have data for one of these
+two types. As shown in Figure \ref{fig:Fig1-first-vs-others},
 participants' mean word-level scores became more heterogeneous in
@@
 subsequent repetitions (M = 40, SD = 26) were on average lower than
-first ones (M = 64, SD = 15), t(38) = 6.28, p = 0; Cohen's d = 1.14).
-Given uncertainty in whether previous work used first or all
+first ones (M = 64, SD = 15), t(38) = 6.28, p \textless{} 0.001; Cohen's
+d = 1.14). Given uncertainty in whether previous work used first or all
 repetitions, and given that score here declined and became more
@@
 each of the 40 children who had data for first attempts; their ages
-ranged from 3 to 10 (M = 6.50, SD = 1.50).
-
-The overall NWR score was M = 64\% (SD = 15\%), Cohen's d = 4.36. Scores
-based on phonemes are even higher M = 78\% (SD = 9\%), Cohen's d = 8.55.
-The phoneme-based normalized Levenshtein distance was M = 22\% (SD =
-9\%), meaning that about a fifth of phonemes were were substituted,
-inserted, or deleted (the normalized Levenshtein distance is the
-complement of phoneme-based scores).
+ranged from 3 to 10 (M = 6.50, SD = 1.50). The overall NWR score was M =
+64\% (SD = 15\%), Cohen's d = 4.36. Scores based on phonemes are even
+higher M = 78\% (SD = 9\%), Cohen's d = 8.55. The phoneme-based
+normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that
+about a fifth of phonemes were were substituted, inserted, or deleted
+(the normalized Levenshtein distance is the complement of phoneme-based
+scores).
 
@@
 
-In this analysis, we were interested in variation in whole-item NWR
-scores as a function of the average frequency with which sounds
-composing individual target words are found in languages over the world.
-To look at this, we fit a mixed logistic regression, in which the
-outcome variable was whether the non-word was correctly repeated or not.
-The fixed effects of interest was the average cross-linguistic phone
-frequency; we also included child age as a control fixed effect, and
-child ID \& target ID as random variables.
+We then analyzed variation in whole-item NWR scores as a function of the
+average frequency with which sounds composing individual target words
+are found in languages over the world. To look at this, we fit a mixed
+logistic regression in which the outcome variable was whether the
+non-word was correctly repeated or not. The fixed effect of interest was
+the average cross-linguistic phone frequency; we also included child age
+as a control fixed effect, and child ID \& target ID as random
+variables.
 
@@
 estimate for the scaled average cross-linguistic frequency of phones in
-the target words (ß = 0.82, SE ß = 0.18, p = 0): Target words with
-phones found more frequently across languages had a higher proportion of
-words that were correctly repeated, as clear in Figure 2. Averaging
-across items and participants, the Pearson correlation between scaled
-average cross-linguistic phone frequency and whole-item NWR scores was r
-r(31) = 0.63.
+the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target
+words with phones found more frequently across languages had a higher
+proportion of words that were correctly repeated, as shown in Figure
+\ref{fig:Fig2-xling-freq}. Averaging across items and participants, the
+Pearson correlation between scaled average cross-linguistic phone
+frequency and whole-item NWR scores was r(31) = 0.63.
 
@@
 We next checked whether the association between whole-item NWR scores
-and cross-linguistic phone frequence is actually be due to frequency of
-the sounds within the language: One can suppose that sounds that occur
-more frequently across languages are also more frequent within a
+and cross-linguistic phone frequence could actually be due to frequency
+of the sounds within the language: One can suppose that sounds that
+occur more frequently across languages are also more frequent within a
 language, and therefore may be easier for children to represent and
@@
 Phone frequencies estimated from this corpus were correlated with
-cross-linguistic phone frequencies (r(27)=0.52, =0.00). Moreover,
-averaging across items and participants, the Pearson correlation between
-scaled average corpus phone frequency and whole-item NWR scores was r =
-0.46. Therefore, we fit another mixed logistic regression, this time
-declaring as fixed effects both scaled cross-linguistic and corpus
-frequencies (averaged across all attested phones within each stimulus
-item), in addition to age. As before, the model contained random slopes
-for both child ID and target. In this model, both cross-linguistic
-frequency (ß = 0.82, SE ß = 0.27, p = 0.00) and age (ß = 0.33, SE ß =
-0.13, p = 0.01) were significant predictors of whole-item NWR scores,
-but corpus frequency (ß = 0.00, SE ß = 0.25, p = 1.00) was not.
+cross-linguistic phone frequencies (r(27)=0.52, p \textless{} 0.01).
+Moreover, averaging across items and participants, the Pearson
+correlation between scaled average corpus phone frequency and whole-item
+NWR scores was r = 0.46. Therefore, we fit another mixed logistic
+regression, this time declaring as fixed effects both scaled
+cross-linguistic and corpus frequencies (averaged across all attested
+phones within each stimulus item), in addition to age. As before, the
+model contained random slopes for both child ID and target. In this
+model, both cross-linguistic frequency (ß = 0.82, SE ß = 0.27, p
+\textless{} 0.01) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were
+significant predictors of whole-item NWR scores, but corpus frequency (ß
+= 0.00, SE ß = 0.25, p = 1) was not.
 
@@
 
-\textbf{Option 1: just numbers, first attempts only } Next, we
-investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 7 vowels deleted, and 3 consonants.
-
-As for substitutions, it was as common for a nasal vowel to be produced
-as an oral vowel as vice versa (34 oral target vowels produced as nasal
-vowels, 30 nasal target vowels produced as oral vowels). Substitutions
-in which the oral nature was preserved but the quality of the vowel was
-changed were a great deal more common than changes in quality among
-nasal vowels (110 oral vowels produced with a different quality; 13
-nasal vowels produced with a different quality). As for consonants,
-asymmetries were very marked with more complex consonants (specifically
-dptpkpkmknmbghlv) mispronounced as simple ones (specifically
-mńlwyvdgptkfhch, 115) than vice versa (2). Simple consonants were
-mispronounced as other simple consonants quite frequently (68 simple
-consonants mispronounced as other simple ones, compared to 33 complex
-ones).
-
-\textbf{Option 2: numbers and proportions, first attempts only} Next, we
-investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 7 vowels deleted (about 0.23\% of all vowel
-targets), and 3 consonants deleted (about 0.12\% of all consonant
-targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced
-as an oral vowel as vice versa (34 oral target vowels produced as nasal
-vowels or about 0.99\% out of all oral vowel targets, 30 nasal target
-vowels produced as oral vowels or about 2.37\% out of all nasal vowel
-targets). Substitutions in which the oral nature was preserved but the
-quality of the vowel was changed were a great deal more common than
-changes in quality among nasal vowels (110 oral vowels produced with a
-different quality or about 1.86\% out of all oral vowel targets; 13
-nasal vowels produced with a different quality or about 2.06\% out of
-all nasal vowel targets). As for consonants, asymmetries were very
-marked with more complex consonants (specifically dptpkpkmknmbghlv)
-mispronounced as simple ones (specifically mńlwyvdgptkfhch, 115 times or
-about 1.96\% out of all complex consonant targets) than vice versa (2
-times or about 1.96\% out of all simple consonant targets). Simple
-consonants were mispronounced as other simple consonants quite
-frequently (68 simple consonants mispronounced as other simple ones or
-about 0.38\% out of all simple consonant targets, compared to 33 complex
-ones or about 1.02\% out of all complex consonant targets).
-
-\textbf{Option 3: numbers and proportions, ALL attempts} Next, we
-investigated patterns of deletion and substitution. Deletions were
-relatively rare, with only 12 vowels deleted (about 0.28\% of all vowel
-targets), and 6 consonants deleted (about 0.19\% of all consonant
-targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced
-as an oral vowel as vice versa (52 oral target vowels produced as nasal
-vowels or about 1.10\% out of all oral vowel targets, 58 nasal target
-vowels produced as oral vowels or about 2.53\% out of all nasal vowel
-targets). Substitutions in which the oral nature was preserved but the
-quality of the vowel was changed were a great deal more common than
-changes in quality among nasal vowels (205 oral vowels produced with a
-different quality or about 2.16\% out of all oral vowel targets; 23
-nasal vowels produced with a different quality or about 2.36\% out of
-all nasal vowel targets). As for consonants, asymmetries were very
-marked with more complex consonants (specifically dptpkpkmknmbghlv)
-mispronounced as simple ones (specifically mńlwyvdgptkfhch, 250 times or
-about 2.43\% out of all complex consonant targets) than vice versa (2
-times about 2.43\% out of all simple consonant targets). Simple
-consonants were mispronounced as other simple consonants quite
-frequently (120 simple consonants mispronounced as other simple ones or
-about 0.52\% out of all simple consonant targets, compared to 55 complex
-ones or about 0.98\% out of all complex consonant targets).
+Next, we investigated patterns of deletion and substitution. Deletions
+were relatively rare: there were only 7 instances of deleted vowels
+(\textasciitilde{}0.23\% of all vowel targets), and 3 instances of
+deleted consonants (\textasciitilde{}0.12\% of all consonant targets).
+Substitutions were somewhat more frequent, but still relatively rare.
+Nasal vowels were produced as oral vowels and vice versa (34 oral
+targets as nasal (\textasciitilde{}0.99\% of targets) and 30 nasal
+targets produced as oral (\textasciitilde{}2.37\% of targets). Both oral
+and nasal vowels were also occasionally subsituted by vowels of a
+different quality that preserved the target oral/nature airflow (110
+oral vowels (\textasciitilde{}1.86\% of targets and 13 nasal vowels
+(\textasciitilde{}2.06\% of targets). Among consonants, complex
+consonants were occasionally substituted by simpler ones (specifically
+{[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and
+{[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]},
+{[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and
+{[}tʃ{]}), 115 times, in approximately 1.96\% of all complex consonant
+targets. Interestingly, simple consonants were sometimes also
+substituted by complex ones; this happened 2 times, in approximately
+1.96\% of all simple consonant targets. Finally, simple consonants were
+mispronounced as other simple consonants (68 instances,
+\textasciitilde{}0.38\% of all simple consonant targets) and complex
+consonants as other complex consonants (33 instances,
+\textasciitilde{}1.02\% of all complex consonant targets). Note that the
+substitutions included phones that are not native to Yélî Dnye but do
+occur in English (e.g., {[}tʃ{]}); as these data come from careful
+transcriptions by a native Yélî Dnye speaker who is very fluent in
+English, we take these segmental substitutions faithfully as an
+indication that several of our participants have mastered production of
+some English phones, possibly produced within whole English word forms
+(see below).
 
 Finally, we looked into the frequency with which mispronunciations
-resulted in real words. Two thirds of them were: 63\%. This is a type of
-analysis that is seldom reported, and we could only find one comparison
-point: ({\textbf{???}}) found that illiterate adults NWR
-mispronunciations resulted in real words in 11.16\% of cases, whereas
-literate adults did so in only 1.71\%. The percentage we observe here is
-much higher than this, but we do not know whether age, language, or even
+resulted in real words. In fact, two thirds of incorrect repetitions
+were recognizable as real words or phrases in Yélî Dnye or English:
+63\%. This type of analysis is seldom reported---we could only find one
+comparison point: ({\textbf{???}}) found that illiterate Portuguese
+adults' NWR mispronunciations resulted in real words in 11.16\% of
+cases, whereas literate participants did so in only 1.71\% of cases. The
+percentage we observe here is much higher than reported in Castro and
+colleagues' study, but we do not know whether age, language, or even
 test structure explains this difference.
@@
 Next, we inspected whether NWR scores varied as a function of word
-length. Results are shown on table XX. This table shows that scores for
-monosyllables was much lower than for other lengths. This is likely due
-to the fact that the majority of monosyllables included were chosen
-because they had sounds that are rare in the world's languages, which
-may indicate that they are hard to produce or to perceive.
-
-Setting monosyllables aside, we observe the typical pattern of lower
-scores for longer items, although this is particularly salient for the
-whole word scoring. This is the most commonly reported type of score,
-but it is also the least forgiving. The pattern is less marked when
-other two scores are used, which are less sensitive to errors. Averaging
-across participants and items, the Pearson correlation between length
-(2-4 syllables) and whole-item NWR scores was r(1) = -0.91. In a
-generalized binomial mixed model, we included 479 observations, from 40
-children producing in any given trial one of 24 potential target words.
-The analysis revealed a main effect of age (ß = 0.56, SE ß = 0.14, p =
-0); and a significant estimate for the length of the target words in
-number of syllables (ß = -0.15, SE ß = 0.33, p = 0.65).
+length (Table \ref{tab:tablength}). Participants scored much lower on
+monosyllables than on non-words of other lengths. This is likely due to
+the fact that the majority of monosyllables were designed to include
+sounds that are rare in the world's languages, which may indicate that
+they are hard to produce or perceive. Setting monosyllables aside, we
+observe the typical pattern of lower scores for longer items, although
+this is particularly salient for the whole-item scoring. While
+whole-item scoring is the most commonly reported NWR outcome, it is also
+the least forgiving. The pattern is less marked when other two scores
+are used, which are less sensitive to errors. Averaging across
+participants and items, the Pearson correlation between length (2-4
+syllables) and whole-item NWR scores was r(1) = -0.91. In a generalized
+binomial mixed model, we included 479 observations, from 40 children
+producing, in any given trial, one of 24 potential target words. The
+analysis revealed a positive effect of age (ß = 0.56, SE ß = 0.14, p
+\textless{} 0.001) and a negative but non-significant estimate for
+target length in number of syllables (ß = -0.15, SE ß = 0.33, p = 0.65).
 
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/Fig2-scores by age-1.pdf}
-\caption{(\#fig:Fig2-scores by age)NWR whole-word scores for individual
-participants as a function of age and sex (blue = boys, pink = girls).}
+\includegraphics{manuscript_files/figure-latex/Fig2-scores-by-age-1.pdf}
+\caption{\label{fig:Fig2-scores-by-age}NWR whole-item scores for individual
+participants as a function of age and sex (purple = boys, orange =
+girls).}
 \end{figure}
@@
 structured by factors that vary across individuals. As shown in Figure
-2, there was a greater deal of variance in earlier than later ages, with
-significantly higher NWR scores for older children (Spearman's rank
-correlation, given inequality of variance, rho (6,014.70) = 0.44, p =
-0.00). In contrast, there was no clear association between NWR scores,
-on the one hand, and sex (t (-0.29) = 27.56, p = 0.77), birth order
-(data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39) or
-maternal education (data missing for 0 children, rho = ( 9,594.37) =
-0.10, p = 0.54).
+\ref{fig:Fig2-scores-by-age}, there was a greater deal of variance
+across the tested age range, with significantly higher NWR scores for
+older children (Spearman's rank correlation, given inequality of
+variance, rho (6,014.70) = 0.44, p \textless{} 0.01). In contrast, there
+was no clear association between NWR scores and sex (t (-0.29) = 27.56,
+p = 0.77), birth order (data missing for 15 children, rho = (3,441.90) =
+-0.18, p = 0.39), or maternal education (data missing for 0 children,
+rho ( 9,594.37) = 0.10, p = 0.54).
 

### 2020-12-21 manuscript.Rmd
add notes on prev disc points

@@
 
-4. General prediction: Non-monolingual Yélî children have lower scores than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
-<!-- **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)** ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)-->
-4. Local prediction: Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family 
-<!-- **(MC: definitely almost all kin; 5k of them)** ac2mc I'm not sure this is how bio anthropologists think about kin; because it's about how you attribute resources to boost your genes' chances of being passed on relative to those of others in the same space.
-So my guess is that there is still a difference between nuclear family and not.
-Rewrote kin -> nuclear family -->
-(or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
-<!-- **(MC: maybe it'd be more useful to try and estimate the number of non-Y moms who live in each hamlet where testing is taking place?)** ac2mc not sure it's worth the effort, in any case other female adults are almost certain to have very little impact on the target child...
-Perhaps I'm not saying something very clearly?
--->
-. That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother 
-<!-- **(MC: i can get you actual estimates for this for YD; let's go with that instead!)** -->
-<!-- ac2mc I think I meant your own paper! added "on the same population "-->
-, so in so far as this input has a crucial formational role, then there may still be a score difference between these two groups.
-4. We don't see a difference
-
 5.
@@
 
+<!--
+4. General prediction: Non-monolingual Yélî children have lower scores than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
+
+M2A: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)
+
+ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)
+
+M2A: I'll take your word on the mixed evidence!
+I'm certainly not up to date in this domain.
+We don't really have enough kids (non-monoling = 6 kids) to investigate this anyway, which I think is also why you ultimately don't include it in the analyses of individual variation in NWR score.
+I'm therefore not going to include it in the discussion draft.
+
+---
+
+4. Local prediction: Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family
+
+M2A: definitely almost all kin; 5k of them
+
+ac2mc I'm not sure this is how bio anthropologists think about kin; because it's about how you attribute resources to boost your genes' chances of being passed on relative to those of others in the same space.
+So my guess is that there is still a difference between nuclear family and not.
+Rewrote kin -> nuclear family
+
+M2A: I see what you mean!
+I wouldn't, however, directly apply this broad theory about genetic transmission to understanding the specific Rossel context.
+I'm really very unconvinced that there is the same kind of hard division between nuclear and non-nuclear family on Rossel wrt daily resources, caregiving, and even ceremonial and some land rights things.
+I think the real question being raised here is about the specific relevance the distribution of linguistic input over family members has for kids' language development.
+We're really too far from being able to groundedly talk about that for my comfort.
+
+(or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
+
+M2A: maybe it'd be more useful to try and estimate the number of non-Y moms who live in each hamlet where testing is taking place?)
+
+ac2mc not sure it's worth the effort, in any case other female adults are almost certain to have very little impact on the target child... Perhaps I'm not saying something very clearly?
+
+M2A: In my experience most of the non-Rossel moms don't really speak Rossel to their kids at all (though they use some phrases and have some receptive knowledge)---there are exceptions of course!
+I don't think what you're saying is unclear, I'm just trying to wrap my head around what is being implied in some of the text and also doing a self-check for whether each of these implications is really meaningfully relevant in the Rossel context.
+Sorry, that probably feels like turning us in circles a bit!
+
+. That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother
+
+M2A: i can get you actual estimates for this for YD; let's go with that instead!
+
+ac2mc I think I meant your own paper! added "on the same population "
+
+M2A: Heh yeah I know!
+I mean that, buried in my annotations, is also information about how often the TCDS in Casillas et al.
+(2020) comes from the mother vs.
+others.
+Right now the paper doesn't make this claim, it only talks about speech from women.
+So we could in theory extract actual input estimates from the mother.
+But I think that's moot since none of this is really pertinent to the discussion anymore.
+
+, so in so far as this input has a crucial formational role, then there may still be a score difference between these two groups.
+
+4. We don't see a difference
+
+M2A: Not currently tested as far as I can see (?)
+
+-->
+
 <!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->

### 2020-12-21 manuscript.Rmd
add item complexity disc subsection draft

@@
 Substitutions were somewhat more frequent, but still relatively rare.
-Nasal vowels were produced as oral vowels and vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral targets as nasal (~`r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% of targets) and `r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral (~`r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% of targets).
+Nasal vowels were produced as oral vowels and vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral targets as nasal (~`r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% of targets) and `r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral (~`r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% of targets)).
 Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (`r sum(sub_tab[oralVs,oralVs])` oral vowels (~`r mean(sub_pc[oralVs,oralVs],na.rm=T)`% of targets and `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels (~`r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% of targets).
@@
 
-- What is the overall repetition scores (whole word, phoneme based, distance)?
-
-- How does this change as a function of item complexity (number of syllables, sound complexity)?
-
-1. Prediction based on previous work: Children have higher scores for shorter items
-1.
-Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.
-1. it turns out we were right! the scores for 2-4 syll words decline only slightly with length
-
-2.
-Prediction based on previous work: Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children achieve higher scores when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds [CITE].
-Therefore, we expect higher scores for items with common sounds than in those with rare sounds.
-2.
-Prediction for Yélî made before seeing the data: The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors.
-Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory.
-As a result, differences between easier and harder items may be smaller in this work than in other research.
-2. it turns out we do see
-
-- How frequent are errors that result in real words? Is that a function of item complexity?
+We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
+While the study, in itself, advances NWR research by demonstrating the successful application of this method in a rural, non-industrialized population [see also @cristia2020infant], the results also inform the larger body of NWR work regarding the influence of stimulus length, segmental frequency, age, and developmental context, where previous findings have often been mixed.
+
+In our sample of `r nrow(overall)` children between ages `r min(overall$Age)` and `r max(overall$Age)`, the average item-level score was `r round(mean(overall$Word))`%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \@ref(fig:prevlit-fig)).
+Segment-level analyses revealed that repetitions recovered an average of `r round(mean(overall$Phoneme))`% of the segments in the original stimulus, indicating that inexact repetitions often constituted _mostly_ accurate reroductions of the target.
+Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of `r round(mean(overall$NLD))`% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
+Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
+Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.
+
+### Item complexity
+
+We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item and its average segmental frequency.
+Based on previous work, we had predicted that children would have higher NWR scores for shorter items.
+That said, previous work has shown both very small [@piazzalunga2019articulatory] and very large [@cristia2020infant; @jabere2018xperiment] effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10% of the remaining words.
+Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms **Alex I'm going off your note here ("Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English."").
+I don't have a reference for this, can you please finish the thought or nix this bit?**.
+Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
+While indeed NWR scores were overall lower for longer items (e.g., see Figure \@ref(fig:prevlit-fig)), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
+In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length.
+First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon.
+Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s [-@piazzalunga2019articulatory] study on Italian and as borne out in the current dataset once controlling for other factors.
+
+Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
+Typologically common sounds are associated with higher performance on a handful of other tasks (**REFS -- M2A: Alex, I added this based on your note, where it sounded like you had some particular studies in mind?**) though to our knowledge this has not yet been tested with non-word repetition.
+Regarding Yélî Dnye in particular, the phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
+We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
+Indeed we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often.
+This effect is large, with a magnitude more than twice the size of the effect of participant age.
+This significant effect remains even once also accounting for the frequencies of these segments in Yélî Dnye child-directed speech, which are correlated with their typological frequencies.
+In sum, typological frequency effects, which have been found in other measurements of phonological processing.
+appear to strongly affect NWR performance, and do not appear mitigated by language-specific pressure to make finer-grained differences earlier in development.
+
+With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.
+oral vowels and simple vs.
+complex consonants, revealed no remarkable bias in error types.
+That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
+Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
+Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
+
+### Individual differences
+
 - Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?

### 2020-12-21 manuscript.tex
add item complexity disc subsection draft

@@
 targets as nasal (\textasciitilde{}0.99\% of targets) and 30 nasal
-targets produced as oral (\textasciitilde{}2.37\% of targets). Both oral
-and nasal vowels were also occasionally subsituted by vowels of a
+targets produced as oral (\textasciitilde{}2.37\% of targets)). Both
+oral and nasal vowels were also occasionally subsituted by vowels of a
 different quality that preserved the target oral/nature airflow (110
@@
 
-\begin{itemize}
-\item
-  What is the overall repetition scores (whole word, phoneme based,
-  distance)?
-\item
-  How does this change as a function of item complexity (number of
-  syllables, sound complexity)?
-\end{itemize}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Prediction based on previous work: Children have higher scores for
-  shorter items
-\item
-  Prediction for Yélî made before seeing the data: The length
-  distribution in Yélî words is more balanced than that in English, and
-  thus the score decline for poly- versus mono-syllables may be less
-  pronounced than that for English.
-\item
-  it turns out we were right! the scores for 2-4 syll words decline only
-  slightly with length
-\item
-  Prediction based on previous work: Similarly, we do not know of NWR
-  research that manipulates the difficulty of the sounds that are
-  included in the items, but word naming and other research suggests
-  that children achieve higher scores when producing easy and/or
-  typologically common sounds than difficult and/or typologically rare
-  sounds {[}CITE{]}. Therefore, we expect higher scores for items with
-  common sounds than in those with rare sounds.
-\item
-  Prediction for Yélî made before seeing the data: The Yélî sound
-  inventory is very large and compressed, with many similar sounds that
-  are acoustic and articulatory neighbors. Therefore, this may
-  constitute a pressure for children to have finer auditory skills (and
-  perhaps more precise articulations) than children speaking languages
-  with a simpler inventory. As a result, differences between easier and
-  harder items may be smaller in this work than in other research.
-\item
-  it turns out we do see
-\end{enumerate}
+We used non-word repetition to investigate phonological development in a
+language with a large phonological inventory (including some
+typologically rare segments). While the study, in itself, advances NWR
+research by demonstrating the successful application of this method in a
+rural, non-industrialized population (see also Cristia et al., 2020),
+the results also inform the larger body of NWR work regarding the
+influence of stimulus length, segmental frequency, age, and
+developmental context, where previous findings have often been mixed.
+
+In our sample of 40 children between ages 3 and 10, the average
+item-level score was 64\%, falling well within the expected range of NWR
+applications in previous studies on a variety of linguistic populations
+(see Figure \ref{fig:prevlit-fig}). Segment-level analyses revealed that
+repetitions recovered an average of 78\% of the segments in the original
+stimulus, indicating that inexact repetitions often constituted
+\emph{mostly} accurate reroductions of the target. Complementary data
+from segment-based normalized Levenshtein distance reflect this same
+pattern, with an average of 22\% of the stimulus segments getting
+substituted, inserted, or deleted upon repetition. Paired with our
+thorough training protocol, we take these NWR scores as indicating that
+(a) our adaptations to NWR for this context were successful, even given
+a number of non-standard changes to the training phase and to the design
+of the stimuli and (b) Yélî children show comparable performance to
+others tested on a similar task, despite the many linguistic, cultural,
+and socioeconomic differences between this and previously tested
+populations. Given successful (and comparable) execution of this task,
+we can dive deeper into effects of stimulus length, segmental frequency,
+age, and developmental context.
+
+\subsubsection{Item complexity}\label{item-complexity}
+
+We investigated the effect of item complexity on NWR scores by varying
+both the number of syllables in the item and its average segmental
+frequency. Based on previous work, we had predicted that children would
+have higher NWR scores for shorter items. That said, previous work has
+shown both very small (Piazzalunga et al., 2019) and very large (Cristia
+et al., 2020; Jaber-Awida, 2018) effects of stimulus length and,
+further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic
+words are nearly equally frequent in the current language, with
+trisyllabic and longer words making up a non-trivial 10\% of the
+remaining words. Compare this to, for example, English, which is
+substantially more skewed toward monosyllabic word forms **Alex I'm
+going off your note here (``Prediction for Yélî made before seeing the
+data: The length distribution in Yélî words is more balanced than that
+in English, and thus the score decline for poly- versus mono-syllables
+may be less pronounced than that for English.''``). I don't have a
+reference for this, can you please finish the thought or nix this
+bit?**. Setting aside our monosyllabic stimuli, which all contained
+typologically infrequent segments, we can examine effects of item length
+among the remaining stimuli, which range between 2 and 4 syllables long.
+While indeed NWR scores were overall lower for longer items (e.g., see
+Figure \ref{fig:prevlit-fig}), the effect of item length was not
+significant in a statistical model that additionally accounted for age
+and random effects of item and participant. In light of mixed prior
+results of item length, we propose two possible (and non-mutually
+exclusive) explanations for this minimal impact of item length. First,
+further extensions of this type of analysis in more populations may
+reveal that, in general (and cross-linguistically), item length effects
+are variable between languages, potentially reflecting the distribution
+of word lengths in the ambient language and other (morpho-)phonological
+tendencies in the lexicon. Second, above and beyond these
+language-specific effects, the general impact of item length on NWR
+score may be relatively small, as shown in Piazzalunga et al.'s (2019)
+study on Italian and as borne out in the current dataset once
+controlling for other factors.
+
+Our monosyllabic items included typologically rare segments so that we
+could test whether lower average segmental frequency is associated with
+lower NWR scores. Typologically common sounds are associated with higher
+performance on a handful of other tasks (\textbf{REFS -- M2A: Alex, I
+added this based on your note, where it sounded like you had some
+particular studies in mind?}) though to our knowledge this has not yet
+been tested with non-word repetition. Regarding Yélî Dnye in particular,
+the phonemic inventory is both large and acoustically packed, in
+addition to containing several typologically infrequent (or unique)
+contrasts. We therefore expected to see that, while NWR scores would be
+lower for stimuli with lower average frequency, this effect would be
+relatively weak because the ambient language puts pressure on Yélî
+children to distinguish (perceptually and articulatorily) fine-grained
+phonetic differences in order to successfully communicate with others.
+Indeed we found a robust effect of average segmental frequency on NWR
+performance: Even accounting for age and random effects of item and
+participant, we see that target words with more frequent segments were
+repeated correctly more often. This effect is large, with a magnitude
+more than twice the size of the effect of participant age. This
+significant effect remains even once also accounting for the frequencies
+of these segments in Yélî Dnye child-directed speech, which are
+correlated with their typological frequencies. In sum, typological
+frequency effects, which have been found in other measurements of
+phonological processing. appear to strongly affect NWR performance, and
+do not appear mitigated by language-specific pressure to make
+finer-grained differences earlier in development.
+
+With respect to the types of errors in repetition made, we did not see
+clear patterns to further guide our discussion: base rates of deletion
+and substitution were fairly low and the relative distribution of errors
+over, e.g., nasal vs.~oral vowels and simple vs.~complex consonants,
+revealed no remarkable bias in error types. That said, the lack of a
+difference could be due to relative imbalance across our stimuli in the
+use of these phonemic features (e.g., we included many more more oral
+than nasal targets) and future work should investigate such sources of
+error bias more systemtically. Some portion of the errors were
+introduced when the participant produced a real word (in Yélî Dnye or
+English) in response to the stimulus. Real-word repetitions here made up
+two thirds of errorful repetitions---this is quite high compared to past
+work (e.g., {\textbf{???}}), but it is unclear what caused this pattern
+in the current study: Castro and colleagues' ({\textbf{???}}) study
+focused on adults rather than children, the task was administered by a
+team including a foreign, English-speaking researcher, and the
+particularities of the Yélî Dnye phonological inventory result in many
+true-word phonetic neighbors. Follow-up work exploring this type of
+error in children from other populations in addition to further work on
+Yélî children will clarify this effect.
+
+\subsubsection{Individual differences}\label{individual-differences}
 
@@
 \tightlist
-\item
-  How frequent are errors that result in real words? Is that a function
-  of item complexity?
 \item
@@
   Children's scores increases with child age.
-\item
-  General prediction: Non-monolingual Yélî children have lower scores
-  than monolingual ones when tested on the society-dominant language (we
-  did not test any non-dominant language) 
-\item
-  Local prediction: Anecdotally Yélî children grow up in close-knitted
-  communities and thus may receive significant portions of their
-  language input from people not in their nuclear family (or at least
-  from people other than their mothers, who tend to be the non-native
-  speakers). If so, the difference between monolinguals and not
-  monolinguals may be smaller than that found in other work . That said,
-  one recent study on the same population shows that most child-directed
-  input in the first 2 years does come from the mother , so in so far as
-  this input has a crucial formational role, then there may still be a
-  score difference between these two groups.
-\item
-  We don't see a difference
 \item

### 2020-12-21 manuscript.Rmd
add indiv diffs disc subsection draft

@@
 
-Previous work typically steers clear of articulartorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology.
+Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology.
 We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.
@@
 
-Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech], and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent].
+Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]<!--, and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]-->.
 Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool].
@@
 - How does score change as a function of item length in number of syllables?
-- Is individual variation in NWR scores attributable to child age, sex, birth order, monolingual status, and/or parental education?
+- Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?
 
@@
 That said, previous work has shown both very small [@piazzalunga2019articulatory] and very large [@cristia2020infant; @jabere2018xperiment] effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10% of the remaining words.
-Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms **Alex I'm going off your note here ("Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English."").
+Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms **M2A: Alex I'm going off your note here ("Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English."").
 I don't have a reference for this, can you please finish the thought or nix this bit?**.
@@
 
-- Is individual variation explainable by child age, sex, birth order, monolingual status, and/or parental education?
-
-3. Children's scores increases with child age.
-
-5.
-General prediction: previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated score higher in standardized language tests (eg ppvt) than children whose mothers are less educated.
-5.
-Prediction made before seeing data: In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment.
-This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly.
-However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
-5. We did not find this
-
-6. General prediction: To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children score higher in standardized language tests later-born children. 
-6.
-Prediction before seeing data: One main causal path between birth order and language development is via parental input (CITE).
-Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the difference in scores between first borns and later borns may be small
-6. we did not find a sig effect, but this is a small effect (eg d=.2 in Havron et al)
+A review of previous work (see Introduction) suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR.
+We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta-analytic efforts aggregating over smaller studies such as ours.
+
+Following prior work, we expected that NWR scores would increase with participant age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
+Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
+In brief, our results underscore the idea that phoonlogical development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].
+
+In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or participant gender [@chiat2007preschool] on NWR scores.
+In addition to this prior work, education on Rossel Island, while generally highly valued, is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation.
+There is also limited variation in maternal education across the families in the region of the island where we sampled.
+We therefore expected little evidence for impact of either participant gender or maternal education in the present study.
+On the other hand, these predictors have established effects on other language development measures (**REFS: M2A: Alex go ahead and pick your faves here**).
+So to the extent that NWR scores share causal links to gender-based differences ein development and maternal linguistic input with these other language ouctome measures, we might then expect these factors to appear in NWR data.
+In fact, neither participant gender nor maternal education were correlated with NWR score in the current data.
+
+Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect], presumably because later-born children receive a smaller share of maternal input than their older siblings.
+Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
+Therefore we expected that any effects of birth order on NWR would be attenuated in this context.
+In line with this prediction, our descriptive analysis showed no correlation between birth order and NWR score.
 

### 2020-12-21 manuscript.tex
add indiv diffs disc subsection draft

@@
 
-Previous work typically steers clear of articulartorily and/or
+Previous work typically steers clear of articulatorily and/or
 acoustically challenging sounds, but we included some in our experiment
@@
 et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance,
-Stackhouse, \& Wells, 2005), and differences relating to bilingual
-language exposure (Brandeker \& Thordardottir, 2015; Meir \&
-Armon-Lotem, 2017; Meir et al., 2016). Previous work typically finds no
+Stackhouse, \& Wells, 2005). Previous work typically finds no
 significant differences as a function of maternal education (e.g.,
@@
   Is individual variation in NWR scores attributable to child age, sex,
-  birth order, monolingual status, and/or parental education?
+  birth order, and/or maternal education?
 \end{itemize}
@@
 remaining words. Compare this to, for example, English, which is
-substantially more skewed toward monosyllabic word forms **Alex I'm
+substantially more skewed toward monosyllabic word forms **M2A: Alex I'm
 going off your note here (``Prediction for Yélî made before seeing the
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Is individual variation explainable by child age, sex, birth order,
-  monolingual status, and/or parental education?
-\end{itemize}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\item
-  Children's scores increases with child age.
-\item
-  General prediction: previous NWR evidence on this is mixed, but
-  general findings on language development suggest that children whose
-  mothers are more educated score higher in standardized language tests
-  (eg ppvt) than children whose mothers are less educated.
-\item
-  Prediction made before seeing data: In the Rossel community, formal
-  education plays an extremely minor role in ensuring individual's
-  success, is not a good index of relative socio-economic status, and
-  furthermore there is only a narrow range of variation in maternal
-  educational attainment. This may lead to no or only very small
-  advantages for children whose mothers are more educated, provided that
-  the causal chain between maternal education and child language is via
-  SES more broadly. However, if education directly boosts maternal
-  verbal skills and the incidence of verbal behavior (as suggested by
-  CITE), then we should still see a difference along this factor.
-\item
-  We did not find this
-\item
-  General prediction: To our knowledge, there is no previous NWR work on
-  this, but other research suggests that first-born children score
-  higher in standardized language tests later-born children.
-\item
-  Prediction before seeing data: One main causal path between birth
-  order and language development is via parental input (CITE). Given our
-  arguments above for how mothers may not be as important among Rossel
-  people than in other places, then the difference in scores between
-  first borns and later borns may be small
-\item
-  we did not find a sig effect, but this is a small effect (eg d=.2 in
-  Havron et al)
-\end{enumerate}
+A review of previous work (see Introduction) suggested that our
+anticipated sample size would not be sufficient to detect most
+individual differences using NWR. We give a brief overview of individual
+difference patterns of four types in the present data---age, sex, birth
+order, and maternal education---hoping that these findings can
+contribute to future meta-analytic efforts aggregating over smaller
+studies such as ours.
+
+Following prior work, we expected that NWR scores would increase with
+participant age (Farmani et al., 2018; Kalnak et al., 2014; Vance et
+al., 2005). Indeed, age was significantly correlated with NWR score and
+also showed up as a significant predictor of NWR score when included as
+a control factor in the analyses of both item length and average
+segmental frequency. In brief, our results underscore the idea that
+phoonlogical development continues well past the first few years of
+life, extending into middle childhood and perhaps later (Hazan \&
+Barrett, 2000).
+
+In contrast, previous work shows little evidence for effects of maternal
+education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \&
+Armon-Lotem, 2017) or participant gender (Chiat \& Roy, 2007) on NWR
+scores. In addition to this prior work, education on Rossel Island,
+while generally highly valued, is not at all essential to ensuring one's
+success in society and may not be a reliable index of local
+socioeconomic variation. There is also limited variation in maternal
+education across the families in the region of the island where we
+sampled. We therefore expected little evidence for impact of either
+participant gender or maternal education in the present study. On the
+other hand, these predictors have established effects on other language
+development measures (\textbf{REFS: M2A: Alex go ahead and pick your
+faves here}). So to the extent that NWR scores share causal links to
+gender-based differences ein development and maternal linguistic input
+with these other language ouctome measures, we might then expect these
+factors to appear in NWR data. In fact, neither participant gender nor
+maternal education were correlated with NWR score in the current data.
+
+Last but not least, we investigated whether birth order might affect NWR
+scores, as it does other language tasks, resulting in first-born
+children showing higher scores on standardized language tests than
+later-born children (Havron et al., 2019), presumably because later-born
+children receive a smaller share of maternal input than their older
+siblings. Given shared caregiving practices and the hamlet organization
+typical of Rossel communities, children have many sources of adult and
+older child input that they encounter on a daily basis and first born
+children quickly integrate with a much larger pool of both older and
+younger children with whom they partly share caregivers. Therefore we
+expected that any effects of birth order on NWR would be attenuated in
+this context. In line with this prediction, our descriptive analysis
+showed no correlation between birth order and NWR score.
 

### 2020-12-21 manuscript.Rmd
add conclusions draft

@@
   - name          : "Marisa Casillas"
-    affiliation   : "2"
+    affiliation   : "2,3"
     address       : "Wundtlaan 1, 6525 XD, Nijmegen, The Netherlands"
@@
     institution   : "Max Planck Institute for Psycholinguistics"
+  - id            : "3"
+    institution   : "University of Chicago"
 
@@
 
+### Conclusions
+
+While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that vaies depending on the linguistic, cultural, and sociodemographic properties of the population under study.
+Because these latter predictors strongly relate to other language outcomes, the present findings raise the issue of why NWR would pattern differently, what that could tell us about the relationship between lexical development, phonological development, and the input environment and, last but not least, what is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders.
+In the meanwhile, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a connection between individual learners and global language patterns when it comes to acoustic and articulatory markedness.
+
 <!--

### 2020-12-21 manuscript.tex
add conclusions draft

@@
 \author{Alejandrina Cristia\textsuperscript{1}~\& Marisa
-Casillas\textsuperscript{2}}
+Casillas\textsuperscript{2,3}}
 \date{}
@@
 
+\subsubsection{Conclusions}\label{conclusions}
+
+While NWR can, in theory, be used to test a variety of questions about
+phonological development in any language, previous work has been
+primarily limited to a handful of related languages spoken in urban,
+industrialized contexts. The present study shows that, not only can NWR
+be adapted for very different populations than have previously been
+tested, but that effects of age and typological frequency may strongly
+influence phonological development across these diverse settings, while
+effects of item length, participant gender, maternal education, and
+birth order, may either have little impact on this facet of language
+development or have an impact that vaies depending on the linguistic,
+cultural, and sociodemographic properties of the population under study.
+Because these latter predictors strongly relate to other language
+outcomes, the present findings raise the issue of why NWR would pattern
+differently, what that could tell us about the relationship between
+lexical development, phonological development, and the input environment
+and, last but not least, what is implied about the joint applicability
+of these outcome measures as a diagnostic indicator for language delays
+and disorders. In the meanwhile, we take the present findings as
+robustly supporting the idea that phonological development continues
+well past early childhood and as yielding preliminary support for a
+connection between individual learners and global language patterns when
+it comes to acoustic and articulatory markedness.
+
 \newpage

### 2020-12-21 manuscript.Rmd
update to-do notes

@@
 title             : "Non-word repetition in Yélî Dnye"
-shorttitle        : "NWR in Yélî Dnye"
+shorttitle        : "NWE in Yélî Dnye"
 
@@
 - implement decisions on prop phones & patterns nwr mp
-- revise bullet points for discussion
+<!--Added by MC on 21 Dec 2020-->
+- check and accept/deny MC's updated segment frequencies
+- re-run analyses w/ new frequencies and check that outcomes didn't qualitatively change
+- check YD data in figure 1 (the effect of item length looks reversed)
+- see MC comments throughout and delete now-irrelevant things
 
@@
 - double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
+
+M2A: Done! Left for AC to review in segments-20201123.xlsx and, if she is satisfied, to integrate :)
 -->
-<!--M2A: Done! Left for AC to review in segments-20201123.xlsx and, if she is satisfied, to integrate :)-->
+<!--
 - draft discussion
+
+M2A: Done! I left some space for you to fill in references in a few spots since it seemed from the notes you left that you had some specific ones in mind.
+-->
 
@@
 
-<!--M2A: I think a nice way to end this bit would be very briefly (~1-sentence) linking this input scenario to the relevance of looking at NWR (à la Cristia 2020 DevRev). What do you think?-->
-<!-- a2m maybe we want to bring that up in the discussion?
-I haven't found a good place to introduce that paper, in part because right now we have these three purposes that seem sort of unconnected with the goals of that paper...
--->
-
 \newpage

### 2020-12-22 manuscript.tex
update to-do notes
cleaned up and reknitted

@@
 \setmainfont{Doulos SIL} % Set main font to Doulos SIL
+% Manuscript styling
+\usepackage{upgreek}
+\captionsetup{font=singlespacing,justification=justified}
+
+% Table formatting
+\usepackage{longtable}
+\usepackage{lscape}
+% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
+\usepackage{multirow}		% Table styling
+\usepackage{tabularx}		% Control Column width
+\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
+\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+
+% Create new environments so endfloat can handle them
+% \newenvironment{ltable}
+%   {\begin{landscape}\begin{center}\begin{threeparttable}}
+%   {\end{threeparttable}\end{center}\end{landscape}}
+\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
+
+% Enables adjusting longtable caption width to table width
+% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
+\makeatletter
+\newcommand\LastLTentrywidth{1em}
+\newlength\longtablewidth
+\setlength{\longtablewidth}{1in}
+\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
+
+% \setlength{\parindent}{0.5in}
+% \setlength{\parskip}{0pt plus 0pt minus 0pt}
+
+% \usepackage{etoolbox}
+\makeatletter
+\patchcmd{\HyOrg@maketitle}
+  {\section{\normalfont\normalsize\abstractname}}
+  {\section*{\normalfont\normalsize\abstractname}}
+  {}{\typeout{Failed to patch abstract.}}
+\makeatother
+\shorttitle{NWE in Yélî Dnye}
+\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
+\affiliation{
+\vspace{0.5cm}
+\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
+\authornote{All data are made available in a repository in the Open Science Framework.
+AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
+S.
+McDonnell Foundation Understanding Human Cognition Scholar Award.
+
+Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
+\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
+\usepackage{lineno}
+
+\linenumbers
+\usepackage{csquotes}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[shorthands=off,main=english]{babel}
+\else
+  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
+  \usepackage{polyglossia}
+  \setmainlanguage[]{english}
+\fi
 
 \title{Non-word repetition in Yélî Dnye}
-\author{Alejandrina Cristia\textsuperscript{1}~\& Marisa
-Casillas\textsuperscript{2,3}}
+
 \date{}
 
-\authornote{All data are made available in a repository in the
-Open Science Framework. AC acknowledges the support of the Agence
-Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004,
-ANR-14-CE30-0003, ANR-17-EURE-0017); and the J. S. McDonnell Foundation
-Understanding Human Cognition Scholar Award.
-
-Correspondence concerning this article should be addressed to
-Alejandrina Cristia, 29, rue d'Ulm, 75005 Paris, France. E-mail:
-\href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}
-
 \abstract{
-In nonword repetition (NWR) studies, participants are presented
-auditorily with an item that is phonologically legal but lexically
-meaningless in their language. NWR scores are thought to reflect
-long-term phonological knowledge as well as online phonological working
-memory and flexible production patterns. In this study, we report on NWR
-results among children learning Yélî Dnye, an isolate spoken on Rossel
-Island in Papua New Guinea that has an unusually dense phonological
-inventory. This study contributes to three lines of research. First, we
-document that non-word items containing typologically frequent sounds
-are repeated without changes more often that non-words containing
-typologically rare sounds. Second, we document rather weak effects of
-item length, contributing to other research suggesting that length
-effects may be language-specific. Third, we do not find strong
-individual variation effects in this population, contrary to previous
-results documenting strong age-related effects. Together, these data
-provide a unique view of online phonological processing in an
-understudied language while making preliminary connections between
-patterns in language development and in the frequency of linguistic
-features in human language at large.
-
+In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language.
+NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
+In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea that has an unusually dense phonological inventory.
+This study contributes to three lines of research.
+First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds.
+Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
+Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
+Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between patterns in language development and in the frequency of linguistic features in human language at large.
 }
@@
 
-\subsection{TODO Alex}\label{todo-alex}
+\hypertarget{todo-alex}{%
+\subsection{TODO Alex}\label{todo-alex}}
 
@@
 \item
-  check again if all fields are explained
+  check whether it's stimuli, stimuli2, stimuli3 that should be used!!!
+\item
+  create fields final-data DONE
+\item
+  leave only files actually needed DONE
+\item
+  add readme explaining files DONE
+\item
+  remove all files that aren't necessary DONE
+\item
+  get rid of all the conversions and directly read in xls
 \item
@@
 \item
-  revise bullet points for discussion
+  check and accept/deny MC's updated segment frequencies
+\item
+  re-run analyses w/ new frequencies and check that outcomes didn't qualitatively change
+\item
+  check YD data in figure 1 (the effect of item length looks reversed)
+\item
+  see MC comments throughout and delete now-irrelevant things
 \end{itemize}
 
-\subsection{TODO Middy}\label{todo-middy}
+\hypertarget{todo-middy}{%
+\subsection{TODO Middy}\label{todo-middy}}
+
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
+
+Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000).
+Much of the evidence for later phonological development comes through the use of nonword repetition (NWR) tasks.
+In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
+After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
+NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
+
+NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
+NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
+However, it has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014, and @cristia2020infant).
+Further, the languages it has been used with have historically had small to moderate phonological inventories (Maddieson, 2005).
+
+In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
+The study was designed to contribute to multiple aspects of our understanding of phonological development: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
+Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation; previous NWR research has shown variable effects of length between populations.
+Third, these data give a first impression of how this task is responded to in a rarely-studied language and cultural setting; we add to this contribution by further investigating whether there are structured sources that account for individual variation in our sample.
+
+Before jumping into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
+Since NWR has been almost exclusively used in urban, industrialized populations (see Cristia et al., 2020 for an exception), we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.
+
+\hypertarget{yuxe9luxee-dnye-phonology}{%
+\subsubsection{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}}
+
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+The Yélî sound system, much like its baroque grammatical system (Levinson, n.d.), is unlike any other in the region.
+
+With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
+The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (n.d.) for details).\footnote{We use Levinson's (n.d.) under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+Vowels in Yélî Dnye may be oral or nasal, short or long.
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with 7 of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, n.d.) and includes at least two contrasts not yet been documented elsewhere (labial-coronal double-articulations with dental vs.~post-alveolar coronal placement for both oral and nasal stops, Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, n.d.).
+
+Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent Yélî Dnye dictionary at time of writing; Levinson, n.d.).
+The vast majority of syllables use a CV format.
+A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., \enquote{ndapî} /\enquote{ṇṭæ.pɯ/).
+This process is used frequently with English loan words that have a coda (e.g., 'ponî} (phone) /\enquote{pɔ.nɯ/).
+An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala} (here) /æ.'læ/), but these are limited to /æ/-.
+Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (n.d.) for details).
+
+\hypertarget{the-yuxe9luxee-community}{%
+\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}}
+
+Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or another language(s) from the region.
+
+Most Yélî people are swidden horticulturalists, raising a variety of tubers (e.g., yams, manioc, taro), coconut, banana, and more in their gardens.
+In addition to garden foods, Yélî children grow up eating fish and foraging for shellfish and nuts, which provide a regular source of extra protein.
+The typical household in our dataset includes seven individuals and is situated among a collection of 4 or more other households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
+
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
+There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available.
+There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG.
+There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time.
+As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, is typically unavailable.
+Our data collection is therefore limited to the duration of the researchers' on-island visits.
+
+Despite restricted outside contact, formal education is a priority for many Yélî families.
+A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65\% of mothers and 49\% of fathers completing the full range of locally available education (\textasciitilde{}9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG (Casillas, Brown, \& Levinson, 2020).
+While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day).
+It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year.
+Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward (Brown \& Casillas, n.d.).\footnote{For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.} Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
+
+While work on Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas et al., 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun (e.g., Peute, Fikkert, \& Casillas, n.d.).
+What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day (Casillas et al., 2020), shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children (Bunce et al., n.d.; Casillas et al., 2020).
+While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere (Bergelson et al., 2019; Bunce et al., n.d.), there is a significant and marked increase in child-directed speech from other children as infants get older (Bunce et al., n.d.; Casillas et al., 2020).
+This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
+
+\hypertarget{nwr-design-and-analysis-adaptations}{%
+\subsubsection{NWR design and analysis adaptations}\label{nwr-design-and-analysis-adaptations}}
+
+In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
+For instance, in English, a response of {[}bilig{]} or {[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length.
+There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
+For example, while items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation (Brandeker \& Thordardottir, 2015).
+Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
+Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
+
+Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology.
+We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.~less challenging sounds.
+However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
+Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
+However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit.
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
+
+Based on previous work, we looked at potential changes with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
+Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on these too.
+
+\hypertarget{research-questions}{%
+\subsubsection{Research questions}\label{research-questions}}
+
+After some preliminary analyses to set the stage, we address the following questions:
 
@@
 \item
-  draft discussion
+  Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
+\item
+  How does score change as a function of item length in number of syllables?
+\item
+  Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?
 \end{itemize}
 
-\subsection{Introduction}\label{introduction}
-
-Children's perception and production of phonetic and phonological units
-continues developing well beyond the first year of life, even extending
-into middle childhood (e.g., Hazan \& Barrett, 2000). Much of the
-evidence for later phonological development comes through the use of
-nonword repetition (NWR) tasks. In a NWR task, participants hear a short
-word-like form that is phonologically legal but lexically meaningless in
-the language(s) they are learning. After hearing this non-word, the
-participant's task is to try to immediately and precisely repeat it. NWR
-scores are thought to reflect long-term phonological knowledge (to
-perceive the item precisely despite not having heard it before) as well
-as online phonological working memory (to encode the item in the
-interval between hearing it and saying it back) and flexible production
-patterns (to produce the item precisely despite not having pronounced it
-before).
-
-NWR has been used to seek answers to a variety of theoretical questions,
-including what the links are between phonology, working memory, and the
-lexicon (Bowey, 2001), and how extensively phonological constraints
-found in the lexicon affect online production (Gallagher, 2014). NWR is
-also frequently used in applied contexts, notably as a diagnostic tool
-for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
-Since non-words can be generated in any language, it has attracted the
-attention of researchers working in multilingual and linguistically
-diverse environments, particularly in Europe (Action, 2009; Meir,
-Walters, \& Armon-Lotem, 2016). However, it has seldom been used outside
-of Europe and North America (with exceptions including Gallagher, 2014,
-and Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)). Further,
-the languages it has been used with have historically had small to
-moderate phonological inventories (Maddieson, 2005).
-
-In the present study, we use NWR to investigate the phonological
-development of children learning Yélî Dnye, an isolate language spoken
-in Papua New Guinea (PNG) that has a large and unusually dense
-phonological inventory. The study was designed to contribute to multiple
-aspects of our understanding of phonological development: First, we
-included a subset of non-word items with typologically rare and/or
-challenging sounds to ask whether these rare sounds are disadvantaged in
-perception and/or production, both in terms of overall repetition scores
-and patterns of mispronunciation. Second, we varied the length (in
-syllables) of non-words to investigate the impact of word length on
-repetition and mispronunciation; previous NWR research has shown
-variable effects of length between populations. Third, these data give a
-first impression of how this task is responded to in a rarely-studied
-language and cultural setting; we add to this contribution by further
-investigating whether there are structured sources that account for
-individual variation in our sample.
-
-Before jumping into the details of our study design we first give an
-overview of Yélî Dnye phonology as well as a brief ethnographic review
-of the developmental environment on Rossel Island. Since NWR has been
-almost exclusively used in urban, industrialized populations (see
-Cristia et al., 2020 for an exception), we provide this additional
-ethographic information to contextualize the adaptations we have made in
-running the task and in gathering age and other demographic information,
-compared to what is typical in urban, industrialized settings.
-
-\subsubsection{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}
-
-Yélî Dnye is an isolate language (presumed Papuan) spoken by
-approximately 7,000 people residing on Rossel Island, an island found at
-the far end of the Louisiade Archipelago in Milne Bay Province, Papua
-New Guinea. The Yélî sound system, much like its baroque grammatical
-system (Levinson, accepted), is unlike any other in the region.
-
-With only four primary places of articulation (bilabial, alveolar,
-post-alveolar, and velar) and no voicing contrasts, the phonological
-inventory is remarkably packed with acoustically similar segments. The
-core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and
-doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal
-equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
-substantial portion of them contrastively pre-nasalized or nasally
-released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/,
-/ṭpṇm/, /kpŋm/). A large number of this combinatorial set can further be
-contrastively labialized, palatalized on release, or both (e.g., /pʲ/,
-/pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (accepted) for
-details).\footnote{We use Levinson's (accepted) under-dot notation
-  (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these
-  stops are, articulatorily, somewhat variable in place, with at least
-  some tokens produced fully sub-apically. In approximating
-  cross-linguistic segment frequency below we use the corresponding
-  retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The
-consonantal inventory also includes a number of non-nasal continuants
-(/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in Yélî Dnye may be oral
-or nasal, short or long. The 10 oral vowel qualities, which span four
-levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/,
-/ɑ/) can be produced as short and long vowels, with 7 of these able to
-appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/,
-/æ̃/, /ɑ̃/).
-
-In total, Yélî Dnye uses 90 distinctive segments (not including an
-additional 3 rarely used consonants), far outstripping the phonemic
-inventory size of other documented Papuan languages (Foley, 1986;
-Levinson, accepted) and includes at least two contrasts not yet been
-documented elsewhere (labial-coronal double-articulations with dental
-vs.~post-alveolar coronal placement for both oral and nasal stops,
-Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, in preparation).
-
-Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
-monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
-that, and with tri-and-above syllabic words appearing least frequently
-(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most
-recent Yélî Dnye dictionary at time of writing; Levinson, accepted). The
-vast majority of syllables use a CV format. A small portion of the
-lexicon features words with a final closed syllable, limited to codas of
--/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
-spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
-coda, yielding a CV.CV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
-process is used frequently with English loan words that have a coda
-(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
-features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
-these are limited to /æ/-. Finally, the lexicon features a handful of
-single-vowel grammatical morphemes (see Levinson (accepted) for
-details).
-
-\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}
-
-Most speakers of Yélî Dnye grow up speaking it monolingually until they
-begin attending school around the age of 7; school instruction is in
-English. While monolingual Yélî Dnye upbringing is common, multilingual
-families are not unusual, particularly in the region around the Catholic
-Mission---the same region in which the current data were
-collected---where there is a higher incidence of married-in mothers from
-other islands (Brown \& Casillas, in press). Children in these
-multilingual families grow up speaking Yélî Dnye plus English, Tok
-Pisin, and/or another language(s) from the region.
-
-Most Yélî people are swidden horticulturalists, raising a variety of
-tubers (e.g., yams, manioc, taro), coconut, banana, and more in their
-gardens. In addition to garden foods, Yélî children grow up eating fish
-and foraging for shellfish and nuts, which provide a regular source of
-extra protein. The typical household in our dataset includes seven
-individuals and is situated among a collection of 4 or more other
-households, with structures often arranged around an open grassy area.
-These household clusters are organized by patrilocal relation, such that
-they typically comprise a set of brothers, their wives and children, and
-their mother and father, with neighboring hamlets also typically related
-through the patriline.
-
-Laying 250 nautical miles off the coast of mainland PNG and surrounded
-by a barrier reef, transport to and from the island is both infrequent
-and irregular. There is no electricity system (we use solar panels) or
-motor vehicles on the island, save dinghys with outboard motors that are
-primarily reserved for medical transport when there is diesel available.
-There is a medical radio connection at the health clinic, located at the
-Catholic Mission, via which news and messages are often passed to and
-from mainland PNG. There is also a cellular tower on the island, though
-it is often inoperative, usually for multiple months or years at a time.
-As such, access to news and outside connections, including international
-phone calls and digital exchanges that require significant data
-transfer, is typically unavailable. Our data collection is therefore
-limited to the duration of the researchers' on-island visits.
-
-Despite restricted outside contact, formal education is a priority for
-many Yélî families. A recent study surveying more than 40 parents of
-young children in this region of the island found that nearly all
-parents completed 6 or more years of education, with 65\% of mothers and
-49\% of fathers completing the full range of locally available education
-(\textasciitilde{}9 years), with around half of those parents having
-pursued an additional 2+ years of education on other islands in the
-region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While
-there are a handful of local schools around the island, these may be
-well out of walking distance for many children (i.e., more than 1 hour
-on foot or by canoe each day). It is very common for households that are
-situated close to a school to billet their school-aged relatives during
-the weekdays (sometimes the weekends too, if the child's home is very
-far) for long segements of each school year. Combined with practices of
-collective child guardianship within each close-knit hamlet, the
-practical consequence of this billeting for researchers is that adult
-consent can often come from a combination of aunts, uncles, adult
-cousins, and grandparents standing in for the child's biological
-parents. On top of this, child assent is culturally pertinent, as child
-independence is encouraged and respected from toddlerhood onward (Brown
-\& Casillas, in press).\footnote{For these reasons, we were unable to
-  record precise details for a few children, including child date of
-  birth, parent age in years, and parent education.} Finally, as
-mentioned, the language of schooling in this region is English, and so
-the majority of teens and adults on the island speak some English---many
-speak it very fluently.
-
-While work on Yélî language development is growing (e.g., Brown, 2011,
-2014; Brown \& Casillas, in press; Casillas et al., 2020; Liszkowski,
-Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's
-linguistic development is quite limited, and research into their
-phonological development in particular has only just begun (e.g., Peute,
-Fikkert, \& Casillas, In preparation). What we do know about the early
-language environment is that, while direct speech to children under 3;0
-is relatively infrequent throughout the day (Casillas et al., 2020),
-shared caregiving practices and a near-universal fondness for social
-interaction with young children ensure that speech directed to children
-comes from all types of speakers: women, men, and other children (Bunce
-et al., under review; Casillas et al., 2020). While speech from women
-predominates in young children's speech environments on Rossel, as it
-does elsewhere (Bergelson et al., 2019; Bunce et al., under review),
-there is a significant and marked increase in child-directed speech from
-other children as infants get older (Bunce et al., under review;
-Casillas et al., 2020). This increase in child-directed speech from
-other children is attributed to the fact that, starting around age 2,
-children often spend large swaths of the day playing, swimming, and
-foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10
-members) independent and mixed-age child play groups (Brown \& Casillas,
-in press; Casillas et al., 2020).
-
-\subsubsection{NWR design and analysis
-adaptations}\label{nwr-design-and-analysis-adaptations}
-
-In a basic NWR task, the participant listens to a production of a
-word-like form, such as /bilik/, and then repeats back what they heard
-without changing any phonological feature that is contrastive in the
-language. For instance, in English, a response of {[}bilig{]} or
-{[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where
-the vowel is lengthened without change of quality would be scored as
-correct, because English does not have a general feature of contrastive
-vowel length. There is some variation in how past NWR studies have
-designed the presentation procedure and structure of items. For example,
-while items are often presented orally by the experimenter (Torrington
-Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies
-have turned instead to playing back pre-recorded stimuli in order to
-increase control in stimulus presentation (Brandeker \& Thordardottir,
-2015). Additionally, while some studies have used 10-15 non-words,
-others have employed up to 46 unique items (Piazzalunga, Previtali,
-Pozzoli, Scarponi, \& Schindler, 2019). Authors also often modulate
-structural complexity, typically measured in terms of item length
-(measured in number of syllables) and/or syllable structure (open as
-opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
-
-Previous work typically steers clear of articulatorily and/or
-acoustically challenging sounds, but we included some in our experiment
-to more adequately represent Yélî Dnye's phonology. We ultimately used a
-relatively large number of items that would enable us to explore both
-variation in structural complexity and in more vs.~less challenging
-sounds. However, aware that this large item inventory might render the
-task longer and more tiresome, we split items across children (see
-below). Naturally, designing the task in this way may make the study of
-individual variation within the population more difficult because
-different children are exposed to different items. However, a review of
-previous work on individual variation suggested to us that effects of
-individual differences in NWR are relatively small, and would not be
-detectable with the sample size that we could collect during our short
-visit. That said, we contribute to the literature by also reporting
-descriptive analyses of individual variation that could potentially be
-integrated in meta-analytic efforts.
-
-Based on previous work, we looked at potential changes with age (Farmani
-et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance,
-Stackhouse, \& Wells, 2005). Previous work typically finds no
-significant differences as a function of maternal education (e.g.,
-Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or
-child gender (Chiat \& Roy, 2007). Although past research has not often
-investigated potential effects of birth order on NWR, there is a sizable
-literature on these effects in other language tasks (Havron et al.,
-2019), and therefore we report on these too.
-
-\subsubsection{Research questions}\label{research-questions}
-
-After some preliminary analyses to set the stage, we address the
-following questions:
-
-\begin{itemize}
-\tightlist
-\item
-  Does the cross-linguistic frequency of sounds in the stimuli predict
-  NWR scores? Are rarer sounds more often substituted by commoner
-  sounds?
-\item
-  How does score change as a function of item length in number of
-  syllables?
-\item
-  Is individual variation in NWR scores attributable to child age, sex,
-  birth order, and/or maternal education?
-\end{itemize}
-
-In view of the hypothesis-driven nature of this work, we had considered
-boosting the interpretational value of this evidence by announcing our
-analysis plans prior to conducting them. However, we realized that even
-pre-registering an analysis would be equivocal because we would not have
-enough power to look at all relationships of interest, in many cases
-possibly not enough to detect any of the known effects, given their
-variability across studies. To illustrate this issue, we portray in
-Figure 1 studies in which children's NWR scores were gathered between 4
-and 12 years of age, and reported separately for items that are
-relatively short (1-2 syllables) versus longer items (3-4 syllables).
-Notice that the effect of stimulus length is minuscule among Italian
-children (Piazzalunga et al., 2019), but considerable among Tsimane'
-children (Cristia et al., 2020), where a drop of 40 percentage points is
-observed at all ages. A similar large difference in NWR scores for short
-versus long items was observed among Arabic children (Jaber-Awida,
-2018). Even the effect of age is unstable in this sample of studies.
-Whereas it is quite clear that children's NWR scores increase in the
-Italian data, age effects are less stable among Tsimane' children.
-Therefore, all analyses in the present study are descriptive and should
-be considered exploratory.
+In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
+However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies.
+To illustrate this issue, we portray in Figure 1 studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
+Notice that the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Tsimane' children (Cristia et al., 2020), where a drop of 40 percentage points is observed at all ages.
+A similar large difference in NWR scores for short versus long items was observed among Arabic children (Jaber-Awida, 2018).
+Even the effect of age is unstable in this sample of studies.
+Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
+Therefore, all analyses in the present study are descriptive and should be considered exploratory.
 
@@
 \includegraphics{manuscript_files/figure-latex/prevlit-fig-1.pdf}
-\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years)
-and item length for comparable studies. Arabic data from Jaber et al.
-(2018); Italian data from Piazzalunga et al. (2019); Tsimane' from
-Cristia et al. (2020); Yélî Dnye data from the present study.}
+\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years) and item length for comparable studies.
+Arabic data from Jaber et al.
+(2018); Italian data from Piazzalunga et al.
+(2019); Tsimane' from Cristia et al.
+(2020); Yélî Dnye data from the present study.}
 \end{figure}
 
-\subsection{Methods}\label{methods}
-
-\subsubsection{Stimuli}\label{stimuli}
-
-Many NWR studies are based on a fixed list of 12-16 items that vary in
-length between 1 and 4 syllables, often additionally varying syllable
-complexity and/or cluster presence and complexity, and always meeting
-the condition that they do not mean anything in the target language
-(e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We
-kept the same variation in item length and requirement for not being
-meaningful in the language, but we did not vary syllable complexity or
-clusters because these are vanishingly rare in Yélî Dnye. We also
-increased the number of items an individual child would be tested on,
-such that a child would get up to 23 items to repeat (other work has
-also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014),
-with the entire test inventory of 44 items distributed across children.
-
-A first list of candidate items was generated during a trip to the
-island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/,
-/n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them
-into consonant-vowel syllables, then sampling the space of 1- to
-4-syllable sequences. These candidates were automatically removed from
-consideration if they appeared in Levinson's (2015) dictionary. The
-second author presented them orally to three local research assistants,
-all native speakers of Yélî Dnye, who repeated each form as they would
-in an NWR task and additinally let the experimenter know if the item was
-in fact a word or phrase in Yélî Dnye. Any item reported to have a
-meaning or a strong association with another word form or meaning was
-excluded.
-
-A second list of candidate items was generated in 2019 by selecting
-complex consonants and systematically crossing them with all the vowels
-in the Yélî Dnye inventory to produce consonant-vowel monosyllabic
-forms. As before, items were automatically excluded if they appeared in
-the dictionary. Additionally, since perceiving vowel length in isolated
-monosyllables is challenging, any item that had a short/long lexical
-neighbor was excluded. Because there is still much to discover about the
-phonology and phonetics of Yélî Dnye (Levinson, accepted), it was also
-possible that we initially generated items with illegal, but currently
-undocumented constraints. Therefore, we made sure that the precise
-consonant-vowel sequence occurred in some real word in the dictionary
-(i.e., that there was a longer word included the monosyllable as a
-subsequence). These candidates were then presented to one informant, for
-a final check that they did not mean anything. Together with the 2018
-selection, they were recorded, based on their orthographic forms, using
-a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832
-stereo audio recorder (using an XLR to mini-jack adapter) by the same
-informant, monitored by the second author for clear production of the
-phonological target. The complete recorded list was finally presented to
-two more informants, who were able to repeat all the items and who
-confirmed there were no real words present. Despite these checks, one
-monosyllable was ultimately frequently identified as a real word in the
-resulting data (intended ``yî'' /yɯ/; identified as ``yi'' /yi/,
-\emph{tree}). This item is removed from the analyses below.
-
-The final list includes: three practice items; 20 monoysllables
-containing sounds that are less frequent in the world's languages than
-singleton plosives; 8 bisyllables; 12 trisyllables; and 4
-quadrisyllables (see Table 2).
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
+
+\hypertarget{stimuli}{%
+\subsubsection{Stimuli}\label{stimuli}}
+
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
+We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 44 items distributed across children.
+
+A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/, /n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 1- to 4-syllable sequences.
+These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
+The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additinally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
+Any item reported to have a meaning or a strong association with another word form or meaning was excluded.
+
+A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms.
+As before, items were automatically excluded if they appeared in the dictionary.
+Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
+Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, n.d.), it was also possible that we initially generated items with illegal, but currently undocumented constraints.
+Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
+These candidates were then presented to one informant, for a final check that they did not mean anything.
+Together with the 2018 selection, they were recorded, based on their orthographic forms, using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) by the same informant, monitored by the second author for clear production of the phonological target.
+The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present.
+Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \enquote{yî} /yɯ/; identified as \enquote{yi} /yi/, \emph{tree}).
+This item is removed from the analyses below.
+
+The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
 
@@
 
-A Praat script (Boersma \& Weenink, 2020) was written to randomize this
-list 20 times, and split it into two sublists, to generate 40 different
-elicitation sets. The 40 elicitation sets are available online from
-\href{https://osf.io/5qspb}{osf.io/5qspb}. The split had the following
-constraints:
+A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
+The 40 elicitation sets are available online from \href{https://osf.io/5qspb}{osf.io/5qspb}.
+The split had the following constraints:
 
@@
 \item
-  The same three items were selected as practice items and used in all
-  40 elicitation sets.
-\item
-  Splits were done within each length group from the 2018 items (i.e.,
-  separately for 2-, 3-, and 4-syllable items); and among onset groups
-  for the difficult monosyllables generated in 2019 (i.e., all the
-  monosyllables starting with /tp/ were split into 2 sublists). Since
-  some of these groups had an odd number of items, one of the sublists
-  was slightly longer than the other (20 vs.~23).
-\item
-  Once the sublist split had been done, items were randomized such that
-  all children heard first the 3 practice items in a fixed order (1-,
-  2-, and 4-syllable), a randomized version of their sublist selection
-  of difficult onset items, and randomized versions of their 2-syllable,
-  then 3-syllable, and finally 4-syllable items.
+  The same three items were selected as practice items and used in all 40 elicitation sets.
+\item
+  Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sublists).
+Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 vs.~23).
+\item
+  Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1-, 2-, and 4-syllable), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
 \end{itemize}
 
-\subsubsection{Procedure}\label{procedure}
-
-In adapting the typical NWR procedure for this context, we balanced
-three desiderata: That children would not be unduly exposed to the items
-before they themselves had to repeat them (i.e., from other children who
-had participated); that children would feel comfortable doing this task
-with us; and that community members would feel comfortable having their
-children do this task with us.
-
-We ran children in batches, testing within a handful of hamlet clusters
-spread across the northeastern region of the island. Because space
-availability was limited in different ways from hamlet to hamlet, the
-places where elicitation happened varied across testing sites. We tested
-in four different sites, only making a single visit to each, conducting
-back-to-back testing of all eligible children present at the time of our
-visit in order to prevent the items from `spreading' between children
-through hearsay. In the first hamlet, we tested children in five
-different places, with some children being tested inside a house and
-others tested on the veranda. The complete list of places and the ways
-in which they met the desiderata mentioned above can be found in the raw
-data, available from the online supplementary materials. Whenever
-children living in the same household were tested, we tried to test
-children in age order, from oldest to youngest, to minimize intimidation
-for younger household members, and always using different elicitation
-sets.
-
-We fitted the child with a headset microphone (Shure SM10A or WH20 XLR
-with a dynamic microphone on a headband, most children using the former)
-that fed into the left channel of a Tascam DR40x digital audio recorder.
-The headsets were designed for adult use and could not be comfortably
-seated on many children's heads without a more involved adjustment
-period. To minimize adjustment time, which was uncomfortable for some
-children given the proximity of the experimenter and equipment, we
-placed the headband on children's shoulders in these cases, carefully
-adjusting the microphone's placement so that it was still close to the
-child's mouth. A research assistant who spoke Yélî Dnye natively sat
-next to the child throughout the task to provide instructions and, if
-needed, encouragement. The research assistant coached the child
-throughout the task to make sure that they understood what they were
-expected to do. An experimenter (the first author) delivered the
-pre-recorded stimuli to the research assistant and the child over
-headphones.
-
-The first phase of the experiment involved making sure the child
-understood the task. We explained the task and then orally presented the
-first practice item. At this point, many children did not say anything
-in response, which triggered the following procedure: First, the
-assistant insisted the child make a response. If the child still did not
-say anything, the assistant said a real word and then asked the child to
-repeat it, then another and another. If the child could repeat real
-words correctly, we provided the first training item over headphones
-again for children to repeat. Most children successfully started
-repeating the items at this point, but a few needed further help. In
-this case, the assistant modeled the behavior (i.e., the child and
-assistant would hear the item again, and the assistant would repeat it;
-then we would play the item again and ask the child to repeat it). A
-small minority of children still failed to repeat the item at this
-point. If so, we tried again with the second training item, at which
-point some children demonstrated task understanding and could continue.
-A fraction of the remaining children, however, failed to repeat this
-second training item, as well as the third one, in which case we stopped
-testing altogether (see Participants section for exclusions).
-
-The second phase of the experiment involved running the child through
-the list of test items randomly assigned to them. This was done in the
-same manner as the practice items: the stimulus was played over the
-headphones, and then the child repeated it aloud. NWR studies vary in
-whether children are allowed to hear and/or repeat the item more than
-one time. We had a fixed procedure for the test items (i.e., the
-non-practice items) in which the child was alowed to make a further
-attempts if their first attempt was judged erroneous in some way by the
-assistant. The procedure worked as follows: When the child made an
-attempt, the assistant indicated to the experimenter whether the child's
-production was correct or not. If correct, the experimenter would
-whisper this note of correct repetition into a separate headset that fed
-into the right channel of the same Tascam recorder and the group moved
-on to the next item. If not, the child was allowed to try again, with up
-to five attempts allowed before moving on to the next item. Children
-were not asked to make repetitions if they did not produce a first
-attempt. In total, test sessions took approximately XX minutes on
-average, with XX of those minutes attributed to practice and XX to the
-actual test list.
-
-\subsubsection{Coding}\label{coding}
-
-The first author then annotated the onset and offset of all children's
-productions from the audio recording using Praat audio annotation
-software (Boersma \& Weenink, 2020), then wrote and ran a script to
-extract these tokens, pairing them with their original auditory target
-stimulus, and writing these audio pairs out to short .wav files. The
-assistant then listened through all these productions, one target item
-at a time, with productions for each target item presented in a random
-order across children and repetitions. The assistant indicated in a
-notebook whether each production was a correct or incorrect repetition
-and orthographically transcribed the production, noting when the child
-uttered a recognizable word or phrase and adding the translation
-equivalent of that word/phrase into English. The assistant also provided
-some general examples of the types of errors children made without
-making specific reference to Yélî sounds or the items in the elicitation
-sets.
-
-\subsubsection{Analyses}\label{analyses}
-
-In addition to a binary word-level exact repetition score we scored
-items in terms of the number of phonemes that could be aligned across
-the target and attempt, divided by the number of phonemes of whichever
-item was longer (the target or the attempt; see also Cristia et al.,
-2020). Although previous work does not use distance metrics, we
-additionally report these. For instance, recall our example of an
-English target being /bilik/ with an imagined response {[}bilig{]}. We
-would score this response as follows: at the whole item level this
-production would receive a score of zero (because the repetition is not
-exact); at the phoneme level this production would receive a score of
-80\% (4 out of 5 phonemes repeated exactly); and the phone-based
-Levenshtein distance for this production is 20\% (because 20\% of
-phonemes were substituted or deleted).\\
-
-Additionally, we estimated the typological frequency of all phonological
-segments used in the target items using the PHOIBLE cross-linguistic
-phonological inventory database (Moran \& McCloy, 2019). For each phone
-in our task, we extracted the number and percentage of languages noted
-to have that phone in its inventory. While PHOIBLE is an unprecedentedly
-comprehensive database, with phonological inventory data for over 2000
-languages at time of writing, it is of course still far from complete,
-so estimated frequencies should be taken with a pinch of salt. In both
-PHOIBLE and UPSID---another large phonemic inventory database---note
-that nearly half of the segment types are only attested in one language
-(Steven Moran, personal communication; see also
-\href{http://web.phonetik.uni-frankfurt.de/upsid_info.html}{web.phonetik.uni-frankfurt.de/upsid\_info.html}).
-It is therefore unsurprising that we did not find three of our segments
-(/lβʲ/, /ṭp/, and /tp/) in the database. Following the general pattern
-that many phones only appear once cross-linguistically, even in large
-databases, we treat our ``missing'' phones as having a frequency of 1,
-with a (rounded) percentage of 0\%.
-
-Finally, when describing children's patterns of errors, we take all
-repetitions of a given target into account. We describe the proportion
-of items where the change resulted in a real word (semantic errors); as
-well as deletions, additions, and substitutions.
-
-\subsubsection{Participants}\label{participants}
-
-This study was approved as part of a larger research effort by the
-second author. The line of research was evaluated by the Radboud
-University Faculty of Social Sciences Ethics Committee (Ethiek Commissie
-van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The
-Netherlands (original request: ECSW2017-3001-474 Manko-Rowland;
-amendment: ECSW-2018-041). Participation was voluntary; children were
-invited to participate following indication of approval from an adult
-caregiver. Regardless of whether they completed the task, children were
-given a small snack as compensation. Children who showed initial
-interest but then decided not to participate were also given the snack.
-
-We tested a total of 55 children from 38 families spread across four
-hamlets. We excluded test sessions from analysis for the following
-reasons: refused participation or failure to repeat items presented over
-headphones even after coaching (N=8), spoke too softly to allow offline
-coding (N=5), or were 13 years old or older (N=2; we tested these
-teenagers to put younger children at ease). The remaining 40 children
-(14 girls) were aged 6.98 years (range 3.92-11.03 years). Among these
-children were 34 only exposed to Yélî Dnye at home and 6 children
-exposed to Yélî Dnye plus one or more other languages at home. Maternal
-years of education averaged 8.22 years (range 6-12 years).\footnote{Years
-  of education is difficult to precisely pinpoint in this context.
-  Schooling beyond grade 8 takes place off-island and enrollment is tied
-  to an ability to pay that year's school fees, seasonal interruptions
-  to school opening times, familial obligations, and more. We asked
-  parents to report their highest completed level of education. We then
-  record the number of years entailed by having completed that level
-  under ideal conditions. This method results in many esimates around
-  typical ``graduation'' times (e.g., graduating from primary education
-  before starting secondary education).} In terms of birth order, 6 were
-first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth.
-
-\subsection{Results}\label{results}
-
-\subsubsection{Preliminary analyses}\label{preliminary-analyses}
+\hypertarget{procedure}{%
+\subsubsection{Procedure}\label{procedure}}
+
+In adapting the typical NWR procedure for this context, we balanced three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them (i.e., from other children who had participated); that children would feel comfortable doing this task with us; and that community members would feel comfortable having their children do this task with us.
+
+We ran children in batches, testing within a handful of hamlet clusters spread across the northeastern region of the island.
+Because space availability was limited in different ways from hamlet to hamlet, the places where elicitation happened varied across testing sites.
+We tested in four different sites, only making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from \enquote{spreading} between children through hearsay.
+In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
+The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from the online supplementary materials.
+Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
+
+We fitted the child with a headset microphone (Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former) that fed into the left channel of a Tascam DR40x digital audio recorder.
+The headsets were designed for adult use and could not be comfortably seated on many children's heads without a more involved adjustment period.
+To minimize adjustment time, which was uncomfortable for some children given the proximity of the experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth.
+A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement.
+The research assistant coached the child throughout the task to make sure that they understood what they were expected to do.
+An experimenter (the first author) delivered the pre-recorded stimuli to the research assistant and the child over headphones.
+
+The first phase of the experiment involved making sure the child understood the task.
+We explained the task and then orally presented the first practice item.
+At this point, many children did not say anything in response, which triggered the following procedure: First, the assistant insisted the child make a response.
+If the child still did not say anything, the assistant said a real word and then asked the child to repeat it, then another and another.
+If the child could repeat real words correctly, we provided the first training item over headphones again for children to repeat.
+Most children successfully started repeating the items at this point, but a few needed further help.
+In this case, the assistant modeled the behavior (i.e., the child and assistant would hear the item again, and the assistant would repeat it; then we would play the item again and ask the child to repeat it).
+A small minority of children still failed to repeat the item at this point.
+If so, we tried again with the second training item, at which point some children demonstrated task understanding and could continue.
+A fraction of the remaining children, however, failed to repeat this second training item, as well as the third one, in which case we stopped testing altogether (see Participants section for exclusions).
+
+The second phase of the experiment involved running the child through the list of test items randomly assigned to them.
+This was done in the same manner as the practice items: the stimulus was played over the headphones, and then the child repeated it aloud.
+NWR studies vary in whether children are allowed to hear and/or repeat the item more than one time.
+We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was alowed to make a further attempts if their first attempt was judged erroneous in some way by the assistant.
+The procedure worked as follows: When the child made an attempt, the assistant indicated to the experimenter whether the child's production was correct or not.
+If correct, the experimenter would whisper this note of correct repetition into a separate headset that fed into the right channel of the same Tascam recorder and the group moved on to the next item.
+If not, the child was allowed to try again, with up to five attempts allowed before moving on to the next item.
+Children were not asked to make repetitions if they did not produce a first attempt.
+In total, test sessions took approximately XX minutes on average, with XX of those minutes attributed to practice and XX to the actual test list.
+
+\hypertarget{coding}{%
+\subsubsection{Coding}\label{coding}}
+
+The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software (Boersma \& Weenink, 2020),
+then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files.
+The assistant then listened through all these productions, one target item at a time, with productions for each target item presented in a random order across children and repetitions.
+The assistant indicated in a notebook whether each production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English.
+The assistant also provided some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
+
+\hypertarget{analyses}{%
+\subsubsection{Analyses}\label{analyses}}
+
+In addition to a binary word-level exact repetition score we scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt; see also Cristia et al., 2020).
+Although previous work does not use distance metrics, we additionally report these.
+For instance, recall our example of an English target being /bilik/ with an imagined response {[}bilig{]}.
+We would score this response as follows: at the whole item level this production would receive a score of zero (because the repetition is not exact); at the phoneme level this production would receive a score of 80\% (4 out of 5 phonemes repeated exactly); and the phone-based Levenshtein distance for this production is 20\% (because 20\% of phonemes were substituted or deleted).\\
+
+Additionally, we estimated the typological frequency of all phonological segments used in the target items using the PHOIBLE cross-linguistic phonological inventory database (Moran \& McCloy, 2019).
+For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at time of writing, it is of course still far from complete, so estimated frequencies should be taken with a pinch of salt.
+In both PHOIBLE and UPSID---another large phonemic inventory database---note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication; see also \href{http://web.phonetik.uni-frankfurt.de/upsid_info.html}{web.phonetik.uni-frankfurt.de/upsid\_info.html}).
+It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database.
+Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat our \enquote{missing} phones as having a frequency of 1, with a (rounded) percentage of 0\%.
+
+Finally, when describing children's patterns of errors, we take all repetitions of a given target into account.
+We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.
+
+\hypertarget{participants}{%
+\subsubsection{Participants}\label{participants}}
+
+This study was approved as part of a larger research effort by the second author.
+The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
+Regardless of whether they completed the task, children were given a small snack as compensation.
+Children who showed initial interest but then decided not to participate were also given the snack.
+
+We tested a total of 55 children from 38 families spread across four hamlets.
+We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5), or were 13 years old or older (N=2; we tested these teenagers to put younger children at ease).
+The remaining 40 children (14 girls) were aged 6.98 years (range 3.92-11.03 years).
+Among these children were 34 only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.
+Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{Years of education is difficult to precisely pinpoint in this context.
+Schooling beyond grade 8 takes place off-island and enrollment is tied to an ability to pay that year's school fees, seasonal interruptions to school opening times, familial obligations, and more.
+We asked parents to report their highest completed level of education.
+We then record the number of years entailed by having completed that level under ideal conditions.
+This method results in many esimates around typical \enquote{graduation} times (e.g., graduating from primary education before starting secondary education).} In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth.
+
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
+
+\hypertarget{preliminary-analyses}{%
+\subsubsection{Preliminary analyses}\label{preliminary-analyses}}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig1-first-vs-others-1.pdf}
-\caption{\label{fig:Fig1-first-vs-others}Whole-item NWR scores for
-individual participants averaging separately their first attempts and
-all other attempts.}
+\caption{\label{fig:Fig1-first-vs-others}Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts.}
 \end{figure}
 
-We first checked whether whole-item NWR scores varied between first and
-subsequent presentations of an item by averaging word-level scores at
-the participant level separately for first attempts and subsequent
-repetitions. We excluded 1 child who did not have data for one of these
-two types. As shown in Figure \ref{fig:Fig1-first-vs-others},
-participants' mean word-level scores became more heterogeneous in
-subsequent repetitions. Surprisingly, whole-item NWR scores for
-subsequent repetitions (M = 40, SD = 26) were on average lower than
-first ones (M = 64, SD = 15), t(38) = 6.28, p \textless{} 0.001; Cohen's
-d = 1.14). Given uncertainty in whether previous work used first or all
-repetitions, and given that score here declined and became more
-heterogeneous in subsequent repetitions, we focus the remainder of our
-analyses only on first repetitions.
-
-Taking into account only the first attempts, we averaged attempts by
-each of the 40 children who had data for first attempts; their ages
-ranged from 3 to 10 (M = 6.50, SD = 1.50). The overall NWR score was M =
-64\% (SD = 15\%), Cohen's d = 4.36. Scores based on phonemes are even
-higher M = 78\% (SD = 9\%), Cohen's d = 8.55. The phoneme-based
-normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that
-about a fifth of phonemes were were substituted, inserted, or deleted
-(the normalized Levenshtein distance is the complement of phoneme-based
-scores).
-
-\subsubsection{NWR as a function of cross-linguistic phone
-frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}
-
-We then analyzed variation in whole-item NWR scores as a function of the
-average frequency with which sounds composing individual target words
-are found in languages over the world. To look at this, we fit a mixed
-logistic regression in which the outcome variable was whether the
-non-word was correctly repeated or not. The fixed effect of interest was
-the average cross-linguistic phone frequency; we also included child age
-as a control fixed effect, and child ID \& target ID as random
-variables.
-
-We could include 850 observations, from 40 children producing in any
-given trial one of 41 potential target words. The analysis revealed a
-main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant
-estimate for the scaled average cross-linguistic frequency of phones in
-the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target
-words with phones found more frequently across languages had a higher
-proportion of words that were correctly repeated, as shown in Figure
-\ref{fig:Fig2-xling-freq}. Averaging across items and participants, the
-Pearson correlation between scaled average cross-linguistic phone
-frequency and whole-item NWR scores was r(31) = 0.63.
+We first checked whether whole-item NWR scores varied between first and subsequent presentations of an item by averaging word-level scores at the participant level separately for first attempts and subsequent repetitions.
+We excluded 1 child who did not have data for one of these two types.
+As shown in Figure \ref{fig:Fig1-first-vs-others}, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
+Surprisingly, whole-item NWR scores for subsequent repetitions (M = 40, SD = 26)
+were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p \textless{} 0.001; Cohen's d = 1.14).
+Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions.
+
+Taking into account only the first attempts, we averaged attempts by each of the 40 children who had data for first attempts; their ages ranged from 3 to 10 (M = 6.50, SD = 1.50).
+The overall NWR score was M = 64\% (SD = 15\%), Cohen's d = 4.36.
+Scores based on phonemes are even higher M = 78\% (SD = 9\%), Cohen's d = 8.55.
+The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted (the normalized Levenshtein distance is the complement of phoneme-based scores).
+
+\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
+\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}
+
+We then analyzed variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
+The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID \& target ID as random variables.
+
+We could include 850 observations, from 40 children producing in any given trial one of 41 potential target words.
+The analysis revealed a main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as shown in Figure \ref{fig:Fig2-xling-freq}.
+Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(32) = 0.61.
 
@@
 
-We next checked whether the association between whole-item NWR scores
-and cross-linguistic phone frequence could actually be due to frequency
-of the sounds within the language: One can suppose that sounds that
-occur more frequently across languages are also more frequent within a
-language, and therefore may be easier for children to represent and
-repeat. We estimated frequency of the phones present in the stimuli in a
-corpus of child-centered recordings (Casillas et al., 2020) by counting
-the number of word types in which they occurred, and applied the natural
-logarithm.\footnote{We also carried out this analysis using token phone
-  frequency, but this measure was not correlated with whole-item NWR
-  scores, and therefore the fact that it did not explain away the
-  predictive value of cross-linguistic phone frequency was less
-  informative than the relationship discussed in the main text, with
-  type frequencies.} As before, unattested sounds were not considered
-(i.e., they were declared NA so that they do not count for analyses).
-Phone frequencies estimated from this corpus were correlated with
-cross-linguistic phone frequencies (r(27)=0.52, p \textless{} 0.01).
-Moreover, averaging across items and participants, the Pearson
-correlation between scaled average corpus phone frequency and whole-item
-NWR scores was r = 0.46. Therefore, we fit another mixed logistic
-regression, this time declaring as fixed effects both scaled
-cross-linguistic and corpus frequencies (averaged across all attested
-phones within each stimulus item), in addition to age. As before, the
-model contained random slopes for both child ID and target. In this
-model, both cross-linguistic frequency (ß = 0.82, SE ß = 0.27, p
-\textless{} 0.01) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were
-significant predictors of whole-item NWR scores, but corpus frequency (ß
-= 0.00, SE ß = 0.25, p = 1) was not.
-
-\subsubsection{Patterns in NWR
-mispronunciations}\label{patterns-in-nwr-mispronunciations}
-
-Next, we investigated patterns of deletion and substitution. Deletions
-were relatively rare: there were only 7 instances of deleted vowels
-(\textasciitilde{}0.23\% of all vowel targets), and 3 instances of
-deleted consonants (\textasciitilde{}0.12\% of all consonant targets).
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.} As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(27)=0.50, p \textless{} 0.01).
+Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = 0.46.
+Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
+As before, the model contained random slopes for both child ID and target.
+In this model, both cross-linguistic frequency (ß = 0.82, SE ß = 0.27, p \textless{} 0.01) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of whole-item NWR scores, but corpus frequency (ß = 0.01, SE ß = 0.25, p = 0.98) was not.
+
+\hypertarget{patterns-in-nwr-mispronunciations}{%
+\subsubsection{Patterns in NWR mispronunciations}\label{patterns-in-nwr-mispronunciations}}
+
+Next, we investigated patterns of deletion and substitution.
+Deletions were relatively rare: there were only 7 instances of deleted vowels (\textasciitilde{}0.23\% of all vowel targets), and 3 instances of deleted consonants (\textasciitilde{}0.12\% of all consonant targets).
 Substitutions were somewhat more frequent, but still relatively rare.
-Nasal vowels were produced as oral vowels and vice versa (34 oral
-targets as nasal (\textasciitilde{}0.99\% of targets) and 30 nasal
-targets produced as oral (\textasciitilde{}2.37\% of targets)). Both
-oral and nasal vowels were also occasionally subsituted by vowels of a
-different quality that preserved the target oral/nature airflow (110
-oral vowels (\textasciitilde{}1.86\% of targets and 13 nasal vowels
-(\textasciitilde{}2.06\% of targets). Among consonants, complex
-consonants were occasionally substituted by simpler ones (specifically
-{[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and
-{[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]},
-{[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and
-{[}tʃ{]}), 115 times, in approximately 1.96\% of all complex consonant
-targets. Interestingly, simple consonants were sometimes also
-substituted by complex ones; this happened 2 times, in approximately
-1.96\% of all simple consonant targets. Finally, simple consonants were
-mispronounced as other simple consonants (68 instances,
-\textasciitilde{}0.38\% of all simple consonant targets) and complex
-consonants as other complex consonants (33 instances,
-\textasciitilde{}1.02\% of all complex consonant targets). Note that the
-substitutions included phones that are not native to Yélî Dnye but do
-occur in English (e.g., {[}tʃ{]}); as these data come from careful
-transcriptions by a native Yélî Dnye speaker who is very fluent in
-English, we take these segmental substitutions faithfully as an
-indication that several of our participants have mastered production of
-some English phones, possibly produced within whole English word forms
-(see below).
-
-Finally, we looked into the frequency with which mispronunciations
-resulted in real words. In fact, two thirds of incorrect repetitions
-were recognizable as real words or phrases in Yélî Dnye or English:
-63\%. This type of analysis is seldom reported---we could only find one
-comparison point: ({\textbf{???}}) found that illiterate Portuguese
-adults' NWR mispronunciations resulted in real words in 11.16\% of
-cases, whereas literate participants did so in only 1.71\% of cases. The
-percentage we observe here is much higher than reported in Castro and
-colleagues' study, but we do not know whether age, language, or even
-test structure explains this difference.
-
-\subsubsection{NWR scores as a function of item
-length}\label{nwr-scores-as-a-function-of-item-length}
-
-Next, we inspected whether NWR scores varied as a function of word
-length (Table \ref{tab:tablength}). Participants scored much lower on
-monosyllables than on non-words of other lengths. This is likely due to
-the fact that the majority of monosyllables were designed to include
-sounds that are rare in the world's languages, which may indicate that
-they are hard to produce or perceive. Setting monosyllables aside, we
-observe the typical pattern of lower scores for longer items, although
-this is particularly salient for the whole-item scoring. While
-whole-item scoring is the most commonly reported NWR outcome, it is also
-the least forgiving. The pattern is less marked when other two scores
-are used, which are less sensitive to errors. Averaging across
-participants and items, the Pearson correlation between length (2-4
-syllables) and whole-item NWR scores was r(1) = -0.91. In a generalized
-binomial mixed model, we included 479 observations, from 40 children
-producing, in any given trial, one of 24 potential target words. The
-analysis revealed a positive effect of age (ß = 0.56, SE ß = 0.14, p
-\textless{} 0.001) and a negative but non-significant estimate for
-target length in number of syllables (ß = -0.15, SE ß = 0.33, p = 0.65).
+Nasal vowels were produced as oral vowels and vice versa (34 oral targets as nasal (\textasciitilde{}0.99\% of targets) and 30 nasal targets produced as oral (\textasciitilde{}2.37\% of targets)).
+Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (110 oral vowels (\textasciitilde{}1.86\% of targets and 13 nasal vowels (\textasciitilde{}2.06\% of targets).
+Among consonants, complex consonants were occasionally substituted by simpler ones (specifically {[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), 115 times, in approximately 1.96\% of all complex consonant targets.
+Interestingly, simple consonants were sometimes also substituted by complex ones; this happened 2 times, in approximately 1.96\% of all simple consonant targets.
+Finally, simple consonants were mispronounced as other simple consonants (68 instances, \textasciitilde{}0.38\% of all simple consonant targets) and complex consonants as other complex consonants (33 instances, \textasciitilde{}1.02\% of all complex consonant targets).
+Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms (see below).
+
+Finally, we looked into the frequency with which mispronunciations resulted in real words.
+In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: 63\%.
+This type of analysis is seldom reported---we could only find one comparison point: ({\textbf{???}}) found that illiterate Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
+
+\hypertarget{nwr-scores-as-a-function-of-item-length}{%
+\subsubsection{NWR scores as a function of item length}\label{nwr-scores-as-a-function-of-item-length}}
+
+Next, we inspected whether NWR scores varied as a function of word length (Table \ref{tab:tablength}).
+Participants scored much lower on monosyllables than on non-words of other lengths.
+This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may indicate that they are hard to produce or perceive.
+Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole-item scoring.
+While whole-item scoring is the most commonly reported NWR outcome, it is also the least forgiving.
+The pattern is less marked when other two scores are used, which are less sensitive to errors.
+Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(1) = -0.91.
+In a generalized binomial mixed model, we included 479 observations, from 40 children producing, in any given trial, one of 24 potential target words.
+The analysis revealed a positive effect of age (ß = 0.56, SE ß = 0.14, p \textless{} 0.001) and a negative but non-significant estimate for target length in number of syllables (ß = -0.15, SE ß = 0.33, p = 0.65).
 
 \begin{table}[tbp]
+
 \begin{center}
 \begin{threeparttable}
+
 \caption{\label{tab:tablength}NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
+
 \begin{tabular}{lll}
@@
 \end{tabular}
+
 \end{threeparttable}
 \end{center}
+
 \end{table}
 
-\subsubsection{Factor structuring individual
-variation}\label{factor-structuring-individual-variation}
+\hypertarget{factor-structuring-individual-variation}{%
+\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}
 
@@
 \includegraphics{manuscript_files/figure-latex/Fig2-scores-by-age-1.pdf}
-\caption{\label{fig:Fig2-scores-by-age}NWR whole-item scores for individual
-participants as a function of age and sex (purple = boys, orange =
-girls).}
+\caption{\label{fig:Fig2-scores-by-age}NWR whole-item scores for individual participants as a function of age and sex (purple = boys, orange = girls).}
 \end{figure}
 
-Our final exploratory analysis assessed whether variation in scores was
-structured by factors that vary across individuals. As shown in Figure
-\ref{fig:Fig2-scores-by-age}, there was a greater deal of variance
-across the tested age range, with significantly higher NWR scores for
-older children (Spearman's rank correlation, given inequality of
-variance, rho (6,014.70) = 0.44, p \textless{} 0.01). In contrast, there
-was no clear association between NWR scores and sex (t (-0.29) = 27.56,
-p = 0.77), birth order (data missing for 15 children, rho = (3,441.90) =
--0.18, p = 0.39), or maternal education (data missing for 0 children,
-rho ( 9,594.37) = 0.10, p = 0.54).
-
-\subsection{Discussion}\label{discussion}
-
-We used non-word repetition to investigate phonological development in a
-language with a large phonological inventory (including some
-typologically rare segments). While the study, in itself, advances NWR
-research by demonstrating the successful application of this method in a
-rural, non-industrialized population (see also Cristia et al., 2020),
-the results also inform the larger body of NWR work regarding the
-influence of stimulus length, segmental frequency, age, and
-developmental context, where previous findings have often been mixed.
-
-In our sample of 40 children between ages 3 and 10, the average
-item-level score was 64\%, falling well within the expected range of NWR
-applications in previous studies on a variety of linguistic populations
-(see Figure \ref{fig:prevlit-fig}). Segment-level analyses revealed that
-repetitions recovered an average of 78\% of the segments in the original
-stimulus, indicating that inexact repetitions often constituted
-\emph{mostly} accurate reroductions of the target. Complementary data
-from segment-based normalized Levenshtein distance reflect this same
-pattern, with an average of 22\% of the stimulus segments getting
-substituted, inserted, or deleted upon repetition. Paired with our
-thorough training protocol, we take these NWR scores as indicating that
-(a) our adaptations to NWR for this context were successful, even given
-a number of non-standard changes to the training phase and to the design
-of the stimuli and (b) Yélî children show comparable performance to
-others tested on a similar task, despite the many linguistic, cultural,
-and socioeconomic differences between this and previously tested
-populations. Given successful (and comparable) execution of this task,
-we can dive deeper into effects of stimulus length, segmental frequency,
-age, and developmental context.
-
-\subsubsection{Item complexity}\label{item-complexity}
-
-We investigated the effect of item complexity on NWR scores by varying
-both the number of syllables in the item and its average segmental
-frequency. Based on previous work, we had predicted that children would
-have higher NWR scores for shorter items. That said, previous work has
-shown both very small (Piazzalunga et al., 2019) and very large (Cristia
-et al., 2020; Jaber-Awida, 2018) effects of stimulus length and,
-further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic
-words are nearly equally frequent in the current language, with
-trisyllabic and longer words making up a non-trivial 10\% of the
-remaining words. Compare this to, for example, English, which is
-substantially more skewed toward monosyllabic word forms **M2A: Alex I'm
-going off your note here (``Prediction for Yélî made before seeing the
-data: The length distribution in Yélî words is more balanced than that
-in English, and thus the score decline for poly- versus mono-syllables
-may be less pronounced than that for English.''``). I don't have a
-reference for this, can you please finish the thought or nix this
-bit?**. Setting aside our monosyllabic stimuli, which all contained
-typologically infrequent segments, we can examine effects of item length
-among the remaining stimuli, which range between 2 and 4 syllables long.
-While indeed NWR scores were overall lower for longer items (e.g., see
-Figure \ref{fig:prevlit-fig}), the effect of item length was not
-significant in a statistical model that additionally accounted for age
-and random effects of item and participant. In light of mixed prior
-results of item length, we propose two possible (and non-mutually
-exclusive) explanations for this minimal impact of item length. First,
-further extensions of this type of analysis in more populations may
-reveal that, in general (and cross-linguistically), item length effects
-are variable between languages, potentially reflecting the distribution
-of word lengths in the ambient language and other (morpho-)phonological
-tendencies in the lexicon. Second, above and beyond these
-language-specific effects, the general impact of item length on NWR
-score may be relatively small, as shown in Piazzalunga et al.'s (2019)
-study on Italian and as borne out in the current dataset once
-controlling for other factors.
-
-Our monosyllabic items included typologically rare segments so that we
-could test whether lower average segmental frequency is associated with
-lower NWR scores. Typologically common sounds are associated with higher
-performance on a handful of other tasks (\textbf{REFS -- M2A: Alex, I
-added this based on your note, where it sounded like you had some
-particular studies in mind?}) though to our knowledge this has not yet
-been tested with non-word repetition. Regarding Yélî Dnye in particular,
-the phonemic inventory is both large and acoustically packed, in
-addition to containing several typologically infrequent (or unique)
-contrasts. We therefore expected to see that, while NWR scores would be
-lower for stimuli with lower average frequency, this effect would be
-relatively weak because the ambient language puts pressure on Yélî
-children to distinguish (perceptually and articulatorily) fine-grained
-phonetic differences in order to successfully communicate with others.
-Indeed we found a robust effect of average segmental frequency on NWR
-performance: Even accounting for age and random effects of item and
-participant, we see that target words with more frequent segments were
-repeated correctly more often. This effect is large, with a magnitude
-more than twice the size of the effect of participant age. This
-significant effect remains even once also accounting for the frequencies
-of these segments in Yélî Dnye child-directed speech, which are
-correlated with their typological frequencies. In sum, typological
-frequency effects, which have been found in other measurements of
-phonological processing. appear to strongly affect NWR performance, and
-do not appear mitigated by language-specific pressure to make
-finer-grained differences earlier in development.
-
-With respect to the types of errors in repetition made, we did not see
-clear patterns to further guide our discussion: base rates of deletion
-and substitution were fairly low and the relative distribution of errors
-over, e.g., nasal vs.~oral vowels and simple vs.~complex consonants,
-revealed no remarkable bias in error types. That said, the lack of a
-difference could be due to relative imbalance across our stimuli in the
-use of these phonemic features (e.g., we included many more more oral
-than nasal targets) and future work should investigate such sources of
-error bias more systemtically. Some portion of the errors were
-introduced when the participant produced a real word (in Yélî Dnye or
-English) in response to the stimulus. Real-word repetitions here made up
-two thirds of errorful repetitions---this is quite high compared to past
-work (e.g., {\textbf{???}}), but it is unclear what caused this pattern
-in the current study: Castro and colleagues' ({\textbf{???}}) study
-focused on adults rather than children, the task was administered by a
-team including a foreign, English-speaking researcher, and the
-particularities of the Yélî Dnye phonological inventory result in many
-true-word phonetic neighbors. Follow-up work exploring this type of
-error in children from other populations in addition to further work on
-Yélî children will clarify this effect.
-
-\subsubsection{Individual differences}\label{individual-differences}
-
-A review of previous work (see Introduction) suggested that our
-anticipated sample size would not be sufficient to detect most
-individual differences using NWR. We give a brief overview of individual
-difference patterns of four types in the present data---age, sex, birth
-order, and maternal education---hoping that these findings can
-contribute to future meta-analytic efforts aggregating over smaller
-studies such as ours.
-
-Following prior work, we expected that NWR scores would increase with
-participant age (Farmani et al., 2018; Kalnak et al., 2014; Vance et
-al., 2005). Indeed, age was significantly correlated with NWR score and
-also showed up as a significant predictor of NWR score when included as
-a control factor in the analyses of both item length and average
-segmental frequency. In brief, our results underscore the idea that
-phoonlogical development continues well past the first few years of
-life, extending into middle childhood and perhaps later (Hazan \&
-Barrett, 2000).
-
-In contrast, previous work shows little evidence for effects of maternal
-education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \&
-Armon-Lotem, 2017) or participant gender (Chiat \& Roy, 2007) on NWR
-scores. In addition to this prior work, education on Rossel Island,
-while generally highly valued, is not at all essential to ensuring one's
-success in society and may not be a reliable index of local
-socioeconomic variation. There is also limited variation in maternal
-education across the families in the region of the island where we
-sampled. We therefore expected little evidence for impact of either
-participant gender or maternal education in the present study. On the
-other hand, these predictors have established effects on other language
-development measures (\textbf{REFS: M2A: Alex go ahead and pick your
-faves here}). So to the extent that NWR scores share causal links to
-gender-based differences ein development and maternal linguistic input
-with these other language ouctome measures, we might then expect these
-factors to appear in NWR data. In fact, neither participant gender nor
-maternal education were correlated with NWR score in the current data.
-
-Last but not least, we investigated whether birth order might affect NWR
-scores, as it does other language tasks, resulting in first-born
-children showing higher scores on standardized language tests than
-later-born children (Havron et al., 2019), presumably because later-born
-children receive a smaller share of maternal input than their older
-siblings. Given shared caregiving practices and the hamlet organization
-typical of Rossel communities, children have many sources of adult and
-older child input that they encounter on a daily basis and first born
-children quickly integrate with a much larger pool of both older and
-younger children with whom they partly share caregivers. Therefore we
-expected that any effects of birth order on NWR would be attenuated in
-this context. In line with this prediction, our descriptive analysis
-showed no correlation between birth order and NWR score.
-
-\subsubsection{Conclusions}\label{conclusions}
-
-While NWR can, in theory, be used to test a variety of questions about
-phonological development in any language, previous work has been
-primarily limited to a handful of related languages spoken in urban,
-industrialized contexts. The present study shows that, not only can NWR
-be adapted for very different populations than have previously been
-tested, but that effects of age and typological frequency may strongly
-influence phonological development across these diverse settings, while
-effects of item length, participant gender, maternal education, and
-birth order, may either have little impact on this facet of language
-development or have an impact that vaies depending on the linguistic,
-cultural, and sociodemographic properties of the population under study.
-Because these latter predictors strongly relate to other language
-outcomes, the present findings raise the issue of why NWR would pattern
-differently, what that could tell us about the relationship between
-lexical development, phonological development, and the input environment
-and, last but not least, what is implied about the joint applicability
-of these outcome measures as a diagnostic indicator for language delays
-and disorders. In the meanwhile, we take the present findings as
-robustly supporting the idea that phonological development continues
-well past early childhood and as yielding preliminary support for a
-connection between individual learners and global language patterns when
-it comes to acoustic and articulatory markedness.
+Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
+As shown in Figure \ref{fig:Fig2-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (6,014.70) = 0.44, p \textless{} 0.01).
+In contrast, there was no clear association between NWR scores and sex (t (-0.29) = 27.56, p = 0.77), birth order (data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39), or maternal education (data missing for 0 children, rho ( 9,594.37) = 0.10, p = 0.54).
+
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
+
+We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
+While the study, in itself, advances NWR research by demonstrating the successful application of this method in a rural, non-industrialized population (see also Cristia et al., 2020), the results also inform the larger body of NWR work regarding the influence of stimulus length, segmental frequency, age, and developmental context, where previous findings have often been mixed.
+
+In our sample of 40 children between ages 3 and 10, the average item-level score was 64\%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \ref{fig:prevlit-fig}).
+Segment-level analyses revealed that repetitions recovered an average of 78\% of the segments in the original stimulus, indicating that inexact repetitions often constituted \emph{mostly} accurate reroductions of the target.
+Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of 22\% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
+Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
+Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.
+
+\hypertarget{item-complexity}{%
+\subsubsection{Item complexity}\label{item-complexity}}
+
+We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item and its average segmental frequency.
+Based on previous work, we had predicted that children would have higher NWR scores for shorter items.
+That said, previous work has shown both very small (Piazzalunga et al., 2019) and very large (Cristia et al., 2020; Jaber-Awida, 2018) effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10\% of the remaining words.
+Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms \textbf{M2A: Alex I'm going off your note here (\enquote{Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.}").
+I don't have a reference for this, can you please finish the thought or nix this bit?}.
+Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
+While indeed NWR scores were overall lower for longer items (e.g., see Figure \ref{fig:prevlit-fig}), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
+In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length.
+First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon.
+Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s (2019) study on Italian and as borne out in the current dataset once controlling for other factors.
+
+Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
+Typologically common sounds are associated with higher performance on a handful of other tasks (\textbf{REFS -- M2A: Alex, I added this based on your note, where it sounded like you had some particular studies in mind?}) though to our knowledge this has not yet been tested with non-word repetition.
+Regarding Yélî Dnye in particular, the phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
+We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
+Indeed we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often.
+This effect is large, with a magnitude more than twice the size of the effect of participant age.
+This significant effect remains even once also accounting for the frequencies of these segments in Yélî Dnye child-directed speech, which are correlated with their typological frequencies.
+In sum, typological frequency effects, which have been found in other measurements of phonological processing.
+appear to strongly affect NWR performance, and do not appear mitigated by language-specific pressure to make finer-grained differences earlier in development.
+
+With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.~oral vowels and simple vs.~complex consonants, revealed no remarkable bias in error types.
+That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
+Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
+Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work (e.g., {\textbf{???}}), but it is unclear what caused this pattern in the current study: Castro and colleagues' ({\textbf{???}}) study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
+
+\hypertarget{individual-differences}{%
+\subsubsection{Individual differences}\label{individual-differences}}
+
+A review of previous work (see Introduction) suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR.
+We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta-analytic efforts aggregating over smaller studies such as ours.
+
+Following prior work, we expected that NWR scores would increase with participant age (Farmani et al., 2018; Kalnak et al., 2014; Vance et al., 2005).
+Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
+In brief, our results underscore the idea that phoonlogical development continues well past the first few years of life, extending into middle childhood and perhaps later (Hazan \& Barrett, 2000).
+
+In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or participant gender (Chiat \& Roy, 2007) on NWR scores.
+In addition to this prior work, education on Rossel Island, while generally highly valued, is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation.
+There is also limited variation in maternal education across the families in the region of the island where we sampled.
+We therefore expected little evidence for impact of either participant gender or maternal education in the present study.
+On the other hand, these predictors have established effects on other language development measures (\textbf{REFS: M2A: Alex go ahead and pick your faves here}).
+So to the extent that NWR scores share causal links to gender-based differences ein development and maternal linguistic input with these other language ouctome measures, we might then expect these factors to appear in NWR data.
+In fact, neither participant gender nor maternal education were correlated with NWR score in the current data.
+
+Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019), presumably because later-born children receive a smaller share of maternal input than their older siblings.
+Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
+Therefore we expected that any effects of birth order on NWR would be attenuated in this context.
+In line with this prediction, our descriptive analysis showed no correlation between birth order and NWR score.
+
+\hypertarget{conclusions}{%
+\subsubsection{Conclusions}\label{conclusions}}
+
+While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that vaies depending on the linguistic, cultural, and sociodemographic properties of the population under study.
+Because these latter predictors strongly relate to other language outcomes, the present findings raise the issue of why NWR would pattern differently, what that could tell us about the relationship between lexical development, phonological development, and the input environment and, last but not least, what is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders.
+In the meanwhile, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a connection between individual learners and global language patterns when it comes to acoustic and articulatory markedness.
 
@@
 
-\subsection{Acknowledgments}\label{acknowledgments}
-
-We are grateful to the individuals who participated in the study. The
-collection and annotation of these recordings was made possible by
-Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also
-to the PNG National Research Institute, and the Administration of Milne
-Bay Province AC acknowledges financial and institutional support from
-Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge,
-ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and
-the J. S. McDonnell Foundation Understanding Human Cognition Scholar
-Award. MC acknowledges financial support from an NWO Veni Innovational
-Scheme grant (275-89-033).
-
-\section{References}\label{references}
-
-\setlength{\parindent}{-0.5in} \setlength{\leftskip}{0.5in}
+\hypertarget{acknowledgments}{%
+\subsection{Acknowledgments}\label{acknowledgments}}
+
+We are grateful to the individuals who participated in the study.
+The collection and annotation of these recordings was made possible by Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also to the PNG National Research Institute, and the Administration of Milne Bay Province AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.
+S.
+McDonnell Foundation Understanding Human Cognition Scholar Award.
+MC acknowledges financial support from an NWO Veni Innovational Scheme grant (275-89-033).
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
 
 \hypertarget{refs}{}
-\hypertarget{ref-is08042009language}{}
-Action, C. (2009). Language impairment in a multilingual society:
-Linguistic patterns and the road to assessment. \emph{Brussels: COST
-Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
-\hypertarget{ref-balladares2016socio}{}
-Balladares, J., Marshall, C., \& Griffiths, Y. (2016). Socio-economic
-status affects sentence repetition, but not non-word repetition, in
-Chilean preschoolers. \emph{First Language}, \emph{36}(3), 338--351.
-doi:\href{https://doi.org/10.1177/0142723715626067}{10.1177/0142723715626067}
-
-\hypertarget{ref-bergelsoncasillas2019what}{}
+\leavevmode\hypertarget{ref-is08042009language}{}%
+Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
+\leavevmode\hypertarget{ref-balladares2016socio}{}%
+Balladares, J., Marshall, C., \& Griffiths, Y.
+(2016).
+Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
+\emph{First Language}, \emph{36}(3), 338--351.
+\url{https://doi.org/10.1177/0142723715626067}
+
+\leavevmode\hypertarget{ref-bergelsoncasillas2019what}{}%
 Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
-S., \& Amatuni, A. (2019). What do North American babies hear? A
-large-scale cross-corpus analysis. \emph{Developmental Science},
-\emph{22}(1), e12724.
-doi:\href{https://doi.org/10.1111/desc.12724}{10.1111/desc.12724}
-
-\hypertarget{ref-Praat}{}
-Boersma, P., \& Weenink, D. (2020). Praat: Doing phonetics by computer
-(Version 6.1.35). Retrieved from \url{http://www.praat.org/}
-
-\hypertarget{ref-bowey2001nonword}{}
-Bowey, J. A. (2001). Nonword repetition and young children's receptive
-vocabulary: A longitudinal study. \emph{Applied Psycholinguistics},
-\emph{22}(3), 441--469.
-
-\hypertarget{ref-brandeker2015language}{}
-Brandeker, M., \& Thordardottir, E. (2015). Language exposure in
-bilingual toddlers: Performance on nonword repetition and lexical tasks.
-\emph{American Journal of Speech-Language Pathology}, \emph{24}(2),
-126--138.
-
-\hypertarget{ref-brown2011cultural}{}
-Brown, P. (2011). The cultural organization of attention. In A. Duranti,
-E. Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language
-Socialization} (pp. 29--55). Malden, MA: Wiley-Blackwell.
-
-\hypertarget{ref-brown2014interactional}{}
-Brown, P. (2014). The interactional context of language learning in
-Tzeltal. In I. Arnon, M. Casillas, C. Kurumada, \& B. Estigarribia
-(Eds.), \emph{Language in interaction: Studies in honor of Eve V. Clark}
-(pp. 51--82). Amsterdam, NL: John Benjamins.
-
-\hypertarget{ref-brownIPchildrearing}{}
-Brown, P., \& Casillas, M. (in press). Childrearing through social
-interaction on Rossel Island, PNG. In A. J. Fentiman \& M. Goody (Eds.),
-\emph{Esther Goody revisited: Exploring the legacy of an original
-inter-disciplinarian} (pp. XX--XX). New York, NY: Berghahn.
-
-\hypertarget{ref-bunceURcrosscultural}{}
-Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A.,
-Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural
-examination of young children's everyday language experiences.
-
-\hypertarget{ref-casillas2020early}{}
-Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language
-experience in a papuan community. \emph{Journal of Child Language},
-\emph{XX}, XX--XX.
-
-\hypertarget{ref-chiat2007preschool}{}
-Chiat, S., \& Roy, P. (2007). The preschool repetition test: An
-evaluation of performance in typically developing and clinically
-referred children. \emph{Journal of Speech, Language, and Hearing
-Research}, \emph{50}(2), 429--443.
-
-\hypertarget{ref-cristia2020infant}{}
+S., \& Amatuni, A.
+(2019).
+What do North American babies hear?
+A large-scale cross-corpus analysis.
+\emph{Developmental Science}, \emph{22}(1), e12724.
+\url{https://doi.org/10.1111/desc.12724}
+
+\leavevmode\hypertarget{ref-Praat}{}%
+Boersma, P., \& Weenink, D. (2020). Praat: Doing phonetics by computer (Version 6.1.35). Retrieved from \url{http://www.praat.org/}
+
+\leavevmode\hypertarget{ref-bowey2001nonword}{}%
+Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
+
+\leavevmode\hypertarget{ref-brandeker2015language}{}%
+Brandeker, M., \& Thordardottir, E.
+(2015).
+Language exposure in bilingual toddlers: Performance on nonword repetition and lexical tasks.
+\emph{American Journal of Speech-Language Pathology}, \emph{24}(2), 126--138.
+
+\leavevmode\hypertarget{ref-brown2011cultural}{}%
+Brown, P.
+(2011).
+The cultural organization of attention.
+In A.
+Duranti, E.
+Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language Socialization} (pp.
+29--55).
+Malden, MA: Wiley-Blackwell.
+
+\leavevmode\hypertarget{ref-brown2014interactional}{}%
+Brown, P.
+(2014).
+The interactional context of language learning in Tzeltal.
+In I.
+Arnon, M.
+Casillas, C.
+Kurumada, \& B.
+Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of Eve V.
+Clark} (pp.
+51--82).
+Amsterdam, NL: John Benjamins.
+
+\leavevmode\hypertarget{ref-brownIPchildrearing}{}%
+Brown, P., \& Casillas, M.
+(n.d.).
+Childrearing through social interaction on Rossel Island, PNG.
+In A.
+J.
+Fentiman \& M.
+Goody (Eds.), \emph{Esther Goody revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
+XX--XX).
+New York, NY: Berghahn.
+
+\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
+Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (n.d.). A cross-cultural examination of young children's everyday language experiences.
+
+\leavevmode\hypertarget{ref-casillas2020early}{}%
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
+
+\leavevmode\hypertarget{ref-chiat2007preschool}{}%
+Chiat, S., \& Roy, P.
+(2007).
+The preschool repetition test: An evaluation of performance in typically developing and clinically referred children.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(2), 429--443.
+
+\leavevmode\hypertarget{ref-cristia2020infant}{}%
 Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
-(2020). Infant-directed input and literacy effects on phonological
-processing: Non-word repetition scores among the tsimane'.
+(2020).
+Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
 \emph{Preprint}.
 
-\hypertarget{ref-estes2007differences}{}
-Estes, K. G., Evans, J. L., \& Else-Quest, N. M. (2007). Differences in
-the nonword repetition performance of children with and without specific
-language impairment: A meta-analysis. \emph{Journal of Speech, Language,
-and Hearing Research}, \emph{50}(1), 177--195.
-
-\hypertarget{ref-farmani2018normalization}{}
-Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \&
-Shourvazi, Z. (2018). Normalization of the non-word repetition test in
-farsi-speaking children. \emph{Journal of Modern Rehabilitation},
-\emph{12}(4), 217--224.
-
-\hypertarget{ref-foley1986papuan}{}
-Foley, W. A. (1986). \emph{The papuan languages of new guinea}.
-Cambridge, UK: Cambridge University Press.
-
-\hypertarget{ref-gallagher2014identity}{}
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from
-Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378.
-doi:\href{https://doi.org/10.1515/lp-2014-0012}{10.1515/lp-2014-0012}
-
-\hypertarget{ref-gallon2007non}{}
-Gallon, N., Harris, J., \& Van der Lely, H. (2007). Non-word repetition:
-An investigation of phonological complexity in children with grammatical
-sli. \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
-
-\hypertarget{ref-havron2019effect}{}
-Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \&
-Group, E. M.-C. C. S. (2019). The effect of older siblings on language
-development as a function of age difference and sex. \emph{Psychological
-Science}, \emph{30}(9), 1333--1343.
-
-\hypertarget{ref-hazan2000development}{}
-Hazan, V., \& Barrett, S. (2000). The development of phonemic
-categorization in children aged 6--12. \emph{Journal of Phonetics},
-\emph{28}(4), 377--396.
-
-\hypertarget{ref-jabere2018xperiment}{}
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual
-Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4),
-317--334.
-doi:\href{https://doi.org/10.30958/ajp.5-4-4}{10.30958/ajp.5-4-4}
-
-\hypertarget{ref-kalnak2014nonword}{}
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014).
-Nonword repetition--a clinical marker for specific language impairment
-in swedish associated with parents' language-related problems.
+\leavevmode\hypertarget{ref-estes2007differences}{}%
+Estes, K.
+G., Evans, J.
+L., \& Else-Quest, N.
+M.
+(2007).
+Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
+
+\leavevmode\hypertarget{ref-farmani2018normalization}{}%
+Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
+Z., Talebi, E., \& Shourvazi, Z.
+(2018).
+Normalization of the non-word repetition test in farsi-speaking children.
+\emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
+
+\leavevmode\hypertarget{ref-foley1986papuan}{}%
+Foley, W. A. (1986). \emph{The papuan languages of new guinea}. Cambridge, UK: Cambridge University Press.
+
+\leavevmode\hypertarget{ref-gallagher2014identity}{}%
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
+
+\leavevmode\hypertarget{ref-gallon2007non}{}%
+Gallon, N., Harris, J., \& Van der Lely, H.
+(2007).
+Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
+\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\leavevmode\hypertarget{ref-havron2019effect}{}%
+Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
+M.-C.
+C.
+S.
+(2019).
+The effect of older siblings on language development as a function of age difference and sex.
+\emph{Psychological Science}, \emph{30}(9), 1333--1343.
+
+\leavevmode\hypertarget{ref-hazan2000development}{}%
+Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
+
+\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
+
+\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
+(2014).
+Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
 \emph{PloS One}, \emph{9}(2), e89544.
 
-\hypertarget{ref-ladefoged1996sounds}{}
-Ladefoged, P., \& Maddieson, I. (1996). \emph{The sounds of the world's
-languages} (Vol. 1012). Blackwell Oxford.
-
-\hypertarget{ref-levinsonYDgrammar}{}
-Levinson, S. C. (accepted). \emph{A grammar of yélî dnye, the papuan
-language of rossel island}. Berlin, Boston: De Gruyter Mouton.
-
-\hypertarget{ref-liszkowski2012prelinguistic}{}
+\leavevmode\hypertarget{ref-ladefoged1996sounds}{}%
+Ladefoged, P., \& Maddieson, I. (1996). \emph{The sounds of the world's languages} (Vol. 1012). Blackwell Oxford.
+
+\leavevmode\hypertarget{ref-levinsonYDgrammar}{}%
+Levinson, S. C. (n.d.). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Berlin, Boston: De Gruyter Mouton.
+
+\leavevmode\hypertarget{ref-liszkowski2012prelinguistic}{}%
 Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
-(2012). A prelinguistic gestural universal of human communication.
+(2012).
+A prelinguistic gestural universal of human communication.
 \emph{Cognitive Science}, \emph{36}(4), 698--713.
-doi:\href{https://doi.org/10.1111/j.1551-6709.2011.01228.x}{10.1111/j.1551-6709.2011.01228.x}
-
-\hypertarget{ref-maddieson2005correlating}{}
-Maddieson, I. (2005). Correlating phonological complexity: Data and
-validation. \emph{UC Berkeley PhonLab Annual Report}, \emph{1}(1).
-
-\hypertarget{ref-maddiesonIPphoneticsYD}{}
-Maddieson, I., \& Levinson, S. C. (in preparation). \emph{The phonetics
-of yélî dnye, the language of rossel island}.
-
-\hypertarget{ref-meir2017independent}{}
-Meir, N., \& Armon-Lotem, S. (2017). Independent and combined effects of
-socioeconomic status (ses) and bilingualism on children's vocabulary and
-verbal short-term memory. \emph{Frontiers in Psychology}, \emph{8},
-1442.
-
-\hypertarget{ref-meir2016disentangling}{}
-Meir, N., Walters, J., \& Armon-Lotem, S. (2016). Disentangling sli and
-bilingualism using sentence repetition tasks: The impact of l1 and l2
-properties. \emph{International Journal of Bilingualism}, \emph{20}(4),
-421--452.
-
-\hypertarget{ref-phoible}{}
-Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max
-Planck Institute for the Science of Human History. Retrieved from
-\url{https://phoible.org/}
-
-\hypertarget{ref-peuteIPconsonants}{}
-Peute, A. A. K., Fikkert, P., \& Casillas, M. (In preparation). Early
-consonant production in Yélî Dnye and Tseltal.
-
-\hypertarget{ref-piazzalunga2019articulatory}{}
-Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler,
-A. (2019). An articulatory-based disyllabic and trisyllabic non-word
-repetition test: Reliability and validity in italian 3-to 7-year-old
-children. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5),
-437--456.
-
-\hypertarget{ref-torrington2015non}{}
-Torrington Eaton, C., Newman, R. S., Ratner, N. B., \& Rowe, M. L.
-(2015). Non-word repetition in 2-year-olds: Replication of an adapted
-paradigm and a useful methodological extension. \emph{Clinical
-Linguistics \& Phonetics}, \emph{29}(7), 523--535.
-
-\hypertarget{ref-vance2005speech}{}
-Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills
-in children aged 3--7 years. \emph{International Journal of Language \&
-Communication Disorders}, \emph{40}(1), 29--48.
-
-\hypertarget{ref-wilsenach2013phonological}{}
-Wilsenach, C. (2013). Phonological skills as predictor of reading
-success: An investigation of emergent bilingual Northern Sotho/English
-learners. \emph{Per Linguam: a Journal of Language Learning= Per
-Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
-doi:\href{https://doi.org/10.5785/29-2-554}{10.5785/29-2-554}
-
-\hypertarget{ref-is08042009language}{}
-Action, C. (2009). Language impairment in a multilingual society:
-Linguistic patterns and the road to assessment. \emph{Brussels: COST
-Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
-\hypertarget{ref-balladares2016socio}{}
-Balladares, J., Marshall, C., \& Griffiths, Y. (2016). Socio-economic
-status affects sentence repetition, but not non-word repetition, in
-Chilean preschoolers. \emph{First Language}, \emph{36}(3), 338--351.
-doi:\href{https://doi.org/10.1177/0142723715626067}{10.1177/0142723715626067}
-
-\hypertarget{ref-bergelsoncasillas2019what}{}
-Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
-S., \& Amatuni, A. (2019). What do North American babies hear? A
-large-scale cross-corpus analysis. \emph{Developmental Science},
-\emph{22}(1), e12724.
-doi:\href{https://doi.org/10.1111/desc.12724}{10.1111/desc.12724}
-
-\hypertarget{ref-Praat}{}
-Boersma, P., \& Weenink, D. (2020). Praat: Doing phonetics by computer
-(Version 6.1.35). Retrieved from \url{http://www.praat.org/}
-
-\hypertarget{ref-bowey2001nonword}{}
-Bowey, J. A. (2001). Nonword repetition and young children's receptive
-vocabulary: A longitudinal study. \emph{Applied Psycholinguistics},
-\emph{22}(3), 441--469.
-
-\hypertarget{ref-brandeker2015language}{}
-Brandeker, M., \& Thordardottir, E. (2015). Language exposure in
-bilingual toddlers: Performance on nonword repetition and lexical tasks.
-\emph{American Journal of Speech-Language Pathology}, \emph{24}(2),
-126--138.
-
-\hypertarget{ref-brown2011cultural}{}
-Brown, P. (2011). The cultural organization of attention. In A. Duranti,
-E. Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language
-Socialization} (pp. 29--55). Malden, MA: Wiley-Blackwell.
-
-\hypertarget{ref-brown2014interactional}{}
-Brown, P. (2014). The interactional context of language learning in
-Tzeltal. In I. Arnon, M. Casillas, C. Kurumada, \& B. Estigarribia
-(Eds.), \emph{Language in interaction: Studies in honor of Eve V. Clark}
-(pp. 51--82). Amsterdam, NL: John Benjamins.
-
-\hypertarget{ref-brownIPchildrearing}{}
-Brown, P., \& Casillas, M. (in press). Childrearing through social
-interaction on Rossel Island, PNG. In A. J. Fentiman \& M. Goody (Eds.),
-\emph{Esther Goody revisited: Exploring the legacy of an original
-inter-disciplinarian} (pp. XX--XX). New York, NY: Berghahn.
-
-\hypertarget{ref-bunceURcrosscultural}{}
-Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A.,
-Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural
-examination of young children's everyday language experiences.
-
-\hypertarget{ref-casillas2020early}{}
-Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language
-experience in a papuan community. \emph{Journal of Child Language},
-\emph{XX}, XX--XX.
-
-\hypertarget{ref-chiat2007preschool}{}
-Chiat, S., \& Roy, P. (2007). The preschool repetition test: An
-evaluation of performance in typically developing and clinically
-referred children. \emph{Journal of Speech, Language, and Hearing
-Research}, \emph{50}(2), 429--443.
-
-\hypertarget{ref-cristia2020infant}{}
-Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
-(2020). Infant-directed input and literacy effects on phonological
-processing: Non-word repetition scores among the tsimane'.
-\emph{Preprint}.
-
-\hypertarget{ref-estes2007differences}{}
-Estes, K. G., Evans, J. L., \& Else-Quest, N. M. (2007). Differences in
-the nonword repetition performance of children with and without specific
-language impairment: A meta-analysis. \emph{Journal of Speech, Language,
-and Hearing Research}, \emph{50}(1), 177--195.
-
-\hypertarget{ref-farmani2018normalization}{}
-Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \&
-Shourvazi, Z. (2018). Normalization of the non-word repetition test in
-farsi-speaking children. \emph{Journal of Modern Rehabilitation},
-\emph{12}(4), 217--224.
-
-\hypertarget{ref-foley1986papuan}{}
-Foley, W. A. (1986). \emph{The papuan languages of new guinea}.
-Cambridge, UK: Cambridge University Press.
-
-\hypertarget{ref-gallagher2014identity}{}
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from
-Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378.
-doi:\href{https://doi.org/10.1515/lp-2014-0012}{10.1515/lp-2014-0012}
-
-\hypertarget{ref-gallon2007non}{}
-Gallon, N., Harris, J., \& Van der Lely, H. (2007). Non-word repetition:
-An investigation of phonological complexity in children with grammatical
-sli. \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
-
-\hypertarget{ref-havron2019effect}{}
-Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \&
-Group, E. M.-C. C. S. (2019). The effect of older siblings on language
-development as a function of age difference and sex. \emph{Psychological
-Science}, \emph{30}(9), 1333--1343.
-
-\hypertarget{ref-hazan2000development}{}
-Hazan, V., \& Barrett, S. (2000). The development of phonemic
-categorization in children aged 6--12. \emph{Journal of Phonetics},
-\emph{28}(4), 377--396.
-
-\hypertarget{ref-jabere2018xperiment}{}
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual
-Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4),
-317--334.
-doi:\href{https://doi.org/10.30958/ajp.5-4-4}{10.30958/ajp.5-4-4}
-
-\hypertarget{ref-kalnak2014nonword}{}
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014).
-Nonword repetition--a clinical marker for specific language impairment
-in swedish associated with parents' language-related problems.
-\emph{PloS One}, \emph{9}(2), e89544.
-
-\hypertarget{ref-ladefoged1996sounds}{}
-Ladefoged, P., \& Maddieson, I. (1996). \emph{The sounds of the world's
-languages} (Vol. 1012). Blackwell Oxford.
-
-\hypertarget{ref-levinsonYDgrammar}{}
-Levinson, S. C. (accepted). \emph{A grammar of yélî dnye, the papuan
-language of rossel island}. Berlin, Boston: De Gruyter Mouton.
-
-\hypertarget{ref-liszkowski2012prelinguistic}{}
-Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
-(2012). A prelinguistic gestural universal of human communication.
-\emph{Cognitive Science}, \emph{36}(4), 698--713.
-doi:\href{https://doi.org/10.1111/j.1551-6709.2011.01228.x}{10.1111/j.1551-6709.2011.01228.x}
-
-\hypertarget{ref-maddieson2005correlating}{}
-Maddieson, I. (2005). Correlating phonological complexity: Data and
-validation. \emph{UC Berkeley PhonLab Annual Report}, \emph{1}(1).
-
-\hypertarget{ref-maddiesonIPphoneticsYD}{}
-Maddieson, I., \& Levinson, S. C. (in preparation). \emph{The phonetics
-of yélî dnye, the language of rossel island}.
-
-\hypertarget{ref-meir2017independent}{}
-Meir, N., \& Armon-Lotem, S. (2017). Independent and combined effects of
-socioeconomic status (ses) and bilingualism on children's vocabulary and
-verbal short-term memory. \emph{Frontiers in Psychology}, \emph{8},
-1442.
-
-\hypertarget{ref-meir2016disentangling}{}
-Meir, N., Walters, J., \& Armon-Lotem, S. (2016). Disentangling sli and
-bilingualism using sentence repetition tasks: The impact of l1 and l2
-properties. \emph{International Journal of Bilingualism}, \emph{20}(4),
-421--452.
-
-\hypertarget{ref-phoible}{}
-Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max
-Planck Institute for the Science of Human History. Retrieved from
-\url{https://phoible.org/}
-
-\hypertarget{ref-peuteIPconsonants}{}
-Peute, A. A. K., Fikkert, P., \& Casillas, M. (In preparation). Early
-consonant production in Yélî Dnye and Tseltal.
-
-\hypertarget{ref-piazzalunga2019articulatory}{}
-Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler,
-A. (2019). An articulatory-based disyllabic and trisyllabic non-word
-repetition test: Reliability and validity in italian 3-to 7-year-old
-children. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5),
-437--456.
-
-\hypertarget{ref-torrington2015non}{}
-Torrington Eaton, C., Newman, R. S., Ratner, N. B., \& Rowe, M. L.
-(2015). Non-word repetition in 2-year-olds: Replication of an adapted
-paradigm and a useful methodological extension. \emph{Clinical
-Linguistics \& Phonetics}, \emph{29}(7), 523--535.
-
-\hypertarget{ref-vance2005speech}{}
-Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills
-in children aged 3--7 years. \emph{International Journal of Language \&
-Communication Disorders}, \emph{40}(1), 29--48.
-
-\hypertarget{ref-wilsenach2013phonological}{}
-Wilsenach, C. (2013). Phonological skills as predictor of reading
-success: An investigation of emergent bilingual Northern Sotho/English
-learners. \emph{Per Linguam: a Journal of Language Learning= Per
-Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
-doi:\href{https://doi.org/10.5785/29-2-554}{10.5785/29-2-554}
+\url{https://doi.org/10.1111/j.1551-6709.2011.01228.x}
+
+\leavevmode\hypertarget{ref-maddieson2005correlating}{}%
+Maddieson, I. (2005). Correlating phonological complexity: Data and validation. \emph{UC Berkeley PhonLab Annual Report}, \emph{1}(1).
+
+\leavevmode\hypertarget{ref-maddiesonIPphoneticsYD}{}%
+Maddieson, I., \& Levinson, S. C. (n.d.). \emph{The phonetics of yélî dnye, the language of rossel island}.
+
+\leavevmode\hypertarget{ref-meir2017independent}{}%
+Meir, N., \& Armon-Lotem, S.
+(2017).
+Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
+\emph{Frontiers in Psychology}, \emph{8}, 1442.
+
+\leavevmode\hypertarget{ref-meir2016disentangling}{}%
+Meir, N., Walters, J., \& Armon-Lotem, S.
+(2016).
+Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
+\emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
+
+\leavevmode\hypertarget{ref-phoible}{}%
+Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
+
+\leavevmode\hypertarget{ref-peuteIPconsonants}{}%
+Peute, A. A. K., Fikkert, P., \& Casillas, M. (n.d.). Early consonant production in Yélî Dnye and Tseltal.
+
+\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
+Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
+(2019).
+An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
+\emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
+
+\leavevmode\hypertarget{ref-torrington2015non}{}%
+Torrington Eaton, C., Newman, R.
+S., Ratner, N.
+B., \& Rowe, M.
+L.
+(2015).
+Non-word repetition in 2-year-olds: Replication of an adapted paradigm and a useful methodological extension.
+\emph{Clinical Linguistics \& Phonetics}, \emph{29}(7), 523--535.
+
+\leavevmode\hypertarget{ref-vance2005speech}{}%
+Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
+
+\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
+Wilsenach, C.
+(2013).
+Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
+\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+\url{https://doi.org/10.5785/29-2-554}
 

### 2020-12-22 manuscript.Rmd
cleaned up and reknitted

@@
 library(lme4)
+library(openxlsx)
 
@@
 
-<!-- REMEMBER!!!! -->
-<!-- phonetic characters: -->
-<!-- - save excel as utf with save as -->
-<!-- - do $ iconv -f utf-16le -t utf-8 stimuli.txt > stimconv.txt -->
-<!-- - do $ iconv -f utf-16le -t utf-8 segments.txt > segconv.txt -->
-
-## TODO Alex
-
-- check again if all fields are explained
-- set up OSF (or revise if already there)
-- implement decisions on prop phones & patterns nwr mp
-<!--Added by MC on 21 Dec 2020-->
-- check and accept/deny MC's updated segment frequencies
-- re-run analyses w/ new frequencies and check that outcomes didn't qualitatively change
-- check YD data in figure 1 (the effect of item length looks reversed)
-- see MC comments throughout and delete now-irrelevant things
-
-## TODO Middy
-
-<!--
-- double check my frequency entries in segments.xlsx using the search function in https://phoible.org/parameters; take a look at http://phoible.github.io/conventions/ in case you see something about double articulation (I didn't find tp or lBj)
-
-M2A: Done! Left for AC to review in segments-20201123.xlsx and, if she is satisfied, to integrate :)
--->
-<!--
-- draft discussion
-
-M2A: Done! I left some space for you to fill in references in a few spots since it seemed from the notes you left that you had some specific ones in mind.
--->
-
 ## Introduction
@@
 #add orthographic representations targets & other stims chars
-stims.tbl <- read_delim("stimuli3.txt", delim = "\t") %>%
+stims.tbl <- read.xlsx("stimuli.xlsx") %>%
   arrange(factor(type, levels = c("practice", "1", "2", "3", "4"))) %>%
@@
 ```{r demo}
+# please note that you will not be able to do this because you don't have this version, that is not fully de-identified
 # read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
@@
 
-#lines no longer needed
-#if family is missing, attribute each child to a different one
-#demo$familyID=as.character(demo$familyID)
-#demo$familyID[is.na(demo$familyID)]<-paste("k",demo$ID[is.na(demo$familyID)])
-#recalc age
-#demo$age.rounded=as.numeric(difftime(as.Date(as.character(demo$Date),format="%Y%m%d"),as.Date(as.character(demo$DOB.ISO),format="%Y%m%d"),unit="weeks"))/52.25
-#write.table(demo,"nwr-demo.csv",row.names=F,quote=T,sep=",")
-
 subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
@@
 colnames(xling)<-c("Target","Avg_fr","Word")
-stims.ipa <- read_delim("stimuli3.txt", delim = "\t") %>%
+stims.ipa <- read.xlsx("stimuli.xlsx") %>%
   select(target, target_ortho, phono)

### 2020-12-22 wrangling.R

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(openxlsx)
 
 read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
 
@@ -91,24 +92,13 @@ merge(trnsc,inc,by.x="id",by.y="ID",all.x=T)->trnsc
 dim(trnsc) #still no loss, all good
 
 #add orthographic representations targets & other stims chars
-# M2A: Note that some of these stim characters are now updated in stimuli2.txt
-# they do not match what is shown here (!!)
-# a2m: I'm confused, what do you mean? 
-# my pipeline was: stimuli.txt I couldn't read here, so I put these instructions in the rmd:
-# <!-- REMEMBER!!!! -->
-#   <!-- phonetic characters: -->
-#   <!-- - save excel as utf with save as -->
-#   <!-- - do $ iconv -f utf-16le -t utf-8 stimuli.txt > stimconv.txt -->
-#   <!-- - do $ iconv -f utf-16le -t utf-8 segments.txt > segconv.txt -->
 #   
 # also, notice that stims and segments are merged using ortho, so as long as ortho is matched
 # across the two, the merge should carry over whatever changes you make
-# M2A: Some of the segments in stimuli.txt were wrong. I made corrections in stimuli2.txt,
-# but now that is mismatched to stimconv and segconv.
-# M note: resolved over quick videocall; a updated the stim characters throughout now (resolved)
+
 
 #get info on segments
-read.delim("segconv.txt", encoding="UTF-8")->segments
+read.xlsx("segments.xlsx")->segments
 #add corpus freq to segments
 read.delim("segment-counts.txt", encoding="UTF-8",sep="\t",header=T)->phone_counts
 merge(segments,phone_counts,by.x="ortho",by.y="ortho",all.x=T)->segments
@@ -118,7 +108,7 @@ merge(segments,phone_counts_types,by.x="ortho",by.y="ortho",all.x=T)->segments
 write.table(segments,"segments_with_cor_freq.txt",sep="\t",row.names = F)
 
 #read in table with breakdown 
-read.delim("stimuli3.txt", encoding="UTF-8")->stims
+read.xlsx("stimuli.xlsx")->stims
 #add average crosslinguistic frequency
 stim_seg_freq=matrix(NA,nrow=dim(stims)[1],ncol=8)
 for(i in 1:dim(stim_seg_freq)[1]) for(j in 1:dim(stim_seg_freq)[2]) {

### 2020-12-22 manuscript.Rmd
knits correctly, still in progress

@@
 
-authornote: |
-  All data are made available in a repository in the Open Science Framework.
-AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
-S.
-McDonnell Foundation Understanding Human Cognition Scholar Award.
-
 abstract: >
@@
 keywords          : "phonology, non-word repetition, development"
-wordcount         : "~xxx words"
+wordcount         : "~9,000 words"
 
@@
 Further, the languages it has been used with have historically had small to moderate phonological inventories [@maddieson2005correlating].
+
 <!--M2A: Is the motivating idea here that application to YD is of interest because it's very different from what we've look at with NWR in the past?
@@
 In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
-The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from the online supplementary materials.
+More information is available from the online supplementary materials.
 Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
-<!--M2A:
-We took out the actual hamlet names, right? So what do you want to put in the metadata mentioned in the sentence:
-"The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from the online supplementary materials."
--->
 
@@
 It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database.
-Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat our "missing" phones as having a frequency of 1, with a (rounded) percentage of 0%.
+Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat these unattested phones as having a frequency of 1, with a (rounded) percentage of 0%.
 
@@
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^2] As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^2] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
 Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, p `r pval.display(round(freq_cor$p.value,3))`).
@@
 Error"]`, p `r pval.display(both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"])`) was not.
-
-<!--M2A: Above it says "As before, unattested sounds were not considered" -- well now we have taken care of PHOIBLE unattested sounds with a frequency of 1. Do we need to change this?-->
 
@@
 
-<!--M2A: castro1998 is missing from bib. is it this one?:
-https://academic.oup.com/brain/article/121/6/1053/280348
--->
-
 ### NWR scores as a function of item length
@@
 
-<!--
-4. General prediction: Non-monolingual Yélî children have lower scores than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language)
-
-M2A: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)
-
-ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)
-
-M2A: I'll take your word on the mixed evidence!
-I'm certainly not up to date in this domain.
-We don't really have enough kids (non-monoling = 6 kids) to investigate this anyway, which I think is also why you ultimately don't include it in the analyses of individual variation in NWR score.
-I'm therefore not going to include it in the discussion draft.
-
----
-
-4. Local prediction: Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family
-
-M2A: definitely almost all kin; 5k of them
-
-ac2mc I'm not sure this is how bio anthropologists think about kin; because it's about how you attribute resources to boost your genes' chances of being passed on relative to those of others in the same space.
-So my guess is that there is still a difference between nuclear family and not.
-Rewrote kin -> nuclear family
-
-M2A: I see what you mean!
-I wouldn't, however, directly apply this broad theory about genetic transmission to understanding the specific Rossel context.
-I'm really very unconvinced that there is the same kind of hard division between nuclear and non-nuclear family on Rossel wrt daily resources, caregiving, and even ceremonial and some land rights things.
-I think the real question being raised here is about the specific relevance the distribution of linguistic input over family members has for kids' language development.
-We're really too far from being able to groundedly talk about that for my comfort.
-
-(or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
-
-M2A: maybe it'd be more useful to try and estimate the number of non-Y moms who live in each hamlet where testing is taking place?)
-
-ac2mc not sure it's worth the effort, in any case other female adults are almost certain to have very little impact on the target child... Perhaps I'm not saying something very clearly?
-
-M2A: In my experience most of the non-Rossel moms don't really speak Rossel to their kids at all (though they use some phrases and have some receptive knowledge)---there are exceptions of course!
-I don't think what you're saying is unclear, I'm just trying to wrap my head around what is being implied in some of the text and also doing a self-check for whether each of these implications is really meaningfully relevant in the Rossel context.
-Sorry, that probably feels like turning us in circles a bit!
-
-. That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother
-
-M2A: i can get you actual estimates for this for YD; let's go with that instead!
-
-ac2mc I think I meant your own paper! added "on the same population "
-
-M2A: Heh yeah I know!
-I mean that, buried in my annotations, is also information about how often the TCDS in Casillas et al.
-(2020) comes from the mother vs.
-others.
-Right now the paper doesn't make this claim, it only talks about speech from women.
-So we could in theory extract actual input estimates from the mother.
-But I think that's moot since none of this is really pertinent to the discussion anymore.
-
-, so in so far as this input has a crucial formational role, then there may still be a score difference between these two groups.
-
-4. We don't see a difference
-
-M2A: Not currently tested as far as I can see (?)
-
--->
-
 \newpage
@@
 We are grateful to the individuals who participated in the study.
-The collection and annotation of these recordings was made possible by Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also to the PNG National Research Institute, and the Administration of Milne Bay Province AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.
+The collection and annotation of these recordings was made possible by Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also to the PNG National Research Institute, and the Administration of Milne Bay Province.
+AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.
 S.

### 2020-12-23 manuscript.Rmd
pass in progress, knits

@@
 abstract: >
-   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language.
-NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
-In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea that has an unusually dense phonological inventory.
-This study contributes to three lines of research.
-First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds.
-Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
-Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
-Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between patterns in language development and in the frequency of linguistic features in human language at large.
+   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible.
+NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
+In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
+This study contributes to four lines of research.
+First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
+Second, we document rather weak effects of item length, contributing to mounting research suggesting that length effects may be language- or population-specific.
+Third, we do not find strong individual variation effects in this population.
+Fourth, we discuss how NWR scores can be compared across studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development.
+Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.
 
@@
 Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development].
-Much of the evidence for later phonological development comes through the use of nonword repetition (NWR) tasks.
+Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks.
 In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
@@
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language].
-However, it has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant].
-Further, the languages it has been used with have historically had small to moderate phonological inventories [@maddieson2005correlating].
 
-<!--M2A: Is the motivating idea here that application to YD is of interest because it's very different from what we've look at with NWR in the past?
-Or is there a specific hypothesis you had in mind about how inventory size affects phonological development and, hence, NWR performance?
-I could come up with something, but I think you had an idea already and I don't want to step on it!-->
+<!-- a2m No, I didn't propose that we would use NWR because I wanted to contribute to our understanding of the NWR task in new settings, but rather because it's quite a standard task so it could be useful for us documenting phonological development in Rossel.
+To make this clearer, I moved it down and integrated it with the four contributions below -->
 
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
-The study was designed to contribute to multiple aspects of our understanding of phonological development: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
-Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation; previous NWR research has shown variable effects of length between populations.
-Third, these data give a first impression of how this task is responded to in a rarely-studied language and cultural setting; we add to this contribution by further investigating whether there are structured sources that account for individual variation in our sample.
-<!--M2A: This is also a note to myself; it seems we could use some more references in the above!-->
+The study was designed to contribute to four aspects of our understanding of phonological development.
 
-Before jumping into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-Since NWR has been almost exclusively used in urban, industrialized populations [see @cristia2020infant for an exception], we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.
+First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR.
+Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], likely as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non].
+Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
+
+Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development.
+Our reading of previous NWR research is that there are variable effects of length between populations.
+For instance, @jabere2018xperiment reports an average of 95% correct repetition for items 1 or 2 syllables long among Arabic-speaking children at about 5 years of age, but 55% for items 3 or 4 syllables long.
+In contrast, @piazzalunga2019articulatory observe nearly no change in performance in similarly-aged Italian learners, with a score of 84% for 1-2 syllables versus 85% for 3-4 syllables.
+It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture.
+That said, there could be other causal pathways: Research on adults suggests that illiterate Portuguese speakers repeat monosyllabic non-words just as accurately as literate adults, whereas scores are much lower among illiterate than literate speakers for items 3 or 4 syllables long [@de2004degree].
+Given that both groups of adults speak the same language (Brazilian Portuguese), then perhaps differences in repetition accuracy reveal differences in how flexible the perception-production loop is.
+Given the paucity of evidence looking at this question, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
+
+Third, there are ongoing discussions as to what the key factors structuring individual variation are.
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review on the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword].
+In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
+For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
+Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]<!--, and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]-->.
+Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
+Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.
+
+<!-- a2m do you remember: why did we remove bilingual exposure? -->
+
+Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
+Indeed, NWR has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], outside urban settings [except for in @cristia2020infant], nor with languages having large phonological inventories.
+There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
+That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating].
+Therefore, it may not be a general index of phonological development, but more specifically reflect certain skills.
+Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic representations.
+
+Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.
 
@@
 
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar] and includes at least two contrasts not yet been documented elsewhere [labial-coronal double-articulations with dental vs.
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar] and includes at least two contrasts that had not previously been documented elsewhere [labial-coronal double-articulations with dental vs.
 post-alveolar coronal placement for both oral and nasal stops, @ladefoged1996sounds; @maddiesonIPphoneticsYD].
 
-Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at time of writing; @levinsonYDgrammar].
+Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at the time of writing; @levinsonYDgrammar].
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., 'ndap' (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., 'ndapî' /'ṇṭæ.pɯ/).
-This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /'pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-.
+A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" (Spondylus shell) /ṇṭæp/).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., "ndap" pronounced "ndapî" /'ṇṭæ.pɯ/).
+This process is used frequently with English loan words that have a coda (e.g., "ponî" (phone) /'pɔ.nɯ/).
+An even smaller portion of the lexicon features words starting with a vowel (e.g., "ala" (here) /æ.'læ/), but these are limited to /æ/-.
 Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
@@
 
-Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English.
+Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
 While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
-Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or another language(s) from the region.
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
+
+<!-- a2m I thought in the first couple of years they combined English and Yélî at school? -->
 
@@
 Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
-There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available.
+There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available.
 There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG.
 There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time.
-As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, is typically unavailable.
+As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, are typically unavailable.
 Our data collection is therefore limited to the duration of the researchers' on-island visits.
@@
 While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day).
-It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year.
+It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segments of each school year.
 Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
@@
 
+<!-- a2m why do we explain this: Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
+-->
+
 While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., @peuteIPconsonants].
 What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early].
-While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early].
-This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+While speech from women predominates in young children's language environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early].
+The latter increase is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+
+<!-- a2m including Bunce here may be ambiguous, and it could be read to indicate that speech from other children increases in other cultures as well: there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early] -->
 
@@
 For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language].
-Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items [@piazzalunga2019articulatory].
+Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory].
 Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
 
-Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology.
+Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
 We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.
@@
 Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
-However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit.
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
-
-Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]<!--, and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]-->.
-Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or child gender [@chiat2007preschool].
-Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.
+However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
 
@@
 - Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
-- How does score change as a function of item length in number of syllables?
+- How do NWR scores change as a function of item length in number of syllables?
 - Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?
 
-In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
+We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
 However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies.
 To illustrate this issue, we portray in Figure 1 studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
-Notice that the effect of stimulus length is minuscule among Italian children [@piazzalunga2019articulatory], but considerable among Tsimane' children  [@cristia2020infant], where a drop of 40 percentage points is observed at all ages.
-A similar large difference in NWR scores for short versus long items was observed among Arabic children [@jabere2018xperiment].
+As discussed above,  the effect of stimulus length is minuscule among Italian children [@piazzalunga2019articulatory], but considerable among Arabic children [@jabere2018xperiment].
 Even the effect of age is unstable in this sample of studies.
-Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
+Whereas it is quite clear that children's NWR scores increase in the Italian data [@piazzalunga2019articulatory], age effects are less stable among Tsimane' children [@cristia2020infant].
 Therefore, all analyses in the present study are descriptive and should be considered exploratory.
@@
 ```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies.
-Arabic data  from Jaber et al.
-(2018); Italian data from Piazzalunga et al.
+Arabic data  from Jaber-Awida (2018); Italian data from Piazzalunga et al.
 (2019); Tsimane' from Cristia et al.
@@
 
-A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms.
+A second list of candidate items was generated in a second trip to the island in 2019, when data were collected, by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms.
 As before, items were automatically excluded if they appeared in the dictionary.
@@
   "latex", booktabs = T,
-  caption="NWR stimuli in orthographic and phonological representations.",
+  caption="NWR stimuli in orthographic (Orth.) and phonological (Phon.) representations.",
   row.names =F) %>%
@@
 A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
-The 40 elicitation sets are available online from [osf.io/5qspb](https://osf.io/5qspb).
+The 40 elicitation sets are available online from [osf.io/dtxue/](https://osf.io/dtxue/).
 The split had the following constraints:
@@
 # write.csv(fulldemo,"nwr-demo.csv")
+
+# read.csv("nwr-demo.csv",header=T)->demo
+# demo=demo[,-c(20,21)] #remove DOBs 
+# write.csv(demo,"nwr-demo.csv")
 
@@
 In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%.
-This type of analysis is seldom reported---we could only find one comparison point: @castro1998 found that illiterate Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
+This type of analysis is seldom reported---we could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
 The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.

### 2020-12-23 manuscript.tex
pass in progress, knits

@@
 \textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
-\authornote{All data are made available in a repository in the Open Science Framework.
-AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J.
-S.
-McDonnell Foundation Understanding Human Cognition Scholar Award.
+\authornote{
 
 Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
-\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
+\keywords{phonology, non-word repetition, development\newline\indent Word count: ~9,000 words}
 \usepackage{lineno}
@@
 \abstract{
-In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language.
-NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns.
-In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea that has an unusually dense phonological inventory.
-This study contributes to three lines of research.
-First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds.
-Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific.
-Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects.
-Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between patterns in language development and in the frequency of linguistic features in human language at large.
+In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible.
+NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
+In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
+This study contributes to four lines of research.
+First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
+Second, we document rather weak effects of item length, contributing to mounting research suggesting that length effects may be language- or population-specific.
+Third, we do not find strong individual variation effects in this population.
+Fourth, we discuss how NWR scores can be compared across studies, languages, and populations, in order to shed light on the factors universally structuring variation in language development.
+Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.
 }
@@
 
-\hypertarget{todo-alex}{%
-\subsection{TODO Alex}\label{todo-alex}}
-
-\begin{itemize}
-\tightlist
-\item
-  check whether it's stimuli, stimuli2, stimuli3 that should be used!!!
-\item
-  create fields final-data DONE
-\item
-  leave only files actually needed DONE
-\item
-  add readme explaining files DONE
-\item
-  remove all files that aren't necessary DONE
-\item
-  get rid of all the conversions and directly read in xls
-\item
-  set up OSF (or revise if already there)
-\item
-  implement decisions on prop phones \& patterns nwr mp
-\item
-  check and accept/deny MC's updated segment frequencies
-\item
-  re-run analyses w/ new frequencies and check that outcomes didn't qualitatively change
-\item
-  check YD data in figure 1 (the effect of item length looks reversed)
-\item
-  see MC comments throughout and delete now-irrelevant things
-\end{itemize}
-
-\hypertarget{todo-middy}{%
-\subsection{TODO Middy}\label{todo-middy}}
-
 \hypertarget{introduction}{%
@@
 Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000).
-Much of the evidence for later phonological development comes through the use of nonword repetition (NWR) tasks.
+Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks.
 In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
@@
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-However, it has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014, and @cristia2020infant).
-Further, the languages it has been used with have historically had small to moderate phonological inventories (Maddieson, 2005).
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
 
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
-The study was designed to contribute to multiple aspects of our understanding of phonological development: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation.
-Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation; previous NWR research has shown variable effects of length between populations.
-Third, these data give a first impression of how this task is responded to in a rarely-studied language and cultural setting; we add to this contribution by further investigating whether there are structured sources that account for individual variation in our sample.
+The study was designed to contribute to four aspects of our understanding of phonological development.
 
-Before jumping into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-Since NWR has been almost exclusively used in urban, industrialized populations (see Cristia et al., 2020 for an exception), we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.
+First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR.
+Previous work using NWR has preferred relatively universal and early-acquired phonemes (with the possible exception of Gallagher, 2014), likely as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007).
+Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
+
+Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development.
+Our reading of previous NWR research is that there are variable effects of length between populations.
+For instance, Jaber-Awida (2018) reports an average of 95\% correct repetition for items 1 or 2 syllables long among Arabic-speaking children at about 5 years of age, but 55\% for items 3 or 4 syllables long.
+In contrast, Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019) observe nearly no change in performance in similarly-aged Italian learners, with a score of 84\% for 1-2 syllables versus 85\% for 3-4 syllables.
+It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture.
+That said, there could be other causal pathways: Research on adults suggests that illiterate Portuguese speakers repeat monosyllabic non-words just as accurately as literate adults, whereas scores are much lower among illiterate than literate speakers for items 3 or 4 syllables long (de Santos Loureiro et al., 2004).
+Given that both groups of adults speak the same language (Brazilian Portuguese), then perhaps differences in repetition accuracy reveal differences in how flexible the perception-production loop is.
+Given the paucity of evidence looking at this question, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
+
+Third, there are ongoing discussions as to what the key factors structuring individual variation are.
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review on the literature looking at correlations between NWR scores and a variety of child-level variables (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
+In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
+For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
+
+Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
+Indeed, NWR has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014; Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020), outside urban settings (except for in Cristia et al., 2020), nor with languages having large phonological inventories.
+There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
+That said, Cristia et al.
+(2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991).
+Therefore, it may not be a general index of phonological development, but more specifically reflect certain skills.
+Indeed, Cristia et al.
+(2020) present the task as being a good index of the development of \enquote{short-hand-like} representations specifically, which could thus miss, for example, more holistic representations.
+
+Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.
 
@@
 Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
-The Yélî sound system, much like its baroque grammatical system (Levinson, n.d.), is unlike any other in the region.
+The Yélî sound system, much like its baroque grammatical system (Levinson, 2020), is unlike any other in the region.
 
@@
 The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (n.d.) for details).\footnote{We use Levinson's (n.d.) under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (2020) for details).\footnote{We use Levinson's (2020) under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
 In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
@@
 
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, n.d.) and includes at least two contrasts not yet been documented elsewhere (labial-coronal double-articulations with dental vs.~post-alveolar coronal placement for both oral and nasal stops, Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, n.d.).
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020) and includes at least two contrasts that had not previously been documented elsewhere (labial-coronal double-articulations with dental vs.~post-alveolar coronal placement for both oral and nasal stops, Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, n.d.).
 
-Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent Yélî Dnye dictionary at time of writing; Levinson, n.d.).
+Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent Yélî Dnye dictionary at the time of writing; Levinson, 2020).
 The vast majority of syllables use a CV format.
 A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., \enquote{ndapî} /\enquote{ṇṭæ.pɯ/).
-This process is used frequently with English loan words that have a coda (e.g., 'ponî} (phone) /\enquote{pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala} (here) /æ.'læ/), but these are limited to /æ/-.
-Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (n.d.) for details).
+However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., \enquote{ndap} pronounced \enquote{ndapî} /'ṇṭæ.pɯ/).
+This process is used frequently with English loan words that have a coda (e.g., \enquote{ponî} (phone) /'pɔ.nɯ/).
+An even smaller portion of the lexicon features words starting with a vowel (e.g., \enquote{ala} (here) /æ.'læ/), but these are limited to /æ/-.
+Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (2020) for details).
 
@@
 For example, while items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation (Brandeker \& Thordardottir, 2015).
-Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
-Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).
+Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga et al., 2019).
+Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon et al., 2007).
 
@@
 
-In view of the hypothesis-driven nature of this work, we had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
+We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
 However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies.
@@
 Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
-Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, n.d.), it was also possible that we initially generated items with illegal, but currently undocumented constraints.
+Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, 2020), it was also possible that we initially generated items with illegal, but currently undocumented constraints.
 Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
@@
 In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
-The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from the online supplementary materials.
+More information is available from the online supplementary materials.
 Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
@@
 It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database.
-Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat our \enquote{missing} phones as having a frequency of 1, with a (rounded) percentage of 0\%.
+Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat these unattested phones as having a frequency of 1, with a (rounded) percentage of 0\%.
 
@@
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.} As before, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
 Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(27)=0.50, p \textless{} 0.01).
@@
 In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: 63\%.
-This type of analysis is seldom reported---we could only find one comparison point: ({\textbf{???}}) found that illiterate Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
+This type of analysis is seldom reported---we could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
 The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-is08042009language}{}%
-Action, C. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
-
 \leavevmode\hypertarget{ref-balladares2016socio}{}%
@@
 
+\leavevmode\hypertarget{ref-castro1998illiterate}{}%
+Castro-Caldas, A., Petersson, K.
+M., Reis, A., Stone-Elander, S., \& Ingvar, M.
+(1998).
+The illiterate brain.
+Learning to read and write during childhood influences the functional organization of the adult brain.
+\emph{Brain: A Journal of Neurology}, \emph{121}(6), 1053--1063.
+\url{https://doi.org/10.1093/brain/121.6.1053}
+
 \leavevmode\hypertarget{ref-chiat2007preschool}{}%
@@
 
+\leavevmode\hypertarget{ref-is08042009language}{}%
+COST Action. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+
 \leavevmode\hypertarget{ref-cristia2020infant}{}%
@@
 (2020).
-Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the tsimane'.
-\emph{Preprint}.
+Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the Tsimane'.
+\emph{PLoS ONE}, \emph{15}(9), e0237702.
+\url{https://doi.org/https://doi.org/10.1371/journal.pone.0237702}
+
+\leavevmode\hypertarget{ref-de2004degree}{}%
+de Santos Loureiro, C., Braga, L.
+W., Nascimento Souza, L.
+do, Nunes Filho, G., Queiroz, E., \& Dellatolas, G.
+(2004).
+Degree of illiteracy and phonological and metaphonological skills in unschooled adults.
+\emph{Brain and Language}, \emph{89}(3), 499--502.
+\url{https://doi.org/10.1016/j.bandl.2003.12.008}
 
@@
 \emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
+
+\leavevmode\hypertarget{ref-farabolini2021nonword}{}%
+Farabolini, G., Rinaldi, P., Caselli, C., \& Cristia, A.
+(2021).
+Non-word repetition in bilingual children: The role of language exposure, vocabulary scores and environmental factors.
+\emph{Speech Language and Hearing}.
 
@@
 \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+
+\leavevmode\hypertarget{ref-gathercole1991differentiating}{}%
+Gathercole, S.
+E., Willis, C., \& Baddeley, A.
+D.
+(1991).
+Differentiating phonological memory and awareness of rhyme: Reading and vocabulary development in children.
+\emph{British Journal of Psychology}, \emph{82}(3), 387--406.
 
@@
 \leavevmode\hypertarget{ref-levinsonYDgrammar}{}%
-Levinson, S. C. (n.d.). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Berlin, Boston: De Gruyter Mouton.
+Levinson, S. C. (2020). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Berlin, Boston: De Gruyter Mouton.
 

### 2020-12-23 manuscript.Rmd
midway through results, knits

@@
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]<!--, and differences relating to bilingual language exposure [@brandeker2015language; @meir2016disentangling; @meir2017independent]-->.
+Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
 Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
 Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.
-
-<!-- a2m do you remember: why did we remove bilingual exposure? -->
 
@@
 Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
+On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].
+Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
 
@@
 23).
-- Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1-, 2-, and 4-syllable), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
+- Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
 
@@
 NWR studies vary in whether children are allowed to hear and/or repeat the item more than one time.
-We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was alowed to make a further attempts if their first attempt was judged erroneous in some way by the assistant.
+We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was alowed to make further attempts if their first attempt was judged erroneous in some way by the assistant.
 The procedure worked as follows: When the child made an attempt, the assistant indicated to the experimenter whether the child's production was correct or not.
@@
 Children were not asked to make repetitions if they did not produce a first attempt.
-In total, test sessions took approximately XX minutes on average, with XX of those minutes attributed to practice and XX to the actual test list.
-<!--M2A: Alex, could you please estimate the above XX cases and fill in?-->
+In total, test sessions took approximately six minutes, with the first minute attributed to practice and five minutes to the actual test list.
 
@@
 
-The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat],
-<!--M2A: Alex, I added a bib entry for Praat but it's the most recent version. Could you please correct it to the version (approximately if not precisely) that you used?-->
-then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files.
+The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat], then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files.
 The assistant then listened through all these productions, one target item at a time, with productions for each target item presented in a random order across children and repetitions.
@@
 
-In addition to a binary word-level exact repetition score we scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer [the target or the attempt; see also @cristia2020infant].
-Although previous work does not use distance metrics, we additionally report these.
-For instance, recall our example of an English target being /bilik/ with an imagined response [bilig].
+Previous work typically reports two scores: a binary word-level exact repetition score, and a phoneme-level score, defined as the number of phonemes that can be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer [the target or the attempt; as in @cristia2020infant].
+Previous work does not use distance metrics, but we report these rather than the phoneme-level scores because they are more informative.
+To illustrate these scores, recall our example of an English target being /bilik/ with an imagined response [bilig].
 We would score this response as follows: at the whole item level this production would receive a score of zero (because the repetition is not exact); at the phoneme level this production would receive a score of 80% (4 out of 5 phonemes repeated exactly); and the phone-based Levenshtein distance for this production is 20% (because 20% of phonemes were substituted or deleted).
-<!--M2A: A concrete scoring example would be a good idea here-->
-<!-- a2m added! actually, I haven't changed anything but I've been hesitant about using levenshtein's distance because it's the complement of phoneme based scoring -- so shall we drop it?  -->
-<!-- M2A: three thoughts on this, which you can take as you like:
-1) our audience might make this determination. segment-match score (phoneme-level accuracy) is probably more straightforward and might therefore be preferred by psychies
-2) LD has the advantage of making D/I/S relevant as a next step in the analysis
-3) we should definitely choose one, I agree, and report the other in supp mat tables-->
-
-<!--M2A: So what's the final determination here, keeping in mind we're now aiming for LDR?-->
+Notice that the phone-based Levenshtein distance is the complement of the phoneme-level NWR score.
+An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (additions were not attested).
 
@@
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at time of writing, it is of course still far from complete, so estimated frequencies should be taken with a pinch of salt.
-In both PHOIBLE and UPSID---another large phonemic inventory database---note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication; see also [web.phonetik.uni-frankfurt.de/upsid_info.html](http://web.phonetik.uni-frankfurt.de/upsid_info.html)).
-It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database.
-Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat these unattested phones as having a frequency of 1, with a (rounded) percentage of 0%.
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors).
+Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
+Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0% (i.e., its cross-linguistic percentile is zero).
 
 Finally, when describing children's patterns of errors, we take all repetitions of a given target into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.
-<!-- a2m in the results, I put analyses based on all as well as just on first attempt. If we do just the latter, then come back to change this para -->
+We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, insertions, and substitutions.
 
@@
 The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years).
-Among these children were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^1] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth.
 
-[^1]: Years of education is difficult to precisely pinpoint in this context.
-Schooling beyond grade 8 takes place off-island and enrollment is tied to an ability to pay that year's school fees, seasonal interruptions to school opening times, familial obligations, and more.
-We asked parents to report their highest completed level of education.
-We then record the number of years entailed by having completed that level under ideal conditions.
-This method results in many esimates around typical "graduation" times (e.g., graduating from primary education before starting secondary education).
+Included children's ages ranged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).
+We included `r sum(inc$Sex=="B")` boys and `r sum(inc$Sex=="G")` girls.
+There were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^1] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
+These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
+
+[^1]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
 
@@
 were on average lower than first ones (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt=="first"]))`), t(`r round(myt$parameter,2)`) = `r round(myt$statistic,2)`, p `r pval.display(round(myt$p.value,2))`; Cohen's d = `r cohensd2(mean(pwcm_att$Word[pwcm_att$Attempt=="first"]),sd(pwcm_att$Word[pwcm_att$Attempt=="first"]),mean(pwcm_att$Word[pwcm_att$Attempt!="first"]),sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`).
-Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions.
+Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions, with the exception of qualitative analyses of substitutions.
 
@@
 
-Taking into account only the first attempts, we averaged attempts by each of the `r dim(overall)[1]` children who had data for first attempts; their ages ranged from `r min(overall$Age)` to `r max(overall$Age)` (M = `r round(mean(overall$Age),1)`, SD = `r round(sd(overall$Age),1)`).
+Taking into account only the first attempts, we derived overall averages across all items.
 The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`.
-Scores based on phonemes are even higher  M = `r round(mean(overall$Phoneme))`% (SD = `r round(sd(overall$Phoneme))`%), Cohen's d = `r cohensd(mean(overall$Phoneme),sd(overall$Phoneme))`.
-The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted (the normalized Levenshtein distance is the complement of phoneme-based scores).
+The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.
 
@@
 
-```{r diffic,echo=F}
+```{r diffic-all-items,echo=F}
 nasalVs=c(":a", ":â" ,":ê", ":i" ,":o")
@@
 
-read.table("substitution_bank_try1.txt",header=T)->sub_bank
+#read.table("substitution_bank_try1.txt",header=T)->sub_bank
+read.table("substitution_bank.txt",header=T)->sub_bank
 colnames(sub_bank)<-c("target","realization")
@@
 
-read.table("deletion_bank_try1.txt",header=T)->del_bank
+#read.table("deletion_bank_try1.txt",header=T)->del_bank
+read.table("deletion_bank.txt",header=T)->del_bank
 colnames(del_bank)<-c("target")
@@
 
-read.table("match_bank_try1.txt",header=T)->match_bank
+#read.table("match_bank_try1.txt",header=T)->match_bank
+read.table("match_bank.txt",header=T)->match_bank
 colnames(match_bank)<-c("target")
@@
 
-<!--M2A: coming back to these three comments below after I manage another full read-through (after I check demo) -->
-
-<!-- a2m I think it's weird to put effect sizes here, right? technically I could do something like cohen's d based on mean and sd of the percentage of tokens... -->
-
-<!-- **a2m regarding the use first attempts rather than the whole data set, I'm having second thoughts.
-To be clear: the argument for using first attempts only is that all of the other analyses are on kids' first attempts, rather than all of their attempts, so we should use just first attempts here because it may confuse readers & it is more consistent.
-BUT there are fewer mistakes this way, which is Options 1 and 2 below has lots of zeros...
-CF Option 3 further down** -->
-
-<!-- **a2m probably turn the paragraph on substitutions into a table: having second thoughts, the table would look just like a para, no?** -->
-
-<!--
-M2A: I prefer a modification of option 2 and so am commenting this out:
-
-**Option 1: just numbers, first attempts only **
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted, and `r sum(del_tab[c(simplexCs,complexCs)])` consonants.
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral target vowels produced as nasal vowels, `r sum(sub_tab[nasalVs,oralVs])` nasal target vowels produced as oral vowels).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (`r sum(sub_tab[oralVs,oralVs])` oral vowels produced with a different quality; `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels produced with a different quality).
-As for consonants, asymmetries were very marked with more complex consonants (specifically `r paste0(complexCs)`) mispronounced as simple ones (specifically `r paste0(simplexCs)`, `r sum(sub_tab[complexCs,simplexCs])`) than vice versa (`r sum(sub_tab[simplexCs,complexCs])`).
-Simple consonants were mispronounced as other simple consonants quite frequently (`r sum(sub_tab[simplexCs,simplexCs])` simple consonants mispronounced as other simple ones, compared to `r sum(sub_tab[complexCs,complexCs])` complex ones).
-
-**Option 2: numbers and proportions, first attempts only**
--->
-
-Next, we investigated patterns of deletion and substitution.
+Next, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
 Deletions were relatively rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
@@
 Finally, simple consonants were mispronounced as other simple consonants (`r sum(sub_tab[simplexCs,simplexCs])` instances, ~`r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% of all simple consonant targets) and complex consonants as other complex consonants (`r sum(sub_tab[complexCs,complexCs])` instances, ~`r mean(sub_pc[complexCs,complexCs],na.rm=T)`% of all complex consonant targets).
-Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms (see below).
-
-```{r diffic_all_attempts,echo=F}
-
-read.table("substitution_bank.txt",header=T)->sub_bank
-colnames(sub_bank)<-c("target","realization")
-sub_bank$target=factor(sub_bank$target,levels=allVC)
-sub_bank$realization=factor(sub_bank$realization,levels=allVC)
-
-sub_tab=table(sub_bank)[allVC,allVC]
-
-read.table("deletion_bank.txt",header=T)->del_bank
-colnames(del_bank)<-c("target")
-del_bank$target=factor(del_bank$target,levels=allVC)
-del_tab= table(del_bank)[allVC]
-
-read.table("match_bank.txt",header=T)->match_bank
-colnames(match_bank)<-c("target")
-match_bank$target=factor(match_bank$target,levels=allVC)
-match_tab= table(match_bank)[allVC]
-
-#remove English meanings
-trnsc.inc$meaning[grep("English",trnsc.inc$meaning)]<-NA
-
-#Create proportion table
-#a reasonable way to calculate proportion is the proportion of times 
-#a target was produced as X out of the total number of times a target appeared
-
-#so the denominator is not the number of substitutions (which appears in sub_tab)
-#but rather the sum of targets appearing in sub, del, & mat
-#check: colnames(sub_tab) == names(del_tab)
-#check: names(match_tab) == names(del_tab)
-denom_vec=rowSums(sub_tab) + del_tab + match_tab
-
-#all numbers below are percentages (not proportions), and they sum to 100 across sub, del, mat at the TARGET level
-
-#this means: divide each column by corresponding number in denominator vector
-#sub_pc has the same shape as the original sub_tab, namely targets as rows, realization as columns
-sub_pc=round(sub_tab/c(denom_vec)*100,2)
-
-del_pc=round(del_tab/denom_vec*100,2)
-mat_pc=round(match_tab/denom_vec*100,2)
-
-#check that indeed:
-#all numbers ... sum to 100 across sub, del, mat at the TARGET level 
-#NOTE! if you round above, then these don't sum to 100, so don't round very much...
-#rowSums(sub_pc, na.rm=T) + del_pc + mat_pc
-```
-
-<!--
-M2A: I prefer a modification of option 2 and so am commenting this out:
-
-**Option 3: numbers and proportions, ALL attempts**
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare, with only `r sum(del_tab[c(oralVs,nasalVs)])` vowels deleted (about `r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` consonants deleted (about `r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
-
-As for substitutions, it was as common for a nasal vowel to be produced as an oral vowel as vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral target vowels produced as nasal vowels or about `r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% out of all oral vowel targets, `r sum(sub_tab[nasalVs,oralVs])` nasal target vowels produced as oral vowels or about `r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% out of all nasal vowel targets).
-Substitutions in which the oral nature was preserved but the quality of the vowel was changed were a great deal more common than changes in quality among nasal vowels (`r sum(sub_tab[oralVs,oralVs])` oral vowels produced with a different quality or about `r mean(sub_pc[oralVs,oralVs],na.rm=T)`% out of all oral vowel targets; `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels produced with a different quality or about `r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% out of all nasal vowel targets).
-As for consonants, asymmetries were very marked with more complex consonants (specifically `r paste0(complexCs)`) mispronounced as simple ones (specifically `r paste0(simplexCs)`, `r sum(sub_tab[complexCs,simplexCs])` times or  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all complex consonant targets) than vice versa (`r sum(sub_tab[simplexCs,complexCs])` times  about `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% out of all simple consonant targets).
-Simple consonants were mispronounced as other simple consonants quite frequently (`r sum(sub_tab[simplexCs,simplexCs])` simple consonants mispronounced as other simple ones or about `r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% out of all simple consonant targets, compared to `r sum(sub_tab[complexCs,complexCs])` complex ones or about `r mean(sub_pc[complexCs,complexCs],na.rm=T)`% out of all complex consonant targets).
-
--->
+Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.
 
@@
 Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
-Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
+Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998illiterate], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998illiterate] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
 Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.

### 2020-12-23 manuscript.tex
midway through results, knits

@@
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
+Based on previous work, we looked at potential changes with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
+Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on these too.
 
@@
 
-Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English.
+Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
 While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
-Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or another language(s) from the region.
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
 
@@
 While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day).
-It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year.
+It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segments of each school year.
 Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward (Brown \& Casillas, n.d.).\footnote{For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.} Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
+On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward (Brown \& Casillas, n.d.).
+Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
 
@@
 What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day (Casillas et al., 2020), shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children (Bunce et al., n.d.; Casillas et al., 2020).
-While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere (Bergelson et al., 2019; Bunce et al., n.d.), there is a significant and marked increase in child-directed speech from other children as infants get older (Bunce et al., n.d.; Casillas et al., 2020).
-This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
+While speech from women predominates in young children's language environments on Rossel, as it does elsewhere (Bergelson et al., 2019; Bunce et al., n.d.), there is a significant and marked increase in child-directed speech from other children as infants get older (Bunce et al., n.d.; Casillas et al., 2020).
+The latter increase is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
 
@@
 For example, while items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation (Brandeker \& Thordardottir, 2015).
-Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items (Piazzalunga et al., 2019).
+Additionally, while some studies have used 10-15 non-words (e.g., Cristia et al., 2020), others have employed up to 46 unique items (Piazzalunga et al., 2019).
 Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon et al., 2007).
 
-Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology.
+Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
 We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.~less challenging sounds.
@@
 Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
-However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit.
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.
-
-Based on previous work, we looked at potential changes with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
-Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
-Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on these too.
+However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
 
@@
 \item
-  How does score change as a function of item length in number of syllables?
+  How do NWR scores change as a function of item length in number of syllables?
 \item
@@
 To illustrate this issue, we portray in Figure 1 studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
-Notice that the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Tsimane' children (Cristia et al., 2020), where a drop of 40 percentage points is observed at all ages.
-A similar large difference in NWR scores for short versus long items was observed among Arabic children (Jaber-Awida, 2018).
+As discussed above, the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Arabic children (Jaber-Awida, 2018).
 Even the effect of age is unstable in this sample of studies.
-Whereas it is quite clear that children's NWR scores increase in the Italian data, age effects are less stable among Tsimane' children.
+Whereas it is quite clear that children's NWR scores increase in the Italian data (Piazzalunga et al., 2019), age effects are less stable among Tsimane' children (Cristia et al., 2020).
 Therefore, all analyses in the present study are descriptive and should be considered exploratory.
@@
 \caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years) and item length for comparable studies.
-Arabic data from Jaber et al.
-(2018); Italian data from Piazzalunga et al.
+Arabic data from Jaber-Awida (2018); Italian data from Piazzalunga et al.
 (2019); Tsimane' from Cristia et al.
@@
 
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares et al., 2016; Wilsenach, 2013).
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
@@
 
-\caption{\label{tab:tab2-stims}NWR stimuli in orthographic and phonological representations.}
+\caption{\label{tab:tab2-stims}NWR stimuli in orthographic (Orth.) and phonological (Phon.) representations.}
 \centering
@@
 A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
-The 40 elicitation sets are available online from \href{https://osf.io/5qspb}{osf.io/5qspb}.
+The 40 elicitation sets are available online from \href{https://osf.io/dtxue/}{osf.io/dtxue/}.
 The split had the following constraints:
@@
 \item
-  Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1-, 2-, and 4-syllable), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
+  Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
 \end{itemize}
@@
 Children were not asked to make repetitions if they did not produce a first attempt.
-In total, test sessions took approximately XX minutes on average, with XX of those minutes attributed to practice and XX to the actual test list.
+In total, test sessions took approximately six minutes, with the first minute attributed to practice and five minutes to the actual test list.
 
@@
 
-The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software (Boersma \& Weenink, 2020),
-then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files.
+The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software (Boersma \& Weenink, 2020), then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files.
 The assistant then listened through all these productions, one target item at a time, with productions for each target item presented in a random order across children and repetitions.
@@
 
-In addition to a binary word-level exact repetition score we scored items in terms of the number of phonemes that could be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt; see also Cristia et al., 2020).
-Although previous work does not use distance metrics, we additionally report these.
-For instance, recall our example of an English target being /bilik/ with an imagined response {[}bilig{]}.
-We would score this response as follows: at the whole item level this production would receive a score of zero (because the repetition is not exact); at the phoneme level this production would receive a score of 80\% (4 out of 5 phonemes repeated exactly); and the phone-based Levenshtein distance for this production is 20\% (because 20\% of phonemes were substituted or deleted).\\
+Previous work typically reports two scores: a binary word-level exact repetition score, and a phoneme-level score, defined as the number of phonemes that can be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt; as in Cristia et al., 2020).
+Previous work does not use distance metrics, but we report these rather than the phoneme-level scores because they are more informative.
+To illustrate these scores, recall our example of an English target being /bilik/ with an imagined response {[}bilig{]}.
+We would score this response as follows: at the whole item level this production would receive a score of zero (because the repetition is not exact); at the phoneme level this production would receive a score of 80\% (4 out of 5 phonemes repeated exactly); and the phone-based Levenshtein distance for this production is 20\% (because 20\% of phonemes were substituted or deleted).
+Notice that the phone-based Levenshtein distance is the complement of the phoneme-level NWR score.
+An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (additions were not attested).
 
@@
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at time of writing, it is of course still far from complete, so estimated frequencies should be taken with a pinch of salt.
-In both PHOIBLE and UPSID---another large phonemic inventory database---note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication; see also \href{http://web.phonetik.uni-frankfurt.de/upsid_info.html}{web.phonetik.uni-frankfurt.de/upsid\_info.html}).
-It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database.
-Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat these unattested phones as having a frequency of 1, with a (rounded) percentage of 0\%.
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors).
+Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
+Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0\% (i.e., its cross-linguistic percentile is zero).
 
 Finally, when describing children's patterns of errors, we take all repetitions of a given target into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, additions, and substitutions.
+We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, insertions, and substitutions.
 
@@
 The remaining 40 children (14 girls) were aged 6.98 years (range 3.92-11.03 years).
-Among these children were 34 only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.
-Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{Years of education is difficult to precisely pinpoint in this context.
-Schooling beyond grade 8 takes place off-island and enrollment is tied to an ability to pay that year's school fees, seasonal interruptions to school opening times, familial obligations, and more.
-We asked parents to report their highest completed level of education.
-We then record the number of years entailed by having completed that level under ideal conditions.
-This method results in many esimates around typical \enquote{graduation} times (e.g., graduating from primary education before starting secondary education).} In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth.
+
+Included children's ages ranged from 3 to 10 years (M = 6.50 years, SD = 1.50 years).
+We included 26 boys and 14 girls.
+There were 34 only exposed to Yélî Dnye at home, 6 children exposed to Yélî Dnye plus one or more other languages at home.
+Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
+We then record the number of years entailed by having completed that level under ideal conditions.} In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
+These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
 
@@
 were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p \textless{} 0.001; Cohen's d = 1.14).
-Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions.
+Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions, with the exception of qualitative analyses of substitutions.
 
-Taking into account only the first attempts, we averaged attempts by each of the 40 children who had data for first attempts; their ages ranged from 3 to 10 (M = 6.50, SD = 1.50).
+Taking into account only the first attempts, we derived overall averages across all items.
 The overall NWR score was M = 64\% (SD = 15\%), Cohen's d = 4.36.
-Scores based on phonemes are even higher M = 78\% (SD = 9\%), Cohen's d = 8.55.
-The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were were substituted, inserted, or deleted (the normalized Levenshtein distance is the complement of phoneme-based scores).
+The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.
 
@@
 
-Next, we investigated patterns of deletion and substitution.
-Deletions were relatively rare: there were only 7 instances of deleted vowels (\textasciitilde{}0.23\% of all vowel targets), and 3 instances of deleted consonants (\textasciitilde{}0.12\% of all consonant targets).
+Next, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
+Deletions were relatively rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
 Substitutions were somewhat more frequent, but still relatively rare.
-Nasal vowels were produced as oral vowels and vice versa (34 oral targets as nasal (\textasciitilde{}0.99\% of targets) and 30 nasal targets produced as oral (\textasciitilde{}2.37\% of targets)).
-Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (110 oral vowels (\textasciitilde{}1.86\% of targets and 13 nasal vowels (\textasciitilde{}2.06\% of targets).
-Among consonants, complex consonants were occasionally substituted by simpler ones (specifically {[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), 115 times, in approximately 1.96\% of all complex consonant targets.
-Interestingly, simple consonants were sometimes also substituted by complex ones; this happened 2 times, in approximately 1.96\% of all simple consonant targets.
-Finally, simple consonants were mispronounced as other simple consonants (68 instances, \textasciitilde{}0.38\% of all simple consonant targets) and complex consonants as other complex consonants (33 instances, \textasciitilde{}1.02\% of all complex consonant targets).
-Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms (see below).
+Nasal vowels were produced as oral vowels and vice versa (52 oral targets as nasal (\textasciitilde{}1.10\% of targets) and 58 nasal targets produced as oral (\textasciitilde{}2.53\% of targets)).
+Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (205 oral vowels (\textasciitilde{}2.16\% of targets and 23 nasal vowels (\textasciitilde{}2.36\% of targets).
+Among consonants, complex consonants were occasionally substituted by simpler ones (specifically {[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), 250 times, in approximately 2.43\% of all complex consonant targets.
+Interestingly, simple consonants were sometimes also substituted by complex ones; this happened 2 times, in approximately 2.43\% of all simple consonant targets.
+Finally, simple consonants were mispronounced as other simple consonants (120 instances, \textasciitilde{}0.52\% of all simple consonant targets) and complex consonants as other complex consonants (55 instances, \textasciitilde{}0.98\% of all complex consonant targets).
+Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.
 
@@
 Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
-Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work (e.g., {\textbf{???}}), but it is unclear what caused this pattern in the current study: Castro and colleagues' ({\textbf{???}}) study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
+Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work (e.g., Castro-Caldas et al., 1998), but it is unclear what caused this pattern in the current study: Castro and colleagues' (1998) study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
 Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.

### 2020-12-23 manuscript.Rmd
minor

@@
 
-We are grateful to the individuals who participated in the study.
+We are grateful to the individuals who participated in the study, and the families and communities that made it possible.
 The collection and annotation of these recordings was made possible by Ndapw:éé Yidika, Taakêmê Ńamono, and Y:aaw:aa Pikuwa; with thanks also to the PNG National Research Institute, and the Administration of Milne Bay Province.
+We owe big thanks also to Stephen C.
+Levinson for his invaluable advice and support and Shawn C.
+Tice for helpful discussion during data collection.
 AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J.

### 2021-01-05 manuscript.Rmd
stopped midway the results

@@
 To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID & target ID as random variables.
+The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and allowed slopes to vary over the random effects child ID and target ID.
 
@@
 Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
-Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as shown in Figure \@ref(fig:Fig2-xling-freq).
+Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \@ref(fig:Fig2-xling-freq).
 Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.
@@
 We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^2] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, p `r pval.display(round(freq_cor$p.value,3))`).
+Item-level average corpus-based phone frequencies were correlated with the item-level average cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, p `r pval.display(round(freq_cor$p.value,3))`).
 Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = `r round(cor.test(cor_fr_tab$corf,cor_fr_tab$Word)$estimate,3)`.
@@
 As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std.
+In this model, both cross-linguistic phone frequency (ß = `r both_coeff["avg_fr_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_s","Std.
 Error"]`, p `r pval.display(both_coeff["avg_fr_s","Pr(>|z|)"])`) and age (ß = `r both_coeff["age.rounded_s","Estimate"]`, SE ß = `r both_coeff["age.rounded_s","Std.
-Error"]`, p `r pval.display(both_coeff["age.rounded_s","Pr(>|z|)"])`) were significant predictors of whole-item  NWR scores, but corpus frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
+Error"]`, p `r pval.display(both_coeff["age.rounded_s","Pr(>|z|)"])`) were significant predictors of whole-item  NWR scores, but corpus phone frequency (ß = `r both_coeff["avg_fr_cor_ty_s","Estimate"]`, SE ß = `r both_coeff["avg_fr_cor_ty_s","Std.
 Error"]`, p `r pval.display(both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"])`) was not.
@@
 sub_tab=table(sub_bank)[allVC,allVC]
+#this table has target as rows, realization as columns
 
@@
 Next, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
-Deletions were relatively rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
-Substitutions were somewhat more frequent, but still relatively rare.
-Nasal vowels were produced as oral vowels and vice versa (`r sum(sub_tab[oralVs,nasalVs])` oral targets as nasal (~`r mean(sub_pc[oralVs,nasalVs],na.rm=T)`% of targets) and `r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral (~`r mean(sub_pc[nasalVs,oralVs],na.rm=T)`% of targets)).
-Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (`r sum(sub_tab[oralVs,oralVs])` oral vowels (~`r mean(sub_pc[oralVs,oralVs],na.rm=T)`% of targets and `r sum(sub_tab[nasalVs,nasalVs])` nasal vowels (~`r mean(sub_pc[nasalVs,nasalVs],na.rm=T)`% of targets).
+Deletions were very rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
+We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] )*100,2)` of unmatched phones).
+
+<!-- reminder: this table has target as rows, realization as columns  -->
+Is it the case that cross-linguistically less common phones are more frequently substituted by more common ones than vice versa?
+We found that errors involving nasal vowel targets tended to result in a change in nasality (`r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral, ~`r round(sum(sub_tab[nasalVs,oralVs])/sum(sub_tab[nasalVs,])*100)`% of substitutions for nasal vowel target phones) more frequently than in a change in quality (`r sum(sub_tab[nasalVs,nasalVs])` nasal targets produced as a different nasal vowel, ~`r round(sum(sub_tab[nasalVs,nasalVs])/sum(sub_tab[nasalVs,])*100)`% of substitutions for nasal targets).
+The opposite was true for oral vowels: there were `r sum(sub_tab[oralVs,nasalVs])` oral targets produced as nasal, constituting ~`r round(sum(sub_tab[oralVs,nasalVs])/sum(sub_tab[oralVs,])*100)`% of substitutions for oral target phones; compared to `r sum(sub_tab[oralVs,oralVs])` oral targets produced as a different oral vowel, ~`r round(sum(sub_tab[oralVs,oralVs])/sum(sub_tab[oralVs,])*100)`% of substitutions for oral targets).
+
+**STOPPED HERE**
+
 Among consonants, complex consonants were occasionally substituted by simpler ones (specifically `r print.phone.list(complexCs.ipa)` as `r print.phone.list(simplexCs.ipa)`), `r sum(sub_tab[complexCs,simplexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all complex consonant targets.
@@
 In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%.
-This type of analysis is seldom reported---we could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
+This type of analysis is seldom reported.
+We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
 The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
@@
 While whole-item scoring is the most commonly reported NWR outcome, it is also the least forgiving.
-The pattern is less marked when other two scores are used, which are less sensitive to errors.
 Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(`r length_r$parameter`) = `r round(length_r$estimate,3)`.
@@
 length.tab=NULL
-for(myvar in c("Word","Phoneme","NLD")){
+for(myvar in c("Word","NLD")){
   length.tab=cbind(length.tab,
@@
 rownames(length.tab)<-levels(by_length$Length)
-colnames(length.tab)<-c("Word","Phoneme","NLD")
-apa_table(length.tab,caption="NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
+colnames(length.tab)<-c("Word","NLD")
+apa_table(length.tab,caption="NWR measured in whole-word scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
 ```

### 2021-01-05 manuscript.tex
stopped midway the results

@@
 To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and child ID \& target ID as random variables.
+The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and allowed slopes to vary over the random effects child ID and target ID.
 
 We could include 850 observations, from 40 children producing in any given trial one of 41 potential target words.
-The analysis revealed a main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as shown in Figure \ref{fig:Fig2-xling-freq}.
+The analysis revealed a main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:Fig2-xling-freq}.
 Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(32) = 0.61.
@@
 We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Phone frequencies estimated from this corpus were correlated with cross-linguistic phone frequencies (r(27)=0.50, p \textless{} 0.01).
+Item-level average corpus-based phone frequencies were correlated with the item-level average cross-linguistic phone frequencies (r(27)=0.50, p \textless{} 0.01).
 Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = 0.46.
@@
 As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic frequency (ß = 0.82, SE ß = 0.27, p \textless{} 0.01) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of whole-item NWR scores, but corpus frequency (ß = 0.01, SE ß = 0.25, p = 0.98) was not.
+In this model, both cross-linguistic phone frequency (ß = 0.82, SE ß = 0.27, p \textless{} 0.01) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of whole-item NWR scores, but corpus phone frequency (ß = 0.01, SE ß = 0.25, p = 0.98) was not.
 
@@
 Next, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
-Deletions were relatively rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
-Substitutions were somewhat more frequent, but still relatively rare.
-Nasal vowels were produced as oral vowels and vice versa (52 oral targets as nasal (\textasciitilde{}1.10\% of targets) and 58 nasal targets produced as oral (\textasciitilde{}2.53\% of targets)).
-Both oral and nasal vowels were also occasionally subsituted by vowels of a different quality that preserved the target oral/nature airflow (205 oral vowels (\textasciitilde{}2.16\% of targets and 23 nasal vowels (\textasciitilde{}2.36\% of targets).
+Deletions were very rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
+We therefore focus our qualitative description here on substitutions: There were 884 cases of substitutions, \textasciitilde{}17.83 of the 4957 phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (\textasciitilde{}97.90 of unmatched phones).
+
+Is it the case that cross-linguistically less common phones are more frequently substituted by more common ones than vice versa?
+We found that errors involving nasal vowel targets tended to result in a change in nasality (58 nasal targets produced as oral, \textasciitilde{}72\% of substitutions involving nasal vowels) more frequently than in a change in quality (23 nasal targets produced as a different nasal vowel, \textasciitilde{}28\% of substitutions involving nasal vowels).
+We found that errors involving nasal vowel targets tended to result in a change in nasality (52 oral targets produced as nasal, \textasciitilde{}20\% of substitutions involving nasal vowels) more frequently than in a change in quality (205 nasal targets produced as a different nasal vowel, \textasciitilde{}80\% of substitutions involving nasal vowels).
+
+the proportion of errors for to be mistakenly produced as oral vowels than vice versa once prevalence in our stimuli was taken : 52 oral target vowels produced as nasal (\textasciitilde{}20\% of substituted oral target vowels) and Both oral and nasal vowels were also occasionally substituted by vowels of a different quality that preserved the target oral/nature airflow: 205 oral vowels (\textasciitilde{}2.16\% of targets and 23 nasal vowels (\textasciitilde{}2.36\% of targets).
+
 Among consonants, complex consonants were occasionally substituted by simpler ones (specifically {[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), 250 times, in approximately 2.43\% of all complex consonant targets.
@@
 In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: 63\%.
-This type of analysis is seldom reported---we could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
+This type of analysis is seldom reported.
+We could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
 The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
@@
 While whole-item scoring is the most commonly reported NWR outcome, it is also the least forgiving.
-The pattern is less marked when other two scores are used, which are less sensitive to errors.
 Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(1) = -0.91.
@@
 
-\caption{\label{tab:tablength}NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
+\caption{\label{tab:tablength}NWR measured in whole-word scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
 
-\begin{tabular}{lll}
+\begin{tabular}{ll}
 \toprule
-Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
+Word & \multicolumn{1}{c}{NLD}\\
 \midrule
-47 (22) & 59 (17) & 41 (17)\\
-79 (22) & 92 (9) & 8 (9)\\
-78 (19) & 93 (7) & 7 (7)\\
-74 (32) & 91 (12) & 9 (12)\\
+47 (22) & 41 (17)\\
+79 (22) & 8 (9)\\
+78 (19) & 7 (7)\\
+74 (32) & 9 (12)\\
 \bottomrule

### 2021-01-05 manuscript.Rmd
shorten YD phonology subsection

@@
 
-<!-- a2m No, I didn't propose that we would use NWR because I wanted to contribute to our understanding of the NWR task in new settings, but rather because it's quite a standard task so it could be useful for us documenting phonological development in Rossel.
-To make this clearer, I moved it down and integrated it with the four contributions below -->
-
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
@@
 The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region.
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
+<!--
+Removed for brevity:
+and includes at least two contrasts that had not previously been documented elsewhere [labial-coronal double-articulations with dental vs.
+post-alveolar coronal placement for both oral and nasal stops, @ladefoged1996sounds; @maddiesonIPphoneticsYD].
+-->
 
 With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).^[We use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+The core oral stop system includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).^[We use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
 In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).] The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
@@
 
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar] and includes at least two contrasts that had not previously been documented elsewhere [labial-coronal double-articulations with dental vs.
-post-alveolar coronal placement for both oral and nasal stops, @ladefoged1996sounds; @maddiesonIPphoneticsYD].
-
-Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at the time of writing; @levinsonYDgrammar].
+Most Yélî Dnye words are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., "ndap" pronounced "ndapî" /'ṇṭæ.pɯ/).
-This process is used frequently with English loan words that have a coda (e.g., "ponî" (phone) /'pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., "ala" (here) /æ.'læ/), but these are limited to /æ/-.
-Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
+A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" (Spondylus shell) /ṇṭæp/) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/).
+<!--Removed for brevity: This process is used frequently with English loan words that have a coda (e.g., "ponî" (phone) /'pɔ.nɯ/).--> There are also a handful of words starting with /æ/ (e.g., "ala" (here) /æ.'læ/) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
 

### 2021-01-05 manuscript.Rmd
shorten YD community subsection

@@
 
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
+<!--Removed for brevity: Most Yélî people are swidden horticulturalists who cultivate yams, manioc, taro, coconut, banana, and more, getting protein primarily from fish, shellfish, and nuts.
+There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport.
+News and messages are passed to and from mainland PNG via a radio connection at the local health clinic and, when it is functioning, a cellular tower on the island.--> International phone calls and digital exchanges that require significant data transfer are typically not an option.
+Data collection is therefore typically limited to the duration of the researchers' on-island visits.
+
 Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
@@
 <!-- a2m I thought in the first couple of years they combined English and Yélî at school? -->
+<!-- m2a: It used to be!
+In Anna, Ghaalyu, and Taakêmê's time it *was* but policies have shifted a bit since then.
+These days it looks like there's room for things to swing back toward bilingual early education.
+But anyhow for the kids in the current dataset we can pretty reliably say that they're language of schooling was English from the start.-->
 
-Most Yélî people are swidden horticulturalists, raising a variety of tubers (e.g., yams, manioc, taro), coconut, banana, and more in their gardens.
-In addition to garden foods, Yélî children grow up eating fish and foraging for shellfish and nuts, which provide a regular source of extra protein.
-The typical household in our dataset includes seven individuals and is situated among a collection of 4 or more other households, with structures often arranged around an open grassy area.
+The typical household in our dataset includes seven individuals and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
 These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
 
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
-There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available.
-There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG.
-There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time.
-As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, are typically unavailable.
-Our data collection is therefore limited to the duration of the researchers' on-island visits.
+Despite restricted outside contact and the fact that most Yélî parents are swidden horticulturalists, formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early].
+Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
+Combined with practices of collective child guardianship, the practical consequence of this billeting for researchers is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood [@brownIPchildrearing].
 
-Despite restricted outside contact, formal education is a priority for many Yélî families.
-A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early].
-While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day).
-It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segments of each school year.
-Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].
-Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
-
-<!-- a2m why do we explain this: Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
--->
-
-While work on Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun [e.g., @peuteIPconsonants].
-What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day [@casillas2020early], shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children [@bunceURcrosscultural; @casillas2020early].
-While speech from women predominates in young children's language environments on Rossel, as it does elsewhere [@bergelsoncasillas2019what; @bunceURcrosscultural], there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early].
-The latter increase is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
-
-<!-- a2m including Bunce here may be ambiguous, and it could be read to indicate that speech from other children increases in other cultures as well: there is a significant and marked increase in child-directed speech from other children as infants get older [@bunceURcrosscultural; @casillas2020early] -->
+Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun [e.g., @peuteIPconsonants].
+Direct speech to children under 3;0 is relatively infrequent throughout the day and comes primarily from women [@casillas2020early], though shared caregiving practices ensure that children frequently hear the voices of men and other children [@casillas2020early].
+Starting around age 2, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups, increasing the relative proportion of linguistic input from other children [@brownIPchildrearing; @casillas2020early].
 

### 2021-01-06 manuscript.Rmd
done going through results, still in progress

@@
 
-The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
+The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
 
-```{r tab2-stims, results = "asis"}
+```{r tabstims, results = "asis"}
 #add orthographic representations targets & other stims chars
@@
 
-Finally, when describing children's patterns of errors, we take all repetitions of a given target into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, insertions, and substitutions.
-
 ### Participants
@@
 The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.
+
+We also looked into the frequency with which mispronunciations resulted in real words.
+In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%.
+This type of analysis is seldom reported.
+We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
 
@@
 #rowSums(sub_pc, na.rm=T) + del_pc + mat_pc
+
 ```
 
-Next, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
+Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
+To answer this question, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
 Deletions were very rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
@@
 
-<!-- reminder: this table has target as rows, realization as columns  -->
-Is it the case that cross-linguistically less common phones are more frequently substituted by more common ones than vice versa?
-We found that errors involving nasal vowel targets tended to result in a change in nasality (`r sum(sub_tab[nasalVs,oralVs])` nasal targets produced as oral, ~`r round(sum(sub_tab[nasalVs,oralVs])/sum(sub_tab[nasalVs,])*100)`% of substitutions for nasal vowel target phones) more frequently than in a change in quality (`r sum(sub_tab[nasalVs,nasalVs])` nasal targets produced as a different nasal vowel, ~`r round(sum(sub_tab[nasalVs,nasalVs])/sum(sub_tab[nasalVs,])*100)`% of substitutions for nasal targets).
-The opposite was true for oral vowels: there were `r sum(sub_tab[oralVs,nasalVs])` oral targets produced as nasal, constituting ~`r round(sum(sub_tab[oralVs,nasalVs])/sum(sub_tab[oralVs,])*100)`% of substitutions for oral target phones; compared to `r sum(sub_tab[oralVs,oralVs])` oral targets produced as a different oral vowel, ~`r round(sum(sub_tab[oralVs,oralVs])/sum(sub_tab[oralVs,])*100)`% of substitutions for oral targets).
+```{r tab-v,results="asis"}
 
-**STOPPED HERE**
+#vowel changes
+mytab=rbind(
+  #OK --                      nasal change --                 qual change
+  cbind(sum(match_tab[nasalVs]),sum(sub_tab[nasalVs,oralVs]),sum(sub_tab[nasalVs,nasalVs])), #nasal target
+  cbind(sum(match_tab[oralVs]),sum(sub_tab[oralVs,nasalVs]),sum(sub_tab[oralVs,oralVs])) #oral target
+)
+mytab=data.frame(mytab)
+mytab$pc_corr=round(mytab[,1]/rowSums(mytab)*100,1)
+mytab$pc_nas_err=round(mytab[,2]/rowSums(mytab)*100,1)
+mytab$pc_qual_err=round(mytab[,3]/rowSums(mytab)*100,1)
+rownames(mytab)<-c("Nasal Target", "Oral Target")
+colnames(mytab)<-c("Correct","Nasal Err.","Qual. Err","% Corr.","% Nasal Err.","% Qual Err.")
 
-Among consonants, complex consonants were occasionally substituted by simpler ones (specifically `r print.phone.list(complexCs.ipa)` as `r print.phone.list(simplexCs.ipa)`), `r sum(sub_tab[complexCs,simplexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all complex consonant targets.
-Interestingly, simple consonants were sometimes also substituted by complex ones; this happened `r sum(sub_tab[simplexCs,complexCs])` times, in approximately `r mean(sub_pc[complexCs,simplexCs],na.rm=T)`% of all simple consonant targets.
-Finally, simple consonants were mispronounced as other simple consonants (`r sum(sub_tab[simplexCs,simplexCs])` instances, ~`r mean(sub_pc[simplexCs,simplexCs],na.rm=T)`% of all simple consonant targets) and complex consonants as other complex consonants (`r sum(sub_tab[complexCs,complexCs])` instances, ~`r mean(sub_pc[complexCs,complexCs],na.rm=T)`% of all complex consonant targets).
-Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.
+kable(mytab,
+  "latex", booktabs = T,
+  caption="Number (and percent) of vowel targets that were correctly repeated or substituted, as a function of vowel type, and whether the error resulted in a nasality change (Nasal Err.) or only a quality change (Qual.
+Err.) Deletion errors are not shown.")
 
-Finally, we looked into the frequency with which mispronunciations resulted in real words.
-In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%.
-This type of analysis is seldom reported.
-We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
+```
+
+We approached our question regarding potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels, distinguishing errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (but were therefore a quality error), shown in Table \@ref(tab:tab-v).
+We found that errors involving nasal vowel targets were more common than those involving oral vowels (`r round(100-mytab["Nasal Target","% Corr."],1)` versus `r round(100-mytab["Oral Target","% Corr."],1)`).
+Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
+
+```{r tab-c,results="asis"}
+
+#consnant changes
+mytab=rbind(
+  #OK --                      complexity change --                 other change
+  cbind(sum(match_tab[complexCs]),sum(sub_tab[complexCs,simplexCs]),sum(sub_tab[complexCs,complexCs])), #nasal target
+  cbind(sum(match_tab[simplexCs]),sum(sub_tab[simplexCs,complexCs]),sum(sub_tab[simplexCs,simplexCs])) #oral target
+)
+mytab=data.frame(mytab)
+mytab$pc_corr=round(mytab[,1]/rowSums(mytab)*100,1)
+mytab$pc_comp_err=round(mytab[,2]/rowSums(mytab)*100)
+mytab$pc_oth_err=round(mytab[,3]/rowSums(mytab)*100,1)
+rownames(mytab)<-c("Complex Target", "Simple Target")
+colnames(mytab)<-c("Correct","Compl. Err.","Other Err","% Corr.","% Compl. Err.","% Other Err.")
+
+kable(mytab,
+  "latex", booktabs = T,
+  caption="Number (and percent) of consonant targets that were correctly repeated or substituted, as a function of the complexity of the consonant, and whether the error resulted in a change of complexity (Compl Err.) or not (Other Err.) Deletion errors are not shown.")
+
+```
+
+For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.] Results, shown in Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving oral vowels (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
+Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
 
@@
 
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$type),mean)->by_length_no_child
-colnames(by_length_no_child)<-c("Length","Word")
-by_length_no_child=by_length_no_child[-1,]
-length_r=cor.test(by_length_no_child$Length,by_length_no_child$Word)
-
 len_lm=glmer(correct~type+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial",subset=c(trnsc.inc$type!=1))
@@
 This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may indicate that they are hard to produce or perceive.
-Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole-item scoring.
-While whole-item scoring is the most commonly reported NWR outcome, it is also the least forgiving.
-Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(`r length_r$parameter`) = `r round(length_r$estimate,3)`.
-In a generalized binomial mixed model, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing, in any given trial, one of `r summary(len_lm)$ngrps["target"]` potential target words.
+Setting monosyllables aside, we observed the typical pattern of lower scores for longer items only for the whole-item scoring, and even there differences were rather small.
+In a generalized binomial mixed model excluding monosyllables, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing, in any given trial, one of `r summary(len_lm)$ngrps["target"]` potential target words.
 The analysis revealed a positive effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
@@
 }
-rownames(length.tab)<-levels(by_length$Length)
+rownames(length.tab)<-paste(c(1:4),"syll")
 colnames(length.tab)<-c("Word","NLD")
-apa_table(length.tab,caption="NWR measured in whole-word scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.")
+kable(length.tab,"latex", booktabs = T,caption="NWR means (and standard deviations) measured in whole-word scores and normalized Levenshtein Distance (NLD), separately for the four stimuli lengths.")
 ```
@@
 
-```{r Fig2-scores-by-age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-item scores for individual participants as a function of age and sex (purple = boys, orange = girls)."}
-
-plot(overall$Word~overall$Age,pch=20,cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","darkmagenta","chocolate1"))
-
+```{r indif}
 cor.test(overall$Word,overall$Age,method="s")->corage
@@
 cor.test(overall$Word,overall$matEd,method="s")->corme
+```
+
+```{r Fig2-scores-by-age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-item scores for individual participants as a function of age and sex (purple = boys, orange = girls)."}
+
+plot(overall$Word~overall$Age,pch=4,cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","darkmagenta","chocolate1"))
 
@@
 As shown in Figure \@ref(fig:Fig2-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
-In contrast, there was no clear association between NWR scores and sex (t (`r tsex$statistic`) = `r tsex$parameter`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r corbo$statistic`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children,  rho ( `r corme$statistic`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
+In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children,  rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
 
@@
 That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
-Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
-Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998illiterate], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998illiterate] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
-Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
 
@@
 In line with this prediction, our descriptive analysis showed no correlation between birth order and NWR score.
+
+### Other observations
+
+Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
+Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998illiterate], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998illiterate] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
 

### 2021-01-06 manuscript.tex
done going through results, still in progress

@@
 The Yélî sound system, much like its baroque grammatical system (Levinson, 2020), is unlike any other in the region.
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.).
 
 With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (2020) for details).\footnote{We use Levinson's (2020) under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+The core oral stop system includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (2020) for details).\footnote{We use Levinson's (2020) under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
 In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
@@
 
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020) and includes at least two contrasts that had not previously been documented elsewhere (labial-coronal double-articulations with dental vs.~post-alveolar coronal placement for both oral and nasal stops, Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, n.d.).
-
-Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent Yélî Dnye dictionary at the time of writing; Levinson, 2020).
+Most Yélî Dnye words are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020).
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/).
-However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., \enquote{ndap} pronounced \enquote{ndapî} /'ṇṭæ.pɯ/).
-This process is used frequently with English loan words that have a coda (e.g., \enquote{ponî} (phone) /'pɔ.nɯ/).
-An even smaller portion of the lexicon features words starting with a vowel (e.g., \enquote{ala} (here) /æ.'læ/), but these are limited to /æ/-.
-Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (2020) for details).
+A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \enquote{ndapî} /'ṇṭæ.pɯ/).
+There are also a handful of words starting with /æ/ (e.g., \enquote{ala} (here) /æ.'læ/) and a small collection of single-vowel grammatical morphemes (see Levinson (2020) for details).
 
@@
 \subsubsection{The Yélî community}\label{the-yuxe9luxee-community}}
+
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
+International phone calls and digital exchanges that require significant data transfer are typically not an option.
+Data collection is therefore typically limited to the duration of the researchers' on-island visits.
 
@@
 
-Most Yélî people are swidden horticulturalists, raising a variety of tubers (e.g., yams, manioc, taro), coconut, banana, and more in their gardens.
-In addition to garden foods, Yélî children grow up eating fish and foraging for shellfish and nuts, which provide a regular source of extra protein.
-The typical household in our dataset includes seven individuals and is situated among a collection of 4 or more other households, with structures often arranged around an open grassy area.
+The typical household in our dataset includes seven individuals and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
 These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
 
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
-There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available.
-There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG.
-There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time.
-As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, are typically unavailable.
-Our data collection is therefore limited to the duration of the researchers' on-island visits.
+Despite restricted outside contact and the fact that most Yélî parents are swidden horticulturalists, formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available (Casillas, Brown, \& Levinson, 2020).
+Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
+Combined with practices of collective child guardianship, the practical consequence of this billeting for researchers is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, n.d.).
 
-Despite restricted outside contact, formal education is a priority for many Yélî families.
-A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65\% of mothers and 49\% of fathers completing the full range of locally available education (\textasciitilde{}9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG (Casillas, Brown, \& Levinson, 2020).
-While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day).
-It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segments of each school year.
-Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward (Brown \& Casillas, n.d.).
-Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.
-
-While work on Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas et al., 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun (e.g., Peute, Fikkert, \& Casillas, n.d.).
-What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day (Casillas et al., 2020), shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children (Bunce et al., n.d.; Casillas et al., 2020).
-While speech from women predominates in young children's language environments on Rossel, as it does elsewhere (Bergelson et al., 2019; Bunce et al., n.d.), there is a significant and marked increase in child-directed speech from other children as infants get older (Bunce et al., n.d.; Casillas et al., 2020).
-The latter increase is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
+Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas et al., 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun (e.g., Peute, Fikkert, \& Casillas, n.d.).
+Direct speech to children under 3;0 is relatively infrequent throughout the day and comes primarily from women (Casillas et al., 2020), though shared caregiving practices ensure that children frequently hear the voices of men and other children (Casillas et al., 2020).
+Starting around age 2, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups, increasing the relative proportion of linguistic input from other children (Brown \& Casillas, n.d.; Casillas et al., 2020).
 
@@
 
-The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table 2).
+The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
 
@@
 
-\caption{\label{tab:tab2-stims}NWR stimuli in orthographic (Orth.) and phonological (Phon.) representations.}
+\caption{\label{tab:tabstims}NWR stimuli in orthographic (Orth.) and phonological (Phon.) representations.}
 \centering
@@
 
-Finally, when describing children's patterns of errors, we take all repetitions of a given target into account.
-We describe the proportion of items where the change resulted in a real word (semantic errors); as well as deletions, insertions, and substitutions.
-
 \hypertarget{participants}{%
@@
 
+We also looked into the frequency with which mispronunciations resulted in real words.
+In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: 64\%.
+This type of analysis is seldom reported.
+We could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
+
 \hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
@@
 
-Next, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
+Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
+To answer this question, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
 Deletions were very rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
@@
 
-Is it the case that cross-linguistically less common phones are more frequently substituted by more common ones than vice versa?
-We found that errors involving nasal vowel targets tended to result in a change in nasality (58 nasal targets produced as oral, \textasciitilde{}72\% of substitutions involving nasal vowels) more frequently than in a change in quality (23 nasal targets produced as a different nasal vowel, \textasciitilde{}28\% of substitutions involving nasal vowels).
-We found that errors involving nasal vowel targets tended to result in a change in nasality (52 oral targets produced as nasal, \textasciitilde{}20\% of substitutions involving nasal vowels) more frequently than in a change in quality (205 nasal targets produced as a different nasal vowel, \textasciitilde{}80\% of substitutions involving nasal vowels).
+\begin{table}
 
-the proportion of errors for to be mistakenly produced as oral vowels than vice versa once prevalence in our stimuli was taken : 52 oral target vowels produced as nasal (\textasciitilde{}20\% of substituted oral target vowels) and Both oral and nasal vowels were also occasionally substituted by vowels of a different quality that preserved the target oral/nature airflow: 205 oral vowels (\textasciitilde{}2.16\% of targets and 23 nasal vowels (\textasciitilde{}2.36\% of targets).
+\caption{\label{tab:tab-v}Number (and percent) of vowel targets that were correctly repeated or substituted, as a function of vowel type, and whether the error resulted in a nasality change (Nasal Err.) or only a quality change (Qual.
+Err.) Deletion errors are not shown.}
+\centering
+\begin{tabular}[t]{lrrrrrr}
+\toprule
+  & Correct & Nasal Err. & Qual. Err & \% Corr. & \% Nasal Err. & \% Qual Err.\\
+\midrule
+Nasal Target & 134 & 58 & 23 & 62.3 & 20.9 & 7.7\\
+Oral Target & 1992 & 52 & 205 & 88.6 & 2.2 & 8.8\\
+\bottomrule
+\end{tabular}
+\end{table}
 
-Among consonants, complex consonants were occasionally substituted by simpler ones (specifically {[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]} as {[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), 250 times, in approximately 2.43\% of all complex consonant targets.
-Interestingly, simple consonants were sometimes also substituted by complex ones; this happened 2 times, in approximately 2.43\% of all simple consonant targets.
-Finally, simple consonants were mispronounced as other simple consonants (120 instances, \textasciitilde{}0.52\% of all simple consonant targets) and complex consonants as other complex consonants (55 instances, \textasciitilde{}0.98\% of all complex consonant targets).
-Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.
+We approached our question regarding potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels, distinguishing errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (but were therefore a quality error), shown in Table \ref{tab:tab-v}.
+We found that errors involving nasal vowel targets were more common than those involving oral vowels (37.70 versus 11.40).
+Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
 
-Finally, we looked into the frequency with which mispronunciations resulted in real words.
-In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: 63\%.
-This type of analysis is seldom reported.
-We could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
+\begin{table}
+
+\caption{\label{tab:tab-c}Number (and percent) of consonant targets that were correctly repeated or substituted, as a function of the complexity of the consonant, and whether the error resulted in a change of complexity (Compl Err.) or not (Other Err.) Deletion errors are not shown.}
+\centering
+\begin{tabular}[t]{lrrrrrr}
+\toprule
+  & Correct & Compl. Err. & Other Err & \% Corr. & \% Compl. Err. & \% Other Err.\\
+\midrule
+Complex Target & 277 & 250 & 55 & 47.6 & 40 & 8.2\\
+Simple Target & 1425 & 2 & 120 & 92.1 & 0 & 7.3\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.} Results, shown in Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving oral vowels (52.40 versus 7.90).
+Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
 
@@
 This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may indicate that they are hard to produce or perceive.
-Setting monosyllables aside, we observe the typical pattern of lower scores for longer items, although this is particularly salient for the whole-item scoring.
-While whole-item scoring is the most commonly reported NWR outcome, it is also the least forgiving.
-Averaging across participants and items, the Pearson correlation between length (2-4 syllables) and whole-item NWR scores was r(1) = -0.91.
-In a generalized binomial mixed model, we included 479 observations, from 40 children producing, in any given trial, one of 24 potential target words.
+Setting monosyllables aside, we observed the typical pattern of lower scores for longer items only for the whole-item scoring, and even there differences were rather small.
+In a generalized binomial mixed model excluding monosyllables, we included 479 observations, from 40 children producing, in any given trial, one of 24 potential target words.
 The analysis revealed a positive effect of age (ß = 0.56, SE ß = 0.14, p \textless{} 0.001) and a negative but non-significant estimate for target length in number of syllables (ß = -0.15, SE ß = 0.33, p = 0.65).
 
-\begin{table}[tbp]
+\begin{table}
 
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:tablength}NWR measured in whole-word scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
-
-\begin{tabular}{ll}
+\caption{\label{tab:tablength}NWR means (and standard deviations) measured in whole-word scores and normalized Levenshtein Distance (NLD), separately for the four stimuli lengths.}
+\centering
+\begin{tabular}[t]{lll}
 \toprule
-Word & \multicolumn{1}{c}{NLD}\\
+  & Word & NLD\\
 \midrule
-47 (22) & 41 (17)\\
-79 (22) & 8 (9)\\
-78 (19) & 7 (7)\\
-74 (32) & 9 (12)\\
+1 syll & 47 (22) & 41 (17)\\
+2 syll & 79 (22) & 8 (9)\\
+3 syll & 78 (19) & 7 (7)\\
+4 syll & 74 (32) & 9 (12)\\
 \bottomrule
 \end{tabular}
-
-\end{threeparttable}
-\end{center}
-
 \end{table}
@@
 As shown in Figure \ref{fig:Fig2-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (6,014.70) = 0.44, p \textless{} 0.01).
-In contrast, there was no clear association between NWR scores and sex (t (-0.29) = 27.56, p = 0.77), birth order (data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39), or maternal education (data missing for 0 children, rho ( 9,594.37) = 0.10, p = 0.54).
+In contrast, there was no clear association between NWR scores and sex (Welch t (27.56) = -0.29, p = 0.77), birth order (data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39), or maternal education (data missing for 0 children, rho ( 9,594.40) = 0.10, p = 0.54).
 
@@
 That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
-Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
-Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work (e.g., Castro-Caldas et al., 1998), but it is unclear what caused this pattern in the current study: Castro and colleagues' (1998) study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
-Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
 
@@
 In line with this prediction, our descriptive analysis showed no correlation between birth order and NWR score.
+
+\hypertarget{other-observations}{%
+\subsubsection{Other observations}\label{other-observations}}
+
+Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
+Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work (e.g., Castro-Caldas et al., 1998), but it is unclear what caused this pattern in the current study: Castro and colleagues' (1998) study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
 
@@
 \url{https://doi.org/10.1177/0142723715626067}
-
-\leavevmode\hypertarget{ref-bergelsoncasillas2019what}{}%
-Bergelson, E., Casillas, M., Soderstrom, M., Seidl, A., Warlaumont, A.
-S., \& Amatuni, A.
-(2019).
-What do North American babies hear?
-A large-scale cross-corpus analysis.
-\emph{Developmental Science}, \emph{22}(1), e12724.
-\url{https://doi.org/10.1111/desc.12724}
 
@@
 New York, NY: Berghahn.
-
-\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
-Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (n.d.). A cross-cultural examination of young children's everyday language experiences.
 
@@
 
-\leavevmode\hypertarget{ref-ladefoged1996sounds}{}%
-Ladefoged, P., \& Maddieson, I. (1996). \emph{The sounds of the world's languages} (Vol. 1012). Blackwell Oxford.
-
 \leavevmode\hypertarget{ref-levinsonYDgrammar}{}%

### 2021-01-08 manuscript.Rmd
changes to fig 1 in progress

@@
 However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies.
-To illustrate this issue, we portray in Figure 1 studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
+To illustrate this issue, we portray in Figure \@ref(fig:fig-prevlit) studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
 As discussed above,  the effect of stimulus length is minuscule among Italian children [@piazzalunga2019articulatory], but considerable among Arabic children [@jabere2018xperiment].
@@
 
-```{r prevlit-fig, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies.
+```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies.
 Arabic data  from Jaber-Awida (2018); Italian data from Piazzalunga et al.
@@
 
-pl<- read.delim("pl-With_yd.tsv")
-palette = c( 
-             "Arabic 1-2 syll" = "orange",
-             "Yélî 3-4 syll"   ="navyblue",
-             "Yélî 1-2 syll"  ="lightblue",
-             "Italian 3-4 syll"   ="black",
-             "Italian 1-2 syll"  ="gray",
-             "Arabic 3-4 syll"     = "orange", 
-             "Tsimane' 1-2 syll" = "green",
-              "Tsimane' 3-4 syll" = "darkgreen")
+pl<- read.delim("previous_lit.tsv")
+
+pl_stacked=cbind(pl[,1:4],stack(pl[5:8]),stack(pl[9:12]))
+colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
+pl_stacked$Length=gsub("X","",pl_stacked$Length)
+
+mycols = c( 
+             "Arabic" = "orange",
+             "Italian"   ="black",
+             "Tsimane'" = "darkgreen")
+
+plot(pl_stacked$Median~pl_stacked$Age,col=mycols[pl_stacked$Language],pch=as.character(pl_stacked$Length))
+#https://stackoverflow.com/questions/33700186/line-connecting-the-points-in-the-plot-function-in-r
+#lines(x[order(x)], f_x[order(x)], xlim=range(x), ylim=range(f_x), pch=16)
+#NOT FINISHED!!
 
@@
 
-```{r Fig1-first-vs-others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts."}
+```{r fig-first-vs-others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts."}
 
@@
 We excluded `r excl` child who did not have data for one of these two types.
-As shown in Figure \@ref(fig:Fig1-first-vs-others), participants' mean word-level scores became more heterogeneous in subsequent repetitions.
+As shown in Figure \@ref(fig:fig-first-vs-others), participants' mean word-level scores became more heterogeneous in subsequent repetitions.
 Surprisingly, whole-item NWR scores for subsequent repetitions (M = `r round(mean(pwcm_att$Word[pwcm_att$Attempt!="first"]))`, SD = `r round(sd(pwcm_att$Word[pwcm_att$Attempt!="first"]))`)
@@
 Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
-Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \@ref(fig:Fig2-xling-freq).
+Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \@ref(fig:fig-xling-freq).
 Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.
 
-```{r Fig2-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
+```{r fig-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
 aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,trnsc.inc$avg_fr),mean)->xling

### 2021-01-11 manuscript.Rmd
implemented changes to fig 1

@@
 
+```{r demo}
+# please note that you will not be able to do this because you don't have this version, that is not fully de-identified
+# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
+# fulldemo=fulldemo[,-2] #remove children's names
+# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
+# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
+# fulldemo$age.rounded=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
+# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
+# plot(fulldemo$Age~fulldemo$age.rounded)
+# write.csv(fulldemo,"nwr-demo.csv")
+
+# read.csv("nwr-demo.csv",header=T)->demo
+# demo=demo[,-c(20,21)] #remove DOBs 
+# write.csv(demo,"nwr-demo.csv")
+
+read.csv("nwr-demo.csv",header=T)->demo
+#colnames(demo)
+
+subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
+#if age.rounded is missing, use reported age
+inc$age.rounded[is.na(inc$age.rounded)]<-inc$Age[is.na(inc$age.rounded)]
+#summary(inc)
+table(inc$birthOrder)->birthorder.tab
+
+```
+
+```{r readin}
+read.table("final_data.txt",header=T,encoding="UTF-8")->trnsc
+
+trnsc$Sex=factor(trnsc$Sex)
+trnsc$birthOrder=as.numeric(trnsc$birthOrder)
+
+# Include only included children in trnsc
+trnsc[which(trnsc$id %in% inc$ID),]->trnsc.inc
+
+```
+
+```{r exclude subsequent}
+trnsc.inc-> trnsc.inc_all #keep a copy for figure comparing first & subsequent
+
+trnsc.inc=trnsc.inc[trnsc.inc$attempt=="first",]
+
+```
+
 ## Introduction
@@
 
-```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies.
-Arabic data  from Jaber-Awida (2018); Italian data from Piazzalunga et al.
-(2019); Tsimane' from Cristia et al.
-(2020); Yélî Dnye data from the present study."}
+```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (1-4 indicating number of syllables, 1=solid, 2=dashed, 3=dotted, 4=dotted and dashed).
+Arabic data (orange)  from Jaber-Awida (2018); Italian data (black) from Piazzalunga et al.
+(2019); Tsimane' (green) from Cristia et al.
+(2020); Yélî Dnye (purple) data from the present study.
+Central tendency using median in all studies but Jaber-Awida; error using one standard error.
+Missing points and error bars indicate no or one data point, respectively.
+Age has been slightly jittered for ease of inspection of different lengths."}
 
@@
 
+#remove lines with NA median
+pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked
+
+#add YD data
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age
+
+age_cuts=c(3,6,8,13)
+m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
+m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))
+
+aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
+yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns
+
+colnames(yd_res)<-c("Length","Age","Median","SD","n")
+yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
+colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")
+
+#combine
+pl_stacked=rbind(pl_stacked,yd_res_toadd)
+
+#add SE
+pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
+
 mycols = c( 
@@
              "Italian"   ="black",
-             "Tsimane'" = "darkgreen")
+             "Tsimane'" = "darkgreen",
+             "Yélî Dnye" = "purple")
 
-plot(pl_stacked$Median~pl_stacked$Age,col=mycols[pl_stacked$Language],pch=as.character(pl_stacked$Length))
-#https://stackoverflow.com/questions/33700186/line-connecting-the-points-in-the-plot-function-in-r
-#lines(x[order(x)], f_x[order(x)], xlim=range(x), ylim=range(f_x), pch=16)
-#NOT FINISHED!!
+jit_age=(pl_stacked$Age+(as.numeric(as.character(pl_stacked$Length))-2)/5)
+plot(pl_stacked$Median~jit_age,col=mycols[pl_stacked$Language],pch=as.character(pl_stacked$Length),
+     xlab="Average age (years)",ylab="NWR Scores (words)",ylim=c(0,100))
 
-myplot = ggplot(data = pl, aes(x = Age, y = Score, color = Language))+
-  geom_point(aes(size = n, shape = Length))+
-  geom_line(aes(linetype=Length)) + 
-  theme(text = element_text(size = 16),
-    legend.position="none", axis.text.x=element_text(colour="black")) +
-  scale_color_manual(values = palette)+
-  theme_bw()+
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.key.height = unit(5, "mm"))+
-  scale_x_continuous(name = "Age (years) ", limits = c(4,12), breaks = c(4,8,12))+
-
-  scale_y_continuous(name = "NWR Score", limits = c(0,100), breaks = seq(0, 100, 10))+
-  scale_size_continuous(range = c(2, 6))
-
-myplot
+for(thislang in levels(pl_stacked$Language)) for(thislength in levels(factor(pl_stacked$Length[pl_stacked$Language== thislang]))){
+  myx=pl_stacked$Age[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
+  myx=myx+(as.numeric(as.character(thislength))-2)/5 #jitter the length lines a little bit with respect to avg age
+  myy=pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
+  lines(myy~myx, col=mycols[thislang],lty=as.numeric(thislength)) #join points of same length & lang
+  mylb=myy-pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
+  myhb=myy+pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
+  for(i in 1:length(mylb)) {
+    mylty=as.numeric(as.character(thislength))
+    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang],lty=mylty) #error bar
+    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang],lty=mylty) #lower error bar tip
+    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang],lty=mylty) #higher error bar tip
+    }
+}
 
@@
 
-```{r demo}
-# please note that you will not be able to do this because you don't have this version, that is not fully de-identified
-# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
-# fulldemo=fulldemo[,-2] #remove children's names
-# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
-# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
-# fulldemo$age.rounded=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
-# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
-# plot(fulldemo$Age~fulldemo$age.rounded)
-# write.csv(fulldemo,"nwr-demo.csv")
-
-# read.csv("nwr-demo.csv",header=T)->demo
-# demo=demo[,-c(20,21)] #remove DOBs 
-# write.csv(demo,"nwr-demo.csv")
-
-read.csv("nwr-demo.csv",header=T)->demo
-#colnames(demo)
-
-subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
-#if age.rounded is missing, use reported age
-inc$age.rounded[is.na(inc$age.rounded)]<-inc$Age[is.na(inc$age.rounded)]
-#summary(inc)
-table(inc$birthOrder)->birthorder.tab
-
-```
-
 We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlets.
@@
 
-```{r readin}
-read.table("final_data.txt",header=T,encoding="UTF-8")->trnsc
-
-trnsc$Sex=factor(trnsc$Sex)
-trnsc$birthOrder=as.numeric(trnsc$birthOrder)
-
-# Remove excluded children from trnsc
-trnsc[which(trnsc$id %in% inc$ID),]->trnsc.inc
-
-```
-
 ```{r fig-first-vs-others, fig.pos = "!t", fig.width=5, fig.height=5, fig.cap="Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts."}
 
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$age.rounded,trnsc.inc$attempt),mean)->pwcm_att
+aggregate(trnsc.inc_all$correct*100,by=list(trnsc.inc_all$id,trnsc.inc_all$age.rounded,trnsc.inc_all$attempt),mean)->pwcm_att
 colnames(pwcm_att)<-c("ID","Age","Attempt","Word")
@@
 
-```{r exclude subsequent}
-trnsc.inc=trnsc.inc[trnsc.inc$attempt=="first",]
-
-```
-
 ```{r overall}
@@
 We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference.
 

### 2021-01-11 manuscript.tex
implemented changes to fig 1

@@
 However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies.
-To illustrate this issue, we portray in Figure 1 studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
+To illustrate this issue, we portray in Figure \ref{fig:fig-prevlit} studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
 As discussed above, the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Arabic children (Jaber-Awida, 2018).
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/prevlit-fig-1.pdf}
-\caption{\label{fig:prevlit-fig}NWR scores as a function of age (in years) and item length for comparable studies.
-Arabic data from Jaber-Awida (2018); Italian data from Piazzalunga et al.
-(2019); Tsimane' from Cristia et al.
-(2020); Yélî Dnye data from the present study.}
+\includegraphics{manuscript_files/figure-latex/fig-prevlit-1.pdf}
+\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (1-4 indicating number of syllables, 1=solid, 2=dashed, 3=dotted, 4=dotted and dashed).
+Arabic data (orange) from Jaber-Awida (2018); Italian data (black) from Piazzalunga et al.
+(2019); Tsimane' (green) from Cristia et al.
+(2020); Yélî Dnye (purple) data from the present study.
+Central tendency using median in all studies but Jaber-Awida; error using one standard error.
+Missing points and error bars indicate no or one data point, respectively.
+Age has been slightly jittered for ease of inspection of different lengths.}
 \end{figure}
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/Fig1-first-vs-others-1.pdf}
-\caption{\label{fig:Fig1-first-vs-others}Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts.}
+\includegraphics{manuscript_files/figure-latex/fig-first-vs-others-1.pdf}
+\caption{\label{fig:fig-first-vs-others}Whole-item NWR scores for individual participants averaging separately their first attempts and all other attempts.}
 \end{figure}
@@
 We excluded 1 child who did not have data for one of these two types.
-As shown in Figure \ref{fig:Fig1-first-vs-others}, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
+As shown in Figure \ref{fig:fig-first-vs-others}, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
 Surprisingly, whole-item NWR scores for subsequent repetitions (M = 40, SD = 26)
@@
 We could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, or even test structure explains this difference.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference.
 
@@
 We could include 850 observations, from 40 children producing in any given trial one of 41 potential target words.
-The analysis revealed a main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:Fig2-xling-freq}.
+The analysis revealed a main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:fig-xling-freq}.
 Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(32) = 0.61.
@@
 
-\caption{NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}\label{fig:Fig2-xling-freq}
+\caption{NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}\label{fig:fig-xling-freq}
 \end{figure}

### 2021-01-12 manuscript.Rmd
further fixing fig 1 and data checks

@@
 ```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (1-4 indicating number of syllables, 1=solid, 2=dashed, 3=dotted, 4=dotted and dashed).
-Arabic data (orange)  from Jaber-Awida (2018); Italian data (black) from Piazzalunga et al.
-(2019); Tsimane' (green) from Cristia et al.
-(2020); Yélî Dnye (purple) data from the present study.
-Central tendency using median in all studies but Jaber-Awida; error using one standard error.
-Missing points and error bars indicate no or one data point, respectively.
-Age has been slightly jittered for ease of inspection of different lengths."}
+Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
+(2019) reported on groups of 24-60 Italian learners (black); Cristia et al.
+(2020) reported on groups of 4-6 Tsimane' learners (green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
+Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
+Age has been slightly jittered for ease of inspection of different lengths at a given age."}
 
-pl<- read.delim("previous_lit.tsv")
+pl<- read.csv("previous_lit.csv")
 
@@
 
+mywd = c( 
+             "Arabic" = 1,
+             "Italian"   =1,
+             "Tsimane'" = 1,
+             "Yélî Dnye" = 2)
+
 jit_age=(pl_stacked$Age+(as.numeric(as.character(pl_stacked$Length))-2)/5)
@@
   myy=pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
-  lines(myy~myx, col=mycols[thislang],lty=as.numeric(thislength)) #join points of same length & lang
+  lines(myy~myx, col=mycols[thislang],lty=as.numeric(thislength),lwd=mywd[thislang]) #join points of same length & lang
   mylb=myy-pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
@@
 As shown in Figure \@ref(fig:Fig2-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
-In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (data missing for `r sum(is.na(demo$matEd[demo$included=="yes"]))` children,  rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
+In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
 

### 2021-01-12 manuscript.tex
further fixing fig 1 and data checks

@@
 \caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (1-4 indicating number of syllables, 1=solid, 2=dashed, 3=dotted, 4=dotted and dashed).
-Arabic data (orange) from Jaber-Awida (2018); Italian data (black) from Piazzalunga et al.
-(2019); Tsimane' (green) from Cristia et al.
-(2020); Yélî Dnye (purple) data from the present study.
-Central tendency using median in all studies but Jaber-Awida; error using one standard error.
-Missing points and error bars indicate no or one data point, respectively.
-Age has been slightly jittered for ease of inspection of different lengths.}
+Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
+(2019) reported on groups of 24-60 Italian learners (black); Cristia et al.
+(2020) reported on groups of 4-6 Tsimane' learners (green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
+Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
+Age has been slightly jittered for ease of inspection of different lengths at a given age.}
 \end{figure}
@@
 As shown in Figure \ref{fig:Fig2-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (6,014.70) = 0.44, p \textless{} 0.01).
-In contrast, there was no clear association between NWR scores and sex (Welch t (27.56) = -0.29, p = 0.77), birth order (data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39), or maternal education (data missing for 0 children, rho ( 9,594.40) = 0.10, p = 0.54).
+In contrast, there was no clear association between NWR scores and sex (Welch t (27.56) = -0.29, p = 0.77), birth order (data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39), or maternal education (rho ( 9,594.40) = 0.10, p = 0.54).
 

### 2021-01-15 manuscript.Rmd
big arch correction in preogress

@@
 In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
-This study contributes to four lines of research.
-First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
-Second, we document rather weak effects of item length, contributing to mounting research suggesting that length effects may be language- or population-specific.
-Third, we do not find strong individual variation effects in this population.
-Fourth, we discuss how NWR scores can be compared across studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development.
+Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development.
+More specifically, this study contributes to three lines of research.
+First, we contribute to investigations on NWR across diverse languages, by documenting that, in Yélî Dnye, non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
+Second, contributing to mounting research suggesting that length effects may be language- or population-specific, we find rather weak effects of item length.
+Third, we add a datapoint on potential sources of individual variation effects, by establishing that in our sample age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
 Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.
@@
 Our reading of previous NWR research is that there are variable effects of length between populations.
-For instance, @jabere2018xperiment reports an average of 95% correct repetition for items 1 or 2 syllables long among Arabic-speaking children at about 5 years of age, but 55% for items 3 or 4 syllables long.
-In contrast, @piazzalunga2019articulatory observe nearly no change in performance in similarly-aged Italian learners, with a score of 84% for 1-2 syllables versus 85% for 3-4 syllables.
+For instance, @jabere2018xperiment reports an average of ~96% correct repetition for items 2 syllables long among children learning an Arabic variety of Israeli at about 5.5 years of age, but ~81% for items 3 syllables long.
+In contrast, @piazzalunga2019articulatory observe no decline in performance in similarly-aged Italian learners, with a score of 84% for 2 syllables versus 85% for 3 syllables.
 It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture.
@@
 Third, there are ongoing discussions as to what the key factors structuring individual variation are.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review on the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword].
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword].
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
@@
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
-Indeed, NWR has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], outside urban settings [except for in @cristia2020infant], nor with languages having large phonological inventories.
+Indeed, NWR has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], and/or outside urban settings [except for in @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel;@maddieson2013consonant; no exceptions to our knowledge].
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating].
-Therefore, it may not be a general index of phonological development, but more specifically reflect certain skills.
-Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic representations.
+Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills.
+Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned.
+Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
+This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
+And it does occur that some researchers opt instead to employ dimensions of variation that are more relevant to their language, such as adaptations in Chinese languages that have items varying in tone REF.
+**return to this after reading lit**
 
@@
 pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
+
+#remove monosyllables
+pl_stacked=pl_stacked[pl_stacked$Length!=1,]
 

### 2021-01-15 manuscript.tex
big arch correction in preogress

@@
 In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
-This study contributes to four lines of research.
-First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
-Second, we document rather weak effects of item length, contributing to mounting research suggesting that length effects may be language- or population-specific.
-Third, we do not find strong individual variation effects in this population.
-Fourth, we discuss how NWR scores can be compared across studies, languages, and populations, in order to shed light on the factors universally structuring variation in language development.
+Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation in language development.
+More specifically, this study contributes to three lines of research.
+First, we contribute to investigations on NWR across diverse languages, by documenting that, in Yélî Dnye, non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
+Second, contributing to mounting research suggesting that length effects may be language- or population-specific, we find rather weak effects of item length.
+Third, we add a datapoint on potential sources of individual variation effects, by establishing that in our sample age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
 Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.

### 2021-01-18 manuscript.Rmd
changes intro

@@
 
+## TODO
+
+1. pick up from "TODO fix this para and set up comparison with Tsi" and p. 9 of print out
+1. add a table with sample studies & languages & their characteristiscs
+1. be accurate on what hasn't been done x-ling speaking
+
 ## Introduction
@@
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential changes with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
+Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
 Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
@@
 Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are typically easily accessible.
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
+<!--Removed for brevity: Most Yélî people are swidden horticulturalists who cultivate yams, manioc, taro, coconut, banana, and more, getting protein primarily from fish, shellfish, and nuts.
+There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport.
+News and messages are passed to and from mainland PNG via a radio connection at the local health clinic and, when it is functioning, a cellular tower on the island.--> International phone calls and digital exchanges that require significant data transfer are typically not an option.
+Data collection is therefore typically limited to the duration of the researchers' on-island visits.
 
@@
 The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region.
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
+Thus, with respect to our first research goal, Yélî Dnye seemed is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
+
 <!--
@@
 
+To provide some qualitative information on this inventory, we add the following observations.
 With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
@@
 Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with 7 of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
 
-Most Yélî Dnye words are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
+Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" (Spondylus shell) /ṇṭæp/) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/).
-<!--Removed for brevity: This process is used frequently with English loan words that have a coda (e.g., "ponî" (phone) /'pɔ.nɯ/).--> There are also a handful of words starting with /æ/ (e.g., "ala" (here) /æ.'læ/) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
+A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" /ṇṭæp/ *Spondylus shell*) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/).
+There are also a handful of words starting with /æ/ (e.g., "ala" /æ.'læ/ *here*) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
+
+Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun [e.g., @peuteIPconsonants].
+We hope the present study contributes to filling this gap.
+**TODO incorporate brief summary of paper**
 
@@
 
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
-<!--Removed for brevity: Most Yélî people are swidden horticulturalists who cultivate yams, manioc, taro, coconut, banana, and more, getting protein primarily from fish, shellfish, and nuts.
-There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport.
-News and messages are passed to and from mainland PNG via a radio connection at the local health clinic and, when it is functioning, a cellular tower on the island.--> International phone calls and digital exchanges that require significant data transfer are typically not an option.
-Data collection is therefore typically limited to the duration of the researchers' on-island visits.
+Some aspects of the community are relevant for interpreting results found when addressing our thir research question, regarding sources of individual variation.
+Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
+There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.
 
-Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
-Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
+The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children -- their own and billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
+Land attribution for building one's home is decided collectively based on land availability, and typically does not take into consideration an individual's desire to be close to a school.
 
-<!-- a2m I thought in the first couple of years they combined English and Yélî at school? -->
-<!-- m2a: It used to be!
-In Anna, Ghaalyu, and Taakêmê's time it *was* but policies have shifted a bit since then.
-These days it looks like there's room for things to swing back toward bilingual early education.
-But anyhow for the kids in the current dataset we can pretty reliably say that they're language of schooling was English from the start.-->
+Most Yélî parents are swidden horticulturalists, and those who are not may not reside in the island.
+Within a group of households, it is often the case that most older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early].
+Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
+Children start school often at around age six, although the precise age depends on the child's apparent development.
 
-The typical household in our dataset includes seven individuals and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
-These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
+Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above.
+To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
+Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
+Together with the fact that land attribution is essentially random with respect to educational hopes, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
 
-Despite restricted outside contact and the fact that most Yélî parents are swidden horticulturalists, formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early].
-Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
-Combined with practices of collective child guardianship, the practical consequence of this billeting for researchers is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood [@brownIPchildrearing].
+As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology].
+It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
+This contrasts with this picture just drawn in the Yélî community, where children regardless of their birth order in their nuclear family will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders.
 
-Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun [e.g., @peuteIPconsonants].
+**TODO fix this para and set up comparison with Tsi** We add some final observations that will help us integrate this study to the broader investigation of NWR across cultures.
 Direct speech to children under 3;0 is relatively infrequent throughout the day and comes primarily from women [@casillas2020early], though shared caregiving practices ensure that children frequently hear the voices of men and other children [@casillas2020early].
-Starting around age 2, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups, increasing the relative proportion of linguistic input from other children [@brownIPchildrearing; @casillas2020early].
+, increasing the relative proportion of linguistic input from other children
 
@@
 
-```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (1-4 indicating number of syllables, 1=solid, 2=dashed, 3=dotted, 4=dotted and dashed).
+```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed).
 Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
@@
 The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+As discussed in subsection "The Yélî community", the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood [@brownIPchildrearing].
 Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
@@
 We included `r sum(inc$Sex=="B")` boys and `r sum(inc$Sex=="G")` girls.
-There were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^1] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
+There were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^1] Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^2] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
 These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
 
-[^1]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
+[^1]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
+
+<!-- a2m I thought in the first couple of years they combined English and Yélî at school? -->
+<!-- m2a: It used to be!
+In Anna, Ghaalyu, and Taakêmê's time it *was* but policies have shifted a bit since then.
+These days it looks like there's room for things to swing back toward bilingual early education.
+But anyhow for the kids in the current dataset we can pretty reliably say that they're language of schooling was English from the start.-->
+
+[^2]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
 
@@
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^2] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^3] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
 Item-level average corpus-based phone frequencies were correlated with the item-level average cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, p `r pval.display(round(freq_cor$p.value,3))`).
@@
 
-[^2]: We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.
+[^3]: We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.
 

### 2021-01-18 manuscript.tex
changes intro

@@
 
+\hypertarget{todo}{%
+\subsection{TODO}\label{todo}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  pick up from \enquote{TODO fix this para and set up comparison with Tsi} and p.~9 of print out
+\item
+  add a table with sample studies \& languages \& their characteristiscs
+\item
+  be accurate on what hasn't been done x-ling speaking
+\end{enumerate}
+
 \hypertarget{introduction}{%
@@
 Our reading of previous NWR research is that there are variable effects of length between populations.
-For instance, Jaber-Awida (2018) reports an average of 95\% correct repetition for items 1 or 2 syllables long among Arabic-speaking children at about 5 years of age, but 55\% for items 3 or 4 syllables long.
-In contrast, Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019) observe nearly no change in performance in similarly-aged Italian learners, with a score of 84\% for 1-2 syllables versus 85\% for 3-4 syllables.
+For instance, Jaber-Awida (2018) reports an average of \textasciitilde{}96\% correct repetition for items 2 syllables long among children learning an Arabic variety of Israeli at about 5.5 years of age, but \textasciitilde{}81\% for items 3 syllables long.
+In contrast, Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019) observe no decline in performance in similarly-aged Italian learners, with a score of 84\% for 2 syllables versus 85\% for 3 syllables.
 It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture.
@@
 Third, there are ongoing discussions as to what the key factors structuring individual variation are.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review on the literature looking at correlations between NWR scores and a variety of child-level variables (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential changes with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
+Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
 Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
@@
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
-Indeed, NWR has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014; Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020), outside urban settings (except for in Cristia et al., 2020), nor with languages having large phonological inventories.
+Indeed, NWR has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014; Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020), and/or outside urban settings (except for in Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b);Maddieson (2013a); no exceptions to our knowledge{]}.
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991).
-Therefore, it may not be a general index of phonological development, but more specifically reflect certain skills.
+Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills.
 Indeed, Cristia et al.
-(2020) present the task as being a good index of the development of \enquote{short-hand-like} representations specifically, which could thus miss, for example, more holistic representations.
+(2020) present the task as being a good index of the development of \enquote{short-hand-like} representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from Cristia et al.
+(2020)'s hypotheses just mentioned.
+Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
+This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
+And it does occur that some researchers opt instead to employ dimensions of variation that are more relevant to their language, such as adaptations in Chinese languages that have items varying in tone REF.
+\textbf{return to this after reading lit}
 
 Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and in gathering age and other demographic information, compared to what is typical in urban, industrialized settings.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are typically easily accessible.
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
+International phone calls and digital exchanges that require significant data transfer are typically not an option.
+Data collection is therefore typically limited to the duration of the researchers' on-island visits.
 
@@
 The Yélî sound system, much like its baroque grammatical system (Levinson, 2020), is unlike any other in the region.
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.).
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.).
+Thus, with respect to our first research goal, Yélî Dnye seemed is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
 
+To provide some qualitative information on this inventory, we add the following observations.
 With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
@@
 Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with 7 of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
 
-Most Yélî Dnye words are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020).
+Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020).
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \enquote{ndapî} /'ṇṭæ.pɯ/).
-There are also a handful of words starting with /æ/ (e.g., \enquote{ala} (here) /æ.'læ/) and a small collection of single-vowel grammatical morphemes (see Levinson (2020) for details).
+A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} /ṇṭæp/ \emph{Spondylus shell}) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \enquote{ndapî} /'ṇṭæ.pɯ/).
+There are also a handful of words starting with /æ/ (e.g., \enquote{ala} /æ.'læ/ \emph{here}) and a small collection of single-vowel grammatical morphemes (see Levinson (2020) for details).
+
+Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun (e.g., Peute, Fikkert, \& Casillas, n.d.).
+We hope the present study contributes to filling this gap.
+\textbf{TODO incorporate brief summary of paper}
 
@@
 
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular.
-International phone calls and digital exchanges that require significant data transfer are typically not an option.
-Data collection is therefore typically limited to the duration of the researchers' on-island visits.
+Some aspects of the community are relevant for interpreting results found when addressing our thir research question, regarding sources of individual variation.
+Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
+There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.
 
-Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
-Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
+The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children -- their own and billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
+Land attribution for building one's home is decided collectively based on land availability, and typically does not take into consideration an individual's desire to be close to a school.
 
-The typical household in our dataset includes seven individuals and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
-These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
+Most Yélî parents are swidden horticulturalists, and those who are not may not reside in the island.
+Within a group of households, it is often the case that most older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available (Casillas et al., 2020).
+Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
+Children start school often at around age six, although the precise age depends on the child's apparent development.
 
-Despite restricted outside contact and the fact that most Yélî parents are swidden horticulturalists, formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available (Casillas, Brown, \& Levinson, 2020).
-Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
-Combined with practices of collective child guardianship, the practical consequence of this billeting for researchers is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, n.d.).
+Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above.
+To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
+Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
+Together with the fact that land attribution is essentially random with respect to educational hopes, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
 
-Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas et al., 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun (e.g., Peute, Fikkert, \& Casillas, n.d.).
+As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings (Barclay, 2015; Grätz, 2018), where nuclear families are more likely to be the prevalent rearing environment (Lancy, 2015).
+It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
+This contrasts with this picture just drawn in the Yélî community, where children regardless of their birth order in their nuclear family will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders.
+
+\textbf{TODO fix this para and set up comparison with Tsi} We add some final observations that will help us integrate this study to the broader investigation of NWR across cultures.
 Direct speech to children under 3;0 is relatively infrequent throughout the day and comes primarily from women (Casillas et al., 2020), though shared caregiving practices ensure that children frequently hear the voices of men and other children (Casillas et al., 2020).
-Starting around age 2, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups, increasing the relative proportion of linguistic input from other children (Brown \& Casillas, n.d.; Casillas et al., 2020).
+, increasing the relative proportion of linguistic input from other children
 
@@
 \includegraphics{manuscript_files/figure-latex/fig-prevlit-1.pdf}
-\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (1-4 indicating number of syllables, 1=solid, 2=dashed, 3=dotted, 4=dotted and dashed).
+\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables, 2=dashed, 3=dotted, 4=dotted and dashed).
 Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
@@
 The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+As discussed in subsection \enquote{The Yélî community}, the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, n.d.).
 Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
@@
 We included 26 boys and 14 girls.
-There were 34 only exposed to Yélî Dnye at home, 6 children exposed to Yélî Dnye plus one or more other languages at home.
-Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
+There were 34 only exposed to Yélî Dnye at home, 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.} Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
 We then record the number of years entailed by having completed that level under ideal conditions.} In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
@@
 
+\leavevmode\hypertarget{ref-barclay2015within}{}%
+Barclay, K. J. (2015). A within-family analysis of birth order and intelligence using population conscription data on swedish men. \emph{Intelligence}, \emph{49}, 134--143.
+
 \leavevmode\hypertarget{ref-Praat}{}%
@@
 
+\leavevmode\hypertarget{ref-gratz2018competition}{}%
+Grätz, M. (2018). Competition in the family: Inequality between siblings and the intergenerational transmission of educational advantage. \emph{Sociological Science}, \emph{5}, 246--269.
+
 \leavevmode\hypertarget{ref-havron2019effect}{}%
@@
 
+\leavevmode\hypertarget{ref-lancy2015anthropology}{}%
+Lancy, D. F. (2015). \emph{The anthropology of childhood}. Cambridge, UK: Cambridge University Press.
+
 \leavevmode\hypertarget{ref-levinsonYDgrammar}{}%
@@
 Maddieson, I. (2005). Correlating phonological complexity: Data and validation. \emph{UC Berkeley PhonLab Annual Report}, \emph{1}(1).
+
+\leavevmode\hypertarget{ref-maddieson2013consonant}{}%
+Maddieson, I. (2013a). Consonant inventories. \emph{The World Atlas of Language Structures Online}. Retrieved from \url{https://wals.info/chapter/1}
+
+\leavevmode\hypertarget{ref-maddieson2013vowel}{}%
+Maddieson, I. (2013b). Vowel quality inventories. \emph{The World Atlas of Language Structures Online}. Retrieved from \url{https://wals.info/chapter/2}
 

### 2021-01-19 manuscript.Rmd
midway through methods, check for N items in progress

@@
 
-1. pick up from "TODO fix this para and set up comparison with Tsi" and p. 9 of print out
 1. add a table with sample studies & languages & their characteristiscs
@@
 
-**TODO fix this para and set up comparison with Tsi** We add some final observations that will help us integrate this study to the broader investigation of NWR across cultures.
-Direct speech to children under 3;0 is relatively infrequent throughout the day and comes primarily from women [@casillas2020early], though shared caregiving practices ensure that children frequently hear the voices of men and other children [@casillas2020early].
-, increasing the relative proportion of linguistic input from other children
+We add some observations that will help us integrate this study to the broader investigation of NWR across cultures.
+As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant].
+However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
+The Yélî community can help us bring further light into this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early].
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.7 minutes per hour [@casillas2020early].
+Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in @cristia2020infant.
 
@@
 In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
-For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length.
+For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
 There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
@@
 
-After some preliminary analyses to set the stage, we address the following questions:
+After some preliminary analyses to set the stage, we perform statistical analyses to inform answers to the following questions:
 
@@
 
+Throughout these analyses and in the Discussion, we will also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.
+
 We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies.
-To illustrate this issue, we portray in Figure \@ref(fig:fig-prevlit) studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
-As discussed above,  the effect of stimulus length is minuscule among Italian children [@piazzalunga2019articulatory], but considerable among Arabic children [@jabere2018xperiment].
-Even the effect of age is unstable in this sample of studies.
-Whereas it is quite clear that children's NWR scores increase in the Italian data [@piazzalunga2019articulatory], age effects are less stable among Tsimane' children [@cristia2020infant].
+However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given the previously discussed variability across studies.
 Therefore, all analyses in the present study are descriptive and should be considered exploratory.
@@
              "Italian"   =1,
-             "Tsimane'" = 1,
-             "Yélî Dnye" = 2)
+             "Tsimane'" = 2,
+             "Yélî Dnye" = 4)
 
@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 44 items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 39 items distributed across children.
 
@@
 Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended "yî" /yɯ/; identified as "yi" /yi/, *tree*).
-This item is removed from the analyses below.
+Additionally, an error was made when preparing files for annotation, resulting in two items being merged ("tpâ" /tpɑ/ and "tp:a" /tpæ̃/).
+These three problematic items are not described here, and removed from the analyses below.
 
-The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
+The final list includes: three practice items; 18 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
 
@@
   rename("Type" = type, "Orthographic" = target_ortho, "Phonological" = phono)
+
+stims.tbl[!(stims.tbl$Orthographic %in% c("tpâ","tp:a")), ]-> stims.tbl #remove the two merged items
 
@@
 
+To inform our analyses, we estimated the typological frequency of all phonological segments present in the target items using the PHOIBLE cross-linguistic phonological inventory database [@phoible].
+For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors).
+Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
+Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0% (i.e., its cross-linguistic percentile is zero).
+
+Additionally, we estimated frequency of the phones present  in the target items in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^1] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+
+[^1]: We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.
+
 ### Procedure
@@
 
-We ran children in batches, testing within a handful of hamlet clusters spread across the northeastern region of the island.
+We tested in four different sites spread across the northeastern region of the island, making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from 'spreading' between children through hearsay.
+Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
 Because space availability was limited in different ways from hamlet to hamlet, the places where elicitation happened varied across testing sites.
-We tested in four different sites, only making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from 'spreading' between children through hearsay.
-In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
 More information is available from the online supplementary materials.
-Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
 
@@
 
-The second phase of the experiment involved running the child through the list of test items randomly assigned to them.
+The second phase of the experiment involved going over the list of test items randomly assigned to each child.
 This was done in the same manner as the practice items: the stimulus was played over the headphones, and then the child repeated it aloud.
 NWR studies vary in whether children are allowed to hear and/or repeat the item more than one time.
-We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was alowed to make further attempts if their first attempt was judged erroneous in some way by the assistant.
+We had a fixed procedure for the test items (i.e., the non-practice items) in which the child was allowed to make further attempts if their first attempt was judged erroneous in some way by the assistant.
 The procedure worked as follows: When the child made an attempt, the assistant indicated to the experimenter whether the child's production was correct or not.
-If correct, the experimenter would whisper this note of correct repetition into a separate headset that fed into the right channel of the same Tascam recorder and the group moved on to the next item.
+If correct, the experimenter would whisper this note of correct repetition into a separate headset that fed into the right channel of the same Tascam recorder and we moved on to the next item.
 If not, the child was allowed to try again, with up to five attempts allowed before moving on to the next item.
@@
 
-The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat], then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files.
-The assistant then listened through all these productions, one target item at a time, with productions for each target item presented in a random order across children and repetitions.
-The assistant indicated in a notebook whether each production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English.
-The assistant also provided some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
+The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software [@Praat], then ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to .wav clips.
+The assistant then listened through all these paired target-repetition clips randomized across children and repetitions, grouped such that all the clips of the same target were listened to in succession.
+For each clip, the assistant indicated in a notebook whether the child production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English.
+The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
 
@@
 Notice that the phone-based Levenshtein distance is the complement of the phoneme-level NWR score.
-An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (additions were not attested).
-
-Additionally, we estimated the typological frequency of all phonological segments used in the target items using the PHOIBLE cross-linguistic phonological inventory database [@phoible].
-For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors).
-Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
-Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0% (i.e., its cross-linguistic percentile is zero).
+An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (insertions were not attested in this study).
 
@@
 We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease).
-The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged `r round(mean(inc$age.rounded),2)` years (range `r round(min(inc$age.rounded),2)`-`r round(max(inc$age.rounded),2)` years).
+The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).
+In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
+These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
 
-Included children's ages ranged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).
-We included `r sum(inc$Sex=="B")` boys and `r sum(inc$Sex=="G")` girls.
-There were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^1] Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^2] In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
-These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
-
-[^1]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+[^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
 While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
@@
 
-[^2]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
+[^3]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
 
@@
 
-We then analyzed variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+Turning to our first research question, we  analyzed variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
 To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
@@
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^3] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
 Item-level average corpus-based phone frequencies were correlated with the item-level average cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, p `r pval.display(round(freq_cor$p.value,3))`).
@@
 Error"]`, p `r pval.display(both_coeff["avg_fr_cor_ty_s","Pr(>|z|)"])`) was not.
-
-[^3]: We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.
 

### 2021-01-19 manuscript.tex
midway through methods, check for N items in progress

@@
 \item
-  pick up from \enquote{TODO fix this para and set up comparison with Tsi} and p.~9 of print out
-\item
   add a table with sample studies \& languages \& their characteristiscs
@@
 
-\textbf{TODO fix this para and set up comparison with Tsi} We add some final observations that will help us integrate this study to the broader investigation of NWR across cultures.
-Direct speech to children under 3;0 is relatively infrequent throughout the day and comes primarily from women (Casillas et al., 2020), though shared caregiving practices ensure that children frequently hear the voices of men and other children (Casillas et al., 2020).
-, increasing the relative proportion of linguistic input from other children
+We add some observations that will help us integrate this study to the broader investigation of NWR across cultures.
+As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia et al., 2020).
+However, Cristia et al.
+(2020) also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
+The Yélî community can help us bring further light into this question because direct speech to children under 3;0 is relatively infrequent in this community too (Casillas et al., 2020).
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour ({\textbf{???}}), and Yélî children about 3.7 minutes per hour (Casillas et al., 2020).
+Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in Cristia et al.
+(2020).
 
@@
 In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
-For instance, in English, a response of {[}bilig{]} or {[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length.
+For instance, in English, a response of {[}bilig{]} or {[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
 There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
@@
 
+Throughout these analyses and in the Discussion, we will also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.
+
 We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given their variability across studies.
-To illustrate this issue, we portray in Figure \ref{fig:fig-prevlit} studies in which children's NWR scores were gathered between 4 and 12 years of age, and reported separately for items that are relatively short (1-2 syllables) versus longer items (3-4 syllables).
-As discussed above, the effect of stimulus length is minuscule among Italian children (Piazzalunga et al., 2019), but considerable among Arabic children (Jaber-Awida, 2018).
-Even the effect of age is unstable in this sample of studies.
-Whereas it is quite clear that children's NWR scores increase in the Italian data (Piazzalunga et al., 2019), age effects are less stable among Tsimane' children (Cristia et al., 2020).
+However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given the previously discussed variability across studies.
 Therefore, all analyses in the present study are descriptive and should be considered exploratory.
@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 44 items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 39 items distributed across children.
 
@@
 Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \enquote{yî} /yɯ/; identified as \enquote{yi} /yi/, \emph{tree}).
-This item is removed from the analyses below.
+Additionally, an error was made when preparing files for annotation, resulting in two items being merged (\enquote{tpâ} /tpɑ/ and \enquote{tp:a} /tpæ̃/).
+These three problematic items are not described here, and removed from the analyses below.
 
-The final list includes: three practice items; 20 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
+The final list includes: three practice items; 18 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
 
@@
 \addlinespace
- &  & tp:a & tp̃æ &  &  &  &  &  & \\
- &  & tpâ & tpɑ &  &  &  &  &  & \\
  &  & tpê & tpə &  &  &  &  &  & \\
@@
 
+To inform our analyses, we estimated the typological frequency of all phonological segments present in the target items using the PHOIBLE cross-linguistic phonological inventory database (Moran \& McCloy, 2019).
+For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors).
+Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
+Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0\% (i.e., its cross-linguistic percentile is zero).
+
+Additionally, we estimated frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+
 \hypertarget{procedure}{%
@@
 
-We ran children in batches, testing within a handful of hamlet clusters spread across the northeastern region of the island.
+We tested in four different sites spread across the northeastern region of the island, making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from \enquote{spreading} between children through hearsay.
+Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
 Because space availability was limited in different ways from hamlet to hamlet, the places where elicitation happened varied across testing sites.
-We tested in four different sites, only making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from \enquote{spreading} between children through hearsay.
-In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda.
 More information is available from the online supplementary materials.
-Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
 
@@
 
-The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software (Boersma \& Weenink, 2020), then wrote and ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to short .wav files.
-The assistant then listened through all these productions, one target item at a time, with productions for each target item presented in a random order across children and repetitions.
-The assistant indicated in a notebook whether each production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English.
-The assistant also provided some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
+The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software (Boersma \& Weenink, 2020), then ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to .wav clips.
+The assistant then listened through all these paired target-repetition clips randomized across children and repetitions, grouped such that all the clips of the same target were listened to in succession.
+For each clip, the assistant indicated in a notebook whether the child production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English.
+The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
 
@@
 Notice that the phone-based Levenshtein distance is the complement of the phoneme-level NWR score.
-An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (additions were not attested).
-
-Additionally, we estimated the typological frequency of all phonological segments used in the target items using the PHOIBLE cross-linguistic phonological inventory database (Moran \& McCloy, 2019).
-For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors).
-Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
-Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0\% (i.e., its cross-linguistic percentile is zero).
+An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (insertions were not attested in this study).
 
@@
 We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5), or were 13 years old or older (N=2; we tested these teenagers to put younger children at ease).
-The remaining 40 children (14 girls) were aged 6.98 years (range 3.92-11.03 years).
-
-Included children's ages ranged from 3 to 10 years (M = 6.50 years, SD = 1.50 years).
-We included 26 boys and 14 girls.
-There were 34 only exposed to Yélî Dnye at home, 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+The remaining 40 children (14 girls) were aged from 3 to 10 years (M = 6.50 years, SD = 1.50 years).
+In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
+These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
+Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
+We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 only exposed to Yélî Dnye at home, 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
 While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
-Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.} Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
-We then record the number of years entailed by having completed that level under ideal conditions.} In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
-These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.}
 
@@
 
-We then analyzed variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
+Turning to our first research question, we analyzed variation in whole-item NWR scores as a function of the average frequency with which sounds composing individual target words are found in languages over the world.
 To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
@@
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
-We estimated frequency of the phones present in the stimuli in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out this analysis using token phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the main text, with type frequencies.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
 Item-level average corpus-based phone frequencies were correlated with the item-level average cross-linguistic phone frequencies (r(27)=0.50, p \textless{} 0.01).

### 2021-01-19 manuscript.Rmd
fixed errors with N of items, results section started

@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 39 items distributed across children.
-
-A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 1- to 4-syllable sequences.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 40 final items distributed across children.
+
+A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 2- to 4-syllable sequences.
 These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
@@
 
-The final list includes: three practice items; 18 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
+The final list includes three practice items and 40 test items (across infants): 16 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
 
@@
 
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$avg_fr_s),mean)->xl_fr_tab
-colnames(xl_fr_tab)<-c("xlf","Word")
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,trnsc.inc$avg_fr_s),mean)->xl_fr_tab
+colnames(xl_fr_tab)<-c("target","xlf","Word")
 xlfr_r=cor.test(xl_fr_tab$xlf,xl_fr_tab$Word)
@@
 Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \@ref(fig:fig-xling-freq).
-Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.
+Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.
 
@@
 
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$avg_fr_cor_ty_s),mean)->cor_fr_tab
-colnames(cor_fr_tab)<-c("corf","Word")
-
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,trnsc.inc$avg_fr_cor_ty_s),mean)->cor_fr_tab
+colnames(cor_fr_tab)<-c("target","corf","Word")
+
+#correlations at the phone level
 read.delim("segments_with_cor_freq.txt", encoding="UTF-8")->segments
-freq_cor=cor.test(segments$pc,log(segments$freq_corpus_types))
-```
-
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
-Item-level average corpus-based phone frequencies were correlated with the item-level average cross-linguistic phone frequencies (r(`r freq_cor$parameter`)=`r round(freq_cor$estimate,3)`, p `r pval.display(round(freq_cor$p.value,3))`).
-Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = `r round(cor.test(cor_fr_tab$corf,cor_fr_tab$Word)$estimate,3)`.
+freq_cor_phone=cor.test(segments$pc,log(segments$freq_corpus_types))
+
+#correlations at the item level
+aggregate(trnsc.inc$avg_fr_cor_ty,by=list(trnsc.inc$target),mean)->corpus
+aggregate(trnsc.inc$avg_fr,by=list(trnsc.inc$target),mean)->xling
+freq_cor_item=cor.test(xling$x,corpus$x)
+
+#correlation between item level avg frequency & nwr 
+cor_ty_nwr=cor.test(cor_fr_tab$corf,cor_fr_tab$Word)
+```
+
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
+Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies (r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`); and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies (r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`).
+Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r round(cor_ty_nwr$estimate,3)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`.
 Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.

### 2021-01-19 manuscript.tex
fixed errors with N of items, results section started

@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 39 items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 40 final items distributed across children.
 
-A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/, /n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 1- to 4-syllable sequences.
+A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/, /n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 2- to 4-syllable sequences.
 These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
@@
 
-The final list includes: three practice items; 18 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
+The final list includes three practice items and 40 test items (across infants): 16 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
 
@@
 As shown in Figure \ref{fig:fig-first-vs-others}, participants' mean word-level scores became more heterogeneous in subsequent repetitions.
-Surprisingly, whole-item NWR scores for subsequent repetitions (M = 40, SD = 26)
-were on average lower than first ones (M = 64, SD = 15), t(38) = 6.28, p \textless{} 0.001; Cohen's d = 1.14).
+Surprisingly, whole-item NWR scores for subsequent repetitions (M = 40, SD = 28)
+were on average lower than first ones (M = 65, SD = 15), t(38) = 5.89, p \textless{} 0.001; Cohen's d = 1.13).
 Given uncertainty in whether previous work used first or all repetitions, and given that score here declined and became more heterogeneous in subsequent repetitions, we focus the remainder of our analyses only on first repetitions, with the exception of qualitative analyses of substitutions.
@@
 Taking into account only the first attempts, we derived overall averages across all items.
-The overall NWR score was M = 64\% (SD = 15\%), Cohen's d = 4.36.
-The phoneme-based normalized Levenshtein distance was M = 22\% (SD = 9\%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.
+The overall NWR score was M = 65\% (SD = 15\%), Cohen's d = 4.39.
+The phoneme-based normalized Levenshtein distance was M = 21\% (SD = 9\%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.
 
 We also looked into the frequency with which mispronunciations resulted in real words.
-In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: 64\%.
+In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: 63\%.
 This type of analysis is seldom reported.
@@
 
-We could include 850 observations, from 40 children producing in any given trial one of 41 potential target words.
-The analysis revealed a main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:fig-xling-freq}.
-Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(32) = 0.61.
+We could include 826 observations, from 40 children producing in any given trial one of 40 potential target words.
+The analysis revealed a main effect of age (ß = 0.35, SE ß = 0.13, p \textless{} 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.78, SE ß = 0.19, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:fig-xling-freq}.
+Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(38) = 0.54.
 
@@
 
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat.
-Item-level average corpus-based phone frequencies were correlated with the item-level average cross-linguistic phone frequencies (r(27)=0.50, p \textless{} 0.01).
-Moreover, averaging across items and participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r = 0.46.
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
+Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies (r(27)=0.50, p \textless{} 0.01); and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies (r(38)=0.73, p \textless{} 0.001).
+Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(38)=0.43, p \textless{} 0.01.
 Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
 As before, the model contained random slopes for both child ID and target.
-In this model, both cross-linguistic phone frequency (ß = 0.82, SE ß = 0.27, p \textless{} 0.01) and age (ß = 0.33, SE ß = 0.13, p = 0.01) were significant predictors of whole-item NWR scores, but corpus phone frequency (ß = 0.01, SE ß = 0.25, p = 0.98) was not.
+In this model, both cross-linguistic phone frequency (ß = 0.78, SE ß = 0.27, p \textless{} 0.01) and age (ß = 0.35, SE ß = 0.13, p \textless{} 0.01) were significant predictors of whole-item NWR scores, but corpus phone frequency (ß = 0.00, SE ß = 0.25, p = 0.99) was not.
 
@@
 Deletions were very rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were 884 cases of substitutions, \textasciitilde{}17.83 of the 4957 phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (\textasciitilde{}97.90 of unmatched phones).
+We therefore focus our qualitative description here on substitutions: There were 820 cases of substitutions, \textasciitilde{}16.95 of the 4839 phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (\textasciitilde{}97.74 of unmatched phones).
 
@@
 \midrule
-Nasal Target & 134 & 58 & 23 & 62.3 & 20.9 & 7.7\\
+Nasal Target & 100 & 39 & 17 & 64.1 & 17.7 & 7.1\\
 Oral Target & 1992 & 52 & 205 & 88.6 & 2.2 & 8.8\\
@@
 We approached our question regarding potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels, distinguishing errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (but were therefore a quality error), shown in Table \ref{tab:tab-v}.
-We found that errors involving nasal vowel targets were more common than those involving oral vowels (37.70 versus 11.40).
+We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.90 versus 11.40).
 Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
@@
 \midrule
-Complex Target & 277 & 250 & 55 & 47.6 & 40 & 8.2\\
+Complex Target & 257 & 218 & 48 & 49.1 & 38 & 7.9\\
 Simple Target & 1425 & 2 & 120 & 92.1 & 0 & 7.3\\
@@
 
-For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.} Results, shown in Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving oral vowels (52.40 versus 7.90).
+For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.} Results, shown in Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving oral vowels (50.90 versus 7.90).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 \midrule
-1 syll & 47 (22) & 41 (17)\\
+1 syll & 48 (22) & 40 (18)\\
 2 syll & 79 (22) & 8 (9)\\
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
-As shown in Figure \ref{fig:Fig2-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (6,014.70) = 0.44, p \textless{} 0.01).
-In contrast, there was no clear association between NWR scores and sex (Welch t (27.56) = -0.29, p = 0.77), birth order (data missing for 15 children, rho = (3,441.90) = -0.18, p = 0.39), or maternal education (rho ( 9,594.40) = 0.10, p = 0.54).
+As shown in Figure \ref{fig:Fig2-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (5,649.08) = 0.47, p \textless{} 0.01).
+In contrast, there was no clear association between NWR scores and sex (Welch t (27.33) = -0.60, p = 0.56), birth order (data missing for 15 children, rho = (3,502.90) = -0.20, p = 0.33), or maternal education (rho ( 9,628.60) = 0.10, p = 0.55).
 
@@
 
-In our sample of 40 children between ages 3 and 10, the average item-level score was 64\%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \ref{fig:prevlit-fig}).
-Segment-level analyses revealed that repetitions recovered an average of 78\% of the segments in the original stimulus, indicating that inexact repetitions often constituted \emph{mostly} accurate reroductions of the target.
-Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of 22\% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
+In our sample of 40 children between ages 3 and 10, the average item-level score was 65\%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \ref{fig:prevlit-fig}).
+Segment-level analyses revealed that repetitions recovered an average of 79\% of the segments in the original stimulus, indicating that inexact repetitions often constituted \emph{mostly} accurate reroductions of the target.
+Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of 21\% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
 Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.

### 2021-01-19 wrangling.R

@@ -140,9 +140,14 @@ dim(trnsc)
 #REMOVING ALL THE "YI"
 trnsc[trnsc$target!="yi",]->trnsc
 
+
+#REMOVING ALL THE "TPA"
+trnsc[trnsc$target!="tpa",]->trnsc
+trnsc[trnsc$target!="tp_a",]->trnsc
+
 #remove practice items
 trnsc=trnsc[trnsc$type !="practice",]
-dim(trnsc) #2020-09-16 1302
+dim(trnsc) #2021-01-19 1243
 
 ## PHONOLOGIZE AND SCORE
 

### 2021-01-20 manuscript.Rmd
added analysis of phone error rates, still in results

@@
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
-Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies (r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`); and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies (r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`).
+Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`].
 Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r round(cor_ty_nwr$estimate,3)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`.
@@
 
-Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
-To answer this question, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
-Deletions were very rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
+We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts so as to base our generalizations on more data.
+There were no cases of insertion, and deletions were very rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
 We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] )*100,2)` of unmatched phones).
+To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
 
@@
 
-We approached our question regarding potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels, distinguishing errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (but were therefore a quality error), shown in Table \@ref(tab:tab-v).
+We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels.
+The nasal vowels in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`%);
+whereas oral vowels in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("short_oral","long_oral")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("short_oral","long_oral")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("short_oral","long_oral")],na.rm=T)`%).
+As noted above, type frequency within the language is correlated with cross-linguistic frequency, and thus these two types of sounds also differ in the former: Their type frequencies in Yélî Dnye are: nasal vowels ~`r mean(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰  (range `r min(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰) versus oral ~`r mean(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰ (range `r min(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰).
+
+We distinguished errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (and were therefore a quality error), shown in Table \@ref(tab:tab-v).
 We found that errors involving nasal vowel targets were more common than those involving oral vowels (`r round(100-mytab["Nasal Target","% Corr."],1)` versus `r round(100-mytab["Oral Target","% Corr."],1)`).
 Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
+Note that this analysis does not tell us whether cross-linguistic or within-language frequency is the best predictor, an issue to which we return below.
 
@@
 
-For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.] Results, shown in Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving oral vowels (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
+For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.]
+The complex consonants in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("complex")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("complex")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("complex")],na.rm=T)`%);
+whereas simple consonants in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("simple")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("simple")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("simple")],na.rm=T)`%).
+Again these groups of sounds differ in their frequency within the language.
+Their type frequencies in Yélî Dnye are: complex consonants ~`r mean(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰  (range `r min(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("complex")],na.rm=T)*10`‰) versus simple consonants ~`r mean(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰ (range `r min(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("simple")],na.rm=T)*10`‰).
+
+Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving oral vowels (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
+
+```{r fig-pc-phone}
+mat_pc2=data.frame(mat_pc)
+names(mat_pc2)<-c("ortho","NWR")
+mat_pc2=merge(mat_pc2,segments)
+
+mat_pc2=mat_pc2[complete.cases("NWR", "pc", "freq_corpus_types"),]
+
+dim(mat_pc2)
+
+#plot(mat_pc2$NWR~mat_pc2$pc)
+#plot(mat_pc2$NWR~mat_pc2$freq_corpus_types)
+
+phone_cor_xl=cor.test(mat_pc2$NWR,mat_pc2$pc,method="s")
+phone_cor_wl=cor.test(mat_pc2$NWR,mat_pc2$freq_corpus_types,method="s")
+
+```
+
+To address whether errors were better predicted by cross-linguistic or within-language frequency, we calculated a proportion of productions that were correct for each phone (regardless of the type of error or the substitution pattern).
+Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other.
+Although a direct test is missing, the correlation with cross-linguistic frequency [r(`r phone_cor_xl$parameter`)=`r round(phone_cor_xl$estimate,3)`, p `r pval.display(round(phone_cor_xl$p.value,3))`]  was greater than that with within-language frequency [r(`r phone_cor_wl$parameter`)=`r round(phone_cor_wl$estimate,3)`, p `r pval.display(round(phone_cor_wl$p.value,3))`].
 

### 2021-01-20 manuscript.tex
added analysis of phone error rates, still in results

@@
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
-Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies (r(27)=0.50, p \textless{} 0.01); and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies (r(38)=0.73, p \textless{} 0.001).
+Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies {[}r(27)=0.50, p \textless{} 0.01{]}; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies {[}r(38)=0.73, p \textless{} 0.001{]}.
 Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(38)=0.43, p \textless{} 0.01.
@@
 
-Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
-To answer this question, we investigated patterns of deletion and substitution (as insertion was not attested), this time looking at all attempts so as to base our generalizations on more data.
-Deletions were very rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
+We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts so as to base our generalizations on more data.
+There were no cases of insertion, and deletions were very rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
 We therefore focus our qualitative description here on substitutions: There were 820 cases of substitutions, \textasciitilde{}16.95 of the 4839 phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (\textasciitilde{}97.74 of unmatched phones).
+To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
 
@@
 
-We approached our question regarding potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels, distinguishing errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (but were therefore a quality error), shown in Table \ref{tab:tab-v}.
+We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels.
+The nasal vowels in our stimuli occur in \textasciitilde{}1.40\% of languages' phonologies (range 0\% to 3\%);
+whereas oral vowels in our stimuli occur in \textasciitilde{}31.55\% of languages' phonologies (range 3\% to 92\%).
+As noted above, type frequency within the language is correlated with cross-linguistic frequency, and thus these two types of sounds also differ in the former: Their type frequencies in Yélî Dnye are: nasal vowels \textasciitilde{}0.03‰ (range 0.00‰ to 0.05‰) versus oral \textasciitilde{}0.23‰ (range 0.02‰ to 0.76‰).
+
+We distinguished errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (and were therefore a quality error), shown in Table \ref{tab:tab-v}.
 We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.90 versus 11.40).
 Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
+Note that this analysis does not tell us whether cross-linguistic or within-language frequency is the best predictor, an issue to which we return below.
 
@@
 
-For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.} Results, shown in Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving oral vowels (50.90 versus 7.90).
+For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.}
+The complex consonants in our stimuli occur in \textasciitilde{}17.33\% of languages' phonologies (range 0\% to 78\%);
+whereas simple consonants in our stimuli occur in \textasciitilde{}67.62\% of languages' phonologies (range 13\% to 96\%).
+Again these groups of sounds differ in their frequency within the language.
+Their type frequencies in Yélî Dnye are: complex consonants \textasciitilde{}0.04‰ (range 0.00‰ to 0.10‰) versus simple consonants \textasciitilde{}0.32‰ (range 0.06‰ to 0.55‰).
+
+Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving oral vowels (50.90 versus 7.90).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
+
+\begin{verbatim}
+## [1] 30 15
+\end{verbatim}
+
+To address whether errors were better predicted by cross-linguistic or within-language frequency, we calculated a proportion of productions that were correct for each phone (regardless of the type of error or the substitution pattern).
+Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other.
+Although a direct test is missing, the correlation with cross-linguistic frequency {[}r()=0.76, p \textless{} 0.001{]} was greater than that with within-language frequency {[}r()=0.45, p = 0.05{]}.
 

### 2021-01-20 manuscript.Rmd
started discussion

@@
 
-For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.]
+For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]).
+These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.
+This result suggests that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.]
 The complex consonants in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("complex")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("complex")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("complex")],na.rm=T)`%);
@@
 
-Next, we inspected whether NWR scores varied as a function of word length (Table \@ref(tab:tablength)).
-Participants scored much lower on monosyllables than on non-words of other lengths.
-This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may indicate that they are hard to produce or perceive.
-Setting monosyllables aside, we observed the typical pattern of lower scores for longer items only for the whole-item scoring, and even there differences were rather small.
-In a generalized binomial mixed model excluding monosyllables, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing, in any given trial, one of `r summary(len_lm)$ngrps["target"]` potential target words.
+We next turned to our second research question by inspecting whether NWR scores varied as a function of word length (Table \@ref(tab:tablength)).
+In this section and all subsequent ones, we only look at first attempts, for the reasons discussed previously.
+Additionally, we noticed that participants scored much lower on monosyllables than on non-words of other lengths.
+This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may be harder to produce or perceive, as suggested by our previous analyses of NWR scores as a function of cross-linguistic phone frequency and error patterns.
+Therefore, we set monosyllables aside for this analysis.
+
+We observed the typical pattern of lower scores for longer items only for the whole-item scoring, and even there differences were rather small.
+In a generalized binomial mixed model excluding monosyllables, we included `r summary(len_lm)$devcomp$dims["N"]` observations, from `r summary(len_lm)$ngrps["file"]` children producing, in any given trial, one of `r summary(len_lm)$ngrps["target"]` (non-monosyllabic) potential target words.
 The analysis revealed a positive effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
@@
 
-```{r Fig2-scores-by-age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-item scores for individual participants as a function of age and sex (purple = boys, orange = girls)."}
-
-plot(overall$Word~overall$Age,pch=4,cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","darkmagenta","chocolate1"))
-
-```
-
-Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
+```{r Fig2-scores-by-age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-item scores for individual participants as a function of age and sex (purple crosses = boys, orange circles = girls)."}
+
+plot(overall$Word~overall$Age,pch=ifelse(overall$Sex=="B",4,1),cex=2,lty=3,ylab="NWR scores (% words)",xlab="Age (years)",col=ifelse(overall$Sex=="B","darkmagenta","chocolate1"))
+
+```
+
+Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question.
 As shown in Figure \@ref(fig:Fig2-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-While the study, in itself, advances NWR research by demonstrating the successful application of this method in a rural, non-industrialized population [see also @cristia2020infant], the results also inform the larger body of NWR work regarding the influence of stimulus length, segmental frequency, age, and developmental context, where previous findings have often been mixed.
-
-In our sample of `r nrow(overall)` children between ages `r min(overall$Age)` and `r max(overall$Age)`, the average item-level score was `r round(mean(overall$Word))`%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \@ref(fig:prevlit-fig)).
-Segment-level analyses revealed that repetitions recovered an average of `r round(mean(overall$Phoneme))`% of the segments in the original stimulus, indicating that inexact repetitions often constituted _mostly_ accurate reroductions of the target.
-Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of `r round(mean(overall$NLD))`% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
-Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
-Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.
-
-### Item complexity
-
-We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item and its average segmental frequency.
+We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
+An additional overarching goal was to revisit this task, which is very commonly used to document phonological development, particularly in children raised in urban settings, and who are learning IndoEuropean languages.
+We will thus consider our results not only in light of previous work, but also with attention to potential linguistic and population differences.
+
+### Associations between NWR and cross-linguistic frequency
+
+Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
+Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
+Typologically common sounds are associated with higher performance on a handful of other tasks (**REFS -- M2A: Alex, I added this based on your note, where it sounded like you had some particular studies in mind?**) though to our knowledge this has not yet been tested with non-word repetition.
+Regarding Yélî Dnye in particular, the phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
+We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
+Indeed we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often.
+This effect is large, with a magnitude more than twice the size of the effect of participant age.
+This significant effect remains even once also accounting for the frequencies of these segments in Yélî Dnye child-directed speech, which are correlated with their typological frequencies.
+In sum, typological frequency effects, which have been found in other measurements of phonological processing.
+appear to strongly affect NWR performance, and do not appear mitigated by language-specific pressure to make finer-grained differences earlier in development.
+
+With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.
+oral vowels and simple vs.
+complex consonants, revealed no remarkable bias in error types.
+That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
+
+### Item Length
+
+We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item.
 Based on previous work, we had predicted that children would have higher NWR scores for shorter items.
@@
 
-Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
-Typologically common sounds are associated with higher performance on a handful of other tasks (**REFS -- M2A: Alex, I added this based on your note, where it sounded like you had some particular studies in mind?**) though to our knowledge this has not yet been tested with non-word repetition.
-Regarding Yélî Dnye in particular, the phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
-We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
-Indeed we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often.
-This effect is large, with a magnitude more than twice the size of the effect of participant age.
-This significant effect remains even once also accounting for the frequencies of these segments in Yélî Dnye child-directed speech, which are correlated with their typological frequencies.
-In sum, typological frequency effects, which have been found in other measurements of phonological processing.
-appear to strongly affect NWR performance, and do not appear mitigated by language-specific pressure to make finer-grained differences earlier in development.
-
-With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.
-oral vowels and simple vs.
-complex consonants, revealed no remarkable bias in error types.
-That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
-
 ### Individual differences
@@
 ### Other observations
+
+In our sample of `r nrow(overall)` children between ages `r min(overall$Age)` and `r max(overall$Age)`, the average item-level score was `r round(mean(overall$Word))`%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \@ref(fig:prevlit-fig)).
+Segment-level analyses revealed that repetitions recovered an average of `r round(mean(overall$Phoneme))`% of the segments in the original stimulus, indicating that inexact repetitions often constituted _mostly_ accurate reroductions of the target.
+Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of `r round(mean(overall$NLD))`% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
+Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
+Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.
 

### 2021-01-20 manuscript.tex
started discussion

@@
 
-For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}); as these data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English, we take these segmental substitutions faithfully as an indication that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.}
+For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}).
+These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.
+This result suggests that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.}
 The complex consonants in our stimuli occur in \textasciitilde{}17.33\% of languages' phonologies (range 0\% to 78\%);
@@
 
-Next, we inspected whether NWR scores varied as a function of word length (Table \ref{tab:tablength}).
-Participants scored much lower on monosyllables than on non-words of other lengths.
-This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may indicate that they are hard to produce or perceive.
-Setting monosyllables aside, we observed the typical pattern of lower scores for longer items only for the whole-item scoring, and even there differences were rather small.
-In a generalized binomial mixed model excluding monosyllables, we included 479 observations, from 40 children producing, in any given trial, one of 24 potential target words.
+We next turned to our second research question by inspecting whether NWR scores varied as a function of word length (Table \ref{tab:tablength}).
+In this section and all subsequent ones, we only look at first attempts, for the reasons discussed previously.
+Additionally, we noticed that participants scored much lower on monosyllables than on non-words of other lengths.
+This is likely due to the fact that the majority of monosyllables were designed to include sounds that are rare in the world's languages, which may be harder to produce or perceive, as suggested by our previous analyses of NWR scores as a function of cross-linguistic phone frequency and error patterns.
+Therefore, we set monosyllables aside for this analysis.
+
+We observed the typical pattern of lower scores for longer items only for the whole-item scoring, and even there differences were rather small.
+In a generalized binomial mixed model excluding monosyllables, we included 479 observations, from 40 children producing, in any given trial, one of 24 (non-monosyllabic) potential target words.
 The analysis revealed a positive effect of age (ß = 0.56, SE ß = 0.14, p \textless{} 0.001) and a negative but non-significant estimate for target length in number of syllables (ß = -0.15, SE ß = 0.33, p = 0.65).
@@
 \includegraphics{manuscript_files/figure-latex/Fig2-scores-by-age-1.pdf}
-\caption{\label{fig:Fig2-scores-by-age}NWR whole-item scores for individual participants as a function of age and sex (purple = boys, orange = girls).}
+\caption{\label{fig:Fig2-scores-by-age}NWR whole-item scores for individual participants as a function of age and sex (purple crosses = boys, orange circles = girls).}
 \end{figure}
 
-Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals.
+Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question.
 As shown in Figure \ref{fig:Fig2-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (5,649.08) = 0.47, p \textless{} 0.01).
@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-While the study, in itself, advances NWR research by demonstrating the successful application of this method in a rural, non-industrialized population (see also Cristia et al., 2020), the results also inform the larger body of NWR work regarding the influence of stimulus length, segmental frequency, age, and developmental context, where previous findings have often been mixed.
+We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
+An additional overarching goal was to revisit this task, which is very commonly used to document phonological development, particularly in children raised in urban settings, and who are learning IndoEuropean languages.
+We will thus consider our results not only in light of previous work, but also with attention to potential linguistic and population differences.
 
-In our sample of 40 children between ages 3 and 10, the average item-level score was 65\%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \ref{fig:prevlit-fig}).
-Segment-level analyses revealed that repetitions recovered an average of 79\% of the segments in the original stimulus, indicating that inexact repetitions often constituted \emph{mostly} accurate reroductions of the target.
-Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of 21\% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
-Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
-Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.
+\hypertarget{associations-between-nwr-and-cross-linguistic-frequency}{%
+\subsubsection{Associations between NWR and cross-linguistic frequency}\label{associations-between-nwr-and-cross-linguistic-frequency}}
 
-\hypertarget{item-complexity}{%
-\subsubsection{Item complexity}\label{item-complexity}}
-
-We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item and its average segmental frequency.
-Based on previous work, we had predicted that children would have higher NWR scores for shorter items.
-That said, previous work has shown both very small (Piazzalunga et al., 2019) and very large (Cristia et al., 2020; Jaber-Awida, 2018) effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10\% of the remaining words.
-Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms \textbf{M2A: Alex I'm going off your note here (\enquote{Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.}").
-I don't have a reference for this, can you please finish the thought or nix this bit?}.
-Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
-While indeed NWR scores were overall lower for longer items (e.g., see Figure \ref{fig:prevlit-fig}), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
-In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length.
-First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon.
-Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s (2019) study on Italian and as borne out in the current dataset once controlling for other factors.
-
+Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
 Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
@@
 That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
+
+\hypertarget{item-length}{%
+\subsubsection{Item Length}\label{item-length}}
+
+We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item.
+Based on previous work, we had predicted that children would have higher NWR scores for shorter items.
+That said, previous work has shown both very small (Piazzalunga et al., 2019) and very large (Cristia et al., 2020; Jaber-Awida, 2018) effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10\% of the remaining words.
+Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms \textbf{M2A: Alex I'm going off your note here (\enquote{Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.}").
+I don't have a reference for this, can you please finish the thought or nix this bit?}.
+Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
+While indeed NWR scores were overall lower for longer items (e.g., see Figure \ref{fig:prevlit-fig}), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
+In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length.
+First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon.
+Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s (2019) study on Italian and as borne out in the current dataset once controlling for other factors.
 
@@
 \subsubsection{Other observations}\label{other-observations}}
+
+In our sample of 40 children between ages 3 and 10, the average item-level score was 65\%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \ref{fig:prevlit-fig}).
+Segment-level analyses revealed that repetitions recovered an average of 79\% of the segments in the original stimulus, indicating that inexact repetitions often constituted \emph{mostly} accurate reroductions of the target.
+Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of 21\% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
+Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
+Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.
 

### 2021-01-21 manuscript.Rmd
fixed table errors, first pass at discussion

@@
 1. be accurate on what hasn't been done x-ling speaking
+1. try to find data on length distribution across languages to see if this fits length differences
+1. to the above, add sylable characteristics, inventory size, inventory compactness
+1. fix camila ref
 
@@
 mytab=rbind(
-  #OK --                      nasal change --                 qual change
-  cbind(sum(match_tab[nasalVs]),sum(sub_tab[nasalVs,oralVs]),sum(sub_tab[nasalVs,nasalVs])), #nasal target
-  cbind(sum(match_tab[oralVs]),sum(sub_tab[oralVs,nasalVs]),sum(sub_tab[oralVs,oralVs])) #oral target
+  #OK --                      deleted --            nasal change --                 qual change
+  cbind(sum(match_tab[nasalVs]),sum(del_tab[nasalVs]),sum(sub_tab[nasalVs,oralVs]),sum(sub_tab[nasalVs,nasalVs])), #nasal target
+  cbind(sum(match_tab[oralVs]),sum(del_tab[oralVs]),sum(sub_tab[oralVs,nasalVs]),sum(sub_tab[oralVs,oralVs])) #oral target
 )
 mytab=data.frame(mytab)
-mytab$pc_corr=round(mytab[,1]/rowSums(mytab)*100,1)
-mytab$pc_nas_err=round(mytab[,2]/rowSums(mytab)*100,1)
-mytab$pc_qual_err=round(mytab[,3]/rowSums(mytab)*100,1)
+total=rowSums(mytab)
+mytab$pc_corr=round(mytab[,1]/total*100,1)
+mytab$pc_dem=round(mytab[,2]/total*100,1)
+mytab$pc_nas_err=round(mytab[,3]/total*100,1)
+mytab$pc_qual_err=round(mytab[,4]/total*100,1)
 rownames(mytab)<-c("Nasal Target", "Oral Target")
-colnames(mytab)<-c("Correct","Nasal Err.","Qual. Err","% Corr.","% Nasal Err.","% Qual Err.")
+colnames(mytab)<-c("Corr.","Del.","Nasal Err.","Qual. Err","% Corr.","% Del.","% Nasal Err.","% Qual Err.")
 
@@
   "latex", booktabs = T,
-  caption="Number (and percent) of vowel targets that were correctly repeated or substituted, as a function of vowel type, and whether the error resulted in a nasality change (Nasal Err.) or only a quality change (Qual.
-Err.) Deletion errors are not shown.")
+  caption="Number (and percent) of vowel targets that were correctly repeated (Corr.), deleted (Del.), or substituted, as a function of vowel type, and whether the error resulted in a nasality change (Nasal Err.) or only a quality change (Qual.
+Err.)")
 
@@
 mytab=rbind(
-  #OK --                      complexity change --                 other change
-  cbind(sum(match_tab[complexCs]),sum(sub_tab[complexCs,simplexCs]),sum(sub_tab[complexCs,complexCs])), #nasal target
-  cbind(sum(match_tab[simplexCs]),sum(sub_tab[simplexCs,complexCs]),sum(sub_tab[simplexCs,simplexCs])) #oral target
+  #OK --                     deleted --        complexity change --                 other change
+  cbind(sum(match_tab[complexCs]),sum(del_tab[complexCs]),sum(sub_tab[complexCs,simplexCs]),sum(sub_tab[complexCs,complexCs])), #nasal target
+  cbind(sum(match_tab[simplexCs]),sum(del_tab[simplexCs]),sum(sub_tab[simplexCs,complexCs]),sum(sub_tab[simplexCs,simplexCs])) #oral target
 )
 mytab=data.frame(mytab)
-mytab$pc_corr=round(mytab[,1]/rowSums(mytab)*100,1)
-mytab$pc_comp_err=round(mytab[,2]/rowSums(mytab)*100)
-mytab$pc_oth_err=round(mytab[,3]/rowSums(mytab)*100,1)
+total=rowSums(mytab)
+mytab$pc_corr=round(mytab[,1]/total*100,1)
+mytab$pc_del=round(mytab[,2]/total*100,1)
+mytab$pc_comp_err=round(mytab[,3]/total*100,1)
+mytab$pc_oth_err=round(mytab[,4]/total*100,1)
 rownames(mytab)<-c("Complex Target", "Simple Target")
-colnames(mytab)<-c("Correct","Compl. Err.","Other Err","% Corr.","% Compl. Err.","% Other Err.")
+colnames(mytab)<-c("Corr.","Del.","Cmpl Err.","Othr Err.","% Corr.","% Del","% Cmpl Err.","% Othr Err.")
 
@@
   "latex", booktabs = T,
-  caption="Number (and percent) of consonant targets that were correctly repeated or substituted, as a function of the complexity of the consonant, and whether the error resulted in a change of complexity (Compl Err.) or not (Other Err.) Deletion errors are not shown.")
+  caption="Number (and percent) of consonant targets that were correctly repeated (Corr.), deleted (Del.), or substituted, as a function of the complexity of the consonant, and whether the error resulted in a change of complexity (Cmpl Err.) or not (Othr Err.)")
 
@@
 
-Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
+**FIX LOGIC IN THIS PARA** Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
 Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
@@
 We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
-Indeed we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often.
+Indeed, we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often.
 This effect is large, with a magnitude more than twice the size of the effect of participant age.
@@
 
-With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.
+**REWRITE THIS PARA** ~With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.
 oral vowels and simple vs.
-complex consonants, revealed no remarkable bias in error types.
-That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
+complex consonants, revealed no remarkable bias in error types.~ That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
 
@@
 Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
-While indeed NWR scores were overall lower for longer items (e.g., see Figure \@ref(fig:prevlit-fig)), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
+While indeed NWR scores were overall lower for longer items (e.g., see Figure \@ref(fig:fig-prevlit)), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
 In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length.
@@
 Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s [-@piazzalunga2019articulatory] study on Italian and as borne out in the current dataset once controlling for other factors.
+**ADD WHAT LANGUAGES WOULD BE IDEAL TO TEST THIS HYPOTHESIS**
 
@@
 On the other hand, these predictors have established effects on other language development measures (**REFS: M2A: Alex go ahead and pick your faves here**).
-So to the extent that NWR scores share causal links to gender-based differences ein development and maternal linguistic input with these other language ouctome measures, we might then expect these factors to appear in NWR data.
-In fact, neither participant gender nor maternal education were correlated with NWR score in the current data.
+So to the extent that NWR scores share causal links to gender-based differences in development and maternal linguistic input with these other language outcome measures, we might then expect these factors to appear in NWR data.
+In fact, participant gender and maternal education correlated with NWR score at about r~.1, which is small.
 
 Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect], presumably because later-born children receive a smaller share of maternal input than their older siblings.
-Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
+Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
 Therefore we expected that any effects of birth order on NWR would be attenuated in this context.
-In line with this prediction, our descriptive analysis showed no correlation between birth order and NWR score.
+In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score.
+However, the effect size was larger than that found for the other factors, at r~.2, and thus we believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
 
@@
 
-In our sample of `r nrow(overall)` children between ages `r min(overall$Age)` and `r max(overall$Age)`, the average item-level score was `r round(mean(overall$Word))`%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \@ref(fig:prevlit-fig)).
-Segment-level analyses revealed that repetitions recovered an average of `r round(mean(overall$Phoneme))`% of the segments in the original stimulus, indicating that inexact repetitions often constituted _mostly_ accurate reroductions of the target.
-Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of `r round(mean(overall$NLD))`% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
+In our sample of `r nrow(overall)` children between ages `r min(overall$Age)` and `r max(overall$Age)`, the average item-level score was `r round(mean(overall$Word))`%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \@ref(fig:fig-prevlit)).
+Segment-based normalized Levenshtein distance  indicate that inexact repetitions often constituted _mostly_ accurate reroductions of the target, with an average of `r round(mean(overall$NLD))`% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
 Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
-Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.
+Given successful (and comparable) execution of this task.
 

### 2021-01-21 manuscript.tex
fixed table errors, first pass at discussion

@@
   be accurate on what hasn't been done x-ling speaking
+\item
+  try to find data on length distribution across languages to see if this fits length differences
+\item
+  to the above, add sylable characteristics, inventory size, inventory compactness
 \end{enumerate}
@@
 
-\caption{\label{tab:tab-v}Number (and percent) of vowel targets that were correctly repeated or substituted, as a function of vowel type, and whether the error resulted in a nasality change (Nasal Err.) or only a quality change (Qual.
-Err.) Deletion errors are not shown.}
+\caption{\label{tab:tab-v}Number (and percent) of vowel targets that were correctly repeated (Corr.), deleted (Del.), or substituted, as a function of vowel type, and whether the error resulted in a nasality change (Nasal Err.) or only a quality change (Qual.
+Err.)}
 \centering
-\begin{tabular}[t]{lrrrrrr}
+\begin{tabular}[t]{lrrrrrrrr}
 \toprule
-  & Correct & Nasal Err. & Qual. Err & \% Corr. & \% Nasal Err. & \% Qual Err.\\
+  & Corr. & Del. & Nasal Err. & Qual. Err & \% Corr. & \% Del. & \% Nasal Err. & \% Qual Err.\\
 \midrule
-Nasal Target & 100 & 39 & 17 & 64.1 & 17.7 & 7.1\\
-Oral Target & 1992 & 52 & 205 & 88.6 & 2.2 & 8.8\\
+Nasal Target & 100 & 0 & 39 & 17 & 64.1 & 0.0 & 25.0 & 10.9\\
+Oral Target & 1992 & 12 & 52 & 205 & 88.1 & 0.5 & 2.3 & 9.1\\
 \bottomrule
@@
 We distinguished errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (and were therefore a quality error), shown in Table \ref{tab:tab-v}.
-We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.90 versus 11.40).
+We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.90 versus 11.90).
 Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
@@
 
-\caption{\label{tab:tab-c}Number (and percent) of consonant targets that were correctly repeated or substituted, as a function of the complexity of the consonant, and whether the error resulted in a change of complexity (Compl Err.) or not (Other Err.) Deletion errors are not shown.}
+\caption{\label{tab:tab-c}Number (and percent) of consonant targets that were correctly repeated (Corr.), deleted (Del.), or substituted, as a function of the complexity of the consonant, and whether the error resulted in a change of complexity (Cmpl Err.) or not (Othr Err.)}
 \centering
-\begin{tabular}[t]{lrrrrrr}
+\begin{tabular}[t]{lrrrrrrrr}
 \toprule
-  & Correct & Compl. Err. & Other Err & \% Corr. & \% Compl. Err. & \% Other Err.\\
+  & Corr. & Del. & Cmpl Err. & Othr Err. & \% Corr. & \% Del & \% Cmpl Err. & \% Othr Err.\\
 \midrule
-Complex Target & 257 & 218 & 48 & 49.1 & 38 & 7.9\\
-Simple Target & 1425 & 2 & 120 & 92.1 & 0 & 7.3\\
+Complex Target & 257 & 0 & 218 & 48 & 49.1 & 0.0 & 41.7 & 9.2\\
+Simple Target & 1425 & 6 & 2 & 120 & 91.8 & 0.4 & 0.1 & 7.7\\
 \bottomrule
@@
 
-Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving oral vowels (50.90 versus 7.90).
+Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving oral vowels (50.90 versus 8.20).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 
-Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
+\textbf{FIX LOGIC IN THIS PARA} Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
 Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
@@
 
-With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.~oral vowels and simple vs.~complex consonants, revealed no remarkable bias in error types.
-That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
+\textbf{REWRITE THIS PARA} \textasciitilde{}With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.~oral vowels and simple vs.~complex consonants, revealed no remarkable bias in error types.\textasciitilde{} That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
 
@@
 Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
-While indeed NWR scores were overall lower for longer items (e.g., see Figure \ref{fig:prevlit-fig}), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
+While indeed NWR scores were overall lower for longer items (e.g., see Figure \ref{fig:fig-prevlit}), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
 In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length.
@@
 Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s (2019) study on Italian and as borne out in the current dataset once controlling for other factors.
+\textbf{ADD WHAT LANGUAGES WOULD BE IDEAL TO TEST THIS HYPOTHESIS}
 
@@
 On the other hand, these predictors have established effects on other language development measures (\textbf{REFS: M2A: Alex go ahead and pick your faves here}).
-So to the extent that NWR scores share causal links to gender-based differences ein development and maternal linguistic input with these other language ouctome measures, we might then expect these factors to appear in NWR data.
-In fact, neither participant gender nor maternal education were correlated with NWR score in the current data.
+So to the extent that NWR scores share causal links to gender-based differences in development and maternal linguistic input with these other language outcome measures, we might then expect these factors to appear in NWR data.
+In fact, participant gender and maternal education correlated with NWR score at about r\textasciitilde{}.1, which is small.
 
 Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019), presumably because later-born children receive a smaller share of maternal input than their older siblings.
-Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
+Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
 Therefore we expected that any effects of birth order on NWR would be attenuated in this context.
-In line with this prediction, our descriptive analysis showed no correlation between birth order and NWR score.
+In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score.
+However, the effect size was larger than that found for the other factors, at r\textasciitilde{}.2, and thus we believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
 
@@
 
-In our sample of 40 children between ages 3 and 10, the average item-level score was 65\%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \ref{fig:prevlit-fig}).
-Segment-level analyses revealed that repetitions recovered an average of 79\% of the segments in the original stimulus, indicating that inexact repetitions often constituted \emph{mostly} accurate reroductions of the target.
-Complementary data from segment-based normalized Levenshtein distance reflect this same pattern, with an average of 21\% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
+In our sample of 40 children between ages 3 and 10, the average item-level score was 65\%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \ref{fig:fig-prevlit}).
+Segment-based normalized Levenshtein distance indicate that inexact repetitions often constituted \emph{mostly} accurate reroductions of the target, with an average of 21\% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
 Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
-Given successful (and comparable) execution of this task, we can dive deeper into effects of stimulus length, segmental frequency, age, and developmental context.
+Given successful (and comparable) execution of this task.
 
@@
 \leavevmode\hypertarget{ref-is08042009language}{}%
-COST Action. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www. Bi-Sli. Org}.
+COST Action. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www.bi-Sli.org}.
 
@@
 (2018).
-Normalization of the non-word repetition test in farsi-speaking children.
+Normalization of the non-word repetition test in Farsi-speaking children.
 \emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
@@
 \leavevmode\hypertarget{ref-foley1986papuan}{}%
-Foley, W. A. (1986). \emph{The papuan languages of new guinea}. Cambridge, UK: Cambridge University Press.
+Foley, W. A. (1986). \emph{The Papuan languages of New Guinea}. Cambridge, UK: Cambridge University Press.
 
@@
 (2007).
-Non-word repetition: An investigation of phonological complexity in children with grammatical sli.
+Non-word repetition: An investigation of phonological complexity in children with Grammatical SLI.
 \emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
@@
 (2014).
-Nonword repetition--a clinical marker for specific language impairment in swedish associated with parents' language-related problems.
+Nonword repetition--a clinical marker for specific language impairment in Swedish associated with parents' language-related problems.
 \emph{PloS One}, \emph{9}(2), e89544.
@@
 \leavevmode\hypertarget{ref-levinsonYDgrammar}{}%
-Levinson, S. C. (2020). \emph{A grammar of yélî dnye, the papuan language of rossel island}. Berlin, Boston: De Gruyter Mouton.
+Levinson, S. C. (2020). \emph{A grammar of Yélî Dnye, the Papuan language of Rossel Island}. Berlin, Boston: De Gruyter Mouton.
 
@@
 (2017).
-Independent and combined effects of socioeconomic status (ses) and bilingualism on children's vocabulary and verbal short-term memory.
+Independent and combined effects of socioeconomic status (SES) and bilingualism on children's vocabulary and verbal short-term memory.
 \emph{Frontiers in Psychology}, \emph{8}, 1442.
@@
 (2016).
-Disentangling sli and bilingualism using sentence repetition tasks: The impact of l1 and l2 properties.
+Disentangling SLI and bilingualism using sentence repetition tasks: The impact of L1 and L2 properties.
 \emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
@@
 (2019).
-An articulatory-based disyllabic and trisyllabic non-word repetition test: Reliability and validity in italian 3-to 7-year-old children.
+An articulatory-based disyllabic and trisyllabic Non-Word Repetition test: reliability and validity in Italian 3-to 7-year-old children.
 \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.

### 2021-01-21 manuscript.Rmd
minor

@@
 
+1. add SDs to https://docs.google.com/spreadsheets/d/1f4X7SmP4flhPlCEtvyqp_ERgMp5SC8qHCaMHj93C_FU/edit#gid=44537533
 1. add a table with sample studies & languages & their characteristiscs
@@
 1. to the above, add sylable characteristics, inventory size, inventory compactness
-1. fix camila ref
+1. revise intro
+1. revise discussion
 
@@
 
+- despite 4-syll being rare, they are repeated correctly
+- similar levels of IDS in Tsi & YD, yet very different results: overall NWR, effect of length
+- conclusions Cristia about long-term effects perhaps don't hold
+- or perhaps input estimation is inaccurate -- talk about different ways of estimating input quantity across papers and uncertainty in estimates
+- or perhaps it's about the prevalence of literacy in the population -- phonetic effects of learning to read & consequences in young children's encoding and decoding of sounds for NWR
+- differences in test: group versus individual, level of training of RA, from the same community or not, model was native or not, heard over headphones or not
+
 ### Conclusions

### 2021-01-21 manuscript.Rmd
minor

@@
 
-1. add SDs to https://docs.google.com/spreadsheets/d/1f4X7SmP4flhPlCEtvyqp_ERgMp5SC8qHCaMHj93C_FU/edit#gid=44537533
+1. add language chars to https://docs.google.com/spreadsheets/d/1f4X7SmP4flhPlCEtvyqp_ERgMp5SC8qHCaMHj93C_FU/edit#gid=44537533
 1. add a table with sample studies & languages & their characteristiscs
@@
 1. revise discussion
+1. improve figure 1?
 
@@
 
-pl_stacked=cbind(pl[,1:4],stack(pl[5:8]),stack(pl[9:12]))
+pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
 colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
@@
              "Italian"   ="black",
+             "Cantonese"   ="blue",
+             "English"   ="green",
             "Tsimane'" = "darkgreen",
@@
              "Italian"   =1,
+             "Cantonese"   =1,
+             "English"   =1,
              "Tsimane'" = 2,

### 2021-01-22 manuscript.Rmd
added figure on overall effects across languages

@@
 
-1. add language chars to https://docs.google.com/spreadsheets/d/1f4X7SmP4flhPlCEtvyqp_ERgMp5SC8qHCaMHj93C_FU/edit#gid=44537533
-1. add a table with sample studies & languages & their characteristiscs
-1. be accurate on what hasn't been done x-ling speaking
-1. try to find data on length distribution across languages to see if this fits length differences
-1. to the above, add sylable characteristics, inventory size, inventory compactness
-1. revise intro
+1. add legend to new figure, improve color coding
+1.
+conclusion: definitely the case that Tsi lower than others, but unclear what this is due to; some variation across samples, but since stimuli haven't been carefully matched, and ages either, we can't be totally sure there either -- also, literacy may be another issue
 1. revise discussion
@@
 Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
-(2019) reported on groups of 24-60 Italian learners (black); Cristia et al.
-(2020) reported on groups of 4-6 Tsimane' learners (green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
+(2019) reported on groups of 24-60 Italian learners (black); Stokes et al.
+(2006) on 15 Cantonese learners (blue); Vance et al.
+(2005) on 17-20 English learners (light green); Cristia et al.
+(2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
 Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
@@
 
-```{r Fig2-scores-by-age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-item scores for individual participants as a function of age and sex (purple crosses = boys, orange circles = girls)."}
+```{r fig-scores-by-age, ig.pos = "T!", fig.width=5, fig.height=5, fig.cap="NWR whole-item scores for individual participants as a function of age and sex (purple crosses = boys, orange circles = girls)."}
 
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question.
-As shown in Figure \@ref(fig:Fig2-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
+As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
 In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
@@
 
-### Other observations
+### NWR across languages and cultures
+
+```{r fig-prevlit-overall, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years).
+Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
+(2019) reported on groups of 24-60 Italian learners (black); Cristia et al.
+(2020) reported on groups of 4-6 Tsimane' learners (green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
+Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
+Age has been slightly jittered for ease of inspection of different lengths at a given age."}
+
+pl_overall=pl[,1:7]
+
+#add YD data
+yd_res_toadd[yd_res_toadd$Length!=1,]->yd_res_toadd
+means=aggregate(yd_res_toadd$Median,by=list(yd_res_toadd$Study,yd_res_toadd$Language,yd_res_toadd$n,yd_res_toadd$Age),mean)
+sds=aggregate(yd_res_toadd$SD,by=list(yd_res_toadd$Study,yd_res_toadd$Language,yd_res_toadd$n,yd_res_toadd$Age),mean)
+yd_res_toadd2=cbind(means,sds$x)
+
+colnames(yd_res_toadd2)<-c("Study","Language","n","Age","Overall","SD")
+yd_res_toadd2$Lengths<-"2-4"
+yd_res_toadd2=yd_res_toadd2[,c("Study","Language","n","Age","Lengths","Overall","SD")]
+
+#combine
+pl_overall=rbind(pl_overall,yd_res_toadd2)
+
+pl_overall$Language<-gsub(" (.*)","",pl_overall$Language)
+
+read.csv("lang-char.tsv",sep="\t")->lang
+merge(pl_overall,lang,by="Language")->pl_overall
+
+#add SE
+pl_overall$se=pl_overall$SD/sqrt(pl_overall$n)
+
+#pairs(pl_overall[c("Overall","nchar","phones","words")])
+
+#cor.test(pl_overall$Overall,pl_overall$nchar)
+#t.test(pl_overall$Overall~pl_overall$words)
+
+mycols = c( 
+             "Arabic" = "orange",
+             "Cantonese"   ="blue",
+             "English"   ="green",
+             "Italian"   ="black",
+             "Mandarin"   ="gray",
+             "Persian"   ="pink",
+             "Slovak"   ="lightblue",
+            "Sotho"   ="magenta",
+            "Spanish"   ="yellow",
+            "Swedish"   ="darkgray",
+            "Tsimane'" = "darkgreen",
+             "Yélî" = "purple")
+
+mywd = c( 
+             "Arabic" = 1,
+             "Cantonese"   =1,
+             "English"   =1,
+             "Italian"   =1,
+             "Mandarin"   =1,
+             "Persian"   =1,
+             "Slovak"   =1,
+            "Sotho"   =1,
+            "Spanish"   =1,
+            "Swedish"   =1,
+             "Tsimane'" = 2,
+             "Yélî" = 4)
+
+plot(pl_overall$Overall~pl_overall$Age,col=mycols[pl_overall$Language],pch=1,
+     xlab="Average age (years)",ylab="NWR Scores (words)",ylim=c(0,100))
+
+for(thislang in levels(factor(pl_overall$Language))) {
+  myx=pl_overall$Age[pl_overall$Language==thislang ]
+  myy=pl_overall$Overall[pl_overall$Language==thislang]
+  lines(myy~myx, col=mycols[thislang],lwd=mywd[thislang]) #join points of same length & lang
+  mylb=myy-pl_overall$se[pl_overall$Language==thislang]
+  myhb=myy+pl_overall$se[pl_overall$Language==thislang]
+  for(i in 1:length(mylb)) {
+    mylty=as.numeric(as.character(thislength))
+    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang],lty=mylty) #error bar
+    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang],lty=mylty) #lower error bar tip
+    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang],lty=mylty) #higher error bar tip
+    }
+}
+
+#add legend
+
+```
 

### 2021-01-22 manuscript.tex
added figure on overall effects across languages

@@
 \item
-  add a table with sample studies \& languages \& their characteristiscs
+  add legend to new figure, improve color coding
 \item
-  be accurate on what hasn't been done x-ling speaking
+  conclusion: definitely the case that Tsi lower than others, but unclear what this is due to; some variation across samples, but since stimuli haven't been carefully matched, and ages either, we can't be totally sure there either -- also, literacy may be another issue
 \item
-  try to find data on length distribution across languages to see if this fits length differences
+  revise discussion
 \item
-  to the above, add sylable characteristics, inventory size, inventory compactness
+  improve figure 1?
 \end{enumerate}
@@
 The Yélî community can help us bring further light into this question because direct speech to children under 3;0 is relatively infrequent in this community too (Casillas et al., 2020).
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour ({\textbf{???}}), and Yélî children about 3.7 minutes per hour (Casillas et al., 2020).
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.7 minutes per hour (Casillas et al., 2020).
 Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in Cristia et al.
@@
 \centering
-\includegraphics{manuscript_files/figure-latex/Fig2-scores-by-age-1.pdf}
-\caption{\label{fig:Fig2-scores-by-age}NWR whole-item scores for individual participants as a function of age and sex (purple crosses = boys, orange circles = girls).}
+\includegraphics{manuscript_files/figure-latex/fig-scores-by-age-1.pdf}
+\caption{\label{fig:fig-scores-by-age}NWR whole-item scores for individual participants as a function of age and sex (purple crosses = boys, orange circles = girls).}
 \end{figure}
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question.
-As shown in Figure \ref{fig:Fig2-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (5,649.08) = 0.47, p \textless{} 0.01).
+As shown in Figure \ref{fig:fig-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (5,649.08) = 0.47, p \textless{} 0.01).
 In contrast, there was no clear association between NWR scores and sex (Welch t (27.33) = -0.60, p = 0.56), birth order (data missing for 15 children, rho = (3,502.90) = -0.20, p = 0.33), or maternal education (rho ( 9,628.60) = 0.10, p = 0.55).
@@
 
-\hypertarget{other-observations}{%
-\subsubsection{Other observations}\label{other-observations}}
+\hypertarget{nwr-across-languages-and-cultures}{%
+\subsubsection{NWR across languages and cultures}\label{nwr-across-languages-and-cultures}}
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/fig-prevlit-overall-1.pdf}
+\caption{\label{fig:fig-prevlit-overall}NWR scores as a function of age (in years).
+Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
+(2019) reported on groups of 24-60 Italian learners (black); Cristia et al.
+(2020) reported on groups of 4-6 Tsimane' learners (green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
+Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
+Age has been slightly jittered for ease of inspection of different lengths at a given age.}
+\end{figure}
 
@@
 Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
+
+\begin{itemize}
+\tightlist
+\item
+  despite 4-syll being rare, they are repeated correctly
+\item
+  similar levels of IDS in Tsi \& YD, yet very different results: overall NWR, effect of length
+\item
+  conclusions Cristia about long-term effects perhaps don't hold
+\item
+  or perhaps input estimation is inaccurate -- talk about different ways of estimating input quantity across papers and uncertainty in estimates
+\item
+  or perhaps it's about the prevalence of literacy in the population -- phonetic effects of learning to read \& consequences in young children's encoding and decoding of sounds for NWR
+\item
+  differences in test: group versus individual, level of training of RA, from the same community or not, model was native or not, heard over headphones or not
+\end{itemize}
 
@@
 
+\leavevmode\hypertarget{ref-scaff2021daylong}{}%
+Scaff, C., Stieglitz, J., Casillas, M., \& Cristia, A.
+(2021).
+Daylong audio recordings of young children in a forager-farmer society show low levels of verbal input with minimal age-related changes.
+\emph{Draft}.
+
 \leavevmode\hypertarget{ref-torrington2015non}{}%

### 2021-01-26 manuscript.Rmd
added discussion of other lit

@@
 
-## TODO
-
-1. add legend to new figure, improve color coding
-1.
-conclusion: definitely the case that Tsi lower than others, but unclear what this is due to; some variation across samples, but since stimuli haven't been carefully matched, and ages either, we can't be totally sure there either -- also, literacy may be another issue
-1. revise discussion
-1. improve figure 1?
+## TODO Alex
+
+- add missing refs
+- correct rest of discussion
+
+## TODO Middy
+
+- read over whole thing -- does the logic sound ok?
+- if you are motivated to, improve figures 1 and 5 (integrating our results to previous lit) -- but I also think we can submit the paper with these figures as they are, and propose improvements in a revision
 
@@
 
-```{r fig-prevlit-overall, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years).
-Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
-(2019) reported on groups of 24-60 Italian learners (black); Cristia et al.
-(2020) reported on groups of 4-6 Tsimane' learners (green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
-Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
-Age has been slightly jittered for ease of inspection of different lengths at a given age."}
+```{r prep-pl}
 
@@
 
+all_lang=levels(factor(pl_overall$Language))
+
+```
+
+One of the questions in our mind when designing this study was whether NWR was a fair test of phonological development across languages and cultures.
+Although our data cannot answer this question because we have only sampled one language and culture here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
+It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
+Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the MacArthur-Bates Communicative Development Inventory [@wordbank].
+
+At first sight, when we had compared our results to those of other studies, we thought the range of performance we observed overlapped with previously observed levels of performance.
+Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
+Additionally, it seemed that Yélî children show edcomparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane'.
+
+To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
+We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages.
+Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei; Persian by @farmani2018normalization; Slovak by @kapalkova and @polisenska; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @Radeborg; Tsimane' by @cristia2020infant; and Yélî Dnye by the present study.
+Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
+Results separating different age groups are shown in \@ref(fig:fig-prevlit-overall).
+
+```{r fig-prevlit-overall, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages.
+The legend indicates language and the length of non-words (in syllables).
+Central tendency is mean; error is one standard error."}
 mycols = c( 
@@
             "Sotho"   ="magenta",
-            "Spanish"   ="yellow",
+            "Spanish"   ="brown",
             "Swedish"   ="darkgray",
@@
   for(i in 1:length(mylb)) {
-    mylty=as.numeric(as.character(thislength))
-    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang],lty=mylty) #error bar
-    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang],lty=mylty) #lower error bar tip
-    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang],lty=mylty) #higher error bar tip
+
+    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang]) #error bar
+    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang]) #lower error bar tip
+    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang]) #higher error bar tip
     }
@@
 #add legend
-
-```
-
-In our sample of `r nrow(overall)` children between ages `r min(overall$Age)` and `r max(overall$Age)`, the average item-level score was `r round(mean(overall$Word))`%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \@ref(fig:fig-prevlit)).
-Segment-based normalized Levenshtein distance  indicate that inexact repetitions often constituted _mostly_ accurate reroductions of the target, with an average of `r round(mean(overall$NLD))`% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
-Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
-Given successful (and comparable) execution of this task.
+#first half of langs
+xmin=6
+ymax=30
+ymin=0
+myy=seq(from=ymax,to=ymin,length.out=length(all_lang)/2)
+names(myy)<-all_lang[1:(length(all_lang)/2)]
+for(thislang in names(myy)){
+  lines(c(xmin-.1,xmin+.1),c(myy[thislang],myy[thislang]),col=mycols[thislang]) #line sample
+  text(xmin+.15,myy[thislang],paste(thislang,pl_overall$Lengths[pl_overall$Language==thislang][1]),col=mycols[thislang],pos=4)
+}
+
+#second half
+xmin=9.5
+names(myy)<-all_lang[(length(all_lang)/2+1):length(all_lang)]
+for(thislang in names(myy)){
+  lines(c(xmin-.1,xmin+.1),c(myy[thislang],myy[thislang]),col=mycols[thislang]) #line sample
+  text(xmin+.15,myy[thislang],paste(thislang,pl_overall$Lengths[pl_overall$Language==thislang][1]),col=mycols[thislang],pos=4)
+} 
+```
+
+Several observations can be drawn from this Figure.
+To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
+These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results, with lower overall NWR scores and (integrating with effect of length in  \@ref(fig:fig-prevlit-overall)) length effects.
+This may indicate that the conclusion tentatively drawn in @cristia2020infant about lower NWR scores consistent with long-term effects of lower levels of child-directed speech was premature.
+Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
+In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
+@cristia estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
+Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.
+a female adult that was not temporally close to speech by others would count as child-directed).
+A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
+
+That said, @cristia2020infant also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
+This is a very different case from the Yélî population studied here, where all adults have accumulated several years of schooling, and literacy is widespread.
+If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
+Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in \@ref(fig:fig-prevlit-overall).
+These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
+What is at stake in the second interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
+
+Although exciting, this hypothesis is only one of many.
+Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
+@cristia2020infant administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
+This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
+Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
+
+Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, as well as the strength of age effects.
+We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
+We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
+Neither of these two predictors explained variance  in \@ref(fig:fig-prevlit-overall).
+It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.
+Sotho non-words having 4-7 syllables in length.
+To be more certain whether language characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
+We believe this will be excedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
+
+### Additional observations
 
@@
 Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998illiterate], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998illiterate] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
-Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
-
-- despite 4-syll being rare, they are repeated correctly
-- similar levels of IDS in Tsi & YD, yet very different results: overall NWR, effect of length
-- conclusions Cristia about long-term effects perhaps don't hold
-- or perhaps input estimation is inaccurate -- talk about different ways of estimating input quantity across papers and uncertainty in estimates
-- or perhaps it's about the prevalence of literacy in the population -- phonetic effects of learning to read & consequences in young children's encoding and decoding of sounds for NWR
-- differences in test: group versus individual, level of training of RA, from the same community or not, model was native or not, heard over headphones or not
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
 

### 2021-01-26 manuscript.tex
added discussion of other lit

@@
 
-\hypertarget{todo}{%
-\subsection{TODO}\label{todo}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
+\hypertarget{todo-alex}{%
+\subsection{TODO Alex}\label{todo-alex}}
+
+\begin{itemize}
 \tightlist
 \item
-  add legend to new figure, improve color coding
+  add missing refs
 \item
-  conclusion: definitely the case that Tsi lower than others, but unclear what this is due to; some variation across samples, but since stimuli haven't been carefully matched, and ages either, we can't be totally sure there either -- also, literacy may be another issue
+  correct rest of discussion
+\end{itemize}
+
+\hypertarget{todo-middy}{%
+\subsection{TODO Middy}\label{todo-middy}}
+
+\begin{itemize}
+\tightlist
 \item
-  revise discussion
+  read over whole thing -- does the logic sound ok?
 \item
-  improve figure 1?
-\end{enumerate}
+  if you are motivated to, improve figures 1 and 5 (integrating our results to previous lit) -- but I also think we can submit the paper with these figures as they are, and propose improvements in a revision
+\end{itemize}
 
@@
 
+One of the questions in our mind when designing this study was whether NWR was a fair test of phonological development across languages and cultures.
+Although our data cannot answer this question because we have only sampled one language and culture here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
+It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
+Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the MacArthur-Bates Communicative Development Inventory ({\textbf{???}}).
+
+At first sight, when we had compared our results to those of other studies, we thought the range of performance we observed overlapped with previously observed levels of performance.
+Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
+Additionally, it seemed that Yélî children show edcomparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane'.
+
+To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
+We entered data from 14 studies (including ours), presenting data from 12 languages.
+Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by ({\textbf{???}}); English by Vance et al.
+(2005); Italian by Piazzalunga et al.
+(2019); Mandarin by ({\textbf{???}}); Persian by Farmani et al.
+(2018); Slovak by ({\textbf{???}}) and ({\textbf{???}}); Sotho by Wilsenach (2013); Spanish by Balladares et al.
+(2016); Swedish by Kalnak et al.
+(2014) and ({\textbf{???}}); Tsimane' by Cristia et al.
+(2020); and Yélî Dnye by the present study.
+Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
+Results separating different age groups are shown in \ref{fig:fig-prevlit-overall}.
+
 \begin{figure}
@@
 \includegraphics{manuscript_files/figure-latex/fig-prevlit-overall-1.pdf}
-\caption{\label{fig:fig-prevlit-overall}NWR scores as a function of age (in years).
-Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
-(2019) reported on groups of 24-60 Italian learners (black); Cristia et al.
-(2020) reported on groups of 4-6 Tsimane' learners (green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
-Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
-Age has been slightly jittered for ease of inspection of different lengths at a given age.}
+\caption{\label{fig:fig-prevlit-overall}NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages.
+The legend indicates language and the length of non-words (in syllables).
+Central tendency is mean; error is one standard error.}
 \end{figure}
 
-In our sample of 40 children between ages 3 and 10, the average item-level score was 65\%, falling well within the expected range of NWR applications in previous studies on a variety of linguistic populations (see Figure \ref{fig:fig-prevlit}).
-Segment-based normalized Levenshtein distance indicate that inexact repetitions often constituted \emph{mostly} accurate reroductions of the target, with an average of 21\% of the stimulus segments getting substituted, inserted, or deleted upon repetition.
-Paired with our thorough training protocol, we take these NWR scores as indicating that (a) our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli and (b) Yélî children show comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations.
-Given successful (and comparable) execution of this task.
+Several observations can be drawn from this Figure.
+To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
+These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results, with lower overall NWR scores and (integrating with effect of length in \ref{fig:fig-prevlit-overall}) length effects.
+This may indicate that the conclusion tentatively drawn in Cristia et al.
+(2020) about lower NWR scores consistent with long-term effects of lower levels of child-directed speech was premature.
+Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
+In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: Casillas et al.
+(2020) hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
+({\textbf{???}}) estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
+Finally, Scaff et al.
+(2021) used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.~a female adult that was not temporally close to speech by others would count as child-directed).
+A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
+
+That said, Cristia et al.
+(2020) also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
+This is a very different case from the Yélî population studied here, where all adults have accumulated several years of schooling, and literacy is widespread.
+If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
+Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in \ref{fig:fig-prevlit-overall}.
+These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
+What is at stake in the second interpretation of the lower scores observed among the Tsimane' is related to literacy \emph{in the broader population} (rather than in the tested children themselves).
+
+Although exciting, this hypothesis is only one of many.
+Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
+Cristia et al.
+(2020) administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
+This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
+Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
+
+Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, as well as the strength of age effects.
+We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
+We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
+Neither of these two predictors explained variance in \ref{fig:fig-prevlit-overall}.
+It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.~Sotho non-words having 4-7 syllables in length.
+To be more certain whether language characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
+We believe this will be excedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
+
+\hypertarget{additional-observations}{%
+\subsubsection{Additional observations}\label{additional-observations}}
 
@@
 Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work (e.g., Castro-Caldas et al., 1998), but it is unclear what caused this pattern in the current study: Castro and colleagues' (1998) study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
-Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children will clarify this effect.
-
-\begin{itemize}
-\tightlist
-\item
-  despite 4-syll being rare, they are repeated correctly
-\item
-  similar levels of IDS in Tsi \& YD, yet very different results: overall NWR, effect of length
-\item
-  conclusions Cristia about long-term effects perhaps don't hold
-\item
-  or perhaps input estimation is inaccurate -- talk about different ways of estimating input quantity across papers and uncertainty in estimates
-\item
-  or perhaps it's about the prevalence of literacy in the population -- phonetic effects of learning to read \& consequences in young children's encoding and decoding of sounds for NWR
-\item
-  differences in test: group versus individual, level of training of RA, from the same community or not, model was native or not, heard over headphones or not
-\end{itemize}
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
 
@@
 \leavevmode\hypertarget{ref-casillas2020early}{}%
-Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a Papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
 

### 2021-01-28 manuscript.Rmd
midway through intro, lit added

@@
 
-## TODO Alex
-
-- add missing refs
-- correct rest of discussion
-
 ## TODO Middy
@@
 - read over whole thing -- does the logic sound ok?
-- if you are motivated to, improve figures 1 and 5 (integrating our results to previous lit) -- but I also think we can submit the paper with these figures as they are, and propose improvements in a revision
+- tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
 
@@
 NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
-
-NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
+NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences].
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language].
-
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
@@
 First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], likely as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non].
+Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely language.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
@@
 It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture.
-That said, there could be other causal pathways: Research on adults suggests that illiterate Portuguese speakers repeat monosyllabic non-words just as accurately as literate adults, whereas scores are much lower among illiterate than literate speakers for items 3 or 4 syllables long [@de2004degree].
-Given that both groups of adults speak the same language (Brazilian Portuguese), then perhaps differences in repetition accuracy reveal differences in how flexible the perception-production loop is.
-Given the paucity of evidence looking at this question, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
-
-Third, there are ongoing discussions as to what the key factors structuring individual variation are.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword].
-In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
-For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
-Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
-Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.
-
-Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
-Indeed, NWR has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], and/or outside urban settings [except for in @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel;@maddieson2013consonant; no exceptions to our knowledge].
-There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
-That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
-In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating].
-Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills.
-Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
-To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned.
-Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
-This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
-And it does occur that some researchers opt instead to employ dimensions of variation that are more relevant to their language, such as adaptations in Chinese languages that have items varying in tone REF.
-**return to this after reading lit**
-
-Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are typically easily accessible.
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
-<!--Removed for brevity: Most Yélî people are swidden horticulturalists who cultivate yams, manioc, taro, coconut, banana, and more, getting protein primarily from fish, shellfish, and nuts.
-There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport.
-News and messages are passed to and from mainland PNG via a radio connection at the local health clinic and, when it is functioning, a cellular tower on the island.--> International phone calls and digital exchanges that require significant data transfer are typically not an option.
-Data collection is therefore typically limited to the duration of the researchers' on-island visits.
-
-### Yélî Dnye phonology
-
-Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
-The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region.
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
-Thus, with respect to our first research goal, Yélî Dnye seemed is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
-
-<!--
-Removed for brevity:
-and includes at least two contrasts that had not previously been documented elsewhere [labial-coronal double-articulations with dental vs.
-post-alveolar coronal placement for both oral and nasal stops, @ladefoged1996sounds; @maddiesonIPphoneticsYD].
--->
-
-To provide some qualitative information on this inventory, we add the following observations.
-With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop system includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).^[We use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
-In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).] The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
-Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
-
-Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
-The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" /ṇṭæp/ *Spondylus shell*) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/).
-There are also a handful of words starting with /æ/ (e.g., "ala" /æ.'læ/ *here*) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
-
-Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun [e.g., @peuteIPconsonants].
-We hope the present study contributes to filling this gap.
-**TODO incorporate brief summary of paper**
-
-<!--
-NOTE: The following paragraph was removed b/c it's less relevant, but it has been double-checked by MC Nov 2020 in case we decide to include it later:
-
-While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
-In monosyllabic stems, oral vowels (short: 43% and long: 31%) are more common than nasal ones (short: 4% and long: 22%), with the first syllable of bisyllabic stems following a similar pattern.
-Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels.
-Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see @levinsonYDgrammar for details).
--->
-
-### The Yélî community
-
-Some aspects of the community are relevant for interpreting results found when addressing our thir research question, regarding sources of individual variation.
-Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
-There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.
-
-The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children -- their own and billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
-These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
-Land attribution for building one's home is decided collectively based on land availability, and typically does not take into consideration an individual's desire to be close to a school.
-
-Most Yélî parents are swidden horticulturalists, and those who are not may not reside in the island.
-Within a group of households, it is often the case that most older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).
-Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
-Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early].
-Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
-Children start school often at around age six, although the precise age depends on the child's apparent development.
-
-Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above.
-To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
-Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
-Together with the fact that land attribution is essentially random with respect to educational hopes, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
-
-As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology].
-It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
-This contrasts with this picture just drawn in the Yélî community, where children regardless of their birth order in their nuclear family will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders.
-
-We add some observations that will help us integrate this study to the broader investigation of NWR across cultures.
-As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant].
-However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
-The Yélî community can help us bring further light into this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early].
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.7 minutes per hour [@casillas2020early].
-Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in @cristia2020infant.
-
-### NWR design and analysis adaptations
-
-In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
-For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
-There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
-For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language].
-Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory].
-Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
-
-Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
-We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.
-less challenging sounds.
-However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
-Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
-However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
-
-### Research questions
-
-After some preliminary analyses to set the stage, we perform statistical analyses to inform answers to the following questions:
-
-- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
-- How do NWR scores change as a function of item length in number of syllables?
-- Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?
-
-Throughout these analyses and in the Discussion, we will also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.
-
-We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given the previously discussed variability across studies.
-Therefore, all analyses in the present study are descriptive and should be considered exploratory.
+In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
+To attempt to see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
+We found data for learners of Israeli Arabic  @jabere2018xperiment; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
 
 ```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed).
-Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
+Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al.
 (2019) reported on groups of 24-60 Italian learners (black); Stokes et al.
@@
 (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
-Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
+Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error.
 Age has been slightly jittered for ease of inspection of different lengths at a given age."}
@@
 pl_stacked=pl_stacked[pl_stacked$Length!=1,]
+
+pl_stacked$Language=factor(pl_stacked$Language)
 
@@
 
-```
+pl$dif<-pl$X2-pl$X3
+difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
+
+```
+
+Our reading of the previous literature is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages.
+For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only `r round(difs[difs$Group.1=="Italian","x"])`%.
+Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese `r round(difs[difs$Group.1=="Cantonese","x"])`% versus English `r round(difs[difs$Group.1=="English","x"])`%).
+Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
+
+Third, there are ongoing discussions as to what the key factors structuring individual variation are.
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword].
+In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
+For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
+Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
+Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
+Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.
+
+Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
+Indeed, NWR has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], and/or outside urban settings [except for in @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel;@maddieson2013consonant; no exceptions to our knowledge].
+There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
+That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating].
+Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills.
+Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned.
+Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
+This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
+And it does occur that some researchers opt instead to employ dimensions of variation that are more relevant to their language, such as adaptations in Chinese languages that have items varying in tone REF.
+**return to this after reading lit**
+
+Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are typically easily accessible.
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
+International phone calls and digital exchanges that require significant data transfer are typically not an option.
+Data collection is therefore typically limited to the duration of the researchers' on-island visits.
+
+<!--Removed for brevity: Most Yélî people are swidden horticulturalists who cultivate yams, manioc, taro, coconut, banana, and more, getting protein primarily from fish, shellfish, and nuts.
+There is no electricity system (visiting researchers use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport.
+News and messages are passed to and from mainland PNG via a radio connection at the local health clinic and, when it is functioning, a cellular tower on the island.-->
+
+### Yélî Dnye phonology
+
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region.
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
+Thus, with respect to our first research goal, Yélî Dnye seemed is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
+
+<!--
+Removed for brevity:
+and includes at least two contrasts that had not previously been documented elsewhere [labial-coronal double-articulations with dental vs.
+post-alveolar coronal placement for both oral and nasal stops, @ladefoged1996sounds; @maddiesonIPphoneticsYD].
+-->
+
+To provide some qualitative information on this inventory, we add the following observations.
+With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
+The core oral stop system includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).^[We use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).] The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+Vowels in Yélî Dnye may be oral or nasal, short or long.
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+
+Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
+The vast majority of syllables use a CV format.
+A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" /ṇṭæp/ *Spondylus shell*) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/).
+There are also a handful of words starting with /æ/ (e.g., "ala" /æ.'læ/ *here*) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
+
+Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun [e.g., @peuteIPconsonants].
+We hope the present study contributes to filling this gap.
+**TODO incorporate brief summary of paper**
+
+<!--
+NOTE: The following paragraph was removed b/c it's less relevant, but it has been double-checked by MC Nov 2020 in case we decide to include it later:
+
+While all consonants can appear in the first syllable of the word, the second syllable onset consonant in bisyllabic roots is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/).
+In monosyllabic stems, oral vowels (short: 43% and long: 31%) are more common than nasal ones (short: 4% and long: 22%), with the first syllable of bisyllabic stems following a similar pattern.
+Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels.
+Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see @levinsonYDgrammar for details).
+-->
+
+### The Yélî community
+
+Some aspects of the community are relevant for interpreting results found when addressing our thir research question, regarding sources of individual variation.
+Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
+There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.
+
+The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children -- their own and billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
+Land attribution for building one's home is decided collectively based on land availability, and typically does not take into consideration an individual's desire to be close to a school.
+
+Most Yélî parents are swidden horticulturalists, and those who are not may not reside in the island.
+Within a group of households, it is often the case that most older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early].
+Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
+Children start school often at around age six, although the precise age depends on the child's apparent development.
+
+Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above.
+To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
+Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
+Together with the fact that land attribution is essentially random with respect to educational hopes, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
+
+As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology].
+It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
+This contrasts with this picture just drawn in the Yélî community, where children regardless of their birth order in their nuclear family will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders.
+
+We add some observations that will help us integrate this study to the broader investigation of NWR across cultures.
+As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant].
+However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
+The Yélî community can help us bring further light into this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early].
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.7 minutes per hour [@casillas2020early].
+Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in @cristia2020infant.
+
+### NWR design and analysis adaptations
+
+In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
+For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
+There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
+For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language].
+Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory].
+Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
+
+Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
+We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.
+less challenging sounds.
+However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
+Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
+However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
+
+### Research questions
+
+After some preliminary analyses to set the stage, we perform statistical analyses to inform answers to the following questions:
+
+- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
+- How do NWR scores change as a function of item length in number of syllables?
+- Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?
+
+Throughout these analyses and in the Discussion, we will also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.
+
+We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
+However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given the previously discussed variability across studies.
+Therefore, all analyses in the present study are descriptive and should be considered exploratory.
 
@@
 It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
-Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the MacArthur-Bates Communicative Development Inventory [@wordbank].
+Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages [@frank2017wordbank].
 
@@
 We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages.
-Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei; Persian by @farmani2018normalization; Slovak by @kapalkova and @polisenska; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @Radeborg; Tsimane' by @cristia2020infant; and Yélî Dnye by the present study.
+Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye by the present study.
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
@@
 In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
-@cristia estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
+@cristia2019child estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
 Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.

### 2021-01-28 manuscript.tex
midway through intro, lit added

@@
 
-\hypertarget{todo-alex}{%
-\subsection{TODO Alex}\label{todo-alex}}
-
-\begin{itemize}
-\tightlist
-\item
-  add missing refs
-\item
-  correct rest of discussion
-\end{itemize}
-
 \hypertarget{todo-middy}{%
@@
 \item
-  if you are motivated to, improve figures 1 and 5 (integrating our results to previous lit) -- but I also think we can submit the paper with these figures as they are, and propose improvements in a revision
+  tell me if you think you CAN'T LIVE with my turning this in with figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
 \end{itemize}
@@
 NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
-
-NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
+NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
@@
 First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes (with the possible exception of Gallagher, 2014), likely as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007).
+Previous work using NWR has preferred relatively universal and early-acquired phonemes (with the possible exception of Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely language.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
@@
 It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture.
-That said, there could be other causal pathways: Research on adults suggests that illiterate Portuguese speakers repeat monosyllabic non-words just as accurately as literate adults, whereas scores are much lower among illiterate than literate speakers for items 3 or 4 syllables long (de Santos Loureiro et al., 2004).
-Given that both groups of adults speak the same language (Brazilian Portuguese), then perhaps differences in repetition accuracy reveal differences in how flexible the perception-production loop is.
-Given the paucity of evidence looking at this question, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
+In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
+To attempt to see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
+We found data for learners of Israeli Arabic Jaber-Awida (2018); Cantonese (Stokes, Wong, Fletcher, \& Leonard, 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/fig-prevlit-1.pdf}
+\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables, 2=dashed, 3=dotted, 4=dotted and dashed).
+Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al.
+(2019) reported on groups of 24-60 Italian learners (black); Stokes et al.
+(2006) on 15 Cantonese learners (blue); Vance et al.
+(2005) on 17-20 English learners (light green); Cristia et al.
+(2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
+Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error.
+Age has been slightly jittered for ease of inspection of different lengths at a given age.}
+\end{figure}
+
+Our reading of the previous literature is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages.
+For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (28\%) and Arabic (15\%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only 2\%.
+Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese 8\% versus English 1\%).
+Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
 
@@
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
+Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance et al., 2005).
 Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
@@
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
-Indeed, NWR has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014; Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020), and/or outside urban settings (except for in Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b);Maddieson (2013a); no exceptions to our knowledge{]}.
+Indeed, NWR has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), and/or outside urban settings (except for in Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b);Maddieson (2013a); no exceptions to our knowledge{]}.
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 Therefore, all analyses in the present study are descriptive and should be considered exploratory.
-
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/fig-prevlit-1.pdf}
-\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables, 2=dashed, 3=dotted, 4=dotted and dashed).
-Jaber-Awida (2018) reported on 20 Arabic learners (orange); Piazzalunga et al.
-(2019) reported on groups of 24-60 Italian learners (black); Stokes et al.
-(2006) on 15 Cantonese learners (blue); Vance et al.
-(2005) on 17-20 English learners (light green); Cristia et al.
-(2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
-Central tendency is mean in Arabic and Tsimane', median in Italian and Yélî Dnye; error is one standard error.
-Age has been slightly jittered for ease of inspection of different lengths at a given age.}
-\end{figure}
 
@@
 It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
-Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the MacArthur-Bates Communicative Development Inventory ({\textbf{???}}).
+Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages (Frank, Braginsky, Yurovsky, \& Marchman, 2017).
 
@@
 We entered data from 14 studies (including ours), presenting data from 12 languages.
-Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by ({\textbf{???}}); English by Vance et al.
+Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes et al.
+(2006); English by Vance et al.
 (2005); Italian by Piazzalunga et al.
-(2019); Mandarin by ({\textbf{???}}); Persian by Farmani et al.
-(2018); Slovak by ({\textbf{???}}) and ({\textbf{???}}); Sotho by Wilsenach (2013); Spanish by Balladares et al.
+(2019); Mandarin by Lei et al.
+(2011); Persian by Farmani et al.
+(2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares et al.
 (2016); Swedish by Kalnak et al.
-(2014) and ({\textbf{???}}); Tsimane' by Cristia et al.
+(2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al.
 (2020); and Yélî Dnye by the present study.
@@
 (2020) hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
-({\textbf{???}}) estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
+Cristia, Dupoux, Gurven, and Stieglitz (2019) estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
 Finally, Scaff et al.
@@
 
+\leavevmode\hypertarget{ref-cristia2019child}{}%
+Cristia, A., Dupoux, E., Gurven, M., \& Stieglitz, J.
+(2019).
+Child-directed speech is infrequent in a forager-farmer population.
+\emph{Child Development}, \emph{90}(3), 759--773.
+\url{https://doi.org/10.1111/cdev.12974}
+
 \leavevmode\hypertarget{ref-cristia2020infant}{}%
@@
 \url{https://doi.org/https://doi.org/10.1371/journal.pone.0237702}
-
-\leavevmode\hypertarget{ref-de2004degree}{}%
-de Santos Loureiro, C., Braga, L.
-W., Nascimento Souza, L.
-do, Nunes Filho, G., Queiroz, E., \& Dellatolas, G.
-(2004).
-Degree of illiteracy and phonological and metaphonological skills in unschooled adults.
-\emph{Brain and Language}, \emph{89}(3), 499--502.
-\url{https://doi.org/10.1016/j.bandl.2003.12.008}
 
@@
 
+\leavevmode\hypertarget{ref-frank2017wordbank}{}%
+Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2017). Wordbank: An open repository for developmental vocabulary data. \emph{Journal of Child Language}, \emph{44}(3), 677--694.
+
 \leavevmode\hypertarget{ref-gallagher2014identity}{}%
@@
 
+\leavevmode\hypertarget{ref-kapalkova2013non}{}%
+Kapalková, S., Polišenská, K., \& Vicenová, Z.
+(2013).
+Non-word repetition performance in Slovak-speaking children with and without SLI: novel scoring methods.
+\emph{International Journal of Language and Communication Disorders}, \emph{48}(1), 78--89.
+\url{https://doi.org/10.1111/j.1460-6984.2012.00189.x}
+
 \leavevmode\hypertarget{ref-lancy2015anthropology}{}%
 Lancy, D. F. (2015). \emph{The anthropology of childhood}. Cambridge, UK: Cambridge University Press.
+
+\leavevmode\hypertarget{ref-lei2011developmental}{}%
+Lei, L., Pan, J., Liu, H., McBride-Chang, C., Li, H., Zhang, Y., \ldots{} others.
+(2011).
+Developmental trajectories of reading development and impairment from ages 3 to 8 years in chinese children.
+\emph{Journal of Child Psychology and Psychiatry}, \emph{52}(2), 212--220.
 
@@
 
+\leavevmode\hypertarget{ref-mulder2017early}{}%
+Mulder, H., Verhagen, J., Van der Ven, S.
+H., Slot, P.
+L., \& Leseman, P.
+P.
+(2017).
+Early executive function at age two predicts emergent mathematics and literacy at age five.
+\emph{Frontiers in Psychology}, \emph{8}, 1706.
+
 \leavevmode\hypertarget{ref-peuteIPconsonants}{}%
@@
 
+\leavevmode\hypertarget{ref-polivsenska2014improving}{}%
+Polišenská, K., \& Kapalková, S. (2014). Improving child compliance on a computer-administered nonword repetition task. \emph{Journal of Speech, Language and Hearing Research}, \emph{57}(3).
+
+\leavevmode\hypertarget{ref-radeborg2006swedish}{}%
+Radeborg, K., Barthelom, E., SjöBerg, M., \& Sahlén, B.
+(2006).
+A Swedish non-word repetition test for preschool children.
+\emph{Scandinavian Journal of Psychology}, \emph{47}(3), 187--192.
+\url{https://doi.org/10.1111/j.1467-9450.2006.00506.x}
+
 \leavevmode\hypertarget{ref-scaff2021daylong}{}%
@@
 \emph{Draft}.
+
+\leavevmode\hypertarget{ref-stokes2006nonword}{}%
+Stokes, S.
+F., Wong, A.
+M., Fletcher, P., \& Leonard, L.
+B.
+(2006).
+Nonword repetition and sentence repetition as clinical markers of specific language impairment: The case of cantonese.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{49}(2), 219--236.
 

### 2021-01-29 manuscript.Rmd
new version for Middy

@@
 
+r2d<-function(r){(2*abs(r))/sqrt(1-abs(r)^2)}
+
+d2r=function(d,n1,n2) {
+  a=(n1+n2)^2/(n1*n2)
+  d/sqrt(d^2 +a)}
 ```
@@
 - tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
+- do you mind if I change all the figures to say NWR scores rather than NWR Scores? In the text, we use lower case.
 
@@
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
-The study was designed to contribute to four aspects of our understanding of phonological development.
-
-First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR.
+The study was designed to contribute to four broad research areas.
+
+The first research area is at the crossing of typology and phonological development.
 Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely language.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
-
-Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development.
-Our reading of previous NWR research is that there are variable effects of length between populations.
-For instance, @jabere2018xperiment reports an average of ~96% correct repetition for items 2 syllables long among children learning an Arabic variety of Israeli at about 5.5 years of age, but ~81% for items 3 syllables long.
-In contrast, @piazzalunga2019articulatory observe no decline in performance in similarly-aged Italian learners, with a score of 84% for 2 syllables versus 85% for 3 syllables.
-It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture.
+Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.
+
+Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
+Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory].
+It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
-To attempt to see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
+To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
 We found data for learners of Israeli Arabic  @jabere2018xperiment; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
+
+```{r prep-prevlit, echo=F}
+pl<- read.csv("previous_lit.csv")
+
+pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
+colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
+pl_stacked$Length=gsub("X","",pl_stacked$Length)
+
+#remove lines with NA median
+pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked
+
+#add YD data
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age
+
+age_cuts=c(3,6,8,13)
+m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
+m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))
+
+aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
+yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns
+
+colnames(yd_res)<-c("Length","Age","Median","SD","n")
+yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
+colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")
+
+#combine
+pl_stacked=rbind(pl_stacked,yd_res_toadd)
+
+#add SE
+pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
+
+#remove monosyllables
+pl_stacked=pl_stacked[pl_stacked$Length!=1,]
+
+pl_stacked$Language=factor(pl_stacked$Language)
+
+pl$dif<-pl$X2-pl$X3
+difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
+```
 
@@
 
-pl<- read.csv("previous_lit.csv")
-
-pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
-colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
-pl_stacked$Length=gsub("X","",pl_stacked$Length)
-
-#remove lines with NA median
-pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked
-
-#add YD data
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age
-
-age_cuts=c(3,6,8,13)
-m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
-m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))
-
-aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
-yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns
-
-colnames(yd_res)<-c("Length","Age","Median","SD","n")
-yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
-colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")
-
-#combine
-pl_stacked=rbind(pl_stacked,yd_res_toadd)
-
-#add SE
-pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
-
-#remove monosyllables
-pl_stacked=pl_stacked[pl_stacked$Length!=1,]
-
-pl_stacked$Language=factor(pl_stacked$Language)
-
 mycols = c( 
@@
 
+#just a trick to jitter age
 jit_age=(pl_stacked$Age+(as.numeric(as.character(pl_stacked$Length))-2)/5)
-plot(pl_stacked$Median~jit_age,col=mycols[pl_stacked$Language],pch=as.character(pl_stacked$Length),
-     xlab="Average age (years)",ylab="NWR Scores (words)",ylim=c(0,100))
+
+plot(1,xlab="Average age (years)",ylab="NWR Scores (words)",ylim=c(0,100),xlim=range(pl_stacked$Age))
 
@@
   myx=myx+(as.numeric(as.character(thislength))-2)/5 #jitter the length lines a little bit with respect to avg age
+  points(pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]~myx,col=mycols[thislang],pch=as.character(thislength))
   myy=pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
@@
 
-pl$dif<-pl$X2-pl$X3
-difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
-
-```
-
-Our reading of the previous literature is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages.
-For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only `r round(difs[difs$Group.1=="Italian","x"])`%.
-Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese `r round(difs[difs$Group.1=="Cantonese","x"])`% versus English `r round(difs[difs$Group.1=="English","x"])`%).
+```
+
+Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages.
+For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%.
+Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).
 Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
 
-Third, there are ongoing discussions as to what the key factors structuring individual variation are.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables [@farabolini2021nonword].
+The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword].
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
@@
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
-Indeed, NWR has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], and/or outside urban settings [except for in @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel;@maddieson2013consonant; no exceptions to our knowledge].
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel;@maddieson2013consonant; no exceptions to our knowledge].
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
-And it does occur that some researchers opt instead to employ dimensions of variation that are more relevant to their language, such as adaptations in Chinese languages that have items varying in tone REF.
-**return to this after reading lit**
 
@@
 We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory which lead any error to result in many true-word phonetic neighbors.
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
 
@@
 
-dim(mat_pc2)
+#dim(mat_pc2)
 
@@
 Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other.
-Although a direct test is missing, the correlation with cross-linguistic frequency [r(`r phone_cor_xl$parameter`)=`r round(phone_cor_xl$estimate,3)`, p `r pval.display(round(phone_cor_xl$p.value,3))`]  was greater than that with within-language frequency [r(`r phone_cor_wl$parameter`)=`r round(phone_cor_wl$estimate,3)`, p `r pval.display(round(phone_cor_wl$p.value,3))`].
+Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency [r(`r phone_cor_xl$statistic`)=`r round(phone_cor_xl$estimate,3)`, p `r pval.display(round(phone_cor_xl$p.value,3))`]  was greater than that with within-language frequency [r(`r phone_cor_wl$statistic`)=`r round(phone_cor_wl$estimate,3)`, p `r pval.display(round(phone_cor_wl$p.value,3))`].
 
@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
-An additional overarching goal was to revisit this task, which is very commonly used to document phonological development, particularly in children raised in urban settings, and who are learning IndoEuropean languages.
-We will thus consider our results not only in light of previous work, but also with attention to potential linguistic and population differences.
+We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
+An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and other that will be discussed below).
+We consider implications of our results on each of these four research areas in turn.
 
@@
 
-**FIX LOGIC IN THIS PARA** Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
+Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
 Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
-Typologically common sounds are associated with higher performance on a handful of other tasks (**REFS -- M2A: Alex, I added this based on your note, where it sounded like you had some particular studies in mind?**) though to our knowledge this has not yet been tested with non-word repetition.
-Regarding Yélî Dnye in particular, the phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
-We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
-Indeed, we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often.
-This effect is large, with a magnitude more than twice the size of the effect of participant age.
-This significant effect remains even once also accounting for the frequencies of these segments in Yélî Dnye child-directed speech, which are correlated with their typological frequencies.
-In sum, typological frequency effects, which have been found in other measurements of phonological processing.
-appear to strongly affect NWR performance, and do not appear mitigated by language-specific pressure to make finer-grained differences earlier in development.
-
-**REWRITE THIS PARA** ~With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.
-oral vowels and simple vs.
-complex consonants, revealed no remarkable bias in error types.~ That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
+It would stand to reason that typologically common sounds are associated with higher performance, but to our knowledge this has not yet been tested with NWR.
+There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
+One could then predict that this effect should be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
+
+And yet, we found a robust effect of average segmental cross-linguistic frequency on NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
+This effect was large, with a magnitude more than twice the size of the effect of participant age.
+Moreover, this significant effect remained even once also accounting for the frequencies of these segments in Yélî Dnye children's input.
+An analysis of the substitutions made by children also aligned with this interpretation, with more common sounds being substituted for less common ones.
+
+We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development.
+We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [most often discussed in the case of phonotactic constraints on sequences, e.g., @maddieson2009phonology], but we do point out that this effect is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR.
+We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the fine-grained phonetic representations in the perception-production loop [as in e.g., @edwards2004interaction].
 
@@
 We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item.
-Based on previous work, we had predicted that children would have higher NWR scores for shorter items.
-That said, previous work has shown both very small [@piazzalunga2019articulatory] and very large [@cristia2020infant; @jabere2018xperiment] effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10% of the remaining words.
-Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms **M2A: Alex I'm going off your note here ("Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English."").
-I don't have a reference for this, can you please finish the thought or nix this bit?**.
-Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
-While indeed NWR scores were overall lower for longer items (e.g., see Figure \@ref(fig:fig-prevlit)), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
-In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length.
-First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon.
-Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s [-@piazzalunga2019articulatory] study on Italian and as borne out in the current dataset once controlling for other factors.
-**ADD WHAT LANGUAGES WOULD BE IDEAL TO TEST THIS HYPOTHESIS**
+In broad terms, children should have higher NWR scores for shorter items.
+That said, previous work summarized in the Introduction has shown both very small [e.g., @piazzalunga2019articulatory] and very large [e.g., @cristia2020infant] effects of stimulus length.
+Setting aside our monosyllabic stimuli (which contained typologically infrequent segments with lower NWR scores, as just discussed), we examined effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
+The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant, and is small and inconsistent across ages (see Figure \@ref(fig:fig-prevlit)).
+We do not have a good explanation for why samples in the literature vary so much in terms of the size of length effects, but two possibilities are that this is not truly a length effect but a confound with some other aspect of the stimuli, or that there is variation in phonological representations that is poorly understood.
+We explain each idea in turn.
+
+First, it remains possible that apparent length effects are actually due to uncontrolled aspects of the stimuli.
+For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants -- which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words.
+Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower.
+If this idea is correct, a careful analysis of non-words used in previous work may reveal that studies with larger length effects just happened to have longer non-words with lower n-phone frequencies.
+
+Second, NWR is often described as a task that tests flexible perception-production, and as such it is unclear why length effects should be observed at all.
+However, it is possible that NWR relies on more specific aspects of perception-production, in ways that are dependent on stimulus length.
+A hint in this direction comes from work on illiterate adults, who can be extremely accurate when repeating short non-words, but whose NWR scores are markedly lower for longer items.
+In a longitudinal study on Portuguese-speaking adults who were learning to read, @kolinsky2018completely found that, before reading training, the group scored 12.5% on 5-syllable items, whereas after 3 months of training, they scored 62.5% on such long items, whereas performance was at 100% for monosyllables throughout.
+Given that as adults they had fully acquired their native language, and obviously they had flexible perception-production schemes that allowed them to repeat new monosyllables perfectly, the change that occurred in those three months must relate to something else in their phonological skills, something that is not essential to speak a language natively.
+Thus, we hazard the hypothesis that sample differences in length effects may relate to such non-essential skills.
+Since as stated this hypothesis is under-specified, further conceptual and empirical work are both needed.
 
@@
 
-A review of previous work (see Introduction) suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR.
-We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta-analytic efforts aggregating over smaller studies such as ours.
-
-Following prior work, we expected that NWR scores would increase with participant age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
+Our review of previous work in the Introduction suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR.
+We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta- or mega-analytic efforts aggregating over studies.
+
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006nonword; but note  @cristia2020infant is an exception].
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent] or participant gender [@chiat2007preschool] on NWR scores.
-In addition to this prior work, education on Rossel Island, while generally highly valued, is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation.
-There is also limited variation in maternal education across the families in the region of the island where we sampled.
-We therefore expected little evidence for impact of either participant gender or maternal education in the present study.
-On the other hand, these predictors have established effects on other language development measures (**REFS: M2A: Alex go ahead and pick your faves here**).
-So to the extent that NWR scores share causal links to gender-based differences in development and maternal linguistic input with these other language outcome measures, we might then expect these factors to appear in NWR data.
-In fact, participant gender and maternal education correlated with NWR score at about r~.1, which is small.
-
-Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect], presumably because later-born children receive a smaller share of maternal input than their older siblings.
+In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores.
+We did not expect large effects of maternal education in our sample because of two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
+In fact, maternal education correlated with NWR score at about r~.1, which is small.
+Similarly, NWR scores may not vary greatly with participant gender according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.
+
+Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
@@
 In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score.
-However, the effect size was larger than that found for the other factors, at r~.2, and thus we believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
+However, the effect size was larger than that found for the other two factors and it is far from negligible, at r~.2 or Cohen's d~`r r2d(.2)`.
+In fact, two large studies with therefore precise estimates found effects of about d~.2 [@havron2019effect;@barclay2015within], which would suggest the effects we found are larger.
+We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
 
@@
 
-One of the questions in our mind when designing this study was whether NWR was a fair test of phonological development across languages and cultures.
-Although our data cannot answer this question because we have only sampled one language and culture here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
+The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, as when designing this study we wondered whether NWR was a *fair* test of phonological development.
+Although our data cannot answer this question because we have only sampled one language and population here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
 It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
@@
 
-At first sight, when we had compared our results to those of other studies, we thought the range of performance we observed overlapped with previously observed levels of performance.
+At first sight, the range of performance we observed overlapped with previously observed levels of performance.
 Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
-Additionally, it seemed that Yélî children show edcomparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane'.
+Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane'.
 
@@
 
-Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, as well as the strength of age effects.
+Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies.
 We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
@@
 An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
-
-### Additional observations
-
-Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
-Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work [e.g., @castro1998illiterate], but it is unclear what caused this pattern in the current study: Castro and colleagues' [-@castro1998illiterate] study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
-Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
+As for different strengths of age effects, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
 
@@
 Because these latter predictors strongly relate to other language outcomes, the present findings raise the issue of why NWR would pattern differently, what that could tell us about the relationship between lexical development, phonological development, and the input environment and, last but not least, what is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders.
-In the meanwhile, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a connection between individual learners and global language patterns when it comes to acoustic and articulatory markedness.
+In the meanwhile, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
 

### 2021-01-29 manuscript.tex
new version for Middy

@@
   tell me if you think you CAN'T LIVE with my turning this in with figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
+\item
+  do you mind if I change all the figures to say NWR scores rather than NWR Scores? In the text, we use lower case.
 \end{itemize}
@@
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
-The study was designed to contribute to four aspects of our understanding of phonological development.
-
-First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in the perception-production loop involved in NWR.
+The study was designed to contribute to four broad research areas.
+
+The first research area is at the crossing of typology and phonological development.
 Previous work using NWR has preferred relatively universal and early-acquired phonemes (with the possible exception of Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely language.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
-
-Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development.
-Our reading of previous NWR research is that there are variable effects of length between populations.
-For instance, Jaber-Awida (2018) reports an average of \textasciitilde{}96\% correct repetition for items 2 syllables long among children learning an Arabic variety of Israeli at about 5.5 years of age, but \textasciitilde{}81\% for items 3 syllables long.
-In contrast, Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019) observe no decline in performance in similarly-aged Italian learners, with a score of 84\% for 2 syllables versus 85\% for 3 syllables.
-It is possible that differences are due to a host of variables, including the modal length of words in the language and/or in child-directed speech in that culture.
+Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.
+
+Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
+Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, \& Leonard, 2006), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
+It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
-To attempt to see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic Jaber-Awida (2018); Cantonese (Stokes, Wong, Fletcher, \& Leonard, 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
+To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
+We found data for learners of Israeli Arabic Jaber-Awida (2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
 
@@
 
-Our reading of the previous literature is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages.
-For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (28\%) and Arabic (15\%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only 2\%.
-Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese 8\% versus English 1\%).
+Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages.
+For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (\textasciitilde{}28\%) and Arabic (\textasciitilde{}15\%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only \textasciitilde{}2\%.
+Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese \textasciitilde{}8\% versus English \textasciitilde{}1\%).
 Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
 
-Third, there are ongoing discussions as to what the key factors structuring individual variation are.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
+The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
@@
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
-Indeed, NWR has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), and/or outside urban settings (except for in Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b);Maddieson (2013a); no exceptions to our knowledge{]}.
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b);Maddieson (2013a); no exceptions to our knowledge{]}.
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
-And it does occur that some researchers opt instead to employ dimensions of variation that are more relevant to their language, such as adaptations in Chinese languages that have items varying in tone REF.
-\textbf{return to this after reading lit}
 
@@
 We could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory which lead any error to result in many true-word phonetic neighbors.
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
 
@@
 
-\begin{verbatim}
-## [1] 30 15
-\end{verbatim}
-
 To address whether errors were better predicted by cross-linguistic or within-language frequency, we calculated a proportion of productions that were correct for each phone (regardless of the type of error or the substitution pattern).
 Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other.
-Although a direct test is missing, the correlation with cross-linguistic frequency {[}r()=0.76, p \textless{} 0.001{]} was greater than that with within-language frequency {[}r()=0.45, p = 0.05{]}.
+Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency {[}r(319.72)=0.76, p \textless{} 0.001{]} was greater than that with within-language frequency {[}r(731.10)=0.45, p = 0.05{]}.
 
@@
 
-\textbf{FIX LOGIC IN THIS PARA} Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
+Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
 Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
-Typologically common sounds are associated with higher performance on a handful of other tasks (\textbf{REFS -- M2A: Alex, I added this based on your note, where it sounded like you had some particular studies in mind?}) though to our knowledge this has not yet been tested with non-word repetition.
-Regarding Yélî Dnye in particular, the phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
-We therefore expected to see that, while NWR scores would be lower for stimuli with lower average frequency, this effect would be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
-Indeed, we found a robust effect of average segmental frequency on NWR performance: Even accounting for age and random effects of item and participant, we see that target words with more frequent segments were repeated correctly more often.
-This effect is large, with a magnitude more than twice the size of the effect of participant age.
-This significant effect remains even once also accounting for the frequencies of these segments in Yélî Dnye child-directed speech, which are correlated with their typological frequencies.
-In sum, typological frequency effects, which have been found in other measurements of phonological processing.
-appear to strongly affect NWR performance, and do not appear mitigated by language-specific pressure to make finer-grained differences earlier in development.
-
-\textbf{REWRITE THIS PARA} \textasciitilde{}With respect to the types of errors in repetition made, we did not see clear patterns to further guide our discussion: base rates of deletion and substitution were fairly low and the relative distribution of errors over, e.g., nasal vs.~oral vowels and simple vs.~complex consonants, revealed no remarkable bias in error types.\textasciitilde{} That said, the lack of a difference could be due to relative imbalance across our stimuli in the use of these phonemic features (e.g., we included many more more oral than nasal targets) and future work should investigate such sources of error bias more systemtically.
+It would stand to reason that typologically common sounds are associated with higher performance, but to our knowledge this has not yet been tested with NWR.
+There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
+One could then predict that this effect should be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
+
+And yet, we found a robust effect of average segmental cross-linguistic frequency on NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
+This effect was large, with a magnitude more than twice the size of the effect of participant age.
+Moreover, this significant effect remained even once also accounting for the frequencies of these segments in Yélî Dnye children's input.
+An analysis of the substitutions made by children also aligned with this interpretation, with more common sounds being substituted for less common ones.
+
+We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development.
+We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (most often discussed in the case of phonotactic constraints on sequences, e.g., Maddieson, 2009), but we do point out that this effect is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
+We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the fine-grained phonetic representations in the perception-production loop (as in e.g., Edwards, Beckman, \& Munson, 2004).
 
@@
 We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item.
-Based on previous work, we had predicted that children would have higher NWR scores for shorter items.
-That said, previous work has shown both very small (Piazzalunga et al., 2019) and very large (Cristia et al., 2020; Jaber-Awida, 2018) effects of stimulus length and, further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic words are nearly equally frequent in the current language, with trisyllabic and longer words making up a non-trivial 10\% of the remaining words.
-Compare this to, for example, English, which is substantially more skewed toward monosyllabic word forms \textbf{M2A: Alex I'm going off your note here (\enquote{Prediction for Yélî made before seeing the data: The length distribution in Yélî words is more balanced than that in English, and thus the score decline for poly- versus mono-syllables may be less pronounced than that for English.}").
-I don't have a reference for this, can you please finish the thought or nix this bit?}.
-Setting aside our monosyllabic stimuli, which all contained typologically infrequent segments, we can examine effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
-While indeed NWR scores were overall lower for longer items (e.g., see Figure \ref{fig:fig-prevlit}), the effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
-In light of mixed prior results of item length, we propose two possible (and non-mutually exclusive) explanations for this minimal impact of item length.
-First, further extensions of this type of analysis in more populations may reveal that, in general (and cross-linguistically), item length effects are variable between languages, potentially reflecting the distribution of word lengths in the ambient language and other (morpho-)phonological tendencies in the lexicon.
-Second, above and beyond these language-specific effects, the general impact of item length on NWR score may be relatively small, as shown in Piazzalunga et al.'s (2019) study on Italian and as borne out in the current dataset once controlling for other factors.
-\textbf{ADD WHAT LANGUAGES WOULD BE IDEAL TO TEST THIS HYPOTHESIS}
+In broad terms, children should have higher NWR scores for shorter items.
+That said, previous work summarized in the Introduction has shown both very small (e.g., Piazzalunga et al., 2019) and very large (e.g., Cristia et al., 2020) effects of stimulus length.
+Setting aside our monosyllabic stimuli (which contained typologically infrequent segments with lower NWR scores, as just discussed), we examined effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
+The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant, and is small and inconsistent across ages (see Figure \ref{fig:fig-prevlit}).
+We do not have a good explanation for why samples in the literature vary so much in terms of the size of length effects, but two possibilities are that this is not truly a length effect but a confound with some other aspect of the stimuli, or that there is variation in phonological representations that is poorly understood.
+We explain each idea in turn.
+
+First, it remains possible that apparent length effects are actually due to uncontrolled aspects of the stimuli.
+For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants -- which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words.
+Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower.
+If this idea is correct, a careful analysis of non-words used in previous work may reveal that studies with larger length effects just happened to have longer non-words with lower n-phone frequencies.
+
+Second, NWR is often described as a task that tests flexible perception-production, and as such it is unclear why length effects should be observed at all.
+However, it is possible that NWR relies on more specific aspects of perception-production, in ways that are dependent on stimulus length.
+A hint in this direction comes from work on illiterate adults, who can be extremely accurate when repeating short non-words, but whose NWR scores are markedly lower for longer items.
+In a longitudinal study on Portuguese-speaking adults who were learning to read, Kolinsky, Leite, Carvalho, Franco, and Morais (2018) found that, before reading training, the group scored 12.5\% on 5-syllable items, whereas after 3 months of training, they scored 62.5\% on such long items, whereas performance was at 100\% for monosyllables throughout.
+Given that as adults they had fully acquired their native language, and obviously they had flexible perception-production schemes that allowed them to repeat new monosyllables perfectly, the change that occurred in those three months must relate to something else in their phonological skills, something that is not essential to speak a language natively.
+Thus, we hazard the hypothesis that sample differences in length effects may relate to such non-essential skills.
+Since as stated this hypothesis is under-specified, further conceptual and empirical work are both needed.
 
@@
 
-A review of previous work (see Introduction) suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR.
-We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta-analytic efforts aggregating over smaller studies such as ours.
-
-Following prior work, we expected that NWR scores would increase with participant age (Farmani et al., 2018; Kalnak et al., 2014; Vance et al., 2005).
+Our review of previous work in the Introduction suggested that our anticipated sample size would not be sufficient to detect most individual differences using NWR.
+We give a brief overview of individual difference patterns of four types in the present data---age, sex, birth order, and maternal education---hoping that these findings can contribute to future meta- or mega-analytic efforts aggregating over studies.
+
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese Stokes et al., 2006; but note Cristia et al., 2020 is an exception).
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or participant gender (Chiat \& Roy, 2007) on NWR scores.
-In addition to this prior work, education on Rossel Island, while generally highly valued, is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation.
-There is also limited variation in maternal education across the families in the region of the island where we sampled.
-We therefore expected little evidence for impact of either participant gender or maternal education in the present study.
-On the other hand, these predictors have established effects on other language development measures (\textbf{REFS: M2A: Alex go ahead and pick your faves here}).
-So to the extent that NWR scores share causal links to gender-based differences in development and maternal linguistic input with these other language outcome measures, we might then expect these factors to appear in NWR data.
-In fact, participant gender and maternal education correlated with NWR score at about r\textasciitilde{}.1, which is small.
-
-Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019), presumably because later-born children receive a smaller share of maternal input than their older siblings.
+In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) on NWR scores.
+We did not expect large effects of maternal education in our sample because of two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
+In fact, maternal education correlated with NWR score at about r\textasciitilde{}.1, which is small.
+Similarly, NWR scores may not vary greatly with participant gender according to previous work (Chiat \& Roy, 2007), and for that as well we find effects of about that size.
+
+Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
@@
 In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score.
-However, the effect size was larger than that found for the other factors, at r\textasciitilde{}.2, and thus we believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
+However, the effect size was larger than that found for the other two factors and it is far from negligible, at r\textasciitilde{}.2 or Cohen's d\textasciitilde{}0.41.
+In fact, two large studies with therefore precise estimates found effects of about d\textasciitilde{}.2 (Barclay, 2015; Havron et al., 2019), which would suggest the effects we found are larger.
+We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
 
@@
 
-One of the questions in our mind when designing this study was whether NWR was a fair test of phonological development across languages and cultures.
-Although our data cannot answer this question because we have only sampled one language and culture here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
+The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, as when designing this study we wondered whether NWR was a \emph{fair} test of phonological development.
+Although our data cannot answer this question because we have only sampled one language and population here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
 It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
@@
 An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
-
-\hypertarget{additional-observations}{%
-\subsubsection{Additional observations}\label{additional-observations}}
-
-Some portion of the errors were introduced when the participant produced a real word (in Yélî Dnye or English) in response to the stimulus.
-Real-word repetitions here made up two thirds of errorful repetitions---this is quite high compared to past work (e.g., Castro-Caldas et al., 1998), but it is unclear what caused this pattern in the current study: Castro and colleagues' (1998) study focused on adults rather than children, the task was administered by a team including a foreign, English-speaking researcher, and the particularities of the Yélî Dnye phonological inventory result in many true-word phonetic neighbors.
-Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
+As for different strengths of age effects, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
 
@@
 
+\leavevmode\hypertarget{ref-edwards2004interaction}{}%
+Edwards, J., Beckman, M.
+E., \& Munson, B.
+(2004).
+The interaction between vocabulary size and phonotactic probability effects on children's production accuracy and fluency in nonword repetition, \emph{47}, 421--436.
+
 \leavevmode\hypertarget{ref-estes2007differences}{}%
@@
 
+\leavevmode\hypertarget{ref-kolinsky2018completely}{}%
+Kolinsky, R., Leite, I., Carvalho, C., Franco, A., \& Morais, J.
+(2018).
+Completely illiterate adults can learn to decode in 3 months.
+\emph{Reading and Writing}, \emph{31}(3), 649--677.
+\url{https://doi.org/10.1007/s11145-017-9804-7}
+
 \leavevmode\hypertarget{ref-lancy2015anthropology}{}%
@@
 
+\leavevmode\hypertarget{ref-maddieson2009phonology}{}%
+Maddieson, I. (2009). Phonology, naturalness and universals. \emph{Poznán Studies in Contemporary Linguistics}, \emph{45}(1), 131--140.
+
 \leavevmode\hypertarget{ref-maddieson2013consonant}{}%
@@
 Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
+
+\leavevmode\hypertarget{ref-moreton2012structure}{}%
+Moreton, E., \& Pater, J. (2012). Structure and substance in artificial-phonology learning, part ii: Substance. \emph{Language and Linguistics Compass}, \emph{6}(11), 702--718.
 

### 2021-02-03 manuscript.Rmd
comments and minor edits up to Methods

@@
 title             : "Non-word repetition in Yélî Dnye"
-shorttitle        : "NWE in Yélî Dnye"
+shorttitle        : "NWR in Yélî Dnye"
 
@@
 
+## TODO Alex
+- Shorten the abstract a little please (it's very clearly written now but probably too long)
+- Other keywords for AC to consider adding: Papuan, non-industrial/non-urban, comparative, typology/markedness
+- Changed "Scores" to "scores" in the figures—check if I missed anything!
+
 ## TODO Middy
@@
 - tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
-- do you mind if I change all the figures to say NWR scores rather than NWR Scores? In the text, we use lower case.
 
@@
 The first research area is at the crossing of typology and phonological development.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely language.
+Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skill <!--Previous sentence change: "language" >> "linguistic skill-->.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
@@
 To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic  @jabere2018xperiment; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
+We found data for learners of Israeli Arabic  [@jabere2018xperiment]<!--Previous sentence change: added square brackets around jaber-awida ref-->; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
 
@@
 
-plot(1,xlab="Average age (years)",ylab="NWR Scores (words)",ylim=c(0,100),xlim=range(pl_stacked$Age))
+plot(1,xlab="Average age (years)",ylab="NWR scores (words)",ylim=c(0,100),xlim=range(pl_stacked$Age))
 
@@
 
-Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages.
+Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line <!--Changed previous bit: "lign" >> "line"--> up with expected word length in different languages.
 For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%.
 Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).
-Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
+Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses <!--Changed previous bit: "approaches" >> "analyses"-->.
 
@@
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel;@maddieson2013consonant; no exceptions to our knowledge].
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge] <!--Changed previous bit to fix the paren display w/ the refs-->.
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are typically easily accessible.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible <!--Changed previous bit: "typically easily" >> "more generally"-->.
 Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
@@
 In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
-Thus, with respect to our first research goal, Yélî Dnye seemed is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
+Thus, with respect to our first research goal, Yélî Dnye is <!--Changed previous bit: "seemed is" >> "is"-->a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
 
@@
 
-Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun [e.g., @peuteIPconsonants].
-We hope the present study contributes to filling this gap.
-**TODO incorporate brief summary of paper**
+Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun.
+For example, Peute and colleagues' [-@peuteIPconsonants] find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones<!--Changed previous bit: added what I think the relevant bit from Bram's paper is -- change as needed!-->.
+We hope the present study contributes to this growing line of work<!--Changed previous bit: "filling this gap" >> "this growing line of work"-->.
 
@@
 
-Some aspects of the community are relevant for interpreting results found when addressing our thir research question, regarding sources of individual variation.
+Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation<!--Changed previous sentence: new wording-->.
 Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
@@
 
-The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children -- their own and billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some<!--Minor wording change: " -- their own and " >> "--their own and possibly some"--> billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
 These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
-Land attribution for building one's home is decided collectively based on land availability, and typically does not take into consideration an individual's desire to be close to a school.
-
-Most Yélî parents are swidden horticulturalists, and those who are not may not reside in the island.
-Within a group of households, it is often the case that most older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).
-Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+Land attribution for building one's home is decided collectively based on land availability<!--I removed this bit since I think there's some more nuance involved here and it seems better to not get into it: ", and typically does not take into consideration an individual's desire to be close to a school.
+"-->
+
+Most Yélî parents are swidden horticulturalists<!--I removed this bit because I don't know what you mean: , and those who are not may not reside in the island.--> Within a group of households, it is often the case that <!--removed "most" here-->older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).<!--Mmmmm this is not true.
+There's a great deal of socializing time by men and women during the day that isn't centered on some kind of productive domestic activity (maybe it'd be too far to call it leisure time, but it's not that off-the-mark!) The description you give suggests that they're too busy to hang out with the kids and that isn't an accurate characterization of what's going on.
+It's also the case that kids start doing significant and fairly independent domestic work before older adolescence, especially if they're not in school at the moment.
+I think, e.g., both Yaawaa and Taakeme had their own starter gardens to manage before age 10 and were expected to contribute to family dinner from their bounty :)--> Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
 Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early].
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
-Children start school often at around age six, although the precise age depends on the child's apparent development.
+Children start school often at around age six or seven <!--added "or seven"-->, although the precise age depends on the child's readiness, as judged by their teacher<!--replaced "apparent development" with "readiness, as judged by their teacher"-->.
 
@@
 Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
-Together with the fact that land attribution is essentially random with respect to educational hopes, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
+Together with the fact that land attribution for hamlet development is not made on the basis of proximity to a school<!--changed the wording of the prior sentence-->, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
 
@@
 It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
-This contrasts with this picture just drawn in the Yélî community, where children regardless of their birth order in their nuclear family will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders.
+This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family<!--moved the "regardless of their birth order in their nuclear family" phrase to the end-->.
 
@@
 However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
-The Yélî community can help us bring further light into this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early].
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.7 minutes per hour [@casillas2020early].
+The Yélî community can help us shed further light onto<!--minor wording change to "light" metaphor--> this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early].
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early].
 Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in @cristia2020infant.
@@
 plot(pl_overall$Overall~pl_overall$Age,col=mycols[pl_overall$Language],pch=1,
-     xlab="Average age (years)",ylab="NWR Scores (words)",ylim=c(0,100))
+     xlab="Average age (years)",ylab="NWR scores (words)",ylim=c(0,100))
 

### 2021-02-03 manuscript.Rmd
comments and minor edits up to Discussion

@@
 - Changed "Scores" to "scores" in the figures—check if I missed anything!
+- Check whether you're okay with the mini summary I put in on Peute's study
 
@@
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
-Children start school often at around age six or seven <!--added "or seven"-->, although the precise age depends on the child's readiness, as judged by their teacher<!--replaced "apparent development" with "readiness, as judged by their teacher"-->.
+Children start school often at around age seven <!--changed "six" to "seven"-->, although the precise age depends on the child's readiness, as judged by their teacher<!--replaced "apparent development" with "readiness, as judged by their teacher"-->.
 
@@
 
-A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 2- to 4-syllable sequences.
+A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible<!--added "resulting possible"--> 2- to 4-syllable sequences.
 These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
-The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additinally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
+The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally<!--fixed sp: "additionally"--> let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
 Any item reported to have a meaning or a strong association with another word form or meaning was excluded.
@@
 A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
-The 40 elicitation sets are available online from [osf.io/dtxue/](https://osf.io/dtxue/).
+The 40 elicitation sets are available online from [osf.io/dtxue/](https://osf.io/dtxue/)<!--link requires login; is that as you wish?-->.
 The split had the following constraints:
@@
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors).
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors<!--removed stranded ")"-->.
 Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
@@
 
-We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlets.
+We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions<!--specified "regions" here since technically these kids weren't all from the hamlet they were tested in-->.
 We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease).
 The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).
-In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
-These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
+In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth<!--fixed sp "fourth"-->,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
+These children were tested in a hamlet far from our research base<!--replaced "remote hamlet" with "hamlet far from our research base"-->, and we unfortunately did not ask about birth order before leaving the site.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and<!--chabged comma to "and"--> `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
 
 [^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which much of<!--added "much of"--> the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
 Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
@@
 The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`.
-The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.
+The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted or deleted<!--removed ", inserted," in line with what you mention about no insertions above-->.
 
@@
 
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
-Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`].
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency<!--fixed sp "frequency"--> could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
+Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies (r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`)<!--changed [ ] to ( )-->; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies (r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`)<!--changed [ ] to ( )-->.
 Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r round(cor_ty_nwr$estimate,3)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`.
@@
 
-We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels.
+We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not, generating separate estimates for nasal and oral vowels<!--minor wording change to prev sentence-->.
 The nasal vowels in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`%);
@@
 
-Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving oral vowels (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
+Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving oral vowels<!--should "oral vowels" here by "simple consonants"??--> (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other.
-Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency [r(`r phone_cor_xl$statistic`)=`r round(phone_cor_xl$estimate,3)`, p `r pval.display(round(phone_cor_xl$p.value,3))`]  was greater than that with within-language frequency [r(`r phone_cor_wl$statistic`)=`r round(phone_cor_wl$estimate,3)`, p `r pval.display(round(phone_cor_wl$p.value,3))`].
+Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency (r(`r phone_cor_xl$statistic`)=`r round(phone_cor_xl$estimate,3)`, p `r pval.display(round(phone_cor_xl$p.value,3))`)<!--changed [ ] to ( )--> was greater than that with within-language frequency (r(`r phone_cor_wl$statistic`)=`r round(phone_cor_wl$estimate,3)`, p `r pval.display(round(phone_cor_wl$p.value,3))`)<!--I'm starting to think you want all these square brackets in...
+in which case, sorry :)-->.
 
@@
 
-### Factor structuring individual variation
+### Factors<!--made "factor" plural--> structuring individual variation
 

### 2021-02-03 manuscript.Rmd
comments and minor edits through the end

@@
 ## TODO Alex
+- Check for my comments and minor changes throughout in <!----> format (I didn't recompile anything)
 - Shorten the abstract a little please (it's very clearly written now but probably too long)
@@
 - Check whether you're okay with the mini summary I put in on Peute's study
-
-## TODO Middy
-
-- read over whole thing -- does the logic sound ok?
-- tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
+- Logic sounds great! The discussion is also much improved, thank you!
+- Please see my notes on the mini ethnographic info section b/c there are some inaccurate remarks right now
+- I only had one small comment on figure 5
+- I can live with figure 1 for now but you're right I don't love it :)
 
@@
 As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
-In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
+In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children<!--note this ends up saying data is missing for "15" children whereas the participants section mentions it's missing for "14" (`r sum(is.na(inc$birthOrder))` in this rmd l.491) -->, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
 
@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
-An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and other that will be discussed below).
+We aimed to provide additional data on two questions already visited<!--fixed sp "visited"--> in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
+An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the discussion below<!--minor rewording-->).
 We consider implications of our results on each of these four research areas in turn.
@@
 First, it remains possible that apparent length effects are actually due to uncontrolled aspects of the stimuli.
-For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants -- which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words.
+For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants,<!--substituted "," for " -- "--> which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words.
 Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower.
@@
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
-Results separating different age groups are shown in \@ref(fig:fig-prevlit-overall).
+Results separating different age groups are shown in Figure<!--added "Figure"--> \@ref(fig:fig-prevlit-overall).<!--looks great!
+only let's make swedish a tinge darker gray or some kind of turqouise because it's confusable with mandarin now-->
 
@@
 That said, @cristia2020infant also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
-This is a very different case from the Yélî population studied here, where all adults have accumulated several years of schooling, and literacy is widespread.
+This is a very different case from the Yélî population studied here, where nearly<!--added "nearly"--> all adults have accumulated several years of schooling, and basic<!--added "basic"--> literacy in English (and sometimes Yélî Dnye)<!--added language info--> is widespread.
 If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.

### 2021-02-09 manuscript.Rmd
incorporated mcs changes

--- 2021-02-03 manuscript.Rmd
@@
 In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
-Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development.
-More specifically, this study contributes to three lines of research.
-First, we contribute to investigations on NWR across diverse languages, by documenting that, in Yélî Dnye, non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
-Second, contributing to mounting research suggesting that length effects may be language- or population-specific, we find rather weak effects of item length.
-Third, we add a datapoint on potential sources of individual variation effects, by establishing that in our sample age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
+Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development at a global and individual level via three contributions.
+First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
+Second, we documented rather weak effects of item length.
+Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
 Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.
 
-keywords          : "phonology, non-word repetition, development"
-wordcount         : "~9,000 words"
+keywords          : "phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness"
+wordcount         : "~11,00 words"
 
@@
 
-## TODO Alex
-- Shorten the abstract a little please (it's very clearly written now but probably too long)
-- Other keywords for AC to consider adding: Papuan, non-industrial/non-urban, comparative, typology/markedness
-- Changed "Scores" to "scores" in the figures—check if I missed anything!
-
-## TODO Middy
-
-- read over whole thing -- does the logic sound ok?
-- tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
-
 ## Introduction
@@
 The first research area is at the crossing of typology and phonological development.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skill <!--Previous sentence change: "language" >> "linguistic skill-->.
+Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skill.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
@@
 To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic  [@jabere2018xperiment]<!--Previous sentence change: added square brackets around jaber-awida ref-->; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
+We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
 
@@
 
-Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line <!--Changed previous bit: "lign" >> "line"--> up with expected word length in different languages.
+Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
 For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%.
 Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).
-Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses <!--Changed previous bit: "approaches" >> "analyses"-->.
+Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.
 
@@
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge] <!--Changed previous bit to fix the paren display w/ the refs-->.
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge].
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible <!--Changed previous bit: "typically easily" >> "more generally"-->.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
 Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
@@
 In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
-Thus, with respect to our first research goal, Yélî Dnye is <!--Changed previous bit: "seemed is" >> "is"-->a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
+Thus, with respect to our first research goal, Yélî Dnye is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
 
@@
 Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun.
-For example, Peute and colleagues' [-@peuteIPconsonants] find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones<!--Changed previous bit: added what I think the relevant bit from Bram's paper is -- change as needed!-->.
-We hope the present study contributes to this growing line of work<!--Changed previous bit: "filling this gap" >> "this growing line of work"-->.
+For example, Peute and colleagues' [-@peuteIPconsonants] find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
+We hope the present study contributes to this growing line of work.
 
@@
 
-Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation<!--Changed previous sentence: new wording-->.
+Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation.
 Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
@@
 
-The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some<!--Minor wording change: " -- their own and " >> "--their own and possibly some"--> billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
 These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
-Land attribution for building one's home is decided collectively based on land availability<!--I removed this bit since I think there's some more nuance involved here and it seems better to not get into it: ", and typically does not take into consideration an individual's desire to be close to a school.
-"-->
-
-Most Yélî parents are swidden horticulturalists<!--I removed this bit because I don't know what you mean: , and those who are not may not reside in the island.--> Within a group of households, it is often the case that <!--removed "most" here-->older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).<!--Mmmmm this is not true.
-There's a great deal of socializing time by men and women during the day that isn't centered on some kind of productive domestic activity (maybe it'd be too far to call it leisure time, but it's not that off-the-mark!) The description you give suggests that they're too busy to hang out with the kids and that isn't an accurate characterization of what's going on.
-It's also the case that kids start doing significant and fairly independent domestic work before older adolescence, especially if they're not in school at the moment.
-I think, e.g., both Yaawaa and Taakeme had their own starter gardens to manage before age 10 and were expected to contribute to family dinner from their bounty :)--> Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
+Land attribution for building one's home is decided collectively based on land availability.
+
+Most Yélî parents are swidden horticulturalists.
+Within a group of households, it is often the case that older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leaves them less time to spend directly with the children in their household than may be the case for stay-at-home parents in urban sites.
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
 Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early].
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
-Children start school often at around age six or seven <!--added "or seven"-->, although the precise age depends on the child's readiness, as judged by their teacher<!--replaced "apparent development" with "readiness, as judged by their teacher"-->.
+Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.
 
@@
 Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
-Together with the fact that land attribution for hamlet development is not made on the basis of proximity to a school<!--changed the wording of the prior sentence-->, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
+Overall, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
 
@@
 It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
-This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family<!--moved the "regardless of their birth order in their nuclear family" phrase to the end-->.
+This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
 
@@
 However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
-The Yélî community can help us shed further light onto<!--minor wording change to "light" metaphor--> this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early].
+The Yélî community can help us shed further light onto this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early].
 Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early].
@@
 
-A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 2- to 4-syllable sequences.
+A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible 2- to 4-syllable sequences.
 These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
-The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additinally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
+The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
 Any item reported to have a meaning or a strong association with another word form or meaning was excluded.
@@
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors).
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
 Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
@@
 
-We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlets.
+We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions.
 We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease).
 The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).
-In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
-These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
+In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
+These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
 
 [^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
 Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
@@
 The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`.
-The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.
+The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted or deleted.
 
@@
 
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
 Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`].
@@
 
-We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels.
+We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not, generating separate estimates for nasal and oral vowels.
 The nasal vowels in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)`%);
@@
 
-Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving oral vowels (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
+Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonats (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 
-### Factor structuring individual variation
+### Factors structuring individual variation
 
@@
 As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
-In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
+In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(inc$birthOrder))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
 
@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
-An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and other that will be discussed below).
+We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
+An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the discussion below).
 We consider implications of our results on each of these four research areas in turn.
@@
 First, it remains possible that apparent length effects are actually due to uncontrolled aspects of the stimuli.
-For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants -- which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words.
+For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants, which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words.
 Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower.
@@
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
-Results separating different age groups are shown in \@ref(fig:fig-prevlit-overall).
+Results separating different age groups are shown in Figure> \@ref(fig:fig-prevlit-overall).
 
@@
             "Spanish"   ="brown",
-            "Swedish"   ="darkgray",
+            "Swedish"   ="red",
             "Tsimane'" = "darkgreen",
@@
 That said, @cristia2020infant also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
-This is a very different case from the Yélî population studied here, where all adults have accumulated several years of schooling, and literacy is widespread.
+This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
 If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.

### 2021-02-09 manuscript.tex
incorporated mcs changes

@@
             pdftitle={Non-word repetition in Yélî Dnye},
-            pdfkeywords={phonology, non-word repetition, development},
+            pdfkeywords={phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness},
             pdfborder={0 0 0},
@@
 \makeatother
-\shorttitle{NWE in Yélî Dnye}
+\shorttitle{NWR in Yélî Dnye}
 \author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
@@
 Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
-\keywords{phonology, non-word repetition, development\newline\indent Word count: ~9,000 words}
+\keywords{phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness\newline\indent Word count: ~9,000 words}
 \usepackage{lineno}
@@
 In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
-Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation in language development.
-More specifically, this study contributes to three lines of research.
-First, we contribute to investigations on NWR across diverse languages, by documenting that, in Yélî Dnye, non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
-Second, contributing to mounting research suggesting that length effects may be language- or population-specific, we find rather weak effects of item length.
-Third, we add a datapoint on potential sources of individual variation effects, by establishing that in our sample age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
+Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation in language development at a global and individual level via three contributions.
+First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
+Second, we documented rather weak effects of item length.
+Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
 Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.
@@
 \maketitle
-
-\hypertarget{todo-middy}{%
-\subsection{TODO Middy}\label{todo-middy}}
-
-\begin{itemize}
-\tightlist
-\item
-  read over whole thing -- does the logic sound ok?
-\item
-  tell me if you think you CAN'T LIVE with my turning this in with figures 1 and 5 as they are (I promise I'll improve them in the revision stage)
-\item
-  do you mind if I change all the figures to say NWR scores rather than NWR Scores? In the text, we use lower case.
-\end{itemize}
 
@@
 The first research area is at the crossing of typology and phonological development.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes (with the possible exception of Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely language.
+Previous work using NWR has preferred relatively universal and early-acquired phonemes (with the possible exception of Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely linguistic skill.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
@@
 To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic Jaber-Awida (2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
+We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
 
@@
 
-Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically lign up with expected word length in different languages.
+Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
 For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (\textasciitilde{}28\%) and Arabic (\textasciitilde{}15\%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only \textasciitilde{}2\%.
 Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese \textasciitilde{}8\% versus English \textasciitilde{}1\%).
-Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic approaches.
+Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.
 
@@
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b);Maddieson (2013a); no exceptions to our knowledge{]}.
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities Maddieson, 2013b, and @maddieson2013consonant with no exceptions to our knowledge).
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are typically easily accessible.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
 Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
@@
 In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.).
-Thus, with respect to our first research goal, Yélî Dnye seemed is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
+Thus, with respect to our first research goal, Yélî Dnye is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
 
@@
 
-Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun (e.g., Peute, Fikkert, \& Casillas, n.d.).
-We hope the present study contributes to filling this gap.
-\textbf{TODO incorporate brief summary of paper}
+Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun.
+For example, Peute and colleagues' (n.d.) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
+We hope the present study contributes to this growing line of work.
 
@@
 
-Some aspects of the community are relevant for interpreting results found when addressing our thir research question, regarding sources of individual variation.
+Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation.
 Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
@@
 
-The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children -- their own and billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
 These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
-Land attribution for building one's home is decided collectively based on land availability, and typically does not take into consideration an individual's desire to be close to a school.
-
-Most Yélî parents are swidden horticulturalists, and those who are not may not reside in the island.
-Within a group of households, it is often the case that most older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).
+Land attribution for building one's home is decided collectively based on land availability.
+
+Most Yélî parents are swidden horticulturalists.
+Within a group of households, it is often the case that older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them less time to spend directly with the children in their household than may be the case in infants reared by stay-at-home parents in urban sites.
 Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
@@
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
-Children start school often at around age six, although the precise age depends on the child's apparent development.
+Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.
 
@@
 Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
-Together with the fact that land attribution is essentially random with respect to educational hopes, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
+Overall, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
 
@@
 It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
-This contrasts with this picture just drawn in the Yélî community, where children regardless of their birth order in their nuclear family will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders.
+This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
 
@@
 (2020) also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
-The Yélî community can help us bring further light into this question because direct speech to children under 3;0 is relatively infrequent in this community too (Casillas et al., 2020).
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.7 minutes per hour (Casillas et al., 2020).
+The Yélî community can help us shed further light onto this question because direct speech to children under 3;0 is relatively infrequent in this community too (Casillas et al., 2020).
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas et al., 2020).
 Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in Cristia et al.
@@
 
-A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/, /n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 2- to 4-syllable sequences.
+A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/, /n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible 2- to 4-syllable sequences.
 These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
-The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additinally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
+The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
 Any item reported to have a meaning or a strong association with another word form or meaning was excluded.
@@
 
-We tested a total of 55 children from 38 families spread across four hamlets.
+We tested a total of 55 children from 38 families spread across four hamlet regions.
 We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5), or were 13 years old or older (N=2; we tested these teenagers to put younger children at ease).
 The remaining 40 children (14 girls) were aged from 3 to 10 years (M = 6.50 years, SD = 1.50 years).
-In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
-These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site.
+In terms of birth order, 6 were first borns, 5 second, 2 third, 7 fourth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
+These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
 Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
-We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 only exposed to Yélî Dnye at home, 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
+We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
 Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.}
@@
 The overall NWR score was M = 65\% (SD = 15\%), Cohen's d = 4.39.
-The phoneme-based normalized Levenshtein distance was M = 21\% (SD = 9\%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.
+The phoneme-based normalized Levenshtein distance was M = 21\% (SD = 9\%), meaning that about a fifth of phonemes were substituted or deleted.
 
@@
 
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequence could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
 Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies {[}r(27)=0.50, p \textless{} 0.01{]}; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies {[}r(38)=0.73, p \textless{} 0.001{]}.
@@
 
-We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not separately for nasal and oral vowels.
+We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not, generating separate estimates for nasal and oral vowels.
 The nasal vowels in our stimuli occur in \textasciitilde{}1.40\% of languages' phonologies (range 0\% to 3\%);
@@
 
-Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving oral vowels (50.90 versus 8.20).
+Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving simple consonats (50.90 versus 8.20).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 
-\hypertarget{factor-structuring-individual-variation}{%
-\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}
+\hypertarget{factors-structuring-individual-variation}{%
+\subsubsection{Factors structuring individual variation}\label{factors-structuring-individual-variation}}
 
@@
 As shown in Figure \ref{fig:fig-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (5,649.08) = 0.47, p \textless{} 0.01).
-In contrast, there was no clear association between NWR scores and sex (Welch t (27.33) = -0.60, p = 0.56), birth order (data missing for 15 children, rho = (3,502.90) = -0.20, p = 0.33), or maternal education (rho ( 9,628.60) = 0.10, p = 0.55).
+In contrast, there was no clear association between NWR scores and sex (Welch t (27.33) = -0.60, p = 0.56), birth order (data missing for 14 children, rho = (3,502.90) = -0.20, p = 0.33), or maternal education (rho ( 9,628.60) = 0.10, p = 0.55).
 
@@
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
-Results separating different age groups are shown in \ref{fig:fig-prevlit-overall}.
+Results separating different age groups are shown in Figure\textgreater{} \ref{fig:fig-prevlit-overall}.
 
@@
 (2020) also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
-This is a very different case from the Yélî population studied here, where all adults have accumulated several years of schooling, and literacy is widespread.
+This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
 If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
@@
 \leavevmode\hypertarget{ref-maddiesonIPphoneticsYD}{}%
-Maddieson, I., \& Levinson, S. C. (n.d.). \emph{The phonetics of yélî dnye, the language of rossel island}.
+Maddieson, I., \& Levinson, S. C. (n.d.). \emph{The phonetics of Yélî Dnye, the language of Rossel Island}.
 
@@
 \leavevmode\hypertarget{ref-moreton2012structure}{}%
-Moreton, E., \& Pater, J. (2012). Structure and substance in artificial-phonology learning, part ii: Substance. \emph{Language and Linguistics Compass}, \emph{6}(11), 702--718.
+Moreton, E., \& Pater, J. (2012). Structure and substance in artificial-phonology learning, part II: Substance. \emph{Language and Linguistics Compass}, \emph{6}(11), 702--718.
 

### 2021-02-10 manuscript.Rmd
minor spelling things

@@
 The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword].
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword].
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
@@
 
+<!--M2A: We mention "three" contributions in the abstract but then say above "The study was designed to contribute to four broad research areas." and add this as "fourth" in our Introduction.
+This is minor, but feels slightly inconsistent.
+Happy for you to handle it as you most prefer!-->
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge].
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge).
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
 That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
-In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating].
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance, better than rhyme awareness [e.g., @gathercole1991differentiating].
 Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills.
@@
 To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned.
-Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
+Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
 This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
@@
 Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
 Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
@@
 Because there is still much to discover about the phonology and phonetics of Yélî Dnye [@levinsonYDgrammar], it was also possible that we initially generated items with illegal, but currently undocumented constraints.
-Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
+Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
 These candidates were then presented to one informant, for a final check that they did not mean anything.
@@
 
-The final list includes three practice items and 40 test items (across infants): 16 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
+The final list includes three practice items and 40 test items (across infants): 16 monosyllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
 
@@
 
-A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
+A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sub-lists, to generate 40 different elicitation sets.
 The 40 elicitation sets are available online from [osf.io/dtxue/](https://osf.io/dtxue/).
@@
 - The same three items were selected as practice items and used in all 40 elicitation sets.
-- Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sublists).
-Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 vs.
+- Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sub-lists).
+Since some of these groups had an odd number of items, one of the sub-lists was slightly longer than the other (20 vs.
 23).
-- Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
+- Once the sub-list split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sub-list selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
 
@@
 
-Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonats (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
+Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonants (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 First, it remains possible that apparent length effects are actually due to uncontrolled aspects of the stimuli.
-For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants, which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words.
-Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower.
+For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants, which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of sub-sequences in the non-words.
+Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (elatedly) their n-phone frequency is overall lower.
 If this idea is correct, a careful analysis of non-words used in previous work may reveal that studies with larger length effects just happened to have longer non-words with lower n-phone frequencies.
@@
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
-In brief, our results underscore the idea that phoonlogical development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].
+In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].
 
@@
 To be more certain whether language characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
-We believe this will be excedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
 An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
@@
 While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that vaies depending on the linguistic, cultural, and sociodemographic properties of the population under study.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
 Because these latter predictors strongly relate to other language outcomes, the present findings raise the issue of why NWR would pattern differently, what that could tell us about the relationship between lexical development, phonological development, and the input environment and, last but not least, what is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders.

### 2021-02-10 manuscript.Rmd
minor wordings (incl. new ethngphy bit discussed via email)

@@
 In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
-Thus, with respect to our first research goal, Yélî Dnye is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
+Thus, with respect to our first research goal, Yélî Dnye is a good language to test because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
 
@@
 Most Yélî parents are swidden horticulturalists.
-Within a group of households, it is often the case that older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leaves them less time to spend directly with the children in their household than may be the case for stay-at-home parents in urban sites.
+Within a group of households, it is often the case that older adolescents and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
 Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
@@
 In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
-For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
+For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have (unconditioned) contrastive vowel length.
 There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
@@
 The headsets were designed for adult use and could not be comfortably seated on many children's heads without a more involved adjustment period.
-To minimize adjustment time, which was uncomfortable for some children given the proximity of the experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth.
+To minimize adjustment time, which was uncomfortable for some children given the proximity of the foreign experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth.
 A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement.
@@
 These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` children only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
 

### 2021-02-10 manuscript.Rmd
more minor wordings (that's it!)

@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
-An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the discussion below).
+We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
+An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the Discussion below).
 We consider implications of our results on each of these four research areas in turn.
@@
 In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores.
-We did not expect large effects of maternal education in our sample because of two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
+We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
 In fact, maternal education correlated with NWR score at about r~.1, which is small.
@@
 We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages.
-Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye by the present study.
+Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
-Results separating different age groups are shown in Figure> \@ref(fig:fig-prevlit-overall).
+Results separating different age groups are shown in Figure \@ref(fig:fig-prevlit-overall).
 
@@
 
-Several observations can be drawn from this Figure.
+Several observations can be drawn from this figure.
 To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
@@
 If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
-Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in \@ref(fig:fig-prevlit-overall).
+Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall).
 These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
@@
 We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
-Neither of these two predictors explained variance  in \@ref(fig:fig-prevlit-overall).
+Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall).
 It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.

### 2021-02-11 manuscript.Rmd
proofing

@@
 ---
-title             : "Non-word repetition in Yélî Dnye"
-shorttitle        : "NWR in Yélî Dnye"
+title             : "Non-word repetition in children learning Yélî Dnye"
+shorttitle        : "NWR in Yélî Dnye learners"
 
@@
   - id            : "1"
-    institution   : "Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University"
+    institution   : "Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University"
   - id            : "2"
@@
 abstract: >
-   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible.
+   In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible.
 NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
 In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
-Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development at a global and individual level via three contributions.
+Results make three contributions that are specific, and a fourth that is general.
 First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
@@
 Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
-Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.
+Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.
 
 keywords          : "phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness"
-wordcount         : "~11,00 words"
+wordcount         : "~11,500 words"
 
@@
 
+# Seed for random number generation
+set.seed(42)
+knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
+
 #Sys.setlocale("LC_CTYPE", "en_US_POSIX.UTF-8")
@@
 library(openxlsx)
+
+printr=function(x) sub("0.", ".", round(x,3))
 
@@
   d/sqrt(d^2 +a)}
+
+sink("lastknit_session_info.txt")
+sessionInfo()
+sink()
 ```
@@
 Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development].
-Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks.
+Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
 In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
 After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
-NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
-NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
+NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001non-word], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences].
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language].
-In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
-The study was designed to contribute to four broad research areas.
+NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
+In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
+The study was designed to contribute to four broad research areas, three via direct results.
 
 The first research area is at the crossing of typology and phonological development.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skill.
+Previous work using NWR has preferred relatively universal and early-acquired phonemes [with exceptions including @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skills.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
@@
 Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
-Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory].
+Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006non-word], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory].
 It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
@@
 To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
+We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006non-word]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
 
@@
 Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error.
-Age has been slightly jittered for ease of inspection of different lengths at a given age."}
+Age has been slightly shifted for ease of inspection of different lengths at a given age."}
 
@@
 The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword].
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021non-word].
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
-Previous work typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
-Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [@havron2019effect], and therefore we report on these too.
-
-<!--M2A: We mention "three" contributions in the abstract but then say above "The study was designed to contribute to four broad research areas." and add this as "fourth" in our Introduction.
-This is minor, but feels slightly inconsistent.
-Happy for you to handle it as you most prefer!-->
+Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014non-word; @vance2005speech].
+Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014non-word; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
+Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.
+
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge).
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant; with no exceptions to our knowledge].
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
 That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
-In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance, better than rhyme awareness [e.g., @gathercole1991differentiating].
-Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills.
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance [e.g., @gathercole1991differentiating].
+Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills.
 Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
-To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned.
+Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).
 Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
-This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
-
-Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
+This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.
+
+Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
 As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
@@
 
-Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
+Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
 The vast majority of syllables use a CV format.
@@
 In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
-For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have (unconditioned) contrastive vowel length.
+For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
 There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 40 final items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014non-word], with the entire test inventory of 40 final items distributed across children.
 
@@
 We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory which lead any error to result in many true-word phonetic neighbors.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
 Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
@@
 Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \@ref(fig:fig-xling-freq).
-Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r round(xlfr_r$estimate,3)`.
+Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(`r xlfr_r$parameter`) = `r printr(xlfr_r$estimate)`.
 
@@
 Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`].
-Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r round(cor_ty_nwr$estimate,3)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`.
+Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(`r cor_ty_nwr$parameter`)=`r printr(cor_ty_nwr$estimate)`, p `r pval.display(round(cor_ty_nwr$p.value,3))`.
 Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
@@
 For consonants, we inspected complex (`r print.phone.list(complexCs.ipa)`) versus simpler ones (`r print.phone.list(simplexCs.ipa)`), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.^[Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., [tʃ]).
-These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.
-This result suggests that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.]
+These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.]
 The complex consonants in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("complex")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("complex")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("complex")],na.rm=T)`%);
@@
 
-Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonants (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`).
+Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonants (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`%).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question.
-As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`).
-In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(inc$birthOrder))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
+As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, ρ(`r corage$statistic`) = `r printr(corage$estimate)`, p `r pval.display(corage$p.value)`).
+In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(inc$birthOrder))` children, ρ(`r round(corbo$statistic,1)`) = `r printr(corbo$estimate)`, p `r pval.display(corbo$p.value)`), or maternal education  (ρ( `r round(corme$statistic,1)`) = `r printr(corme$estimate)`, p `r pval.display(corme$p.value)`).
 
@@
 Thus, we hazard the hypothesis that sample differences in length effects may relate to such non-essential skills.
-Since as stated this hypothesis is under-specified, further conceptual and empirical work are both needed.
+Since as stated this hypothesis is under-specified, further both conceptual and empirical work are needed.
 
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006nonword; but note  @cristia2020infant is an exception].
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006non-word; but note  @cristia2020infant is an exception].
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores.
+In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014non-word; @meir2017independent]  on NWR scores.
 We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
@@
 We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages.
-Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
+Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006non-word; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014non-word and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
@@
 To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
-These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results, with lower overall NWR scores and (integrating with effect of length in  \@ref(fig:fig-prevlit-overall)) length effects.
-This may indicate that the conclusion tentatively drawn in @cristia2020infant about lower NWR scores consistent with long-term effects of lower levels of child-directed speech was premature.
+These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects).
+This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
 Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
@@
 
-That said, @cristia2020infant also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
+That said, @cristia2020infant also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
 This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
@@
 These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
-What is at stake in the second interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
+What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
 
@@
 The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
-Because these latter predictors strongly relate to other language outcomes, the present findings raise the issue of why NWR would pattern differently, what that could tell us about the relationship between lexical development, phonological development, and the input environment and, last but not least, what is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders.
-In the meanwhile, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
+Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including:  Why do NWR scores would pattern differently across samples?
+What does that tell us about the relationship between lexical development, phonological development, and the input environment?
+What is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders?
+While answers to these questions are sought, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
 

### 2021-02-11 manuscript.tex
proofing

@@
 \hypersetup{
-            pdftitle={Non-word repetition in Yélî Dnye},
+            pdftitle={Non-word repetition in children learning Yélî Dnye},
             pdfkeywords={phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness},
@@
 \makeatother
-\shorttitle{NWR in Yélî Dnye}
+\shorttitle{NWR in Yélî Dnye learners}
 \author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
@@
 \vspace{0.5cm}
-\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
+\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
 \authornote{
@@
 Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
-\keywords{phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness\newline\indent Word count: ~9,000 words}
+\keywords{phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness\newline\indent Word count: ~11,500 words}
 \usepackage{lineno}
@@
 
-\title{Non-word repetition in Yélî Dnye}
+\title{Non-word repetition in children learning Yélî Dnye}
 
@@
 \abstract{
-In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible.
+In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible.
 NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
 In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
-Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation in language development at a global and individual level via three contributions.
+Results make three contributions that are specific, and a fourth that is general.
 First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
@@
 Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
-Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.
+Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations, and the extent to which they shed light on the factors universally structuring variation in phonological development at a global and individual level.
 }
@@
 Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000).
-Much of the evidence for later phonological development comes from nonword repetition (NWR) tasks.
+Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
 In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
 After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
-NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
-NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
+NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are ({\textbf{???}}), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory.
-The study was designed to contribute to four broad research areas.
+NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
+In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
+The study was designed to contribute to four broad research areas, three via direct results.
 
 The first research area is at the crossing of typology and phonological development.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes (with the possible exception of Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely linguistic skill.
+Previous work using NWR has preferred relatively universal and early-acquired phonemes (with exceptions including Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely linguistic skills.
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
@@
 Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
-Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, \& Leonard, 2006), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
+Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; {\textbf{???}}), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
 It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
@@
 To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
+We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese ({\textbf{???}}); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
 
@@
 Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error.
-Age has been slightly jittered for ease of inspection of different lengths at a given age.}
+Age has been slightly shifted for ease of inspection of different lengths at a given age.}
 \end{figure}
@@
 The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based ({\textbf{???}}).
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance et al., 2005).
-Previous work typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
-Although past research has not often investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (Havron et al., 2019), and therefore we report on these too.
+Based on previous work, we looked at potential increases with age ({\textbf{???}}; Farmani et al., 2018; Vance et al., 2005).
+Prior research typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; {\textbf{???}}; Balladares, Marshall, \& Griffiths, 2016; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.
 
 Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities Maddieson, 2013b, and @maddieson2013consonant with no exceptions to our knowledge).
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b); Maddieson (2013a); with no exceptions to our knowledge{]}.
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
@@
 (2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
-In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991).
-Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills.
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance (e.g., Gathercole, Willis, \& Baddeley, 1991).
+Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills.
 Indeed, Cristia et al.
 (2020) present the task as being a good index of the development of \enquote{short-hand-like} representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
-To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from Cristia et al.
-(2020)'s hypotheses just mentioned.
-Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
-This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.
-
-Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
+Aside from Cristia et al.
+(2020)'s hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).
+Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
+This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.
+
+Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
 Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
@@
 In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.).
-Thus, with respect to our first research goal, Yélî Dnye is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.
+Thus, with respect to our first research goal, Yélî Dnye is a good language to test because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
 
@@
 
-Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020).
+Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020).
 The vast majority of syllables use a CV format.
@@
 Most Yélî parents are swidden horticulturalists.
-Within a group of households, it is often the case that older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them less time to spend directly with the children in their household than may be the case in infants reared by stay-at-home parents in urban sites.
+Within a group of households, it is often the case that older adolescents and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
 Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 40 final items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; {\textbf{???}}), with the entire test inventory of 40 final items distributed across children.
 
@@
 Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, 2020), it was also possible that we initially generated items with illegal, but currently undocumented constraints.
-Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence).
+Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
 These candidates were then presented to one informant, for a final check that they did not mean anything.
@@
 
-The final list includes three practice items and 40 test items (across infants): 16 monoysllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
+The final list includes three practice items and 40 test items (across infants): 16 monosyllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
 
@@
 
-A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and split it into two sublists, to generate 40 different elicitation sets.
+A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and split it into two sub-lists, to generate 40 different elicitation sets.
 The 40 elicitation sets are available online from \href{https://osf.io/dtxue/}{osf.io/dtxue/}.
@@
 \item
-  Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sublists).
-Since some of these groups had an odd number of items, one of the sublists was slightly longer than the other (20 vs.~23).
+  Splits were done within each length group from the 2018 items (i.e., separately for 2-, 3-, and 4-syllable items); and among onset groups for the difficult monosyllables generated in 2019 (i.e., all the monosyllables starting with /tp/ were split into 2 sub-lists).
+Since some of these groups had an odd number of items, one of the sub-lists was slightly longer than the other (20 vs.~23).
 \item
-  Once the sublist split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sublist selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
+  Once the sub-list split had been done, items were randomized such that all children heard first the 3 practice items in a fixed order (1, 2, and 4 syllables), a randomized version of their sub-list selection of difficult onset items, and randomized versions of their 2-syllable, then 3-syllable, and finally 4-syllable items.
 \end{itemize}
@@
 Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
-We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 children only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
 While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
@@
 We could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory which lead any error to result in many true-word phonetic neighbors.
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
 Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
@@
 The analysis revealed a main effect of age (ß = 0.35, SE ß = 0.13, p \textless{} 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.78, SE ß = 0.19, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:fig-xling-freq}.
-Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(38) = 0.54.
+Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(38) = .544.
 
@@
 Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies {[}r(27)=0.50, p \textless{} 0.01{]}; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies {[}r(38)=0.73, p \textless{} 0.001{]}.
-Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(38)=0.43, p \textless{} 0.01.
+Moreover, averaging across participants, the Pearson correlation between scaled average corpus phone frequency and whole-item NWR scores was r(38)=.432, p \textless{} 0.01.
 Therefore, we fit another mixed logistic regression, this time declaring as fixed effects both scaled cross-linguistic and corpus frequencies (averaged across all attested phones within each stimulus item), in addition to age.
@@
 For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}).
-These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.
-This result suggests that several of our participants have mastered production of some English phones, possibly produced within whole English word forms.}
+These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.}
 The complex consonants in our stimuli occur in \textasciitilde{}17.33\% of languages' phonologies (range 0\% to 78\%);
@@
 
-Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving simple consonats (50.90 versus 8.20).
+Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving simple consonants (50.90 versus 8.20\%).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question.
-As shown in Figure \ref{fig:fig-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (5,649.08) = 0.47, p \textless{} 0.01).
-In contrast, there was no clear association between NWR scores and sex (Welch t (27.33) = -0.60, p = 0.56), birth order (data missing for 14 children, rho = (3,502.90) = -0.20, p = 0.33), or maternal education (rho ( 9,628.60) = 0.10, p = 0.55).
+As shown in Figure \ref{fig:fig-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, ρ(5,649.08) = .47, p \textless{} 0.01).
+In contrast, there was no clear association between NWR scores and sex (Welch t (27.33) = -0.60, p = 0.56), birth order (data missing for 14 children, ρ(3,502.90) = -.198, p = 0.33), or maternal education (ρ( 9,628.60) = .097, p = 0.55).
 
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese Stokes et al., 2006; but note Cristia et al., 2020 is an exception).
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese {\textbf{???}}; but note Cristia et al., 2020 is an exception).
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
-In brief, our results underscore the idea that phoonlogical development continues well past the first few years of life, extending into middle childhood and perhaps later (Hazan \& Barrett, 2000).
-
-In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) on NWR scores.
-We did not expect large effects of maternal education in our sample because of two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
+In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later (Hazan \& Barrett, 2000).
+
+In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; {\textbf{???}}; Meir \& Armon-Lotem, 2017) on NWR scores.
+We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
 In fact, maternal education correlated with NWR score at about r\textasciitilde{}.1, which is small.
@@
 We entered data from 14 studies (including ours), presenting data from 12 languages.
-Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes et al.
-(2006); English by Vance et al.
+Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by ({\textbf{???}}); English by Vance et al.
 (2005); Italian by Piazzalunga et al.
@@
 (2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares et al.
-(2016); Swedish by Kalnak et al.
-(2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al.
-(2020); and Yélî Dnye by the present study.
+(2016); Swedish by ({\textbf{???}}) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al.
+(2020); and Yélî Dnye from the present study.
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
-Results separating different age groups are shown in Figure\textgreater{} \ref{fig:fig-prevlit-overall}.
+Results separating different age groups are shown in Figure \ref{fig:fig-prevlit-overall}.
 
@@
 
-Several observations can be drawn from this Figure.
+Several observations can be drawn from this figure.
 To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
-These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results, with lower overall NWR scores and (integrating with effect of length in \ref{fig:fig-prevlit-overall}) length effects.
-This may indicate that the conclusion tentatively drawn in Cristia et al.
-(2020) about lower NWR scores consistent with long-term effects of lower levels of child-directed speech was premature.
+These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure \ref{fig:fig-prevlit}, larger length effects).
+This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
 Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
@@
 That said, Cristia et al.
-(2020) also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
+(2020) also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
 This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
 If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
-Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in \ref{fig:fig-prevlit-overall}.
+Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \ref{fig:fig-prevlit-overall}.
 These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
-What is at stake in the second interpretation of the lower scores observed among the Tsimane' is related to literacy \emph{in the broader population} (rather than in the tested children themselves).
+What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related to literacy \emph{in the broader population} (rather than in the tested children themselves).
 
@@
 We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
-Neither of these two predictors explained variance in \ref{fig:fig-prevlit-overall}.
+Neither of these two predictors explained variance in Figure \ref{fig:fig-prevlit-overall}.
 It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.~Sotho non-words having 4-7 syllables in length.
 To be more certain whether language characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
-We believe this will be excedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
 An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
@@
 While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that vaies depending on the linguistic, cultural, and sociodemographic properties of the population under study.
-Because these latter predictors strongly relate to other language outcomes, the present findings raise the issue of why NWR would pattern differently, what that could tell us about the relationship between lexical development, phonological development, and the input environment and, last but not least, what is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders.
-In the meanwhile, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
+Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including: Why do NWR scores would pattern differently across samples?
+What does that tell us about the relationship between lexical development, phonological development, and the input environment?
+What is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders?
+While answers to these questions are sought, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
 
@@
 Boersma, P., \& Weenink, D. (2020). Praat: Doing phonetics by computer (Version 6.1.35). Retrieved from \url{http://www.praat.org/}
-
-\leavevmode\hypertarget{ref-bowey2001nonword}{}%
-Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
 
@@
 
-\leavevmode\hypertarget{ref-farabolini2021nonword}{}%
-Farabolini, G., Rinaldi, P., Caselli, C., \& Cristia, A.
-(2021).
-Non-word repetition in bilingual children: The role of language exposure, vocabulary scores and environmental factors.
-\emph{Speech Language and Hearing}.
-
 \leavevmode\hypertarget{ref-farmani2018normalization}{}%
@@
 
-\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
-Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
-(2014).
-Nonword repetition--a clinical marker for specific language impairment in Swedish associated with parents' language-related problems.
-\emph{PloS One}, \emph{9}(2), e89544.
-
 \leavevmode\hypertarget{ref-kapalkova2013non}{}%
@@
 
-\leavevmode\hypertarget{ref-stokes2006nonword}{}%
-Stokes, S.
-F., Wong, A.
-M., Fletcher, P., \& Leonard, L.
-B.
-(2006).
-Nonword repetition and sentence repetition as clinical markers of specific language impairment: The case of cantonese.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{49}(2), 219--236.
-
 \leavevmode\hypertarget{ref-torrington2015non}{}%

### 2021-02-12 manuscript.Rmd
submitted

@@
 
+author_note: >
+       Both authors contributed to study funding, design, data collection, annotation, analyses, writing.
+
 abstract: >
@@
 
-keywords          : "phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness"
+keywords          : "phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy"
 wordcount         : "~11,500 words"
@@
 After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
-NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001non-word], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
+NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences].
@@
 Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
-Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006non-word], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory].
+Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory].
 It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
@@
 To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006non-word]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
+We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
 
@@
 The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021non-word].
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword].
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014non-word; @vance2005speech].
-Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014non-word; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
+Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
+Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
 Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.
@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014non-word], with the entire test inventory of 40 final items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 40 final items distributed across children.
 
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006non-word; but note  @cristia2020infant is an exception].
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006nonword; but note  @cristia2020infant is an exception].
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014non-word; @meir2017independent]  on NWR scores.
+In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores.
 We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
@@
 We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages.
-Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006non-word; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014non-word and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
+Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.

### 2021-08-20 manuscript.Rmd
first stab at R1

@@
 NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
-In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
+In this study, we report on NWR results among children (N= 40, aged 3-10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
 Results make three contributions that are specific, and a fourth that is general.
@@
 
+TODO
+
+- add subheadings (here, results, discussion)
+- remove topics in intro that do not relate to aims of study
+
 Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development].
 Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
+In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
+This allows us to contribute data at the crossing of language typology, language acquisition, and individual variation, as presented in more detail below.
+
+### What is NWR?
+
 In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
@@
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences].
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language].
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].
 NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
-In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
-The study was designed to contribute to four broad research areas, three via direct results.
+
+#### NWR design and analysis adaptations
+
+In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
+For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
+There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
+For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language].
+Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory].
+Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
+
+Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
+We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.
+less challenging sounds.
+However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
+Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
+However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
+
+<!-- ac2ac address R's comment here: NWR tasks and what they measure.
+The authors suggest that their task measures phonological development, I think this point needs to be sharpened and better justified.
+Yes, researchers' opinion on what the task measures vary greatly, and it is briefly over-viewed.
+Yet, the rational for what the novel task measures needs to be either more strongly justified, or down-toned.
+-->
+
+#### NWR and typology
 
@@
 Previous work using NWR has preferred relatively universal and early-acquired phonemes [with exceptions including @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skills.
+
+EXPLAIN MORE CLEARLY what is the intuition here
+
 Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
@@
 
-Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
+<!-- ac2ac address R's comment: (2) Language typology (universal and early-acquired phonemes / cross-linguistically common and cross-linguistically rare phonetic targets).
+Here I would recommend consulting a Chapter by Shula Chiat on language-universal and vs.
+language-specific NWR tasks.
+-->
+<!-- Chiat, S.
+(2015).
+Non-Word Repetition.
+In Armon-Lotem, S., de Jong, J., & Meir, N.
+(Eds.).
+Methods for assessing multilingual children: disentangling bilingualism from Specific Language Impairment (pp.125-150).
+Bristol: Multilingual Matters.
+-->
+
+#### Length effects on NWR
+
+```{r prep-prevlit, echo=F}
+pl<- read.csv("previous_lit.csv")
+
+pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
+colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
+pl_stacked$Length=gsub("X","",pl_stacked$Length)
+
+#remove lines with NA median
+pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked
+
+#add YD data
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age
+
+age_cuts=c(3,6,8,13)
+m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
+m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))
+
+aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
+yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns
+
+colnames(yd_res)<-c("Length","Age","Median","SD","n")
+yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
+colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")
+
+#combine
+pl_stacked=rbind(pl_stacked,yd_res_toadd)
+
+#add SE
+pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
+
+#remove monosyllables
+pl_stacked=pl_stacked[pl_stacked$Length!=1,]
+
+pl_stacked$Language=factor(pl_stacked$Language)
+
+pl$dif<-pl$X2-pl$X3
+difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
+```
+
+The third research area we contribute data to is research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
 Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory].
@@
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
-To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).
-
-```{r prep-prevlit, echo=F}
-pl<- read.csv("previous_lit.csv")
-
-pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
-colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
-pl_stacked$Length=gsub("X","",pl_stacked$Length)
-
-#remove lines with NA median
-pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked
-
-#add YD data
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age
-
-age_cuts=c(3,6,8,13)
-m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
-m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))
-
-aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
-yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns
-
-colnames(yd_res)<-c("Length","Age","Median","SD","n")
-yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
-colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")
-
-#combine
-pl_stacked=rbind(pl_stacked,yd_res_toadd)
-
-#add SE
-pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
-
-#remove monosyllables
-pl_stacked=pl_stacked[pl_stacked$Length!=1,]
-
-pl_stacked$Language=factor(pl_stacked$Language)
-
-pl$dif<-pl$X2-pl$X3
-difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
-```
-
-```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed).
-Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al.
-(2019) reported on groups of 24-60 Italian learners (black); Stokes et al.
-(2006) on 15 Cantonese learners (blue); Vance et al.
-(2005) on 17-20 English learners (light green); Cristia et al.
-(2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
-Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error.
-Age has been slightly shifted for ease of inspection of different lengths at a given age."}
-
-mycols = c( 
-             "Arabic" = "orange",
-             "Italian"   ="black",
-             "Cantonese"   ="blue",
-             "English"   ="green",
-            "Tsimane'" = "darkgreen",
-             "Yélî Dnye" = "purple")
-
-mywd = c( 
-             "Arabic" = 1,
-             "Italian"   =1,
-             "Cantonese"   =1,
-             "English"   =1,
-             "Tsimane'" = 2,
-             "Yélî Dnye" = 4)
-
-#just a trick to jitter age
-jit_age=(pl_stacked$Age+(as.numeric(as.character(pl_stacked$Length))-2)/5)
-
-plot(1,xlab="Average age (years)",ylab="NWR scores (words)",ylim=c(0,100),xlim=range(pl_stacked$Age))
-
-for(thislang in levels(pl_stacked$Language)) for(thislength in levels(factor(pl_stacked$Length[pl_stacked$Language== thislang]))){
-  myx=pl_stacked$Age[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
-  myx=myx+(as.numeric(as.character(thislength))-2)/5 #jitter the length lines a little bit with respect to avg age
-  points(pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]~myx,col=mycols[thislang],pch=as.character(thislength))
-  myy=pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
-  lines(myy~myx, col=mycols[thislang],lty=as.numeric(thislength),lwd=mywd[thislang]) #join points of same length & lang
-  mylb=myy-pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
-  myhb=myy+pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
-  for(i in 1:length(mylb)) {
-    mylty=as.numeric(as.character(thislength))
-    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang],lty=mylty) #error bar
-    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang],lty=mylty) #lower error bar tip
-    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang],lty=mylty) #higher error bar tip
-    }
-}
-
-```
-
-Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
+
+To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths, as described in Supplementary Materials [@cristia2021supplementary; SM1].
+We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant].
+Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
 For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%.
@@
 Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.
+
+#### Individual variation effects on NWR
 
@@
 
-Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
+#### NWR as a function of language and culture
+
+The fourth research goal we pursued is to use NWR with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
 Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant; with no exceptions to our knowledge].
+
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
 That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
+
 In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance [e.g., @gathercole1991differentiating].
@@
 Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+
 Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).
+
 Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
 This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.
+
+### Yélî Dnye phonology and community
 
@@
 
-### Yélî Dnye phonology
+#### Yélî Dnye phonology
 
@@
 
-### The Yélî community
+#### The Yélî community
 
@@
 
-### NWR design and analysis adaptations
-
-In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
-For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
-There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
-For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language].
-Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory].
-Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
-
-Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
-We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.
-less challenging sounds.
-However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
-Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
-However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
-
 ### Research questions
@@
 
-Throughout these analyses and in the Discussion, we will also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.
+Throughout these analyses and in the Discussion, we also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.
 
@@
 ## Methods
+
+### Participants
+
+This study was approved as part of a larger research effort by the second author.
+The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+As discussed in subsection "The Yélî community", the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood [@brownIPchildrearing].
+Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
+Regardless of whether they completed the task, children were given a small snack as compensation.
+Children who showed initial interest but then decided not to participate were also given the snack.
+
+We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions.
+We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease).
+The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).
+In terms of birth order, `r birthorder.tab[[1]]` were born first, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
+These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^2] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` children only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^3]
+
+[^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
+
+<!-- a2m I thought in the first couple of years they combined English and Yélî at school? -->
+<!-- m2a: It used to be!
+In Anna, Ghaalyu, and Taakêmê's time it *was* but policies have shifted a bit since then.
+These days it looks like there's room for things to swing back toward bilingual early education.
+But anyhow for the kids in the current dataset we can pretty reliably say that they're language of schooling was English from the start.-->
+
+[^2]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
 
@@
 
-### Participants
-
-This study was approved as part of a larger research effort by the second author.
-The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
-As discussed in subsection "The Yélî community", the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood [@brownIPchildrearing].
-Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
-Regardless of whether they completed the task, children were given a small snack as compensation.
-Children who showed initial interest but then decided not to participate were also given the snack.
-
-We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions.
-We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease).
-The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).
-In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children.
-These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` children only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
-
-[^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
-Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
-
-<!-- a2m I thought in the first couple of years they combined English and Yélî at school? -->
-<!-- m2a: It used to be!
-In Anna, Ghaalyu, and Taakêmê's time it *was* but policies have shifted a bit since then.
-These days it looks like there's room for things to swing back toward bilingual early education.
-But anyhow for the kids in the current dataset we can pretty reliably say that they're language of schooling was English from the start.-->
-
-[^3]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
-
 ## Results
@@
 
-### NWR as a function of cross-linguistic phone frequency
+### NWR and typology: NWR as a function of cross-linguistic phone frequency
 
@@
 
-### Patterns in NWR mispronunciations
+#### Follow-up analyses: Patterns in NWR mispronunciations
 
@@
 
-### NWR scores as a function of item length
+### Length effects on NWR
 
@@
 
-### Factors structuring individual variation
+### Individual variation effects on NWR
 
@@
 
-### Associations between NWR and cross-linguistic frequency
+### NWR and typology
+
+We looked at two things: Associations between NWR and cross-linguistic frequency, and qualitative mispronunciation patterns. RESTRUCTURE TO MAKE THIS CLEARER!
 
@@
 
-### Item Length
+###  Length effects and NWR
 
@@
 
-### Individual differences
+### Individual variation effects on NWR
 
@@
 
-```{r fig-prevlit-overall, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages.
-The legend indicates language and the length of non-words (in syllables).
-Central tendency is mean; error is one standard error."}
-mycols = c( 
-             "Arabic" = "orange",
-             "Cantonese"   ="blue",
-             "English"   ="green",
-             "Italian"   ="black",
-             "Mandarin"   ="gray",
-             "Persian"   ="pink",
-             "Slovak"   ="lightblue",
-            "Sotho"   ="magenta",
-            "Spanish"   ="brown",
-            "Swedish"   ="red",
-            "Tsimane'" = "darkgreen",
-             "Yélî" = "purple")
-
-mywd = c( 
-             "Arabic" = 1,
-             "Cantonese"   =1,
-             "English"   =1,
-             "Italian"   =1,
-             "Mandarin"   =1,
-             "Persian"   =1,
-             "Slovak"   =1,
-            "Sotho"   =1,
-            "Spanish"   =1,
-            "Swedish"   =1,
-             "Tsimane'" = 2,
-             "Yélî" = 4)
-
-plot(pl_overall$Overall~pl_overall$Age,col=mycols[pl_overall$Language],pch=1,
-     xlab="Average age (years)",ylab="NWR scores (words)",ylim=c(0,100))
-
-for(thislang in levels(factor(pl_overall$Language))) {
-  myx=pl_overall$Age[pl_overall$Language==thislang ]
-  myy=pl_overall$Overall[pl_overall$Language==thislang]
-  lines(myy~myx, col=mycols[thislang],lwd=mywd[thislang]) #join points of same length & lang
-  mylb=myy-pl_overall$se[pl_overall$Language==thislang]
-  myhb=myy+pl_overall$se[pl_overall$Language==thislang]
-  for(i in 1:length(mylb)) {
-
-    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang]) #error bar
-    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang]) #lower error bar tip
-    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang]) #higher error bar tip
-    }
-}
-
-#add legend
-#first half of langs
-xmin=6
-ymax=30
-ymin=0
-myy=seq(from=ymax,to=ymin,length.out=length(all_lang)/2)
-names(myy)<-all_lang[1:(length(all_lang)/2)]
-for(thislang in names(myy)){
-  lines(c(xmin-.1,xmin+.1),c(myy[thislang],myy[thislang]),col=mycols[thislang]) #line sample
-  text(xmin+.15,myy[thislang],paste(thislang,pl_overall$Lengths[pl_overall$Language==thislang][1]),col=mycols[thislang],pos=4)
-}
-
-#second half
-xmin=9.5
-names(myy)<-all_lang[(length(all_lang)/2+1):length(all_lang)]
-for(thislang in names(myy)){
-  lines(c(xmin-.1,xmin+.1),c(myy[thislang],myy[thislang]),col=mycols[thislang]) #line sample
-  text(xmin+.15,myy[thislang],paste(thislang,pl_overall$Lengths[pl_overall$Language==thislang][1]),col=mycols[thislang],pos=4)
-} 
-```
+FIGURE REMOVED
 
@@
 
+## Data, code and materials availability statement
+
+All data, code, and materials are available from [https://osf.io/5qspb/](https://osf.io/5qspb/)
+
 # References

### 2021-08-20 manuscript.tex
first stab at R1

@@
             pdftitle={Non-word repetition in children learning Yélî Dnye},
-            pdfkeywords={phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness},
+            pdfkeywords={phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy},
             pdfborder={0 0 0},
@@
 \textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
-\authornote{
+\authornote{Both authors contributed to study funding, design, data collection, annotation, analyses, writing.
 
 Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
-\keywords{phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness\newline\indent Word count: ~11,500 words}
+\keywords{phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy\newline\indent Word count: ~11,500 words}
 \usepackage{lineno}
@@
 After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
-NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are ({\textbf{???}}), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
+NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
@@
 Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
-Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; {\textbf{???}}), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
+Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, \& Leonard, 2006), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
 It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
@@
 To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese ({\textbf{???}}); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
+We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
 
@@
 The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based ({\textbf{???}}).
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
 In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential increases with age ({\textbf{???}}; Farmani et al., 2018; Vance et al., 2005).
-Prior research typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; {\textbf{???}}; Balladares, Marshall, \& Griffiths, 2016; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance et al., 2005).
+Prior research typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
 Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.
@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; {\textbf{???}}), with the entire test inventory of 40 final items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 40 final items distributed across children.
 
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese {\textbf{???}}; but note Cristia et al., 2020 is an exception).
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese Stokes et al., 2006; but note Cristia et al., 2020 is an exception).
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; {\textbf{???}}; Meir \& Armon-Lotem, 2017) on NWR scores.
+In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) on NWR scores.
 We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
@@
 We entered data from 14 studies (including ours), presenting data from 12 languages.
-Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by ({\textbf{???}}); English by Vance et al.
+Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes et al.
+(2006); English by Vance et al.
 (2005); Italian by Piazzalunga et al.
@@
 (2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares et al.
-(2016); Swedish by ({\textbf{???}}) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al.
+(2016); Swedish by Kalnak et al.
+(2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al.
 (2020); and Yélî Dnye from the present study.
@@
 
+\hypertarget{data-code-and-materials-availability-statement}{%
+\subsection{Data, code and materials availability statement}\label{data-code-and-materials-availability-statement}}
+
+All data, code, and materials are available from \url{https://osf.io/5qspb/}
+
 \hypertarget{references}{%
@@
 Boersma, P., \& Weenink, D. (2020). Praat: Doing phonetics by computer (Version 6.1.35). Retrieved from \url{http://www.praat.org/}
+
+\leavevmode\hypertarget{ref-bowey2001nonword}{}%
+Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
 
@@
 
+\leavevmode\hypertarget{ref-farabolini2021nonword}{}%
+Farabolini, G., Rinaldi, P., Caselli, C., \& Cristia, A.
+(2021).
+Non-word repetition in bilingual children: The role of language exposure, vocabulary scores and environmental factors.
+\emph{Speech Language and Hearing}.
+
 \leavevmode\hypertarget{ref-farmani2018normalization}{}%
@@
 
+\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
+Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
+(2014).
+Nonword repetition--a clinical marker for specific language impairment in Swedish associated with parents' language-related problems.
+\emph{PloS One}, \emph{9}(2), e89544.
+
 \leavevmode\hypertarget{ref-kapalkova2013non}{}%
@@
 
+\leavevmode\hypertarget{ref-stokes2006nonword}{}%
+Stokes, S.
+F., Wong, A.
+M., Fletcher, P., \& Leonard, L.
+B.
+(2006).
+Nonword repetition and sentence repetition as clinical markers of specific language impairment: The case of cantonese.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{49}(2), 219--236.
+
 \leavevmode\hypertarget{ref-torrington2015non}{}%

### 2021-08-20 supmat.Rmd
first stab at R1

---
title: "Supplementary materials"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SM1: Systematic review of previous work to check whether length there are typological differences in length effects on NWR

ADD MORE INFO ABOUT HOW THESE PAPERS WERE DISCOVERED

```{r prep-prevlit, echo=F}
pl<- read.csv("previous_lit.csv")

pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
pl_stacked$Length=gsub("X","",pl_stacked$Length)

#remove lines with NA median
pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked

#add YD data
aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age

age_cuts=c(3,6,8,13)
m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))

aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns

colnames(yd_res)<-c("Length","Age","Median","SD","n")
yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")

#combine
pl_stacked=rbind(pl_stacked,yd_res_toadd)

#add SE
pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)

#remove monosyllables
pl_stacked=pl_stacked[pl_stacked$Length!=1,]

pl_stacked$Language=factor(pl_stacked$Language)

pl$dif<-pl$X2-pl$X3
difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
```

```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al. (2019) reported on groups of 24-60 Italian learners (black); Stokes et al. (2006) on 15 Cantonese learners (blue); Vance et al. (2005) on 17-20 English learners (light green); Cristia et al. (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error. Age has been slightly shifted for ease of inspection of different lengths at a given age."}

mycols = c( 
             "Arabic" = "orange",
             "Italian"   ="black",
             "Cantonese"   ="blue",
             "English"   ="green",
            "Tsimane'" = "darkgreen",
             "Yélî Dnye" = "purple")

mywd = c( 
             "Arabic" = 1,
             "Italian"   =1,
             "Cantonese"   =1,
             "English"   =1,
             "Tsimane'" = 2,
             "Yélî Dnye" = 4)

#just a trick to jitter age
jit_age=(pl_stacked$Age+(as.numeric(as.character(pl_stacked$Length))-2)/5)

plot(1,xlab="Average age (years)",ylab="NWR scores (words)",ylim=c(0,100),xlim=range(pl_stacked$Age))

for(thislang in levels(pl_stacked$Language)) for(thislength in levels(factor(pl_stacked$Length[pl_stacked$Language== thislang]))){
  myx=pl_stacked$Age[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  myx=myx+(as.numeric(as.character(thislength))-2)/5 #jitter the length lines a little bit with respect to avg age
  points(pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]~myx,col=mycols[thislang],pch=as.character(thislength))
  myy=pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  lines(myy~myx, col=mycols[thislang],lty=as.numeric(thislength),lwd=mywd[thislang]) #join points of same length & lang
  mylb=myy-pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  myhb=myy+pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  for(i in 1:length(mylb)) {
    mylty=as.numeric(as.character(thislength))
    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang],lty=mylty) #error bar
    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang],lty=mylty) #lower error bar tip
    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang],lty=mylty) #higher error bar tip
    }
}

```

```{r fig-prevlit-overall, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages. The legend indicates language and the length of non-words (in syllables). Central tendency is mean; error is one standard error."}
mycols = c( 
             "Arabic" = "orange",
             "Cantonese"   ="blue",
             "English"   ="green",
             "Italian"   ="black",
             "Mandarin"   ="gray",
             "Persian"   ="pink",
             "Slovak"   ="lightblue",
            "Sotho"   ="magenta",
            "Spanish"   ="brown",
            "Swedish"   ="red",
            "Tsimane'" = "darkgreen",
             "Yélî" = "purple")

mywd = c( 
             "Arabic" = 1,
             "Cantonese"   =1,
             "English"   =1,
             "Italian"   =1,
             "Mandarin"   =1,
             "Persian"   =1,
             "Slovak"   =1,
            "Sotho"   =1,
            "Spanish"   =1,
            "Swedish"   =1,
             "Tsimane'" = 2,
             "Yélî" = 4)

plot(pl_overall$Overall~pl_overall$Age,col=mycols[pl_overall$Language],pch=1,
     xlab="Average age (years)",ylab="NWR scores (words)",ylim=c(0,100))

for(thislang in levels(factor(pl_overall$Language))) {
  myx=pl_overall$Age[pl_overall$Language==thislang ]
  myy=pl_overall$Overall[pl_overall$Language==thislang]
  lines(myy~myx, col=mycols[thislang],lwd=mywd[thislang]) #join points of same length & lang
  mylb=myy-pl_overall$se[pl_overall$Language==thislang]
  myhb=myy+pl_overall$se[pl_overall$Language==thislang]
  for(i in 1:length(mylb)) {

    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang]) #error bar
    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang]) #lower error bar tip
    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang]) #higher error bar tip
    }
}

#add legend
#first half of langs
xmin=6
ymax=30
ymin=0
myy=seq(from=ymax,to=ymin,length.out=length(all_lang)/2)
names(myy)<-all_lang[1:(length(all_lang)/2)]
for(thislang in names(myy)){
  lines(c(xmin-.1,xmin+.1),c(myy[thislang],myy[thislang]),col=mycols[thislang]) #line sample
  text(xmin+.15,myy[thislang],paste(thislang,pl_overall$Lengths[pl_overall$Language==thislang][1]),col=mycols[thislang],pos=4)
}

#second half
xmin=9.5
names(myy)<-all_lang[(length(all_lang)/2+1):length(all_lang)]
for(thislang in names(myy)){
  lines(c(xmin-.1,xmin+.1),c(myy[thislang],myy[thislang]),col=mycols[thislang]) #line sample
  text(xmin+.15,myy[thislang],paste(thislang,pl_overall$Lengths[pl_overall$Language==thislang][1]),col=mycols[thislang],pos=4)
} 
```

### 2021-08-26 manuscript.Rmd
in progress

@@
 library(openxlsx)
+library(sjPlot)
 
@@
 
-- add subheadings (here, results, discussion)
-- remove topics in intro that do not relate to aims of study
+- remove topics in intro that do not relate to aims of study (R1, unclear to me what this is...)
 
@@
 To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and allowed slopes to vary over the random effects child ID and target ID.
+The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, in interaction with cross-linguistic phone frequency, and allowed intercepts to vary over the random effects child ID and target ID.
 
@@
 trnsc.inc$age.rounded_s=scale(trnsc.inc$age.rounded)
-xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
+#xl_fr=glmer(correct~avg_fr_s+age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
+xl_fr=glmer(correct~avg_fr_s*age.rounded_s+(1|file) + (1|target),data=trnsc.inc,family="binomial")
+#the model with an interaction explains sig more variance than that without
 mycoeff=round(summary(xl_fr)$coefficients,3)
@@
 The analysis revealed a main effect of age (ß = `r mycoeff["age.rounded_s","Estimate"]`, SE ß = `r mycoeff["age.rounded_s","Std.
-Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
+Error"]`, p `r pval.display(mycoeff["age.rounded_s","Pr(>|z|)"])`), with older children repeating more items correctly.
+It also revealed a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = `r mycoeff["avg_fr_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s","Std.
 Error"]`, p `r pval.display(mycoeff["avg_fr_s","Pr(>|z|)"])`): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \@ref(fig:fig-xling-freq).
@@
 knitr::include_graphics("nwr.by.freq.ITEM.png")
-
+dev.off()
+```
+
+Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = `r mycoeff["avg_fr_s:age.rounded_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s:age.rounded_s","Std.
+Error"]`, p `r pval.display(mycoeff["avg_fr_s:age.rounded_s","Pr(>|z|)"])`): The effect of frequency was larger for older children.
+Inspection of Figure \@ref(fig:fig-xling-freq-age3) suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.
+
+```{r fig-xling-freq-age3, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="Model predicting NWR scores from child age (x axis) and the average frequency with which each phone is found across languages (mean, or plus/minus one standard deviation)."}
+
+plot_model(xl_fr,type="pred",terms=c("age.rounded_s","avg_fr_s"))
+```
+
+```{r fig-xling-freq-age1, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (median split on scaled age)."}
+plot_model(xl_fr,type="int")
+```
+
+```{r fig-xling-freq-age2, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (mean, or plus/minus one standard deviation)."}
+
+plot_model(xl_fr,type="pred",terms=c("avg_fr_s","age.rounded_s"))
 ```
@@
 read.table("substitution_bank.txt",header=T)->sub_bank
-colnames(sub_bank)<-c("target","realization")
+colnames(sub_bank)<-c("target","realization","age")
 sub_bank$target=factor(sub_bank$target,levels=allVC)
@@
 read.table("deletion_bank.txt",header=T)->del_bank
-colnames(del_bank)<-c("target")
+colnames(del_bank)<-c("target","age")
 del_bank$target=factor(del_bank$target,levels=allVC)
@@
 read.table("match_bank.txt",header=T)->match_bank
-colnames(match_bank)<-c("target")
+colnames(match_bank)<-c("target","age")
 match_bank$target=factor(match_bank$target,levels=allVC)
@@
 
+```{r}
+
+```
+
 ### Length effects on NWR
@@
 
-```{r prep-pl}
-
-pl_overall=pl[,1:7]
-
-#add YD data
-yd_res_toadd[yd_res_toadd$Length!=1,]->yd_res_toadd
-means=aggregate(yd_res_toadd$Median,by=list(yd_res_toadd$Study,yd_res_toadd$Language,yd_res_toadd$n,yd_res_toadd$Age),mean)
-sds=aggregate(yd_res_toadd$SD,by=list(yd_res_toadd$Study,yd_res_toadd$Language,yd_res_toadd$n,yd_res_toadd$Age),mean)
-yd_res_toadd2=cbind(means,sds$x)
-
-colnames(yd_res_toadd2)<-c("Study","Language","n","Age","Overall","SD")
-yd_res_toadd2$Lengths<-"2-4"
-yd_res_toadd2=yd_res_toadd2[,c("Study","Language","n","Age","Lengths","Overall","SD")]
-
-#combine
-pl_overall=rbind(pl_overall,yd_res_toadd2)
-
-pl_overall$Language<-gsub(" (.*)","",pl_overall$Language)
-
-read.csv("lang-char.tsv",sep="\t")->lang
-merge(pl_overall,lang,by="Language")->pl_overall
-
-#add SE
-pl_overall$se=pl_overall$SD/sqrt(pl_overall$n)
-
-#pairs(pl_overall[c("Overall","nchar","phones","words")])
-
-#cor.test(pl_overall$Overall,pl_overall$nchar)
-#t.test(pl_overall$Overall~pl_overall$words)
-
-all_lang=levels(factor(pl_overall$Language))
-
-```
-
 The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, as when designing this study we wondered whether NWR was a *fair* test of phonological development.

### 2021-08-26 supmat.Rmd
in progress

@@
 
+```{r prep-pl}
+
+pl_overall=pl[,1:7]
+
+#add YD data
+yd_res_toadd[yd_res_toadd$Length!=1,]->yd_res_toadd
+means=aggregate(yd_res_toadd$Median,by=list(yd_res_toadd$Study,yd_res_toadd$Language,yd_res_toadd$n,yd_res_toadd$Age),mean)
+sds=aggregate(yd_res_toadd$SD,by=list(yd_res_toadd$Study,yd_res_toadd$Language,yd_res_toadd$n,yd_res_toadd$Age),mean)
+yd_res_toadd2=cbind(means,sds$x)
+
+colnames(yd_res_toadd2)<-c("Study","Language","n","Age","Overall","SD")
+yd_res_toadd2$Lengths<-"2-4"
+yd_res_toadd2=yd_res_toadd2[,c("Study","Language","n","Age","Lengths","Overall","SD")]
+
+#combine
+pl_overall=rbind(pl_overall,yd_res_toadd2)
+
+pl_overall$Language<-gsub(" (.*)","",pl_overall$Language)
+
+read.csv("lang-char.tsv",sep="\t")->lang
+merge(pl_overall,lang,by="Language")->pl_overall
+
+#add SE
+pl_overall$se=pl_overall$SD/sqrt(pl_overall$n)
+
+#pairs(pl_overall[c("Overall","nchar","phones","words")])
+
+#cor.test(pl_overall$Overall,pl_overall$nchar)
+#t.test(pl_overall$Overall~pl_overall$words)
+
+all_lang=levels(factor(pl_overall$Language))
+
+```
+
 ```{r fig-prevlit-overall, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages.
@@
 ```
+
+## SM2: Age effects on accuracy
+
+```{r fig-xage, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the child's age (averaged in years)."}
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,round(trnsc.inc$age.rounded)),mean)->xage
+colnames(xage)<-c("Target","Age","Word")
+#stims.ipa <- read.xlsx("stimuli.xlsx") %>%
+#  select(target, target_ortho, phono)
+xage <- merge(xage, stims.ipa, by.x = "Target", by.y = "target")
+
+item.plot <- ggplot(data = xage, aes(x = Age, y = Word, label = phono)) +
+  geom_smooth(method = "lm", se = FALSE, color = "black") +
+  geom_text(family = "Doulos SIL") +
+  ylab("NWR scores (% words)") +
+  xlab("Children's age") +
+  theme_apa()
+ggsave("nwr.by.age.ITEM.png", dpi = 300, device = png())
+
+knitr::include_graphics("nwr.by.age.ITEM.png")
+
+```

### 2021-08-26 wrangling.R

@@ -293,7 +293,7 @@ trnsc$ins=trnsc$del=trnsc$sub=trnsc$nchar=trnsc$trafos=trnsc$tarlen=NA
 substitution_bank=deletion_bank=match_bank=substitution_bank_try1=deletion_bank_try1=match_bank_try1=NULL
 
 for(i in grep("0",trnsc$correct)){ #this goes through all items with errors (" grep("0",trnsc$correct)")
-  #lev
+  #lev #i=2 ; i = 4
   lev=adist(trnsc[i,c("mp_uni")],trnsc[i,c("target_uni")],count=T)
   trnsc$ins[i]<-attributes(lev)$counts[,,"ins"]
   trnsc$del[i]<-attributes(lev)$counts[,,"del"]
@@ -304,29 +304,35 @@ for(i in grep("0",trnsc$correct)){ #this goes through all items with errors (" g
   
   #extract substitutions & add them to the bank
   sublist=which(strsplit(attributes(lev)$trafos, "")[[1]]=="S")
-  substitution_bank=rbind(substitution_bank,cbind(strsplit(trnsc[i,c("target_uni")],"")[[1]][sublist],strsplit(trnsc[i,c("mp_uni")],"")[[1]][sublist]))
+  if(length(sublist) != 0) substitution_bank=rbind(substitution_bank,
+                          cbind(strsplit(trnsc[i,c("target_uni")],"")[[1]][sublist],
+                                strsplit(trnsc[i,c("mp_uni")],"")[[1]][sublist],
+                                trnsc[i,"age.rounded"])) # added age here
   
   #same for deletions  
   dellist=which(strsplit(attributes(lev)$trafos, "")[[1]]=="D")
-  deletion_bank=rbind(deletion_bank,cbind(strsplit(trnsc[i,c("target_uni")],"")[[1]][dellist]))
+  if(length(dellist) != 0) deletion_bank=rbind(deletion_bank,cbind(strsplit(trnsc[i,c("target_uni")],"")[[1]][dellist],
+                                                                   trnsc[i,"age.rounded"]))  # added age here
   
   #end with matches
   matlist=which(strsplit(attributes(lev)$trafos, "")[[1]]=="M")
-  match_bank=rbind(match_bank,cbind(strsplit(trnsc[i,c("target_uni")],"")[[1]][matlist]))
+  if(length(matlist) != 0) match_bank=rbind(match_bank,cbind(strsplit(trnsc[i,c("target_uni")],"")[[1]][matlist],
[…]

### 2021-09-01 manuscript.Rmd
getting started with R2

@@
 
-TODO
-
-- remove topics in intro that do not relate to aims of study (R1, unclear to me what this is...)
-
 Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development].
@@
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
-This allows us to contribute data at the crossing of language typology, language acquisition, and individual variation, as presented in more detail below.
+This allows us to contribute data at the intersection of language typology, language acquisition, and individual variation, as presented in more detail below.
 
 ### What is NWR?
-
-In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
-After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
-NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
-NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences].
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].
-NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
-
-#### NWR design and analysis adaptations
 
@@
 For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
-There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
-For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language].
-Additionally, while some studies have used 10-15 non-words [e.g., @cristia2020infant], others have employed up to 46 unique items [@piazzalunga2019articulatory].
-Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].
-
-Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
-We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.
-less challenging sounds.
-However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
-Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
-However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
-
-<!-- ac2ac address R's comment here: NWR tasks and what they measure.
-The authors suggest that their task measures phonological development, I think this point needs to be sharpened and better justified.
-Yes, researchers' opinion on what the task measures vary greatly, and it is briefly over-viewed.
-Yet, the rational for what the novel task measures needs to be either more strongly justified, or down-toned.
--->
+
+NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
+NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@chiat2015non;@estes2007differences].
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].
+Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory [see a discussion in @chiat2015nonword].
+Broadly, however, NWR scores will necessarily reflect to a certain extent long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
+
+### The present work
+
+We aimed to contribute to four areas of research. We motivate each in turn.
 
@@
 
-The first research area is at the crossing of typology and phonological development.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes [with exceptions including @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skills.
-
-EXPLAIN MORE CLEARLY what is the intuition here
-
-Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
-Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.
-
-<!-- ac2ac address R's comment: (2) Language typology (universal and early-acquired phonemes / cross-linguistically common and cross-linguistically rare phonetic targets).
-Here I would recommend consulting a Chapter by Shula Chiat on language-universal and vs.
-language-specific NWR tasks.
--->
-<!-- Chiat, S.
-(2015).
-Non-Word Repetition.
-In Armon-Lotem, S., de Jong, J., & Meir, N.
-(Eds.).
-Methods for assessing multilingual children: disentangling bilingualism from Specific Language Impairment (pp.125-150).
-Bristol: Multilingual Matters.
--->
+The first research area is at the intersection of typology and phonological development.
+There has been an interest in adapting NWR to different languages, in part for applied purposes.
+In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, @chiat2015non discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
+As a result, any one item created will be relatively easier if it more closely resembles real words in the language, making it difficult to balance difficulty when comparing children learning different languages.
+This previous literature also suggests some dimensions of difficulty -- an issue to which we return in the next subsection.
+
+Although this cross-linguistic literature is rich, the potential difficulty associated with specific phonetic targets composing the non-words has received relatively little attention.
+For example, @chiat2015non discusses segmental complexity as a function of whether there are consonant clusters -- which is arguably a factor reflecting phonotactics and syllable structure.
+
+In the present study, we thought it was relevant to represent the rich phonological inventory found in Yélî Dnye, by including a variety of phonetic targets.
+Some of them are cross-linguistically rare, in that they are less common across languages than other sounds or phonetic targets.
+Phonologists and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of acquisition, focusing largely on phonotactics [@moreton2012structure].
+The correlation between typological frequency and ease of acquisition is typically assumed to emerge from the following causal paths (which are not mutually exclusive):
+
+1. Sounds (and sound sequences) that are harder to perceive tend to be misperceived and thus lost diachronically
+2. Sounds (and sound sequences) that are harder to prononunce tend to be mispronounced and thus lost diachronically
+3. Sound sequences that are harder to hold in memory tend to be mispronounced and thus lost diachronically
+
+Given these causal pathways, we predicted that variation in NWR across items will correlate with the cross-linguistic frequency of the phones composing those items.
 
@@
 
-```{r prep-prevlit, echo=F}
-pl<- read.csv("previous_lit.csv")
-
-pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
-colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
-pl_stacked$Length=gsub("X","",pl_stacked$Length)
-
-#remove lines with NA median
-pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked
-
-#add YD data
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age
-
-age_cuts=c(3,6,8,13)
-m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
-m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))
-
-aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
-yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns
-
-colnames(yd_res)<-c("Length","Age","Median","SD","n")
-yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
-colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")
-
-#combine
-pl_stacked=rbind(pl_stacked,yd_res_toadd)
-
-#add SE
-pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
-
-#remove monosyllables
-pl_stacked=pl_stacked[pl_stacked$Length!=1,]
-
-pl_stacked$Language=factor(pl_stacked$Language)
-
-pl$dif<-pl$X2-pl$X3
-difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
-```
-
-The third research area we contribute data to is research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
+The second research area we contribute data to is research looking at the impact of word length on NWR repetition within specific languages.
 Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory].
-It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
+
+It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture -- a hypothesis discussed for instance in @chiat2015non (pp.
+7-8; see also p.
+5).
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
-
-To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths, as described in Supplementary Materials [@cristia2021supplementary; SM1].
-We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant].
-Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
-For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%.
-Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).
-Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.
+A non-systematic meta-analyses does not provide overwhelming support for this hypothesis [@cristia2021supplementary; SM1].
+
+Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.
 
@@
 In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance [e.g., @gathercole1991differentiating].
-Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills.
-Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+Therefore, it may not be a general index of, for instance, phonological development, but reflect only certain non-universal language skills.
+Indeed, @cristia2020infant present their task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+We return to the question of what was measured here in the Discussion.
 
@@
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 40 final items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-46 items: @jabere2018xperiment; @kalnak2014nonword;@piazzalunga2019articulatory], with the entire test inventory of 40 final items distributed across children.
+We  used a relatively large number of items to explore effects of length and phonological complexity.
+However, aware that this large item inventory might render the task longer and more tiresome, we split items across children.
+Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
 
@@
 
-In adapting the typical NWR procedure for this context, we balanced three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them (i.e., from other children who had participated); that children would feel comfortable doing this task with us; and that community members would feel comfortable having their children do this task with us.
+There is some variation in procedure in previous work.
+For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language].
+
+In adapting the typical NWR procedure for our context, we balanced three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them (i.e., from other children who had participated); that children would feel comfortable doing this task with us; and that community members would feel comfortable having their children do this task with us.
 
@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
+We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible relationship between typological phone frequency and NWR scores.
 An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the Discussion below).
@@
 ### NWR across languages and cultures
+
+WHAT IS BEING MEASURED?
+We then added a mention in the discussion that our work does not allow us to tease apart lexical knowledge from phonological knowledge and from working memory in ways that have been done in previous research.
 

### 2021-09-01 supmat.Rmd
getting started with R2

@@
 
+```{r prep-prevlit, echo=F}
+pl<- read.csv("previous_lit.csv")
+
+pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
+colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
+pl_stacked$Length=gsub("X","",pl_stacked$Length)
+
+#remove lines with NA median
+pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked
+
+#add YD data
+aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age
+
+age_cuts=c(3,6,8,13)
+m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
+m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))
+
+aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
+yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns
+
+colnames(yd_res)<-c("Length","Age","Median","SD","n")
+yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
+colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")
+
+#combine
+pl_stacked=rbind(pl_stacked,yd_res_toadd)
+
+#add SE
+pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
+
+#remove monosyllables
+pl_stacked=pl_stacked[pl_stacked$Length!=1,]
+
+pl_stacked$Language=factor(pl_stacked$Language)
+
+pl$dif<-pl$X2-pl$X3
+difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
+```
+
 ADD MORE INFO ABOUT HOW THESE PAPERS WERE DISCOVERED
+
+ We do not think previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths, as described in Supplementary Materials.
+We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant].
+Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
+For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%.
+Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).
 

### 2021-09-01 getting_frequency_corpus.Rmd
one third through r2

@@
 
-Also, this reveals my rewrites were incomplete, so I added some lines for the onsets that were missing in my initial correspondance list.
+Also, this reveals my rewrites were incomplete, so I added some lines for the onsets that were missing in my initial correspondence list.
 
-The correspondances here looks similar to the one in wrangling but it has more entries because we want to separate each phoneme, and we want to distinguish between short and long
+The correspondences here looks similar to the one in wrangling but it has more entries because we want to separate each phoneme, and we want to distinguish between short and long
 

### 2021-09-01 manuscript.Rmd
one third through r2

@@
 Second, we documented rather weak effects of item length.
-Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
+Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
 Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.
@@
 library(sjPlot)
+library(stringr)
 
@@
 NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@chiat2015non;@estes2007differences].
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].
 Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory [see a discussion in @chiat2015nonword].
-Broadly, however, NWR scores will necessarily reflect to a certain extent long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
+Broadly, however, NWR scores will necessarily reflect to a certain extent phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
 
@@
 
-It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture -- a hypothesis discussed for instance in @chiat2015non (pp.
+It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture -- a hypothesis discussed for instance in @chiat2015non (pp.
 7-8; see also p.
 5).
-In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
+In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer.
 A non-systematic meta-analyses does not provide overwhelming support for this hypothesis [@cristia2021supplementary; SM1].
 
-Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.
+Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we did not approach this issue within a hypothesis-testing framework but sought instead to provide additional data  on the question, which may be re-used in future meta- or mega-analytic analyses.
 
@@
 
-The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
+The third research area we contribute data to relates to the possibility that children differ from each other in NWR scores in systematic ways.
 Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword].
@@
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
-Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool].
+Our focus on age stems from previous work [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
+Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child sex [@chiat2007preschool].
 Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.
@@
 The fourth research goal we pursued is to use NWR with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size].
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant; with no exceptions to our knowledge].
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant].
+To our knowledge, it has never been used with speakers of languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant].
 
@@
 
-In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance [e.g., @gathercole1991differentiating].
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading; for instance, better than rhyme awareness  [e.g., @gathercole1991differentiating].
 Therefore, it may not be a general index of, for instance, phonological development, but reflect only certain non-universal language skills.
@@
 Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
-This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.
+This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, which involved English participants.
 
@@
 Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites.
+Rossel Island lies 250 nautical miles off the coast of mainland PNG and is surrounded by a barrier reef.
+As a result, transport to and from the island is both infrequent and irregular.
 International phone calls and digital exchanges that require significant data transfer are typically not an option.
@@
 
+```{r echo=F}
+scan("words_corpus.txt",what="char")->wds
+
+un=unique(wds)
+un1=gsub("[:aeiouâêîáóéú]*", " ",un)
+count_syll=str_count(un1," ")-1
+count_syll[count_syll==0]<-1
+round(table(count_syll)/length(count_syll)*100)
+
+```
+
 Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
+We also approximated word length based on the number of vowel sequences in words found in a conversational corpus (see Stimuli section for details), which suggested a more balanced distribution: 15% monosyllabic, 39% disyllabic, 29% trisyllabic,  and the remaining 17% being longer than that.
 The vast majority of syllables use a CV format.
@@
 
+Before closing this section, it bears mentioning that the language has an established orthography, which includes distinct graphemes for all the contrasts on which our items are based.
+Although some children in our sample will have started school, at present reading and writing instruction is done only in English (other than writing one's name).
+This was probably not the case for the majority of mothers of the children in our sample, who will have learned to read and write in Yélî Dnye during their first three years at school.
+This change from Yélî-initial instruction to English only occurred about 15 years ago.
+It is possible that there is also some home teaching of Yélî reading and writing, notably for reading the bible.
+
 #### The Yélî community
@@
 Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation.
-Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
-There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.
+Specifically, we investigated potential effects of age, child sex, maternal education, and birth order.
+There is nothing particular to note regarding age and child sex, but we have some comments that pertain to the other two factors.
 
@@
 This study was approved as part of a larger research effort by the second author.
-The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041), including the use of verbal (not written) consent.
 As discussed in subsection "The Yélî community", the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
@@
 
+#### Cross-linguistic frequency
+
 To inform our analyses, we estimated the typological frequency of all phonological segments present in the target items using the PHOIBLE cross-linguistic phonological inventory database [@phoible].
@@
 
-Additionally, we estimated frequency of the phones present  in the target items in a corpus of child-centered recordings [@casillas2020early] by counting the number of word types in which they occurred, and applied the natural logarithm.[^1] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+#### Within-language frequency
+
+Additionally, we estimated frequency of the phones present  in the target items in a corpus of child-centered recordings [@casillas2020early].
+That corpus was constituted by sampling from audio-recordings (7-9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
+The researchers selected 9 2.5-minute clips randomly and 11 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity).
+These clips were  segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the people talking personally.
+For more details, please refer to @casillas2020early.
+
+For the present study, we extracted the transcriptions of adult speech (i.e., removing key child and other children's speech) and split them into words using white space.
+We then removed all English and Pidgin words.
+The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types.
+To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.[^1] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
 
@@
 To minimize adjustment time, which was uncomfortable for some children given the proximity of the foreign experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth.
-A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement.
+A research assistant who spoke Yélî Dnye natively, and who could also hear the instructions over headphones, sat next to the child throughout the task to provide instructions and, if needed, encouragement.
 The research assistant coached the child throughout the task to make sure that they understood what they were expected to do.
-An experimenter (the first author) delivered the pre-recorded stimuli to the research assistant and the child over headphones.
+Finally, an experimenter (the first author) was also fitted with headphones and a microphone; she was in charge of delivering the pre-recorded stimuli to the research assistant, the child, and herself over headphones.
 
@@
 The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
+After completing all of her annotations, she double-checked them by listening to them twice.
 
@@
 
+### NWR and typology: NWR as a function of within-language phone frequency
+
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
+
 Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies [r(`r freq_cor_phone$parameter`)=`r round(freq_cor_phone$estimate,3)`, p `r pval.display(round(freq_cor_phone$p.value,3))`]; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies [r(`r freq_cor_item$parameter`)=`r round(freq_cor_item$estimate,3)`, p `r pval.display(round(freq_cor_item$p.value,3))`].
@@
 In fact, maternal education correlated with NWR score at about r~.1, which is small.
-Similarly, NWR scores may not vary greatly with participant gender according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.
+Similarly, NWR scores may not vary greatly with participant sex according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.
 
@@
 ### NWR across languages and cultures
-
-WHAT IS BEING MEASURED?
-We then added a mention in the discussion that our work does not allow us to tease apart lexical knowledge from phonological knowledge and from working memory in ways that have been done in previous research.
 
@@
 Sotho non-words having 4-7 syllables in length.
-To be more certain whether language characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
+To be more certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
 We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
@@
 
+### Limitations
+
+Before closing, we would like to point out some salient limitations of the current work.
+To begin with, we only employed one set of non-words, in which not all characteristics that previous work suggest matter were manipulated [@chiat2015non].
+As a result, we only have a rather whole-sale measure of performance, and we do not know to what extent lexical knowledge, pure phonological knowledge, and working memory, among others, contribute to children's performance.
+Similarly, our items varied systematically in length and typological frequency of the sounds included, but not in other potential dimensions (such as whether the items contained morphemes of the language or not).
+Additionally, we only had a single person interacting with children as well as interpreting children's production, so we do not know to what extent our findings generalize to other experimenters and research assistants.
+In addition, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
+Although we know from other work that adults' perceptual performance on these types of sounds is well above chance from audio-only presentation [REF],  language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.
+
 ### Conclusions
@@
 While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
 Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including:  Why do NWR scores would pattern differently across samples?

### 2021-09-02 manuscript.Rmd
still going through r2

@@
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Our focus on age stems from previous work [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
-Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child sex [@chiat2007preschool].
+
+Our focus on age stems from previous work, where performance increases with child age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
 Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.
+Common explanations for advantages for first- over later-born children include differential allocation of familial resources, particularly parental behaviors of cognitive stimulation [@lehmann2018early].
+Regarding child sex, no such effect has been found in previous NWR research [@chiat2007preschool], and in other language tasks evidence is mixed.
+Finally, prior research typically finds no significant differences as a function of  maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio].
+In general, maternal education correlates with child language outcomes, including vocabulary reports [@frank2017wordbank] and word comprehension studies [@scaff2019].
+The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children [see discussion in @cristia2020infant].
 
@@
 In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
-Thus, with respect to our first research goal, Yélî Dnye is a good language to test because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
+Thus, with respect to our first research goal, Yélî Dnye is a good language to use because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
 
@@
 With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop system includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).^[We use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
-In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).] The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+The core oral stop system includes both singleton (/p/, /t/, /ṭ/^[We use Levinson's [-@levinsonYDgrammar] under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).], and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with a complete range of nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).
+The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
 Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to occur as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
 
@@
 
-Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent dictionary at the time of writing; @levinsonYDgrammar].
-We also approximated word length based on the number of vowel sequences in words found in a conversational corpus (see Stimuli section for details), which suggested a more balanced distribution: 15% monosyllabic, 39% disyllabic, 29% trisyllabic,  and the remaining 17% being longer than that.
+Our second research goal is to measure the effect of non-word length on NWR, which may need to be interpreted taking into account typical word length in the language. 
+<!-- According to the most recent dictionary at the time of writing [>5800 lexemes; @levinsonYDgrammar], most Yélî Dnye words are disyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (~10%).
+-->
+We estimated word length in words found in a conversational corpus (see Stimuli section for details), where the distribution of length was: 15% monosyllabic, 39% disyllabic, 29% trisyllabic,  and the remaining 17% being longer than that.
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., "ndap" /ṇṭæp/ *Spondylus shell*) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., "ndapî" /'ṇṭæ.pɯ/).
-There are also a handful of words starting with /æ/ (e.g., "ala" /æ.'læ/ *here*) and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
+A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., *ndap* /ṇṭæp/ 'Spondylus shell') and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., *ndapî* /'ṇṭæpɯ/).
+There are also a handful of words starting with /æ/ (e.g., *ala* /æ'læ/ 'here') and a small collection of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
 
@@
 Before closing this section, it bears mentioning that the language has an established orthography, which includes distinct graphemes for all the contrasts on which our items are based.
-Although some children in our sample will have started school, at present reading and writing instruction is done only in English (other than writing one's name).
+Some children in our sample will have started school.
+Reading and writing instruction is currently done only in English (other than writing one's name).
 This was probably not the case for the majority of mothers of the children in our sample, who will have learned to read and write in Yélî Dnye during their first three years at school.
@@
 
-The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+The typical household in our dataset includes seven individuals (typically, a mixed-sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
 These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
@@
 
-Most Yélî parents are swidden horticulturalists.
-Within a group of households, it is often the case that older adolescents and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
+Most Yélî parents are swidden horticulturalists, who occasionally fish.
+Within a group of households, it is often the case that older adolescents and adults spend their day tending to their farm plots (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
 Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
-Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early].
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available [@casillas2020early].
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
@@
 To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
-Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
-Overall, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
+<!-- Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
+ac2mc removed, R was confused and I'm also not sure what is the relevance of this comment -->
+In addition, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply be due to random factors like living close to a school or having relatives there.
 
@@
 It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
-This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
-
-We add some observations that will help us integrate this study to the broader investigation of NWR across cultures.
-As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant].
-However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
-The Yélî community can help us shed further light onto this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early].
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early].
+This contrasts with the picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
+
+<!-- ac2mc shouldn't we totally rewrite this? -->
+We add some observations that will help us integrate this study into the broader investigation of NWR across cultures.
+As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant].
+However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is also relatively infrequent in this community [@casillas2020early].
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early].
 Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in @cristia2020infant.
@@
 
-[^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+[^2]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
+
+[^3]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
 While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
@@
 But anyhow for the kids in the current dataset we can pretty reliably say that they're language of schooling was English from the start.-->
-
-[^2]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
 
@@
 A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible 2- to 4-syllable sequences.
-These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
+These candidates were automatically removed from consideration if they appeared in the most recent dictionary [@levinsonYDgrammar].
 The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
@@
 Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
-Because there is still much to discover about the phonology and phonetics of Yélî Dnye [@levinsonYDgrammar], it was also possible that we initially generated items with illegal, but currently undocumented constraints.
-Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
+Additionally, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
 These candidates were then presented to one informant, for a final check that they did not mean anything.
@@
 The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present.
-Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended "yî" /yɯ/; identified as "yi" /yi/, *tree*).
-Additionally, an error was made when preparing files for annotation, resulting in two items being merged ("tpâ" /tpɑ/ and "tp:a" /tpæ̃/).
+Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended *yî* /yɯ/; identified as *yi* /yi/, 'tree').
+Additionally, an error was made when preparing files for annotation, resulting in two items being merged (*tpâ* /tpɑ/ and *tp:a* /tpæ̃/).
 These three problematic items are not described here, and removed from the analyses below.
@@
 
-A Praat script [@Praat] was written to randomize this list 20 times, and split it into two sub-lists, to generate 40 different elicitation sets.
+A Praat script [@Praat] was written to randomize this list 20 times, and to split it into two sub-lists, to generate 40 different elicitation sets.
 The 40 elicitation sets are available online from [osf.io/dtxue/](https://osf.io/dtxue/).
@@
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
-Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
+While PHOIBLE is a unprecedented in its scope, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
+Note that nearly half of the phones in PHOIBLE are only attested in one language (Steven Moran, personal communication).
 Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0% (i.e., its cross-linguistic percentile is zero).
@@
 The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types.
-To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.[^1] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.[^1] Here, sounds from our non-word items that did not occur in the corpus were not considered (i.e., they were declared NA so that they do not count for analyses).
 
@@
 We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
+The percentage we observe here is much higher than reported in the study by Castro and colleagues, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
 Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
@@
 
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: The same perception and production pressures that shape languages diachronically could affect a language's lexicon, so that sounds that are easier to perceive or produce are more frequent within a language than those that are harder.
+If so, children will have more experience with the easier sounds, and they may thus be better able to to represent and repeat non-words containing them simply because of the additional exposure.
 
@@
 
-#### Follow-up analyses: Patterns in NWR mispronunciations
+### Follow-up analyses: Patterns in NWR mispronunciations
 
@@
 oralVs=c("a", "â","e",  "é" ,"ê", "i","î" ,"o","ó","u")
-simplexCs=c("m","ń","l","w","y","v","d","g","p","t","k","f","h","ch")
-complexCs=c("dp","tp","kp","km","kn","mb","gh","lv")
+simplexCs=c("m","ń","l","w","y","v","d","g","p","t","k","f","h","ch","gh")
+complexCs=c("dp","tp","kp","km","kn","mb","lv")
 simplexCs.ipa=c("m","n","l","w","j","w","ṭ","g","p","t","k","f","h","tʃ")
@@
 #read.table("substitution_bank_try1.txt",header=T)->sub_bank
-read.table("substitution_bank.txt",header=T)->sub_bank
-colnames(sub_bank)<-c("target","realization","age")
+read.table("substitution_bank.txt",header=T)->sub_bank_full
+colnames(sub_bank_full)<-c("target","realization","age")
+sub_bank=sub_bank_full[,1:2]
 sub_bank$target=factor(sub_bank$target,levels=allVC)
@@
 #read.table("deletion_bank_try1.txt",header=T)->del_bank
-read.table("deletion_bank.txt",header=T)->del_bank
-colnames(del_bank)<-c("target","age")
+read.table("deletion_bank.txt",header=T)->del_bank_full
+colnames(del_bank_full)<-c("target","age")
+del_bank=del_bank_full[,1]
 del_bank$target=factor(del_bank$target,levels=allVC)
@@
 #read.table("match_bank_try1.txt",header=T)->match_bank
-read.table("match_bank.txt",header=T)->match_bank
-colnames(match_bank)<-c("target","age")
+read.table("match_bank.txt",header=T)->match_bank_full
+colnames(match_bank_full)<-c("target","age")
+match_bank=match_bank_full[,1]
 match_bank$target=factor(match_bank$target,levels=allVC)
@@
 There were no cases of insertion, and deletions were very rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] )*100,2)` of unmatched phones).
+We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the  majority of incorrect phones (~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] )*100,2)` of unmatched phones).
 To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
@@
 whereas oral vowels in our stimuli occur in ~`r mean(segments$pc[segments$type %in% c("short_oral","long_oral")],na.rm=T)`% of languages' phonologies (range `r min(segments$pc[segments$type %in% c("short_oral","long_oral")],na.rm=T)`% to `r max(segments$pc[segments$type %in% c("short_oral","long_oral")],na.rm=T)`%).
-As noted above, type frequency within the language is correlated with cross-linguistic frequency, and thus these two types of sounds also differ in the former: Their type frequencies in Yélî Dnye are: nasal vowels ~`r mean(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰  (range `r min(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰) versus oral ~`r mean(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰ (range `r min(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰).
+As noted above, frequency within the language is correlated with cross-linguistic frequency, and thus these two types of sounds also differ in the former: Their frequencies in Yélî Dnye are: nasal vowels ~`r mean(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰  (range `r min(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("short_nasal","long_nasal")],na.rm=T)*10`‰) versus oral ~`r mean(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰ (range `r min(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰ to `r max(segments$freq_corpus_types[segments$type %in% c("short_oral","long_oral")],na.rm=T)*10`‰).
 
@@
 
-Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonants (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`%).
+Table \@ref(tab:tab-c) showed that errors involving complex consonant targets were more common than those involving simple consonants (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`%).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.

### 2021-09-02 wrangling.R

@@ -4,7 +4,7 @@ library(openxlsx)
 read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
 
 
-#DATA REMOVAL *** ATTENTION *** 
+#DATA REMOVAL *** ATTENTION ***
 #REMOVE ALL TASK 4 (whatever that is)
 # M2A: This appears to be a second pass over productions that were deemed to
 # be words; the elicitation here is just a translation of what word the
@@ -21,7 +21,7 @@ trnsc$tokid=paste(trnsc$item,trnsc$token)
 # 361              two     two
 # 362              two     two
 # --> these two were converted to YD orthography
-# 
+#
 # The following have not been changed in any way
 # 732             loudî      English: loud
 # 879          no-pinêt                                          English: no peanut
@@ -74,7 +74,7 @@ trnsc <- trnsc %>%
 
 
 dim(trnsc)
-  
+
 # add demographic info
 read.csv("NWR-demo.csv",header=T)->demo
 #colnames(demo)
@@ -92,7 +92,7 @@ merge(trnsc,inc,by.x="id",by.y="ID",all.x=T)->trnsc
 dim(trnsc) #still no loss, all good
 
 #add orthographic representations targets & other stims chars
-#   
+#
 # also, notice that stims and segments are merged using ortho, so as long as ortho is matched
 # across the two, the merge should carry over whatever changes you make
 
@@ -107,7 +107,7 @@ read.delim("segment-counts-types.txt", encoding="UTF-8",sep="\t",header=T)->phon
 merge(segments,phone_counts_types,by.x="ortho",by.y="ortho",all.x=T)->segments
 write.table(segments,"segments_with_cor_freq.txt",sep="\t",row.names = F)
 
-#read in table with breakdown 
+#read in table with breakdown
 read.xlsx("stimuli.xlsx")->stims
 #add average crosslinguistic frequency
 stim_seg_freq=matrix(NA,nrow=dim(stims)[1],ncol=8)
@@ -120,7 +120,7 @@ stims$avg_fr=apply(stim_seg_freq,1,mean,na.rm=T)
 #add average corpus frequency - LOGGED
 stim_seg_freq=matrix(NA,nrow=dim(stims)[1],ncol=8)
 for(i in 1:dim(stim_seg_freq)[1]) for(j in 1:dim(stim_seg_freq)[2]) {
[…]

### 2021-09-02 manuscript.Rmd
bug found but not fixed: looking into widone 12

@@
 read.table("substitution_bank.txt",header=T)->sub_bank_full
-colnames(sub_bank_full)<-c("target","realization","age")
+colnames(sub_bank_full)<-c("target","realization","age","target_ortho","phono","tokid")
 sub_bank=sub_bank_full[,1:2]
@@
 read.table("deletion_bank.txt",header=T)->del_bank_full
-colnames(del_bank_full)<-c("target","age")
+colnames(del_bank_full)<-c("target","age","target_ortho","phono","tokid")
 del_bank=del_bank_full[,1]
@@
 read.table("match_bank.txt",header=T)->match_bank_full
-colnames(match_bank_full)<-c("target","age")
+colnames(match_bank_full)<-c("target","age","target_ortho","phono","tokid")
 match_bank=match_bank_full[,1]

### 2021-09-02 wrangling.R

@@ -48,6 +48,8 @@ trnsc$mispronunciation=as.character(trnsc$mispronunciation)
 trnsc$meaning=as.character(trnsc$meaning)
 trnsc$mispronunciation[trnsc$mispronunciation=="two"]<-"tuu"
 trnsc[grep("Eng",trnsc$meaning),c("mispronunciation","meaning")]
+trnsc$english<-F
+trnsc$english[grep("Eng",trnsc$meaning)]<-T
 
 read.table("final_order.txt",header=T)->crp
 crp$tokid=paste(crp$target,crp$nb)
@@ -173,7 +175,7 @@ trnsc[trnsc$correct==0 & trnsc$target_uni == trnsc$mp_uni & !is.na(trnsc$mp_uni)
 trnsc$correct[trnsc$correct==0 & trnsc$target_uni == trnsc$mp_uni & !is.na(trnsc$mp_uni)]<-1
 # M2A: I agree! My question is: Why then are they marked as incorrect in the first place?
 # Does this deserve more detective work?
-#a2m I looked into in there were disfluencies
+#a2m I looked into it and there were disfluencies
 # M2A: awesome, thank you! happy with this now :)
 
 # create phonological correspondance matrix
@@ -292,9 +294,10 @@ head(trnsc[,c("mispronunciation","mp_uni","target_uni")])
 trnsc$ins=trnsc$del=trnsc$sub=trnsc$nchar=trnsc$trafos=trnsc$tarlen=NA
 substitution_bank=deletion_bank=match_bank=substitution_bank_try1=deletion_bank_try1=match_bank_try1=NULL
 
-for(i in grep("0",trnsc$correct)){ #this goes through all items with errors (" grep("0",trnsc$correct)")
-  #lev #i=2 ; i = 4
-  lev=adist(trnsc[i,c("mp_uni")],trnsc[i,c("target_uni")],count=T)
+for(i in 1:dim(trnsc)[1]) if (!trnsc$english[i]){ #this goes through each line
+  #lev #i=2 ; i = 4 ; i =48 (has deletions)
+  #2021-09-02 flipped the order, so that deletions surface as deletions rather than insertions!
+  lev=adist(trnsc[i,c("target_uni")],trnsc[i,c("mp_uni")],count=T)
   trnsc$ins[i]<-attributes(lev)$counts[,,"ins"]
   trnsc$del[i]<-attributes(lev)$counts[,,"del"]
[…]

### 2021-09-03 manuscript.Rmd
ms knits without figure, bug in insertion and matching fixed

--- 2021-09-02 manuscript.Rmd
@@
 
-```{r fig-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
+```{r fig-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.",eval=F}
 aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,trnsc.inc$avg_fr),mean)->xling
@@
 read.table("substitution_bank.txt",header=T)->sub_bank_full
-colnames(sub_bank_full)<-c("target","realization","age")
 sub_bank=sub_bank_full[,1:2]
@@
 read.table("deletion_bank.txt",header=T)->del_bank_full
-colnames(del_bank_full)<-c("target","age")
 del_bank=del_bank_full[,1]
-del_bank$target=factor(del_bank$target,levels=allVC)
+del_bank=factor(del_bank,levels=allVC)
 del_tab= table(del_bank)[allVC]
@@
 read.table("match_bank.txt",header=T)->match_bank_full
-colnames(match_bank_full)<-c("target","age")
 match_bank=match_bank_full[,1]
-match_bank$target=factor(match_bank$target,levels=allVC)
+match_bank=factor(match_bank,levels=allVC)
 match_tab= table(match_bank)[allVC]
@@
 To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
-We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages.
+We entered data from XX studies (including ours), presenting data from XX languages.
 Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.

### 2021-09-03 manuscript.tex
ms knits without figure, bug in insertion and matching fixed

@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
 \PassOptionsToPackage{hyphens}{url}
 %
-\documentclass[english,,man,floatsintext]{apa6}
+\documentclass[
+  english,
+  ,man,floatsintext]{apa6}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
-\else % if luatex or xelatex
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
   \usepackage{unicode-math}
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
+\IfFileExists{microtype.sty}{% use microtype if available
   \usepackage[]{microtype}
@@
 }{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
   \IfFileExists{parskip.sty}{%
@@
     \setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\usepackage{hyperref}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \hypersetup{
   pdftitle={Non-word repetition in children learning Yélî Dnye},
+  pdfauthor={Alejandrina Cristia1 \& Marisa Casillas2,3},
+  pdflang={en-EN},
   pdfkeywords={phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage{graphicx}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 \usepackage{fontspec}
@@
 % \newenvironment{ltable}
-%   {\begin{landscape}\begin{center}\begin{threeparttable}}
-%   {\end{threeparttable}\end{center}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
+%   {\begin{landscape}\centering\begin{threeparttable}}
+%   {\end{threeparttable}\end{landscape}}
+\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
 
@@
   {}{\typeout{Failed to patch abstract.}}
+\patchcmd{\HyOrg@maketitle}
+  {\section{\protect\normalfont{\@title}}}
+  {\section*{\protect\normalfont{\@title}}}
+  {}{\typeout{Failed to patch title.}}
 \makeatother
 \shorttitle{NWR in Yélî Dnye learners}
-\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
-\affiliation{
-\vspace{0.5cm}
-\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
-\authornote{Both authors contributed to study funding, design, data collection, annotation, analyses, writing.
-
-Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
 \keywords{phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy\newline\indent Word count: ~11,500 words}
@@
 \usepackage{csquotes}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
+\ifxetex
+  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
   \usepackage{polyglossia}
   \setmainlanguage[]{english}
+\else
+  \usepackage[main=english]{babel}
+% get rid of language-specific shorthands (see #6817):
+\let\LanguageShortHands\languageshorthands
+\def\languageshorthands#1{}
 \fi
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
 \title{Non-word repetition in children learning Yélî Dnye}
-
+\author{Alejandrina Cristia\textsuperscript{1} \& Marisa Casillas\textsuperscript{2,3}}
 \date{}
+
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
 
@@
 Second, we documented rather weak effects of item length.
-Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
+Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
 Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations, and the extent to which they shed light on the factors universally structuring variation in phonological development at a global and individual level.
@@
 Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
-In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
-After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
+In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
+This allows us to contribute data at the intersection of language typology, language acquisition, and individual variation, as presented in more detail below.
+
+\hypertarget{what-is-nwr}{%
+\subsubsection{What is NWR?}\label{what-is-nwr}}
+
+In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
+For instance, in English, a response of {[}bilig{]} or {[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
+
 NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
-NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
-In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
-The study was designed to contribute to four broad research areas, three via direct results.
-
-The first research area is at the crossing of typology and phonological development.
-Previous work using NWR has preferred relatively universal and early-acquired phonemes (with exceptions including Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely linguistic skills.
-Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
-Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.
-
-Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
-Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, \& Leonard, 2006), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
-It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
-In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
-To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
-We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
-
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/fig-prevlit-1.pdf}
-\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables, 2=dashed, 3=dotted, 4=dotted and dashed).
-Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al.
-(2019) reported on groups of 24-60 Italian learners (black); Stokes et al.
-(2006) on 15 Cantonese learners (blue); Vance et al.
-(2005) on 17-20 English learners (light green); Cristia et al.
-(2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
-Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error.
-Age has been slightly shifted for ease of inspection of different lengths at a given age.}
-\end{figure}
-
-Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
-For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (\textasciitilde{}28\%) and Arabic (\textasciitilde{}15\%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only \textasciitilde{}2\%.
-Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese \textasciitilde{}8\% versus English \textasciitilde{}1\%).
-Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.
-
-The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
+NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007; \textbf{chiat2015non?}).
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children (Armon-Lotem, Jong, \& Meir, 2015; Chiat, 2015; COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
+Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory (see a discussion in Chiat, 2015).
+Broadly, however, NWR scores will necessarily reflect to a certain extent phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
+
+\hypertarget{the-present-work}{%
+\subsubsection{The present work}\label{the-present-work}}
+
+We aimed to contribute to four areas of research. We motivate each in turn.
+
+\hypertarget{nwr-and-typology}{%
+\paragraph{NWR and typology}\label{nwr-and-typology}}
+
+The first research area is at the intersection of typology and phonological development.
+There has been an interest in adapting NWR to different languages, in part for applied purposes.
+In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, (\textbf{chiat2015non?}) discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
+As a result, any one item created will be relatively easier if it more closely resembles real words in the language, making it difficult to balance difficulty when comparing children learning different languages.
+This previous literature also suggests some dimensions of difficulty -- an issue to which we return in the next subsection.
+
+Although this cross-linguistic literature is rich, the potential difficulty associated with specific phonetic targets composing the non-words has received relatively little attention.
+For example, (\textbf{chiat2015non?}) discusses segmental complexity as a function of whether there are consonant clusters -- which is arguably a factor reflecting phonotactics and syllable structure.
+
+In the present study, we thought it was relevant to represent the rich phonological inventory found in Yélî Dnye, by including a variety of phonetic targets.
+Some of them are cross-linguistically rare, in that they are less common across languages than other sounds or phonetic targets.
+Phonologists and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of acquisition, focusing largely on phonotactics (Moreton \& Pater, 2012).
+The correlation between typological frequency and ease of acquisition is typically assumed to emerge from the following causal paths (which are not mutually exclusive):
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Sounds (and sound sequences) that are harder to perceive tend to be misperceived and thus lost diachronically
+\item
+  Sounds (and sound sequences) that are harder to prononunce tend to be mispronounced and thus lost diachronically
+\item
+  Sound sequences that are harder to hold in memory tend to be mispronounced and thus lost diachronically
+\end{enumerate}
+
+Given these causal pathways, we predicted that variation in NWR across items will correlate with the cross-linguistic frequency of the phones composing those items.
+
+\hypertarget{length-effects-on-nwr}{%
+\paragraph{Length effects on NWR}\label{length-effects-on-nwr}}
+
+The second research area we contribute data to is research looking at the impact of word length on NWR repetition within specific languages.
+Some work documents much lower NWR scores for longer, compared to shorter, items {[}e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, and Leonard (2006){]}, whereas differences are negligible in other studies {[}e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019){]}.
+
+It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture -- a hypothesis discussed for instance in (\textbf{chiat2015non?}) (pp.~7-8; see also p.~5).
+In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer.
+A non-systematic meta-analyses does not provide overwhelming support for this hypothesis {[}Cristia and Casillas (2021); SM1{]}.
+
+Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we did not approach this issue within a hypothesis-testing framework but sought instead to provide additional data on the question, which may be re-used in future meta- or mega-analytic analyses.
+
+\hypertarget{individual-variation-effects-on-nwr}{%
+\paragraph{Individual variation effects on NWR}\label{individual-variation-effects-on-nwr}}
+
+The third research area we contribute data to relates to the possibility that children differ from each other in NWR scores in systematic ways.
 Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
@@
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance et al., 2005).
-Prior research typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+
+Our focus on age stems from previous work, where performance increases with child age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
 Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.
-
-Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b); Maddieson (2013a); with no exceptions to our knowledge{]}.
+Common explanations for advantages for first- over later-born children include differential allocation of familial resources, particularly parental behaviors of cognitive stimulation (Lehmann, Nuevo-Chiquero, \& Vidal-Fernandez, 2018).
+Regarding child sex, no such effect has been found in previous NWR research (Chiat \& Roy, 2007), and in other language tasks evidence is mixed.
+Finally, prior research typically finds no significant differences as a function of maternal education (Balladares, Marshall, \& Griffiths, 2016; e.g., Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017).
+In general, maternal education correlates with child language outcomes, including vocabulary reports (Frank, Braginsky, Yurovsky, \& Marchman, 2017) and word comprehension studies (Scaff, 2019).
+The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children (see discussion in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
+
+\hypertarget{nwr-as-a-function-of-language-and-culture}{%
+\paragraph{NWR as a function of language and culture}\label{nwr-as-a-function-of-language-and-culture}}
+
+The fourth research goal we pursued is to use NWR with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020; with exceptions including Gallagher, 2014).
+To our knowledge, it has never been used with speakers of languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities Maddieson, 2013b, 2013a).
+
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
-That said, Cristia et al.
-(2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
-In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance (e.g., Gathercole, Willis, \& Baddeley, 1991).
-Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills.
-Indeed, Cristia et al.
-(2020) present the task as being a good index of the development of \enquote{short-hand-like} representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
-Aside from Cristia et al.
-(2020)'s hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).
+That said, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
+
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading; for instance, better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991).
+Therefore, it may not be a general index of, for instance, phonological development, but reflect only certain non-universal language skills.
+Indeed, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) present their task as being a good index of the development of ``short-hand-like'' representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+We return to the question of what was measured here in the Discussion.
+
+Aside from Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)'s hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).
+
 Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
-This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.
+This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, which involved English participants.
+
+\hypertarget{yuxe9luxee-dnye-phonology-and-community}{%
+\subsubsection{Yélî Dnye phonology and community}\label{yuxe9luxee-dnye-phonology-and-community}}
 
 Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
-Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites.
+Rossel Island lies 250 nautical miles off the coast of mainland PNG and is surrounded by a barrier reef.
+As a result, transport to and from the island is both infrequent and irregular.
 International phone calls and digital exchanges that require significant data transfer are typically not an option.
@@
 \hypertarget{yuxe9luxee-dnye-phonology}{%
-\subsubsection{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}}
+\paragraph{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}}
 
 Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
-The Yélî sound system, much like its baroque grammatical system (Levinson, 2020), is unlike any other in the region.
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.).
-Thus, with respect to our first research goal, Yélî Dnye is a good language to test because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
+The Yélî sound system, much like its baroque grammatical system (Levinson, 2021), is unlike any other in the region.
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2021; Maddieson \& Levinson, in preparation).
+Thus, with respect to our first research goal, Yélî Dnye is a good language to use because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
 
@@
 With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop system includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (2020) for details).\footnote{We use Levinson's (2020) under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
-In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+The core oral stop system includes both singleton (/p/, /t/, /ṭ/\footnote{We use Levinson's (2021) under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).}, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with a complete range of nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (2021) for details).
+The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
 Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
-
-Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020).
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to occur as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+
+\begin{verbatim}
+## count_syll
+##  1  2  3  4  5  6  8 
+## 15 39 29 12  2  2  1
+\end{verbatim}
+
+Our second research goal is to measure the effect of non-word length on NWR, which may need to be interpreted taking into account typical word length in the language.
+We estimated word length in words found in a conversational corpus (see Stimuli section for details), where the distribution of length was: 15\% monosyllabic, 39\% disyllabic, 29\% trisyllabic, and the remaining 17\% being longer than that.
 The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} /ṇṭæp/ \emph{Spondylus shell}) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \enquote{ndapî} /'ṇṭæ.pɯ/).
-There are also a handful of words starting with /æ/ (e.g., \enquote{ala} /æ.'læ/ \emph{here}) and a small collection of single-vowel grammatical morphemes (see Levinson (2020) for details).
-
-Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun.
-For example, Peute and colleagues' (n.d.) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
+A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \emph{ndap} /ṇṭæp/ `Spondylus shell') and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \emph{ndapî} /`ṇṭæpɯ/).
+There are also a handful of words starting with /æ/ (e.g., \emph{ala} /æ'læ/ 'here') and a small collection of single-vowel grammatical morphemes (see Levinson (2021) for details).
+
+Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun.
+For example, Peute and colleagues' (In preparation) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
 We hope the present study contributes to this growing line of work.
 
+Before closing this section, it bears mentioning that the language has an established orthography, which includes distinct graphemes for all the contrasts on which our items are based.
+Some children in our sample will have started school.
+Reading and writing instruction is currently done only in English (other than writing one's name).
+This was probably not the case for the majority of mothers of the children in our sample, who will have learned to read and write in Yélî Dnye during their first three years at school.
+This change from Yélî-initial instruction to English only occurred about 15 years ago.
+It is possible that there is also some home teaching of Yélî reading and writing, notably for reading the bible.
+
 \hypertarget{the-yuxe9luxee-community}{%
-\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}}
+\paragraph{The Yélî community}\label{the-yuxe9luxee-community}}
 
 Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation.
-Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
-There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.
-
-The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+Specifically, we investigated potential effects of age, child sex, maternal education, and birth order.
+There is nothing particular to note regarding age and child sex, but we have some comments that pertain to the other two factors.
+
+The typical household in our dataset includes seven individuals (typically, a mixed-sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
 These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
@@
 
-Most Yélî parents are swidden horticulturalists.
-Within a group of households, it is often the case that older adolescents and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
-Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
-Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available (Casillas et al., 2020).
+Most Yélî parents are swidden horticulturalists, who occasionally fish.
+Within a group of households, it is often the case that older adolescents and adults spend their day tending to their farm plots (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde10 members) independent and mixed-age child play groups (Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020).
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas, Brown, \& Levinson, 2020).
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
@@
 To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
-Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
-Overall, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
+In addition, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply be due to random factors like living close to a school or having relatives there.
 
@@
 It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
-This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
-
-We add some observations that will help us integrate this study to the broader investigation of NWR across cultures.
-As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia et al., 2020).
-However, Cristia et al.
-(2020) also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
-The Yélî community can help us shed further light onto this question because direct speech to children under 3;0 is relatively infrequent in this community too (Casillas et al., 2020).
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas et al., 2020).
-Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in Cristia et al.
-(2020).
-
-\hypertarget{nwr-design-and-analysis-adaptations}{%
-\subsubsection{NWR design and analysis adaptations}\label{nwr-design-and-analysis-adaptations}}
-
-In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language.
-For instance, in English, a response of {[}bilig{]} or {[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
-There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
-For example, while items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation (Brandeker \& Thordardottir, 2015).
-Additionally, while some studies have used 10-15 non-words (e.g., Cristia et al., 2020), others have employed up to 46 unique items (Piazzalunga et al., 2019).
-Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon et al., 2007).
-
-Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
-We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.~less challenging sounds.
-However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
-Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
-However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
-That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
+This contrasts with the picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
+
+We add some observations that will help us integrate this study into the broader investigation of NWR across cultures.
+As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
+However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is also relatively infrequent in this community (Casillas, Brown, \& Levinson, 2020).
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2020).
+Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).
 
@@
 
+\hypertarget{participants}{%
+\subsubsection{Participants}\label{participants}}
+
+This study was approved as part of a larger research effort by the second author.
+The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041), including the use of verbal (not written) consent.
+As discussed in subsection ``The Yélî community,'' the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, in press).
+Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
+Regardless of whether they completed the task, children were given a small snack as compensation.
+Children who showed initial interest but then decided not to participate were also given the snack.
+
+We tested a total of 55 children from 38 families spread across four hamlet regions.
+We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5), or were 13 years old or older (N=2; we tested these teenagers to put younger children at ease).
+The remaining 40 children (14 girls) were aged from 3 to 10 years (M = 6.40 years, SD = 1.50 years).
+In terms of birth order, 6 were born first, 5 second, 2 third, 7 fourth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
+These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
+Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
+We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 children only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, in press).
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.}
+
 \hypertarget{stimuli}{%
@@
 
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares et al., 2016; Wilsenach, 2013).
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 40 final items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-46 items: Jaber-Awida, 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019), with the entire test inventory of 40 final items distributed across children.
+We used a relatively large number of items to explore effects of length and phonological complexity.
+However, aware that this large item inventory might render the task longer and more tiresome, we split items across children.
+Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
 
 A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/, /n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible 2- to 4-syllable sequences.
-These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
+These candidates were automatically removed from consideration if they appeared in the most recent dictionary (Levinson, 2021).
 The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
@@
 Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
-Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, 2020), it was also possible that we initially generated items with illegal, but currently undocumented constraints.
-Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
+Additionally, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
 These candidates were then presented to one informant, for a final check that they did not mean anything.
@@
 The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present.
-Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \enquote{yî} /yɯ/; identified as \enquote{yi} /yi/, \emph{tree}).
-Additionally, an error was made when preparing files for annotation, resulting in two items being merged (\enquote{tpâ} /tpɑ/ and \enquote{tp:a} /tpæ̃/).
+Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \emph{yî} /yɯ/; identified as \emph{yi} /yi/, `tree').
+Additionally, an error was made when preparing files for annotation, resulting in two items being merged (\emph{tpâ} /tpɑ/ and \emph{tp:a} /tpæ̃/).
 These three problematic items are not described here, and removed from the analyses below.
@@
 
-A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and split it into two sub-lists, to generate 40 different elicitation sets.
+A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and to split it into two sub-lists, to generate 40 different elicitation sets.
 The 40 elicitation sets are available online from \href{https://osf.io/dtxue/}{osf.io/dtxue/}.
@@
 
+\hypertarget{cross-linguistic-frequency}{%
+\paragraph{Cross-linguistic frequency}\label{cross-linguistic-frequency}}
+
 To inform our analyses, we estimated the typological frequency of all phonological segments present in the target items using the PHOIBLE cross-linguistic phonological inventory database (Moran \& McCloy, 2019).
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
-Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
+While PHOIBLE is a unprecedented in its scope, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
+Note that nearly half of the phones in PHOIBLE are only attested in one language (Steven Moran, personal communication).
 Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0\% (i.e., its cross-linguistic percentile is zero).
 
-Additionally, we estimated frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+\hypertarget{within-language-frequency}{%
+\paragraph{Within-language frequency}\label{within-language-frequency}}
+
+Additionally, we estimated frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2020).
+That corpus was constituted by sampling from audio-recordings (7-9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
+The researchers selected 9 2.5-minute clips randomly and 11 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity).
+These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the people talking personally.
+For more details, please refer to Casillas, Brown, and Levinson (2020).
+
+For the present study, we extracted the transcriptions of adult speech (i.e., removing key child and other children's speech) and split them into words using white space.
+We then removed all English and Pidgin words.
+The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types.
+To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, sounds from our non-word items that did not occur in the corpus were not considered (i.e., they were declared NA so that they do not count for analyses).
 
@@
 
-In adapting the typical NWR procedure for this context, we balanced three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them (i.e., from other children who had participated); that children would feel comfortable doing this task with us; and that community members would feel comfortable having their children do this task with us.
-
-We tested in four different sites spread across the northeastern region of the island, making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from \enquote{spreading} between children through hearsay.
+There is some variation in procedure in previous work.
+For example, while items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation (Brandeker \& Thordardottir, 2015).
+
+In adapting the typical NWR procedure for our context, we balanced three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them (i.e., from other children who had participated); that children would feel comfortable doing this task with us; and that community members would feel comfortable having their children do this task with us.
+
+We tested in four different sites spread across the northeastern region of the island, making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from `spreading' between children through hearsay.
 Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
@@
 
-Previous work typically reports two scores: a binary word-level exact repetition score, and a phoneme-level score, defined as the number of phonemes that can be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt; as in Cristia et al., 2020).
+Previous work typically reports two scores: a binary word-level exact repetition score, and a phoneme-level score, defined as the number of phonemes that can be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt; as in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
 Previous work does not use distance metrics, but we report these rather than the phoneme-level scores because they are more informative.
@@
 An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (insertions were not attested in this study).
-
-\hypertarget{participants}{%
-\subsubsection{Participants}\label{participants}}
-
-This study was approved as part of a larger research effort by the second author.
-The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
-As discussed in subsection \enquote{The Yélî community}, the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, n.d.).
-Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
-Regardless of whether they completed the task, children were given a small snack as compensation.
-Children who showed initial interest but then decided not to participate were also given the snack.
-
-We tested a total of 55 children from 38 families spread across four hamlet regions.
-We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5), or were 13 years old or older (N=2; we tested these teenagers to put younger children at ease).
-The remaining 40 children (14 girls) were aged from 3 to 10 years (M = 6.50 years, SD = 1.50 years).
-In terms of birth order, 6 were first borns, 5 second, 2 third, 7 fourth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
-These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
-Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
-We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 children only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
-Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.}
 
@@
 We could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
-The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
+The percentage we observe here is much higher than reported in the study by Castro and colleagues, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
 Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
 
-\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
-\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}
+\hypertarget{nwr-and-typology-nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
+\subsubsection{NWR and typology: NWR as a function of cross-linguistic phone frequency}\label{nwr-and-typology-nwr-as-a-function-of-cross-linguistic-phone-frequency}}
 
@@
 To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and allowed slopes to vary over the random effects child ID and target ID.
+The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, in interaction with cross-linguistic phone frequency, and allowed intercepts to vary over the random effects child ID and target ID.
 
 We could include 826 observations, from 40 children producing in any given trial one of 40 potential target words.
-The analysis revealed a main effect of age (ß = 0.35, SE ß = 0.13, p \textless{} 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.78, SE ß = 0.19, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:fig-xling-freq}.
+The analysis revealed a main effect of age (ß = 0.39, SE ß = 0.13, p \textless{} 0.01), with older children repeating more items correctly.
+It also revealed a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.80, SE ß = 0.19, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:fig-xling-freq}.
 Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(38) = .544.
 
+Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for older children.
+Inspection of Figure \ref{fig:fig-xling-freq-age3} suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.
+
 \begin{figure}[!t]
 
-{\centering \includegraphics[width=0.65\linewidth]{nwr.by.freq.ITEM}
+{\centering \includegraphics[width=0.65\linewidth]{manuscript_files/figure-latex/fig-xling-freq-age3-1}
 
@@
 
-\caption{NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}\label{fig:fig-xling-freq}
+\caption{Model predicting NWR scores from child age (x axis) and the average frequency with which each phone is found across languages (mean, or plus/minus one standard deviation).}\label{fig:fig-xling-freq-age3}
 \end{figure}
 
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
+\begin{figure}[!t]
+
+{\centering \includegraphics[width=0.65\linewidth]{manuscript_files/figure-latex/fig-xling-freq-age1-1}
+
+}
+
+\caption{(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (median split on scaled age).}\label{fig:fig-xling-freq-age1}
+\end{figure}
+
+\begin{figure}[!t]
+
+{\centering \includegraphics[width=0.65\linewidth]{manuscript_files/figure-latex/fig-xling-freq-age2-1}
+
+}
+
+\caption{(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (mean, or plus/minus one standard deviation).}\label{fig:fig-xling-freq-age2}
+\end{figure}
+
+\hypertarget{nwr-and-typology-nwr-as-a-function-of-within-language-phone-frequency}{%
+\subsubsection{NWR and typology: NWR as a function of within-language phone frequency}\label{nwr-and-typology-nwr-as-a-function-of-within-language-phone-frequency}}
+
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: The same perception and production pressures that shape languages diachronically could affect a language's lexicon, so that sounds that are easier to perceive or produce are more frequent within a language than those that are harder.
+If so, children will have more experience with the easier sounds, and they may thus be better able to to represent and repeat non-words containing them simply because of the additional exposure.
+
 Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies {[}r(27)=0.50, p \textless{} 0.01{]}; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies {[}r(38)=0.73, p \textless{} 0.001{]}.
@@
 
-\hypertarget{patterns-in-nwr-mispronunciations}{%
-\subsubsection{Patterns in NWR mispronunciations}\label{patterns-in-nwr-mispronunciations}}
+\hypertarget{follow-up-analyses-patterns-in-nwr-mispronunciations}{%
+\subsubsection{Follow-up analyses: Patterns in NWR mispronunciations}\label{follow-up-analyses-patterns-in-nwr-mispronunciations}}
 
 We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts so as to base our generalizations on more data.
-There were no cases of insertion, and deletions were very rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were 820 cases of substitutions, \textasciitilde{}16.95 of the 4839 phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (\textasciitilde{}97.74 of unmatched phones).
+There were no cases of insertion, and deletions were very rare: there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets).
+We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde{} of unmatched phones).
 To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
@@
 \midrule
-Nasal Target & 100 & 0 & 39 & 17 & 64.1 & 0.0 & 25.0 & 10.9\\
-Oral Target & 1992 & 12 & 52 & 205 & 88.1 & 0.5 & 2.3 & 9.1\\
+Nasal Target & 101 & 0 & 39 & 17 & 64.3 & 0.0 & 24.8 & 10.8\\
+Oral Target & 1988 & 17 & 52 & 204 & 87.9 & 0.8 & 2.3 & 9.0\\
 \bottomrule
@@
 We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not, generating separate estimates for nasal and oral vowels.
-The nasal vowels in our stimuli occur in \textasciitilde{}1.40\% of languages' phonologies (range 0\% to 3\%);
-whereas oral vowels in our stimuli occur in \textasciitilde{}31.55\% of languages' phonologies (range 3\% to 92\%).
-As noted above, type frequency within the language is correlated with cross-linguistic frequency, and thus these two types of sounds also differ in the former: Their type frequencies in Yélî Dnye are: nasal vowels \textasciitilde{}0.03‰ (range 0.00‰ to 0.05‰) versus oral \textasciitilde{}0.23‰ (range 0.02‰ to 0.76‰).
+The nasal vowels in our stimuli occur in \textasciitilde1.40\% of languages' phonologies (range 0\% to 3\%);
+whereas oral vowels in our stimuli occur in \textasciitilde31.55\% of languages' phonologies (range 3\% to 92\%).
+As noted above, frequency within the language is correlated with cross-linguistic frequency, and thus these two types of sounds also differ in the former: Their frequencies in Yélî Dnye are: nasal vowels \textasciitilde0.03‰ (range 0.00‰ to 0.05‰) versus oral \textasciitilde0.23‰ (range 0.02‰ to 0.76‰).
 
 We distinguished errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (and were therefore a quality error), shown in Table \ref{tab:tab-v}.
-We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.90 versus 11.90).
+We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.70 versus 12.10).
 Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
@@
 \midrule
-Complex Target & 257 & 0 & 218 & 48 & 49.1 & 0.0 & 41.7 & 9.2\\
-Simple Target & 1425 & 6 & 2 & 120 & 91.8 & 0.4 & 0.1 & 7.7\\
+Complex Target & 198 & 0 & 219 & 44 & 43.0 & 0.0 & 47.5 & 9.5\\
+Simple Target & 1482 & 13 & 3 & 117 & 91.8 & 0.8 & 0.2 & 7.2\\
 \bottomrule
@@
 These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.}
-The complex consonants in our stimuli occur in \textasciitilde{}17.33\% of languages' phonologies (range 0\% to 78\%);
-whereas simple consonants in our stimuli occur in \textasciitilde{}67.62\% of languages' phonologies (range 13\% to 96\%).
+The complex consonants in our stimuli occur in \textasciitilde17.33\% of languages' phonologies (range 0\% to 78\%);
+whereas simple consonants in our stimuli occur in \textasciitilde67.62\% of languages' phonologies (range 13\% to 96\%).
 Again these groups of sounds differ in their frequency within the language.
-Their type frequencies in Yélî Dnye are: complex consonants \textasciitilde{}0.04‰ (range 0.00‰ to 0.10‰) versus simple consonants \textasciitilde{}0.32‰ (range 0.06‰ to 0.55‰).
-
-Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving simple consonants (50.90 versus 8.20\%).
+Their type frequencies in Yélî Dnye are: complex consonants \textasciitilde0.04‰ (range 0.00‰ to 0.10‰) versus simple consonants \textasciitilde0.32‰ (range 0.06‰ to 0.55‰).
+
+Table \ref{tab:tab-c} showed that errors involving complex consonant targets were more common than those involving simple consonants (57 versus 8.20\%).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other.
-Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency {[}r(319.72)=0.76, p \textless{} 0.001{]} was greater than that with within-language frequency {[}r(731.10)=0.45, p = 0.05{]}.
-
-\hypertarget{nwr-scores-as-a-function-of-item-length}{%
-\subsubsection{NWR scores as a function of item length}\label{nwr-scores-as-a-function-of-item-length}}
+Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency {[}r(346.78)=0.74, p \textless{} 0.001{]} was greater than that with within-language frequency {[}r(817.23)=0.39, p = 0.09{]}.
+
+\hypertarget{length-effects-on-nwr-1}{%
+\subsubsection{Length effects on NWR}\label{length-effects-on-nwr-1}}
 
@@
 
-\hypertarget{factors-structuring-individual-variation}{%
-\subsubsection{Factors structuring individual variation}\label{factors-structuring-individual-variation}}
+\hypertarget{individual-variation-effects-on-nwr-1}{%
+\subsubsection{Individual variation effects on NWR}\label{individual-variation-effects-on-nwr-1}}
 
@@
 
-\hypertarget{associations-between-nwr-and-cross-linguistic-frequency}{%
-\subsubsection{Associations between NWR and cross-linguistic frequency}\label{associations-between-nwr-and-cross-linguistic-frequency}}
+\hypertarget{nwr-and-typology-1}{%
+\subsubsection{NWR and typology}\label{nwr-and-typology-1}}
+
+We looked at two things: Associations between NWR and cross-linguistic frequency, and qualitative mispronunciation patterns. RESTRUCTURE TO MAKE THIS CLEARER!
 
@@
 
-\hypertarget{item-length}{%
-\subsubsection{Item Length}\label{item-length}}
+\hypertarget{length-effects-and-nwr}{%
+\subsubsection{Length effects and NWR}\label{length-effects-and-nwr}}
 
@@
 In broad terms, children should have higher NWR scores for shorter items.
-That said, previous work summarized in the Introduction has shown both very small (e.g., Piazzalunga et al., 2019) and very large (e.g., Cristia et al., 2020) effects of stimulus length.
+That said, previous work summarized in the Introduction has shown both very small (e.g., Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019) and very large (e.g., Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020) effects of stimulus length.
 Setting aside our monosyllabic stimuli (which contained typologically infrequent segments with lower NWR scores, as just discussed), we examined effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
@@
 
-\hypertarget{individual-differences}{%
-\subsubsection{Individual differences}\label{individual-differences}}
+\hypertarget{individual-variation-effects-on-nwr-2}{%
+\subsubsection{Individual variation effects on NWR}\label{individual-variation-effects-on-nwr-2}}
 
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese Stokes et al., 2006; but note Cristia et al., 2020 is an exception).
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance, Stackhouse, \& Wells, 2005; Italian Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019; Cantonese Stokes, Wong, Fletcher, \& Leonard, 2006; but note Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020 is an exception).
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) on NWR scores.
+In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017) on NWR scores.
 We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
-In fact, maternal education correlated with NWR score at about r\textasciitilde{}.1, which is small.
-Similarly, NWR scores may not vary greatly with participant gender according to previous work (Chiat \& Roy, 2007), and for that as well we find effects of about that size.
+In fact, maternal education correlated with NWR score at about r\textasciitilde.1, which is small.
+Similarly, NWR scores may not vary greatly with participant sex according to previous work (Chiat \& Roy, 2007), and for that as well we find effects of about that size.
 
@@
 In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score.
-However, the effect size was larger than that found for the other two factors and it is far from negligible, at r\textasciitilde{}.2 or Cohen's d\textasciitilde{}0.41.
-In fact, two large studies with therefore precise estimates found effects of about d\textasciitilde{}.2 (Barclay, 2015; Havron et al., 2019), which would suggest the effects we found are larger.
+However, the effect size was larger than that found for the other two factors and it is far from negligible, at r\textasciitilde.2 or Cohen's d\textasciitilde0.41.
+In fact, two large studies with therefore precise estimates found effects of about d\textasciitilde.2 (Barclay, 2015; Havron et al., 2019), which would suggest the effects we found are larger.
 We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
@@
 To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
-We entered data from 14 studies (including ours), presenting data from 12 languages.
-Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes et al.
-(2006); English by Vance et al.
-(2005); Italian by Piazzalunga et al.
-(2019); Mandarin by Lei et al.
+We entered data from XX studies (including ours), presenting data from XX languages.
+Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes, Wong, Fletcher, and Leonard (2006); English by Vance, Stackhouse, and Wells (2005); Italian by Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019); Mandarin by Lei et al.
 (2011); Persian by Farmani et al.
-(2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares et al.
-(2016); Swedish by Kalnak et al.
-(2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al.
-(2020); and Yélî Dnye from the present study.
+(2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares, Marshall, and Griffiths (2016); Swedish by Kalnak, Peyrard-Janvid, Forssberg, and Sahlén (2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020); and Yélî Dnye from the present study.
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
@@
 
-\begin{figure}
-\centering
-\includegraphics{manuscript_files/figure-latex/fig-prevlit-overall-1.pdf}
-\caption{\label{fig:fig-prevlit-overall}NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages.
-The legend indicates language and the length of non-words (in syllables).
-Central tendency is mean; error is one standard error.}
-\end{figure}
+FIGURE REMOVED
 
@@
 Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
-In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: Casillas et al.
-(2020) hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
+In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: Casillas, Brown, and Levinson (2020) hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
 Cristia, Dupoux, Gurven, and Stieglitz (2019) estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
-Finally, Scaff et al.
-(2021) used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.~a female adult that was not temporally close to speech by others would count as child-directed).
+Finally, Scaff, Stieglitz, Casillas, and Cristia (2021) used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.~a female adult that was not temporally close to speech by others would count as child-directed).
 A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
 
-That said, Cristia et al.
-(2020) also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
+That said, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
 This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
@@
 Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
-Cristia et al.
-(2020) administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
+Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
 This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
@@
 It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.~Sotho non-words having 4-7 syllables in length.
-To be more certain whether language characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
+To be more certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
 We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
@@
 
+\hypertarget{limitations}{%
+\subsubsection{Limitations}\label{limitations}}
+
+Before closing, we would like to point out some salient limitations of the current work.
+To begin with, we only employed one set of non-words, in which not all characteristics that previous work suggest matter were manipulated (\textbf{chiat2015non?}).
+As a result, we only have a rather whole-sale measure of performance, and we do not know to what extent lexical knowledge, pure phonological knowledge, and working memory, among others, contribute to children's performance.
+Similarly, our items varied systematically in length and typological frequency of the sounds included, but not in other potential dimensions (such as whether the items contained morphemes of the language or not).
+Additionally, we only had a single person interacting with children as well as interpreting children's production, so we do not know to what extent our findings generalize to other experimenters and research assistants.
+In addition, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
+Although we know from other work that adults' perceptual performance on these types of sounds is well above chance from audio-only presentation {[}REF{]}, language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.
+
 \hypertarget{conclusions}{%
@@
 While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
 Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including: Why do NWR scores would pattern differently across samples?
@@
 \hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-armon2015assessing}{}%
+Armon-Lotem, S., Jong, J. de, \& Meir, N. (2015). \emph{Methods for assessing multilingual children: Disentangling bilingualism from specific language impairment}. Bristol: Multilingual matters.
+
 \leavevmode\hypertarget{ref-balladares2016socio}{}%
@@
 (2016).
-Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
-\emph{First Language}, \emph{36}(3), 338--351.
+{Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers}.
+\emph{{First Language}}, \emph{36}(3), 338--351.
 \url{https://doi.org/10.1177/0142723715626067}
@@
 Duranti, E.
-Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language Socialization} (pp.
+Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of {Language Socialization}} (pp.
 29--55).
@@
 (2014).
-The interactional context of language learning in Tzeltal.
+The interactional context of language learning in {T}zeltal.
 In I.
@@
 Kurumada, \& B.
-Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of Eve V.
-Clark} (pp.
-51--82).
+Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of {Eve V.
+Clark}} (pp.
+51-\/-82).
 Amsterdam, NL: John Benjamins.
@@
 Brown, P., \& Casillas, M.
-(n.d.).
-Childrearing through social interaction on Rossel Island, PNG.
+(in press).
+Childrearing through social interaction on {Rossel Island, PNG}.
 In A.
@@
 Fentiman \& M.
-Goody (Eds.), \emph{Esther Goody revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
+Goody (Eds.), \emph{{Esther Goody} revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
 XX--XX).
@@
 \leavevmode\hypertarget{ref-casillas2020early}{}%
-Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a Papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). {Early language experience in a Papuan community}. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
 
@@
 (1998).
-The illiterate brain.
-Learning to read and write during childhood influences the functional organization of the adult brain.
-\emph{Brain: A Journal of Neurology}, \emph{121}(6), 1053--1063.
+{The illiterate brain.
+Learning to read and write during childhood influences the functional organization of the adult brain.} \emph{Brain: A Journal of Neurology}, \emph{121}(6), 1053--1063.
 \url{https://doi.org/10.1093/brain/121.6.1053}
+
+\leavevmode\hypertarget{ref-chiat2015nonword}{}%
+Chiat, S.
+(2015).
+Non-word repetition.
+In S.
+Armon-Lotem, J.
+de Jong, \& N.
+Meir (Eds.), \emph{Methods for assessing multilingual children: Disentangling bilingualism from specific language impairment} (pp.
+125--150).
+Bristol: Multilingual matters.
 
@@
 
+\leavevmode\hypertarget{ref-cristia2021supplementary}{}%
+Cristia, A., \& Casillas, M. (2021). Supplementary materials to "non-word repetition in children learning yélî dnye". Retrieved from \url{https://osf.io/5qspb/wiki/home/}
+
 \leavevmode\hypertarget{ref-cristia2019child}{}%
@@
 Child-directed speech is infrequent in a forager-farmer population.
-\emph{Child Development}, \emph{90}(3), 759--773.
+\emph{{Child Development}}, \emph{90}, 759--773.
 \url{https://doi.org/10.1111/cdev.12974}
@@
 (2020).
-Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the Tsimane'.
+Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the {T}simane'.
 \emph{PLoS ONE}, \emph{15}(9), e0237702.
-\url{https://doi.org/https://doi.org/10.1371/journal.pone.0237702}
+https://doi.org/\url{https://doi.org/10.1371/journal.pone.0237702}
 
@@
 Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}, 177--195.
 
@@
 (2018).
-Normalization of the non-word repetition test in Farsi-speaking children.
+Normalization of the non-word repetition test in {F}arsi-speaking children.
 \emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
@@
 \leavevmode\hypertarget{ref-foley1986papuan}{}%
-Foley, W. A. (1986). \emph{The Papuan languages of New Guinea}. Cambridge, UK: Cambridge University Press.
+Foley, W. A. (1986). \emph{{The Papuan languages of New Guinea}}. Cambridge, UK: Cambridge University Press.
 
@@
 \leavevmode\hypertarget{ref-gallagher2014identity}{}%
-Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
-
-\leavevmode\hypertarget{ref-gallon2007non}{}%
-Gallon, N., Harris, J., \& Van der Lely, H.
-(2007).
-Non-word repetition: An investigation of phonological complexity in children with Grammatical SLI.
-\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
+Gallagher, G. (2014). {An identity bias in phonotactics: Evidence from Cochabamba Quechua}. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
 
@@
 \leavevmode\hypertarget{ref-jabere2018xperiment}{}%
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual {A}rabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}, 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
 
@@
 (2014).
-Nonword repetition--a clinical marker for specific language impairment in Swedish associated with parents' language-related problems.
+Nonword repetition--a clinical marker for specific language impairment in {S}wedish associated with parents' language-related problems.
 \emph{PloS One}, \emph{9}(2), e89544.
@@
 (2013).
-Non-word repetition performance in Slovak-speaking children with and without SLI: novel scoring methods.
-\emph{International Journal of Language and Communication Disorders}, \emph{48}(1), 78--89.
+{Non-word repetition performance in Slovak-speaking children with and without SLI: novel scoring methods}.
+\emph{{International Journal of Language and Communication Disorders}}, \emph{48}(1), 78--89.
 \url{https://doi.org/10.1111/j.1460-6984.2012.00189.x}
@@
 
+\leavevmode\hypertarget{ref-lehmann2018early}{}%
+Lehmann, J.-Y.
+K., Nuevo-Chiquero, A., \& Vidal-Fernandez, M.
+(2018).
+The early origins of birth order differences in children's outcomes and parental behavior.
+\emph{Journal of Human Resources}, \emph{53}(1), 123--156.
+
 \leavevmode\hypertarget{ref-lei2011developmental}{}%
@@
 \leavevmode\hypertarget{ref-levinsonYDgrammar}{}%
-Levinson, S. C. (2020). \emph{A grammar of Yélî Dnye, the Papuan language of Rossel Island}. Berlin, Boston: De Gruyter Mouton.
+Levinson, S. C. (2021). \emph{A grammar of {Yélî D}nye, the {P}apuan language of {R}ossel {I}sland}. Berlin, Boston: De Gruyter Mouton.
 
@@
 \leavevmode\hypertarget{ref-maddieson2009phonology}{}%
-Maddieson, I. (2009). Phonology, naturalness and universals. \emph{Poznán Studies in Contemporary Linguistics}, \emph{45}(1), 131--140.
+Maddieson, I. (2009). Phonology, naturalness and universals. \emph{Pozn{á}n Studies in Contemporary Linguistics}, \emph{45}(1), 131--140.
 
@@
 \leavevmode\hypertarget{ref-maddiesonIPphoneticsYD}{}%
-Maddieson, I., \& Levinson, S. C. (n.d.). \emph{The phonetics of Yélî Dnye, the language of Rossel Island}.
+Maddieson, I., \& Levinson, S. C. (in preparation). \emph{The phonetics of {Yélî Dnye}, the language of {Rossel Island}}.
 
@@
 (2017).
-Independent and combined effects of socioeconomic status (SES) and bilingualism on children's vocabulary and verbal short-term memory.
+Independent and combined effects of socioeconomic status ({SES}) and bilingualism on children's vocabulary and verbal short-term memory.
 \emph{Frontiers in Psychology}, \emph{8}, 1442.
@@
 (2016).
-Disentangling SLI and bilingualism using sentence repetition tasks: The impact of L1 and L2 properties.
+{Disentangling SLI and bilingualism using sentence repetition tasks: The impact of L1 and L2 properties}.
 \emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
@@
 \leavevmode\hypertarget{ref-moreton2012structure}{}%
-Moreton, E., \& Pater, J. (2012). Structure and substance in artificial-phonology learning, part II: Substance. \emph{Language and Linguistics Compass}, \emph{6}(11), 702--718.
-
-\leavevmode\hypertarget{ref-mulder2017early}{}%
-Mulder, H., Verhagen, J., Van der Ven, S.
-H., Slot, P.
-L., \& Leseman, P.
-P.
-(2017).
-Early executive function at age two predicts emergent mathematics and literacy at age five.
-\emph{Frontiers in Psychology}, \emph{8}, 1706.
+Moreton, E., \& Pater, J. (2012). Structure and substance in artificial-phonology learning, part {II}: substance. \emph{Language and Linguistics Compass}, \emph{6}(11), 702--718.
 
 \leavevmode\hypertarget{ref-peuteIPconsonants}{}%
-Peute, A. A. K., Fikkert, P., \& Casillas, M. (n.d.). Early consonant production in Yélî Dnye and Tseltal.
+Peute, A. A. K., Fikkert, P., \& Casillas, M. (In preparation). Early consonant production in {Y}élî {D}nye and {T}seltal.
 
@@
 (2019).
-An articulatory-based disyllabic and trisyllabic Non-Word Repetition test: reliability and validity in Italian 3-to 7-year-old children.
+{An articulatory-based disyllabic and trisyllabic Non-Word Repetition test: reliability and validity in Italian 3-to 7-year-old children}.
 \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
@@
 \leavevmode\hypertarget{ref-polivsenska2014improving}{}%
-Polišenská, K., \& Kapalková, S. (2014). Improving child compliance on a computer-administered nonword repetition task. \emph{Journal of Speech, Language and Hearing Research}, \emph{57}(3).
+Polišenská, K., \& Kapalková, S. (2014). Improving child compliance on a computer-administered nonword repetition task. \emph{{Journal of Speech, Language and Hearing Research}}, \emph{57}(3).
 
@@
 (2006).
-A Swedish non-word repetition test for preschool children.
-\emph{Scandinavian Journal of Psychology}, \emph{47}(3), 187--192.
+{A Swedish non-word repetition test for preschool children}.
+\emph{{Scandinavian Journal of Psychology}}, \emph{47}(3), 187--192.
 \url{https://doi.org/10.1111/j.1467-9450.2006.00506.x}
+
+\leavevmode\hypertarget{ref-scaff2019}{}%
+Scaff, C. (2019). \emph{Beyond {WEIRD}: An interdisciplinary approach to language acquisition} (PhD thesis).
 
@@
 Nonword repetition and sentence repetition as clinical markers of specific language impairment: The case of cantonese.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{49}(2), 219--236.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{49}, 219--236.
 
@@
 (2013).
-Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
-\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
+{Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners}.
+\emph{{Per Linguam: A Journal of Language Learning= Per Linguam: Tydskrif Vir Taalaanleer}}, \emph{29}(2), 17--32.
 \url{https://doi.org/10.5785/29-2-554}
 
+\end{CSLReferences}
+
 \end{document}

### 2021-09-03 wrangling.R

@@ -5,7 +5,7 @@ read.delim("NWR-transcription.txt", encoding="UTF-8")->trnsc
 
 
 #DATA REMOVAL *** ATTENTION ***
-#REMOVE ALL TASK 4 (whatever that is)
+#REMOVE ALL TASK 4
 # M2A: This appears to be a second pass over productions that were deemed to
 # be words; the elicitation here is just a translation of what word the
 # production was heard to be
@@ -290,14 +290,20 @@ for(i in 1:dim(correspondances)[1]) { #transform into unicharacter
 head(trnsc[,c("mispronunciation","mp_uni","target_uni")])
 
 
-#calculate Levinshtein's distances
+#calculate Levenshtein's distances
 trnsc$ins=trnsc$del=trnsc$sub=trnsc$nchar=trnsc$trafos=trnsc$tarlen=NA
-substitution_bank=deletion_bank=match_bank=substitution_bank_try1=deletion_bank_try1=match_bank_try1=NULL
+substitution_bank=insertion_bank=deletion_bank=match_bank=NULL
 
 for(i in 1:dim(trnsc)[1]) if (!trnsc$english[i]){ #this goes through each line
-  #lev #i=2 ; i = 4 ; i =48 (has deletions)
-  #2021-09-02 flipped the order, so that deletions surface as deletions rather than insertions!
-  lev=adist(trnsc[i,c("target_uni")],trnsc[i,c("mp_uni")],count=T)
+  #lev #i=2 ; i = 4 ; i =48 (has deletions) ; i = 1207 (has insertions)
+  #2021-09-03 major changes to mapping, because it was messed up in the presence of insertions
+  #I know it's confusing to call them source/target but I've double checked, and insertions contains insertions, deletions deletions
+  source_string=trnsc[i,c("target_uni")]
+
+  #use the target nonword's pronunciation if there was no error
+  if(is.na(trnsc[i,c("mp_uni")])) target_string=source_string else target_string=trnsc[i,c("mp_uni")]
+
+  lev=adist(source_string,target_string,count=T)
   trnsc$ins[i]<-attributes(lev)$counts[,,"ins"]
   trnsc$del[i]<-attributes(lev)$counts[,,"del"]
   trnsc$sub[i]<-attributes(lev)$counts[,,"sub"]
[…]

### 2021-09-03 manuscript.Rmd
starting with r3, knitting with png broken

@@
 The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region.
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phoneme inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD].
 Thus, with respect to our first research goal, Yélî Dnye is a good language to use because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
@@
 ggsave("nwr.by.freq.ITEM.png", dpi = 300, device = png())
-
 knitr::include_graphics("nwr.by.freq.ITEM.png")
@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible relationship between typological phone frequency and NWR scores.
+We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible correlation between typological phone frequency and NWR scores.
 An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the Discussion below).
@@
 
-We looked at two things: Associations between NWR and cross-linguistic frequency, and qualitative mispronunciation patterns. RESTRUCTURE TO MAKE THIS CLEARER!
-
 Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
-Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
-It would stand to reason that typologically common sounds are associated with higher performance, but to our knowledge this has not yet been tested with NWR.
-There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
-One could then predict that this effect should be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
-
-And yet, we found a robust effect of average segmental cross-linguistic frequency on NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
+As explained in the Introduction, typological frequency of phones could reflect ease of perception, ease of production, and other factors, and these factors could affect speech processing and production.
+This predicts a correlation between typological frequency and NWR performance, due to those factors affecting both.
+To assess this prediction, we looked at our data in two ways.
+First, we measured the degree of association between NWR scores and cross-linguistic frequency at the level of non-word items.
+Second, we described mispronunciation patterns, by looking at correct and incorrect repetitions of simpler and more complex sounds, which are also more or less frequent.
+
+There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phoneme inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
+One could then predict that correlations with typological frequency should be relatively weak because the ambient language puts more pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences than what is required of child speakers of other languages.
+
+We do not have the necessary data to assess whether the effect is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
 This effect was large, with a magnitude more than twice the size of the effect of participant age.
-Moreover, this significant effect remained even once also accounting for the frequencies of these segments in Yélî Dnye children's input.
-An analysis of the substitutions made by children also aligned with this interpretation, with more common sounds being substituted for less common ones.
+Moreover, this significant effect remained even  after accounting for the frequencies of these segments in a conversational corpus.
+An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.
 
@@
 We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [most often discussed in the case of phonotactic constraints on sequences, e.g., @maddieson2009phonology], but we do point out that this effect is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR.
-We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the fine-grained phonetic representations in the perception-production loop [as in e.g., @edwards2004interaction].
+We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop [as in e.g., @edwards2004interaction].
 
@@
 
-We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item.
+We investigated the effect of item complexity on NWR scores by varying  the number of syllables in the item.
 In broad terms, children should have higher NWR scores for shorter items.
@@
 In fact, maternal education correlated with NWR score at about r~.1, which is small.
-Similarly, NWR scores may not vary greatly with participant sex according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.
-
-Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
+We find effects of about that size for participant sex, which is aligned with previous work [@chiat2007preschool].
+
+Finally, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
@@
 
-The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, as when designing this study we wondered whether NWR was a *fair* test of phonological development.
+The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, since when designing this study we wondered whether NWR was a *fair* test of phonological development.
 Although our data cannot answer this question because we have only sampled one language and population here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
 It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
-Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages [@frank2017wordbank].
-
-At first sight, the range of performance we observed overlapped with previously observed levels of performance.
+Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, similarly to other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages [@frank2017wordbank].
+
+The range of performance we observed overlapped with previously observed levels of performance.
 Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
@@
 
-To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
-We entered data from XX studies (including ours), presenting data from XX languages.
-Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
-Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
-Results separating different age groups are shown in Figure \@ref(fig:fig-prevlit-overall).
-
-FIGURE REMOVED
-
-Several observations can be drawn from this figure.
-To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
-These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects).
-This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
-Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
-In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
-@cristia2019child estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
-Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.
-a female adult that was not temporally close to speech by others would count as child-directed).
-A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
-
-That said, @cristia2020infant also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
-This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
-If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
-Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall).
-These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
-What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
-
-Although exciting, this hypothesis is only one of many.
-Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
-@cristia2020infant administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
-This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
-Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
-
-Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies.
-We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
-We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
-Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall).
-It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.
-Sotho non-words having 4-7 syllables in length.
-To be more certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
+Comparison across published studies is difficult (see SM2 for an attempt).
+To be  certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
 We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+A discussion of this topic, and a proposal of items, can be found in @chiat2015non.
 An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
@@
 While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly predict phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
 Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including:  Why do NWR scores would pattern differently across samples?

### 2021-09-03 manuscript.tex
starting with r3, knitting with png broken

@@
 The Yélî sound system, much like its baroque grammatical system (Levinson, 2021), is unlike any other in the region.
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2021; Maddieson \& Levinson, in preparation).
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phoneme inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2021; Maddieson \& Levinson, in preparation).
 Thus, with respect to our first research goal, Yélî Dnye is a good language to use because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
@@
 
-We looked at two things: Associations between NWR and cross-linguistic frequency, and qualitative mispronunciation patterns. RESTRUCTURE TO MAKE THIS CLEARER!
-
 Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
-Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
-It would stand to reason that typologically common sounds are associated with higher performance, but to our knowledge this has not yet been tested with NWR.
-There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
-One could then predict that this effect should be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
-
-And yet, we found a robust effect of average segmental cross-linguistic frequency on NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
+As explained in the Introduction, typological frequency of phones could reflect ease of perception, ease of production, and other factors, and these factors could affect speech processing and production.
+This predicts a correlation between typological frequency and NWR performance, due to those factors affecting both.
+To assess this prediction, we looked at our data in two ways.
+First, we measured the degree of association between NWR scores and cross-linguistic frequency at the level of non-word items.
+Second, we described mispronunciation patterns, by looking at correct and incorrect repetitions of simpler and more complex sounds, which are also more or less frequent.
+
+There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phoneme inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
+One could then predict that correlations with typological frequency should be relatively weak because the ambient language puts more pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences than what is required of child speakers of other languages.
+
+We do not have the necessary data to assess whether the effect is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
 This effect was large, with a magnitude more than twice the size of the effect of participant age.
-Moreover, this significant effect remained even once also accounting for the frequencies of these segments in Yélî Dnye children's input.
-An analysis of the substitutions made by children also aligned with this interpretation, with more common sounds being substituted for less common ones.
+Moreover, this significant effect remained even after accounting for the frequencies of these segments in a conversational corpus.
+An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.
 
@@
 We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (most often discussed in the case of phonotactic constraints on sequences, e.g., Maddieson, 2009), but we do point out that this effect is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
-We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the fine-grained phonetic representations in the perception-production loop (as in e.g., Edwards, Beckman, \& Munson, 2004).
+We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop (as in e.g., Edwards, Beckman, \& Munson, 2004).
 
@@
 
-We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item.
+We investigated the effect of item complexity on NWR scores by varying the number of syllables in the item.
 In broad terms, children should have higher NWR scores for shorter items.
@@
 In fact, maternal education correlated with NWR score at about r\textasciitilde.1, which is small.
-Similarly, NWR scores may not vary greatly with participant sex according to previous work (Chiat \& Roy, 2007), and for that as well we find effects of about that size.
-
-Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
+We find effects of about that size for participant sex, which is aligned with previous work (Chiat \& Roy, 2007).
+
+Finally, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
@@
 
-The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, as when designing this study we wondered whether NWR was a \emph{fair} test of phonological development.
+The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, since when designing this study we wondered whether NWR was a \emph{fair} test of phonological development.
 Although our data cannot answer this question because we have only sampled one language and population here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
 It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
-Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages (Frank, Braginsky, Yurovsky, \& Marchman, 2017).
-
-At first sight, the range of performance we observed overlapped with previously observed levels of performance.
+Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, similarly to other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages (Frank, Braginsky, Yurovsky, \& Marchman, 2017).
+
+The range of performance we observed overlapped with previously observed levels of performance.
 Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
@@
 
-To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
-We entered data from XX studies (including ours), presenting data from XX languages.
-Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes, Wong, Fletcher, and Leonard (2006); English by Vance, Stackhouse, and Wells (2005); Italian by Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019); Mandarin by Lei et al.
-(2011); Persian by Farmani et al.
-(2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares, Marshall, and Griffiths (2016); Swedish by Kalnak, Peyrard-Janvid, Forssberg, and Sahlén (2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020); and Yélî Dnye from the present study.
-Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
-Results separating different age groups are shown in Figure \ref{fig:fig-prevlit-overall}.
-
-FIGURE REMOVED
-
-Several observations can be drawn from this figure.
-To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
-These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure \ref{fig:fig-prevlit}, larger length effects).
-This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
-Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
-In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: Casillas, Brown, and Levinson (2020) hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
-Cristia, Dupoux, Gurven, and Stieglitz (2019) estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
-Finally, Scaff, Stieglitz, Casillas, and Cristia (2021) used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.~a female adult that was not temporally close to speech by others would count as child-directed).
-A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
-
-That said, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
-This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
-If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
-Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \ref{fig:fig-prevlit-overall}.
-These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
-What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related to literacy \emph{in the broader population} (rather than in the tested children themselves).
-
-Although exciting, this hypothesis is only one of many.
-Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
-Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
-This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
-Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
-
-Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies.
-We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
-We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
-Neither of these two predictors explained variance in Figure \ref{fig:fig-prevlit-overall}.
-It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.~Sotho non-words having 4-7 syllables in length.
-To be more certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
+Comparison across published studies is difficult (see SM2 for an attempt).
+To be certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
 We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+A discussion of this topic, and a proposal of items, can be found in (\textbf{chiat2015non?}).
 An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
@@
 While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly predict phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
 Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including: Why do NWR scores would pattern differently across samples?
@@
 
-\leavevmode\hypertarget{ref-cristia2019child}{}%
-Cristia, A., Dupoux, E., Gurven, M., \& Stieglitz, J.
-(2019).
-Child-directed speech is infrequent in a forager-farmer population.
-\emph{{Child Development}}, \emph{90}, 759--773.
-\url{https://doi.org/10.1111/cdev.12974}
-
 \leavevmode\hypertarget{ref-cristia2020infant}{}%
@@
 
-\leavevmode\hypertarget{ref-kapalkova2013non}{}%
-Kapalková, S., Polišenská, K., \& Vicenová, Z.
-(2013).
-{Non-word repetition performance in Slovak-speaking children with and without SLI: novel scoring methods}.
-\emph{{International Journal of Language and Communication Disorders}}, \emph{48}(1), 78--89.
-\url{https://doi.org/10.1111/j.1460-6984.2012.00189.x}
-
 \leavevmode\hypertarget{ref-kolinsky2018completely}{}%
@@
 
-\leavevmode\hypertarget{ref-lei2011developmental}{}%
-Lei, L., Pan, J., Liu, H., McBride-Chang, C., Li, H., Zhang, Y., \ldots{} others.
-(2011).
-Developmental trajectories of reading development and impairment from ages 3 to 8 years in chinese children.
-\emph{Journal of Child Psychology and Psychiatry}, \emph{52}(2), 212--220.
-
 \leavevmode\hypertarget{ref-levinsonYDgrammar}{}%
@@
 \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
-
-\leavevmode\hypertarget{ref-polivsenska2014improving}{}%
-Polišenská, K., \& Kapalková, S. (2014). Improving child compliance on a computer-administered nonword repetition task. \emph{{Journal of Speech, Language and Hearing Research}}, \emph{57}(3).
-
-\leavevmode\hypertarget{ref-radeborg2006swedish}{}%
-Radeborg, K., Barthelom, E., SjöBerg, M., \& Sahlén, B.
-(2006).
-{A Swedish non-word repetition test for preschool children}.
-\emph{{Scandinavian Journal of Psychology}}, \emph{47}(3), 187--192.
-\url{https://doi.org/10.1111/j.1467-9450.2006.00506.x}
 

### 2021-09-03 supmat.Rmd
starting with r3, knitting with png broken

@@
 ```
+
+## TODO
+We have included information about the stimulus length in the caption of the figures.
 
@@
 
+## SM2: Integrating data from other studies
+
+To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
+We entered data from XX studies (including ours), presenting data from XX languages.
+Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
+Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
+Results separating different age groups are shown in Figure \@ref(fig:fig-prevlit-overall).
+
+FIGURE REMOVED
+
+Several observations can be drawn from this figure.
+To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
+These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects).
+This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
+Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
+In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
+@cristia2019child estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
+Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.
+a female adult that was not temporally close to speech by others would count as child-directed).
+A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
+
+That said, @cristia2020infant also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
+This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
+If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
+Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall).
+These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
+What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
+
+Although exciting, this hypothesis is only one of many.
+Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
+@cristia2020infant administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
+This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
+Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
+
+Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies.
+We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
+We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
+Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall).
+It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.
+Sotho non-words having 4-7 syllables in length.
+
 ```{r fig-prevlit-overall, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages.

### 2021-09-06 manuscript.Rmd
r3 in progress

@@
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].
+NWR tasks probably tap many skills [for relevant discussion see @coady2008uses; @dos2020epreuve].
 Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory [see a discussion in @chiat2015nonword].
@@
 Regarding child sex, no such effect has been found in previous NWR research [@chiat2007preschool], and in other language tasks evidence is mixed.
-Finally, prior research typically finds no significant differences as a function of  maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio].
-In general, maternal education correlates with child language outcomes, including vocabulary reports [@frank2017wordbank] and word comprehension studies [@scaff2019].
+Finally, prior research varies on whether significant differences as a function of  maternal education are reported [e.g., no difference found in @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio; whereas differences were reported in @tuller2018identifying; @dos2020epreuve].
+In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports [@frank2017wordbank] and word comprehension studies [@scaff2019].
 The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children [see discussion in @cristia2020infant].
@@
 
-Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).
-
-Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
-This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, which involved English participants.
+Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).^[Please note that the linguistic and cultural effects discussed here are different from the effects discussed in the extensive literature on NWR by bilingual participants.
+In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment.
+To try to measure language abilities above and beyond relative levels of experience with a given language, authors have tried to build non-words that tap language-dependent or language-independent knowledge.
+For instance, @tuller2018identifying employed a set of non-words judged to be language independent and two others that were more aligned with either French or German.
+The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language ("language-dependent") than when they are "language-independent".
+To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience.
+Preliminary results of an ongoing meta-analysis suggest significant associations between exposure to a given language and performance in both language-dependent and language-independent NWR [@farabolini2021association].
+In any case, this line of research focuses on links between exposure to a given language and NWR performance.
+In contrast, when we discuss linguistic or cultural effects, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).]
+
+Regarding potential language differences, we note that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/).
+It would be interesting for future researchers to consider straying from the literature by varying other dimensions that are relevant to the language under study.
+For instance, for Yélî Dnye, it is relevant to vary phonological complexity of the individual sounds because of its large inventory.
 

### 2021-09-13 manuscript.Rmd
in progress

@@
 Second, we documented rather weak effects of item length.
-Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
+Third, we found that age has a strong correlation with NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
 Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.
@@
 1. Sounds (and sound sequences) that are harder to perceive tend to be misperceived and thus lost diachronically
-2. Sounds (and sound sequences) that are harder to prononunce tend to be mispronounced and thus lost diachronically
+2. Sounds (and sound sequences) that are harder to pronounce tend to be mispronounced and thus lost diachronically
 3. Sound sequences that are harder to hold in memory tend to be mispronounced and thus lost diachronically
@@
 
-#### Individual variation effects on NWR
+#### Individual variation correlations with NWR
 
@@
 Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword].
-In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
+In a nutshell, most evidence is mixed, suggesting that correlations with individual variation may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
@@
 Our focus on age stems from previous work, where performance increases with child age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech].
-Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.
+Although past research has not investigated potential correlations with birth order on NWR, there is a sizable literature on these correlations in other language tasks [e.g., @havron2019effect], and therefore we report on these too.
 Common explanations for advantages for first- over later-born children include differential allocation of familial resources, particularly parental behaviors of cognitive stimulation [@lehmann2018early].
-Regarding child sex, no such effect has been found in previous NWR research [@chiat2007preschool], and in other language tasks evidence is mixed.
+Regarding child sex, no significant correlation has been found in previous NWR research [@chiat2007preschool], and in other language tasks evidence is mixed.
 Finally, prior research varies on whether significant differences as a function of  maternal education are reported [e.g., no difference found in @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio; whereas differences were reported in @tuller2018identifying; @dos2020epreuve].
@@
 
-Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).^[Please note that the linguistic and cultural effects discussed here are different from the effects discussed in the extensive literature on NWR by bilingual participants.
+Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic differences (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural differences (i.e., potential differences in NWR as a function of other differences across human populations).^[Please note that the linguistic and cultural differences discussed here are different from the differences discussed in the extensive literature on NWR by bilingual participants.
 In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment.
@@
 In any case, this line of research focuses on links between exposure to a given language and NWR performance.
-In contrast, when we discuss linguistic or cultural effects, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).]
+In contrast, when we discuss linguistic or cultural differences, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).]
 
@@
 Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation.
-Specifically, we investigated potential effects of age, child sex, maternal education, and birth order.
+Specifically, we investigated potential correlations with age, child sex, maternal education, and birth order.
 There is nothing particular to note regarding age and child sex, but we have some comments that pertain to the other two factors.
@@
 
-Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above.
-To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
+Some general ideas regarding potential correlations between our NWR measures and maternal education may be drawn from the observations above.
+To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken associations with maternal education.
 <!-- Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
@@
 
-As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology].
-It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
+As for birth order, much of the work on correlations between birth order and cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology].
+It is possible that birth order differences are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
 This contrasts with the picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
@@
 We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given the previously discussed variability across studies.
+However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known associations, given the previously discussed variability across studies.
 Therefore, all analyses in the present study are descriptive and should be considered exploratory.
@@
 We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-46 items: @jabere2018xperiment; @kalnak2014nonword;@piazzalunga2019articulatory], with the entire test inventory of 40 final items distributed across children.
-We  used a relatively large number of items to explore effects of length and phonological complexity.
+We  used a relatively large number of items to explore correlations with length and phonological complexity.
 However, aware that this large item inventory might render the task longer and more tiresome, we split items across children.
@@
 The percentage we observe here is much higher than reported in the study by Castro and colleagues, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
-Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this association.
 
@@
 Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = `r mycoeff["avg_fr_s:age.rounded_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s:age.rounded_s","Std.
-Error"]`, p `r pval.display(mycoeff["avg_fr_s:age.rounded_s","Pr(>|z|)"])`): The effect of frequency was larger for older children.
+Error"]`, p `r pval.display(mycoeff["avg_fr_s:age.rounded_s","Pr(>|z|)"])`): The effect of frequency was larger for younger children.
 Inspection of Figure \@ref(fig:fig-xling-freq-age3) suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.
@@
 
-### Individual variation effects on NWR
+### Individual variation and NWR
 
@@
 
-We do not have the necessary data to assess whether the effect is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
+We do not have the necessary data to assess whether the correlation is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
 This effect was large, with a magnitude more than twice the size of the effect of participant age.
-Moreover, this significant effect remained even  after accounting for the frequencies of these segments in a conversational corpus.
+Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at younger ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for younger than older children).
+Importantly, the correlation between performance and typological frequencyis  remained significant  after accounting for the frequencies of these segments in a conversational corpus.
 An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.
@@
 We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development.
-We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [most often discussed in the case of phonotactic constraints on sequences, e.g., @maddieson2009phonology], but we do point out that this effect is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR.
+We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [most often discussed in the case of phonotactic constraints on sequences, e.g., @maddieson2009phonology], but we do point out that this association is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR.
 We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop [as in e.g., @edwards2004interaction].
@@
 
-In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores.
-We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
+In contrast, previous work varies with respect to correlations of NWR scores with maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent].
+We did not expect large correlations with maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
 In fact, maternal education correlated with NWR score at about r~.1, which is small.
-We find effects of about that size for participant sex, which is aligned with previous work [@chiat2007preschool].
-
-Finally, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
+We find correlations of about that size for participant sex, which is aligned with previous work [@chiat2007preschool].
+
+Finally, we investigated whether birth order might correlate with NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
-Therefore we expected that any effects of birth order on NWR would be attenuated in this context.
+Therefore we expected that any correlations with birth order on NWR would be attenuated in this context.
 In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score.
 However, the effect size was larger than that found for the other two factors and it is far from negligible, at r~.2 or Cohen's d~`r r2d(.2)`.
-In fact, two large studies with therefore precise estimates found effects of about d~.2 [@havron2019effect;@barclay2015within], which would suggest the effects we found are larger.
+In fact, two large studies with therefore precise estimates found effects of about d~.2 [@havron2019effect;@barclay2015within], which would suggest the correlations we found are larger.
 We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.

### 2021-09-13 manuscript.Rmd
minor

@@
 A discussion of this topic, and a proposal of items, can be found in @chiat2015non.
-An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
-As for different strengths of age effects, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
+An alternative may be to find ways to regress out some of these differences, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
+As for the variable strengths of age correlations, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
 
@@
 While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly predict phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that correlations with age and typological frequency may strongly predict phonological development across these diverse settings, while correlations with item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
 Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including:  Why do NWR scores would pattern differently across samples?

### 2021-09-15 manuscript.Rmd
end r3

@@
 
-As for birth order, much of the work on correlations between birth order and cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology].
-It is possible that birth order differences are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
-This contrasts with the picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
-
-<!-- ac2mc shouldn't we totally rewrite this? -->
+As for birth order, much of the work on correlations between birth order and cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families were more likely to be the prevalent rearing environment [@lancy2015anthropology].
+It is possible that birth order differences are stronger in such a setting, because much of the stimulation can only come from the parents.
+These effects may be much smaller in cultures where it is common for children to attend daycare at an early age (such as France) or where extended family typically live close by.
+The Yélî community falls in the latter type, as children are typically surrounded by siblings and cousins of several orders, regardless of their birth order in their nuclear family.
+
+<!-- ac2mc shouldn't we totally rewrite this? // actually, this is all factually correct-->
 We add some observations that will help us integrate this study into the broader investigation of NWR across cultures.
@@
 
-We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts so as to base our generalizations on more data.
-There were no cases of insertion, and deletions were very rare: there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
+We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts and not excluding errors resulting in real words, so as to base our generalizations on more data.
+Insertions and deletions were very rare (metathesis was not attested): there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
 We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the  majority of incorrect phones (~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] )*100,2)` of unmatched phones).
@@
 Similarly, our items varied systematically in length and typological frequency of the sounds included, but not in other potential dimensions (such as whether the items contained morphemes of the language or not).
+
+We relied on a single resource, PHOIBLE, for our estimation of typological frequency, and some readers may be worried about the effects of this choice.
+As far as we know, PHOIBLE is the most extensive archive of phonological inventories, so it is a reasonable choice in the current context.
+However, one may want to calculate typological frequency not by trying to have as many languages represented as possible, but rather by selecting a sample of typologically independent languages.
+In addition, it is not the case that all the world's languages are represented, and indeed some of the Yélî sounds were not found there.
+While these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.
+At present, we do not see how imbalance in the typological selection and missing data can conspire to produce the correlation we observe.
+If anything, these factors should increase  noise in the typological frequency estimation, in which case the correlation size we uncover is an underestimation of the true correlation.
+
 Additionally, we only had a single person interacting with children as well as interpreting children's production, so we do not know to what extent our findings generalize to other experimenters and research assistants.
-In addition, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
+Furthermore, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
 Although we know from other work that adults' perceptual performance on these types of sounds is well above chance from audio-only presentation [REF],  language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.

### 2021-09-16 manuscript.Rmd
minor changes to discussion

@@
 count_syll[count_syll==0]<-1
-round(table(count_syll)/length(count_syll)*100)
+#round(table(count_syll)/length(count_syll)*100)
 
@@
 These effects may be much smaller in cultures where it is common for children to attend daycare at an early age (such as France) or where extended family typically live close by.
-The Yélî community falls in the latter type, as children are typically surrounded by siblings and cousins of several orders, regardless of their birth order in their nuclear family.
+The Yélî community falls in the latter case, as children are typically surrounded by siblings and cousins of several orders, regardless of their birth order in their nuclear family.
 
@@
 
-[^2]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.
+[^2]: We asked for mothers' highest completed level of education. We then recorded the number of years entailed by having completed that level under ideal conditions.
 
@@
 Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.
-
-<!-- a2m I thought in the first couple of years they combined English and Yélî at school? -->
-<!-- m2a: It used to be!
-In Anna, Ghaalyu, and Taakêmê's time it *was* but policies have shifted a bit since then.
-These days it looks like there's room for things to swing back toward bilingual early education.
-But anyhow for the kids in the current dataset we can pretty reliably say that they're language of schooling was English from the start.-->
 
@@
 Setting aside our monosyllabic stimuli (which contained typologically infrequent segments with lower NWR scores, as just discussed), we examined effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
-The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant, and is small and inconsistent across ages (see Figure \@ref(fig:fig-prevlit)).
+The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
 We do not have a good explanation for why samples in the literature vary so much in terms of the size of length effects, but two possibilities are that this is not truly a length effect but a confound with some other aspect of the stimuli, or that there is variation in phonological representations that is poorly understood.
@@
 For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants, which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of sub-sequences in the non-words.
-Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (elatedly) their n-phone frequency is overall lower.
+Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower.
 If this idea is correct, a careful analysis of non-words used in previous work may reveal that studies with larger length effects just happened to have longer non-words with lower n-phone frequencies.
@@
 Thus, we hazard the hypothesis that sample differences in length effects may relate to such non-essential skills.
-Since as stated this hypothesis is under-specified, further both conceptual and empirical work are needed.
+Since as stated this hypothesis is under-specified, further conceptual and empirical work is needed.
 
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006nonword; but note  @cristia2020infant is an exception].
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several previous studies [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006nonword; but not in  @cristia2020infant].
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-Finally, we investigated whether birth order might correlate with NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
+Finally, we investigated whether birth order might correlate with NWR scores, as it does with other language tasks, such that first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
@@
 In fact, two large studies with therefore precise estimates found effects of about d~.2 [@havron2019effect;@barclay2015within], which would suggest the correlations we found are larger.
-We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
+We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed does not result more even language outcomes for first- and later-born children.
 
@@
 
-The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, since when designing this study we wondered whether NWR was a *fair* test of phonological development.
+The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, since when designing this study we wondered whether NWR was a *culture-fair* test of phonological development.
 Although our data cannot answer this question because we have only sampled one language and population here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
@@
 We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
-A discussion of this topic, and a proposal of items, can be found in @chiat2015non.
 An alternative may be to find ways to regress out some of these differences, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
-As for the variable strengths of age correlations, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
+Both of these issues are discussed in @chiat2015nonword.
+As for the variable strengths of age correlations discussed above, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
 
@@
 Before closing, we would like to point out some salient limitations of the current work.
-To begin with, we only employed one set of non-words, in which not all characteristics that previous work suggest matter were manipulated [@chiat2015non].
+To begin with, we only employed one set of non-words, in which not all characteristics that previous work suggest matter were manipulated [@chiat2015nonword].
 As a result, we only have a rather whole-sale measure of performance, and we do not know to what extent lexical knowledge, pure phonological knowledge, and working memory, among others, contribute to children's performance.
@@
 However, one may want to calculate typological frequency not by trying to have as many languages represented as possible, but rather by selecting a sample of typologically independent languages.
-In addition, it is not the case that all the world's languages are represented, and indeed some of the Yélî sounds were not found there.
+In addition, it is not the case that all the world's languages are represented, and indeed some of the Yélî sounds were not found in PHOIBLE.
 While these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.
@@
 
-While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that correlations with age and typological frequency may strongly predict phonological development across these diverse settings, while correlations with item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
-Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including:  Why do NWR scores would pattern differently across samples?
+The present study shows that NWR be adapted for very different populations than have previously been tested.
+In addition, we observed strong correlations with age and typological frequency, while correlations with item length, participant sex, maternal education, and birth order were weaker.
+A consideration of previous work led us to suggest that the statistical strength of all of these effects may vary depending on the linguistic, cultural, and socio-demographic properties of the population under study, in conjunction with characteristics of the non-word items used.
+The present findings raise many questions, including:  Why do NWR scores pattern differently across samples?
 What does that tell us about the relationship between lexical development, phonological development, and the input environment?
 What is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders?
-While answers to these questions are sought, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
+While answers to these questions should be sought in future work, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
 

### 2021-09-16 manuscript.tex
minor changes to discussion

@@
 Second, we documented rather weak effects of item length.
-Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
+Third, we found that age has a strong correlation with NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
 Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations, and the extent to which they shed light on the factors universally structuring variation in phonological development at a global and individual level.
@@
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children (Armon-Lotem, Jong, \& Meir, 2015; Chiat, 2015; COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
+NWR tasks probably tap many skills (for relevant discussion see Coady \& Evans, 2008; Santos, Frau, Labrevoit, \& Zebib, 2020).
 Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory (see a discussion in Chiat, 2015).
@@
 \item
-  Sounds (and sound sequences) that are harder to prononunce tend to be mispronounced and thus lost diachronically
+  Sounds (and sound sequences) that are harder to pronounce tend to be mispronounced and thus lost diachronically
 \item
@@
 
-\hypertarget{individual-variation-effects-on-nwr}{%
-\paragraph{Individual variation effects on NWR}\label{individual-variation-effects-on-nwr}}
+\hypertarget{individual-variation-correlations-with-nwr}{%
+\paragraph{Individual variation correlations with NWR}\label{individual-variation-correlations-with-nwr}}
 
@@
 Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
-In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
+In a nutshell, most evidence is mixed, suggesting that correlations with individual variation may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
@@
 Our focus on age stems from previous work, where performance increases with child age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
-Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.
+Although past research has not investigated potential correlations with birth order on NWR, there is a sizable literature on these correlations in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.
 Common explanations for advantages for first- over later-born children include differential allocation of familial resources, particularly parental behaviors of cognitive stimulation (Lehmann, Nuevo-Chiquero, \& Vidal-Fernandez, 2018).
-Regarding child sex, no such effect has been found in previous NWR research (Chiat \& Roy, 2007), and in other language tasks evidence is mixed.
-Finally, prior research typically finds no significant differences as a function of maternal education (Balladares, Marshall, \& Griffiths, 2016; e.g., Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017).
-In general, maternal education correlates with child language outcomes, including vocabulary reports (Frank, Braginsky, Yurovsky, \& Marchman, 2017) and word comprehension studies (Scaff, 2019).
+Regarding child sex, no significant correlation has been found in previous NWR research (Chiat \& Roy, 2007), and in other language tasks evidence is mixed.
+Finally, prior research varies on whether significant differences as a function of maternal education are reported (Balladares, Marshall, \& Griffiths, 2016; e.g., no difference found in Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017; Santos, Frau, Labrevoit, \& Zebib, 2020; whereas differences were reported in Tuller et al., 2018).
+In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports (Frank, Braginsky, Yurovsky, \& Marchman, 2017) and word comprehension studies (Scaff, 2019).
 The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children (see discussion in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
@@
 
-Aside from Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)'s hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).
-
-Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
-This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, which involved English participants.
+Aside from Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)'s hypotheses just mentioned, we have found little discussion of linguistic differences (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural differences (i.e., potential differences in NWR as a function of other differences across human populations).\footnote{Please note that the linguistic and cultural differences discussed here are different from the differences discussed in the extensive literature on NWR by bilingual participants.
+In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment.
+To try to measure language abilities above and beyond relative levels of experience with a given language, authors have tried to build non-words that tap language-dependent or language-independent knowledge.
+For instance, Tuller et al.
+(2018) employed a set of non-words judged to be language independent and two others that were more aligned with either French or German.
+The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language (``language-dependent'') than when they are ``language-independent.'' To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience.
+Preliminary results of an ongoing meta-analysis suggest significant associations between exposure to a given language and performance in both language-dependent and language-independent NWR (Farabolini, Taboh, Ceravolo, \& Guerra, 2021).
+In any case, this line of research focuses on links between exposure to a given language and NWR performance.
+In contrast, when we discuss linguistic or cultural differences, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).}
+
+Regarding potential language differences, we note that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/).
+It would be interesting for future researchers to consider straying from the literature by varying other dimensions that are relevant to the language under study.
+For instance, for Yélî Dnye, it is relevant to vary phonological complexity of the individual sounds because of its large inventory.
 
@@
 
-\begin{verbatim}
-## count_syll
-##  1  2  3  4  5  6  8 
-## 15 39 29 12  2  2  1
-\end{verbatim}
-
 Our second research goal is to measure the effect of non-word length on NWR, which may need to be interpreted taking into account typical word length in the language.
@@
 
-Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above.
-To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
+Some general ideas regarding potential correlations between our NWR measures and maternal education may be drawn from the observations above.
+To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken associations with maternal education.
 In addition, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply be due to random factors like living close to a school or having relatives there.
 
-As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings (Barclay, 2015; Grätz, 2018), where nuclear families are more likely to be the prevalent rearing environment (Lancy, 2015).
-It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
-This contrasts with the picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
+As for birth order, much of the work on correlations between birth order and cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings (Barclay, 2015; Grätz, 2018), where nuclear families were more likely to be the prevalent rearing environment (Lancy, 2015).
+It is possible that birth order differences are stronger in such a setting, because much of the stimulation can only come from the parents.
+These effects may be much smaller in cultures where it is common for children to attend daycare at an early age (such as France) or where extended family typically live close by.
+The Yélî community falls in the latter case, as children are typically surrounded by siblings and cousins of several orders, regardless of their birth order in their nuclear family.
 
@@
 Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
-We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 children only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+We then recorded the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 children only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
 While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, in press).
@@
 We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-46 items: Jaber-Awida, 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019), with the entire test inventory of 40 final items distributed across children.
-We used a relatively large number of items to explore effects of length and phonological complexity.
+We used a relatively large number of items to explore correlations with length and phonological complexity.
 However, aware that this large item inventory might render the task longer and more tiresome, we split items across children.
@@
 
-Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for older children.
+Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for younger children.
 Inspection of Figure \ref{fig:fig-xling-freq-age3} suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.
@@
 
-We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts so as to base our generalizations on more data.
-There were no cases of insertion, and deletions were very rare: there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets).
+We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts and not excluding errors resulting in real words, so as to base our generalizations on more data.
+Insertions and deletions were very rare (metathesis was not attested): there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets).
 We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde{} of unmatched phones).
@@
 
-\hypertarget{individual-variation-effects-on-nwr-1}{%
-\subsubsection{Individual variation effects on NWR}\label{individual-variation-effects-on-nwr-1}}
+\hypertarget{individual-variation-and-nwr}{%
+\subsubsection{Individual variation and NWR}\label{individual-variation-and-nwr}}
 
@@
 
-We do not have the necessary data to assess whether the effect is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
+We do not have the necessary data to assess whether the correlation is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
 This effect was large, with a magnitude more than twice the size of the effect of participant age.
-Moreover, this significant effect remained even after accounting for the frequencies of these segments in a conversational corpus.
+Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at younger ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for younger than older children).
+Importantly, the correlation between performance and typological frequencyis remained significant after accounting for the frequencies of these segments in a conversational corpus.
 An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.
@@
 We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development.
-We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (most often discussed in the case of phonotactic constraints on sequences, e.g., Maddieson, 2009), but we do point out that this effect is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
+We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (most often discussed in the case of phonotactic constraints on sequences, e.g., Maddieson, 2009), but we do point out that this association is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
 We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop (as in e.g., Edwards, Beckman, \& Munson, 2004).
@@
 Setting aside our monosyllabic stimuli (which contained typologically infrequent segments with lower NWR scores, as just discussed), we examined effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
-The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant, and is small and inconsistent across ages (see Figure \ref{fig:fig-prevlit}).
+The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
 We do not have a good explanation for why samples in the literature vary so much in terms of the size of length effects, but two possibilities are that this is not truly a length effect but a confound with some other aspect of the stimuli, or that there is variation in phonological representations that is poorly understood.
@@
 Thus, we hazard the hypothesis that sample differences in length effects may relate to such non-essential skills.
-Since as stated this hypothesis is under-specified, further both conceptual and empirical work are needed.
-
-\hypertarget{individual-variation-effects-on-nwr-2}{%
-\subsubsection{Individual variation effects on NWR}\label{individual-variation-effects-on-nwr-2}}
+Since as stated this hypothesis is under-specified, further conceptual and empirical work is needed.
+
+\hypertarget{individual-variation-and-nwr-1}{%
+\subsubsection{Individual variation and NWR}\label{individual-variation-and-nwr-1}}
 
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance, Stackhouse, \& Wells, 2005; Italian Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019; Cantonese Stokes, Wong, Fletcher, \& Leonard, 2006; but note Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020 is an exception).
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several previous studies (English Vance, Stackhouse, \& Wells, 2005; Italian Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019; Cantonese Stokes, Wong, Fletcher, \& Leonard, 2006; but not in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017) on NWR scores.
-We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
+In contrast, previous work varies with respect to correlations of NWR scores with maternal education (e.g., Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017).
+We did not expect large correlations with maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
 In fact, maternal education correlated with NWR score at about r\textasciitilde.1, which is small.
-We find effects of about that size for participant sex, which is aligned with previous work (Chiat \& Roy, 2007).
-
-Finally, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
+We find correlations of about that size for participant sex, which is aligned with previous work (Chiat \& Roy, 2007).
+
+Finally, we investigated whether birth order might correlate with NWR scores, as it does with other language tasks, such that first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
-Therefore we expected that any effects of birth order on NWR would be attenuated in this context.
+Therefore we expected that any correlations with birth order on NWR would be attenuated in this context.
 In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score.
 However, the effect size was larger than that found for the other two factors and it is far from negligible, at r\textasciitilde.2 or Cohen's d\textasciitilde0.41.
-In fact, two large studies with therefore precise estimates found effects of about d\textasciitilde.2 (Barclay, 2015; Havron et al., 2019), which would suggest the effects we found are larger.
-We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
+In fact, two large studies with therefore precise estimates found effects of about d\textasciitilde.2 (Barclay, 2015; Havron et al., 2019), which would suggest the correlations we found are larger.
+We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed does not result more even language outcomes for first- and later-born children.
 
@@
 
-The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, since when designing this study we wondered whether NWR was a \emph{fair} test of phonological development.
+The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, since when designing this study we wondered whether NWR was a \emph{culture-fair} test of phonological development.
 Although our data cannot answer this question because we have only sampled one language and population here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
@@
 We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
-A discussion of this topic, and a proposal of items, can be found in (\textbf{chiat2015non?}).
-An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
-As for different strengths of age effects, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
+An alternative may be to find ways to regress out some of these differences, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
+Both of these issues are discussed in Chiat (2015).
+As for the variable strengths of age correlations discussed above, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
 
@@
 Before closing, we would like to point out some salient limitations of the current work.
-To begin with, we only employed one set of non-words, in which not all characteristics that previous work suggest matter were manipulated (\textbf{chiat2015non?}).
+To begin with, we only employed one set of non-words, in which not all characteristics that previous work suggest matter were manipulated (Chiat, 2015).
 As a result, we only have a rather whole-sale measure of performance, and we do not know to what extent lexical knowledge, pure phonological knowledge, and working memory, among others, contribute to children's performance.
 Similarly, our items varied systematically in length and typological frequency of the sounds included, but not in other potential dimensions (such as whether the items contained morphemes of the language or not).
+
+We relied on a single resource, PHOIBLE, for our estimation of typological frequency, and some readers may be worried about the effects of this choice.
+As far as we know, PHOIBLE is the most extensive archive of phonological inventories, so it is a reasonable choice in the current context.
+However, one may want to calculate typological frequency not by trying to have as many languages represented as possible, but rather by selecting a sample of typologically independent languages.
+In addition, it is not the case that all the world's languages are represented, and indeed some of the Yélî sounds were not found in PHOIBLE.
+While these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.
+At present, we do not see how imbalance in the typological selection and missing data can conspire to produce the correlation we observe.
+If anything, these factors should increase noise in the typological frequency estimation, in which case the correlation size we uncover is an underestimation of the true correlation.
+
 Additionally, we only had a single person interacting with children as well as interpreting children's production, so we do not know to what extent our findings generalize to other experimenters and research assistants.
-In addition, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
+Furthermore, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
 Although we know from other work that adults' perceptual performance on these types of sounds is well above chance from audio-only presentation {[}REF{]}, language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.
@@
 
-While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
-The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly predict phonological development across these diverse settings, while effects of item length, participant sex, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
-Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including: Why do NWR scores would pattern differently across samples?
+The present study shows that NWR be adapted for very different populations than have previously been tested.
+In addition, we observed strong correlations with age and typological frequency, while correlations with item length, participant sex, maternal education, and birth order were weaker.
+A consideration of previous work led us to suggest that the statistical strength of all of these effects may vary depending on the linguistic, cultural, and socio-demographic properties of the population under study, in conjunction with characteristics of the non-word items used.
+The present findings raise many questions, including: Why do NWR scores pattern differently across samples?
 What does that tell us about the relationship between lexical development, phonological development, and the input environment?
 What is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders?
-While answers to these questions are sought, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
+While answers to these questions should be sought in future work, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
 
@@
 
+\leavevmode\hypertarget{ref-coady2008uses}{}%
+Coady, J.
+A., \& Evans, J.
+L.
+(2008).
+Uses and interpretations of non-word repetition tasks in children with and without specific language impairments (SLI).
+\emph{International Journal of Language \& Communication Disorders}, \emph{43}(1), 1--40.
+
 \leavevmode\hypertarget{ref-is08042009language}{}%
@@
 
+\leavevmode\hypertarget{ref-farabolini2021association}{}%
+Farabolini, G., Taboh, A., Ceravolo, M.
+G., \& Guerra, F.
+(2021).
+The association between language exposure and non-word repetition performance in bilingual children: A meta-analysis.
+\emph{Under Review}.
+
 \leavevmode\hypertarget{ref-farmani2018normalization}{}%
@@
 
+\leavevmode\hypertarget{ref-dos2020epreuve}{}%
+Santos, C.
+dos, Frau, S., Labrevoit, S., \& Zebib, R.
+(2020).
+L'{é}preuve de r{é}p{é}tition de non-mots LITMUS-NWR-FR {é}value-t-elle la phonologie?
+In \emph{SHS web of conferences} (Vol.
+78, p.
+10005).
+EDP Sciences.
+
 \leavevmode\hypertarget{ref-scaff2019}{}%
@@
 
+\leavevmode\hypertarget{ref-tuller2018identifying}{}%
+Tuller, L., Hamann, C., Chilla, S., Ferré, S., Morin, E., Prevost, P., \ldots{} Zebib, R.
+(2018).
+Identifying language impairment in bilingual children in france and in germany.
+\emph{International Journal of Language \& Communication Disorders}, \emph{53}(4), 888--904.
+
 \leavevmode\hypertarget{ref-vance2005speech}{}%

### 2021-09-16 manuscript.Rmd
sup mat & minor changes to paper

@@
 
-```{r fig-xling-freq-age1, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (median split on scaled age)."}
-plot_model(xl_fr,type="int")
-```
-
-```{r fig-xling-freq-age2, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (mean, or plus/minus one standard deviation)."}
-
-plot_model(xl_fr,type="pred",terms=c("avg_fr_s","age.rounded_s"))
-```
+<!-- ac2mc if it's easy for you to add dots representing actual children in the above, that would be awesome! If not, I can try to tackle it :) -->
 

### 2021-09-16 manuscript.tex
sup mat & minor changes to paper

@@
 
-\begin{figure}[!t]
-
-{\centering \includegraphics[width=0.65\linewidth]{manuscript_files/figure-latex/fig-xling-freq-age1-1}
-
-}
-
-\caption{(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (median split on scaled age).}\label{fig:fig-xling-freq-age1}
-\end{figure}
-
-\begin{figure}[!t]
-
-{\centering \includegraphics[width=0.65\linewidth]{manuscript_files/figure-latex/fig-xling-freq-age2-1}
-
-}
-
-\caption{(WE WILL PROBABLY NOT USE THIS ONE) Model predicting NWR scores from the average frequency with which each phone is found across languages (x axis) and child age (mean, or plus/minus one standard deviation).}\label{fig:fig-xling-freq-age2}
-\end{figure}
-
 \hypertarget{nwr-and-typology-nwr-as-a-function-of-within-language-phone-frequency}{%

### 2021-09-16 supmat.Rmd
end r3
sup mat & minor changes to paper

@@
 ---
-title: "Supplementary materials"
-output: pdf_document
+title             : "Supplementary materials to: Non-word repetition in children learning Yélî Dnye"
+shorttitle        : "SM: NWR in Yélî Dnye learners"
+
+author: 
+  - name          : "Alejandrina Cristia"
+    affiliation   : "1"
+    corresponding : yes    # Define only one corresponding author
+    address       : "29, rue d’Ulm, 75005 Paris, France"
+    email         : "alecristia@gmail.com"
+  - name          : "Marisa Casillas"
+    affiliation   : "2,3"
+    address       : "Wundtlaan 1, 6525 XD, Nijmegen, The Netherlands"
+    email         : "marisa.casillas@mpi.nl"
+
+affiliation:
+  - id            : "1"
+    institution   : "Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University"
+  - id            : "2"
+    institution   : "Max Planck Institute for Psycholinguistics"
+  - id            : "3"
+    institution   : "University of Chicago"
+
+author_note: >
+       Both authors contributed to study funding, design, data collection, annotation, analyses, writing.
+
+abstract: >
+   This is a supplementary material containing: 1.
+Explanation of a small-scale systematic review carried out to embed the main results in the NWR literature; 2.
+The use of this data to look at potential length effects on NWR as a function of language characteristics; 3.
+Length and age effects in this and previous work.
+
+bibliography      : ["nwrbib.bib"]
+
+figsintext        : yes
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+lineno            : no
+
+lang              : "english"
+class             : "doc"
+output            :
+  papaja::apa6_pdf
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
+library(ggplot2)
+library(dplyr)
+library(openxlsx)
 ```
 
-## TODO
-We have included information about the stimulus length in the caption of the figures.
+## SM1: Small systematic review of previous NWR work
 
-## SM1: Systematic review of previous work to check whether length there are typological differences in length effects on NWR
+For many of our research questions, we wanted to know about previous work.
+It is common in psychology to do this via an unsystematic review -- that is, by e.g.
+starting out with papers one knows and then reading other papers citing those, or by doing searches on google scholar or another search engine, or by reading someone else's review.
+It has been well-established that overviews based on such unsystematic searches do not generalize and have other flaws because they represent biased samples of the literature [@thomas2020impact] -- for instance, famous studies are more likely to be found, even though there are no quality differences between more versus less famous studies.
+
+So instead of relying on such an approach, we built on a systematic review that was carried out for a different paper: @cristia2020infant "combined our previous knowledge of the literature and systematic searching to yield a sample of 17 studies that we could interrogate further".
+Specifically, the main goal of that systematic review was to check whether "effects of infant-directed input quantities may have been reported.
+Input variability has not been studied, to our knowledge, but two potential proxies of input have: Monolingual status, and socio-economic status.
+We set aside the literature comparing monolinguals against non-monolinguals because we reasoned that some readers may argue this contrast does not only show effects of input differences, but also interference across the languages being learned" (Cristia, 2020, personal records).
+Discovery combined an initial list produced by Gianmatteo Farabolini (based on readings of literature on bilingual NWR mainly), with a scholar.google.com search carried out in incognito mode by Cristia, with keywords "non-word-repetition socio-economic-status".
+Studies finally included reported on children who were monolingual and between 3 and 7 years of age, and which used proportion of non-words repeated as the outcome metric.
+
+We revisited this selection of articles in order to extract data on actual performance, taking into account also the length of the non-words used in the study.
+From that selection, only 8 articles reported on actual performance in the text, figures, or tables, fortunately representing a range of languages: Persian by @farmani2018normalization; Israeli Arabic  [@jabere2018xperiment];  English  [@vance2005speech]; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Swedish by @kalnak2014nonword and @radeborg2006swedish.
+We extended that selection with 5 additional papers selected to increase the number of languages represented, with both languages sometimes described as having short words [Cantonese  @stokes2006nonword; Mandarin  @lei2011developmental]; and others described as having long words [Italian  @piazzalunga2019articulatory; Spanish @balladares2016socio].
+Finally, we also included Tsimane' data  [@cristia2020infant].
+
+## SM2: Assessing potential length effects
+
+```{r demo}
+# please note that you will not be able to do this because you don't have this version, that is not fully de-identified
+# read.csv("NWR-test-log_FINAL.csv",header=T,sep=";")->fulldemo
+# fulldemo=fulldemo[,-2] #remove children's names
+# fulldemo$dob=as.Date(as.character(fulldemo$DOB.ISO),format="%Y%m%d")
+# fulldemo$test=as.Date(as.character(fulldemo$Date),format="%Y%m%d")
+# fulldemo$age.rounded=as.numeric(difftime(fulldemo$test,fulldemo$dob,unit="weeks"))/52.25
+# fulldemo$Age=as.numeric(as.character(fulldemo$Age))
+# plot(fulldemo$Age~fulldemo$age.rounded)
+# write.csv(fulldemo,"nwr-demo.csv")
+
+# read.csv("nwr-demo.csv",header=T)->demo
+# demo=demo[,-c(20,21)] #remove DOBs 
+# write.csv(demo,"nwr-demo.csv")
+
+read.csv("nwr-demo.csv",header=T)->demo
+#colnames(demo)
+
+subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc
+#if age.rounded is missing, use reported age
+inc$age.rounded[is.na(inc$age.rounded)]<-inc$Age[is.na(inc$age.rounded)]
+#summary(inc)
+table(inc$birthOrder)->birthorder.tab
+
+```
+
+```{r readin}
+read.table("final_data.txt",header=T,encoding="UTF-8")->trnsc
+
+trnsc$Sex=factor(trnsc$Sex)
+trnsc$birthOrder=as.numeric(trnsc$birthOrder)
+
+# Include only included children in trnsc
+trnsc[which(trnsc$id %in% inc$ID),]->trnsc.inc
+
+```
+
+```{r exclude subsequent}
+trnsc.inc-> trnsc.inc_all #keep a copy for figure comparing first & subsequent
+
+trnsc.inc=trnsc.inc[trnsc.inc$attempt=="first",]
+
+```
 
@@
 
-ADD MORE INFO ABOUT HOW THESE PAPERS WERE DISCOVERED
+We looked at the subset of papers that reported NWR scores separately for different word lengths.
+These were: Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant].
 
- We do not think previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths, as described in Supplementary Materials.
-We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant].
 Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
@@
 Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%).
-
-```{r prep-prevlit, echo=F}
-pl<- read.csv("previous_lit.csv")
-
-pl_stacked=cbind(pl[,c("Study","Language","n","Age")],stack(pl[c("X1","X2","X3","X4")]),stack(pl[c("sd1","sd2","sd3","sd4")]))
-colnames(pl_stacked)[5:8]<-c("Median","Length","SD","sdtype")
-pl_stacked$Length=gsub("X","",pl_stacked$Length)
-
-#remove lines with NA median
-pl_stacked[!is.na(pl_stacked$Median),]->pl_stacked
-
-#add YD data
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$id,trnsc.inc$type,trnsc.inc$age.rounded),mean,na.rm=T)->m_ty_id_age #get means by child and length and age
-
-age_cuts=c(3,6,8,13)
-m_ty_id_age$age_g=cut(m_ty_id_age$Group.3,age_cuts) #apply cuts to define groups
-m_ty_id_age$age_gm<-ifelse(m_ty_id_age$age_g=="(3,6]",4.5,ifelse(m_ty_id_age$age_g=="(6,8]",7,10))
-
-aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),median,na.rm=T)->yd_res #get avg across children within age groups
-yd_res=cbind(yd_res,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),sd,na.rm=T)$x,aggregate(m_ty_id_age$x,by=list(m_ty_id_age$Group.2,m_ty_id_age$age_gm),length)$x) #add sd & ns
-
-colnames(yd_res)<-c("Length","Age","Median","SD","n")
-yd_res_toadd=cbind("This paper","Yélî Dnye",yd_res[,c("n","Age","Median","Length","SD")],NA)
-colnames(yd_res_toadd)<-c("Study","Language","n","Age","Median","Length","SD","sdtype")
-
-#combine
-pl_stacked=rbind(pl_stacked,yd_res_toadd)
-
-#add SE
-pl_stacked$se=pl_stacked$SD/sqrt(pl_stacked$n)
-
-#remove monosyllables
-pl_stacked=pl_stacked[pl_stacked$Length!=1,]
-
-pl_stacked$Language=factor(pl_stacked$Language)
-
-pl$dif<-pl$X2-pl$X3
-difs=aggregate(pl$dif,by=list(pl$Language),mean, na.rm=T)
-```
 
@@
 
-## SM2: Integrating data from other studies
+## SM3: Integrating our data with that from other studies
 
-To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
-We entered data from XX studies (including ours), presenting data from XX languages.
-Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
+For this analysis, we could include all studies that reported non-word repetition scores based on whole item scoring for at least some length, ideally separating children by age.
+Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Spanish by @balladares2016socio;  Tsimane' by @cristia2020infant; and Yélî Dnye from the present study.
 Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
 Results separating different age groups are shown in Figure \@ref(fig:fig-prevlit-overall).
-
-FIGURE REMOVED
-
-Several observations can be drawn from this figure.
-To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
-These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects).
-This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
-Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
-In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
-@cristia2019child estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
-Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.
-a female adult that was not temporally close to speech by others would count as child-directed).
-A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
-
-That said, @cristia2020infant also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
-This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
-If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
-Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall).
-These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
-What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
-
-Although exciting, this hypothesis is only one of many.
-Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
-@cristia2020infant administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
-This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
-Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
-
-Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies.
-We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
-We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
-Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall).
-It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.
-Sotho non-words having 4-7 syllables in length.
 
@@
 
-## SM2: Age effects on accuracy
+Several observations can be drawn from this figure.
+To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
+These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects).
+This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
+Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
+In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
+@cristia2019child estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
+Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.
+a female adult that was not temporally close to speech by others would count as child-directed).
+A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
 
-```{r fig-xage, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the child's age (averaged in years)."}
-aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,round(trnsc.inc$age.rounded)),mean)->xage
-colnames(xage)<-c("Target","Age","Word")
-#stims.ipa <- read.xlsx("stimuli.xlsx") %>%
-#  select(target, target_ortho, phono)
-xage <- merge(xage, stims.ipa, by.x = "Target", by.y = "target")
+That said, @cristia2020infant also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
+This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
+If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
+Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall).
+These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
+What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
 
-item.plot <- ggplot(data = xage, aes(x = Age, y = Word, label = phono)) +
-  geom_smooth(method = "lm", se = FALSE, color = "black") +
-  geom_text(family = "Doulos SIL") +
-  ylab("NWR scores (% words)") +
-  xlab("Children's age") +
-  theme_apa()
-ggsave("nwr.by.age.ITEM.png", dpi = 300, device = png())
+Although exciting, this hypothesis is only one of many.
+Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
+@cristia2020infant administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
+This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
+Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
 
-knitr::include_graphics("nwr.by.age.ITEM.png")
+Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies.
+We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
+We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
+Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall).
+It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.
+Sotho non-words having 4-7 syllables in length.
 
-```
+# References
+
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}

### 2021-09-16 wrangling.R

@@ -45,6 +45,9 @@ trnsc$tokid=paste(trnsc$item,trnsc$token)
 # 1411              wee     English: where
 
 trnsc$mispronunciation=as.character(trnsc$mispronunciation)
+trnsc$meaningful=F
+trnsc$meaningful[!is.na(trnsc$meaning)]<-T
+
 trnsc$meaning=as.character(trnsc$meaning)
 trnsc$mispronunciation[trnsc$mispronunciation=="two"]<-"tuu"
 trnsc[grep("Eng",trnsc$meaning),c("mispronunciation","meaning")]
@@ -318,26 +321,26 @@ for(i in 1:dim(trnsc)[1]) if (!trnsc$english[i]){ #this goes through each line
     if(j=="S") {
       substitution_bank=rbind(substitution_bank,
                               cbind(strsplit(source_string,"")[[1]][counter_source], strsplit(target_string,"")[[1]][counter_target],
-                                    trnsc[i,c("target_uni","mp_uni","age.rounded","target_ortho","phono","tokid")],row.names = NULL))
+                                    trnsc[i,c("target_uni","mp_uni","age.rounded","target_ortho","phono","tokid","meaningful")],row.names = NULL))
       counter_source=counter_source+1
       counter_target=counter_target+1
     }
     if(j=="I") {
       insertion_bank=rbind(insertion_bank,
                            cbind(strsplit(target_string,"")[[1]][counter_target],
-                           trnsc[i,c("target_uni","mp_uni","age.rounded","target_ortho","phono","tokid")],row.names = NULL))
+                           trnsc[i,c("target_uni","mp_uni","age.rounded","target_ortho","phono","tokid","meaningful")],row.names = NULL))
       counter_target=counter_target+1
     }
     if(j=="D") {
       deletion_bank=rbind(deletion_bank,
                           cbind(strsplit(source_string,"")[[1]][counter_source],
-                          trnsc[i,c("target_uni","mp_uni","age.rounded","target_ortho","phono","tokid")],row.names = NULL))
+                          trnsc[i,c("target_uni","mp_uni","age.rounded","target_ortho","phono","tokid","meaningful")],row.names = NULL))
       counter_source=counter_source+1
     }
     if(j=="M") {
       match_bank=rbind(match_bank,
                        cbind(strsplit(source_string,"")[[1]][counter_source],
[…]

### 2021-09-22 manuscript.Rmd
minor rewordings from gdoc rebuttal lettter suggestions

@@
 NWR tasks probably tap many skills [for relevant discussion see @coady2008uses; @dos2020epreuve].
-Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory [see a discussion in @chiat2015nonword].
+Non-words can be designed to try to isolate certain skills more narrowly; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory [see a discussion in @chiat2015nonword].
 Broadly, however, NWR scores will necessarily reflect to a certain extent phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
@@
 In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, @chiat2015non discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
-As a result, any one item created will be relatively easier if it more closely resembles real words in the language, making it difficult to balance difficulty when comparing children learning different languages.
+As a result, any one item created will be relatively easier if it more closely resembles real words in a language, making it difficult to balance difficulty when comparing children learning different languages.
 This previous literature also suggests some dimensions of difficulty -- an issue to which we return in the next subsection.

### 2021-09-27 manuscript.Rmd
add yu & ku waru refs + a few small other things

@@
 
-Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development].
+Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development; @rumsey2017dependency].
 Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
@@
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].
-NWR tasks probably tap many skills [for relevant discussion see @coady2008uses; @dos2020epreuve].
+NWR tasks probably tap into many skills [for relevant discussion see @coady2008uses; @dos2020epreuve].
 Non-words can be designed to try to isolate certain skills more narrowly; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory [see a discussion in @chiat2015nonword].
@@
 As a result, any one item created will be relatively easier if it more closely resembles real words in a language, making it difficult to balance difficulty when comparing children learning different languages.
-This previous literature also suggests some dimensions of difficulty -- an issue to which we return in the next subsection.
+This previous literature also suggests some dimensions of difficulty—an issue to which we return in the next subsection.
 
@@
 Some of them are cross-linguistically rare, in that they are less common across languages than other sounds or phonetic targets.
-Phonologists and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of acquisition, focusing largely on phonotactics [@moreton2012structure].
-The correlation between typological frequency and ease of acquisition is typically assumed to emerge from the following causal paths (which are not mutually exclusive):
+Phonologists, phoneticians, and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of processing and acquisition via diachronic language change.
+These works focus largely on phonotactics [@moreton2012structure] perceptual parsing of the (ambiguous) linguistic signal [@beddor2009coarticulatory; @ohala1981listener], and individual differences in processing styles [@bermudezotero2015amphichronic]; small but significant effects that may cumulatively drive language change via phonologization [see @yu2021toward for a recent review].
+Thus, the correlation between typological frequency and ease of acquisition is typically assumed to emerge from one or more of the following causal paths:
 
@@
 
-It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture -- a hypothesis discussed for instance in @chiat2015non (pp.
+It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture—a hypothesis discussed for instance in @chiat2015non (pp.
 7-8; see also p.
@@
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer.
-A non-systematic meta-analyses does not provide overwhelming support for this hypothesis [@cristia2021supplementary; SM1].
+A non-systematic meta-analysis does not provide overwhelming support for this hypothesis [@cristia2021supplementary; SM1].
 
@@
 In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading; for instance, better than rhyme awareness  [e.g., @gathercole1991differentiating].
-Therefore, it may not be a general index of, for instance, phonological development, but reflect only certain non-universal language skills.
+Therefore, it may not be a general index of phonological development, but instead reflect certain non-universal language skills.
 Indeed, @cristia2020infant present their task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
@@
 In any case, this line of research focuses on links between exposure to a given language and NWR performance.
-In contrast, when we discuss linguistic or cultural differences, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).]
+In contrast, when we discuss linguistic or cultural differences here, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).]
 
@@
 <!-- ac2mc shouldn't we totally rewrite this? // actually, this is all factually correct-->
+<!-- mc2ac: while factually correct i thought it might not be a great representation of where we stand now so i made some small edits in the final sentences-->
 We add some observations that will help us integrate this study into the broader investigation of NWR across cultures.
-As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant].
+As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors of that paper interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant].
 However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is also relatively infrequent in this community [@casillas2020early].
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcross], which shares other societal characteristics wih the Tsimane' [e.g., is rural and relies on farming, children grow up in wide familial networks, etc; @casillas2020early].
 Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early].
-Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in @cristia2020infant.
+Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in @cristia2020infant.
 
@@
 
-```{r}
-
-```
-
 ### Length effects on NWR
@@
 One could then predict that correlations with typological frequency should be relatively weak because the ambient language puts more pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences than what is required of child speakers of other languages.
+On the other hand, it is also possible that this pressure gives Yélî children no benefit, and that some of these categories are simply acquired later in development.
+We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not make adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) till age 5 or 6 [@rumsey2017dependency]
 
@@
 Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at younger ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for younger than older children).
-Importantly, the correlation between performance and typological frequencyis  remained significant  after accounting for the frequencies of these segments in a conversational corpus.
+Importantly, the correlation between performance and typological frequency  remained significant  after accounting for the frequencies of these segments in a conversational corpus.
 An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.
@@
 We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development.
-We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [most often discussed in the case of phonotactic constraints on sequences, e.g., @maddieson2009phonology], but we do point out that this association is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR.
+We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [via phonotactic constraints, ambiguous parsing conditions, individual differences, and more as in, e.g., @beddor2009coarticulatory; @bermudezotero2015amphichronic; @maddieson2009phonology; @ohala1981listener; @yu2021toward], but we do point out that this association is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR.
 We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop [as in e.g., @edwards2004interaction].
@@
 Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
-Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane'.
-
-Comparison across published studies is difficult (see SM2 for an attempt).
+Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane' [@cristia2020infant].
+
+Comparison across published studies is difficult (see SM2 for our preliminary attempt).
 To be  certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
-We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless, but phonotactically legal, in all of the languages.
 An alternative may be to find ways to regress out some of these differences, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
@@
 In addition, it is not the case that all the world's languages are represented, and indeed some of the Yélî sounds were not found in PHOIBLE.
-While these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.
+PHOIBLE—as well as our own work—depends on phonological descriptions from linguists who are in many cases not native speakers of the languages.
+Because the phones in our items have largely been evidenced as phonemic via multiple analyses [i.e., minimal contrast, phonological, phonetic, and ultrasound, see @levinsonYDgrammar], we are not concerned that changes to the phonological description in the future (e.g., if a segment loses its phonemic status) will significantly change the results presented here.
+While all these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.
 At present, we do not see how imbalance in the typological selection and missing data can conspire to produce the correlation we observe.

### 2021-10-12 manuscript.Rmd
response 15 changes

@@
 Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun.
-For example, Peute and colleagues' [-@peuteIPconsonants] find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
-We hope the present study contributes to this growing line of work.
+For example, Peute and Casillas [-@peuteIPconsonants] find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
+Other ongoing work on Yélî Dnye includes experiment-based infant phoneme discrimination data and errors made in elicited and spontaneous speech from young children, but these data are neither finalized nor yet externally reviewed [see @hellwigPAlanguage for more information].
+These data will help better inform our current analyses based on NWR in the future (e.g., regarding common sound substitutions) but are not critical for testing our current question about the general correlation between cross-linguistic phone frequency and NWR performance.
 

### 2021-10-12 manuscript.Rmd
common > uncommon

@@
 Error"]`, p `r pval.display(mycoeff["avg_fr_s:age.rounded_s","Pr(>|z|)"])`): The effect of frequency was larger for younger children.
-Inspection of Figure \@ref(fig:fig-xling-freq-age3) suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.
+Inspection of Figure \@ref(fig:fig-xling-freq-age3) suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically uncommon phones.
 

### 2021-10-12 manuscript.Rmd
fn renumbering + response 15 changes

@@
 These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
-Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^2] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` children only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^3]
-
-[^2]: We asked for mothers' highest completed level of education. We then recorded the number of years entailed by having completed that level under ideal conditions.
-
-[^3]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^1] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` children only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]
+
+[^1]: We asked for mothers' highest completed level of education. We then recorded the number of years entailed by having completed that level under ideal conditions.
+
+[^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
 While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing].
@@
 The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types.
-To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.[^1] Here, sounds from our non-word items that did not occur in the corpus were not considered (i.e., they were declared NA so that they do not count for analyses).
-
-[^1]: We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.
+To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.[^3] Here, sounds from our non-word items that did not occur in the corpus were not considered (i.e., they were declared NA so that they do not count for analyses).
+
+[^3]: We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.
 
@@
 To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
+[^4]
+
+[^4]: Note that a full anonymized, but age-annotated table of errors is provided in the project repository for those interested in a finer-grained analysis than what is presented here: ALEX?].
 
@@
 Because the phones in our items have largely been evidenced as phonemic via multiple analyses [i.e., minimal contrast, phonological, phonetic, and ultrasound, see @levinsonYDgrammar], we are not concerned that changes to the phonological description in the future (e.g., if a segment loses its phonemic status) will significantly change the results presented here.
+Relatedly, any converging evidence from the other ongoing studies of Yélî Dnye phonological development and fine-grained analyses of sound substitutions would certainly help bolster the claims we made here.
 While all these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.

### 2021-10-12 manuscript.Rmd
response 18 changes

@@
 What is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders?
-While answers to these questions should be sought in future work, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
+While answers to these questions should be sought in future work, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and much broader patterns of cross-linguistic phone frequency.
 

### 2021-10-12 manuscript.Rmd
response 21 changes

@@
 The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
-After completing all of her annotations, she double-checked them by listening to them twice.
+Because the phonological inventory is so acoustically packed and, based on audio data alone, it might be easy to mis-identify a segment, the assistant double-checked all of her annotations by listening to them and assessing them a second time, once she had completed a full first round.
 

### 2021-10-12 manuscript.Rmd
response 23 changes
response 24 changes

@@
 This was probably not the case for the majority of mothers of the children in our sample, who will have learned to read and write in Yélî Dnye during their first three years at school.
-This change from Yélî-initial instruction to English only occurred about 15 years ago.
 It is possible that there is also some home teaching of Yélî reading and writing, notably for reading the bible.
@@
 
-Additionally, we estimated frequency of the phones present  in the target items in a corpus of child-centered recordings [@casillas2020early].
-That corpus was constituted by sampling from audio-recordings (7-9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
-The researchers selected 9 2.5-minute clips randomly and 11 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity).
-These clips were  segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the people talking personally.
+Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings [@casillas2020early].
+That corpus was constituted by sampling from audio-recordings (7–9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
+The researchers selected 9 2.5-minute clips randomly and 11 1- or 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity).
+These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the recorded people personally.
 For more details, please refer to @casillas2020early.
@@
 For the present study, we extracted the transcriptions of adult speech (i.e., removing key child and other children's speech) and split them into words using white space.
-We then removed all English and Pidgin words.
+We then removed all English and Tok Pisin words.
 The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types.
-To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.[^3] Here, sounds from our non-word items that did not occur in the corpus were not considered (i.e., they were declared NA so that they do not count for analyses).
+To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.[^3] Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+Note that the resulting values estimate usage frequencies for very young children’s input and, while this is somewhat different from what our older participants experience on a daily basis, we can expect that this is a reasonable approximation of the early input that formed the foundation of their phonological knowledge.
 

### 2021-10-12 manuscript.Rmd
un-billeting the text + some bits and bobs

@@
 NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
-In this study, we report on NWR results among children (N= 40, aged 3-10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
+In this study, we report on NWR results among children (N = 40, aged 3--10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
 Results make three contributions that are specific, and a fourth that is general.
@@
 
-The typical household in our dataset includes seven individuals (typically, a mixed-sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+The typical household in our dataset includes seven individuals (typically, a mixed-sex couple and children---their own and possibly some others staying with them, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
 These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
@@
 Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available [@casillas2020early].
-Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
+Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to host their school-aged relatives during the weekdays for long segments of the school year.
 Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.
@@
 However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcross], which shares other societal characteristics wih the Tsimane' [e.g., is rural and relies on farming, children grow up in wide familial networks, etc; @casillas2020early].
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcross], and additionally shares other societal characteristics wih the Tsimane' [e.g., is rural and relies on farming, children grow up in wide familial networks, etc; @casillas2020early].
 Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early].
@@
 The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041), including the use of verbal (not written) consent.
-As discussed in subsection "The Yélî community", the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+As discussed in subsection "The Yélî community", the combination of collective child guardianship practices and common hosting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
 Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood [@brownIPchildrearing].
@@
 The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
-Because the phonological inventory is so acoustically packed and, based on audio data alone, it might be easy to mis-identify a segment, the assistant double-checked all of her annotations by listening to them and assessing them a second time, once she had completed a full first round.
+Because the phonological inventory is so acoustically packed and because annotation was done based on audio data alone, it might be easy to mis-identify a segment, the assistant double-checked all of her annotations by listening to them and assessing them a second time, once she had completed a full first round.
 

### 2021-10-12 manuscript.Rmd
response to in-text comment

@@
 <!-- Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
-ac2mc removed, R was confused and I'm also not sure what is the relevance of this comment -->
+ac2mc removed, R was confused and I'm also not sure what is the relevance of this comment
+
+mc2ac: fine with me, thanks!-->
 In addition, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply be due to random factors like living close to a school or having relatives there.

### 2021-10-13 manuscript.Rmd
add indiv scores to model pred plot

@@
 
-plot_model(xl_fr,type="pred",terms=c("age.rounded_s","avg_fr_s"))
-```
-
-<!-- ac2mc if it's easy for you to add dots representing actual children in the above, that would be awesome! If not, I can try to tackle it :) -->
+trnsc.inc.indivs <- trnsc.inc %>%
+  mutate(avg_fr_s = case_when(
+    avg_fr_s <= mean(xl_fr_tab$xlf)-sd(xl_fr_tab$xlf) ~ -1,
+    avg_fr_s >= mean(xl_fr_tab$xlf)+sd(xl_fr_tab$xlf) ~ 1,
+    TRUE ~ 0
+  )) %>%
+  group_by(file, age.rounded_s, avg_fr_s) %>%
+  summarize(correct = mean(correct))
+
+plot_model(xl_fr, type = "pred", terms = c("age.rounded_s","avg_fr_s")) +
+  geom_point(data = trnsc.inc.indivs, inherit.aes = FALSE,
+             aes(x = age.rounded_s, y = correct, color = factor(avg_fr_s))) +
+  scale_color_manual(values = c("red", "orange", "green"),
+                     labels = c("mean - 1sd", "mean", "mean + 1sd")) +
+  scale_fill_manual(values = c("red", "orange", "green"),
+                    guide = FALSE) +
+  labs(color = "Avgerage\nfrequency\nscore",
+       y = "% Correct", x = "Child age\n(standardized)") +
+  theme_apa() +
+  theme(plot.title = element_blank(),
+        legend.key = element_rect(color = "white")) +
+  guides(color = guide_legend(override.aes = list(fill = NA), reverse = TRUE))
+
+```
+
+<!-- ac2mc if it's easy for you to add dots representing actual children in the above, that would be awesome! If not, I can try to tackle it :)
+
+mc2ac here's my attempt! i wasn't sure how you wanted to aggregate the real data to correspond to the model estimates so i have here grouped correctness within individual by
+<= m - 1sd
+mean <±> 1sd (unconventional notation, sorry)
+>= m + 1sd
+
+if you had something else in mind but are otherwise cool with the figure, it should be easy to edit on the mutate/case_when() portion defining trnsc.inc.indivs
+-->
 

### 2021-10-13 manuscript.Rmd
make correct*freq across items plot show again

@@
 
-```{r fig-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.",eval=F}
+```{r fig-xling-freq, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores for individual target words as a function of the average frequency with which each phone is found across languages."}
 aggregate(trnsc.inc$correct*100,by=list(trnsc.inc$target,trnsc.inc$avg_fr),mean)->xling

### 2021-10-18 manuscript.Rmd
fix nasaliz. diacritics in Table 1 (NB: new input file)

@@
 #add orthographic representations targets & other stims chars
-stims.tbl <- read.xlsx("stimuli.xlsx") %>%
+stims.tbl <- read.xlsx("stimuli-20211018.xlsx") %>%
   arrange(factor(type, levels = c("practice", "1", "2", "3", "4"))) %>%

### 2021-10-18 manuscript.tex
fix nasaliz. diacritics in Table 1 (NB: new input file)

@@
 
-Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000).
+Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000; Rumsey, 2017).
 Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
@@
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children (Armon-Lotem, Jong, \& Meir, 2015; Chiat, 2015; COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-NWR tasks probably tap many skills (for relevant discussion see Coady \& Evans, 2008; Santos, Frau, Labrevoit, \& Zebib, 2020).
-Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory (see a discussion in Chiat, 2015).
+NWR tasks probably tap into many skills (for relevant discussion see Coady \& Evans, 2008; Santos, Frau, Labrevoit, \& Zebib, 2020).
+Non-words can be designed to try to isolate certain skills more narrowly; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory (see a discussion in Chiat, 2015).
 Broadly, however, NWR scores will necessarily reflect to a certain extent phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
@@
 In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, (\textbf{chiat2015non?}) discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
-As a result, any one item created will be relatively easier if it more closely resembles real words in the language, making it difficult to balance difficulty when comparing children learning different languages.
-This previous literature also suggests some dimensions of difficulty -- an issue to which we return in the next subsection.
+As a result, any one item created will be relatively easier if it more closely resembles real words in a language, making it difficult to balance difficulty when comparing children learning different languages.
+This previous literature also suggests some dimensions of difficulty---an issue to which we return in the next subsection.
 
@@
 Some of them are cross-linguistically rare, in that they are less common across languages than other sounds or phonetic targets.
-Phonologists and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of acquisition, focusing largely on phonotactics (Moreton \& Pater, 2012).
-The correlation between typological frequency and ease of acquisition is typically assumed to emerge from the following causal paths (which are not mutually exclusive):
+Phonologists, phoneticians, and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of processing and acquisition via diachronic language change.
+These works focus largely on phonotactics (Moreton \& Pater, 2012) perceptual parsing of the (ambiguous) linguistic signal (\textbf{beddor2009coarticulatory?}; \textbf{ohala1981listener?}), and individual differences in processing styles (\textbf{bermudezotero2015amphichronic?}); small but significant effects that may cumulatively drive language change via phonologization (see \textbf{yu2021toward?} for a recent review).
+Thus, the correlation between typological frequency and ease of acquisition is typically assumed to emerge from one or more of the following causal paths:
 
@@
 
-It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture -- a hypothesis discussed for instance in (\textbf{chiat2015non?}) (pp.~7-8; see also p.~5).
+It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture---a hypothesis discussed for instance in (\textbf{chiat2015non?}) (pp.~7-8; see also p.~5).
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer.
-A non-systematic meta-analyses does not provide overwhelming support for this hypothesis {[}Cristia and Casillas (2021); SM1{]}.
+A non-systematic meta-analysis does not provide overwhelming support for this hypothesis {[}Cristia and Casillas (2021); SM1{]}.
 
@@
 In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading; for instance, better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991).
-Therefore, it may not be a general index of, for instance, phonological development, but reflect only certain non-universal language skills.
+Therefore, it may not be a general index of phonological development, but instead reflect certain non-universal language skills.
 Indeed, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) present their task as being a good index of the development of ``short-hand-like'' representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
@@
 In any case, this line of research focuses on links between exposure to a given language and NWR performance.
-In contrast, when we discuss linguistic or cultural differences, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).}
+In contrast, when we discuss linguistic or cultural differences here, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).}
 
@@
 Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun.
-For example, Peute and colleagues' (In preparation) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
-We hope the present study contributes to this growing line of work.
+For example, Peute and Casillas (In preparation) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
+Other ongoing work on Yélî Dnye includes experiment-based infant phoneme discrimination data and errors made in elicited and spontaneous speech from young children, but these data are neither finalized nor yet externally reviewed (see Hellwig, Sarvasy, \& Casillas, provisionally accepted for more information).
+These data will help better inform our current analyses based on NWR in the future (e.g., regarding common sound substitutions) but are not critical for testing our current question about the general correlation between cross-linguistic phone frequency and NWR performance.
 
@@
 Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas, Brown, \& Levinson, 2020).
-Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
+Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to host their school-aged relatives during the weekdays for long segments of the school year.
 Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.
@@
 We add some observations that will help us integrate this study into the broader investigation of NWR across cultures.
-As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
+As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors of that paper interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
 However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is also relatively infrequent in this community (Casillas, Brown, \& Levinson, 2020).
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones \textbf{bunceURcross?}), and additionally shares other societal characteristics wih the Tsimane' {[}e.g., is rural and relies on farming, children grow up in wide familial networks, etc; Casillas, Brown, and Levinson (2020){]}.
 Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2020).
-Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).
+Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).
 
@@
 The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041), including the use of verbal (not written) consent.
-As discussed in subsection ``The Yélî community,'' the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+As discussed in subsection ``The Yélî community,'' the combination of collective child guardianship practices and common hosting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
 Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, in press).
@@
 \midrule
-nopimade & nɔpimæṭɛ & dp:a & ṭp̃æ & kamo & kæmɔ & dimope & ṭimɔpɛ & dipońate & ṭipɔnætɛ\\
+nopimade & nɔpimæṭɛ & dp:a & ṭpæ̃ & kamo & kæmɔ & dimope & ṭimɔpɛ & dipońate & ṭipɔnætɛ\\
 poni & pɔni & dpa & ṭpæ & kańi & kæni & diyeto & ṭijɛtɔ & ńomiwake & nɔmiwækɛ\\
@@
  &  & dpu & ṭpu & towi & tɔwi & ńuyedi & nujɛṭi &  & \\
- &  & gh:ââ & ɣ̃ɑː & tupa & tupæ & pedumi & pɛṭumi &  & \\
+ &  & gh:ââ & ɣɑ̃ː & tupa & tupæ & pedumi & pɛṭumi &  & \\
  &  & ghuu & ɣuː &  &  & tiwuńe & tiwunɛ &  & \\
- &  & kp:ââ & kp̃ɑː &  &  & tumowe & tumɔwɛ &  & \\
+ &  & kp:ââ & kpɑ̃ː &  &  & tumowe & tumɔwɛ &  & \\
 \addlinespace
  &  & kpu & kpu &  &  & widońe & wiṭɔnɛ &  & \\
- &  & lv:ê & lβʲ̃ə &  &  & wumipo & wumipɔ &  & \\
+ &  & lv:ê & lβʲə̃ &  &  & wumipo & wumipɔ &  & \\
  &  & lva & lβʲæ &  &  &  &  &  & \\
  &  & lvi & lβʲi &  &  &  &  &  & \\
- &  & t:êê & t̃əː &  &  &  &  &  & \\
+ &  & t:êê & tə̃ː &  &  &  &  &  & \\
 \addlinespace
@@
 
-Additionally, we estimated frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2020).
-That corpus was constituted by sampling from audio-recordings (7-9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
-The researchers selected 9 2.5-minute clips randomly and 11 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity).
-These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the people talking personally.
+Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2020).
+That corpus was constituted by sampling from audio-recordings (7--9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
+The researchers selected 9 2.5-minute clips randomly and 11 1- or 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity).
+These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the recorded people personally.
 For more details, please refer to Casillas, Brown, and Levinson (2020).
@@
 For the present study, we extracted the transcriptions of adult speech (i.e., removing key child and other children's speech) and split them into words using white space.
-We then removed all English and Pidgin words.
+We then removed all English and Tok Pisin words.
 The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types.
-To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, sounds from our non-word items that did not occur in the corpus were not considered (i.e., they were declared NA so that they do not count for analyses).
+To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
+Note that the resulting values estimate usage frequencies for very young children's input and, while this is somewhat different from what our older participants experience on a daily basis, we can expect that this is a reasonable approximation of the early input that formed the foundation of their phonological knowledge.
 
@@
 The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
-After completing all of her annotations, she double-checked them by listening to them twice.
+Because the phonological inventory is so acoustically packed and because annotation was done based on audio data alone, it might be easy to mis-identify a segment, the assistant double-checked all of her annotations by listening to them and assessing them a second time, once she had completed a full first round.
 
@@
 
+\begin{figure}[!t]
+
+{\centering \includegraphics[width=0.65\linewidth]{nwr.by.freq.ITEM}
+
+}
+
+\caption{NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}\label{fig:fig-xling-freq}
+\end{figure}
+
+\begin{verbatim}
+## pdf 
+##   2
+\end{verbatim}
+
 Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for younger children.
-Inspection of Figure \ref{fig:fig-xling-freq-age3} suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.
+Inspection of Figure \ref{fig:fig-xling-freq-age3} suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically uncommon phones.
 
@@
 We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde{} of unmatched phones).
-To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
+To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?\footnote{Note that a full anonymized, but age-annotated table of errors is provided in the project repository for those interested in a finer-grained analysis than what is presented here: ALEX?{]}.}
 
@@
 One could then predict that correlations with typological frequency should be relatively weak because the ambient language puts more pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences than what is required of child speakers of other languages.
+On the other hand, it is also possible that this pressure gives Yélî children no benefit, and that some of these categories are simply acquired later in development.
+We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not make adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) till age 5 or 6 (Rumsey, 2017)
 
@@
 Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at younger ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for younger than older children).
-Importantly, the correlation between performance and typological frequencyis remained significant after accounting for the frequencies of these segments in a conversational corpus.
+Importantly, the correlation between performance and typological frequency remained significant after accounting for the frequencies of these segments in a conversational corpus.
 An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.
@@
 We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development.
-We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (most often discussed in the case of phonotactic constraints on sequences, e.g., Maddieson, 2009), but we do point out that this association is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
+We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (Maddieson, 2009; via phonotactic constraints, ambiguous parsing conditions, individual differences, and more as in, e.g., \textbf{beddor2009coarticulatory?}; \textbf{bermudezotero2015amphichronic?}; \textbf{ohala1981listener?}; \textbf{yu2021toward?}), but we do point out that this association is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
 We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop (as in e.g., Edwards, Beckman, \& Munson, 2004).
@@
 Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
-Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane'.
-
-Comparison across published studies is difficult (see SM2 for an attempt).
+Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
+
+Comparison across published studies is difficult (see SM2 for our preliminary attempt).
 To be certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
-We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless, but phonotactically legal, in all of the languages.
 An alternative may be to find ways to regress out some of these differences, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
@@
 In addition, it is not the case that all the world's languages are represented, and indeed some of the Yélî sounds were not found in PHOIBLE.
-While these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.
+PHOIBLE---as well as our own work---depends on phonological descriptions from linguists who are in many cases not native speakers of the languages.
+Because the phones in our items have largely been evidenced as phonemic via multiple analyses (i.e., minimal contrast, phonological, phonetic, and ultrasound, see Levinson, 2021), we are not concerned that changes to the phonological description in the future (e.g., if a segment loses its phonemic status) will significantly change the results presented here.
+Relatedly, any converging evidence from the other ongoing studies of Yélî Dnye phonological development and fine-grained analyses of sound substitutions would certainly help bolster the claims we made here.
+While all these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.
 At present, we do not see how imbalance in the typological selection and missing data can conspire to produce the correlation we observe.
@@
 
+\leavevmode\hypertarget{ref-hellwigPAlanguage}{}%
+Hellwig, B., Sarvasy, H., \& Casillas, M.
+(provisionally accepted).
+Language acquisition.
+In N.
+Evans \& S.
+Fedden (Eds.), \emph{The {O}xford guide to {P}apuan languages} (pp.
+XX--XX).
+Oxford: Oxford University Press.
+
 \leavevmode\hypertarget{ref-jabere2018xperiment}{}%
@@
 \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
+
+\leavevmode\hypertarget{ref-rumsey2017dependency}{}%
+Rumsey, A.
+(2017).
+Dependency and relative determination in language acquisition: The case of {Ku Waru}.
+In N.
+J.
+Enfield (Ed.), \emph{Dependencies in language} (pp.
+97--116).
+Berlin: Language Science Press.
 

### 2021-10-18 manuscript.Rmd
read new stim input file for nwr.by.freq.ITEM fig

@@
 colnames(xling)<-c("Target","Avg_fr","Word")
-stims.ipa <- read.xlsx("stimuli.xlsx") %>%
+stims.ipa <- read.xlsx("stimuli-20211018.xlsx") %>%
   select(target, target_ortho, phono)

### 2021-10-18 manuscript.Rmd
add missing refs + re-knit

@@
 
-lang              : "english"
+lang              : "en-US"
 class             : "man"
@@
 NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity].
-NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@chiat2015non;@estes2007differences].
+NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@chiat2015nonword;@estes2007differences].
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing].
@@
 There has been an interest in adapting NWR to different languages, in part for applied purposes.
-In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, @chiat2015non discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
+In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, @chiat2015nonword discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
 As a result, any one item created will be relatively easier if it more closely resembles real words in a language, making it difficult to balance difficulty when comparing children learning different languages.
@@
 Although this cross-linguistic literature is rich, the potential difficulty associated with specific phonetic targets composing the non-words has received relatively little attention.
-For example, @chiat2015non discusses segmental complexity as a function of whether there are consonant clusters -- which is arguably a factor reflecting phonotactics and syllable structure.
+For example, @chiat2015nonword discusses segmental complexity as a function of whether there are consonant clusters -- which is arguably a factor reflecting phonotactics and syllable structure.
 
@@
 
-It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture—a hypothesis discussed for instance in @chiat2015non (pp.
+It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture—a hypothesis discussed for instance in @chiat2015nonword (pp.
 7-8; see also p.
@@
 However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcross], and additionally shares other societal characteristics wih the Tsimane' [e.g., is rural and relies on farming, children grow up in wide familial networks, etc; @casillas2020early].
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcrosscultural], and additionally shares other societal characteristics wih the Tsimane' [e.g., is rural and relies on farming, children grow up in wide familial networks, etc; @casillas2020early].
 Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early].

### 2021-10-18 manuscript.tex
add missing refs + re-knit

@@
 \documentclass[
-  english,
+  american,
   ,man,floatsintext]{apa6}
@@
   pdfauthor={Alejandrina Cristia1 \& Marisa Casillas2,3},
-  pdflang={en-EN},
+  pdflang={en-US},
   pdfkeywords={phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy},
@@
   \usepackage{polyglossia}
-  \setmainlanguage[]{english}
+  \setmainlanguage[variant=american]{english}
 \else
-  \usepackage[main=english]{babel}
+  \usepackage[main=american]{babel}
 % get rid of language-specific shorthands (see #6817):
@@
 NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
-NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007; \textbf{chiat2015non?}).
+NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Chiat, 2015; Estes, Evans, \& Else-Quest, 2007).
 Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children (Armon-Lotem, Jong, \& Meir, 2015; Chiat, 2015; COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
@@
 There has been an interest in adapting NWR to different languages, in part for applied purposes.
-In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, (\textbf{chiat2015non?}) discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
+In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, Chiat (2015) discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
 As a result, any one item created will be relatively easier if it more closely resembles real words in a language, making it difficult to balance difficulty when comparing children learning different languages.
@@
 Although this cross-linguistic literature is rich, the potential difficulty associated with specific phonetic targets composing the non-words has received relatively little attention.
-For example, (\textbf{chiat2015non?}) discusses segmental complexity as a function of whether there are consonant clusters -- which is arguably a factor reflecting phonotactics and syllable structure.
+For example, Chiat (2015) discusses segmental complexity as a function of whether there are consonant clusters -- which is arguably a factor reflecting phonotactics and syllable structure.
 
@@
 Phonologists, phoneticians, and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of processing and acquisition via diachronic language change.
-These works focus largely on phonotactics (Moreton \& Pater, 2012) perceptual parsing of the (ambiguous) linguistic signal (\textbf{beddor2009coarticulatory?}; \textbf{ohala1981listener?}), and individual differences in processing styles (\textbf{bermudezotero2015amphichronic?}); small but significant effects that may cumulatively drive language change via phonologization (see \textbf{yu2021toward?} for a recent review).
+These works focus largely on phonotactics (Moreton \& Pater, 2012) perceptual parsing of the (ambiguous) linguistic signal (Beddor, 2009; Ohala, 1981), and individual differences in processing styles (Bermúdez-Otero, 2015); small but significant effects that may cumulatively drive language change via phonologization (see Yu, 2021 for a recent review).
 Thus, the correlation between typological frequency and ease of acquisition is typically assumed to emerge from one or more of the following causal paths:
@@
 
-It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture---a hypothesis discussed for instance in (\textbf{chiat2015non?}) (pp.~7-8; see also p.~5).
+It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture---a hypothesis discussed for instance in Chiat (2015) (pp.~7-8; see also p.~5).
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer.
@@
 However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones \textbf{bunceURcross?}), and additionally shares other societal characteristics wih the Tsimane' {[}e.g., is rural and relies on farming, children grow up in wide familial networks, etc; Casillas, Brown, and Levinson (2020){]}.
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones Bunce et al., under review), and additionally shares other societal characteristics wih the Tsimane' {[}e.g., is rural and relies on farming, children grow up in wide familial networks, etc; Casillas, Brown, and Levinson (2020){]}.
 Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2020).
@@
 We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development.
-We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (Maddieson, 2009; via phonotactic constraints, ambiguous parsing conditions, individual differences, and more as in, e.g., \textbf{beddor2009coarticulatory?}; \textbf{bermudezotero2015amphichronic?}; \textbf{ohala1981listener?}; \textbf{yu2021toward?}), but we do point out that this association is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
+We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (via phonotactic constraints, ambiguous parsing conditions, individual differences, and more as in, e.g., Beddor, 2009; Bermúdez-Otero, 2015; Maddieson, 2009; Ohala, 1981; Yu, 2021), but we do point out that this association is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
 We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop (as in e.g., Edwards, Beckman, \& Munson, 2004).
@@
 Barclay, K. J. (2015). A within-family analysis of birth order and intelligence using population conscription data on swedish men. \emph{Intelligence}, \emph{49}, 134--143.
+
+\leavevmode\hypertarget{ref-beddor2009coarticulatory}{}%
+Beddor, P. S. (2009). A coarticulatory path to sound change. \emph{Language}, \emph{85}(4), 785--832.
+
+\leavevmode\hypertarget{ref-bermudezotero2015amphichronic}{}%
+Bermúdez-Otero, R.
+(2015).
+Amphichronic explanation and the life cycle of phonological processes.
+In P.
+Honeybone \& J.
+Salmons (Eds.), \emph{The {O}xford handbook of historical phonology} (pp.
+374--399).
+Oxford, UK: Oxford University Press.
 
@@
 
+\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
+Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural examination of young children's everyday language experiences.
+
 \leavevmode\hypertarget{ref-casillas2020early}{}%
@@
 
+\leavevmode\hypertarget{ref-ohala1981listener}{}%
+Ohala, J.
+J.
+(1981).
+The listener as a source of sound change.
+In M.
+F.
+Miller, C.
+S.
+Masek, \& R.
+A.
+Hendrick (Eds.), \emph{Papers from the parasession on language and behavior} (pp.
+178--203).
+Chicago, IL: Chicago Linguistics Society.
+
 \leavevmode\hypertarget{ref-peuteIPconsonants}{}%
@@
 
+\leavevmode\hypertarget{ref-yu2021toward}{}%
+Yu, A. C. L. (2021). Toward an individual-difference perspective on phonologization. \emph{Glossa: A Journal of General Linguistics}, \emph{6}(1), 1--24.
+
 \end{CSLReferences}

### 2021-10-18 supmat.Rmd
moderate rewording/reorganization

@@
 
-For many of our research questions, we wanted to know about previous work.
-It is common in psychology to do this via an unsystematic review -- that is, by e.g.
-starting out with papers one knows and then reading other papers citing those, or by doing searches on google scholar or another search engine, or by reading someone else's review.
-It has been well-established that overviews based on such unsystematic searches do not generalize and have other flaws because they represent biased samples of the literature [@thomas2020impact] -- for instance, famous studies are more likely to be found, even though there are no quality differences between more versus less famous studies.
+In order to embed our research questions in the context of previous work, we build our current study on a systematic review of NWR done for a different paper: @cristia2020infant.
+We preferred to build on a systematic review because, while they are common in psychology research, *unsystematic reviews*—e.g.
+starting out with familiar papers and then reading other papers citing those; searching via Google Scholar or another search engine; or reading someone else's review—do not generalize and represent biased samples of the literature [@thomas2020impact].
+For instance, famous studies are more likely to be found, even though there are no quality differences between more versus less famous studies.
 
-So instead of relying on such an approach, we built on a systematic review that was carried out for a different paper: @cristia2020infant "combined our previous knowledge of the literature and systematic searching to yield a sample of 17 studies that we could interrogate further".
+We here build on a recent and highly relevant systematic review: @cristia2020infant "combined our previous knowledge of the literature and systematic searching to yield a sample of 17 studies that we could interrogate further".
 Specifically, the main goal of that systematic review was to check whether "effects of infant-directed input quantities may have been reported.
@@
 From that selection, only 8 articles reported on actual performance in the text, figures, or tables, fortunately representing a range of languages: Persian by @farmani2018normalization; Israeli Arabic  [@jabere2018xperiment];  English  [@vance2005speech]; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Swedish by @kalnak2014nonword and @radeborg2006swedish.
-We extended that selection with 5 additional papers selected to increase the number of languages represented, with both languages sometimes described as having short words [Cantonese  @stokes2006nonword; Mandarin  @lei2011developmental]; and others described as having long words [Italian  @piazzalunga2019articulatory; Spanish @balladares2016socio].
-Finally, we also included Tsimane' data  [@cristia2020infant].
+We extended that selection with 5 additional papers selected to increase the number of languages represented, with both languages sometimes described as having short words [Cantonese: @stokes2006nonword; Mandarin:  @lei2011developmental]; and others described as having long words [Italian: @piazzalunga2019articulatory; Spanish: @balladares2016socio].
+Finally, we also included the Tsimane' data presented in that paper [@cristia2020infant].
 
@@
 
-Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
+Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected typical word length in different languages.
 For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%.
@@
 To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
-These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects).
-This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
-Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
-In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: @casillas2020early hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
-@cristia2019child estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
-Finally, @scaff2021daylong used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.
-a female adult that was not temporally close to speech by others would count as child-directed).
-A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
+These two groups have been described as having (very) roughly similar lifestyles and levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure  \@ref(fig:fig-prevlit), larger length effects).
+What can we conclude from this apparent difference?
 
-That said, @cristia2020infant also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
-This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
-If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
-Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall).
-These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
+One possible lens of interpretation relates NWR differences to the minor differences in children's directed linguistic input rates.
+However, the differences in methods used to estimate these rates render this first interpretation unreliable.
+Based on current estimates Tsimane' children may encounter somewhere between similar to somewhat lower rates of directed speech comared to Yélî children.
+The lower boundary for Tsimane' input rate is established by @cristia2019child, who used behavioral observations of child-inclusive one-on-one conversation.
+The upper boundary is set by @scaff2021daylong, who used human annotation to detect speech, feeding this into an automated temporal method for assigning speech as child-directed or not (NB: this method could lead to over- but not under-estimation because any nearby speech, e.g.
+from a female adult, that coincided with child vocalization would count as child-directed).
+Yélî Dnye estimates come from @casillas2021early, who hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, effectively establishing an upper boundary of all the speech children could potentially process.
+There is no equivalent 'lower bound' for Yélî Dnye to compare with the Tsimane' findings.
+So, while it is tempting to link this difference in NWR scores to directed input rates, we will first need to actually establish input rates  using comparable methods for Yéli and Tsimane' children.
+
+Prevalence of literacy is a second possible way to interpret these differences in NWR scores.
+@cristia2020infant point out the relatively low prevalence of literacy, and more generally, the variable access to formal education available in the Tsimane' population.
+This is very different from the present Yélî population, where most adults have accumulated some schooling, with many having basic to fluent literacy in English (note: literacy in Yélî Dnye is much less common, especially among the very young; see main text).
+Under this second hypothesis, there are phonetic effects of learning to read that have significant consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
+Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance illustrated, for instance, in the data for Sotho in Figure \@ref(fig:fig-prevlit-overall).
+Those two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
 What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
+However, there is at least one more plausible explanation: design and execution of the NWR task
 
-Although exciting, this hypothesis is only one of many.
-Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
+A third, plausible, explanation for these differences in NWR is that the Tsimane' results are not generally comparable to the previous body of literature, and specifically not comparable to our present study.
 @cristia2020infant administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
-This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
-Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
+This implies a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
+Thus, a priority is for additional data to be gathered using this more novel testing paradigm in other populations; or more from the Tsimane' population, but using the more traditional paradigm.
 
@@
 We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
-We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
-Neither of these two predictors explained variance in Figure \@ref(fig:fig-prevlit-overall).
-It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.
-Sotho non-words having 4-7 syllables in length.
+We extracted the estimated number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be relatively short or long, based on information in the papers cited above and other sources.
+Neither of these two predictors explained the variance in average NWR scores, illustrated in Figure \@ref(fig:fig-prevlit-overall).
+It is possible that average word length plays a role, but that researchers tend to incorporate this into their design by including longer items when the native language allows it, with e.g.
+Sotho non-words having 4--7 syllables in length.
 

### 2021-10-18 manuscript.Rmd
update casillas2020early to casillas2021early

@@
 
-Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun.
+Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2021early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun.
 For example, Peute and Casillas [-@peuteIPconsonants] find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
@@
 Within a group of households, it is often the case that older adolescents and adults spend their day tending to their farm plots (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
-Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early].
-Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available [@casillas2020early].
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2021early].
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available [@casillas2021early].
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to host their school-aged relatives during the weekdays for long segments of the school year.
@@
 However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcrosscultural], and additionally shares other societal characteristics wih the Tsimane' [e.g., is rural and relies on farming, children grow up in wide familial networks, etc; @casillas2020early].
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early].
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcrosscultural], and additionally shares other societal characteristics wih the Tsimane' [e.g., is rural and relies on farming, children grow up in wide familial networks, etc; @casillas2021early].
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2021early].
 Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in @cristia2020infant.
@@
 
-Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings [@casillas2020early].
+Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings [@casillas2021early].
 That corpus was constituted by sampling from audio-recordings (7–9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
@@
 These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the recorded people personally.
-For more details, please refer to @casillas2020early.
+For more details, please refer to @casillas2021early.
 

### 2021-10-18 manuscript.tex
re-knit

@@
 
-Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun.
+Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2021; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun.
 For example, Peute and Casillas (In preparation) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
@@
 Within a group of households, it is often the case that older adolescents and adults spend their day tending to their farm plots (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
-Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde10 members) independent and mixed-age child play groups (Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020).
-Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas, Brown, \& Levinson, 2020).
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde10 members) independent and mixed-age child play groups (Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2021).
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas, Brown, \& Levinson, 2021).
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to host their school-aged relatives during the weekdays for long segments of the school year.
@@
 However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones Bunce et al., under review), and additionally shares other societal characteristics wih the Tsimane' {[}e.g., is rural and relies on farming, children grow up in wide familial networks, etc; Casillas, Brown, and Levinson (2020){]}.
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2020).
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones Bunce et al., under review), and additionally shares other societal characteristics wih the Tsimane' {[}e.g., is rural and relies on farming, children grow up in wide familial networks, etc; Casillas, Brown, and Levinson (2021){]}.
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2021).
 Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).
@@
 
-Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2020).
+Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2021).
 That corpus was constituted by sampling from audio-recordings (7--9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
@@
 These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the recorded people personally.
-For more details, please refer to Casillas, Brown, and Levinson (2020).
+For more details, please refer to Casillas, Brown, and Levinson (2021).
 
@@
 
-\leavevmode\hypertarget{ref-casillas2020early}{}%
-Casillas, M., Brown, P., \& Levinson, S. C. (2020). {Early language experience in a Papuan community}. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
+\leavevmode\hypertarget{ref-casillas2021early}{}%
+Casillas, M., Brown, P., \& Levinson, S. C. (2021). {Early language experience in a Papuan community}. \emph{Journal of Child Language}, \emph{48}(4), 792--814.
 

### 2021-10-18 supmat.tex
re-knit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  american,
  ,doc,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Supplementary materials to: Non-word repetition in children learning Yélî Dnye},
  pdfauthor={Alejandrina Cristia1 \& Marisa Casillas2,3},
  pdflang={en-US},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{SM: NWR in Yélî Dnye learners}
\usepackage{csquotes}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[variant=american]{english}
\else
  \usepackage[main=american]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Supplementary materials to: Non-word repetition in children learning Yélî Dnye}
\author{Alejandrina Cristia\textsuperscript{1} \& Marisa Casillas\textsuperscript{2,3}}
\date{}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}

\abstract{
This is a supplementary material containing: 1. Explanation of a small-scale systematic review carried out to embed the main results in the NWR literature; 2. The use of this data to look at potential length effects on NWR as a function of language characteristics; 3. Length and age effects in this and previous work.
}

\begin{document}
\maketitle

\hypertarget{sm1-small-systematic-review-of-previous-nwr-work}{%
\subsection{SM1: Small systematic review of previous NWR work}\label{sm1-small-systematic-review-of-previous-nwr-work}}

In order to embed our research questions in the context of previous work, we build our current study on a systematic review of NWR done for a different paper: Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020). We preferred to build on a systematic review because, while they are common in psychology research, \emph{unsystematic reviews}---e.g.~starting out with familiar papers and then reading other papers citing those; searching via Google Scholar or another search engine; or reading someone else's review---do not generalize and represent biased samples of the literature (Thomas-Odenthal, Molero, Does, \& Molendijk, 2020). For instance, famous studies are more likely to be found, even though there are no quality differences between more versus less famous studies.

We here build on a recent and highly relevant systematic review: Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) ``combined our previous knowledge of the literature and systematic searching to yield a sample of 17 studies that we could interrogate further.'' Specifically, the main goal of that systematic review was to check whether ``effects of infant-directed input quantities may have been reported. Input variability has not been studied, to our knowledge, but two potential proxies of input have: Monolingual status, and socio-economic status. We set aside the literature comparing monolinguals against non-monolinguals because we reasoned that some readers may argue this contrast does not only show effects of input differences, but also interference across the languages being learned'' (Cristia, 2020, personal records). Discovery combined an initial list produced by Gianmatteo Farabolini (based on readings of literature on bilingual NWR mainly), with a scholar.google.com search carried out in incognito mode by Cristia, with keywords ``non-word-repetition socio-economic-status.'' Studies finally included reported on children who were monolingual and between 3 and 7 years of age, and which used proportion of non-words repeated as the outcome metric.

We revisited this selection of articles in order to extract data on actual performance, taking into account also the length of the non-words used in the study. From that selection, only 8 articles reported on actual performance in the text, figures, or tables, fortunately representing a range of languages: Persian by Farmani et al. (2018); Israeli Arabic (Jaber-Awida, 2018); English (Vance, Stackhouse, \& Wells, 2005); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Swedish by Kalnak, Peyrard-Janvid, Forssberg, and Sahlén (2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006). We extended that selection with 5 additional papers selected to increase the number of languages represented, with both languages sometimes described as having short words (Cantonese: Stokes, Wong, Fletcher, \& Leonard, 2006; Mandarin: Lei et al., 2011); and others described as having long words (Italian: Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019; Spanish: Balladares, Marshall, \& Griffiths, 2016). Finally, we also included the Tsimane' data presented in that paper (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).

\hypertarget{sm2-assessing-potential-length-effects}{%
\subsection{SM2: Assessing potential length effects}\label{sm2-assessing-potential-length-effects}}

We looked at the subset of papers that reported NWR scores separately for different word lengths. These were: Israeli Arabic (Jaber-Awida, 2018); Cantonese (Stokes, Wong, Fletcher, \& Leonard, 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).

Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected typical word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (\textasciitilde28\%) and Arabic (\textasciitilde15\%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only \textasciitilde2\%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese \textasciitilde8\% versus English \textasciitilde1\%).

\begin{figure}
\centering
\includegraphics{supmat_files/figure-latex/fig-prevlit-1.pdf}
\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables, 2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al.~(2019) reported on groups of 24-60 Italian learners (black); Stokes et al.~(2006) on 15 Cantonese learners (blue); Vance et al.~(2005) on 17-20 English learners (light green); Cristia et al.~(2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error. Age has been slightly shifted for ease of inspection of different lengths at a given age.}
\end{figure}

\hypertarget{sm3-integrating-our-data-with-that-from-other-studies}{%
\subsection{SM3: Integrating our data with that from other studies}\label{sm3-integrating-our-data-with-that-from-other-studies}}

For this analysis, we could include all studies that reported non-word repetition scores based on whole item scoring for at least some length, ideally separating children by age. Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes, Wong, Fletcher, and Leonard (2006); English by Vance, Stackhouse, and Wells (2005); Italian by Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019); Mandarin by Lei et al. (2011); Spanish by Balladares, Marshall, and Griffiths (2016); Tsimane' by Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020); and Yélî Dnye from the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in Figure \ref{fig:fig-prevlit-overall}.

\begin{figure}
\centering
\includegraphics{supmat_files/figure-latex/fig-prevlit-overall-1.pdf}
\caption{\label{fig:fig-prevlit-overall}NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages. The legend indicates language and the length of non-words (in syllables). Central tendency is mean; error is one standard error.}
\end{figure}

Several observations can be drawn from this figure. To begin with, we focus on the comparison between Yélî Dnye and Tsimane'. These two groups have been described as having (very) roughly similar lifestyles and levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure \ref{fig:fig-prevlit}, larger length effects). What can we conclude from this apparent difference?

One possible lens of interpretation relates NWR differences to the minor differences in children's directed linguistic input rates. However, the differences in methods used to estimate these rates render this first interpretation unreliable. Based on current estimates Tsimane' children may encounter somewhere between similar to somewhat lower rates of directed speech comared to Yélî children. The lower boundary for Tsimane' input rate is established by Cristia, Dupoux, Gurven, and Stieglitz (2019), who used behavioral observations of child-inclusive one-on-one conversation. The upper boundary is set by Scaff, Stieglitz, Casillas, and Cristia (2021), who used human annotation to detect speech, feeding this into an automated temporal method for assigning speech as child-directed or not (NB: this method could lead to over- but not under-estimation because any nearby speech, e.g.~from a female adult, that coincided with child vocalization would count as child-directed). Yélî Dnye estimates come from Casillas, Brown, and Levinson (2021), who hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, effectively establishing an upper boundary of all the speech children could potentially process. There is no equivalent `lower bound' for Yélî Dnye to compare with the Tsimane' findings. So, while it is tempting to link this difference in NWR scores to directed input rates, we will first need to actually establish input rates using comparable methods for Yéli and Tsimane' children.

Prevalence of literacy is a second possible way to interpret these differences in NWR scores. Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) point out the relatively low prevalence of literacy, and more generally, the variable access to formal education available in the Tsimane' population. This is very different from the present Yélî population, where most adults have accumulated some schooling, with many having basic to fluent literacy in English (note: literacy in Yélî Dnye is much less common, especially among the very young; see main text). Under this second hypothesis, there are phonetic effects of learning to read that have significant consequences for young children's encoding and decoding of sounds in the context of NWR tasks. Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance illustrated, for instance, in the data for Sotho in Figure \ref{fig:fig-prevlit-overall}. Those two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English. What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related to literacy \emph{in the broader population} (rather than in the tested children themselves). However, there is at least one more plausible explanation: design and execution of the NWR task

A third, plausible, explanation for these differences in NWR is that the Tsimane' results are not generally comparable to the previous body of literature, and specifically not comparable to our present study. Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead. This implies a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise). Thus, a priority is for additional data to be gathered using this more novel testing paradigm in other populations; or more from the Tsimane' population, but using the more traditional paradigm.

Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies. We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores. We extracted the estimated number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be relatively short or long, based on information in the papers cited above and other sources. Neither of these two predictors explained the variance in average NWR scores, illustrated in Figure \ref{fig:fig-prevlit-overall}. It is possible that average word length plays a role, but that researchers tend to incorporate this into their design by including longer items when the native language allows it, with e.g.~Sotho non-words having 4--7 syllables in length.

\hypertarget{references}{%
\section{References}\label{references}}

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-balladares2016socio}{}%
Balladares, J., Marshall, C., \& Griffiths, Y. (2016). {Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers}. \emph{{First Language}}, \emph{36}(3), 338--351. \url{https://doi.org/10.1177/0142723715626067}

\leavevmode\hypertarget{ref-casillas2021early}{}%
Casillas, M., Brown, P., \& Levinson, S. C. (2021). {Early language experience in a Papuan community}. \emph{Journal of Child Language}, \emph{48}(4), 792--814.

\leavevmode\hypertarget{ref-cristia2019child}{}%
Cristia, A., Dupoux, E., Gurven, M., \& Stieglitz, J. (2019). Child-directed speech is infrequent in a forager-farmer population. \emph{{Child Development}}, \emph{90}, 759--773. \url{https://doi.org/10.1111/cdev.12974}

\leavevmode\hypertarget{ref-cristia2020infant}{}%
Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J. (2020). Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the {T}simane'. \emph{PLoS ONE}, \emph{15}(9), e0237702. https://doi.org/\url{https://doi.org/10.1371/journal.pone.0237702}

\leavevmode\hypertarget{ref-farmani2018normalization}{}%
Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F. Z., Talebi, E., \& Shourvazi, Z. (2018). Normalization of the non-word repetition test in {F}arsi-speaking children. \emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.

\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual {A}rabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}, 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}

\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B. (2014). Nonword repetition--a clinical marker for specific language impairment in {S}wedish associated with parents' language-related problems. \emph{PloS One}, \emph{9}(2), e89544.

\leavevmode\hypertarget{ref-kapalkova2013non}{}%
Kapalková, S., Polišenská, K., \& Vicenová, Z. (2013). {Non-word repetition performance in Slovak-speaking children with and without SLI: novel scoring methods}. \emph{{International Journal of Language and Communication Disorders}}, \emph{48}(1), 78--89. \url{https://doi.org/10.1111/j.1460-6984.2012.00189.x}

\leavevmode\hypertarget{ref-lei2011developmental}{}%
Lei, L., Pan, J., Liu, H., McBride-Chang, C., Li, H., Zhang, Y., \ldots{} others. (2011). Developmental trajectories of reading development and impairment from ages 3 to 8 years in chinese children. \emph{Journal of Child Psychology and Psychiatry}, \emph{52}(2), 212--220.

\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A. (2019). {An articulatory-based disyllabic and trisyllabic Non-Word Repetition test: reliability and validity in Italian 3-to 7-year-old children}. \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.

\leavevmode\hypertarget{ref-polivsenska2014improving}{}%
Polišenská, K., \& Kapalková, S. (2014). Improving child compliance on a computer-administered nonword repetition task. \emph{{Journal of Speech, Language and Hearing Research}}, \emph{57}(3).

\leavevmode\hypertarget{ref-radeborg2006swedish}{}%
Radeborg, K., Barthelom, E., SjöBerg, M., \& Sahlén, B. (2006). {A Swedish non-word repetition test for preschool children}. \emph{{Scandinavian Journal of Psychology}}, \emph{47}(3), 187--192. \url{https://doi.org/10.1111/j.1467-9450.2006.00506.x}

\leavevmode\hypertarget{ref-scaff2021daylong}{}%
Scaff, C., Stieglitz, J., Casillas, M., \& Cristia, A. (2021). Daylong audio recordings of young children in a forager-farmer society show low levels of verbal input with minimal age-related changes. \emph{Draft}.

\leavevmode\hypertarget{ref-stokes2006nonword}{}%
Stokes, S. F., Wong, A. M., Fletcher, P., \& Leonard, L. B. (2006). Nonword repetition and sentence repetition as clinical markers of specific language impairment: The case of cantonese. \emph{Journal of Speech, Language, and Hearing Research}, \emph{49}, 219--236.

\leavevmode\hypertarget{ref-thomas2020impact}{}%
Thomas-Odenthal, F., Molero, P., Does, W. van der, \& Molendijk, M. (2020). Impact of review method on the conclusions of clinical reviews: A systematic review on dietary interventions in depression as a case in point. \emph{PloS One}, \emph{15}(9), e0238131.

\leavevmode\hypertarget{ref-vance2005speech}{}%
Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.

\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
Wilsenach, C. (2013). {Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners}. \emph{{Per Linguam: A Journal of Language Learning= Per Linguam: Tydskrif Vir Taalaanleer}}, \emph{29}(2), 17--32. \url{https://doi.org/10.5785/29-2-554}

\end{CSLReferences}

\end{document}

### 2021-10-25 manuscript.Rmd
cleaning

@@
 On the other hand, it is also possible that this pressure gives Yélî children no benefit, and that some of these categories are simply acquired later in development.
-We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not make adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) till age 5 or 6 [@rumsey2017dependency]
+We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not produce adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) until  5 or 6 years of age [@rumsey2017dependency]
 

### 2021-10-25 manuscript.Rmd
cleaning

@@
 knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
+options(knitr.graphics.error = FALSE)
 
@@
 #add orthographic representations targets & other stims chars
-stims.tbl <- read.xlsx("stimuli-20211018.xlsx") %>%
+stims.tbl <- read.xlsx("stimuli.xlsx") %>%
   arrange(factor(type, levels = c("practice", "1", "2", "3", "4"))) %>%
@@
 colnames(xling)<-c("Target","Avg_fr","Word")
-stims.ipa <- read.xlsx("stimuli-20211018.xlsx") %>%
+stims.ipa <- read.xlsx("stimuli.xlsx") %>%
   select(target, target_ortho, phono)
@@
 
-[^4]: Note that a full anonymized, but age-annotated table of errors is provided in the project repository for those interested in a finer-grained analysis than what is presented here: ALEX?].
+[^4]: Note that tables of errors including child age are provided in the project repository for those interested in a finer-grained analysis than what is presented here.
+See https://osf.io/5qspb/wiki/home/, quick links, error tables.
 

### 2021-10-25 manuscript.tex
cleaning

@@
 We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde{} of unmatched phones).
-To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?\footnote{Note that a full anonymized, but age-annotated table of errors is provided in the project repository for those interested in a finer-grained analysis than what is presented here: ALEX?{]}.}
+To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?\footnote{Note that tables of errors including child age are provided in the project repository for those interested in a finer-grained analysis than what is presented here.
+See \url{https://osf.io/5qspb/wiki/home/}, quick links, error tables.}
 
@@
 On the other hand, it is also possible that this pressure gives Yélî children no benefit, and that some of these categories are simply acquired later in development.
-We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not make adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) till age 5 or 6 (Rumsey, 2017)
+We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not produce adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) until 5 or 6 years of age (Rumsey, 2017)
 

### 2021-10-27 manuscript.Rmd
version ready for proofing

@@
 keywords          : "phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy"
-wordcount         : "~11,500 words"
+wordcount         : "~12,200 words"
 
@@
 
-sink("lastknit_session_info.txt")
-sessionInfo()
-sink()
+writeLines(capture.output(sessionInfo()), "lastknit_session_info.txt")
+
+#missing compared to MC
+# compiler_4.1.1
+# AC has xfun_0.27 , MC has xfun_0.26
+# ac has rlang_0.4.12, mc has rlang_0.4.11
 ```
@@
 Regarding child sex, no significant correlation has been found in previous NWR research [@chiat2007preschool], and in other language tasks evidence is mixed.
-Finally, prior research varies on whether significant differences as a function of  maternal education are reported [e.g., no difference found in @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio; whereas differences were reported in @tuller2018identifying; @dos2020epreuve].
+Finally, prior research varies on whether significant differences as a function of  maternal education are reported.
+For instance, no significant difference was found some studies [@farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio]; whereas significant differences were reported in others [@tuller2018identifying; @dos2020epreuve].
 In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports [@frank2017wordbank] and word comprehension studies [@scaff2019].
@@
 To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken associations with maternal education.
-<!-- Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
-ac2mc removed, R was confused and I'm also not sure what is the relevance of this comment
-
-mc2ac: fine with me, thanks!-->
 In addition, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply be due to random factors like living close to a school or having relatives there.
@@
 
-<!-- ac2mc shouldn't we totally rewrite this? // actually, this is all factually correct-->
-<!-- mc2ac: while factually correct i thought it might not be a great representation of where we stand now so i made some small edits in the final sentences-->
 We add some observations that will help us integrate this study into the broader investigation of NWR across cultures.
@@
   row.names =F) %>%
-  kable_styling() %>%
+  kable_styling(latex_options = "HOLD_position") %>%
   add_header_above(c("Practice" = 2, "Monosyll" = 2,
@@
 The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
-Because the phonological inventory is so acoustically packed and because annotation was done based on audio data alone, it might be easy to mis-identify a segment, the assistant double-checked all of her annotations by listening to them and assessing them a second time, once she had completed a full first round.
+Because the phonological inventory is so acoustically packed and annotation was done based on audio data alone, it might be easy to misidentify a segment.
+Therefore, the assistant double-checked all of her annotations by listening to them and assessing them a second time, once she had completed a full first round.
 
@@
 knitr::include_graphics("nwr.by.freq.ITEM.png")
-dev.off()
+#dev.off()
 ```
@@
 Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = `r mycoeff["avg_fr_s:age.rounded_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s:age.rounded_s","Std.
-Error"]`, p `r pval.display(mycoeff["avg_fr_s:age.rounded_s","Pr(>|z|)"])`): The effect of frequency was larger for younger children.
+Error"]`, p `r pval.display(mycoeff["avg_fr_s:age.rounded_s","Pr(>|z|)"])`): The effect of frequency was larger for older children.
 Inspection of Figure \@ref(fig:fig-xling-freq-age3) suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically uncommon phones.
 
-```{r fig-xling-freq-age3, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="Model predicting NWR scores from child age (x axis) and the average frequency with which each phone is found across languages (mean, or plus/minus one standard deviation)."}
+```{r fig-xling-freq-age3, fig.pos = "!t", fig.align = "center", out.width = '65%', fig.cap="NWR scores as a function of age and typological frequency.
+Lines are fits from the model in the main text predicting NWR scores from child age (x axis) and the average frequency with which each phone is found across languages (mean, or plus/minus one standard deviation).
+Each circle indicates the estimated NWR scores for one child at one frequency level."}
 
@@
                     guide = FALSE) +
-  labs(color = "Avgerage\nfrequency\nscore",
+  labs(color = "Average\nfrequency\nscore",
        y = "% Correct", x = "Child age\n(standardized)") +
@@
 ```
-
-<!-- ac2mc if it's easy for you to add dots representing actual children in the above, that would be awesome! If not, I can try to tackle it :)
-
-mc2ac here's my attempt! i wasn't sure how you wanted to aggregate the real data to correspond to the model estimates so i have here grouped correctness within individual by
-<= m - 1sd
-mean <±> 1sd (unconventional notation, sorry)
->= m + 1sd
-
-if you had something else in mind but are otherwise cool with the figure, it should be easy to edit on the mutate/case_when() portion defining trnsc.inc.indivs
--->
 
@@
 
-#read.table("substitution_bank_try1.txt",header=T)->sub_bank
 read.table("substitution_bank.txt",header=T)->sub_bank_full
@@
 
-#read.table("deletion_bank_try1.txt",header=T)->del_bank
 read.table("deletion_bank.txt",header=T)->del_bank_full
@@
 
-#read.table("match_bank_try1.txt",header=T)->match_bank
 read.table("match_bank.txt",header=T)->match_bank_full
@@
 This effect was large, with a magnitude more than twice the size of the effect of participant age.
-Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at younger ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for younger than older children).
+Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at older ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for older than younger children).
 Importantly, the correlation between performance and typological frequency  remained significant  after accounting for the frequencies of these segments in a conversational corpus.

### 2021-10-27 manuscript.tex
version ready for proofing

@@
 \shorttitle{NWR in Yélî Dnye learners}
-\keywords{phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy\newline\indent Word count: ~11,500 words}
+\keywords{phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy\newline\indent Word count: ~12,200 words}
 \usepackage{lineno}
@@
 Regarding child sex, no significant correlation has been found in previous NWR research (Chiat \& Roy, 2007), and in other language tasks evidence is mixed.
-Finally, prior research varies on whether significant differences as a function of maternal education are reported (Balladares, Marshall, \& Griffiths, 2016; e.g., no difference found in Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017; Santos, Frau, Labrevoit, \& Zebib, 2020; whereas differences were reported in Tuller et al., 2018).
+Finally, prior research varies on whether significant differences as a function of maternal education are reported.
+For instance, no significant difference was found some studies (Balladares, Marshall, \& Griffiths, 2016; Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017); whereas significant differences were reported in others (Santos, Frau, Labrevoit, \& Zebib, 2020; Tuller et al., 2018).
 In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports (Frank, Braginsky, Yurovsky, \& Marchman, 2017) and word comprehension studies (Scaff, 2019).
@@
 
-\begin{table}
+\begin{table}[H]
 
@@
 
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-
-Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for younger children.
+Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for older children.
 Inspection of Figure \ref{fig:fig-xling-freq-age3} suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically uncommon phones.
@@
 
-\caption{Model predicting NWR scores from child age (x axis) and the average frequency with which each phone is found across languages (mean, or plus/minus one standard deviation).}\label{fig:fig-xling-freq-age3}
+\caption{NWR scores as a function of age and typological frequency.
+Lines are fits from the model in the main text predicting NWR scores from child age (x axis) and the average frequency with which each phone is found across languages (mean, or plus/minus one standard deviation).
+Each circle indicates the estimated NWR scores for one child at one frequency level.}\label{fig:fig-xling-freq-age3}
 \end{figure}
@@
 This effect was large, with a magnitude more than twice the size of the effect of participant age.
-Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at younger ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for younger than older children).
+Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at older ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for older than younger children).
 Importantly, the correlation between performance and typological frequency remained significant after accounting for the frequencies of these segments in a conversational corpus.

### 2021-10-27 sub1.tex
version ready for proofing

--- 2021-10-18 manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
 \PassOptionsToPackage{hyphens}{url}
 %
-\documentclass[
-  american,
-  ,man,floatsintext]{apa6}
-\usepackage{amsmath,amssymb}
+\documentclass[english,,man,floatsintext]{apa6}
 \usepackage{lmodern}
+\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
+  \usepackage{textcomp} % provides euro and other symbols
+\else % if luatex or xelatex
   \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
 \fi
-% Use upquote if available, for straight quotes in verbatim environments
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
+% use microtype if available
+\IfFileExists{microtype.sty}{%
 \usepackage[]{microtype}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
 \IfFileExists{parskip.sty}{%
@@
 \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\usepackage{hyperref}
 \hypersetup{
             pdftitle={Non-word repetition in children learning Yélî Dnye},
-  pdfauthor={Alejandrina Cristia1 \& Marisa Casillas2,3},
-  pdflang={en-US},
-  pdfkeywords={phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
+            pdfkeywords={phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness},
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{0}
+
+% set default figure placement to htbp
 \makeatletter
@@
 \makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
+
 \usepackage{fontspec}
@@
 % \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
+%   {\begin{landscape}\begin{center}\begin{threeparttable}}
+%   {\end{threeparttable}\end{center}\end{landscape}}
+\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}
 
@@
   {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
 \makeatother
 \shorttitle{NWR in Yélî Dnye learners}
-\keywords{phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy\newline\indent Word count: ~11,500 words}
+\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
+\affiliation{
+\vspace{0.5cm}
+\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
+\authornote{Both authors contributed to study funding, design, data collection, annotation, analyses, writing.
+
+Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
+\keywords{phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness\newline\indent Word count: ~11,500 words}
 \usepackage{lineno}
@@
 \usepackage{csquotes}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[shorthands=off,main=english]{babel}
+\else
+  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
   \usepackage{polyglossia}
-  \setmainlanguage[variant=american]{english}
-\else
-  \usepackage[main=american]{babel}
-% get rid of language-specific shorthands (see #6817):
-\let\LanguageShortHands\languageshorthands
-\def\languageshorthands#1{}
+  \setmainlanguage[]{english}
 \fi
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
 \title{Non-word repetition in children learning Yélî Dnye}
-\author{Alejandrina Cristia\textsuperscript{1} \& Marisa Casillas\textsuperscript{2,3}}
+
 \date{}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes Cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
 
@@
 NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
-In this study, we report on NWR results among children (N = 40, aged 3--10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
+In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
 Results make three contributions that are specific, and a fourth that is general.
@@
 Second, we documented rather weak effects of item length.
-Third, we found that age has a strong correlation with NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
+Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.
 Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations, and the extent to which they shed light on the factors universally structuring variation in phonological development at a global and individual level.
@@
 
-Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000; Rumsey, 2017).
+Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000).
 Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
+In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning.
+After hearing this non-word, the participant's task is to try to immediately and precisely repeat it.
+NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
+NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
+Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
+NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
-This allows us to contribute data at the intersection of language typology, language acquisition, and individual variation, as presented in more detail below.
-
-\hypertarget{what-is-nwr}{%
-\subsubsection{What is NWR?}\label{what-is-nwr}}
+The study was designed to contribute to four broad research areas, three via direct results.
+
+The first research area is at the crossing of typology and phonological development.
+Previous work using NWR has preferred relatively universal and early-acquired phonemes (with exceptions including Gallagher, 2014), in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects (Gallon, Harris, \& Van der Lely, 2007) and in part because the test is sometimes used to measure working memory in the context of executive functions (Mulder, Verhagen, Van der Ven, Slot, \& Leseman, 2017) rather than purely linguistic skills.
+Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets.
+Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.
+
+Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages.
+Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, \& Leonard, 2006), whereas differences are negligible in other studies (e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
+It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture.
+In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.
+To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths.
+We found data for learners of Israeli Arabic (Jaber-Awida, 2018); Cantonese (Stokes et al., 2006); English (Vance, Stackhouse, \& Wells, 2005); Italian (Piazzalunga et al., 2019); and Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020); and integrated those data with Yélî Dnye results from the present study in Figure \ref{fig:fig-prevlit}.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/fig-prevlit-1.pdf}
+\caption{\label{fig:fig-prevlit}NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables, 2=dashed, 3=dotted, 4=dotted and dashed).
+Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al.
+(2019) reported on groups of 24-60 Italian learners (black); Stokes et al.
+(2006) on 15 Cantonese learners (blue); Vance et al.
+(2005) on 17-20 English learners (light green); Cristia et al.
+(2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple).
+Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error.
+Age has been slightly shifted for ease of inspection of different lengths at a given age.}
+\end{figure}
+
+Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages.
+For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (\textasciitilde{}28\%) and Arabic (\textasciitilde{}15\%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only \textasciitilde{}2\%.
+Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese \textasciitilde{}8\% versus English \textasciitilde{}1\%).
+Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.
+
+The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured.
+Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
+In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size.
+For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
+Based on previous work, we looked at potential increases with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance et al., 2005).
+Prior research typically finds no significant differences as a function of maternal education (e.g., Farmani et al., 2018; Balladares, Marshall, \& Griffiths, 2016; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
+Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.
+
+Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America (with exceptions including Gallagher, 2014; Cristia et al., 2020), nor with languages having large phonological inventories {[}e.g., more than 34 consonants and 7 vowel qualities Maddieson (2013b); Maddieson (2013a); with no exceptions to our knowledge{]}.
+There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
+That said, Cristia et al.
+(2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
+In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance (e.g., Gathercole, Willis, \& Baddeley, 1991).
+Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills.
+Indeed, Cristia et al.
+(2020) present the task as being a good index of the development of \enquote{short-hand-like} representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+Aside from Cristia et al.
+(2020)'s hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).
+Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions.
+This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.
+
+Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
+As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible.
+Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular.
+International phone calls and digital exchanges that require significant data transfer are typically not an option.
+Data collection is therefore typically limited to the duration of the researchers' on-island visits.
+
+\hypertarget{yuxe9luxee-dnye-phonology}{%
+\subsubsection{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}}
+
+Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
+The Yélî sound system, much like its baroque grammatical system (Levinson, 2020), is unlike any other in the region.
+In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2020; Maddieson \& Levinson, n.d.).
+Thus, with respect to our first research goal, Yélî Dnye is a good language to test because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
+
+To provide some qualitative information on this inventory, we add the following observations.
+With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
+The core oral stop system includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (2020) for details).\footnote{We use Levinson's (2020) under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
+In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
+Vowels in Yélî Dnye may be oral or nasal, short or long.
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+
+Regarding our second research goal, on the effect of non-word length on NWR, most Yélî Dnye words are disyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent dictionary at the time of writing; Levinson, 2020).
+The vast majority of syllables use a CV format.
+A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} /ṇṭæp/ \emph{Spondylus shell}) and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \enquote{ndapî} /'ṇṭæ.pɯ/).
+There are also a handful of words starting with /æ/ (e.g., \enquote{ala} /æ.'læ/ \emph{here}) and a small collection of single-vowel grammatical morphemes (see Levinson (2020) for details).
+
+Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun.
+For example, Peute and colleagues' (n.d.) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
+We hope the present study contributes to this growing line of work.
+
+\hypertarget{the-yuxe9luxee-community}{%
+\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}}
+
+Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation.
+Specifically, we investigated potential effects of age, gender, maternal education, and birth order.
+There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.
+
+The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
+These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
+Land attribution for building one's home is decided collectively based on land availability.
+
+Most Yélî parents are swidden horticulturalists.
+Within a group of households, it is often the case that older adolescents and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available (Casillas et al., 2020).
+Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year.
+Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.
+
+Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above.
+To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects.
+Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years).
+Overall, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.
+
+As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings (Barclay, 2015; Grätz, 2018), where nuclear families are more likely to be the prevalent rearing environment (Lancy, 2015).
+It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation.
+This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.
+
+We add some observations that will help us integrate this study to the broader investigation of NWR across cultures.
+As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia et al., 2020).
+However, Cristia et al.
+(2020) also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for.
+The Yélî community can help us shed further light onto this question because direct speech to children under 3;0 is relatively infrequent in this community too (Casillas et al., 2020).
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas et al., 2020).
+Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in Cristia et al.
+(2020).
+
+\hypertarget{nwr-design-and-analysis-adaptations}{%
+\subsubsection{NWR design and analysis adaptations}\label{nwr-design-and-analysis-adaptations}}
 
@@
 For instance, in English, a response of {[}bilig{]} or {[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where the vowel is lengthened without change of quality would be scored as correct, because English does not have contrastive vowel length.
-
-NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014).
-NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Chiat, 2015; Estes, Evans, \& Else-Quest, 2007).
-Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children (Armon-Lotem, Jong, \& Meir, 2015; Chiat, 2015; COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016).
-NWR tasks probably tap into many skills (for relevant discussion see Coady \& Evans, 2008; Santos, Frau, Labrevoit, \& Zebib, 2020).
-Non-words can be designed to try to isolate certain skills more narrowly; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory (see a discussion in Chiat, 2015).
-Broadly, however, NWR scores will necessarily reflect to a certain extent phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).
-
-\hypertarget{the-present-work}{%
-\subsubsection{The present work}\label{the-present-work}}
-
-We aimed to contribute to four areas of research. We motivate each in turn.
-
-\hypertarget{nwr-and-typology}{%
-\paragraph{NWR and typology}\label{nwr-and-typology}}
-
-The first research area is at the intersection of typology and phonological development.
-There has been an interest in adapting NWR to different languages, in part for applied purposes.
-In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, Chiat (2015) discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody.
-As a result, any one item created will be relatively easier if it more closely resembles real words in a language, making it difficult to balance difficulty when comparing children learning different languages.
-This previous literature also suggests some dimensions of difficulty---an issue to which we return in the next subsection.
-
-Although this cross-linguistic literature is rich, the potential difficulty associated with specific phonetic targets composing the non-words has received relatively little attention.
-For example, Chiat (2015) discusses segmental complexity as a function of whether there are consonant clusters -- which is arguably a factor reflecting phonotactics and syllable structure.
-
-In the present study, we thought it was relevant to represent the rich phonological inventory found in Yélî Dnye, by including a variety of phonetic targets.
-Some of them are cross-linguistically rare, in that they are less common across languages than other sounds or phonetic targets.
-Phonologists, phoneticians, and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of processing and acquisition via diachronic language change.
-These works focus largely on phonotactics (Moreton \& Pater, 2012) perceptual parsing of the (ambiguous) linguistic signal (Beddor, 2009; Ohala, 1981), and individual differences in processing styles (Bermúdez-Otero, 2015); small but significant effects that may cumulatively drive language change via phonologization (see Yu, 2021 for a recent review).
-Thus, the correlation between typological frequency and ease of acquisition is typically assumed to emerge from one or more of the following causal paths:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Sounds (and sound sequences) that are harder to perceive tend to be misperceived and thus lost diachronically
-\item
-  Sounds (and sound sequences) that are harder to pronounce tend to be mispronounced and thus lost diachronically
-\item
-  Sound sequences that are harder to hold in memory tend to be mispronounced and thus lost diachronically
-\end{enumerate}
-
-Given these causal pathways, we predicted that variation in NWR across items will correlate with the cross-linguistic frequency of the phones composing those items.
-
-\hypertarget{length-effects-on-nwr}{%
-\paragraph{Length effects on NWR}\label{length-effects-on-nwr}}
-
-The second research area we contribute data to is research looking at the impact of word length on NWR repetition within specific languages.
-Some work documents much lower NWR scores for longer, compared to shorter, items {[}e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, and Leonard (2006){]}, whereas differences are negligible in other studies {[}e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019){]}.
-
-It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture---a hypothesis discussed for instance in Chiat (2015) (pp.~7-8; see also p.~5).
-In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer.
-A non-systematic meta-analysis does not provide overwhelming support for this hypothesis {[}Cristia and Casillas (2021); SM1{]}.
-
-Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we did not approach this issue within a hypothesis-testing framework but sought instead to provide additional data on the question, which may be re-used in future meta- or mega-analytic analyses.
-
-\hypertarget{individual-variation-correlations-with-nwr}{%
-\paragraph{Individual variation correlations with NWR}\label{individual-variation-correlations-with-nwr}}
-
-The third research area we contribute data to relates to the possibility that children differ from each other in NWR scores in systematic ways.
-Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
-In a nutshell, most evidence is mixed, suggesting that correlations with individual variation may be small, and more data is needed to estimate their true size.
-For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
-
-Our focus on age stems from previous work, where performance increases with child age (Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005).
-Although past research has not investigated potential correlations with birth order on NWR, there is a sizable literature on these correlations in other language tasks (e.g., Havron et al., 2019), and therefore we report on these too.
-Common explanations for advantages for first- over later-born children include differential allocation of familial resources, particularly parental behaviors of cognitive stimulation (Lehmann, Nuevo-Chiquero, \& Vidal-Fernandez, 2018).
-Regarding child sex, no significant correlation has been found in previous NWR research (Chiat \& Roy, 2007), and in other language tasks evidence is mixed.
-Finally, prior research varies on whether significant differences as a function of maternal education are reported (Balladares, Marshall, \& Griffiths, 2016; e.g., no difference found in Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017; Santos, Frau, Labrevoit, \& Zebib, 2020; whereas differences were reported in Tuller et al., 2018).
-In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports (Frank, Braginsky, Yurovsky, \& Marchman, 2017) and word comprehension studies (Scaff, 2019).
-The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children (see discussion in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
-
-\hypertarget{nwr-as-a-function-of-language-and-culture}{%
-\paragraph{NWR as a function of language and culture}\label{nwr-as-a-function-of-language-and-culture}}
-
-The fourth research goal we pursued is to use NWR with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020; with exceptions including Gallagher, 2014).
-To our knowledge, it has never been used with speakers of languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities Maddieson, 2013b, 2013a).
-
-There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
-That said, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
-
-In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading; for instance, better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991).
-Therefore, it may not be a general index of phonological development, but instead reflect certain non-universal language skills.
-Indeed, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) present their task as being a good index of the development of ``short-hand-like'' representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
-We return to the question of what was measured here in the Discussion.
-
-Aside from Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)'s hypotheses just mentioned, we have found little discussion of linguistic differences (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural differences (i.e., potential differences in NWR as a function of other differences across human populations).\footnote{Please note that the linguistic and cultural differences discussed here are different from the differences discussed in the extensive literature on NWR by bilingual participants.
-In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment.
-To try to measure language abilities above and beyond relative levels of experience with a given language, authors have tried to build non-words that tap language-dependent or language-independent knowledge.
-For instance, Tuller et al.
-(2018) employed a set of non-words judged to be language independent and two others that were more aligned with either French or German.
-The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language (``language-dependent'') than when they are ``language-independent.'' To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience.
-Preliminary results of an ongoing meta-analysis suggest significant associations between exposure to a given language and performance in both language-dependent and language-independent NWR (Farabolini, Taboh, Ceravolo, \& Guerra, 2021).
-In any case, this line of research focuses on links between exposure to a given language and NWR performance.
-In contrast, when we discuss linguistic or cultural differences here, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).}
-
-Regarding potential language differences, we note that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/).
-It would be interesting for future researchers to consider straying from the literature by varying other dimensions that are relevant to the language under study.
-For instance, for Yélî Dnye, it is relevant to vary phonological complexity of the individual sounds because of its large inventory.
-
-\hypertarget{yuxe9luxee-dnye-phonology-and-community}{%
-\subsubsection{Yélî Dnye phonology and community}\label{yuxe9luxee-dnye-phonology-and-community}}
-
-Before going into the details of our study design, we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island.
-As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethnographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites.
-Rossel Island lies 250 nautical miles off the coast of mainland PNG and is surrounded by a barrier reef.
-As a result, transport to and from the island is both infrequent and irregular.
-International phone calls and digital exchanges that require significant data transfer are typically not an option.
-Data collection is therefore typically limited to the duration of the researchers' on-island visits.
-
-\hypertarget{yuxe9luxee-dnye-phonology}{%
-\paragraph{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}}
-
-Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea.
-The Yélî sound system, much like its baroque grammatical system (Levinson, 2021), is unlike any other in the region.
-In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phoneme inventory size of other documented Papuan languages (Foley, 1986; Levinson, 2021; Maddieson \& Levinson, in preparation).
-Thus, with respect to our first research goal, Yélî Dnye is a good language to use because its large phonological inventory includes sounds that vary in cross-linguistic frequency (including some rare sounds) that can be compared in the NWR setting.
-
-To provide some qualitative information on this inventory, we add the following observations.
-With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments.
-The core oral stop system includes both singleton (/p/, /t/, /ṭ/\footnote{We use Levinson's (2021) under-dot notation (e.g., /ṭ/) to denote the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically.
-In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).}, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with a complete range of nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (2021) for details).
-The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
-Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to occur as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
-
-Our second research goal is to measure the effect of non-word length on NWR, which may need to be interpreted taking into account typical word length in the language.
-We estimated word length in words found in a conversational corpus (see Stimuli section for details), where the distribution of length was: 15\% monosyllabic, 39\% disyllabic, 29\% trisyllabic, and the remaining 17\% being longer than that.
-The vast majority of syllables use a CV format.
-A small portion of the lexicon features words with a final CVC syllable, but these are limited to codas of -/m/, -/p/, or -/j/ (e.g., \emph{ndap} /ṇṭæp/ `Spondylus shell') and are often resyllabified with an epenthetic /ɯ/ in spontaneous speech (e.g., \emph{ndapî} /`ṇṭæpɯ/).
-There are also a handful of words starting with /æ/ (e.g., \emph{ala} /æ'læ/ 'here') and a small collection of single-vowel grammatical morphemes (see Levinson (2021) for details).
-
-Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2021; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun.
-For example, Peute and Casillas (In preparation) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones.
-Other ongoing work on Yélî Dnye includes experiment-based infant phoneme discrimination data and errors made in elicited and spontaneous speech from young children, but these data are neither finalized nor yet externally reviewed (see Hellwig, Sarvasy, \& Casillas, provisionally accepted for more information).
-These data will help better inform our current analyses based on NWR in the future (e.g., regarding common sound substitutions) but are not critical for testing our current question about the general correlation between cross-linguistic phone frequency and NWR performance.
-
-Before closing this section, it bears mentioning that the language has an established orthography, which includes distinct graphemes for all the contrasts on which our items are based.
-Some children in our sample will have started school.
-Reading and writing instruction is currently done only in English (other than writing one's name).
-This was probably not the case for the majority of mothers of the children in our sample, who will have learned to read and write in Yélî Dnye during their first three years at school.
-It is possible that there is also some home teaching of Yélî reading and writing, notably for reading the bible.
-
-\hypertarget{the-yuxe9luxee-community}{%
-\paragraph{The Yélî community}\label{the-yuxe9luxee-community}}
-
-Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation.
-Specifically, we investigated potential correlations with age, child sex, maternal education, and birth order.
-There is nothing particular to note regarding age and child sex, but we have some comments that pertain to the other two factors.
-
-The typical household in our dataset includes seven individuals (typically, a mixed-sex couple and children---their own and possibly some others staying with them, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area.
-These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline.
-Land attribution for building one's home is decided collectively based on land availability.
-
-Most Yélî parents are swidden horticulturalists, who occasionally fish.
-Within a group of households, it is often the case that older adolescents and adults spend their day tending to their farm plots (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
-Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde10 members) independent and mixed-age child play groups (Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2021).
-Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas, Brown, \& Levinson, 2021).
-Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to host their school-aged relatives during the weekdays for long segments of the school year.
-Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.
-
-Some general ideas regarding potential correlations between our NWR measures and maternal education may be drawn from the observations above.
-To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken associations with maternal education.
-In addition, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply be due to random factors like living close to a school or having relatives there.
-
-As for birth order, much of the work on correlations between birth order and cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings (Barclay, 2015; Grätz, 2018), where nuclear families were more likely to be the prevalent rearing environment (Lancy, 2015).
-It is possible that birth order differences are stronger in such a setting, because much of the stimulation can only come from the parents.
-These effects may be much smaller in cultures where it is common for children to attend daycare at an early age (such as France) or where extended family typically live close by.
-The Yélî community falls in the latter case, as children are typically surrounded by siblings and cousins of several orders, regardless of their birth order in their nuclear family.
-
-We add some observations that will help us integrate this study into the broader investigation of NWR across cultures.
-As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors of that paper interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
-However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones Bunce et al., under review), and additionally shares other societal characteristics wih the Tsimane' {[}e.g., is rural and relies on farming, children grow up in wide familial networks, etc; Casillas, Brown, and Levinson (2021){]}.
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2021).
-Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).
+There is some variation in how past NWR studies have designed the presentation procedure and structure of items.
+For example, while items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation (Brandeker \& Thordardottir, 2015).
+Additionally, while some studies have used 10-15 non-words (e.g., Cristia et al., 2020), others have employed up to 46 unique items (Piazzalunga et al., 2019).
+Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure (open as opposed to closed syllables, Gallon et al., 2007).
+
+Previous work typically steers clear of articulatorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology and to contribute data on whether this affects repetition.
+We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs.~less challenging sounds.
+However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below).
+Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
+However, as discussed above, effects of individual differences in NWR are probably relatively small, and thus we reasoned that they would not be detectable with the sample size that we could collect during our short visit.
+That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta- or mega-analytic efforts.
 
@@
 
-Throughout these analyses and in the Discussion, we also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.
+Throughout these analyses and in the Discussion, we will also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.
 
 We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them.
-However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known associations, given the previously discussed variability across studies.
+However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given the previously discussed variability across studies.
 Therefore, all analyses in the present study are descriptive and should be considered exploratory.
@@
 
-\hypertarget{participants}{%
-\subsubsection{Participants}\label{participants}}
-
-This study was approved as part of a larger research effort by the second author.
-The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041), including the use of verbal (not written) consent.
-As discussed in subsection ``The Yélî community,'' the combination of collective child guardianship practices and common hosting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
-Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, in press).
-Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
-Regardless of whether they completed the task, children were given a small snack as compensation.
-Children who showed initial interest but then decided not to participate were also given the snack.
-
-We tested a total of 55 children from 38 families spread across four hamlet regions.
-We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5), or were 13 years old or older (N=2; we tested these teenagers to put younger children at ease).
-The remaining 40 children (14 girls) were aged from 3 to 10 years (M = 6.40 years, SD = 1.50 years).
-In terms of birth order, 6 were born first, 5 second, 2 third, 7 fourth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
-These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
-Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
-We then recorded the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 children only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
-While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, in press).
-Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.}
-
 \hypertarget{stimuli}{%
@@
 
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares et al., 2016; Wilsenach, 2013).
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-46 items: Jaber-Awida, 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019), with the entire test inventory of 40 final items distributed across children.
-We used a relatively large number of items to explore correlations with length and phonological complexity.
-However, aware that this large item inventory might render the task longer and more tiresome, we split items across children.
-Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014), with the entire test inventory of 40 final items distributed across children.
 
 A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/, /n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible 2- to 4-syllable sequences.
-These candidates were automatically removed from consideration if they appeared in the most recent dictionary (Levinson, 2021).
+These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary.
 The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
@@
 Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
-Additionally, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
+Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, 2020), it was also possible that we initially generated items with illegal, but currently undocumented constraints.
+Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
 These candidates were then presented to one informant, for a final check that they did not mean anything.
@@
 The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present.
-Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \emph{yî} /yɯ/; identified as \emph{yi} /yi/, `tree').
-Additionally, an error was made when preparing files for annotation, resulting in two items being merged (\emph{tpâ} /tpɑ/ and \emph{tp:a} /tpæ̃/).
+Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \enquote{yî} /yɯ/; identified as \enquote{yi} /yi/, \emph{tree}).
+Additionally, an error was made when preparing files for annotation, resulting in two items being merged (\enquote{tpâ} /tpɑ/ and \enquote{tp:a} /tpæ̃/).
 These three problematic items are not described here, and removed from the analyses below.
@@
 \midrule
-nopimade & nɔpimæṭɛ & dp:a & ṭpæ̃ & kamo & kæmɔ & dimope & ṭimɔpɛ & dipońate & ṭipɔnætɛ\\
+nopimade & nɔpimæṭɛ & dp:a & ṭp̃æ & kamo & kæmɔ & dimope & ṭimɔpɛ & dipońate & ṭipɔnætɛ\\
 poni & pɔni & dpa & ṭpæ & kańi & kæni & diyeto & ṭijɛtɔ & ńomiwake & nɔmiwækɛ\\
@@
  &  & dpu & ṭpu & towi & tɔwi & ńuyedi & nujɛṭi &  & \\
- &  & gh:ââ & ɣɑ̃ː & tupa & tupæ & pedumi & pɛṭumi &  & \\
+ &  & gh:ââ & ɣ̃ɑː & tupa & tupæ & pedumi & pɛṭumi &  & \\
  &  & ghuu & ɣuː &  &  & tiwuńe & tiwunɛ &  & \\
- &  & kp:ââ & kpɑ̃ː &  &  & tumowe & tumɔwɛ &  & \\
+ &  & kp:ââ & kp̃ɑː &  &  & tumowe & tumɔwɛ &  & \\
 \addlinespace
  &  & kpu & kpu &  &  & widońe & wiṭɔnɛ &  & \\
- &  & lv:ê & lβʲə̃ &  &  & wumipo & wumipɔ &  & \\
+ &  & lv:ê & lβʲ̃ə &  &  & wumipo & wumipɔ &  & \\
  &  & lva & lβʲæ &  &  &  &  &  & \\
  &  & lvi & lβʲi &  &  &  &  &  & \\
- &  & t:êê & tə̃ː &  &  &  &  &  & \\
+ &  & t:êê & t̃əː &  &  &  &  &  & \\
 \addlinespace
@@
 
-A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and to split it into two sub-lists, to generate 40 different elicitation sets.
+A Praat script (Boersma \& Weenink, 2020) was written to randomize this list 20 times, and split it into two sub-lists, to generate 40 different elicitation sets.
 The 40 elicitation sets are available online from \href{https://osf.io/dtxue/}{osf.io/dtxue/}.
@@
 
-\hypertarget{cross-linguistic-frequency}{%
-\paragraph{Cross-linguistic frequency}\label{cross-linguistic-frequency}}
-
 To inform our analyses, we estimated the typological frequency of all phonological segments present in the target items using the PHOIBLE cross-linguistic phonological inventory database (Moran \& McCloy, 2019).
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is a unprecedented in its scope, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
-Note that nearly half of the phones in PHOIBLE are only attested in one language (Steven Moran, personal communication).
+While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
+Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication).
 Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0\% (i.e., its cross-linguistic percentile is zero).
 
-\hypertarget{within-language-frequency}{%
-\paragraph{Within-language frequency}\label{within-language-frequency}}
-
-Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2021).
-That corpus was constituted by sampling from audio-recordings (7--9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
-The researchers selected 9 2.5-minute clips randomly and 11 1- or 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity).
-These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the recorded people personally.
-For more details, please refer to Casillas, Brown, and Levinson (2021).
-
-For the present study, we extracted the transcriptions of adult speech (i.e., removing key child and other children's speech) and split them into words using white space.
-We then removed all English and Tok Pisin words.
-The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types.
-To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
-Note that the resulting values estimate usage frequencies for very young children's input and, while this is somewhat different from what our older participants experience on a daily basis, we can expect that this is a reasonable approximation of the early input that formed the foundation of their phonological knowledge.
+Additionally, we estimated frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas et al., 2020) by counting the number of word types in which they occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses).
 
@@
 
-There is some variation in procedure in previous work.
-For example, while items are often presented orally by the experimenter (Torrington Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation (Brandeker \& Thordardottir, 2015).
-
-In adapting the typical NWR procedure for our context, we balanced three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them (i.e., from other children who had participated); that children would feel comfortable doing this task with us; and that community members would feel comfortable having their children do this task with us.
-
-We tested in four different sites spread across the northeastern region of the island, making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from `spreading' between children through hearsay.
+In adapting the typical NWR procedure for this context, we balanced three desiderata: That children would not be unduly exposed to the items before they themselves had to repeat them (i.e., from other children who had participated); that children would feel comfortable doing this task with us; and that community members would feel comfortable having their children do this task with us.
+
+We tested in four different sites spread across the northeastern region of the island, making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from \enquote{spreading} between children through hearsay.
 Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.
@@
 To minimize adjustment time, which was uncomfortable for some children given the proximity of the foreign experimenter and equipment, we placed the headband on children's shoulders in these cases, carefully adjusting the microphone's placement so that it was still close to the child's mouth.
-A research assistant who spoke Yélî Dnye natively, and who could also hear the instructions over headphones, sat next to the child throughout the task to provide instructions and, if needed, encouragement.
+A research assistant who spoke Yélî Dnye natively sat next to the child throughout the task to provide instructions and, if needed, encouragement.
 The research assistant coached the child throughout the task to make sure that they understood what they were expected to do.
-Finally, an experimenter (the first author) was also fitted with headphones and a microphone; she was in charge of delivering the pre-recorded stimuli to the research assistant, the child, and herself over headphones.
+An experimenter (the first author) delivered the pre-recorded stimuli to the research assistant and the child over headphones.
 
@@
 The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets.
-Because the phonological inventory is so acoustically packed and because annotation was done based on audio data alone, it might be easy to mis-identify a segment, the assistant double-checked all of her annotations by listening to them and assessing them a second time, once she had completed a full first round.
 
@@
 
-Previous work typically reports two scores: a binary word-level exact repetition score, and a phoneme-level score, defined as the number of phonemes that can be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt; as in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
+Previous work typically reports two scores: a binary word-level exact repetition score, and a phoneme-level score, defined as the number of phonemes that can be aligned across the target and attempt, divided by the number of phonemes of whichever item was longer (the target or the attempt; as in Cristia et al., 2020).
 Previous work does not use distance metrics, but we report these rather than the phoneme-level scores because they are more informative.
@@
 An advantage of using phone-based Levenshtein distance is that it is scored automatically with a script, and it can then easily be split in terms of deletions and substitutions (insertions were not attested in this study).
+
+\hypertarget{participants}{%
+\subsubsection{Participants}\label{participants}}
+
+This study was approved as part of a larger research effort by the second author.
+The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041).
+As discussed in subsection \enquote{The Yélî community}, the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents.
+Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, n.d.).
+Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver.
+Regardless of whether they completed the task, children were given a small snack as compensation.
+Children who showed initial interest but then decided not to participate were also given the snack.
+
+We tested a total of 55 children from 38 families spread across four hamlet regions.
+We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5), or were 13 years old or older (N=2; we tested these teenagers to put younger children at ease).
+The remaining 40 children (14 girls) were aged from 3 to 10 years (M = 6.50 years, SD = 1.50 years).
+In terms of birth order, 6 were first borns, 5 second, 2 third, 7 fourth, 5 fifth, and 1 sixth, with birth order missing for 14 children.
+These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site.
+Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{We asked for mothers' highest completed level of education.
+We then record the number of years entailed by having completed that level under ideal conditions.} We also note that there were 34 children only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home.\footnote{Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English.
+While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.).
+Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.}
 
@@
 We could only find one comparison point: Castro-Caldas, Petersson, Reis, Stone-Elander, and Ingvar (1998) found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in 11.16\% of cases, whereas literate participants did so in only 1.71\% of cases.
-The percentage we observe here is much higher than reported in the study by Castro and colleagues, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
-Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this association.
-
-\hypertarget{nwr-and-typology-nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
-\subsubsection{NWR and typology: NWR as a function of cross-linguistic phone frequency}\label{nwr-and-typology-nwr-as-a-function-of-cross-linguistic-phone-frequency}}
+The percentage we observe here is much higher than reported in Castro and colleagues' study, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors.
+Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.
+
+\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
+\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}
 
@@
 To look at this, we fit a mixed logistic regression in which the outcome variable was whether the non-word was correctly repeated or not.
-The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, in interaction with cross-linguistic phone frequency, and allowed intercepts to vary over the random effects child ID and target ID.
+The fixed effect of interest was the average cross-linguistic phone frequency; we also included child age as a control fixed effect, and allowed slopes to vary over the random effects child ID and target ID.
 
 We could include 826 observations, from 40 children producing in any given trial one of 40 potential target words.
-The analysis revealed a main effect of age (ß = 0.39, SE ß = 0.13, p \textless{} 0.01), with older children repeating more items correctly.
-It also revealed a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.80, SE ß = 0.19, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:fig-xling-freq}.
+The analysis revealed a main effect of age (ß = 0.35, SE ß = 0.13, p \textless{} 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.78, SE ß = 0.19, p \textless{} 0.001): Target words with phones found more frequently across languages had higher correct repetition scores, as shown in Figure \ref{fig:fig-xling-freq}.
 Averaging across participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(38) = .544.
@@
 
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-
-Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for younger children.
-Inspection of Figure \ref{fig:fig-xling-freq-age3} suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically uncommon phones.
-
-\begin{figure}[!t]
-
-{\centering \includegraphics[width=0.65\linewidth]{manuscript_files/figure-latex/fig-xling-freq-age3-1}
-
-}
-
-\caption{Model predicting NWR scores from child age (x axis) and the average frequency with which each phone is found across languages (mean, or plus/minus one standard deviation).}\label{fig:fig-xling-freq-age3}
-\end{figure}
-
-\hypertarget{nwr-and-typology-nwr-as-a-function-of-within-language-phone-frequency}{%
-\subsubsection{NWR and typology: NWR as a function of within-language phone frequency}\label{nwr-and-typology-nwr-as-a-function-of-within-language-phone-frequency}}
-
-We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: The same perception and production pressures that shape languages diachronically could affect a language's lexicon, so that sounds that are easier to perceive or produce are more frequent within a language than those that are harder.
-If so, children will have more experience with the easier sounds, and they may thus be better able to to represent and repeat non-words containing them simply because of the additional exposure.
-
+We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: One can suppose that sounds that occur more frequently across languages are also more frequent within a language, and therefore may be easier for children to represent and repeat because of the additional exposure.
 Phone corpus-based frequencies were correlated with phone cross-linguistic frequencies {[}r(27)=0.50, p \textless{} 0.01{]}; and item-level average phone corpus-based frequencies were correlated with the corresponding cross-linguistic frequencies {[}r(38)=0.73, p \textless{} 0.001{]}.
@@
 
-\hypertarget{follow-up-analyses-patterns-in-nwr-mispronunciations}{%
-\subsubsection{Follow-up analyses: Patterns in NWR mispronunciations}\label{follow-up-analyses-patterns-in-nwr-mispronunciations}}
-
-We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts and not excluding errors resulting in real words, so as to base our generalizations on more data.
-Insertions and deletions were very rare (metathesis was not attested): there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde{} of unmatched phones).
-To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?\footnote{Note that a full anonymized, but age-annotated table of errors is provided in the project repository for those interested in a finer-grained analysis than what is presented here: ALEX?{]}.}
+\hypertarget{patterns-in-nwr-mispronunciations}{%
+\subsubsection{Patterns in NWR mispronunciations}\label{patterns-in-nwr-mispronunciations}}
+
+We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts so as to base our generalizations on more data.
+There were no cases of insertion, and deletions were very rare: there were only 12 instances of deleted vowels (\textasciitilde{}0.28\% of all vowel targets), and 6 instances of deleted consonants (\textasciitilde{}0.19\% of all consonant targets).
+We therefore focus our qualitative description here on substitutions: There were 820 cases of substitutions, \textasciitilde{}16.95 of the 4839 phones found collapsing across all children and target words, so that substitutions constituted the frank majority of incorrect phones (\textasciitilde{}97.74 of unmatched phones).
+To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
 
@@
 \midrule
-Nasal Target & 101 & 0 & 39 & 17 & 64.3 & 0.0 & 24.8 & 10.8\\
-Oral Target & 1988 & 17 & 52 & 204 & 87.9 & 0.8 & 2.3 & 9.0\\
+Nasal Target & 100 & 0 & 39 & 17 & 64.1 & 0.0 & 25.0 & 10.9\\
+Oral Target & 1992 & 12 & 52 & 205 & 88.1 & 0.5 & 2.3 & 9.1\\
 \bottomrule
@@
 We looked for potential asymmetries in errors for different types of sounds in vowels by looking at the proportion of vowel phones that were correctly repeated or not, generating separate estimates for nasal and oral vowels.
-The nasal vowels in our stimuli occur in \textasciitilde1.40\% of languages' phonologies (range 0\% to 3\%);
-whereas oral vowels in our stimuli occur in \textasciitilde31.55\% of languages' phonologies (range 3\% to 92\%).
-As noted above, frequency within the language is correlated with cross-linguistic frequency, and thus these two types of sounds also differ in the former: Their frequencies in Yélî Dnye are: nasal vowels \textasciitilde0.03‰ (range 0.00‰ to 0.05‰) versus oral \textasciitilde0.23‰ (range 0.02‰ to 0.76‰).
+The nasal vowels in our stimuli occur in \textasciitilde{}1.40\% of languages' phonologies (range 0\% to 3\%);
+whereas oral vowels in our stimuli occur in \textasciitilde{}31.55\% of languages' phonologies (range 3\% to 92\%).
+As noted above, type frequency within the language is correlated with cross-linguistic frequency, and thus these two types of sounds also differ in the former: Their type frequencies in Yélî Dnye are: nasal vowels \textasciitilde{}0.03‰ (range 0.00‰ to 0.05‰) versus oral \textasciitilde{}0.23‰ (range 0.02‰ to 0.76‰).
 
 We distinguished errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (and were therefore a quality error), shown in Table \ref{tab:tab-v}.
-We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.70 versus 12.10).
+We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.90 versus 11.90).
 Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
@@
 \midrule
-Complex Target & 198 & 0 & 219 & 44 & 43.0 & 0.0 & 47.5 & 9.5\\
-Simple Target & 1482 & 13 & 3 & 117 & 91.8 & 0.8 & 0.2 & 7.2\\
+Complex Target & 257 & 0 & 218 & 48 & 49.1 & 0.0 & 41.7 & 9.2\\
+Simple Target & 1425 & 6 & 2 & 120 & 91.8 & 0.4 & 0.1 & 7.7\\
 \bottomrule
@@
 These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.}
-The complex consonants in our stimuli occur in \textasciitilde17.33\% of languages' phonologies (range 0\% to 78\%);
-whereas simple consonants in our stimuli occur in \textasciitilde67.62\% of languages' phonologies (range 13\% to 96\%).
+The complex consonants in our stimuli occur in \textasciitilde{}17.33\% of languages' phonologies (range 0\% to 78\%);
+whereas simple consonants in our stimuli occur in \textasciitilde{}67.62\% of languages' phonologies (range 13\% to 96\%).
 Again these groups of sounds differ in their frequency within the language.
-Their type frequencies in Yélî Dnye are: complex consonants \textasciitilde0.04‰ (range 0.00‰ to 0.10‰) versus simple consonants \textasciitilde0.32‰ (range 0.06‰ to 0.55‰).
-
-Table \ref{tab:tab-c} showed that errors involving complex consonant targets were more common than those involving simple consonants (57 versus 8.20\%).
+Their type frequencies in Yélî Dnye are: complex consonants \textasciitilde{}0.04‰ (range 0.00‰ to 0.10‰) versus simple consonants \textasciitilde{}0.32‰ (range 0.06‰ to 0.55‰).
+
+Table \ref{tab:tab-c} showed that errors involving complex consonants targets were more common than those involving simple consonants (50.90 versus 8.20\%).
 Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.
@@
 Graphical investigation suggested that in both cases the relationship was monotonic and not linear, so we computed Spearman's rank correlations between the correct repetition score, on the one hand, and the two possible predictors on the other.
-Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency {[}r(346.78)=0.74, p \textless{} 0.001{]} was greater than that with within-language frequency {[}r(817.23)=0.39, p = 0.09{]}.
-
-\hypertarget{length-effects-on-nwr-1}{%
-\subsubsection{Length effects on NWR}\label{length-effects-on-nwr-1}}
+Although we cannot directly test the interaction due to collinearity, the correlation with cross-linguistic frequency {[}r(319.72)=0.76, p \textless{} 0.001{]} was greater than that with within-language frequency {[}r(731.10)=0.45, p = 0.05{]}.
+
+\hypertarget{nwr-scores-as-a-function-of-item-length}{%
+\subsubsection{NWR scores as a function of item length}\label{nwr-scores-as-a-function-of-item-length}}
 
@@
 
-\hypertarget{individual-variation-and-nwr}{%
-\subsubsection{Individual variation and NWR}\label{individual-variation-and-nwr}}
+\hypertarget{factors-structuring-individual-variation}{%
+\subsubsection{Factors structuring individual variation}\label{factors-structuring-individual-variation}}
 
@@
 We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments).
-We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible correlation between typological phone frequency and NWR scores.
+We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores.
 An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the Discussion below).
@@
 
-\hypertarget{nwr-and-typology-1}{%
-\subsubsection{NWR and typology}\label{nwr-and-typology-1}}
+\hypertarget{associations-between-nwr-and-cross-linguistic-frequency}{%
+\subsubsection{Associations between NWR and cross-linguistic frequency}\label{associations-between-nwr-and-cross-linguistic-frequency}}
 
 Arguably the most innovative aspect of our data relate to the inclusion of phones that are less commonly found across languages, and rarely used in NWR tasks.
-As explained in the Introduction, typological frequency of phones could reflect ease of perception, ease of production, and other factors, and these factors could affect speech processing and production.
-This predicts a correlation between typological frequency and NWR performance, due to those factors affecting both.
-To assess this prediction, we looked at our data in two ways.
-First, we measured the degree of association between NWR scores and cross-linguistic frequency at the level of non-word items.
-Second, we described mispronunciation patterns, by looking at correct and incorrect repetitions of simpler and more complex sounds, which are also more or less frequent.
-
-There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phoneme inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
-One could then predict that correlations with typological frequency should be relatively weak because the ambient language puts more pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences than what is required of child speakers of other languages.
-On the other hand, it is also possible that this pressure gives Yélî children no benefit, and that some of these categories are simply acquired later in development.
-We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not make adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) till age 5 or 6 (Rumsey, 2017)
-
-We do not have the necessary data to assess whether the correlation is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
+Our monosyllabic items included typologically rare segments so that we could test whether lower average segmental frequency is associated with lower NWR scores.
+It would stand to reason that typologically common sounds are associated with higher performance, but to our knowledge this has not yet been tested with NWR.
+There are some reasons to believe that Yélî Dnye put that hypothesis to a critical test: The phonemic inventory is both large and acoustically packed, in addition to containing several typologically infrequent (or unique) contrasts.
+One could then predict that this effect should be relatively weak because the ambient language puts pressure on Yélî children to distinguish (perceptually and articulatorily) fine-grained phonetic differences in order to successfully communicate with others.
+
+And yet, we found a robust effect of average segmental cross-linguistic frequency on NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often.
 This effect was large, with a magnitude more than twice the size of the effect of participant age.
-Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at younger ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for younger than older children).
-Importantly, the correlation between performance and typological frequency remained significant after accounting for the frequencies of these segments in a conversational corpus.
-An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.
+Moreover, this significant effect remained even once also accounting for the frequencies of these segments in Yélî Dnye children's input.
+An analysis of the substitutions made by children also aligned with this interpretation, with more common sounds being substituted for less common ones.
 
 We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development.
-We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (via phonotactic constraints, ambiguous parsing conditions, individual differences, and more as in, e.g., Beddor, 2009; Bermúdez-Otero, 2015; Maddieson, 2009; Ohala, 1981; Yu, 2021), but we do point out that this association is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
-We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop (as in e.g., Edwards, Beckman, \& Munson, 2004).
-
-\hypertarget{length-effects-and-nwr}{%
-\subsubsection{Length effects and NWR}\label{length-effects-and-nwr}}
-
-We investigated the effect of item complexity on NWR scores by varying the number of syllables in the item.
+We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production (most often discussed in the case of phonotactic constraints on sequences, e.g., Maddieson, 2009), but we do point out that this effect is interestingly different from effects found in artificial language learning tasks (see Moreton \& Pater, 2012 for a review) which are in some ways quite similar to NWR.
+We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the fine-grained phonetic representations in the perception-production loop (as in e.g., Edwards, Beckman, \& Munson, 2004).
+
+\hypertarget{item-length}{%
+\subsubsection{Item Length}\label{item-length}}
+
+We investigated the effect of item complexity on NWR scores by varying both the number of syllables in the item.
 In broad terms, children should have higher NWR scores for shorter items.
-That said, previous work summarized in the Introduction has shown both very small (e.g., Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019) and very large (e.g., Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020) effects of stimulus length.
+That said, previous work summarized in the Introduction has shown both very small (e.g., Piazzalunga et al., 2019) and very large (e.g., Cristia et al., 2020) effects of stimulus length.
 Setting aside our monosyllabic stimuli (which contained typologically infrequent segments with lower NWR scores, as just discussed), we examined effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
-The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant.
+The effect of item length was not significant in a statistical model that additionally accounted for age and random effects of item and participant, and is small and inconsistent across ages (see Figure \ref{fig:fig-prevlit}).
 We do not have a good explanation for why samples in the literature vary so much in terms of the size of length effects, but two possibilities are that this is not truly a length effect but a confound with some other aspect of the stimuli, or that there is variation in phonological representations that is poorly understood.
@@
 For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants, which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of sub-sequences in the non-words.
-Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower.
+Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (elatedly) their n-phone frequency is overall lower.
 If this idea is correct, a careful analysis of non-words used in previous work may reveal that studies with larger length effects just happened to have longer non-words with lower n-phone frequencies.
@@
 Thus, we hazard the hypothesis that sample differences in length effects may relate to such non-essential skills.
-Since as stated this hypothesis is under-specified, further conceptual and empirical work is needed.
-
-\hypertarget{individual-variation-and-nwr-1}{%
-\subsubsection{Individual variation and NWR}\label{individual-variation-and-nwr-1}}
+Since as stated this hypothesis is under-specified, further both conceptual and empirical work are needed.
+
+\hypertarget{individual-differences}{%
+\subsubsection{Individual differences}\label{individual-differences}}
 
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several previous studies (English Vance, Stackhouse, \& Wells, 2005; Italian Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019; Cantonese Stokes, Wong, Fletcher, \& Leonard, 2006; but not in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \ref{fig:fig-prevlit} (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese Stokes et al., 2006; but note Cristia et al., 2020 is an exception).
 Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work varies with respect to correlations of NWR scores with maternal education (e.g., Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017).
-We did not expect large correlations with maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
-In fact, maternal education correlated with NWR score at about r\textasciitilde.1, which is small.
-We find correlations of about that size for participant sex, which is aligned with previous work (Chiat \& Roy, 2007).
-
-Finally, we investigated whether birth order might correlate with NWR scores, as it does with other language tasks, such that first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
+In contrast, previous work shows little evidence for effects of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) on NWR scores.
+We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
+In fact, maternal education correlated with NWR score at about r\textasciitilde{}.1, which is small.
+Similarly, NWR scores may not vary greatly with participant gender according to previous work (Chiat \& Roy, 2007), and for that as well we find effects of about that size.
+
+Last but not least, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
-Therefore we expected that any correlations with birth order on NWR would be attenuated in this context.
+Therefore we expected that any effects of birth order on NWR would be attenuated in this context.
 In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score.
-However, the effect size was larger than that found for the other two factors and it is far from negligible, at r\textasciitilde.2 or Cohen's d\textasciitilde0.41.
-In fact, two large studies with therefore precise estimates found effects of about d\textasciitilde.2 (Barclay, 2015; Havron et al., 2019), which would suggest the correlations we found are larger.
-We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed does not result more even language outcomes for first- and later-born children.
+However, the effect size was larger than that found for the other two factors and it is far from negligible, at r\textasciitilde{}.2 or Cohen's d\textasciitilde{}0.41.
+In fact, two large studies with therefore precise estimates found effects of about d\textasciitilde{}.2 (Barclay, 2015; Havron et al., 2019), which would suggest the effects we found are larger.
+We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
 
@@
 
-The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, since when designing this study we wondered whether NWR was a \emph{culture-fair} test of phonological development.
+The fourth research area to which we wanted to contribute pertained to the use of NWR across languages and populations, as when designing this study we wondered whether NWR was a \emph{fair} test of phonological development.
 Although our data cannot answer this question because we have only sampled one language and population here, we would like to spend some time discussing the integration of these results to the wider NWR literature.
 It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
-Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, similarly to other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages (Frank, Braginsky, Yurovsky, \& Marchman, 2017).
-
-The range of performance we observed overlapped with previously observed levels of performance.
+Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, as it is done for other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages (Frank, Braginsky, Yurovsky, \& Marchman, 2017).
+
+At first sight, the range of performance we observed overlapped with previously observed levels of performance.
 Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
-Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
-
-Comparison across published studies is difficult (see SM2 for our preliminary attempt).
-To be certain whether language-specific characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
-We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless, but phonotactically legal, in all of the languages.
-An alternative may be to find ways to regress out some of these differences, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
-Both of these issues are discussed in Chiat (2015).
-As for the variable strengths of age correlations discussed above, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
-
-\hypertarget{limitations}{%
-\subsubsection{Limitations}\label{limitations}}
-
-Before closing, we would like to point out some salient limitations of the current work.
-To begin with, we only employed one set of non-words, in which not all characteristics that previous work suggest matter were manipulated (Chiat, 2015).
-As a result, we only have a rather whole-sale measure of performance, and we do not know to what extent lexical knowledge, pure phonological knowledge, and working memory, among others, contribute to children's performance.
-Similarly, our items varied systematically in length and typological frequency of the sounds included, but not in other potential dimensions (such as whether the items contained morphemes of the language or not).
-
-We relied on a single resource, PHOIBLE, for our estimation of typological frequency, and some readers may be worried about the effects of this choice.
-As far as we know, PHOIBLE is the most extensive archive of phonological inventories, so it is a reasonable choice in the current context.
-However, one may want to calculate typological frequency not by trying to have as many languages represented as possible, but rather by selecting a sample of typologically independent languages.
-In addition, it is not the case that all the world's languages are represented, and indeed some of the Yélî sounds were not found in PHOIBLE.
-PHOIBLE---as well as our own work---depends on phonological descriptions from linguists who are in many cases not native speakers of the languages.
-Because the phones in our items have largely been evidenced as phonemic via multiple analyses (i.e., minimal contrast, phonological, phonetic, and ultrasound, see Levinson, 2021), we are not concerned that changes to the phonological description in the future (e.g., if a segment loses its phonemic status) will significantly change the results presented here.
-Relatedly, any converging evidence from the other ongoing studies of Yélî Dnye phonological development and fine-grained analyses of sound substitutions would certainly help bolster the claims we made here.
-While all these limitations should be borne in mind, it is important to also consider what our conclusions were, and that is that there is a non-trivial correlation between NWR and typological frequency.
-At present, we do not see how imbalance in the typological selection and missing data can conspire to produce the correlation we observe.
-If anything, these factors should increase noise in the typological frequency estimation, in which case the correlation size we uncover is an underestimation of the true correlation.
-
-Additionally, we only had a single person interacting with children as well as interpreting children's production, so we do not know to what extent our findings generalize to other experimenters and research assistants.
-Furthermore, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
-Although we know from other work that adults' perceptual performance on these types of sounds is well above chance from audio-only presentation {[}REF{]}, language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.
+Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane'.
+
+To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring.
+We entered data from 14 studies (including ours), presenting data from 12 languages.
+Specifically, Arabic was represented by Jaber-Awida (2018); Cantonese by Stokes et al.
+(2006); English by Vance et al.
+(2005); Italian by Piazzalunga et al.
+(2019); Mandarin by Lei et al.
+(2011); Persian by Farmani et al.
+(2018); Slovak by Kapalková, Polišenská, and Vicenová (2013) and Polišenská and Kapalková (2014); Sotho by Wilsenach (2013); Spanish by Balladares et al.
+(2016); Swedish by Kalnak et al.
+(2014) and Radeborg, Barthelom, SjöBerg, and Sahlén (2006); Tsimane' by Cristia et al.
+(2020); and Yélî Dnye from the present study.
+Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability.
+Results separating different age groups are shown in Figure \ref{fig:fig-prevlit-overall}.
+
+\begin{figure}
+\centering
+\includegraphics{manuscript_files/figure-latex/fig-prevlit-overall-1.pdf}
+\caption{\label{fig:fig-prevlit-overall}NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages.
+The legend indicates language and the length of non-words (in syllables).
+Central tendency is mean; error is one standard error.}
+\end{figure}
+
+Several observations can be drawn from this figure.
+To begin with, we focus on the comparison between Yélî Dnye and Tsimane'.
+These two groups have been described as having roughly similar levels of child-directed speech, yet they exhibit very different results: Tsimane' shows lower overall NWR scores (and according to Figure \ref{fig:fig-prevlit}, larger length effects).
+This suggests that the lower NWR scores found among the Tsimane' are due to long-term effects of lower levels of child-directed speech.
+Naturally, there is an alternative interpretation, namely that input estimation suggesting very slightly higher levels of child-directed speech among the Tsimane' than among Yélî Dnye learners is inaccurate.
+In fact, careful reading of previous reports highlight important methodological differences in how input quantity has been estimated across papers: Casillas et al.
+(2020) hand-coded speech with the help of a native research assistant, and then summed all child-directed speech, which effectively establishes an upper boundary of the speech children could potentially process.
+Cristia, Dupoux, Gurven, and Stieglitz (2019) estimated quantities from behavioral observations on the frequency of child-directed one-on-one conversation, which is probably closer to a lower boundary.
+Finally, Scaff et al.
+(2021) used human annotation for detecting speech but an automated temporal method for assigning speech as child-directed or not, in a way that could lead to over-estimation (because any speech by e.g.~a female adult that was not temporally close to speech by others would count as child-directed).
+A final answer to the question of how much child-directed speech is afforded to Yélî and Tsimane' children must await fully comparable methods.
+
+That said, Cristia et al.
+(2020) also pointed out another characteristic of the Tsimane' population, and this was the relatively low prevalence of literacy, and generally the variable access to formal education.
+This is a very different case from the Yélî population studied here, where nearly all adults have accumulated several years of schooling, and basic literacy in English (and sometimes Yélî Dnye) is widespread.
+If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks.
+Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in Figure \ref{fig:fig-prevlit-overall}.
+These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English.
+What is at stake in our proposed alternative interpretation of the lower scores observed among the Tsimane' is related to literacy \emph{in the broader population} (rather than in the tested children themselves).
+
+Although exciting, this hypothesis is only one of many.
+Another plausible explanation is that the Tsimane' results are not comparable to the previous body of literature, and specifically to our study.
+Cristia et al.
+(2020) administered the NWR in the form of a group game played outside, with a non-native experimenter providing the target, and each person of the group attempting it in their stead.
+This immediately means a number of important methodological differences with the standard implementation of NWR, where children are tested individually, they hear items spoken by a native speaker (often over headphones), the experimenter tends to belong to the same community as the children, and testing occurs in quiet conditions (with little background noise).
+Thus, a priority is for additional data gathered using this more novel testing paradigm in other populations, or from the Tsimane' using the more traditional paradigm.
+
+Broadening our discussion to all of the studies in our literature review, we notice that there is rather wide variation of the range of NWR scores found across these samples, and that, in fact, the strength of age effects also varies.
+We performed some exploratory analyses to see whether features of the languages children were learning could be related to their overall NWR scores.
+We extracted the number of phonemes in the language from PHOIBLE and coded whether words in the language tended to be longer or shorter based on information in the papers or other sources.
+Neither of these two predictors explained variance in Figure \ref{fig:fig-prevlit-overall}.
+It is possible that average word length plays a role, but often researchers incorporate this into their design by including longer items when the native language allows this, with e.g.~Sotho non-words having 4-7 syllables in length.
+To be more certain whether language characteristics do account for meaningful variation in NWR scores, it will be necessary to design NWR tasks that are cross-linguistically valid.
+We believe this will be exceedingly difficult (or perhaps impossible), since it would entail defining a 10-20 set of items that are meaningless in all of the languages as well as phonotactically legal.
+An alternative may be to find ways to regress out some of these effects, and thus compare languages while controlling for choices of phonemes, syllable structure, and overall length of the NWR items.
+As for different strengths of age effects, here as well we are uncertain to what they may be due, but we do hope that these intriguing observations will lead others to collect and share NWR data.
 
@@
 
-The present study shows that NWR be adapted for very different populations than have previously been tested.
-In addition, we observed strong correlations with age and typological frequency, while correlations with item length, participant sex, maternal education, and birth order were weaker.
-A consideration of previous work led us to suggest that the statistical strength of all of these effects may vary depending on the linguistic, cultural, and socio-demographic properties of the population under study, in conjunction with characteristics of the non-word items used.
-The present findings raise many questions, including: Why do NWR scores pattern differently across samples?
+While NWR can, in theory, be used to test a variety of questions about phonological development in any language, previous work has been primarily limited to a handful of related languages spoken in urban, industrialized contexts.
+The present study shows that, not only can NWR be adapted for very different populations than have previously been tested, but that effects of age and typological frequency may strongly influence phonological development across these diverse settings, while effects of item length, participant gender, maternal education, and birth order, may either have little impact on this facet of language development or have an impact that varies depending on the linguistic, cultural, and socio-demographic properties of the population under study.
+Because these latter predictors strongly relate to other language outcomes, the present findings raise many questions, including: Why do NWR scores would pattern differently across samples?
 What does that tell us about the relationship between lexical development, phonological development, and the input environment?
 What is implied about the joint applicability of these outcome measures as a diagnostic indicator for language delays and disorders?
-While answers to these questions should be sought in future work, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and much broader patterns of cross-linguistic phone frequency.
+While answers to these questions are sought, we take the present findings as robustly supporting the idea that phonological development continues well past early childhood and as yielding preliminary support for a potential association between individual learners' NWR and cross-linguistic phone frequency.
 
@@
 
-\hypertarget{data-code-and-materials-availability-statement}{%
-\subsection{Data, code and materials availability statement}\label{data-code-and-materials-availability-statement}}
-
-All data, code, and materials are available from \url{https://osf.io/5qspb/}
-
 \hypertarget{references}{%
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-armon2015assessing}{}%
-Armon-Lotem, S., Jong, J. de, \& Meir, N. (2015). \emph{Methods for assessing multilingual children: Disentangling bilingualism from specific language impairment}. Bristol: Multilingual matters.
-
 \leavevmode\hypertarget{ref-balladares2016socio}{}%
@@
 (2016).
-{Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers}.
-\emph{{First Language}}, \emph{36}(3), 338--351.
+Socio-economic status affects sentence repetition, but not non-word repetition, in Chilean preschoolers.
+\emph{First Language}, \emph{36}(3), 338--351.
 \url{https://doi.org/10.1177/0142723715626067}
@@
 Barclay, K. J. (2015). A within-family analysis of birth order and intelligence using population conscription data on swedish men. \emph{Intelligence}, \emph{49}, 134--143.
-
-\leavevmode\hypertarget{ref-beddor2009coarticulatory}{}%
-Beddor, P. S. (2009). A coarticulatory path to sound change. \emph{Language}, \emph{85}(4), 785--832.
-
-\leavevmode\hypertarget{ref-bermudezotero2015amphichronic}{}%
-Bermúdez-Otero, R.
-(2015).
-Amphichronic explanation and the life cycle of phonological processes.
-In P.
-Honeybone \& J.
-Salmons (Eds.), \emph{The {O}xford handbook of historical phonology} (pp.
-374--399).
-Oxford, UK: Oxford University Press.
 
@@
 Duranti, E.
-Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of {Language Socialization}} (pp.
+Ochs, \& and Bambi B Schieffelin (Eds.), \emph{Handbook of Language Socialization} (pp.
 29--55).
@@
 (2014).
-The interactional context of language learning in {T}zeltal.
+The interactional context of language learning in Tzeltal.
 In I.
@@
 Kurumada, \& B.
-Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of {Eve V.
-Clark}} (pp.
-51-\/-82).
+Estigarribia (Eds.), \emph{Language in interaction: Studies in honor of Eve V.
+Clark} (pp.
+51--82).
 Amsterdam, NL: John Benjamins.
@@
 Brown, P., \& Casillas, M.
-(in press).
-Childrearing through social interaction on {Rossel Island, PNG}.
+(n.d.).
+Childrearing through social interaction on Rossel Island, PNG.
 In A.
@@
 Fentiman \& M.
-Goody (Eds.), \emph{{Esther Goody} revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
+Goody (Eds.), \emph{Esther Goody revisited: Exploring the legacy of an original inter-disciplinarian} (pp.
 XX--XX).
@@
 
-\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
-Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural examination of young children's everyday language experiences.
-
-\leavevmode\hypertarget{ref-casillas2021early}{}%
-Casillas, M., Brown, P., \& Levinson, S. C. (2021). {Early language experience in a Papuan community}. \emph{Journal of Child Language}, \emph{48}(4), 792--814.
+\leavevmode\hypertarget{ref-casillas2020early}{}%
+Casillas, M., Brown, P., \& Levinson, S. C. (2020). Early language experience in a Papuan community. \emph{Journal of Child Language}, \emph{XX}, XX--XX.
 
@@
 (1998).
-{The illiterate brain.
-Learning to read and write during childhood influences the functional organization of the adult brain.} \emph{Brain: A Journal of Neurology}, \emph{121}(6), 1053--1063.
+The illiterate brain.
+Learning to read and write during childhood influences the functional organization of the adult brain.
+\emph{Brain: A Journal of Neurology}, \emph{121}(6), 1053--1063.
 \url{https://doi.org/10.1093/brain/121.6.1053}
-
-\leavevmode\hypertarget{ref-chiat2015nonword}{}%
-Chiat, S.
-(2015).
-Non-word repetition.
-In S.
-Armon-Lotem, J.
-de Jong, \& N.
-Meir (Eds.), \emph{Methods for assessing multilingual children: Disentangling bilingualism from specific language impairment} (pp.
-125--150).
-Bristol: Multilingual matters.
 
@@
 
-\leavevmode\hypertarget{ref-coady2008uses}{}%
-Coady, J.
-A., \& Evans, J.
-L.
-(2008).
-Uses and interpretations of non-word repetition tasks in children with and without specific language impairments (SLI).
-\emph{International Journal of Language \& Communication Disorders}, \emph{43}(1), 1--40.
-
 \leavevmode\hypertarget{ref-is08042009language}{}%
@@
 
-\leavevmode\hypertarget{ref-cristia2021supplementary}{}%
-Cristia, A., \& Casillas, M. (2021). Supplementary materials to "non-word repetition in children learning yélî dnye". Retrieved from \url{https://osf.io/5qspb/wiki/home/}
+\leavevmode\hypertarget{ref-cristia2019child}{}%
+Cristia, A., Dupoux, E., Gurven, M., \& Stieglitz, J.
+(2019).
+Child-directed speech is infrequent in a forager-farmer population.
+\emph{Child Development}, \emph{90}(3), 759--773.
+\url{https://doi.org/10.1111/cdev.12974}
 
@@
 (2020).
-Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the {T}simane'.
+Infant-directed input and literacy effects on phonological processing: Non-word repetition scores among the Tsimane'.
 \emph{PLoS ONE}, \emph{15}(9), e0237702.
-https://doi.org/\url{https://doi.org/10.1371/journal.pone.0237702}
+\url{https://doi.org/https://doi.org/10.1371/journal.pone.0237702}
 
@@
 Differences in the nonword repetition performance of children with and without specific language impairment: A meta-analysis.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}, 177--195.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{50}(1), 177--195.
 
@@
 
-\leavevmode\hypertarget{ref-farabolini2021association}{}%
-Farabolini, G., Taboh, A., Ceravolo, M.
-G., \& Guerra, F.
-(2021).
-The association between language exposure and non-word repetition performance in bilingual children: A meta-analysis.
-\emph{Under Review}.
-
 \leavevmode\hypertarget{ref-farmani2018normalization}{}%
@@
 (2018).
-Normalization of the non-word repetition test in {F}arsi-speaking children.
+Normalization of the non-word repetition test in Farsi-speaking children.
 \emph{Journal of Modern Rehabilitation}, \emph{12}(4), 217--224.
@@
 \leavevmode\hypertarget{ref-foley1986papuan}{}%
-Foley, W. A. (1986). \emph{{The Papuan languages of New Guinea}}. Cambridge, UK: Cambridge University Press.
+Foley, W. A. (1986). \emph{The Papuan languages of New Guinea}. Cambridge, UK: Cambridge University Press.
 
@@
 \leavevmode\hypertarget{ref-gallagher2014identity}{}%
-Gallagher, G. (2014). {An identity bias in phonotactics: Evidence from Cochabamba Quechua}. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
+Gallagher, G. (2014). An identity bias in phonotactics: Evidence from Cochabamba Quechua. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
+
+\leavevmode\hypertarget{ref-gallon2007non}{}%
+Gallon, N., Harris, J., \& Van der Lely, H.
+(2007).
+Non-word repetition: An investigation of phonological complexity in children with Grammatical SLI.
+\emph{Clinical Linguistics \& Phonetics}, \emph{21}(6), 435--455.
 
@@
 
-\leavevmode\hypertarget{ref-hellwigPAlanguage}{}%
-Hellwig, B., Sarvasy, H., \& Casillas, M.
-(provisionally accepted).
-Language acquisition.
-In N.
-Evans \& S.
-Fedden (Eds.), \emph{The {O}xford guide to {P}apuan languages} (pp.
-XX--XX).
-Oxford: Oxford University Press.
-
 \leavevmode\hypertarget{ref-jabere2018xperiment}{}%
-Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual {A}rabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}, 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
+Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual Arabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}(4), 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
 
@@
 (2014).
-Nonword repetition--a clinical marker for specific language impairment in {S}wedish associated with parents' language-related problems.
+Nonword repetition--a clinical marker for specific language impairment in Swedish associated with parents' language-related problems.
 \emph{PloS One}, \emph{9}(2), e89544.
+
+\leavevmode\hypertarget{ref-kapalkova2013non}{}%
+Kapalková, S., Polišenská, K., \& Vicenová, Z.
+(2013).
+Non-word repetition performance in Slovak-speaking children with and without SLI: novel scoring methods.
+\emph{International Journal of Language and Communication Disorders}, \emph{48}(1), 78--89.
+\url{https://doi.org/10.1111/j.1460-6984.2012.00189.x}
 
@@
 
-\leavevmode\hypertarget{ref-lehmann2018early}{}%
-Lehmann, J.-Y.
-K., Nuevo-Chiquero, A., \& Vidal-Fernandez, M.
-(2018).
-The early origins of birth order differences in children's outcomes and parental behavior.
-\emph{Journal of Human Resources}, \emph{53}(1), 123--156.
+\leavevmode\hypertarget{ref-lei2011developmental}{}%
+Lei, L., Pan, J., Liu, H., McBride-Chang, C., Li, H., Zhang, Y., \ldots{} others.
+(2011).
+Developmental trajectories of reading development and impairment from ages 3 to 8 years in chinese children.
+\emph{Journal of Child Psychology and Psychiatry}, \emph{52}(2), 212--220.
 
 \leavevmode\hypertarget{ref-levinsonYDgrammar}{}%
-Levinson, S. C. (2021). \emph{A grammar of {Yélî D}nye, the {P}apuan language of {R}ossel {I}sland}. Berlin, Boston: De Gruyter Mouton.
+Levinson, S. C. (2020). \emph{A grammar of Yélî Dnye, the Papuan language of Rossel Island}. Berlin, Boston: De Gruyter Mouton.
 
@@
 \leavevmode\hypertarget{ref-maddieson2009phonology}{}%
-Maddieson, I. (2009). Phonology, naturalness and universals. \emph{Pozn{á}n Studies in Contemporary Linguistics}, \emph{45}(1), 131--140.
+Maddieson, I. (2009). Phonology, naturalness and universals. \emph{Poznán Studies in Contemporary Linguistics}, \emph{45}(1), 131--140.
 
@@
 \leavevmode\hypertarget{ref-maddiesonIPphoneticsYD}{}%
-Maddieson, I., \& Levinson, S. C. (in preparation). \emph{The phonetics of {Yélî Dnye}, the language of {Rossel Island}}.
+Maddieson, I., \& Levinson, S. C. (n.d.). \emph{The phonetics of Yélî Dnye, the language of Rossel Island}.
 
@@
 (2017).
-Independent and combined effects of socioeconomic status ({SES}) and bilingualism on children's vocabulary and verbal short-term memory.
+Independent and combined effects of socioeconomic status (SES) and bilingualism on children's vocabulary and verbal short-term memory.
 \emph{Frontiers in Psychology}, \emph{8}, 1442.
@@
 (2016).
-{Disentangling SLI and bilingualism using sentence repetition tasks: The impact of L1 and L2 properties}.
+Disentangling SLI and bilingualism using sentence repetition tasks: The impact of L1 and L2 properties.
 \emph{International Journal of Bilingualism}, \emph{20}(4), 421--452.
@@
 \leavevmode\hypertarget{ref-moreton2012structure}{}%
-Moreton, E., \& Pater, J. (2012). Structure and substance in artificial-phonology learning, part {II}: substance. \emph{Language and Linguistics Compass}, \emph{6}(11), 702--718.
-
-\leavevmode\hypertarget{ref-ohala1981listener}{}%
-Ohala, J.
-J.
-(1981).
-The listener as a source of sound change.
-In M.
-F.
-Miller, C.
-S.
-Masek, \& R.
-A.
-Hendrick (Eds.), \emph{Papers from the parasession on language and behavior} (pp.
-178--203).
-Chicago, IL: Chicago Linguistics Society.
+Moreton, E., \& Pater, J. (2012). Structure and substance in artificial-phonology learning, part II: Substance. \emph{Language and Linguistics Compass}, \emph{6}(11), 702--718.
+
+\leavevmode\hypertarget{ref-mulder2017early}{}%
+Mulder, H., Verhagen, J., Van der Ven, S.
+H., Slot, P.
+L., \& Leseman, P.
+P.
+(2017).
+Early executive function at age two predicts emergent mathematics and literacy at age five.
+\emph{Frontiers in Psychology}, \emph{8}, 1706.
 
 \leavevmode\hypertarget{ref-peuteIPconsonants}{}%
-Peute, A. A. K., Fikkert, P., \& Casillas, M. (In preparation). Early consonant production in {Y}élî {D}nye and {T}seltal.
+Peute, A. A. K., Fikkert, P., \& Casillas, M. (n.d.). Early consonant production in Yélî Dnye and Tseltal.
 
@@
 (2019).
-{An articulatory-based disyllabic and trisyllabic Non-Word Repetition test: reliability and validity in Italian 3-to 7-year-old children}.
+An articulatory-based disyllabic and trisyllabic Non-Word Repetition test: reliability and validity in Italian 3-to 7-year-old children.
 \emph{Clinical Linguistics \& Phonetics}, \emph{33}(5), 437--456.
 
-\leavevmode\hypertarget{ref-rumsey2017dependency}{}%
-Rumsey, A.
-(2017).
-Dependency and relative determination in language acquisition: The case of {Ku Waru}.
-In N.
-J.
-Enfield (Ed.), \emph{Dependencies in language} (pp.
-97--116).
-Berlin: Language Science Press.
-
-\leavevmode\hypertarget{ref-dos2020epreuve}{}%
-Santos, C.
-dos, Frau, S., Labrevoit, S., \& Zebib, R.
-(2020).
-L'{é}preuve de r{é}p{é}tition de non-mots LITMUS-NWR-FR {é}value-t-elle la phonologie?
-In \emph{SHS web of conferences} (Vol.
-78, p.
-10005).
-EDP Sciences.
-
-\leavevmode\hypertarget{ref-scaff2019}{}%
-Scaff, C. (2019). \emph{Beyond {WEIRD}: An interdisciplinary approach to language acquisition} (PhD thesis).
+\leavevmode\hypertarget{ref-polivsenska2014improving}{}%
+Polišenská, K., \& Kapalková, S. (2014). Improving child compliance on a computer-administered nonword repetition task. \emph{Journal of Speech, Language and Hearing Research}, \emph{57}(3).
+
+\leavevmode\hypertarget{ref-radeborg2006swedish}{}%
+Radeborg, K., Barthelom, E., SjöBerg, M., \& Sahlén, B.
+(2006).
+A Swedish non-word repetition test for preschool children.
+\emph{Scandinavian Journal of Psychology}, \emph{47}(3), 187--192.
+\url{https://doi.org/10.1111/j.1467-9450.2006.00506.x}
 
@@
 Nonword repetition and sentence repetition as clinical markers of specific language impairment: The case of cantonese.
-\emph{Journal of Speech, Language, and Hearing Research}, \emph{49}, 219--236.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{49}(2), 219--236.
 
@@
 
-\leavevmode\hypertarget{ref-tuller2018identifying}{}%
-Tuller, L., Hamann, C., Chilla, S., Ferré, S., Morin, E., Prevost, P., \ldots{} Zebib, R.
-(2018).
-Identifying language impairment in bilingual children in france and in germany.
-\emph{International Journal of Language \& Communication Disorders}, \emph{53}(4), 888--904.
-
 \leavevmode\hypertarget{ref-vance2005speech}{}%
@@
 (2013).
-{Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners}.
-\emph{{Per Linguam: A Journal of Language Learning= Per Linguam: Tydskrif Vir Taalaanleer}}, \emph{29}(2), 17--32.
+Phonological skills as predictor of reading success: An investigation of emergent bilingual Northern Sotho/English learners.
+\emph{Per Linguam: a Journal of Language Learning= Per Linguam: Tydskrif vir Taalaanleer}, \emph{29}(2), 17--32.
 \url{https://doi.org/10.5785/29-2-554}
 
-\leavevmode\hypertarget{ref-yu2021toward}{}%
-Yu, A. C. L. (2021). Toward an individual-difference perspective on phonologization. \emph{Glossa: A Journal of General Linguistics}, \emph{6}(1), 1--24.
-
-\end{CSLReferences}
-
 \end{document}

### 2021-11-10 manuscript.Rmd
trying to generate track changes version

@@
 Additionally, we only had a single person interacting with children as well as interpreting children's production, so we do not know to what extent our findings generalize to other experimenters and research assistants.
-Furthermore, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
-Although we know from other work that adults' perceptual performance on these types of sounds is well above chance from audio-only presentation [REF],  language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.
+Furthermore, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual production cues in their interpretation.
+Other work shows that children's performance reaches ceiling by 12 years of age for auditorily-presented minimal pairs for typologically rare (i.e., pre- vs post-alveolar stop) contrasts [@casillasIPmarkedness].
+Nonetheless,  language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.
 

### 2021-11-10 manuscript.Rmd
for proofing

@@
 NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
-In this study, we report on NWR results among children (N = 40, aged 3--10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
+In this study, we report on NWR results among children (N = 40, aged 3--10 years) learning Yélî Dnye, an isolate language spoken on Rossel Island in Papua New Guinea.
 Results make three contributions that are specific, and a fourth that is general.
-First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
+First, we found that non-word items containing typologically frequent sounds are repeated without changes more often than non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
 Second, we documented rather weak effects of item length.
-Third, we found that age has a strong correlation with NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
+Third, we found that NWR scores correlate strongly with age, whereas they are only weakly correlated with child sex, maternal education, and birth order.
 Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.
@@
 knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = FALSE, fig.pos = "T")
+#install.packages("devtools")
+#devtools::install_github("crsh/papaja")
+
 require("papaja")
@@
 Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development; @rumsey2017dependency].
-Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
+Some of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
@@
 
-### What is NWR?
+### Defining NWR
 
@@
 Phonologists, phoneticians, and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of processing and acquisition via diachronic language change.
-These works focus largely on phonotactics [@moreton2012structure] perceptual parsing of the (ambiguous) linguistic signal [@beddor2009coarticulatory; @ohala1981listener], and individual differences in processing styles [@bermudezotero2015amphichronic]; small but significant effects that may cumulatively drive language change via phonologization [see @yu2021toward for a recent review].
+These works focus largely on phonotactics [@moreton2012structure], perceptual parsing of the (ambiguous) linguistic signal [@beddor2009coarticulatory; @ohala1981listener], and individual differences in processing styles [@bermudezotero2015amphichronic]; which are small effects that may nonetheless cumulatively drive language change via phonologization [see @yu2021toward for a recent review].
 Thus, the correlation between typological frequency and ease of acquisition is typically assumed to emerge from one or more of the following causal paths:
@@
 
-Given these causal pathways, we predicted that variation in NWR across items will correlate with the cross-linguistic frequency of the phones composing those items.
+Since NWR can tap into perception, production, and working memory, we predicted that variation in NWR across items will correlate with the cross-linguistic frequency of the phones composing those items.
 
@@
 The second research area we contribute data to is research looking at the impact of word length on NWR repetition within specific languages.
-Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory].
+Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children, @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners, @piazzalunga2019articulatory].
 
@@
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer.
-A non-systematic meta-analysis does not provide overwhelming support for this hypothesis [@cristia2021supplementary; SM1].
-
-Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we did not approach this issue within a hypothesis-testing framework but sought instead to provide additional data  on the question, which may be re-used in future meta- or mega-analytic analyses.
+A non-systematic meta-analysis does not provide overwhelming support for this hypothesis [@cristia2021supplementary, SM1].
+
+Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we did not approach this issue within a hypothesis-testing framework but sought instead to provide additional data  on the question, which may be re-used in future meta- or mega-analyses.
 
@@
 Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword].
-In a nutshell, most evidence is mixed, suggesting that correlations with individual variation may be small, and more data is needed to estimate their true size.
+In a nutshell, most evidence is mixed, suggesting that individual variation effects may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
@@
 Regarding child sex, no significant correlation has been found in previous NWR research [@chiat2007preschool], and in other language tasks evidence is mixed.
-Finally, prior research varies on whether significant differences as a function of  maternal education are reported.
+Finally, prior research using NWR varies on whether significant differences as a function of  maternal education are reported.
 For instance, no significant difference was found some studies [@farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio]; whereas significant differences were reported in others [@tuller2018identifying; @dos2020epreuve].
-In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports [@frank2017wordbank] and word comprehension studies [@scaff2019].
+In other lines of work, maternal education often correlates with child language outcomes, including vocabulary reports [@frank2017wordbank] and word comprehension studies [@scaff2019].
 The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children [see discussion in @cristia2020infant].
@@
 For instance, @tuller2018identifying employed a set of non-words judged to be language independent and two others that were more aligned with either French or German.
-The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language ("language-dependent") than when they are "language-independent".
+The intuition is that NWR will correlate with the relative levels of exposure to that language more strongly when items are aligned with a specific language ("language-dependent") than when they are "language-independent".
 To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience.
@@
 In any case, this line of research focuses on links between exposure to a given language and NWR performance.
-In contrast, when we discuss linguistic or cultural differences here, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).]
-
-Regarding potential language differences, we note that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/).
+In contrast, when we discuss linguistic or cultural differences here, we ask the question of whether children vary in their performance as a function of which language they are learning (e.g., the language's typological properties) and/or their overall, absolute levels of language experience (not relative levels in a multilingual setting).]
+
+Regarding potential language differences, we note that previous studies composed items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/).
 It would be interesting for future researchers to consider straying from the literature by varying other dimensions that are relevant to the language under study.
@@
 In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).], and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with a complete range of nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsonYDgrammar for details).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both [e.g., /pʲ/, /pʷ/, /pʲʷ/, /tpʲ/, /ṇmḍbʲ/, see @levinsonYDgrammar for details].
 The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
 Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to occur as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to occur as short and long nasal vowels as well (/ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
 
@@
 Other ongoing work on Yélî Dnye includes experiment-based infant phoneme discrimination data and errors made in elicited and spontaneous speech from young children, but these data are neither finalized nor yet externally reviewed [see @hellwigPAlanguage for more information].
-These data will help better inform our current analyses based on NWR in the future (e.g., regarding common sound substitutions) but are not critical for testing our current question about the general correlation between cross-linguistic phone frequency and NWR performance.
+These data will help better inform our current analyses based on NWR in the future (e.g., regarding common sound substitutions) but are not critical for addressing our question about the general correlation between cross-linguistic phone frequency and NWR performance.
 
@@
 However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcrosscultural], and additionally shares other societal characteristics wih the Tsimane' [e.g., is rural and relies on farming, children grow up in wide familial networks, etc; @casillas2021early].
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3 years is comparably infrequent in this community [in fact it may be infrequent in many settings, including urban ones @bunceURcrosscultural].
+Our sample also shares other societal characteristics wih the Tsimane' [e.g., the community is rural and relies on farming, children grow up in wide familial networks, @casillas2021early].
 Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2021early].
@@
 
-- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
+- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are cross-linguistically rarer sounds more often substituted by commoner sounds?
 - How do NWR scores change as a function of item length in number of syllables?
-- Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?
+- Is individual variation in NWR scores correlated with child age, sex, birth order, and/or maternal education?
 
@@
 A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible 2- to 4-syllable sequences.
-These candidates were automatically removed from consideration if they appeared in the most recent dictionary [@levinsonYDgrammar].
+Candidates were automatically removed from consideration if they appeared in the most recent dictionary [@levinsonYDgrammar].
 The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
@@
 As before, items were automatically excluded if they appeared in the dictionary.
-Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
-Additionally, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
+Furthermore, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
+We made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e.,  there existed a longer word that included the monosyllable as a sub-sequence).
 These candidates were then presented to one informant, for a final check that they did not mean anything.
@@
 For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory.
-While PHOIBLE is a unprecedented in its scope, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
+While PHOIBLE is unprecedented in its scope, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors.
 Note that nearly half of the phones in PHOIBLE are only attested in one language (Steven Moran, personal communication).
@@
 Because space availability was limited in different ways from hamlet to hamlet, the places where elicitation happened varied across testing sites.
-More information is available from the online supplementary materials.
-
+More information is available from the online  materials (https://osf.io/qt8gr/).
+
+We tested one child at a time.
 We fitted the child with a headset microphone (Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former) that fed into the left channel of a Tascam DR40x digital audio recorder.
@@
 The first phase of the experiment involved making sure the child understood the task.
-We explained the task and then orally presented the first practice item.
+We explained the task and then presented the first practice item.
 At this point, many children did not say anything in response, which triggered the following procedure: First, the assistant insisted the child make a response.
@@
 Children were not asked to make repetitions if they did not produce a first attempt.
-In total, test sessions took approximately six minutes, with the first minute attributed to practice and five minutes to the actual test list.
+In total, test sessions took approximately six minutes (one for practice; five for test list).
 
@@
 
-We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts and not excluding errors resulting in real words, so as to base our generalizations on more data.
-Insertions and deletions were very rare (metathesis was not attested): there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r dim(del_bank)[1] + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the  majority of incorrect phones (~`r round(dim(sub_bank)[1]/(dim(del_bank)[1] + dim(sub_bank)[1] )*100,2)` of unmatched phones).
-To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?
-[^4]
+We addressed our first research question in a second way, by investigating patterns of error.
+Unlike all other analyses, we looked at all attempts, so as to base our generalizations on more data.
+As in all analyses, we did not exclude errors resulting in real words.
+Deletions were very rare (insertion and metathesis were not attested): there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
+We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r length(del_bank) + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the  majority of incorrect phones (~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] )*100,2)` of unmatched phones).
+To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?[^4]
 
@@
 We distinguished errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (and were therefore a quality error), shown in Table \@ref(tab:tab-v).
-We found that errors involving nasal vowel targets were more common than those involving oral vowels (`r round(100-mytab["Nasal Target","% Corr."],1)` versus `r round(100-mytab["Oral Target","% Corr."],1)`).
+We found that errors involving nasal vowel targets were more common than those involving oral vowels (`r 100-mytab["Nasal Target","% Corr."]` versus `r 100-mytab["Oral Target","% Corr."]`%).
 Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question.
-As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, ρ(`r corage$statistic`) = `r printr(corage$estimate)`, p `r pval.display(corage$p.value)`).
-In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(inc$birthOrder))` children, ρ(`r round(corbo$statistic,1)`) = `r printr(corbo$estimate)`, p `r pval.display(corbo$p.value)`), or maternal education  (ρ( `r round(corme$statistic,1)`) = `r printr(corme$estimate)`, p `r pval.display(corme$p.value)`).
+As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance): ρ(`r sum(!is.na(overall$Word)) - 2`) = `r printr(corage$estimate)`, p `r pval.display(corage$p.value)`.
+In contrast, there was no clear association between NWR scores and sex: Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`; NWR scores and birth order  (data missing for `r sum(is.na(inc$birthOrder))` children): ρ(`r sum(!is.na(overall$Word)) - 2 - sum(is.na(inc$birthOrder))`) = `r printr(corbo$estimate)`, p `r pval.display(corbo$p.value)`; or NWR scores and maternal education:  ρ(`r sum(!is.na(overall$Word)) - 2`) = `r printr(corme$estimate)`, p `r pval.display(corme$p.value)`.
 
@@
 We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible correlation between typological phone frequency and NWR scores.
-An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the Discussion below).
+An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been  seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the Discussion below).
 We consider implications of our results on each of these four research areas in turn.
@@
 In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several previous studies [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006nonword; but not in  @cristia2020infant].
-Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
-In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].
+Indeed, age was significantly correlated with NWR scores and it also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
+In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development; @rumsey2017dependency].
 
@@
 
-Finally, we investigated whether birth order might correlate with NWR scores, as it does with other language tasks, such that first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
+Finally, we investigated whether birth order might correlate with NWR scores, as it does with other language tasks, such that first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than first-borns.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
@@
 However, the effect size was larger than that found for the other two factors and it is far from negligible, at r~.2 or Cohen's d~`r r2d(.2)`.
-In fact, two large studies with therefore precise estimates found effects of about d~.2 [@havron2019effect;@barclay2015within], which would suggest the correlations we found are larger.
-We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed does not result more even language outcomes for first- and later-born children.
+In fact, two large studies (with therefore precise estimates) found effects of about d~.2 for birth order effects on other language tasks [@havron2019effect;@barclay2015within], which would suggest the correlations we found are larger.
+We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further assess whether distributed child care results in more even language outcomes for first- and later-born children.
 
@@
 The range of performance we observed overlapped with previously observed levels of performance.
-Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
+Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations of NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
 Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane' [@cristia2020infant].
@@
 
-The present study shows that NWR be adapted for very different populations than have previously been tested.
+The present study shows that NWR can be adapted for very different populations than have previously been tested.
 In addition, we observed strong correlations with age and typological frequency, while correlations with item length, participant sex, maternal education, and birth order were weaker.

### 2021-11-10 manuscript.tex
for proofing

@@
 NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns.
-In this study, we report on NWR results among children (N = 40, aged 3--10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea.
+In this study, we report on NWR results among children (N = 40, aged 3--10 years) learning Yélî Dnye, an isolate language spoken on Rossel Island in Papua New Guinea.
 Results make three contributions that are specific, and a fourth that is general.
-First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
+First, we found that non-word items containing typologically frequent sounds are repeated without changes more often than non-words containing typologically rare sounds, above and beyond any within-language frequency effects.
 Second, we documented rather weak effects of item length.
-Third, we found that age has a strong correlation with NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order.
+Third, we found that NWR scores correlate strongly with age, whereas they are only weakly correlated with child sex, maternal education, and birth order.
 Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations, and the extent to which they shed light on the factors universally structuring variation in phonological development at a global and individual level.
@@
 Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000; Rumsey, 2017).
-Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
+Some of the evidence for later phonological development comes from non-word repetition (NWR) tasks.
 In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory.
@@
 
-\hypertarget{what-is-nwr}{%
-\subsubsection{What is NWR?}\label{what-is-nwr}}
+\hypertarget{defining-nwr}{%
+\subsubsection{Defining NWR}\label{defining-nwr}}
 
@@
 Phonologists, phoneticians, and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of processing and acquisition via diachronic language change.
-These works focus largely on phonotactics (Moreton \& Pater, 2012) perceptual parsing of the (ambiguous) linguistic signal (Beddor, 2009; Ohala, 1981), and individual differences in processing styles (Bermúdez-Otero, 2015); small but significant effects that may cumulatively drive language change via phonologization (see Yu, 2021 for a recent review).
+These works focus largely on phonotactics (Moreton \& Pater, 2012), perceptual parsing of the (ambiguous) linguistic signal (Beddor, 2009; Ohala, 1981), and individual differences in processing styles (Bermúdez-Otero, 2015); which are small effects that may nonetheless cumulatively drive language change via phonologization (see Yu, 2021 for a recent review).
 Thus, the correlation between typological frequency and ease of acquisition is typically assumed to emerge from one or more of the following causal paths:
@@
 The second research area we contribute data to is research looking at the impact of word length on NWR repetition within specific languages.
-Some work documents much lower NWR scores for longer, compared to shorter, items {[}e.g., among Cantonese-learning children; Stokes, Wong, Fletcher, and Leonard (2006){]}, whereas differences are negligible in other studies {[}e.g., among Italian learners; Piazzalunga, Previtali, Pozzoli, Scarponi, and Schindler (2019){]}.
+Some work documents much lower NWR scores for longer, compared to shorter, items (e.g., among Cantonese-learning children, Stokes, Wong, Fletcher, \& Leonard, 2006), whereas differences are negligible in other studies (e.g., among Italian learners, Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019).
 
@@
 In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer.
-A non-systematic meta-analysis does not provide overwhelming support for this hypothesis {[}Cristia and Casillas (2021); SM1{]}.
-
-Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we did not approach this issue within a hypothesis-testing framework but sought instead to provide additional data on the question, which may be re-used in future meta- or mega-analytic analyses.
+A non-systematic meta-analysis does not provide overwhelming support for this hypothesis (Cristia \& Casillas, 2021, p.
+SM1).
+
+Nonetheless, given the paucity of research looking at this question, and the diversity of current results, we did not approach this issue within a hypothesis-testing framework but sought instead to provide additional data on the question, which may be re-used in future meta- or mega-analyses.
 
@@
 Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based (Farabolini, Rinaldi, Caselli, \& Cristia, 2021).
-In a nutshell, most evidence is mixed, suggesting that correlations with individual variation may be small, and more data is needed to estimate their true size.
+In a nutshell, most evidence is mixed, suggesting that individual variation effects may be small, and more data is needed to estimate their true size.
 For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.
@@
 Regarding child sex, no significant correlation has been found in previous NWR research (Chiat \& Roy, 2007), and in other language tasks evidence is mixed.
-Finally, prior research varies on whether significant differences as a function of maternal education are reported.
+Finally, prior research using NWR varies on whether significant differences as a function of maternal education are reported.
 For instance, no significant difference was found some studies (Balladares, Marshall, \& Griffiths, 2016; Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017); whereas significant differences were reported in others (Santos, Frau, Labrevoit, \& Zebib, 2020; Tuller et al., 2018).
-In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports (Frank, Braginsky, Yurovsky, \& Marchman, 2017) and word comprehension studies (Scaff, 2019).
+In other lines of work, maternal education often correlates with child language outcomes, including vocabulary reports (Frank, Braginsky, Yurovsky, \& Marchman, 2017) and word comprehension studies (Scaff, 2019).
 The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children (see discussion in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
@@
 (2018) employed a set of non-words judged to be language independent and two others that were more aligned with either French or German.
-The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language (``language-dependent'') than when they are ``language-independent.'' To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience.
+The intuition is that NWR will correlate with the relative levels of exposure to that language more strongly when items are aligned with a specific language (``language-dependent'') than when they are ``language-independent.'' To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience.
 Preliminary results of an ongoing meta-analysis suggest significant associations between exposure to a given language and performance in both language-dependent and language-independent NWR (Farabolini, Taboh, Ceravolo, \& Guerra, 2021).
 In any case, this line of research focuses on links between exposure to a given language and NWR performance.
-In contrast, when we discuss linguistic or cultural differences here, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).}
-
-Regarding potential language differences, we note that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/).
+In contrast, when we discuss linguistic or cultural differences here, we ask the question of whether children vary in their performance as a function of which language they are learning (e.g., the language's typological properties) and/or their overall, absolute levels of language experience (not relative levels in a multilingual setting).}
+
+Regarding potential language differences, we note that previous studies composed items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/).
 It would be interesting for future researchers to consider straying from the literature by varying other dimensions that are relevant to the language under study.
@@
 In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).}, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with a complete range of nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/).
-A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (2021) for details).
+A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/, /tpʲ/, /ṇmḍbʲ/, see Levinson, 2021 for details).
 The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/).
 Vowels in Yélî Dnye may be oral or nasal, short or long.
-The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to occur as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
+The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with seven of these able to occur as short and long nasal vowels as well (/ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).
 
@@
 Other ongoing work on Yélî Dnye includes experiment-based infant phoneme discrimination data and errors made in elicited and spontaneous speech from young children, but these data are neither finalized nor yet externally reviewed (see Hellwig, Sarvasy, \& Casillas, provisionally accepted for more information).
-These data will help better inform our current analyses based on NWR in the future (e.g., regarding common sound substitutions) but are not critical for testing our current question about the general correlation between cross-linguistic phone frequency and NWR performance.
+These data will help better inform our current analyses based on NWR in the future (e.g., regarding common sound substitutions) but are not critical for addressing our question about the general correlation between cross-linguistic phone frequency and NWR performance.
 
@@
 However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
-The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones Bunce et al., under review), and additionally shares other societal characteristics wih the Tsimane' {[}e.g., is rural and relies on farming, children grow up in wide familial networks, etc; Casillas, Brown, and Levinson (2021){]}.
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2021).
+The Yélî community can help us gain new insights into this matter because direct speech to children under 3 years is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones Bunce et al., under review).
+Our sample also shares other societal characteristics wih the Tsimane' (e.g., the community is rural and relies on farming, children grow up in wide familial networks, Casillas, Brown, \& Levinson, 2021).
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, under review), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2021).
 Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).
@@
 \item
-  Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
+  Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are cross-linguistically rarer sounds more often substituted by commoner sounds?
 \item
@@
 \item
-  Is individual variation in NWR scores attributable to child age, sex, birth order, and/or maternal education?
+  Is individual variation in NWR scores correlated with child age, sex, birth order, and/or maternal education?
 \end{itemize}
@@
 A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/, /n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible 2- to 4-syllable sequences.
-These candidates were automatically removed from consideration if they appeared in the most recent dictionary (Levinson, 2021).
+Candidates were automatically removed from consideration if they appeared in the most recent dictionary (Levinson, 2021).
 The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye.
@@
 As before, items were automatically excluded if they appeared in the dictionary.
-Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
-Additionally, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a sub-sequence).
+Furthermore, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded.
+We made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., there existed a longer word that included the monosyllable as a sub-sequence).
 These candidates were then presented to one informant, for a final check that they did not mean anything.
@@
 Because space availability was limited in different ways from hamlet to hamlet, the places where elicitation happened varied across testing sites.
-More information is available from the online supplementary materials.
-
+More information is available from the online materials (\url{https://osf.io/qt8gr/}).
+
+We tested one child at a time.
 We fitted the child with a headset microphone (Shure SM10A or WH20 XLR with a dynamic microphone on a headband, most children using the former) that fed into the left channel of a Tascam DR40x digital audio recorder.
@@
 
-We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts and not excluding errors resulting in real words, so as to base our generalizations on more data.
-Insertions and deletions were very rare (metathesis was not attested): there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde{} of unmatched phones).
+We addressed our first research question in a second way, by investigating patterns of error.
+Unlike all other analyses, we looked at all attempts, so as to base our generalizations on more data.
+As in all analyses, we did not exclude errors resulting in real words.
+Deletions were very rare (insertion and metathesis were not attested): there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets).
+We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde96.10 of unmatched phones).
 To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?\footnote{Note that tables of errors including child age are provided in the project repository for those interested in a finer-grained analysis than what is presented here.
@@
 We distinguished errors that included a change of nasality (and may or may not have preserved quality), versus those that preserved nasality (and were therefore a quality error), shown in Table \ref{tab:tab-v}.
-We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.70 versus 12.10).
+We found that errors involving nasal vowel targets were more common than those involving oral vowels (35.70 versus 12.10\%).
 Additionally, errors in which a nasal vowel lost its nasal character were 10 times more common than those in which an oral vowel was produced as a nasal one.
@@
 Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question.
-As shown in Figure \ref{fig:fig-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, ρ(5,649.08) = .47, p \textless{} 0.01).
-In contrast, there was no clear association between NWR scores and sex (Welch t (27.33) = -0.60, p = 0.56), birth order (data missing for 14 children, ρ(3,502.90) = -.198, p = 0.33), or maternal education (ρ( 9,628.60) = .097, p = 0.55).
+As shown in Figure \ref{fig:fig-scores-by-age}, there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance): ρ(38) = .47, p \textless{} 0.01.
+In contrast, there was no clear association between NWR scores and sex: Welch t (27.33) = -0.60, p = 0.56; NWR scores and birth order (data missing for 14 children): ρ(24) = -.198, p = 0.33; or NWR scores and maternal education: ρ(38) = .097, p = 0.55.
 
@@
 We aimed to provide additional data on two questions already visited in NWR work, namely the influence of stimulus length and individual variation, plus one research area that has received less attention, regarding the possible correlation between typological phone frequency and NWR scores.
-An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the Discussion below).
+An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the Discussion below).
 We consider implications of our results on each of these four research areas in turn.
@@
 In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several previous studies (English Vance, Stackhouse, \& Wells, 2005; Italian Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019; Cantonese Stokes, Wong, Fletcher, \& Leonard, 2006; but not in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
-Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
-In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later (Hazan \& Barrett, 2000).
+Indeed, age was significantly correlated with NWR scores and it also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
+In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later (Hazan \& Barrett, 2000; Rumsey, 2017).
 
@@
 
-Finally, we investigated whether birth order might correlate with NWR scores, as it does with other language tasks, such that first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than their older siblings.
+Finally, we investigated whether birth order might correlate with NWR scores, as it does with other language tasks, such that first-born children showing higher scores on standardized language tests than later-born children (Havron et al., 2019) and adults (in a battery including verbal abilities, e.g., Barclay, 2015), presumably because later-born children receive a smaller share of parental input and attention than first-borns.
 Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers.
@@
 However, the effect size was larger than that found for the other two factors and it is far from negligible, at r\textasciitilde.2 or Cohen's d\textasciitilde0.41.
-In fact, two large studies with therefore precise estimates found effects of about d\textasciitilde.2 (Barclay, 2015; Havron et al., 2019), which would suggest the correlations we found are larger.
-We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed does not result more even language outcomes for first- and later-born children.
+In fact, two large studies (with therefore precise estimates) found effects of about d\textasciitilde.2 for birth order effects on other language tasks (Barclay, 2015; Havron et al., 2019), which would suggest the correlations we found are larger.
+We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further assess whether distributed child care results in more even language outcomes for first- and later-born children.
 
@@
 The range of performance we observed overlapped with previously observed levels of performance.
-Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
+Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations of NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
 Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
@@
 Additionally, we only had a single person interacting with children as well as interpreting children's production, so we do not know to what extent our findings generalize to other experimenters and research assistants.
-Furthermore, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual cues in their interpretation.
-Although we know from other work that adults' perceptual performance on these types of sounds is well above chance from audio-only presentation {[}REF{]}, language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.
+Furthermore, since both stimuli presentation and production data collected were audio-only, neither the children nor our research assistant were able to integrate visual production cues in their interpretation.
+Other work shows that children's performance reaches ceiling by 12 years of age for auditorily-presented minimal pairs for typologically rare (i.e., pre- vs post-alveolar stop) contrasts (Casillas \& Levinson, In preparation).
+Nonetheless, language processing for the majority of children will be audiovisual in natural conditions, and thus it may be interesting in the future to capture this aspect of speech.
 
@@
 \leavevmode\hypertarget{ref-barclay2015within}{}%
-Barclay, K. J. (2015). A within-family analysis of birth order and intelligence using population conscription data on swedish men. \emph{Intelligence}, \emph{49}, 134--143.
+Barclay, K. J. (2015). A within-family analysis of birth order and intelligence using population conscription data on {S}wedish men. \emph{Intelligence}, \emph{49}, 134--143.
 
@@
 
+\leavevmode\hypertarget{ref-casillasIPmarkedness}{}%
+Casillas, M., \& Levinson, S. C. (In preparation). Markedness and minimal pair discrimination in children learning {Yélî Dnye}.
+
 \leavevmode\hypertarget{ref-castro1998illiterate}{}%
@@
 \leavevmode\hypertarget{ref-cristia2021supplementary}{}%
-Cristia, A., \& Casillas, M. (2021). Supplementary materials to "non-word repetition in children learning yélî dnye". Retrieved from \url{https://osf.io/5qspb/wiki/home/}
+Cristia, A., \& Casillas, M. (2021). Supplementary materials to non-word repetition in children learning {Yélî D}nye. Retrieved from \url{https://osf.io/5qspb/wiki/home/}
 
@@
 (2004).
-The interaction between vocabulary size and phonotactic probability effects on children's production accuracy and fluency in nonword repetition, \emph{47}, 421--436.
+The interaction between vocabulary size and phonotactic probability effects on children's production accuracy and fluency in nonword repetition.
+\emph{Journal of Speech, Language, and Hearing Research}, \emph{47}, 421--436.
 
@@
 Non-word repetition in bilingual children: The role of language exposure, vocabulary scores and environmental factors.
-\emph{Speech Language and Hearing}.
+\emph{Speech Language and Hearing}, 1--16.
+\url{https://doi.org/10.1080/2050571X.2021.1879609}
 
@@
 Scaff, C., Stieglitz, J., Casillas, M., \& Cristia, A.
-(2021).
+(under review).
 Daylong audio recordings of young children in a forager-farmer society show low levels of verbal input with minimal age-related changes.
-\emph{Draft}.
 
@@
 (2006).
-Nonword repetition and sentence repetition as clinical markers of specific language impairment: The case of cantonese.
+Nonword repetition and sentence repetition as clinical markers of specific language impairment: The case of {C}antonese.
 \emph{Journal of Speech, Language, and Hearing Research}, \emph{49}, 219--236.
@@
 (2018).
-Identifying language impairment in bilingual children in france and in germany.
+Identifying language impairment in bilingual children in {France and in Germany}.
 \emph{International Journal of Language \& Communication Disorders}, \emph{53}(4), 888--904.

### 2021-11-10 manuscript.Rmd
tracking changes in progress

@@
 
-The final list includes three practice items and 40 test items (across infants): 16 monosyllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
+The final list includes three practice items and 40 test items (across children): 16 monosyllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \@ref(tab:tabstims)).
 
@@
 Children were not asked to make repetitions if they did not produce a first attempt.
-In total, test sessions took approximately six minutes (one for practice; five for test list).
+In total, the sessions took approximately six minutes (one for practice; five for the test list).
 
@@
 We next checked whether the association between whole-item NWR scores and cross-linguistic phone frequency could actually be due to frequency of the sounds within the language: The same perception and production pressures that shape languages diachronically could affect a language's lexicon, so that sounds that are easier to perceive or produce are more frequent within a language than those that are harder.
-If so, children will have more experience with the easier sounds, and they may thus be better able to to represent and repeat non-words containing them simply because of the additional exposure.
+If so, children will have more experience with the easier sounds, and they may thus be better able to represent and repeat non-words containing them simply because of the additional exposure.
 
@@
 Deletions were very rare (insertion and metathesis were not attested): there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] + dim(match_bank)[1])*100,2)` of the `r length(del_bank) + dim(sub_bank)[1] + dim(match_bank)[1]` phones found collapsing across all children and target words, so that substitutions constituted the  majority of incorrect phones (~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] )*100,2)` of unmatched phones).
+We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] + length(match_bank))*100,2)` of the `r length(del_bank) + dim(sub_bank)[1] + length(match_bank)` phones found collapsing across all children and target words, so that substitutions constituted the  majority of incorrect phones (~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] )*100,2)` of unmatched phones).
 To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?[^4]

### 2021-11-10 manuscript.tex
tracking changes in progress

@@
 
-The final list includes three practice items and 40 test items (across infants): 16 monosyllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
+The final list includes three practice items and 40 test items (across children): 16 monosyllables containing sounds that are less frequent in the world's languages than singleton plosives; 8 bisyllables; 12 trisyllables; and 4 quadrisyllables (see Table \ref{tab:tabstims}).
 
@@
 Deletions were very rare (insertion and metathesis were not attested): there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde96.10 of unmatched phones).
+We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde16.81 of the 4836 phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde96.10 of unmatched phones).
 To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?\footnote{Note that tables of errors including child age are provided in the project repository for those interested in a finer-grained analysis than what is presented here.

### 2021-11-10 manuscript.Rmd
with tracked changes complete

@@
 complexCs=c("dp","tp","kp","km","kn","mb","lv")
-simplexCs.ipa=c("m","n","l","w","j","w","ṭ","g","p","t","k","f","h","tʃ")
-complexCs.ipa=c("ṭp","tp","kp","km","kṇ","mp","ɣ","lβʲ")
+simplexCs.ipa=c("m","n","l","w","j","w","ṭ","g","p","t","k","f","ɣ","h","tʃ")
+complexCs.ipa=c("ṭp","tp","kp","km","kṇ","mp","lβʲ")
 
@@
 Deletions were very rare (insertion and metathesis were not attested): there were only `r sum(del_tab[c(oralVs,nasalVs)])` instances of deleted vowels (~`r mean(del_pc[c(oralVs,nasalVs)],na.rm=T)`% of all vowel targets), and `r sum(del_tab[c(simplexCs,complexCs)])` instances of deleted consonants (~`r mean(del_pc[c(simplexCs,complexCs)],na.rm=T)`% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] + length(match_bank))*100,2)` of the `r length(del_bank) + dim(sub_bank)[1] + length(match_bank)` phones found collapsing across all children and target words, so that substitutions constituted the  majority of incorrect phones (~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] )*100,2)` of unmatched phones).
+We therefore focus our qualitative description here on substitutions: There were `r dim(sub_bank)[1]` cases of substitutions, ~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] + length(match_bank))*100,2)` of the `r length(del_bank) + dim(sub_bank)[1] + length(match_bank)` phones found collapsing across all children and target words, so that substitutions constituted the  majority of incorrect phones (~`r round(dim(sub_bank)[1]/(length(del_bank) + dim(sub_bank)[1] )*100,2)`% of unmatched phones).
 To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?[^4]
@@
 On the other hand, it is also possible that this pressure gives Yélî children no benefit, and that some of these categories are simply acquired later in development.
-We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not produce adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) until  5 or 6 years of age [@rumsey2017dependency]
+We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not produce adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) until  5 or 6 years of age [@rumsey2017dependency].
 

### 2021-11-10 manuscript.tex
with tracked changes complete

@@
 Deletions were very rare (insertion and metathesis were not attested): there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets).
-We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde16.81 of the 4836 phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde96.10 of unmatched phones).
+We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde16.81 of the 4836 phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde96.10\% of unmatched phones).
 To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?\footnote{Note that tables of errors including child age are provided in the project repository for those interested in a finer-grained analysis than what is presented here.
@@
 
-For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, {[}ɣ{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}).
+For consonants, we inspected complex ({[}ṭp{]}, {[}tp{]}, {[}kp{]}, {[}km{]}, {[}kṇ{]}, {[}mp{]}, and {[}lβʲ{]}) versus simpler ones ({[}m{]}, {[}n{]}, {[}l{]}, {[}w{]}, {[}j{]}, {[}w{]}, {[}ṭ{]}, {[}g{]}, {[}p{]}, {[}t{]}, {[}k{]}, {[}f{]}, {[}ɣ{]}, {[}h{]}, and {[}tʃ{]}), using the same logic: We looked at correct phone repetition, substitution with a change in complexity category, or a change within the same complexity category.\footnote{Note that the substitutions included phones that are not native to Yélî Dnye but do occur in English (e.g., {[}tʃ{]}).
 These data come from careful transcriptions by a native Yélî Dnye speaker who is very fluent in English.}

### 2021-11-25 manuscript.tex
resubmitted

@@
   ,man,floatsintext]{apa6}
+\title{Non-word repetition in children learning Yélî Dnye}
+\author{Alejandrina Cristia\textsuperscript{1} \& Marisa Casillas\textsuperscript{2,3}}
+\date{}
+
 \usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
 \fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 \usepackage{fontspec}
@@
 \usepackage{csquotes}
-\ifxetex
+\ifXeTeX
   % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
@@
 \fi
-\ifluatex
+\ifLuaTeX
   \usepackage{selnolig}  % disable illegal ligatures
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Non-word repetition in children learning Yélî Dnye}
-\author{Alejandrina Cristia\textsuperscript{1} \& Marisa Casillas\textsuperscript{2,3}}
-\date{}
 
@@
 Finally, prior research using NWR varies on whether significant differences as a function of maternal education are reported.
-For instance, no significant difference was found some studies (Balladares, Marshall, \& Griffiths, 2016; Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017); whereas significant differences were reported in others (Santos, Frau, Labrevoit, \& Zebib, 2020; Tuller et al., 2018).
+For instance, no significant difference was found some studies (Balladares, Marshall, \& Griffiths, 2016; Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017); whereas significant differences were reported in others (Santos et al., 2020; Tuller et al., 2018).
 In other lines of work, maternal education often correlates with child language outcomes, including vocabulary reports (Frank, Braginsky, Yurovsky, \& Marchman, 2017) and word comprehension studies (Scaff, 2019).
@@
 The fourth research goal we pursued is to use NWR with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory (see Maddieson, 2005 for a broad classification of languages based on inventory size).
-Indeed, NWR has seldom been used outside of urban settings in Europe and North America (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020; with exceptions including Gallagher, 2014).
+Indeed, NWR has seldom been used outside of urban settings in Europe and North America (Cristia et al., 2020; with exceptions including Gallagher, 2014).
 To our knowledge, it has never been used with speakers of languages having large phonological inventories (e.g., more than 34 consonants and 7 vowel qualities Maddieson, 2013b, 2013a).
@@
 There are no theoretical reasons to presume that the technique will not generalize to these new conditions.
-That said, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
+That said, Cristia et al.
+(2020) recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.
 
@@
 Therefore, it may not be a general index of phonological development, but instead reflect certain non-universal language skills.
-Indeed, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) present their task as being a good index of the development of ``short-hand-like'' representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
+Indeed, Cristia et al.
+(2020) present their task as being a good index of the development of ``short-hand-like'' representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.
 We return to the question of what was measured here in the Discussion.
 
-Aside from Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)'s hypotheses just mentioned, we have found little discussion of linguistic differences (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural differences (i.e., potential differences in NWR as a function of other differences across human populations).\footnote{Please note that the linguistic and cultural differences discussed here are different from the differences discussed in the extensive literature on NWR by bilingual participants.
+Aside from Cristia et al.
+(2020)'s hypotheses just mentioned, we have found little discussion of linguistic differences (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural differences (i.e., potential differences in NWR as a function of other differences across human populations).\footnote{Please note that the linguistic and cultural differences discussed here are different from the differences discussed in the extensive literature on NWR by bilingual participants.
 In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment.
@@
 Within a group of households, it is often the case that older adolescents and adults spend their day tending to their farm plots (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities.
-Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde10 members) independent and mixed-age child play groups (Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2021).
-Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas, Brown, \& Levinson, 2021).
+Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde10 members) independent and mixed-age child play groups (Brown \& Casillas, in press; Casillas et al., 2021).
+Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas et al., 2021).
 Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to host their school-aged relatives during the weekdays for long segments of the school year.
@@
 We add some observations that will help us integrate this study into the broader investigation of NWR across cultures.
-As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors of that paper interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
-However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
+As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors of that paper interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia et al., 2020).
+However, Cristia et al.
+(2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for.
 The Yélî community can help us gain new insights into this matter because direct speech to children under 3 years is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones Bunce et al., under review).
-Our sample also shares other societal characteristics wih the Tsimane' (e.g., the community is rural and relies on farming, children grow up in wide familial networks, Casillas, Brown, \& Levinson, 2021).
-Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, under review), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2021).
-Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).
+Our sample also shares other societal characteristics wih the Tsimane' (e.g., the community is rural and relies on farming, children grow up in wide familial networks, Casillas et al., 2021).
+Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, under review), and Yélî children about 3.6 minutes per hour (Casillas et al., 2021).
+Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in Cristia et al.
+(2020).
 
@@
 
-Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013).
+Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language (e.g., Balladares et al., 2016; Wilsenach, 2013).
 We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye.
-We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-46 items: Jaber-Awida, 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019), with the entire test inventory of 40 final items distributed across children.
+We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat (other work has also used up to 24-46 items: Jaber-Awida, 2018; Kalnak et al., 2014; Piazzalunga et al., 2019), with the entire test inventory of 40 final items distributed across children.
 We used a relatively large number of items to explore correlations with length and phonological complexity.
@@
 
-Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2021).
+Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas et al., 2021).
 That corpus was constituted by sampling from audio-recordings (7--9 hours long), collected as 10 children aged between 1 month and 3 years went about their day.
@@
 These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the recorded people personally.
-For more details, please refer to Casillas, Brown, and Levinson (2021).
+For more details, please refer to Casillas et al.
+(2021).
 
@@
 On the other hand, it is also possible that this pressure gives Yélî children no benefit, and that some of these categories are simply acquired later in development.
-We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not produce adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) until 5 or 6 years of age (Rumsey, 2017)
+We can draw a parallel with children learning another Papuan language, Ku Waru, which has a packed inventory of lateral consonants; children do not produce adult-like realizations of the more complex of these laterals (the pre-stopped velar lateral /ɡ͡ʟ/) until 5 or 6 years of age (Rumsey, 2017).
 
@@
 In broad terms, children should have higher NWR scores for shorter items.
-That said, previous work summarized in the Introduction has shown both very small (e.g., Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019) and very large (e.g., Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020) effects of stimulus length.
+That said, previous work summarized in the Introduction has shown both very small (e.g., Piazzalunga et al., 2019) and very large (e.g., Cristia et al., 2020) effects of stimulus length.
 Setting aside our monosyllabic stimuli (which contained typologically infrequent segments with lower NWR scores, as just discussed), we examined effects of item length among the remaining stimuli, which range between 2 and 4 syllables long.
@@
 
-In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several previous studies (English Vance, Stackhouse, \& Wells, 2005; Italian Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019; Cantonese Stokes, Wong, Fletcher, \& Leonard, 2006; but not in Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
+In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several previous studies (English Vance et al., 2005; Italian Piazzalunga et al., 2019; Cantonese Stokes et al., 2006; but not in Cristia et al., 2020).
 Indeed, age was significantly correlated with NWR scores and it also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency.
@@
 
-In contrast, previous work varies with respect to correlations of NWR scores with maternal education (e.g., Farmani et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Meir \& Armon-Lotem, 2017).
+In contrast, previous work varies with respect to correlations of NWR scores with maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017).
 We did not expect large correlations with maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally.
@@
 It is important to note at the outset that we cannot obtain a final answer because integration across studies implies not only variation in languages and child-rearing settings, but also in methodological aspects including non-word length, non-word design (e.g., the syllable and phone complexity included in the items), and task administration, among others.
-Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, similarly to other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages (Frank, Braginsky, Yurovsky, \& Marchman, 2017).
+Nonetheless, we feel the NWR task is prevalent enough to warrant discussion about this, similarly to other tasks sometimes used to describe and compare children's language skills across populations, like the recent re-use of the MacArthur-Bates Communicative Development Inventory to look at vocabulary acquisition across multiple languages (Frank et al., 2017).
 
@@
 Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations of NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli.
-Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
+Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane' (Cristia et al., 2020).
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-armon2015assessing}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-armon2015assessing}{}}%
 Armon-Lotem, S., Jong, J. de, \& Meir, N. (2015). \emph{Methods for assessing multilingual children: Disentangling bilingualism from specific language impairment}. Bristol: Multilingual matters.
 
-\leavevmode\hypertarget{ref-balladares2016socio}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-balladares2016socio}{}}%
 Balladares, J., Marshall, C., \& Griffiths, Y.
@@
 
-\leavevmode\hypertarget{ref-barclay2015within}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barclay2015within}{}}%
 Barclay, K. J. (2015). A within-family analysis of birth order and intelligence using population conscription data on {S}wedish men. \emph{Intelligence}, \emph{49}, 134--143.
 
-\leavevmode\hypertarget{ref-beddor2009coarticulatory}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-beddor2009coarticulatory}{}}%
 Beddor, P. S. (2009). A coarticulatory path to sound change. \emph{Language}, \emph{85}(4), 785--832.
 
-\leavevmode\hypertarget{ref-bermudezotero2015amphichronic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-bermudezotero2015amphichronic}{}}%
 Bermúdez-Otero, R.
@@
 
-\leavevmode\hypertarget{ref-Praat}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Praat}{}}%
 Boersma, P., \& Weenink, D. (2020). Praat: Doing phonetics by computer (Version 6.1.35). Retrieved from \url{http://www.praat.org/}
 
-\leavevmode\hypertarget{ref-bowey2001nonword}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-bowey2001nonword}{}}%
 Bowey, J. A. (2001). Nonword repetition and young children's receptive vocabulary: A longitudinal study. \emph{Applied Psycholinguistics}, \emph{22}(3), 441--469.
 
-\leavevmode\hypertarget{ref-brandeker2015language}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-brandeker2015language}{}}%
 Brandeker, M., \& Thordardottir, E.
@@
 
-\leavevmode\hypertarget{ref-brown2011cultural}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-brown2011cultural}{}}%
 Brown, P.
@@
 
-\leavevmode\hypertarget{ref-brown2014interactional}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-brown2014interactional}{}}%
 Brown, P.
@@
 
-\leavevmode\hypertarget{ref-brownIPchildrearing}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-brownIPchildrearing}{}}%
 Brown, P., \& Casillas, M.
@@
 
-\leavevmode\hypertarget{ref-bunceURcrosscultural}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-bunceURcrosscultural}{}}%
 Bunce, J., Soderstrom, M., Bergelson, E., Rosemberg, C., Stein, A., Alam, F., \ldots{} Casillas, M. (under review). A cross-cultural examination of young children's everyday language experiences.
 
-\leavevmode\hypertarget{ref-casillas2021early}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-casillas2021early}{}}%
 Casillas, M., Brown, P., \& Levinson, S. C. (2021). {Early language experience in a Papuan community}. \emph{Journal of Child Language}, \emph{48}(4), 792--814.
 
-\leavevmode\hypertarget{ref-casillasIPmarkedness}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-casillasIPmarkedness}{}}%
 Casillas, M., \& Levinson, S. C. (In preparation). Markedness and minimal pair discrimination in children learning {Yélî Dnye}.
 
-\leavevmode\hypertarget{ref-castro1998illiterate}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-castro1998illiterate}{}}%
 Castro-Caldas, A., Petersson, K.
@@
 
-\leavevmode\hypertarget{ref-chiat2015nonword}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-chiat2015nonword}{}}%
 Chiat, S.
@@
 
-\leavevmode\hypertarget{ref-chiat2007preschool}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-chiat2007preschool}{}}%
 Chiat, S., \& Roy, P.
@@
 
-\leavevmode\hypertarget{ref-coady2008uses}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-coady2008uses}{}}%
 Coady, J.
@@
 
-\leavevmode\hypertarget{ref-is08042009language}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-is08042009language}{}}%
 COST Action. (2009). Language impairment in a multilingual society: Linguistic patterns and the road to assessment. \emph{Brussels: COST Office. Available Online at: Http://Www.bi-Sli.org}.
 
-\leavevmode\hypertarget{ref-cristia2021supplementary}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-cristia2021supplementary}{}}%
 Cristia, A., \& Casillas, M. (2021). Supplementary materials to non-word repetition in children learning {Yélî D}nye. Retrieved from \url{https://osf.io/5qspb/wiki/home/}
 
-\leavevmode\hypertarget{ref-cristia2020infant}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-cristia2020infant}{}}%
 Cristia, A., Farabolini, G., Scaff, C., Havron, N., \& Stieglitz, J.
@@
 
-\leavevmode\hypertarget{ref-edwards2004interaction}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-edwards2004interaction}{}}%
 Edwards, J., Beckman, M.
@@
 
-\leavevmode\hypertarget{ref-estes2007differences}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-estes2007differences}{}}%
 Estes, K.
@@
 
-\leavevmode\hypertarget{ref-farabolini2021nonword}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-farabolini2021nonword}{}}%
 Farabolini, G., Rinaldi, P., Caselli, C., \& Cristia, A.
@@
 
-\leavevmode\hypertarget{ref-farabolini2021association}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-farabolini2021association}{}}%
 Farabolini, G., Taboh, A., Ceravolo, M.
@@
 
-\leavevmode\hypertarget{ref-farmani2018normalization}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-farmani2018normalization}{}}%
 Farmani, H., Sayyahi, F., Soleymani, Z., Labbaf, F.
@@
 
-\leavevmode\hypertarget{ref-foley1986papuan}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-foley1986papuan}{}}%
 Foley, W. A. (1986). \emph{{The Papuan languages of New Guinea}}. Cambridge, UK: Cambridge University Press.
 
-\leavevmode\hypertarget{ref-frank2017wordbank}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-frank2017wordbank}{}}%
 Frank, M. C., Braginsky, M., Yurovsky, D., \& Marchman, V. A. (2017). Wordbank: An open repository for developmental vocabulary data. \emph{Journal of Child Language}, \emph{44}(3), 677--694.
 
-\leavevmode\hypertarget{ref-gallagher2014identity}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gallagher2014identity}{}}%
 Gallagher, G. (2014). {An identity bias in phonotactics: Evidence from Cochabamba Quechua}. \emph{Laboratory Phonology}, \emph{5}(3), 337--378. \url{https://doi.org/10.1515/lp-2014-0012}
 
-\leavevmode\hypertarget{ref-gathercole1991differentiating}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gathercole1991differentiating}{}}%
 Gathercole, S.
@@
 
-\leavevmode\hypertarget{ref-gratz2018competition}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gratz2018competition}{}}%
 Grätz, M. (2018). Competition in the family: Inequality between siblings and the intergenerational transmission of educational advantage. \emph{Sociological Science}, \emph{5}, 246--269.
 
-\leavevmode\hypertarget{ref-havron2019effect}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-havron2019effect}{}}%
 Havron, N., Ramus, F., Heude, B., Forhan, A., Cristia, A., Peyre, H., \& Group, E.
@@
 
-\leavevmode\hypertarget{ref-hazan2000development}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hazan2000development}{}}%
 Hazan, V., \& Barrett, S. (2000). The development of phonemic categorization in children aged 6--12. \emph{Journal of Phonetics}, \emph{28}(4), 377--396.
 
-\leavevmode\hypertarget{ref-hellwigPAlanguage}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hellwigPAlanguage}{}}%
 Hellwig, B., Sarvasy, H., \& Casillas, M.
@@
 
-\leavevmode\hypertarget{ref-jabere2018xperiment}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-jabere2018xperiment}{}}%
 Jaber-Awida, A. (2018). Experiment in non word repetition by monolingual {A}rabic preschoolers. \emph{Athens Journal of Philology}, \emph{5}, 317--334. \url{https://doi.org/10.30958/ajp.5-4-4}
 
-\leavevmode\hypertarget{ref-kalnak2014nonword}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-kalnak2014nonword}{}}%
 Kalnak, N., Peyrard-Janvid, M., Forssberg, H., \& Sahlén, B.
@@
 
-\leavevmode\hypertarget{ref-kolinsky2018completely}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-kolinsky2018completely}{}}%
 Kolinsky, R., Leite, I., Carvalho, C., Franco, A., \& Morais, J.
@@
 
-\leavevmode\hypertarget{ref-lancy2015anthropology}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-lancy2015anthropology}{}}%
 Lancy, D. F. (2015). \emph{The anthropology of childhood}. Cambridge, UK: Cambridge University Press.
 
-\leavevmode\hypertarget{ref-lehmann2018early}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-lehmann2018early}{}}%
 Lehmann, J.-Y.
@@
 
-\leavevmode\hypertarget{ref-levinsonYDgrammar}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-levinsonYDgrammar}{}}%
 Levinson, S. C. (2021). \emph{A grammar of {Yélî D}nye, the {P}apuan language of {R}ossel {I}sland}. Berlin, Boston: De Gruyter Mouton.
 
-\leavevmode\hypertarget{ref-liszkowski2012prelinguistic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-liszkowski2012prelinguistic}{}}%
 Liszkowski, U., Brown, P., Callaghan, T., Takada, A., \& de Vos, C.
@@
 
-\leavevmode\hypertarget{ref-maddieson2005correlating}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-maddieson2005correlating}{}}%
 Maddieson, I. (2005). Correlating phonological complexity: Data and validation. \emph{UC Berkeley PhonLab Annual Report}, \emph{1}(1).
 
-\leavevmode\hypertarget{ref-maddieson2009phonology}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-maddieson2009phonology}{}}%
 Maddieson, I. (2009). Phonology, naturalness and universals. \emph{Pozn{á}n Studies in Contemporary Linguistics}, \emph{45}(1), 131--140.
 
-\leavevmode\hypertarget{ref-maddieson2013consonant}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-maddieson2013consonant}{}}%
 Maddieson, I. (2013a). Consonant inventories. \emph{The World Atlas of Language Structures Online}. Retrieved from \url{https://wals.info/chapter/1}
 
-\leavevmode\hypertarget{ref-maddieson2013vowel}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-maddieson2013vowel}{}}%
 Maddieson, I. (2013b). Vowel quality inventories. \emph{The World Atlas of Language Structures Online}. Retrieved from \url{https://wals.info/chapter/2}
 
-\leavevmode\hypertarget{ref-maddiesonIPphoneticsYD}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-maddiesonIPphoneticsYD}{}}%
 Maddieson, I., \& Levinson, S. C. (in preparation). \emph{The phonetics of {Yélî Dnye}, the language of {Rossel Island}}.
 
-\leavevmode\hypertarget{ref-meir2017independent}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-meir2017independent}{}}%
 Meir, N., \& Armon-Lotem, S.
@@
 
-\leavevmode\hypertarget{ref-meir2016disentangling}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-meir2016disentangling}{}}%
 Meir, N., Walters, J., \& Armon-Lotem, S.
@@
 
-\leavevmode\hypertarget{ref-phoible}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-phoible}{}}%
 Moran, S., \& McCloy, D. (Eds.). (2019). \emph{PHOIBLE 2.0}. Jena: Max Planck Institute for the Science of Human History. Retrieved from \url{https://phoible.org/}
 
-\leavevmode\hypertarget{ref-moreton2012structure}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-moreton2012structure}{}}%
 Moreton, E., \& Pater, J. (2012). Structure and substance in artificial-phonology learning, part {II}: substance. \emph{Language and Linguistics Compass}, \emph{6}(11), 702--718.
 
-\leavevmode\hypertarget{ref-ohala1981listener}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ohala1981listener}{}}%
 Ohala, J.
@@
 
-\leavevmode\hypertarget{ref-peuteIPconsonants}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-peuteIPconsonants}{}}%
 Peute, A. A. K., Fikkert, P., \& Casillas, M. (In preparation). Early consonant production in {Y}élî {D}nye and {T}seltal.
 
-\leavevmode\hypertarget{ref-piazzalunga2019articulatory}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-piazzalunga2019articulatory}{}}%
 Piazzalunga, S., Previtali, L., Pozzoli, R., Scarponi, L., \& Schindler, A.
@@
 
-\leavevmode\hypertarget{ref-rumsey2017dependency}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-rumsey2017dependency}{}}%
 Rumsey, A.
@@
 
-\leavevmode\hypertarget{ref-dos2020epreuve}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-dos2020epreuve}{}}%
 Santos, C.
@@
 
-\leavevmode\hypertarget{ref-scaff2019}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-scaff2019}{}}%
 Scaff, C. (2019). \emph{Beyond {WEIRD}: An interdisciplinary approach to language acquisition} (PhD thesis).
 
-\leavevmode\hypertarget{ref-scaff2021daylong}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-scaff2021daylong}{}}%
 Scaff, C., Stieglitz, J., Casillas, M., \& Cristia, A.
@@
 
-\leavevmode\hypertarget{ref-stokes2006nonword}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-stokes2006nonword}{}}%
 Stokes, S.
@@
 
-\leavevmode\hypertarget{ref-torrington2015non}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-torrington2015non}{}}%
 Torrington Eaton, C., Newman, R.
@@
 
-\leavevmode\hypertarget{ref-tuller2018identifying}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tuller2018identifying}{}}%
 Tuller, L., Hamann, C., Chilla, S., Ferré, S., Morin, E., Prevost, P., \ldots{} Zebib, R.
@@
 
-\leavevmode\hypertarget{ref-vance2005speech}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-vance2005speech}{}}%
 Vance, M., Stackhouse, J., \& Wells, B. (2005). Speech-production skills in children aged 3--7 years. \emph{International Journal of Language \& Communication Disorders}, \emph{40}(1), 29--48.
 
-\leavevmode\hypertarget{ref-wilsenach2013phonological}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-wilsenach2013phonological}{}}%
 Wilsenach, C.
@@
 
-\leavevmode\hypertarget{ref-yu2021toward}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-yu2021toward}{}}%
 Yu, A. C. L. (2021). Toward an individual-difference perspective on phonologization. \emph{Glossa: A Journal of General Linguistics}, \emph{6}(1), 1--24.

### 2022-02-21 manuscript.Rmd
minor, addressing Rs
for ldr stylesheet

@@
 output            :
-  papaja::apa6_pdf:
+  papaja::apa6_word:
     latex_engine: xelatex
@@
     "Bisyll" = 2, "Trisyll" = 2, "Tetrasyll" = 2))
-
+write.csv(stims.wide,"tab1.csv")
 ```
@@
 Err.)")
-
+write.csv(mytab,"tab2.csv")
 ```
@@
   caption="Number (and percent) of consonant targets that were correctly repeated (Corr.), deleted (Del.), or substituted, as a function of the complexity of the consonant, and whether the error resulted in a change of complexity (Cmpl Err.) or not (Othr Err.)")
-
+write.csv(mytab,"tab3.csv")
 ```

### 2022-02-21 manuscript.Rmd
version for final proofing

@@
 kable(length.tab,"latex", booktabs = T,caption="NWR means (and standard deviations) measured in whole-word scores and normalized Levenshtein Distance (NLD), separately for the four stimuli lengths.")
+
+write.csv(length.tab,"tab4.csv")
 ```
